### 2019-06-11 2017_AFP_AMBRR proofs/document/root.tex
Further work on the article on transactions

\documentclass[11pt,a4paper]{article}
\usepackage{isabelle,isabellesym}
\usepackage{amsfonts, amsmath, amssymb}

% this should be the last package used
\usepackage{pdfsetup}

% urls in roman style, theory text in math-similar italics
\urlstyle{rm}
\isabellestyle{it}

\begin{document}
\title{Proofs for Preserving Invariant Rules in Relation Algebra}
\author{Stef Joosten (0000-0001-8308-0189)}
%orcid.org/0000-0001-8308-0189
\maketitle

\begin{abstract}
	Relation Algebra can be used as a programming language for building information systems.
	This claim was made in a paper~\cite{JoostenRAMiCS2017} at the conference
	``Relational and Algebraic Methods in Computer Science''.
	An entirely generated information system was demonstrated and presented in Lyon on May 17th of 2017 at the 16th edition of the conference.
	It has been implemented in a tool called Ampersand,
	in which programming is done by specifying invariant rules.
	An invariant is a rule that must remain satisfied as long as it is valid in its context.
	This paper contains the proofs that are needed to build the code for such software.
	It is a companion paper to the extended version of the RAMiCS-2017 paper~\cite{Joosten-AFP17}.
\end{abstract}

\tableofcontents

\section{Introduction}
\label{sct:Introduction}
	This paper presents proofs that are used for programming information systems in relation algebra.
	A compiler, Ampersand~\cite{Michels2011}, is used to compile concepts, relations and rules into a working database-application.
	Ampersand is a syntactically sugared version of heterogeneous relation algebra~\cite{Schmidt1997}.

	The user may regard Ampersand as a programming language, especially suited for designing the back-end of information systems.
	The axioms of Tarski can be used to manipulate expressions in a way that preserves meaning~\cite{vdWoude2011}.
	This makes Ampersand a declarative language.

	This paper contributes by publishing the required proofs in the Archive of Formal Proofs\footnote{https://www.isa-afp.org/}.
	It means that every proof presented here has been validated in the theorem prover Isabelle~\cite{Nipkow2002}.
	These proofs contribute to a theory to assist in preserving invariants.

\input{session}

\bibliographystyle{abbrv}
\bibliography{root}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2019-06-11 2017_JLAMP/articleRamics.tex
Further work on the article on transactions

\documentclass{elsarticle}
\usepackage{graphicx}
%\usepackage{multicol}
%\usepackage{footmisc}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[english]{babel}
%\usepackage[official,right]{eurosym}
\selectlanguage{english}
\include{preamramics}
\hyphenation{ExecEngine}
\begin{document}

\title{Relation Algebra as Programming Language using the Ampersand Compiler}
\author[ou,ordina]{Stef Joosten\fnref{fn1}}
\ead{stef.joosten@ou.nl}
\address[ou]{Open Universiteit Nederland, Postbus 2960, 6401 DL Heerlen, the Netherlands}
\address[ordina]{Ordina NV,  Nieuwegein, the Netherlands}
\fntext[fn1]{ORCID 0000-0001-8308-0189}

\begin{abstract}
	Relation Algebra can be used as a programming language for building information systems.
	This paper demonstrates the principle by presenting a case study together with the theory behind programming in Relation Algebra.
	As a case study, we have developed a database application for legal reasoning.
	We discuss a small part of it to illustrate the mechanisms of programming in Relation Algebra.
	Beside being declarative, relation algebra comes with attractive prospects for developing software.
	The compiler that was used for this case study, Ampersand, is the result of an open source project.
	Ampersand has been tried and tested in practice and is available as free open source software\footnote{\tt https://github.com/AmpersandTarski/Ampersand}.
\end{abstract}

\begin{keyword}
relation algebra\sep software development\sep legal reasoning\sep information systems design\sep Ampersand\sep MirrorMe\sep reactive programming
\end{keyword}
\maketitle

\section{Introduction}
\label{sct:Introduction}
	This paper investigates how relation algebra can be used as a programming language for information systems.
	A compiler, Ampersand~\cite{Michels2011}, is used to compile concepts, relations and rules into a working database application.
	Ampersand is a syntactically sugared version of heterogeneous relation algebra \cite{Schmidt1997}.
	We present a case study to demonstrate programming in relation algebra and its impact on the software development process.
	The case study takes the reader by the hand in the thinking process of a software developer who programs with relation algebra.

	The use of relation algebra as a programming language is rooted in relation algebra~\cite{Maddux06}, logic programming~\cite{Lloyd1984},
	database application programming~\cite{Codd70}, and formal specification of software~\cite{Dahl93}.
	Ampersand uses these ideas to compile relation algebra to working software, an idea which is also found in RelView~\cite{Berghammer2005} by Berghammer (Univ. of Kiel).
 	In Ampersand the user specifies constraints. The compiler then generates an information system that respects these constraints.
	RelView also generates software but uses relational algebra operators in an otherwise imperative programming language.
	This makes RelView stronger in specifying complex computational problems.
	Some more differences with earlier programming languages are discussed in section~\ref{sct:Comparison}.

	The author shares the conviction that a provably correct implementation of requirements
	is essential to the software development process~\cite{Boehm1981}.
	Even though formal methods are not widely used in computer science,
        Ampersand is built on the foundations laid by formal specification methods such as Z~\cite{Z} and Alloy~\cite{Alloy2006}.
	In contrast with such methods,
	Ampersand is equipped with a software generator that generates information systems,
	each of which consists of a database with web-based user interfaces.
	Thus, specifying in Ampersand is developing software at the same time.

	The user may regard Ampersand as a programming language, especially suited for designing the back-end of information systems.
	The axioms of Tarski can be used to manipulate expressions in a way that preserves meaning~\cite{vdWoude2011}.
	This makes Ampersand a declarative language.

	Our case study shows an argument assistant for legal professionals, which was built as innovation project at Ordina.
	The purpose of this argument assistant is to support legal professionals in constructing a legal brief%
\footnote{A \define{brief} is a document that is meant to summarize a lawsuit for the judge and counterparty.
	It provides legal reasons for claims in a lawsuit based on regulations, precedents, and other legally acceptable sources.
	It shows how the reasoning applies to facts from the case.}.
	To the programmer, this offers the challenge to represent the problem of legal reasoning in relation algebra.
	This challenge is by no means simple, but outside the scope of this paper.
	In the Ampersand project at the Open University of the Netherlands,
	we are studying practical and didactic issues of software development in relation algebra~\cite{Michels2015}.

	The contribution of this paper is twofold.
	The case study contributes to the idea that programming can be done in relation algebra.
	The theory in section~\ref{sct:Code Fragments} contributes a theory to assist in preserving invariants.
	Section~\ref{sct:Code Fragments} extends a paper presented at RAMiCS-2017 \cite{JoostenRAMiCS2017}
	It provides derivations of program code,
	all of which have been checked using the theorem prover Isabelle~\cite{Nipkow2002}.

	Section~\ref{sct:Ampersand} introduces Ampersand and its semantics.
	Section~\ref{sct:Conceptual analysis} introduces part of a theory of legal reasoning,
	which was developed for argument assistance.
	Section~\ref{sct:Programming in RA} discusses the programming mechanism in the application,
	and section~\ref{sct:Programming} visualizes that mechanism.
	Section~\ref{sct:Code Fragments} presents one way (of infinitely many) to derive code for arbitrary rules in relation algebra.
	Section~\ref{sct:Reflection} reflects on software development in Ampersand.
	It also provides an overview of the use of Ampersand in practice and an outlook to its further development.

\section{Ampersand}
\label{sct:Ampersand}
	In this section we explain the basics of Ampersand.
	The reader is expected to have sufficient background in relation algebra to understand the remainder of this paper.

	The core of an Ampersand script is a triple $\la\rules,\rels,\concepts\ra$,
	which consists of a set of rules $\rules$, relations $\rels$, and concepts $\concepts$.
	An Ampersand script is interpreted by the compiler as an information system.
	The rules constitute a theory in heterogeneous relation algebra.
	They constrain a body of data that resides in a database.

	The Ampersand compiler generates a database from relations in the script.
	A database application%
\footnote{Ampersand generates an application that consists of a relational database and interface components.
	Currently this application runs server-side on a PHP/MySQL platform and on a web-browser on the client-side.}
	assists users to keep rules satisfied throughout the lifetime of the database. It is also generated by Ampersand.

	A rule is a constraint in the form of an equality between two terms.
	Terms are built from relations.
	Ampersand interprets every relation as a finite set of pairs, which are stored in the database.
	The phase in which Ampersand takes a script and turns it into a database, is what we will refer to as \define{compile-time}.
	The phase in which a user interacts with the database, is what we will refer to as \define{run-time}.
	At run-time, Ampersand can decide which rules are satisfied by querying the database.
	The compiler generates all software needed to keep rules satisfied at run-time.
	If one of the rules is not satisfied as a result of data that has changed,
	that change is reverted (rolled back) to the original state.
	Either way, Ampersand ensures that all rules remain satisfied.

	Atoms are values that have no internal structure, meant to represent data elements in a database.
	From a business perspective, atoms are used to represent concrete items of the world,
	such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
	By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ are used to represent \emph{atoms}.
	The set of all atoms is called $\atoms$.
        Each atom is an instance of a \emph{concept}.

	Concepts (from set $\concepts$) are names we use to classify atoms in a meaningful way.
	For example, you might choose to classify \atom{Peter} as a person, and \atom{074238991} as a telephone number.
        We will use variables $A$, $B$, $C$, $D$ to represent concepts.
	The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.
	The expression $a \in A$ means that atom $a$ is an \emph{instance} of concept $A$.
	The declaration of $A\isa B$ (pronounce: $A$ is a $B$)
	in an Ampersand script states that any instance of $A$ is an instance of $B$ as well.
	We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
	Specialization is needed to allow statements such as: ``An orange is a fruit that ....''.
	Specialization is not used in the remainder of this paper.

	Relations (from set $\rels$) are used in information systems to store data.
%	A \define{fact} is a statement that is true in a business context.
%	Facts are stored and kept as data in a computer.
	As data changes over time, so do the contents of these relations.
	In this paper relations are represented by variables $r$, $s$, and $d$.
	We represent the declaration of a relation $r$ by $\declare{nm}{A}{B}$,
	in which \id{nm} is a name and $A$ and $B$ are concepts.
	We call $A$ the source concept and $B$ the target concept of the relation.
	The pair $\pair{A}{B}$ is called the \emph{type} of the relation.
	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.

	The meaning of relations in Ampersand is defined by an interpretation function $\mathfrak{I}$.
	It maps each relation to a set of pairs.
	The type system of Ampersand guarantees that each pair in $r$ respects the type of $r$:
\begin{equation}
	\pair{a}{b}\in\ti{\declare{nm}{A}{B}} \Rightarrow\ a \in A \wedge b \in B \label{typing of relations}
\end{equation}
	The type system is strong, which means that every atom has a type.
	The type system is static, which means that type checking is entirely done at compile-time.

	Terms are used to combine relations using operators.
	The set of terms is called $\terms$.
\begin{definition}[terms]
\label{def:terms}
\item   The set of terms, $\terms$, is the smallest set that satisfies, for all $r,s \in \terms$, $d\in\rels$ and $A,B \in \concepts$: 
\begin{eqnarray}
	d&\in&\terms         \quad\quad\text{(every relation is a term)}\\
	(r \cup s)&\in&\terms\quad\quad\text{(union)}\\
	(r \cap s)&\in&\terms\quad\quad\text{(intersection)}\\
	(r-s)&\in&\terms     \quad\quad\text{(difference)}\label{def:difference}\\
	(r;s)&\in&\terms     \quad\quad\text{(composition)}\\
	(r\backslash s)&\in&\terms     \quad\quad\text{(right residual)}\\
	(r\slash s)&\in&\terms     \quad\quad\text{(left residual)}\\
	\flip r&\in&\terms   \quad\quad\text{(converse)}\\
	\cmpl{r}&\in&\terms   \quad\quad\text{(complement)}\\
%	\kleeneplus r&\in&\terms   \quad\quad\text{(Kleene closure)}\\
%	\kleenestar r&\in&\terms   \quad\quad\text{(Kleene closure)}\\
	\ident{A}&\in&\terms \quad\quad\text{(identity)}\\
	\fullt{A}{B}&\in&\terms \quad\quad\text{(full set)}
\end{eqnarray}
\end{definition}
	Throughout the remainder of this paper,	terms are represented by variables $r$, $s$, $d$, and $t$.
	The \define{type} of a term $r$ is a pair of concepts, given by $\tf{r}$.
	$\mathfrak{T}$ is a partial function that maps terms to types.
	If a term has one type, it is called \define{type correct}.
	The Ampersand compiler requires all terms to be type correct.
	If no type can be assigned, the compiler gives an error message.
	If multiple types are assignable to a term, the ambiguity is signalled by the compiler.
	In that case it will also emit an error message and prompt the programmer to disambiguate the term by adding type information.
	In practice, the derivation of types can disambiguate most terms.
	So, the programmer experiences some freedom to denote distinct relations by the same name without the obligation to specify their types at all occurrences.
	The compiler does not generate any code if there is a term that is not type correct.
	The type function and its restrictions are discussed in~\cite{Joosten2015}
	and have no consequences for the remainder of this paper.

 	The meaning of terms in Ampersand is given by an interpretation function $\mathfrak{I}$.
	Let $A$ and $B$ be finite sets of atoms, then $\mathfrak{I}$ maps each term to the set of pairs for which that term stands.
\begin{definition}[interpretation of terms]
\label{interpretation of terms}
\item   For every $A,B\in\concepts$ and $r,s\in\terms$
\begin{eqnarray}
	\ti{r}		 &=&\{\pair{a}{b}|\ a\ r\ b\}	\\
	\ti{r \cup s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{or }\ \pair{a}{b}\in\ti{s}\}	\\
	\ti{r \cap s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{a}{b}\in\ti{s}\}	\\
	\ti{r-s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{a}{b}\notin\ti{s}\}	\\
	\ti{r;s}	 &=&\{\pair{a}{c}|\ \text{for some}\ b,\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{b}{c}\in\ti{s}\}	\\
	\ti{r\backslash s}	 &=&\{\pair{b}{c}|\ \text{for all}\ a,\ \pair{a}{b}\in\ti{r}\ \text{implies}\ \pair{a}{c}\in\ti{s}\}	\\
	\ti{r\slash s}	 &=&\{\pair{a}{b}|\ \text{for all}\ c,\ \pair{b}{c}\in\ti{s}\ \text{implies}\ \pair{a}{c}\in\ti{r}\}	\\
	\ti{\flip{r}}	 &=&\{\pair{b}{a}|\ \pair{a}{b}\in\ti{r}    \}	\\
	\ti{\cmpl{\declare{r}{A}{B}}}	 &=&\fullt{A}{B}-r	\\
%	\ti{\kleeneplus{r}}	 &=&\text{the smallest set that satisfies }\kleeneplus{r}=r\ \cup\ \kleeneplus{r};\kleeneplus{r}	\\
%	\ti{\kleenestar{\declare{r}{A}{A}}}	 &=&\ident{A}\cup\kleeneplus{r}\\
	\ti{\ident{A}} 	 &=&\{\pair{a}{a}|\ a\in A\}	\\
	\ti{\fullt{A}{B}}&=&\{\pair{a}{b}|\ a\in A, b\in B\}
\end{eqnarray}
\end{definition}
%	The Kleene closure operators (postfix $\kleeneplus{\ }$ and $\kleenestar{\ }$) have been implemented partially in the current Ampersand implementation.

	A \define{rule} is a pair of terms $r,s\in\terms$ with $\tf{r}=\tf{s}$, which is syntactically recognizable as a rule.
\[\text{RULE}\ r = s\]
	This means \(\ti{r} = \ti{s}\). In practice, many rules are written as:
\[\text{RULE}\ r\subs s\]
	This is a shorthand for 
\[\text{RULE}\ r\cap s = r\]
	We have enhanced the type function $\mathfrak{T}$ and the interpretation function $\mathfrak{I}$ to cover rules as well.
	If $\tf{r}=\tf{s}$ and $\tf{s}=\pair{A}{B}$:
\begin{eqnarray}
	\tf{\text{RULE}\ r = s}   &=&\pair{A}{B}\\
	\tf{\text{RULE}\ r\subs s}&=&\pair{A}{B}\\
 resolved this \end{eqnarray}

	We call a rule $\text{RULE}\ r = s$ \define{satisfied} when its interpretation equals $\ti{\fullt{A}{B}}$.
	As the population of relations used in $r$ changes with time, the satisfaction of the rule changes accordingly.
	A software developer, who conceives these rules, must consider how to keep each one of them satisfied.
	We call a rule \define{violated} if it is not satisfied.
	The set $\ti{(s-r)\cup(r-s)}$ is called the \emph{violation set} of \(\text{RULE}\ r = s\).
	To \define{resolve} violations means to change the contents of relations such that the rule is satisfied%
\footnote{To \define{restore invariance} is sometimes used as a synonym to resolving violations.
	To \define{preserve invariance} is sometimes used as a synonym to keeping a rule satisfied.}.
	Consequently, such a rule may also be referred to as \define{invariant}.
	Some ways of resolving violations are presented in section~\ref{sct:Code Fragments}.
	Each pair in the violation set of a rule is called a violation of that rule.

	The software developer must define how to resolve violations when they occur.
	She does so by inserting and/or deleting pairs in appropriately chosen relations.
	Whatever choice she makes, she must ensure that her code yields data that satisfies the rules.
	When we say: ``rule $r$ specifies this action'' we mean that satisfaction of rule $r$ is a postcondition of any action specified by rule $r$.

\section{Software for Legal Reasoning}
\label{sct:Conceptual analysis}
	As a case study, an argument assistance system, MirrorMe, was built.
	We have chosen to implement the ideas of Toulmin \cite{Toulmin1958},
	because his book ``The uses of Arguments'' is still one of the most influential works in the area of legal argument management%
\footnote{Thanks to Elleke Baecke for pointing us towards this source.}.
	Recent work typically draws on Toulmin, so his doctrine offers a good starting point.
	Toulmin's ideas have been implemented before, for example in a tool called ArguMed \cite{Verheij1999}.

	Software systems that support legal arguments have been around for many decades.
	Legal reasoning differs from logical reasoning because of the human judgments that are involved.
	Verheij~\cite{Verheij2003} distinguishes between argument assistance systems and automated reasoning systems.
	An automated reasoning system (ARS) mimics human reasoning in an attempt to automate a decision.
	An argument assisance system (AAS) helps its user to construct a decision.
	The literature on legal reasoning~\cite{Lind2007} makes it quite clear why mathematical logic alone does not suffice.
	Legal reasoning puts a human decision at the root of its notion of truth,
	whereas mathematical logic eradicates any human influence.
	For this reason, attempts to apply mathematical logic to legal judgments have had limited impact~\cite{Prakken2005}.
	Argument assistance systems~\cite{Verheij2005} respect the professional freedom of each legal professional
	in constructing her own line of argumentation.
	AAS offer help in many different ways.
	For instance by looking up legal references, jurisdiction from the past, scholarly works etc.
	They can also help to construct and validate arguments by keeping arguments and evidence organized.
	They can store, disclose and share legal evidence.
	The MirrorMe system lets each user (whether a plaintiff, judge, or defendant) decide individually on the truth of every statement.
	Differences in truth assessments can then be made visible and manageable.

	Software development is founded on mathematical reasoning, whereas legal work is built on defeasible reasoning.
	So we have built an AAS in the mathematical formalism of relation algebra
	by \emph{implementing} Toulmin's principles%
\footnote{We deliberately did not want to squeeze legal reasoning into relation algebra, or any other objective formalism.}.
	The structure of MirrorMe was designed by conceptually analysing ideas of Toulmin,
	such as claim, warrant, argument, and rebuttal. They appear in MirrorMe as relations.
	The Ampersand compiler generates a conceptual data model
	to help the software developer to oversee all relations.
	Even though our case study yields a model that is a bit too large for this paper,
	we have printed a relevant part of it at the end of this paper in figure~\ref{fig:conceptual model}.
	This diagram gives an impression of what it looks like.
	It shows entities such as Statement, Argument, Binding, Template, Case, and Claim, together with their attributes.
	Attributes are the univalent and injective relations, which Ampersand derives from the script.
	Relations that are neither univalent nor injective are n-m relations between entities.
	In practice, the conceptual models have shown to be readable without much explanation for experienced database developers.

\section{Programming in Relation Algebra}
\label{sct:Programming in RA}
	This section illustrates the execution model of Ampersand by means of an example.
	A demonstration of this example was given at RAMiCS-2017.
	The example covers a tiny part of the whole MirrorMe program,
	which is enough to give a taste of software development in relation algebra.

	The use case for this example is the substitution of placeholders in text.
	For instance, the user might want to replace \stmtText{The employee, [emp], is entitled to [raise]} by
		\stmtText{The employee, James, is entitled to 50 Euros.}.
	The first text is called a \define{template}, because it contains placeholders in square brackets.
	The strings \stmtText{[emp]} and \stmtText{[raise]} are the placeholders.
	The second text is called a \define{statement},
	because it contains no placeholders and can be read as a statement in natural language.
	So, a statement can be either true or false.

	A user makes such replacements when formulating arguments and reasoning with them.
	MirrorMe helps her by making replacements consistently throughout each case.
	In MirrorMe, a user can define templates.
	The reason for using placeholders is that legal rules are stated in general terms, e.g.
	\stmtText{Every employee is entitled to a raise in salary}.
	A user will pick a legal text (from a source he trusts) and replace parts of that text by placeholders.

	The substitution of a placeholder by a phrase is called a \define{binding}.
	A binding assigns a piece of text to a placeholder.
	That text is the \define{phrase of} the binding and
	also the \define{value} of the placeholder.
	When all placeholders in a template have been substituted, a statement remains.

	The process of substitution is embodied by a sequence of events
	that are triggered by rule violations.
	Events may resolve violations in one rule, but cause violations in others.
	In this manner, rule violations and events to resolve them alternate.
	That is shown in figure~\ref{fig:log file}.

\begin{figure}[htb]
\begin{verbatim}
 ExecEngine run started
 ExecEngine satisfying rule 'signal phrase update'
 InsPair(resetS,Statement,Stat623,Statement,Stat623)
 ExecEngine satisfying rule 'flush substitutions'
 DelPair(substituted,Binding,Bind625,Statement,Stat623)
 ExecEngine satisfying rule 'reset statement text'
 InsPair(stmtShowText,Statement,Stat623,StmtText,
         The employee, [emp], is entitled to [raise].)
 ExecEngine satisfying rule 'done initializing'
 DelPair(resetS,Statement,Stat623,Statement,Stat623)
 ExecEngine satisfying rule 'substitute'
 InsPair(stmtShowText,Statement,Stat623,StmtText,
         The employee, James, is entitled to [raise].)
 InsPair(substituted,Binding,Bind625,Statement,Stat623)
 ExecEngine satisfying rule 'fill shownPhrase'
 InsPair(shownPhrase,Binding,Bind625,Phrase,James)
 ExecEngine run completed
\end{verbatim}
\caption{Log file of a substitution}
\label{fig:log file}
\end{figure}
	The log file of figure~\ref{fig:log file} has been taken from a computer that carries out the procedure to satisfy all rules.
	It illustrates how the machine is busy satisfying one rule after another by inserting or deleting elements in different relations.
	This example follows the event flow in a dependency graph, figure~\ref{fig:event flow}, which is discussed in section~\ref{sct:Programming}.
	The machine will only act on rules that are violated.
	The first rule violated is ``signal phrase update'', in which the computer signals a change in relation \id{phrase}.
	That change is caused by a user who has added or deleted pairs in this relation,
	The computer will react by changing other relations such that the rule is satisfied again.
	This may then violate other rules, which have to be restored in turn.
	The procedure ends by doing the necessary substitutions,
	ensuring that all statements have the actual phrase of a placeholder in their text.

	Before looking at the actual rules,
	let us introduce an auxiliary relation, \id{differB}, to represent bindings that must be updated:
\[\id{differB} = \ident{Binding}\cap\id{shownPhrase};\cmpl{\ident{Phrase}};\flip{\id{phrase}}\]
	The relation \id{phrase} links bindings to texts.
	It contains the actual phrase of each binding.
	This is also known as the value of a placeholder%
\footnote{A phrase and a placeholder are linked by a pair from $\flip{\id{phrase}};\id{placeholder}$.}.
	The relation \id{shownPhrase} links bindings to text fields in the user interface.
	So \id{differB} contains the bindings of which the phrase differs from what is shown on the screen.
	Whether a placeholder has been edited can be observed by comparing its shown phrase to the phrase.
	Bindings in \id{differB} are to be updated,
	because the value of placeholders should always correspond with the text on the screen.

	Let us now study the rules to see how these rules cause the right actions to take place in the correct order.
	We will follow the log file from figure~\ref{fig:log file} in reverse direction,
	reasoning backwards from the result.

\subsection{rule: fill shownPhrase}
\label{rule: fill shownPhrase}
	The sole purpose of having the relation \id{shownPhrase} is to detect a change
	and propagate it as the new value of a placeholder.
	Then, the new value of the placeholder must be propagated in every statement in which the placeholder was used.
	That update action is specified in section~\ref{rule: substitute}.
	The following rule specifies how \id{shownPhrase} is made equal to \id{phrase},
	after all necessary substitutions are done.

\[\text{RULE}\ (\ident{Binding}\cap\id{substituted}/\id{inStatement} );\id{phrase}\ \subs\ \id{shownPhrase} \]
        This rule says that far a binding with all substitutions made, the \id{phrase} must be equal to the \id{shownPhrase}.
	When violated, the software inserts all violations into \id{shownPhrase} to satisfy the rule.

\subsection{rule: substitute}
\label{rule: substitute}
	Let us now look into the process of substituting placeholders by phrases.
	The relation \id{tmplParsedText} contains the template text,
	e.g. \stmtText{The employee, [emp], is entitled to [raise]}.
	The text in which a placeholder has been substituted by a phrase,
	e.g. \stmtText{The employee, John Brown, is entitled to 50 Euros},
	is kept in relation \id{stmtShowText}.
	Each substitution that has been done in a statement corresponds to a binding-statement pair in the relation \id{substituted}.
	This relation keeps track of all substitutions.
	After a placeholder has been substituted, it no longer occurs in \id{stmtShowText}.
	The absence of placeholders in \id{stmtShowText} poses a problem if we want to substitute a new phrase in \id{stmtShowText}.
	That substitution must be carried out in the original template,
	which contains all of the original placeholders.
	So, the text in \id{stmtShowText} must first be reset to the original text from \id{tmplParsedText}.
	To keep track of substitutions correctly, all corresponding binding-statement pairs must be removed from the relation \id{substituted}.
	Only after resetting is done, the substitutions can be put back in place with the new phrases filled in.

	We define a relation \id{resetS} to register the statements that are being reset.
    In statements that are not being reset, $\ident{Statement}-\id{resetS}$, substitutions can take place.
	All placeholders that have a binding with a phrase can be substituted. 
	The following rule specifies the action of substituting placeholders.
\begin{displaymath}
\begin{array}{rl}
\multicolumn{2}{l}{\text{RULE}}\\
&(\ident{Binding}\cap \id{phrase} ;\flip{\id{phrase}});\id{inStatement} ;(\ident{Statement} - \id{resetS})\\
\subs\\
&\id{substituted}
\end{array}
\end{displaymath}
	Violations of this rule are binding-statement pairs, of which the binding has a phrase and the statement is not being reset.
	Hence, this rule can be satisfied by inserting every violation into the relation substituted.

\subsection{rule: done initializing}
\label{rule: done initializing}
	Resetting a statement is done when two conditions are met.
	First, every statement that is (still) being reset may have no bindings in the relation \id{substituted}.
	Second, the text in \id{stmtShowText} corresponds to the text in \id{tmplParsedText}.
	So the rule that specifies the action is:

\begin{displaymath}
\begin{array}{rl}
\multicolumn{2}{l}{\text{RULE}}\\
&(\id{resetS} - \flip{\id{inStatement}};\id{substituted})\ \cap\\
&\id{template} ;\id{tmplParsedText} ;\flip{\id{stmtShowText}}\\
\subs\\
&\overline{\id{resetS}}
\end{array}
\end{displaymath}
	Violations of this rule are statements that are no longer being reset, but are still in the relation \id{resetS}.
	The appropriate action is to remove them from \id{resetS}.

\subsection{rule: reset statement text}
\label{rule: reset statement text}
	To satisfy one condition from section~\ref{rule: done initializing},
	the text in \id{stmtShowText} must be made equal to the original text in the template.
	That text is obtained from the relation \id{descriptor}.
	The action is specified by the following rule:
\[\text{RULE}\ \id{resetS} ;\id{template} ;\id{descriptor}\ \subs\ \id{stmtShowText} \]
	Violations of this rule are descriptors of templates that belong to statements that are being reset.
	These violations are resolved by inserting them in \id{stmtShowText}.

\subsection{rule: flush substitutions}
\label{rule: flush substitutions}
	To satisfy the other condition from section~\ref{rule: done initializing},
	the following rule specifies the action to be taken:
\[\text{RULE}\ \fullt{Binding}{Binding};\id{inStatement} ;\id{resetS}\ \subs\ \overline{\id{substituted} }\]
	Every binding in a statement that is being reset needs to be removed from the relation \id{substituted}.
	The software developer can implement this by deleting all violations of this rule from the relation \id{substituted}.

\subsection{rule: signal phrase update}
\label{rule: signal phrase update}
	When the phrase in a binding is edited and the new phrase differs from the shown phrase,
	this signals that substitutions must be flushed (see~\ref{rule: flush substitutions}),
	that the statement text must be reset to the original text (see~\ref{rule: reset statement text}),
	that the reset-state must be revoked (see~\ref{rule: done initializing}),
	that substitution must take place (see~\ref{rule: substitute}),
	and finally that the phrase detection is switched off again (see~\ref{rule: fill shownPhrase}).
	The initial condition occurs if a binding has been used (substituted) in a statement,
	and the binding satisfies \id{differB}.
	The following rule specifies the action that resetting a statment can start:
\[\text{RULE}\ \ident{Statement}\cap \flip{\id{inStatement}};\id{{\it differB}};\id{substituted}\ \subs\ \id{resetS} \]
	Violations of this rule are statements of which a substitution must be re-done.
	The software developer can have these violations added to \id{resetS} to satisfy this rule.
	In doing so, the chain of events is triggered that ends when all rules are satisfied.

\section{Programming in the Small}
\label{sct:Programming}
	Let us take a closer look at the programming process.
	Working in relation algebra, a software developer must think about satisfying constraints.
	She considers how violations arise from changing content of relations.
	She also thinks about insert and delete actions to resolve these violations.
	In our case study, we have reasoned with the event types: Ins $\la\id{relation}\ra$ and Del $\la\id{relation}\ra$%
\footnote{Ins and Del are called each others duals.}.
\begin{table}[htb]
{\small\begin{tabular}{|l|p{3.2in}|}\hline
rule&event types\\ \hline
signal phrase update&\small Ins \id{inStatement}, Ins \id{differB}, Ins \id{substituted}, Del \id{resetS}\\
flush substitutions&\small Ins \id{inStatement}, Ins \id{resetS}, Del \id{substituted}\\
reset statement text&\small Ins \id{resetS}, Ins \id{template}, Ins \id{descriptor}, Del \id{stmtShowText}\\
done initializing&\small Ins \id{resetS}, Del \id{inStatement}, Del \id{substituted}, Ins \id{template}, Ins \id{tmplParsedText}, Ins \id{stmtShowText}\\
substitute&\small Ins \id{phrase}, Ins \id{inStatement}, Del \id{resetS}, Del \id{substituted}\\
fill shownPhrase&\small Ins \id{substituted}, Del \id{inStatement}, Del \id{shownPhrase}, Ins \id{phrase}\\ \hline
\end{tabular}}
\caption{By which type of events can rules be violated?}
\label{fig:violation of rules}
\end{table}
	Table~\ref{fig:violation of rules} shows which event types may violate which rules.
	Recall section~\ref{sct:Programming in RA}, where the process of substitution started by updating the value of a placeholder.
	This meant doing an insert after a delete on the relation \id{phrase}, causing the updated phrase to appear in \id{differB}.
	That was signaled by the rule ``signal phrase update''.
	In general, the software developer must decide how to resolve violations that occur as a result of some event.
	This can be done by choosing among the duals of the event types from table~\ref{fig:violation of rules}.
	By systematically swapping Ins for Del and vice-versa,
	the table shows by which type of events violations can be resolved.
	To ensure progress, the software developer will pick a different relation.
	For resolving the violations in the same relation would resolve the violations by undoing the change.
    Typically, preserving invariance of a rule is done by changing one or more other relations than the one that caused the violations.
	The software developer can draw a graph that contains all information from table~\ref{fig:violation of rules}
	and the dual information.
\begin{figure}[bht]
\begin{center}
  \includegraphics[scale=.55]{eventFlow.png}
\end{center}
\caption{event flow}
\label{fig:event flow}
\end{figure}
	Figure~\ref{fig:event flow} shows part of that graph.
	A circle represents a rule, a rectangle represents an event type, and the edges correspond to the flow of values between them.
	The propagation of changes across the graph is initiated by the event sources (e.g., a user changing the set of pairs in a relation).
	The software developer can work her way through the graph, to write code to resolve violations for every event type that might occur.
	Figure~\ref{fig:event flow} shows just that part of the graph that corresponds with the case study in this article.
	In actual programs the complete graph can be much larger.

\section{Code Fragments}
\label{sct:Code Fragments}
	After the previous section, one question remains open: which code can preserve invariance and can we produce that code for arbitrary rules?
	We shall provide an answer to this question not just for the case study at hand, but for all conceivable cases.
	This section derives such code for every term in relation algebra.
	The problem of generating such code suffers from arbitrariness: there are many ways to generate correct code.
	We have not explored all possibilities, because there are too many.
	We do give one implementation for every conceivable term, pointing out where choices are to be made.

	The purpose of the code is to preserve invariance of rules by insertion and deletion into and from relations.
	Inserting (a set of) pairs $\Delta$ into a relation $r$ yields $r\cup\Delta$.
	Likewise, deleting pairs $\Delta$ from a relation $r$ yields $r-\Delta$.
	In this section we derive code for computers,
	which is built around two basic operations: insert and delete.
	The insert operation can be represented in terms of pre- and postconditions\footnote{This notation
	was conceived by Floyd and Hoare~\cite{Floyd1967,Hoare1969}.
	It is well known for making assertions about computer programs and reasoning about them.}.
\[\begin{array}{r@{~}l}
&\{\text{pre: }\code{P}=p\}\\
\code{INSERT}&\Delta\code{ IN P}\\
&\{\text{post: }\code{P}=p\cup\Delta\}\\
\end{array}\]

	The delete operation is introduced similarly:
\[\begin{array}{r@{~}l}
&\{\text{pre: }\code{P}=p\}\\
\code{DELETE}&\Delta\code{ FROM P}\\
&\{\text{post: }\code{P}=p-\Delta\}\\
\end{array}\]

	Let the following example introduce the idea of breaking and restoring invariance.
	Suppose the rule $\code{P}\subs\code{S}\cup\code{T}$ is to be kept satisfied by a computer,
	and \code{P}, \code{S}, and \code{T} are variables (memory locations) in that computer,
	which represent relations.
	We assume that before executing any code, we have $\code{P}=p$, $\code{S}=s$, $\code{T}=t$ and $\code{P}\subs\code{S}\cup\code{T}$.
	The following example shows that inserting a set of pairs, $\Delta$, in \code{P}
	may violate the rule.
	Inserting the same $\Delta$ in \code{S} restores invariance.
	In terms of code, we write:
\[\begin{array}{r@{~}l}
&\{\text{Assume $\code{P}=p$, $\code{S}=s$, $\code{T}=t$ and $\code{P}\subs\code{S}\cup\code{T}$}\}\\
\code{INSERT}&\Delta\code{ IN P}\\
&\{\text{$\code{P}=p\cup\Delta$, $\code{S}=s$, and $\code{T}=t$, and invariance is broken: $\code{P}\not\subs\code{S}\cup\code{T}$}\}\\
\code{INSERT}&\Delta\code{ IN S}\\
&\{\text{postcondition: $\code{P}=p\cup\Delta$, $\code{S}=s\cup\Delta$, $\code{T}=t$, and $\code{P}\subs\code{S}\cup\code{T}$}\}\\
\end{array}\]

	To prove that rule $\code{P}\subs\code{S}\cup\code{T}$ is kept satisfied,
	we must show that its value afterwards is equal to $\full$,
	assuming that its initial value is $\full$.
\[\begin{array}{cl}
&(p\cup\Delta)\subs(s\cup\Delta)\cup t\\
=&\\
&\cmpl{p\cup\Delta}\ \cup\ s\ \cup\ \Delta\ \cup\ t\\
=&\hspace{1in}\{\hbox{De Morgan's law}\}\\
&(\cmpl{p}\cap\cmpl{\Delta})\ \cup\ s\ \cup\ \Delta\ \cup\ t\\
=&\hspace{1in}\{\hbox{Distribute $\cup$ over $\cap$}\}\\
&(\cmpl{p}\cup s\cup\Delta\cup t)\ \cap\ (\cmpl{\Delta}\cup s\cup\Delta\cup t)\\
=&\hspace{1in}\{\cmpl{\Delta}\cup\Delta=\full\}\\
&\cmpl{p}\cup s\cup\Delta\cup t\\
=&\hspace{1in}\{\full\ =\ \cmpl{p}\cup s\cup t\}\\
&\full
\end{array}\]
	From this derivation we may conclude that invariance of $\code{P}\subs\code{S}\cup\code{T}$ is preserved.
	From this example, we can see that $\code{INSERT }\Delta\code{ IN T}$ would have worked just as well.
	In general, there are alternative ways of preserving invariance.
	This calls for a systematic analysis.
	The following sections analyze the various terms in Relation Algebra and provides some code to restore invariance of rules containing such terms.
	We start with an analysis of rules, then we proceed with boolean operators, and finally we arrive at relational operators.

\subsection{Implication Rules}
\label{ssct:Implication Rules}
	We will start by studying insertions and deletions in the simplest of rules:
\begin{equation}
\code{RULE }r\subs s\label{rule:r subs s}
\end{equation}
	Recall that $\subs$ is monotonous, so:
\begin{eqnarray}
r\subs s&\Rightarrow&r\cup\Delta\subs s\cup \Delta\label{eqn:r subs s}
\end{eqnarray}
	It immediately follows that inserting $\Delta$ into \code{S} is an adequate reaction to
	an insertion of $\Delta$ into \code{R}:
\[\begin{array}{r@{~}l}
&\{\text{Assume $\code{R}=r$, $\code{S}=s$, and $\code{R}\subs\code{S}$}\}\\
\code{INSERT}&\Delta\code{ INTO R};\\
&\{\text{$\code{R}=r\cup\Delta$, $\code{S}=s$, and $\code{R}\subs\code{S}$ is not necessarily true.}\}\\
\code{INSERT}&\Delta\code{ INTO S}\\
&\{\text{postcondition: $\code{R}=r\cup\Delta$, $\code{S}=s\cup\Delta$, and $\code{R}\subs\code{S}$}\}
\end{array}\]

	A similar line of reasoning can be used for deletion as well:
\[\begin{array}{r@{~}l}
&\{\text{Assume $\code{R}=r$, $\code{S}=s$, and $\code{R}\subs\code{S}$}\}\\
\code{DELETE}&\Delta\code{ FROM S};\\
&\{\text{$\code{R}=r$, $\code{S}=s-\Delta$, and $\code{R}\subs\code{S}$ is not necessarily true.}\}\\
\code{DELETE}&\Delta\code{ FROM R}\\
&\{\text{postcondition: $\code{R}=r-\Delta$, $\code{S}=s-\Delta$, and $\code{R}\subs\code{S}$}\}
\end{array}\]
	The proof is noteworthy similar and equally trivial:
\begin{eqnarray}
r\subs s&\Rightarrow&r-\Delta\subs s-\Delta\label{eqn:r subs s by deletion}
\end{eqnarray}
	If the event is a deletion of $\Delta$ from $r$ or an insertion into $s$, no reaction is required.
	Such events cannot violate the invariance of rule~\ref{rule:r subs s},
	because equations~\ref{eqn:r subs s delfrom r} and~\ref{eqn:r subs s insinto s} hold for any $\Delta$:
\begin{eqnarray}
r\subs s&\Rightarrow&r-\Delta\subs s\label{eqn:r subs s delfrom r}\\
r\subs s&\Rightarrow&r\subs s\cup\Delta\label{eqn:r subs s insinto s}
\end{eqnarray}

Summarizing, some code for preserving invariance of rules of form $\code{R}\subs\code{S}$ is:
\[\begin{array}{|l|l|l|l|}
\hline
\text{invariant}&\text{action}&\text{reaction}&\text{equation}\\ \hline
\full=\code{R}\subs\code{S}&\code{INSERT}\ \Delta\code{ INTO R}&\code{INSERT }\Delta\code{ INTO S}&\ref{eqn:r subs s}\\
\full=\code{R}\subs\code{S}&\code{DELETE}\ \Delta\code{ FROM S}&\code{DELETE }\Delta\code{ FROM R}&\ref{eqn:r subs s by deletion}\\
\full=\code{R}\subs\code{S}&\code{DELETE}\ \Delta\code{ FROM R}&\code{Nop}&\ref{eqn:r subs s delfrom r}\\
\full=\code{R}\subs\code{S}&\code{INSERT}\ \Delta\code{ INTO S}&\code{Nop}&\ref{eqn:r subs s insinto s}\\ \hline
\end{array}\]
Preserving an invariant of the form $\code{R}=\code{S}$ is equivalent to preserving both $\code{R}\subs\code{S}$ and $\code{S}\subs\code{R}$ simultaneously.

\subsection{Rules with Boolean Operators}
	To write code for boolean expressions,
	we revert to the following lemma's:
\begin{eqnarray}
r=\cmpl{s}&\Rightarrow&r\cup\Delta=\cmpl{s-\Delta}\label{lbl:Preserve r = -s with Ins Delta into r}\\
r=\cmpl{s}&\Rightarrow&r-\Delta=\cmpl{s\cup\Delta}\label{lbl:Preserve r = -s with Del Delta from r}\\
t=r\cap s&\Rightarrow&t\cup\Delta=(r\cup\Delta)\cap(s\cup\Delta)\label{lbl:Preserve t=r inter s with Ins Delta into t}\\
t=r\cup s&\Rightarrow&t-\Delta=(r-\Delta)\cup(s-\Delta)\label{lbl:Preserve t=r union s with Del Delta from t}\\
t=r\cap s&\Rightarrow&t-\Delta=(r-\Delta)\cap s\label{lbl:Preserve t=r inter s with Del Delta from t}\\
t=r\cup s&\Rightarrow&t\cup\Delta=(r\cup\Delta)\cup s\label{lbl:Preserve t=r union s with Ins Delta into t}
\end{eqnarray}
	In proof assistant Isabelle, these lemmas are proven in zero steps, so we omit these proofs here.

	Equations~\ref{lbl:Preserve r = -s with Ins Delta into r} and~\ref{lbl:Preserve r = -s with Del Delta from r}
	show that insertion in a term means deletion from its complement and vice versa.
	So the programmer can write the following code:
\[\begin{array}{r@{~}l}
&\{\text{Assume $\code{R}=r$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$}\}\\
\code{INSERT}&\Delta\code{ INTO R};\\
&\{\text{$\code{R}=r\cup\Delta$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$ is not necessarily true.}\}\\
\code{DELETE}&\Delta\code{ FROM S}\\
&\{\text{$\code{R}=r\cup\Delta$ and $\code{S}=s-\Delta$, so by equation~\ref{lbl:Preserve r = -s with Ins Delta into r} we have $\code{R}=\cmpl{\code{S}}$}\}
\end{array}\]
	The proofs for the following table are similar and equally simple,
	so we omit them for the sake of brevity.
% 	Likewise, the reaction on a deletion will be an insertion:
% \[\begin{array}{r@{~}l}
% &\{\text{Assume $\code{R}=r$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$}\}\\
% \code{DELETE}&\Delta\code{ FROM R};\\
% &\{\text{$\code{R}=r-\Delta$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$ is not necessarily true.}\}\\
% \code{INSERT}&\Delta\code{ INTO S}\\
% &\{\code{R}=r-\Delta\text{ and }\code{S}=s\cup\Delta\text{, so by equation~\ref{lbl:Preserve r = -s with Del Delta from r} we have }\code{R}=\cmpl{\code{S}}\}
% \end{array}\]
% 
% 	Now consider equations~\ref{lbl:Preserve t=r inter s with Ins Delta into t} and~\ref{lbl:Preserve t=r union s with Del Delta from t}.
% 	If \code{T} changes, both \code{R} and \code{S} must be changed accordingly.
% 	Since $\cup$ and $\cap$ are associative operators, the arguments can be evaluated with no interaction between them.
% 	This allows us to define a primitive, \code{ALL}, without specifying in which order it processes its arguments\footnote{... leaving room to take advantage of parallelism.}.
% 	We define \code{ALL} to perform all actions from a given list of actions \code{AS}:
% \[\begin{array}{r@{~}l}
% &\{\text{Assume }\forall\code{A}\in\code{AS}: (\{\id{pre}_{\code{\tiny A}}\}\ \code{A}\ \{\id{post}_{\code{\tiny A}}\})\text{ and } \id{pre}_{\code{\tiny A}} \}\\
% \code{ALL}&\ \code{AS}\\
% &\{\forall\code{A}: \id{post}_{\code{\tiny A}}\}
% \end{array}\]
% 	With this primitive, we can write code for inserting pairs in \code{T},
% 	using equation~\ref{lbl:Preserve t=r inter s with Ins Delta into t}:
% \[\begin{array}{r@{~}l}
% &\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cap\code{S}$}\}\\
% \code{INSERT}&\Delta\code{ INTO T};\\
% &\{\text{$\code{T}=t\cup\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cap\code{S}$}\}\\
% \code{ALL}&[\code{INSERT}\ \Delta\code{ INTO R},\code{INSERT}\ \Delta\code{ INTO S}]\\
% &\{\code{R}=r\cup\Delta\text{ and }\code{S}=s\cup\Delta\text{, so by equation~\ref{lbl:Preserve t=r inter s with Ins Delta into t} we have }\code{T}=\code{R}\cap\code{S}\}
% \end{array}\]
% 	Similarly, we can derive code for equation~\ref{lbl:Preserve t=r union s with Del Delta from t}:
% \[\begin{array}{r@{~}l}
% &\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cup\code{S}$}\}\\
% \code{DELETE}&\Delta\code{ FROM T};\\
% &\{\text{$\code{T}=t-\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cup\code{S}$}\}\\
% \code{ALL}&[\code{DELETE}\ \Delta\code{ FROM R},\code{DELETE}\ \Delta\code{ FROM S}]\\
% &\{\code{R}=r-\Delta\text{ and }\code{S}=s-\Delta\text{, so by equation~\ref{lbl:Preserve t=r union s with Del Delta from t} we have }\code{T}=\code{R}\cup\code{S}\}
% \end{array}\]
% 
% 	In equations~\ref{lbl:Preserve t=r inter s with Del Delta from t} and~\ref{lbl:Preserve t=r union s with Ins Delta into t},
% 	a choice must be made between $r$ and $s$. For that purpose we need a primitive \code{CHC} that chooses one action from a list $\code{AS}$, say $\code{A}$:
% \[\begin{array}{r@{~}l}
% &\{ (\{\id{pre}_{\code{\tiny A}}\}\ \code{A}\ \{\id{post}_{\code{\tiny A}}\})\text{ and } \id{pre}_{\code{\tiny A}} \}\\
% \code{CHC}&\ \code{AS}\\
% &\{\id{post}_{\code{\tiny A}}\}
% \end{array}\]
% 	Note that this definition leaves open which term is selected.
% 	The following code shows how this works out for equation~\ref{lbl:Preserve t=r inter s with Del Delta from t}:
% \[\begin{array}{r@{~}l}
% &\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cap\code{S}$}\}\\
% \code{DELETE}&\Delta\code{ FROM T};\\
% &\{\text{$\code{T}=t-\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cap\code{S}$}\}\\
% \code{CHC}&[\code{DELETE}\ \Delta\code{ FROM R},\code{DELETE}\ \Delta\code{ FROM S}]\\
% &\{\text{Either }\code{R}=r-\Delta\text{ or }\code{S}=s-\Delta\text{, so by equation~\ref{lbl:Preserve t=r inter s with Del Delta from t} we have }\code{T}=\code{R}\cap\code{S}\}
% \end{array}\]
% 
% 	The following code shows how this works out for equation~\ref{lbl:Preserve t=r union s with Ins Delta into t}:
% \[\begin{array}{r@{~}l}
% &\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cup\code{S}$}\}\\
% \code{INSERT}&\Delta\code{ INTO T};\\
% &\{\text{$\code{T}=t\cup\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cup\code{S}$}\}\\
% \code{CHC}&[\code{INSERT}\ \Delta\code{ INTO R},\code{INSERT}\ \Delta\code{ INTO S}]\\
% &\{\text{Either }\code{R}=r\cup\Delta\text{ or }\code{S}=s\cup\Delta\text{, so by equation~\ref{lbl:Preserve t=r union s with Ins Delta into t} we have }\code{T}=\code{R}\cup\code{S}\}
% \end{array}\]
% 
% 	Section~\ref{ssct:Implication Rules} treated rules of the form $\code{R}\subs\code{S}$.
% 	Note that this rule can be rewritten to $\cmpl{\code{R}}\cup\code{S}=\full$.
% 	Consequently, rules did not have to be treated separately.
% 	We could have treated them as a special case of equations with boolean operators only.
% 	In fact, every rule with boolean operators $\cup$, $\cap$, and $-$ can be rewritten as a number of rules in disjunctive normal form\footnote{%
% 	Normalizing a boolean algebra is existing knowledge, and is considered outside the scope of this paper.} (DNF~\cite{davey1990}).
% 	We can represent a rule in DNF informally by:
% \begin{eqnarray}
% \code{RULE }\cmpl{r_0}\cup\ldots\cup\cmpl{r_n}\cup s_0\cup\ldots\cup s_m
% \end{eqnarray}
% 	or more formally by
% \begin{eqnarray}
% \full=\cmpl{\bigcap\id{antecedents}}\cup\bigcup\id{consequents}
% \end{eqnarray}
% 	in which \id{antecedents} are the complemented (negative) terms and \id{consequents} are the non-complemented (positive) terms.
% 
	To preserve invariants in a boolean algebra is done as follows:
\[\begin{array}{|l|l|l|l|}
\hline
\text{invariant}&\text{action}&\text{reaction}&\text{equation}\\ \hline
\code{R}=\cmpl{\code{S}}&\code{INSERT}\ \Delta\code{ INTO R}&\code{DELETE }\Delta\code{ FROM S}&\ref{lbl:Preserve r = -s with Ins Delta into r}\\
\code{R}=\cmpl{\code{S}}&\code{DELETE}\ \Delta\code{ FROM R}&\code{INSERT }\Delta\code{ INTO S}&\ref{lbl:Preserve r = -s with Del Delta from r}\\
\code{T}=\bigcap\code{TERMS}&\code{INSERT}\ \Delta\code{ INTO T}&\code{ALL }[\code{INSERT}\ \Delta\code{ INTO X}|\code{X}\in\code{TERMS}]&\ref{lbl:Preserve t=r inter s with Ins Delta into t}\\
\code{T}=\bigcup\code{TERMS}&\code{DELETE}\ \Delta\code{ FROM T}&\code{ALL }[\code{DELETE}\ \Delta\code{ FROM X}|\code{X}\in\code{TERMS}]&\ref{lbl:Preserve t=r union s with Del Delta from t}\\
\code{T}=\bigcap\code{TERMS}&\code{DELETE}\ \Delta\code{ FROM T}&\code{CHC }[\code{DELETE}\ \Delta\code{ FROM X}|\code{X}\in\code{TERMS}]&\ref{lbl:Preserve t=r inter s with Del Delta from t}\\
\code{T}=\bigcup\code{TERMS}&\code{INSERT}\ \Delta\code{ INTO T}&\code{CHC }[\code{INSERT}\ \Delta\code{ INTO X}|\code{X}\in\code{TERMS}]&\ref{lbl:Preserve t=r union s with Ins Delta into t}\\ \hline
\end{array}\]
	So let us now turn to code that is needed when relational operators are involved.

\subsection{Rules with relational operators}
	This section studies how to insert pairs into a term $r;s$.
	To formulate this precisely, we introduce an auxiliary relation $t$ to represent the term $r;s$.
	First we will insert pairs in $t$ while preserving invariance of $t=r;s$
	Then we will insert pairs in $r$ preserving the same rule.

	To insert pairs $\Delta$ into $t$ involves finding relations $\Delta_r$ and $\Delta_s$
	that satisfy equation~\ref{eqn:t subs r;s}.
\begin{equation}
t=r;s\ \Rightarrow\ t\cup\Delta=(r\cup \Delta_r);(s\cup \Delta_s)\label{eqn:t subs r;s}
\end{equation}
	Let us define a software primitive to compute both $\Delta_r$ and $\Delta_s$ and call it \code{NEWPAIRS}.
	It computes $\Delta_r$ and $\Delta_s$ from $r$, $s$, and $\Delta$.
	We can derive its postconditions from equation~\ref{eqn:t subs r;s} by means of the following derivation:
\[\begin{array}{r@{~}l}
&t\cup\Delta\\
=&\hspace{1in}\{\text{since }t\cup\Delta=t\cup(\Delta-t)\}\\
&t\cup(\Delta-t)\\
=&\hspace{1in}\{t=r;s\text{ and let \code{NEWPAIRS} ensure that }\Delta_r;\Delta_s=\Delta-t\}\\
&r;s\cup\Delta_r;\Delta_s\\
=&\hspace{1in}\{\text{let \code{NEWPAIRS} ensure: }r;\Delta_s\subs t\cup\Delta\text{ and }\Delta_r;s\subs t\cup\Delta\}\\
&r;s\cup r;\Delta_s\cup \Delta_r;s\cup \Delta_r;\Delta_s\\
=&\hspace{1in}\{\text{distributivity}\}\\
&(r\cup \Delta_r);(s\cup \Delta_s)
\end{array}\]
	The conditions in this derivation must be satisfied by \code{NEWPAIRS}. 
	This yields the following specification of \code{NEWPAIRS}:
\[\begin{array}{r@{~}l}
&\{\code{R}=r\text{ and }\code{S}=s\text{ and }\code{DELTA}=\Delta\}\\
\code{NEWPAIRS}&(\code{R}, \code{S}, \code{DELTA}): (\code{DELTAR}, \code{DELTAS})\\
&\begin{array}[t]{@{}l@{}l}
        \{&\code{DELTR};\code{DELTAS}=\Delta - r;s\text{ and }\\
          &\code{R};\code{DELTAS}\subs r;s\cup\Delta\text{ and }\code{DELTAR};\code{S}\subs r;s\cup\Delta\}
\end{array}
\end{array}\]
	The pre- and postconditions of \code{NEWPAIRS} are sufficient to satisfy equation~\ref{eqn:t subs r;s}.
	Since $t$ does not occur in the pre- and postconditions, \code{NEWPAIRS} implements equation~\ref{eqn:ins delta r;s}:
\begin{equation}
r;s\cup\Delta=(r\cup \Delta_r);(s\cup \Delta_s)\label{eqn:ins delta r;s}
\end{equation}
	Note that the specification of \code{NEWPAIRS} does not determine its implementation uniquely.
	There are multiple ways to implement \code{NEWPAIRS}.
	If \code{NEWPAIRS} creates one new pair for every pair in $\Delta-t$,
	we get $r;\Delta_s=\emptyset$ and $\Delta_r;s=\emptyset$.
	The weaker condition from the derivation, i.e. $r;\Delta_s\subs t\cup\Delta$ and $\Delta_r;s\subs t\cup\Delta$ allows some reuse of existing pairs.
	However, any implementation that preserves invariance of equation~\ref{eqn:ins delta r;s} will do.

	At this point we have the ingredients for code that preserves invariance of $\code{T}=\code{R};\code{S}$
\[\begin{array}{l}
\hspace{0.5in}\{\text{Assume }\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=t\text{ and }\code{T}=\code{R};\code{S}\}\\
\code{INSERT }\Delta\code{ INTO T};\\
\hspace{0.5in}\{\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=t\cup\Delta\text{, so }\code{T}\neq\code{R};\code{S}\}\\
(\code{DELTAR},\code{DELTAS})\code{ := }\code{NEWPAIRS}(\code{R}, \code{S}, \Delta-\code{T});\\
\hspace{0.5in}\{\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=t\cup\Delta\text{ and }\\
\hspace{0.575in}\code{DELTAR};\code{DELTAS}=\Delta-\code{T}\text{ and }\\
\hspace{0.575in}\code{R};\code{DELTAS}\subs\code{T}\text{ and }\code{DELTAR};\code{S}\subs\code{T}\text{, so equation~\ref{eqn:ins delta r;s} is satisfied}\}\\
\code{INSERT DELTAR INTO R};\\
\hspace{0.5in}\{\code{R}=r\cup\code{DELTAR}\text{ and }\code{S}=s\text{ and }\code{T}=t\cup\Delta\}\\
\code{INSERT DELTAS INTO S}\\
\hspace{0.5in}\{\code{R}=r\cup\code{DELTAR}\text{ and }\code{S}=s\cup\code{DELTAS}\text{ and }\code{T}=t\cup\Delta\text{, so }\code{T}=\code{R};\code{S}\}
\end{array}\]
    This shows how to satisfy invariant $\code{T}=\code{R};\code{S}$ when pairs $\Delta$ are inserted into \code{T}.

	Let us now study how to keep $r;s$ constant when $r$ changes.
	Inserting pairs $\Delta$ into $r$ involves deleting pairs from $s$ such that:
\begin{equation}
(r\cup \Delta);(s-\Delta_s)=r;s\label{eqn:ins r del s}
\end{equation}
	In order to solve equation~\ref{eqn:ins r del s} towards $\Delta_s$,
	we derive\footnote{Thanks to Peter Jipsen, who pointed out this line of reasoning.}:
\begin{eqnarray}
&&(r\cup \Delta);(s-\Delta_s)\ \subs\ r;s\label{eqn:rs subs t by ins r del s}\\
&\Leftrightarrow&\nonumber\\
&&s-\Delta_s\ \subs\ (r\cup \Delta)\backslash(r;s)\nonumber\\
&\Leftrightarrow&\hspace{1in}\{\text{apply }x-y=z\ \Leftrightarrow\ x=z\cup y\text{ twice.}\}\nonumber\\
&&s-(r\cup \Delta)\backslash(r;s)\ \subs\ \Delta_s\nonumber
\end{eqnarray}
	This derivation shows that every solution $\Delta_s$ contains $s-(r\cup \Delta)\backslash(r;s)$.
	So the latter, $s-(r\cup \Delta)\backslash(r;s)$, is the smallest solution.

	The smallest solution might not contain all pairs from $r;s$.
	The missing pairs are $r;s-(r\cup\Delta);(s-(r\cup \Delta)\backslash(r;s))$.
	Therefore it may be necessary to reintroduce these missing pairs afterwards.
	Fortunately, we have just shown how to do just that with \code{NEWPAIRS}.
	This provides us with all the ingredients for code that preserves invariance of $\code{R};\code{S}=\code{T}$
\[\begin{array}{l}
\hspace{0.5in}\{\text{Assume }\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=r;s\text{, so }\code{R};\code{S}=\code{T}\}\\
\code{INSERT }\Delta\code{ INTO R};\\
\hspace{0.5in}\{\code{R}=r\cup\Delta\text{ and }\code{S}=s\text{ and }\code{T}=r;s\text{, so }\code{R};\code{S}\neq\code{T}\}\\
\code{DELETE R}\backslash\code{T FROM S};\\
\hspace{0.5in}\{\text{Since }\code{R}=r\cup\Delta\text{ and }\code{T}=r;s\text{, we have }\code{R}\backslash\code{T}=(r\cup \Delta)\backslash(r;s)\\
\hspace{0.575in}\text{So }\code{S}=s-(r\cup \Delta)\backslash(r;s)\text{ and therefore}\\
\hspace{0.575in}\code{R};\code{S}\subs r;s\}\\
(\code{DELTAR},\code{DELTAS})\code{ := }\code{NEWPAIRS}(\code{R},\ \code{S},\ \code{T}-\code{R};\code{S});\\
\code{INSERT DELTAR INTO R};\\
\code{INSERT DELTAS INTO S}\\
\hspace{0.5in}\{\code{R};\code{S}=r;s\cup(\code{T}-\code{R};\code{S})=r;s\\
\hspace{0.575in}\text{so }\code{R};\code{S}=\code{T}\}
\end{array}\]

	Let us now study how to delete pairs $\Delta$ from $r$ while keeping $r;s$ constant.
	This involves inserting pairs into $s$ such that:
\begin{equation}
(r-\Delta);(s\cup\Delta_s)=r;s\label{eqn:del r ins s}
\end{equation}
	The largest $\Delta_s$ follows from the law $x;(x\backslash y)\subs y$ for $x=r-\Delta$ and $y=r;s$:
\begin{eqnarray}
&&(r-\Delta);((r-\Delta)\backslash(r;s))\ \subs\ r;s\nonumber\\
&\Rightarrow&\hspace{1in}\{\text{Let }\Delta_s\ =\ (r-\Delta)\backslash(r;s)\}\nonumber\\
&&(r-\Delta);\Delta_s\ \subs\ r;s\nonumber\\
&\Leftrightarrow&\hspace{1in}\{p;x\subs p;q\ \Leftrightarrow\ p;(q\cup x)\subs p;q\}\nonumber\\
&&(r-\Delta);(s\cup\Delta_s)\ \subs\ r;s\nonumber
\end{eqnarray}
	This derivation shows that we can choose $\Delta_s=(r-\Delta)\backslash(r;s)$ as the largest set that preserves invariant~\ref{eqn:del r ins s}.

	Once again, we might not have all pairs from $r;s$,
	so equation~\ref{eqn:del r ins s} needs to be completed with the missing pairs.
	The missing pairs are $r;s-(r-\Delta);(s\cup(r-\Delta)\backslash(r;s))$.
	In this case too, these missing pairs can be reintroduced afterwards.
	This provides us with all the ingredients for code that preserves invariance of $\code{R};\code{S}=\code{T}$
\[\begin{array}{l}
\hspace{0.5in}\{\text{Assume }\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=r;s\text{, so }\code{R};\code{S}=\code{T}\}\\
\code{DELETE }\Delta\code{ FROM R};\\
\hspace{0.5in}\{\code{R}=r-\Delta\text{ and }\code{S}=s\text{ and }\code{T}=r;s\text{, so possibly }\code{R};\code{S}\neq\code{T}\}\\
\code{INSERT R}\backslash\code{T INTO S};\\
\hspace{0.5in}\{\text{Since }\code{R}=r-\Delta\text{ and }\code{T}=r;s\text{, we have }\code{R}\backslash\code{T}=(r-\Delta)\backslash(r;s)\\
\hspace{0.575in}\text{So }\code{S}=s\cup(r-\Delta)\backslash(r;s)\text{ and therefore}\\
\hspace{0.575in}\code{R};\code{S}\subs r;s\}\\
(\code{DELTAR},\code{DELTAS})\code{ := }\code{NEWPAIRS}(\code{R},\ \code{S},\ \code{T}-\code{R};\code{S});\\
\code{INSERT DELTAR INTO R};\\
\code{INSERT DELTAS INTO S}\\
\hspace{0.5in}\{\code{R};\code{S}=r;s\cup(\code{T}-\code{R};\code{S})=r;s\\
\hspace{0.575in}\text{so }\code{R};\code{S}=\code{T}\}
\end{array}\]

	Summarizing, we have obtained some more code to preserve invariants in a relation algebra:
\[\begin{array}{|l|l|l|l|}
\hline
\text{invariants}&\text{action}&\text{reaction}&\text{equation}\\ \hline
\begin{array}{l}
\code{R};\code{S}\subs\code{T}\\
\code{S}\subs\code{R}\backslash\code{T}\\
\code{R}\subs\code{T}\slash\code{S}
\end{array}&\code{INSERT}\ \Delta\code{ INTO R}&\code{DELETE R}\backslash\code{T FROM S}&\ref{eqn:rs subs t by ins r del s}\\ \hline
\begin{array}{l}
\code{R};\code{S}\subs\code{T}\\
\code{S}\subs\code{R}\backslash\code{T}\\
\code{R}\subs\code{T}\slash\code{S}
\end{array}&\code{DELETE}\ \Delta\code{ FROM R}&\code{Nop}&\\ \hline
\code{T}\subs\code{R};\code{S}&\code{INSERT}\ \Delta\code{ INTO T}&
\begin{array}{l}
(\code{DELTAR},\code{DELTAS})\code{:=NEWPAIRS}(\code{R},\ \code{S},\ \Delta-\code{T});\\
\code{INSERT DELTAR INTO R};\\
\code{INSERT DELTAS INTO S}
\end{array}&\ref{eqn:ins delta r;s}\\ \hline
\code{T}\subs\code{R};\code{S}&\code{DELETE}\ \Delta\code{ FROM T}&\code{Nop}&\\ \hline
\code{T}\subs\code{R};\code{S}&\code{INSERT}\ \Delta\code{ INTO R}&\code{Nop}&\\ \hline
\code{T}\subs\code{R};\code{S}&\code{DELETE}\ \Delta\code{ FROM R}&\code{INSERT R}\backslash\code{T INTO S}&\\ \hline
\code{R};\code{S}=\code{T}&\code{INSERT}\ \Delta\code{ INTO R}&
\begin{array}{l}
\code{DELETE R}\backslash\code{T FROM S};\\
(\code{DELTAR},\code{DELTAS})\code{:=NEWPAIRS}(\code{R},\ \code{S},\ \code{T}-\code{R};\code{S});\\
\code{INSERT DELTAR INTO R};\\
\code{INSERT DELTAS INTO S}
\end{array}&\ref{eqn:ins r del s}\\ \hline
\code{R};\code{S}=\code{T}&\code{DELETE}\ \Delta\code{ FROM R}&
\begin{array}{l}
\code{INSERT R}\backslash\code{T INTO S};\\
(\code{DELTAR},\code{DELTAS})\code{:=NEWPAIRS}(\code{R},\ \code{S},\ \code{T}-\code{R};\code{S});\\
\code{INSERT DELTAR INTO R};\\
\code{INSERT DELTAS INTO S}
\end{array}&\ref{eqn:del r ins s}\\ \hline
\end{array}\]
	The solutions given here are far from being exhaustive.
	We illustrate this point by giving an example.
	Suppose we have invariant $\code{T}\subs\code{R};\code{S}$ and action $\code{DELETE}\ \Delta\code{ FROM R}$.
	So the question is which pairs must be added to $s$ after pairs $\Delta$ have been removed from $r$.
	In normal circumstances, we might just look it up in the table above, picking $\code{INSERT R}\backslash\code{T INTO S}$ as a reaction.
	In the circumstance that initially $\code{T}=r;s$, we might reason differently.
	We must still find a smallest value for $\Delta_s$ that satisfies:
\begin{equation}
r;s\ \subs\ (r-\Delta);(s\cup\Delta_s)\label{lbl:nonshrinking r;s when Del from r}
\end{equation}
	The following derivation shows that $\flip{(r-\Delta)};(r\cap\Delta);s$ satisfies this requirement conditionally.
	Here is the derivation:
\begin{eqnarray}
&&\flip{(r-\Delta)};(r\cap\Delta);s\ \subs\ \Delta_s\nonumber\\
&\Rightarrow\nonumber\\
&&(r-\Delta);\flip{(r-\Delta)};(r\cap\Delta);s\ \subs\ (r-\Delta);\Delta_s\nonumber\\
&\Rightarrow&\hspace{1in}\{r;s;\flip{s};\flip{r} \cap\iden\ \ \subs\ (r-\Delta);\flip{(r-\Delta)}\}\nonumber\\
&&(r\cap\Delta);s\ \subs\ (r-\Delta);\Delta_s\nonumber\\
&\Leftrightarrow\nonumber\\
&&(r-(r-\Delta));s\ \subs\ (r-\Delta);\Delta_s\nonumber\\
&\Rightarrow\nonumber\\
&&(r;s)-((r-\Delta);s)\ \subs\ (r-\Delta);\Delta_s\nonumber\\
&\Leftrightarrow\nonumber\\
&&r;s\ \subs\ (r-\Delta);s \cup (r-\Delta);\Delta_s\nonumber\\
&\Leftrightarrow\nonumber\\
&&r;s\ \subs\ (r-\Delta);(s \cup \Delta_s)\nonumber
\end{eqnarray}
	This derivation justifies the reaction $\code{INSERT }\flip{(r-\Delta)};(r\cap\Delta);s\code{ INTO S}$,
	but only if $r;s;\flip{s};\flip{r} \cap\iden\ \ \subs\ (r-\Delta);\flip{(r-\Delta)}$.
	The part $r;s;\flip{s};\flip{r}$ is not necessary in this proof, but it is useful in practice.
	It says that $r-\Delta$ must contain a pair for every element in the domain of $r;s$,
	instead of every element in $I$.
	This makes this result applicable in more situations.

\section{Reflection}
\label{sct:Reflection}
	Ampersand is being built to automate software development.
	The holy grail is to derive deterministic ways to preserve invariants, to allow programming by constraints.
	This is motivated by the benefits of describing a result in terms of constraints,
	rather than having to invent all the steps that lead to that result.
	For example, describing the situation that any financial transaction must preserve the total amount of money is simpler
	than having to describe all the steps involved in such a transaction.
	The attraction is not only for the programmer, but even more so for auditors and users for who the constraints are easier to grasp and validate.
	In this paper we have shown how to derive specific steps,
	helping the programmer to invent code and providing mathematical assurance of correctness (compliance-by-design).

	The analysis in this paper shows once more that inventing algorithms for preserving invariants is far from trivial.
	Algorithms cannot be derived from constraints alone. The choices that remain must be resolved first.
	This paper does show, however, that these choices occur in predictable ways.
	That gives opportunities, for instance to make these choices explicit to a user, to optimize the number of choices,
	or to use external knowledge to resolve the remaining choices at run time.

	In practice we have experienced that programming by constraints in relation algebra yields an unconventional way of programming.
	This experience consists of inventing rules and choosing event types for resolving violations, as illustrated by the case study.
	To relate this experience to programming as we know it, section~\ref{sct:Comparison} makes a comparison with established programming languages.
	Section~\ref{sct:Contribution} summarizes the contributions to software development claimed by Ampersand.
	Section~\ref{sct:Ampersand in practice} summarizes the use of Ampersand in practice and
	section~\ref{sct:Further Research} gives an outlook on research that is required in the near future.

\subsection{Comparison}
\label{sct:Comparison}
	This section compares Ampersand with existing programming paradigms by mentioning the most important differences and similarities.

	The first to compare with is the object oriented programming paradigm, known from popular languages such as Java~\cite{Java} and C++~\cite{Stroustrup97a}.
	Ampersand cannot be called object oriented, because the notion of control flow in object oriented languages is very different.
	In those languages, the threads in the control flow are controlled by the software developer.
	As shown in our case study, the control flow in an Ampersand script emerges as a result of changes in relation content as illustrated
	by figure~\ref{fig:event flow}.

	Ampersand also differs from the logic programming paradigm of Prolog~\cite{Lloyd1984}
	and of all rule engines that can be considered to be offspring of Prolog.
	A difference lies in the way rules are interpreted.
	In logic programming, a program consists of Horn-clauses and a resolution-proof is constructed at run time.
	The software developer works with notions such as backtracking (backward chaining) and unification, which are absent in Ampersand.
	On the other hand, Ampersand is not restricted to implications with one predicate on the consequent side, as Prolog does.
        Ampersand allows rules, which are equations with arbitrary relational expressions on either side of the equals sign.

	Ampersand also differs from functional programming, of which Haskell and Scala are prominent representants.
	A core idea in functional programming is to evaluate a program as a function from input to output~\cite{Backus1978}.
	A functional program consists of function definitions, that are evaluated by a term- or graph-rewriter,
	using various strategies such as lazy or eager evaluation.
	In contrast, our case study illustrates that not everything has to be a function.
	In relation algebra, everything is a relation and a function is a restricted form of a relation.
	A similarity to functional programming is shown by the declarative style,
	because substitution of equal terms without changing the semantics is a property we see in both worlds.

	A difference with database programming is found in the type of algebra that is used.
	Conventional databases are typically programmed in SQL, which is based on n-ary relations in relational algebra~\cite{Codd70}.
	In contrast, Ampersand uses binary relations%
\footnote{At runtime, Ampersand uses the efficiency of n-ary tables. It does so by translating relation algebra to efficient SQL-code by avoiding as many joins as possible.}
	based on heterogeneous relation algebra~\cite{Schmidt1997}.
	In this comparison, Ampersand is closer to graph databases~\cite{Vicknair2010} than to relational databases.

	In the tradition of formal specification, there are many relational approaches, such as Z~\cite{Z}, CSP~\cite{CSP}, LOTOS~\cite{LOTOS}, VDM~\cite{VDM}.
	Where formal specification techniques typically analyse and diagnose specifications,
	Ampersand actually synthesizes (generates) information systems.

	The similarity with the reactive programming paradigm is obvious from figure~\ref{fig:event flow}.
	This graph is part of the dependency graph, which is a familiar phenomenon in reactive programming.
	Ampersand may be seen as a reactive programming language on binary relations.
	It has a unidirectional push based evaluation model, static typing, implicit lifting, and glitch avoidance~\cite{Bainomugisha2013}.
	Even though Ampersand generates code to execute scripts,
	code fragments described in section~\ref{sct:Code Fragments}
	might just as well be generated towards a mature reactive programming language and executed there%
\footnote{Currently Ampersand generates towards classical PHP+HTML+JavaScript code,
	but generation to a mature reactive programming language is definitely on the wishlist.}.
	A distinguishing feature of Ampersand is that it treats constraints (rules) as its central notion,
	which is also seen in TBAG~\cite{Elliott94} and FHM~\cite{Giorgidze2011} but not common in most reactive programming languages.
	In the author's opinion a constraint based approach is preferrable, because the correctness of a program is within closer range.

	If Ampersand represents a programming style at all, we might call it a relation-algebraic style of programming or maybe constraint programming\footnote{https://en.wikipedia.org/wiki/Constraint\_programming}.
	Ampersand's style would be characterized by a programmer who specifies constraints and a computer trying to satisfy these constraints by resolving violations.

\subsection{Contribution}
\label{sct:Contribution}
	Contributions of Ampersand to the software development process are:
\begin{itemize}
\item	Ampersand has the usual benefits of a declarative language:
	This means that terms can be manipulated by Tarski's axioms without changing their semantics~\cite{vdWoude2011}.
	It also means that the order in which rules are written has no consequence for their semantics.
\item	Heterogeneous relation algebra has a straightforward interpretation in natural language~\cite{RBD}.
	We have used that to formalize business requirements without exposing business stakeholders to any formal notation.
\item	Ampersand is statically typed~\cite{Joosten2015}, which is an obvious consequence of being heterogeneous.
	There is much evidence that static typing yields significantly lower software maintenance cost as opposed to dynamic typing~\cite{HanenbergKRTS14,Petersen2014}.
\item	Heterogeneous relation algebra is well studied.
	As a consequence, many tools that are readily available in the public domain can be put to good use.
	For executives of large organizations it can be reassuring that the formalism is well studied and understood.
\item	Relation algebra facilitates composing software from reusable components, because a program consists of rules.
	Since the union of sets of rules is a set of rules, compositionality comes from the union operator.
	In practice, when components are brought together in larger assemblies, hardly any adjustments have to be made%
\footnote{This is (unsubstantiated) experience collected from projects we have done with Ampersand.}.
\end{itemize}

	Ampersand also has a disadvantage. It appears to be difficult to learn for large groups of software professionals.
	Research~\cite{Michels2015} shows that this is largely due to deficits in prerequisite knowledge, especially skills in discrete mathematics.
%	For this reason, we have made an addendum to the course we teach at the Open University of the Netherlands.
	Also, programming appears to be difficult in practice.
	Ongoing research tries to find better ways for learning, which is necessary for large scale use.

\subsection{Ampersand in Practice}
\label{sct:Ampersand in practice}
	Ampersand has been used in practice both in education (Open University of the Netherlands)
	and in industry (Ordina and TNO).
	For example, Ordina designed a proof-of-concept in 2007 of the INDIGO-system,
	which is still in use today as the core information system of the Dutch immigration authority, IND.
	The design of INDIGO was made with Ampersand, to obtain correct, detailed results in the least amount of time.
	More recently, Ampersand was used to design an information system called DTV for the Dutch food authority, NVWA.
	A prototype of DTV was built by Ampersand and was used as a model to build the actual system.
	TNO, a major Dutch industrial research laboratory, is using Ampersand for research purposes.
	For example, TNO did a study of international standardizations efforts such as
	RBAC (Role Based Access Control) in 2003 and architecture (IEEE 1471-2000)~\cite{IEEE1471} in 2004.
	Several inconsistencies were found in the last (draft) RBAC standard~\cite{RBAC}.
	TNO has also used the technique in conceiving several patents%
\footnote{e.g. patents DE60218042D, WO2006126875, EP1727327, WO2004046848, EP1563361, NL1023394C, EP1420323, WO03007571, and NL1013450C.}.
	At the Open University of the Netherlands, Ampersand is being taught in a course called Rule Based Design~\cite{RBD}.
	In this course, students use a platform called RAP, which has been built in Ampersand~\cite{Michels2015}.
	RAP has been the first Ampersand-application that has run in production.

\subsection{Further Research}
\label{sct:Further Research}
	Further research on this topic is required to bring relation algebra still closer to the community of practitioners.
	Further use of relation algebra can be made by incorporating a model checker, such as the Alloy analyser~\cite{Alloy2006}, to detect inconsistent rules.
	An exciting new development is Amperspiegel~\cite{Amperspiegel}, which brings notational flexibility to the fingertips of the user.
	Developments in the Ampersand compiler are going towards a rule-repository (written in Ampersand itself).
	This will make collaborative information systems development in Ampersand easier, because the repository can assist in automating
	the software development process further.
	Other research is needed towards a comprehensive theory of information systems.
	Currently, there is no theory (in the mathematical meaning of the word) for information systems.
	In the Ampersand project, a sub-project called ``Formal Ampersand'' is being conducted to achieve this goal.
\subsection{Acknowledgements}
\label{sct:Acknowledgements}
	Firstly, I wish to thank Sebastiaan Joosten for introducing me to Isabelle and for inspiring me with his challenging critiques.
	I owe many thanks to Wolfram Kahl, Michiel Stornebrink, Han Joosten, and the anonymous reviewers of RAMiCS for their useful feedback.
	And I am very grateful to the Open University and Ordina for funding this research.

\begin{figure}[htb]
\begin{center}
  \includegraphics[angle=90,scale=.357]{LogicalDataModel.pdf}
\end{center}
\caption{Conceptual data model (from section~\ref{sct:Conceptual analysis})}
\label{fig:conceptual model}
\end{figure}

\section{Bibliography}
\bibliographystyle{elsarticle-harv}
\bibliography{doc}

\end{document}

### 2019-06-11 2017_RAMiCS/articleRamics.tex
Further work on the article on transactions

--- 2019-06-11 2017_JLAMP/articleRamics.tex
@@
-\documentclass{elsarticle}
+\documentclass{svproc}
 \usepackage{graphicx}
-%\usepackage{multicol}
-%\usepackage{footmisc}
+\usepackage{multicol}
+\usepackage{footmisc}
 \usepackage{amstext}
@@
 \usepackage[english]{babel}
-%\usepackage[official,right]{eurosym}
+\usepackage[official,right]{eurosym}
 \selectlanguage{english}
 \include{preamramics}
-\hyphenation{ExecEngine}
 \begin{document}
 
-\title{Relation Algebra as Programming Language using the Ampersand Compiler}
-\author[ou,ordina]{Stef Joosten\fnref{fn1}}
-\ead{stef.joosten@ou.nl}
-\address[ou]{Open Universiteit Nederland, Postbus 2960, 6401 DL Heerlen, the Netherlands}
-\address[ordina]{Ordina NV,  Nieuwegein, the Netherlands}
-\fntext[fn1]{ORCID 0000-0001-8308-0189}
+\title{Software Development in Relation Algebra with Ampersand}
+\toctitle{Software Development in Relation Algebra with Ampersand}
+%\titlerunning{Ampersand} % (for running head)
+\author{Stef Joosten\inst{1,}\inst{2} (0000-0001-8308-0189)}
+%orcid.org/0000-0001-8308-0189
+\authorrunning{Stef Joosten} % (for running head)
+\tocauthor{Stef Joosten} % for table of contents
+\institute{Open Universiteit Nederland,\\  Postbus 2960,   6401 DL Heerlen, the Netherlands\\
+\and Ordina NV,  Nieuwegein, the Netherlands\\
+\email{stef.joosten@ou.nl} % optional, same for URL of homepage
+} % end of address field
+\maketitle
 
-\begin{abstract}
-	Relation Algebra can be used as a programming language for building information systems.
-	This paper demonstrates the principle by presenting a case study together with the theory behind programming in Relation Algebra.
-	As a case study, we have developed a database application for legal reasoning.
-	We discuss a small part of it to illustrate the mechanisms of programming in Relation Algebra.
-	Beside being declarative, relation algebra comes with attractive prospects for developing software.
+\abstract{Relation Algebra can be used as a programming language for building information systems.
+	This paper presents a case study to demonstrate this principle.
+	We have developed a database-application for legal reasoning as a case study,
+	of which a small part is discussed in this paper to illustrate the mechanisms of programming in Relation Algebra.
+	Beside being declarative, relation algebra comes with attractive promises for developing big software.
 	The compiler that was used for this case study, Ampersand, is the result of an open source project.
-	Ampersand has been tried and tested in practice and is available as free open source software\footnote{\tt https://github.com/AmpersandTarski/Ampersand}.
-\end{abstract}
-
-\begin{keyword}
-relation algebra\sep software development\sep legal reasoning\sep information systems design\sep Ampersand\sep MirrorMe\sep reactive programming
-\end{keyword}
-\maketitle
+	Ampersand has been tried and tested in practice and is available as free open source software.
+\keywords{relation algebra, software development, legal reasoning, information systems design, Ampersand, MirrorMe, big software}
+	}% abstract contains 102 words
 
@@
 	This paper investigates how relation algebra can be used as a programming language for information systems.
-	A compiler, Ampersand~\cite{Michels2011}, is used to compile concepts, relations and rules into a working database application.
+	A compiler, Ampersand~\cite{Michels2011}, is used to compile concepts, relations and rules into a working database-application.
 	Ampersand is a syntactically sugared version of heterogeneous relation algebra \cite{Schmidt1997}.
@@
 
-	The use of relation algebra as a programming language is rooted in relation algebra~\cite{Maddux06}, logic programming~\cite{Lloyd1984},
-	database application programming~\cite{Codd70}, and formal specification of software~\cite{Dahl93}.
+	The use of relation algebra as a programming language is not new.
+	It stands in the tradition of relation algebra~\cite{Maddux06}, logic programming~\cite{Lloyd1984},
+	database application programming~\cite{Codd70}, and formal specification of software.
 	Ampersand uses these ideas to compile relation algebra to working software, an idea which is also found in RelView~\cite{Berghammer2005} by Berghammer (Univ. of Kiel).
- 	In Ampersand the user specifies constraints. The compiler then generates an information system that respects these constraints.
-	RelView also generates software but uses relational algebra operators in an otherwise imperative programming language.
-	This makes RelView stronger in specifying complex computational problems.
-	Some more differences with earlier programming languages are discussed in section~\ref{sct:Comparison}.
-
-	The author shares the conviction that a provably correct implementation of requirements
-	is essential to the software development process~\cite{Boehm1981}.
-	Even though formal methods are not widely used in computer science,
-        Ampersand is built on the foundations laid by formal specification methods such as Z~\cite{Z} and Alloy~\cite{Alloy2006}.
-	In contrast with such methods,
-	Ampersand is equipped with a software generator that generates information systems,
-	each of which consists of a database with web-based user interfaces.
-	Thus, specifying in Ampersand is developing software at the same time.
+	Some differences with earlier programming languages are discussed in section~\ref{sct:Reflection}, after presenting the case study.
 
@@
 	It shows how the reasoning applies to facts from the case.}.
-	To the programmer, this offers the challenge to represent the problem of legal reasoning in relation algebra.
-	This challenge is by no means simple, but outside the scope of this paper.
-	In the Ampersand project at the Open University of the Netherlands,
-	we are studying practical and didactic issues of software development in relation algebra~\cite{Michels2015}.
+	The challenge is to create a program that consists of relation algebra as much as possible.
+	In doing so, we hope to learn more about software development in relation algebra.
 
-	The contribution of this paper is twofold.
-	The case study contributes to the idea that programming can be done in relation algebra.
-	The theory in section~\ref{sct:Code Fragments} contributes a theory to assist in preserving invariants.
-	Section~\ref{sct:Code Fragments} extends a paper presented at RAMiCS-2017 \cite{JoostenRAMiCS2017}
-	It provides derivations of program code,
-	all of which have been checked using the theorem prover Isabelle~\cite{Nipkow2002}.
-
-	Section~\ref{sct:Ampersand} introduces Ampersand and its semantics.
-	Section~\ref{sct:Conceptual analysis} introduces part of a theory of legal reasoning,
+	Section~\ref{sct:Ampersand} introduces Ampersand and its computational semantics.
+	Section~\ref{sct:Conceptual analysis} introduces a theory of legal reasoning,
 	which was developed for argument assistance.
@@
 	and section~\ref{sct:Programming} visualizes that mechanism.
-	Section~\ref{sct:Code Fragments} presents one way (of infinitely many) to derive code for arbitrary rules in relation algebra.
 	Section~\ref{sct:Reflection} reflects on software development in Ampersand.
@@
 	In this section we explain the basics of Ampersand.
-	The reader is expected to have sufficient background in relation algebra to understand the remainder of this paper.
+	The reader is expected to have sufficient background in relation algebra,
+	in order to understand the remainder of this paper.
 
-	The core of an Ampersand script is a triple $\la\rules,\rels,\concepts\ra$,
-	which consists of a set of rules $\rules$, relations $\rels$, and concepts $\concepts$.
-	An Ampersand script is interpreted by the compiler as an information system.
+	The core of an Ampersand-script is a tuple $\la\rules,\rels,\concepts,\mathfrak{T}\ra$, which consists of a set of rules $\rules$, relations $\rels$, concepts $\concepts$, and a type function $\mathfrak{T}$.
+	Ampersand-scripts are interpreted by the compiler as an information system.
 	The rules constitute a theory in heterogeneous relation algebra.
 	They constrain a body of data that resides in a database.
-
-	The Ampersand compiler generates a database from relations in the script.
-	A database application%
+	The Ampersand-compiler generates a database from relations in the script.
+	A database-application%
 \footnote{Ampersand generates an application that consists of a relational database and interface components.
@@
 
-	A rule is a constraint in the form of an equality between two terms.
+	A rule is an equality between two terms.
 	Terms are built from relations.
 	Ampersand interprets every relation as a finite set of pairs, which are stored in the database.
-	The phase in which Ampersand takes a script and turns it into a database, is what we will refer to as \define{compile-time}.
+	The phase in which Ampersand takes a script, and turns it into a database, is what we will refer to as \define{compile-time}.
 	The phase in which a user interacts with the database, is what we will refer to as \define{run-time}.
 	At run-time, Ampersand can decide which rules are satisfied by querying the database.
-	The compiler generates all software needed to keep rules satisfied at run-time.
-	If one of the rules is not satisfied as a result of data that has changed,
-	that change is reverted (rolled back) to the original state.
-	Either way, Ampersand ensures that all rules remain satisfied.
+	The compiler generates all software needed to maintain rules at run-time.
+	If a rule is not satisfied as a result of data that has changed, that change is reverted (rolled back) to
+	maintain a state in which all rules are satisfied.
+	Changes to the database are not specified by the software developer, but generated by the compiler.
+	Rules in Ampersand are maintained rather than executed directly.
 
@@
 	The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.
-	The expression $a \in A$ means that atom $a$ is an \emph{instance} of concept $A$.
-	The declaration of $A\isa B$ (pronounce: $A$ is a $B$)
-	in an Ampersand script states that any instance of $A$ is an instance of $B$ as well.
-	We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
+	The expression $a \in A$ means that atom $a$ is an instance of concept $A$.
+	In the syntax of Ampersand, concepts form a separate syntactic category, allowing a parser to recognize them as concepts.
+%	The declaration of $A\isa B$ (pronounce: $A$ is a $B$)
+%	in an Ampersand-script states that any instance of $A$ is an instance of $B$ as well.
+%	We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
+	Ampersand also features specialization.
 	Specialization is needed to allow statements such as: ``An orange is a fruit that ....''.
-	Specialization is not used in the remainder of this paper.
+	Specialization is not relevant for the remainder of this paper.
 
-	Relations (from set $\rels$) are used in information systems to store data.
-%	A \define{fact} is a statement that is true in a business context.
-%	Facts are stored and kept as data in a computer.
+	Relations (from set $\rels$) are used in information systems to store facts.
+	A \define{fact} is a statement that is true in a business context.
+	Facts are stored and kept as data in a computer.
 	As data changes over time, so do the contents of these relations.
 	In this paper relations are represented by variables $r$, $s$, and $d$.
-	We represent the declaration of a relation $r$ by $\declare{nm}{A}{B}$,
-	in which \id{nm} is a name and $A$ and $B$ are concepts.
+	We represent the declaration of a relation $r$ by $\declare{nm}{A}{B}$, in which \id{nm} is a name and $A$ and $B$ are concepts.
 	We call $A$ the source concept and $B$ the target concept of the relation.
-	The pair $\pair{A}{B}$ is called the \emph{type} of the relation.
 	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.
@@
         The meaning of relations in Ampersand is defined by an interpretation function $\mathfrak{I}$.
-	It maps each relation to a set of pairs.
-	The type system of Ampersand guarantees that each pair in $r$ respects the type of $r$:
+	It maps each relation to a set of facts.
+	Furthermore, it is a run-time requirement that the pairs in $r$ are contained in its type:
 \begin{equation}
@@
 \end{equation}
-	The type system is strong, which means that every atom has a type.
-	The type system is static, which means that type checking is entirely done at compile-time.
 
 	Terms are used to combine relations using operators.
-	The set of terms is called $\terms$.
+	The set of terms is called $\terms$. It is defined by:
 \begin{definition}[terms]
@@
 	d&\in&\terms         \quad\quad\text{(every relation is a term)}\\
-	(r \cup s)&\in&\terms\quad\quad\text{(union)}\\
 	(r \cap s)&\in&\terms\quad\quad\text{(intersection)}\\
@@
 	(r;s)&\in&\terms     \quad\quad\text{(composition)}\\
-	(r\backslash s)&\in&\terms     \quad\quad\text{(right residual)}\\
-	(r\slash s)&\in&\terms     \quad\quad\text{(left residual)}\\
 	\flip r&\in&\terms   \quad\quad\text{(converse)}\\
-	\cmpl{r}&\in&\terms   \quad\quad\text{(complement)}\\
-%	\kleeneplus r&\in&\terms   \quad\quad\text{(Kleene closure)}\\
-%	\kleenestar r&\in&\terms   \quad\quad\text{(Kleene closure)}\\
 	\ident{A}&\in&\terms \quad\quad\text{(identity)}\\
@@
 	Throughout the remainder of this paper,	terms are represented by variables $r$, $s$, $d$, and $t$.
-	The \define{type} of a term $r$ is a pair of concepts, given by $\tf{r}$.
+	The \define{type} of a term $r$ is a pair of concepts given by $\tf{r}$.
 	$\mathfrak{T}$ is a partial function that maps terms to types.
-	If a term has one type, it is called \define{type correct}.
-	The Ampersand compiler requires all terms to be type correct.
-	If no type can be assigned, the compiler gives an error message.
-	If multiple types are assignable to a term, the ambiguity is signalled by the compiler.
-	In that case it will also emit an error message and prompt the programmer to disambiguate the term by adding type information.
-	In practice, the derivation of types can disambiguate most terms.
-	So, the programmer experiences some freedom to denote distinct relations by the same name without the obligation to specify their types at all occurrences.
-	The compiler does not generate any code if there is a term that is not type correct.
-	The type function and its restrictions are discussed in~\cite{Joosten2015}
-	and have no consequences for the remainder of this paper.
+	If term $r$ has a type, this term is called \define{type correct}.
+	The Ampersand compiler requires all terms to be type correct, or else it will not generate any code.
+	The type function and the restrictions it suffers are discussed in~\cite{Joosten2015}.
+	However, for the remainder of this paper this is irrelevant.
 
- 	The meaning of terms in Ampersand is given by an interpretation function $\mathfrak{I}$.
+    The meaning of terms in Ampersand is an extension of interpretation function $\mathfrak{I}$.
 	Let $A$ and $B$ be finite sets of atoms, then $\mathfrak{I}$ maps each term to the set of pairs for which that term stands.
@@
 	\ti{r}		 &=&\{\pair{a}{b}|\ a\ r\ b\}	\\
-	\ti{r \cup s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{or }\ \pair{a}{b}\in\ti{s}\}	\\
+	%\ti{r \cup s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{or }\ \pair{a}{b}\in\ti{s}\}	\\
 	\ti{r \cap s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{a}{b}\in\ti{s}\}	\\
@@
 	\ti{r;s}	 &=&\{\pair{a}{c}|\ \text{for some}\ b,\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{b}{c}\in\ti{s}\}	\\
-	\ti{r\backslash s}	 &=&\{\pair{b}{c}|\ \text{for all}\ a,\ \pair{a}{b}\in\ti{r}\ \text{implies}\ \pair{a}{c}\in\ti{s}\}	\\
-	\ti{r\slash s}	 &=&\{\pair{a}{b}|\ \text{for all}\ c,\ \pair{b}{c}\in\ti{s}\ \text{implies}\ \pair{a}{c}\in\ti{r}\}	\\
 	\ti{\flip{r}}	 &=&\{\pair{b}{a}|\ \pair{a}{b}\in\ti{r}    \}	\\
-	\ti{\cmpl{\declare{r}{A}{B}}}	 &=&\fullt{A}{B}-r	\\
-%	\ti{\kleeneplus{r}}	 &=&\text{the smallest set that satisfies }\kleeneplus{r}=r\ \cup\ \kleeneplus{r};\kleeneplus{r}	\\
-%	\ti{\kleenestar{\declare{r}{A}{A}}}	 &=&\ident{A}\cup\kleeneplus{r}\\
 	\ti{\ident{A}} 	 &=&\{\pair{a}{a}|\ a\in A\}	\\
@@
 \end{definition}
-%	The Kleene closure operators (postfix $\kleeneplus{\ }$ and $\kleenestar{\ }$) have been implemented partially in the current Ampersand implementation.
+	Ampersand has more operators than the ones introduced in definition~\ref{interpretation of terms}:
+	the complement (prefix unary $-$),
+	Kleene closure operators (postfix $\kleeneplus{\ }$ and $\kleenestar{\ }$),
+	left- and right residuals (infix $\backslash$ and $/$),
+	relational addition (infix $\dagger$),
+	and product (infix $\times$).
+	These are all expressible in the definitions above,
+	so we have limited this exposition to the operators introduced above.
+
+	The complement operator is defined by means of the binary difference operator (Equation~\ref{def:difference}).
+\begin{equation}
+	\tf{r}=\pair{A}{B}\ \Rightarrow\ \cmpl{r}\ =\ \fullt{A}{B}-r
+\label{eqn:complement}
+\end{equation}
+	This definition is elaborated in~\cite{vdWoude2011}.
 
@@
 	\tf{\text{RULE}\ r\subs s}&=&\pair{A}{B}\\
- resolved this \end{eqnarray}
+	\ti{\text{RULE}\ r = s}   &=&\ti{\fullt{A}{B}-((s-r)\cup(r-s))}\\
+	\ti{\text{RULE}\ r\subs s}&=&\ti{\fullt{A}{B}-(r-s)}
+\end{eqnarray}
 
-	We call a rule $\text{RULE}\ r = s$ \define{satisfied} when its interpretation equals $\ti{\fullt{A}{B}}$.
+	We call rule $r$ \define{satisfied} when $\ti{\text{RULE}\ r = s}\ =\ \ti{\fullt{A}{B}}$.
 	As the population of relations used in $r$ changes with time, the satisfaction of the rule changes accordingly.
@@
 \footnote{To \define{restore invariance} is sometimes used as a synonym to resolving violations.
-	To \define{preserve invariance} is sometimes used as a synonym to keeping a rule satisfied.}.
-	Consequently, such a rule may also be referred to as \define{invariant}.
-	Some ways of resolving violations are presented in section~\ref{sct:Code Fragments}.
+	Consequently, a rule is sometimes called \define{invariant}.}.
 	Each pair in the violation set of a rule is called a violation of that rule.
@@
 	Whatever choice she makes, she must ensure that her code yields data that satisfies the rules.
-	When we say: ``rule $r$ specifies this action'' we mean that satisfaction of rule $r$ is a postcondition of any action specified by rule $r$.
+	When we say: ``rule $r$ specifies this action'' we mean that satisfaction of rule $r$ is the goal of any action specified by rule $r$.
 
-\section{Software for Legal Reasoning}
+\section{Conceptual Analysis}
 \label{sct:Conceptual analysis}
@@
 \footnote{Thanks to Elleke Baecke for pointing us towards this source.}.
+	Toulmin is regarded as the first scholar in modern history to come up with a usable theory of argumentation.
 	Recent work typically draws on Toulmin, so his doctrine offers a good starting point.
@@
 	Software systems that support legal arguments have been around for many decades.
-	Legal reasoning differs from logical reasoning because of the human judgments that are involved.
 	Verheij~\cite{Verheij2003} distinguishes between argument assistance systems and automated reasoning systems.
-	An automated reasoning system (ARS) mimics human reasoning in an attempt to automate a decision.
-	An argument assisance system (AAS) helps its user to construct a decision.
+	Automated reasoning has never gained wide acceptance for making legal decisions,
+	because lawyers and judges alike feel that human judgment should be at the core of any legal decision.
+	Attempts to apply mathematical logic to legal judgments have had limited impact for similar reasons~\cite{Prakken2005}.
+	Legal reasoning differs from logic reasoning because of the human judgments that are involved.
 	The literature on legal reasoning~\cite{Lind2007} makes it quite clear why mathematical logic alone does not suffice.
-	Legal reasoning puts a human decision at the root of its notion of truth,
-	whereas mathematical logic eradicates any human influence.
-	For this reason, attempts to apply mathematical logic to legal judgments have had limited impact~\cite{Prakken2005}.
-	Argument assistance systems~\cite{Verheij2005} respect the professional freedom of each legal professional
-	in constructing her own line of argumentation.
-	AAS offer help in many different ways.
-	For instance by looking up legal references, jurisdiction from the past, scholarly works etc.
+	Argument assistance systems~\cite{Verheij2005} have been more successful,
+	because they respect the professional freedom of legal professionals to construct their own line of argumentation.
+	Such systems offer help in many different ways.
+	They can help by looking up legal references, jurisdiction from the past, scholarly works etc.
 	They can also help to construct and validate arguments by keeping arguments and evidence organized.
 	They can store, disclose and share legal evidence.
-	The MirrorMe system lets each user (whether a plaintiff, judge, or defendant) decide individually on the truth of every statement.
-	Differences in truth assessments can then be made visible and manageable.
 
-	Software development is founded on mathematical reasoning, whereas legal work is built on defeasible reasoning.
-	So we have built an AAS in the mathematical formalism of relation algebra
-	by \emph{implementing} Toulmin's principles%
-\footnote{We deliberately did not want to squeeze legal reasoning into relation algebra, or any other objective formalism.}.
+	In our case study we use logic to reason about the correctness of a program.
+	The argumentation principles of Toulmin are \emph{implemented in} logic rather than \emph{replaced by} logic.
 	The structure of MirrorMe was designed by conceptually analysing ideas of Toulmin,
@@
 	Even though our case study yields a model that is a bit too large for this paper,
-	we have printed a relevant part of it at the end of this paper in figure~\ref{fig:conceptual model}.
-	This diagram gives an impression of what it looks like.
-	It shows entities such as Statement, Argument, Binding, Template, Case, and Claim, together with their attributes.
-	Attributes are the univalent and injective relations, which Ampersand derives from the script.
-	Relations that are neither univalent nor injective are n-m relations between entities.
-	In practice, the conceptual models have shown to be readable without much explanation for experienced database developers.
+	figure~\ref{fig:conceptual model} gives a good impression of what it looks like.
+\begin{figure}[htb]
+\begin{center}
+  \includegraphics[scale=.23]{LogicalDataModel.pdf}
+\end{center}
+\caption{Conceptual data model}
+\label{fig:conceptual model}
+\end{figure}
+
+	To convey the flavour of software development in relation algebra, it suffices to discuss a tiny part of the whole program.
+	Therefore, we shall discuss the part of the conceptual model that is used in the sequel.
+	The context in which a user creates arguments and reasons about them is a legal case.
+	For this reason, validity, falsehood and truth of statements are related to the case.
+	A \define{statement} is a phrase in natural language that can be either true or false.
+	An example is \stmtText{The employee, John Brown, is entitled to 50 Euros}%
+\footnote{This type of statement is typical for cases. It is valid only in the case where this particular John Brown is known.
+	In other cases, where John Brown is unknown, this statement is meaningless}.
+	In MirrorMe, a user can define a template, such as \stmtText{The employee, [emp], is entitled to [increase]}.
+	The strings \stmtText{[emp]} and \stmtText{[increase]} are called placeholders.
+	The reason for using placeholders is that legal rules are stated in general terms, e.g.
+	\stmtText{Every employee is entitled to an increase in salary}.
+	The user of MirrorMe will pick a legal text (from a source he trusts) and substitute parts of that text by placeholders.
+	When the facts are known, the argument can be completed by substituting placeholders by actual phrases.
+
+	It is precisely this substitution process that we have chosen to describe in this paper as a case study in
+	programming with relation algebra.
 
@@
 \label{sct:Programming in RA}
-	This section illustrates the execution model of Ampersand by means of an example.
-	A demonstration of this example was given at RAMiCS-2017.
-	The example covers a tiny part of the whole MirrorMe program,
-	which is enough to give a taste of software development in relation algebra.
+	At this point we have reached the core of this case study.
+	We focus on the substitution of placeholders when their values change.
+	By focusing on this tiny detail, we can discuss the mechanics ``under the hood'' of the application generated by Ampersand.
 
-	The use case for this example is the substitution of placeholders in text.
-	For instance, the user might want to replace \stmtText{The employee, [emp], is entitled to [raise]} by
-		\stmtText{The employee, James, is entitled to 50 Euros.}.
-	The first text is called a \define{template}, because it contains placeholders in square brackets.
-	The strings \stmtText{[emp]} and \stmtText{[raise]} are the placeholders.
-	The second text is called a \define{statement},
-	because it contains no placeholders and can be read as a statement in natural language.
-	So, a statement can be either true or false.
-
-	A user makes such replacements when formulating arguments and reasoning with them.
-	MirrorMe helps her by making replacements consistently throughout each case.
-	In MirrorMe, a user can define templates.
-	The reason for using placeholders is that legal rules are stated in general terms, e.g.
-	\stmtText{Every employee is entitled to a raise in salary}.
-	A user will pick a legal text (from a source he trusts) and replace parts of that text by placeholders.
-
-	The substitution of a placeholder by a phrase is called a \define{binding}.
-	A binding assigns a piece of text to a placeholder.
-	That text is the \define{phrase of} the binding and
-	also the \define{value} of the placeholder.
-	When all placeholders in a template have been substituted, a statement remains.
-
-	The process of substitution is embodied by a sequence of events
-	that are triggered by rule violations.
-	Events may resolve violations in one rule, but cause violations in others.
-	In this manner, rule violations and events to resolve them alternate.
-	That is shown in figure~\ref{fig:log file}.
+	First we show how the computer solves the issue by looking at an exerpt of a log file (figure~\ref{fig:log file}).
+	It shows an alternating sequence of the computer (ExecEngine) mentioning a rule,
+	followed by insert or delete actions to satisfy that rule.
+	Then we zoom in further, one rule at a time, to explain precisely what each rule looks like and how programming is done.
+	We then discuss the same flow of events by means of a graph (figure~\ref{fig:event flow}) in which these rules and actions are nodes.
+	This graph serves as an event flow diagram to illustrate the process behind figure~\ref{fig:log file}.
 
@@
  InsPair(stmtShowText,Statement,Stat623,StmtText,
-         The employee, [emp], is entitled to [raise].)
+         The employee, [emp], is entitled to [increase].)
  ExecEngine satisfying rule 'done initializing'
@@
  InsPair(stmtShowText,Statement,Stat623,StmtText,
-         The employee, James, is entitled to [raise].)
+         The employee, James, is entitled to [increase].)
  InsPair(substituted,Binding,Bind625,Statement,Stat623)
@@
 	The log file of figure~\ref{fig:log file} has been taken from a computer that carries out the procedure to satisfy all rules.
-	It illustrates how the machine is busy satisfying one rule after another by inserting or deleting elements in different relations.
-	This example follows the event flow in a dependency graph, figure~\ref{fig:event flow}, which is discussed in section~\ref{sct:Programming}.
 	The machine will only act on rules that are violated.
-	The first rule violated is ``signal phrase update'', in which the computer signals a change in relation \id{phrase}.
-	That change is caused by a user who has added or deleted pairs in this relation,
-	The computer will react by changing other relations such that the rule is satisfied again.
-	This may then violate other rules, which have to be restored in turn.
+	The first rule to be violated is ``signal phrase update'', in which the computer signals that something or someone has made a change
+	in relation \id{phrase}.
 	The procedure ends by doing the necessary substitutions,
@@
 
-	Before looking at the actual rules,
-	let us introduce an auxiliary relation, \id{differB}, to represent bindings that must be updated:
-\[\id{differB} = \ident{Binding}\cap\id{shownPhrase};\cmpl{\ident{Phrase}};\flip{\id{phrase}}\]
-	The relation \id{phrase} links bindings to texts.
-	It contains the actual phrase of each binding.
-	This is also known as the value of a placeholder%
-\footnote{A phrase and a placeholder are linked by a pair from $\flip{\id{phrase}};\id{placeholder}$.}.
-	The relation \id{shownPhrase} links bindings to text fields in the user interface.
-	So \id{differB} contains the bindings of which the phrase differs from what is shown on the screen.
-	Whether a placeholder has been edited can be observed by comparing its shown phrase to the phrase.
-	Bindings in \id{differB} are to be updated,
-	because the value of placeholders should always correspond with the text on the screen.
-
-	Let us now study the rules to see how these rules cause the right actions to take place in the correct order.
+	Let us now study the actual rules to see how these rules cause the right actions to take place in the correct order.
 	We will follow the log file from figure~\ref{fig:log file} in reverse direction,
@@
 
+	Whether a placeholder has been edited can be observed by comparing its new phrase to the shown phrase%
+\footnote{Note that we use the notion ``the phrase of a placeholder'' to indicate a pair from $\flip{\id{phrase}};\id{placeholder}$.}.
+	The phrase of a placeholder is kept in the relation \id{phrase}.
+	The shown phrase is kept in the relation \id{shownPhrase}.
+	The sole purpose for having the relation \id{shownPhrase} is to detect a change in \id{phrase}.
+	Let us introduce \id{differB} to represent the bindings with an updated phrase:
+\[\id{differB} = \ident{Binding}\cap\id{shownPhrase};\cmpl{\ident{Phrase}};\flip{\id{phrase}}\]
+	When the phrase of a placeholder changes,
+	that phrase must be updated in every statement in which the placeholder was used.
+	That update action is specified in section~\ref{rule: substitute}.
+	Section~\ref{rule: fill shownPhrase} specifies how \id{shownPhrase} is made equal to \id{phrase},
+	after all necessary substitutions are done.
+
 \subsection{rule: fill shownPhrase}
 \label{rule: fill shownPhrase}
-	The sole purpose of having the relation \id{shownPhrase} is to detect a change
-	and propagate it as the new value of a placeholder.
-	Then, the new value of the placeholder must be propagated in every statement in which the placeholder was used.
-	That update action is specified in section~\ref{rule: substitute}.
-	The following rule specifies how \id{shownPhrase} is made equal to \id{phrase},
-	after all necessary substitutions are done.
 
 \[\text{RULE}\ (\ident{Binding}\cap\id{substituted}/\id{inStatement} );\id{phrase}\ \subs\ \id{shownPhrase} \]
-        This rule says that far a binding with all substitutions made, the \id{phrase} must be equal to the \id{shownPhrase}.
-	When violated, the software inserts all violations into \id{shownPhrase} to satisfy the rule.
+        This rule says that for each binding that has been substituted in every statement it is used in,
+	the \id{phrase} must be equal to the \id{shownPhrase}.
+	When violated, it is satisfied by inserting all violations into \id{shownPhrase}.
 
@@
 	Let us now look into the process of substituting placeholders by phrases.
-	The relation \id{tmplParsedText} contains the template text,
-	e.g. \stmtText{The employee, [emp], is entitled to [raise]}.
+	The relation \id{tmplParsedText} contains the original text, provided by a user.
+	Placeholders are specified by enclosing them in brackets,
+	e.g. \stmtText{The employee, [emp], is entitled to [increase]}.
 	The text in which a placeholder has been substituted by a phrase,
@@
 	After a placeholder has been substituted, it no longer occurs in \id{stmtShowText}.
-	The absence of placeholders in \id{stmtShowText} poses a problem if we want to substitute a new phrase in \id{stmtShowText}.
-	That substitution must be carried out in the original template,
-	which contains all of the original placeholders.
+	This poses a problem if we want to substitute the new phrase in \id{stmtShowText}.
+    For the placeholder that defines the place in the text where to substitute, is no longer in that text.
+	Therefore, substitutions must be done in the original text of the statement.
+	All placeholders in that text must then be substituted again.
 	So, the text in \id{stmtShowText} must first be reset to the original text from \id{tmplParsedText}.
@@
 	the text in \id{stmtShowText} must be made equal to the original text in the template.
-	That text is obtained from the relation \id{descriptor}.
 	The action is specified by the following rule:
@@
 	Violations of this rule are descriptors of templates that belong to statements that are being reset.
-	These violations are resolved by inserting them in \id{stmtShowText}.
+	These violations can be resolved by inserting them in \id{stmtShowText}.
 
@@
 	She considers how violations arise from changing content of relations.
-	She also thinks about insert and delete actions to resolve these violations.
+	And she thinks about insert and delete actions to restore these violations.
 	In our case study, we have reasoned with the event types: Ins $\la\id{relation}\ra$ and Del $\la\id{relation}\ra$%
@@
 \begin{table}[htb]
-{\small\begin{tabular}{|l|p{3.2in}|}\hline
+{\small\begin{tabular}{|l|p{3.3in}|}\hline
 rule&event types\\ \hline
@@
 	That was signaled by the rule ``signal phrase update''.
-	In general, the software developer must decide how to resolve violations that occur as a result of some event.
-	This can be done by choosing among the duals of the event types from table~\ref{fig:violation of rules}.
-	By systematically swapping Ins for Del and vice-versa,
-	the table shows by which type of events violations can be resolved.
-	To ensure progress, the software developer will pick a different relation.
-	For resolving the violations in the same relation would resolve the violations by undoing the change.
-    Typically, preserving invariance of a rule is done by changing one or more other relations than the one that caused the violations.
-	The software developer can draw a graph that contains all information from table~\ref{fig:violation of rules}
-	and the dual information.
 \begin{figure}[bht]
@@
 \end{figure}
+	In general, the software developer must decide how to resolve violations that occur as a result of some event.
+	This can be done by choosing among the duals of the event types from table~\ref{fig:violation of rules}.
+	By systematically swapping Ins for Del and vice-versa,
+	the table shows by which type of events violations can be restored.
+	To ensure progress, the software developer will pick a different relation for restoring than the relation that causes the violation.
+	The software developer can draw a graph that contains all information from table~\ref{fig:violation of rules}
+	and the dual information.
 	Figure~\ref{fig:event flow} shows part of that graph.
-	A circle represents a rule, a rectangle represents an event type, and the edges correspond to the flow of values between them.
-	The propagation of changes across the graph is initiated by the event sources (e.g., a user changing the set of pairs in a relation).
-	The software developer can work her way through the graph, to write code to resolve violations for every event type that might occur.
+	A circle represents a rule, a rectangle represents an event type, and the arrows connect them.
+	The software developer can work her way through the graph, to write code to restore invariants for every event type that might cause violations.
 	Figure~\ref{fig:event flow} shows just that part of the graph that corresponds with the case study in this article.
-	In actual programs the complete graph can be much larger.
-
-\section{Code Fragments}
-\label{sct:Code Fragments}
-	After the previous section, one question remains open: which code can preserve invariance and can we produce that code for arbitrary rules?
-	We shall provide an answer to this question not just for the case study at hand, but for all conceivable cases.
-	This section derives such code for every term in relation algebra.
-	The problem of generating such code suffers from arbitrariness: there are many ways to generate correct code.
-	We have not explored all possibilities, because there are too many.
-	We do give one implementation for every conceivable term, pointing out where choices are to be made.
-
-	The purpose of the code is to preserve invariance of rules by insertion and deletion into and from relations.
-	Inserting (a set of) pairs $\Delta$ into a relation $r$ yields $r\cup\Delta$.
-	Likewise, deleting pairs $\Delta$ from a relation $r$ yields $r-\Delta$.
-	In this section we derive code for computers,
-	which is built around two basic operations: insert and delete.
-	The insert operation can be represented in terms of pre- and postconditions\footnote{This notation
-	was conceived by Floyd and Hoare~\cite{Floyd1967,Hoare1969}.
-	It is well known for making assertions about computer programs and reasoning about them.}.
-\[\begin{array}{r@{~}l}
-&\{\text{pre: }\code{P}=p\}\\
-\code{INSERT}&\Delta\code{ IN P}\\
-&\{\text{post: }\code{P}=p\cup\Delta\}\\
-\end{array}\]
-
-	The delete operation is introduced similarly:
-\[\begin{array}{r@{~}l}
-&\{\text{pre: }\code{P}=p\}\\
-\code{DELETE}&\Delta\code{ FROM P}\\
-&\{\text{post: }\code{P}=p-\Delta\}\\
-\end{array}\]
-
-	Let the following example introduce the idea of breaking and restoring invariance.
-	Suppose the rule $\code{P}\subs\code{S}\cup\code{T}$ is to be kept satisfied by a computer,
-	and \code{P}, \code{S}, and \code{T} are variables (memory locations) in that computer,
-	which represent relations.
-	We assume that before executing any code, we have $\code{P}=p$, $\code{S}=s$, $\code{T}=t$ and $\code{P}\subs\code{S}\cup\code{T}$.
-	The following example shows that inserting a set of pairs, $\Delta$, in \code{P}
-	may violate the rule.
-	Inserting the same $\Delta$ in \code{S} restores invariance.
-	In terms of code, we write:
-\[\begin{array}{r@{~}l}
-&\{\text{Assume $\code{P}=p$, $\code{S}=s$, $\code{T}=t$ and $\code{P}\subs\code{S}\cup\code{T}$}\}\\
-\code{INSERT}&\Delta\code{ IN P}\\
-&\{\text{$\code{P}=p\cup\Delta$, $\code{S}=s$, and $\code{T}=t$, and invariance is broken: $\code{P}\not\subs\code{S}\cup\code{T}$}\}\\
-\code{INSERT}&\Delta\code{ IN S}\\
-&\{\text{postcondition: $\code{P}=p\cup\Delta$, $\code{S}=s\cup\Delta$, $\code{T}=t$, and $\code{P}\subs\code{S}\cup\code{T}$}\}\\
-\end{array}\]
-
-	To prove that rule $\code{P}\subs\code{S}\cup\code{T}$ is kept satisfied,
-	we must show that its value afterwards is equal to $\full$,
-	assuming that its initial value is $\full$.
-\[\begin{array}{cl}
-&(p\cup\Delta)\subs(s\cup\Delta)\cup t\\
-=&\\
-&\cmpl{p\cup\Delta}\ \cup\ s\ \cup\ \Delta\ \cup\ t\\
-=&\hspace{1in}\{\hbox{De Morgan's law}\}\\
-&(\cmpl{p}\cap\cmpl{\Delta})\ \cup\ s\ \cup\ \Delta\ \cup\ t\\
-=&\hspace{1in}\{\hbox{Distribute $\cup$ over $\cap$}\}\\
-&(\cmpl{p}\cup s\cup\Delta\cup t)\ \cap\ (\cmpl{\Delta}\cup s\cup\Delta\cup t)\\
-=&\hspace{1in}\{\cmpl{\Delta}\cup\Delta=\full\}\\
-&\cmpl{p}\cup s\cup\Delta\cup t\\
-=&\hspace{1in}\{\full\ =\ \cmpl{p}\cup s\cup t\}\\
-&\full
-\end{array}\]
-	From this derivation we may conclude that invariance of $\code{P}\subs\code{S}\cup\code{T}$ is preserved.
-	From this example, we can see that $\code{INSERT }\Delta\code{ IN T}$ would have worked just as well.
-	In general, there are alternative ways of preserving invariance.
-	This calls for a systematic analysis.
-	The following sections analyze the various terms in Relation Algebra and provides some code to restore invariance of rules containing such terms.
-	We start with an analysis of rules, then we proceed with boolean operators, and finally we arrive at relational operators.
-
-\subsection{Implication Rules}
-\label{ssct:Implication Rules}
-	We will start by studying insertions and deletions in the simplest of rules:
-\begin{equation}
-\code{RULE }r\subs s\label{rule:r subs s}
-\end{equation}
-	Recall that $\subs$ is monotonous, so:
-\begin{eqnarray}
-r\subs s&\Rightarrow&r\cup\Delta\subs s\cup \Delta\label{eqn:r subs s}
-\end{eqnarray}
-	It immediately follows that inserting $\Delta$ into \code{S} is an adequate reaction to
-	an insertion of $\Delta$ into \code{R}:
-\[\begin{array}{r@{~}l}
-&\{\text{Assume $\code{R}=r$, $\code{S}=s$, and $\code{R}\subs\code{S}$}\}\\
-\code{INSERT}&\Delta\code{ INTO R};\\
-&\{\text{$\code{R}=r\cup\Delta$, $\code{S}=s$, and $\code{R}\subs\code{S}$ is not necessarily true.}\}\\
-\code{INSERT}&\Delta\code{ INTO S}\\
-&\{\text{postcondition: $\code{R}=r\cup\Delta$, $\code{S}=s\cup\Delta$, and $\code{R}\subs\code{S}$}\}
-\end{array}\]
-
-	A similar line of reasoning can be used for deletion as well:
-\[\begin{array}{r@{~}l}
-&\{\text{Assume $\code{R}=r$, $\code{S}=s$, and $\code{R}\subs\code{S}$}\}\\
-\code{DELETE}&\Delta\code{ FROM S};\\
-&\{\text{$\code{R}=r$, $\code{S}=s-\Delta$, and $\code{R}\subs\code{S}$ is not necessarily true.}\}\\
-\code{DELETE}&\Delta\code{ FROM R}\\
-&\{\text{postcondition: $\code{R}=r-\Delta$, $\code{S}=s-\Delta$, and $\code{R}\subs\code{S}$}\}
-\end{array}\]
-	The proof is noteworthy similar and equally trivial:
-\begin{eqnarray}
-r\subs s&\Rightarrow&r-\Delta\subs s-\Delta\label{eqn:r subs s by deletion}
-\end{eqnarray}
-	If the event is a deletion of $\Delta$ from $r$ or an insertion into $s$, no reaction is required.
-	Such events cannot violate the invariance of rule~\ref{rule:r subs s},
-	because equations~\ref{eqn:r subs s delfrom r} and~\ref{eqn:r subs s insinto s} hold for any $\Delta$:
-\begin{eqnarray}
-r\subs s&\Rightarrow&r-\Delta\subs s\label{eqn:r subs s delfrom r}\\
-r\subs s&\Rightarrow&r\subs s\cup\Delta\label{eqn:r subs s insinto s}
-\end{eqnarray}
-
-Summarizing, some code for preserving invariance of rules of form $\code{R}\subs\code{S}$ is:
-\[\begin{array}{|l|l|l|l|}
-\hline
-\text{invariant}&\text{action}&\text{reaction}&\text{equation}\\ \hline
-\full=\code{R}\subs\code{S}&\code{INSERT}\ \Delta\code{ INTO R}&\code{INSERT }\Delta\code{ INTO S}&\ref{eqn:r subs s}\\
-\full=\code{R}\subs\code{S}&\code{DELETE}\ \Delta\code{ FROM S}&\code{DELETE }\Delta\code{ FROM R}&\ref{eqn:r subs s by deletion}\\
-\full=\code{R}\subs\code{S}&\code{DELETE}\ \Delta\code{ FROM R}&\code{Nop}&\ref{eqn:r subs s delfrom r}\\
-\full=\code{R}\subs\code{S}&\code{INSERT}\ \Delta\code{ INTO S}&\code{Nop}&\ref{eqn:r subs s insinto s}\\ \hline
-\end{array}\]
-Preserving an invariant of the form $\code{R}=\code{S}$ is equivalent to preserving both $\code{R}\subs\code{S}$ and $\code{S}\subs\code{R}$ simultaneously.
-
-\subsection{Rules with Boolean Operators}
-	To write code for boolean expressions,
-	we revert to the following lemma's:
-\begin{eqnarray}
-r=\cmpl{s}&\Rightarrow&r\cup\Delta=\cmpl{s-\Delta}\label{lbl:Preserve r = -s with Ins Delta into r}\\
-r=\cmpl{s}&\Rightarrow&r-\Delta=\cmpl{s\cup\Delta}\label{lbl:Preserve r = -s with Del Delta from r}\\
-t=r\cap s&\Rightarrow&t\cup\Delta=(r\cup\Delta)\cap(s\cup\Delta)\label{lbl:Preserve t=r inter s with Ins Delta into t}\\
-t=r\cup s&\Rightarrow&t-\Delta=(r-\Delta)\cup(s-\Delta)\label{lbl:Preserve t=r union s with Del Delta from t}\\
-t=r\cap s&\Rightarrow&t-\Delta=(r-\Delta)\cap s\label{lbl:Preserve t=r inter s with Del Delta from t}\\
-t=r\cup s&\Rightarrow&t\cup\Delta=(r\cup\Delta)\cup s\label{lbl:Preserve t=r union s with Ins Delta into t}
-\end{eqnarray}
-	In proof assistant Isabelle, these lemmas are proven in zero steps, so we omit these proofs here.
-
-	Equations~\ref{lbl:Preserve r = -s with Ins Delta into r} and~\ref{lbl:Preserve r = -s with Del Delta from r}
-	show that insertion in a term means deletion from its complement and vice versa.
-	So the programmer can write the following code:
-\[\begin{array}{r@{~}l}
-&\{\text{Assume $\code{R}=r$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$}\}\\
-\code{INSERT}&\Delta\code{ INTO R};\\
-&\{\text{$\code{R}=r\cup\Delta$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$ is not necessarily true.}\}\\
-\code{DELETE}&\Delta\code{ FROM S}\\
-&\{\text{$\code{R}=r\cup\Delta$ and $\code{S}=s-\Delta$, so by equation~\ref{lbl:Preserve r = -s with Ins Delta into r} we have $\code{R}=\cmpl{\code{S}}$}\}
-\end{array}\]
-	The proofs for the following table are similar and equally simple,
-	so we omit them for the sake of brevity.
-% 	Likewise, the reaction on a deletion will be an insertion:
-% \[\begin{array}{r@{~}l}
-% &\{\text{Assume $\code{R}=r$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$}\}\\
-% \code{DELETE}&\Delta\code{ FROM R};\\
-% &\{\text{$\code{R}=r-\Delta$, $\code{S}=s$, and $\code{R}=\cmpl{\code{S}}$ is not necessarily true.}\}\\
-% \code{INSERT}&\Delta\code{ INTO S}\\
-% &\{\code{R}=r-\Delta\text{ and }\code{S}=s\cup\Delta\text{, so by equation~\ref{lbl:Preserve r = -s with Del Delta from r} we have }\code{R}=\cmpl{\code{S}}\}
-% \end{array}\]
-% 
-% 	Now consider equations~\ref{lbl:Preserve t=r inter s with Ins Delta into t} and~\ref{lbl:Preserve t=r union s with Del Delta from t}.
-% 	If \code{T} changes, both \code{R} and \code{S} must be changed accordingly.
-% 	Since $\cup$ and $\cap$ are associative operators, the arguments can be evaluated with no interaction between them.
-% 	This allows us to define a primitive, \code{ALL}, without specifying in which order it processes its arguments\footnote{... leaving room to take advantage of parallelism.}.
-% 	We define \code{ALL} to perform all actions from a given list of actions \code{AS}:
-% \[\begin{array}{r@{~}l}
-% &\{\text{Assume }\forall\code{A}\in\code{AS}: (\{\id{pre}_{\code{\tiny A}}\}\ \code{A}\ \{\id{post}_{\code{\tiny A}}\})\text{ and } \id{pre}_{\code{\tiny A}} \}\\
-% \code{ALL}&\ \code{AS}\\
-% &\{\forall\code{A}: \id{post}_{\code{\tiny A}}\}
-% \end{array}\]
-% 	With this primitive, we can write code for inserting pairs in \code{T},
-% 	using equation~\ref{lbl:Preserve t=r inter s with Ins Delta into t}:
-% \[\begin{array}{r@{~}l}
-% &\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cap\code{S}$}\}\\
-% \code{INSERT}&\Delta\code{ INTO T};\\
-% &\{\text{$\code{T}=t\cup\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cap\code{S}$}\}\\
-% \code{ALL}&[\code{INSERT}\ \Delta\code{ INTO R},\code{INSERT}\ \Delta\code{ INTO S}]\\
-% &\{\code{R}=r\cup\Delta\text{ and }\code{S}=s\cup\Delta\text{, so by equation~\ref{lbl:Preserve t=r inter s with Ins Delta into t} we have }\code{T}=\code{R}\cap\code{S}\}
-% \end{array}\]
-% 	Similarly, we can derive code for equation~\ref{lbl:Preserve t=r union s with Del Delta from t}:
-% \[\begin{array}{r@{~}l}
-% &\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cup\code{S}$}\}\\
-% \code{DELETE}&\Delta\code{ FROM T};\\
-% &\{\text{$\code{T}=t-\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cup\code{S}$}\}\\
-% \code{ALL}&[\code{DELETE}\ \Delta\code{ FROM R},\code{DELETE}\ \Delta\code{ FROM S}]\\
-% &\{\code{R}=r-\Delta\text{ and }\code{S}=s-\Delta\text{, so by equation~\ref{lbl:Preserve t=r union s with Del Delta from t} we have }\code{T}=\code{R}\cup\code{S}\}
-% \end{array}\]
-% 
-% 	In equations~\ref{lbl:Preserve t=r inter s with Del Delta from t} and~\ref{lbl:Preserve t=r union s with Ins Delta into t},
-% 	a choice must be made between $r$ and $s$. For that purpose we need a primitive \code{CHC} that chooses one action from a list $\code{AS}$, say $\code{A}$:
-% \[\begin{array}{r@{~}l}
-% &\{ (\{\id{pre}_{\code{\tiny A}}\}\ \code{A}\ \{\id{post}_{\code{\tiny A}}\})\text{ and } \id{pre}_{\code{\tiny A}} \}\\
-% \code{CHC}&\ \code{AS}\\
-% &\{\id{post}_{\code{\tiny A}}\}
-% \end{array}\]
-% 	Note that this definition leaves open which term is selected.
-% 	The following code shows how this works out for equation~\ref{lbl:Preserve t=r inter s with Del Delta from t}:
-% \[\begin{array}{r@{~}l}
-% &\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cap\code{S}$}\}\\
-% \code{DELETE}&\Delta\code{ FROM T};\\
-% &\{\text{$\code{T}=t-\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cap\code{S}$}\}\\
-% \code{CHC}&[\code{DELETE}\ \Delta\code{ FROM R},\code{DELETE}\ \Delta\code{ FROM S}]\\
-% &\{\text{Either }\code{R}=r-\Delta\text{ or }\code{S}=s-\Delta\text{, so by equation~\ref{lbl:Preserve t=r inter s with Del Delta from t} we have }\code{T}=\code{R}\cap\code{S}\}
-% \end{array}\]
-% 
-% 	The following code shows how this works out for equation~\ref{lbl:Preserve t=r union s with Ins Delta into t}:
-% \[\begin{array}{r@{~}l}
-% &\{\text{Assume $\code{T}=t$, $\code{R}=r$, $\code{S}=s$, and $\code{T}=\code{R}\cup\code{S}$}\}\\
-% \code{INSERT}&\Delta\code{ INTO T};\\
-% &\{\text{$\code{T}=t\cup\Delta$, $\code{R}=r$, $\code{S}=s$ are true, but not necessarily $\code{T}=\code{R}\cup\code{S}$}\}\\
-% \code{CHC}&[\code{INSERT}\ \Delta\code{ INTO R},\code{INSERT}\ \Delta\code{ INTO S}]\\
-% &\{\text{Either }\code{R}=r\cup\Delta\text{ or }\code{S}=s\cup\Delta\text{, so by equation~\ref{lbl:Preserve t=r union s with Ins Delta into t} we have }\code{T}=\code{R}\cup\code{S}\}
-% \end{array}\]
-% 
-% 	Section~\ref{ssct:Implication Rules} treated rules of the form $\code{R}\subs\code{S}$.
-% 	Note that this rule can be rewritten to $\cmpl{\code{R}}\cup\code{S}=\full$.
-% 	Consequently, rules did not have to be treated separately.
-% 	We could have treated them as a special case of equations with boolean operators only.
-% 	In fact, every rule with boolean operators $\cup$, $\cap$, and $-$ can be rewritten as a number of rules in disjunctive normal form\footnote{%
-% 	Normalizing a boolean algebra is existing knowledge, and is considered outside the scope of this paper.} (DNF~\cite{davey1990}).
-% 	We can represent a rule in DNF informally by:
-% \begin{eqnarray}
-% \code{RULE }\cmpl{r_0}\cup\ldots\cup\cmpl{r_n}\cup s_0\cup\ldots\cup s_m
-% \end{eqnarray}
-% 	or more formally by
-% \begin{eqnarray}
-% \full=\cmpl{\bigcap\id{antecedents}}\cup\bigcup\id{consequents}
-% \end{eqnarray}
-% 	in which \id{antecedents} are the complemented (negative) terms and \id{consequents} are the non-complemented (positive) terms.
-% 
-	To preserve invariants in a boolean algebra is done as follows:
-\[\begin{array}{|l|l|l|l|}
-\hline
-\text{invariant}&\text{action}&\text{reaction}&\text{equation}\\ \hline
-\code{R}=\cmpl{\code{S}}&\code{INSERT}\ \Delta\code{ INTO R}&\code{DELETE }\Delta\code{ FROM S}&\ref{lbl:Preserve r = -s with Ins Delta into r}\\
-\code{R}=\cmpl{\code{S}}&\code{DELETE}\ \Delta\code{ FROM R}&\code{INSERT }\Delta\code{ INTO S}&\ref{lbl:Preserve r = -s with Del Delta from r}\\
-\code{T}=\bigcap\code{TERMS}&\code{INSERT}\ \Delta\code{ INTO T}&\code{ALL }[\code{INSERT}\ \Delta\code{ INTO X}|\code{X}\in\code{TERMS}]&\ref{lbl:Preserve t=r inter s with Ins Delta into t}\\
-\code{T}=\bigcup\code{TERMS}&\code{DELETE}\ \Delta\code{ FROM T}&\code{ALL }[\code{DELETE}\ \Delta\code{ FROM X}|\code{X}\in\code{TERMS}]&\ref{lbl:Preserve t=r union s with Del Delta from t}\\
-\code{T}=\bigcap\code{TERMS}&\code{DELETE}\ \Delta\code{ FROM T}&\code{CHC }[\code{DELETE}\ \Delta\code{ FROM X}|\code{X}\in\code{TERMS}]&\ref{lbl:Preserve t=r inter s with Del Delta from t}\\
-\code{T}=\bigcup\code{TERMS}&\code{INSERT}\ \Delta\code{ INTO T}&\code{CHC }[\code{INSERT}\ \Delta\code{ INTO X}|\code{X}\in\code{TERMS}]&\ref{lbl:Preserve t=r union s with Ins Delta into t}\\ \hline
-\end{array}\]
-	So let us now turn to code that is needed when relational operators are involved.
-
-\subsection{Rules with relational operators}
-	This section studies how to insert pairs into a term $r;s$.
-	To formulate this precisely, we introduce an auxiliary relation $t$ to represent the term $r;s$.
-	First we will insert pairs in $t$ while preserving invariance of $t=r;s$
-	Then we will insert pairs in $r$ preserving the same rule.
-
-	To insert pairs $\Delta$ into $t$ involves finding relations $\Delta_r$ and $\Delta_s$
-	that satisfy equation~\ref{eqn:t subs r;s}.
-\begin{equation}
-t=r;s\ \Rightarrow\ t\cup\Delta=(r\cup \Delta_r);(s\cup \Delta_s)\label{eqn:t subs r;s}
-\end{equation}
-	Let us define a software primitive to compute both $\Delta_r$ and $\Delta_s$ and call it \code{NEWPAIRS}.
-	It computes $\Delta_r$ and $\Delta_s$ from $r$, $s$, and $\Delta$.
-	We can derive its postconditions from equation~\ref{eqn:t subs r;s} by means of the following derivation:
-\[\begin{array}{r@{~}l}
-&t\cup\Delta\\
-=&\hspace{1in}\{\text{since }t\cup\Delta=t\cup(\Delta-t)\}\\
-&t\cup(\Delta-t)\\
-=&\hspace{1in}\{t=r;s\text{ and let \code{NEWPAIRS} ensure that }\Delta_r;\Delta_s=\Delta-t\}\\
-&r;s\cup\Delta_r;\Delta_s\\
-=&\hspace{1in}\{\text{let \code{NEWPAIRS} ensure: }r;\Delta_s\subs t\cup\Delta\text{ and }\Delta_r;s\subs t\cup\Delta\}\\
-&r;s\cup r;\Delta_s\cup \Delta_r;s\cup \Delta_r;\Delta_s\\
-=&\hspace{1in}\{\text{distributivity}\}\\
-&(r\cup \Delta_r);(s\cup \Delta_s)
-\end{array}\]
-	The conditions in this derivation must be satisfied by \code{NEWPAIRS}. 
-	This yields the following specification of \code{NEWPAIRS}:
-\[\begin{array}{r@{~}l}
-&\{\code{R}=r\text{ and }\code{S}=s\text{ and }\code{DELTA}=\Delta\}\\
-\code{NEWPAIRS}&(\code{R}, \code{S}, \code{DELTA}): (\code{DELTAR}, \code{DELTAS})\\
-&\begin{array}[t]{@{}l@{}l}
-        \{&\code{DELTR};\code{DELTAS}=\Delta - r;s\text{ and }\\
-          &\code{R};\code{DELTAS}\subs r;s\cup\Delta\text{ and }\code{DELTAR};\code{S}\subs r;s\cup\Delta\}
-\end{array}
-\end{array}\]
-	The pre- and postconditions of \code{NEWPAIRS} are sufficient to satisfy equation~\ref{eqn:t subs r;s}.
-	Since $t$ does not occur in the pre- and postconditions, \code{NEWPAIRS} implements equation~\ref{eqn:ins delta r;s}:
-\begin{equation}
-r;s\cup\Delta=(r\cup \Delta_r);(s\cup \Delta_s)\label{eqn:ins delta r;s}
-\end{equation}
-	Note that the specification of \code{NEWPAIRS} does not determine its implementation uniquely.
-	There are multiple ways to implement \code{NEWPAIRS}.
-	If \code{NEWPAIRS} creates one new pair for every pair in $\Delta-t$,
-	we get $r;\Delta_s=\emptyset$ and $\Delta_r;s=\emptyset$.
-	The weaker condition from the derivation, i.e. $r;\Delta_s\subs t\cup\Delta$ and $\Delta_r;s\subs t\cup\Delta$ allows some reuse of existing pairs.
-	However, any implementation that preserves invariance of equation~\ref{eqn:ins delta r;s} will do.
-
-	At this point we have the ingredients for code that preserves invariance of $\code{T}=\code{R};\code{S}$
-\[\begin{array}{l}
-\hspace{0.5in}\{\text{Assume }\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=t\text{ and }\code{T}=\code{R};\code{S}\}\\
-\code{INSERT }\Delta\code{ INTO T};\\
-\hspace{0.5in}\{\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=t\cup\Delta\text{, so }\code{T}\neq\code{R};\code{S}\}\\
-(\code{DELTAR},\code{DELTAS})\code{ := }\code{NEWPAIRS}(\code{R}, \code{S}, \Delta-\code{T});\\
-\hspace{0.5in}\{\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=t\cup\Delta\text{ and }\\
-\hspace{0.575in}\code{DELTAR};\code{DELTAS}=\Delta-\code{T}\text{ and }\\
-\hspace{0.575in}\code{R};\code{DELTAS}\subs\code{T}\text{ and }\code{DELTAR};\code{S}\subs\code{T}\text{, so equation~\ref{eqn:ins delta r;s} is satisfied}\}\\
-\code{INSERT DELTAR INTO R};\\
-\hspace{0.5in}\{\code{R}=r\cup\code{DELTAR}\text{ and }\code{S}=s\text{ and }\code{T}=t\cup\Delta\}\\
-\code{INSERT DELTAS INTO S}\\
-\hspace{0.5in}\{\code{R}=r\cup\code{DELTAR}\text{ and }\code{S}=s\cup\code{DELTAS}\text{ and }\code{T}=t\cup\Delta\text{, so }\code{T}=\code{R};\code{S}\}
-\end{array}\]
-    This shows how to satisfy invariant $\code{T}=\code{R};\code{S}$ when pairs $\Delta$ are inserted into \code{T}.
-
-	Let us now study how to keep $r;s$ constant when $r$ changes.
-	Inserting pairs $\Delta$ into $r$ involves deleting pairs from $s$ such that:
-\begin{equation}
-(r\cup \Delta);(s-\Delta_s)=r;s\label{eqn:ins r del s}
-\end{equation}
-	In order to solve equation~\ref{eqn:ins r del s} towards $\Delta_s$,
-	we derive\footnote{Thanks to Peter Jipsen, who pointed out this line of reasoning.}:
-\begin{eqnarray}
-&&(r\cup \Delta);(s-\Delta_s)\ \subs\ r;s\label{eqn:rs subs t by ins r del s}\\
-&\Leftrightarrow&\nonumber\\
-&&s-\Delta_s\ \subs\ (r\cup \Delta)\backslash(r;s)\nonumber\\
-&\Leftrightarrow&\hspace{1in}\{\text{apply }x-y=z\ \Leftrightarrow\ x=z\cup y\text{ twice.}\}\nonumber\\
-&&s-(r\cup \Delta)\backslash(r;s)\ \subs\ \Delta_s\nonumber
-\end{eqnarray}
-	This derivation shows that every solution $\Delta_s$ contains $s-(r\cup \Delta)\backslash(r;s)$.
-	So the latter, $s-(r\cup \Delta)\backslash(r;s)$, is the smallest solution.
-
-	The smallest solution might not contain all pairs from $r;s$.
-	The missing pairs are $r;s-(r\cup\Delta);(s-(r\cup \Delta)\backslash(r;s))$.
-	Therefore it may be necessary to reintroduce these missing pairs afterwards.
-	Fortunately, we have just shown how to do just that with \code{NEWPAIRS}.
-	This provides us with all the ingredients for code that preserves invariance of $\code{R};\code{S}=\code{T}$
-\[\begin{array}{l}
-\hspace{0.5in}\{\text{Assume }\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=r;s\text{, so }\code{R};\code{S}=\code{T}\}\\
-\code{INSERT }\Delta\code{ INTO R};\\
-\hspace{0.5in}\{\code{R}=r\cup\Delta\text{ and }\code{S}=s\text{ and }\code{T}=r;s\text{, so }\code{R};\code{S}\neq\code{T}\}\\
-\code{DELETE R}\backslash\code{T FROM S};\\
-\hspace{0.5in}\{\text{Since }\code{R}=r\cup\Delta\text{ and }\code{T}=r;s\text{, we have }\code{R}\backslash\code{T}=(r\cup \Delta)\backslash(r;s)\\
-\hspace{0.575in}\text{So }\code{S}=s-(r\cup \Delta)\backslash(r;s)\text{ and therefore}\\
-\hspace{0.575in}\code{R};\code{S}\subs r;s\}\\
-(\code{DELTAR},\code{DELTAS})\code{ := }\code{NEWPAIRS}(\code{R},\ \code{S},\ \code{T}-\code{R};\code{S});\\
-\code{INSERT DELTAR INTO R};\\
-\code{INSERT DELTAS INTO S}\\
-\hspace{0.5in}\{\code{R};\code{S}=r;s\cup(\code{T}-\code{R};\code{S})=r;s\\
-\hspace{0.575in}\text{so }\code{R};\code{S}=\code{T}\}
-\end{array}\]
-
-	Let us now study how to delete pairs $\Delta$ from $r$ while keeping $r;s$ constant.
-	This involves inserting pairs into $s$ such that:
-\begin{equation}
-(r-\Delta);(s\cup\Delta_s)=r;s\label{eqn:del r ins s}
-\end{equation}
-	The largest $\Delta_s$ follows from the law $x;(x\backslash y)\subs y$ for $x=r-\Delta$ and $y=r;s$:
-\begin{eqnarray}
-&&(r-\Delta);((r-\Delta)\backslash(r;s))\ \subs\ r;s\nonumber\\
-&\Rightarrow&\hspace{1in}\{\text{Let }\Delta_s\ =\ (r-\Delta)\backslash(r;s)\}\nonumber\\
-&&(r-\Delta);\Delta_s\ \subs\ r;s\nonumber\\
-&\Leftrightarrow&\hspace{1in}\{p;x\subs p;q\ \Leftrightarrow\ p;(q\cup x)\subs p;q\}\nonumber\\
-&&(r-\Delta);(s\cup\Delta_s)\ \subs\ r;s\nonumber
-\end{eqnarray}
-	This derivation shows that we can choose $\Delta_s=(r-\Delta)\backslash(r;s)$ as the largest set that preserves invariant~\ref{eqn:del r ins s}.
-
-	Once again, we might not have all pairs from $r;s$,
-	so equation~\ref{eqn:del r ins s} needs to be completed with the missing pairs.
-	The missing pairs are $r;s-(r-\Delta);(s\cup(r-\Delta)\backslash(r;s))$.
-	In this case too, these missing pairs can be reintroduced afterwards.
-	This provides us with all the ingredients for code that preserves invariance of $\code{R};\code{S}=\code{T}$
-\[\begin{array}{l}
-\hspace{0.5in}\{\text{Assume }\code{R}=r\text{ and }\code{S}=s\text{ and }\code{T}=r;s\text{, so }\code{R};\code{S}=\code{T}\}\\
-\code{DELETE }\Delta\code{ FROM R};\\
-\hspace{0.5in}\{\code{R}=r-\Delta\text{ and }\code{S}=s\text{ and }\code{T}=r;s\text{, so possibly }\code{R};\code{S}\neq\code{T}\}\\
-\code{INSERT R}\backslash\code{T INTO S};\\
-\hspace{0.5in}\{\text{Since }\code{R}=r-\Delta\text{ and }\code{T}=r;s\text{, we have }\code{R}\backslash\code{T}=(r-\Delta)\backslash(r;s)\\
-\hspace{0.575in}\text{So }\code{S}=s\cup(r-\Delta)\backslash(r;s)\text{ and therefore}\\
-\hspace{0.575in}\code{R};\code{S}\subs r;s\}\\
-(\code{DELTAR},\code{DELTAS})\code{ := }\code{NEWPAIRS}(\code{R},\ \code{S},\ \code{T}-\code{R};\code{S});\\
-\code{INSERT DELTAR INTO R};\\
-\code{INSERT DELTAS INTO S}\\
-\hspace{0.5in}\{\code{R};\code{S}=r;s\cup(\code{T}-\code{R};\code{S})=r;s\\
-\hspace{0.575in}\text{so }\code{R};\code{S}=\code{T}\}
-\end{array}\]
-
-	Summarizing, we have obtained some more code to preserve invariants in a relation algebra:
-\[\begin{array}{|l|l|l|l|}
-\hline
-\text{invariants}&\text{action}&\text{reaction}&\text{equation}\\ \hline
-\begin{array}{l}
-\code{R};\code{S}\subs\code{T}\\
-\code{S}\subs\code{R}\backslash\code{T}\\
-\code{R}\subs\code{T}\slash\code{S}
-\end{array}&\code{INSERT}\ \Delta\code{ INTO R}&\code{DELETE R}\backslash\code{T FROM S}&\ref{eqn:rs subs t by ins r del s}\\ \hline
-\begin{array}{l}
-\code{R};\code{S}\subs\code{T}\\
-\code{S}\subs\code{R}\backslash\code{T}\\
-\code{R}\subs\code{T}\slash\code{S}
-\end{array}&\code{DELETE}\ \Delta\code{ FROM R}&\code{Nop}&\\ \hline
-\code{T}\subs\code{R};\code{S}&\code{INSERT}\ \Delta\code{ INTO T}&
-\begin{array}{l}
-(\code{DELTAR},\code{DELTAS})\code{:=NEWPAIRS}(\code{R},\ \code{S},\ \Delta-\code{T});\\
-\code{INSERT DELTAR INTO R};\\
-\code{INSERT DELTAS INTO S}
-\end{array}&\ref{eqn:ins delta r;s}\\ \hline
-\code{T}\subs\code{R};\code{S}&\code{DELETE}\ \Delta\code{ FROM T}&\code{Nop}&\\ \hline
-\code{T}\subs\code{R};\code{S}&\code{INSERT}\ \Delta\code{ INTO R}&\code{Nop}&\\ \hline
-\code{T}\subs\code{R};\code{S}&\code{DELETE}\ \Delta\code{ FROM R}&\code{INSERT R}\backslash\code{T INTO S}&\\ \hline
-\code{R};\code{S}=\code{T}&\code{INSERT}\ \Delta\code{ INTO R}&
-\begin{array}{l}
-\code{DELETE R}\backslash\code{T FROM S};\\
-(\code{DELTAR},\code{DELTAS})\code{:=NEWPAIRS}(\code{R},\ \code{S},\ \code{T}-\code{R};\code{S});\\
-\code{INSERT DELTAR INTO R};\\
-\code{INSERT DELTAS INTO S}
-\end{array}&\ref{eqn:ins r del s}\\ \hline
-\code{R};\code{S}=\code{T}&\code{DELETE}\ \Delta\code{ FROM R}&
-\begin{array}{l}
-\code{INSERT R}\backslash\code{T INTO S};\\
-(\code{DELTAR},\code{DELTAS})\code{:=NEWPAIRS}(\code{R},\ \code{S},\ \code{T}-\code{R};\code{S});\\
-\code{INSERT DELTAR INTO R};\\
-\code{INSERT DELTAS INTO S}
-\end{array}&\ref{eqn:del r ins s}\\ \hline
-\end{array}\]
-	The solutions given here are far from being exhaustive.
-	We illustrate this point by giving an example.
-	Suppose we have invariant $\code{T}\subs\code{R};\code{S}$ and action $\code{DELETE}\ \Delta\code{ FROM R}$.
-	So the question is which pairs must be added to $s$ after pairs $\Delta$ have been removed from $r$.
-	In normal circumstances, we might just look it up in the table above, picking $\code{INSERT R}\backslash\code{T INTO S}$ as a reaction.
-	In the circumstance that initially $\code{T}=r;s$, we might reason differently.
-	We must still find a smallest value for $\Delta_s$ that satisfies:
-\begin{equation}
-r;s\ \subs\ (r-\Delta);(s\cup\Delta_s)\label{lbl:nonshrinking r;s when Del from r}
-\end{equation}
-	The following derivation shows that $\flip{(r-\Delta)};(r\cap\Delta);s$ satisfies this requirement conditionally.
-	Here is the derivation:
-\begin{eqnarray}
-&&\flip{(r-\Delta)};(r\cap\Delta);s\ \subs\ \Delta_s\nonumber\\
-&\Rightarrow\nonumber\\
-&&(r-\Delta);\flip{(r-\Delta)};(r\cap\Delta);s\ \subs\ (r-\Delta);\Delta_s\nonumber\\
-&\Rightarrow&\hspace{1in}\{r;s;\flip{s};\flip{r} \cap\iden\ \ \subs\ (r-\Delta);\flip{(r-\Delta)}\}\nonumber\\
-&&(r\cap\Delta);s\ \subs\ (r-\Delta);\Delta_s\nonumber\\
-&\Leftrightarrow\nonumber\\
-&&(r-(r-\Delta));s\ \subs\ (r-\Delta);\Delta_s\nonumber\\
-&\Rightarrow\nonumber\\
-&&(r;s)-((r-\Delta);s)\ \subs\ (r-\Delta);\Delta_s\nonumber\\
-&\Leftrightarrow\nonumber\\
-&&r;s\ \subs\ (r-\Delta);s \cup (r-\Delta);\Delta_s\nonumber\\
-&\Leftrightarrow\nonumber\\
-&&r;s\ \subs\ (r-\Delta);(s \cup \Delta_s)\nonumber
-\end{eqnarray}
-	This derivation justifies the reaction $\code{INSERT }\flip{(r-\Delta)};(r\cap\Delta);s\code{ INTO S}$,
-	but only if $r;s;\flip{s};\flip{r} \cap\iden\ \ \subs\ (r-\Delta);\flip{(r-\Delta)}$.
-	The part $r;s;\flip{s};\flip{r}$ is not necessary in this proof, but it is useful in practice.
-	It says that $r-\Delta$ must contain a pair for every element in the domain of $r;s$,
-	instead of every element in $I$.
-	This makes this result applicable in more situations.
 
@@
 \label{sct:Reflection}
-	Ampersand is being built to automate software development.
-	The holy grail is to derive deterministic ways to preserve invariants, to allow programming by constraints.
-	This is motivated by the benefits of describing a result in terms of constraints,
-	rather than having to invent all the steps that lead to that result.
-	For example, describing the situation that any financial transaction must preserve the total amount of money is simpler
-	than having to describe all the steps involved in such a transaction.
-	The attraction is not only for the programmer, but even more so for auditors and users for who the constraints are easier to grasp and validate.
-	In this paper we have shown how to derive specific steps,
-	helping the programmer to invent code and providing mathematical assurance of correctness (compliance-by-design).
+	Ampersand is built on the belief that software development should be automated.
+	It uses rules to represent requirements, in the firm belief that consistent requirements are essential to the software development process~\cite{Boehm1981}.
+	Being fully aware of the fate of formal methods in computer science, Ampersand is founded on the belief that software development must be done more formally.
+	So Ampersand is building further on the foundations laid by formal specification methods
+	such as Z~\cite{Z} and Alloy~\cite{Alloy2006}.
+	In contrast however with such methods, Ampersand is equipped with a software generator that generates an information system.
+	Thus, specifying in Ampersand is developing software at the same time.
+	Ampersand complements the RelView approach~\cite{Berghammer2005}, which also generates software but is stronger in specifying complex computational problems.
 
-	The analysis in this paper shows once more that inventing algorithms for preserving invariants is far from trivial.
-	Algorithms cannot be derived from constraints alone. The choices that remain must be resolved first.
-	This paper does show, however, that these choices occur in predictable ways.
-	That gives opportunities, for instance to make these choices explicit to a user, to optimize the number of choices,
-	or to use external knowledge to resolve the remaining choices at run time.
-
-	In practice we have experienced that programming by constraints in relation algebra yields an unconventional way of programming.
+	In retrospect we see that programming in relation algebra yields an unconventional programming experience.
 	This experience consists of inventing rules and choosing event types for resolving violations, as illustrated by the case study.
-	To relate this experience to programming as we know it, section~\ref{sct:Comparison} makes a comparison with established programming languages.
+	To relate this experience to programming as we know it, section~\ref{sct:Comparison} makes a comparison with established programming languages (prior art).
 	Section~\ref{sct:Contribution} summarizes the contributions to software development claimed by Ampersand.
@@
 
-	The first to compare with is the object oriented programming paradigm, known from popular languages such as Java~\cite{Java} and C++~\cite{Stroustrup97a}.
-	Ampersand cannot be called object oriented, because the notion of control flow in object oriented languages is very different.
-	In those languages, the threads in the control flow are controlled by the software developer.
-	As shown in our case study, the control flow in an Ampersand script emerges as a result of changes in relation content as illustrated
+	The first to compare with is the imperative programming paradigm, known from popular languages such as Java~\cite{Java} and C++~\cite{Stroustrup97a}.
+	Our case study cannot be called imperative, because the notion of control flow in imperative languages is very different.
+	In imperative languages, the control flow is defined by the software developer.
+	In this case study, the control flow emerges as a result of changes in relation content as illustrated
 	by figure~\ref{fig:event flow}.
 
-	Ampersand also differs from the logic programming paradigm of Prolog~\cite{Lloyd1984}
-	and of all rule engines that can be considered to be offspring of Prolog.
+	The case study also differs from the logic programming paradigm of Prolog~\cite{Lloyd1984} and of all rule engines that can be considered to be an offspring of Prolog.
 	A difference lies in the way rules are interpreted.
-	In logic programming, a program consists of Horn-clauses and a resolution-proof is constructed at run time.
+	In logic programming, a program consists of Horn-clauses and a resolution-proof is constructed on run-time.
 	The software developer works with notions such as backtracking (backward chaining) and unification, which are absent in Ampersand.
-	On the other hand, Ampersand is not restricted to implications with one predicate on the consequent side, as Prolog does.
-        Ampersand allows rules, which are equations with arbitrary relational expressions on either side of the equals sign.
+	Ampersand is not restricted to Horn-clauses; any relation-algebraic equation over relations can be used as a rule.
 
-	Ampersand also differs from functional programming, of which Haskell and Scala are prominent representants.
+	The case study also differs from functional programming, of which Haskell and Scala are prominent representants.
 	A core idea in functional programming is to evaluate a program as a function from input to output~\cite{Backus1978}.
-	A functional program consists of function definitions, that are evaluated by a term- or graph-rewriter,
-	using various strategies such as lazy or eager evaluation.
-	In contrast, our case study illustrates that not everything has to be a function.
+	A functional program consists of function definitions, that are evaluated by a term- or graph-rewriter, using various strategies such as lazy or eager evaluation.
+	In contrast, one might interpret our case study as a relaxation of the constraint that everything is a function.
 	In relation algebra, everything is a relation and a function is a restricted form of a relation.
-	A similarity to functional programming is shown by the declarative style,
-	because substitution of equal terms without changing the semantics is a property we see in both worlds.
+	A similarity to functional programming is the declarative style, because substitution of equal terms without changing the semantics is a property we see in both worlds.
 
 	A difference with database programming is found in the type of algebra that is used.
-	Conventional databases are typically programmed in SQL, which is based on n-ary relations in relational algebra~\cite{Codd70}.
-	In contrast, Ampersand uses binary relations%
-\footnote{At runtime, Ampersand uses the efficiency of n-ary tables. It does so by translating relation algebra to efficient SQL-code by avoiding as many joins as possible.}
-	based on heterogeneous relation algebra~\cite{Schmidt1997}.
-	In this comparison, Ampersand is closer to graph databases~\cite{Vicknair2010} than to relational databases.
+	Relational databases are founded on relational algebra~\cite{Codd70}. They are typically programmed in SQL.
+	In contrast with database programming, Ampersand implements heterogeneous relation algebra~\cite{Schmidt1997}.
+	A software developer working with relational algebra sees n-ary tables, while Ampersand is restricted to binary relations.
+	The comparison between relation{\it al} algebra and relation algebra might relate to comparing relational databases with graph databases~\cite{Vicknair2010},
+	although no literature was found to corroborate this.
 
@@
 
-	The similarity with the reactive programming paradigm is obvious from figure~\ref{fig:event flow}.
-	This graph is part of the dependency graph, which is a familiar phenomenon in reactive programming.
-	Ampersand may be seen as a reactive programming language on binary relations.
-	It has a unidirectional push based evaluation model, static typing, implicit lifting, and glitch avoidance~\cite{Bainomugisha2013}.
-	Even though Ampersand generates code to execute scripts,
-	code fragments described in section~\ref{sct:Code Fragments}
-	might just as well be generated towards a mature reactive programming language and executed there%
-\footnote{Currently Ampersand generates towards classical PHP+HTML+JavaScript code,
-	but generation to a mature reactive programming language is definitely on the wishlist.}.
-	A distinguishing feature of Ampersand is that it treats constraints (rules) as its central notion,
-	which is also seen in TBAG~\cite{Elliott94} and FHM~\cite{Giorgidze2011} but not common in most reactive programming languages.
-	In the author's opinion a constraint based approach is preferrable, because the correctness of a program is within closer range.
-
-	If Ampersand represents a programming style at all, we might call it a relation-algebraic style of programming or maybe constraint programming\footnote{https://en.wikipedia.org/wiki/Constraint\_programming}.
-	Ampersand's style would be characterized by a programmer who specifies constraints and a computer trying to satisfy these constraints by resolving violations.
+	If Ampersand represents a programming style at all, we might call it ``a relation-algebraic style of programming''.
+	That style would be characterized by a programmer who specifies constraints and a computer trying to satisfy these constraints by resolving violations.
 
@@
 	We have used that to formalize business requirements without exposing business stakeholders to any formal notation.
-\item	Ampersand is statically typed~\cite{Joosten2015}, which is an obvious consequence of being heterogeneous.
-	There is much evidence that static typing yields significantly lower software maintenance cost as opposed to dynamic typing~\cite{HanenbergKRTS14,Petersen2014}.
+\item	Heterogeneous relation algebra in Ampersand is statically typed~\cite{Joosten2015}.
+	There is much evidence for significantly lower software maintenance cost due to static typing as opposed to dynamic typing~\cite{HanenbergKRTS14,Petersen2014}.
 \item	Heterogeneous relation algebra is well studied.
 	As a consequence, many tools that are readily available in the public domain can be put to good use.
-	For executives of large organizations it can be reassuring that the formalism is well studied and understood.
+	For executives of large organizations it can be reassuring that the formalism is free of childhood diseases.
 \item	Relation algebra facilitates composing software from reusable components, because a program consists of rules.
@@
 
-	Ampersand also has a disadvantage. It appears to be difficult to learn for large groups of software professionals.
+	Ampersand also has disadvantages. It appears to be difficult to learn for large groups of software professionals.
 	Research~\cite{Michels2015} shows that this is largely due to deficits in prerequisite knowledge, especially skills in discrete mathematics.
@@
 	Also, programming appears to be difficult in practice.
-	Ongoing research tries to find better ways for learning, which is necessary for large scale use.
 
@@
 	Ampersand has been used in practice both in education (Open University of the Netherlands)
-	and in industry (Ordina and TNO).
-	For example, Ordina designed a proof-of-concept in 2007 of the INDIGO-system,
-	which is still in use today as the core information system of the Dutch immigration authority, IND.
-	The design of INDIGO was made with Ampersand, to obtain correct, detailed results in the least amount of time.
+	and in industry (Ordina and TNO-ICT).
+	For example, Ordina designed a proof-of-concept in 2007 of the INDIGO-system.
+	This design was based on Ampersand, to obtain correct, detailed results in the least amount of time.
+	Today INDIGO is in use as the core information system of the Dutch immigration authority, IND.
 	More recently, Ampersand was used to design an information system called DTV for the Dutch food authority, NVWA.
-	A prototype of DTV was built by Ampersand and was used as a model to build the actual system.
-	TNO, a major Dutch industrial research laboratory, is using Ampersand for research purposes.
-	For example, TNO did a study of international standardizations efforts such as
+	A prototype of DTV was built in Ampersand and was used as a model to build the actual system.
+	TNO-ICT, a major Dutch industrial research laboratory, is using Ampersand for research purposes.
+	For example, TNO-ICT did a study of international standardizations efforts such as
 	RBAC (Role Based Access Control) in 2003 and architecture (IEEE 1471-2000)~\cite{IEEE1471} in 2004.
 	Several inconsistencies were found in the last (draft) RBAC standard~\cite{RBAC}.
-	TNO has also used the technique in conceiving several patents%
+	TNO-ICT has also used the technique in conceiving several patents%
 \footnote{e.g. patents DE60218042D, WO2006126875, EP1727327, WO2004046848, EP1563361, NL1023394C, EP1420323, WO03007571, and NL1013450C.}.
@@
 	Further use of relation algebra can be made by incorporating a model checker, such as the Alloy analyser~\cite{Alloy2006}, to detect inconsistent rules.
-	An exciting new development is Amperspiegel~\cite{Amperspiegel}, which brings notational flexibility to the fingertips of the user.
-	Developments in the Ampersand compiler are going towards a rule-repository (written in Ampersand itself).
+	An exciting new development is Amperspiegel~\cite{Amperspiegel}, which brings notational flexibility at the fingertips of the user.
+	Developments in the Ampersand-compiler are going towards a rule-repository (written in Ampersand itself).
 	This will make collaborative information systems development in Ampersand easier, because the repository can assist in automating
@@
 	In the Ampersand project, a sub-project called ``Formal Ampersand'' is being conducted to achieve this goal.
-\subsection{Acknowledgements}
-\label{sct:Acknowledgements}
-	Firstly, I wish to thank Sebastiaan Joosten for introducing me to Isabelle and for inspiring me with his challenging critiques.
-	I owe many thanks to Wolfram Kahl, Michiel Stornebrink, Han Joosten, and the anonymous reviewers of RAMiCS for their useful feedback.
-	And I am very grateful to the Open University and Ordina for funding this research.
-
-\begin{figure}[htb]
-\begin{center}
-  \includegraphics[angle=90,scale=.357]{LogicalDataModel.pdf}
-\end{center}
-\caption{Conceptual data model (from section~\ref{sct:Conceptual analysis})}
-\label{fig:conceptual model}
-\end{figure}
-
-\section{Bibliography}
-\bibliographystyle{elsarticle-harv}
+\bibliographystyle{spmpsci}
 \bibliography{doc}

### 2019-06-11 2018_Transactions/articleTransactions.tex
Further work on the article on transactions

\documentclass{elsarticle}
\usepackage{graphicx}
%\usepackage{multicol}
%\usepackage{footmisc}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage[english]{babel}
%\usepackage[official,right]{eurosym}
\selectlanguage{english}
\include{preamramics}
\hyphenation{ExecEngine}
\newtheorem{lemma}{Lemma}
\def\id#1{\text{\em #1\/}}
\def\Events{{\mathit E}}
\begin{document}

\title{A Theory for Distributed Database Transactions}
\author[ou,ordina]{Stef Joosten\fnref{fn1}}
\ead{stef.joosten@ou.nl}
\author[utwente]{Sebastiaan Joosten\fnref{fn2}}
\address[ou]{Open Universiteit Nederland, Postbus 2960, 6401 DL Heerlen, the Netherlands}
\address[ordina]{Ordina NV, Nieuwegein, the Netherlands}
\address[utwente]{University of Twente, Enschede, the Netherlands}
\fntext[fn1]{ORCID 0000-0001-8308-0189}
\fntext[fn2]{ORCID 0000-0002-6590-6220}

\begin{abstract}
	A shared data space can be problematic in globally distributed information systems,
	because it can become a performance bottleneck or a single point of failure.
	Yet, transaction theory as we know it assumes a shared data space.
	Several solutions exist, such as nested transactions and distributed transactions.
	Practitioners tend to avoid them when possible becaus the are complicated and error prone.

	In this contribution we let go of the shared data space requirement.
	We propose a transaction theory based on events rather than states.
	By expressing the required properties A (atomic), C (consistent), I (isolated), and D (durable)
	in terms of events we can avoid the assumption of a shared state space.

	This makes transactions suitable for more databases than just relational ones.
	The soundness of the theory has been established by means of formal proof.
	To show that the theory works, the theory is demonstrated by a prototype developed in Ampersand\footnote{\tt https://github.com/AmpersandTarski/Ampersand}.
	To demonstrate that the theory is about real transactions, this contribution defines semantics of conventional database transactions.
	To fulfill our promise, we also gives semantics of transactions on the semantic web.
\end{abstract}

\begin{keyword}
relation algebra\sep software development\sep legal reasoning\sep information systems design\sep Ampersand\sep MirrorMe\sep reactive programming
\end{keyword}
\maketitle

\section{Introduction}
\label{sct:Introduction}
	A distributed information system (DIS) serves its users with things such as
	registering data, guarding consistency, protecting that data,
	and providing it accurately and meaningfully for its intended purpose to various users.
	These users want predictable behaviour, even when they change a shared state,
	e.g. when reserving a seat in a global flight reservation system like Galileo.
	Although we cannot always avoid sharing a state,
	we usually can avoid sharing an entire database.
	Globally distributed applications such as Netflix and LinkedIn cannot afford to rely on a shared database for obvious reasons of performance and reliability.
	Many cloud applications use distributed storage models instead.
	Examples are event stores (e.g.\ Kafka), triple stores (e.g.\ semantic web), graph databases (e.g. Neo4J) and others.
	Many so-called NoSQL databases lack functionality for transactions,
	leaving the problems for the user to solve.
	As a result, architectural patterns exist for the purpose of doing transactions in a distributed setting.
	Some of these patterns simply circumvent the problem.
	Others will isolate transactionality into a smaller sub-problem.
	Still others introduce locking mechanisms to do the job.

	Our work is inspired by the observation that the merge of Git corresponds to the commit that we know from transactional databases.
	We show this correspondence by expressing the characteristic properties of transactions in a theory of time.
	For this purpose we build on a well-known theory of time~\cite{DBLP:journals/ipl/FosterCWZ18},
	which is generally applicable,
	well-established, and founded in a solid mathematical structure~\cite{Foster14}.
	This means we have to redefine Atomicity, Consistency, Isolation, and Durability in terms of events.
	Informally, {\em Atomicity} means that a sequence of primitive actions can be executed as though it were a single action.
	{\em Consistency} means that integrity rules are warranted.
	{\em Isolation} means that actions can be done without any knowledge of what other users are doing.
	{\em Durability} means that history does not change.
	The result of this work contributes to the body of foundational theory of information systems
	by defining transactions in a theory of time.
	By showing that transactions come quite naturally from an event perspective,
	we give formal substance under the observation that event systems like Git do not need new features to behave transactionally.

	\section{Related work}
\label{sct:Related work}
	Distributed information systems are dynamic systems,
	so time is the first aspect to focus on.
	We will define transactions in terms of traces,
	which draws on the seminal work of Hoare on communicating sequential processes~\cite{HoareHe98},
	work on process algebras by Bergstra and Klop~\cite{DBLP:journals/iandc/BergstraK84}, and
	Robin Milner's work on CCS~\cite{Milner1989}.
	Trace theory is today the usual way of defining distributed event-based systems.
	Trace algebras also have sound interpretations for analog real-time systems and hybrid systems~\cite{Foster17b}.
	This paper, however, sticks to the interpretation of event sequences, because distributed information systems are event-based.
	Yet, we define the theory in trace algebra so as to impose as few restrictions as possible.
	We have used Isabelle to verify the required proofs,
	to increase our confidence in this theory.

\cite{Foster17c}
	In the nineties, scientists saw that category theory can be used to formalize database schemas~\cite{Frederiks1997},
	which was advocated as a means to further foundational theories of information systems.
	An interesting read is~\cite{Foster17c}\footnote{\tt\tiny https://towardsdatascience.com/distributed-transactions-and-why-you-should-care-116b6da8d72}.
	It explains how modern distributed database cope with ACID properties in a way that makes sense.

	ACID 2.0 - Associative, Commutative, Idempotent, Distributed

\section{Traces}
\label{sct:Traces}
	We need the notion of {\em event}\footnote{Cambridge Dictionary defines ``event'' as: anything that happens} to describe things that happen in a distributed information system.
	Being distributed means that we must assume that the temporal order of events is partial:
	We do not necessarily know in which order two events have happened.
	To describe this, we define traces algebraically corresponding to Foster~\cite{Foster17b}.

	We define traces as a set ${\mathcal T}$ with
	a concatenation operator $+:\mathcal T\rightarrow\mathcal T\rightarrow\mathcal T$ and
	the empty trace $0:\mathcal T$, which obey the following axioms.
\begin{definition}[Trace Algebra]
\label{def:Trace Algebra}
\item   A trace algebra $(\mathcal T, +, 0)$, is an algebra that satisfies: 
\begin{eqnarray}
	x+(y+z)&=&(x+y)+z\label{axi:associative}\\
	0+x&=&x+0\label{axi:lr-identity}\\
	x+y\ =\ x+z&\Rightarrow&y=z\label{axi:right cancellation}\\
	x+z\ =\ y+z&\Rightarrow&x=y\label{axi:left cancellation}\\
	x+z\ =\ 0&\Rightarrow&x=0\label{axi:no inverse}
\end{eqnarray}
\end{definition}
	In terms of universal algebra, equations~\ref{axi:associative} and~\ref{axi:lr-identity} state that a trace algebra is a {\em monoid}.
	Equations~\ref{axi:right cancellation} and~\ref{axi:left cancellation} state that $+$ is {\em cancellative}.
	Equation~\ref{axi:no inverse} makes the cancellative monoid into a trace algebra.
	It implies that there is no inverse.

	In terms of information systems, a trace represents history.
	For that purpose it makes sense to define the notion of prefix, so we can define notions like shared history.
\begin{definition}[Prefix]
\label{def:Prefix}
\begin{eqnarray}
	x\leq y&\Rightarrow&\exists z:\ x+z=y
\end{eqnarray}
\end{definition}
	We also define subtraction of traces as a way to "cut off" a prefix from a trace.
\begin{definition}[Subtract]
\label{def:Subtract}
\label{def:Subtract}
\begin{array}[t]{rcll}
	y-x&=&\text{if }x\leq y:&\exists z: y=x+z\Rightarrow z\\
	   & &\text{otherwise}:&0
\end{array}
\end{definition}

\section{History}
\label{sct:History}
	The history of a distributed information system (DIS) consists of events and an ordering on these events.
	Because a DIS is distributed,
	we may not know the temporal order of two events in the history structure.
	For this reason we describe the history structure of a DIS as a partial order of events.

	We use variables $e$, $e_1$, and $e_2$ to represent events.
	Let $\mathbb E$ represent the set of all conceivable events.
	We also introduce the notion of {\em history structure}
	to represent all events that have happened up to some point in time and the precedences between those events.
	In this paper, variables $h$, $h'$, $h_1$, and $h_2$ represent history structures.
	Let $\mathbb H$ be the set of all conceivable history structures.

	A {\em history structure} is a tuple $\la \Events, \id{prec}\ra$ in which
\begin{eqnarray}
	&&\Events\ \subseteq\ \mathbb E\\
	&&\id{prec}\ \subseteq\ \Events\times\Events\label{req:precInE}\\
	&&(e,e)\notin\id{prec}\label{req:prec irreflexive}\\
	&&(e_1,e_2)\in\id{prec}^+\ \wedge\ (e_2,e_1)\in\id{prec}^+\ \Rightarrow\ e_1=e_2\label{req:precPlus asy}
\end{eqnarray}
	In a history structure $\la \Events, \id{prec}\ra$ set $\Events$ represents the events that have been registered up to some point in time.
	For every event $e\in\Events$ we can say that it has happened (at some time in the past).
	Let $e_1\ \id{prec}\ e_2$ represent the fact that event $e_1$ has happened before event $e_2$,
	making $\id{prec}$ a relation on events in $\Events$.
	Requirement~\ref{req:prec irreflexive} excludes the situation that $e$ has happened before itself.
	To represent time faithfully, we require that $\la\Events,\id{prec}^*\ra$ is a partially ordered set,
	meaning that $\id{prec}^*$ is reflexive, transitive, and antisymmetric.
	For this reason we have requirement~\ref{req:precPlus asy}, which says that the transitive closure $\id{prec}^+$ must be antisymmetric.

	In a history structure $\la \Events, \id{prec}\ra$, an event $e$ that precedes another event is called a {\em past} event,
	which we define as \ $\exists e_1\in\Events: e\ \id{prec}\ e_1$.
	All other events are called {\em current} events in that history structure.
	Function $\id{current}:\mathbb H\rightarrow\powerset{\Events}$ computes the current events in a history structure.
	Events in $\Events$ that have no preceding events are called {\em initial} events.
	Function $\id{past}$ computes the past events and $\id{initial}$ computes the initial events.
\begin{eqnarray}
	\id{current}(\la \Events, \id{prec}\ra)&=&\{ c\in\Events|\ \neg\exists e_1\in\Events: e\ \id{prec}\ e_1 \}\\
	\id{initial}(\la \Events, \id{prec}\ra)&=&\{ c\in\Events|\ \neg\exists e_1\in\Events: e_1\ \id{prec}\ e \}
\end{eqnarray}
	We define the {\em closure} of a history structure as the history structure whose order is the reflexive, transitive closure of the original ordering relation.
\begin{eqnarray}
	\la \Events, \id{prec}\ra^*&=&\la \Events, \id{prec}^*\ra\label{def:closure}
\end{eqnarray}
	Equation~\ref{def:closure} allows us to write $h^*$ for elements of $\mathbb H$.
	The order of $h^*$ is reflexive, transitive, and antisymmetric.

	The special case in which $\la\Events,\id{prec}^*\ra$ is a total order is called a {\em sequential history structure}.
	Note that any nonempty finite sequential history structure has one current event and one first event.
	Other finite history structures may have more.
	In a sequential history structure, every event except the first one has precisely one preceding event.
	In general, however, events may have multiple preceding events and multiple succeeding events.

	We extend the notion of subset to historic structures and use the subset symbol $\subseteq$ for that purpose.
	We use $h\subseteq h'$ to state that $h$ lies in the past of $h'$.
	Let $h=\la \Events, \id{prec}\ra$ and $h'=\la \Events', \id{prec}'\ra$ in the definition of $h\subseteq h'$:
\begin{equation}
\label{def:lies in the past}
	h\subseteq h'\ \Leftrightarrow\ 
	\Events\subseteq\Events'\ \wedge\ \id{initial}(h)\subseteq\id{initial}(h')\ \wedge\ \id{prec}^*\subseteq\id{prec}'^*
\end{equation}

	Durability is meant to express that an action preserves history:
\begin{definition}[Durability]
\item An action $\id{act}:\mathbb H\rightarrow\mathbb H$ is durable iff for every $h\in\mathbb H$:
\label{def:Durability}
\begin{eqnarray}
	h\subseteq\id{act}(h)
\end{eqnarray}
\end{definition}

	Isolation is meant to express that the effect of an action depends only on its current state.
\begin{definition}[Isolation]
\item An action $\id{act}:\mathbb H\rightarrow\mathbb H$ works in isolation iff there is a function $f:\mathbb S\rightarrow\mathbb S$ such that:
\label{def:Isolation}
\begin{eqnarray}
	(h',e')=\id{act}(h,e)&\Rightarrow&\id{stat}(h',e')=f(\id{stat}(h,e))
\end{eqnarray}
\end{definition}

	Atomicity expresses that a sequence of multiple actions can be treated as one.
\begin{definition}[Atomicity]
\item An action $\id{act}:\mathbb H\rightarrow\mathbb H$ is atomic iff its constituent actions ...
\label{def:Atomicity}
\begin{eqnarray}
	(h',e')=\id{act}(h,e)&\Rightarrow&e\ \id{pred}^*\ e'
\end{eqnarray}
\end{definition}

	Consistency of an action means that it preserves truth.
\begin{definition}[Consistency]
\item An action $\id{act}:\mathbb H\rightarrow\mathbb H$ is consistent iff it preserves the integrity constraints of all contexts in the system.
\label{def:Consistency}
\begin{eqnarray}
	(h',e',r')=\id{act}(h,e,r)\ \wedge\ r(\id{stat}(h,e))&\Rightarrow&r'(\id{stat}(h',e'))
\end{eqnarray}
\end{definition}

\section{Progress}
\label{sct:Progress}

	A DIS grows by adding events to it.
	We describe each time a new event $e$ happens as a mapping, $\id{happens}_{P,e}$, from one history structure to another:
\begin{equation}
\begin{array}[b]{l}
	P\subseteq\Events\ \wedge\ e\notin E\\
	\ \ \Rightarrow\ \ \id{happens}_{P,e}(\la \Events, \id{prec}\ra)\ =\ \la \Events\cup\{e\}, \id{prec}\cup\{(e_1,e) | e_1\in P\}\ra
\end{array}
\label{req:def happens}
\end{equation}
	This happening adds one event to $\Events$ and enlarges relation $\id{prec}$ with one pair for every immediate predecessor of $e$.
	To understand why $P\subseteq\Events$ is necessary,
	consider $\la \Events', \id{prec}'\ra=\id{happens}_{P,e}(h)$.
	Now suppose event $e_1\in P$ but $e_1\notin\Events$.
	Equation~\ref{req:def happens} tells us that $(e_1,e)\in\id{prec}'$.
	However, this violates requirement~\ref{req:precInE}.
	So it is necessary that every $e_1\in P$ is an element of $\Events$.
	The immediate predecessors of $e$ are specified in set $P$, which we call the predecessor set.
	We need the predecessor set to cater for the distributed nature of a DIS.
\begin{lemma}
\label{lemma:happens preserves history structure}
	If $\la \Events, \id{prec}\ra$ is a history structure and $P\subseteq\Events$ and $e\notin\Events$
	then $\id{happens}_{P,e}(\la \Events, \id{prec}\ra)$ is a history structure.
\end{lemma}
	Lemma~\ref{lemma:happens preserves history structure} is a preservation property.
	It says that $\id{happens}_{P,e}$ preserves the history structure,
	provided that $P\subseteq\Events$ and $e\notin\Events$.
	This lemma has been proven in proof assistant Isabelle/HOL.
	Condition $P\subseteq\Events$ is necessary to satisfy equation~\ref{req:precInE}.
	Condition $e\notin\Events$ is necessary too,
	because any $e\in\Events$ violates the antisymmetry of $\id{prec}^+$ (requirement~\ref{req:precPlus asy}).

	With lemma~\ref{lemma:happens preserves history structure} we can prove that $\id{happens}_{P,e}(h)=h'$ implies that $h\subseteq h'$.

	Let us define predicate $\id{HS}\la \Events, \id{prec}\ra$
	to denote that $\la \Events, \id{prec}\ra$ is a history structure.
\begin{eqnarray}
	&&\id{HS}\la \emptyset, \emptyset\ra\label{def:zeroHS}\\
&&\begin{array}{@{}l}
	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{HS}\la \Events, \id{prec}\ra\ \Rightarrow\ \id{HS}(\id{happens}_{P,e}\la \Events, \id{prec}\ra)
\end{array}
\label{def:HS}
\end{eqnarray}
	Equation~\ref{def:zeroHS} says that $\la \emptyset, \emptyset\ra$ is a history structure.
	Equation~\ref{def:HS} specifies recursively that $\id{happens}_{P,e}(h)$ is a history structure iff
	$h$ is a history structure and if $P$ and $e$ satisfy the specified conditions.
	This definition allows us to prove properties of finite history structures by induction.
	In practical situations, every history structure is finite.

	Note that $\id{happens}_{P,e}$ can make current events become past events.
	If $h'=\id{happens}_{P,e}(h)$, every event in $P$ is a past event in $h'$ and event $e$ becomes current in $h'$.

\section{States}
	This section defines the requirements for predictable behavior of a distributed information system (DIS).
	We consider behavior to be predictable if every state can be computed unambiguously from its history.
	For that purpose we introduce the notion of state.
	Let $\mathbb S$ represent the set of all conceivable states.

	This section expands history structures to stateful history substructures
	to represent the (global) state of an information system at any point in time.
	Since this paper is about distributed systems,
	we may not assume that the global state of a DIS at some point in time is observable.
	The state of a DIS is observable only in its local parts.
	So we shall define multiple states, each of which may be interpreted locally.

	We define a {\em stateful history structure} as a triple $\la\Events, \id{prec}, \id{state}\ra$ in which
	$\la \Events, \id{prec}\ra$ is a history structure and $\id{state}:\Events\rightarrow{\mathbb S}$ is a function
	that maps every event in $\Events$ to a state.
	We will use predicate $\id{SHS}\la \Events, \id{prec}, \id{state}\ra$
	to express that $\la \Events, \id{prec}, \id{state}\ra$ is a stateful history structure.
	Let $s=\id{state}(e)$ represent the fact that $s$ is the state immediately after event $e$ has happened.
	Or shorter: $s$ is the state after event $e$.
%	Instead of $s=\id{state}(e)$, we may also write $(s,e)\in\id{state}$ because a function is a univalent and total relation
%	(and therefore a set of pairs).
	A {\em current state} is the state after a current event
	and a {\em past state} is the state after a past event.

	In a stateful history structure, the occurrence of an event $e$ defines a new history structure.
	This is the way systems evolve over time.
	So we define the occurrence of an event as a mapping, $\id{occurs}_{P,e,s}$, from one SHS to another SHS:
\begin{equation}
\begin{array}{l}
	P\subseteq\Events\ \wedge\ e\notin E\\
	\ \ \Rightarrow\ \ 
\begin{array}[t]{l}
	\id{occurs}_{P,e,s}(\la\Events, \id{prec}, \id{state}\ra)\\
	\ \ =\ \begin{array}[t]{@{}l@{\hspace{.2em}}c@{\hspace{.2em}}ll}
		\text{Let }\id{newState}(x)&=&s\text{,}&\text{if }x=e\\
				&=&\id{state}(x)\text{,}&\text{otherwise}\\
		\multicolumn{4}{@{}l}{\text{in }\la\Events\cup\{e\}, \id{prec}\cup\{(e_2,e)|e_2\in P\}, \id{newState}\}\ra}
		\end{array}
\end{array}
\end{array}
\label{req:def occurs}
\end{equation}
	This definition extends the definition of $\id{happens}$ (eqn.~\ref{req:def happens}).
	By using lemma~\ref{lemma:happens preserves history structure} we can prove that $\id{occurs}_{P,e,s}$ preserves
	the stateful history structure.
	Also, we will use the subset symbol $\subseteq$ to represent the fact that $h$ is a {\em stateful historic substructure} of $h'$.
	And it will come as no surprise that $\id{occurs}_{P,e,s}(h)=h'$ implies that $h\subseteq h'$.
	We will use predicate $\id{SHS}\la \Events, \id{prec}, \id{state}\ra$
	to express that $\la \Events, \id{prec}, \id{state}\ra$ is a stateful history structure.
	To describe the growth of a stateful history structure by the event,
	we define $\id{SHS}$ recursively using $\id{occurs}_{P,e,s}$ as recurring step:
\begin{eqnarray}
	&&\id{SHS}\la \emptyset, \emptyset, \emptyset\ra\\
&&\begin{array}{@{}l}
	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{SHS}\la \Events, \id{prec}, \id{state}\ra\\
	\ \ \Rightarrow\ \id{SHS}(\id{occurs}_{P,e,s}\la \Events, \id{prec}, \id{state}\ra)
\end{array}
\label{req:recursive def occurs}
\end{eqnarray}

\section{Transactions}
\label{sct:Transactions}
	In practice, a new state is computed from the current state by specifying a change, $\Delta$.
	Such changes are called {\em transactions}.
	To describe a DIS in terms of changes, we need operators to compute a new state ($s_2$) from a given state ($s_1$).
	In fact, we need two operators, $\oplus$ and $\ominus$, such that
\begin{equation}
	s_2\ =\ s_1\oplus(s_2\ominus s_1)
\label{req:oplus and ominus}
\end{equation}
	Let us call $s_2\ominus s_1$ the difference ($\Delta$) between states $s_2$ and $s_1$.
	We use $s_1\oplus\Delta$ to compute a new state from an existing state $s_1$ and a change $\Delta$.

	We can define a stateful history structure,
	in which the state after every event is computed from a change, $\Delta$.
	For this purpose we must designate one specific state, $\mathcal I\in\mathbb S$, as the initial state.
	We must additionally specify an existing event $e_1$ in order to compute $\id{state}(e_1)\oplus\Delta$.
	The following mapping $\id{occurByDelta}_{P,e_1,\Delta,e}$ defines how an SHS grows by one event ($e$)
	using only the change between the state of event $e_1$ to the new event $e$:
\begin{eqnarray}
&&\begin{array}{l}
	\id{occurByDelta}_{P,e_1,\Delta,e}(\la\Events, \id{prec}, \id{state}\ra)\\
	\ \ =\ \begin{array}[t]{@{}l@{\hspace{.2em}}c@{\hspace{.2em}}ll}
		\text{Let }s_1&=&\id{state}(e_1)\text{,}&\text{if }e_1\in P\\
				&=&\mathcal I\text{,}&\text{otherwise}\\
		\text{Let }\id{newState}(x)&=&s_1\text{,}&\text{if }x=e\\
		&=&\id{state}(x)\text{,}&\text{otherwise}\\
		\multicolumn{4}{@{}l}{\text{in }\la\Events\cup\{e\}, \id{prec}\cup\{(e_2,e)|e_2\in P\}, \id{newState}\}\ra}
		\end{array}
\end{array}
\label{req:SHS by Delta}
\end{eqnarray}
	This mapping serves as recursion step in the following definition of a stateful history structure
\begin{eqnarray}
	&&\id{SHS}\la \emptyset, \emptyset, \emptyset\ra\\
&&\begin{array}{@{}l}
	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{SHS}\la \Events, \id{prec}, \id{state}\ra\\
	\ \ \Rightarrow\ \id{SHS}(\id{occurByDelta}_{P,e_1,\Delta,e}\la \Events, \id{prec}, \id{state}\ra)
\end{array}
\label{req:recursive def SHS by Delta}
\end{eqnarray}
	In practice, information systems build states incrementally.
	Each event comes with an increment ($\Delta$) that changes the current state.
	For this purpose we define an incremental history structure (IHS).
	A {\em incremental history structure} is a quadruple $\la\Events, \id{prec}, \delta, \iota\ra$ in which:
\begin{itemize}
\item	$\la \Events, \id{prec}\ra$ is a history structure;
\item	$\delta(e_1,e_2)$ represents a change, $\id{state}(e_2)\ominus\id{state}(e_1)$, for every pair $(e_1,e_2)\in\id{prec}$;
\item	$\iota(e)$ represents the change $\id{state}(e_1)\ominus\mathcal I$ for every event $e\in\Events$ that has no predecessor in $\id{prec}$.
\end{itemize}
	Since histories are built event-by-event, we define IHSs incrementally.
	For that purpose we use the following mapping, $\id{increment}_{P,e_1,e,\Delta,\iota}$,
	which adds one event ($e$) to an IHS:
\begin{eqnarray}
\begin{array}{l@{}l}
	\multicolumn{2}{l}{\id{increment}_{P,e_1,e,\Delta}\la\Events, \id{prec}, \delta, \iota\ra}\\
	\ \ =\ &\begin{array}[t]{@{}l@{\hspace{0.2cm}}c@{\hspace{0.2cm}}ll}
		\text{Let }\delta'(x,y)&=&\Delta\text{, }&\text{if }x=e_1\wedge y=e\wedge x\in P\\
				       &=&(\id{state}(e_1)\oplus\Delta)\ominus\id{state}(x)\text{, }&\text{if }x\neq e_1\wedge y=e\wedge x\in P\\
				       &=&\delta(x,y)\text{, }&\text{if }(x,y)\in\id{prec}
		\end{array}\\
		&\begin{array}[t]{@{}l@{\hspace{0.2cm}}c@{\hspace{0.2cm}}ll}
			\text{Let }\iota'(x)&=&\Delta\text{, }&\text{if }x=e\wedge P=\emptyset\\
				    &=&\iota(x)\text{, }&\text{if }x\neq e\wedge P=\emptyset
		\end{array}\\
		&\text{in }\la\Events\cup\{e\}, \id{prec}\cup\{(e_2,e)|e_2\in P\}, \delta', \iota'\}\ra
\end{array}
\label{req:increment}
\end{eqnarray}
	This definition inserts a new event, $e$, into a history structure with $e_1$ as its predecessor.
	The new state, $\id{state}(e)$, is computed by $\id{state}(e_1)\oplus\Delta$; let us call it $s$.
	Beside $e_1$, there may be other events preceding $e$ in the predecessor set $P$.
	The change from every predecessor $x$ can be computed by $s\ominus\id{state}(x)$.
	All these changes are added to the function $\delta$.
	The function $\iota$ is there for initialization purposes.
	To get an initial history structure (with no events in it) going,
	we must add events while there are no other events available as predecessor.
	We use the empty predecessor set as condition to signal that there is no previous state.
	The mapping $\id{increment}_{P,e_1,e,\Delta}$ can be used as recursive step in defining predicate $\id{IHS}$,
	which characterizes an incremental history structure:
\begin{eqnarray}
	&&\id{IHS}\la \emptyset, \emptyset, \emptyset, \emptyset\ra\\
&&\begin{array}{@{}l}
	P\subseteq\Events\ \wedge\ e\notin E\ \wedge\ \id{IHS}\la \Events, \id{prec}, \delta, \iota\ra\\
	\ \ \Rightarrow\ \id{IHS}(\id{increment}_{P,e_1,e,\Delta}\la \Events, \id{prec}, \delta, \iota\ra)
\end{array}
\label{req:recursive def IHS}
\end{eqnarray}
The following mapping transforms a stateful history structure to an incremental history structure:
\begin{eqnarray}
&&\begin{array}{l}
	\id{shs2ihs}(\la\Events, \id{prec}, \id{state}\ra)\\
	\ \ =\ \begin{array}[t]{@{}l}
		\text{Let }\forall(e_1,e_2)\in\id{prec}:\ \delta(e_1,e_2)=\id{state}(e_2)\ominus\id{state}(e_1)\\
		\text{Let }\id{initial}=\{e\in\Events|\forall e_1\in\Events: (e_1,e)\notin\id{prec}\}\\
		\text{Let }\forall e\in\id{initial}:\ \iota(e)=\id{state}(e)\ominus\mathcal I\\
		\text{in }\la\Events, \id{prec}, \delta, \iota\ra
		\end{array}
\end{array}
\label{req:shs2ihs}
\end{eqnarray}
	Vice versa, we can also map an incremental history structure to a stateful history structure:
\begin{eqnarray}
&&\begin{array}{l}
	\id{ihs2shs}(\la\Events, \id{prec}, \delta, \iota\ra)\\
	\ \ =\ \begin{array}[t]{@{}l}
		\text{Let }\forall(e_1,e_2)\in\id{prec}:\ \id{state}(e_2)=\id{state}(e_1)\oplus\delta(e_1,e_2)\\
		\text{Let }\id{initial}=\{e\in\Events|\forall e_1\in\Events: (e_1,e)\notin\id{prec}\}\\
		\text{Let }\forall e\in\id{initial}:\ \id{state}(e)=\mathcal I\oplus\iota(e)\\
		\text{in }\la\Events, \id{prec}, \id{state}\ra
		\end{array}
\end{array}
\label{req:ihs2shs}
\end{eqnarray}
	The\marginpar{claim} fact that $\id{shs2ihs}$ and $\id{ihs2shs}$ are each other's inverse proves that the choice to represent
	history by an IHS or SHS can be made on practical grounds.

	We must prove\marginpar{claim} that
\begin{equation}
	\id{ihs2shs}(\id{increment}_{P,e_1,e,\Delta}(\id{shs2ihs}(h)))\ =\ \id{occurs}_{P,e,\id{state}(e_1)\oplus\Delta}(h)
\end{equation}

\section{Isolation}
\label{sct:Isolation}
	How does a user perceive a distributed information system (DIS)?
	There are different options.
	An example is an information screen with flight arrival times in an airport,
	where a user sees the current state of flight arrivals and sees it change each time a new event occurs.
	This is an example of non-transactional use: the user consumes information passively.
	However, the topic of this paper is transactional use,
	an example of which is to make a flight reservation.
	In that example, the user interacts with the system for the purpose of reserving a seat on a flight.
	She sees a current state and provides events to the system by filling out fields in a reservation screen.
	These events cause changes in the state (e.g.\ from {\em The flight is not paid} to {\em The flight has been paid}).
	With this type of use, i.e transactional use, the user expects a predictable result.
	That is: if the current state is $s$ and the user specifies a change $\Delta$,
	the resulting state must be $s\oplus\Delta$.
	This requirement is the incremental variant of requirement~\ref{req:transaction}:
\begin{equation}
	h'=\id{occurByDelta}_{P,e_1,\Delta,e}(h)\ \Rightarrow\ \id{state}(e)\ =\ \id{state}(e_1)\oplus\Delta
\label{req:occurByDelta}
\end{equation}
	Since $\id{occurByDelta}$ preserves a stateful history structure (equation~\ref{req:recursive def SHS by Delta}),
	the user can interact repeatedly with the system and update the state predictably in a sequence of changes.
	This yields a user experience of being the only user of the system.
	That property is called {\em isolation}.

\section{Consistency}
\label{sct:Consistency}
	From a user perspective, {\em consistency} means that the system is free of contradictions.%
\footnote{A user might have wanted the system to provide the truth itself,
	but that is beyond what technical systems can guarantee. For people can feed the system with lies.
	However, freedom of contradictions is quite desirable as second-best, because truth is free of contradictions.}
	We require a predicate on states to determine whether a state is consistent.
	So we extend the notion of stateful historic structure with that predicate.

	A {\em nonconflicting history structure} is a quadruple $\la\Events, \id{prec}, \id{state}, \id{consistent}\ra$ in which
	$\la \Events, \id{prec}, \id{state}\ra$ is a SHS and $\id{consistent}$ is a predicate on $\mathbb S$,
	which holds for every state in the SHS.
	We will use predicate $\id{NHS}$ as a definition:
\begin{equation}
\begin{array}{l@{}ll}
	\multicolumn{3}{l}{\id{NHS}\la \Events, \id{prec}, \id{state}, \id{consistent}\ra}\\
	\ \Leftrightarrow\ &\id{SHS}\la \Events, \id{prec}, \id{state}\ra&\wedge\\
	&\id{consistent}\subseteq\mathbb S&\wedge\\
	&\forall e\in\Events:\ \id{consistent}(\id{state}(e))&
\end{array}
\label{def:NHS}
\end{equation}
	Equation~\ref{def:NHS} defines a nonconflicting history structure:
	For practical use, we require $\id{consistent}$ to be decidable.

\section{Atomicity}
\label{sct:Atomicity}
	{\em Atomicity} guarantees all-or-nothing outcomes.
	Let an example clarify this.
	If I purchase a song on the internet, I must pay.
	As a result, it is acceptable that the payment is done and the song is delivered.
	It is also acceptable that the payment is not done and the song is not delivered.
	However, it is unacceptable that the payment is done and the song is not delivered or vice-versa.

	Consider this:
	\[h'\ =\ \id{increment}_{P,e_2,e_3,\Delta_2}(\id{increment}_{P,e_1,e_2,\Delta_1}(h))\]
	Imagine that $\Delta_1$ represents my payment and $\Delta_2$ represents the delivery of my song.
	In the corresponding history structure, we will see that
	\[\{(e_1,e_2), (e_2,e_3)\}\subseteq\id{prec} \]
	It means that the overall state change caused by a sequence of events can be committed as a whole or rejected completely.

	So let us first discuss event sequences.
	Let us define an event sequence as a totally ordered history structure.
	Note that a history structure that is not totally ordered may contain multiple historic substructures
	that are totally ordered.
	Typically, event sequences occur multiply in a history structure.

\section{Durability}
\label{sct:Durability}
	In the eyes of the user, durability means that history is not changed.
	In the previous definitions durability has been secured by defining every change as an expansion of an existing system.
	Events are added only. Once added they are never removed or replaced.
	This corresponds to working practices in financial ledgers.
	Transactions are never removed, only compensated.

\section{Bibliography}
\bibliographystyle{elsarticle-harv}
\bibliography{doc}

\end{document}

\section{Threads}
	So how does a stateful history structure represent a distributed information system?
\marginpar{claim}
	Part of the answer is that a substructure of a stateful history structure is a stateful history structure by itself.

	The answer is that we have made no assumptions about location,
	like for instance that there is a shared memory.
	Even if we assume that each state exists in one location,
	a stateful history structure can still have multiple current states.
	Each of these states can exist in a separate location.

	The purpose of threads is to support deterministic computations.
	We consider any nonempty history structure that is totally ordered to be a thread.
	A {\em thread} is a history structure $\la \Events, \id{prec}\ra$ that satisfies
\begin{eqnarray}
	\forall e_1,e_2&:&e_1\in\Events\ \wedge\ e_2\in\Events\ \Rightarrow\ e_1\ \id{prec}+\ e_2\vee e_2\ \id{prec}^+\ e_1\label{eqn:order is total}\\
	\exists e&:&e\in\Events\label{eqn:nonempty thread}
\end{eqnarray}
	The total order (\ref{eqn:order is total}) is required to get deterministic results from a computation in a thread.
	In order to talk about the first and the last event in a thread, we require at least one event (\ref{eqn:nonempty thread}) in a thread.
	We shall use variable $t$ to represent threads.
	The first event in a thread $t$ is the infimum of the events in $t$;
	the last event is the supremum.

	We need the notion of state%
	\footnote{Cambridge Dictionary meaning of ``state'':
	a condition or way of being that exists at a particular time}
	to describe things that change over time.
	We shall use variables $s$, $s_1$, and $s_2$ to denote states.
	Let $\id{state}(e)$ represent the state after event $e$ has happened.

\section{Distributed Information System}

	Now equation~\ref{eqn:eval step} is the recursion step and the initial state is
	the starting point of the recursive computation of the final state of $t$.
	We consider a distributed information system as a set of processes that share a history.
	The system is distributed because processes need not share all histories.
	We need the notion of process%
	\footnote{Cambridge Dictionary meaning of ``process'':
		a series of actions that you take in order to achieve a result}
	to describe how components of a DIS interact.
	Let $\mathbb P$ be the set of all processes.
	We have formulated the following requirements:
	\begin{itemize}
	\item Belong/Contain\\
		To represent the fact that an event {\em belongs to} a process or
		(equivalently) that a process {\em contains} certain events,
		let $e\in p$ represent the statement that event $e$ belongs to process $p$.
	\item Shared history\\
		To abstract away from location, we make no assumptions with respect to the place where a process lives.
		However, to partake in a distributed system, processes must be able to share history.
		For that purpose we define what it means to be part of a history:
		\begin{equation}
			p\ \id{partof}\ h\ \Leftrightarrow\ (\forall e: e\in p\Rightarrow e\in h)
		\end{equation}
		We define an order between processes that share a history.
		\begin{equation}
			p_1\ \preceq\ p_2\ \Leftrightarrow(\forall e: e\in p_1\Rightarrow e\in p_2)
		\end{equation}
		We can prove that
		\begin{equation}
			p_1\ \preceq\ p_2\ \Rightarrow\ (\exists h:p_1\ \id{partof}\ h\wedge p_2\ \id{partof}\ h)
		\end{equation}
	\item Total order\\
		We require a process to have predictable behaviour, for which we need the process to contain a series (sequence) of events.
		So we require that $(\{e|e\in p\}, \preceq)$ is a total order.
		Apart from being a partial order, $\preceq$ must satisfy (for every $e_1$, $e_2$, and $p$):
		\begin{equation}
			e_1\in p\ \wedge\ e_2\in p\ \Rightarrow\ e_1\preceq e_2\vee e_2\preceq e_1
		\end{equation}
	\end{itemize}

\section{States}
	\begin{itemize}
	\item Revert\\
		A change can be used to compute a previous state.
		For this purpose we require an operator $\ominus$ such that
	\begin{equation}
		s_1\oplus\Delta=s_2\ \Leftrightarrow\ s_1=s_2\ominus\Delta
	\end{equation}
		As a consequence, we have for every event $e$:
	\begin{equation}
		\id{state}(\id{past}(e))=\id{state}(\id{pres}(e))\ominus\id{delta}(e)
	\end{equation}
		With $\preceq$ being reflexive, antisymmetric, and transitive,
		it follows immediately that $\oplus$ is idempotent, commutative, and associative.
	\end{itemize}

\section{Isolation}
\label{sct:Isolation}
{\em Isolation} means that actions can be done without any knowledge of what other users are doing simultaneously.

\section{Durability}
\label{sct:Durability}
{\em Durability} means that the contents of a database cannot change unintendedly.

\section{Bibliography}
\bibliographystyle{elsarticle-harv}
\bibliography{doc}

\end{document}

%% Obsolete

\section{Consistency}
\label{sct:Consistency}
Each user of a database issues a stream of events.
	The order (in time) of events is given by a relation\footnote{%
A relation $\declare{r}{A}{B}$ is a set of pairs. Each pair $\pair{a}{b}$ in that set satisfies $a\in A$ and $b\in B$}
	$\declare{pred}{Event}{Event}$,
	where $\pair{e1}{e2}\in\id{pred}$ means that event $e1$ precedes event $e2$.
	From a user perspective this stream is a sequence, so every event except the first has exactly one predecessor.

	For simplicity's sake, we assume the user sees a finite set of items at any given moment.
	The items seen by the user at any moment are given by a relation $\declare{state}{Event}{Item}$,
	where $\pair{e}{i}\in\id{state}$ means that item $i$ exists immediately after $e$ has happened.
	So the set of items visible to the user immediately after $e$ has happened can be computed by $\{ i | \pair{e}{i}\in\id{state} \}$.
	Note that the relation $\id{state}$ is interpreted in an open world.
	If $\pair{e}{i}\not\in\id{state}$ this does not mean $i$ does not exist.
	It just means that the pair $\pair{e}{i}$ is not in $\id{state}$, so the user will not see it.

	To cope with change, a user may insert and/or delete items at will.
	Let us specify these deletions and insertions by two relations, $\declare{ins}{Event}{Item}$ and $\declare{del}{Event}{Item}$,
	where $\pair{e}{i}\in\id{ins}$ means that pair $\pair{e}{i}$ is to be added to $\id{state}$,
	and $\pair{e}{i}\in\id{del}$ means that pair $\pair{e}{i}$ is to be removed from $\id{state}$.
	More precisely, let us consider an event $e$.
	The items $\{ i | \pair{e}{i}\in\id{state} \}$ are the items that exist immediately after the preceding event has happened,
	minus the items to be removed, plus the items to be inserted.
	The following equation defines this \define{behaviour}:
\begin{equation}
	\id{state}\ =\ (\id{pred};\id{state}-\id{del})\cup\id{ins}
\label{behaviour}
\end{equation}
	Equation~\ref{behaviour} says that the items visible to the user immediately after an event $e$
	are all items that are inserted together with items that were visible immediately after the preceding event, except those that were deleted.

	We define $A_{(\id{ins},\id{del})}$ as the function\footnote{The name $A$ is chosen as a reminder to the word ``action''.}
	that embodies this change:
\begin{equation}
	A_{(\id{ins},\id{del})}(x)\ =\ (x-\id{del})\cup\id{ins}
\end{equation}

	This raises the question whether we can compose two actions into a single action.
	For this, we derive:
\[\begin{array}{r@{~}l}
	&(((x-\id{del})\cup\id{ins})-\id{del}')\cup\id{ins}'\\
	=&\hspace{1in}\{x-y=x\cap\overline{y}\}\\
	&(((x\cap\overline{\id{del}})\cup\id{ins})\cap\overline{\id{del}'})\cup\id{ins}'\\
	=&\hspace{1in}\{\text{distribute $\cap$ over $\cup$}\}\\
	&(x\cap\overline{\id{del}}\cap\overline{\id{del}'})\cup(\id{ins}\cap\overline{\id{del}'})\cup\id{ins}'\\
	=&\hspace{1in}\{\text{De Morgan}\}\\
	&(x-(\id{del}\cup\id{del}'))\cup(\id{ins}-\id{del}')\cup\id{ins}'\\
	=&\hspace{1in}\{\text{Assume}\ \id{ins}\cap\id{del}=\emptyset\ \text{and}\ \id{ins}'\cap\id{del}'=\emptyset\}\\
	&(x-(\id{del}\cup\id{del}'))\cup(\id{ins}\cup\id{ins}')
\end{array}\]
	This yields composition ($;$) of $A$:
\begin{equation}
	A_{(\id{ins},\id{del})}\ ;\ A_{(\id{ins}',\id{del}')}\ =\ A_{((\id{ins}-\id{del}')\cup\id{ins}', \id{del}\cup\id{del}')}
\label{composition of A}
\end{equation}

	The composition operator turns out to be commutative, but under a condition.
	This will appear to be useful later on in this paper.
	Equation~\ref{composition of A} shows that we only need to investigate commutativity in the first argument of $A$: 
\[\begin{array}{r@{~}rcl}
	&A_{(\id{ins},\id{del})}\ ;\ A_{(\id{ins}',\id{del}')}&=&A_{(\id{ins}',\id{del}')}\ ;\ A_{(\id{ins},\id{del})}\\
	\Leftrightarrow\\
	&A_{((\id{ins}-\id{del}')\cup\id{ins}', \id{del}\cup\id{del}')}&=&A_{((\id{ins}'-\id{del})\cup\id{ins}, \id{del}'\cup\id{del})}\\
	\Leftrightarrow\\
	&(\id{ins}-\id{del}')\cup\id{ins}'&=&(\id{ins}'-\id{del})\cup\id{ins}\\
	\Leftrightarrow\\
	&(\id{ins}\cap\overline{\id{del}'})\cup\id{ins}'&=&(\id{ins}'\cap\overline{\id{del}})\cup\id{ins}\\
	\Leftrightarrow\\
	&(\id{ins}\cup\id{ins}')\cap(\overline{\id{del}'}\cup\id{ins}')&=&(\id{ins}\cup\id{ins}')\cap(\overline{\id{del}}\cup\id{ins})\\
	\Leftrightarrow\\
	&(\id{ins}\cup\id{ins}')-\overline{\overline{\id{del}'}\cup\id{ins}'}&=&(\id{ins}\cup\id{ins}')-\overline{\overline{\id{del}}\cup\id{ins}}\\
	\Leftrightarrow\\
	&\id{del}'\cap\overline{\id{ins}'}&=&\id{del}\cap\overline{\id{ins}}\\
	\Leftrightarrow\\
	&\id{del}'-\id{ins}'&=&\id{del}-\id{ins}
\end{array}\]
	For the purpose of making commutativity independent of its arguments,
	we must check the condition $\id{del}'-\id{ins}'=\id{del}-\id{ins}$.

\section{Isolation}
\label{sct:Isolation}
	If a user were alone,
	The moment a transaction starts, its user is isolated from other users.
	That moment is represented by a relation $\declare{beginTransaction}{Event}{Transaction}$,
	where $\pair{e}{t}\in\id{beginTransaction}$ means that event $e$ marks the start of transaction $t$.

\section{Ampersand}
\label{sct:Ampersand}
	In this section we explain the basics of Ampersand.
	The reader is expected to have sufficient background in relation algebra to understand the remainder of this paper.

	The core of an Ampersand script is a triple $\la\rules,\rels,\concepts\ra$,
	which consists of a set of rules $\rules$, relations $\rels$, and concepts $\concepts$.
	An Ampersand script is interpreted by the compiler as an information system.
	The rules constitute a theory in heterogeneous relation algebra.
	They constrain a body of data that resides in a database.

	The Ampersand compiler generates a database from relations in the script.
	A database application%
\footnote{Ampersand generates an application that consists of a relational database and interface components.
	Currently this application runs server-side on a PHP/MySQL platform and on a web-browser on the client-side.}
	assists users to keep rules satisfied throughout the lifetime of the database. It is also generated by Ampersand.

	A rule is a constraint in the form of an equality between two terms.
	Terms are built from relations.
	Ampersand interprets every relation as a finite set of pairs, which are stored in the database.
	The phase in which Ampersand takes a script and turns it into a database, is what we will refer to as \define{compile-time}.
	The phase in which a user interacts with the database, is what we will refer to as \define{run-time}.
	At run-time, Ampersand can decide which rules are satisfied by querying the database.
	The compiler generates all software needed to keep rules satisfied at run-time.
	If one of the rules is not satisfied as a result of data that has changed,
	that change is reverted (rolled back) to the original state.
	Either way, Ampersand ensures that all rules remain satisfied.

	Atoms are values that have no internal structure, meant to represent data elements in a database.
	From a business perspective, atoms are used to represent concrete items of the world,
	such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
	By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ are used to represent \emph{atoms}.
	The set of all atoms is called $\atoms$.
        Each atom is an instance of a \emph{concept}.

	Concepts (from set $\concepts$) are names we use to classify atoms in a meaningful way.
	For example, you might choose to classify \atom{Peter} as a person, and \atom{074238991} as a telephone number.
        We will use variables $A$, $B$, $C$, $D$ to represent concepts.
	The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.
	The expression $a \in A$ means that atom $a$ is an \emph{instance} of concept $A$.
	The declaration of $A\isa B$ (pronounce: $A$ is a $B$)
	in an Ampersand script states that any instance of $A$ is an instance of $B$ as well.
	We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
	Specialization is needed to allow statements such as: ``An orange is a fruit that ....''.
	Specialization is not used in the remainder of this paper.

	Relations (from set $\rels$) are used in information systems to store data.
%	A \define{fact} is a statement that is true in a business context.
%	Facts are stored and kept as data in a computer.
	As data changes over time, so do the contents of these relations.
	In this paper relations are represented by variables $r$, $s$, and $d$.
	We represent the declaration of a relation $r$ by $\declare{nm}{A}{B}$,
	in which \id{nm} is a name and $A$ and $B$ are concepts.
	We call $A$ the source concept and $B$ the target concept of the relation.
	The pair $\pair{A}{B}$ is called the \emph{type} of the relation.
	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.

	The meaning of relations in Ampersand is defined by an interpretation function $\mathfrak{I}$.
	It maps each relation to a set of pairs.
	The type system of Ampersand guarantees that each pair in $r$ respects the type of $r$:
\begin{equation}
	\pair{a}{b}\in\ti{\declare{nm}{A}{B}} \Rightarrow\ a \in A \wedge b \in B \label{typing of relations}
\end{equation}
	The type system is strong, which means that every atom has a type.
	The type system is static, which means that type checking is entirely done at compile-time.

	Terms are used to combine relations using operators.
	The set of terms is called $\terms$.
\begin{definition}[terms]
\label{def:terms}
\item   The set of terms, $\terms$, is the smallest set that satisfies, for all $r,s \in \terms$, $d\in\rels$ and $A,B \in \concepts$: 
\begin{eqnarray}
	d&\in&\terms         \quad\quad\text{(every relation is a term)}\\
	(r \cup s)&\in&\terms\quad\quad\text{(union)}\\
	(r \cap s)&\in&\terms\quad\quad\text{(intersection)}\\
	(r-s)&\in&\terms     \quad\quad\text{(difference)}\label{def:difference}\\
	(r;s)&\in&\terms     \quad\quad\text{(composition)}\\
	(r\backslash s)&\in&\terms     \quad\quad\text{(right residual)}\\
	(r\slash s)&\in&\terms     \quad\quad\text{(left residual)}\\
	\flip r&\in&\terms   \quad\quad\text{(converse)}\\
	\cmpl{r}&\in&\terms   \quad\quad\text{(complement)}\\
%	\kleeneplus r&\in&\terms   \quad\quad\text{(Kleene closure)}\\
%	\kleenestar r&\in&\terms   \quad\quad\text{(Kleene closure)}\\
	\ident{A}&\in&\terms \quad\quad\text{(identity)}\\
	\fullt{A}{B}&\in&\terms \quad\quad\text{(full set)}
\end{eqnarray}
\end{definition}
	Throughout the remainder of this paper,	terms are represented by variables $r$, $s$, $d$, and $t$.
	The \define{type} of a term $r$ is a pair of concepts, given by $\tf{r}$.
	$\mathfrak{T}$ is a partial function that maps terms to types.
	If a term has one type, it is called \define{type correct}.
	The Ampersand compiler requires all terms to be type correct.
	If no type can be assigned, the compiler gives an error message.
	If multiple types are assignable to a term, the ambiguity is signalled by the compiler.
	In that case it will also emit an error message and prompt the programmer to disambiguate the term by adding type information.
	In practice, the derivation of types can disambiguate most terms.
	So, the programmer experiences some freedom to denote distinct relations by the same name without the obligation to specify their types at all occurrences.
	The compiler does not generate any code if there is a term that is not type correct.
	The type function and its restrictions are discussed in~\cite{Joosten2015}
	and have no consequences for the remainder of this paper.

 	The meaning of terms in Ampersand is given by an interpretation function $\mathfrak{I}$.
	Let $A$ and $B$ be finite sets of atoms, then $\mathfrak{I}$ maps each term to the set of pairs for which that term stands.
\begin{definition}[interpretation of terms]
\label{interpretation of terms}
\item   For every $A,B\in\concepts$ and $r,s\in\terms$
\begin{eqnarray}
	\ti{r}		 &=&\{\pair{a}{b}|\ a\ r\ b\}	\\
	\ti{r \cup s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{or }\ \pair{a}{b}\in\ti{s}\}	\\
	\ti{r \cap s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{a}{b}\in\ti{s}\}	\\
	\ti{r-s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{a}{b}\notin\ti{s}\}	\\
	\ti{r;s}	 &=&\{\pair{a}{c}|\ \text{for some}\ b,\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{b}{c}\in\ti{s}\}	\\
	\ti{r\backslash s}	 &=&\{\pair{b}{c}|\ \text{for all}\ a,\ \pair{a}{b}\in\ti{r}\ \text{implies}\ \pair{a}{c}\in\ti{s}\}	\\
	\ti{r\slash s}	 &=&\{\pair{a}{b}|\ \text{for all}\ c,\ \pair{b}{c}\in\ti{s}\ \text{implies}\ \pair{a}{c}\in\ti{r}\}	\\
	\ti{\flip{r}}	 &=&\{\pair{b}{a}|\ \pair{a}{b}\in\ti{r}    \}	\\
	\ti{\cmpl{\declare{r}{A}{B}}}	 &=&\fullt{A}{B}-r	\\
%	\ti{\kleeneplus{r}}	 &=&\text{the smallest set that satisfies }\kleeneplus{r}=r\ \cup\ \kleeneplus{r};\kleeneplus{r}	\\
%	\ti{\kleenestar{\declare{r}{A}{A}}}	 &=&\ident{A}\cup\kleeneplus{r}\\
	\ti{\ident{A}} 	 &=&\{\pair{a}{a}|\ a\in A\}	\\
	\ti{\fullt{A}{B}}&=&\{\pair{a}{b}|\ a\in A, b\in B\}
\end{eqnarray}
\end{definition}
%	The Kleene closure operators (postfix $\kleeneplus{\ }$ and $\kleenestar{\ }$) have been implemented partially in the current Ampersand implementation.

	A \define{rule} is a pair of terms $r,s\in\terms$ with $\tf{r}=\tf{s}$, which is syntactically recognizable as a rule.
\[\text{RULE}\ r = s\]
	This means \(\ti{r} = \ti{s}\). In practice, many rules are written as:
\[\text{RULE}\ r\subs s\]
	This is a shorthand for 
\[\text{RULE}\ r\cap s = r\]
	We have enhanced the type function $\mathfrak{T}$ and the interpretation function $\mathfrak{I}$ to cover rules as well.
	If $\tf{r}=\tf{s}$ and $\tf{s}=\pair{A}{B}$:
\begin{eqnarray}
	\tf{\text{RULE}\ r = s}   &=&\pair{A}{B}\\
	\tf{\text{RULE}\ r\subs s}&=&\pair{A}{B}\\
 resolved this \end{eqnarray}

	We call a rule $\text{RULE}\ r = s$ \define{satisfied} when its interpretation equals $\ti{\fullt{A}{B}}$.
	As the population of relations used in $r$ changes with time, the satisfaction of the rule changes accordingly.
	A software developer, who conceives these rules, must consider how to keep each one of them satisfied.
	We call a rule \define{violated} if it is not satisfied.
	The set $\ti{(s-r)\cup(r-s)}$ is called the \emph{violation set} of \(\text{RULE}\ r = s\).
	To \define{resolve} violations means to change the contents of relations such that the rule is satisfied%
\footnote{To \define{restore invariance} is sometimes used as a synonym to resolving violations.
	To \define{preserve invariance} is sometimes used as a synonym to keeping a rule satisfied.}.
	Consequently, such a rule may also be referred to as \define{invariant}.
	Some ways of resolving violations are presented in section~\ref{sct:Code Fragments}.
	Each pair in the violation set of a rule is called a violation of that rule.

	The software developer must define how to resolve violations when they occur.
	She does so by inserting and/or deleting pairs in appropriately chosen relations.
	Whatever choice she makes, she must ensure that her code yields data that satisfies the rules.
	When we say: ``rule $r$ specifies this action'' we mean that satisfaction of rule $r$ is a postcondition of any action specified by rule $r$.

\section{Software for Legal Reasoning}
\label{sct:Conceptual analysis}

% \begin{figure}[htb]
% \begin{center}
%   \includegraphics[angle=90,scale=.357]{LogicalDataModel.pdf}
% \end{center}
% \caption{Conceptual data model (from section~\ref{sct:Conceptual analysis})}
% \label{fig:conceptual model}
% \end{figure}

\section{Bibliography}
\bibliographystyle{elsarticle-harv}
\bibliography{doc}

\end{document}

### 2019-06-12 2018_Transactions/articleTransactions.tex
Simplified the definition of history to one relation, prec.

@@
 
+\section{History}
+\label{sct:History}
+	A treatment of time lays the foundation of our theory of distributed information systems (DIS).
+	Because of the distributed nature of a DIS,
+	we may not know in which order two events\footnote{Cambridge Dictionary defines ``event'' as: anything that happens} have happened.
+	For this reason we describe the history of a DIS as a partially ordered relation of events.
+
+	We use variables $e$, $e_1$, and $e_2$ to represent events.
+	Let $\mathbb E$ represent the set of all conceivable events.
+%	We also introduce the notion of {\em history structure}
+%	to represent all events that have happened up to some point in time and the precedences between those events.
+%	In this paper, variables $h$, $h'$, $h_1$, and $h_2$ represent history structures.
+
+	We represent the ``history of a DIS up to some point in time'' by a relation $\id{prec}$ (pronounce: {\em precedes}),
+	being a set of pairs from $\mathbb E$.
+\[\id{prec}\subseteq\mathbb E\times\mathbb E\]
+	Let $e_1\ \id{prec}\ e_2$ represent the fact that event $e_1$ has happened before event $e_2$,
+	making $\id{prec}$ a relation on events in $\mathbb E$.
+	We require:
+\begin{eqnarray}
+	&&(e,e)\notin\id{prec}\label{req:prec irreflexive}\\
+	&&(e_1,e_2)\in\id{prec}^+\ \wedge\ (e_2,e_1)\in\id{prec}^+\ \Rightarrow\ e_1=e_2\label{req:precPlus asy}
+\end{eqnarray}
+	The events in the DIS, of which $\id{prec}$ represents the history, are 
+\begin{eqnarray}
+	&&\id{events}(\id{prec})\ =\ \{e_1|(e_1,e_2)\in\id{prec}\}\cup\{e_2|(e_1,e_2)\in\id{prec}\}\label{req:events}
+\end{eqnarray}
+	Requirement~\ref{req:prec irreflexive} excludes the situation that $e$ has happened before itself.
+	To represent time faithfully, equation~\ref{req:precPlus asy} requires that the transitive closure $\id{prec}^+$ is antisymmetric.
+	This makes $\id{prec}^*$ a reflexive, transitive, and antisymmetric relation, which lets us use it as a temporal order between events.
+	So we shall call $\id{prec}$ "a history" because it represents all events that have happened up to a certain point in time and the (partial) order between them.
+	Let $\mathbb H$ be the set of all conceivable histories, meaning that every $\id{prec}\in\mathbb H$.
+
+	In a history $\id{prec}$, an event $e$ that precedes another event is called a {\em past} event,
+	which we define as \ $\exists e_1:\ e\ \id{prec}\ e_1$.
+	All other events are called {\em current} events in $\id{prec}$.
+	Function $\id{current}$ computes the current events in history $\id{prec}$.
+	Function $\id{initial}$ computes the initial events, which are the events without preceding events.
+\begin{eqnarray}
+	\id{current}(\id{prec})&=&\{e |\ \exists e_1: (e_1,e)\in\id{prec}\ \wedge\ \neg\exists e_2: (e,e_2)\in\id{prec}\}\\
+	\id{initial}(\id{prec})&=&\{e |\ \exists e_1: (e,e_1)\in\id{prec}\ \wedge\ \neg\exists e_2: (e_2,e)\in\id{prec}\}
+\end{eqnarray}
+
+	The special case in which $\id{prec}^*$ is a total order is called a {\em sequential history}.
+	Note that any nonempty finite sequential history has one current event and one initial event.
+	Other finite histories may have more.
+	In a sequential history structure, every event except the initial one has precisely one preceding event.
+	In general, however, events may have multiple preceding events and multiple succeeding events.
+
+	We define $\leq$ to represent the fact that history $\id{prec}$ has preceded history $\id{prec}'$ in time.
+	It is defined by:
+\begin{equation}
+\label{def:lies in the past}
+	\id{prec}\leq\id{prec}'\ \Leftrightarrow\ \id{prec}^*\subseteq\id{prec}'^*
+\end{equation}
+
 \section{Traces}
 \label{sct:Traces}
-	We need the notion of {\em event}\footnote{Cambridge Dictionary defines ``event'' as: anything that happens} to describe things that happen in a distributed information system.
-	Being distributed means that we must assume that the temporal order of events is partial:
-	We do not necessarily know in which order two events have happened.
-	To describe this, we define traces algebraically corresponding to Foster~\cite{Foster17b}.
+	A transaction in a DIS may be perceived as a distributed action,
+	which leaves a trace of events in the history of a DIS.
+	Let $\mathbb T$ be the set of all conceivable traces.
+	With \id{prec} being a history, we define that every pair $(e_1,e_2)\in\id{prec}^*$ is a trace.
+	So the traces of history $\id{prec}$ are denoted ${\mathcal T}_{\id{prec}}$ with ${\mathcal T}_{\id{prec}}\subseteq\mathbb T$.
+\begin{definition}[Traces]
+\label{def:Traces}
+\item   The set of traces of history $\text{prec}$ is defined by ${\mathcal T}_{\text{prec}}\ =\ \text{prec}^*$.
+\end{definition}
+	Pair $(e,e)$ represents the empty trace for any $e\in\id{events}(\id{prec})$ because $(e,e)\in\id{prec}^*$.
+	We will denote the empty trace as $0$ and define $(e_1,e_1)=0=(e_2,e_2)$.
+	If $(e_1,e)\in\id{prec}^*$ and $(e,e_2)\in\id{prec}'^*$ then
+	we define the concatenation operator $+:\mathbb T\rightarrow\mathbb T\rightarrow\mathbb T$ by
+\begin{definition}[Concatenation of traces]
+\label{def:concatenation}
+\item   $(e_1,e)+(e,e_2)=(e_1,e_2)$
+\end{definition}
+	Note that the definition of $+$ is partial, because $(e_1,e_2)+(e_3,e_4)$ is not defined unless $e_2=e_3$.
+	Also, $(e_1,e)+(e,e_2)$ is defined in the combined history $\id{prec}^*\cup\id{prec}'^*$
+	because $(e_1,e_2)\in(\id{prec}\cup\id{prec}')^*$.
 
-	We define traces as a set ${\mathcal T}$ with
-	a concatenation operator $+:\mathcal T\rightarrow\mathcal T\rightarrow\mathcal T$ and
-	the empty trace $0:\mathcal T$, which obey the following axioms.
-\begin{definition}[Trace Algebra]
-\label{def:Trace Algebra}
-\item   A trace algebra $(\mathcal T, +, 0)$, is an algebra that satisfies: 
+	This definition of trace is an interpretation of trace algebra~\cite{Foster17b}, which describes discrete-event systems.
+	Other interpretations of this algebra comprise real-time systems and hybrid systems, but that is beyond the scope of this paper.
+
+	The concatenation operator $+$ and empty trace $0$ satisfy the axioms of trace algebras: (proof needed)
 \begin{eqnarray}
@@
 	0+x&=&x+0\label{axi:lr-identity}\\
-	x+y\ =\ x+z&\Rightarrow&y=z\label{axi:right cancellation}\\
-	x+z\ =\ y+z&\Rightarrow&x=y\label{axi:left cancellation}\\
+	x+y\ =\ x+z&\Rightarrow&y=z\label{axi:left cancellation}\\
+	x+z\ =\ y+z&\Rightarrow&x=y\label{axi:right cancellation}\\
 	x+z\ =\ 0&\Rightarrow&x=0\label{axi:no inverse}
 \end{eqnarray}
-\end{definition}
 	In terms of universal algebra, equations~\ref{axi:associative} and~\ref{axi:lr-identity} state that a trace algebra is a {\em monoid}.
-	Equations~\ref{axi:right cancellation} and~\ref{axi:left cancellation} state that $+$ is {\em cancellative}.
+	Equations~\ref{axi:left cancellation} and~\ref{axi:right cancellation} state that $+$ is {\em cancellative}.
 	Equation~\ref{axi:no inverse} makes the cancellative monoid into a trace algebra.
@@
 
-	In terms of information systems, a trace represents history.
-	For that purpose it makes sense to define the notion of prefix, so we can define notions like shared history.
+	It makes sense to define the notion of prefix, so we can define notions like shared history.
 \begin{definition}[Prefix]
@@
 \label{def:Subtract}
-\label{def:Subtract}
 \begin{array}[t]{rcll}
@@
 \end{definition}
-
-\section{History}
-\label{sct:History}
-	The history of a distributed information system (DIS) consists of events and an ordering on these events.
-	Because a DIS is distributed,
-	we may not know the temporal order of two events in the history structure.
-	For this reason we describe the history structure of a DIS as a partial order of events.
-
-	We use variables $e$, $e_1$, and $e_2$ to represent events.
-	Let $\mathbb E$ represent the set of all conceivable events.
-	We also introduce the notion of {\em history structure}
-	to represent all events that have happened up to some point in time and the precedences between those events.
-	In this paper, variables $h$, $h'$, $h_1$, and $h_2$ represent history structures.
-	Let $\mathbb H$ be the set of all conceivable history structures.
-
-	A {\em history structure} is a tuple $\la \Events, \id{prec}\ra$ in which
-\begin{eqnarray}
-	&&\Events\ \subseteq\ \mathbb E\\
-	&&\id{prec}\ \subseteq\ \Events\times\Events\label{req:precInE}\\
-	&&(e,e)\notin\id{prec}\label{req:prec irreflexive}\\
-	&&(e_1,e_2)\in\id{prec}^+\ \wedge\ (e_2,e_1)\in\id{prec}^+\ \Rightarrow\ e_1=e_2\label{req:precPlus asy}
-\end{eqnarray}
-	In a history structure $\la \Events, \id{prec}\ra$ set $\Events$ represents the events that have been registered up to some point in time.
-	For every event $e\in\Events$ we can say that it has happened (at some time in the past).
-	Let $e_1\ \id{prec}\ e_2$ represent the fact that event $e_1$ has happened before event $e_2$,
-	making $\id{prec}$ a relation on events in $\Events$.
-	Requirement~\ref{req:prec irreflexive} excludes the situation that $e$ has happened before itself.
-	To represent time faithfully, we require that $\la\Events,\id{prec}^*\ra$ is a partially ordered set,
-	meaning that $\id{prec}^*$ is reflexive, transitive, and antisymmetric.
-	For this reason we have requirement~\ref{req:precPlus asy}, which says that the transitive closure $\id{prec}^+$ must be antisymmetric.
-
-	In a history structure $\la \Events, \id{prec}\ra$, an event $e$ that precedes another event is called a {\em past} event,
-	which we define as \ $\exists e_1\in\Events: e\ \id{prec}\ e_1$.
-	All other events are called {\em current} events in that history structure.
-	Function $\id{current}:\mathbb H\rightarrow\powerset{\Events}$ computes the current events in a history structure.
-	Events in $\Events$ that have no preceding events are called {\em initial} events.
-	Function $\id{past}$ computes the past events and $\id{initial}$ computes the initial events.
-\begin{eqnarray}
-	\id{current}(\la \Events, \id{prec}\ra)&=&\{ c\in\Events|\ \neg\exists e_1\in\Events: e\ \id{prec}\ e_1 \}\\
-	\id{initial}(\la \Events, \id{prec}\ra)&=&\{ c\in\Events|\ \neg\exists e_1\in\Events: e_1\ \id{prec}\ e \}
-\end{eqnarray}
-	We define the {\em closure} of a history structure as the history structure whose order is the reflexive, transitive closure of the original ordering relation.
-\begin{eqnarray}
-	\la \Events, \id{prec}\ra^*&=&\la \Events, \id{prec}^*\ra\label{def:closure}
-\end{eqnarray}
-	Equation~\ref{def:closure} allows us to write $h^*$ for elements of $\mathbb H$.
-	The order of $h^*$ is reflexive, transitive, and antisymmetric.
-
-	The special case in which $\la\Events,\id{prec}^*\ra$ is a total order is called a {\em sequential history structure}.
-	Note that any nonempty finite sequential history structure has one current event and one first event.
-	Other finite history structures may have more.
-	In a sequential history structure, every event except the first one has precisely one preceding event.
-	In general, however, events may have multiple preceding events and multiple succeeding events.
-
-	We extend the notion of subset to historic structures and use the subset symbol $\subseteq$ for that purpose.
-	We use $h\subseteq h'$ to state that $h$ lies in the past of $h'$.
-	Let $h=\la \Events, \id{prec}\ra$ and $h'=\la \Events', \id{prec}'\ra$ in the definition of $h\subseteq h'$:
-\begin{equation}
-\label{def:lies in the past}
-	h\subseteq h'\ \Leftrightarrow\ 
-	\Events\subseteq\Events'\ \wedge\ \id{initial}(h)\subseteq\id{initial}(h')\ \wedge\ \id{prec}^*\subseteq\id{prec}'^*
-\end{equation}
 

### 2019-06-13 2018_Transactions/articleTransactions.tex
improved the definition of isolation

@@
 
-	We define $\leq$ to represent the fact that history $\id{prec}$ has preceded history $\id{prec}'$ in time.
-	It is defined by:
-\begin{equation}
+	We define $\leq$ to represent the fact that history $\id{prec}$ lies in the past of history $\id{prec}'$.
+\begin{definition}[$\leq$]
 \label{def:lies in the past}
-	\id{prec}\leq\id{prec}'\ \Leftrightarrow\ \id{prec}^*\subseteq\id{prec}'^*
-\end{equation}
+\begin{displaymath}
+	\id{prec}\leq\id{prec}'\ \Leftrightarrow\ \id{prec}^*\subseteq\id{prec}'^*\ \wedge\ \id{initial}(\id{prec})\subseteq\id{initial}(\id{prec}')
+\end{displaymath}
+\end{definition}
 
@@
 	Let $\mathbb T$ be the set of all conceivable traces.
-	With \id{prec} being a history, we define that every pair $(e_1,e_2)\in\id{prec}^*$ is a trace.
-	So the traces of history $\id{prec}$ are denoted ${\mathcal T}_{\id{prec}}$ with ${\mathcal T}_{\id{prec}}\subseteq\mathbb T$.
+	With \id{prec} being a history, we define a trace $(e_1,e_2)_{\id{prec}}$ as follows:
 \begin{definition}[Traces]
 \label{def:Traces}
-\item   The set of traces of history $\text{prec}$ is defined by ${\mathcal T}_{\text{prec}}\ =\ \text{prec}^*$.
+\item   $(e_1,e_2)_{\id{prec}}\in\mathbb T\ \Leftrightarrow\ (e_1,e_2)\in\text{prec}^*$
 \end{definition}
-	Pair $(e,e)$ represents the empty trace for any $e\in\id{events}(\id{prec})$ because $(e,e)\in\id{prec}^*$.
-	We will denote the empty trace as $0$ and define $(e_1,e_1)=0=(e_2,e_2)$.
-	If $(e_1,e)\in\id{prec}^*$ and $(e,e_2)\in\id{prec}'^*$ then
+	We use variables $x$, $y$, and $z$ to represent traces.
+	We 
+\begin{definition}[Empty trace]
+\label{def:Empty trace}
+\item   $\forall e: 0_{\id{prec}}\ =\ (e,e)_{\id{prec}}$
+\end{definition}
+	We denote the empty trace as $0$ and define $(e_1,e_1)=0=(e_2,e_2)$.
+	If $(e_1,e)_{\id{prec}}$ and $(e,e_2)_{\id{prec}'}$ are traces then
 	we define the concatenation operator $+:\mathbb T\rightarrow\mathbb T\rightarrow\mathbb T$ by
@@
 \label{def:concatenation}
-\item   $(e_1,e)+(e,e_2)=(e_1,e_2)$
+\item   $(e_1,e)_{\id{prec}}+(e,e_2)_{\id{prec}'}=(e_1,e_2)_{\id{prec}\cup\id{prec}'}$
 \end{definition}
-	Note that the definition of $+$ is partial, because $(e_1,e_2)+(e_3,e_4)$ is not defined unless $e_2=e_3$.
-	Also, $(e_1,e)+(e,e_2)$ is defined in the combined history $\id{prec}^*\cup\id{prec}'^*$
-	because $(e_1,e_2)\in(\id{prec}\cup\id{prec}')^*$.
+	Note that the definition of $+$ is partial, because $(e_1,e_2)_{\id{prec}}+(e_3,e_4)_{\id{prec}'}$ is not defined unless $e_2=e_3$.
+	Also, the result is defined in the combined history $\id{prec}^*\cup\id{prec}'^*$
+	because $(e_1,e_2)\in(\id{prec}\cup\id{prec}')^*$. (proof needed)
 
@@
 
-	It makes sense to define the notion of prefix, so we can define notions like shared history.
-\begin{definition}[Prefix]
-\label{def:Prefix}
+	The operator $\leq$ prefix, so we can define notions like shared history.
+$
+(e_1,e_2)\leq(e_3,e_4)
+\Leftrightarrow
+(\exists\id{prec}:(e_1,e_2)\in\id{prec}^*)\ \wedge\ (\exists\id{prec}':(e_3,e_4)\in\id{prec}'^*)\ \wedge\ \id{prec}\leq\id{prec}'
+$
 \begin{eqnarray}
-	x\leq y&\Rightarrow&\exists z:\ x+z=y
+	x\leq y&\Rightarrow&\exists z:\ x+z=y\label{def:Prefix}
 \end{eqnarray}
-\end{definition}
 	We also define subtraction of traces as a way to "cut off" a prefix from a trace.
@@
 
-	Isolation is meant to express that the effect of an action depends only on its current state.
+	An action $\id{act}_\id{prec}(e_1,e_2)$ is performed in isolation if it does not affect any other action in the system.
+	This means that the only event in trace $(e_1,e_2)_\id{prec}$ that precedes events outside that trace is e_1.
+	Likewise, the only event in trace $(e_1,e_2)_\id{prec}$ that is preceded by events outside that trace is e_2.
+	To define this, we need an auxiliary function
+	$\id{atomize}:\mathbb E\times\mathbb E\times\mathbb H\rightarrow\mathbb H$,
+	which replaces a trace $(e_1,e_2)_\id{prec}$ by the pair $(e_1,e_2)$
+	in a history $\id{prec}$.
+\begin{eqnarray}
+	\begin{array}[t]{r@{}l@{}l}
+	\multicolumn{3}{l}{\id{atomize}(e_1,e_2,\id{prec})}\\
+	   \ =\ \id{prec}-&(&\{ (e_1,e)\in\id{prec}| e_1\ \id{prec}^*\ e\ \wedge\ e\ \id{prec}^*\ e_2\}\ \cup\\
+			  & &\{ (e,e_2)\in\id{prec}| e_1\ \id{prec}^*\ e\ \wedge\ e\ \id{prec}^*\ e_2\}\\
+			  &)\\
+		\cup&&\{ (e_1,e_2) \}\\
+	\end{array}
+\end{eqnarray}
+	We can now define isolation as the property that the traces of the "atomized" history
+	are the same as the atomization of all traces in that history.
 \begin{definition}[Isolation]
-\item An action $\id{act}:\mathbb H\rightarrow\mathbb H$ works in isolation iff there is a function $f:\mathbb S\rightarrow\mathbb S$ such that:
+\item An action $\id{act}_{\id{prec}}(e_1,e_2)$ works in isolation iff:
 \label{def:Isolation}
 \begin{eqnarray}
-	(h',e')=\id{act}(h,e)&\Rightarrow&\id{stat}(h',e')=f(\id{stat}(h,e))
+	(\id{atomize}(e_1,e_2,\id{prec}))^*&=&\id{atomize}(e_1,e_2,\id{prec}^*)
 \end{eqnarray}
@@
 	Atomicity expresses that a sequence of multiple actions can be treated as one.
-\begin{definition}[Atomicity]
-\item An action $\id{act}:\mathbb H\rightarrow\mathbb H$ is atomic iff its constituent actions ...
+
+\begin{definition}[Atomic]
+\item An action $\id{act}_{\id{prec}}(e_1,e_2)$ is atomic iff
 \label{def:Atomicity}
-\begin{eqnarray}
-	(h',e')=\id{act}(h,e)&\Rightarrow&e\ \id{pred}^*\ e'
-\end{eqnarray}
 \end{definition}

### 2019-06-14 2018_Transactions/articleTransactions.tex
Redone the storyline. Defined "action" and "system".

@@
 \newtheorem{lemma}{Lemma}
-\def\id#1{\text{\em #1\/}}
+\def\id#1{\text{\it #1\/}}
 \def\Events{{\mathit E}}
@@
 	This makes $\id{prec}^*$ a reflexive, transitive, and antisymmetric relation, which lets us use it as a temporal order between events.
-	So we shall call $\id{prec}$ "a history" because it represents all events that have happened up to a certain point in time and the (partial) order between them.
+	We call $\id{prec}$ ``a history'' because it represents all events that have happened up to a certain point in time and the (partial) order between them.
 	Let $\mathbb H$ be the set of all conceivable histories, meaning that every $\id{prec}\in\mathbb H$.
@@
 
-	The special case in which $\id{prec}^*$ is a total order is called a {\em sequential history}.
+	A history $\id{prec}$ where $\id{prec}^*$ is a total order is a special case, which we call a {\em sequential history}.
 	Note that any nonempty finite sequential history has one current event and one initial event.
@@
 
-	We define $\leq$ to represent the fact that history $\id{prec}$ lies in the past of history $\id{prec}'$.
+	We define $\leq$ (called ``prefix'') to represent the fact that history $\id{prec}$ lies in the past of history $\id{prec}'$.
 \begin{definition}[$\leq$]
@@
 \label{sct:Traces}
-	A transaction in a DIS may be perceived as a distributed action,
-	which leaves a trace of events in the history of a DIS.
+	Transactions leave traces in the history of a DIS, so let us define traces.
+	For this purpose we have chosen definitions that satisfy trace algebra
+	and represent traces in discrete-event systems.
+	Other interpretations of the same trace algebra comprise real-time systems and hybrid systems,
+	which is proven by Foster~\cite{Foster17b}.
+	We prove (TBD) that our definitions comply with Foster's axioms of trace algebra,
+	but stick to one interpretation (discrete event systems).
+
 	Let $\mathbb T$ be the set of all conceivable traces.
-	With \id{prec} being a history, we define a trace $(e_1,e_2)_{\id{prec}}$ as follows:
+	With \id{prec} being a history, we define a trace $(e_1,e_2)_\id{prec}$ as follows:
 \begin{definition}[Traces]
 \label{def:Traces}
-\item   $(e_1,e_2)_{\id{prec}}\in\mathbb T\ \Leftrightarrow\ (e_1,e_2)\in\text{prec}^*$
+\item   $(e_1,e_2)_\id{prec}\in\mathbb T\ \Leftrightarrow\ (e_1,e_2)\in\id{prec}^*$
 \end{definition}
-	We use variables $x$, $y$, and $z$ to represent traces.
-	We 
-\begin{definition}[Empty trace]
-\label{def:Empty trace}
-\item   $\forall e: 0_{\id{prec}}\ =\ (e,e)_{\id{prec}}$
-\end{definition}
-	We denote the empty trace as $0$ and define $(e_1,e_1)=0=(e_2,e_2)$.
-	If $(e_1,e)_{\id{prec}}$ and $(e,e_2)_{\id{prec}'}$ are traces then
-	we define the concatenation operator $+:\mathbb T\rightarrow\mathbb T\rightarrow\mathbb T$ by
-\begin{definition}[Concatenation of traces]
-\label{def:concatenation}
-\item   $(e_1,e)_{\id{prec}}+(e,e_2)_{\id{prec}'}=(e_1,e_2)_{\id{prec}\cup\id{prec}'}$
-\end{definition}
-	Note that the definition of $+$ is partial, because $(e_1,e_2)_{\id{prec}}+(e_3,e_4)_{\id{prec}'}$ is not defined unless $e_2=e_3$.
-	Also, the result is defined in the combined history $\id{prec}^*\cup\id{prec}'^*$
-	because $(e_1,e_2)\in(\id{prec}\cup\id{prec}')^*$. (proof needed)
-
-	This definition of trace is an interpretation of trace algebra~\cite{Foster17b}, which describes discrete-event systems.
-	Other interpretations of this algebra comprise real-time systems and hybrid systems, but that is beyond the scope of this paper.
-
-	The concatenation operator $+$ and empty trace $0$ satisfy the axioms of trace algebras: (proof needed)
+	If it is obvious from the context which history ($\id{prec}$) is meant, we may write $(e_1,e_2)$ rather than $(e_1,e_2)_\id{prec}$.
+	A trace algebra is a triple $\la\mathbb T,+,0\ra$,
+	where the operator $+$ (concatenation) and empty trace $0$ satisfy the axioms of trace algebras:
 \begin{eqnarray}
@@
 \end{eqnarray}
-	In terms of universal algebra, equations~\ref{axi:associative} and~\ref{axi:lr-identity} state that a trace algebra is a {\em monoid}.
+	In terms of universal algebra,
+	equations~\ref{axi:associative} and~\ref{axi:lr-identity} state that a trace algebra is a {\em monoid}.
+	Note that $+$ is {\em not} commutative.
 	Equations~\ref{axi:left cancellation} and~\ref{axi:right cancellation} state that $+$ is {\em cancellative}.
-	Equation~\ref{axi:no inverse} makes the cancellative monoid into a trace algebra.
-	It implies that there is no inverse.
+	Equation~\ref{axi:no inverse} implies that there is no inverse.
+	It makes the cancellative monoid into a trace algebra.
 
-	The operator $\leq$ prefix, so we can define notions like shared history.
-$
-(e_1,e_2)\leq(e_3,e_4)
-\Leftrightarrow
-(\exists\id{prec}:(e_1,e_2)\in\id{prec}^*)\ \wedge\ (\exists\id{prec}':(e_3,e_4)\in\id{prec}'^*)\ \wedge\ \id{prec}\leq\id{prec}'
-$
+	For showing that definition~\ref{def:Traces} satisfies the axioms of trace algebra,
+	we define the empty trace ($0$) and the trace concatenation ($+$):
+\begin{definition}[Empty trace]
+\label{def:Empty trace}
+\item   $\forall e: 0_\id{prec}\ =\ (e,e)_\id{prec}$
+\end{definition}
+	We denote the empty trace as $0$ and define $(e_1,e_1)=0=(e_2,e_2)$.
+
+	If $(e_1,e)_\id{prec}$ and $(e,e_2)_{\id{prec}'}$ are traces then
+	we define the concatenation operator $+:\mathbb T\rightarrow\mathbb T\rightarrow\mathbb T$ by
+\begin{definition}[Concatenation of traces]
+\label{def:concatenation}
+\item   $(e_1,e)_\id{prec}+(e,e_2)_{\id{prec}'}=(e_1,e_2)_{\id{prec}\cup\id{prec}'}$
+\end{definition}
+	Note that the definition of $+$ is partial, because $(e_1,e_2)_\id{prec}+(e_3,e_4)_{\id{prec}'}$ is not defined unless $e_2=e_3$.
+	Nevertheless, the $+$ operator is closed: the result of composing two traces is always a trace.
+	(proof needed)
+
+	Let us define prefix for traces ($\leq$), to correspond closely to the prefix for histories
+	(as given in definition~\ref{def:lies in the past}):
+\begin{definition}[Prefix]
+\label{def:Prefix of traces}
+\item $\begin{array}[t]{rcl}
+	(e,e_1)_\id{prec}\leq(e,e_2)_{\id{prec}'}&\Leftrightarrow&\id{prec}\leq\id{prec}'\ \wedge\ e_1\ \id{prec}'\ e_2
+\end{array}$
+\end{definition}
+	We have defined this operator such that it satisfies Foster's algebraic definition of prefix~\cite{Foster17b}.
+	(proof needed)
 \begin{eqnarray}
@@
 \end{eqnarray}
-	We also define subtraction of traces as a way to "cut off" a prefix from a trace.
+	The prefix allows us to define subtraction of traces, to have a way to ``cut off'' an ``older'' part of a trace.
 \begin{definition}[Subtract]
 \label{def:Subtract}
+\item $\begin{array}[t]{rcll}
+	(e,e_1)_\id{prec}-(e,e_2)_{\id{prec}'}&=&(e_1,e_2)_{\id{prec}'}&\text{if }(e,e_1)_\id{prec}\leq(e,e_2)_{\id{prec}'}\\
+	   &=&0&\text{otherwise}
+\end{array}$
+\end{definition}
+	This definition satisfies Foster's definition of subtract (eqn.~\ref{axi:Foster subtract})
+\begin{equation}
+\label{axi:Foster subtract}
 \begin{array}[t]{rcll}
-	y-x&=&\text{if }x\leq y:&\exists z: y=x+z\Rightarrow z\\
-	   & &\text{otherwise}:&0
+	y-x&=&z&\text{if }x\leq y\ \wedge\ \exists z: y=x+z\\
+	   &=&0&\text{otherwise}
 \end{array}
-\end{definition}
+\end{equation}
 
-	Durability is meant to express that an action preserves history:
+\section{State}
+\label{sct:State}
+	Typically, distributed information systems are stateful.
+	Every event in the history of a DIS may cause a different (local) state.
+	(The distributed nature of a DIS means that states are local.)
+	Let $\mathbb S$ be the set of all conceivable states.
+	So we assume a function $\id{state}_\id{prec}:\mathbb E\rightarrow\mathbb S$ that assigns one state to every event $e$ in history $\id{prec}$.
+	State $\id{state}_\id{prec}(e)$ represents the state immediately after $e$ has happened.
+	We assume an empty state, $\emptyset$, which is the state immediately before an initial event has happened.
+	The global state of the system, if relevant, can be computed by $\{\id{state}_\id{prec}(e) | e\in\id{current}(\id{prec})\}$.
+	Locally we work with one specific current event, $e$, with $\id{state}_\id{prec}(e)$ representing the local current state.
+
+	In a DIS, new states typically emerge as a change on an existing state.
+	So we want to describe a DIS in terms of changes.
+	Let $\mathbb D$ be the set of all conceivable changes and let $\delta_\id{prec}:\mathbb E\times\mathbb E\rightarrow\mathbb D$
+	be a function that maps a pair of events to the corresponding change.
+	The change between the states of events $e_1$ and $e_2$ is given by $\delta_\id{prec}(e_1,e_2)$.
+	Let us define a DIS as a tuple $(\id{prec},\delta_\id{prec})$,
+	in which $\id{prec}$ is a history and
+	$\delta_\id{prec}$ represents the changes of state that have taken place throughout the history.
+
+	We need operators to compute a new state ($s_2$) from a given state ($s_1$).
+	In fact, we need two operators, $\oplus:\mathbb S\times\mathbb D\rightarrow\mathbb S$ and $\ominus:\mathbb S\times\mathbb S\rightarrow\mathbb D$, such that
+\begin{equation}
+	s_2\ =\ s_1\oplus(s_2\ominus s_1)
+\label{req:oplus and ominus}
+\end{equation}
+	Let us call $s_2\ominus s_1$ the difference between states $s_2$ and $s_1$.
+	The following equations represent the laws about state.
+	They must be kept true all the time in a DIS.
+\begin{eqnarray}
+	\delta_\id{prec}(e_1,e_2)&=&\id{state}_\id{prec}(e_2)\ominus\id{state}_\id{prec}(e_1)\\
+	e_1\ \id{prec}\ e_2&\Rightarrow&\id{state}_\id{prec}(e_2)\ =\ \id{state}_\id{prec}(e_1)\oplus\delta_\id{prec}(e_1,e_2)
+\end{eqnarray}
+
+\section{Transactions}
+\label{sct:Transactions}
+	With the theory in place, let us proceed to define transactions.
+	Let us first define the notion of action ($\id{act}$)
+	and then define the classical properties of transactions: atomic, consistent, isolated, and durable.
+
+	An action is denoted as $\id{act}_{e_1,e_2,\Delta}$, which is a mapping from one DIS to another
+\begin{equation}
+\begin{array}{rl}
+	\multicolumn{2}{l}{e_2\not\in\id{events}(\id{prec})\ \wedge\ (e_1,e_2)\in\id{prec}'^*\ \Rightarrow}\\
+	&\id{act}_{e_1,e_2,\Delta}(\id{prec},\delta_\id{prec})\ =\ (\id{prec}',\delta'_{\id{prec}'})\\
+	&\text{where}\\
+	&\ \ \begin{array}[t]{rcll}
+		\delta'_{\id{prec}'}(e,e')&=&\Delta&\text{if }(e,e')=(e_1,e_2)\\
+		\delta'_{\id{prec}'}(e,e')&=&\delta_\id{prec}(e,e')&\text{otherwise}\\
+		\end{array}
+\end{array}
+\end{equation}
+	The state function can be computed recursively as follows:
+\begin{equation}
+\begin{array}{rcll}
+	\id{state}_\id{prec}(e)&=&\id{state}_\id{prec}(e_1)\oplus\delta(e_1,e)&\text{if }(e_1,e)\in\id{prec}\\
+	                       &=&\emptyset&\text{otherwise}
+\end{array}
+\end{equation}
+
+	Durability is meant to express that an action preserves history.
 \begin{definition}[Durability]
@@
 \begin{eqnarray}
-	h\subseteq\id{act}(h)
+	\id{prec}^*\subseteq\id{prec}'^*
 \end{eqnarray}
@@
 	An action $\id{act}_\id{prec}(e_1,e_2)$ is performed in isolation if it does not affect any other action in the system.
-	This means that the only event in trace $(e_1,e_2)_\id{prec}$ that precedes events outside that trace is e_1.
-	Likewise, the only event in trace $(e_1,e_2)_\id{prec}$ that is preceded by events outside that trace is e_2.
+	This means that the only event in trace $(e_1,e_2)_\id{prec}$ that precedes events outside that trace is $e_1$.
+	Likewise, the only event in trace $(e_1,e_2)_\id{prec}$ that is preceded by events outside that trace is $e_2$.
 	To define this, we need an auxiliary function
@@
 \begin{definition}[Isolation]
-\item An action $\id{act}_{\id{prec}}(e_1,e_2)$ works in isolation iff:
+\item An action $\id{act}_\id{prec}(e_1,e_2)$ works in isolation iff:
 \label{def:Isolation}
@@
 \begin{definition}[Atomic]
-\item An action $\id{act}_{\id{prec}}(e_1,e_2)$ is atomic iff
+\item An action $\id{act}_\id{prec}(e_1,e_2)$ is atomic iff
 \label{def:Atomicity}
@@
 	Such changes are called {\em transactions}.
-	To describe a DIS in terms of changes, we need operators to compute a new state ($s_2$) from a given state ($s_1$).
-	In fact, we need two operators, $\oplus$ and $\ominus$, such that
-\begin{equation}
-	s_2\ =\ s_1\oplus(s_2\ominus s_1)
-\label{req:oplus and ominus}
-\end{equation}
-	Let us call $s_2\ominus s_1$ the difference ($\Delta$) between states $s_2$ and $s_1$.
 	We use $s_1\oplus\Delta$ to compute a new state from an existing state $s_1$ and a change $\Delta$.

### 2019-06-14 2018_Transactions/articleTransactions.tex
language and explanation

@@
 	Several solutions exist, such as nested transactions and distributed transactions.
-	Practitioners tend to avoid them when possible becaus the are complicated and error prone.
+	Practitioners tend to avoid them when possible because they are complicated and error-prone.
 
@@
 	These users want predictable behaviour, even when they change a shared state,
-	e.g. when reserving a seat in a global flight reservation system like Galileo.
+	e.g. when reserving a seat in a global flight reservation system like Galileo or Amadeus.
 	Although we cannot always avoid sharing a state,
@@
 	{\em Durability} means that history does not change.
+
 	The result of this work contributes to the body of foundational theory of information systems
@@
 
-	We use variables $e$, $e_1$, and $e_2$ to represent events.
+	We use variables $e$, $e'$, $e_1$, and $e_2$ to represent events.
 	Let $\mathbb E$ represent the set of all conceivable events.
-%	We also introduce the notion of {\em history structure}
-%	to represent all events that have happened up to some point in time and the precedences between those events.
-%	In this paper, variables $h$, $h'$, $h_1$, and $h_2$ represent history structures.
 
@@
 \label{sct:Traces}
-	Transactions leave traces in the history of a DIS, so let us define traces.
+	Transactions leave {\em traces} in the history of a DIS, so let us define these traces.
 	For this purpose we have chosen definitions that satisfy trace algebra
@@
 	Other interpretations of the same trace algebra comprise real-time systems and hybrid systems,
-	which is proven by Foster~\cite{Foster17b}.
+	which has been  proven by Foster~\cite{Foster17b}.
 	We prove (TBD) that our definitions comply with Foster's axioms of trace algebra,
@@
 \end{definition}
-	This definition satisfies Foster's definition of subtract (eqn.~\ref{axi:Foster subtract})
+	Definition~\ref{def:Subtract} satisfies Foster's definition of subtract:
 \begin{equation}
@@
 	Let $\mathbb S$ be the set of all conceivable states.
-	So we assume a function $\id{state}_\id{prec}:\mathbb E\rightarrow\mathbb S$ that assigns one state to every event $e$ in history $\id{prec}$.
+	So we need a function $\id{state}_\id{prec}:\mathbb E\rightarrow\mathbb S$ that assigns one state to every event $e$ in history $\id{prec}$.
 	State $\id{state}_\id{prec}(e)$ represents the state immediately after $e$ has happened.
+	If the history $\id{prec}$ is obvious from the context, we may write $\id{state}(e)$ rather than $\id{state}_\id{prec}(e)$.
 	We assume an empty state, $\emptyset$, which is the state immediately before an initial event has happened.
-	The global state of the system, if relevant, can be computed by $\{\id{state}_\id{prec}(e) | e\in\id{current}(\id{prec})\}$.
+	The global state of the system, if relevant, can be computed by $\{\id{state}(e) | e\in\id{current}(\id{prec})\}$.
 	Locally we work with one specific current event, $e$, with $\id{state}_\id{prec}(e)$ representing the local current state.
@@
 \begin{equation}
+\label{def:action}
 \begin{array}{rl}
@@
 
-	Durability is meant to express that an action preserves history.
+	Now let us define the classical characteristic properties of transactions.
+	We start with the easiest one: durability.
+	This property expresses that an action must preserve history.
 \begin{definition}[Durability]
-\item An action $\id{act}:\mathbb H\rightarrow\mathbb H$ is durable iff for every $h\in\mathbb H$:
+\item An action $\id{act}_{e_1,e_2,\Delta}$ is durable iff:
 \label{def:Durability}
-\begin{eqnarray}
-	\id{prec}^*\subseteq\id{prec}'^*
-\end{eqnarray}
-\end{definition}
-
-	An action $\id{act}_\id{prec}(e_1,e_2)$ is performed in isolation if it does not affect any other action in the system.
+\[\begin{array}{rl}
+	&\id{act}_{e_1,e_2,\Delta}(\id{prec},\delta_\id{prec})\ =\ (\id{prec}',\delta'_{\id{prec}'})\\
+	\Rightarrow\\
+	&\id{prec}^*\subseteq\id{prec}'^*\ \wedge\ \delta_\id{prec}\subseteq\delta'_{\id{prec}'}
+\end{array}\]
+\end{definition}
+	Note that durability is already satisfied in the definition of action (equation~\ref{def:action}). (proof needed)
+	Such is not the case with isolation.
+	An action $\id{act}_\id{prec}(e_1,e_2)$ is isolated, i.e. performed in isolation, if it does not affect any other action in the system.
 	This means that the only event in trace $(e_1,e_2)_\id{prec}$ that precedes events outside that trace is $e_1$.
@@
 \end{eqnarray}
-	We can now define isolation as the property that the traces of the "atomized" history
+	We can now define isolation as the property that the traces of the ``atomized'' history
 	are the same as the atomization of all traces in that history.
 \begin{definition}[Isolation]
-\item An action $\id{act}_\id{prec}(e_1,e_2)$ works in isolation iff:
+\item An action $\id{act}_{e_1,e_2,\Delta}$ works isolated in history $\id{prec}$ iff:
 \label{def:Isolation}
@@
 
-	Atomicity expresses that a sequence of multiple actions can be treated as one.
+	Atomicity is related to isolation because it expresses that a sequence of multiple actions can be treated as one.
 
 \begin{definition}[Atomic]
-\item An action $\id{act}_\id{prec}(e_1,e_2)$ is atomic iff
+\item An action $\id{act}_{e_1,e_2,\Delta}$ is atomic in history $\id{prec}$ iff:
+\[\begin{array}{rl}
+	&\id{act}_{e_1,e_2,\Delta}(\id{prec},\delta_\id{prec})\ =\ (\id{prec}',\delta'_{\id{prec}'})\\
+	\Rightarrow\\
+	&\id{atomize}(e_1,e_2,\id{prec})\ \leq\ \id{prec}'
+\end{array}\]
 \label{def:Atomicity}

### 2019-06-14 2018_Transactions/articleTransactions.tex
WIP

@@
 	With the theory in place, let us proceed to define transactions.
-	Let us first define the notion of action ($\id{act}$)
+	Let us first define the notion of transaction
 	and then define the classical properties of transactions: atomic, consistent, isolated, and durable.
 
-	An action is denoted as $\id{act}_{e_1,e_2,\Delta}$, which is a mapping from one DIS to another
-\begin{equation}
-\label{def:action}
+	A transaction is denoted as $\id{act}_{e_1,e_2,\Delta}$, which is a mapping from one DIS to another
+\begin{equation}
+\label{def:transaction}
 \begin{array}{rl}
@@
 	Now let us define the classical characteristic properties of transactions.
-	We start with the easiest one: durability.
-	This property expresses that an action must preserve history.
+	Let us start with atomic.
+	Atomicity expresses that a trace of multiple events can be treated as a single pair $\in\id{prec}$.
+	To define this, we need an auxiliary function
+	$\id{atomize}:\mathbb E\times\mathbb E\times\mathbb H\rightarrow\mathbb H$,
+	which replaces a trace $(e_1,e_2)_\id{prec}$ by the pair $(e_1,e_2)$
+	in a history $\id{prec}$.
+\begin{eqnarray}
+	\begin{array}[t]{r@{}l@{}l}
+	\multicolumn{3}{l}{\id{atomize}(e_1,e_2,\id{prec})}\\
+	   \ =\ \id{prec}-&(&\{ (e',e)\in\id{prec}| e_1\ \id{prec}^*\ e\ \wedge\ e\ \id{prec}^*\ e_2\}\ \cup\\
+			  & &\{ (e,e')\in\id{prec}| e_1\ \id{prec}^*\ e\ \wedge\ e\ \id{prec}^*\ e_2\}\\
+			  &)\\
+		\cup&&\{ (e_1,e_2) \}\\
+	\end{array}
+\end{eqnarray}
+
+\begin{definition}[Atomic]
+\item A transaction $\id{act}_{e_1,e_2,\Delta}$ is atomic in history $\id{prec}$ iff:
+\[\begin{array}{rl}
+	&\id{act}_{e_1,e_2,\Delta}(\id{prec},\delta_\id{prec})\ =\ (\id{prec}',\delta'_{\id{prec}'})\\
+	\Rightarrow\\
+	&\id{atomize}(e_1,e_2,\id{prec})\ \leq\ \id{prec}'
+\end{array}\]
+\label{def:Atomicity}
+\end{definition}
+
+	Now let us turn to consistency.
+	This property means that a transaction preserves truth of the state.
+	We need a predicate $\mathcal P\{\mathbb S\}$ to express that a state is consistent.
+\begin{definition}[Consistency]
+\item A transaction $\id{act}_{e_1,e_2,\Delta}$ is consistent w.r.t. $\mathcal P$ iff:
+\[\mathcal P\{\id{state}_\id{prec}(e_1)\}\ \Rightarrow\ \mathcal P\{\id{state}_\id{prec}(e_2)\}\]
+\label{def:Consistency}
+\end{definition}
+
+	The third property is isolation.
+	A transaction $\id{act}_\id{prec}(e_1,e_2)$ is isolated, i.e. performed in isolation, if it does not affect any other transaction in the system.
+	This means that the only event in trace $(e_1,e_2)_\id{prec}$ that precedes events outside that trace is $e_1$.
+	Likewise, the only event in trace $(e_1,e_2)_\id{prec}$ that is preceded by events outside that trace is $e_2$.
+	We can now define isolation as the property that the traces of the ``atomized'' history
+	are the same as the atomization of all traces in that history.
+\begin{definition}[Isolation]
+\item A transaction $\id{act}_{e_1,e_2,\Delta}$ works isolated in history $\id{prec}$ iff:
+\label{def:Isolation}
+\begin{eqnarray}
+	(\id{atomize}(e_1,e_2,\id{prec}))^*&=&\id{atomize}(e_1,e_2,\id{prec}^*)
+\end{eqnarray}
+\end{definition}
+
+	Then we have durability.
+	This property expresses that a transaction must preserve history.
 \begin{definition}[Durability]
-\item An action $\id{act}_{e_1,e_2,\Delta}$ is durable iff:
+\item A transaction $\id{act}_{e_1,e_2,\Delta}$ is durable iff:
 \label{def:Durability}
@@
 \end{definition}
-	Note that durability is already satisfied in the definition of action (equation~\ref{def:action}). (proof needed)
-	Such is not the case with isolation.
-	An action $\id{act}_\id{prec}(e_1,e_2)$ is isolated, i.e. performed in isolation, if it does not affect any other action in the system.
-	This means that the only event in trace $(e_1,e_2)_\id{prec}$ that precedes events outside that trace is $e_1$.
-	Likewise, the only event in trace $(e_1,e_2)_\id{prec}$ that is preceded by events outside that trace is $e_2$.
-	To define this, we need an auxiliary function
-	$\id{atomize}:\mathbb E\times\mathbb E\times\mathbb H\rightarrow\mathbb H$,
-	which replaces a trace $(e_1,e_2)_\id{prec}$ by the pair $(e_1,e_2)$
-	in a history $\id{prec}$.
-\begin{eqnarray}
-	\begin{array}[t]{r@{}l@{}l}
-	\multicolumn{3}{l}{\id{atomize}(e_1,e_2,\id{prec})}\\
-	   \ =\ \id{prec}-&(&\{ (e_1,e)\in\id{prec}| e_1\ \id{prec}^*\ e\ \wedge\ e\ \id{prec}^*\ e_2\}\ \cup\\
-			  & &\{ (e,e_2)\in\id{prec}| e_1\ \id{prec}^*\ e\ \wedge\ e\ \id{prec}^*\ e_2\}\\
-			  &)\\
-		\cup&&\{ (e_1,e_2) \}\\
-	\end{array}
-\end{eqnarray}
-	We can now define isolation as the property that the traces of the ``atomized'' history
-	are the same as the atomization of all traces in that history.
-\begin{definition}[Isolation]
-\item An action $\id{act}_{e_1,e_2,\Delta}$ works isolated in history $\id{prec}$ iff:
-\label{def:Isolation}
-\begin{eqnarray}
-	(\id{atomize}(e_1,e_2,\id{prec}))^*&=&\id{atomize}(e_1,e_2,\id{prec}^*)
-\end{eqnarray}
-\end{definition}
-
-	Atomicity is related to isolation because it expresses that a sequence of multiple actions can be treated as one.
-
-\begin{definition}[Atomic]
-\item An action $\id{act}_{e_1,e_2,\Delta}$ is atomic in history $\id{prec}$ iff:
-\[\begin{array}{rl}
-	&\id{act}_{e_1,e_2,\Delta}(\id{prec},\delta_\id{prec})\ =\ (\id{prec}',\delta'_{\id{prec}'})\\
-	\Rightarrow\\
-	&\id{atomize}(e_1,e_2,\id{prec})\ \leq\ \id{prec}'
-\end{array}\]
-\label{def:Atomicity}
-\end{definition}
-
-	Consistency of an action means that it preserves truth.
-\begin{definition}[Consistency]
-\item An action $\id{act}:\mathbb H\rightarrow\mathbb H$ is consistent iff it preserves the integrity constraints of all contexts in the system.
-\label{def:Consistency}
-\begin{eqnarray}
-	(h',e',r')=\id{act}(h,e,r)\ \wedge\ r(\id{stat}(h,e))&\Rightarrow&r'(\id{stat}(h',e'))
-\end{eqnarray}
-\end{definition}
+	Note that durability is already satisfied in the definition of transaction (equation~\ref{def:transaction}). (proof needed)
+
+\section{Bibliography}
+\bibliographystyle{elsarticle-harv}
+\bibliography{doc}
+
+\end{document}
 

### 2019-06-15 2018_Transactions/articleTransactions.tex
repair mistakes

@@
 	we may not know in which order two events\footnote{Cambridge Dictionary defines ``event'' as: anything that happens} have happened.
-	For this reason we describe the history of a DIS as a partially ordered relation of events.
+	For this reason a partial order is used to describe the history of a DIS.
 
@@
 \begin{eqnarray}
-	&&(e,e)\notin\id{prec}\label{req:prec irreflexive}\\
-	&&(e_1,e_2)\in\id{prec}^+\ \wedge\ (e_2,e_1)\in\id{prec}^+\ \Rightarrow\ e_1=e_2\label{req:precPlus asy}
+	&&\id{prec}\ \subseteq\ \overline{I}\label{req:prec irreflexive}\\
+	&&\id{prec}^+\cap{\flip{\id{prec}}}^+\ \subseteq\ I\label{req:precPlus asy}
 \end{eqnarray}
@@
 \begin{eqnarray}
-	&&\id{events}(\id{prec})\ =\ \{e_1|(e_1,e_2)\in\id{prec}\}\cup\{e_2|(e_1,e_2)\in\id{prec}\}\label{req:events}
+	\id{events}(\id{prec})&=&(V;\id{prec}\cap I)\ \cup\ (I\cap\id{prec};V)
 \end{eqnarray}
@@
 \begin{eqnarray}
-	\id{current}(\id{prec})&=&\{e |\ \exists e_1: (e_1,e)\in\id{prec}\ \wedge\ \neg\exists e_2: (e,e_2)\in\id{prec}\}\\
-	\id{initial}(\id{prec})&=&\{e |\ \exists e_1: (e,e_1)\in\id{prec}\ \wedge\ \neg\exists e_2: (e_2,e)\in\id{prec}\}
+	\id{current}(\id{prec})&=&\overline{\id{prec}}^+;\id{prec}^+\cap I\\
+	\id{initial}(\id{prec})&=&\id{prec}^+;\overline{\id{prec}}^+\cap I
 \end{eqnarray}
@@
 \label{def:Traces}
-\item   $(e_1,e_2)_\id{prec}\in\mathbb T\ \Leftrightarrow\ (e_1,e_2)\in\id{prec}^*$
+\item $(e_1,e_2)_\id{prec}\ =\ e_1;\id{prec}^*\cap\id{prec}^*;e_2$
 \end{definition}
@@
 	For showing that definition~\ref{def:Traces} satisfies the axioms of trace algebra,
-	we define the empty trace ($0$) and the trace concatenation ($+$):
-\begin{definition}[Empty trace]
-\label{def:Empty trace}
-\item   $\forall e: 0_\id{prec}\ =\ (e,e)_\id{prec}$
+	we define the empty trace ($0$) and the trace concatenation ($+$) as $I_\mathbb E$ and $;$ (relation composition).
+\begin{definition}[trace operators]
+\item $0\ =\ I$
+\item $x+y\ =\ x;y$
 \end{definition}
-	We denote the empty trace as $0$ and define $(e_1,e_1)=0=(e_2,e_2)$.
-
-	If $(e_1,e)_\id{prec}$ and $(e,e_2)_{\id{prec}'}$ are traces then
-	we define the concatenation operator $+:\mathbb T\rightarrow\mathbb T\rightarrow\mathbb T$ by
-\begin{definition}[Concatenation of traces]
-\label{def:concatenation}
-\item   $(e_1,e)_\id{prec}+(e,e_2)_{\id{prec}'}=(e_1,e_2)_{\id{prec}\cup\id{prec}'}$
-\end{definition}
-	Note that the definition of $+$ is partial, because $(e_1,e_2)_\id{prec}+(e_3,e_4)_{\id{prec}'}$ is not defined unless $e_2=e_3$.
-	Nevertheless, the $+$ operator is closed: the result of composing two traces is always a trace.
-	(proof needed)
 
@@
 	Let $\mathbb S$ be the set of all conceivable states.
-	So we need a function $\id{state}_\id{prec}:\mathbb E\rightarrow\mathbb S$ that assigns one state to every event $e$ in history $\id{prec}$.
-	State $\id{state}_\id{prec}(e)$ represents the state immediately after $e$ has happened.
-	If the history $\id{prec}$ is obvious from the context, we may write $\id{state}(e)$ rather than $\id{state}_\id{prec}(e)$.
+	So we need a function $\id{state}:\mathbb E\rightarrow\mathbb S$ that assigns one state to every event $e$ in history $\id{prec}$.
+	State $\id{state}(e)$ represents the state immediately after $e$ has happened.
+	If the history $\id{prec}$ is obvious from the context, we may write $\id{state}(e)$ rather than $\id{state}(e)$.
 	We assume an empty state, $\emptyset$, which is the state immediately before an initial event has happened.
 	The global state of the system, if relevant, can be computed by $\{\id{state}(e) | e\in\id{current}(\id{prec})\}$.
-	Locally we work with one specific current event, $e$, with $\id{state}_\id{prec}(e)$ representing the local current state.
+	Locally we work with one specific current event, $e$, with $\id{state}(e)$ representing the local current state.
 
@@
 \begin{eqnarray}
-	\delta_\id{prec}(e_1,e_2)&=&\id{state}_\id{prec}(e_2)\ominus\id{state}_\id{prec}(e_1)\\
-	e_1\ \id{prec}\ e_2&\Rightarrow&\id{state}_\id{prec}(e_2)\ =\ \id{state}_\id{prec}(e_1)\oplus\delta_\id{prec}(e_1,e_2)
+	\delta_\id{prec}(e_1,e_2)&=&\id{state}(e_2)\ominus\id{state}(e_1)\\
+	e_1\ \id{prec}\ e_2&\Rightarrow&\id{state}(e_2)\ =\ \id{state}(e_1)\oplus\delta_\id{prec}(e_1,e_2)
 \end{eqnarray}
@@
 \begin{array}{rcll}
-	\id{state}_\id{prec}(e)&=&\id{state}_\id{prec}(e_1)\oplus\delta(e_1,e)&\text{if }(e_1,e)\in\id{prec}\\
+	\id{state}(e)&=&\id{state}(e_1)\oplus\delta(e_1,e)&\text{if }(e_1,e)\in\id{prec}\\
 	                       &=&\emptyset&\text{otherwise}
@@
 	Let us start with atomic.
+
+	In his paper "Splitting Atoms Safely"~\cite{JONES2007109} Cliff Jones\footnote{%
+	Jones is author of the Atomicity Manifesto and author of the formal method VDM.}
+	describes the intended meaning of atomicity in natural language:
+	``If an action is said to be executed “atomically”, it is assumed that it will not be affected by interference
+	and that the environment will not be able to observe intermediate steps of the action in question.''
+	Jones argues that interference must be tolerated, however.
+	With shared state programs, he says, an action must achieve some required result even though its state can be changed by other interfering processes.
+
+	The purpose of this paper is to propose a useful formal definition of atomicity.
 	Atomicity expresses that a trace of multiple events can be treated as a single pair $\in\id{prec}$.
@@
 \item A transaction $\id{act}_{e_1,e_2,\Delta}$ is consistent w.r.t. $\mathcal P$ iff:
-\[\mathcal P\{\id{state}_\id{prec}(e_1)\}\ \Rightarrow\ \mathcal P\{\id{state}_\id{prec}(e_2)\}\]
+\[\mathcal P\{\id{state}(e_1)\}\ \Rightarrow\ \mathcal P\{\id{state}(e_2)\}\]
 \label{def:Consistency}
@@
 	The third property is isolation.
-	A transaction $\id{act}_\id{prec}(e_1,e_2)$ is isolated, i.e. performed in isolation, if it does not affect any other transaction in the system.
+	A transaction $\id{act}(e_1,e_2)$ is isolated, i.e. performed in isolation, if it does not affect any other transaction in the system.
 	This means that the only event in trace $(e_1,e_2)_\id{prec}$ that precedes events outside that trace is $e_1$.

### 2019-11-04 2018_Transactions/articleTransactions.tex
Improve language and arguments

@@
 	To demonstrate that the theory is about real transactions, this contribution defines semantics of conventional database transactions.
-	To fulfill our promise, we also gives semantics of transactions on the semantic web.
 \end{abstract}
@@
 	leaving the problems for the user to solve.
-	As a result, architectural patterns exist for the purpose of doing transactions in a distributed setting.
-	Some of these patterns simply circumvent the problem.
-	Others will isolate transactionality into a smaller sub-problem.
+	As a result, many architectural patterns have emerged for the purpose of doing transactions in a distributed setting.
+	Some of these patterns simply advise how to avoid or circumvent the problem.
+	Others will isolate transactionality into a smaller sub-problem,
+        downsizing the problem and hoping to deal with the sub-problem in a conventional transactional database.
 	Still others introduce locking mechanisms to do the job.
-
-	Our work is inspired by the observation that the merge of Git corresponds to the commit that we know from transactional databases.
+        However, sometimes we must avoid locking for performance reasons and avoid a globally shared database for reasons of size.
+        As is the case with large, globally distributed information systems such as Facebook, LinkedIn, and Netflix.
+        In case we want full transactional behaviour in such systems, we have a problem.
+
+        This problem seems to be solved by Git.
+	Our work is inspired by the observation that the merge of Git corresponds to a transactional commit.
 	We show this correspondence by expressing the characteristic properties of transactions in a theory of time.
@@
 	well-established, and founded in a solid mathematical structure~\cite{Foster14}.
-	This means we have to redefine Atomicity, Consistency, Isolation, and Durability in terms of events.
+	For expressing transactions in terms of events we have to redefine Atomicity, Consistency, Isolation, and Durability.
 	Informally, {\em Atomicity} means that a sequence of primitive actions can be executed as though it were a single action.
@@
 	{\em Durability} means that history does not change.
+
+We define the history of a distributed information system as a directed acyclig graph $\la\mathcal E,\mathcal C\ra$ with
+\begin{definition}
+\begin{eqnarray}\mathcal C\subseteq\mathcal E\times\mathcal E\\
+\mathcal C^+\cap I_{\mathcal E}=\emptyset
+\end{eqnarray}
+\end{definition}
+A transaction is a set of events.
+\begin{enumerate}
+\item   Atomic\\
+A transaction is atomic if it can be replaced by a single event without consequences for other parts of the system.
+\item   Consistent\\
+If the initial states of a transaction satisfy a given set of rules, we say that a transaction maintains consistency if its final state satisfies that same set of rules.
+\item   Isolated\\
+A transaction is isolated if its events are not shared by other transactions.
+\item   Durable\\
+A transaction is durable if it is part of every future history.
+\end{enumerate}
 
@@
 \label{sct:State}
-	Typically, distributed information systems are stateful.
-	Every event in the history of a DIS may cause a different (local) state.
-	(The distributed nature of a DIS means that states are local.)
+	Typically, distributed information systems may have states.
+%        Consider for example Twitter: the entire timeline on your smartphone represents the state.
+	Every new event in the history of a DIS may cause a different state.
+	The distributed nature of a DIS means that states are local.
 	Let $\mathbb S$ be the set of all conceivable states.
-	So we need a function $\id{state}:\mathbb E\rightarrow\mathbb S$ that assigns one state to every event $e$ in history $\id{prec}$.
+	We need a function $\id{state}:\mathbb E\rightarrow\mathbb S$ that assigns one state to every event $e$ in history $\id{prec}$.
 	State $\id{state}(e)$ represents the state immediately after $e$ has happened.
@@
 	We assume an empty state, $\emptyset$, which is the state immediately before an initial event has happened.
-	The global state of the system, if relevant, can be computed by $\{\id{state}(e) | e\in\id{current}(\id{prec})\}$.
+	The global state of the system, if deemed relevant, can be computed by $\{\id{state}(e) | e\in\id{current}(\id{prec})\}$.
 	Locally we work with one specific current event, $e$, with $\id{state}(e)$ representing the local current state.
@@
 
-	In his paper "Splitting Atoms Safely"~\cite{JONES2007109} Cliff Jones\footnote{%
+	In his paper ``Splitting Atoms Safely''~\cite{JONES2007109} Cliff Jones\footnote{%
 	Jones is author of the Atomicity Manifesto and author of the formal method VDM.}

### 2022-08-21 2022 Migration/articleMigration.tex
Making the scaffold for a new paper about migration.

\documentclass{elsarticle}
\usepackage{graphicx}
%\usepackage{multicol}
%\usepackage{footmisc}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage[english]{babel}
%\usepackage[official,right]{eurosym}
\selectlanguage{english}
\include{preamramics}
\hyphenation{ExecEngine}
\newtheorem{lemma}{Lemma}
\def\id#1{\text{\it #1\/}}
\def\Events{{\mathit E}}
\begin{document}

\title{A Theory for Migration of Information Systems}
\author[ou,ordina]{Stef Joosten\fnref{fn1}}
\ead{stef.joosten@ou.nl}
\author[umn]{Sebastiaan Joosten\fnref{fn2}}
\address[ou]{Open Universiteit Nederland, Heerlen, the Netherlands}
\address[ordina]{Ordina NV, Nieuwegein, the Netherlands}
\address[umn]{University of Minnesota, Minneapolis, USA}
\fntext[fn1]{ORCID 0000-0001-8308-0189}
\fntext[fn2]{ORCID 0000-0002-6590-6220}

\begin{abstract}
	The Ampersand project has provided the theory and tools to generate semantic information systems from an algebraic specification.
	However, information systems in practice may change repeatedly after their maiden deployment.
	Changes that affect the data model typically result in a data migration.
	In such cases, simply regenerating the system is not enough.

	In this contribution we develop a theory for data migration that aims at automating the migration,
	to enable more frequent migrations.
	A theory is needed to do this reliably, without mistakes.
	This paper proposes such a theory.

	The ultimate target is to generate the entire migration from two specifications: the as-is specification and the to-be specification.
	The theory for doing that is the topic of this paper.
	The software generator that embodies this theory is subject of future research.
\end{abstract}

\begin{keyword}
relation algebra\sep software development\sep data migration\sep software migration\sep Ampersand
\end{keyword}
\maketitle

\section{Introduction}
\label{sct:Introduction}

\section{Bibliography}
\bibliographystyle{elsarticle-harv}
\bibliography{doc}

\end{document}

### 2022-08-27 2022 Migration/articleMigration.tex
Introduce the notion of dataset in the Migration paper.
Migration of data means to preserve as much as possible from an existing dataset into the new one.

@@
 
+\section{Basics}
+\label{sct:Basics}
+	In this section we explain the basic notions to describe migration.
+	We introduce the notion of dataset.
+	A dataset is a triple $\la\rules,\rels,\concepts,\pairs,\atoms\ra$,
+	which consists of a set of rules $\rules$, a set of relations $\rels$, a set of concepts $\concepts$, a set of pairs $\pairs$, and a set of atoms $\atoms$.
+	Let us introduce these five notions one by one.
+
+	Atoms are values that have no internal structure, meant to represent data elements in a database.
+	From a business perspective, atoms are used to represent concrete items of the world,
+	such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
+	By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ are used to represent \emph{atoms}.
+	The set of all atoms is called $\atoms$.
+    Each atom is an instance of a \emph{concept}.
+
+	Concepts (from set $\concepts$) are names we use to classify atoms in a meaningful way.
+	For example, you might choose to classify \atom{Peter} as a \concept{Person}, and \atom{074238991} as a \concept{TelephoneNumber}.
+	In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
+    We will use variables $A$, $B$, $C$, $D$ to represent concepts.
+	The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.
+	The expression $a \in A$ means that atom $a$ is an \emph{instance} of concept $A$.
+	The declaration of $A\isa B$ (pronounce: $A$ is a $B$)
+	in an Ampersand script states that any instance of $A$ is an instance of $B$ as well.
+	We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
+	Specialization is needed to allow statements such as: ``An orange is a fruit''.
+
+	Pairs (from set $\pairs$) are used to relate two atoms.
+	A pair $\pair{\atom{Peter}}{\atom{074238991}}$ might bear the meaning that the ``thing'' that \atom{Peter} refers to
+	has \atom{074238991} as a telephone number.
+	In the formal world, we do not care about that meaning,
+	leaving it entirely up to a user which meaning (s)he attaches to this pair.
+
+	Relations (from set $\rels$) are used in information systems to store data as pairs.
+%	A \define{fact} is a statement that is true in a business context.
+%	Facts are stored and kept as data in a computer.
+	In this paper relations are represented by variables $r$, $s$, and $d$.
+	We represent the declaration of a relation $r$ by $\declare{nm}{A}{B}$,
+	in which \id{nm} is a name and $A$ and $B$ are concepts.
+	We call $A$ the source concept and $B$ the target concept of the relation.
+	The pair $\pair{A}{B}$ is called the \emph{type} of the relation.
+	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.
+
+	The meaning of relations in Ampersand is defined by an interpretation function $\mathfrak{I}$.
+	It maps each relation to a set of pairs.
+	The type system of Ampersand guarantees that each pair in $r$ respects the type of $r$:
+\begin{equation}
+	\pair{a}{b}\in\ti{\declare{nm}{A}{B}} \Rightarrow\ a \in A \wedge b \in B \label{typing of relations}
+\end{equation}
+
+	A rule is an equality between two terms that constrains the data stored in relations,
+	ensuring that the dataset is at all times consistent with all rules in the dataset ($\rules$).
+	Terms are used to combine relations using operators.
+	The set of terms is called $\terms$.
+\begin{definition}[terms]
+\label{def:terms}
+\item   The set of terms, $\terms$, is the smallest set that satisfies, for all $r,s \in \terms$, $d\in\rels$ and $A,B \in \concepts$: 
+\begin{eqnarray}
+	d&\in&\terms         \quad\quad\text{(every relation is a term)}\\
+	(r \cup s)&\in&\terms\quad\quad\text{(union)}\\
+	(r \cap s)&\in&\terms\quad\quad\text{(intersection)}\\
+	(r-s)&\in&\terms     \quad\quad\text{(difference)}\label{def:difference}\\
+	(r;s)&\in&\terms     \quad\quad\text{(composition)}\\
+	(r\backslash s)&\in&\terms     \quad\quad\text{(right residual)}\\
+	(r\slash s)&\in&\terms     \quad\quad\text{(left residual)}\\
+	\flip r&\in&\terms   \quad\quad\text{(converse)}\\
+	\cmpl{r}&\in&\terms   \quad\quad\text{(complement)}\\
+%	\kleeneplus r&\in&\terms   \quad\quad\text{(Kleene closure)}\\
+%	\kleenestar r&\in&\terms   \quad\quad\text{(Kleene closure)}\\
+	\ident{A}&\in&\terms \quad\quad\text{(identity)}\\
+	\fullt{A}{B}&\in&\terms \quad\quad\text{(full set)}
+\end{eqnarray}
+\end{definition}
+	Throughout the remainder of this paper,	terms are represented by variables $r$, $s$, $d$, and $t$.
+	The \define{type} of a term $r$ is a pair of concepts, given by $\tf{r}$.
+	$\mathfrak{T}$ is a partial function that maps terms to types.
+	If a term has one type, it is called \define{type correct}.
+	The Ampersand compiler requires all terms to be type correct.
+	If no type can be assigned, the compiler gives an error message.
+	If multiple types are assignable to a term, the ambiguity is signalled by the compiler.
+	In that case it will also emit an error message and prompt the programmer to disambiguate the term by adding type information.
+	In practice, the derivation of types can disambiguate most terms.
+	So, the programmer experiences some freedom to denote distinct relations by the same name without the obligation to specify their types at all occurrences.
+	The compiler does not generate any code if there is a term that is not type correct.
+	The type function and its restrictions are discussed in~\cite{Joosten2015}
+	and have no consequences for the remainder of this paper.
+
+ 	The meaning of terms in Ampersand is given by an interpretation function $\mathfrak{I}$.
+	Let $A$ and $B$ be finite sets of atoms, then $\mathfrak{I}$ maps each term to the set of pairs for which that term stands.
+\begin{definition}[interpretation of terms]
+\label{interpretation of terms}
+\item   For every $A,B\in\concepts$ and $r,s\in\terms$
+\begin{eqnarray}
+	\ti{r}		 &=&\{\pair{a}{b}|\ a\ r\ b\}	\\
+	\ti{r \cup s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{or }\ \pair{a}{b}\in\ti{s}\}	\\
+	\ti{r \cap s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{a}{b}\in\ti{s}\}	\\
+	\ti{r-s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{a}{b}\notin\ti{s}\}	\\
+	\ti{r;s}	 &=&\{\pair{a}{c}|\ \text{for some}\ b,\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{b}{c}\in\ti{s}\}	\\
+	\ti{r\backslash s}	 &=&\{\pair{b}{c}|\ \text{for all}\ a,\ \pair{a}{b}\in\ti{r}\ \text{implies}\ \pair{a}{c}\in\ti{s}\}	\\
+	\ti{r\slash s}	 &=&\{\pair{a}{b}|\ \text{for all}\ c,\ \pair{b}{c}\in\ti{s}\ \text{implies}\ \pair{a}{c}\in\ti{r}\}	\\
+	\ti{\flip{r}}	 &=&\{\pair{b}{a}|\ \pair{a}{b}\in\ti{r}    \}	\\
+	\ti{\cmpl{\declare{r}{A}{B}}}	 &=&\fullt{A}{B}-r	\\
+%	\ti{\kleeneplus{r}}	 &=&\text{the smallest set that satisfies }\kleeneplus{r}=r\ \cup\ \kleeneplus{r};\kleeneplus{r}	\\
+%	\ti{\kleenestar{\declare{r}{A}{A}}}	 &=&\ident{A}\cup\kleeneplus{r}\\
+	\ti{\ident{A}} 	 &=&\{\pair{a}{a}|\ a\in A\}	\\
+	\ti{\fullt{A}{B}}&=&\{\pair{a}{b}|\ a\in A, b\in B\}
+\end{eqnarray}
+\end{definition}
+%	The Kleene closure operators (postfix $\kleeneplus{\ }$ and $\kleenestar{\ }$) have been implemented partially in the current Ampersand implementation.
+
+	A \define{rule} is a pair of terms $r,s\in\terms$ with $\tf{r}=\tf{s}$, which is syntactically recognizable as a rule.
+\[\text{RULE}\ r = s\]
+	This means \(\ti{r} = \ti{s}\). In practice, many rules are written as:
+\[\text{RULE}\ r\subs s\]
+	This is a shorthand for 
+\[\text{RULE}\ r\cap s = r\]
+	We have enhanced the type function $\mathfrak{T}$ and the interpretation function $\mathfrak{I}$ to cover rules as well.
+	If $\tf{r}=\tf{s}$ and $\tf{s}=\pair{A}{B}$:
+\begin{eqnarray}
+	\tf{\text{RULE}\ r = s}   &=&\pair{A}{B}\\
+	\tf{\text{RULE}\ r\subs s}&=&\pair{A}{B}\\
+ resolved this \end{eqnarray}
+
+	We call a rule $\text{RULE}\ r = s$ \define{satisfied} when its interpretation equals $\ti{\fullt{A}{B}}$.
+	As the population of relations used in $r$ changes with time, the satisfaction of the rule changes accordingly.
+	A software developer, who conceives these rules, must consider how to keep each one of them satisfied.
+	We call a rule \define{violated} if it is not satisfied.
+	The set $\ti{(s-r)\cup(r-s)}$ is called the \emph{violation set} of \(\text{RULE}\ r = s\).
+	To \define{resolve} violations means to change the contents of relations such that the rule is satisfied%
+\footnote{To \define{restore invariance} is sometimes used as a synonym to resolving violations.
+	To \define{preserve invariance} is sometimes used as a synonym to keeping a rule satisfied.}.
+	Consequently, such a rule may also be referred to as \define{invariant}.
+	Some ways of resolving violations are presented in section~\ref{sct:Code Fragments}.
+	Each pair in the violation set of a rule is called a violation of that rule.
+
+	The software developer must define how to resolve violations when they occur.
+	She does so by inserting and/or deleting pairs in appropriately chosen relations.
+	Whatever choice she makes, she must ensure that her code yields data that satisfies the rules.
+	When we say: ``rule $r$ specifies this action'' we mean that satisfaction of rule $r$ is a postcondition of any action specified by rule $r$.
+
+	The Ampersand compiler generates a database from relations in the script.
+	A database application%
+\footnote{Ampersand generates an application that consists of a relational database and interface components.
+	Currently this application runs server-side on a PHP/MySQL platform and on a web-browser on the client-side.}
+	assists users to keep rules satisfied throughout the lifetime of the database. It is also generated by Ampersand.
+
 \section{Bibliography}

### 2022-08-27 2022 Migration/articleMigration.tex
I left the chapter on datasets in a state that is ready for review.

@@
 	We introduce the notion of dataset.
-	A dataset is a triple $\la\rules,\rels,\concepts,\pairs,\atoms\ra$,
+	A dataset is a quintuple $\la\rules,\rels,\concepts,\pairs,\atoms\ra$,
 	which consists of a set of rules $\rules$, a set of relations $\rels$, a set of concepts $\concepts$, a set of pairs $\pairs$, and a set of atoms $\atoms$.
@@
 	The expression $a \in A$ means that atom $a$ is an \emph{instance} of concept $A$.
-	The declaration of $A\isa B$ (pronounce: $A$ is a $B$)
-	in an Ampersand script states that any instance of $A$ is an instance of $B$ as well.
+	The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
 	We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
-	Specialization is needed to allow statements such as: ``An orange is a fruit''.
+	Specialization is needed to allow statements such as: ``A woman is a human.'' or ``A human is a mammal.''.
 
@@
 	Relations (from set $\rels$) are used in information systems to store data as pairs.
-%	A \define{fact} is a statement that is true in a business context.
-%	Facts are stored and kept as data in a computer.
 	In this paper relations are represented by variables $r$, $s$, and $d$.
-	We represent the declaration of a relation $r$ by $\declare{nm}{A}{B}$,
-	in which \id{nm} is a name and $A$ and $B$ are concepts.
-	We call $A$ the source concept and $B$ the target concept of the relation.
-	The pair $\pair{A}{B}$ is called the \emph{type} of the relation.
+	Every relation has a source concept and a target concept.
+	If we need to be specific about the source and target concepts, we write $\declare{r}{A}{B}$ instead of just $r$.
+	The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
 	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.
+	The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.
+	The meaning of relations is defined by an interpretation function $\mathfrak{I}$.
+	It maps each relation to a set of pairs.
 
-	The meaning of relations in Ampersand is defined by an interpretation function $\mathfrak{I}$.
-	It maps each relation to a set of pairs.
-	The type system of Ampersand guarantees that each pair in $r$ respects the type of $r$:
+	The typing rule guarantees that each pair in $r$ respects the type of $r$:
 \begin{equation}
-	\pair{a}{b}\in\ti{\declare{nm}{A}{B}} \Rightarrow\ a \in A \wedge b \in B \label{typing of relations}
+	\tag{typing rule}
+	\pair{a}{b}\in\ti{\declare{r}{A}{B}} \Rightarrow\ a \in A \wedge b \in B \label{typing of relations}
+\end{equation}
+
+\begin{equation}
+	\tag{population}
+	\pairs\ =\ \{ \pair{a}{b}|\ r\in\rels,\ \pair{a}{b}\in\ti{r}\} \label{def. population}
+\end{equation}
+
+\begin{equation}
+	\tag{atoms}
+	\atoms\ =\ \{ a |\ \pair{b}{c}\in\pairs,\ a\in\{b,c\}\ \} \label{def atoms}
 \end{equation}
@@
 	A rule is an equality between two terms that constrains the data stored in relations,
-	ensuring that the dataset is at all times consistent with all rules in the dataset ($\rules$).
+	to ensure that the dataset is at all times consistent with all rules in the dataset ($\rules$).
 	Terms are used to combine relations using operators.
@@
 	$\mathfrak{T}$ is a partial function that maps terms to types.
-	If a term has one type, it is called \define{type correct}.
-	The Ampersand compiler requires all terms to be type correct.
-	If no type can be assigned, the compiler gives an error message.
-	If multiple types are assignable to a term, the ambiguity is signalled by the compiler.
-	In that case it will also emit an error message and prompt the programmer to disambiguate the term by adding type information.
-	In practice, the derivation of types can disambiguate most terms.
-	So, the programmer experiences some freedom to denote distinct relations by the same name without the obligation to specify their types at all occurrences.
-	The compiler does not generate any code if there is a term that is not type correct.
 	The type function and its restrictions are discussed in~\cite{Joosten2015}
 	and have no consequences for the remainder of this paper.
+	(\emph{\@Bas, komen we ermee weg als we de definitie van $\tf{r}$ hier weglaten?})
 
- 	The meaning of terms in Ampersand is given by an interpretation function $\mathfrak{I}$.
+ 	The meaning of terms is given by an interpretation function $\mathfrak{I}$.
 	Let $A$ and $B$ be finite sets of atoms, then $\mathfrak{I}$ maps each term to the set of pairs for which that term stands.
@@
 
-	A \define{rule} is a pair of terms $r,s\in\terms$ with $\tf{r}=\tf{s}$, which is syntactically recognizable as a rule.
+	A \define{rule} is a pair of terms $r,s\in\terms$ with the same types ($\tf{r}=\tf{s}$).
 \[\text{RULE}\ r = s\]
@@
 \[\text{RULE}\ r\cap s = r\]
-	We have enhanced the type function $\mathfrak{T}$ and the interpretation function $\mathfrak{I}$ to cover rules as well.
+	We have enhanced the type function $\mathfrak{T}$ to cover rules as well.
 	If $\tf{r}=\tf{s}$ and $\tf{s}=\pair{A}{B}$:
@@
 	\tf{\text{RULE}\ r = s}   &=&\pair{A}{B}\\
-	\tf{\text{RULE}\ r\subs s}&=&\pair{A}{B}\\
- resolved this \end{eqnarray}
+	\tf{\text{RULE}\ r\subs s}&=&\pair{A}{B}
+ \end{eqnarray}
 
-	We call a rule $\text{RULE}\ r = s$ \define{satisfied} when its interpretation equals $\ti{\fullt{A}{B}}$.
+	We call a rule $\text{RULE}\ r = s$ \define{satisfied} if \(\ti{r} = \ti{s}\).
+
+	The data in a data set $\la\rules,\rels,\concepts,\pairs,\atoms\ra$ are the pairs in the interpretation of all relations in $\rels$.
+	This constitutes the set $\pairs$, also known as the population of the dataset.
+	In response to messages such as
+\[\text{INSERT}\ <3,{\tt Brown}> \text{INTO}\ \declare{r}{Age}{Name}\]
+ 	the population of a dataset changes. 
 	As the population of relations used in $r$ changes with time, the satisfaction of the rule changes accordingly.
@@
 	Consequently, such a rule may also be referred to as \define{invariant}.
-	Some ways of resolving violations are presented in section~\ref{sct:Code Fragments}.
 	Each pair in the violation set of a rule is called a violation of that rule.
-
-	The software developer must define how to resolve violations when they occur.
-	She does so by inserting and/or deleting pairs in appropriately chosen relations.
-	Whatever choice she makes, she must ensure that her code yields data that satisfies the rules.
-	When we say: ``rule $r$ specifies this action'' we mean that satisfaction of rule $r$ is a postcondition of any action specified by rule $r$.
-
-	The Ampersand compiler generates a database from relations in the script.
-	A database application%
-\footnote{Ampersand generates an application that consists of a relational database and interface components.
-	Currently this application runs server-side on a PHP/MySQL platform and on a web-browser on the client-side.}
-	assists users to keep rules satisfied throughout the lifetime of the database. It is also generated by Ampersand.
 

### 2022-08-28 2022 Migration/articleMigration.tex
Further refinement on the definition of datasets

--- 2022-08-21 2022 Migration/articleMigration.tex
@@
 \selectlanguage{english}
-\include{preamramics}
 \hyphenation{ExecEngine}
 \newtheorem{lemma}{Lemma}
-\def\id#1{\text{\it #1\/}}
 \def\Events{{\mathit E}}
 \begin{document}
+\include{preamramics}
 
@@
 
+\section{Basics}
+\label{sct:Basics}
+	In this section we explain the basic notions to describe migration.
+	We introduce the notion of dataset.
+	A dataset is a quintuple $\la\rules,\rels,\concepts,\pairs,\atoms\ra$,
+	which consists of a set of rules $\rules$, a set of relations $\rels$, a set of concepts $\concepts$, a set of pairs $\pairs$, and a set of atoms $\atoms$.
+	Let us introduce these five notions one by one.
+\subsection{Atoms}
+	Atoms (from set $\atoms$) are values that have no internal structure, meant to represent data elements in a database.
+	From a business perspective, atoms are used to represent concrete items of the world,
+	such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
+	By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ are used to represent \emph{atoms}.
+	The set of all atoms is called $\atoms$.
+    Each atom is an instance of a \emph{concept}.
+\subsection{Concepts}
+	Concepts (from set $\concepts$) are names we use to classify atoms in a meaningful way.
+	For example, you might choose to classify \atom{Peter} as a \concept{Person}, and \atom{074238991} as a \concept{TelephoneNumber}.
+	In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
+    We will use variables $A$, $B$, $C$, $D$ to represent concepts.
+	The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.
+	The expression $a \in A$ means that atom $a$ is an \emph{instance} of concept $A$.
+	The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
+	We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
+	Specialization is needed to allow statements such as: ``A woman is a human.'' or ``A human is a mammal.''.
+\subsection{Pairs}
+	Pairs (from set $\pairs$) are used to relate two atoms.
+	A pair $\pair{\text{\atom{Peter}}}{\text{\atom{074238991}}}$ might bear the meaning that the ``thing'' that \atom{Peter} refers to
+	has \atom{074238991} as a telephone number.
+	In the formal world, we do not care about that meaning,
+	leaving it entirely up to a user which meaning (s)he attaches to this pair.
+\subsection{Relations}
+	Relations (from set $\rels$) are used in information systems to store data as pairs.
+	In this paper relations are represented by variables $r$, $s$, and $d$.
+	Every relation $r$ has a set of pairs, $\pop{r}$, which we call the population of $r$.
+	Every relation has a source concept and a target concept.
+	If we need to be specific about the source and target concepts, we write $\declare{r}{A}{B}$ instead of just $r$.
+	The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
+	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.
+	The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.
+
+	The signature rule guarantees that each pair in $r$ respects the signature of $r$:
+\begin{equation}
+	\tag{signature rule}
+	\pair{a}{b}\in\pop{\declare{r}{A}{B}} \Rightarrow\ a \in A \wedge b \in B \label{typing of relations}
+\end{equation}
+
+	The population rule defines the population of a dataset:
+\begin{equation}
+	\tag{population rule}
+	\pairs\ =\ \{ \pair{a}{b}|\ r\in\rels,\ \pair{a}{b}\in\pop{r}\} \label{def. population}
+\end{equation}
+
+	The atoms rule defines all atoms in a dataset:
+\begin{equation}
+	\tag{atoms rule}
+	\atoms\ =\ \{ a |\ \pair{b}{c}\in\pairs,\ a\in\{b,c\}\ \} \label{def atoms}
+\end{equation}
+\subsection{Terms}
+	A term is a combination of relations (from $\rels$), which we use to express constraints from the real world.
+	For this purpose we use the operators from relation algebra~\cite{BrinkKahlSchmidt1997}.
+\begin{definition}[terms]
+\label{def:terms}
+\item   The set of terms, $\terms$, is the smallest set that satisfies, for all $r,s \in \terms$, $d\in\rels$ and $A,B \in \concepts$: 
+\begin{eqnarray*}
+	d&\in&\terms         \quad\quad\text{(every relation is a term)}\\
+	(r \cup s)&\in&\terms\quad\quad\text{(union)}\\
+	(r \cap s)&\in&\terms\quad\quad\text{(intersection)}\\
+	(r-s)&\in&\terms     \quad\quad\text{(difference)}\label{def:difference}\\
+	(r;s)&\in&\terms     \quad\quad\text{(composition)}\\
+	(r\backslash s)&\in&\terms     \quad\quad\text{(right residual)}\\
+	(r\slash s)&\in&\terms     \quad\quad\text{(left residual)}\\
+	\flip r&\in&\terms   \quad\quad\text{(converse)}\\
+	\cmpl{r}&\in&\terms   \quad\quad\text{(complement)}\\
+%	\kleeneplus r&\in&\terms   \quad\quad\text{(Kleene closure)}\\
+%	\kleenestar r&\in&\terms   \quad\quad\text{(Kleene closure)}\\
+	\ident{A}&\in&\terms \quad\quad\text{(identity)}\\
+	\fullt{A}{B}&\in&\terms \quad\quad\text{(full set)}
+\end{eqnarray*}
+\end{definition}
+	Throughout the remainder of this paper,	terms are represented by variables $r$, $s$, $d$, and $t$.
+
+	The relations used in a term $t$ are given by $\relsIn{t}$:
+\begin{definition}[relations in terms]
+	\label{relations in terms}
+	\item   For every $A,B\in\concepts$ and $r,s\in\terms$
+	\begin{eqnarray*}
+		\relsIn{r}		 &=&\{r\}	\\
+		\relsIn{r \cup s}	 &=&\relsIn{r}\cup\relsIn{s}\\
+		\relsIn{r \cap s}	 &=&\relsIn{r}\cup\relsIn{s}\\
+		\relsIn{r-s}	 &=&\relsIn{r}\cup\relsIn{s}\\
+		\relsIn{r;s}	 &=&\relsIn{r}\cup\relsIn{s}\\
+		\relsIn{r\backslash s}	 &=&\relsIn{r}\cup\relsIn{s}\\
+		\relsIn{r\slash s}	 &=&\relsIn{r}\cup\relsIn{s}\\
+		\relsIn{\flip{r}}	 &=&\relsIn{r}\\
+		\relsIn{\cmpl{\declare{r}{A}{B}}}	 &=&\relsIn{r}\\
+	%	\relsIn{\kleeneplus{r}}	 &=&\relsIn{r}\\
+	%	\relsIn{\kleenestar{r}}	 &=&\relsIn{r}\\
+		\relsIn{\ident{A}} 	 &=&\emptyset\\
+		\relsIn{\fullt{A}{B}}&=&\emptyset\\
+	\end{eqnarray*}
+\end{definition}
+	The \define{signature} of a term $r$ is a pair of concepts, given by $\tf{r}$.
+	$\mathfrak{T}$ is a function that maps terms to signatures.
+	This function and its peculiarities are discussed in~\cite{Joosten2015}
+	and have no consequences for the remainder of this paper.
+	(\emph{\@Bas, komen we ermee weg als we de definitie van $\tf{r}$ hier weglaten?})
+
+	Every term $t$ has a set of pairs $\pop{t}$, which we call the population of that term.
+ 	Let $A$ and $B$ be finite sets of atoms, then the population of term is defined as follows:
+\begin{definition}[population of terms]
+\label{population of terms}
+\item   For every $A,B\in\concepts$ and $r,s\in\terms$
+\begin{eqnarray*}
+	\pop{r}		 &=&\{\pair{a}{b}|\ a\ r\ b\}	\\
+	\pop{r \cup s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\pop{r}\ \text{or }\ \pair{a}{b}\in\pop{s}\}	\\
+	\pop{r \cap s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\pop{r}\ \text{and}\ \pair{a}{b}\in\pop{s}\}	\\
+	\pop{r-s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\pop{r}\ \text{and}\ \pair{a}{b}\notin\pop{s}\}	\\
+	\pop{r;s}	 &=&\{\pair{a}{c}|\ \text{for some}\ b,\ \pair{a}{b}\in\pop{r}\ \text{and}\ \pair{b}{c}\in\pop{s}\}	\\
+	\pop{r\backslash s}	 &=&\{\pair{b}{c}|\ \text{for all}\ a,\ \pair{a}{b}\in\pop{r}\ \text{implies}\ \pair{a}{c}\in\pop{s}\}	\\
+	\pop{r\slash s}	 &=&\{\pair{a}{b}|\ \text{for all}\ c,\ \pair{b}{c}\in\pop{s}\ \text{implies}\ \pair{a}{c}\in\pop{r}\}	\\
+	\pop{\flip{r}}	 &=&\{\pair{b}{a}|\ \pair{a}{b}\in\pop{r}    \}	\\
+	\pop{\cmpl{\declare{r}{A}{B}}}	 &=&\fullt{A}{B}-r	\\
+%	\pop{\kleeneplus{r}}	 &=&\text{the smallest set that satisfies }\kleeneplus{r}=r\ \cup\ \kleeneplus{r};\kleeneplus{r}	\\
+%	\pop{\kleenestar{\declare{r}{A}{A}}}	 &=&\ident{A}\cup\kleeneplus{r}\\
+	\pop{\ident{A}} 	 &=&\{\pair{a}{a}|\ a\in A\}	\\
+	\pop{\fullt{A}{B}}&=&\{\pair{a}{b}|\ a\in A, b\in B\}
+\end{eqnarray*}
+\end{definition}
+
+\subsection{Rules}
+	A rule (from set $\rules$) is an equality between two terms that constrains the data stored in relations,
+	to ensure that all rules in the dataset are satisfied.
+	A \define{rule} is a pair of terms $r,s\in\terms$ with the same types ($\tf{r}=\tf{s}$), which we denote as:
+\[\text{RULE}\ r = s\]
+	This means that the constraint \(\pop{r} = \pop{s}\) is satisfied throughout the dataset.
+	In practice, we will also use \(\text{RULE}\ r\subs s\) as a shorthand for \(\text{RULE}\ r\cap s = r\).
+
+	We call a rule $\text{RULE}\ r = s$ \define{satisfied} if \(\pop{r} = \pop{s}\).
+	Each relation that is used in a term of a dataset $\la\rules,\rels,\concepts,\pairs,\atoms\ra$ is an element of $\rels$ as well:
+\begin{equation}
+	\tag{relations in dataset}
+	\{ r|\ \text{RULE}\ t = s\in\rules,\ r\in\relsIn{t}\cup\relsIn{s}\ \}\subseteq\rels\label{def relations in dataset}
+\end{equation}
+
+	The data in a data set $\la\rules,\rels,\concepts,\pairs,\atoms\ra$ are the pairs in the population of all relations in $\rels$.
+	This constitutes the set $\pairs$, also known as the population of the dataset.
+	In response to messages such as
+\[\text{INSERT}\ <3,{\tt Brown}> \text{INTO}\ \declare{r}{Age}{Name}\]
+ 	the population of a dataset changes. 
+	As the population of relations used in $r$ changes with time, the satisfaction of the rule changes accordingly.
+	A software developer, who conceives these rules, must consider how to keep each one of them satisfied.
+	We call a rule \define{violated} if it is not satisfied.
+	The set $\pop{(s-r)\cup(r-s)}$ is called the \emph{violation set} of \(\text{RULE}\ r = s\).
+	To \define{resolve} violations means to change the contents of relations such that the rule is satisfied%
+\footnote{To \define{restore invariance} is sometimes used as a synonym to resolving violations.
+	To \define{preserve invariance} is sometimes used as a synonym to keeping a rule satisfied.}.
+	Consequently, such a rule may also be referred to as \define{invariant}.
+	Each pair in the violation set of a rule is called a violation of that rule.
+
 \section{Bibliography}

### 2022-08-28 2022 Migration/articleMigration.tex
Processing the changes discussed on 2022/08/28

--- 2022-08-27 2022 Migration/articleMigration.tex
@@
 \selectlanguage{english}
-\include{preamramics}
 \hyphenation{ExecEngine}
 \newtheorem{lemma}{Lemma}
-\def\id#1{\text{\it #1\/}}
 \def\Events{{\mathit E}}
 \begin{document}
+\include{preambleMigrations}
 
@@
 \label{sct:Basics}
-	In this section we explain the basic notions to describe migration.
-	We introduce the notion of dataset.
-	A dataset is a quintuple $\la\rules,\rels,\concepts,\pairs,\atoms\ra$,
-	which consists of a set of rules $\rules$, a set of relations $\rels$, a set of concepts $\concepts$, a set of pairs $\pairs$, and a set of atoms $\atoms$.
-	Let us introduce these five notions one by one.
-
+	To describe migration, we first introduce the notion of dataset.
+	A dataset describes a set of data, which is typically stored persistently in a database of some kind.
+	Before defining datasets, we must first define the constituent notions of atom, concept, relation, rule, and role.
+\subsection{Atoms}
 	Atoms are values that have no internal structure, meant to represent data elements in a database.
@@
 	The set of all atoms is called $\atoms$.
-    Each atom is an instance of a \emph{concept}.
 
-	Concepts (from set $\concepts$) are names we use to classify atoms in a meaningful way.
+\subsection{Concepts}
+	Concepts are names we use to classify atoms in a meaningful way.
 	For example, you might choose to classify \atom{Peter} as a \concept{Person}, and \atom{074238991} as a \concept{TelephoneNumber}.
@@
 	We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
-	Specialization is needed to allow statements such as: ``A woman is a human.'' or ``A human is a mammal.''.
+	Specialization is needed to allow statements such as: ``A woman is a human'' or ``A human is a mammal''.
+	The set of all concepts is called $\concepts$.
 
-	Pairs (from set $\pairs$) are used to relate two atoms.
-	A pair $\pair{\atom{Peter}}{\atom{074238991}}$ might bear the meaning that the ``thing'' that \atom{Peter} refers to
+\subsection{Triples}
+	A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean that the ``thing'' that \atom{Peter} refers to
 	has \atom{074238991} as a telephone number.
-	In the formal world, we do not care about that meaning,
-	leaving it entirely up to a user which meaning (s)he attaches to this pair.
+	Triples are used to relate two atoms and a relation.
+	In the formal world, we leave it entirely up to a user to attach meaning to a triple.
+	The practical meaning has no consequences in the formal world.
+	The set of all triples is called $\triples$.
 
-	Relations (from set $\rels$) are used in information systems to store data as pairs.
+\subsection{Relations}
+	Relations are used in information systems to store data.
 	In this paper relations are represented by variables $r$, $s$, and $d$.
-	Every relation has a source concept and a target concept.
-	If we need to be specific about the source and target concepts, we write $\declare{r}{A}{B}$ instead of just $r$.
+
+	Every relation $r$ has a set of pairs, which we call the population of $r$:
+\begin{equation}
+	\pairs{r}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\popR\}
+\end{equation}
+	Every relation has a name, a source concept, and a target concept.
+	We write $\declare{nm}{A}{B}$ to denote a relation with name \id{nm}, source concept $A$, and target concept $B$.
+	% To disassemble a relation in its name, source and target, there exist three functions:
+% \[\begin{array}[3]{rcl}
+	% \id{relname}&:&\rels\rightarrow\id{RelationIdentifier}\\
+	% \id{src}&:&\rels\rightarrow\concepts\\
+	% \id{tgt}&:&\rels\rightarrow\concepts
+% \end{array}\]
+	% Or, stated otherwise, the denotation $\declare{nm}{A}{B}$ stands for a relation with
+	% \(\relname{\declare{nm}{A}{B}} = \id{nm}\),
+	% \(\src{\declare{nm}{A}{B}}=A\), and
+	% \(\tgt{\declare{nm}{A}{B}}=B\).
 	The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
-	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.
+	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concept symbols $A$ and $B$.
 	The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.
-	The meaning of relations is defined by an interpretation function $\mathfrak{I}$.
-	It maps each relation to a set of pairs.
 
-	The typing rule guarantees that each pair in $r$ respects the type of $r$:
-\begin{equation}
-	\tag{typing rule}
-	\pair{a}{b}\in\ti{\declare{r}{A}{B}} \Rightarrow\ a \in A \wedge b \in B \label{typing of relations}
-\end{equation}
+\subsection{Roles}
+	A role is a name that identifies a group of users that get the same privileges in a data set.
+	The set of all roles is called $\roles$.
 
-\begin{equation}
-	\tag{population}
-	\pairs\ =\ \{ \pair{a}{b}|\ r\in\rels,\ \pair{a}{b}\in\ti{r}\} \label{def. population}
-\end{equation}
+\subsection{Constraints}
+	Each constraint $u\in\rules$ has a signature $\signat{u}$, which is a pair of concepts.
+	Each constraint $u$ also has a function $\id{viol}_u:\powerset{\atoms\times\rels\times\atoms}\ \rightarrow\ \powerset{\atoms\times\atoms}$.
+	We call the set $\viol{\popR}{u}$ the \define{violation set} of constraint $u$ in population $\popR$.
+	We call every element $\pair{a}{b}\in\viol{\popR}{u}$ a \define{violation} of constraint $u$ in $\popR$.
+	Note that the signature of $u$ is the same for every population, whereas the violation set may be different for every $\popR$.
 
-\begin{equation}
-	\tag{atoms}
-	\atoms\ =\ \{ a |\ \pair{b}{c}\in\pairs,\ a\in\{b,c\}\ \} \label{def atoms}
-\end{equation}
-
-	A rule is an equality between two terms that constrains the data stored in relations,
-	to ensure that the dataset is at all times consistent with all rules in the dataset ($\rules$).
-	Terms are used to combine relations using operators.
-	The set of terms is called $\terms$.
-\begin{definition}[terms]
-\label{def:terms}
-\item   The set of terms, $\terms$, is the smallest set that satisfies, for all $r,s \in \terms$, $d\in\rels$ and $A,B \in \concepts$: 
-\begin{eqnarray}
-	d&\in&\terms         \quad\quad\text{(every relation is a term)}\\
-	(r \cup s)&\in&\terms\quad\quad\text{(union)}\\
-	(r \cap s)&\in&\terms\quad\quad\text{(intersection)}\\
-	(r-s)&\in&\terms     \quad\quad\text{(difference)}\label{def:difference}\\
-	(r;s)&\in&\terms     \quad\quad\text{(composition)}\\
-	(r\backslash s)&\in&\terms     \quad\quad\text{(right residual)}\\
-	(r\slash s)&\in&\terms     \quad\quad\text{(left residual)}\\
-	\flip r&\in&\terms   \quad\quad\text{(converse)}\\
-	\cmpl{r}&\in&\terms   \quad\quad\text{(complement)}\\
-%	\kleeneplus r&\in&\terms   \quad\quad\text{(Kleene closure)}\\
-%	\kleenestar r&\in&\terms   \quad\quad\text{(Kleene closure)}\\
-	\ident{A}&\in&\terms \quad\quad\text{(identity)}\\
-	\fullt{A}{B}&\in&\terms \quad\quad\text{(full set)}
-\end{eqnarray}
+\subsection{Datasets}
+	We can now define the notion of dataset.
+	Let
+	$\concepts$ be a set of concept symbols,
+	$\rels$ be a set of relation symbols,
+	$\roles$ be a set of roles,
+	$\rules$ be a set of constraints, and
+	$\atoms$ be a set of atoms.
+\begin{definition}[dataset]
+	\label{def:dataset}
+	\item A dataset is a tuple $\la\concepts,\rels,\roles,\rules,\maintain,\atoms,\popR,\popC\ra$ with
+	\[\begin{array}[3]{rcl}
+		\id{maintain}&:&\roles\times\rules\\
+		\id{popR}&:&\atoms\times\rels\times\atoms\\
+		\id{popC}&:&\concepts\times\atoms\\
+		\multicolumn{3}{l}{\triple{a}{r}{b}\in\popR\ \Rightarrow\ \pair{\src{r}}{a}\in\popC \wedge \pair{\tgt{r}}{b}\in\popC}\\
+		\forall a\in\atoms\ \exists A\in\concepts&:&\pair{A}{a}\in\popC
+	\end{array}\]
 \end{definition}
-	Throughout the remainder of this paper,	terms are represented by variables $r$, $s$, $d$, and $t$.
-	The \define{type} of a term $r$ is a pair of concepts, given by $\tf{r}$.
-	$\mathfrak{T}$ is a partial function that maps terms to types.
-	The type function and its restrictions are discussed in~\cite{Joosten2015}
-	and have no consequences for the remainder of this paper.
-	(\emph{\@Bas, komen we ermee weg als we de definitie van $\tf{r}$ hier weglaten?})
-
- 	The meaning of terms is given by an interpretation function $\mathfrak{I}$.
-	Let $A$ and $B$ be finite sets of atoms, then $\mathfrak{I}$ maps each term to the set of pairs for which that term stands.
-\begin{definition}[interpretation of terms]
-\label{interpretation of terms}
-\item   For every $A,B\in\concepts$ and $r,s\in\terms$
-\begin{eqnarray}
-	\ti{r}		 &=&\{\pair{a}{b}|\ a\ r\ b\}	\\
-	\ti{r \cup s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{or }\ \pair{a}{b}\in\ti{s}\}	\\
-	\ti{r \cap s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{a}{b}\in\ti{s}\}	\\
-	\ti{r-s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{a}{b}\notin\ti{s}\}	\\
-	\ti{r;s}	 &=&\{\pair{a}{c}|\ \text{for some}\ b,\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{b}{c}\in\ti{s}\}	\\
-	\ti{r\backslash s}	 &=&\{\pair{b}{c}|\ \text{for all}\ a,\ \pair{a}{b}\in\ti{r}\ \text{implies}\ \pair{a}{c}\in\ti{s}\}	\\
-	\ti{r\slash s}	 &=&\{\pair{a}{b}|\ \text{for all}\ c,\ \pair{b}{c}\in\ti{s}\ \text{implies}\ \pair{a}{c}\in\ti{r}\}	\\
-	\ti{\flip{r}}	 &=&\{\pair{b}{a}|\ \pair{a}{b}\in\ti{r}    \}	\\
-	\ti{\cmpl{\declare{r}{A}{B}}}	 &=&\fullt{A}{B}-r	\\
-%	\ti{\kleeneplus{r}}	 &=&\text{the smallest set that satisfies }\kleeneplus{r}=r\ \cup\ \kleeneplus{r};\kleeneplus{r}	\\
-%	\ti{\kleenestar{\declare{r}{A}{A}}}	 &=&\ident{A}\cup\kleeneplus{r}\\
-	\ti{\ident{A}} 	 &=&\{\pair{a}{a}|\ a\in A\}	\\
-	\ti{\fullt{A}{B}}&=&\{\pair{a}{b}|\ a\in A, b\in B\}
-\end{eqnarray}
-\end{definition}
-%	The Kleene closure operators (postfix $\kleeneplus{\ }$ and $\kleenestar{\ }$) have been implemented partially in the current Ampersand implementation.
-
-	A \define{rule} is a pair of terms $r,s\in\terms$ with the same types ($\tf{r}=\tf{s}$).
-\[\text{RULE}\ r = s\]
-	This means \(\ti{r} = \ti{s}\). In practice, many rules are written as:
-\[\text{RULE}\ r\subs s\]
-	This is a shorthand for 
-\[\text{RULE}\ r\cap s = r\]
-	We have enhanced the type function $\mathfrak{T}$ to cover rules as well.
-	If $\tf{r}=\tf{s}$ and $\tf{s}=\pair{A}{B}$:
-\begin{eqnarray}
-	\tf{\text{RULE}\ r = s}   &=&\pair{A}{B}\\
-	\tf{\text{RULE}\ r\subs s}&=&\pair{A}{B}
- \end{eqnarray}
-
-	We call a rule $\text{RULE}\ r = s$ \define{satisfied} if \(\ti{r} = \ti{s}\).
-
-	The data in a data set $\la\rules,\rels,\concepts,\pairs,\atoms\ra$ are the pairs in the interpretation of all relations in $\rels$.
-	This constitutes the set $\pairs$, also known as the population of the dataset.
-	In response to messages such as
-\[\text{INSERT}\ <3,{\tt Brown}> \text{INTO}\ \declare{r}{Age}{Name}\]
- 	the population of a dataset changes. 
-	As the population of relations used in $r$ changes with time, the satisfaction of the rule changes accordingly.
-	A software developer, who conceives these rules, must consider how to keep each one of them satisfied.
-	We call a rule \define{violated} if it is not satisfied.
-	The set $\ti{(s-r)\cup(r-s)}$ is called the \emph{violation set} of \(\text{RULE}\ r = s\).
-	To \define{resolve} violations means to change the contents of relations such that the rule is satisfied%
-\footnote{To \define{restore invariance} is sometimes used as a synonym to resolving violations.
-	To \define{preserve invariance} is sometimes used as a synonym to keeping a rule satisfied.}.
-	Consequently, such a rule may also be referred to as \define{invariant}.
-	Each pair in the violation set of a rule is called a violation of that rule.
 

### 2022-08-29 2022 Migration/articleMigration.tex
WIP on datasets

@@
 \label{sct:Basics}
-	To describe migration, we first introduce the notion of dataset.
-	A dataset describes a set of data, which is typically stored persistently in a database of some kind.
+	To define migration, let us first introduce the notion of dataset.
+	A dataset describes a set of structured data, which is typically stored persistently in a database of some kind.
+	To abstract away from a particular database implementation,
+	we represent data elements as triples.
+	This makes our theory valid for any kind of database that triple stores can represent,
+	such as SQL databases, graph databases and other no-SQL databases
+
 	Before defining datasets, we must first define the constituent notions of atom, concept, relation, rule, and role.
@@
 	By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ are used to represent \emph{atoms}.
-	The set of all atoms is called $\atoms$.
+	The set of atoms is called $\atoms$.
 
@@
 	The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.
-	The expression $a \in A$ means that atom $a$ is an \emph{instance} of concept $A$.
+	The expression $a\ \inst\ A$ means that atom $a$ is an \emph{instance} of concept $A$.
 	The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
@@
 	Specialization is needed to allow statements such as: ``A woman is a human'' or ``A human is a mammal''.
-	The set of all concepts is called $\concepts$.
+	The set of concepts is called $\concepts$.
 
@@
 	The practical meaning has no consequences in the formal world.
-	The set of all triples is called $\triples$.
+	The set of triples is called $\triples$.
 
 \subsection{Relations}
-	Relations are used in information systems to store data.
+	Relations are used in datasets to store data.
 	In this paper relations are represented by variables $r$, $s$, and $d$.
@@
 \begin{equation}
-	\pairs{r}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\popR\}
+	\pop{r}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\triples\}
 \end{equation}
@@
 	% \(\tgt{\declare{nm}{A}{B}}=B\).
+
 	The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
@@
 	The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.
+	The set of relations is called $\rels$.
 
-\subsection{Roles}
-	A role is a name that identifies a group of users that get the same privileges in a data set.
-	The set of all roles is called $\roles$.
+% \subsection{Roles}
+% 	A role is a name that identifies a group of users that get the same privileges in a data set.
+% 	The set of roles is called $\roles$.
 
@@
 	Each constraint $u$ also has a function $\id{viol}_u:\powerset{\atoms\times\rels\times\atoms}\ \rightarrow\ \powerset{\atoms\times\atoms}$.
-	We call the set $\viol{\popR}{u}$ the \define{violation set} of constraint $u$ in population $\popR$.
-	We call every element $\pair{a}{b}\in\viol{\popR}{u}$ a \define{violation} of constraint $u$ in $\popR$.
-	Note that the signature of $u$ is the same for every population, whereas the violation set may be different for every $\popR$.
+	We call the set $\viol{\triples}{u}$ the \define{violation set} of constraint $u$ in population $\triples$.
+	We call every element $\pair{a}{b}\in\viol{\triples}{u}$ a \define{violation} of constraint $u$ in $\triples$.
+	Note that the signature of $u$ does not depend on the population, whereas the violation set of $u$ may be different for every $\triples$.
+	The set of constraints is called $\rules$.
 
@@
 	We can now define the notion of dataset.
-	Let
-	$\concepts$ be a set of concept symbols,
-	$\rels$ be a set of relation symbols,
-	$\roles$ be a set of roles,
-	$\rules$ be a set of constraints, and
-	$\atoms$ be a set of atoms.
+	Let	$\concepts$ be a set of concept symbols,
+	$\rels$ be a set of relations,
+	$\atoms$ be a set of atoms,
+	let $\inst$ be a relation $\atoms\times\concepts$ that represents the instance relation between concepts and atoms,
+	% let $\roles$ be a set of roles,
+	and $\rules$ be a set of constraints.
 \begin{definition}[dataset]
 	\label{def:dataset}
-	\item A dataset is a tuple $\la\concepts,\rels,\roles,\rules,\maintain,\atoms,\popR,\popC\ra$ with
+	\item A dataset is a tuple $\la\concepts,\rels,\atoms,\inst,\triples,\rules\ra$ with
 	\[\begin{array}[3]{rcl}
-		\id{maintain}&:&\roles\times\rules\\
-		\id{popR}&:&\atoms\times\rels\times\atoms\\
-		\id{popC}&:&\concepts\times\atoms\\
-		\multicolumn{3}{l}{\triple{a}{r}{b}\in\popR\ \Rightarrow\ \pair{\src{r}}{a}\in\popC \wedge \pair{\tgt{r}}{b}\in\popC}\\
-		\forall a\in\atoms\ \exists A\in\concepts&:&\pair{A}{a}\in\popC
+		% \id{maintain}&:&\roles\times\rules\\
+		\forall\triple{a}{r}{b}\in\triples&:&a\ \inst\ \src{r}\ \wedge\ b\ \inst\ \tgt{r}\\
+		\forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\\
+		\forall u\in\rules&:&\viol{\triples}{u}=\emptyset
 	\end{array}\]

### 2022-09-03 2022 Migration/articleMigration.tex
tussentijdse push

@@
 \label{sct:Basics}
-	To define migration, let us first introduce the notion of dataset.
+	To define migration, we must introduce the notion of dataset because migration aims to preserve data whilst changing the software that manipulates data.
 	A dataset describes a set of structured data, which is typically stored persistently in a database of some kind.
@@
 	we represent data elements as triples.
-	This makes our theory valid for any kind of database that triple stores can represent,
-	such as SQL databases, graph databases and other no-SQL databases
+	This makes our theory valid for any kind of database that triples can represent,
+	such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases
 
@@
 	In the formal world, we leave it entirely up to a user to attach meaning to a triple.
-	The practical meaning has no consequences in the formal world.
+	This ``meaning from practice'' has no consequences in the formal world.
 	The set of triples is called $\triples$.
@@
 \subsection{Relations}
-	Relations are used in datasets to store data.
+	Relations in datasets are used to store data.
 	In this paper relations are represented by variables $r$, $s$, and $d$.
 
-	Every relation $r$ has a set of pairs, which we call the population of $r$:
+	Every relation $r$ contains a set of pairs, which we call the population of $r$:
 \begin{equation}
@@
 	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concept symbols $A$ and $B$.
+	It contains all pairs that can be made of elements of $A$ and $B$:
+\[\pop{\fullt{A}{B}} = \{ \pair{a}{b}| a\ \inst\ A, \}\]
 	The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.
@@
 	\label{def:dataset}
-	\item A dataset is a tuple $\la\concepts,\rels,\atoms,\inst,\triples,\rules\ra$ with
+	\item A dataset is a tuple $\la\concepts,\isa,\rels,\atoms,\inst,\triples,\rules\ra$ with
 	\[\begin{array}[3]{rcl}
 		% \id{maintain}&:&\roles\times\rules\\
+		\isa\text{\ is a partial order}\\
 		\forall\triple{a}{r}{b}\in\triples&:&a\ \inst\ \src{r}\ \wedge\ b\ \inst\ \tgt{r}\\
@@
 
+\subsection{Motivations}
+	We have already motivated the use of triples in a dataset:
+	It is a generic way of representing structured data, which maps to most (if not all) existing database technologies.
+	Changes in the data means changing the set of triples, adding or removing triples from the set.
+	That is not migration; that is regular operational behaviour.
+	If the software changes, this means that the concepts, relations, and constraints change as well.
+	However, a dataset can contain atoms that do not partake in any triple.
+	That is why we need a set of atoms, $\atoms$ too.
+
+	Atoms serve as data elements. Changes in the dataset are reflected in $\atoms$.
+	Concepts serve to classify atoms in a partial order (the classification) so we can implement a typing regime on the dataset.
+	This type system ensures that the dataset can be interpreted as a meaningful database.
+	As a consequence, the relations $\isa$ and $\inst$
+
+	The choice of elements in the tuple that represents a dataset is determined by the dynamics of a dataset.
+	Anything that can change needs to be in that tuple.
+	It is not only the data that changes. Rules can change, relations and concepts can change, the taxonomy can change,
+	rules can change.
+	In migrations we want to cope with any conceivable change,
+	ensuring that the dataset satisfies its defining properties (from definition~\ref{def:dataset}).
+	The migration must also ensure that the population remains intact as much as possible.
+
 \section{Bibliography}

### 2022-09-04 2022 Migration/articleMigration.tex
On what needs to be done during migration

@@
 
+\section{Migration steps}
+	A smooth migration from an old system to a new system would go as follows:
+	Launch the new system in parallel to the old, copy data from the old to the new system, and have everyone use the new system.
+	We identify the following possible issues in this approach:
+
+	\begin{enumerate}
+	\item Data required in the new system is missing in the old system.
+	There may be no way in the old system to enter that data.
+	An example could be that every reimbursement form needs to have an address associated to it to mail the check to, but address information is not stored in the old system:
+	The old system required the reimbursement office to look up employee's addresses from a hand-written list they had on their desk.
+	\item Data in the old system is wrong but cannot be corrected there due to how the old system was designed.
+	An example is if the old system only allows approvals to be entered as the current user, and the CEO insists that her administrative staff enters his approvals into the system for her.
+	This may result in approvals being entered as admin staff, where it was really the CEO making the approval.
+	\item Data in the old system does not satisfy invariants of the new system.
+	There may be no way in the old system of making the data satisfy those invariants.
+We can use the same example as in the previous bullet point, but add the requirement (in the new system) that every purchase above a certain amount needs to be approved by the CEO.
+	\item A way of entering data into the old system is missing in the new system.
+	People or automated processes might rely on these ways of entering data.
+	An example could be that when employees turned their computers on or off, an ad-hoc script would automatically clock them in- and out to determine the number of hours they worked.
+	A handful of employees still relies on this.
+	\item Data present in the old system cannot be stored in the new system.
+	An example could be that references to physical locations where original receipts are kept are stored in the old system, but the new system relies on scans of receipts and allows the originals to be destroyed or not submitted.
+	Until the original receipts are scanned, the old data should be kept.
+	\end{enumerate}
+
+	To mitigate these issues, we:
+
+	\begin{enumerate}
+	\item Allow `missing' data-points requirements to be ignored during migration.
+	\item Allow data-points to be migrated while being marked as needing correction.
+	\item Allow invariants in the new system to be ignored for certain data-points during migration.
+	\item Allow continued use of interfaces of the old system, data entered into the old system via those interfaces needs to be continuously copied to the new system.
+	\item Retain data in the old system until it can be marked as ready to be phased out.
+	\end{enumerate}
+
 \section{Bibliography}

### 2022-09-04 2022 Migration/articleMigration.tex
WIP

@@
 \label{sct:Basics}
-	To define migration, we must introduce the notion of dataset because migration aims to preserve data whilst changing the software that manipulates data.
+	To define migration, we must introduce the notion of dataset
+	because migration aims to preserve data while changing the software that manipulates data.
 	A dataset describes a set of structured data, which is typically stored persistently in a database of some kind.
@@
 \subsection{Atoms}
-	Atoms are values that have no internal structure, meant to represent data elements in a database.
+	Atoms serve as data elements.
+	They are values without internal structure of interest, meant to represent data elements in a database.
 	From a business perspective, atoms are used to represent concrete items of the world,
@@
 \subsection{Triples}
+	Triples are used to represent data.
+	Each triple relates two atoms to a relation.
 	A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean that the ``thing'' that \atom{Peter} refers to
 	has \atom{074238991} as a telephone number.
-	Triples are used to relate two atoms and a relation.
 	In the formal world, we leave it entirely up to a user to attach meaning to a triple.
@@
 	It contains all pairs that can be made of elements of $A$ and $B$:
-\[\pop{\fullt{A}{B}} = \{ \pair{a}{b}| a\ \inst\ A, \}\]
+\[\pop{\fullt{A}{B}} = \{ \pair{a}{b}| a\ \inst\ A,\ b\ \inst\ B \}\]
 	The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.
@@
 \subsection{Constraints}
+	Constraints are the invariants of a dataset: restrictions that are kept true at all times.
+	Their purpose is to ensure the semantics of a dataset, i.e. the preservation of meaning as data changes.
 	Each constraint $u\in\rules$ has a signature $\signat{u}$, which is a pair of concepts.
@@
 	Let	$\concepts$ be a set of concept symbols,
-	$\rels$ be a set of relations,
-	$\atoms$ be a set of atoms,
+	let $\isa$ be a partial order on $\concepts$,
+	let $\rels$ be a set of relations,
+	let $\atoms$ be a set of atoms,
 	let $\inst$ be a relation $\atoms\times\concepts$ that represents the instance relation between concepts and atoms,
 	% let $\roles$ be a set of roles,
-	and $\rules$ be a set of constraints.
+	and let $\rules$ be a set of constraints.
 \begin{definition}[dataset]
@@
 	\item A dataset is a tuple $\la\concepts,\isa,\rels,\atoms,\inst,\triples,\rules\ra$ with
-	\[\begin{array}[3]{rcl}
+	\begin{eqnarray}
+		\forall r\in\rels&:&\src{r}\in\concepts\ \wedge\ \tgt{r}\in\concepts\\
+		\forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\\
 		% \id{maintain}&:&\roles\times\rules\\
-		\isa\text{\ is a partial order}\\
-		\forall\triple{a}{r}{b}\in\triples&:&a\ \inst\ \src{r}\ \wedge\ b\ \inst\ \tgt{r}\\
-		\forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\\
-		\forall u\in\rules&:&\viol{\triples}{u}=\emptyset
-	\end{array}\]
+		\forall A,B\in\concepts, a\in\atoms&:&A\isa B\wedge a\ \inst\ A\ \Rightarrow\ a\ \inst\ B\\
+		\forall\triple{a}{r}{b}\in\triples&:&r\in\rels\ \wedge\ a\ \inst\ \src{r}\ \wedge\ b\ \inst\ \tgt{r}\\
+		\forall u\in\rules&:&\viol{\triples}{u}=\emptyset\label{violation free}
+	\end{eqnarray}
 \end{definition}
 
-\subsection{Motivations}
-	We have already motivated the use of triples in a dataset:
-	It is a generic way of representing structured data, which maps to most (if not all) existing database technologies.
-	Changes in the data means changing the set of triples, adding or removing triples from the set.
-	That is not migration; that is regular operational behaviour.
-	If the software changes, this means that the concepts, relations, and constraints change as well.
-	However, a dataset can contain atoms that do not partake in any triple.
-	That is why we need a set of atoms, $\atoms$ too.
-
-	Atoms serve as data elements. Changes in the dataset are reflected in $\atoms$.
-	Concepts serve to classify atoms in a partial order (the classification) so we can implement a typing regime on the dataset.
-	This type system ensures that the dataset can be interpreted as a meaningful database.
-	As a consequence, the relations $\isa$ and $\inst$
-
-	The choice of elements in the tuple that represents a dataset is determined by the dynamics of a dataset.
-	Anything that can change needs to be in that tuple.
-	It is not only the data that changes. Rules can change, relations and concepts can change, the taxonomy can change,
-	rules can change.
-	In migrations we want to cope with any conceivable change,
-	ensuring that the dataset satisfies its defining properties (from definition~\ref{def:dataset}).
-	The migration must also ensure that the population remains intact as much as possible.
+\subsection{Motivation}
+% The purpose of this section is to explain why a dataset is structured the way it is.
+	The challenge of migration is to preserve meaning rather than just copy data from one place to another.
+	The bearer of meaning to users of a dataset is the data.
+	That is represented by the population, i.e. the sets of triples and atoms, $\triples$ and $\atoms$.
+	How a user interprets the data is up to the user and may therefore have no consequence to the dataset.
+	So we attach no intrinsic meaning to a triple and take $\triples$ and $\atoms$ for granted.
+	In a migration we seek to preserve that data where we can.
+	The other elements in a dataset, $\isa$, $\concepts$, $\rels$, and $\rules$ do have consequences, however.
+	Definition~\ref{def:dataset} implies that changes to data may cause violations of rules,
+	so measures must be taken to get rid of those violations (eqn.~\ref{violation free}).
+	Changes to $\isa$, $\concepts$, $\rels$ change the data structure in a dataset.
+	Changes to $\rules$ change the invariants of the dataset.
+	This paper studies those changes and seeks to preserve the rules in definition~\ref{def:dataset} in such a way that large portions of migration can be automated.
 

### 2022-09-04 2022 Migration/articleMigration.tex
First attempt at defining information system.

@@
 \label{sct:Introduction}
-
-\section{Basics}
-\label{sct:Basics}
-	To define migration, we must introduce the notion of dataset
-	because migration aims to preserve data while changing the software that manipulates data.
-	A dataset describes a set of structured data, which is typically stored persistently in a database of some kind.
-	To abstract away from a particular database implementation,
-	we represent data elements as triples.
-	This makes our theory valid for any kind of database that triples can represent,
-	such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases
-
-	Before defining datasets, we must first define the constituent notions of atom, concept, relation, rule, and role.
-\subsection{Atoms}
-	Atoms serve as data elements.
-	They are values without internal structure of interest, meant to represent data elements in a database.
-	From a business perspective, atoms are used to represent concrete items of the world,
-	such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
-	By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ are used to represent \emph{atoms}.
-	The set of atoms is called $\atoms$.
-
-\subsection{Concepts}
-	Concepts are names we use to classify atoms in a meaningful way.
-	For example, you might choose to classify \atom{Peter} as a \concept{Person}, and \atom{074238991} as a \concept{TelephoneNumber}.
-	In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
-    We will use variables $A$, $B$, $C$, $D$ to represent concepts.
-	The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.
-	The expression $a\ \inst\ A$ means that atom $a$ is an \emph{instance} of concept $A$.
-	The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
-	We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
-	Specialization is needed to allow statements such as: ``A woman is a human'' or ``A human is a mammal''.
-	The set of concepts is called $\concepts$.
-
-\subsection{Triples}
-	Triples are used to represent data.
-	Each triple relates two atoms to a relation.
-	A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean that the ``thing'' that \atom{Peter} refers to
-	has \atom{074238991} as a telephone number.
-	In the formal world, we leave it entirely up to a user to attach meaning to a triple.
-	This ``meaning from practice'' has no consequences in the formal world.
-	The set of triples is called $\triples$.
-
-\subsection{Relations}
-	Relations in datasets are used to store data.
-	In this paper relations are represented by variables $r$, $s$, and $d$.
-
-	Every relation $r$ contains a set of pairs, which we call the population of $r$:
-\begin{equation}
-	\pop{r}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\triples\}
-\end{equation}
-	Every relation has a name, a source concept, and a target concept.
-	We write $\declare{nm}{A}{B}$ to denote a relation with name \id{nm}, source concept $A$, and target concept $B$.
-	% To disassemble a relation in its name, source and target, there exist three functions:
-% \[\begin{array}[3]{rcl}
-	% \id{relname}&:&\rels\rightarrow\id{RelationIdentifier}\\
-	% \id{src}&:&\rels\rightarrow\concepts\\
-	% \id{tgt}&:&\rels\rightarrow\concepts
-% \end{array}\]
-	% Or, stated otherwise, the denotation $\declare{nm}{A}{B}$ stands for a relation with
-	% \(\relname{\declare{nm}{A}{B}} = \id{nm}\),
-	% \(\src{\declare{nm}{A}{B}}=A\), and
-	% \(\tgt{\declare{nm}{A}{B}}=B\).
-
-	The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
-	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concept symbols $A$ and $B$.
-	It contains all pairs that can be made of elements of $A$ and $B$:
-\[\pop{\fullt{A}{B}} = \{ \pair{a}{b}| a\ \inst\ A,\ b\ \inst\ B \}\]
-	The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.
-	The set of relations is called $\rels$.
-
-% \subsection{Roles}
-% 	A role is a name that identifies a group of users that get the same privileges in a data set.
-% 	The set of roles is called $\roles$.
-
-\subsection{Constraints}
-	Constraints are the invariants of a dataset: restrictions that are kept true at all times.
-	Their purpose is to ensure the semantics of a dataset, i.e. the preservation of meaning as data changes.
-	Each constraint $u\in\rules$ has a signature $\signat{u}$, which is a pair of concepts.
-	Each constraint $u$ also has a function $\id{viol}_u:\powerset{\atoms\times\rels\times\atoms}\ \rightarrow\ \powerset{\atoms\times\atoms}$.
-	We call the set $\viol{\triples}{u}$ the \define{violation set} of constraint $u$ in population $\triples$.
-	We call every element $\pair{a}{b}\in\viol{\triples}{u}$ a \define{violation} of constraint $u$ in $\triples$.
-	Note that the signature of $u$ does not depend on the population, whereas the violation set of $u$ may be different for every $\triples$.
-	The set of constraints is called $\rules$.
-
-\subsection{Datasets}
-	We can now define the notion of dataset.
-	Let	$\concepts$ be a set of concept symbols,
-	let $\isa$ be a partial order on $\concepts$,
-	let $\rels$ be a set of relations,
-	let $\atoms$ be a set of atoms,
-	let $\inst$ be a relation $\atoms\times\concepts$ that represents the instance relation between concepts and atoms,
-	% let $\roles$ be a set of roles,
-	and let $\rules$ be a set of constraints.
-\begin{definition}[dataset]
-	\label{def:dataset}
-	\item A dataset is a tuple $\la\concepts,\isa,\rels,\atoms,\inst,\triples,\rules\ra$ with
-	\begin{eqnarray}
-		\forall r\in\rels&:&\src{r}\in\concepts\ \wedge\ \tgt{r}\in\concepts\\
-		\forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\\
-		% \id{maintain}&:&\roles\times\rules\\
-		\forall A,B\in\concepts, a\in\atoms&:&A\isa B\wedge a\ \inst\ A\ \Rightarrow\ a\ \inst\ B\\
-		\forall\triple{a}{r}{b}\in\triples&:&r\in\rels\ \wedge\ a\ \inst\ \src{r}\ \wedge\ b\ \inst\ \tgt{r}\\
-		\forall u\in\rules&:&\viol{\triples}{u}=\emptyset\label{violation free}
-	\end{eqnarray}
-\end{definition}
-
-\subsection{Motivation}
-% The purpose of this section is to explain why a dataset is structured the way it is.
-	The challenge of migration is to preserve meaning rather than just copy data from one place to another.
-	The bearer of meaning to users of a dataset is the data.
-	That is represented by the population, i.e. the sets of triples and atoms, $\triples$ and $\atoms$.
-	How a user interprets the data is up to the user and may therefore have no consequence to the dataset.
-	So we attach no intrinsic meaning to a triple and take $\triples$ and $\atoms$ for granted.
-	In a migration we seek to preserve that data where we can.
-	The other elements in a dataset, $\isa$, $\concepts$, $\rels$, and $\rules$ do have consequences, however.
-	Definition~\ref{def:dataset} implies that changes to data may cause violations of rules,
-	so measures must be taken to get rid of those violations (eqn.~\ref{violation free}).
-	Changes to $\isa$, $\concepts$, $\rels$ change the data structure in a dataset.
-	Changes to $\rules$ change the invariants of the dataset.
-	This paper studies those changes and seeks to preserve the rules in definition~\ref{def:dataset} in such a way that large portions of migration can be automated.
+	The challenge of migration is to preserve meaning rather than just copy the data from one place to another.
 
@@
 	\begin{enumerate}
-	\item Allow `missing' data-points requirements to be ignored during migration.
-	\item Allow data-points to be migrated while being marked as needing correction.
-	\item Allow invariants in the new system to be ignored for certain data-points during migration.
+	\item Allow `missing' triples requirements to be ignored during migration.
+	\item Allow triples to be migrated while being marked as needing correction.
+	\item Allow invariants in the new system to be ignored for certain triples during migration.
 	\item Allow continued use of interfaces of the old system, data entered into the old system via those interfaces needs to be continuously copied to the new system.
@@
 	\end{enumerate}
+
+\section{Terminology}
+\label{sct:Terminology}
+	Migration is sometimes called data migration, because most people understand that data from "the old system" is fed into "the new system".
+	The idea is that the new system lets the business continu without interruption, after its software has been upgraded.
+	Since data is so central to migration, we start by introducing the notion of dataset.
+	To abstract away from a particular database implementation,
+	we represent data elements as triples.
+	This makes our theory valid for any kind of database that triples can represent,
+	such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases
+
+	Migration is not just about porting data.
+	It is about preserving the meaning of that data.
+	For that purpose, section~\ref{sct:Information Systems} introduces the notion of information system.
+
+\subsection{Datasets}
+	A dataset describes a set of structured data, which is typically stored persistently in a database of some kind.
+
+	Before defining datasets, we must first define the constituent notions of atom, concept, triple, and relation.
+
+	Atoms serve as data elements.
+	They are values without internal structure of interest, meant to represent data elements in a database.
+	From a business perspective, atoms are used to represent concrete items of the world,
+	such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
+	By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ are used to represent \emph{atoms}.
+	The set of atoms is called $\atoms$.
+
+	Concepts are names we use in a conceptual model during the design of a dataset.
+	They are also used to assign signatures to relations, to enable type checking on the dataset.
+	For example, you might choose to classify \atom{Peter} as a \concept{Person}, and \atom{074238991} as a \concept{TelephoneNumber}.
+	In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
+    We will use variables $A$, $B$, $C$, $D$ to represent concepts.
+	The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.
+	The expression $a\ \inst\ A$ means that atom $a$ is an \emph{instance} of concept $A$.
+	The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
+	We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
+	Specialization is needed to allow statements such as: ``A woman is a human'' or ``A human is a mammal''.
+	The set of concepts is called $\concepts$.
+
+	Triples are used to represent data.
+	Each triple relates two atoms to a relation.
+	A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean that the ``thing'' that \atom{Peter} refers to
+	has \atom{074238991} as a telephone number.
+	In the formal world, we leave it entirely up to a user to attach meaning to a triple.
+	This ``meaning from practice'' has no consequences in the formal world.
+	The set of triples is called $\triples$.
+
+	Relations in datasets are used to store data.
+	In this paper relations are represented by variables $r$, $s$, and $d$.
+
+	Every relation $r$ contains a set of pairs, which we call the population of $r$:
+\begin{equation}
+	\pop{r}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\triples\}
+\end{equation}
+	Every relation has a name, a source concept, and a target concept.
+	We write $\declare{nm}{A}{B}$ to denote a relation with name \id{nm}, source concept $A$, and target concept $B$.
+	% To disassemble a relation in its name, source and target, there exist three functions:
+% \[\begin{array}[3]{rcl}
+	% \id{relname}&:&\rels\rightarrow\id{RelationIdentifier}\\
+	% \id{src}&:&\rels\rightarrow\concepts\\
+	% \id{tgt}&:&\rels\rightarrow\concepts
+% \end{array}\]
+	% Or, stated otherwise, the denotation $\declare{nm}{A}{B}$ stands for a relation with
+	% \(\relname{\declare{nm}{A}{B}} = \id{nm}\),
+	% \(\src{\declare{nm}{A}{B}}=A\), and
+	% \(\tgt{\declare{nm}{A}{B}}=B\).
+
+	The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
+	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concept symbols $A$ and $B$.
+	It contains all pairs that can be made of elements of $A$ and $B$:
+\[\pop{\fullt{A}{B}} = \{ \pair{a}{b}| a\ \inst\ A,\ b\ \inst\ B \}\]
+	The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.
+	The set of relations is called $\rels$.
+
+	We can now define the notion of dataset.
+	Let	$\concepts$ be a set of concept symbols,
+	let $\isa$ be a partial order on $\concepts$,
+	let $\rels$ be a set of relations,
+	let $\atoms$ be a set of atoms,
+	and let $\inst$ be a relation $\atoms\times\concepts$ that represents the instance relation between concepts and atoms,
+\begin{definition}[dataset]
+	\label{def:dataset}
+	\item A dataset is a tuple $\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$ with
+	\begin{eqnarray}
+		\forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\\
+		\forall A,B\in\concepts, a\in\atoms&:&A\isa B\wedge a\ \inst\ A\ \Rightarrow\ a\ \inst\ B\\
+		\forall r\in\rels&:&\src{r}\in\concepts\ \wedge\ \tgt{r}\in\concepts\\
+		\forall\triple{a}{r}{b}\in\triples&:&r\in\rels\ \wedge\ a\ \inst\ \src{r}\ \wedge\ b\ \inst\ \tgt{r}\\
+	\end{eqnarray}
+\end{definition}
+
+\subsection{Information Systems}
+\label{sct:Information Systems}
+	Information systems process information from many different users, each with their own tasks and responsibilities.
+	Users work from different locations and on different moments.
+	This collective use serves a purpose which we loosely call "the business".
+	As a consequence, the data in an information system changes continually.
+	To preserve meaning, users are trying to maintain semantic constraints on the data.
+	Before defining information systems for the purpose of data migration,
+	let us first define the constituent notions role and constraint, which are part of that definition.
+
+	A \define{role} is a name that identifies a group of users.
+	The purpose of a role is to mention an individual user without knowing who that user is.
+	In the sequel, when we talk about a role we actually mean a user (an arbitrary one) who fulfills that role in the information system.
+	The set of roles is called $\roles$.
+
+	A \define{constraint} is a restriction on a dataset, to be enforced by a role.
+	The purpose of constraints is to ensure the semantics of a dataset, i.e. the preservation of meaning as data changes.
+	Each constraint $u\in\rules$ has a signature $\signat{u}$, which is a pair of concepts.
+	Each constraint $u$ also has a function $\id{viol}_u:\powerset{\atoms\times\rels\times\atoms}\ \rightarrow\ \powerset{\atoms\times\atoms}$.
+	We call the set $\viol{\triples}{u}$ the \define{violation set} of constraint $u$ in population $\triples$.
+	We call every element $\pair{a}{b}\in\viol{\triples}{u}$ a \define{violation} of constraint $u$ in $\triples$.
+	We say that a constraint \define{is satisfied} if its violation set is empty (i.e. if $\viol{\triples}{u}$ is the empty set).
+	The set of constraints is called $\rules$.
+
+	Note that the signature of $u$ does not depend on the population, whereas the violation set of $u$ may be different for every $\triples$.
+	Since $\triples$ changes continuallyin an information system, the violations are changing along.
+
+	We can now define the notion of information system.
+	Let $\roles$ be a set of roles,
+	let $\rules$ be a set of constraints, 
+	let $\maintain$ be a surjective relation $\roles\times\rules$,
+	and let $\dataset$ be a dataset.
+\begin{definition}[information system]
+	\label{def:information system}
+	\item An information system is a tuple $\la\roles,\rules,\id{maintain},\dataset\ra$\\
+	with dataset $\dataset=\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$
+	in which $o\ \maintain\ u$ means that role $o$ must keep the set $\viol{\triples}{u}$ empty.
+\end{definition}
+	If a constraint $u$ has violations, i.e. $\viol{\triples}{u}$ is not empty, there is a role whose job it is to remove those violations.
+	The relation $\maintain$ being surjective means that all users together are keeping the violation set empty.
+	Which means that all constraints are being kept satisfied in the collective effort of users, even if the data changes.
+	In another section we will give some options on how users can keep rules satisfied.
+
+\subsection{Changes}
+% The purpose of this section is to explain why a dataset is structured the way it is.
+	In the migration of a dataset we deal with changes to the elements that are not data:
+	$\isa$, $\inst$, $\concepts$, and $\rels$.
+	Such changes have further reaching consequences, however.
+	Changes to $\isa$, $\inst$, $\concepts$, $\rels$ change the data structure in a dataset.
+	We define a \define{migration of a dataset} $\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$ as a change in which one or more of $\isa$, $\inst$, $\concepts$, or $\rels$ change.
+	To satisfy definition~\ref{def:dataset}, $\triples$ and $\atoms$ must change too,
+	but the migration should try to preserve the maximal amount of data.
+
+	This paper studies those changes and seeks to preserve the rules in definition~\ref{def:dataset}
+	in such a way that large portions of migration can be automated.
 

### 2022-09-05 2022 Migration/articleMigration.tex
WIP

@@
 		\forall r\in\rels&:&\src{r}\in\concepts\ \wedge\ \tgt{r}\in\concepts\\
-		\forall\triple{a}{r}{b}\in\triples&:&r\in\rels\ \wedge\ a\ \inst\ \src{r}\ \wedge\ b\ \inst\ \tgt{r}\\
+		\forall\triple{a}{r}{b}\in\triples&:&r\in\rels\ \wedge\ a\ \inst\ \src{r}\ \wedge\ b\ \inst\ \tgt{r}
 	\end{eqnarray}
@@
 \label{sct:Information Systems}
-	Information systems process information from many different users, each with their own tasks and responsibilities.
-	Users work from different locations and on different moments.
+	Information systems process information from many different users,
+	all with their own tasks and responsibilities,
+	and all working from different locations and on different moments.
 	This collective use serves a purpose which we loosely call "the business".
@@
 	To preserve meaning, users are trying to maintain semantic constraints on the data.
-	Before defining information systems for the purpose of data migration,
-	let us first define the constituent notions role and constraint, which are part of that definition.
+	So, let us first define the constituent notions role and constraint before defining information systems proper.
 
@@
 	Note that the signature of $u$ does not depend on the population, whereas the violation set of $u$ may be different for every $\triples$.
-	Since $\triples$ changes continuallyin an information system, the violations are changing along.
+	Since $\triples$ changes continually in an information system, the violations are changing along.
 

### 2022-09-05 AMMBR/AMBRR2.tex
Include the AMMBR idea for future publication

\documentclass[11pt,a4paper,fleqn,oneside]{article}
\usepackage{amssymb}
\parskip 10pt plus 2.5pt minus 4pt  % Extra vertical space between paragraphs.
\parindent 0em                      % Width of paragraph indentation.
\addtolength{\textwidth}{.5in}
\newcommand{\tick}{\mbox{\(\surd\)}}
\long\def\@savemarbox#1#2{\global\setbox#1\vtop{\hsize\marginparwidth 
   \@parboxrestore\smal #2}}
\def\bottomfraction{.5}
\renewcommand{\textfraction}{0.2}
\renewcommand{\floatpagefraction}{0.7}
\setcounter{tocdepth}{1}
\makeatletter
\def\@dummydocumentstyle{\@ifnextchar[{\@@dds}{\@@dds[]}}
\def\@@dds[#1]#2{\relax}

\def\getchapter#1{{\def\documentstyle{\@dummydocumentstyle}%
		   \def\document{\relax}%
		   \def\enddocument{\relax}\include{#1}}} 
\makeatother
\newlength{\Example}\setlength{\Example}{\textwidth}
\addtolength{\Example}{-0.5in}
\newlength{\Exinside}\setlength{\Exinside}{\Example}
\addtolength{\Exinside}{-0.5in}
\def\theor{\list{}{\leftmargin .5in}\item[]{\ }\\*[-\itemsep] \rm}
\let\endtheor=\endlist
\newtheorem{lemma}[equation]{Lemma}
\newtheorem{define}[equation]{Definition}
\newtheorem{examp}[equation]{Example}
%% end taken from preamble
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\def\rbracket{\text{]}}
\def\lbracket{\text{[}}
%\newcommand{\powerset}[1]{{\cal P}\{#1\}}
\newcommand{\one}[1]{{\cal\bf 1}_{#1}}
\newcommand{\choice}[2]{#1\ |\ #2}
\newcommand{\all}[2]{#1 + #2}
\newcommand{\seq}[2]{#1 ; #2}
\newcommand{\eca}[2]{{\tt ON}~#1~{\tt DO}~#2}
\newcommand{\subst}[3]{#3_{\tt [#1 := #2]}}
\newcommand{\assign}[2]{#1~{\tt :=}~#2}
\newcommand{\code}[1]{\text{\tt\small #1}}
\newcommand{\relAdd}{\dagger}
\newcommand{\dom}[1]{\id{dom}_#1}
\newcommand{\cod}[1]{\id{cod}_#1}
\newcommand{\conts}[1]{\id{contents}_#1}
\newcommand{\name}[1]{\id{name}_#1}
\def\id#1{\text{\em #1\/}}
\def\define#1{\label{dfn:#1}{\em #1}\index{#1}}
\def\definem#1{\label{dfn:#1}{\em #1}\index{#1}\marginpar{\flushright\small\em #1}}
\newcommand{\marg}[1]{\index{#1}\marginpar{\flushright\small\em #1}}
\newcommand{\uni}{\text{\small UNI}}
\newcommand{\tot}{\text{\small TOT}}
\newcommand{\inj}{\text{\small INJ}}
\newcommand{\sur}{\text{\small SUR}}
\newcommand{\univalent}[1]{\id{univalent}(#1)}
\newcommand{\total}[1]{\id{total}(#1)}
\newcommand{\surjective}[1]{\id{surjective}(#1)}
\newcommand{\injective}[1]{\id{injective}(#1)}
\newcommand{\transitive}[1]{\id{transitive}(#1)}
\newcommand{\symmetric}[1]{\id{symmetric}(#1)}
\newcommand{\antisymmetric}[1]{\id{antisymmetric}(#1)}
\newcommand{\reflexive}[1]{\id{reflexive}(#1)}
\newcommand{\ident}[1]{\mathbb{I}_{#1}}
\newcommand{\iden}{\mathbb{I}}
\newcommand{\full}{\mathbb{V}}
\newcommand{\flip}[1]{{#1}^\smallsmile}
\newcommand{\kleeneplus}[1]{{#1}^+}
\newcommand{\kleenestar}[1]{{#1}^*}
\newcommand{\cmpl}[1]{\overline{#1}}
\newcommand{\rel}{\times}
\newcommand{\compose}{;}
\newcommand{\subs}{\subseteq}
\newcommand{\fun}{\rightarrow}
\newcommand{\N}{\mbox{\msb N}}
\newcommand{\disjn}[1]{\id{disjoint}(#1)}
\newcommand{\fsignat}[3]{\id{#1}:\id{#2}\mbox{$\rightarrow$}\id{#3}}
\newcommand{\signat}[3]{\mbox{${#1}_{[{#2},{#3}]}$}}
\newcommand{\declare}[3]{\id{#1}:\id{#2}\mbox{$\times$}\id{#3}}

\title{AMMBR: A Method to Maintain Business Rules}
\author{Stef Joosten}
\begin{document}
\maketitle
\begin{abstract} % 149 words counted.
%<claim>
	This paper proposes a method, AMMBR, to generate software for maintaining the transactional integrity of a dataset.
%</claim>
	The purpose of this method is to enhance development environments that generate information systems with the feature of built-in compliance.
	Inspiration for this problem came from the author's practice of designing business processes in organisations.
	The idea is to let a dataset represent the facts of an organization,
	and let each business rule correspond to one invariant on that dataset.
	Business rules can be maintained by using software to detect violations of the invariants.
	That software can be generated in full by a generator, which analyses the invariants and compiles them into computer code.
	Thus, AMMBR determines the computer code to restore the invariants, when broken by events that change the dataset.
	This method has been tried and implemented in Ampersand, a tool that generates functional specifications and information systems.
\end{abstract}
keywords: compliance, business rule, relation algebra, software generation, provable software, process control.
\section{Introduction}
	In order to define an information system to support an organization,
	we look at organizations as systems that maintain rules.
	An ideal information infrastructure supports employees and other stakeholders to maintain the rules of the business.
	Yet rules must be concrete if they are to be used by computers.
	The method is therefore restricted to those rules of which
	violation can be established unambiguously and objectively.
	To {\em maintain} a rule means to prevent or correct all violations, either by human or automated actions.
	Temporary violation of a rule is only acceptable for rules that are maintained by humans,
	but the infrastructure must signal these violations and make them available to those people

	The problem addressed in this paper is how to define software services to maintain all rules that apply in a given context.
	This paper proposes a method, AMMBR (pronounce: Amber), to derive software services from a set of rules and
	to assemble these services into an information system (i.e. a system of people and computers) that maintains these rules.
	AMMBR consists of the following steps:
\begin{description}
\item[represent rules]
	Business rules are represented in a relation algebra \cite{Maddux2006}.
	Requirement elicitation and reuse of existing patterns produce the desired set of rules.
	This is the only step in AMMBR that requires human effort.
	The following steps are all purely algorithmic, meaning that a computer can do the work.
\item[determine clauses]
	Once the set of rules is checked and found type consistent,
	it is reduced to clauses.
	A clause is a rule that cannot be broken in smaller pieces without losing meaning.
\item[derive code fragments]
	From the clauses, code fragments are derived for violation detection and for automated actions.
\item[assemble services]
	For every role, a set of services is defined that allows a user (who has that role)
	to change the state of the system while maintaining all rules.
\end{description}
	In AMMBR, human involvement is required only in representing rules.
	So, there is a tool that generates the specification of a service layer directly from the business rules.
	This tool has been built in the form of a compiler, which produces functional specifications.
	It compiles from a language called ADL, which contains a particular relation algebra made suitable for compilation by a computer.
	The functional specification produced by this tool defines a service layer
	that maintains the rules defined in the ADL source code.
	The tool provides physical proof that an information system can be defined up to its service layer by means of rules only.

	Using experimental, earlier versions of AMMBR and ADL, information systems have been defined for banks, insurance companies and governmental organizations.

	The contribution of this paper is that business rules can define software,
	thus providing a piece of the bridge to span the abyss between information technology and business.
	From a computing perspective this is a step towards automating the design of information systems.
	From a business perspective this provides mathematical certainty that business rules are being maintained.

\section{Preliminaries}
	Deze sectie is zwaar in aanbouw.
	Ik heb hier `eventjes' een aantal regels bijeengezocht die verderop in bewijzen gebruikt worden.
	Nog aanvullen en tekst van maken (de hele sectie is nog niet voor review geschikt; doorlezen bij sectie \ref{sct:Clauses}).

	This section defines the formalism in which rules are described.
	The first two laws (\ref{axi:right identity} and \ref{axi:left identity})
	state that an identity relation may be removed
	if it stands left or right of a semicolon (the composition operator).
	Equation \ref{axi:associative compose} states that the semicolon is an associative operator.
	This means that it does not matter how brackets are placed in an expression with two or more semicolons.
	In the sequel brackets are omitted in situations with associative operators.
\begin{eqnarray}
r\compose\iden&=&r\label{axi:right identity}\\
\iden\compose r&=&r\label{axi:left identity}\\
(p\compose q)\compose r&=&p\compose(q\compose r)\ =\ p\compose q\compose r\label{axi:associative compose}\\
(p\relAdd q)\relAdd r&=&p\relAdd(q\relAdd r)\ =\ p\relAdd q\relAdd r\label{axi:associative relAdd}
\end{eqnarray}
	The following laws show how the conversion operator behaves.
\begin{eqnarray}
\flip{\flip{r}}&=&r\label{axi:flipflip}\\
\flip{(r\cup q)}&=&\flip{r}\cup\flip{q}\label{axi:flipcup}\\
\flip{(r\cap q)}&=&\flip{r}\cap\flip{q}\label{axi:flipcap}\\
\flip{(r\compose s)}&=&\flip{s}\compose\flip{r}\label{axi:flipcombi}\\
\flip{(r\relAdd s)}&=&\flip{s}\relAdd\flip{r}\label{axi:fliprelAdd}
\end{eqnarray}
	These laws are useful to show, for instance, that the conversion of an injective relation is univalent.
	Let $r$ be injective.
\[\begin{array}{cl}
&r\compose\flip{r}\ \subs\ \iden\\
\Leftrightarrow&\hspace{1in}\{\text{$r=\flip{\flip{r}}$ by law \ref{axi:flipflip}}\}\\
&\flip{\flip{r}}\compose\flip{r}\ \subs\ \iden
\end{array}\]
	The last line in this derivation says that $\flip{r}$ is univalent (definition \ref{def:univalent}).
	So the derivation proves that
\begin{equation}
\injective{r}\ \Leftrightarrow\ \univalent{\flip{r}}\label{eqn:dual injective univalent}
\end{equation}

	Another useful law is the following:
\begin{eqnarray}
p\subs r&=&\cmpl{p}\cup r\label{eqn:rid of subseteq}
\end{eqnarray}
	It shows how to remove a subset operator ($\subs$) from any equation.
	Of similar use is a law to replace an equality by two subsets.
\begin{eqnarray}
(p=r)&=&p\subs r \cup r\subs p\label{axi:rid of eq}
\end{eqnarray}

	The following law shows how the union operator can be removed from compositions of relations.
\begin{eqnarray}
p\compose(q\cup r)\compose s&=&p\compose q\compose s \cup p\compose r\compose s\label{axi:distributive cup}
\end{eqnarray}
	It would have been nice if the same distribution property holds for the intersect operator, $\cap$,
	but this is not always the case.
	This property holds only if $p$ is univalent and $s$ is injective:
\begin{equation}
p\compose(q\cap r)\compose s\ =\ p\compose q\compose s\cap p\compose r\compose s\label{prop:cap distr compo}
\end{equation}
	Since the identity relation $\iden$ is both injective and univalent,
	this law can also be replaced by the following two laws:
\begin{eqnarray}
\injective{s} & \Rightarrow & (q\cap r)\compose s\ =\ q\compose s\cap r\compose s\\
\univalent{p} & \Rightarrow & p\compose (q\cap r)\ =\ p\compose q\cap p\compose r
\end{eqnarray}
	Another way to formulate these laws in an unconditional way is:
\begin{equation}
	p\compose (q\cap r)\compose s\ \subs\ p\compose q\compose s\cap p\compose r\compose s
\label{axi:distributive cap}
\end{equation}

	Two other useful laws move relations across the $\subs$ symbol.
	However, this law is conditional too.
	If $f$ is a function (that is: a univalent and total relation) then
\begin{eqnarray}
r\subs s\compose\flip{f} & \Leftrightarrow & r\compose f\subs s\label{law:move function over subset right}\\
r\subs f\compose s        & \Leftrightarrow & \flip{f}\compose r\subs s\label{law:move function over subset left}
\end{eqnarray}

	The following equivalences are attributed to the famous mathematician De Morgan:
\begin{equation}
\begin{array}[m]{lcr}
	\cmpl{r}\cup\cmpl{s}&=&\cmpl{r\cap s}\\
	\cmpl{r}\cap\cmpl{s}&=&\cmpl{r\cup s}
\end{array}
\label{eqn:De Morgan}
\end{equation}
	De Morgan's equivalences turn out to be among the most frequently used laws,
	because the allow you to `move around' the complement operator.
	The operators $\compose$ and $\relAdd$ satisfy similar laws:
\begin{equation}
\begin{array}[m]{lcr}
	\cmpl{r}\compose\cmpl{s}&=&\cmpl{r\relAdd s}\\
	\cmpl{r}\relAdd\cmpl{s}&=&\cmpl{r\compose s}
\end{array}
\label{eqn:De Morgan ;}
\end{equation}
	Equations \ref{eqn:De Morgan} and \ref{eqn:De Morgan ;} are known as De Morgan's laws.

	Less well known, but equally important are the following equivalences, which De Morgan called ``Theorem K''.
\begin{equation}
\begin{array}[m]{llcr}
	&p\compose q&\subs&r\\=\\
	&\flip{p}\compose\cmpl{r}&\subs&\cmpl{q}\\=\\
	&\cmpl{r}\compose\flip{q}&\subs&\cmpl{p}
\end{array}
\label{eqn:theorem K}
\end{equation}

	Well known properties of relations are the so called `multiplicity properties', or \define{multiplicities}:
	univalent, total, surjective, and injective.
\begin{itemize}
\item{\bf univalent}\\A relation $\declare{r}{A}{B}$ is \define{univalent}
	if each element of $A$ corresponds to at most one element of $B$.
	This property is defined by:
\begin{equation}
\flip{r}\compose r\ \subs\ \ident{B}\label{def:univalent}
\end{equation}
	The definition says that $a\ r\ b$ and $a\ r\ b'$ imply $b=b'$ for all $a$, $b$, and $b'$.
	To denote that relation $r$ is univalent, we write $\univalent{r}$.
\item{\bf total}\\A relation $\declare{r}{A}{B}$ is \define{total}
	if each element of $A$ corresponds to at least one element of $B$.
	This property is defined by:
\begin{equation}
\ident{A}\ \subs\ r\compose\flip{r}\label{def:total}
\end{equation}
	The definition says that for all $a\in A$ there exists $b\in B$ such that $a\ r\ b$.
	To denote that relation $r$ is univalent, we write $\univalent{r}$.
\item{\bf function}\\
	A relation is a \define{function} if it is both univalent and total.
	That is: a relation $\declare{r}{A}{B}$ is a function
	if every element of $A$ corresponds to precisely one element of $B$.
	If $r$ is univalent but not total then it is called a \define{partial function}.
\item{\bf injective}\\A relation $\declare{r}{A}{B}$ is \define{injective}
	if each element of $B$ corresponds to at most one element of $A$.
	This property is defined by:
\begin{equation}
r\compose\flip{r}\ \subs\ \ident{A}\label{def:injective}
\end{equation}
	The definition says that $b\ r\ a$ and $b\ r\ a'$ imply $a=a'$ for all $a$, $a'$, and $b$.
	Notice that the property injective is defined conversely to the property univalent.
	To denote that relation $r$ is injective, we write $\injective{r}$.
\item{\bf surjective}\\A relation $\declare{r}{A}{B}$ is \define{surjective}
	if each element of $B$ corresponds to at least one element of $A$.
	This property is defined by:
\begin{equation}
\ident{B}\ \subs\ \flip{r}\compose r\label{def:surjective}
\end{equation}
	The definition says that for all $b\in B$ there exists $a\in A$ such that $a\ r\ b$.
\end{itemize}

\section{Clauses}
\label{sct:Clauses}
	This section explains:
\begin{enumerate}
\item	how to derive clauses from a set of rules, by a normalization procedure.
\item	how to transform clauses into left identity clauses and right identity clauses, if possible.
\item	how to derive multiplicity properties from clauses thus obtained.
\end{enumerate}
	A \define{rule} is an expression $e$ that must be kept true.
	To \define{maintain} rule $r$ means to ensure that $\full\subs r$ is true,
	where $\full$ is the complete relation (i.e. everything is an pair of $\full$).
	In order to discuss maintaining rules, we must introduce some vocabulary.
	As long as $\full\subs r$, we say that $r$ is \define{satisfied}, or simply that it \define{holds}.
	If $r$ is not $\full$, we say that \define{invariance is broken},
	or rule $r$ is \define{not satisfied}, or it \define{does not hold}.
	Since $r\cup\cmpl{r}=\full$, the missing pairs are determined by $\cmpl{r}$.
	Each pair of $\cmpl{r}$ is called a \define{violation} of $r$.
	To \define{restore invariance} of rule $r$ means to change the contents of $r$ in such a way that $\full\subs r$.

	Let $r_0$, $r_1$, $...$, $r_n$ be the rules that apply in context $C$.
	Let $R_C$ be defined by $r_0\cap r_1\cap ... \cap r_n$.
	Expression $R_C$ is called the \define{conjunction} of $r_0$, $r_1$, $...$, $r_n$, because all expressions (conjuncts) $r_i$ are separated by the conjunction operator $\cap$.
	Maintaining all rules (every $r_i$) in context $C$ means to ensure that $R_C=\full$ at all times\footnote{Informally: $R_C$ is kept true at all times}.

	A \define{clause} is the smallest expression that is still a rule (i.e. are equal to $\full$).
	Let $e_0$, $e_1$, $...$, $e_m$ be unique expressions such that:
\begin{equation}
	r_0\cap r_1\cap ... \cap r_n\ =\ e_0\cap e_1\cap ... \cap e_m
\end{equation}
	with $m$ the largest possible number and `unique' meaning that for every $i$ and $j$
\begin{equation}
	e_i=e_j\ \Rightarrow\ i=j
\end{equation}
	So each $e_i$ thus defined is a clause.
	Since this definition calls for `as many unique expressions as possible',
	no clause $e$ can be written as a conjunction.
	(If an $a$ and $b$ would exist such that $e=a\cap b$, then we could have had one clause more. This contradicts the definition of clause.)

\subsection{Normalization}
	In order to transform a set of rules into clauses,
	a normalization procedure is performed.
	This procedure transforms a set of rules by meaning preserving transformations.
	Consequently, the result has the appropriate form and still has the exact same meaning.
	All normalization rules in this section have the form $a\rightarrow b$,
	and they all satisfy $a=b$ (which means they are meaning preserving).
\begin{enumerate}
\item	Move all $\flip{}$ operators inward, using the following transformations:
  \[\begin{array}{rcl}
	\flip{(r\cup s)}&\rightarrow&\flip{r}\cup\flip{s}\\
	\flip{(r\cap s)}&\rightarrow&\flip{r}\cap\flip{s}\\
	\flip{(r\compose s)}&\rightarrow&\flip{s}\compose\flip{r}\\
	\flip{(r\dagger s)}&\rightarrow&\flip{s}\dagger\flip{r}\\
	\flip{\cmpl{r}}&\rightarrow&\cmpl{\flip{r}}
  \end{array}\]
\item	Move all $\compose$ operators inwards, as much as possible, using the following transformations:
  \[\begin{array}{rcll}
	q\compose(r\cup s)&\rightarrow&q\compose r\ \cup\ q\compose s\\
	(q\cup r)\compose s&\rightarrow&q\compose s\ \cup\ r\compose s\\
	q\compose(r\cap s)&\rightarrow&q\compose r\ \cap\ q\compose s&\text{if $q$ is a function}\\
	(q\cap r)\compose s&\rightarrow&q\compose s\ \cap\ r\compose s&\text{if $\flip{s}$ is a function}
  \end{array}\]
\item	Distribute $\cup$ over $\cap$, using the following transformations:
  \[\begin{array}{rcl}
	q\cup(r\cap s)&\rightarrow&(q\cup r)\ \cap\ (q\cup s)\\
	(q\cap r)\cup s&\rightarrow&(q\cup s)\ \cap\ (r\cup s)
  \end{array}\]
\item	Move all complement operators inward across $\cap$ and $\cup$, using the following transformations:
  \[\begin{array}{rcl}
	\cmpl{r\cup s}&\rightarrow&\cmpl{r}\ \cap\ \cmpl{s}\\
	\cmpl{r\cap s}&\rightarrow&\cmpl{r}\ \cup\ \cmpl{s}
  \end{array}\]
\item	Remove identical clauses and terms:
  \[\begin{array}{rcl}
	r\cup r&\rightarrow&r\\
	r\cap r&\rightarrow&r
  \end{array}\]
\item	Remove opposite clauses and terms:
  \[\begin{array}{rcl}
	r\cup \cmpl{r}&\rightarrow&\full\\
	r\cap \cmpl{r}&\rightarrow&\cmpl{\full}
  \end{array}\]
\item	Absorb redundant terms:
  \[\begin{array}{rcl}
	r\cap(\cmpl{r}\cup s)&\rightarrow&r\cap s\\
	r\cup(\cmpl{r}\cap s)&\rightarrow&r\cup s\\
	r\cap(r\cup s)&\rightarrow&r\\
	r\cup(r\cap s)&\rightarrow&r
  \end{array}\]
\item	Remove redundant clauses and terms:
  \[\begin{array}{rcl}
	r\cup \full&\rightarrow&\full\\
	r\cup \cmpl{\full}&\rightarrow&r\\
	r\cap \full&\rightarrow&r\\
	r\cap \cmpl{\full}&\rightarrow&\cmpl{\full}
  \end{array}\]
\item	Remove redundant brackets:
  \[\begin{array}{rcl}
	q\cap(r\cap s)&\rightarrow&q\cap r\cap s\\
	(q\cap r)\cap s&\rightarrow&q\cap r\cap s\\
	q\cup(r\cup s)&\rightarrow&q\cup r\cup s\\
	(q\cup r)\cup s&\rightarrow&q\cup r\cup s\\
	q\compose(r\compose s)&\rightarrow&q\compose r\compose s\\
	(q\compose r)\compose s&\rightarrow&q\compose r\compose s\\
	q\dagger(r\dagger s)&\rightarrow&q\dagger r\dagger s\\
	(q\dagger r)\dagger s&\rightarrow&q\dagger r\dagger s
  \end{array}\]
\end{enumerate}
	The normalization procedure continues to apply these transformations until none of them are applicable.
	If the normalization procedure reduces one or more clauses to $\cmpl{\full}$,
	there is an inconsistency.
	Note that this renders the entire set of rules inconsistent (i.e. equivalent to $\cmpl{\full}$).
	Any clause that is reduced to $\full$ is a tautology and is consequently removed by the normalization procedure.

	Without proof, we claim that:
\begin{itemize}
\item	The normalization procedure terminates.
\item	The result is a conjunction of clauses.
\item	Each clause can be written as a disjunction of (zero or more) positive and negative expressions:
\begin{equation}
	\cmpl{a_1}\cup\cmpl{a_2}\cup ... \cup\cmpl{a_k}\ \cup\ c_1\cup c_2\cup ... \cup c_l
\label{eqn:clause}
\end{equation}
\end{itemize}
	Expressions that are separated by disjunctions are called \define{disjuncts} or \define{terms}.
	A term is called \define{negative} if it is an expression enclosed by a complement operator.
	A \define{positive term} is any term that is not negative.
	Each expression $a_i$ in equation \ref{eqn:clause} is called an \define{antecedent} of the clause and 
	each expression $c_j$ is called a \define{consequent} of the clause.

\subsection{Left- and right identity clauses}
	The clauses that are produced by the normalization procedure can be distinguished further into three categories:
\begin{enumerate}
\item	A clause with only one negative term, which is equal to $\iden$ is called a \define{left identity clause}.
\item	A clause with only one positive term, which is equal to $\iden$ is called a \define{right identity clause}.
\item	All other clauses.
\end{enumerate}

	Let $\cmpl{a}\cup c$ be a clause and let $a$ be of the form
\[f_1\compose f_2\compose ... \compose f_n\compose g_1\compose g_2\compose ... \compose g_m\]
	in which every $g_i$ is a function and every $\flip{f_j}$ is a function too.
	In that case, $\cmpl{a}\cup c$ is a left identity clause, which we write as:
\[\cmpl{\iden}\cup \flip{(f_1\compose f_2\compose ... \compose f_n)}\compose c\compose\flip{(g_1\compose g_2\compose ... \compose g_m)}\]

	Let $\cmpl{a}\cup c$ be a clause and let $c$ be of the form
\[t\ =\ f_1\compose f_2\compose ... \compose f_n\compose g_1\compose g_2\compose ... \compose g_m\]
	in which every $f_i$ is a function and every $\flip{g_j}$ is a function too.
	In that case, $\cmpl{a}\cup c$ is a right identity clause, which we write as:
\[\cmpl{\flip{(f_1\compose f_2\compose ... \compose f_n)}\compose c\compose\flip{(g_1\compose g_2\compose ... \compose g_m)}}\ \cup\ \iden\]

	Note that $c$ in the left identity clause and $c$ in the right identity clause may consist of several terms.
	Therefore one clause $\cmpl{a}\cup c$ may yield multiple (or no) right identity clauses and multiple (or no) left identity clauses.

\subsection{Multiplicity derivation}
	Multiplicity information is important for deriving data structures.
	Some multiplicities are given (by the modeler)
	and others can be derived from those already given.
	This section shows how multiplicities can be derived from given ones.

	For deriving multiplicities, we use the following laws.
\begin{eqnarray}
	\injective{r}&=&\univalent{\flip{r}}\label{eqn:flip univalent}\\
	\surjective{r}&=&\total{\flip{r}}\label{eqn:flip total}\\
	\univalent{r}\cap\univalent{s}&\subs&\univalent{r\compose s}\label{eqn:combi univalent}\\
	\surjective{s}\cap \surjective{r}&\subs&\surjective{r\compose s}\label{eqn:combi surjective}\\
	\total{s}\cap\total{r}&\subs&\total{r\compose s}\label{eqn:combi total}\\
	\injective{s}\cap\injective{r}&\subs&\injective{r\compose s}\label{eqn:combi injective}\\
	\univalent{r}\cup\univalent{s}&\subs&\univalent{r\cap s}\label{eqn:cap univalent}\\
	\injective{r}\cup\injective{s}&\subs&\injective{r\cap s}\label{eqn:cap injective}\\
	\total{r}\cap\total{s}&\subs&\total{r\cup s}\label{eqn:cup total}\\
	\univalent{r\cup s}&\subs&\univalent{r}\cap\univalent{s}\label{eqn:cup univalent}\\
	\injective{r\cup s}&\subs&\injective{r}\cap\injective{s}\\
	\surjective{r\cup s}&\subs&\surjective{r}\cap\surjective{s}\label{eqn:cup surjective}\\
	\total{r\cup s}&\subs&\total{r}\cap\total{s}
\end{eqnarray}

\subsection{old stuff}

	After this example, let us analyze how to restore invariants in all possible cases.	
	Suppose we have clause $r$, and an operation $t$.
	We define: operation $t$ \define{respects invariance of $r$} means
\begin{equation}
r\ \subs\ t(r)
\label{axi:invariance}
\end{equation}	
	This definition can be understood by saying that the effect of $t$,
	which is $t(r)$,
	must hold under the condition that $r$ holds.

	In a database both people and computer programs
	can take any action at an arbitrary moment.
	We call that an event.
	An \define{event} is an insertion or deletion of items (i.e. pairs) in a relation.
	If that relation is used in a clause, that clause may be affected by the event.
	Let $a(r)$ be the result of event $a$ on clause $r$.
	That result might not be equal to $\full$, i.e. there may be violations $\cmpl{a(r)}$ of the clause.
	So, in order to restore invariance, a reaction $t$ is required that satisfies $t(a(r))=\full$.
	So if we propose a response $t$ to an event $a$, we must prove for all possible $r$:
\begin{equation}
	r\ \subs\ t(a(r))
\label{axi:code fragment}
\end{equation}

	A \define{code fragment} is a computer program that consists of an action $t$
	that is taken in response to an event $a$.
	In order to derive code fragments, we first define the possible events.
	An event occurs as a result of changing the contents of relations.
	So the relevant ones are those relations $m$ that affect (i.e. are used in) $r$.
	Let $\id{mors}(r)$ be the set of relations that affect $r$.
	The set of possible events consists of insert- and delete events on relations $\id{mors}(r)$.
	Some of these events, however, do not violate clause $e$.
	If absence of violations can be proven for every $\Delta$,
	there is no need to consider the event.
	Therefore, rule $r$ is affected by the following set of possible events
\[\id{events}(r)\ =\ \begin{array}[t]{@{}l}	
       \{\id{insert}_m |\ m\in\id{mors}(r)\text{ and }\forall\Delta: r\ \not\subs\ \id{insert}_m(\Delta,r)\}\\
               \cup\\
       \{\id{delete}_m |\ m\in\id{mors}(r)\text{ and }\forall\Delta: r\ \not\subs\ \id{delete}_m(\Delta,r)\}\\
\end{array}\]

	Any reaction on one of these actions must restore invariance of $e$.
	This allows us to specify all code fragments $f_e$ by the following equation:
\begin{equation}
    f_e\ \in\ \begin{array}[t]{l}
               \{\langle e,\ \id{insert}_m,\ e{\tt :=}e\cup\cmpl{\id{insert}_m(\Delta,e)}\rangle
               |m\in\id{mors}(e)\text{ and }\cmpl{\id{insert}_m(\Delta,e)}\not = \emptyset\}\\
               \cup\\
               \{\langle e,\ \id{delete}_m,\ e{\tt :=}e\cup\cmpl{\id{delete}_m(\Delta,e)}\rangle
               |m\in\id{mors}(e)\text{ and }\cmpl{\id{delete}_m(\Delta,e)}\not = \emptyset\}\\
              \end{array}
\label{axi:code fragments}
\end{equation}
	Note that there are choices involved when inserting violations $\cmpl{a(e)}$ in $e$.
	Any choice that satisfies equation \ref{axi:code fragments} is acceptable,
	because the resulting code fragment will satisfy equation \ref{axi:code fragment}.
	Therefore, the designer of the code fragments is free to make his or her own choices.

	By way of illustration, let us elaborate one possible choice.
	Suppose we restrict insertion of violations to just one of the terms of $e$.
	Let $\id{terms}(e)$ be the set of terms in $e$,
	then the possible reactions are
\[\{t\code{:=}t\cup\cmpl{a(e)} | t\in\id{terms}(e)\}\]
	If actions are restricted to individual terms as well,
	we can enumerate possible fragments.
	Suppose for example, that we have a clause $e$
\[e\ =\ \cmpl{p}\cup\cmpl{q}\cup s\cup t\]
	If compensation is restricted to terms, table \ref{tbl:code fragments} enumerates all possibilities.
\begin{table}[ht]
\begin{center}
\(\begin{array}{|r|r@{~\tt :=~}l@{~}c@{~}c|}
\hline
\text{event $a$}&\multicolumn{4}{|l|}{\text{action $r$}}\\\hline
\id{insert}_p(\Delta,r)&q&q&\cap&\cmpl{a(r)}\\
\id{insert}_p(\Delta,r)&s&s&\cup&a(r)\\
\id{insert}_p(\Delta,r)&t&t&\cup&a(r)\\
\id{insert}_q(\Delta,r)&p&p&\cap&\cmpl{a(r)}\\
\id{insert}_q(\Delta,r)&s&s&\cup&a(r)\\
\id{insert}_q(\Delta,r)&t&t&\cup&a(r)\\
\id{delete}_s(\Delta,r)&p&p&\cap&\cmpl{a(r)}\\
\id{delete}_s(\Delta,r)&q&q&\cap&\cmpl{a(r)}\\
\id{delete}_s(\Delta,r)&t&t&\cup&a(r)\\
\id{delete}_t(\Delta,r)&p&p&\cap&\cmpl{a(r)}\\
\id{delete}_t(\Delta,r)&q&q&\cap&\cmpl{a(r)}\\
\id{delete}_t(\Delta,r)&s&s&\cup&a(r)\\
\hline
\end{array}\)
\end{center}
\caption{Code fragments for $r\ =\ \cmpl{p}\cup\cmpl{q}\cup s\cup t$}
\label{tbl:code fragments}
\end{table}

	Each of the code fragments from table \ref{tbl:code fragments} will preserve invariance of $e$.
	We shall prove this for one code fragment only (we have arbitrarily picked the third code fragment):
	This gives
\[\begin{array}{rl}
&\cmpl{(p\cup\Delta)}\ \cup\ \cmpl{q}\ \cup\ s\ \cup\ (t\cup\cmpl{a(e)})\\
=&~~~~~~~~\{a(e)\ =\ \cmpl{(p\cup\Delta)}\cup\cmpl{q}\cup s\cup t\}\\
&\cmpl{p\cup\Delta}\ \cup\ \cmpl{q}\ \cup\ s\ \cup\ t\ \cup\ \cmpl{\cmpl{(p\cup\Delta)}\cup\cmpl{q}\cup s\cup t}\\
=&~~~~~~~~\{\text{De Morgan (\ref{eqn:De Morgan})}\}\\
&\cmpl{p\cup\Delta}\ \cup\ \cmpl{q}\ \cup\ s\ \cup\ t\ \cup\ (\cmpl{\cmpl{(p\cup\Delta)}}\cap\cmpl{\cmpl{q}}\cap\cmpl{s}\cap\cmpl{t})\\
=&~~~~~~~~\{\text{remove double complement}\}\\
&\cmpl{p\cup\Delta}\ \cup\ \cmpl{q}\ \cup\ s\ \cup\ t\ \cup\ ((p\cup\Delta)\cap q\cap\cmpl{s}\cap\cmpl{t})\\
=&~~~~~~~~\{\text{Distribute $\cap$ over $\cup$}\}\\
&(\cmpl{p\cup\Delta}\ \cup\ \cmpl{q}\ \cup\ s\ \cup\ t\ \cup\ (p\cup\Delta))\ \cap\\
&(\cmpl{p\cup\Delta}\ \cup\ \cmpl{q}\ \cup\ s\ \cup\ t\ \cup\ q)\ \cap\\
&(\cmpl{p\cup\Delta}\ \cup\ \cmpl{q}\ \cup\ s\ \cup\ t\ \cup\ \cmpl{s})\ \cap\\
&(\cmpl{p\cup\Delta}\ \cup\ \cmpl{q}\ \cup\ s\ \cup\ t\ \cup\ \cmpl{t})\\
=&~~~~~~~~\{\cmpl{p\cup\Delta}\ \cup\ p\cup\Delta=\full\text{~and~}\cmpl{q}\cup q=\full\text{~and~}s\ \cup\ \cmpl{s}=\full\text{~and~}t\ \cup\ \cmpl{t}=\full\}\\
&\full\cap \full\cap \full\cap \full\\
=\\
&\full\\
\end{array}\]

	This derivation proves that the action of inserting an arbitrary $\Delta$ in $p$ can be reacted upon by
	inserting the violations $\cmpl{a(e)}$ into $t$.
	Although this particular table is specific for rules of the form $\cmpl{p}\cup\cmpl{q}\cup s\cup t$,
	it can be generalized to clauses with any number (0 or more) negative terms
	and 1 or more positive terms.

	A special situation occurs if a clause contains no negative terms,
	as in:
\[s\cup t\]
	To keep this clause invariant means to keep $\full\subs s\cup t$ equal to $\full$ at all times.
	In a database, a representation of $\full$ might change when
	$\Delta$ contains new instances, i.e. instances that are not (yet) in $\full$.
	Then, either $s$ or $t$ must be extended with $\Delta\compose \full\cup \full\compose\Delta$.

\begin{table}[ht]
\begin{center}
\(\begin{array}{|r@{~\tt :=~}l@{~}c@{~}c|r@{~\tt :=~}l@{~}c@{~}c|}
\hline
\multicolumn{4}{|l|}{\text{action $a$}}&\multicolumn{4}{|l|}{\text{reaction $r$}}\\\hline
\ident{A}&\ident{A}&\cup&\Delta&s&s&\cup&\Delta\compose \full\\
\ident{A}&\ident{A}&\cup&\Delta&t&t&\cup&\Delta\compose \full\\
\ident{B}&\ident{B}&\cup&\Delta&s&s&\cup&\full\compose\Delta\\
\ident{B}&\ident{B}&\cup&\Delta&t&t&\cup&\full\compose\Delta\\
\hline
\end{array}\)
\end{center}
\caption{Code fragments for $s\cup t$}
\end{table}

\begin{eqnarray}
\id{Ins}_\Delta(\cmpl{x})&=&\id{Del}_\Delta(x)\\
\id{Ins}_\Delta(x\cup y)&=&\choice{\id{Ins}_\Delta(x)}{\id{Ins}_\Delta(y)}\\
\id{Ins}_\Delta(x\cap y)&=&\all{\id{Ins}_\Delta(x)}{\id{Ins}_\Delta(y)}\\
\id{Ins}_\Delta(x\compose y)&=&\begin{array}[t]{@{}l}\text{let $n\in\id{target}(x)$ (arbitrarily chosen) in}\\ \all{\id{Ins}_{(\Delta\cap\cmpl{x\compose y})\compose\full\compose\one{n}}(x)}{\id{Ins}_{\one{n}\compose\full\compose(\Delta\cap\cmpl{x\compose y})}(y)}\end{array}\\
\id{Ins}_\Delta(m)&=&\assign{m}{m\cup\Delta}\\
\id{Del}_\Delta(\cmpl{x})&=&\choice{\choice{\id{Ins}_\Delta(x)}{\id{Del}_{(\Delta\compose\full)}(I_{[\id{source}(x)]})}}{\id{Del}_{(\full\compose\flip{\Delta})}(I_{[\id{target}(x)]})}\\
\id{Del}_\Delta(x\cup y)&=&\all{\id{Del}_\Delta(x)}{\id{Del}_\Delta(y)}\\
\id{Del}_\Delta(x\cap y)&=&\choice{\id{Del}_\Delta(x)}{\id{Del}_\Delta(y)}\\
\id{Del}_\Delta(x\compose y)&=&\begin{array}[t]{@{}l}\id{Del}_{\Delta\compose\full\ \cap\ \full\compose\flip{y}}(x)\ \ |\\
\id{Del}_{\flip{x}\compose\full\ \cap\ \full\compose\Delta}(y) \end{array}\\
\id{Del}_\Delta(m)&=&\assign{m}{m\cap\cmpl{\Delta}}
\end{eqnarray}

\section{Services}

\end{document}

### 2022-09-05 AMMBR/patent.tex
Include the AMMBR idea for future publication

--- 2022-09-05 AMMBR/AMBRR2.tex
@@
 \usepackage{amssymb}
+\usepackage{amsmath}
 \parskip 10pt plus 2.5pt minus 4pt  % Extra vertical space between paragraphs.
@@
 \let\endtheor=\endlist
+\newenvironment{derivation}{\begin{tabbing}\advance\mathindent -1em\hspace*{1em}\=\hspace*{\mathindent}\=\hspace*{.6in}\=etc.\+\kill}{\end{tabbing}}
 \newtheorem{lemma}[equation]{Lemma}
@@
 \newcommand{\iden}{\mathbb{I}}
-\newcommand{\full}{\mathbb{V}}
 \newcommand{\flip}[1]{{#1}^\smallsmile}
@@
 \newcommand{\compose}{;}
-\newcommand{\subs}{\subseteq}
+\newcommand{\subs}{\dashv}
 \newcommand{\fun}{\rightarrow}
@@
 \maketitle
-\begin{abstract} % 149 words counted.
-%<claim>
-	This paper proposes a method, AMMBR, to generate software for maintaining the transactional integrity of a dataset.
-%</claim>
-	The purpose of this method is to enhance development environments that generate information systems with the feature of built-in compliance.
-	Inspiration for this problem came from the author's practice of designing business processes in organisations.
-	The idea is to let a dataset represent the facts of an organization,
-	and let each business rule correspond to one invariant on that dataset.
-	Business rules can be maintained by using software to detect violations of the invariants.
-	That software can be generated in full by a generator, which analyses the invariants and compiles them into computer code.
-	Thus, AMMBR determines the computer code to restore the invariants, when broken by events that change the dataset.
-	This method has been tried and implemented in Ampersand, a tool that generates functional specifications and information systems.
+\begin{abstract}
+	Ideally, an information infrastructure supports an organization by maintaining rules that are specific to the business.
+	Such rules are known as \define{business rules} \cite{BRM}.
+	This paper proposes a method to define such an infrastructure.
+	It allows organizations to build information systems that comply to their business rules in a provable way.
+	The method is implemented in a tool that generates functional specifications.
+	The tool is an instrument for designers (information architects) who define processes and information systems.
+
+	This paper is meant for readers who are interested in the mathematical background of
+	deriving software from business rules.
+	It uses a relation algebra as a vehicle for processing business rules.
 \end{abstract}
@@
 \end{enumerate}
-	A \define{rule} is an expression $e$ that must be kept true.
-	To \define{maintain} rule $r$ means to ensure that $\full\subs r$ is true,
-	where $\full$ is the complete relation (i.e. everything is an pair of $\full$).
+	A \define{rule} is an expression $e$ that must be maintained.
+	To \define{maintain} rule $r$ means to ensure that $\full\subs r$ at all times,
+	where $\full$ is the complete relation (i.e. everything is an element of $\full$).
 	In order to discuss maintaining rules, we must introduce some vocabulary.
@@
 	Since $r\cup\cmpl{r}=\full$, the missing pairs are determined by $\cmpl{r}$.
-	Each pair of $\cmpl{r}$ is called a \define{violation} of $r$.
+	Each element of $\cmpl{r}$ is called a \define{violation} of $r$.
 	To \define{restore invariance} of rule $r$ means to change the contents of $r$ in such a way that $\full\subs r$.
@@
 
-\subsection{old stuff}
+\section{Code Fragments}
+	In this section we derive code for computers,
+	which is built around two basic operations: insert and delete.
+	The insert operation can be represented in terms of pre- and postconditions\footnote{this idea
+	was introduced by Floyd and Hoare.
+	It is well known for making assertions about computer programs and reasoning about them.}.
+\[\begin{array}{r@{~}l}
+&\{\text{pre: $\code{P}=p$}\}\\
+\code{INSERT}&\Delta\code{ IN P}\\
+&\{\text{post: $\code{P}=p\cup\Delta$}\}\\
+\end{array}\]
+	The same can be expressed mathematically by the function $\id{insert}_\code{P}$:
+\begin{equation}
+\id{insert}_\code{P}(\Delta,r)\ =\ \subst{P}{P\cup\Delta}{r}
+\end{equation}
+	This function inserts a set of pairs, $\Delta$, into its argument.
+
+	The delete operation is introduced similarly:
+\[\begin{array}{r@{~}l}
+&\{\text{pre: $\code{P}=p$}\}\\
+\code{DELETE}&\Delta\code{ FROM P}\\
+&\{\text{post: $\code{P}=p\cap\cmpl{\Delta}$}\}\\
+\end{array}\]
+	This too is expressed mathematically by the function $\id{delete}_\code{P}$:
+\begin{equation}
+\id{delete}_\code{P}(\Delta,r)\ =\ \subst{P}{P\cap\cmpl{\Delta}}{r}
+\end{equation}
+
+	Let one example introduce the idea of breaking and restoring invariance.
+	Suppose the rule $\code{P}\subs\code{S}\cup\code{T}$ is to be maintained by a computer,
+	and \code{P}, \code{S}, and \code{T} are variables (memory locations) in that computer,
+	which represent relations.
+	We assume that before executing any code, we have $\code{P}=p$, $\code{S}=s$, and $\code{T}=t$.
+	The following example shows that inserting a set of pairs, $\Delta$, in \code{P}
+	may violate the rule.
+	Inserting the same $\Delta$ in \code{S} restores invariance.
+	In terms of code, we write:
+\[\begin{array}{r@{~}l}
+&\{\text{Assume $\code{P}=p$, $\code{S}=s$, $\code{T}=t$ and $\code{P}\subs\code{S}\cup\code{T}$}\}\\
+\code{INSERT}&\Delta\code{ IN P}\\
+&\{\text{$\code{P}=p\cup\Delta$, $\code{S}=s$, and $\code{T}=t$, and invariance is broken: $\code{P}\not\subs\code{S}\cup\code{T}$}\}\\
+\code{INSERT}&\Delta\code{ IN S}\\
+&\{\text{postcondition: $\code{P}=p\cup\Delta$, $\code{S}=s\cup\Delta$, $\code{T}=t$, and $\code{P}\subs\code{S}\cup\code{T}$}\}\\
+\end{array}\]
+	To discuss this example, we start at the top.
+	We may assume $\code{P}\subs\code{S}\cup\code{T}$ that holds.
+	Rephrased in disjunctive form, this is equivalent to $\cmpl{\code{P}}\cup\code{S}\cup\code{T}$.
+	The first line of code inserts a set of pairs, $\Delta$, into relation \code{P},
+	which means that after the first line, we have $\code{P}=p\cup\Delta$.
+	Consequently, $\cmpl{\code{P}}=\cmpl{p}\cap\cmpl{\Delta}$,
+	which means that some elements of $\Delta$ may be missing from the rule, $\cmpl{\code{P}}\cup\code{S}\cup\code{T}$.
+	To neutralize that result,
+	the missing elements must be added to either S or T.
+	That is what happens in the second line.
+
+	To prove that rule $\code{P}\subs\code{S}\cup\code{T}$ is maintained,
+	we must show that its value afterwards is equal to $\full$,
+	under the condition that its initial value equals $\full$.
+	The initial value is $\cmpl{p}\cup s\cup t$, which we may assume.
+	The final value is $(p\cup\Delta)\subs(s\cup\Delta)\cup t$, which must be proven equal to $\full$:
+\[\begin{array}{cl}
+&(p\cup\Delta)\subs(s\cup\Delta)\cup t\\
+=&\hspace{1in}\{\hbox{by law \ref{eqn:rid of subseteq}}\}\\
+&\cmpl{p\cup\Delta}\ \cup\ s\ \cup\ \Delta\ \cup\ t\\
+=&\hspace{1in}\{\hbox{De Morgan's law (\ref{eqn:De Morgan})}\}\\
+&(\cmpl{p}\cap\cmpl{\Delta})\ \cup\ s\ \cup\ \Delta\ \cup\ t\\
+=&\hspace{1in}\{\hbox{Distribute $\cup$ over $\cap$}\}\\
+&(\cmpl{p}\cup s\cup\Delta\cup t)\ \cap\ (\cmpl{\Delta}\cup s\cup\Delta\cup t)\\
+=&\hspace{1in}\{\cmpl{\Delta}\cup\Delta=\full\}\\
+&\cmpl{p}\cup s\cup\Delta\cup t\\
+=&\hspace{1in}\{\full\ =\ \cmpl{p}\cup s\cup t\}\\
+&\full
+\end{array}\]
+	From this derivation we may conclude that invariance of $\code{P}\subs\code{S}\cup\code{T}$ is restored.
 
@@
 
-	Each of the code fragments from table \ref{tbl:code fragments} will preserve invariance of $e$.
+	Each of the code fragments from table \ref{tbl:code fragments} will maintain invariance of $e$.
 	We shall prove this for one code fragment only (we have arbitrarily picked the third code fragment):

### 2022-09-06 2022 Migration/articleMigration.tex
WIP on the summary

@@
 	Changes that affect the data model typically result in a data migration.
-	In such cases, simply regenerating the system is not enough.
+	In such cases, simply regenerating the system is not enough because it would reset the database to its initial state (deleting all data gathered so far).
+	Migration aims at preserving the data as much as possible.
+	Typically, migration involves transferring the old data to the new system while preserving the semantics as much as possible.
 
-	In this contribution we develop a theory for data migration that aims at automating the migration,
+	In this contribution we develop a theory for reliable data migration that aims at automating it as much as possible,
 	to enable more frequent migrations.
-	A theory is needed to do this reliably, without mistakes.
-	This paper proposes such a theory.
-
-	The ultimate target is to generate the entire migration from two specifications: the as-is specification and the to-be specification.
-	The theory for doing that is the topic of this paper.
-	The software generator that embodies this theory is subject of future research.
+	The ultimate target is to generate a migration from two specifications: the as-is specification and the to-be specification.
+	A software generator that embodies this theory is subject of future research.
 \end{abstract}

### 2022-09-10 2022 Migration/articleMigration.tex
wat probeerseltjes...

@@
 	and all working from different locations and on different moments.
-	This collective use serves a purpose which we loosely call "the business".
+	This collective use serves a purpose which we loosely call ``the business''.
 	As a consequence, the data in an information system changes continually.
@@
 	We call every element $\pair{a}{b}\in\viol{\triples}{u}$ a \define{violation} of constraint $u$ in $\triples$.
-	We say that a constraint \define{is satisfied} if its violation set is empty (i.e. if $\viol{\triples}{u}$ is the empty set).
+	We say that a constraint \define{is satisfied} if its violation set is empty.
 	The set of constraints is called $\rules$.
@@
 	\label{def:information system}
-	\item An information system is a tuple $\la\roles,\rules,\id{maintain},\dataset\ra$\\
+	\item An information system $\infsys$ is a tuple $\la\roles,\rules,\maintain,\dataset\ra$\\
 	with dataset $\dataset=\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$
@@
 	In another section we will give some options on how users can keep rules satisfied.
+
+\subsection{Migration}
+	To migrate information system $\infsys$ to $\infsys'$, we take the disjoint union $\infsys\sqcup\infsys'$ of both information systems
+	and add rules to describe the migration.
+	$\infsys$ is the existing system, which contains data to be preserved.
+	$\infsys'$ is the new system, which contains no or hardly any data.
+	It does, however, contain the concepts, relations, rules and roles of the new system.
+
+	We use two datasets: $\dataset=\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$ and $\dataset'=\la\atoms',\concepts',\inst',\isa',\rels',\triples'\ra$.
+	Before doing so, let us first define the disjoint union of two information systems.
+\begin{definition}[disjoint union of datasets]
+	\begin{eqnarray}
+		\dataset\sqcup\dataset'&=&\la\atoms\sqcup\atoms',\ \concepts\sqcup\concepts',\ \inst\sqcup\inst',\ \isa\sqcup\isa',\ \rels\sqcup\rels',\ \triples\sqcup\triples'\ra\\
+		\atoms\sqcup\atoms'&=&\{(a,0)|\ a\in\atoms\}\ \cup\ \{(a',1)|\ a'\in\atoms'\}\\
+		\concepts\sqcup\concepts'&=&\{(c,0)|\ c\in\concepts\}\ \cup\ \{(c',1)|\ c'\in\concepts'\}\\
+		\inst\sqcup\inst'&=&\begin{array}[t]{@{}l}
+							  \{((a,0),(A,0))|\ (a,A)\in\inst\}\ \cup\\
+			                  \{((a',1),(A',1))|\ (a',A')\in\inst'\}
+							\end{array}\\
+		\isa\sqcup\isa'&=&\begin{array}[t]{@{}l}
+							  \{((A,0),(B,0))|\ A\isa B\}\ \cup\\
+			                  \{((A',1),(B',1))|\ A'\isa' B'\}
+							\end{array}\\
+		\rels\sqcup\rels'&=&\{(r,0)|\ r\in\rels\}\ \cup\ \{(r',1)|\ r'\in\rels'\}\\
+		\triples\sqcup\triples'&=&\begin{array}[t]{@{}l}
+			                        \{\triple{(a,0)}{(r,0)}{(b,0)}|\ \triple{a}{r}{b}\in\triples\}\ \cup\\
+			                        \{\triple{(a',1)}{(r',1)}{(b',1)}|\ \triple{a'}{r'}{b'}\in\triples'\}
+								   \end{array}
+	\end{eqnarray}
+\end{definition}
+\begin{definition}[$\doteq$]
+	\begin{eqnarray}
+		\declare{nm}{A}{B}\doteq\declare{nm'}{A'}{B'}&\Leftrightarrow&\id{nm}=\id{nm'}\wedge A=A'\wedge B=B'
+	\end{eqnarray}
+\end{definition}
+\begin{definition}[migration dataset]
+	\begin{eqnarray}
+		&&\{ {\tt ENFORCE\ }r'{\tt := }r| r\in\rels, r'\in\rels', r\doteq r'\}\\
+		&\cup&\begin{array}[t]{@{}l}
+				\{ {\tt ENFORCE\ }\declare{nm'}{A'}{B'}{\tt := }\declare{nm}{A}{C};\ident{B}\\
+				|\ \begin{array}[t]{@{}l}
+					\declare{nm}{A}{C}\in\rels\wedge B\isa C\ \wedge\\
+					\declare{nm'}{A'}{B'}\in\rels'\wedge B'\not\isa C\ \wedge\\
+					\declare{nm}{A}{B}\doteq\declare{nm'}{A'}{B'}\}
+				   \end{array}
+			  \end{array}\\
+		&\cup&\begin{array}[t]{@{}l}
+				\{ {\tt ENFORCE\ }\declare{nm'}{A'}{C'}{\tt := }\ident{A};\declare{nm}{B}{C}\\
+				|\ \begin{array}[t]{@{}l}
+					\declare{nm}{B}{C}\in\rels\wedge A\isa B\ \wedge\\
+					\declare{nm'}{A'}{C'}\in\rels'\wedge A'\not\isa B\ \wedge\\
+					\declare{nm}{A}{C}\doteq\declare{nm'}{A'}{C'}\}
+				   \end{array}
+			  \end{array}
+	\end{eqnarray}
+\end{definition}
+\begin{definition}[disjoint union of information systems]
+	\begin{eqnarray}
+		\infsys\sqcup\infsys'&=&\la\roles\sqcup\roles',\rules\sqcup\rules',\maintain\sqcup\maintain',\dataset\sqcup\dataset'\ra\\
+		\roles\sqcup\roles'&=&\{(r,0)|\ r\in\roles\}\ \cup\ \{(r',1)|\ r'\in\roles'\}\\
+		\rules\sqcup\rules'&=&\{(r,0)|\ r\in\rules\}\ \cup\ \{(r',1)|\ r'\in\rules'\}\\
+		\maintain\sqcup\maintain'&=&\begin{array}[t]{@{}l}
+							  \{((o,0),(u,0))|\ (o,u)\in\maintain\}\ \cup\\
+			                  \{((o',1),(u',1))|\ (o',u')\in\maintain'\}
+							\end{array}
+	\end{eqnarray}
+\end{definition}
 

### 2022-09-16 2022 Migration/articleMigration.tex
Inserted a visualization of the migration
wijzigingen tijdens gesprek op 11 sept 2022
Tekst meer purposeful maken

@@
 
-\title{A Theory for Migration of Information Systems}
+\title{A Theory for Data Migration of Information Systems}
 \author[ou,ordina]{Stef Joosten\fnref{fn1}}
@@
 	In such cases, simply regenerating the system is not enough because it would reset the database to its initial state (deleting all data gathered so far).
-	Migration aims at preserving the data as much as possible.
+	Data migration aims at preserving the data as much as possible.
 	Typically, migration involves transferring the old data to the new system while preserving the semantics as much as possible.
@@
 \label{sct:Introduction}
-	The challenge of migration is to preserve meaning rather than just copy the data from one place to another.
+	Data migration occurs when one stateful information system, $\infsys$, is replaced by another, $\infsys'$,
+	while the data from $\infsys$ must be preserved.
+	Agile teams today need to automate their data migrations, 
+	as they do for all other activities in their software process.
+	By continually automating their own work,
+	teams keep improving their DevOps metrics such as deployment frequency, reliability of deployments, and change failure rate~\cite{DevOps2021}.
+
+	Data migrations can produce headaches, however.
+	The challenge of data migration is to preserve the meaning of data
+	rather than just copy the data from one place to another.
+	Accumulated data pollution can pose extra complications.
+	This makes data migrations far from trivial and a potential impediment for rapid and reliable deployment.
+
+	The benefits of automating data migrations work in two directions.
+	The automation itself saves effort and errors,
+	contributing to more frequent and smaller releases.
+	Smaller and faster releases also mean smaller migrations
+	and therefore a smaller difference between $\infsys$ and $\infsys'$.
+	The smaller that difference, the simpler the migration and the more of it can be automated.
+
+	A formal understanding of data migration will help to automate data migrations correctly and reliably.
+	That is why this paper proposes a theory of migration that is based on the
+	following premises:
+\begin{itemize}
+	\item the as-is data set may be polluted;
+	\item the data migration may require human interaction, which may take time;
+	\item the part of the data that does not change should be migrated automatically;
+	\item the meaning of data must be preserved;
+	\item the business must continu during the migration without interruption;
+	\item the part of the migration that can be automated is usually not sufficient;
+	      it takes additional human creativity to complete the migration specification.
+\end{itemize}
+
+	There have been other attempts to formalize data migration~\cite{Thalheim2013}.
+	(At this point: Elaborate on earlier work)
 
@@
 	Launch the new system in parallel to the old, copy data from the old to the new system, and have everyone use the new system.
-	We identify the following possible issues in this approach:
-
+	However, numerous issues might impede that plan.
+	The following examples illustrate the practical issues that may occur:
 \begin{enumerate}
@@
 
+(@Bas, het volgende zit al in de oplossingen sfeer. Willen we dat hier al doen?)
 	To mitigate these issues, we:
@@
 \label{sct:Terminology}
-	Migration is sometimes called data migration, because most people understand that data from "the old system" is fed into "the new system".
-	The idea is that the new system lets the business continu without interruption, after its software has been upgraded.
-	Since data is so central to migration, we start by introducing the notion of dataset.
-	To abstract away from a particular database implementation,
-	we represent data elements as triples.
-	This makes our theory valid for any kind of database that triples can represent,
-	such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases
-
-	Migration is not just about porting data.
-	It is about preserving the meaning of that data.
-	For that purpose, section~\ref{sct:Information Systems} introduces the notion of information system.
+	To migrate data from one information system to another,
+	we must define ``information system''.
+	In that definition we will single out the dataset as the focal point of a data migration.
+	For this reason, we define datasets first.
 
@@
 
-	Concepts are names we use in a conceptual model during the design of a dataset.
-	They are also used to assign signatures to relations, to enable type checking on the dataset.
+	Concepts are names we use during the design of a dataset
+	to name a group of atoms of the same type,
+	to assign signatures to relations, and thus
+	to enable type checking of the dataset.
 	For example, you might choose to classify \atom{Peter} as a \concept{Person}, and \atom{074238991} as a \concept{TelephoneNumber}.
@@
     We will use variables $A$, $B$, $C$, $D$ to represent concepts.
-	The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.
 	The expression $a\ \inst\ A$ means that atom $a$ is an \emph{instance} of concept $A$.
@@
 	Triples are used to represent data.
+	This makes our theory valid for any kind of database that triples can represent,
+	such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases
 	Each triple relates two atoms to a relation.
@@
 	In the formal world, we leave it entirely up to a user to attach meaning to a triple.
-	This ``meaning from practice'' has no consequences in the formal world.
+	As a consequence, this ``meaning from practice'' may have no consequences in the formal world.
 	The set of triples is called $\triples$.
@@
 	In this paper relations are represented by variables $r$, $s$, and $d$.
-
 	Every relation $r$ contains a set of pairs, which we call the population of $r$:
@@
 \end{equation}
+	The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.
+
 	Every relation has a name, a source concept, and a target concept.
@@
 	The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
-	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concept symbols $A$ and $B$.
+	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.
 	It contains all pairs that can be made of elements of $A$ and $B$:
@@
 	We can now define the notion of dataset.
-	Let	$\concepts$ be a set of concept symbols,
+	Let	$\concepts$ be a set of concepts,
 	let $\isa$ be a partial order on $\concepts$,
@@
 \label{sct:Information Systems}
-	Information systems process information from many different users,
-	all with their own tasks and responsibilities,
-	and all working from different locations and on different moments.
+	The purpose of an information system is to make data meaningful to users.
+	Data migration is all about preserving that meaning.
+	So this section introduces the notion of information system
+	as a system that consist of a dataset and a set constraints that users must keep satisfied.
+	These constraints reflect the meaning that is shared by all users.
+
+	Users have their own tasks and responsibilities
+	and may work from different locations and on different moments.
 	This collective use serves a purpose which we loosely call ``the business''.
 	As a consequence, the data in an information system changes continually.
-	To preserve meaning, users are trying to maintain semantic constraints on the data.
+	To preserve meaning, users are trying to maintain semantic constraints on the data,
+	amidst of all changes that are going on around them.
 	So, let us first define the constituent notions role and constraint before defining information systems proper.
@@
 
-\subsection{Migration}
+\subsection{Data migration}
 	To migrate information system $\infsys$ to $\infsys'$, we take the disjoint union $\infsys\sqcup\infsys'$ of both information systems
@@
 	It does, however, contain the concepts, relations, rules and roles of the new system.
+\begin{figure}[bht]
+	\begin{center}
+	  \includegraphics[scale=.45]{Migration.drawio.png}
+	\end{center}
+\caption{Data migration}
+\label{fig:event flow}
+\end{figure}
 
@@
 	\begin{eqnarray}
-		&&\{ {\tt ENFORCE\ }r'{\tt := }r| r\in\rels, r'\in\rels', r\doteq r'\}\\
-		&\cup&\begin{array}[t]{@{}l}
-				\{ {\tt ENFORCE\ }\declare{nm'}{A'}{B'}{\tt := }\declare{nm}{A}{C};\ident{B}\\
+		&&\begin{array}[t]{@{}l}
+			\{ {\tt ENFORCE\ }\declare{nm'}{A'}{B'}{\tt := }\ident{A'};\declare{nm}{A}{B};\ident{B'}\\
 				|\ \begin{array}[t]{@{}l}
-					\declare{nm}{A}{C}\in\rels\wedge B\isa C\ \wedge\\
-					\declare{nm'}{A'}{B'}\in\rels'\wedge B'\not\isa C\ \wedge\\
-					\declare{nm}{A}{B}\doteq\declare{nm'}{A'}{B'}\}
+					\declare{nm}{A}{B}\in\rels\wedge\declare{nm'}{A'}{B'}\in\rels'\wedge\id{nm}=\id{nm'}\ \wedge\\
+					\{\pair{A}{A'},\pair{B}{B'}\}\subseteq\isa\cup\isa'\cup\flip{\isa}\cup\flip{\isa'}\}
 				   \end{array}
@@
 			  &\cup&\begin{array}[t]{@{}l}
-				\{ {\tt ENFORCE\ }\declare{nm'}{A'}{C'}{\tt := }\ident{A};\declare{nm}{B}{C}\\
-				|\ \begin{array}[t]{@{}l}
-					\declare{nm}{B}{C}\in\rels\wedge A\isa B\ \wedge\\
-					\declare{nm'}{A'}{C'}\in\rels'\wedge A'\not\isa B\ \wedge\\
-					\declare{nm}{A}{C}\doteq\declare{nm'}{A'}{C'}\}
-				   \end{array}
+				\{ {\tt CLASSIFY\ }(C',1){\tt\ IS\ }(C,0)|\ C\in\concepts,C'\in\concepts',C=C'\}\\
 			  \end{array}

### 2022-09-17 2022 Migration/articleMigration.tex
Clarified disjoint union of datasets

@@
 		\forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\\
-		\forall A,B\in\concepts, a\in\atoms&:&A\isa B\wedge a\ \inst\ A\ \Rightarrow\ a\ \inst\ B\\
-		\forall r\in\rels&:&\src{r}\in\concepts\ \wedge\ \tgt{r}\in\concepts\\
-		\forall\triple{a}{r}{b}\in\triples&:&r\in\rels\ \wedge\ a\ \inst\ \src{r}\ \wedge\ b\ \inst\ \tgt{r}
+		%\forall A,B\in\concepts, a\in\atoms&:&A\isa B\wedge a\ \inst\ A\ \Rightarrow\ a\ \inst\ B\\
+		\forall \declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts\\
+		\forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&\begin{array}[t]{@{}l}\declare{n}{A}{B}\in\rels\ \wedge\\ a\ (\inst \compose \kleenestar{\flip{\isa}})\ A\ \wedge\\ b\ (\inst \compose \kleenestar{\flip{\isa}})\ B\end{array}
 	\end{eqnarray}
@@
 	\begin{center}
-	  \includegraphics[scale=.45]{Migration.drawio.png}
+	  \includegraphics[scale=.45]{Migration.png}
 	\end{center}
@@
 	\begin{eqnarray}
-		\dataset\sqcup\dataset'&=&\la\atoms\sqcup\atoms',\ \concepts\sqcup\concepts',\ \inst\sqcup\inst',\ \isa\sqcup\isa',\ \rels\sqcup\rels',\ \triples\sqcup\triples'\ra\\
-		\atoms\sqcup\atoms'&=&\{(a,0)|\ a\in\atoms\}\ \cup\ \{(a',1)|\ a'\in\atoms'\}\\
-		\concepts\sqcup\concepts'&=&\{(c,0)|\ c\in\concepts\}\ \cup\ \{(c',1)|\ c'\in\concepts'\}\\
-		\inst\sqcup\inst'&=&\begin{array}[t]{@{}l}
-							  \{((a,0),(A,0))|\ (a,A)\in\inst\}\ \cup\\
-			                  \{((a',1),(A',1))|\ (a',A')\in\inst'\}
-							\end{array}\\
-		\isa\sqcup\isa'&=&\begin{array}[t]{@{}l}
-							  \{((A,0),(B,0))|\ A\isa B\}\ \cup\\
-			                  \{((A',1),(B',1))|\ A'\isa' B'\}
-							\end{array}\\
-		\rels\sqcup\rels'&=&\{(r,0)|\ r\in\rels\}\ \cup\ \{(r',1)|\ r'\in\rels'\}\\
-		\triples\sqcup\triples'&=&\begin{array}[t]{@{}l}
-			                        \{\triple{(a,0)}{(r,0)}{(b,0)}|\ \triple{a}{r}{b}\in\triples\}\ \cup\\
-			                        \{\triple{(a',1)}{(r',1)}{(b',1)}|\ \triple{a'}{r'}{b'}\in\triples'\}
-								   \end{array}
+		\dataset\sqcup\dataset'&=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\ \inst\uplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \triples\uplus^5\triples'\ra\\
+		X\uplus Y&=&\{(x,0)|\ x\in X\}\ \cup\ \{(y,1)|\ y\in Y\}\\
+		X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))|\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))|\ (y_1,y_2)\in Y\}\end{array}\\
+		X\uplus^3 Y&=&\begin{array}[t]{@{}l}\{\declare{(n,0)}{(A,0)}{(B,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\declare{(n,1)}{(A,1)}{(B,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}\\
+		X\uplus^5 Y&=&\begin{array}[t]{@{}l}\{\triple{(a,0)}{\declare{(n,0)}{(A,0)}{(B,0)}}{(b,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\triple{(a,1)}{\declare{(n,1)}{(A,1)}{(B,1)}}{(b,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}
 	\end{eqnarray}
 \end{definition}
-\begin{definition}[$\doteq$]
-	\begin{eqnarray}
-		\declare{nm}{A}{B}\doteq\declare{nm'}{A'}{B'}&\Leftrightarrow&\id{nm}=\id{nm'}\wedge A=A'\wedge B=B'
-	\end{eqnarray}
-\end{definition}
+
 \begin{definition}[migration dataset]
@@
 		&&\begin{array}[t]{@{}l}
-			\{ {\tt ENFORCE\ }\declare{nm'}{A'}{B'}{\tt := }\ident{A'};\declare{nm}{A}{B};\ident{B'}\\
-				|\ \begin{array}[t]{@{}l}
+			\{ {\tt ENFORCE\ }\declare{(nm',1)}{(A',1)}{(B',1)}\\\quad{\tt := }\ident{(A',1)};\declare{(nm,0)}{(A,0)}{(B,0)};\ident{(B',1)}\\
+				\mid\ \begin{array}[t]{@{}l}
 					\declare{nm}{A}{B}\in\rels\wedge\declare{nm'}{A'}{B'}\in\rels'\wedge\id{nm}=\id{nm'}\ \wedge\\
@@
 	\begin{eqnarray}
-		\infsys\sqcup\infsys'&=&\la\roles\sqcup\roles',\rules\sqcup\rules',\maintain\sqcup\maintain',\dataset\sqcup\dataset'\ra\\
-		\roles\sqcup\roles'&=&\{(r,0)|\ r\in\roles\}\ \cup\ \{(r',1)|\ r'\in\roles'\}\\
-		\rules\sqcup\rules'&=&\{(r,0)|\ r\in\rules\}\ \cup\ \{(r',1)|\ r'\in\rules'\}\\
-		\maintain\sqcup\maintain'&=&\begin{array}[t]{@{}l}
-							  \{((o,0),(u,0))|\ (o,u)\in\maintain\}\ \cup\\
-			                  \{((o',1),(u',1))|\ (o',u')\in\maintain'\}
-							\end{array}
+		\infsys\sqcup\infsys'&=&\la\roles\uplus\roles',\rules\uplus^?\rules',\maintain\uplus^2\maintain',\dataset\sqcup\dataset'\ra
 	\end{eqnarray}

### 2022-09-17 2022 Migration/articleMigration.tex
Text erbij over de ENFORCE en CLASSIFY regels

@@
 \end{definition}
+Note that if $\dataset$ and $\dataset'$ are datasets, then so is $\dataset\sqcup\dataset'$.
 
@@
 		&&\begin{array}[t]{@{}l}
-			\{ {\tt ENFORCE\ }\declare{(nm',1)}{(A',1)}{(B',1)}\\\quad{\tt := }\ident{(A',1)};\declare{(nm,0)}{(A,0)}{(B,0)};\ident{(B',1)}\\
+			\{ {\tt ENFORCE\ }\declare{(nm',1)}{(A',1)}{(B',1)}\\\quad{\tt\ :=\ }\ident{(A',1)};\declare{(nm,0)}{(A,0)}{(B,0)};\ident{(B',1)}\\
 				\mid\ \begin{array}[t]{@{}l}
@@
 \end{definition}
+
+The ${\tt CLASSIFY\ }(C',1){\tt\ IS\ }(C,0)$ statements add $((C',1),(C,0))$ and  $((C,0),(C',1))$ to the $\isa$ relation.
+Note that if $\dataset$ is a dataset, and $\dataset'$ is that dataset with pairs added\footnote{TODO: het woord `add' is informeel, dit kan beter} to the $\isa$ relation, then the result is again a dataset.
+The ${\tt ENFORCE\ }r{\tt\ :=\ }e$ statements add triples $(x,r,y)$ for all $(x,y)\in e$.
+The expressions $e$ in these statements are such that $x\ (\inst \compose \kleenestar{\flip{\isa}})\ (A',1)$ and $y \ (\inst \compose \kleenestar{\flip{\isa}})\ (B',1)$, thus preserving the property that the result is a dataset.
+
 \begin{definition}[disjoint union of information systems]

### 2022-09-17 2022 Migration/articleMigration.tex
WIP

@@
 	as they do for all other activities in their software process.
-	By continually automating their own work,
-	teams keep improving their DevOps metrics such as deployment frequency, reliability of deployments, and change failure rate~\cite{DevOps2021}.
+	Teams continually automate to keep improving their performance without working any harder.
+	They measure the effects in terms of DevOps metrics such as
+	deployment frequency,
+	reliability of deployments, and
+	change failure rate~\cite{DevOps2021}.
 

### 2022-09-17 2022 Migration/articleMigration.tex
Add an example information system

@@
 
+\subsection{Example}
+	Having defined an information system in mathematical terms, let us give an example.
+	For this purpose we use a subset of the language Ampersand~\cite{Joosten-JLAMP2018}
+	because it makes the examples more appealing to read.
+	Let us first define a dataset of eight triples and three relations.
+\begin{verbatim}
+RELATION takes[Student*Course] =
+[ ("Peter", "Management")
+; ("Susan", "Business IT")
+; ("John", "Business IT")
+]
+\end{verbatim}
+	This code snippet introduces a relation with name \verb#takes#,
+	source concept \verb#Student#, and
+	target concept \verb#Course#.
+	It also introduces three triples:
+\[\begin{array}{l}
+	\triple{\text{\tt "Peter"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Management"}}\\
+	\triple{\text{\tt "Susan"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
+	\triple{\text{\tt "John"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
+\end{array}\]
+	The informal meaning of this relation is that it states which students are taking which courses.
+	So, there is a student, Peter, taking the course Management and two students, Susan and John, taking the course Business IT.
+	As stated before, this informal meaning is interesting only for the sake of illustration.
+
+	The example system also has a second relation that states which modules are part of which course:
+
+\begin{verbatim}
+RELATION isPartOf[Module*Course] =
+[ ("Finance", "Management")
+; ("Business Rules", "Business IT")
+; ("Business Analytics", "Business IT")
+; ("IT-Governance", "Business IT")
+; ("IT-Governance", "Management")
+]
+\end{verbatim}
+	The third and last relation states which students are enrolled for which module.
+	We leave it empty for now.
+\begin{verbatim}
+RELATION isEnrolledFor [Student*Module]
+\end{verbatim}
+	Summarizing, here we have a set of atoms, which contains \verb-"Peter"-, \verb-"Finance"-, etc.,
+	a set of concepts $\{$\verb-Student-, \verb-Module-, \verb-Course-$\}$,
+	an instance relation that states which atoms belong to which concept,
+	a specialization relation, which is empty,
+	a set of three relations, and
+	a set of eight triples.
+	The example in Ampersand defines precisely that dataset.
+
+	Now let us define an information system.
+	To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\triples$.
+	We define a constraint that states that a student can enroll for any module that is part of a course that student takes.
+	In logic, this constraint is written as:
+\[\begin{array}{l}\forall s\in\text{\tt Student}, m\in\text{\tt Module}\ \exists c\in\text{\tt Course}:\\
+s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
+\end{array}\]
+	In Ampersand, which is a syntactically sugared form of relation algebra,
+	we give that constraint a name and state a role to maintains it:
+\begin{verbatim}
+RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
+ROLE Administrator MAINTAINS EnrollRule
+\end{verbatim}
+	The signature of this rule is $\pair{\text{\tt Student}}{\text{\tt Module}}$, which is derived by the Ampersand compiler.
+	Ampersand also derives the violation set as:
+\[\begin{array}{l}
+	\viol{\triples}{\text{\tt EnrollRule}}\\
+	\hspace{1cm}=\{\pair{s}{m}|s\ \text{\tt isEnrolledFor}\ m\ \wedge\not\exists c: s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\}
+\end{array}\]
+	So, next to a dataset we have one rule, one role, a relation $\maintain$ that contains one pair.
+	So we have now defined an example information system.
 \subsection{Data migration}

### 2022-09-17 2022 Migration/articleMigration.tex
minor

@@
 \end{array}\]
-	So, next to a dataset we have one rule, one role, a relation $\maintain$ that contains one pair.
-	So we have now defined an example information system.
+	Notice that the current population yields an empty set of violations,
+	so the rule is satisfied.
+	Having one rule, one role, a relation $\maintain$ that contains one pair, and a dataset,
+	we have now defined a complete information system (albeit small).
+
 \subsection{Data migration}

### 2022-09-23 2022 Migration/articleMigration.tex
introduced rules and working on a second example for the to-be system.

@@
 	For this reason, we define datasets first.
+	We need rules to capture the semantics, so we need to define rules as well.
+	After that, we define information systems.
 
 \subsection{Datasets}
+\label{sct:Datasets}
 	A dataset describes a set of structured data, which is typically stored persistently in a database of some kind.
@@
 
+\subsection{Rules}
+\label{sct:Rules}
+	A \define{rule} is used to restrict the populations of a dataset.
+\begin{definition}[rule]
+	\label{def:rule}
+	\item A rule $u$ is a tuple $\la p_u,A,B,\id{viol}_u\ra$
+	in which $p_u$ is a constraint (i.e. a condition that constrains the triples of a dataset),
+	$\pair{A}{B}$ is the signature of $u$, and
+	$\id{viol}_u:\powerset{\triples}\rightarrow\powerset{\atoms\times\atoms}$ is a function that gives the violations (which are pairs) of this rule,
+	in which
+\begin{eqnarray}
+	\pair{a}{b}\in\viol{u}{\triples}&\Rightarrow&a\in A\wedge b\in B\\
+	p_u&\Leftrightarrow&\viol{u}{\triples}=\emptyset
+\end{eqnarray}
+\end{definition}
+	The function $\id{viol}_u$ computes the violations of $u$ for any set of triples $\triples$ in the dataset,
+	which allows us to check at any moment in time whether the current population is free of violations.
+	We say that a rule $u$ is \define{satisfied} in a dataset $\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$ iff $\viol{u}{\triples}=\emptyset$.
+	Since $\triples$ changes continually in an information system, the violations are changing along.
+	As a result, rules that are violated at one moment may be satisfied at another moment.
+
+	The set of triples changes due to events observed by the dataset.
+	These events may be caused by external systems, roles acting on the dataset, or any other force that changes triples.
+	Every rule is assigned to a role, so that role must ensure that the rule returns to a satisfied state after some event has caused the set of triples to change.
+	In other words: If anyone edits in the dataset and breaks a rule, someone has to ensure that the data is changed so that satisfies the rule.
+
 \subsection{Information Systems}
@@
 	Data migration is all about preserving that meaning.
-	So this section introduces the notion of information system
-	as a system that consist of a dataset and a set constraints that users must keep satisfied.
-	These constraints reflect the meaning that is shared by all users.
+	This section introduces the notion of information system
+	as a system that consist of a dataset and a set of rules that constrain that data,
+	together with a set of roles that must keep all rules satisfied.
+	These rules reflect the meaning (semantics) that is shared among users.
 
@@
 	amidst of all changes that are going on around them.
-	So, let us first define the constituent notions role and constraint before defining information systems proper.
+	So, let us first define the constituent notions role and rule before defining information systems proper.
 
@@
 
-	A \define{constraint} is a restriction on a dataset, to be enforced by a role.
-	The purpose of constraints is to ensure the semantics of a dataset, i.e. the preservation of meaning as data changes.
-	Each constraint $u\in\rules$ has a signature $\signat{u}$, which is a pair of concepts.
-	Each constraint $u$ also has a function $\id{viol}_u:\powerset{\atoms\times\rels\times\atoms}\ \rightarrow\ \powerset{\atoms\times\atoms}$.
-	We call the set $\viol{\triples}{u}$ the \define{violation set} of constraint $u$ in population $\triples$.
-	We call every element $\pair{a}{b}\in\viol{\triples}{u}$ a \define{violation} of constraint $u$ in $\triples$.
-	We say that a constraint \define{is satisfied} if its violation set is empty.
-	The set of constraints is called $\rules$.
-
-	Note that the signature of $u$ does not depend on the population, whereas the violation set of $u$ may be different for every $\triples$.
-	Since $\triples$ changes continually in an information system, the violations are changing along.
-
 	We can now define the notion of information system.
 	Let $\roles$ be a set of roles,
-	let $\rules$ be a set of constraints, 
+	let $\rules$ be a set of rules, 
 	let $\maintain$ be a surjective relation $\roles\times\rules$,
@@
 	with dataset $\dataset=\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$
-	in which $o\ \maintain\ u$ means that role $o$ must keep the set $\viol{\triples}{u}$ empty.
+	in which $o\ \maintain\ u$ means that role $o$ must keep the set $\viol{u}{\triples}$ empty.
 \end{definition}
-	If a constraint $u$ has violations, i.e. $\viol{\triples}{u}$ is not empty, there is a role whose job it is to remove those violations.
+	If a rule $u$ has violations, i.e. $\viol{u}{\triples}$ is not empty, there is a role whose job it is to remove those violations.
 	The relation $\maintain$ being surjective means that all users together are keeping the violation set empty.
@@
 \subsection{Example}
+\label{As-is IS}
 	Having defined an information system in mathematical terms, let us give an example.
@@
 	To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\triples$.
-	We define a constraint that states that a student can enroll for any module that is part of a course that student takes.
-	In logic, this constraint is written as:
+	We define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that student takes.
+	Its constraint, $p_{\tt Enroll}$ is written in logic as:
 \[\begin{array}{l}\forall s\in\text{\tt Student}, m\in\text{\tt Module}\ \exists c\in\text{\tt Course}:\\
@@
 	In Ampersand, which is a syntactically sugared form of relation algebra,
-	we give that constraint a name and state a role to maintains it:
+	we give rule a name and declare a role to maintains it:
 \begin{verbatim}
@@
 \end{verbatim}
-	The signature of this rule is $\pair{\text{\tt Student}}{\text{\tt Module}}$, which is derived by the Ampersand compiler.
-	Ampersand also derives the violation set as:
+	The Ampersand compiler derives the signature of every rule.
+	This rule has signature $\pair{\text{\tt Student}}{\text{\tt Module}}$.
+	The Ampersand compiler also derives the violation set as:
 \[\begin{array}{l}
-	\viol{\triples}{\text{\tt EnrollRule}}\\
+	\viol{\text{\tt EnrollRule}}{\triples}\\
 	\hspace{1cm}=\{\pair{s}{m}|s\ \text{\tt isEnrolledFor}\ m\ \wedge\not\exists c: s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\}
@@
 	Having one rule, one role, a relation $\maintain$ that contains one pair, and a dataset,
-	we have now defined a complete information system (albeit small).
+	we have now defined a small information system.
 
@@
 
-	We use two datasets: $\dataset=\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$ and $\dataset'=\la\atoms',\concepts',\inst',\isa',\rels',\triples'\ra$.
+	We use two datasets: $\dataset=$ and $\dataset'=$.
 	Before doing so, let us first define the disjoint union of two information systems.
@@
 \begin{eqnarray}
-		\dataset\sqcup\dataset'&=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\ \inst\uplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \triples\uplus^5\triples'\ra\\
+	\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\triples'\ra
+\end{eqnarray}
+\begin{eqnarray}
+	\omit\rlap{Some text that I want to appear here...}\\
+	&=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\ \inst\uplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \triples\uplus^5\triples'\ra\\
 		X\uplus Y&=&\{(x,0)|\ x\in X\}\ \cup\ \{(y,1)|\ y\in Y\}\\
@@
 		X\uplus^3 Y&=&\begin{array}[t]{@{}l}\{\declare{(n,0)}{(A,0)}{(B,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\declare{(n,1)}{(A,1)}{(B,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}\\
-		X\uplus^5 Y&=&\begin{array}[t]{@{}l}\{\triple{(a,0)}{\declare{(n,0)}{(A,0)}{(B,0)}}{(b,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\triple{(a,1)}{\declare{(n,1)}{(A,1)}{(B,1)}}{(b,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}
+		X\uplus^5 Y&=&\begin{array}[t]{@{}l}\{\triple{(a,0)}{\declare{(n,0)}{(A,0)}{(B,0)}}{(b,0)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in X\}\ \cup\\ \{\triple{(a,1)}{\declare{(n,1)}{(A,1)}{(B,1)}}{(b,1)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in Y\}\end{array}
 \end{eqnarray}
@@
 Note that if $\dataset$ and $\dataset'$ are datasets, then so is $\dataset\sqcup\dataset'$.
+
+\subsection{Example}
+	Let us proceed to specify a to-be information system, by way of example.
+	We will use the example in section~\ref{As-is IS} as the as-is information system.
+	The new information system keeps all three relations, but it constrains one relation, \verb-isPartOf-, to be univalent.
+	This means that every \verb-Module- is part of at most one \verb-Course-.
+	It restricts the relation \verb-isPartOf-,
+	so the migration might yield violations in the to-be information system that were not in the as-is information system.
+\begin{verbatim}
+RELATION takes[Student*Course]
+RELATION isPartOf[Module*Course] [UNI]
+RELATION isEnrolledFor[Student*Module]
+RELATION course[ExamReg*Course] [UNI]
+RELATION student[ExamReg*Student] [UNI]
+RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
+RULE ExamRule1: student~;course |- takes
+RULE ExamRule2: student~;course;isPartOf~ |- isEnrolledFor
+ROLE Administrator MAINTAINS EnrollRule
+\end{verbatim}
+	Part of the change is that exam registrations (concept \verb-ExamReg-) will be involved.
+	In an exam registration, a student registers for the examination of a course.
+	The to-be information information system contains two extra rules: ExamRule1 and ExamRule2.
+	The first one says that an exam registration requires that a student actually takes the course.
+	In logic, ExamRule1 is written as:
+\[\begin{array}{l}\forall s\in\text{\tt Student}, e\in\text{\tt ExamReg}, c\in\text{\tt Course}:\\
+	e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \rightarrow\ s\ \text{\tt takes}\ c\\
+\end{array}\]
+	Another requirement, ExamRule2, is that the student is enrolled for every module that is part of the course.
+	In logic its constraint, $p_{\tt ExamRule2}$ is written as:
+\[\begin{array}{l}\forall s\in\text{\tt Student}, e\in\text{\tt ExamReg}, m\in\text{\tt Module}, c\in\text{\tt Course}:\\
+	e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\ \rightarrow\ s\ \text{\tt isEnrolledFor}\ m
+\end{array}\]
+	Ampersand also derives the following violation sets:
+\[\begin{array}{l}
+	\viol{\text{\tt ExamRule1}}{\triples}\\
+	\hspace{1cm}=\{\pair{s}{c}\mid\exists e:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\neg(s\ \text{\tt takes}\ c)\}\\
+	\viol{\text{\tt ExamRule2}}{\triples}\\
+	\hspace{1cm}=\{\pair{s}{m}\mid\exists e,c:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\ \wedge\neg(s\ \text{\tt isEnrolledFor}\ m)\}
+\end{array}\]
 
@@
 
-\begin{definition}[disjoint union of information systems]
+\begin{definition}[]
 	\begin{eqnarray}
-		\infsys\sqcup\infsys'&=&\la\roles\uplus\roles',\rules\uplus^?\rules',\maintain\uplus^2\maintain',\dataset\sqcup\dataset'\ra
+		\infsys\sqcup\infsys'&=&\la\roles',\rules',\maintain',\dataset\sqcup\dataset'\ra
 	\end{eqnarray}

### 2022-09-25 2022 Migration/articleMigration.tex
minor cleanup

@@
 
-	We use two datasets: $\dataset=$ and $\dataset'=$.
+	We use two datasets: $\dataset$ and $\dataset'$.
 	Before doing so, let us first define the disjoint union of two information systems.
@@
 \begin{eqnarray}
-	\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\triples'\ra
-\end{eqnarray}
-\begin{eqnarray}
-	\omit\rlap{Some text that I want to appear here...}\\
-	&=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\ \inst\uplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \triples\uplus^5\triples'\ra\\
+	\omit\rlap{$\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\triples'\ra$}\notag\\
+	&=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\ \inst\uplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \triples\uplus^5\triples'\ra\notag\\
 		X\uplus Y&=&\{(x,0)|\ x\in X\}\ \cup\ \{(y,1)|\ y\in Y\}\\

### 2022-09-25 2022 Migration/articleMigration.tex
improved the storyline in the running example

@@
 \label{sct:Rules}
-	A \define{rule} is used to restrict the populations of a dataset.
+	Rules are used to restrict the populations of a dataset.
 \begin{definition}[rule]
@@
 	These events may be caused by external systems, roles acting on the dataset, or any other force that changes triples.
-	Every rule is assigned to a role, so that role must ensure that the rule returns to a satisfied state after some event has caused the set of triples to change.
-	In other words: If anyone edits in the dataset and breaks a rule, someone has to ensure that the data is changed so that satisfies the rule.
+	Every rule is assigned to a role.
+	That role must ensure to satisfy that rule after some event has caused violations by changing the set of triples.
+	In other words: If anyone edits in the dataset and breaks a rule, someone or something has to fix the dataset to satisfy that rule again.
+	Example: if a rule says that every request must be acknowledged, the receipt of a request will break this rule.
+	Sending an acknowledgement message will cause the rule to be satisfied again.
 
@@
 	with dataset $\dataset=\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$
-	in which $o\ \maintain\ u$ means that role $o$ must keep the set $\viol{u}{\triples}$ empty.
 \end{definition}
+	The relation $\maintain$ links rules to roles.
+	$o\ \maintain\ u$ means that role $o$ must keep the set $\viol{u}{\triples}$ empty.
 	If a rule $u$ has violations, i.e. $\viol{u}{\triples}$ is not empty, there is a role whose job it is to remove those violations.
 	The relation $\maintain$ being surjective means that all users together are keeping the violation set empty.
-	Which means that all constraints are being kept satisfied in the collective effort of users, even if the data changes.
-	In another section we will give some options on how users can keep rules satisfied.
 
@@
 	because it makes the examples more appealing to read.
-	Let us first define a dataset of eight triples and three relations.
+	Let us first define a dataset of seven triples and three relations.
 \begin{verbatim}
@@
 \end{verbatim}
-	This code snippet introduces a relation with name \verb#takes#,
+	This introduces a relation with name \verb#takes#,
 	source concept \verb#Student#, and
@@
 
-	The example system also has a second relation that states which modules are part of which course:
-
+	The example system also has a second relation that states which modules are part of which course.
+	It is constrained by \verb-[UNI]- to be univalent.
+	This means that every \verb-Module- is part of at most one \verb-Course-.
 \begin{verbatim}
-RELATION isPartOf[Module*Course] =
+RELATION isPartOf[Module*Course] [UNI] =
 [ ("Finance", "Management")
@@
 ; ("Business Analytics", "Business IT")
-; ("IT-Governance", "Business IT")
 ; ("IT-Governance", "Management")
@@
 	In Ampersand, which is a syntactically sugared form of relation algebra,
-	we give rule a name and declare a role to maintains it:
+	we give each rule a name and declare a role to maintain it:
 \begin{verbatim}
@@
 \subsection{Example}
-	Let us proceed to specify a to-be information system, by way of example.
-	We will use the example in section~\ref{As-is IS} as the as-is information system.
-	The new information system keeps all three relations, but it constrains one relation, \verb-isPartOf-, to be univalent.
-	This means that every \verb-Module- is part of at most one \verb-Course-.
-	It restricts the relation \verb-isPartOf-,
-	so the migration might yield violations in the to-be information system that were not in the as-is information system.
+	Let us proceed to specify a to-be information system, $\infsys'$, to illustrate a (toy) migration.
+	We will use the example in section~\ref{As-is IS} as the as-is information system
+	and call that $\infsys$ in this section.
+	Its population reflects the state of the information system just before the migration.
+
+	Information system $\infsys'$, the to-be information system, is specified by:
 \begin{verbatim}
 RELATION takes[Student*Course]
-RELATION isPartOf[Module*Course] [UNI]
+RELATION isPartOf[Module*Course] [UNI] =
+   [ ("IT-Governance", "Business IT") ]
 RELATION isEnrolledFor[Student*Module]
-RELATION course[ExamReg*Course] [UNI]
-RELATION student[ExamReg*Student] [UNI]
+RELATION course[ExamReg*Course] [UNI] =
+   [ ("ER1", "Management")
+   ; ("ER2", "Business IT")
+   ; ("ER3", "Business IT")
+   ]
+RELATION student[ExamReg*Student] [UNI] =
+   [ ("ER1", "Peter")
+   ; ("ER2", "Susan")
+   ]
 RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
@@
 \end{verbatim}
-	Part of the change is that exam registrations (concept \verb-ExamReg-) will be involved.
+	This specification shows that $\infsys'$ gets to keep all three relations and the rule from $\infsys$.
+	Some new population, two new relations, and two new rules have been added to $\infsys'$.
+	$\infsys'$ contains exam registrations (concept \verb-ExamReg-), which is new.
 	In an exam registration, a student registers for the examination of a course.
-	The to-be information information system contains two extra rules: ExamRule1 and ExamRule2.
+	$\infsys'$ contains two extra rules: \verb-ExamRule1- and \verb-ExamRule2-.
 	The first one says that an exam registration requires that a student actually takes the course.
-	In logic, ExamRule1 is written as:
+	In logic, the constraint $p_{\tt ExamRule1}$ is written as:
 \[\begin{array}{l}\forall s\in\text{\tt Student}, e\in\text{\tt ExamReg}, c\in\text{\tt Course}:\\
@@
 \end{array}\]
-	Another requirement, ExamRule2, is that the student is enrolled for every module that is part of the course.
-	In logic its constraint, $p_{\tt ExamRule2}$ is written as:
+	Another requirement, \verb-ExamRule2-, is that the student is enrolled for every module that is part of the course.
+	Its constraint, $p_{\tt ExamRule2}$, is written in logic as:
 \[\begin{array}{l}\forall s\in\text{\tt Student}, e\in\text{\tt ExamReg}, m\in\text{\tt Module}, c\in\text{\tt Course}:\\
@@
 \end{array}\]
+
+	The intention of any migration is to preserve as much of the data from $\infsys$ as possible into $\infsys'$,
+	while adding the new triples to $\infsys'$.
+	Any data migration must anticipate any population in the as-is information system.
+	The only thing we know for sure about the ``old'' population is that it satisfies the rules in $\infsys$.
 

### 2022-09-25 2022 Migration/articleMigration.tex
take disjoint union in the example,
to elaborate this example a little further.

@@
 
+   So let us now turn to the migration.
    The intention of any migration is to preserve as much of the data from $\infsys$ as possible into $\infsys'$,
    while adding the new triples to $\infsys'$.
+   So the first thing to do is to take the disjoint union of both information systems.
+   This yields:
+\begin{verbatim}
+   RELATION takes[Student*Course] =
+      [ ("Peter", "Management")
+      ; ("Susan", "Business IT")
+      ; ("John", "Business IT")
+      ]
+
+   RELATION isPartOf[Module*Course] [UNI]=
+      [ ("Finance", "Management")
+      ; ("Business Rules", "Business IT")
+      ; ("Business Analytics", "Business IT")
+      ; ("IT-Governance", "Management")
+      ; ("IT-Governance", "Business IT")
+      ]
+
+   RELATION isEnrolledFor [Student*Module] =
+      [ ("Susan", "Business Analytics")
+      ; ("Susan", "IT-Governance")
+      ; ("Susan", "Business Rules")
+      ]
+
+   RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
+   ROLE Administrator MAINTAINS EnrollRule
+
+   RELATION course[ExamReg*Course] [UNI] =
+      [ ("ER1", "Management")
+      ; ("ER2", "Business IT")
+      ; ("ER3", "Business IT")
+      ]
+   RELATION student[ExamReg*Student] [UNI] =
+      [ ("ER1", "Peter")
+      ; ("ER2", "Susan")
+      ]
+   RULE ExamRule1: student~;course |- takes
+   RULE ExamRule2: student~;course;isPartOf~ |- isEnrolledFor
+\end{verbatim}
+   If we check this for violations, we get the following results:
+\begin{verbatim}
+   There are 2 violations of RULE "ExamRule2":
+      ("Peter", "IT-Governance")
+      ("Peter", "Finance")
+   ==============================
+   There is one violation of RULE "UNI isPartOf[Module*Course]":
+      ("IT-Governance", "IT-Governance")
+\end{verbatim}
    Any data migration must anticipate any population in the as-is information system.

### 2022-09-25 2022 Migration/articleMigration.tex
minor

@@
    RELATION takes[Student*Course]
-RELATION isPartOf[Module*Course] [UNI] =
+   RELATION isPartOf[Module*Course] [UNI]=
       [ ("IT-Governance", "Business IT") ]
-RELATION isEnrolledFor[Student*Module]
+   RELATION isEnrolledFor [Student*Module] =
+      [ ("Susan", "Business Analytics")
+      ; ("Susan", "IT-Governance")
+      ; ("Susan", "Business Rules")
+      ]
+   RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
+   ROLE Administrator MAINTAINS EnrollRule
+
    RELATION course[ExamReg*Course] [UNI] =
@@
       ]
-RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
    RULE ExamRule1: student~;course |- takes
    RULE ExamRule2: student~;course;isPartOf~ |- isEnrolledFor
-ROLE Administrator MAINTAINS EnrollRule
 \end{verbatim}

### 2022-10-02 2022 Migration/articleMigration.tex
minor
Todo's voor Stef en mij
Isa is a rule

@@
 \include{preambleMigrations}
+
+% TODO Bas:
+% - verhaal over incrementeel fouten oplossen (fouten staan al in artikel)
+
+% TODO Algemeen:
+% - type van triple-set en atom-set definieren zodat viol_u ?->P(?x?)
+% - nieuwe populatie na de disjoint union zou de populatie met ENFORCE regels toegepast moeten zijn
+
+% TODO Stef:
+% - stukje literatuuronderzoek
+% https://ieeexplore.ieee.org/abstract/document/7445334?casa_token=ECzi6XeV2ncAAAAA:KhWzB8XBFOUJ0C6AD-XjX_ryuA9ARvTd3gm6RR-ZNiR8sZ1858FJpQ7zKQhkAZDlv8IjPdgD
+% https://ieeexplore.ieee.org/abstract/document/8549944?casa_token=9qiGqNzh2Q0AAAAA:C-cYogExB35nGxQdxLcdBh4JoLNvM0OHedAMhCbB5V4kb4_6nzHUvc23xSJbeoBu67LSiz-Y
+% https://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.651.9298&rep=rep1&type=pdf
+% https://www.scirp.org/html/4-7800724_106592.htm
+% https://www.researchgate.net/profile/Ranjana-Badre/publication/318665687_GUI_for_Data_Migration_and_Query_Conversion/links/5b45bbea0f7e9b1c722386e5/GUI-for-Data-Migration-and-Query-Conversion.pdf
+% https://journal3.uin-alauddin.ac.id/index.php/literatify/article/view/12567
 
@@
    The expression $a\ \inst\ A$ means that atom $a$ is an \emph{instance} of concept $A$.
-   The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
-   We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
-   Specialization is needed to allow statements such as: ``A woman is a human'' or ``A human is a mammal''.
    The set of concepts is called $\concepts$.
@@
 \begin{equation}
-   \pop{r}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\triples\}
+   \pop{\triples}{r}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\triples\}
 \end{equation}
@@
    It contains all pairs that can be made of elements of $A$ and $B$:
-\[\pop{\fullt{A}{B}} = \{ \pair{a}{b}| a\ \inst\ A,\ b\ \inst\ B \}\]
+\[\pop{\triples}{\fullt{A}{B}} = \{ \pair{a}{b}| a\ \inst\ A,\ b\ \inst\ B \}\]
    The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.
@@
    \label{def:dataset}
-   \item A dataset is a tuple $\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$ with
+   \item A dataset is a tuple $\la\atoms,\concepts,\inst,\rels,\triples\ra$ with
    \begin{eqnarray}
@@
       \forall \declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts\\
-      \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&\begin{array}[t]{@{}l}\declare{n}{A}{B}\in\rels\ \wedge\\ a\ (\inst \compose \kleenestar{\flip{\isa}})\ A\ \wedge\\ b\ (\inst \compose \kleenestar{\flip{\isa}})\ B\end{array}
+      \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&\begin{array}[t]{@{}l}\declare{n}{A}{B}\in\rels\ \wedge\ a\ \inst\ A\ \wedge\ b\ \inst\ B\end{array}
    \end{eqnarray}
@@
 \label{sct:Rules}
-   Rules are used to restrict the populations of a dataset.
+   Rules are used to describe whether certain (assumably desirable) properties hold in an information system.
+
 \begin{definition}[rule]
    \label{def:rule}
-   \item A rule $u$ is a tuple $\la p_u,A,B,\id{viol}_u\ra$
-   in which $p_u$ is a constraint (i.e. a condition that constrains the triples of a dataset),
-   $\pair{A}{B}$ is the signature of $u$, and
-   $\id{viol}_u:\powerset{\triples}\rightarrow\powerset{\atoms\times\atoms}$ is a function that gives the violations (which are pairs) of this rule,
+   \item A rule $u$ is a tuple $\la A,B,\id{viol}_u\ra$.
+   %in which $p_u$ is a constraint (i.e. a condition that constrains the triples of a dataset),
+   $\pair{A}{B}$ is the signature of $u$ and
+   $\id{viol}_u$ is a function that gives the violations (which are pairs) of this rule,
    in which
 \begin{eqnarray}
-   \pair{a}{b}\in\viol{u}{\triples}&\Rightarrow&a\in A\wedge b\in B\\
-   p_u&\Leftrightarrow&\viol{u}{\triples}=\emptyset
+   \pair{a}{b}\in\viol{u}{\triples,\inst}&\Rightarrow&a\ \inst\ A\wedge b\ \inst\ B
 \end{eqnarray}
 \end{definition}
-   The function $\id{viol}_u$ computes the violations of $u$ for any set of triples $\triples$ in the dataset,
+
+   The function $\id{viol}_u$ computes the violations of $u$,
    which allows us to check at any moment in time whether the current population is free of violations.
-   We say that a rule $u$ is \define{satisfied} in a dataset $\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$ iff $\viol{u}{\triples}=\emptyset$.
+   We say that a rule $u$ is \define{satisfied} in a dataset $\la\atoms,\concepts,\inst,\rels,\triples\ra$ iff $\viol{u}{\triples}=\emptyset$.
+   This allows us to define a predicate $p_u$ that indicates if a rule $u$ is satisfied:
+\begin{eqnarray}
+   p_u(\triples,\inst)&\Leftrightarrow&\viol{u}{\triples,\inst}=\emptyset
+\end{eqnarray}
+
    Since $\triples$ changes continually in an information system, the violations are changing along.
    As a result, rules that are violated at one moment may be satisfied at another moment.
-
-   The set of triples changes due to events observed by the dataset.
+   The set of triples changes due to events observed by the information system.
    These events may be caused by external systems, roles acting on the dataset, or any other force that changes triples.
@@
    Sending an acknowledgement message will cause the rule to be satisfied again.
+
+   Ampersand supports a special kind of rule that is used for specialization.
+   The rule $A\isa B = \la A, A, viol_{A\isa B} \ra$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
+   Instances of $A$ that are not an instance of $B$ are the violations of this rule: $viol_{A\isa B}(\triples,\inst) = \{(a,a) \mid a\ \inst\ A \wedge \lnot (a\ \inst\ B)\}$.
+   We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
+   Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.
+
+      Specialization rules are special for two reasons:
+   first, $A\isa B$ is automatically maintained by the information system itself, by changing $\inst$ to include any violating pairs in $B$.
+   This means that an information system can never be brought into a state in which $p_{A\isa B}$ is False if $A\isa B$ is a specialization rule of the information system, allowing for certain run-time optimizations.
+   A second property that sets specialization rules apart, is that they play a role in the Ampersand compiler.
+   For example, code that determines equality between atoms only works for atoms that are an instance of the same concept.
+   To ensure that this does not result in problems at run-time, users are required to write specialization rules that make these comparisons possible, and the type checker in the Ampersand compiler will spot these issues if they occur and refuse to generate the information system.
 
@@
    together with a set of roles that must keep all rules satisfied.
-   These rules reflect the meaning (semantics) that is shared among users.
+   %These rules reflect the meaning (semantics) that is shared among users.
 
@@
    In the sequel, when we talk about a role we actually mean a user (an arbitrary one) who fulfills that role in the information system.
-   The set of roles is called $\roles$.
 
    We can now define the notion of information system.
-   Let $\roles$ be a set of roles,
-   let $\rules$ be a set of rules, 
-   let $\maintain$ be a surjective relation $\roles\times\rules$,
-   and let $\dataset$ be a dataset.
 \begin{definition}[information system]
 \label{def:information system}
-\item An information system $\infsys$ is a tuple $\la\roles,\rules,\maintain,\dataset\ra$\\
-   with dataset $\dataset=\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$
+\item An information system $\infsys$ is a tuple $\la\roles,\rules,\specrules,\maintain,\dataset\ra$.
+   Here $\roles$ is a set of roles,
+   $\rules$ a set of rules,
+   $\specrules$ a set of specialization rules,
+   $\maintain$ is a relation $\roles\times\rules$,
+   and $\dataset$ is a dataset.
 \end{definition}

### 2022-10-06 2022 Migration/articleMigration.tex
Over incrementeel fouten oplossen

@@
 % - type van triple-set en atom-set definieren zodat viol_u ?->P(?x?)
+%   (Bas: ik heb foute types weggehaald, maar dit nog niet gedefinieerd)
 % - nieuwe populatie na de disjoint union zou de populatie met ENFORCE regels toegepast moeten zijn
+%  (en in het bijzonder de ISA regels)
 
@@
 
+\subsubsection{Strategies for dealing with violations}
+
+The violations of ExamRule2 don't pose a problem: they have a role (administrator) assigned to them.
+We allow these violations to occur in the transition to the new system, and expect people with the Administrator role to deal with them.
+
+However, the violations that arise when taking the (non-disjoint) union of the two scripts include violations of rules that the information system should maintain.
+This means that if we cannot generate software based on the disjoint union as it is.
+To resolve this, the simplest solution is to change the roles assigned to the rules that are violated in the disjoint union:
+\begin{verbatim}
+   RELATION isPartOf[Module*Course]
+   RULE isPartOfUNI: isPartOf;isPartOf~ |- I
+   ROLE migrationHelper MAINTAINS isPartOfUNI
+\end{verbatim}
+
+Naturally, this solution requires effort from people with the role migrationHelper.
+This in itself is not an issue: problems need to be solved one way or another.
+However, a problem that may occur is that regular users of the system, who are used to the old system, might be adding data to the system that causes violations faster than they can be solved.
+To mitigate this, we propose two solutions.
+
+One solution is to state that the only violations that may occur in \verb=isPartOfUNI= are those that were present when the migration started:
+\begin{verbatim}
+   RELATION isPartOf[Module*Course]
+   RULE isPartOfUNI: isPartOf;isPartOf~ |- I
+   ROLE migrationHelper MAINTAINS isPartOfUNI
+
+   RELATION isPartOfUNI_violations[Module*Module]
+     = [("IT-Governance", "IT-Governance")]
+   RULE isPartOfUNI_progress: -(isPartOf;isPartOf~) /\ I |- isPartOfUNI_violations
+   SYSTEM MAINTAINS isPartOfUNI_progress
+
+   ENFORCE isPartOfUNI_violations >: isPartOfUNI_progress: -(isPartOf;isPartOf~) /\ I  
+\end{verbatim}
+
+In this first solution, the first rule is the same as with the simple solution.
+The second rule prevents any new violation from occurring.
+Existing violations are recorded in the \verb=isPartOfUNI_violations= relation, and the rule \verb=isPartOfUNI_progress= prevents the set of violations to \verb=isPartOfUNI= from growing beyond this.
+The \verb=ENFORCE= statement at the end of this code snippet states that the relation recording existing violations should be shrunk whenever violations are solved.
+This way, we ensure that violations cannot re-occur.
+As a whole, this means that \verb=migrationHelper= has a finite task.
+
+As a second solution, we observe that the data in the designed system does not have any violations.
+Consequentially, the cause of the violations comes from the triples in the old system.
+Rather than taking the union of all triples, we can take the disjoint union instead;
+\begin{verbatim}
+   RELATION isPartOf_old[Module*Course] [UNI]=
+      [ ("Finance", "Management")
+      ; ("Business Rules", "Business IT")
+      ; ("Business Analytics", "Business IT")
+      ; ("IT-Governance", "Management")
+      ]
+   RELATION isPartOf[Module*Course] [UNI]=
+      [ ("IT-Governance", "Business IT") ]
+   RELATION isPartOf_ignored[Module*Course]
+
+   RULE isPartOf_isUnion:  isPartOf \/ isPartOf_ignored :- isPartOf_old
+   ROLE migrationHelper MAINTAINS isPartOf_isUnion
+
+   ENFORCE isPartOf_ignored :< isPartOf
+\end{verbatim}
+
+In this second solution, we immediately move to a system in which the \verb=UNI= rule is maintained by the system.
+The work for the \verb=migrationHelper= is again finite: the tuples in \verb=isPartOf_old= need to be copied over or explicitly ignored.
+Once a tuple is in \verb=isPartOf=, the tuple is copied over to \verb=isPartOf_ignored= by the \verb=ENFORCE= statement, such that regular users can delete it without adding work to anyone with the \verb=migrationHelper= role.
+Of course, since this solution starts with fewer pairs in the \verb=isPartOf= relation, other rules could be violated as a result of choosing this solution.
+Indeed, ExamRule2 initially has more violations in this scenario.
+
 \begin{definition}[migration dataset]
@@
    In the migration of a dataset we deal with changes to the elements that are not data:
-   $\isa$, $\inst$, $\concepts$, and $\rels$.
+   $\inst$, $\concepts$, and $\rels$.
    Such changes have further reaching consequences, however.
-   Changes to $\isa$, $\inst$, $\concepts$, $\rels$ change the data structure in a dataset.
-   We define a \define{migration of a dataset} $\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$ as a change in which one or more of $\isa$, $\inst$, $\concepts$, or $\rels$ change.
+   Changes to $\inst$, $\concepts$, $\rels$ change the data structure in a dataset.
+   We define a \define{migration of a dataset} $\la\atoms,\concepts,\inst,\rels,\triples\ra$ as a change in which one or more of $\inst$, $\concepts$, or $\rels$ change.
    To satisfy definition~\ref{def:dataset}, $\triples$ and $\atoms$ must change too,

### 2022-10-09 2022 Migration/articleMigration.tex
aantekeningen 9 oktober

@@
 
-   ENFORCE isPartOfUNI_violations >: isPartOfUNI_progress: -(isPartOf;isPartOf~) /\ I  
+   ENFORCE isPartOfUNI_violations :< -(isPartOf;isPartOf~) /\ I  
 \end{verbatim}
@@
 
-   RULE isPartOf_isUnion:  isPartOf \/ isPartOf_ignored :- isPartOf_old
+   RULE isPartOf_isUnion:  isPartOf_old :- isPartOf \/ isPartOf_ignored
    ROLE migrationHelper MAINTAINS isPartOf_isUnion
 
-   ENFORCE isPartOf_ignored :< isPartOf
+   ENFORCE isPartOf_ignored :> isPartOf
 \end{verbatim}
 
+\begin{verbatim}
+   RELATION isPartOf_old[Module*Course] [UNI]=
+      [ ("Finance", "Management")
+      ; ("Business Rules", "Business IT")
+      ; ("Business Analytics", "Business IT")
+      ; ("IT-Governance", "Management")
+      ]
+   RELATION isPartOf[Module*Course] [UNI]=
+      [ ("IT-Governance", "Business IT") ]
+   RELATION isPartOf_ignored[Module*Course]
+
+   ENFORCE isPartOf :> isPartOf_old - isPartOf_ignored
+   ENFORCE isPartOf_ignored :> isPartOf
+\end{verbatim}
 In this second solution, we immediately move to a system in which the \verb=UNI= rule is maintained by the system.

### 2022-10-15 2022 Migration/articleMigration.tex
working on the introduction

@@
 \label{sct:Introduction}
-   Data migration occurs when one stateful information system, $\infsys$, is replaced by another, $\infsys'$,
-   while the data from $\infsys$ must be preserved.
-   Agile teams today need to automate their data migrations, 
-   as they do for all other activities in their software process.
-   Teams continually automate to keep improving their performance without working any harder.
-   They measure the effects in terms of DevOps metrics such as
+   The general aim of this work is to generate information systems incrementally, while in production.
+   Our research focuses on generating information system,
+   and specifically on the data migration aspect of it.
+   The reason that we believe in generating information systems is to prevent human errors as early as possible in the build-chain:
+   A mistake not made cannot propagate into production.
+
+   Incremental generation of information systems poses specific requirements to data migration.
+   For instance, if data migration is not for incremental generation,
+   e.g.~\cite{Gholami2016,Bisbal1999},
+   preserving the functionality can be useful to avoid introducing new errors in an otherwise error-prone process.
+   For incremental generation, however, preserving the functionality is not an option.
+   To change the functionality is the very purpose and the data migration must behave accordingly.
+   So, we must make sure to use methods for data migration that are specifically meant for incrementally generating information systems.
+
+   Data migration occurs when an as-is information system, $\infsys$, is replaced by a to-be information system, $\infsys'$,
+   while preserving the meaning of the data from $\infsys$ as much as possible.
+   Just copying the set of data from $\infsys$ to $\infsys'$ is clearly wrong if the schemas of both information systems differ.
+   Information systems are typically used by distributed actors (both users and computers) who continually work with data.
+   The entire set of data, $\dataset$, in an information system represents the state of the system.
+   Every event that an actor causes may change the state $\dataset$.
+   We will write $\dataset_{\infsys}$ in case we need to be specific about which information the dataset belongs to.
+
+   In this work we will encode the semantics in a schema, $\schema$,
+   so we can be explicit about "preserving the meaning as much as possible".
+   The schema $\schema$ of an information system is basically a set of constraints on its dataset $\dataset$.
+   Let us assume that information systems $\infsys$ and $\infsys'$ have different schemas
+   because the task of migrating data is trivial if the schema doesn’t change.
+   We will denote the schema of an information system $\infsys$ by $\schema_{\infsys}$
+   if we need to be specific about the information system of which $\schema$ is the schema.
+
+   We believe that information systems should be released into production in small increments,
+   to obtain shorter release cycles, more frequent releases, and thus a more agile software development process%
+   \footnote{The effects can be measured in terms of DevOps metrics such as
    deployment frequency,
    reliability of deployments, and
-   change failure rate~\cite{DevOps2021}.
-
-   Data migrations can produce headaches, however.
-   The challenge of data migration is to preserve the meaning of data
-   rather than just copy the data from one place to another.
-   Accumulated data pollution can pose extra complications.
-   This makes data migrations far from trivial and a potential impediment for rapid and reliable deployment.
+   change failure rate~\cite{DevOps2021}.}.
+   And, if we want to deploy software updates more frequently, this requires updates with zero down-time.
+   Extra complications arise from accumulated data pollution in the as-is information system.
+   So, the theory we develop in this paper must lead to tools that satisfy such practical requirements as
+   zero down-time updates and ways to cope with data pollution.
+   That will help agile teams today to automate their software processes further.
 

### 2022-10-15 2022 Migration/articleMigration.tex
explain why a third schema comes into play

@@
 \label{sct:Introduction}
-   The general aim of this work is to generate information systems incrementally, while in production.
-   Our research focuses on generating information system,
-   and specifically on the data migration aspect of it.
-   The reason that we believe in generating information systems is to prevent human errors as early as possible in the build-chain:
+   Our research focuses on generating information systems incrementally,
+   especially on the data migration aspect of it.
+   The reason that we believe in generating information systems is to prevent human errors as early as possible in the build chain:
    A mistake not made cannot propagate into production.
-
    Incremental generation of information systems poses specific requirements to data migration.
-   For instance, if data migration is not for incremental generation,
+   For instance, if a data migration is not for incremental generation,
    e.g.~\cite{Gholami2016,Bisbal1999},
@@
    while preserving the meaning of the data from $\infsys$ as much as possible.
-   Just copying the set of data from $\infsys$ to $\infsys'$ is clearly wrong if the schemas of both information systems differ.
+   Just copying the set of data from $\infsys$ to $\infsys'$ is obviously wrong if the schemas of both information systems differ.
    Information systems are typically used by distributed actors (both users and computers) who continually work with data.
    The entire set of data, $\dataset$, in an information system represents the state of the system.
-   Every event that an actor causes may change the state $\dataset$.
-   We will write $\dataset_{\infsys}$ in case we need to be specific about which information the dataset belongs to.
+   Every event that an actor causes in information system $\infsys$ may change the state $\dataset$.
+   We will write $\dataset_{\infsys}$ in case we need to be specific about which information system the dataset belongs to,
+   but just $\dataset$ if it is obvious or irrelevant to which information system this dataset belongs.
 
-   In this work we will encode the semantics in a schema, $\schema$,
-   so we can be explicit about "preserving the meaning as much as possible".
-   The schema $\schema$ of an information system is basically a set of constraints on its dataset $\dataset$.
-   Let us assume that information systems $\infsys$ and $\infsys'$ have different schemas
-   because the task of migrating data is trivial if the schema doesn’t change.
-   We will denote the schema of an information system $\infsys$ by $\schema_{\infsys}$
-   if we need to be specific about the information system of which $\schema$ is the schema.
+   In this work we will encode the semantics in a schema,
+   so we can be explicit about ``preserving the meaning as much as possible''.
+   The schema of an information system is basically a set of constraints on its dataset $\dataset$.
+   We will denote the schema of an information system $\infsys$ by $\schema_{\infsys}$,
+   or just $\schema$ if it is obvious or irrelevant to which information system this schema belongs.
 
@@
    to obtain shorter release cycles, more frequent releases, and thus a more agile software development process%
-   \footnote{The effects can be measured in terms of DevOps metrics such as
+   \footnote{These effects can be measured in terms of DevOps metrics such as
    deployment frequency,
@@
    change failure rate~\cite{DevOps2021}.}.
-   And, if we want to deploy software updates more frequently, this requires updates with zero down-time.
+   In practice, more frequent updates require that updates have zero down-time.
    Extra complications arise from accumulated data pollution in the as-is information system.
@@
    That will help agile teams today to automate their software processes further.
+
+   Automating data migrations, however, has its limits.
+   If the semantics encoded in the to-be schema $\schema'$ differs from the as-is schema $\schema$,
+   interaction with a migration engineer may be needed to determine how to interpret existing data in the new schema.
+   This may even involve adapting data of which the constraints in the schema have not changed.
+   This requires a third schema, $\schema_m$, which specifies the migration itself.
+   We generate $\schema_m$ in the form of source code, so the migration engineer can change everything she wants to suit the particulars of the data migration.
+   In the sequel, automation of data migration means to generate the schema $\schema_m$.
 
@@
 
+   The contribution of this paper is to propose a formalism in which $\schema_m$ is defined.
    A formal understanding of data migration will help to automate data migrations correctly and reliably.
-   That is why this paper proposes a theory of migration that is based on the
-   following premises:
+   Our theory is based on the following premises:
 \begin{itemize}
-   \item the as-is data set may be polluted;
+   \item the as-is data set may be polluted, but it satisfies its schema;
    \item the data migration may require human interaction, which may take time;
-   \item the part of the data that does not change should be migrated automatically;
+   \item the part of the data that does not change in $\schema'$ can be migrated automatically;
    \item the meaning of data must be preserved;

### 2022-10-15 2022 Migration/articleMigration.tex
mention data pollution in the introduction

@@
    Information systems are typically used by distributed actors (both users and computers) who continually work with data.
-   The entire set of data, $\dataset$, in an information system represents the state of the system.
-   Every event that an actor causes in information system $\infsys$ may change the state $\dataset$.
-   We will write $\dataset_{\infsys}$ in case we need to be specific about which information system the dataset belongs to,
-   but just $\dataset$ if it is obvious or irrelevant to which information system this dataset belongs.
+   Information system $\infsys$ contains a dataset $\dataset_{\infsys}$, which represents the state of the system.
+   (In the sequel we may just write $\dataset$ if the corresponding information system is either obvious or irrelevant.)
+   Every event that an actor causes in information system $\infsys$ may change its state $\dataset$.
 
-   In this work we will encode the semantics in a schema,
+   Our work assumes that the semantics of the as-is dataset is encoded in its schema,
    so we can be explicit about ``preserving the meaning as much as possible''.
-   The schema of an information system is basically a set of constraints on its dataset $\dataset$.
-   We will denote the schema of an information system $\infsys$ by $\schema_{\infsys}$,
-   or just $\schema$ if it is obvious or irrelevant to which information system this schema belongs.
+   The schema $\schema_{\infsys}$ of an information system $\infsys$ is basically a set of constraints on its dataset $\dataset$.
+   (We will denote the schema just as $\schema$ if the information system is obvious or irrelevant to know.)
+   In contrast with earlier work, e.g.~\cite{Thalheim2013}, we assume that the dataset satisfies all semantic constraints in the schema.
+   This is possible because we use a compiler, Ampersand~\cite{Joosten-JLAMP2018},
+   that generates an information system that keeps all semantic constraints satisfied at all times.
+   As a consequence, it contains only data pollution that is not already captured by any of the constraints mentioned in the schema.
+   Remaining data pollution must therefore be captured in constraints in the schema of the to-be information system $\infsys'$,
+   so it can guard those new constraints from the moment $\infsys'$ is in production.
+   In the meantime, the migration engineer must ensure that $\infsys'$ can start with data that satisfies all constraints in its schema.
 
@@
    A formal understanding of data migration will help to automate data migrations correctly and reliably.
-   Our theory is based on the following premises:
+   Our theory is based on the following assumptions:
 \begin{itemize}
@@
    \item the part of the migration that can be automated is usually not sufficient;
-         it takes additional human creativity to complete the migration specification.
+         it takes additional human creativity to complete the migration specification;
+   \item there is a compiler to generate an information system from a given schema.
+         In this work, we use the Ampersand compiler~\cite{Joosten-JLAMP2018} for that purpose.
 \end{itemize}

### 2022-10-15 2022 Migration/articleMigration.tex
WIP

@@
 
-   Our work assumes that the semantics of the as-is dataset is encoded in its schema,
+   Much like~\cite{Spivak2012}, our work assumes that the semantics of the as-is dataset is encoded in its schema,
    so we can be explicit about ``preserving the meaning as much as possible''.
@@
    If the semantics encoded in the to-be schema $\schema'$ differs from the as-is schema $\schema$,
-   interaction with a migration engineer may be needed to determine how to interpret existing data in the new schema.
-   This may even involve adapting data of which the constraints in the schema have not changed.
+   a migration engineer must make choices on how to interpret existing data in the new schema.
+   She may even want to adapt data of which the constraints in the schema have not changed.
    This requires a third schema, $\schema_m$, which specifies the migration itself.
@@
 
-   The contribution of this paper is to propose a formalism in which $\schema_m$ is defined.
+   The contribution of this paper is to derive a migration schema $\schema_m$ from an existing information system $\infsys$
+   (which has its own $\schema$ and $\dataset$) and a to-be schema $\schema'$.
    A formal understanding of data migration will help to automate data migrations correctly and reliably.
@@
    \item there is a compiler to generate an information system from a given schema.
-         In this work, we use the Ampersand compiler~\cite{Joosten-JLAMP2018} for that purpose.
+         In this work, we use the Ampersand compiler for that purpose.
 \end{itemize}
-
-   There have been other attempts to formalize data migration~\cite{Thalheim2013}.
-   (At this point: Elaborate on earlier work)
 
@@
    They are values without internal structure of interest, meant to represent data elements in a database.
-   From a business perspective, atoms are used to represent concrete items of the world,
+   From a business perspective, atoms represent concrete items of the world,
    such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
-   By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ are used to represent \emph{atoms}.
+   By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ represent \emph{atoms}.
    The set of atoms is called $\atoms$.
@@
 
-   Triples are used to represent data.
+   We will use triples to represent data.
    This makes our theory valid for any kind of database that triples can represent,
@@
    Each triple relates two atoms to a relation.
-   A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean that the ``thing'' that \atom{Peter} refers to
-   has \atom{074238991} as a telephone number.
-   In the formal world, we leave it entirely up to a user to attach meaning to a triple.
-   As a consequence, this ``meaning from practice'' may have no consequences in the formal world.
+   A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean in practice that
+   the ``thing'' that \atom{Peter} refers to has \atom{074238991} as a telephone number.
+   This ``meaning from practice'' has no consequences in the formal world.
+   We leave it entirely up to a user to attach a practical meaning to a triple.
+   In this paper, the informal meaning is interesting only for the sake of illustration.
    The set of triples is called $\triples$.
+   To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\triples$.
 
    Relations in datasets are used to store data.
-   In this paper relations are represented by variables $r$, $s$, and $d$.
+   In this paper, variables $r$, $s$, and $d$ represent relations.
    Every relation $r$ contains a set of pairs, which we call the population of $r$:
@@
 \end{equation}
-   The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.
-
    Every relation has a name, a source concept, and a target concept.
-   We write $\declare{nm}{A}{B}$ to denote a relation with name \id{nm}, source concept $A$, and target concept $B$.
+   We write $r=\declare{nm}{A}{B}$ to denote that relation $r$ has name \id{nm}, source concept $A$, and target concept $B$.
    % To disassemble a relation in its name, source and target, there exist three functions:
@@
    % \(\tgt{\declare{nm}{A}{B}}=B\).
-
-   The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
+   The pair $\pair{A}{B}$ is called the \define{signature} of the relation.
    The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.
+% @Bas, de volgende twee definities zitten mij nog niet lekker.
+% Zou het niet $\pop{\dataset}{\fullt{A}{B}}$ en $\pop{\dataset}{\ident{A}}$ moeten zijn (omdat $A$ en $B$ niet uit $\triples$ maar uit $\concepts$ komen)?
+% Alleen,...    $\dataset$ is op deze plek nog niet gedefinieerd...
    It contains all pairs that can be made of elements of $A$ and $B$:
-\[\pop{\triples}{\fullt{A}{B}} = \{ \pair{a}{b}| a\ \inst\ A,\ b\ \inst\ B \}\]
-   The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.
-   The set of relations is called $\rels$.
+\[\pop{\triples}{\fullt{A}{B}}\ =\ \{ \pair{a}{b}|\ a\ \inst\ A,\ b\ \inst\ B \}\]
+   The term $\ident{A}$ represents the \emph{identity relation} of concept $A$:
+\[\pop{\triples}{\ident{A}}\ =\ \{ \pair{a}{a}|\ a\ \inst\ A \}\]
+   The set of relations is called $\rels$ and it contains $\full$, $\iden$, and all relations mentioned in $\triples$.
 
@@
    let $\atoms$ be a set of atoms,
-   and let $\inst$ be a relation $\atoms\times\concepts$ that represents the instance relation between concepts and atoms,
+   and let $\inst$ be a relation $\atoms\times\concepts$ that is not in $\rels$, which represents the instance relation between concepts and atoms,
 \begin{definition}[dataset]
@@
 \end{eqnarray}
-
    Since $\triples$ changes continually in an information system, the violations are changing along.
@@
    These events may be caused by external systems, roles acting on the dataset, or any other force that changes triples.
+
    Every rule is assigned to a role.
@@
    first, $A\isa B$ is automatically maintained by the information system itself, by changing $\inst$ to include any violating pairs in $B$.
-   This means that an information system can never be brought into a state in which $p_{A\isa B}$ is False if $A\isa B$ is a specialization rule of the information system, allowing for certain run-time optimizations.
+   This means that an information system that contains rule $A\isa B$ can never be brought into a state in which $p_{A\isa B}$ is False.
+   This allows for certain run-time optimizations, which are beyond the scope of this paper.
    A second property that sets specialization rules apart, is that they play a role in the Ampersand compiler.
    For example, code that determines equality between atoms only works for atoms that are an instance of the same concept.
-   To ensure that this does not result in problems at run-time, users are required to write specialization rules that make these comparisons possible, and the type checker in the Ampersand compiler will spot these issues if they occur and refuse to generate the information system.
+   To ensure that this does not result in problems at run-time, users are required to write specialization rules that make these comparisons possible.
+   The type checker in the Ampersand compiler will spot violations and refuse to generate an information system if they occur.
 
@@
    Having defined an information system in mathematical terms, let us give an example.
-   For this purpose we use a subset of the language Ampersand~\cite{Joosten-JLAMP2018}
+   For this purpose we use a subset of the language Ampersand
    because it makes the examples more appealing to read.
@@
    So, there is a student, Peter, taking the course Management and two students, Susan and John, taking the course Business IT.
-   As stated before, this informal meaning is interesting only for the sake of illustration.
 
@@
    Now let us define an information system.
-   To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\triples$.
    We define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that student takes.

### 2022-10-17 2022 Migration/articleMigration.tex
improve the reasoning quality in the introduction.

@@
    A mistake not made cannot propagate into production.
+   For this purpose, we use an existing compiler, Ampersand~\cite{Joosten-JLAMP2018},
+   that generates an information system from an Ampersand script.
+   An Ampersand script contains just enough information to generate a complete information system,
+   which means that a classical database schema (i.e. data structure plus semantics) can be extracted from the Ampersand script.
+   For the purpose of this paper we can equate an Ampersand script with the scheme of the generated information system.
+
    Incremental generation of information systems poses specific requirements to data migration.
@@
    e.g.~\cite{Gholami2016,Bisbal1999},
-   preserving the functionality can be useful to avoid introducing new errors in an otherwise error-prone process.
+   migration engineers may want to preserve the functionality to avoid introducing new errors in an otherwise error-prone process.
    For incremental generation, however, preserving the functionality is not an option.
@@
    Information systems are typically used by distributed actors (both users and computers) who continually work with data.
-   Information system $\infsys$ contains a dataset $\dataset_{\infsys}$, which represents the state of the system.
-   (In the sequel we may just write $\dataset$ if the corresponding information system is either obvious or irrelevant.)
+   An information system $\infsys$ contains a dataset $\dataset_{\infsys}$, which represents the state of the system.
+   (In the sequel we may just write $\dataset$
+   if the corresponding information system is either obvious from the context or irrelevant for the argument.)
    Every event that an actor causes in information system $\infsys$ may change its state $\dataset$.
@@
    (We will denote the schema just as $\schema$ if the information system is obvious or irrelevant to know.)
-   In contrast with earlier work, e.g.~\cite{Thalheim2013}, we assume that the dataset satisfies all semantic constraints in the schema.
-   This is possible because we use a compiler, Ampersand~\cite{Joosten-JLAMP2018},
-   that generates an information system that keeps all semantic constraints satisfied at all times.
-   As a consequence, it contains only data pollution that is not already captured by any of the constraints mentioned in the schema.
-   Remaining data pollution must therefore be captured in constraints in the schema of the to-be information system $\infsys'$,
-   so it can guard those new constraints from the moment $\infsys'$ is in production.
-   In the meantime, the migration engineer must ensure that $\infsys'$ can start with data that satisfies all constraints in its schema.
+   In contrast with~\cite{Spivak2012}, our approach does not work with functions (functors)
+   but with relations in a way that bears resemblance with~\cite{Zielinski2013}.
 
@@
 
-   Automating data migrations, however, has its limits.
+   The Ampersand compiler generates information systems that keep all semantic constraints satisfied at all times.
+   This allows us to categorize data pollution in the as-is information system in the following categories:
+   \begin{itemize}
+      \item data pollution that cannot be captured in constraints on the dataset,
+      such as a street address that has become obsolete because the person failed to send a change of address.
+      \item data pollution that is a violation of a constraint in the schema.
+      \item data pollution that is a violation of a constraint that is not in the schema.
+      \item data pollution that is a consequence of an erroneous constraint in the schema.
+   \end{itemize}
+   The first category of pollution must be dealt with by working procedures, to prevent such pollution as much as possible.
+   In some cases, constraints on data can be formulated,
+   so the information system can help the user and simplify (or eliminate) the corresponding working procedure.
+   Such constraints move this type of pollution to one of the other categories.
+   The second category, violations of constraints from the schema, does not occur in our work.
+   Since the as-is information system $\infsys$ was generated by Ampersand,
+   $\infsys$ keeps all constraints from the schema satisfied at all times.
+   This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
+   Data pollution that is not captured by any of the constraints mentioned in the schema
+   can be dealt with by adding constraints in the to-be information system $\infsys'$.
+   In that case, a migration engineer must ensure that $\infsys'$ can start with data that satisfies all constraints in its schema.
+   So some human intervention may be necessary.
+   The last category of data pollution, erroneous constraints, must be fixed by replacing them by the correct constraints in $\infsys'$.
+
+   This illustrates that automating data migrations also has its limits.
+   Some effort of a migration engineer may be required.
    If the semantics encoded in the to-be schema $\schema'$ differs from the as-is schema $\schema$,

### 2022-10-22 2022 Migration/articleMigration.tex
minor changes in abstract and introduction

@@
    In such cases, simply regenerating the system is not enough because it would reset the database to its initial state (deleting all data gathered so far).
-   Data migration aims at preserving the data as much as possible.
-   Typically, migration involves transferring the old data to the new system while preserving the semantics as much as possible.
+   A migration engineer must transfer the old data to the new system while preserving the semantics as much as possible.
 
@@
    which means that a classical database schema (i.e. data structure plus semantics) can be extracted from the Ampersand script.
-   For the purpose of this paper we can equate an Ampersand script with the scheme of the generated information system.
+   For the purpose of this paper we can equate an Ampersand script with the schema of the generated information system.
 
-   Incremental generation of information systems poses specific requirements to data migration.
-   For instance, if a data migration is not for incremental generation,
+   Incremental development of information systems poses specific requirements to data migration.
+   For instance, in data migrations for other purposes,
    e.g.~\cite{Gholami2016,Bisbal1999},
    migration engineers may want to preserve the functionality to avoid introducing new errors in an otherwise error-prone process.
-   For incremental generation, however, preserving the functionality is not an option.
-   To change the functionality is the very purpose and the data migration must behave accordingly.
-   So, we must make sure to use methods for data migration that are specifically meant for incrementally generating information systems.
+   For incremental development, however, preserving the functionality is not an option.
+   To change the functionality is the very purpose of each increment, so the data migration must support just that.
+   So, we must make sure to use methods for data migration that are specifically suitable for incremental development of information systems.
 
@@
    In contrast with~\cite{Spivak2012}, our approach does not work with functions (functors)
-   but with relations in a way that bears resemblance with~\cite{Zielinski2013}.
+   but with relations in a way that bears resemblance with~\cite{Alloy2006} and~\cite{Zielinski2013}.
 

### 2022-10-23 2022 Migration/articleMigration.tex
Introductie 'evolving'. Bewijsverplichtingen in commentaar geschetst.

@@
 
-% TODO Bas:
-% - verhaal over incrementeel fouten oplossen (fouten staan al in artikel)
-
 % TODO Algemeen:
@@
 
-\title{A Theory for Data Migration of Information Systems}
+% Bewijsverplichtingen:
+% -> het migratie-systeem is typefout-vrij (opmerking: als we de type-checker buiten beschouwing willen laten, kunnen we dit niet beschrijven)
+% -> het migratie-systeem is vrij van overtredingen op regels die het migratie-systeem moet bewaken
+% -> het migratie-systeem bevat alle oude data, ihb nog steeds na het toepassen van de enforce regels
+% -> er is een pad naar ingebruikname van het nieuwe systeem (vanaf de initiële toestand van het migratie-systeem)
+% -> corollary: op het moment van ingebruikname van het nieuwe systeem, is het migratie-systeem vrij van overtredingen op regels die het nieuwe systeem moet bewaken
+% -> optioneel: na een begrensd aantal 'voortgangs-stappen' kan het nieuwe systeem in gebruik genomen worden
+% -> optioneel: er bestaat een migratie-systeem dat de oude functionaliteit behoudt (mogelijk uitbreid) totdat het nieuwe systeem in gebruik genomen is?
+
+\title{A Theory for Data Migration of Generated Evolving Information Systems}
 \author[ou,ordina]{Stef Joosten\fnref{fn1}}
@@
 \label{sct:Introduction}
-   Our research focuses on generating information systems incrementally,
-   especially on the data migration aspect of it.
-   The reason that we believe in generating information systems is to prevent human errors as early as possible in the build chain:
+   We believe in generating evolving systems because it addresses human errors.
+   By generating information systems rather than programming them by hand, we avoid introducing errors in the translation from design to concrete system.
    A mistake not made cannot propagate into production.
@@
 
-   Incremental development of information systems poses specific requirements to data migration.
-   For instance, in data migrations for other purposes,
+   We recognise that even designs can have errors in them:
+   either the error was made directly by a designer, the requirements the designer was taking into account were miscommunicated, or the reality changed in a way was not foreseen at the time the design was made.
+   Regardless of the reason, even a generated information system may need to be repaired.
+   For this reason, we focus on evolving systems.
+
+   When a generated system evolves, we are left with the problem of migrating the data from the concrete old system to the newly designed system.
+   This situation poses specific requirements to data migration.
+   For instance, if a data migration is not for evolving systems,
    e.g.~\cite{Gholami2016,Bisbal1999},
    migration engineers may want to preserve the functionality to avoid introducing new errors in an otherwise error-prone process.
-   For incremental development, however, preserving the functionality is not an option.
-   To change the functionality is the very purpose of each increment, so the data migration must support just that.
-   So, we must make sure to use methods for data migration that are specifically suitable for incremental development of information systems.
+   For evolving a system, however, preserving the functionality is not an option.
+   To change the functionality is the very purpose of the evolution, so the data migration must support just that.
+   So, we must make sure to use methods for data migration that are specifically meant for evolving information systems.
 
@@
    \item the meaning of data must be preserved;
-   \item the business must continu during the migration without interruption;
+   \item the business must continue during the migration without interruption;
    \item the part of the migration that can be automated is usually not sufficient;

### 2022-10-24 2022 Migration/articleMigration.tex
working on redefining datasets and schemas

@@
 \newtheorem{lemma}{Lemma}
-\def\Events{{\mathit E}}
 \begin{document}
@@
 \begin{abstract}
-   The Ampersand project has provided the theory and tools to generate semantic information systems from an algebraic specification.
+   The Ampersand project has provided the theory and tools to generate information systems from an algebraic specification.
    However, information systems in practice may change repeatedly after their maiden deployment.
    Changes that affect the data model typically result in a data migration.
-   In such cases, simply regenerating the system is not enough because it would reset the database to its initial state (deleting all data gathered so far).
-   A migration engineer must transfer the old data to the new system while preserving the semantics as much as possible.
+   In such cases, simply regenerating the system is not enough.
+   That would reset the database to its initial state, losing all data gathered so far.
+   To prevent that, a migration engineer must transfer the old data to the new system by hand.
 
-   In this contribution we develop a theory for reliable data migration that aims at automating it as much as possible,
-   to enable more frequent migrations.
-   The ultimate target is to generate a migration from two specifications: the as-is specification and the to-be specification.
+   In this contribution we develop a theory for reliable data migration to help the migration engineer
+   to transfer the data and preserve the semantics as much as possible.
+   We aim to automate the data migration,
+   to prevent mistakes and enable more frequent migrations.
+   The target is to generate a migration script from two specifications: the old specification and the new specification.
    A software generator that embodies this theory is subject of future research.
@@
 \label{sct:Introduction}
-   We believe in generating evolving systems because it addresses human errors.
-   By generating information systems rather than programming them by hand, we avoid introducing errors in the translation from design to concrete system.
+   We believe in generating information systems%
+\footnote{In the sequel, the word ``system'' refers to the phrase ``information system''. This simplifies the language a little. }
+   to prevent human errors.
+   By generating rather than programming systems by hand,
+   we avoid introducing errors in the translation from design to a concrete system.
    A mistake not made cannot propagate into production.
    For this purpose, we use an existing compiler, Ampersand~\cite{Joosten-JLAMP2018},
-   that generates an information system from an Ampersand script.
-   An Ampersand script contains just enough information to generate a complete information system,
+   that generates a system from an Ampersand script.
+   An Ampersand script contains just enough information to generate a complete system,
    which means that a classical database schema (i.e. data structure plus semantics) can be extracted from the Ampersand script.
-   For the purpose of this paper we can equate an Ampersand script with the schema of the generated information system.
+   For the purpose of this paper we can equate an Ampersand script with the schema of the generated system.
 
-   We recognise that even designs can have errors in them:
-   either the error was made directly by a designer, the requirements the designer was taking into account were miscommunicated, or the reality changed in a way was not foreseen at the time the design was made.
-   Regardless of the reason, even a generated information system may need to be repaired.
-   For this reason, we focus on evolving systems.
-
-   When a generated system evolves, we are left with the problem of migrating the data from the concrete old system to the newly designed system.
-   This situation poses specific requirements to data migration.
-   For instance, if a data migration is not for evolving systems,
-   e.g.~\cite{Gholami2016,Bisbal1999},
-   migration engineers may want to preserve the functionality to avoid introducing new errors in an otherwise error-prone process.
-   For evolving a system, however, preserving the functionality is not an option.
-   To change the functionality is the very purpose of the evolution, so the data migration must support just that.
-   So, we must make sure to use methods for data migration that are specifically meant for evolving information systems.
-
-   Data migration occurs when an as-is information system, $\infsys$, is replaced by a to-be information system, $\infsys'$,
-   while preserving the meaning of the data from $\infsys$ as much as possible.
-   Just copying the set of data from $\infsys$ to $\infsys'$ is obviously wrong if the schemas of both information systems differ.
-   Information systems are typically used by distributed actors (both users and computers) who continually work with data.
-   An information system $\infsys$ contains a dataset $\dataset_{\infsys}$, which represents the state of the system.
-   (In the sequel we may just write $\dataset$
-   if the corresponding information system is either obvious from the context or irrelevant for the argument.)
-   Every event that an actor causes in information system $\infsys$ may change its state $\dataset$.
-
-   Much like~\cite{Spivak2012}, our work assumes that the semantics of the as-is dataset is encoded in its schema,
-   so we can be explicit about ``preserving the meaning as much as possible''.
-   The schema $\schema_{\infsys}$ of an information system $\infsys$ is basically a set of constraints on its dataset $\dataset$.
-   (We will denote the schema just as $\schema$ if the information system is obvious or irrelevant to know.)
-   In contrast with~\cite{Spivak2012}, our approach does not work with functions (functors)
-   but with relations in a way that bears resemblance with~\cite{Alloy2006} and~\cite{Zielinski2013}.
-
-   We believe that information systems should be released into production in small increments,
+   We also believe that information systems should be released into production in small increments,
    to obtain shorter release cycles, more frequent releases, and thus a more agile software development process%
@@
    change failure rate~\cite{DevOps2021}.}.
+   By installing upgrades automatically, frequently and in small increments,
+   users experience an evolving system.
+
+   We recognize that designs from which concrete systems are generated can have errors in them:
+   Either a designer was making a mistake directly,
+   or the requirements the designer was taking into account were miscommunicated,
+   or the environment changed in a way that was not foreseen.
+   Regardless of the reason, even a generated system may need to be repaired.
+   For this reason, we focus on evolving systems.
+
+   This paper studies this evolution in increments, in which we distinguish an ``old system'' and a ``new system''.
+   In each increment we need to migrate the data from the old system to the newly designed system.
+   This situation poses specific requirements to data migration.
+   For instance, if a data migration has other purposes than just to evolve a system,
+   e.g.~\cite{Gholami2016,Bisbal1999},
+   migration engineers may want to preserve the functionality to avoid introducing new errors in an otherwise error-prone process.
+   When one system evolves into another, however, preserving the functionality is not an option.
+   To change the functionality is the very purpose of the evolution.
+   So, we must make sure to use methods for data migration that are specifically meant for evolving systems.
+
+   Information systems are typically used by distributed actors (both users and computers) who continually work with data.
+   A system $\infsys$ contains a dataset $\dataset_{\infsys}$, which represents the state of the system.
+   (In the sequel we may just write $\dataset$
+   if the corresponding system is either obvious from the context or irrelevant for the argument.)
+   Every event that an actor causes in system $\infsys$ may change the state $\dataset$ of that system.
+   Data migration occurs when an old system, $\infsys$, is replaced by a new system, $\infsys'$,
+   while preserving the meaning of the data from $\infsys$ as much as possible.
+   Just copying the set of data from $\infsys$ to $\infsys'$ is obviously wrong if the schemas of both systems differ.
+
+   The schema $\schema_{\infsys}$ of a system $\infsys$ is basically a set of semantic constraints on its dataset $\dataset$.
+   (We will denote the schema just as $\schema$ if the system is obvious or irrelevant to know.)
+   The Ampersand compiler generates systems that keep all constraints satisfied at all times.
+   Much like~\cite{Spivak2012}, our work assumes that the semantics of the old dataset is encoded in its schema,
+   so we can be explicit about ``preserving the meaning as much as possible''.
+   In contrast with~\cite{Spivak2012}, our approach does not work with functions (functors)
+   but with relations in a way that bears resemblance with Alloy~\cite{Alloy2006} and allegories%
+   \footnote{Allegories are a specific type of categories, which the reader does not need to understand for the purpose of this paper}~\cite{Zielinski2013}.
+
+   The theory we develop in this paper must lead to tools that satisfy such practical requirements as
+   zero down-time updates and ways to cope with data pollution.
    In practice, more frequent updates require that updates have zero down-time.
-   Extra complications arise from accumulated data pollution in the as-is information system.
-   So, the theory we develop in this paper must lead to tools that satisfy such practical requirements as
-   zero down-time updates and ways to cope with data pollution.
-   That will help agile teams today to automate their software processes further.
+   For our theory, zero down-time means that the release of new functionality may not be delayed by anything a migration engineer must do.
 
-   The Ampersand compiler generates information systems that keep all semantic constraints satisfied at all times.
-   This allows us to categorize data pollution in the as-is information system in the following categories:
+   Extra complications arise from accumulated data pollution in the old system.
+   This allows us to categorize data pollution in the old system in the following categories:
    \begin{itemize}
@@
    In some cases, constraints on data can be formulated,
-   so the information system can help the user and simplify (or eliminate) the corresponding working procedure.
+   so the system can help the user and simplify (or eliminate) the corresponding working procedure.
    Such constraints move this type of pollution to one of the other categories.
    The second category, violations of constraints from the schema, does not occur in our work.
-   Since the as-is information system $\infsys$ was generated by Ampersand,
+   Since the old system $\infsys$ was generated by Ampersand,
    $\infsys$ keeps all constraints from the schema satisfied at all times.
@@
    Data pollution that is not captured by any of the constraints mentioned in the schema
-   can be dealt with by adding constraints in the to-be information system $\infsys'$.
+   can be dealt with by adding constraints in the new system $\infsys'$.
    In that case, a migration engineer must ensure that $\infsys'$ can start with data that satisfies all constraints in its schema.
@@
    Some effort of a migration engineer may be required.
-   If the semantics encoded in the to-be schema $\schema'$ differs from the as-is schema $\schema$,
+   If the semantics encoded in the new schema $\schema'$ differs from the old schema $\schema$,
    a migration engineer must make choices on how to interpret existing data in the new schema.
@@
 
-   The contribution of this paper is to derive a migration schema $\schema_m$ from an existing information system $\infsys$
-   (which has its own $\schema$ and $\dataset$) and a to-be schema $\schema'$.
+   The contribution of this paper is to derive a migration schema $\schema_m$ from an existing system $\infsys$
+   (which has its own schema and dataset) and a new schema $\schema'$.
    A formal understanding of data migration will help to automate data migrations correctly and reliably.
@@
 \begin{itemize}
-   \item the as-is data set may be polluted, but it satisfies its schema;
+   \item the old data set may be polluted, but it satisfies its schema;
    \item the data migration may require human interaction, which may take time;
-   \item the part of the data that does not change in $\schema'$ can be migrated automatically;
-   \item the meaning of data must be preserved;
+   \item the part of the data that does not change can be migrated automatically;
+   \item the meaning of data must be preserved, at least for the data that does not change;
    \item the business must continue during the migration without interruption;
@@
 \section{Migration steps}
-   A smooth migration from an old system to a new system would go as follows:
-   Launch the new system in parallel to the old, copy data from the old to the new system, and have everyone use the new system.
+   A smooth migration would ideally proceed as follows:
+   Launch the new system in parallel to the old one, copy data from the old to the new system, and have everyone use the new system.
    However, numerous issues might impede that plan.
@@
    The old system required the reimbursement office to look up employee's addresses from a hand-written list they had on their desk.
+
+   This issue will require somebody to insert the missing addresses in the new system.
 \item Data in the old system is wrong but cannot be corrected there due to how the old system was designed.
-   An example is if the old system only allows approvals to be entered as the current user, and the CEO insists that her administrative staff enters his approvals into the system for her.
+   An example is if the old system only allows approvals to be entered as the current user, and the CEO has always insisted that her administrative staff enters the approvals into the system for her.
    This may result in approvals being entered as admin staff, where it was really the CEO making the approval.
+
+   This issue will require that the incorrect registration of staff members is corrected in the new system.
 \item Data in the old system does not satisfy invariants of the new system.
    There may be no way in the old system of making the data satisfy those invariants.
-We can use the same example as in the previous bullet point, but add the requirement (in the new system) that every purchase above a certain amount needs to be approved by the CEO.
+   We can use the same example as in the previous bullet point,
+   but add the requirement (in the new system) that every purchase above a certain amount needs to be approved by the CEO.
+
+   This issue will require somebody to change the data to satisfy the broken invariants in the new system.
 \item A way of entering data into the old system is missing in the new system.
    People or automated processes might rely on these ways of entering data.
-   An example could be that when employees turned their computers on or off, an ad-hoc script would automatically clock them in- and out to determine the number of hours they worked.
+   An example could be that when employees turned their computers on or off,
+   an ad-hoc script would automatically check them in- and out to determine the number of hours they worked.
    A handful of employees still relies on this.
+
+   This issue will require that users of obsolete functionality are informed and given a way to cope with the missing functionality where appropriate.
 \item Data present in the old system cannot be stored in the new system.
-   An example could be that references to physical locations where original receipts are kept are stored in the old system, but the new system relies on scans of receipts and allows the originals to be destroyed or not submitted.
-   Until the original receipts are scanned, the old data should be kept.
+   An example could be that references to physical locations where original receipts are kept are stored in the old system,
+   but the new system relies on scans of receipts and allows the originals to be destroyed or not submitted.
+
+   This issue will require that the old data is kept until the original receipts have been scanned.
 \end{enumerate}
@@
 \label{sct:Terminology}
-   To migrate data from one information system to another,
-   we must define ``information system''.
-   In that definition we will single out the dataset as the focal point of a data migration.
-   For this reason, we define datasets first.
-   We need rules to capture the semantics, so we need to define rules as well.
-   After that, we define information systems.
+   To migrate data from one system to another,
+   we define ``information system'' as a combination of dataset and schema.
+   So, let us first define datasets in section~\ref{sct:Datasets} as the focal point of a data migration.
+   Then we define schemas in section~\ref{sct:Schemas}, to capture the semantics of information systems.
 
@@
    Let $\concepts$ be a set of concepts,
-   let $\isa$ be a partial order on $\concepts$,
    let $\rels$ be a set of relations,
    let $\atoms$ be a set of atoms,
-   and let $\inst$ be a relation $\atoms\times\concepts$ that is not in $\rels$, which represents the instance relation between concepts and atoms,
+   and let $\inst$ be a relation $\atoms\times\concepts$ that is not in $\rels$,
+   which represents the instance relation between concepts and atoms:
 \begin{definition}[dataset]
@@
    \end{eqnarray}
+   The set of all conceivable datasets is called $\Dataset$.
 \end{definition}
 
-\subsection{Rules}
-\label{sct:Rules}
+\subsection{Schemas}
+\label{sct:Schemas}
    Rules are used to describe whether certain (assumably desirable) properties hold in an information system.
+   let $\isa$ be a partial order on $\concepts$,
 
@@
    \label{def:rule}
-   \item A rule $u$ is a tuple $\la A,B,\id{viol}_u\ra$.
+   \item A rule $u$ is a tuple $\la \id{sign}_u:\Pair{\Concepts}{\Concepts},\id{viol}_u:\Dataset\rightarrow(\Pair{\Concepts}{\Concepts})\ra$.
    %in which $p_u$ is a constraint (i.e. a condition that constrains the triples of a dataset),
-   $\pair{A}{B}$ is the signature of $u$ and
+   $\id{sign}_u$ is the signature of $u$ and
    $\id{viol}_u$ is a function that gives the violations (which are pairs) of this rule,
@@
 \begin{eqnarray}
-   \pair{a}{b}\in\viol{u}{\triples,\inst}&\Rightarrow&a\ \inst\ A\wedge b\ \inst\ B
+   \begin{array}[t]{rl}
+      &\dataset\ =\ \la\atoms,\concepts,\inst,\rels,\triples\ra\ \wedge\\
+      &\id{sign}_u=\pair{A}{B}\ \wedge\ \pair{a}{b}\in\viol{u}{\dataset}\\
+      \Rightarrow\\
+      &A\in\concepts\wedge B\in\concepts\wedge a\ \inst\ A\wedge b\ \inst\ B
+   \end{array}
 \end{eqnarray}
@@
    which allows us to check at any moment in time whether the current population is free of violations.
-   We say that a rule $u$ is \define{satisfied} in a dataset $\la\atoms,\concepts,\inst,\rels,\triples\ra$ iff $\viol{u}{\triples}=\emptyset$.
+   We say that a rule $u$ is \define{satisfied} in a dataset $\dataset$ iff $\viol{u}{\dataset}=\emptyset$.
    This allows us to define a predicate $p_u$ that indicates if a rule $u$ is satisfied:
 \begin{eqnarray}
-   p_u(\triples,\inst)&\Leftrightarrow&\viol{u}{\triples,\inst}=\emptyset
+   p_u(\dataset)&\Leftrightarrow&\viol{u}{\dataset}=\emptyset
 \end{eqnarray}
-   Since $\triples$ changes continually in an information system, the violations are changing along.
+   Since $\triples$ changes continually in the dataset of an information system, the violations are changing along.
    As a result, rules that are violated at one moment may be satisfied at another moment.
-   The set of triples changes due to events observed by the information system.
+   The set of triples changes due to events observed by the system.
    These events may be caused by external systems, roles acting on the dataset, or any other force that changes triples.
@@
    The rule $A\isa B = \la A, A, viol_{A\isa B} \ra$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
-   Instances of $A$ that are not an instance of $B$ are the violations of this rule: $viol_{A\isa B}(\triples,\inst) = \{(a,a) \mid a\ \inst\ A \wedge \lnot (a\ \inst\ B)\}$.
+   Instances of $A$ that are not an instance of $B$ are the violations of this rule: $viol_{A\isa B}(\dataset) = \{(a,a) \mid a\ \inst\ A \wedge \lnot (a\ \inst\ B)\}$.
    We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
@@
    Specialization rules are special for two reasons:
-   first, $A\isa B$ is automatically maintained by the information system itself, by changing $\inst$ to include any violating pairs in $B$.
-   This means that an information system that contains rule $A\isa B$ can never be brought into a state in which $p_{A\isa B}$ is False.
+   first, $A\isa B$ is automatically maintained by the system itself, by changing $\inst$ to include any violating pairs in $B$.
+   This means that a system that contains rule $A\isa B$ can never be brought into a state in which $p_{A\isa B}$ is False.
    This allows for certain run-time optimizations, which are beyond the scope of this paper.
@@
    To ensure that this does not result in problems at run-time, users are required to write specialization rules that make these comparisons possible.
-   The type checker in the Ampersand compiler will spot violations and refuse to generate an information system if they occur.
+   The type checker in the Ampersand compiler will spot violations and refuse to generate a system if they occur.
 
@@
    This collective use serves a purpose which we loosely call ``the business''.
-   As a consequence, the data in an information system changes continually.
+   As a consequence, the data in a system changes continually.
    To preserve meaning, users are trying to maintain semantic constraints on the data,
@@
    The purpose of a role is to mention an individual user without knowing who that user is.
-   In the sequel, when we talk about a role we actually mean a user (an arbitrary one) who fulfills that role in the information system.
+   In the sequel, when we talk about a role we actually mean a user (an arbitrary one) who fulfills that role in the system.
 
@@
 \subsection{Example}
-\label{As-is IS}
+\label{old IS}
    Having defined an information system in mathematical terms, let us give an example.
@@
    Having one rule, one role, a relation $\maintain$ that contains one pair, and a dataset,
-   we have now defined a small information system.
+   we have now defined a small system.
 
 \subsection{Data migration}
-   To migrate information system $\infsys$ to $\infsys'$, we take the disjoint union $\infsys\sqcup\infsys'$ of both information systems
+   To migrate system $\infsys$ to $\infsys'$, we take the disjoint union $\infsys\sqcup\infsys'$ of both systems
    and add rules to describe the migration.
@@
    We use two datasets: $\dataset$ and $\dataset'$.
-   Before doing so, let us first define the disjoint union of two information systems.
+   Before doing so, let us first define the disjoint union of two systems.
 \begin{definition}[disjoint union of datasets]
@@
 \subsection{Example}
-   Let us proceed to specify a to-be information system, $\infsys'$, to illustrate a (toy) migration.
-   We will use the example in section~\ref{As-is IS} as the as-is information system
+   Let us proceed to specify a new system, $\infsys'$, to illustrate a (toy) migration.
+   We will use the example in section~\ref{old IS} as the old system
    and call that $\infsys$ in this section.
-   Its population reflects the state of the information system just before the migration.
+   Its population reflects the state of the system just before the migration.
 
-   Information system $\infsys'$, the to-be information system, is specified by:
+   Information system $\infsys'$, the new system, is specified by:
 \begin{verbatim}
@@
    while adding the new triples to $\infsys'$.
-   So the first thing to do is to take the disjoint union of both information systems.
+   So the first thing to do is to take the disjoint union of both systems.
    This yields:
@@
 \end{verbatim}
-   Any data migration must anticipate any population in the as-is information system.
+   Any data migration must anticipate any population in the old system.
    The only thing we know for sure about the ``old'' population is that it satisfies the rules in $\infsys$.
@@
 
-However, the violations that arise when taking the (non-disjoint) union of the two scripts include violations of rules that the information system should maintain.
+However, the violations that arise when taking the (non-disjoint) union of the two scripts include violations of rules that the system should maintain.
 This means that if we cannot generate software based on the disjoint union as it is.

### 2022-10-29 2022 Migration/articleMigration.tex
redone sections to fit schemas into the picture.

@@
    We recognize that designs from which concrete systems are generated can have errors in them:
-   Either a designer was making a mistake directly,
-   or the requirements the designer was taking into account were miscommunicated,
+   Either an engineer was making a mistake directly,
+   or the requirements the engineer was taking into account were miscommunicated,
    or the environment changed in a way that was not foreseen.
@@
    (In the sequel we may just write $\dataset$
-   if the corresponding system is either obvious from the context or irrelevant for the argument.)
+   if the corresponding system is either clear from the context or irrelevant for the argument.)
    Every event that an actor causes in system $\infsys$ may change the state $\dataset$ of that system.
@@
       such as a street address that has become obsolete because the person failed to send a change of address.
-      \item data pollution that is a violation of a constraint in the schema.
-      \item data pollution that is a violation of a constraint that is not in the schema.
+      \item data pollution that violates a constraint in the schema.
+      \item data pollution that violates a constraint that is not in the schema.
       \item data pollution that is a consequence of an erroneous constraint in the schema.
@@
 \label{sct:Datasets}
-   A dataset describes a set of structured data, which is typically stored persistently in a database of some kind.
-
-   Before defining datasets, we must first define the constituent notions of atom, concept, triple, and relation.
+   A dataset $\dataset$ describes a set of structured data, which is typically stored persistently in a database of some kind.
+   Before defining datasets, we must first define the constituent notions of atom, concept, relation, and triple.
 
@@
    By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ represent \emph{atoms}.
-   The set of atoms is called $\atoms$.
+   All atoms are taken from an infinite set called $\Atoms$.
 
-   Concepts are names we use during the design of a dataset
-   to name a group of atoms of the same type,
-   to assign signatures to relations, and thus
-   to enable type checking of the dataset.
-   For example, you might choose to classify \atom{Peter} as a \concept{Person}, and \atom{074238991} as a \concept{TelephoneNumber}.
+   Concepts are names that group atoms of the same type.
+   For example, an engineer might choose to classify \atom{Peter} and \atom{Melissa} as \concept{Person},
+   and \atom{074238991} as a \concept{TelephoneNumber}.
    In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
@@
    The expression $a\ \inst\ A$ means that atom $a$ is an \emph{instance} of concept $A$.
-   The set of concepts is called $\concepts$.
+   All concepts are taken from an infinite set $\Concepts$.
+   $\Concepts$ and $\Atoms$ are disjoint sets.
 
-   We will use triples to represent data.
+   Relations serve to organize and store data, to allow an engineer to represent facts.
+   In this paper, variables $r$, $s$, and $d$ represent relations.
+   All relations are taken from an infinite set $\Rels$.
+   $\Rels$ is disjoint from $\Concepts$ and $\Atoms$.
+
+   Triples serve to represent data.
+   A dataset $\dataset$ is a set of triples.
+   A triple is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.
    This makes our theory valid for any kind of database that triples can represent,
@@
    In this paper, the informal meaning is interesting only for the sake of illustration.
-   The set of triples is called $\triples$.
-   To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\triples$.
+   To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\dataset$
+   in those cases where $\dataset$ is either clear from the context or irrelevant for the argument.
+   A relation $r$ can serve as a container of pairs,
+   as defined by the funciton $\id{pop}_r:\Dataset\rightarrow\Pair{\Atoms}{\Atoms}$.
+   It defines a set of pairs, which we call the population of $r$:
+\begin{equation}
+   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\dataset\}
+\end{equation}
 
-   Relations in datasets are used to store data.
-   In this paper, variables $r$, $s$, and $d$ represent relations.
-   Every relation $r$ contains a set of pairs, which we call the population of $r$:
-\begin{equation}
-   \pop{\triples}{r}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\triples\}
-\end{equation}
-   Every relation has a name, a source concept, and a target concept.
-   We write $r=\declare{nm}{A}{B}$ to denote that relation $r$ has name \id{nm}, source concept $A$, and target concept $B$.
-   % To disassemble a relation in its name, source and target, there exist three functions:
-% \[\begin{array}[3]{rcl}
-   % \id{relname}&:&\rels\rightarrow\id{RelationIdentifier}\\
-   % \id{src}&:&\rels\rightarrow\concepts\\
-   % \id{tgt}&:&\rels\rightarrow\concepts
-% \end{array}\]
-   % Or, stated otherwise, the denotation $\declare{nm}{A}{B}$ stands for a relation with
-   % \(\relname{\declare{nm}{A}{B}} = \id{nm}\),
-   % \(\src{\declare{nm}{A}{B}}=A\), and
-   % \(\tgt{\declare{nm}{A}{B}}=B\).
-   The pair $\pair{A}{B}$ is called the \define{signature} of the relation.
-   The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.
-% @Bas, de volgende twee definities zitten mij nog niet lekker.
-% Zou het niet $\pop{\dataset}{\fullt{A}{B}}$ en $\pop{\dataset}{\ident{A}}$ moeten zijn (omdat $A$ en $B$ niet uit $\triples$ maar uit $\concepts$ komen)?
-% Alleen,...    $\dataset$ is op deze plek nog niet gedefinieerd...
-   It contains all pairs that can be made of elements of $A$ and $B$:
-\[\pop{\triples}{\fullt{A}{B}}\ =\ \{ \pair{a}{b}|\ a\ \inst\ A,\ b\ \inst\ B \}\]
-   The term $\ident{A}$ represents the \emph{identity relation} of concept $A$:
-\[\pop{\triples}{\ident{A}}\ =\ \{ \pair{a}{a}|\ a\ \inst\ A \}\]
-   The set of relations is called $\rels$ and it contains $\full$, $\iden$, and all relations mentioned in $\triples$.
-
-   We can now define the notion of dataset.
-   Let $\concepts$ be a set of concepts,
-   let $\rels$ be a set of relations,
-   let $\atoms$ be a set of atoms,
-   and let $\inst$ be a relation $\atoms\times\concepts$ that is not in $\rels$,
-   which represents the instance relation between concepts and atoms:
-\begin{definition}[dataset]
-   \label{def:dataset}
-   \item A dataset is a tuple $\la\atoms,\concepts,\inst,\rels,\triples\ra$ with
-   \begin{eqnarray}
-      \forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\\
-      %\forall A,B\in\concepts, a\in\atoms&:&A\isa B\wedge a\ \inst\ A\ \Rightarrow\ a\ \inst\ B\\
-      \forall \declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts\\
-      \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&\begin{array}[t]{@{}l}\declare{n}{A}{B}\in\rels\ \wedge\ a\ \inst\ A\ \wedge\ b\ \inst\ B\end{array}
-   \end{eqnarray}
-   The set of all conceivable datasets is called $\Dataset$.
-\end{definition}
+   Over time, an information system ``observes'' events that change the dataset by inserting and deleting triples in $\dataset$.
+   By the same account, the population of relations change over time.
+   These events may be caused by external systems, roles acting on the dataset, or any other force that changes triples.
 
@@
 \label{sct:Schemas}
-   Rules are used to describe whether certain (assumably desirable) properties hold in an information system.
-   let $\isa$ be a partial order on $\concepts$,
+   To design an information system, an engineer defines a schema in which she defines concepts, relations, and rules.
+   Rules serve the purpose of invariants,
+   i.e.\ properties that the data in the database must satisfy at all times.
+   In this paper, variables $u$ and $v$ represent rules.
 
-\begin{definition}[rule]
-   \label{def:rule}
-   \item A rule $u$ is a tuple $\la \id{sign}_u:\Pair{\Concepts}{\Concepts},\id{viol}_u:\Dataset\rightarrow(\Pair{\Concepts}{\Concepts})\ra$.
-   %in which $p_u$ is a constraint (i.e. a condition that constrains the triples of a dataset),
-   $\id{sign}_u$ is the signature of $u$ and
-   $\id{viol}_u$ is a function that gives the violations (which are pairs) of this rule,
-   in which
+   A schema $\schema$ is a triple $\triple{\concepts_{\schema}}{\rels_{\schema}}{\rules_{\schema}}$,
+   in which $\concepts_{\schema}$ is a finite set of concepts,
+   $\rels_{\schema}$ is a finite set of relations,
+   and $\rules_{\schema}$ is a finite set of rules.
+   We will simply write $\concepts$, $\rels$, or $\rules$ if the schema is clear from the context or irrelevant for the argument.
+   A rule in a schema serves to constrain the dataset,
+   to ensure the semantic integrity of the dataset.
+   For every rule $u$ in the schema, the Ampersand compiler generates code for the predicate $\id{sat}_u\{\Dataset\}$
+   to determine whether rule $u$ is satisfied in dataset $\dataset$.
+   The Ampersand compiler also generates code for a function $\id{viol}_u : \Dataset\rightarrow\id{Text}$
+   to produce meaningful error messages in case rule $u$ is not satisfied.
+
+\subsection{Information Systems}
+\label{sct:Information Systems}
+   The purpose of an information system is to make data meaningful to its users.
+   This section introduces the notion of information system as a quintuple $\la\dataset_{\infsys},\schema_{\infsys},\inst_{\infsys},\roles_{\infsys},\maintain_{\infsys}\ra$.
+   As before, we will omit the suffixes if $\infsys$ is obvious (i.e. clear from the context or irrelevant for the argument).
+\begin{definition}[information system]
+\label{def:information system}
+\item An information system $\infsys$ is a tuple $\la\dataset,\schema,\inst,\roles,\maintain\ra$, in which
+\begin{itemize}
+   \item dataset $\dataset$ and $\schema$ are defined as in sections~\ref{sct:Schemas} and~\ref{sct:Datasets};
+   \item $\roles$ is a set of roles;
+   \item $\inst : \Atoms\times\Concepts$ is the instance relation between atoms from $\dataset$ and concepts from $\schema$;
+   \item $\maintain : \roles\times\rules$ is the maintainance relation between roles and rules from $\schema$; and
+   \item requirements~\ref{eqn:inst} thru~\ref{eqn:specialization} are satisfied.
+\end{itemize}
+\end{definition}
+   Let us discuss these items one by one.
+
+   The language Ampersand is typed, to prevent a substantial amount of programming mistakes at compile-time.
+   This means that every atom is an instance of a concept:
 \begin{eqnarray}
-   \begin{array}[t]{rl}
-      &\dataset\ =\ \la\atoms,\concepts,\inst,\rels,\triples\ra\ \wedge\\
-      &\id{sign}_u=\pair{A}{B}\ \wedge\ \pair{a}{b}\in\viol{u}{\dataset}\\
-      \Rightarrow\\
-      &A\in\concepts\wedge B\in\concepts\wedge a\ \inst\ A\wedge b\ \inst\ B
-   \end{array}
+   \forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\label{eqn:inst}
 \end{eqnarray}
-\end{definition}
+   To do type checking based on the schema only,
+   all relations in triples from the dataset must be ``known'' in the schema:
+\begin{equation}
+   \forall a,r,b:\ \triple{a}{r}{b}\in\dataset\ \Rightarrow\ r\in\rels
+\end{equation}
 
-   The function $\id{viol}_u$ computes the violations of $u$,
-   which allows us to check at any moment in time whether the current population is free of violations.
-   We say that a rule $u$ is \define{satisfied} in a dataset $\dataset$ iff $\viol{u}{\dataset}=\emptyset$.
-   This allows us to define a predicate $p_u$ that indicates if a rule $u$ is satisfied:
+   In a typed information system, atoms in a relation have a type as declared in a relation.
+   For this reason, every relation $r$ has not only a name, but also a source concept and a target concept.
+   We write $r=\declare{nm}{A}{B}$ to denote that relation $r$ has name \id{nm}, source concept $A$, and target concept $B$.
 \begin{eqnarray}
-   p_u(\dataset)&\Leftrightarrow&\viol{u}{\dataset}=\emptyset
+   \forall\declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts\\
+   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\dataset&:&a\ \inst\ A\ \wedge\ b\ \inst\ B
 \end{eqnarray}
-   Since $\triples$ changes continually in the dataset of an information system, the violations are changing along.
-   As a result, rules that are violated at one moment may be satisfied at another moment.
-   The set of triples changes due to events observed by the system.
-   These events may be caused by external systems, roles acting on the dataset, or any other force that changes triples.
 
-   Every rule is assigned to a role.
-   That role must ensure to satisfy that rule after some event has caused violations by changing the set of triples.
-   In other words: If anyone edits in the dataset and breaks a rule, someone or something has to fix the dataset to satisfy that rule again.
-   Example: if a rule says that every request must be acknowledged, the receipt of a request will break this rule.
-   Sending an acknowledgement message will cause the rule to be satisfied again.
+   To compare atoms for equality,
+   the set $\rels$ must contain an identity relation $\ident{C}$ for every concept $C$ in which all atoms occur.
+   It must satisfy:
+\begin{equation}
+   \forall a,b:\ a\ \inst\ C\ \wedge\ b\ \inst\ C\ \Rightarrow\ (a\ \ident{C}\ b\ \Leftrightarrow\ a=b)
+\end{equation}
+   So, $\id{pop}_{\ident{C}}$ is defined by:
+\begin{eqnarray}
+   \forall C\in\concepts&:&\pop{\ident{C}}{\dataset}=\{\pair{a}{a}| a\ \inst\ C\}
+\end{eqnarray}
 
    Ampersand supports a special kind of rule that is used for specialization.
-   The rule $A\isa B = \la A, A, viol_{A\isa B} \ra$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
-   Instances of $A$ that are not an instance of $B$ are the violations of this rule: $viol_{A\isa B}(\dataset) = \{(a,a) \mid a\ \inst\ A \wedge \lnot (a\ \inst\ B)\}$.
+   The rule $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
+\begin{equation}
+   \label{eqn:specialization}
+   \id{sat}_{A\isa B}\{\dataset\}\ \Leftrightarrow\ \forall a: a\ \inst\ A\rightarrow a\ \inst\ B
+\end{equation}
    We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
@@
 
+   All specialization rules in a schema together form a partial order $\isa : \Pair{\concepts}{\concepts}$.
    Specialization rules are special for two reasons:
-   first, $A\isa B$ is automatically maintained by the system itself, by changing $\inst$ to include any violating pairs in $B$.
-   This means that a system that contains rule $A\isa B$ can never be brought into a state in which $p_{A\isa B}$ is False.
+   Firstly, the system keeps $A\isa B$ satisfied automatically by changing $\inst$ whenever needed.
+   This means that a system that contains rule $A\isa B$ can never be brought into a state in which $\id{sat}_{A\isa B}$ is False.
    This allows for certain run-time optimizations, which are beyond the scope of this paper.
@@
 
-\subsection{Information Systems}
-\label{sct:Information Systems}
-   The purpose of an information system is to make data meaningful to users.
-   Data migration is all about preserving that meaning.
-   This section introduces the notion of information system
-   as a system that consist of a dataset and a set of rules that constrain that data,
-   together with a set of roles that must keep all rules satisfied.
    %These rules reflect the meaning (semantics) that is shared among users.
+
+   Every rule is assigned to a role.
+   That role must ensure to satisfy that rule, after some event has violated the rule by changing the set of triples.
+   In other words: If anyone edits in the dataset and violates a rule, someone or something has to fix the dataset to satisfy that rule again.
+   Example: if a rule says that every request must be acknowledged, the receipt of a request will violate this rule.
+   Sending an acknowledgement message will cause the rule to be satisfied again.
 
@@
 
-   We can now define the notion of information system.
-\begin{definition}[information system]
-\label{def:information system}
-\item An information system $\infsys$ is a tuple $\la\roles,\rules,\specrules,\maintain,\dataset\ra$.
-   Here $\roles$ is a set of roles,
-   $\rules$ a set of rules,
-   $\specrules$ a set of specialization rules,
-   $\maintain$ is a relation $\roles\times\rules$,
-   and $\dataset$ is a dataset.
-\end{definition}
    The relation $\maintain$ links rules to roles.
-   $o\ \maintain\ u$ means that role $o$ must keep the set $\viol{u}{\triples}$ empty.
-   If a rule $u$ has violations, i.e. $\viol{u}{\triples}$ is not empty, there is a role whose job it is to remove those violations.
+   $o\ \maintain\ u$ means that role $o$ must keep the set $\viol{u}{\dataset}$ empty.
+   If a rule $u$ has violations, i.e. $\viol{u}{\dataset}$ is not empty, there is a role whose job it is to remove those violations.
    The relation $\maintain$ being surjective means that all users together are keeping the violation set empty.
@@
 \[\begin{array}{l}
-   \viol{\text{\tt EnrollRule}}{\triples}\\
+   \viol{\text{\tt EnrollRule}}{\dataset}\\
    \hspace{1cm}=\{\pair{s}{m}|s\ \text{\tt isEnrolledFor}\ m\ \wedge\not\exists c: s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\}
@@
 \begin{eqnarray}
-   \omit\rlap{$\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\triples'\ra$}\notag\\
-   &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\ \inst\uplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \triples\uplus^5\triples'\ra\notag\\
+   \omit\rlap{$\la\atoms,\concepts,\inst,\isa,\rels,\dataset\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\dataset'\ra$}\notag\\
+   &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\ \inst\uplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \dataset\uplus^5\dataset'\ra\notag\\
       X\uplus Y&=&\{(x,0)|\ x\in X\}\ \cup\ \{(y,1)|\ y\in Y\}\\
@@
 \[\begin{array}{l}
-   \viol{\text{\tt ExamRule1}}{\triples}\\
+   \viol{\text{\tt ExamRule1}}{\dataset}\\
    \hspace{1cm}=\{\pair{s}{c}\mid\exists e:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\neg(s\ \text{\tt takes}\ c)\}\\
-   \viol{\text{\tt ExamRule2}}{\triples}\\
+   \viol{\text{\tt ExamRule2}}{\dataset}\\
    \hspace{1cm}=\{\pair{s}{m}\mid\exists e,c:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\ \wedge\neg(s\ \text{\tt isEnrolledFor}\ m)\}
@@
    Changes to $\inst$, $\concepts$, $\rels$ change the data structure in a dataset.
-   We define a \define{migration of a dataset} $\la\atoms,\concepts,\inst,\rels,\triples\ra$ as a change in which one or more of $\inst$, $\concepts$, or $\rels$ change.
-   To satisfy definition~\ref{def:dataset}, $\triples$ and $\atoms$ must change too,
+   We define a \define{migration of a dataset} $\la\atoms,\concepts,\inst,\rels,\dataset\ra$ as a change in which one or more of $\inst$, $\concepts$, or $\rels$ change.
+   To satisfy definition~\ref{def:dataset}, $\dataset$ and $\atoms$ must change too,
    but the migration should try to preserve the maximal amount of data.

### 2022-10-30 2022 Migration/articleMigration.tex
Result after 30 Oct 2022 discussion

@@
 \begin{equation}
-   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\dataset\}
+   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}\mid\ \triple{a}{r}{b}\in\dataset\}
 \end{equation}
@@
    We will simply write $\concepts$, $\rels$, or $\rules$ if the schema is clear from the context or irrelevant for the argument.
+
+   In a typed information system, atoms in a relation have a type as declared in a relation.
+   For this reason, every relation $r$ has not only a name, but also a source concept and a target concept.
+   We write $r=\declare{nm}{A}{B}$ to denote that relation $r$ has name \id{nm}, source concept $A$, and target concept $B$.
+   We require that the source and target concepts are in the set of concepts:
+\begin{eqnarray}
+   \forall\declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts
+\end{eqnarray}
    A rule in a schema serves to constrain the dataset,
    to ensure the semantic integrity of the dataset.
-   For every rule $u$ in the schema, the Ampersand compiler generates code for the predicate $\id{sat}_u\{\Dataset\}$
+   For every rule $u$ in the schema, the Ampersand compiler generates code for the predicate $\sat{u}{\Dataset}$
    to determine whether rule $u$ is satisfied in dataset $\dataset$.
@@
    The purpose of an information system is to make data meaningful to its users.
+   Users have their own tasks and responsibilities
+   and may work from different locations and on different moments.
+   This collective use serves a purpose which we loosely call ``the business''.
+   As a consequence, the data in a system changes continually.
+   To preserve meaning, users are trying to maintain semantic constraints on the data,
+   amidst of all changes that are going on around them.
+
    This section introduces the notion of information system as a quintuple $\la\dataset_{\infsys},\schema_{\infsys},\inst_{\infsys},\roles_{\infsys},\maintain_{\infsys}\ra$.
@@
 \begin{itemize}
-   \item dataset $\dataset$ and $\schema$ are defined as in sections~\ref{sct:Schemas} and~\ref{sct:Datasets};
+   \item dataset $\dataset$ is defined as in section~\ref{sct:Datasets};
+   \item schema $\schema$ is defined as in sections~\ref{sct:Schemas};
+   \item $\inst : \Atoms\times\Concepts$ is the instance relation between atoms from $\dataset$ and concepts from $\schema$;
    \item $\roles$ is a set of roles;
-   \item $\inst : \Atoms\times\Concepts$ is the instance relation between atoms from $\dataset$ and concepts from $\schema$;
    \item $\maintain : \roles\times\rules$ is the maintainance relation between roles and rules from $\schema$; and
@@
    This means that every atom is an instance of a concept:
+   To do type checking based on the schema only,
+   all concepts in the relation $\id{inst}$ and all relations in triples from the dataset must be ``known'' in the schema:
 \begin{eqnarray}
-   \forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\label{eqn:inst}
-\end{eqnarray}
-   To do type checking based on the schema only,
-   all relations in triples from the dataset must be ``known'' in the schema:
-\begin{equation}
-   \forall a,r,b:\ \triple{a}{r}{b}\in\dataset\ \Rightarrow\ r\in\rels
-\end{equation}
-
-   In a typed information system, atoms in a relation have a type as declared in a relation.
-   For this reason, every relation $r$ has not only a name, but also a source concept and a target concept.
-   We write $r=\declare{nm}{A}{B}$ to denote that relation $r$ has name \id{nm}, source concept $A$, and target concept $B$.
-\begin{eqnarray}
-   \forall\declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts\\
-   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\dataset&:&a\ \inst\ A\ \wedge\ b\ \inst\ B
+   a\ \inst\ C&\Rightarrow&C\in\concepts\label{eqn:inst}\\
+   \triple{a}{r}{b}\in\dataset&\Rightarrow&r\in\rels\label{eqn:define R}
 \end{eqnarray}
 
-   To compare atoms for equality,
-   the set $\rels$ must contain an identity relation $\ident{C}$ for every concept $C$ in which all atoms occur.
-   It must satisfy:
-\begin{equation}
-   \forall a,b:\ a\ \inst\ C\ \wedge\ b\ \inst\ C\ \Rightarrow\ (a\ \ident{C}\ b\ \Leftrightarrow\ a=b)
-\end{equation}
-   So, $\id{pop}_{\ident{C}}$ is defined by:
 \begin{eqnarray}
-   \forall C\in\concepts&:&\pop{\ident{C}}{\dataset}=\{\pair{a}{a}| a\ \inst\ C\}
+   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\dataset&:&a\ \inst\ A\ \wedge\ b\ \inst\ B\label{eqn:type}
 \end{eqnarray}
@@
    \label{eqn:specialization}
-   \id{sat}_{A\isa B}\{\dataset\}\ \Leftrightarrow\ \forall a: a\ \inst\ A\rightarrow a\ \inst\ B
+   \sat{A\isa B}{\dataset}\ \Leftrightarrow\ \forall a: a\ \inst\ A\rightarrow a\ \inst\ B
 \end{equation}
@@
    Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.
-
    All specialization rules in a schema together form a partial order $\isa : \Pair{\concepts}{\concepts}$.
@@
    For example, code that determines equality between atoms only works for atoms that are an instance of the same concept.
-   To ensure that this does not result in problems at run-time, users are required to write specialization rules that make these comparisons possible.
-   The type checker in the Ampersand compiler will spot violations and refuse to generate a system if they occur.
+   To ensure that this does not result in problems at run-time, users are required to write all specialization rules explicitly.
 
@@
 
-   Every rule is assigned to a role.
-   That role must ensure to satisfy that rule, after some event has violated the rule by changing the set of triples.
+   A \define{role} is a name that identifies a group of users.
+   It serves as a placeholder for a person or a machine (i.e. an actor) who can change the dataset.
+   The purpose of a role is to mention an individual user or an automated actor without knowing who that user is.
+   In the sequel, when we talk about a role we actually mean an actor who fulfills that role in the system.
+
+   Every rule is assigned to at least one role by means of the relation $\maintain$.
+   $o\ \maintain\ u$ means that role $o$ must keep rule $u$ satisfied, i.e.\ to keep $\sat{u}{\dataset}$ true.
+   After some event has violated a rule $u$ (by making changes in the dataset),
+   role $o$ maintains this rule by making other changes to satisfy rule $u$ again.
+   We call this ``to restore the invariance'' of rule $u$.
    In other words: If anyone edits in the dataset and violates a rule, someone or something has to fix the dataset to satisfy that rule again.
    Example: if a rule says that every request must be acknowledged, the receipt of a request will violate this rule.
-   Sending an acknowledgement message will cause the rule to be satisfied again.
+   Sending a message that acknowledges this receipt will cause the rule to be satisfied again.
 
-   Users have their own tasks and responsibilities
-   and may work from different locations and on different moments.
-   This collective use serves a purpose which we loosely call ``the business''.
-   As a consequence, the data in a system changes continually.
-   To preserve meaning, users are trying to maintain semantic constraints on the data,
-   amidst of all changes that are going on around them.
-   So, let us first define the constituent notions role and rule before defining information systems proper.
-
-   A \define{role} is a name that identifies a group of users.
-   The purpose of a role is to mention an individual user without knowing who that user is.
-   In the sequel, when we talk about a role we actually mean a user (an arbitrary one) who fulfills that role in the system.
-
-   The relation $\maintain$ links rules to roles.
-   $o\ \maintain\ u$ means that role $o$ must keep the set $\viol{u}{\dataset}$ empty.
-   If a rule $u$ has violations, i.e. $\viol{u}{\dataset}$ is not empty, there is a role whose job it is to remove those violations.
-   The relation $\maintain$ being surjective means that all users together are keeping the violation set empty.
+   The fact that every rule has at least one role in the relation $\maintain$, means that all roles together are keeping all rules satisfied.
 
@@
    Having defined an information system in mathematical terms, let us give an example.
-   For this purpose we use a subset of the language Ampersand
+   For this purpose we use the language Ampersand
    because it makes the examples more appealing to read.
@@
 \begin{verbatim}
-RELATION isEnrolledFor [Student*Module]
+RELATION isEnrolledFor[Student*Module]
 \end{verbatim}
-   Summarizing, here we have a set of atoms, which contains \verb-"Peter"-, \verb-"Finance"-, etc.,
-   a set of concepts $\{$\verb-Student-, \verb-Module-, \verb-Course-$\}$,
-   an instance relation that states which atoms belong to which concept,
-   a specialization relation, which is empty,
-   a set of three relations, and
-   a set of eight triples.
-   The example in Ampersand defines precisely that dataset.
-
-   Now let us define an information system.
-   We define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that student takes.
+   We also define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that student takes.
    Its constraint, $p_{\tt Enroll}$ is written in logic as:
@@
 \end{verbatim}
-   The Ampersand compiler derives the signature of every rule.
-   This rule has signature $\pair{\text{\tt Student}}{\text{\tt Module}}$.
-   The Ampersand compiler also derives the violation set as:
-\[\begin{array}{l}
-   \viol{\text{\tt EnrollRule}}{\dataset}\\
-   \hspace{1cm}=\{\pair{s}{m}|s\ \text{\tt isEnrolledFor}\ m\ \wedge\not\exists c: s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\}
-\end{array}\]
-   Notice that the current population yields an empty set of violations,
-   so the rule is satisfied.
-   Having one rule, one role, a relation $\maintain$ that contains one pair, and a dataset,
-   we have now defined a small system.
+
+   Summarizing, this example contains a dataset with seven triples
+   and a schema with three concepts, three relations, and one rule.
+   Now let us define an information system.
+   Notice that the current population satisfies the rule.
+   Let us check requirements~\ref{eqn:inst} thru~\ref{eqn:specialization} to verify that this is an information system.
+   Requirement~\ref{eqn:concepts} defines the set $\concepts$.
+   Requirement~\ref{eqn:define R} defines the set $\rels$.
+   Requirement~\ref{eqn:type} defines the relation $\id{inst}$.
+
+%    The language Ampersand is typed, to prevent a substantial amount of programming mistakes at compile-time.
+%    This means that every atom is an instance of a concept:
+% \begin{eqnarray}
+%    \forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\label{eqn:inst}
+% \end{eqnarray}
+%    To do type checking based on the schema only,
+%    all relations in triples from the dataset must be ``known'' in the schema:
+% \begin{equation}
+%    \forall a,r,b:\ \triple{a}{r}{b}\in\dataset\ \Rightarrow\ r\in\rels\label{eqn:define R}
+% \end{equation}
+
+%    In a typed information system, atoms in a relation have a type as declared in a relation.
+%    For this reason, every relation $r$ has not only a name, but also a source concept and a target concept.
+%    We write $r=\declare{nm}{A}{B}$ to denote that relation $r$ has name \id{nm}, source concept $A$, and target concept $B$.
+% \begin{eqnarray}
+%    \forall\declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts\label{eqn:concepts}\\
+%    \forall\triple{a}{\declare{n}{A}{B}}{b}\in\dataset&:&a\ \inst\ A\ \wedge\ b\ \inst\ B\label{eqn:type}
+% \end{eqnarray}
+%    The type checker in the Ampersand compiler will spot violations and refuse to generate a system if they occur.
+
+%    To compare atoms for equality,
+%    the set $\rels$ contains an identity relation $\ident{C}$ for every concept $C$ in which all atoms occur.
+%    It satisfies:
+% \begin{equation}
+%    \forall a,b:\ a\ \inst\ C\ \wedge\ b\ \inst\ C\ \Rightarrow\ (a\ \ident{C}\ b\ \Leftrightarrow\ a=b)
+% \end{equation}
+%    So, $\id{pop}_{\ident{C}}$ is defined by:
+% \begin{eqnarray}
+%    \forall C\in\concepts&:&\pop{\ident{C}}{\dataset}=\{\pair{a}{a}\mid a\ \inst\ C\}
+% \end{eqnarray}
+
+%    Having one rule, one role, a relation $\maintain$ that contains one pair, and a dataset,
+%    we have now defined a small system.
 
@@
    &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\ \inst\uplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \dataset\uplus^5\dataset'\ra\notag\\
-      X\uplus Y&=&\{(x,0)|\ x\in X\}\ \cup\ \{(y,1)|\ y\in Y\}\\
-      X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))|\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))|\ (y_1,y_2)\in Y\}\end{array}\\
+      X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
+      X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}\\
       X\uplus^3 Y&=&\begin{array}[t]{@{}l}\{\declare{(n,0)}{(A,0)}{(B,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\declare{(n,1)}{(A,1)}{(B,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}\\
@@
            &\cup&\begin{array}[t]{@{}l}
-            \{ {\tt CLASSIFY\ }(C',1){\tt\ IS\ }(C,0)|\ C\in\concepts,C'\in\concepts',C=C'\}\\
+            \{ {\tt CLASSIFY\ }(C',1){\tt\ IS\ }(C,0)\mid\ C\in\concepts,C'\in\concepts',C=C'\}\\
            \end{array}
@@
    We define a \define{migration of a dataset} $\la\atoms,\concepts,\inst,\rels,\dataset\ra$ as a change in which one or more of $\inst$, $\concepts$, or $\rels$ change.
-   To satisfy definition~\ref{def:dataset}, $\dataset$ and $\atoms$ must change too,
+   To satisfy the definition of datasets, $\dataset$ and $\atoms$ must change too,
    but the migration should try to preserve the maximal amount of data.
-
-   This paper studies those changes and seeks to preserve the rules in definition~\ref{def:dataset}
-   in such a way that large portions of migration can be automated.
 

### 2022-10-30 2022 Migration/articleMigration.tex
WIP

@@
    We will use variables $A$, $B$, $C$, $D$ to represent concepts.
-   The expression $a\ \inst\ A$ means that atom $a$ is an \emph{instance} of concept $A$.
+   The expression $a\ \inst\ C$ means that atom $a$ is an \emph{instance} of concept $C$.
+   The relation $\inst:\Pair{\Atoms}{\Concepts}$ is the instance relation between atoms and concepts.
    All concepts are taken from an infinite set $\Concepts$.
@@
    Triples serve to represent data.
-   A dataset $\dataset$ is a set of triples.
    A triple is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.
@@
    in those cases where $\dataset$ is either clear from the context or irrelevant for the argument.
+
+   A dataset $\dataset$ is a tuple $\pair{\triples_{\dataset}}{\inst_{\dataset}}$,
+   in which $\triples_{\dataset}$ is a finite set of triples
+   and $\inst_{\dataset}$ is the instance relation.
+   We will simply write $\triples$ or $\inst$ if the dataset is clear from the context or irrelevant for the argument.
+
    A relation $r$ can serve as a container of pairs,
-   as defined by the funciton $\id{pop}_r:\Dataset\rightarrow\Pair{\Atoms}{\Atoms}$.
+   as defined by the function $\id{pop}_r:\Dataset\rightarrow\Pair{\Atoms}{\Atoms}$.
    It defines a set of pairs, which we call the population of $r$:
 \begin{equation}
-   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}\mid\ \triple{a}{r}{b}\in\dataset\}
+   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}\mid\ \triple{a}{r}{b}\in\triples\}
 \end{equation}
@@
 
-   This section introduces the notion of information system as a quintuple $\la\dataset_{\infsys},\schema_{\infsys},\inst_{\infsys},\roles_{\infsys},\maintain_{\infsys}\ra$.
+   This section introduces the notion of information system as a quadrtuple $\la\dataset_{\infsys},\schema_{\infsys},\roles_{\infsys},\maintain_{\infsys}\ra$.
    As before, we will omit the suffixes if $\infsys$ is obvious (i.e. clear from the context or irrelevant for the argument).
@@
 \label{def:information system}
-\item An information system $\infsys$ is a tuple $\la\dataset,\schema,\inst,\roles,\maintain\ra$, in which
+\item An information system $\infsys$ is a tuple $\la\dataset,\schema,\roles,\maintain\ra$, in which
 \begin{itemize}
@@
    \item schema $\schema$ is defined as in sections~\ref{sct:Schemas};
-   \item $\inst : \Atoms\times\Concepts$ is the instance relation between atoms from $\dataset$ and concepts from $\schema$;
    \item $\roles$ is a set of roles;

### 2022-11-02 2022 Migration/articleMigration.tex
Example in sct. 3.4 is OK with the new definitions

@@
 
-   We recognize that designs from which concrete systems are generated can have errors in them:
-   Either an engineer was making a mistake directly,
-   or the requirements the engineer was taking into account were miscommunicated,
-   or the environment changed in a way that was not foreseen.
-   Regardless of the reason, even a generated system may need to be repaired.
-   For this reason, we focus on evolving systems.
-
-   This paper studies this evolution in increments, in which we distinguish an ``old system'' and a ``new system''.
-   In each increment we need to migrate the data from the old system to the newly designed system.
-   This situation poses specific requirements to data migration.
-   For instance, if a data migration has other purposes than just to evolve a system,
+   In each increment, we distinguish an ``old system'' and a ``new system''.
+   The difference between the old system and the new system is not just a difference in two schemas.
+   When deploying the increment, the data from the old system needs to migrate to the new system as well.
+   Incremental evolution poses specific requirements to data migration,
+   which sets it apart from data migrations for other purposes.
+   For instance, if a data migration is done for switching to another platform or to different technology,
    e.g.~\cite{Gholami2016,Bisbal1999},
    migration engineers may want to preserve the functionality to avoid introducing new errors in an otherwise error-prone process.
-   When one system evolves into another, however, preserving the functionality is not an option.
-   To change the functionality is the very purpose of the evolution.
-   So, we must make sure to use methods for data migration that are specifically meant for evolving systems.
+   In our work, however, changing the functionality is the very purpose of an increment.
+   We can also interpret an increment as a variation between two data structures,
+   which is a well studied subject by now, e.g.~\cite{Walkingshaw2014}.
+   For instance, Ataei et.al.~\cite{Ataei2021} show how to unify databases with slight variations.
+   They do this, however, by preserving all variations in one comprehensive database.
+   However, if we want to get rid of the old system, we need to look at other approaches.
+   Such examples have convinced us to define a method for data migration that is specifically meant for systems that evolve incrementally.
 
@@
    The Ampersand compiler generates systems that keep all constraints satisfied at all times.
+   This is why we have done this research with Ampersand.
    Much like~\cite{Spivak2012}, our work assumes that the semantics of the old dataset is encoded in its schema,
@@
    Concepts are names that group atoms of the same type.
+   All concepts are taken from an infinite set $\Concepts$.
+   $\Concepts$ and $\Atoms$ are disjoint sets.
    For example, an engineer might choose to classify \atom{Peter} and \atom{Melissa} as \concept{Person},
@@
    The relation $\inst:\Pair{\Atoms}{\Concepts}$ is the instance relation between atoms and concepts.
-   All concepts are taken from an infinite set $\Concepts$.
-   $\Concepts$ and $\Atoms$ are disjoint sets.
+   This relation is used in the type system.
 
@@
    A triple is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.
-   This makes our theory valid for any kind of database that triples can represent,
-   such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases
-   Each triple relates two atoms to a relation.
    A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean in practice that
@@
    In this paper, the informal meaning is interesting only for the sake of illustration.
-   To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\dataset$
-   in those cases where $\dataset$ is either clear from the context or irrelevant for the argument.
+   This makes our theory valid for any kind of database that triples can represent,
+   such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
 
@@
    We will simply write $\triples$ or $\inst$ if the dataset is clear from the context or irrelevant for the argument.
+   Every dataset is an element of an infinite set called $\Dataset$.
+   To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\triples$.
 
@@
 \begin{equation}
-   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}\mid\ \triple{a}{r}{b}\in\triples\}
+   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}\mid\ \triple{a}{r}{b}\in\triples_{\dataset}\}
+   \label{eqn:pop}
 \end{equation}
 
-   Over time, an information system ``observes'' events that change the dataset by inserting and deleting triples in $\dataset$.
-   By the same account, the population of relations change over time.
-   These events may be caused by external systems, roles acting on the dataset, or any other force that changes triples.
+   Over time, an information system ``observes'' events that change the dataset by inserting and deleting triples.
+   By the same account, the population of relations changes over time.
+   These events may be caused by external systems, roles acting on the dataset, or any other force that changes triples in a dataset.
 
@@
 
-   In a typed information system, atoms in a relation have a type as declared in a relation.
-   For this reason, every relation $r$ has not only a name, but also a source concept and a target concept.
+   In order to attach types to atoms, relations must be declared in the schema.
+   Every relation $r$ has not only a name, but also a source concept and a target concept.
    We write $r=\declare{nm}{A}{B}$ to denote that relation $r$ has name \id{nm}, source concept $A$, and target concept $B$.
@@
    \forall\declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts
+   \label{eqn:relationsIntroduceConcepts}
 \end{eqnarray}
@@
    to ensure the semantic integrity of the dataset.
-   For every rule $u$ in the schema, the Ampersand compiler generates code for the predicate $\sat{u}{\Dataset}$
-   to determine whether rule $u$ is satisfied in dataset $\dataset$.
-   The Ampersand compiler also generates code for a function $\id{viol}_u : \Dataset\rightarrow\id{Text}$
-   to produce meaningful error messages in case rule $u$ is not satisfied.
+   For every rule $u$ in the schema, the Ampersand compiler generates code
+   for the function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$.
+   For any given dataset $\dataset$, $\viol{u}{\dataset}$ represents the set of violations of rule $u$.
+   The type of the atoms in a violation set is defined by the function $\Sign{u} : \Dataset\rightarrow\Pair{\Concepts}{\Concepts}$:
+\begin{eqnarray}
+   \sign{u}{\dataset}=\pair{A}{B}\ \wedge\ \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\ \inst\ A\wedge b\ \inst\ B
+   \label{eqn:typed violations}
+\end{eqnarray}
+To determine whether rule $u$ is satisfied in dataset $\dataset$,
+the Ampersand compiler also generates code for a predicate $\sat{u}{\Dataset}$:
+\begin{eqnarray}
+   \sat{u}{\dataset}&\Leftrightarrow&\viol{u}{\dataset}=\emptyset
+   \label{eqn:sat}
+\end{eqnarray}
+   Violations are used to produce meaningful error messages and to trigger changes in the dataset to satisfy rule $u$ again.
 
@@
 
-   This section introduces the notion of information system as a quadrtuple $\la\dataset_{\infsys},\schema_{\infsys},\roles_{\infsys},\maintain_{\infsys}\ra$.
-   As before, we will omit the suffixes if $\infsys$ is obvious (i.e. clear from the context or irrelevant for the argument).
+   This section introduces the notion of information system as a quadruple $\la\dataset_{\infsys},\schema_{\infsys},\roles_{\infsys},\maintain_{\infsys}\ra$.
+   As before, we will omit the suffixes if $\infsys$ is self-evident.
 \begin{definition}[information system]
@@
    \item dataset $\dataset$ is defined as in section~\ref{sct:Datasets};
-   \item schema $\schema$ is defined as in sections~\ref{sct:Schemas};
+   \item schema $\schema$ is defined as in section~\ref{sct:Schemas};
    \item $\roles$ is a set of roles;
    \item $\maintain : \roles\times\rules$ is the maintainance relation between roles and rules from $\schema$; and
-   \item requirements~\ref{eqn:inst} thru~\ref{eqn:specialization} are satisfied.
+   \item requirements~\ref{eqn:pop} thru~\ref{eqn:specialization} are satisfied.
 \end{itemize}
@@
    The language Ampersand is typed, to prevent a substantial amount of programming mistakes at compile-time.
-   This means that every atom is an instance of a concept:
-   To do type checking based on the schema only,
-   all concepts in the relation $\id{inst}$ and all relations in triples from the dataset must be ``known'' in the schema:
+   Type checking is based on the schema only.
+   So, all concepts in the relation $\id{inst}$ and all relations in triples from the dataset must be ``known'' in the schema:
 \begin{eqnarray}
-   a\ \inst\ C&\Rightarrow&C\in\concepts\label{eqn:inst}\\
-   \triple{a}{r}{b}\in\dataset&\Rightarrow&r\in\rels\label{eqn:define R}
+   a\ \inst\ C&\Rightarrow&C\in\concepts
+   \label{eqn:concepts}\\
+   \triple{a}{r}{b}\in\triples&\Rightarrow&r\in\rels
+   \label{eqn:define R}
 \end{eqnarray}
-
+   Every atom is an instance of a concept:
 \begin{eqnarray}
-   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\dataset&:&a\ \inst\ A\ \wedge\ b\ \inst\ B\label{eqn:type}
+   a\ \inst\ A&\Leftrightarrow&(\exists n,b,B:\ \triple{a}{\declare{n}{A}{B}}{b}\in\triples\ \vee\ \triple{b}{\declare{n}{B}{A}}{a}\in\triples)
+   \label{eqn:instance}
+\end{eqnarray}
+   Every atom has the type of the relation in which it resides:
+\begin{eqnarray}
+   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B
+   \label{eqn:type}
 \end{eqnarray}
@@
 \label{old IS}
-   Having defined an information system in mathematical terms, let us give an example.
+   Having defined an information system in mathematical terms, let us discuss a small example.
    For this purpose we use the language Ampersand
@@
 \end{verbatim}
-   This introduces a relation with name \verb#takes#,
+   This declaration introduces a relation with name \verb#takes#,
    source concept \verb#Student#, and
@@
    \triple{\text{\tt "Susan"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
-   \triple{\text{\tt "John"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
+   \triple{\text{\tt "John"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}
 \end{array}\]
@@
    We also define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that student takes.
-   Its constraint, $p_{\tt Enroll}$ is written in logic as:
-\[\begin{array}{l}\forall s\in\text{\tt Student}, m\in\text{\tt Module}\ \exists c\in\text{\tt Course}:\\
-s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
-\end{array}\]
    In Ampersand, which is a syntactically sugared form of relation algebra,
-   we give each rule a name and declare a role to maintain it:
+   we give each rule a name and declare a role to preserve its invariance:
 \begin{verbatim}
@@
 \end{verbatim}
+   The actual constraint, \verb#isEnrolledFor |- takes;isPartOf~#, is written in logic as:
+\[\begin{array}{l}\forall s\in\text{\tt Student}, m\in\text{\tt Module}\ \exists c\in\text{\tt Course}:\\
+s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
+\end{array}\]
 
@@
    and a schema with three concepts, three relations, and one rule.
-   Now let us define an information system.
-   Notice that the current population satisfies the rule.
-   Let us check requirements~\ref{eqn:inst} thru~\ref{eqn:specialization} to verify that this is an information system.
-   Requirement~\ref{eqn:concepts} defines the set $\concepts$.
-   Requirement~\ref{eqn:define R} defines the set $\rels$.
-   Requirement~\ref{eqn:type} defines the relation $\id{inst}$.
+
+   Now let us check the requirements to verify that this example defines an information system.
+   Requirement~\ref{eqn:pop} results in a set $\triples$ that contains seven triples.
+   Requirements~\ref{eqn:instance} and~\ref{eqn:type} define the relation $\id{inst}$ to contains the pairs:
+   \[\begin{array}{l}
+      \pair{\tt "Finance"}{\tt Module}\\
+      \pair{\tt "Business Rules"}{\tt Module}\\
+      \pair{\tt "Business Analytics"}{\tt Module}\\
+      \pair{\tt "IT-Governance"}{\tt Module}\\
+      \pair{\tt "Management"}{\tt Course}\\
+      \pair{\tt "Business IT"}{\tt Course}\\
+      \pair{\tt "Peter"}{\tt Student}\\
+      \pair{\tt "Susan"}{\tt Student}\\
+      \pair{\tt "John"}{\tt Student}
+   \end{array}\]
+   Requirements~\ref{eqn:concepts} and~\ref{eqn:relationsIntroduceConcepts} both say that 
+   \[\concepts=\{ {\tt Module}, {\tt Course}, {\tt Student}\}\]
+   Requirement~\ref{eqn:define R} says that 
+   \[\begin{array}{rcl}
+      \rels&=&\{\begin{array}[t]{l}
+                  \declare{\tt takes}{\tt Student}{\tt Course},\\
+                  \declare{\tt isPartOf}{\tt Module}{\tt Course},\\
+                  \declare{\tt isEnrolledFor}{\tt Student}{\tt Module}\ \}
+                \end{array}
+     \end{array}
+   \]
+   The set of rules $\rules$ contains just one rule: \verb-EnrollRule-.
+   So $\concepts$, $\relations$, and $\rules$ complete the schema $\schema$ of this example.
+   The dataset $\dataset$ contains the set $\triples$ and the relation $\inst$.
+   The set of roles, $\roles$, contains one role: ${\tt Administrator}$
+   and the relation $\maintain$ relates ${\tt Administrator}$ to rule\verb-EnrollRule-.
+   Requirement~\ref{eqn:typed violations} and~\ref{eqn:sat} are satisfied because $\viol{\tt EnrollRule}{\dataset}$ is empty.
+   Requirement~\ref{eqn:specialization} is satisfied because this example contains no specialization.
+   So we can conclude that this example satisfies definition~\ref{def:information system}.
 
@@
 % \begin{eqnarray}
-%    \forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\label{eqn:inst}
+%    \forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\label{eqn:concepts}
 % \end{eqnarray}
@@
 % \begin{equation}
-%    \forall a,r,b:\ \triple{a}{r}{b}\in\dataset\ \Rightarrow\ r\in\rels\label{eqn:define R}
+%    \forall a,r,b:\ \triple{a}{r}{b}\in\triples\ \Rightarrow\ r\in\rels\label{eqn:define R}
 % \end{equation}
@@
 %    \forall\declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts\label{eqn:concepts}\\
-%    \forall\triple{a}{\declare{n}{A}{B}}{b}\in\dataset&:&a\ \inst\ A\ \wedge\ b\ \inst\ B\label{eqn:type}
+%    \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B\label{eqn:type}
 % \end{eqnarray}

### 2022-11-20 2022 Migration/articleMigration.tex
Trying to make the migration procedure explicit (plus other work)

@@
 \footnote{In the sequel, the word ``system'' refers to the phrase ``information system''. This simplifies the language a little. }
-   to prevent human errors.
+   to prevent human programming errors.
    By generating rather than programming systems by hand,
@@
    e.g.~\cite{Gholami2016,Bisbal1999},
-   migration engineers may want to preserve the functionality to avoid introducing new errors in an otherwise error-prone process.
-   In our work, however, changing the functionality is the very purpose of an increment.
-   We can also interpret an increment as a variation between two data structures,
-   which is a well studied subject by now, e.g.~\cite{Walkingshaw2014}.
-   For instance, Ataei et.al.~\cite{Ataei2021} show how to unify databases with slight variations.
-   They do this, however, by preserving all variations in one comprehensive database.
-   However, if we want to get rid of the old system, we need to look at other approaches.
-   Such examples have convinced us to define a method for data migration that is specifically meant for systems that evolve incrementally.
+   migration engineers may exclude new functionality to avoid introducing new errors in an otherwise error-prone migration process.
+   In our work, however, introducing new functionality and removing obsolete functionality is the very purpose of an increment.
+   Data migration literature sometimes uses a different notion of increment.
+   For example, Walkingshaw et.al.~\cite{Walkingshaw2014} and Ataei~\cite{Ataei2021} define an increment as a variation between two data structures.
+   Ataei et.al. show how to unify databases with slight variations by preserving all variations in one comprehensive database.
+   In our work, however, we want to get rid of the old system and not retain it.
+   So, we need to look at a different approach.
+   Examples like these have convinced us to define a new method for data migration that is specifically meant for systems under incremental development.
 
+   The contribution of this paper is to derive a migration schema from two artifacts: an existing system
+   (which has its own schema and dataset) and a new schema (which specifies the new system),
+   which allows us to facilitate the migration.
+   The purpose of our theory is to develop tools.
+   These tools must satisfy such practical requirements as
+   zero down-time updates and ways to cope with data pollution.
+   In practice, more frequent updates require that updates come with zero down-time.
+   For our theory, zero down-time means that the release of new functionality may not be delayed by anything a migration engineer must do.
+
+   This theory has enhanced our formal understanding of data migration and it will help to automate data migrations correctly and reliably.
+   The benefits work in two directions.
+   Automation data migrations saves effort and errors,
+   contributing to more frequent and smaller releases.
+   Smaller and faster releases also mean smaller migrations
+   and therefore a smaller difference between the old system and the new system.
+   The smaller that difference, the simpler the migration and the more of it can be automated.
+
+   Our theory is based on the following assumptions and requirements:
+\begin{itemize}
+   \item the data migration is solely meant to deploy a software increment in production;
+   \item the old data set may be polluted, but it satisfies its schema;
+   \item the data migration may require human interaction, which may take time;
+   \item the meaning of data must be preserved;
+   \item the business continues during the migration without interruption;
+   \item there is a compiler to generate an information system from a given schema.
+         In this work, we use the Ampersand compiler for that purpose.
+\end{itemize}
+
+\section{Analysis}
    Information systems are typically used by distributed actors (both users and computers) who continually work with data.
-   A system $\infsys$ contains a dataset $\dataset_{\infsys}$, which represents the state of the system.
-   (In the sequel we may just write $\dataset$
-   if the corresponding system is either clear from the context or irrelevant for the argument.)
+   A system $\infsys$ contains a dataset $\dataset$, which represents the state of the system.%
+\footnote{Where ambiguity lures, we write $\dataset_{\infsys}$ instead of $\dataset$, to indicate that we specifically mean the dataset of $\infsys$.}
    Every event that an actor causes in system $\infsys$ may change the state $\dataset$ of that system.
+   To keep our theory technology independent, we will assume that a dataset contains triples.
    Data migration occurs when an old system, $\infsys$, is replaced by a new system, $\infsys'$,
@@
 
-   The schema $\schema_{\infsys}$ of a system $\infsys$ is basically a set of semantic constraints on its dataset $\dataset$.
-   (We will denote the schema just as $\schema$ if the system is obvious or irrelevant to know.)
-   The Ampersand compiler generates systems that keep all constraints satisfied at all times.
+   The schema $\schema$ of a system $\infsys$ is basically a set of semantic constraints on its dataset $\dataset$.%
+\footnote{Where ambiguity lures, we write $\schema_{\infsys}$ instead of $\schema$ to indicate that we specifically mean the schema of $\infsys$.}
+   The Ampersand compiler generates systems, which keep all constraints satisfied at all times.
    This is why we have done this research with Ampersand.
-   Much like~\cite{Spivak2012}, our work assumes that the semantics of the old dataset is encoded in its schema,
-   so we can be explicit about ``preserving the meaning as much as possible''.
+   Much like~\cite{Spivak2012}, in our work the schema encodes the semantics of a system.
+   This will allow us to be explicit about ``preserving the meaning as much as possible''.
    In contrast with~\cite{Spivak2012}, our approach does not work with functions (functors)
@@
 
-   The theory we develop in this paper must lead to tools that satisfy such practical requirements as
-   zero down-time updates and ways to cope with data pollution.
-   In practice, more frequent updates require that updates have zero down-time.
-   For our theory, zero down-time means that the release of new functionality may not be delayed by anything a migration engineer must do.
-
    Extra complications arise from accumulated data pollution in the old system.
-   This allows us to categorize data pollution in the old system in the following categories:
+   We distinguish data pollution in the following categories:
 \begin{itemize}
    \item data pollution that cannot be captured in constraints on the dataset,
-      such as a street address that has become obsolete because the person failed to send a change of address.
+   such as a street address that has become obsolete because the system never registered the change of address.
    \item data pollution that violates a constraint in the schema.
@@
    Such constraints move this type of pollution to one of the other categories.
-   The second category, violations of constraints from the schema, does not occur in our work.
-   Since the old system $\infsys$ was generated by Ampersand,
+   The second category is about violations of constraints from the schema.
+   Such violations do not occur in our theory because the old system $\infsys$ was generated by Ampersand,
    $\infsys$ keeps all constraints from the schema satisfied at all times.
    This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
-   Data pollution that is not captured by any of the constraints mentioned in the schema
+   The third category, data pollution that is not captured by any of the constraints mentioned in the schema,
    can be dealt with by adding constraints in the new system $\infsys'$.
@@
 
-   This illustrates that automating data migrations also has its limits.
+   This illustrates that automating data migrations has its limits.
    Some effort of a migration engineer may be required.
@@
    a migration engineer must make choices on how to interpret existing data in the new schema.
-   She may even want to adapt data of which the constraints in the schema have not changed.
-   This requires a third schema, $\schema_m$, which specifies the migration itself.
-   We generate $\schema_m$ in the form of source code, so the migration engineer can change everything she wants to suit the particulars of the data migration.
-   In the sequel, automation of data migration means to generate the schema $\schema_m$.
+   She may even want to adapt existing data of which the constraints in the schema have not changed.
+   This requires a third schema, $\schema_{\migrsys}$, which specifies the migration itself.
+   It preserves all data from the old system, which may be needed for the migration.
+   It also contains the new schema and possibly some new data, which is specified by the migration engineer.
+   Hence, the migration is based on the disjoint union of $\schema$ and $\schema'$.
+   We generate $\schema_{\migrsys}$ in the form of source code, so the migration engineer can change everything she wants to suit the particulars of the data migration.
 
-   The benefits of automating data migrations work in two directions.
-   The automation itself saves effort and errors,
-   contributing to more frequent and smaller releases.
-   Smaller and faster releases also mean smaller migrations
-   and therefore a smaller difference between $\infsys$ and $\infsys'$.
-   The smaller that difference, the simpler the migration and the more of it can be automated.
-
-   The contribution of this paper is to derive a migration schema $\schema_m$ from an existing system $\infsys$
-   (which has its own schema and dataset) and a new schema $\schema'$.
-   A formal understanding of data migration will help to automate data migrations correctly and reliably.
-   Our theory is based on the following assumptions:
-\begin{itemize}
-   \item the old data set may be polluted, but it satisfies its schema;
-   \item the data migration may require human interaction, which may take time;
-   \item the part of the data that does not change can be migrated automatically;
-   \item the meaning of data must be preserved, at least for the data that does not change;
-   \item the business must continue during the migration without interruption;
-   \item the part of the migration that can be automated is usually not sufficient;
-         it takes additional human creativity to complete the migration specification;
-   \item there is a compiler to generate an information system from a given schema.
-         In this work, we use the Ampersand compiler for that purpose.
-\end{itemize}
-
-\section{Migration steps}
-   A smooth migration would ideally proceed as follows:
+   A naive migration might proceed as follows:
    Launch the new system in parallel to the old one, copy data from the old to the new system, and have everyone use the new system.
@@
 
+   In the following section we introduce the definitions required to migrate data from one system to another.
+
 \section{Terminology}
 \label{sct:Terminology}
-   To migrate data from one system to another,
-   we define ``information system'' as a combination of dataset and schema.
-   So, let us first define datasets in section~\ref{sct:Datasets} as the focal point of a data migration.
-   Then we define schemas in section~\ref{sct:Schemas}, to capture the semantics of information systems.
+   First,
+   we define ``information system'' as a combination of dataset, schema, and functionality.
+   We start with datasets in section~\ref{sct:Datasets}, then we define schemas in section~\ref{sct:Schemas},
+   and then we can define information systems.
 
@@
    A dataset $\dataset$ describes a set of structured data, which is typically stored persistently in a database of some kind.
+   The purpose of a dataset is to describe the data of a system at one point in time, so that data changes are representable. 
    Before defining datasets, we must first define the constituent notions of atom, concept, relation, and triple.
@@
 
-   A dataset $\dataset$ is a tuple $\pair{\triples_{\dataset}}{\inst_{\dataset}}$,
-   in which $\triples_{\dataset}$ is a finite set of triples
-   and $\inst_{\dataset}$ is the instance relation.
-   We will simply write $\triples$ or $\inst$ if the dataset is clear from the context or irrelevant for the argument.
+   A dataset $\dataset$ is a tuple $\pair{\triples}{\inst}$,
+   in which $\triples$ is a finite set of triples
+   and $\inst$ is the instance relation.
+   Where ambiguity lures, we write $\triples_{\dataset}$ and $\inst_{\dataset}$ to indicate that we specifically mean the dataset $\dataset$.
    Every dataset is an element of an infinite set called $\Dataset$.
@@
    To design an information system, an engineer defines a schema in which she defines concepts, relations, and rules.
+   The purpose of a schema is to capture the semantics of an information system.
    Rules serve the purpose of invariants,
@@
 
-   A schema $\schema$ is a triple $\triple{\concepts_{\schema}}{\rels_{\schema}}{\rules_{\schema}}$,
-   in which $\concepts_{\schema}$ is a finite set of concepts,
-   $\rels_{\schema}$ is a finite set of relations,
-   and $\rules_{\schema}$ is a finite set of rules.
-   We will simply write $\concepts$, $\rels$, or $\rules$ if the schema is clear from the context or irrelevant for the argument.
+   A schema $\schema$ is a triple $\triple{\concepts}{\rels}{\rules}$,
+   in which $\concepts$ is a finite set of concepts,
+   $\rels$ is a finite set of relations,
+   and $\rules$ is a finite set of rules.
+   Where ambiguity lures, we write $\concepts_{\schema}$, $rels_{\schema}$, and $\rules_{\schema}$ to indicate that we specifically mean the schema $\schema$.
+%  Every schema is an element of an infinite set called $\Schema$.
 
@@
 
+\subsection{Functionality}
+   The functionality of an information system is an integral part of it.
+   It consists of software.
+   In this paper, we are not interested in the precise nature of the functionality.
+
 \subsection{Information Systems}
@@
 
-   This section introduces the notion of information system as a quadruple $\la\dataset_{\infsys},\schema_{\infsys},\roles_{\infsys},\maintain_{\infsys}\ra$.
-   As before, we will omit the suffixes if $\infsys$ is self-evident.
+   This section introduces the notion of information system.
+   As before, we will suffix the elements of this definition if the situation requires it.
 \begin{definition}[information system]
 \label{def:information system}
-\item An information system $\infsys$ is a tuple $\la\dataset,\schema,\roles,\maintain\ra$, in which
+\item An information system $\infsys$ is a tuple $\la\dataset,\schema,\functionality,\roles,\maintain\ra$, in which
 \begin{itemize}
@@
    \item schema $\schema$ is defined as in section~\ref{sct:Schemas};
+   \item $\functionality$ represents the functionality of the system, which is typically software;
    \item $\roles$ is a set of roles;
@@
 \subsection{Data migration}
-   To migrate system $\infsys$ to $\infsys'$, we take the disjoint union $\infsys\sqcup\infsys'$ of both systems
-   and add rules to describe the migration.
-   $\infsys$ is the existing system, which contains data to be preserved.
-   $\infsys'$ is the new system, which contains no or hardly any data.
-   It does, however, contain the concepts, relations, rules and roles of the new system.
 \begin{figure}[bht]
@@
 \end{figure}
+   To migrate system $\infsys$ to $\infsys'$, we take the following steps:
+\begin{enumerate}
+   \item Let the old system $\infsys$ have state $\dataset$ and schema $\schema$.
+         A developer defines a schema $\schema'$ of the new system as a first step in the deployment of the increment.
+   \item The generator generates a migration system $\migrsys$ based on $\schema$ and $\schema'$,
+         which gives the user the full functionality of the old system and the new system,
+         i.e.\ $\schema_{\migrsys}=\schema\sqcup\schema'$.
+   \item A migration engineer changes $\migrsys$ to accommodate the non-standard requirements of the migration,
+         yielding migration system .
+   \item The production environment switches instantly from $\infsys$ to $\migrsys'$,
+         while $\dataset_{\infsys}\subseteq\dataset_{\migrsys'}$.
+         This means that the data of the old system remains intact.
+   \item After a finite span of time, $\dataset_{\migrsys'}$ satisfies all its enforce rules.
+         This implies that the generated rules plus the changes made by the migration engineer ensure the preservation of the meaning of the old data.
+   \item The system removes the data, the schema, and the functionality of $\migrsys'$ that are not supported by $\schema'$.
+         i.e\ $\schema_{\migrsys''}=\schema'$ and $\dataset_{\migrsys''}=\select{\schema'}{\dataset_{\migrsys''}}$.
+\end{enumerate}
 
-   We use two datasets: $\dataset$ and $\dataset'$.
-   Before doing so, let us first define the disjoint union of two systems.
+   Let us define the instruments needed to describe the derivation of $\schema_{\migrsys}$.
 \begin{definition}[disjoint union of datasets]
 \begin{eqnarray}
-   \omit\rlap{$\la\atoms,\concepts,\inst,\isa,\rels,\dataset\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\dataset'\ra$}\notag\\
-   &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\ \inst\uplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \dataset\uplus^5\dataset'\ra\notag\\
+   \omit\rlap{$\pair{\triples}{\inst}\sqcup\pair{\triples'}{\inst'}$}\notag\\
+   &=&\pair{\triples\uplus\triples'}{\inst\uplus^2\inst'}\notag\\
       X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
       X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}\\
-      X\uplus^3 Y&=&\begin{array}[t]{@{}l}\{\declare{(n,0)}{(A,0)}{(B,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\declare{(n,1)}{(A,1)}{(B,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}\\
-      X\uplus^5 Y&=&\begin{array}[t]{@{}l}\{\triple{(a,0)}{\declare{(n,0)}{(A,0)}{(B,0)}}{(b,0)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in X\}\ \cup\\ \{\triple{(a,1)}{\declare{(n,1)}{(A,1)}{(B,1)}}{(b,1)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in Y\}\end{array}
 \end{eqnarray}
 \end{definition}
+
+\begin{definition}[disjoint union of schemas]
+\begin{eqnarray}
+   \triple{\concepts}{\rels}{\rules}\sqcup\triple{\concepts'}{\rels'}{\rules'}&=&\triple{\concepts\uplus\concepts}{\rels\uplus\rels'}{\rules\uplus\rules'}
+\end{eqnarray}
+\end{definition}
+% \begin{definition}[disjoint union of datasets]
+% \begin{eqnarray}
+%    \omit\rlap{$\la\atoms,\concepts,\inst,\isa,\rels,\dataset\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\dataset'\ra$}\notag\\
+%    &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\ \inst\uplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \dataset\uplus^5\dataset'\ra\notag\\
+%       X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
+%       X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}\\
+%       X\uplus^3 Y&=&\begin{array}[t]{@{}l}\{\declare{(n,0)}{(A,0)}{(B,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\declare{(n,1)}{(A,1)}{(B,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}\\
+%       X\uplus^5 Y&=&\begin{array}[t]{@{}l}\{\triple{(a,0)}{\declare{(n,0)}{(A,0)}{(B,0)}}{(b,0)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in X\}\ \cup\\ \{\triple{(a,1)}{\declare{(n,1)}{(A,1)}{(B,1)}}{(b,1)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in Y\}\end{array}
+% \end{eqnarray}
+% \end{definition}
 Note that if $\dataset$ and $\dataset'$ are datasets, then so is $\dataset\sqcup\dataset'$.
@@
 
+\section{Validation}
+   To validate the theory, we have to prove that:
+\begin{itemize}
+   % -> het migratie-systeem is typefout-vrij (opmerking: als we de type-checker buiten beschouwing willen laten, kunnen we dit niet beschrijven)
+   \item the migration system $\migrsys$ is free of type errors.
+   % -> het migratie-systeem is vrij van overtredingen op regels die het migratie-systeem moet bewaken
+   \item the migration system $\migrsys$ is an information system (definition~\ref{def:information system}),
+         which implies that its population yields no violations.
+   % -> het migratie-systeem bevat alle oude data, ihb nog steeds na het toepassen van de enforce regels
+   \item $\migrsys$ contains the old data, insofar its enforce rules have not deleted it.
+   % -> er is een pad naar ingebruikname van het nieuwe systeem (vanaf de initiële toestand van het migratie-systeem)
+   \item there is a path to releasing the new system, starting with the initial state of the $\migrsys$.
+   % -> corollary: op het moment van ingebruikname van het nieuwe systeem, is het migratie-systeem vrij van overtredingen op regels die het nieuwe systeem moet bewaken
+   \item As a result, the new system $\infsys'$ is void of violations at the start of its production life.
+   % -> optioneel: na een begrensd aantal 'voortgangs-stappen' kan het nieuwe systeem in gebruik genomen worden
+   \item Optionally, the new system can be used in production after a limited number of steps.
+   % -> optioneel: er bestaat een migratie-systeem dat de oude functionaliteit behoudt (mogelijk uitbreidt) totdat het nieuwe systeem in gebruik genomen is?
+   \item Optionally, there is a migration systen that preserves the functionality of the old system until the new system has been put in production.
+\end{itemize}
+% Bewijsverplichtingen:
+
+\section{Conclusions}
+\begin{itemize}
+   \item the part of the data that does not change can be migrated automatically;
+   \item the part of the migration that can be automated is usually not sufficient;
+         it takes additional human creativity to complete the migration specification;
+\end{itemize}
 \section{Bibliography}

### 2022-11-20 2022 Migration/articleMigration.tex
verworvenheden van ons gesprek van vandaag.

@@
    and $\inst$ is the instance relation.
-   Where ambiguity lures, we write $\triples_{\dataset}$ and $\inst_{\dataset}$ to indicate that we specifically mean the dataset $\dataset$.
+   We write $\triples_{\dataset}$ and $\inst_{\dataset}$ to refer to $\triples$ and $\inst$ respectively.
    Every dataset is an element of an infinite set called $\Dataset$.
@@
    Violations are used to produce meaningful error messages and to trigger changes in the dataset to satisfy rule $u$ again.
+
+   A subset of the rules are so called {\em enforce rules}.
+   A developer makes an enforce rule by using some special syntax.
+   Whenever the dataset violates an enforce rule,
+   the system will make other changes to the dataset to satisfy that rule.
+   Hence, an enforce rule $u$ can be violated during a small amount of time in which the system restores its invariance.
 
@@
          which gives the user the full functionality of the old system and the new system,
-         i.e.\ $\schema_{\migrsys}=\schema\sqcup\schema'$.
-   \item A migration engineer changes $\migrsys$ to accommodate the non-standard requirements of the migration,
-         yielding migration system .
+         i.e.\ $\schema_{\migrsys}=(\schema\sqcup\schema')\cup\id{migrationrules}$.
+         The schema $\id{migrationrules}$ represents the enforce rules that migrate the existing data to the new dataset.
+         TODO: $\id{migrationrules}$ bevat alleen tijdelijke datastructuur, indien nodig.
+   \item A migration engineer changes $\id{migrationrules}$ to accommodate the non-standard requirements of the migration,
+         yielding migration system $\migrsys'$.
+         TODO: de migration engineer kan alleen maar in $\dataset'$ dingen (laten) wijzigen, zodat je altijd terug kunt naar het oude systeem.
    \item The production environment switches instantly from $\infsys$ to $\migrsys'$,
@@
          This implies that the generated rules plus the changes made by the migration engineer ensure the preservation of the meaning of the old data.
-   \item The system removes the data, the schema, and the functionality of $\migrsys'$ that are not supported by $\schema'$.
-         i.e\ $\schema_{\migrsys''}=\schema'$ and $\dataset_{\migrsys''}=\select{\schema'}{\dataset_{\migrsys''}}$.
+   \item Users may have to perform manual migration steps.
+         These are specified as process rules in $\id{migrationrules}$.
+         % Dit zijn "kurk" regels.
+         The system ensures that the amount of work needed to perform these steps is finite.
+   \item The patron of the system can decide whether to carry on with the migration or to roll back to the old system.
+         She can do that incrementally by switching off the interfaces one-by-one.
+   \item The system retains the data, the schema, and the functionality that is covered by $\schema'$, dropping everything else.
 \end{enumerate}

### 2022-11-20 2022 Migration/articleMigration.tex
nabrander

@@
    Information systems are typically used by distributed actors (both users and computers) who continually work with data.
-   A system $\infsys$ contains a dataset $\dataset$, which represents the state of the system.%
-\footnote{Where ambiguity lures, we write $\dataset_{\infsys}$ instead of $\dataset$, to indicate that we specifically mean the dataset of $\infsys$.}
+   A system $\infsys$ contains a dataset $\dataset$, which represents the state of the system.
+   We write $\dataset_{\infsys}$ to refer to dataset $\dataset$.
    Every event that an actor causes in system $\infsys$ may change the state $\dataset$ of that system.
@@
 
-   The schema $\schema$ of a system $\infsys$ is basically a set of semantic constraints on its dataset $\dataset$.%
-\footnote{Where ambiguity lures, we write $\schema_{\infsys}$ instead of $\schema$ to indicate that we specifically mean the schema of $\infsys$.}
+   The schema $\schema$ of a system $\infsys$ is basically a set of semantic constraints on its dataset $\dataset$.
+   We write $\schema_{\infsys}$ to refer to schema $\schema$.
    The Ampersand compiler generates systems, which keep all constraints satisfied at all times.
@@
    and $\rules$ is a finite set of rules.
-   Where ambiguity lures, we write $\concepts_{\schema}$, $rels_{\schema}$, and $\rules_{\schema}$ to indicate that we specifically mean the schema $\schema$.
+   We write $\concepts_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$ to refer to $\concepts$, $\rels$, and $\rules$ respectively.
 %  Every schema is an element of an infinite set called $\Schema$.

### 2022-11-22 2022 Migration/articleMigration.tex
cleaning up the analysis and the definitions

@@
 
-\title{A Theory for Data Migration of Generated Evolving Information Systems}
+\title{It's all about Meaning!\\{\normalsize A Theory for Data Migration during Software Development}}
 \author[ou,ordina]{Stef Joosten\fnref{fn1}}
@@
    A mistake not made cannot propagate into production.
-   For this purpose, we use an existing compiler, Ampersand~\cite{Joosten-JLAMP2018},
-   that generates a system from an Ampersand script.
-   An Ampersand script contains just enough information to generate a complete system,
-   which means that a classical database schema (i.e. data structure plus semantics) can be extracted from the Ampersand script.
-   For the purpose of this paper we can equate an Ampersand script with the schema of the generated system.
 
@@
    change failure rate~\cite{DevOps2021}.}.
-   By installing upgrades automatically, frequently and in small increments,
-   users experience an evolving system.
+   By installing upgrades automatically, frequently, in small increments, and without down-time,
+   the users experience a system that evolves organically without them even noticing the updates.
 
@@
    Data migration literature sometimes uses a different notion of increment.
-   For example, Walkingshaw et.al.~\cite{Walkingshaw2014} and Ataei~\cite{Ataei2021} define an increment as a variation between two data structures.
-   Ataei et.al. show how to unify databases with slight variations by preserving all variations in one comprehensive database.
+   For example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014} define an increment as a variation between two data structures.
+   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
    In our work, however, we want to get rid of the old system and not retain it.
    So, we need to look at a different approach.
-   Examples like these have convinced us to define a new method for data migration that is specifically meant for systems under incremental development.
+   Examples like these have convinced us to define a new method for data migration that is specifically meant for systems in production that are
+   continually developed further in small incrementals.
 
@@
 \begin{itemize}
-   \item the data migration is solely meant to deploy a software increment in production;
+   \item the data migration is meant to deploy a software increment in production;
    \item the old data set may be polluted, but it satisfies its schema;
@@
    \item there is a compiler to generate an information system from a given schema.
-         In this work, we use the Ampersand compiler for that purpose.
 \end{itemize}
@@
    Information systems are typically used by distributed actors (both users and computers) who continually work with data.
-   A system $\infsys$ contains a dataset $\dataset$, which represents the state of the system.
-   We write $\dataset_{\infsys}$ to refer to dataset $\dataset$.
-   Every event that an actor causes in system $\infsys$ may change the state $\dataset$ of that system.
+   Every system contains a dataset, which represents the state of the system.
+   Events that the system detects may cause the state to change.
    To keep our theory technology independent, we will assume that a dataset contains triples.
-   Data migration occurs when an old system, $\infsys$, is replaced by a new system, $\infsys'$,
-   while preserving the meaning of the data from $\infsys$ as much as possible.
-   Just copying the set of data from $\infsys$ to $\infsys'$ is obviously wrong if the schemas of both systems differ.
+   This makes our theory valid for any kind of database that triples can represent,
+   such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
 
-   The schema $\schema$ of a system $\infsys$ is basically a set of semantic constraints on its dataset $\dataset$.
-   We write $\schema_{\infsys}$ to refer to schema $\schema$.
-   The Ampersand compiler generates systems, which keep all constraints satisfied at all times.
-   This is why we have done this research with Ampersand.
-   Much like~\cite{Spivak2012}, in our work the schema encodes the semantics of a system.
-   This will allow us to be explicit about ``preserving the meaning as much as possible''.
-   In contrast with~\cite{Spivak2012}, our approach does not work with functions (functors)
-   but with relations in a way that bears resemblance with Alloy~\cite{Alloy2006} and allegories%
+   Data migration occurs when a new system replaces an old one,
+   while preserving the meaning of the data from the old system as much as possible.
+   Just copying the set of data from the old system to the new is obviously wrong if the schemas of both systems differ.
+
+   We will assume that the schema of a system contains constraints on its dataset.
+   These constraints represent the semantics of the system~\cite{Spivak2012}.
+   This allows us to be explicit about ``preserving the meaning as much as possible''.
+   In contrast with~\cite{Spivak2012}, however, our approach does not work with functions (functors)
+   but with relations similar to Alloy~\cite{Alloy2006} and allegories%
 \footnote{Allegories are a specific type of categories, which the reader does not need to understand for the purpose of this paper}~\cite{Zielinski2013}.
 
-   Extra complications arise from accumulated data pollution in the old system.
+   In this work we use the Ampersand compiler~\cite{Joosten-JLAMP2018},
+   to generate information systems.
+   A script in the Ampersand language represents the system's semantics by constraints only.
+   The compiler builds systems that keep these constraints satisfied at all times.
+   We don't have to formalize the semantics from specifications or from source code
+   because the constraints are already explicitly available in the Ampersand script.
+   This makes Ampersand an obvious platform for validating our theory,
+   An Ampersand script contains just enough information to generate a complete system,
+   which means that a classical database schema (i.e. data structure plus semantics) can be extracted from the Ampersand script.
+   For the purpose of this paper we can equate an Ampersand script with the schema of the generated system.
+
+   For data migration, complications arise from accumulated data pollution in the old system.
    We distinguish data pollution in the following categories:
@@
    The second category is about violations of constraints from the schema.
-   Such violations do not occur in our theory because the old system $\infsys$ was generated by Ampersand,
-   $\infsys$ keeps all constraints from the schema satisfied at all times.
+   Such violations do not occur in our theory because the old system was generated by Ampersand,
+   so that system will keep all constraints from the schema satisfied at all times.
    This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
    The third category, data pollution that is not captured by any of the constraints mentioned in the schema,
-   can be dealt with by adding constraints in the new system $\infsys'$.
-   In that case, a migration engineer must ensure that $\infsys'$ can start with data that satisfies all constraints in its schema.
+   can be dealt with by adding constraints in the new system.
+   In that case, a migration engineer must ensure that the new system can start with data that satisfies all constraints in its schema.
    So some human intervention may be necessary.
-   The last category of data pollution, erroneous constraints, must be fixed by replacing them by the correct constraints in $\infsys'$.
+   The last category of data pollution, erroneous constraints, must be fixed by replacing them by the correct constraints in the new system.
 
@@
    Some effort of a migration engineer may be required.
-   If the semantics encoded in the new schema $\schema'$ differs from the old schema $\schema$,
+   If the semantics encoded in the new schema differs from the old schema,
    a migration engineer must make choices on how to interpret existing data in the new schema.
    She may even want to adapt existing data of which the constraints in the schema have not changed.
-   This requires a third schema, $\schema_{\migrsys}$, which specifies the migration itself.
+   This requires a third schema, which specifies the migration itself.
    It preserves all data from the old system, which may be needed for the migration.
    It also contains the new schema and possibly some new data, which is specified by the migration engineer.
-   Hence, the migration is based on the disjoint union of $\schema$ and $\schema'$.
-   We generate $\schema_{\migrsys}$ in the form of source code, so the migration engineer can change everything she wants to suit the particulars of the data migration.
+   Hence, the migration is based on the disjoint union of the old and the new schemas.
+   We generate the schema of the migration system in the form of source code,
+   so the migration engineer can change everything she wants to suit the particulars of the data migration.
 
@@
    A dataset $\dataset$ describes a set of structured data, which is typically stored persistently in a database of some kind.
-   The purpose of a dataset is to describe the data of a system at one point in time, so that data changes are representable. 
+   We write $\dataset_{\infsys}$ to refer to the dataset of a particular information system $\infsys$.
+   The purpose of a dataset is to describe the data of a system at one point in time. 
    Before defining datasets, we must first define the constituent notions of atom, concept, relation, and triple.
@@
    All concepts are taken from an infinite set $\Concepts$.
-   $\Concepts$ and $\Atoms$ are disjoint sets.
-   For example, an engineer might choose to classify \atom{Peter} and \atom{Melissa} as \concept{Person},
+   $\Concepts$ and $\Atoms$ are disjoint.
+   For example, a developer might choose to classify \atom{Peter} and \atom{Melissa} as \concept{Person},
    and \atom{074238991} as a \concept{TelephoneNumber}.
@@
    We will use variables $A$, $B$, $C$, $D$ to represent concepts.
+
+   The relation $\inst:\Pair{\Atoms}{\Concepts}$ is the instance relation between atoms and concepts.
    The expression $a\ \inst\ C$ means that atom $a$ is an \emph{instance} of concept $C$.
-   The relation $\inst:\Pair{\Atoms}{\Concepts}$ is the instance relation between atoms and concepts.
    This relation is used in the type system.
+   In the type system, every atom must have a type,
+   which makes $\inst$ a total relation:
+\begin{eqnarray}
+   \forall a\in\atoms\ \exists C\in\Concepts: a\ \inst\ C
+\label{eqn:concepts}
+\end{eqnarray}
 
-   Relations serve to organize and store data, to allow an engineer to represent facts.
+   Relations serve to organize and store data, to allow a developer to represent facts.
    In this paper, variables $r$, $s$, and $d$ represent relations.
@@
    $\Rels$ is disjoint from $\Concepts$ and $\Atoms$.
+   Every relation $r$ has a name, a source concept, and a target concept.
+   We write $r=\declare{n}{A}{B}$ to denote that relation $r$ has name $n$, source concept $A$, and target concept $B$.
 
@@
    A triple is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.
-   A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean in practice that
-   the ``thing'' that \atom{Peter} refers to has \atom{074238991} as a telephone number.
+   For example, $\triple{\text{\atom{Peter}}}{\declare{\id{phone}}{\tt Person}{\tt TelephoneNumber}}{\text{\atom{074238991}}}$ is a triple.
+   Equation~\ref{eqn:type} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
+   In practice, users can say that Peter has telephone number 074238991.
+   So, the ``thing'' that \atom{Peter} refers to (which is Peter) has \atom{074238991} as a telephone number.
    This ``meaning from practice'' has no consequences in the formal world.
@@
    In this paper, the informal meaning is interesting only for the sake of illustration.
-   This makes our theory valid for any kind of database that triples can represent,
-   such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
 
-   A dataset $\dataset$ is a tuple $\pair{\triples}{\inst}$,
-   in which $\triples$ is a finite set of triples
-   and $\inst$ is the instance relation.
-   We write $\triples_{\dataset}$ and $\inst_{\dataset}$ to refer to $\triples$ and $\inst$ respectively.
+   A dataset $\dataset$ is a tuple $\pair{\triples}{\inst}$ that satisfies:
+\begin{eqnarray}
+   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B
+   \label{eqn:type}
+\end{eqnarray}
+   $\triples$ is a finite set of triples and $\inst$ is the instance relation.
+   In case we need to be explicit about the dataset,
+   we will write $\triples_{\dataset}$ and $\inst_{\dataset}$ to refer to $\triples$ and $\inst$ respectively.
    Every dataset is an element of an infinite set called $\Dataset$.
    To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\triples$.
+   Every atom has the type of the relation in which it resides:
 
    A relation $r$ can serve as a container of pairs,
-   as defined by the function $\id{pop}_r:\Dataset\rightarrow\Pair{\Atoms}{\Atoms}$.
+   as defined by the function $\id{pop}_r:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$.
    It defines a set of pairs, which we call the population of $r$:
@@
 \end{equation}
-
-   Over time, an information system ``observes'' events that change the dataset by inserting and deleting triples.
-   By the same account, the population of relations changes over time.
-   These events may be caused by external systems, roles acting on the dataset, or any other force that changes triples in a dataset.
+   Equation~\ref{eqn:type} implies that for every dataset $\dataset$:
+\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ \inst\ A\ \wedge\ b\ \inst\ B\]
+   For a developer, this means that the type of an atom depends only on the relation in which it resides; not on the actual population of the database.
+   This allows for static typing, which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.
 
@@
 \label{sct:Schemas}
-   To design an information system, an engineer defines a schema in which she defines concepts, relations, and rules.
+   To design an information system, a developer defines a schema in which she defines concepts, relations, and rules.
    The purpose of a schema is to capture the semantics of an information system.
@@
    and $\rules$ is a finite set of rules.
+   We write $\schema_{\infsys}$ to refer to the schema of information system $\infsys$.
    We write $\concepts_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$ to refer to $\concepts$, $\rels$, and $\rules$ respectively.
@@
 
-   In order to attach types to atoms, relations must be declared in the schema.
-   Every relation $r$ has not only a name, but also a source concept and a target concept.
-   We write $r=\declare{nm}{A}{B}$ to denote that relation $r$ has name \id{nm}, source concept $A$, and target concept $B$.
-   We require that the source and target concepts are in the set of concepts:
+   In order to do static type checking,
+   all concepts and relations must be ``known'' in the schema:
 \begin{eqnarray}
-   \forall\declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts
-   \label{eqn:relationsIntroduceConcepts}
+   \declare{n}{A}{B}\in\rels&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
+   \label{eqn:relationsIntroduceConcepts}\\
+   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels\ \wedge\ a\ \inst\ A\wedge b\ \inst\ B
+   \label{eqn:define R}
 \end{eqnarray}
+
    A rule in a schema serves to constrain the dataset,
    to ensure the semantic integrity of the dataset.
-   For every rule $u$ in the schema, the Ampersand compiler generates code
-   for the function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$.
-   For any given dataset $\dataset$, $\viol{u}{\dataset}$ represents the set of violations of rule $u$.
-   The type of the atoms in a violation set is defined by the function $\Sign{u} : \Dataset\rightarrow\Pair{\Concepts}{\Concepts}$:
+   For every rule $u$ in the schema, we assume there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
+   that represents the set of violations of rule $u$ in any dataset.
+   We also assume a function $\Sign{u} : \Dataset\rightarrow\Pair{\Concepts}{\Concepts}$
+   that represents the signature of a rule, i.e. the type of atoms in violations:
 \begin{eqnarray}
@@
    To determine whether rule $u$ is satisfied in dataset $\dataset$,
-   the Ampersand compiler also generates code for a predicate $\sat{u}{\Dataset}$:
+   we define $\sat{u}{\Dataset}$:
 \begin{eqnarray}
@@
 \end{eqnarray}
-   Violations are used to produce meaningful error messages and to trigger changes in the dataset to satisfy rule $u$ again.
+   Violations are used to produce meaningful error messages and to trigger changes in the dataset to restore satisfaction of that rule.
 
-   A subset of the rules are so called {\em enforce rules}.
-   A developer makes an enforce rule by using some special syntax.
-   Whenever the dataset violates an enforce rule,
-   the system will make other changes to the dataset to satisfy that rule.
-   Hence, an enforce rule $u$ can be violated during a small amount of time in which the system restores its invariance.
+   We assume a special kind of rule that is used for specialization.
+   The rule $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
+\begin{equation}
+   \label{eqn:specialization}
+   \sat{A\isa B}{\dataset}\ \Leftrightarrow\ \forall a: a\ \inst\ A\rightarrow a\ \inst\ B
+\end{equation}
+   We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
+   Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.
+   All specialization rules in a schema together form a partial order $\isa : \Pair{\concepts}{\concepts}$.
+   Specialization rules are special for two reasons:
+   Firstly, the system keeps $A\isa B$ satisfied automatically by changing $\inst$ whenever needed.
+   This means that a system that contains rule $A\isa B$ can never be brought into a state in which $\id{sat}_{A\isa B}$ is False.
+   This allows for certain run-time optimizations, which are beyond the scope of this paper.
+   A second property that sets specialization rules apart, is that they play a role in the Ampersand compiler.
+   For example, code that determines equality between atoms only works for atoms that are an instance of the same concept.
+   To ensure that this does not result in problems at run-time, users are required to write all specialization rules explicitly.
+
+   %These rules reflect the meaning (semantics) that is shared among users.
 
@@
    \item $\roles$ is a set of roles;
-   \item $\maintain : \roles\times\rules$ is the maintainance relation between roles and rules from $\schema$; and
-   \item requirements~\ref{eqn:pop} thru~\ref{eqn:specialization} are satisfied.
+   \item $\maintain : \roles\times\rules$ is the maintainance relation between roles and rules from $\schema$.
 \end{itemize}
 \end{definition}
-   Let us discuss these items one by one.
-
-   The language Ampersand is typed, to prevent a substantial amount of programming mistakes at compile-time.
-   Type checking is based on the schema only.
-   So, all concepts in the relation $\id{inst}$ and all relations in triples from the dataset must be ``known'' in the schema:
-\begin{eqnarray}
-   a\ \inst\ C&\Rightarrow&C\in\concepts
-   \label{eqn:concepts}\\
-   \triple{a}{r}{b}\in\triples&\Rightarrow&r\in\rels
-   \label{eqn:define R}
-\end{eqnarray}
-   Every atom is an instance of a concept:
-\begin{eqnarray}
-   a\ \inst\ A&\Leftrightarrow&(\exists n,b,B:\ \triple{a}{\declare{n}{A}{B}}{b}\in\triples\ \vee\ \triple{b}{\declare{n}{B}{A}}{a}\in\triples)
-   \label{eqn:instance}
-\end{eqnarray}
-   Every atom has the type of the relation in which it resides:
-\begin{eqnarray}
-   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B
-   \label{eqn:type}
-\end{eqnarray}
-
-   Ampersand supports a special kind of rule that is used for specialization.
-   The rule $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
-\begin{equation}
-   \label{eqn:specialization}
-   \sat{A\isa B}{\dataset}\ \Leftrightarrow\ \forall a: a\ \inst\ A\rightarrow a\ \inst\ B
-\end{equation}
-   We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
-   Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.
-   All specialization rules in a schema together form a partial order $\isa : \Pair{\concepts}{\concepts}$.
-   Specialization rules are special for two reasons:
-   Firstly, the system keeps $A\isa B$ satisfied automatically by changing $\inst$ whenever needed.
-   This means that a system that contains rule $A\isa B$ can never be brought into a state in which $\id{sat}_{A\isa B}$ is False.
-   This allows for certain run-time optimizations, which are beyond the scope of this paper.
-   A second property that sets specialization rules apart, is that they play a role in the Ampersand compiler.
-   For example, code that determines equality between atoms only works for atoms that are an instance of the same concept.
-   To ensure that this does not result in problems at run-time, users are required to write all specialization rules explicitly.
-
-   %These rules reflect the meaning (semantics) that is shared among users.
-
    A \define{role} is a name that identifies a group of users.
    It serves as a placeholder for a person or a machine (i.e. an actor) who can change the dataset.
-   The purpose of a role is to mention an individual user or an automated actor without knowing who that user is.
+   The purpose of a role is to mention an individual user (human) or an automated actor (bot) without knowing who that user is.
    In the sequel, when we talk about a role we actually mean an actor who fulfills that role in the system.
 
-   Every rule is assigned to at least one role by means of the relation $\maintain$.
-   $o\ \maintain\ u$ means that role $o$ must keep rule $u$ satisfied, i.e.\ to keep $\sat{u}{\dataset}$ true.
-   After some event has violated a rule $u$ (by making changes in the dataset),
-   role $o$ maintains this rule by making other changes to satisfy rule $u$ again.
-   We call this ``to restore the invariance'' of rule $u$.
-   In other words: If anyone edits in the dataset and violates a rule, someone or something has to fix the dataset to satisfy that rule again.
+   Every rule in a schema is meant to be kept satisfied invariably.
+   The only way to break this invariance is to remove the rule from the schema.
+   So the "normal" state of a system is that its dataset $\dataset$ contains no violations of any rule in the schema:
+   $\forall u\in\rules:\ \sat{u}{\dataset}$.
+   However, the system continually registers events that insert or delete triples in its dataset,
+   which may cause rules to be satisfied no longer.
+   This is a temporary breach of the invariance of the affected rules,
+   which requires a reaction to restore invariance.
    Example: if a rule says that every request must be acknowledged, the receipt of a request will violate this rule.
    Sending a message that acknowledges this receipt will cause the rule to be satisfied again.
+
+   In this paper we assume two ways of enforcing satisfaction of rules: automatic or manual.
+   Manual enforcement is specified by the developer by assigning a rule $u$ to a role $o$.
+   Any actor with role $o$, whether person or machine,
+   can restore invariance of rule $u$ by inserting or deleting the right triples to compensate for the violations.
+   For this reason, we assume that every rule is assigned to at least one role by means of the relation $\maintain$.
+\begin{equation}
+   \forall u\in\rules\ \exists o\in\roles:\ o\ \maintain\ u\label{eqn:maintain}
+\end{equation}
+   $o\ \maintain\ u$ means that role $o$ must keep rule $u$ satisfied, i.e.\ to keep $\sat{u}{\dataset}$ true.
+   Not all ways of restoring invariance can be anticipated by an algorithm, which makes manual enforcement a necessary feature.
+   As long as rule $u$ is not satisfied, its violations won't go away, stimulating the actor to do something about it.
+
+   Automatic enforcement is specified by the developer with a special syntax,
+   which describes not only the rule, but also the way to restore invariance.
+   Whenever the dataset violates an enforce rule,
+   the system will make other changes to the dataset to satisfy that rule.
+   Hence, the violation of the rule exists during a small, finite amount of time in which the system restores its invariance.
 

### 2022-11-27 2022 Migration/articleMigration.tex
wip

@@
 \label{sct:Introduction}
+   To automate the datamigrations that come with incrementally developing and deploying information systems.
+   That is the purpose of the theory proposed in this paper.
+
    We believe in generating information systems%
@@
    to prevent human programming errors.
-   By generating rather than programming systems by hand,
-   we avoid introducing errors in the translation from design to a concrete system.
-   A mistake not made cannot propagate into production.
-
+   For, a mistake not made can never propagate into production.
    We also believe that information systems should be released into production in small increments,
@@
    By installing upgrades automatically, frequently, in small increments, and without down-time,
-   the users experience a system that evolves organically without them even noticing the updates.
+   the users experience a system that evolves organically, without even noticing the updates.
+   Thus, we believe our theory will help to automate data migrations correctly and reliably.
+   The benefits work in two directions.
+   Automation data migrations saves effort and errors,
+   contributing to more frequent and smaller releases.
+   Smaller and faster releases also mean smaller migrations
+   and therefore a smaller difference between the old system and the new system.
+   The smaller that difference, the simpler the migration and the more of it can be automated.
 
@@
    Examples like these have convinced us to define a new method for data migration that is specifically meant for systems in production that are
-   continually developed further in small incrementals.
+   continually evolving in small increments.
 
@@
    In practice, more frequent updates require that updates come with zero down-time.
-   For our theory, zero down-time means that the release of new functionality may not be delayed by anything a migration engineer must do.
-
-   This theory has enhanced our formal understanding of data migration and it will help to automate data migrations correctly and reliably.
-   The benefits work in two directions.
-   Automation data migrations saves effort and errors,
-   contributing to more frequent and smaller releases.
-   Smaller and faster releases also mean smaller migrations
-   and therefore a smaller difference between the old system and the new system.
-   The smaller that difference, the simpler the migration and the more of it can be automated.
+   For our theory, zero down-time means that all human activity for the data migration can be done without taking the system out of production.
+   We consider any delay caused by automatically switching from the old system to the new one negligible,
+   assuming a modern deployment platform that does this with zero down-time.
 
@@
    We will assume that the schema of a system contains constraints on its dataset.
-   These constraints represent the semantics of the system~\cite{Spivak2012}.
+   These constraints represent the semantics of the system, similar to the theory developed by Spivak~\cite{Spivak2012}.
    This allows us to be explicit about ``preserving the meaning as much as possible''.
-   In contrast with~\cite{Spivak2012}, however, our approach does not work with functions (functors)
+   In contrast with Spivak, however, our approach does not work with functions (functors)
    but with relations similar to Alloy~\cite{Alloy2006} and allegories%
@@
    A script in the Ampersand language represents the system's semantics by constraints only.
-   The compiler builds systems that keep these constraints satisfied at all times.
    We don't have to formalize the semantics from specifications or from source code
    because the constraints are already explicitly available in the Ampersand script.
-   This makes Ampersand an obvious platform for validating our theory,
+   A system generated by the Ampersand compiler will keep these constraints satisfied at all times.
+   This makes Ampersand an obvious platform for validating our theory.
    An Ampersand script contains just enough information to generate a complete system,
@@
    so the system can help the user and simplify (or eliminate) the corresponding working procedure.
-   Such constraints move this type of pollution to one of the other categories.
+   So, such constraints may help to move this type of pollution to one of the other categories.
    The second category is about violations of constraints from the schema.
@@
 
+\begin{figure}[bht]
+   \begin{center}
+     \includegraphics[scale=.35]{Migration.png}
+   \end{center}
+\caption{Data migration}
+\label{fig:event flow}
+\end{figure}
    This illustrates that automating data migrations has its limits.
-   Some effort of a migration engineer may be required.
+   Some effort of a migration engineer can never be excluded.
    If the semantics encoded in the new schema differs from the old schema,
-   a migration engineer must make choices on how to interpret existing data in the new schema.
-   She may even want to adapt existing data of which the constraints in the schema have not changed.
-   This requires a third schema, which specifies the migration itself.
-   It preserves all data from the old system, which may be needed for the migration.
-   It also contains the new schema and possibly some new data, which is specified by the migration engineer.
-   Hence, the migration is based on the disjoint union of the old and the new schemas.
-   We generate the schema of the migration system in the form of source code,
+   a migration engineer must make choices.
+   If the interpretation of data changes,
+   a migration engineer must remove obsolete constraints from the old system and introduce new constraints in the new system.
+   The consequences of such changes may require changes in data as well.
+   The migration engineer may even want to adapt existing data of which the constraints in the schema have not changed.
+   All of this requires a third schema, which specifies the migration itself.
+   This migration schema preserves data from the old system, it may introduce new data automatically (generated from the old dataset),
+   and it may require users to introduce new data (which will take time). 
+   It must contain the old schema, but it also contains the new schema and possibly some new data, which is specified by the migration engineer.
+   As the migration schema must distinguish between the old dataset and the new dataset,
+   the migration must be based on the disjoint union of the two.
+   So the idea of a migration schema is to compute the disjoint union of the old and the new schemas
+   and to present it to the migration engineer in the form of source code,
    so the migration engineer can change everything she wants to suit the particulars of the data migration.
 
-   A naive migration might proceed as follows:
-   Launch the new system in parallel to the old one, copy data from the old to the new system, and have everyone use the new system.
-   However, numerous issues might impede that plan.
-   The following examples illustrate the practical issues that may occur:
-\begin{enumerate}
-\item Data required in the new system is missing in the old system.
-   There may be no way in the old system to enter that data.
-   An example could be that every reimbursement form needs to have an address associated to it to mail the check to, but address information is not stored in the old system:
-   The old system required the reimbursement office to look up employee's addresses from a hand-written list they had on their desk.
+%    A naive migration might proceed as follows:
+%    Launch the new system in parallel to the old one, copy data from the old to the new system, and have everyone use the new system.
+%    However, numerous issues might impede that plan.
+%    The following examples illustrate the practical issues that may occur:
+% \begin{enumerate}
+% \item Data required in the new system is missing in the old system.
+%    There may be no way in the old system to enter that data.
+%    An example could be that every reimbursement form needs to have an address associated to it to mail the check to, but address information is not stored in the old system:
+%    The old system required the reimbursement office to look up employee's addresses from a hand-written list they had on their desk.
 
-   This issue will require somebody to insert the missing addresses in the new system.
-\item Data in the old system is wrong but cannot be corrected there due to how the old system was designed.
-   An example is if the old system only allows approvals to be entered as the current user, and the CEO has always insisted that her administrative staff enters the approvals into the system for her.
-   This may result in approvals being entered as admin staff, where it was really the CEO making the approval.
+%    This issue will require somebody to insert the missing addresses in the new system.
+% \item Data in the old system is wrong but cannot be corrected there due to how the old system was designed.
+%    An example is if the old system only allows approvals to be entered as the current user, and the CEO has always insisted that her administrative staff enters the approvals into the system for her.
+%    This may result in approvals being entered as admin staff, where it was really the CEO making the approval.
 
-   This issue will require that the incorrect registration of staff members is corrected in the new system.
-\item Data in the old system does not satisfy invariants of the new system.
-   There may be no way in the old system of making the data satisfy those invariants.
-   We can use the same example as in the previous bullet point,
-   but add the requirement (in the new system) that every purchase above a certain amount needs to be approved by the CEO.
+%    This issue will require that the incorrect registration of staff members is corrected in the new system.
+% \item Data in the old system does not satisfy invariants of the new system.
+%    There may be no way in the old system of making the data satisfy those invariants.
+%    We can use the same example as in the previous bullet point,
+%    but add the requirement (in the new system) that every purchase above a certain amount needs to be approved by the CEO.
 
-   This issue will require somebody to change the data to satisfy the broken invariants in the new system.
-\item A way of entering data into the old system is missing in the new system.
-   People or automated processes might rely on these ways of entering data.
-   An example could be that when employees turned their computers on or off,
-   an ad-hoc script would automatically check them in- and out to determine the number of hours they worked.
-   A handful of employees still relies on this.
+%    This issue will require somebody to change the data to satisfy the broken invariants in the new system.
+% \item A way of entering data into the old system is missing in the new system.
+%    People or automated processes might rely on these ways of entering data.
+%    An example could be that when employees turned their computers on or off,
+%    an ad-hoc script would automatically check them in- and out to determine the number of hours they worked.
+%    A handful of employees still relies on this.
 
-   This issue will require that users of obsolete functionality are informed and given a way to cope with the missing functionality where appropriate.
-\item Data present in the old system cannot be stored in the new system.
-   An example could be that references to physical locations where original receipts are kept are stored in the old system,
-   but the new system relies on scans of receipts and allows the originals to be destroyed or not submitted.
+%    This issue will require that users of obsolete functionality are informed and given a way to cope with the missing functionality where appropriate.
+% \item Data present in the old system cannot be stored in the new system.
+%    An example could be that references to physical locations where original receipts are kept are stored in the old system,
+%    but the new system relies on scans of receipts and allows the originals to be destroyed or not submitted.
 
-   This issue will require that the old data is kept until the original receipts have been scanned.
-\end{enumerate}
+%    This issue will require that the old data is kept until the original receipts have been scanned.
+% \end{enumerate}
 
-(@Bas, het volgende zit al in de oplossingen sfeer. Willen we dat hier al doen?)
-   To mitigate these issues, we:
+% (@Bas, het volgende zit al in de oplossingen sfeer. Willen we dat hier al doen?)
+%    To mitigate these issues, we:
 
-   \begin{enumerate}
-   \item Allow `missing' triples requirements to be ignored during migration.
-   \item Allow triples to be migrated while being marked as needing correction.
-   \item Allow invariants in the new system to be ignored for certain triples during migration.
-   \item Allow continued use of interfaces of the old system, data entered into the old system via those interfaces needs to be continuously copied to the new system.
-   \item Retain data in the old system until it can be marked as ready to be phased out.
-   \end{enumerate}
+%    \begin{enumerate}
+%    \item Allow `missing' triples requirements to be ignored during migration.
+%    \item Allow triples to be migrated while being marked as needing correction.
+%    \item Allow invariants in the new system to be ignored for certain triples during migration.
+%    \item Allow continued use of interfaces of the old system, data entered into the old system via those interfaces needs to be continuously copied to the new system.
+%    \item Retain data in the old system until it can be marked as ready to be phased out.
+%    \end{enumerate}
 
@@
    A triple is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.
+   A dataset $\dataset$ is a tuple $\pair{\triples}{\inst}$ that satisfies:
+\begin{eqnarray}
+   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B
+   \label{eqn:type}
+\end{eqnarray}
    For example, $\triple{\text{\atom{Peter}}}{\declare{\id{phone}}{\tt Person}{\tt TelephoneNumber}}{\text{\atom{074238991}}}$ is a triple.
@@
 
-   A dataset $\dataset$ is a tuple $\pair{\triples}{\inst}$ that satisfies:
-\begin{eqnarray}
-   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B
-   \label{eqn:type}
-\end{eqnarray}
    $\triples$ is a finite set of triples and $\inst$ is the instance relation.
@@
 \subsection{Data migration}
-\begin{figure}[bht]
-   \begin{center}
-     \includegraphics[scale=.45]{Migration.png}
-   \end{center}
-\caption{Data migration}
-\label{fig:event flow}
-\end{figure}
    To migrate system $\infsys$ to $\infsys'$, we take the following steps:

### 2022-12-11 2022 Migration/articleMigration.tex
Repair the first example for the new requirements.

@@
 \label{sct:Introduction}
-   To automate the datamigrations that come with incrementally developing and deploying information systems.
-   That is the purpose of the theory proposed in this paper.
+   To automate the datamigrations that come with incrementally developing and deploying information systems;
+   that is the purpose of the theory proposed in this paper.
 
@@
    contributing to more frequent and smaller releases.
-   Smaller and faster releases also mean smaller migrations
-   and therefore a smaller difference between the old system and the new system.
-   The smaller that difference, the simpler the migration and the more of it can be automated.
+   Smaller and faster releases also mean a smaller difference between the old system and the new system.
+   The smaller that difference, the simpler the migration can be and the more of it can be automated.
 
    In each increment, we distinguish an ``old system'' and a ``new system''.
-   The difference between the old system and the new system is not just a difference in two schemas.
-   When deploying the increment, the data from the old system needs to migrate to the new system as well.
-   Incremental evolution poses specific requirements to data migration,
+   A difference of schemas complicates the migration of data between the old system and the new system.
+   The purpose of incremental evolution poses specific requirements to data migration,
    which sets it apart from data migrations for other purposes.
@@
 \section{Analysis}
+   The purpose of an information system is to make data meaningful to its users.
+   Users have their own tasks and responsibilities
+   and may work from different locations and on different moments.
+   This collective use serves a purpose which we loosely call ``the business''.
+   As a consequence, the data in a system changes continually.
+   To preserve meaning, users are trying to maintain semantic constraints on the data,
+   amidst of all changes that are going on around them.
+
    Information systems are typically used by distributed actors (both users and computers) who continually work with data.
@@
    We will assume that the schema of a system contains constraints on its dataset.
-   These constraints represent the semantics of the system, similar to the theory developed by Spivak~\cite{Spivak2012}.
+   These constraints represent the semantics of the system~\cite{Spivak2012}.
    This allows us to be explicit about ``preserving the meaning as much as possible''.
-   In contrast with Spivak, however, our approach does not work with functions (functors)
-   but with relations similar to Alloy~\cite{Alloy2006} and allegories%
+   Our approach works with relations in a way similar to Alloy~\cite{Alloy2006} and allegories%
 \footnote{Allegories are a specific type of categories, which the reader does not need to understand for the purpose of this paper}~\cite{Zielinski2013}.
@@
    An Ampersand script contains just enough information to generate a complete system,
-   which means that a classical database schema (i.e. data structure plus semantics) can be extracted from the Ampersand script.
+   which means that a classical database schema (i.e.\ data structure plus semantics) can be extracted from the Ampersand script.
    For the purpose of this paper we can equate an Ampersand script with the schema of the generated system.
@@
    In that case, a migration engineer must ensure that the new system can start with data that satisfies all constraints in its schema.
-   So some human intervention may be necessary.
+   So some human intervention may be necessary in specific cases.
    The last category of data pollution, erroneous constraints, must be fixed by replacing them by the correct constraints in the new system.
@@
    This illustrates that automating data migrations has its limits.
-   Some effort of a migration engineer can never be excluded.
+   We cannot eliminate the need for interventions by a migration engineer entirely.
    If the semantics encoded in the new schema differs from the old schema,
@@
    and it may require users to introduce new data (which will take time). 
-   It must contain the old schema, but it also contains the new schema and possibly some new data, which is specified by the migration engineer.
+   It must contain the old schema, but it also contains the new schema and possibly some new data, all of which a migration engineer must specify.
    As the migration schema must distinguish between the old dataset and the new dataset,
@@
    Atoms serve as data elements.
-   They are values without internal structure of interest, meant to represent data elements in a database.
+   Atoms are values without internal structure of interest, meant to represent data elements in a database.
    From a business perspective, atoms represent concrete items of the world,
@@
    The expression $a\ \inst\ C$ means that atom $a$ is an \emph{instance} of concept $C$.
-   This relation is used in the type system.
-   In the type system, every atom must have a type,
-   which makes $\inst$ a total relation:
+   This relation is used in the type system, in which $\inst$ assigns a type to every atom.
+   This makes $\inst$ a total relation:
 \begin{eqnarray}
@@
    We leave it entirely up to a user to attach a practical meaning to a triple.
-   In this paper, the informal meaning is interesting only for the sake of illustration.
 
@@
    To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\triples$.
-   Every atom has the type of the relation in which it resides:
 
@@
    \label{eqn:relationsIntroduceConcepts}\\
-   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels\ \wedge\ a\ \inst\ A\wedge b\ \inst\ B
+   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
    \label{eqn:define R}
@@
 
-   A rule in a schema serves to constrain the dataset,
-   to ensure the semantic integrity of the dataset.
+   A rule in a schema serves to constrain the dataset, to ensure its semantic integrity.
    For every rule $u$ in the schema, we assume there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
@@
    Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.
-   All specialization rules in a schema together form a partial order $\isa : \Pair{\concepts}{\concepts}$.
+   The specialization rules in a schema form a partial order $\isa : \Pair{\concepts}{\concepts}$,
+   meaning that $\isa$ is reflexive, transitive, and antisymmetric.
    Specialization rules are special for two reasons:
@@
 \label{sct:Information Systems}
-   The purpose of an information system is to make data meaningful to its users.
-   Users have their own tasks and responsibilities
-   and may work from different locations and on different moments.
-   This collective use serves a purpose which we loosely call ``the business''.
-   As a consequence, the data in a system changes continually.
-   To preserve meaning, users are trying to maintain semantic constraints on the data,
-   amidst of all changes that are going on around them.
-
-   This section introduces the notion of information system.
+   This section defines the notion of information system.
    As before, we will suffix the elements of this definition if the situation requires it.
@@
    So the "normal" state of a system is that its dataset $\dataset$ contains no violations of any rule in the schema:
-   $\forall u\in\rules:\ \sat{u}{\dataset}$.
+\begin{equation}
+   \forall u\in\rules:\ \sat{u}{\dataset}\label{eqn:satisfaction}
+\end{equation}
    However, the system continually registers events that insert or delete triples in its dataset,
@@
    which requires a reaction to restore invariance.
-   Example: if a rule says that every request must be acknowledged, the receipt of a request will violate this rule.
-   Sending a message that acknowledges this receipt will cause the rule to be satisfied again.
+   Example: if a rule says that every request must be acknowledged, the receipt of a request (which is an event) will violate this rule.
+   Sending a message that acknowledges this receipt (another event) will cause the rule to be satisfied again.
+   During the time between the receipt of the request and sending the acknowledgment, the rule is (temporarily) not satisfied.
 
@@
    $o\ \maintain\ u$ means that role $o$ must keep rule $u$ satisfied, i.e.\ to keep $\sat{u}{\dataset}$ true.
-   Not all ways of restoring invariance can be anticipated by an algorithm, which makes manual enforcement a necessary feature.
    As long as rule $u$ is not satisfied, its violations won't go away, stimulating the actor to do something about it.
@@
    Automatic enforcement is specified by the developer with a special syntax,
-   which describes not only the rule, but also the way to restore invariance.
+   which specifies not only the rule, but also the way to restore invariance.
+   The system contains an engine that restores invariance of all automatically enforced rules without unneccessary delay.
    Whenever the dataset violates an enforce rule,
-   the system will make other changes to the dataset to satisfy that rule.
-   Hence, the violation of the rule exists during a small, finite amount of time in which the system restores its invariance.
+   this engine will make other changes to the dataset to satisfy that rule.
+   Hence, the violation of the rule exists during a small, finite amount of time.
 
-   The fact that every rule has at least one role in the relation $\maintain$, means that all roles together are keeping all rules satisfied.
+   The system contains a dedicated role for the enforcement engine,
+   so every rule has at least one role in the relation $\maintain$, means that all roles together are keeping all rules satisfied.
+   (\verb-@sjcjoosten-, what if $\maintain$ relates both a human role and the enforcement engine role to the same rule?)
 
@@
 \end{verbatim}
-   We also define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that student takes.
+   We also define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that a student takes.
    In Ampersand, which is a syntactically sugared form of relation algebra,
@@
    The actual constraint, \verb#isEnrolledFor |- takes;isPartOf~#, is written in logic as:
-\[\begin{array}{l}\forall s\in\text{\tt Student}, m\in\text{\tt Module}\ \exists c\in\text{\tt Course}:\\
+\begin{equation}
+   \begin{array}{l}\forall s\in\text{\tt Student}, m\in\text{\tt Module}\ \exists c\in\text{\tt Course}:\\
 s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
-\end{array}\]
+\end{array}
+\label{eqn:example isEnrolledFor}
+\end{equation}
 
@@
    Now let us check the requirements to verify that this example defines an information system.
-   Requirement~\ref{eqn:pop} results in a set $\triples$ that contains seven triples.
-   Requirements~\ref{eqn:instance} and~\ref{eqn:type} define the relation $\id{inst}$ to contains the pairs:
+   The Ampersand compiler generates a dataset $\dataset$, which contains a set of triples and a relation $\inst$.
+   It defines the set of triples $\triples$ as:
+   \[\begin{array}[t]{l}
+         \triple{\text{\tt "Peter"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Management"}}\\
+         \triple{\text{\tt "Susan"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
+         \triple{\text{\tt "John"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
+         \triple{\text{\tt "Finance"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Management"}}\\
+         \triple{\text{\tt "Business Rules"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
+         \triple{\text{\tt "Business Analytics"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
+         \triple{\text{\tt "IT-Governance"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Management"}}
+   \end{array}\]
+   The relation $\id{inst}$ contains the pairs:
    \[\begin{array}{l}
@@
    \end{array}\]
-   Requirements~\ref{eqn:concepts} and~\ref{eqn:relationsIntroduceConcepts} both say that 
+   Requirement~\ref{eqn:concepts} is satisfied because every atom has a concept.
+   Dataset $\pair{\triples}{\inst}$ satisfies both requirements~\ref{eqn:type} and~\ref{eqn:pop}.
+   So, the dataset $\dataset=\pair{\triples}{\inst}$ satisfies the requirements from section~\ref{sct:Datasets}
+
+   The Ampersand compiler generates a schema $\schema$, which contains concepts, relations, and rules.
+   It defines the set of concepts to satisfy requirement~\ref{eqn:relationsIntroduceConcepts}:
    \[\concepts=\{ {\tt Module}, {\tt Course}, {\tt Student}\}\]
-   Requirement~\ref{eqn:define R} says that 
+   It defines the set of relations to satisfy requirement~\ref{eqn:define R}:
    \[\begin{array}{rcl}
@@
    \]
-   The set of rules $\rules$ contains just one rule: \verb-EnrollRule-.
-   So $\concepts$, $\relations$, and $\rules$ complete the schema $\schema$ of this example.
-   The dataset $\dataset$ contains the set $\triples$ and the relation $\inst$.
-   The set of roles, $\roles$, contains one role: ${\tt Administrator}$
-   and the relation $\maintain$ relates ${\tt Administrator}$ to rule\verb-EnrollRule-.
-   Requirement~\ref{eqn:typed violations} and~\ref{eqn:sat} are satisfied because $\viol{\tt EnrollRule}{\dataset}$ is empty.
+   And, it defines the set of rules $\rules$ to contain just one rule: \verb-EnrollRule-.
+   $\rules$ satisfies requirement~\ref{eqn:satisfaction} because $\viol{\tt EnrollRule}{\dataset}$ is empty.
+   So, the schema $\schema=\triple{\concepts}{\rels}{\rules}$ satisfies the requirements from section~\ref{sct:Schemas}.
+
+   To fulfill the definition of information system, definition~\ref{def:information system},
+   Ampersand generates a set of roles $\roles=\{{\tt Administrator}\}$ and
+   a relation $\maintain$, which contains one pair only: $\pair{{\tt Administrator}}{{\tt EnrollRule}}$.
+   For the purpose of this paper, the functionality $\functionality$ is irrelevant (onderbouwen!).
+   The constraint~\ref{eqn:example isEnrolledFor} of rule {\tt EnrollRule} is satisfied
+   because relation $\declare{\tt isEnrolledFor}{\tt Student}{\tt Module}$ is empty.
+   So, there are no violations, which satisfies requirement~\ref{eqn:typed violations}.
+   With only one rule in the system, requirement~\ref{eqn:sat} is satisfied too.
    Requirement~\ref{eqn:specialization} is satisfied because this example contains no specialization.
-   So we can conclude that this example satisfies definition~\ref{def:information system}.
+   This completes the requirements for the entire quintuple $\la\dataset,\schema,\functionality,\roles,\maintain\ra$.
+   So we can conclude that information system $\infsys$ satisfies definition~\ref{def:information system}.
 

### 2022-12-11 2022 Migration/articleMigration.tex
WIP

@@
    target concept \verb#Course#.
-   It also introduces three triples:
-\[\begin{array}{l}
-   \triple{\text{\tt "Peter"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Management"}}\\
-   \triple{\text{\tt "Susan"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
-   \triple{\text{\tt "John"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}
-\end{array}\]
    The informal meaning of this relation is that it states which students are taking which courses.
@@
 
-   Summarizing, this example contains a dataset with seven triples
+   This example contains a dataset with seven triples
    and a schema with three concepts, three relations, and one rule.
-
    Now let us check the requirements to verify that this example defines an information system.

### 2022-12-11 2022 Migration/articleMigration.tex
changes made together on dec 11th

@@
 \label{sct:Terminology}
-   First,
-   we define ``information system'' as a combination of dataset, schema, and functionality.
+   An ``information system'' is a combination of dataset, schema, and functionality.
+   For the purpose of this paper, we ignore the functionality because it does not impact the migration.
    We start with datasets in section~\ref{sct:Datasets}, then we define schemas in section~\ref{sct:Schemas},
@@
    This relation is used in the type system, in which $\inst$ assigns a type to every atom.
-   This makes $\inst$ a total relation:
+   This makes $\inst$ total on the atoms in the dataset:
 \begin{eqnarray}
    \forall a\in\atoms\ \exists C\in\Concepts: a\ \inst\ C
-\label{eqn:concepts}
+\label{eqn:typedAtom}
 \end{eqnarray}
@@
    \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B
-   \label{eqn:type}
+   \label{eqn:wellTypedEdge}
 \end{eqnarray}
    For example, $\triple{\text{\atom{Peter}}}{\declare{\id{phone}}{\tt Person}{\tt TelephoneNumber}}{\text{\atom{074238991}}}$ is a triple.
-   Equation~\ref{eqn:type} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
+   Equation~\ref{eqn:wellTypedEdge} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
    In practice, users can say that Peter has telephone number 074238991.
@@
 \end{equation}
-   Equation~\ref{eqn:type} implies that for every dataset $\dataset$:
+   Equation~\ref{eqn:wellTypedEdge} implies that for every dataset $\dataset$:
 \[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ \inst\ A\ \wedge\ b\ \inst\ B\]
@@
    that represents the set of violations of rule $u$ in any dataset.
-   We also assume a function $\Sign{u} : \Dataset\rightarrow\Pair{\Concepts}{\Concepts}$
+   We also assume a function $\Sign : \rules\rightarrow\Pair{\Concepts}{\Concepts}$
    that represents the signature of a rule, i.e. the type of atoms in violations:
 \begin{eqnarray}
-   \sign{u}{\dataset}=\pair{A}{B}\ \wedge\ \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\ \inst\ A\wedge b\ \inst\ B
-   \label{eqn:typed violations}
+   \sign{u}=\pair{A}{B}\\
+   \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\ \inst\ A\wedge b\ \inst\ B
+   \label{eqn:wellTypedEdged violations}
 \end{eqnarray}
@@
 
-   %These rules reflect the meaning (semantics) that is shared among users.
-
-\subsection{Functionality}
-   The functionality of an information system is an integral part of it.
-   It consists of software.
-   In this paper, we are not interested in the precise nature of the functionality.
-
 \subsection{Information Systems}
@@
 \label{def:information system}
-\item An information system $\infsys$ is a tuple $\la\dataset,\schema,\functionality,\roles,\maintain\ra$, in which
+\item An information system $\infsys$ is a tuple $\la\dataset,\schema,\roles,\maintain\ra$, in which
 \begin{itemize}
@@
    \item schema $\schema$ is defined as in section~\ref{sct:Schemas};
-   \item $\functionality$ represents the functionality of the system, which is typically software;
    \item $\roles$ is a set of roles;
@@
 \end{verbatim}
-   The actual constraint, \verb#isEnrolledFor |- takes;isPartOf~#, is written in logic as:
+   The semantics of Ampersand defines $\viol{\tt EnrollRule}{\dataset}$ as follows:
 \begin{equation}
-   \begin{array}{l}\forall s\in\text{\tt Student}, m\in\text{\tt Module}\ \exists c\in\text{\tt Course}:\\
+   \viol{\tt EnrollRule}{\dataset}\ =\{\begin{array}{l}
+   \pair{m}{s}\mid\neg\exists c\in\text{\tt Course}:\\
 s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
-\end{array}
+\end{array}\}
 \label{eqn:example isEnrolledFor}
@@
    \end{array}\]
-   Requirement~\ref{eqn:concepts} is satisfied because every atom has a concept.
-   Dataset $\pair{\triples}{\inst}$ satisfies both requirements~\ref{eqn:type} and~\ref{eqn:pop}.
-   So, the dataset $\dataset=\pair{\triples}{\inst}$ satisfies the requirements from section~\ref{sct:Datasets}
+   Requirement~\ref{eqn:typedAtom} is satisfied because every atom has a concept.
+   The tuple $\pair{\triples}{\inst}$ satisfies both requirements~\ref{eqn:wellTypedEdge} and~\ref{eqn:pop}.
+   So, this defines a dataset $\dataset=\pair{\triples}{\inst}$ as introduced in section~\ref{sct:Datasets}.
 
@@
    a relation $\maintain$, which contains one pair only: $\pair{{\tt Administrator}}{{\tt EnrollRule}}$.
-   For the purpose of this paper, the functionality $\functionality$ is irrelevant (onderbouwen!).
    The constraint~\ref{eqn:example isEnrolledFor} of rule {\tt EnrollRule} is satisfied
    because relation $\declare{\tt isEnrolledFor}{\tt Student}{\tt Module}$ is empty.
-   So, there are no violations, which satisfies requirement~\ref{eqn:typed violations}.
+   So, there are no violations, which satisfies requirement~\ref{eqn:wellTypedEdged violations}.
    With only one rule in the system, requirement~\ref{eqn:sat} is satisfied too.
    Requirement~\ref{eqn:specialization} is satisfied because this example contains no specialization.
-   This completes the requirements for the entire quintuple $\la\dataset,\schema,\functionality,\roles,\maintain\ra$.
+   This completes the requirements for $\la\dataset,\schema,\roles,\maintain\ra$.
    So we can conclude that information system $\infsys$ satisfies definition~\ref{def:information system}.
@@
 % \begin{eqnarray}
-%    \forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\label{eqn:concepts}
+%    \forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\label{eqn:typedAtom}
 % \end{eqnarray}
@@
 % \begin{eqnarray}
-%    \forall\declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts\label{eqn:concepts}\\
-%    \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B\label{eqn:type}
+%    \forall\declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts\label{eqn:typedAtom}\\
+%    \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B\label{eqn:wellTypedEdge}
 % \end{eqnarray}

### 2022-12-11 2022 Migration/articleMigration.tex
fixed the signature function

@@
    The expression $a\ \inst\ C$ means that atom $a$ is an \emph{instance} of concept $C$.
-   This relation is used in the type system, in which $\inst$ assigns a type to every atom.
-   This makes $\inst$ total on the atoms in the dataset:
-\begin{eqnarray}
-   \forall a\in\atoms\ \exists C\in\Concepts: a\ \inst\ C
-\label{eqn:typedAtom}
-\end{eqnarray}
+   This relation is used in the type system, in which $\inst$ assigns a type to every atom in the dataset.
 
@@
    A triple is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.
+
    A dataset $\dataset$ is a tuple $\pair{\triples}{\inst}$ that satisfies:
@@
 
-   $\triples$ is a finite set of triples and $\inst$ is the instance relation.
    In case we need to be explicit about the dataset,
@@
    Every dataset is an element of an infinite set called $\Dataset$.
-   To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\triples$.
+   To save writing in the sequel, we may write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\triples$.
 
@@
    Equation~\ref{eqn:wellTypedEdge} implies that for every dataset $\dataset$:
-\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ \inst\ A\ \wedge\ b\ \inst\ B\]
+\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ a\ \inst\ A\ \wedge\ b\ \inst\ B\]
    For a developer, this means that the type of an atom depends only on the relation in which it resides; not on the actual population of the database.
@@
    We write $\schema_{\infsys}$ to refer to the schema of information system $\infsys$.
-   We write $\concepts_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$ to refer to $\concepts$, $\rels$, and $\rules$ respectively.
+   If needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$ to refer to $\concepts$, $\rels$, and $\rules$ respectively.
 %  Every schema is an element of an infinite set called $\Schema$.
@@
    A rule in a schema serves to constrain the dataset, to ensure its semantic integrity.
+   Every rule is an element of an infinite set called $\Rules$,
+   which is disjoint from $\Concepts$, $\Rels$, and $\Atoms$.
    For every rule $u$ in the schema, we assume there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
    that represents the set of violations of rule $u$ in any dataset.
-   We also assume a function $\Sign : \rules\rightarrow\Pair{\Concepts}{\Concepts}$
+   We also assume a function $\id{sign} : \Rules\rightarrow\Pair{\Concepts}{\Concepts}$
    that represents the signature of a rule, i.e. the type of atoms in violations:
 \begin{eqnarray}
-   \sign{u}=\pair{A}{B}\\
-   \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\ \inst\ A\wedge b\ \inst\ B
+   \sign{u}=\pair{A}{B}\ \wedge\ \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\ \inst\ A\wedge b\ \inst\ B
    \label{eqn:wellTypedEdged violations}
@@
 \begin{itemize}
-   \item dataset $\dataset$ is defined as in section~\ref{sct:Datasets};
-   \item schema $\schema$ is defined as in section~\ref{sct:Schemas};
+   \item dataset $\dataset=\pair{\triples}{\inst}$ is defined as in section~\ref{sct:Datasets};
+   \item schema $\schema=\triple{\concepts}{\rels}{\rules}$ is defined as in section~\ref{sct:Schemas};
    \item $\roles$ is a set of roles;
@@
    \end{array}\]
-   Requirement~\ref{eqn:typedAtom} is satisfied because every atom has a concept.
    The tuple $\pair{\triples}{\inst}$ satisfies both requirements~\ref{eqn:wellTypedEdge} and~\ref{eqn:pop}.
@@
    We define a \define{migration of a dataset} $\la\atoms,\concepts,\inst,\rels,\dataset\ra$ as a change in which one or more of $\inst$, $\concepts$, or $\rels$ change.
-   To satisfy the definition of datasets, $\dataset$ and $\atoms$ must change too,
+   To satisfy the definition of datasets, $\dataset$ must change too,
    but the migration should try to preserve the maximal amount of data.

### 2022-12-13 2022 Migration/articleMigration.tex
Aligned the first example to the definitions.

@@
    to generate information systems.
-   A script in the Ampersand language represents the system's semantics by constraints only.
+   A script in the Ampersand language represents the system's semantics by rules,
+   which constrain the dataset.
    We don't have to formalize the semantics from specifications or from source code
-   because the constraints are already explicitly available in the Ampersand script.
+   because the constraints are already explicitly available as rules in the Ampersand script.
    A system generated by the Ampersand compiler will keep these constraints satisfied at all times.
@@
    To design an information system, a developer defines a schema in which she defines concepts, relations, and rules.
-   The purpose of a schema is to capture the semantics of an information system.
-   Rules serve the purpose of invariants,
-   i.e.\ properties that the data in the database must satisfy at all times.
+   Together, the concepts, relations, and rules capture the semantics of an information system~\cite{Spivak2012}.
+   Each rule in a schema serves to constrain the dataset, to ensure its semantic integrity.
+   Every rule is an element of an infinite set called $\Rules$,
+   which is disjoint from $\Atoms$, $\Concepts$, $\Rels$, and $\Dataset$.
    In this paper, variables $u$ and $v$ represent rules.
+   For every rule $u$ in the schema, we assume there is a predicate $\sat{u}{\Dataset}$
+   that says whether dataset $\dataset$ satisfies rule $u$.
+%    For every rule $u$ in the schema, we assume there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
+%    that represents the set of violations%
+% \footnote{The definition of ``violation'' is not needed in this paper. We only use the absence of violations to show satisfaction of a rule.}
+%    of rule $u$ in any dataset.
+%    We also assume a function $\id{sign} : \Rules\rightarrow\Pair{\Concepts}{\Concepts}$
+%    that represents the signature of a rule, i.e. the type of atoms in violations:
+% \begin{eqnarray}
+%    \sign{u}=\pair{A}{B}\ \wedge\ \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\ \inst\ A\wedge b\ \inst\ B
+%    \label{eqn:wellTypedEdged violations}
+% \end{eqnarray}
+%    To determine whether rule $u$ is satisfied in dataset $\dataset$,
+%    we define $\sat{u}{\Dataset}$:
+% \begin{eqnarray}
+%    \sat{u}{\dataset}&\Leftrightarrow&\viol{u}{\dataset}=\emptyset
+%    \label{eqn:sat}
+% \end{eqnarray}
+%    Violations are used to produce meaningful error messages and to trigger changes in the dataset to restore satisfaction of that rule.
+
+   Rules serve as invariants,
+   i.e.\ constraints that the data in the database must satisfy at all times.
+   At times in which the dataset does not satisfy a rule, we say that the rule is {\em broken}.
+   In those cases, something or someone must take action to fix the situation and restore invariance of that rule.
 
@@
 
-   A rule in a schema serves to constrain the dataset, to ensure its semantic integrity.
-   Every rule is an element of an infinite set called $\Rules$,
-   which is disjoint from $\Concepts$, $\Rels$, and $\Atoms$.
-   For every rule $u$ in the schema, we assume there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
-   that represents the set of violations of rule $u$ in any dataset.
-   We also assume a function $\id{sign} : \Rules\rightarrow\Pair{\Concepts}{\Concepts}$
-   that represents the signature of a rule, i.e. the type of atoms in violations:
-\begin{eqnarray}
-   \sign{u}=\pair{A}{B}\ \wedge\ \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\ \inst\ A\wedge b\ \inst\ B
-   \label{eqn:wellTypedEdged violations}
-\end{eqnarray}
-   To determine whether rule $u$ is satisfied in dataset $\dataset$,
-   we define $\sat{u}{\Dataset}$:
-\begin{eqnarray}
-   \sat{u}{\dataset}&\Leftrightarrow&\viol{u}{\dataset}=\emptyset
-   \label{eqn:sat}
-\end{eqnarray}
-   Violations are used to produce meaningful error messages and to trigger changes in the dataset to restore satisfaction of that rule.
-
    We assume a special kind of rule that is used for specialization.
@@
    Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.
-   The specialization rules in a schema form a partial order $\isa : \Pair{\concepts}{\concepts}$,
+   The specialization rules in a schema form a partial order $\isa$,
    meaning that $\isa$ is reflexive, transitive, and antisymmetric.
@@
    \item $\roles$ is a set of roles;
-   \item $\maintain : \roles\times\rules$ is the maintainance relation between roles and rules from $\schema$.
+   \item $\maintain : \roles\times\rules$ is the maintainance relation between roles and rules.
 \end{itemize}
@@
    A \define{role} is a name that identifies a group of users.
-   It serves as a placeholder for a person or a machine (i.e. an actor) who can change the dataset.
+   It serves as a placeholder for a person or a machine (i.e. an actor) that works with the dataset (i.e. create, read, update, or delete triples).
    The purpose of a role is to mention an individual user (human) or an automated actor (bot) without knowing who that user is.
@@
 
-   Every rule in a schema is meant to be kept satisfied invariably.
-   The only way to break this invariance is to remove the rule from the schema.
-   So the "normal" state of a system is that its dataset $\dataset$ contains no violations of any rule in the schema:
+   The "normal" state of a system is that every rule in the schema is satisfied:
 \begin{equation}
@@
    However, the system continually registers events that insert or delete triples in its dataset,
-   which may cause rules to be satisfied no longer.
-   This is a temporary breach of the invariance of the affected rules,
-   which requires a reaction to restore invariance.
+   which may cause rules to be broken.
+   This requires a reaction to restore invariance.
    Example: if a rule says that every request must be acknowledged, the receipt of a request (which is an event) will violate this rule.
@@
    Any actor with role $o$, whether person or machine,
-   can restore invariance of rule $u$ by inserting or deleting the right triples to compensate for the violations.
+   can restore invariance of rule $u$ by inserting or deleting the right triples.
    For this reason, we assume that every rule is assigned to at least one role by means of the relation $\maintain$.
@@
    $o\ \maintain\ u$ means that role $o$ must keep rule $u$ satisfied, i.e.\ to keep $\sat{u}{\dataset}$ true.
-   As long as rule $u$ is not satisfied, its violations won't go away, stimulating the actor to do something about it.
+   As long as rule $u$ is broken, the system should notify actors in the role(s) that are administered in $\maintain$ to do something about it.
 
    Automatic enforcement is specified by the developer with a special syntax,
-   which specifies not only the rule, but also the way to restore invariance.
-   The system contains an engine that restores invariance of all automatically enforced rules without unneccessary delay.
+   the {\em enforce rule}.
+   An enforce rule specifies not only the rule, but also the way to restore invariance.
+   We assume that the system features an engine that restores invariance of all enforce rules without unneccessary delay.
    Whenever the dataset violates an enforce rule,
    this engine will make other changes to the dataset to satisfy that rule.
-   Hence, the violation of the rule exists during a small, finite amount of time.
+   Hence, the rule is not satisfied during a small, finite amount of time.
 
@@
    so every rule has at least one role in the relation $\maintain$, means that all roles together are keeping all rules satisfied.
-   (\verb-@sjcjoosten-, what if $\maintain$ relates both a human role and the enforcement engine role to the same rule?)
 
@@
    because it makes the examples more appealing to read.
-   Let us first define a dataset of seven triples and three relations.
+   Let us first define a dataset of just a handful of triples and three relations.
 \begin{verbatim}
@@
    The informal meaning of this relation is that it states which students are taking which courses.
-   So, there is a student, Peter, taking the course Management and two students, Susan and John, taking the course Business IT.
 
@@
 \end{verbatim}
-   We also define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that a student takes.
+
+   To complete the semantics,
+   we also define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that a student takes.
    In Ampersand, which is a syntactically sugared form of relation algebra,
@@
 \end{verbatim}
-   The semantics of Ampersand defines $\viol{\tt EnrollRule}{\dataset}$ as follows:
+   The semantics of Ampersand defines $\sat{\tt EnrollRule}{\dataset}$ as follows:
 \begin{equation}
-   \viol{\tt EnrollRule}{\dataset}\ =\{\begin{array}{l}
-   \pair{m}{s}\mid\neg\exists c\in\text{\tt Course}:\\
+   \begin{array}{l}
+   \forall \pair{s}{m}\in\pop{\tt isEnrolledFor}{\dataset}\ \exists c\in\text{\tt Course}:\\
 s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
-\end{array}\}
+\end{array}
 \label{eqn:example isEnrolledFor}
 \end{equation}
+   Rule {\tt EnrollRule} is satisfied because relation $\declare{\tt isEnrolledFor}{\tt Student}{\tt Module}$ is empty.
 
-   This example contains a dataset with seven triples
-   and a schema with three concepts, three relations, and one rule.
    Now let us check the requirements to verify that this example defines an information system.
@@
    \end{array}\]
-   The tuple $\pair{\triples}{\inst}$ satisfies both requirements~\ref{eqn:wellTypedEdge} and~\ref{eqn:pop}.
-   So, this defines a dataset $\dataset=\pair{\triples}{\inst}$ as introduced in section~\ref{sct:Datasets}.
+   The tuple $\pair{\triples}{\inst}$ satisfies requirement~\ref{eqn:wellTypedEdge} so this is a dataset $\dataset$ as introduced in section~\ref{sct:Datasets}.
 
@@
    And, it defines the set of rules $\rules$ to contain just one rule: \verb-EnrollRule-.
-   $\rules$ satisfies requirement~\ref{eqn:satisfaction} because $\viol{\tt EnrollRule}{\dataset}$ is empty.
+   Requirement~\ref{eqn:specialization} is satisfied because this example contains no specialization.
    So, the schema $\schema=\triple{\concepts}{\rels}{\rules}$ satisfies the requirements from section~\ref{sct:Schemas}.
 
-   To fulfill the definition of information system, definition~\ref{def:information system},
+   Now let us check the definition of information system.
    Ampersand generates a set of roles $\roles=\{{\tt Administrator}\}$ and
    a relation $\maintain$, which contains one pair only: $\pair{{\tt Administrator}}{{\tt EnrollRule}}$.
-   The constraint~\ref{eqn:example isEnrolledFor} of rule {\tt EnrollRule} is satisfied
-   because relation $\declare{\tt isEnrolledFor}{\tt Student}{\tt Module}$ is empty.
-   So, there are no violations, which satisfies requirement~\ref{eqn:wellTypedEdged violations}.
-   With only one rule in the system, requirement~\ref{eqn:sat} is satisfied too.
-   Requirement~\ref{eqn:specialization} is satisfied because this example contains no specialization.
-   This completes the requirements for $\la\dataset,\schema,\roles,\maintain\ra$.
+   Requirement~\ref{eqn:satisfaction} is satisfied because the only rule, {\tt EnrollRule} is satisfied.
+   This completes the requirements for $\infsys=\la\dataset,\schema,\roles,\maintain\ra$.
    So we can conclude that information system $\infsys$ satisfies definition~\ref{def:information system}.
-
-%    The language Ampersand is typed, to prevent a substantial amount of programming mistakes at compile-time.
-%    This means that every atom is an instance of a concept:
-% \begin{eqnarray}
-%    \forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\label{eqn:typedAtom}
-% \end{eqnarray}
-%    To do type checking based on the schema only,
-%    all relations in triples from the dataset must be ``known'' in the schema:
-% \begin{equation}
-%    \forall a,r,b:\ \triple{a}{r}{b}\in\triples\ \Rightarrow\ r\in\rels\label{eqn:define R}
-% \end{equation}
-
-%    In a typed information system, atoms in a relation have a type as declared in a relation.
-%    For this reason, every relation $r$ has not only a name, but also a source concept and a target concept.
-%    We write $r=\declare{nm}{A}{B}$ to denote that relation $r$ has name \id{nm}, source concept $A$, and target concept $B$.
-% \begin{eqnarray}
-%    \forall\declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts\label{eqn:typedAtom}\\
-%    \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B\label{eqn:wellTypedEdge}
-% \end{eqnarray}
-%    The type checker in the Ampersand compiler will spot violations and refuse to generate a system if they occur.
-
-%    To compare atoms for equality,
-%    the set $\rels$ contains an identity relation $\ident{C}$ for every concept $C$ in which all atoms occur.
-%    It satisfies:
-% \begin{equation}
-%    \forall a,b:\ a\ \inst\ C\ \wedge\ b\ \inst\ C\ \Rightarrow\ (a\ \ident{C}\ b\ \Leftrightarrow\ a=b)
-% \end{equation}
-%    So, $\id{pop}_{\ident{C}}$ is defined by:
-% \begin{eqnarray}
-%    \forall C\in\concepts&:&\pop{\ident{C}}{\dataset}=\{\pair{a}{a}\mid a\ \inst\ C\}
-% \end{eqnarray}
-
-%    Having one rule, one role, a relation $\maintain$ that contains one pair, and a dataset,
-%    we have now defined a small system.
 
@@
 \end{array}\]
-   Ampersand also derives the following violation sets:
-\[\begin{array}{l}
-   \viol{\text{\tt ExamRule1}}{\dataset}\\
-   \hspace{1cm}=\{\pair{s}{c}\mid\exists e:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\neg(s\ \text{\tt takes}\ c)\}\\
-   \viol{\text{\tt ExamRule2}}{\dataset}\\
-   \hspace{1cm}=\{\pair{s}{m}\mid\exists e,c:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\ \wedge\neg(s\ \text{\tt isEnrolledFor}\ m)\}
-\end{array}\]
+   % Ampersand also derives the following violation sets:
+% \[\begin{array}{l}
+   % \viol{\text{\tt ExamRule1}}{\dataset}\\
+   % \hspace{1cm}=\{\pair{s}{c}\mid\exists e:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\neg(s\ \text{\tt takes}\ c)\}\\
+   % \viol{\text{\tt ExamRule2}}{\dataset}\\
+   % \hspace{1cm}=\{\pair{s}{m}\mid\exists e,c:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\ \wedge\neg(s\ \text{\tt isEnrolledFor}\ m)\}
+% \end{array}\]
 
@@
 \end{verbatim}
-   If we check this for violations, we get the following results:
+   If we check this script, we get the following results:
 \begin{verbatim}
@@
 
-\subsubsection{Strategies for dealing with violations}
+\subsubsection{Strategies for restoring invariance}
 
-The violations of ExamRule2 don't pose a problem: they have a role (administrator) assigned to them.
-We allow these violations to occur in the transition to the new system, and expect people with the Administrator role to deal with them.
+We allow ExamRule2 to be broken during the transition to the new system, and expect people with the Administrator role to deal with them.
 

### 2022-12-13 Generative Software/proposalBrief.tex
teaser gemaakt

\documentclass{elsarticle}
% \usepackage{graphicx}
% \usepackage{footmisc}
% \usepackage{amstext}
% \usepackage{amsmath}
% \usepackage{amssymb}
% \usepackage{amsthm}
\usepackage[english]{babel}
% \usepackage[official,right]{eurosym}
\selectlanguage{english}
\begin{document}
% \include{preamble}

\title{Generative Software Development\\{\em\normalsize Research Proposal}}
\author{prof.dr.ir.\ Stef Joosten\footnote{ORCID 0000-0001-8308-0189}}
\author{dr.\ Tim Steenvoorden\footnote{ORCID 0000-0002-8436-2054}}
\address{Open Universiteit Nederland, Heerlen, the Netherlands}

\begin{abstract}
    This paper summarizes the capability of the Open Universiteit in generative software development.
    We offer this enabling technology as a component of a larger effort that involves designing distributed information systems.
    Generative software development offers attractive benefits for security and resilience,
    and enables a product to be demonstrated live every 2 or 3 weeks as it develops into an end product.
\end{abstract}

\begin{keyword}
    generative software, software engineering, security, prototyping, correctness, Ampersand, TopHat
\end{keyword}
\maketitle

\section{Problem area}
   Data driven organizations that need to adapt frequently to changing circumstances face a software dilemma.
   On one hand, flexible and fast adaptation requires immediate deployment of new software components, without interrupting the ongoing business.
   On the other hand, one-of-a-kind organizations require one-of-a-kind components that need to be developed first.
   If the required component is generic enough, commercial off-the-shelf (COTS) components or Software-as-a-Service (SaaS) components will solve that problem.
   However, if very specific components are not on the market, organizations must revert to homebrew components.
   The choice between a readily available solution and a homebrew solution can be a dilemma:
   you either sacrifice the specific support for your business or you sacrifice the required flexibility.
   In practice, we see many organizations that opt for low-code platforms to solve this dilemma.
   Triggered by the advertised prospect that business stakeholders can build the software by themselves,
   organizations can be in for a disappointment.

\section{Vision}
   Generative software development uses a software generator to build information systems from a formal specification.
   Even though the idea is quite old (compilers are a form of software generator),
   this idea has not been picked up in mainstream software development.
   Yet, research so far has yielded promising results with respect to validation and resilience.
   A software generator enables the DevOps way of working, which is to demonstrate the end product at the end of every sprint.
   Thus, a way of software development emerges that allows for very fast incremental development of large information systems.
   Research at the Open Universiteit has produced software generators specifically for information systems.
   This is useful for supporting distributed business processes that require a mix of human and automated activity.

\section{Benefits}
    The Open University offers its expertise~\cite{Joosten-JLAMP2018,JoostenRAMiCS2017,10.1145/3354166.3354182} and tools as a building block for a larger research effort,
    in which generative software development is an enabler for the research proposed.
    This capability offers the following benefits:
\begin{itemize}
    \item By generating prototypes, we can produce live demonstrations periodically to keep the future user population informed of the progress.
    This produces quick results, which keeps funders happy.
    \item By automating software production, we can prevent a large class of mistakes and thereby reduce the project risk.
    \item By formally specifying the functionality, we can give valuable guarantees on the correctness of the results.
    \item Software that is correctly generated needs no testing. This yields significant savings on testing and test maintenance.
\end{itemize}

\section{Researchers}
    The authors are acknowledged scientists, both with a proven track record in generative software development.

    \section{Bibliography}
\bibliographystyle{elsarticle-harv}
\bibliography{doc}

\end{document}

### 2022-12-13 Generative Software/proposalBrief.tex
proposal brief made

@@
    Even though the idea is quite old (compilers are a form of software generator),
-   this idea has not been picked up in mainstream software development.
+   this idea is not being picked up widely in the mainstream of software development.
    Yet, research so far has yielded promising results with respect to validation and resilience.

### 2022-12-13 2022 Migration/articleMigration.tex
wip

@@
    To complete the semantics,
-   we also define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that a student takes.
-   In Ampersand, which is a syntactically sugared form of relation algebra,
+   we also define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that student takes.
+   In Ampersand, which is a syntactically sugared form of relation algebra~\cite{JoostenRAMiCS2017},
    we give each rule a name and declare a role to preserve its invariance:
@@
 \end{verbatim}
-   The semantics of Ampersand defines $\sat{\tt EnrollRule}{\dataset}$ as follows:
+   The semantics of this rule defines $\sat{\tt EnrollRule}{\dataset}$ as:
 \begin{equation}

### 2022-12-13 Generative Software/proposalBrief.tex
small enhancements proposal brief

@@
 \usepackage[english]{babel}
+\usepackage{libertine}
+\usepackage{inconsolata}
 % \usepackage[official,right]{eurosym}
 \selectlanguage{english}
+\frenchspacing
+
 \begin{document}
@@
     We offer this enabling technology as a component of a larger effort that involves designing distributed information systems.
-    Generative software development offers attractive benefits for security and resilience,
-    and enables a product to be demonstrated live every 2 or 3 weeks as it develops into an end product.
+    Generative software development offers attractive benefits for correctness, security, and resilience.
+    Also, it enables a product to be demonstrated live during its creation,
+    hereby supporting agile software development methods.
+    %every 2 or 3 weeks as it develops into an end product.
 \end{abstract}
@@
 \section{Benefits}
-    The Open University offers its expertise~\cite{Joosten-JLAMP2018,JoostenRAMiCS2017,10.1145/3354166.3354182} and tools as a building block for a larger research effort,
+    The Open University offers its expertise~\cite{JoostenRAMiCS2017,Steenvoorden2022} and tools~\cite{Joosten-JLAMP2018,10.1145/3354166.3354182} as a building block for a larger research effort,
     in which generative software development is an enabler for the research proposed.
@@
 \begin{itemize}
-    \item By generating prototypes, we can produce live demonstrations periodically to keep the future user population informed of the progress.
+    \item By generating software, we can produce live demonstrations periodically to keep the future user population informed of the progress.
     This produces quick results, which keeps funders happy.
@@
 \section{Researchers}
-    The authors are acknowledged scientists, both with a proven track record in generative software development.
+    The authors are acknowledged scientists, both with a proven track record in generative software development and workflow modelling.
+
+\newpage
 

### 2022-12-14 2022 Migration/articleMigration.tex
first attempt to define renaming on concepts

@@
 \documentclass{elsarticle}
-\usepackage{graphicx}
+% \usepackage{graphicx}
+\usepackage{hyperref}
 %\usepackage{multicol}
@@
    In this contribution we develop a theory for reliable data migration to help the migration engineer
-   to transfer the data and preserve the semantics as much as possible.
+   to transfer the data and preserve their semantics as much as possible.
    We aim to automate the data migration,
@@
    \begin{center}
-     \includegraphics[scale=.35]{Migration.png}
+     \includegraphics[scale=.35]{Migration.drawio.png}
    \end{center}
@@
    In Ampersand, which is a syntactically sugared form of relation algebra~\cite{JoostenRAMiCS2017},
-   we give each rule a name and declare a role to preserve its invariance:
+   we give each rule a name and declare a role to maintain its invariance:
 \begin{verbatim}
@@
    Requirement~\ref{eqn:satisfaction} is satisfied because the only rule, {\tt EnrollRule} is satisfied.
-   This completes the requirements for $\infsys=\la\dataset,\schema,\roles,\maintain\ra$.
-   So we can conclude that information system $\infsys$ satisfies definition~\ref{def:information system}.
+   This completes the requirements for $\infsys=\la\dataset,\schema,\roles,\maintain\ra$ from definition~\ref{def:information system}.
+   So we can conclude that $\infsys$ is an  information system.
 
 \subsection{Data migration}
+   In a datamigration, we distinguish three moments:
+\begin{enumerate}
+   \item the moment of transition, i.e. the moment the old system is replaced (in production) by the new system;
+   \item the decision point, i.e. the moment a business authority declares the migration successful;
+   \item the point of no return, i.e. the moment that the migration cannot be undone other than by the regular backup-procedure.
+\end{enumerate}
+   The decision point must come after the moment of transition, to allow the business to verify the migration.
+   The point of no return must come after the decision point, to allow for a no-go decision from the business.
+
    To migrate system $\infsys$ to $\infsys'$, we take the following steps:
@@
    \item Let the old system $\infsys$ have state $\dataset$ and schema $\schema$.
-         A developer defines a schema $\schema'$ of the new system as a first step in the deployment of the increment.
-   \item The generator generates a migration system $\migrsys$ based on $\schema$ and $\schema'$,
+         A developer defines a new system $\infsys'$, which contains a dataset $\dataset'$, a schema $\schema'$, and the functionality\footnote{For now, we ignore the functionality.}.
+         In Ampersand, this means that the developer produces an Ampersand script for the new system.
+         The difference between the new system and the old system is the increment.
+         To preserve the semantics of data means that the dataset $\dataset$ must be combined with $\dataset'$ in the new system.
+         Dataset $\dataset'$ is optional, in case the developer wants to add some new data into the system on the moment of transition.
+         If $\schema$ equals $\schema'$, we call the migration trivial because a migration is not necessary.
+   \item The generator generates a migration system $\migrsys$ based on $\schema$ and $\infsys'$,
          which gives the user the full functionality of the old system and the new system,
@@
          yielding migration system $\migrsys'$.
-         TODO: de migration engineer kan alleen maar in $\dataset'$ dingen (laten) wijzigen, zodat je altijd terug kunt naar het oude systeem.
+         The migration engineer does not alter the old system $\infsys$ nor the specification new system $\infsys'$ that the developer has produced.
+         As a result, the migration engineer can always restore the old system because it is part of the migration system.
    \item The production environment switches instantly from $\infsys$ to $\migrsys'$,
@@
          This means that the data of the old system remains intact.
+         This event marks the moment of transition.
+   \item The rules in the migration system that are not in $\schema'$ will remain in the system until they are satisfied,
+         to enforce their execution.
+         Automatic migration rules can be executed at the moment of transition.
+         Migration rules that are enforced manually will be held alive until they are satisfied.
+         All rules from $\schema$ (the rules from the old system) are removed on the moment of transition, because the are satisfied by property~\ref{eqn:satisfaction}.
    \item After a finite span of time, $\dataset_{\migrsys'}$ satisfies all its enforce rules.
@@
 \begin{eqnarray}
-   \omit\rlap{$\pair{\triples}{\inst}\sqcup\pair{\triples'}{\inst'}$}\notag\\
+   \rlap{$\pair{\triples}{\inst}\sqcup\pair{\triples'}{\inst'}$}\notag\\
    &=&\pair{\triples\uplus\triples'}{\inst\uplus^2\inst'}\notag\\
@@
 \end{definition}
+
+\begin{lemma}
+   If $\dataset$ and $\dataset'$ are datasets, then so is $\dataset\sqcup\dataset'$.
+\end{lemma}
+   This lemma is being proven in Isabelle/HOL~\cite{Isabelle}. The proof is published \href{location.domain}{here}
 
@@
 % \end{definition}
-Note that if $\dataset$ and $\dataset'$ are datasets, then so is $\dataset\sqcup\dataset'$.
+\begin{lemma}
+   If $\schema$ and $\schema'$ are schemas, then so is $\schema\sqcup\schema'$.
+\end{lemma}
+The proof of this lemma is published \href{location.domain}{here}
 
+To implement the disjoint union, we need to relabel 
+\begin{definition}[relabel concepts]
+   \[\subst{C}{D}{A}\ =\ \text{\bf if}\ C=A\ \text{\bf then}\ D\ \text{\bf else}\ A\]
+\end{definition}
+\begin{definition}[relabel concepts in triples]
+   \[\subst{C}{D}{\triple{a}{r}{b}} = \triple{a}{\subst{C}{D}{r}}{b}\]
+\end{definition}
+\begin{definition}[relabel concepts in relations]
+   \[\subst{C}{D}{(\declare{n}{A}{B})} = \declare{n}{\subst{C}{D}{A}}{\subst{C}{D}{B}}\]
+\end{definition}
+\begin{definition}[relabel concepts in datasets]
+   \[\begin{array}{rcl}
+      \rlap{$\subst{C}{D}{\pair{\triples}{\inst}}$}\\
+      &=&\pair{\{\subst{C}{D}{t}\mid t\in\triples\}}{\{a\ \inst\ \subst{C}{D}{A}\mid a\ \inst\ A\}}
+   \end{array}\]
+\end{definition}
+\begin{definition}[relabel concepts in rules]
+   \[\sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{C}{D}{u}}{\subst{C}{D}{\dataset}}\]
+\end{definition}
+\begin{definition}[relabel concepts in schemas]
+   \begin{eqnarray}
+      \rlap{$\subst{C}{D}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
+      &=&\begin{array}[t]{l@{}l}
+         \la&\{\subst{C}{D}{t}\mid t\in\triples\}\\
+         ,&\{\subst{C}{D}{r}\mid r\in\rels\}\\
+         ,&\{\subst{C}{D}{u}\mid u\in\rules\}\ \ra\notag
+         \end{array}
+   \end{eqnarray}
+\end{definition}
+
+   The relabeling of concepts is not trivial because of specialization.
+   To illustrate this point, suppose $D\isa A$.
+   Then the substitution $[C\rightarrow D]$ means that every atom $a$ that used to be of concept $C$ in the old system, is now not only a $D$ but also an $A$.
+   For example by relabeling the concept {\tt Hotel} to {\tt Motel}, in a situation where ${\tt Motel}\isa{\tt Parking}$,
+   all atoms that represent hotels might suddenly be expected to have parking lots.
+   So, whereas such examples may fly in a technical sense, the developer must use renaming with care to preserve the intended semantics.
 \subsection{Example}

### 2022-12-18 2022 Migration/articleMigration.tex
fixed subst of concepts in inst
rewritten the abstract

@@
 
-\title{It's all about Meaning!\\{\normalsize A Theory for Data Migration during Software Development}}
+\title{It's all about Meaning!\\{\normalsize A Theory for Data Migration when Deploying Changes}}
 \author[ou,ordina]{Stef Joosten\fnref{fn1}}
@@
 \begin{abstract}
+   To support incremental software development with software generators bears the potential of decreasing the time to deploy changes.
+   However, an increment that changes the database schema causes a nontrivial data migration problem.
+   The problem is how to preserve the semantics of the data as much as possible and satisfy elementary business requirements at the same time.
+
    The Ampersand project has provided the theory and tools to generate information systems from an algebraic specification.
-   However, information systems in practice may change repeatedly after their maiden deployment.
-   Changes that affect the data model typically result in a data migration.
-   In such cases, simply regenerating the system is not enough.
-   That would reset the database to its initial state, losing all data gathered so far.
-   To prevent that, a migration engineer must transfer the old data to the new system by hand.
-
-   In this contribution we develop a theory for reliable data migration to help the migration engineer
-   to transfer the data and preserve their semantics as much as possible.
-   We aim to automate the data migration,
-   to prevent mistakes and enable more frequent migrations.
-   The target is to generate a migration script from two specifications: the old specification and the new specification.
-   A software generator that embodies this theory is subject of future research.
+   This paper develops a theory for making changes to an already deployed information system that was generated by Ampersand.
+   We propose a migration method with zero down-time, and allows the business to undo the change after verification.
+   It also leaves room for the business to finish ongoing work after the moment the change is deployed.
+   We aim to automate the change including the data migration,
+   to prevent mistakes and enable more frequent changes.
 \end{abstract}
@@
 \begin{keyword}
-relation algebra\sep software development\sep data migration\sep software migration\sep Ampersand
+generative software\sep incremental software development\sep data migration\sep relation algebra \sep Ampersand
 \end{keyword}
@@
    \item the decision point, i.e. the moment a business authority declares the migration successful;
-   \item the point of no return, i.e. the moment that the migration cannot be undone other than by the regular backup-procedure.
+   \item the point of no return, i.e. the moment that the migration cannot be undone other than by a backup-procedure,
+         which is outside the scope of this paper.
 \end{enumerate}
@@
    \item Let the old system $\infsys$ have state $\dataset$ and schema $\schema$.
-         A developer defines a new system $\infsys'$, which contains a dataset $\dataset'$, a schema $\schema'$, and the functionality\footnote{For now, we ignore the functionality.}.
+         A developer defines a new system $\infsys'$, which contains a dataset $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.}.
          In Ampersand, this means that the developer produces an Ampersand script for the new system.
          The difference between the new system and the old system is the increment.
-         To preserve the semantics of data means that the dataset $\dataset$ must be combined with $\dataset'$ in the new system.
-         Dataset $\dataset'$ is optional, in case the developer wants to add some new data into the system on the moment of transition.
+         The new system must contain the triples from $\dataset$ and $\dataset'$, to preserve the old data.
+         If the developer wants to add some new data into the system on the moment of transition,
+         she specifies that in $\dataset'$.
+         That might be necessary to initialize specific new features.
          If $\schema$ equals $\schema'$, we call the migration trivial because a migration is not necessary.
@@
 \end{definition}
+\begin{definition}[relabel concepts in $\inst$]
+   \[a\ \subst{C}{D}{\inst}\ A\ =\ \text{\bf if}\ C=A\ \text{\bf then}\ a\ \inst\ D\ \text{\bf else}\ a\ \inst\ A\]
+\end{definition}
 \begin{definition}[relabel concepts in datasets]
@@
       \rlap{$\subst{C}{D}{\pair{\triples}{\inst}}$}\\
-      &=&\pair{\{\subst{C}{D}{t}\mid t\in\triples\}}{\{a\ \inst\ \subst{C}{D}{A}\mid a\ \inst\ A\}}
+      &=&\pair{\{\subst{C}{D}{t}\mid t\in\triples\}}{\subst{C}{D}{\inst}}
    \end{array}\]

### 2022-12-18 2022 Migration/articleMigration.tex
fixed mistake in definition 7 (inst)

@@
 \begin{abstract}
-   To support incremental software development with software generators bears the potential of decreasing the time to deploy changes.
+   To support incremental software development with a software generator bears the potential of decreasing the time to deploy changes.
    However, an increment that changes the database schema causes a nontrivial data migration problem.
@@
 
-   The Ampersand project has provided the theory and tools to generate information systems from an algebraic specification.
-   This paper develops a theory for making changes to an already deployed information system that was generated by Ampersand.
-   We propose a migration method with zero down-time, and allows the business to undo the change after verification.
-   It also leaves room for the business to finish ongoing work after the moment the change is deployed.
-   We aim to automate the change including the data migration,
-   to prevent mistakes and enable more frequent changes.
+   To generate an information system in a non-incremental manner is already
+   provided for by the Ampersand project.
+   This paper is about changing an already generated system in production.
+   It develops a theory for deploying an incremental change,
+   making Ampersand more useful for agile software development.
+   We propose a migration method that features zero down-time,
+   and features an undo after verification by the business.
+   It also leaves room for the business to finish ongoing work
+   after the moment the change has been deployed.
 \end{abstract}
@@
    The benefits work in two directions.
-   Automation data migrations saves effort and errors,
+   Automating data migrations saves effort and errors,
    contributing to more frequent and smaller releases.
@@
    Examples like these have convinced us to define a new method for data migration that is specifically meant for systems in production that are
-   continually evolving in small increments.
+   evolving incrementally.
 
    The contribution of this paper is to derive a migration schema from two artifacts: an existing system
-   (which has its own schema and dataset) and a new schema (which specifies the new system),
-   which allows us to facilitate the migration.
+   (which has its own schema and dataset) and a new schema (which specifies the new system).
    The purpose of our theory is to develop tools.
@@
    and may work from different locations and on different moments.
-   This collective use serves a purpose which we loosely call ``the business''.
-   As a consequence, the data in a system changes continually.
-   To preserve meaning, users are trying to maintain semantic constraints on the data,
+   This collective use by multiple users serves a purpose which we loosely call ``the business''.
+   To preserve meaning, the business maintains semantic constraints on the data,
    amidst of all changes that are going on around them.
@@
    Information systems are typically used by distributed actors (both users and computers) who continually work with data.
+   As a consequence, the data in a system changes continually.
    Every system contains a dataset, which represents the state of the system.
@@
 \begin{itemize}
+   \item data pollution that violates a constraint in the schema.
+   \item data pollution that violates a constraint that is not in the schema.
    \item data pollution that cannot be captured in constraints on the dataset,
    such as a street address that has become obsolete because the system never registered the change of address.
-   \item data pollution that violates a constraint in the schema.
-   \item data pollution that violates a constraint that is not in the schema.
    \item data pollution that is a consequence of an erroneous constraint in the schema.
 \end{itemize}
-   The first category of pollution must be dealt with by working procedures, to prevent such pollution as much as possible.
+   The first category is about violations of constraints from the schema.
+   Such violations do not occur in our theory because the old system was generated by Ampersand,
+   so the old system has kept all constraints from the schema satisfied.
+   This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
+   The second category, data pollution that is not captured by any of the constraints mentioned in the schema,
+   can be dealt with by adding constraints in the new system.
+   In that case, a migration engineer must ensure that the new system can start with data that satisfies all constraints in its schema.
+   So some human intervention may be necessary in specific cases.
+   The third category of pollution must be dealt with by working procedures, to prevent such pollution as much as possible.
    In some cases, constraints on data can be formulated,
@@
    So, such constraints may help to move this type of pollution to one of the other categories.
-   The second category is about violations of constraints from the schema.
-   Such violations do not occur in our theory because the old system was generated by Ampersand,
-   so that system will keep all constraints from the schema satisfied at all times.
-   This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
-   The third category, data pollution that is not captured by any of the constraints mentioned in the schema,
-   can be dealt with by adding constraints in the new system.
-   In that case, a migration engineer must ensure that the new system can start with data that satisfies all constraints in its schema.
-   So some human intervention may be necessary in specific cases.
    The last category of data pollution, erroneous constraints, must be fixed by replacing them by the correct constraints in the new system.
@@
    If the semantics encoded in the new schema differs from the old schema,
-   a migration engineer must make choices.
+   a migration engineer sometimes has to make choices.
    If the interpretation of data changes,
-   a migration engineer must remove obsolete constraints from the old system and introduce new constraints in the new system.
+   a migration engineer drops obsolete constraints from the old system and introduces new constraints in the new system.
    The consequences of such changes may require changes in data as well.
@@
    Atoms serve as data elements.
-   Atoms are values without internal structure of interest, meant to represent data elements in a database.
+   Atoms are values without internal structure of interest, meant to represent atomic data elements (e.g. dates, strings, numbers, etc.) in a database.
    From a business perspective, atoms represent concrete items of the world,
@@
 
-   The "normal" state of a system is that every rule in the schema is satisfied:
+   The system is {\em at rest} when every rule in the schema is satisfied:
 \begin{equation}
@@
 \subsection{Example}
+\label{sct:Example old IS}
 \label{old IS}
@@
 \subsection{Data migration}
+   Sections~\ref{sct:Information Systems} and~\ref{sct:Example old IS} both show a complete information system.
+   This section studies the transition from an old system to a new one, involving a data migration.
+
    In a datamigration, we distinguish three moments:
@@
          In Ampersand, this means that the developer produces an Ampersand script for the new system.
-         The difference between the new system and the old system is the increment.
+         The {\em increment} is the difference between the new system and the old system.
          The new system must contain the triples from $\dataset$ and $\dataset'$, to preserve the old data.
@@
          yielding migration system $\migrsys'$.
-         The migration engineer does not alter the old system $\infsys$ nor the specification new system $\infsys'$ that the developer has produced.
+         The migration engineer does not alter the old system $\infsys$ nor the specification of the new system $\infsys'$ that the developer has produced.
          As a result, the migration engineer can always restore the old system because it is part of the migration system.
@@
 \begin{definition}[relabel concepts in $\inst$]
-   \[a\ \subst{C}{D}{\inst}\ A\ =\ \text{\bf if}\ C=A\ \text{\bf then}\ a\ \inst\ D\ \text{\bf else}\ a\ \inst\ A\]
+   \[a\ \subst{C}{D}{\inst}\ \subst{C}{D}{A}\ \Leftrightarrow\ a\ \inst\ A\]
 \end{definition}

### 2022-12-23 2022 Migration/articleMigration.tex
redone the analysis chapter

@@
 
-\title{It's all about Meaning!\\{\normalsize A Theory for Data Migration when Deploying Changes}}
+\title{Data migration under a Changing Schema}
 \author[ou,ordina]{Stef Joosten\fnref{fn1}}
@@
    We propose a migration method that features zero down-time,
-   and features an undo after verification by the business.
+   and features an undo after deployment.
    It also leaves room for the business to finish ongoing work
@@
 \begin{keyword}
-generative software\sep incremental software development\sep data migration\sep relation algebra \sep Ampersand
+generative software\sep incremental software development\sep data migration\sep relation algebra\sep Ampersand\sep schema change
 \end{keyword}
@@
    contributing to more frequent and smaller releases.
-   Smaller and faster releases also mean a smaller difference between the old system and the new system.
+   Smaller and faster releases also mean a smaller difference between the old system and the desired system.
    The smaller that difference, the simpler the migration can be and the more of it can be automated.
 
-   In each increment, we distinguish an ``old system'' and a ``new system''.
-   A difference of schemas complicates the migration of data between the old system and the new system.
+   In each increment, we distinguish an ``old system'' and a ``desired system''.
+   A difference of schemas complicates the migration of data between the old system and the desired system.
    The purpose of incremental evolution poses specific requirements to data migration,
@@
    The contribution of this paper is to derive a migration schema from two artifacts: an existing system
-   (which has its own schema and dataset) and a new schema (which specifies the new system).
+   (which has its own schema and dataset) and a new schema (which specifies the desired system).
    The purpose of our theory is to develop tools.
@@
    As a consequence, the data in a system changes continually.
+   In practice, actors ``talk to'' the system through an ingress mechanism, which connects each user to the right service(s).
+   We assume that the ingress function is provided by the deployment platform, so it is beyond the scope of this paper.
+   We also assume that every service runs independent of other services,
+   meaning that the service can be stopped, (re)started and substituted without disrupting the other services.
+
    Every system contains a dataset, which represents the state of the system.
+   Every service produces and consumes events that may change the state of the system.
+   This state is represented in a persistent store, aka the database%
+\footnote{Whether the database is local, remote, or globally distributed is immaterial for the theory in this paper.}.
    Events that the system detects may cause the state to change.
+\begin{figure}[bht]
+   \begin{center}
+     \includegraphics[scale=.45]{datamigration-Pre-migration.drawio.png}
+   \end{center}
+\caption{Anatomy of an information system}
+\label{fig:pre-migration}
+\end{figure}
    To keep our theory technology independent, we will assume that a dataset contains triples.
@@
    such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
+   We will also assume that the system's semantics are represented as constraints,
+   which can be implemented most database management systems as integrity rules that keep the constraints satisfied.
 
-   Data migration occurs when a new system replaces an old one,
-   while preserving the meaning of the data from the old system as much as possible.
-   Just copying the set of data from the old system to the new is obviously wrong if the schemas of both systems differ.
-
-   We will assume that the schema of a system contains constraints on its dataset.
-   These constraints represent the semantics of the system~\cite{Spivak2012}.
-   This allows us to be explicit about ``preserving the meaning as much as possible''.
-   Our approach works with relations in a way similar to Alloy~\cite{Alloy2006} and allegories%
+   In this work, we use the Ampersand compiler~\cite{Joosten-JLAMP2018},
+   to validate our theory.
+   Ampersand is a tool to generate information systems.
+   It works with relations in a way similar to Alloy~\cite{Alloy2006} and allegories%
 \footnote{Allegories are a specific type of categories, which the reader does not need to understand for the purpose of this paper}~\cite{Zielinski2013}.
-
-   In this work we use the Ampersand compiler~\cite{Joosten-JLAMP2018},
-   to generate information systems.
-   A script in the Ampersand language represents the system's semantics by rules,
-   which constrain the dataset.
-   We don't have to formalize the semantics from specifications or from source code
-   because the constraints are already explicitly available as rules in the Ampersand script.
-   A system generated by the Ampersand compiler will keep these constraints satisfied at all times.
-   This makes Ampersand an obvious platform for validating our theory.
+   Each rule in Ampersand is a constraint (aka invariant), which the Ampersand run-time engine keeps satisfied as long as that rule lives%
+\footnote{The dotted lines in the diagrams show which set of rules the engine keeps satisfied.}.
+   So, the user spends no effort to extract constraints from specifications or (even more laborious) from code
+   because all constraints are explicitly available as rules in the code.
+   This, and the absence of imperative code in an Ampersand script, makes Ampersand a suitable platform to validate the theory.
+   It also allows us to be explicit about ``preserving the meaning as much as possible''.
    An Ampersand script contains just enough information to generate a complete system,
@@
 
-   For data migration, complications arise from accumulated data pollution in the old system.
+   Data migration occurs when a desired system replaces an old one,
+   while preserving the meaning of the present data as much as possible~\cite{Spivak2012}.
+   Just copying the set of data from the old system to the desired is obviously wrong if the schemas of both systems differ.
+
+   Complications may arise as a result of data pollution,
+   which may require human interventions to resolve.
    We distinguish data pollution in the following categories:
@@
    The second category, data pollution that is not captured by any of the constraints mentioned in the schema,
-   can be dealt with by adding constraints in the new system.
-   In that case, a migration engineer must ensure that the new system can start with data that satisfies all constraints in its schema.
+   can be dealt with by adding constraints in the desired system.
+   In that case, a migration engineer must ensure that the desired system can start with data that satisfies all constraints in its schema.
    So some human intervention may be necessary in specific cases.
@@
    So, such constraints may help to move this type of pollution to one of the other categories.
-   The last category of data pollution, erroneous constraints, must be fixed by replacing them by the correct constraints in the new system.
+   The last category of data pollution, erroneous constraints, must be fixed by replacing them by the correct constraints in the desired system.
 
+   This illustrates that automating data migrations may require user intervention,
+   either by a migration engineer or by end-users.
+   In our approach, such user interventions are given some time by
+   defining two distinct moments:
+\begin{enumerate}
+   \item the moment of transition (the MoT), i.e. the moment the new system is made available to users;
+   \item the moment of no return (the MoN), i.e. the moment that the migration cannot be undone other than by a backup-procedure%
+\footnote{Backup mechanisms are outside the scope of this paper.}.
+\end{enumerate}
+
+   During the time between the MoT and MoN,
+   we propose to keep the old system and the new system alive, side by side, as shown in figure
 \begin{figure}[bht]
    \begin{center}
-     \includegraphics[scale=.35]{Migration.drawio.png}
+     \includegraphics[scale=.35]{datamigration-Migration phase.drawio.png}
    \end{center}
-\caption{Data migration}
-\label{fig:event flow}
+\caption{Migration phase}
+\label{fig:migration phase}
 \end{figure}
-   This illustrates that automating data migrations has its limits.
-   We cannot eliminate the need for interventions by a migration engineer entirely.
-   If the semantics encoded in the new schema differs from the old schema,
-   a migration engineer sometimes has to make choices.
-   If the interpretation of data changes,
-   a migration engineer drops obsolete constraints from the old system and introduces new constraints in the new system.
-   The consequences of such changes may require changes in data as well.
-   The migration engineer may even want to adapt existing data of which the constraints in the schema have not changed.
-   All of this requires a third schema, which specifies the migration itself.
+
+   The migration requires a third schema, which specifies the migration itself.
+   The migration schema comprises the schemas of both the old system and the new system.
+   The rules of the migration schema will cause the correct data to be transported into the new system%
+\footnote{In this study we focus only on the correctness, not on the efficiency of this data transport}.
    This migration schema preserves data from the old system, it may introduce new data automatically (generated from the old dataset),
    and it may require users to introduce new data (which will take time). 
-   It must contain the old schema, but it also contains the new schema and possibly some new data, all of which a migration engineer must specify.
-   As the migration schema must distinguish between the old dataset and the new dataset,
-   the migration must be based on the disjoint union of the two.
    So the idea of a migration schema is to compute the disjoint union of the old and the new schemas
@@
 %    A naive migration might proceed as follows:
-%    Launch the new system in parallel to the old one, copy data from the old to the new system, and have everyone use the new system.
+%    Launch the desired system in parallel to the old one, copy data from the old to the desired system, and have everyone use the desired system.
 %    However, numerous issues might impede that plan.
@@
 % \begin{enumerate}
-% \item Data required in the new system is missing in the old system.
+% \item Data required in the desired system is missing in the old system.
 %    There may be no way in the old system to enter that data.
@@
 
-%    This issue will require somebody to insert the missing addresses in the new system.
+%    This issue will require somebody to insert the missing addresses in the desired system.
 % \item Data in the old system is wrong but cannot be corrected there due to how the old system was designed.
@@
 
-%    This issue will require that the incorrect registration of staff members is corrected in the new system.
-% \item Data in the old system does not satisfy invariants of the new system.
+%    This issue will require that the incorrect registration of staff members is corrected in the desired system.
+% \item Data in the old system does not satisfy invariants of the desired system.
 %    There may be no way in the old system of making the data satisfy those invariants.
 %    We can use the same example as in the previous bullet point,
-%    but add the requirement (in the new system) that every purchase above a certain amount needs to be approved by the CEO.
+%    but add the requirement (in the desired system) that every purchase above a certain amount needs to be approved by the CEO.
 
-%    This issue will require somebody to change the data to satisfy the broken invariants in the new system.
-% \item A way of entering data into the old system is missing in the new system.
+%    This issue will require somebody to change the data to satisfy the broken invariants in the desired system.
+% \item A way of entering data into the old system is missing in the desired system.
 %    People or automated processes might rely on these ways of entering data.
@@
 %    This issue will require that users of obsolete functionality are informed and given a way to cope with the missing functionality where appropriate.
-% \item Data present in the old system cannot be stored in the new system.
+% \item Data present in the old system cannot be stored in the desired system.
 %    An example could be that references to physical locations where original receipts are kept are stored in the old system,
-%    but the new system relies on scans of receipts and allows the originals to be destroyed or not submitted.
+%    but the desired system relies on scans of receipts and allows the originals to be destroyed or not submitted.
 
@@
 %    \item Allow triples to be migrated while being marked as needing correction.
-%    \item Allow invariants in the new system to be ignored for certain triples during migration.
-%    \item Allow continued use of interfaces of the old system, data entered into the old system via those interfaces needs to be continuously copied to the new system.
+%    \item Allow invariants in the desired system to be ignored for certain triples during migration.
+%    \item Allow continued use of interfaces of the old system, data entered into the old system via those interfaces needs to be continuously copied to the desired system.
 %    \item Retain data in the old system until it can be marked as ready to be phased out.
 %    \end{enumerate}
+
+   During the migration phase, transactions in the old system are allowed because the migration engine will transport them to the new system.
+   This allows the business to finish transactions in the old system while the new system is already up and running.
+   Rules in the migration system that require user interaction are given time by requiring that the migration phase ends
+   only after all old rules and migration rules are at rest.
+   At the MoN, the migration engineer takes down everything but the new system,
+   leaving the business with a successful migration.
+\begin{figure}[bht]
+   \begin{center}
+     \includegraphics[scale=.35]{datamigration-Post-migration.drawio.png}
+   \end{center}
+\caption{The system after the data migration}
+\label{fig:post-migration}
+\end{figure}
 
@@
 \end{verbatim}
-   This declaration introduces a relation with name \verb#takes#,
+   This declaration introduces a relation with the name \verb#takes#,
    source concept \verb#Student#, and
@@
    Sections~\ref{sct:Information Systems} and~\ref{sct:Example old IS} both show a complete information system.
-   This section studies the transition from an old system to a new one, involving a data migration.
-
-   In a datamigration, we distinguish three moments:
-\begin{enumerate}
-   \item the moment of transition, i.e. the moment the old system is replaced (in production) by the new system;
-   \item the decision point, i.e. the moment a business authority declares the migration successful;
-   \item the point of no return, i.e. the moment that the migration cannot be undone other than by a backup-procedure,
-         which is outside the scope of this paper.
-\end{enumerate}
-   The decision point must come after the moment of transition, to allow the business to verify the migration.
-   The point of no return must come after the decision point, to allow for a no-go decision from the business.
+   This section studies the transition from an old system to a desired one, involving a data migration.
 
@@
    \item Let the old system $\infsys$ have state $\dataset$ and schema $\schema$.
-         A developer defines a new system $\infsys'$, which contains a dataset $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.}.
-         In Ampersand, this means that the developer produces an Ampersand script for the new system.
-         The {\em increment} is the difference between the new system and the old system.
-         The new system must contain the triples from $\dataset$ and $\dataset'$, to preserve the old data.
-         If the developer wants to add some new data into the system on the moment of transition,
+         A developer defines a desired system $\infsys'$ by specifying an initial dataset $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.}.
+         Schema $\schema'$ is the schema that governs the system after the migration.
+         After the migration, the system must contain the triples from $\dataset$ and $\dataset'$, to preserve the old data.
+         If the developer wants to add some new data into the system on the moment of transition (MoT),
          she specifies that in $\dataset'$.
@@
          If $\schema$ equals $\schema'$, we call the migration trivial because a migration is not necessary.
-   \item The generator generates a migration system $\migrsys$ based on $\schema$ and $\infsys'$,
-         which gives the user the full functionality of the old system and the new system,
-         i.e.\ $\schema_{\migrsys}=(\schema\sqcup\schema')\cup\id{migrationrules}$.
+   \item The generator generates a migration script $\migrsys$ based on $\schema$ and $\infsys'$,
+         such that $\schema_{\migrsys}=(\schema\sqcup\schema')\cup\id{migrationrules}$.
          The schema $\id{migrationrules}$ represents the enforce rules that migrate the existing data to the new dataset.
@@
    \item A migration engineer changes $\id{migrationrules}$ to accommodate the non-standard requirements of the migration,
-         yielding migration system $\migrsys'$.
-         The migration engineer does not alter the old system $\infsys$ nor the specification of the new system $\infsys'$ that the developer has produced.
+         yielding migration script $\migrsys'$.
+         The migration engineer does not alter the old system $\infsys$ nor the specification of the desired system $\infsys'$ that the developer has produced.
          As a result, the migration engineer can always restore the old system because it is part of the migration system.
    \item The production environment switches instantly from $\infsys$ to $\migrsys'$,
-         while $\dataset_{\infsys}\subseteq\dataset_{\migrsys'}$.
-         This means that the data of the old system remains intact.
-         This event marks the moment of transition.
-   \item The rules in the migration system that are not in $\schema'$ will remain in the system until they are satisfied,
-         to enforce their execution.
-         Automatic migration rules can be executed at the moment of transition.
-         Migration rules that are enforced manually will be held alive until they are satisfied.
-         All rules from $\schema$ (the rules from the old system) are removed on the moment of transition, because the are satisfied by property~\ref{eqn:satisfaction}.
+         while the data of the old system remains intact.
+         This can be realized by configuring the ingress system to open the route to the new system,
+         while keeping the old system accessible.
+         This event marks the MoT.
+         From this point onward, both the old functionality and the desired functionality are available to users.
+   \item The business has the opportunity to evaluate the change, which must end in a go/no-go decision.
+   \item In case of a go-decision, the migration engineer can disable the undo-facility.
+         In case of a no-go, the migration engineer restores the old system.
+         This event marks the moment of no return (MoN)
+   % \item The rules in the migration system that are not in $\schema'$ will remain in the system until they are satisfied,
+   %       to enforce their execution.
+   %       Automatic migration rules can be executed at the MoT.
+   %       Migration rules that are enforced manually will be held alive until they are satisfied.
+   %       All rules from $\schema$ (the rules from the old system) are removed on the MoT, because the are satisfied by property~\ref{eqn:satisfaction}.
    \item After a finite span of time, $\dataset_{\migrsys'}$ satisfies all its enforce rules.
@@
 \subsection{Example}
-   Let us proceed to specify a new system, $\infsys'$, to illustrate a (toy) migration.
+   Let us proceed to specify a desired system, $\infsys'$, to illustrate a (toy) migration.
    We will use the example in section~\ref{old IS} as the old system
@@
 
-   Information system $\infsys'$, the new system, is specified by:
+   Information system $\infsys'$, the desired system, is specified by:
 \begin{verbatim}
@@
 
-We allow ExamRule2 to be broken during the transition to the new system, and expect people with the Administrator role to deal with them.
+We allow ExamRule2 to be broken during the transition to the desired system, and expect people with the Administrator role to deal with them.
 
@@
 
-One solution is to state that the only violations that may occur in \verb=isPartOfUNI= are those that were present when the migration started:
+One solution is to state that the only violations that may occur in the relation \verb=isPartOfUNI= are those that were present when the migration started:
 \begin{verbatim}
@@
      = [("IT-Governance", "IT-Governance")]
-   RULE isPartOfUNI_progress: -(isPartOf;isPartOf~) /\ I |- isPartOfUNI_violations
+   RULE isPartOfUNI_progress: -(isPartOf;isPartOf~) /\ I
+         |- isPartOfUNI_violations
    SYSTEM MAINTAINS isPartOfUNI_progress
@@
 
-   RULE isPartOf_isUnion:  isPartOf_old :- isPartOf \/ isPartOf_ignored
+   RULE isPartOf_isUnion:
+         isPartOf_old |- isPartOf \/ isPartOf_ignored
    ROLE migrationHelper MAINTAINS isPartOf_isUnion
@@
 
-The ${\tt CLASSIFY\ }(C',1){\tt\ IS\ }(C,0)$ statements add $((C',1),(C,0))$ and  $((C,0),(C',1))$ to the $\isa$ relation.
-Note that if $\dataset$ is a dataset, and $\dataset'$ is that dataset with pairs added\footnote{TODO: het woord `add' is informeel, dit kan beter} to the $\isa$ relation, then the result is again a dataset.
+The ${\tt CLASSIFY\ }(C',1){\tt\ IS\ }(C,0)$ statement adds two pairs to the $\isa$ relation:
+$\pair{(C',1)}{(C,0)}$ and $\pair{(C,0)}{(C',1)}$.
+Note that if $\dataset$ is a dataset, and $\dataset'$ is that dataset with pairs added%
+\footnote{TODO: het woord `add' is informeel, dit kan beter} to the $\isa$ relation, then the result is again a dataset.
 The ${\tt ENFORCE\ }r{\tt\ :=\ }e$ statements add triples $(x,r,y)$ for all $(x,y)\in e$.
@@
    % -> er is een pad naar ingebruikname van het nieuwe systeem (vanaf de initiële toestand van het migratie-systeem)
-   \item there is a path to releasing the new system, starting with the initial state of the $\migrsys$.
+   \item there is a path to releasing the desired system, starting with the initial state of the $\migrsys$.
    % -> corollary: op het moment van ingebruikname van het nieuwe systeem, is het migratie-systeem vrij van overtredingen op regels die het nieuwe systeem moet bewaken
-   \item As a result, the new system $\infsys'$ is void of violations at the start of its production life.
+   \item As a result, the desired system $\infsys'$ is void of violations at the start of its production life.
    % -> optioneel: na een begrensd aantal 'voortgangs-stappen' kan het nieuwe systeem in gebruik genomen worden
-   \item Optionally, the new system can be used in production after a limited number of steps.
+   \item Optionally, the desired system can be used in production after a limited number of steps.
    % -> optioneel: er bestaat een migratie-systeem dat de oude functionaliteit behoudt (mogelijk uitbreidt) totdat het nieuwe systeem in gebruik genomen is?
-   \item Optionally, there is a migration systen that preserves the functionality of the old system until the new system has been put in production.
+   \item Optionally, there is a migration systen that preserves the functionality of the old system until the desired system has been put in production.
 \end{itemize}

### 2022-12-24 2022 Migration/articleMigration.tex
WIP

@@
    contributing to more frequent and smaller releases.
-   Smaller and faster releases also mean a smaller difference between the old system and the desired system.
+   Smaller and faster releases also mean a smaller difference between the existing system and the desired system.
    The smaller that difference, the simpler the migration can be and the more of it can be automated.
 
-   In each increment, we distinguish an ``old system'' and a ``desired system''.
-   A difference of schemas complicates the migration of data between the old system and the desired system.
+   In each increment, we distinguish an ``existing system'' and a ``desired system''.
+   A difference of schemas complicates the migration of data between the existing system and the desired system.
    The purpose of incremental evolution poses specific requirements to data migration,
@@
    They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
-   In our work, however, we want to get rid of the old system and not retain it.
+   In our work, however, we want to get rid of the existing system and not retain it.
    So, we need to look at a different approach.
@@
    For our theory, zero down-time means that all human activity for the data migration can be done without taking the system out of production.
-   We consider any delay caused by automatically switching from the old system to the new one negligible,
+   We consider any delay caused by automatically switching from the existing system to the desired one negligible,
    assuming a modern deployment platform that does this with zero down-time.
@@
    \item the data migration is meant to deploy a software increment in production;
-   \item the old data set may be polluted, but it satisfies its schema;
+   \item the existing data set may be polluted, but it satisfies its schema;
    \item the data migration may require human interaction, which may take time;
@@
 \end{itemize}
+   The theory we develop focuses on the correctness of the migration.
+   Efficiency of the migration is beyond the scope of this paper.
 
@@
 
-   Data migration occurs when a desired system replaces an old one,
+   Data migration occurs when a desired system replaces an existing one,
    while preserving the meaning of the present data as much as possible~\cite{Spivak2012}.
-   Just copying the set of data from the old system to the desired is obviously wrong if the schemas of both systems differ.
+   Just copying the set of data from the existing system to the desired is obviously wrong if the schemas of both systems differ.
 
@@
    \item data pollution that cannot be captured in constraints on the dataset,
-   such as a street address that has become obsolete because the system never registered the change of address.
+   such as a street address that has become obsolete because someone has moved without notification.
    \item data pollution that is a consequence of an erroneous constraint in the schema.
@@
    The first category is about violations of constraints from the schema.
-   Such violations do not occur in our theory because the old system was generated by Ampersand,
-   so the old system has kept all constraints from the schema satisfied.
+   Such violations do not occur in our theory because the existing system was generated by Ampersand,
+   so the existing system has kept all constraints from the schema satisfied.
    This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
@@
 \begin{enumerate}
-   \item the moment of transition (the MoT), i.e. the moment the new system is made available to users;
+   \item the moment of transition (the MoT), i.e. the moment the desired system is made available to users;
    \item the moment of no return (the MoN), i.e. the moment that the migration cannot be undone other than by a backup-procedure%
@@
    During the time between the MoT and MoN,
-   we propose to keep the old system and the new system alive, side by side, as shown in figure
+   we propose to keep the existing system and the desired system alive, side by side, as shown in figure~\ref{fig:migration phase}.
 \begin{figure}[bht]
@@
    The migration requires a third schema, which specifies the migration itself.
-   The migration schema comprises the schemas of both the old system and the new system.
-   The rules of the migration schema will cause the correct data to be transported into the new system%
-\footnote{In this study we focus only on the correctness, not on the efficiency of this data transport}.
-   This migration schema preserves data from the old system, it may introduce new data automatically (generated from the old dataset),
+   The migration schema comprises the schemas of both the existing system and the desired system.
+   The rules of the migration schema will cause the correct data to be transported into the desired system.
+   This migration schema preserves data from the existing system, it may introduce new data automatically (generated from the existing dataset),
    and it may require users to introduce new data (which will take time). 
-   So the idea of a migration schema is to compute the disjoint union of the old and the new schemas
+   So the idea of a migration schema is to compute the disjoint union of the existing and the desired schemas
    and to present it to the migration engineer in the form of source code,
@@
 %    A naive migration might proceed as follows:
-%    Launch the desired system in parallel to the old one, copy data from the old to the desired system, and have everyone use the desired system.
+%    Launch the desired system in parallel to the existing one, copy data from the existing to the desired system, and have everyone use the desired system.
 %    However, numerous issues might impede that plan.
@@
 % \begin{enumerate}
-% \item Data required in the desired system is missing in the old system.
-%    There may be no way in the old system to enter that data.
-%    An example could be that every reimbursement form needs to have an address associated to it to mail the check to, but address information is not stored in the old system:
-%    The old system required the reimbursement office to look up employee's addresses from a hand-written list they had on their desk.
+% \item Data required in the desired system is missing in the existing system.
+%    There may be no way in the existing system to enter that data.
+%    An example could be that every reimbursement form needs to have an address associated to it to mail the check to, but address information is not stored in the existing system:
+%    The existing system required the reimbursement office to look up employee's addresses from a hand-written list they had on their desk.
 
 %    This issue will require somebody to insert the missing addresses in the desired system.
-% \item Data in the old system is wrong but cannot be corrected there due to how the old system was designed.
-%    An example is if the old system only allows approvals to be entered as the current user, and the CEO has always insisted that her administrative staff enters the approvals into the system for her.
+% \item Data in the existing system is wrong but cannot be corrected there due to how the existing system was designed.
+%    An example is if the existing system only allows approvals to be entered as the current user, and the CEO has always insisted that her administrative staff enters the approvals into the system for her.
 %    This may result in approvals being entered as admin staff, where it was really the CEO making the approval.
@@
 %    This issue will require that the incorrect registration of staff members is corrected in the desired system.
-% \item Data in the old system does not satisfy invariants of the desired system.
-%    There may be no way in the old system of making the data satisfy those invariants.
+% \item Data in the existing system does not satisfy invariants of the desired system.
+%    There may be no way in the existing system of making the data satisfy those invariants.
 %    We can use the same example as in the previous bullet point,
@@
 %    This issue will require somebody to change the data to satisfy the broken invariants in the desired system.
-% \item A way of entering data into the old system is missing in the desired system.
+% \item A way of entering data into the existing system is missing in the desired system.
 %    People or automated processes might rely on these ways of entering data.
@@
 %    This issue will require that users of obsolete functionality are informed and given a way to cope with the missing functionality where appropriate.
-% \item Data present in the old system cannot be stored in the desired system.
-%    An example could be that references to physical locations where original receipts are kept are stored in the old system,
+% \item Data present in the existing system cannot be stored in the desired system.
+%    An example could be that references to physical locations where original receipts are kept are stored in the existing system,
 %    but the desired system relies on scans of receipts and allows the originals to be destroyed or not submitted.
 
-%    This issue will require that the old data is kept until the original receipts have been scanned.
+%    This issue will require that the existing data is kept until the original receipts have been scanned.
 % \end{enumerate}
@@
 %    \item Allow invariants in the desired system to be ignored for certain triples during migration.
-%    \item Allow continued use of interfaces of the old system, data entered into the old system via those interfaces needs to be continuously copied to the desired system.
-%    \item Retain data in the old system until it can be marked as ready to be phased out.
+%    \item Allow continued use of interfaces of the existing system, data entered into the existing system via those interfaces needs to be continuously copied to the desired system.
+%    \item Retain data in the existing system until it can be marked as ready to be phased out.
 %    \end{enumerate}
 
-   During the migration phase, transactions in the old system are allowed because the migration engine will transport them to the new system.
-   This allows the business to finish transactions in the old system while the new system is already up and running.
+   During the migration phase, transactions in the existing system are allowed because the migration engine will transport them to the desired system.
+   This allows the business to finish transactions in the existing system while the desired system is already up and running.
+   Similarly, the existing system must execute transactions from the desired system, just in case the business calls off the migration.
    Rules in the migration system that require user interaction are given time by requiring that the migration phase ends
-   only after all old rules and migration rules are at rest.
-   At the MoN, the migration engineer takes down everything but the new system,
+   only after all migration rules are at rest.
+   At the MoN, the migration engineer takes down everything but the desired system,
    leaving the business with a successful migration.
@@
 \label{sct:Terminology}
-   An ``information system'' is a combination of dataset, schema, and functionality.
+   An {\em information system} is a combination of dataset, schema, and functionality.
    For the purpose of this paper, we ignore the functionality because it does not impact the migration.
@@
 
+   As a result, if an atom is an instance of concept $A$ and $A\isa B$,
+   this atom has all properties that atoms of type $B$ have:
+\begin{equation}
+   \begin{array}[b]{ll}
+      &a\ \inst\ A\wedge b\ \inst\ B\wedge A\isa B\\
+      \wedge&a=b\\
+      \wedge&b\ r\ x
+   \end{array}
+      \ \Rightarrow\ a\ r\ x
+\end{equation}
+
 \subsection{Information Systems}
@@
 \subsection{Example}
-\label{sct:Example old IS}
-\label{old IS}
+\label{sct:Example existing IS}
+\label{existing IS}
    Having defined an information system in mathematical terms, let us discuss a small example.
@@
 \subsection{Data migration}
-   Sections~\ref{sct:Information Systems} and~\ref{sct:Example old IS} both show a complete information system.
-   This section studies the transition from an old system to a desired one, involving a data migration.
+   Sections~\ref{sct:Information Systems} and~\ref{sct:Example existing IS} both show a complete information system.
+   This section studies the transition from an existing system to a desired one, involving a data migration.
 
@@
 \begin{enumerate}
-   \item Let the old system $\infsys$ have state $\dataset$ and schema $\schema$.
+   \item Let the existing system $\infsys$ have state $\dataset$ and schema $\schema$.
          A developer defines a desired system $\infsys'$ by specifying an initial dataset $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.}.
          Schema $\schema'$ is the schema that governs the system after the migration.
-         After the migration, the system must contain the triples from $\dataset$ and $\dataset'$, to preserve the old data.
+         After the migration, the system must contain the triples from $\dataset$ and $\dataset'$, to preserve the existing data.
          If the developer wants to add some new data into the system on the moment of transition (MoT),
@@
          such that $\schema_{\migrsys}=(\schema\sqcup\schema')\cup\id{migrationrules}$.
-         The schema $\id{migrationrules}$ represents the enforce rules that migrate the existing data to the new dataset.
+         The schema $\id{migrationrules}$ represents the enforce rules that migrate the existing data to the desired dataset.
          TODO: $\id{migrationrules}$ bevat alleen tijdelijke datastructuur, indien nodig.
@@
          yielding migration script $\migrsys'$.
-         The migration engineer does not alter the old system $\infsys$ nor the specification of the desired system $\infsys'$ that the developer has produced.
-         As a result, the migration engineer can always restore the old system because it is part of the migration system.
+         The migration engineer does not alter the existing system $\infsys$ nor the specification of the desired system $\infsys'$ that the developer has produced.
+         As a result, the migration engineer can always restore the existing system because it is part of the migration system.
    \item The production environment switches instantly from $\infsys$ to $\migrsys'$,
-         while the data of the old system remains intact.
-         This can be realized by configuring the ingress system to open the route to the new system,
-         while keeping the old system accessible.
+         while the data of the existing system remains intact.
+         This can be realized by configuring the ingress system to open the route to the desired system,
+         while keeping the existing system accessible.
          This event marks the MoT.
-         From this point onward, both the old functionality and the desired functionality are available to users.
+         From this point onward, both the existing functionality and the desired functionality are available to users.
    \item The business has the opportunity to evaluate the change, which must end in a go/no-go decision.
    \item In case of a go-decision, the migration engineer can disable the undo-facility.
-         In case of a no-go, the migration engineer restores the old system.
+         In case of a no-go, the migration engineer restores the existing system.
          This event marks the moment of no return (MoN)
@@
    %       Migration rules that are enforced manually will be held alive until they are satisfied.
-   %       All rules from $\schema$ (the rules from the old system) are removed on the MoT, because the are satisfied by property~\ref{eqn:satisfaction}.
+   %       All rules from $\schema$ (the rules from the existing system) are removed on the MoT, because the are satisfied by property~\ref{eqn:satisfaction}.
    \item After a finite span of time, $\dataset_{\migrsys'}$ satisfies all its enforce rules.
-         This implies that the generated rules plus the changes made by the migration engineer ensure the preservation of the meaning of the old data.
+         This implies that the generated rules plus the changes made by the migration engineer ensure the preservation of the meaning of the existing data.
    \item Users may have to perform manual migration steps.
@@
          The system ensures that the amount of work needed to perform these steps is finite.
-   \item The patron of the system can decide whether to carry on with the migration or to roll back to the old system.
+   \item The patron of the system can decide whether to carry on with the migration or to roll back to the existing system.
          She can do that incrementally by switching off the interfaces one-by-one.
@@
    To illustrate this point, suppose $D\isa A$.
-   Then the substitution $[C\rightarrow D]$ means that every atom $a$ that used to be of concept $C$ in the old system, is now not only a $D$ but also an $A$.
+   Then the substitution $[C\rightarrow D]$ means that every atom $a$ that used to be of concept $C$ in the existing system, is now not only a $D$ but also an $A$.
    For example by relabeling the concept {\tt Hotel} to {\tt Motel}, in a situation where ${\tt Motel}\isa{\tt Parking}$,
@@
    Let us proceed to specify a desired system, $\infsys'$, to illustrate a (toy) migration.
-   We will use the example in section~\ref{old IS} as the old system
+   We will use the example in section~\ref{sct:Example existing IS} as the existing system
    and call that $\infsys$ in this section.
@@
 \end{verbatim}
-   Any data migration must anticipate any population in the old system.
-   The only thing we know for sure about the ``old'' population is that it satisfies the rules in $\infsys$.
+   Any data migration must anticipate any population in the existing system.
+   The only thing we know for sure about the existing population is that it satisfies the rules in $\infsys$.
 
@@
 This in itself is not an issue: problems need to be solved one way or another.
-However, a problem that may occur is that regular users of the system, who are used to the old system, might be adding data to the system that causes violations faster than they can be solved.
+However, a problem that may occur is that regular users of the system, who are used to the existing system, might be adding data to the system that causes violations faster than they can be solved.
 To mitigate this, we propose two solutions.
@@
 As a second solution, we observe that the data in the designed system does not have any violations.
-Consequentially, the cause of the violations comes from the triples in the old system.
+Consequentially, the cause of the violations comes from the triples in the existing system.
 Rather than taking the union of all triples, we can take the disjoint union instead;
@@
    % -> het migratie-systeem bevat alle oude data, ihb nog steeds na het toepassen van de enforce regels
-   \item $\migrsys$ contains the old data, insofar its enforce rules have not deleted it.
+   \item $\migrsys$ contains the existing data, insofar its enforce rules have not deleted it.
    % -> er is een pad naar ingebruikname van het nieuwe systeem (vanaf de initiële toestand van het migratie-systeem)
@@
    % -> optioneel: er bestaat een migratie-systeem dat de oude functionaliteit behoudt (mogelijk uitbreidt) totdat het nieuwe systeem in gebruik genomen is?
-   \item Optionally, there is a migration systen that preserves the functionality of the old system until the desired system has been put in production.
+   \item Optionally, there is a migration systen that preserves the functionality of the existing system until the desired system has been put in production.
 \end{itemize}

### 2022-12-25 2022 Migration/articleMigration.tex
WIP

@@
 \usepackage{amssymb}
+\usepackage{mathrsfs}
 \usepackage{amsthm}
@@
 \label{sct:Example existing IS}
-\label{existing IS}
    Having defined an information system in mathematical terms, let us discuss a small example.
@@
          Schema $\schema'$ is the schema that governs the system after the migration.
-         After the migration, the system must contain the triples from $\dataset$ and $\dataset'$, to preserve the existing data.
+         After the migration, the system must contain the triples from both $\dataset$ and $\dataset'$, to preserve the existing data.
          If the developer wants to add some new data into the system on the moment of transition (MoT),
@@
          If $\schema$ equals $\schema'$, we call the migration trivial because a migration is not necessary.
-   \item The generator generates a migration script $\migrsys$ based on $\schema$ and $\infsys'$,
-         such that $\schema_{\migrsys}=(\schema\sqcup\schema')\cup\id{migrationrules}$.
+   \item The generator generates a migration script $\migrsys$ based on $\schema$ and $\infsys'$
+         to migrate the data and define the semantics during the migration phase as far as a generator can reasonably predict.
          The schema $\id{migrationrules}$ represents the enforce rules that migrate the existing data to the desired dataset.
-         TODO: $\id{migrationrules}$ bevat alleen tijdelijke datastructuur, indien nodig.
    \item A migration engineer changes $\id{migrationrules}$ to accommodate the non-standard requirements of the migration,
          yielding migration script $\migrsys'$.
+         This schema also contains data structures and rules that are needed for the duration of the migration phase.
          The migration engineer does not alter the existing system $\infsys$ nor the specification of the desired system $\infsys'$ that the developer has produced.
@@
          From this point onward, both the existing functionality and the desired functionality are available to users.
+   \item The migration system $\dataset_{\migrsys'}$ must be at rest before the MoN,
+         to ensure that the migration has completed before the migration becomes irreversible.
    \item The business has the opportunity to evaluate the change, which must end in a go/no-go decision.
-   \item In case of a go-decision, the migration engineer can disable the undo-facility.
+   \item In case of a go-decision, the migration engineer can disable the undo-facility as soon as the migration system is at rest.
          In case of a no-go, the migration engineer restores the existing system.
          This event marks the moment of no return (MoN)
-   % \item The rules in the migration system that are not in $\schema'$ will remain in the system until they are satisfied,
-   %       to enforce their execution.
-   %       Automatic migration rules can be executed at the MoT.
-   %       Migration rules that are enforced manually will be held alive until they are satisfied.
-   %       All rules from $\schema$ (the rules from the existing system) are removed on the MoT, because the are satisfied by property~\ref{eqn:satisfaction}.
-   \item After a finite span of time, $\dataset_{\migrsys'}$ satisfies all its enforce rules.
-         This implies that the generated rules plus the changes made by the migration engineer ensure the preservation of the meaning of the existing data.
    \item Users may have to perform manual migration steps.
@@
       &=&\begin{array}[t]{l@{}l}
-         \la&\{\subst{C}{D}{t}\mid t\in\triples\}\\
+         \la&\{\subst{C}{D}{c}\mid c\in\concepts\}\\
          ,&\{\subst{C}{D}{r}\mid r\in\rels\}\\

### 2022-12-27 2022 Migration/articleMigration.tex
refined the migration steps.

@@
    \item Let the existing system $\infsys$ have state $\dataset$ and schema $\schema$.
-         A developer defines a desired system $\infsys'$ by specifying an initial dataset $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.}.
-         Schema $\schema'$ is the schema that governs the system after the migration.
-         After the migration, the system must contain the triples from both $\dataset$ and $\dataset'$, to preserve the existing data.
-         If the developer wants to add some new data into the system on the moment of transition (MoT),
-         she specifies that in $\dataset'$.
-         That might be necessary to initialize specific new features.
+         A developer defines a desired system $\infsys'$, which contains a dataset of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.} as illustrated in figure~\ref{fig:post-migration}.
+         After the migration, the system will contain the triples from $\dataset$, to preserve the existing data.
+         The developer may want to add some new data into the system,
+         for instance to initialize new features on the moment of transition (MoT).
+         She specifies such data in $\dataset'$.
          If $\schema$ equals $\schema'$, we call the migration trivial because a migration is not necessary.
@@
          This event marks the MoT.
-         From this point onward, both the existing functionality and the desired functionality are available to users.
-   \item The migration system $\dataset_{\migrsys'}$ must be at rest before the MoN,
-         to ensure that the migration has completed before the migration becomes irreversible.
-   \item The business has the opportunity to evaluate the change, which must end in a go/no-go decision.
+         From this point onward, both the existing functionality and the desired functionality are available to users as illustrated in figure~\ref{fig:migration phase}.
+   \item To ensure that all data is migrated before the the migration engineer makes the migration irreversible,
+         we require that the rules of the the migration system $\rules_{\migrsys'}$ are at rest before the migration engineer takes down the existing system.
+         Users (i.e.\ the business) must do the work to satisfy the manually enforced rules.
+         Ampersand guarantees that enforce rules will eventually be satisfied.
+   \item To give the business an opportunity to evaluate the change,
+         we can require a go/no-go decision before the the migration engineer makes the migration irreversible.
    \item In case of a go-decision, the migration engineer can disable the undo-facility as soon as the migration system is at rest.
          In case of a no-go, the migration engineer restores the existing system.
-         This event marks the moment of no return (MoN)
-   \item Users may have to perform manual migration steps.
-         These are specified as process rules in $\id{migrationrules}$.
-         % Dit zijn "kurk" regels.
-         The system ensures that the amount of work needed to perform these steps is finite.
-   \item The patron of the system can decide whether to carry on with the migration or to roll back to the existing system.
-         She can do that incrementally by switching off the interfaces one-by-one.
+         This event marks the moment of no return (MoN).
    \item The system retains the data, the schema, and the functionality that is covered by $\schema'$, dropping everything else.
+         This establishes the post-migration situation (figure~\ref{fig:post-migration}).
 \end{enumerate}

### 2022-12-27 2022 Migration/articleMigration.tex
substitutions for relations added

@@
 
-\subsection{Data migration}
-   Sections~\ref{sct:Information Systems} and~\ref{sct:Example existing IS} both show a complete information system.
-   This section studies the transition from an existing system to a desired one, involving a data migration.
-
-   To migrate system $\infsys$ to $\infsys'$, we take the following steps:
+\subsection{Process of Data Migration}
+   Sections~\ref{sct:Information Systems} and~\ref{sct:Example existing IS} introduce a single information system.
+   The transition from an existing system to the desired one is the topic of this section.
+   Let us first sketch the steps to migrate system $\infsys$ to $\infsys'$ informally.
 \begin{enumerate}
@@
          She specifies such data in $\dataset'$.
-         If $\schema$ equals $\schema'$, we call the migration trivial because a migration is not necessary.
+         If $\schema$ equals $\schema'$, we call the migration trivial because a data migration is not necessary.
    \item The generator generates a migration script $\migrsys$ based on $\schema$ and $\infsys'$
          to migrate the data and define the semantics during the migration phase as far as a generator can reasonably predict.
-         The schema $\id{migrationrules}$ represents the enforce rules that migrate the existing data to the desired dataset.
+         It generates a schema $\id{migrationrules}$, which contains the enforce rules that migrate existing data to the desired dataset.
    \item A migration engineer changes $\id{migrationrules}$ to accommodate the non-standard requirements of the migration,
          yielding migration script $\migrsys'$.
-         This schema also contains data structures and rules that are needed for the duration of the migration phase.
+         This script also contains data structures and rules that are needed for the duration of the migration phase.
          The migration engineer does not alter the existing system $\infsys$ nor the specification of the desired system $\infsys'$ that the developer has produced.
-         As a result, the migration engineer can always restore the existing system because it is part of the migration system.
-   \item The production environment switches instantly from $\infsys$ to $\migrsys'$,
-         while the data of the existing system remains intact.
-         This can be realized by configuring the ingress system to open the route to the desired system,
+         As a result, the migration engineer can always restore the existing system.
+   \item The production environment switches instantly from $\infsys$ to $\migrsys'$.
+         This can be realized by configuring the ingress system to open the route to the migration system and the desired system,
          while keeping the existing system accessible.
@@
    \item To ensure that all data is migrated before the the migration engineer makes the migration irreversible,
-         we require that the rules of the the migration system $\rules_{\migrsys'}$ are at rest before the migration engineer takes down the existing system.
+         we require that every rule from $\rules_{\migrsys'}$ is satisfied before the migration engineer takes down the existing system.
          Users (i.e.\ the business) must do the work to satisfy the manually enforced rules.
-         Ampersand guarantees that enforce rules will eventually be satisfied.
+         Ampersand guarantees that enforce rules will be satisfied.
    \item To give the business an opportunity to evaluate the change,
          we can require a go/no-go decision before the the migration engineer makes the migration irreversible.
-   \item In case of a go-decision, the migration engineer can disable the undo-facility as soon as the migration system is at rest.
+         In case of a go-decision, the migration engineer can disable the undo-facility as soon as the migration rules are satisfied.
          In case of a no-go, the migration engineer restores the existing system.
          This event marks the moment of no return (MoN).
-   \item The system retains the data, the schema, and the functionality that is covered by $\schema'$, dropping everything else.
+   \item The system retains the data, the schema, and the functionality that is covered by $\schema'$ by dropping everything else.
          This establishes the post-migration situation (figure~\ref{fig:post-migration}).
@@
 
+\subsection{Formalization}
    Let us define the instruments needed to describe the derivation of $\schema_{\migrsys}$.
@@
          ,&\{\subst{C}{D}{u}\mid u\in\rules\}\ \ra\notag
+         \end{array}
+   \end{eqnarray}
+\end{definition}
+\begin{definition}[relabel relations]
+   \[\subst{\declare{n}{A}{B}}{r}{{\declare{m}{C}{D}}}\ =\ \text{\bf if}\ n=m\wedge A=C\wedge B=D\ \text{\bf then}\ r\ \text{\bf else}\ \declare{m}{C}{D}\]
+\end{definition}
+\begin{definition}[relabel relations in triples]
+   \[\subst{p}{q}{\triple{a}{r}{b}} = \triple{a}{\subst{p}{q}{r}}{b}\]
+\end{definition}
+\begin{definition}[relabel relations in datasets]
+   \[\begin{array}{rcl}
+      \rlap{$\subst{p}{q}{\pair{\triples}{\inst}}$}\\
+      &=&\pair{\{\subst{p}{q}{t}\mid t\in\triples\}}{\inst}
+   \end{array}\]
+\end{definition}
+\begin{definition}[relabel relations in rules]
+   \[\sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{p}{q}{u}}{\subst{p}{q}{\dataset}}\]
+\end{definition}
+\begin{definition}[relabel relations in schemas]
+   \begin{eqnarray}
+      \rlap{$\subst{\declare{n}{A}{B}}{{\declare{m}{C}{D}}}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
+      &=&\begin{array}[t]{l@{}l}
+         \la&\{\subst{B}{D}{{\subst{A}{C}{c}}}\mid c\in\concepts\}\\
+         ,&\{\subst{B}{D}{{\subst{A}{C}{{\subst{\declare{n}{A}{B}}{{\declare{m}{C}{D}}}{r}}}}}\mid r\in\rels\}\\
+         ,&\{\subst{B}{D}{{\subst{A}{C}{{\subst{\declare{n}{A}{B}}{{\declare{m}{C}{D}}}{u}}}}}\mid u\in\rules\}\ \ra\notag
          \end{array}

### 2023-01-06 2022 Migration/articleMigration.tex
fixed rename of relations
Adapted the migration process in section 3.5.

--- 2022-12-27 2022 Migration/articleMigration.tex
@@
 \end{eqnarray}
-
+   If a triple $\triple{a}{r}{b}$ satisfies requirement~\ref{eqn:define R} and its relation $r$ satisfies requirement~{eqn:relationsIntroduceConcepts},
+   we say that it {\em befits} schema $\triple{\concepts}{\rels}{\rules}$.
    We assume a special kind of rule that is used for specialization.
@@
    Now let us check the definition of information system.
-   Ampersand generates a set of roles $\roles=\{{\tt Administrator}\}$ and
-   a relation $\maintain$, which contains one pair only: $\pair{{\tt Administrator}}{{\tt EnrollRule}}$.
+   Ampersand generates a set of roles $\roles=\{{\tt Administrator}\}$.
+   The relation $\maintain$ contains one pair only, $\pair{{\tt Administrator}}{{\tt EnrollRule}}$,
+   instructing the run-time engine to forward all violations of {\tt EnrollRule} to an administrator so she can keep {\tt EnrollRule} satisfied.
    Requirement~\ref{eqn:satisfaction} is satisfied because the only rule, {\tt EnrollRule} is satisfied.
@@
 
-\subsection{Data migration}
-   Sections~\ref{sct:Information Systems} and~\ref{sct:Example existing IS} both show a complete information system.
-   This section studies the transition from an existing system to a desired one, involving a data migration.
-
-   To migrate system $\infsys$ to $\infsys'$, we take the following steps:
+\subsection{Process of Data Migration}
+   Sections~\ref{sct:Information Systems} and~\ref{sct:Example existing IS} introduce a single information system.
+   The transition from an existing system to the desired one is the topic of this section.
+   Let us first sketch the steps to migrate system $\infsys$ to $\infsys'$.
 \begin{enumerate}
@@
          A developer defines a desired system $\infsys'$, which contains a dataset of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.} as illustrated in figure~\ref{fig:post-migration}.
-         After the migration, the system will contain the triples from $\dataset$, to preserve the existing data.
-         The developer may want to add some new data into the system,
+         After the migration, the system will contain the triples from $\dataset$ that befit $\schema'$,
+         except if the migration engineer decides otherwise (see step~\ref{item3}).
+         $\dataset'$ may contain some new data,
          for instance to initialize new features on the moment of transition (MoT).
-         She specifies such data in $\dataset'$.
-         If $\schema$ equals $\schema'$, we call the migration trivial because a migration is not necessary.
-   \item The generator generates a migration script $\migrsys$ based on $\schema$ and $\infsys'$
+         If schemas $\schema$ equals $\schema'$, we call the migration trivial because a data migration is not necessary.
+   \item The generator generates a migration script $\migrsys$
          to migrate the data and define the semantics during the migration phase as far as a generator can reasonably predict.
-         The schema $\id{migrationrules}$ represents the enforce rules that migrate the existing data to the desired dataset.
-   \item A migration engineer changes $\id{migrationrules}$ to accommodate the non-standard requirements of the migration,
+         It generates a schema, $\schema_{\migrsys}$, which contains the enforce rules that migrate existing data to the desired dataset.
+         It also contains data structures and rules that are needed for the duration of the migration phase.
+   \item\label{item3}
+         A migration engineer changes $\migrsys$ to accommodate the non-standard requirements of the migration,
          yielding migration script $\migrsys'$.
-         This schema also contains data structures and rules that are needed for the duration of the migration phase.
-         The migration engineer does not alter the existing system $\infsys$ nor the specification of the desired system $\infsys'$ that the developer has produced.
-         As a result, the migration engineer can always restore the existing system because it is part of the migration system.
-   \item The production environment switches instantly from $\infsys$ to $\migrsys'$,
-         while the data of the existing system remains intact.
-         This can be realized by configuring the ingress system to open the route to the desired system,
-         while keeping the existing system accessible.
+         The migration engineer does not alter the existing system $\infsys$ nor the specification
+         of the desired system $\infsys'$ that the developer has produced.
+   \item The migration engineer deploys system $\migrsys'$.
+         This event starts the copying of data from $\dataset$ to $\dataset'$ (figure~\ref{fig:migration phase}).
+         During this phase, the ingress is still referring users to the existing system,
+         so they will not notice that the desired system is being fired up and loaded with data.
+         Users are still changing $\dataset$ (both inserts and deletes),
+         but these changes are being transferred to $\dataset'$ as well.
+         This step is complete when all data from $\dataset$ has been copied to $\dataset'$ and
+         both datasets are in sync with respect to the ongoing transactions.
+   \item At this point, the ingress switches the traffic from $\infsys$ to $\migrsys'$,
+         so users will now experience the new functionality.
          This event marks the MoT.
-         From this point onward, both the existing functionality and the desired functionality are available to users as illustrated in figure~\ref{fig:migration phase}.
-   \item To ensure that all data is migrated before the the migration engineer makes the migration irreversible,
-         we require that the rules of the the migration system $\rules_{\migrsys'}$ are at rest before the migration engineer takes down the existing system.
-         Users (i.e.\ the business) must do the work to satisfy the manually enforced rules.
-         Ampersand guarantees that enforce rules will eventually be satisfied.
+         From this point onward, users work with the desired functionality in dataset $\dataset'$.
    \item To give the business an opportunity to evaluate the change,
          we can require a go/no-go decision before the the migration engineer makes the migration irreversible.
-   \item In case of a go-decision, the migration engineer can disable the undo-facility as soon as the migration system is at rest.
-         In case of a no-go, the migration engineer restores the existing system.
          This event marks the moment of no return (MoN).
-   \item The system retains the data, the schema, and the functionality that is covered by $\schema'$, dropping everything else.
+         In case of a go-decision, the migration engineer can take down the existing system, which by this time has become unreachable for users.
+         In case of a no-go, the migration engineer restores the existing system by letting the ingress switch traffic back to $\infsys$.
+   \item In more complex cases, the migration system contains rules that must be satisfied during the migration.
+         These rules can be necessary, for example to resolve data pollution that violates a constraint that was not in schema $\schema_{\infsys}$.
+         By putting it in the schema $\schema_{\migrsys'}$,
+         the migration engineer can fix such problems.
+         She can do this in the form of an enforce rule, which means that the problem gets fixed automatically.
+         Or she can use the business to satisfy this rule, in case human intelligence is required.
+         Whatever the choice, we require that every rule from $\rules_{\migrsys'}$ is satisfied before the migration engineer takes down the migration system.
+   \item After taking down the existing system and the migration system, the desired system remains in production and the migration is finished.
          This establishes the post-migration situation (figure~\ref{fig:post-migration}).
@@
 \end{definition}
-\begin{definition}[relabel concepts in rules]
-   \[\sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{C}{D}{u}}{\subst{C}{D}{\dataset}}\]
-\end{definition}
+   The relabeling of concepts in rules must satisfy:
+\begin{equation}
+   \sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{C}{D}{u}}{\subst{C}{D}{\dataset}}
+\end{equation}
 \begin{definition}[relabel concepts in schemas]
@@
          \end{array}
+   \end{eqnarray}
+\end{definition}
+\begin{definition}[relabel relations]
+   \[\subst{\declare{m}{C}{D}}{m'}{{\declare{n}{A}{B}}}\ =\ \text{\bf if}\ n=m\wedge A=C\wedge B=D\ \text{\bf then}\ \declare{m'}{A}{B}\ \text{\bf else}\ \declare{m}{A}{B}\]
+\end{definition}
+\begin{definition}[relabel relations in triples]
+   \[\subst{r'}{m'}{\triple{a}{r}{b}} = \triple{a}{\subst{r'}{m'}{r}}{b}\]
+\end{definition}
+\begin{definition}[relabel relations in datasets]
+   \[\begin{array}{rcl}
+      \rlap{$\subst{r'}{m'}{\pair{\triples}{\inst}}$}\\
+      &=&\pair{\{\subst{r'}{m'}{t}\mid t\in\triples\}}{\inst}
+   \end{array}\]
+\end{definition}
+   The relabeling of relation names in rules must satisfy:
+\begin{equation}
+   \sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{r'}{m'}{u}}{\subst{r'}{m'}{\dataset}}
+\end{equation}
+\begin{definition}[relabel relations in schemas]
+   \begin{eqnarray}
+      \rlap{$\subst{r'}{m'}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
+      &=&\la\concepts,\{\subst{r'}{m'}{r} \mid r\in\rels\}, \{\subst{r'}{m'}{u} \mid u\in\rules\}\ \ra\notag
    \end{eqnarray}
@@
    RELATION takes[Student*Course]
+   RELATION isPartOf[Module*Course] [UNI]
+   RELATION isEnrolledFor [Student*Module]
+   RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
+   ROLE Administrator MAINTAINS EnrollRule
+
+   RELATION course[ExamReg*Course] [UNI] =
+      [ ("ER1", "Management")
+      ; ("ER2", "Business IT")
+      ; ("ER3", "Business IT")
+      ]
+   RELATION student[ExamReg*Student] [UNI] =
+      [ ("ER1", "Peter")
+      ; ("ER2", "Susan")
+      ]
+   RULE ExamRule1: student~;course |- takes
+   RULE ExamRule2: student~;course;isPartOf~ |- isEnrolledFor
+\end{verbatim}
+
+   The generated migration system is:
+\begin{verbatim}
+   RELATION viol_EnrollRule[Student*Module]
+   RELATION diff_takes[Student*Course]
+   RELATION diff_isPartOf[Module*Course] [UNI]
+   RELATION diff_isEnrolledFor [Student*Module]
+
+   ENFORCE diff_takes :< old.takes-new.takes
+
+  -- ENFORCE new.takes >: old.takes  - diff_takes
+  -- ENFORCE new.takes :< old.takes \/ diff_takes
+   ENFORCE old.takes >: new.takes  - diff_takes
+   ENFORCE old.takes :< new.takes \/ diff_takes
+
+   ENFORCE new.isPartOf := old.isPartOf
+   ENFORCE new.isEnrolledFor := old.isEnrolledFor
+\end{verbatim}
+   The prefix `old.` refers to concepts, relations, and rules from the existing information system, $\infsys$, which is defined in section~\ref{sct:Example existing IS}.
+   The prefix `new.` refers to items from the desired information system, $\infsys'$, defined in this section.
+   The population just before the moment of transition is.
+\begin{verbatim}
+   RELATION takes[Student*Course]
    RELATION isPartOf[Module*Course] [UNI]=
@@
       ]
-   RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
-   ROLE Administrator MAINTAINS EnrollRule
-
-   RELATION course[ExamReg*Course] [UNI] =
-      [ ("ER1", "Management")
-      ; ("ER2", "Business IT")
-      ; ("ER3", "Business IT")
-      ]
-   RELATION student[ExamReg*Student] [UNI] =
-      [ ("ER1", "Peter")
-      ; ("ER2", "Susan")
-      ]
-   RULE ExamRule1: student~;course |- takes
-   RULE ExamRule2: student~;course;isPartOf~ |- isEnrolledFor
-\end{verbatim}
+\end{verbatim}
+
    This specification shows that $\infsys'$ gets to keep all three relations and the rule from $\infsys$.

### 2023-01-07 2022 Migration/articleMigration.tex
introduce data loading mechanisme by enforce rules

@@
 \label{sct:Introduction}
-   To automate the datamigrations that come with incrementally developing and deploying information systems;
-   that is the purpose of the theory proposed in this paper.
-
-   We believe in generating information systems%
-\footnote{In the sequel, the word ``system'' refers to the phrase ``information system''. This simplifies the language a little. }
-   to prevent human programming errors.
-   For, a mistake not made can never propagate into production.
+   The purpose of this work is to automate the data migrations that come with incrementally developing and deploying information systems.
+   This paper proposes a theory for such data migrations in the context of generated information systems%
+\footnote{In the sequel, the word ``system'' refers to the phrase ``information system''. This simplifies the language a little. }.
+
+   We believe in generating systems to prevent human programming errors.
+   For, every mistake not made will never propagate into production.
    We also believe that information systems should be released into production in small increments,
@@
    A difference of schemas complicates the migration of data between the existing system and the desired system.
-   The purpose of incremental evolution poses specific requirements to data migration,
+   The purpose of incremental deployment poses specific requirements to data migration,
    which sets it apart from data migrations for other purposes.
@@
    They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
-   In our work, however, we want to get rid of the existing system and not retain it.
+   In our work, however, we want to get rid of the existing system and not retain variations.
    So, we need to look at a different approach.
    Examples like these have convinced us to define a new method for data migration that is specifically meant for systems in production that are
-   evolving incrementally.
+   deployed incrementally.
 
@@
 
-   Information systems are typically used by distributed actors (both users and computers) who continually work with data.
+   Information systems are typically used by actors (both users and computers) who are distributed and work with data all the time.
    As a consequence, the data in a system changes continually.
@@
    We also assume that every service runs independent of other services,
-   meaning that the service can be stopped, (re)started and substituted without disrupting the other services.
+   meaning that each service can be stopped, (re)started and substituted without disrupting the other services.
 
@@
    We will also assume that the system's semantics are represented as constraints,
-   which can be implemented most database management systems as integrity rules that keep the constraints satisfied.
+   which most database management systems refer to as integrity rules.
+   The purpose of integrity rules is to keep the constraints they represent satisfied.
 
@@
    while preserving the meaning of the present data as much as possible~\cite{Spivak2012}.
-   Just copying the set of data from the existing system to the desired is obviously wrong if the schemas of both systems differ.
+   Just copying the set of data from the existing system to the desired system is obviously wrong if the schemas of both systems differ.
 
@@
    The migration schema comprises the schemas of both the existing system and the desired system.
-   The rules of the migration schema will cause the correct data to be transported into the desired system.
+   The rules of the migration schema will cause the right data to be copied correctly into the desired system.
    This migration schema preserves data from the existing system, it may introduce new data automatically (generated from the existing dataset),
@@
    We will use the example in section~\ref{sct:Example existing IS} as the existing system
-   and call that $\infsys$ in this section.
+   and call that $\infsys$.
    Its population reflects the state of the system just before the migration.
@@
    RELATION takes[Student*Course]
-   RELATION isPartOf[Module*Course] [UNI]
-   RELATION isEnrolledFor [Student*Module]
+   RELATION isPartOf[Module*Course]
+   RELATION isEnrolledFor[Student*Module]
    RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
@@
 \end{verbatim}
-
-   The generated migration system is:
-\begin{verbatim}
-   RELATION viol_EnrollRule[Student*Module]
-   RELATION diff_takes[Student*Course]
-   RELATION diff_isPartOf[Module*Course] [UNI]
-   RELATION diff_isEnrolledFor [Student*Module]
-
-   ENFORCE diff_takes :< old.takes-new.takes
-
-  -- ENFORCE new.takes >: old.takes  - diff_takes
-  -- ENFORCE new.takes :< old.takes \/ diff_takes
-   ENFORCE old.takes >: new.takes  - diff_takes
-   ENFORCE old.takes :< new.takes \/ diff_takes
-
-   ENFORCE new.isPartOf := old.isPartOf
-   ENFORCE new.isEnrolledFor := old.isEnrolledFor
-\end{verbatim}
-   The prefix `old.` refers to concepts, relations, and rules from the existing information system, $\infsys$, which is defined in section~\ref{sct:Example existing IS}.
-   The prefix `new.` refers to items from the desired information system, $\infsys'$, defined in this section.
-   The population just before the moment of transition is.
-\begin{verbatim}
-   RELATION takes[Student*Course]
-   RELATION isPartOf[Module*Course] [UNI]=
-      [ ("IT-Governance", "Business IT") ]
-   RELATION isEnrolledFor [Student*Module] =
-      [ ("Susan", "Business Analytics")
-      ; ("Susan", "IT-Governance")
-      ; ("Susan", "Business Rules")
-      ]
-\end{verbatim}
-
-   This specification shows that $\infsys'$ gets to keep all three relations and the rule from $\infsys$.
-   Some new population, two new relations, and two new rules have been added to $\infsys'$.
-   $\infsys'$ contains exam registrations (concept \verb-ExamReg-), which is new.
+   We can see that the relations and the rule of the existing system remain part of the desired system%
+\footnote{Removal or changing of relations is allowed as well.}
+   because the developer has copied these items into the script of the desired system.
+   However, she has lifted the restriction \verb-UNI- on the relation {\tt isPartOf}.
+   Besides, two new relations and two rules have appeared.
+
+   The desired system, $\infsys'$, contains exam registrations (concept \verb-ExamReg-), which is new.
    In an exam registration, a student registers for the examination of a course.
@@
 % \end{array}\]
+
+\subsection{Copying Data}
+   For loading the data in the desired database, we use enforce rules.
+   In Ampersand, an enforce rule comes in one of the following three kinds:
+\begin{eqnarray}
+   \text{\tt ENFORCE}\ r\ \text{\tt :<}\ \id{expression}\label{enforce del}\\
+   \text{\tt ENFORCE}\ r\ \text{\tt :=}\ \id{expression}\label{enforce eq}\\
+   \text{\tt ENFORCE}\ r\ \text{\tt >:}\ \id{expression}\label{enforce ins}
+\end{eqnarray}
+   Every enforce rule is enforced by an engine,
+   which changes the population of relation $r$ to satisfy the rule.
+   An enforce rule with operator {\tt :<} (alternative~\ref{enforce del}) keeps the population of $r$ smaller than or equal to \id{expression}.
+   When the expression is losing pairs that are in $r$, the engine will spontaneously remove such pairs from $r$ to keep $r$ smaller than or equal to \id{expression}.
+   An enforce rule with operator {\tt >:} (alternative~\ref{enforce ins}) keeps the population of $r$ larger than or equal to \id{expression}.
+   When the expression is growing with pairs that are in $r$, the engine will spontaneously insert such pairs in $r$ to keep $r$ greater than or equal to \id{expression}.
+   An enforce rule with operator {\tt :=} (alternative~\ref{enforce eq}) keeps the population of $r$ equal to \id{expression}.
+   It does so by inserting and deleting pairs in $r$ to keep it equal to \id{expression}.
+   Summarizing, an enforce rule can be read as an invariant that keeps one of $r\leq\id{expression}$, $r=\id{expression}$, or $r\geq\id{expression}$ satisfied by changing the population of $r$.
+
+   It would be tempting to state that a relation from the example system, say {\tt takes}, could be copied by the following code fragment%
+\footnote{The prefix {\tt old.} refers to concepts, relations, and rules from the existing system (section~\ref{sct:Example existing IS}).
+The prefix {\tt new.} refers to items from the desired information system (this section).}:
+\begin{verbatim}
+   ENFORCE new.takes >: old.takes
+\end{verbatim}
+   This would mean, however, that all pairs from {\tt old.takes} will be in {\tt new.takes} forever.
+   An extra relation fixes this problem by remembering which pairs have been copied:
+
+\begin{verbatim}
+   RELATION copied_takes[Student*Course]
+   ENFORCE new.takes >: old.takes - copied_takes
+   ENFORCE copied_takes >: new.takes
+\end{verbatim}
+   Initially, when {\tt copied\_takes} is still empty, the difference between {\tt old.takes} and {\tt copied\_takes} equals {\tt old.takes}.
+   So, {\tt new.takes} will start filling up with the pairs of {\tt old.takes}.
+   Then, the second enforce rule will start filling {\tt copied\_takes} with pairs from {\tt new.takes},
+   and the difference between {\tt old.takes} and {\tt copied\_takes} gets smaller and smaller, until {\tt old.takes} and {\tt copied\_takes} are equal.
+   Then, the enforce rule that fills {\tt new.takes} will be satisfied forever and will change nothing more to {\tt new.takes}.
+
+   So, the generator that ``writes'' the code for the migration system must introduce an extra relation for each relation to be copied.
+   After the migration has finished, this extra relation and both enforce rules can be taken out of production.
+   For this purpose, the generator writes a rule that signals the migration engineer when copying is done.
+
+\subsection{Migration system}
+   The generated migration system is:
+\begin{verbatim}
+   RELATION copied_takes[Student*Course]
+   ENFORCE new.takes >: old.takes - copied_takes
+   ENFORCE copied_takes >: new.takes
+
+   RELATION copied_isPartOf[Module*Course]
+   ENFORCE new.isPartOf >: old.isPartOf - copied_isPartOf
+   ENFORCE copied_isPartOf >: new.isPartOf
+
+   RELATION copied_isEnrolledFor[Student*Module]
+   ENFORCE new.isEnrolledFor >: old.isEnrolledFor - copied_isEnrolledFor
+   ENFORCE copied_isEnrolledFor >: new.isEnrolledFor
+
+   ROLE MigrationEngineer MAINTAINS migrating_takes, migrating_isPartOf, migrating_isEnrolledFor
+   RULE migrating_takes : old.takes-copied_takes = -V
+   RULE migrating_isPartOf : old.isPartOf-copied_isPartOf = -V
+   RULE migrating_isEnrolledFor : old.isEnrolledFor-copied_isEnrolledFor = -V
+\end{verbatim}
+
+\subsection{Moment of Transition}
+   Let us assume the population just before the moment of transition is.
+\begin{verbatim}
+   RELATION takes[Student*Course]
+   RELATION isPartOf[Module*Course] [UNI]=
+      [ ("IT-Governance", "Business IT") ]
+   RELATION isEnrolledFor [Student*Module] =
+      [ ("Susan", "Business Analytics")
+      ; ("Susan", "IT-Governance")
+      ; ("Susan", "Business Rules")
+      ]
+\end{verbatim}
 

### 2023-01-08 2022 Migration/articleMigration.tex
changed sct 3.7 Loading data

@@
    The example system also has a second relation that states which modules are part of which course.
-   It is constrained by \verb-[UNI]- to be univalent.
-   This means that every \verb-Module- is part of at most one \verb-Course-.
-\begin{verbatim}
-RELATION isPartOf[Module*Course] [UNI] =
+\begin{verbatim}
+RELATION isPartOf[Module*Course] =
 [ ("Finance", "Management")
@@
          If schemas $\schema$ equals $\schema'$, we call the migration trivial because a data migration is not necessary.
-   \item The generator generates a migration script $\migrsys$
+   \item The generator generates a migration script, $\migrsys$,
          to migrate the data and define the semantics during the migration phase as far as a generator can reasonably predict.
@@
    \item\label{item3}
-         A migration engineer changes $\migrsys$ to accommodate the non-standard requirements of the migration,
-         yielding migration script $\migrsys'$.
-         The migration engineer does not alter the existing system $\infsys$ nor the specification
-         of the desired system $\infsys'$ that the developer has produced.
+         In many cases, the schema $\schema_{\migrsys}$ will be enough to perform the migration.
+         However, a migration engineer may have specific requirements that call for changes in $\migrsys$.
+         For that purpose, the generator generates the migration script as an Ampersand script,
+         to let the migration engineer accommodate $\migrsys$ to the non-standard requirements.
+         The resulting migration script, $\migrsys'$, can be tested separately and can be used to take the desired system to production.
+         The migration engineer does not alter the schemas of the existing system or the desired system.
+         All migration functionality is concentrated in $\migrsys'$.
    \item The migration engineer deploys system $\migrsys'$.
@@
    So, whereas such examples may fly in a technical sense, the developer must use renaming with care to preserve the intended semantics.
+
+\subsection{Loading Data}
+   For loading the data in the desired database, we use enforce rules.
+   In Ampersand, an enforce rule comes in three flavors:
+\begin{eqnarray}
+   \text{\tt ENFORCE}\ r\ \text{\tt :<}\ \id{expression}\label{enforce del}\\
+   \text{\tt ENFORCE}\ r\ \text{\tt :=}\ \id{expression}\label{enforce eq}\\
+   \text{\tt ENFORCE}\ r\ \text{\tt >:}\ \id{expression}\label{enforce ins}
+\end{eqnarray}
+   Every enforce rule is enforced by an engine,
+   which changes the population of relation $r$ to satisfy the rule.
+   An enforce rule with operator {\tt :<} (alternative~\ref{enforce del}) keeps the population of $r$ smaller than or equal to \id{expression}.
+   When the population of this expression is losing pairs that are in $r$, the engine will remove such pairs from $r$ to keep $r$ smaller than or equal to \id{expression}.
+   An enforce rule with operator {\tt >:} (alternative~\ref{enforce ins}) keeps the population of $r$ larger than or equal to \id{expression}.
+   When the population of this expression is growing with pairs that are not in $r$, the engine will insert such pairs in $r$ to keep $r$ greater than or equal to \id{expression}.
+   An enforce rule with operator {\tt :=} (alternative~\ref{enforce eq}) keeps the population of $r$ equal to \id{expression}.
+   It does so by inserting and deleting pairs in $r$ to keep it equal to \id{expression}.
+   Summarizing, an enforce rule can be read as an invariant that keeps one of $r\leq\id{expression}$, $r=\id{expression}$, or $r\geq\id{expression}$ satisfied by changing the population of $r$.
+
+   It would be tempting to state that a relation from the example system, say {\tt takes}, could be copied by the following code fragment%
+\footnote{The prefix {\tt old.} refers to concepts, relations, and rules from the existing system (section~\ref{sct:Example existing IS}).
+The prefix {\tt new.} refers to items from the desired information system.}:
+\begin{verbatim}
+   ENFORCE new.takes >: old.takes
+\end{verbatim}
+   This would mean, however, that all pairs from {\tt old.takes} will be in {\tt new.takes} forever.
+   An extra relation (\ref{extraRelation}) fixes this problem by remembering which pairs have been copied:
+\begin{eqnarray}
+   &&\verb#RELATION copied_takes[Student*Course]#\label{extraRelation}\\
+   &&\verb#ENFORCE new.takes >: old.takes - copied_takes#\label{difference}\\
+   &&\verb#ENFORCE copied_takes := new.takes#\label{fill copies with new.takes}\\
+   &&\verb#ENFORCE copied_takes :< old.takes#\label{del copies from old.takes}
+\end{eqnarray}
+   Initially, when {\tt copied\_takes} is still empty, the difference between {\tt old.takes} and {\tt copied\_takes} equals {\tt old.takes}.
+   So, {\tt new.takes} will start filling up with the pairs of {\tt old.takes} by rule~\ref*{difference}.
+   Then, the engine will start filling {\tt copied\_takes} with pairs from {\tt new.takes} because of rule~\ref*{fill copies with new.takes}.
+   As a result, the difference between {\tt old.takes} and {\tt copied\_takes} gets smaller and smaller, until {\tt old.takes} and {\tt copied\_takes} are equal.
+   While this copying process is going on, {\tt old.takes} may accumulate more pairs or lose pairs.
+\begin{itemize}
+\item If pair $p$ is inserted in {\tt old.takes} and $p$ is not in {\tt copied\_takes},
+      then rule~\ref{difference} causes the engine to insert $p$ in {\tt new.takes}.
+      Subsequently, rule~\ref{fill copies with new.takes} causes the engine to insert $p$ in {\tt copied\_takes} as well.
+\item If pair $p$ is inserted in {\tt old.takes} and $p$ is already in {\tt copied\_takes},
+      then nothing happens.
+\item If pair $p$ is deleted from {\tt old.takes} and $p$ is in {\tt copied\_takes},
+      then rule~\ref{del copies from old.takes} causes the engine to delete $p$ from {\tt copied\_takes}.
+\item If pair $p$ is deleted from {\tt old.takes} and $p$ is not in {\tt copied\_takes},
+      then nothing happens.
+\item If pair $p$ is inserted in {\tt new.takes},
+      rule~\ref{fill copies with new.takes} causes the engine to insert $p$ into {\tt copied\_takes}.
+\item If pair $p$ is deleted from {\tt new.takes},
+      rule~\ref{fill copies with new.takes} causes the engine to delete $p$ from {\tt copied\_takes}.
+      If this pair is in {\tt old.takes}, rule~\ref{difference} causes the engine to insert $p$ back into {\tt new.takes}.
+\end{itemize}
+
+   These new pairs are copied by the same rule.
+   However, {\tt old.takes} may also lose pairs, which are 
+   Then, the enforce rule that fills {\tt new.takes} will be satisfied forever and will change nothing more to {\tt new.takes}.
+
+   But what if (in the meantime) someone removes a pair from {\tt old.takes} that has already been inserted in {\tt new.takes}?
+\begin{verbatim}
+   RELATION added_takes[Student*Course]
+   RELATION deleted_takes[Student*Course]
+   ENFORCE new.takes >: old.takes - added_takes
+   ENFORCE added_takes >: new.takes
+\end{verbatim}
+   So, the generator that ``writes'' the code for the migration system must introduce an extra relation for each relation to be copied.
+   After the migration has finished, this extra relation and both enforce rules can be taken out of production.
+
 \subsection{Example}
@@
    RELATION takes[Student*Course]
-   RELATION isPartOf[Module*Course]
+   RELATION isPartOf[Module*Course] [UNI]
    RELATION isEnrolledFor[Student*Module]
@@
       ]
+   ROLE Student MAINTAINS ExamRule1, ExamRule2
    RULE ExamRule1: student~;course |- takes
@@
    because the developer has copied these items into the script of the desired system.
-   However, she has lifted the restriction \verb-UNI- on the relation {\tt isPartOf}.
+   However, she has imposed the restriction \verb-UNI- on the relation {\tt isPartOf}, stating that this relation is univalent.
+   This means that every \verb-Module- is part of at most one \verb-Course-.
    Besides, two new relations and two rules have appeared.
@@
 \end{array}\]
+   Both rules must be maintained by the student.
+   So, a student can register for an exam only if (s)he is taking the course for which (s)he registers and if (s)he is enrolled for every module of that course.
    % Ampersand also derives the following violation sets:
@@
 
-\subsection{Copying Data}
-   For loading the data in the desired database, we use enforce rules.
-   In Ampersand, an enforce rule comes in one of the following three kinds:
-\begin{eqnarray}
-   \text{\tt ENFORCE}\ r\ \text{\tt :<}\ \id{expression}\label{enforce del}\\
-   \text{\tt ENFORCE}\ r\ \text{\tt :=}\ \id{expression}\label{enforce eq}\\
-   \text{\tt ENFORCE}\ r\ \text{\tt >:}\ \id{expression}\label{enforce ins}
-\end{eqnarray}
-   Every enforce rule is enforced by an engine,
-   which changes the population of relation $r$ to satisfy the rule.
-   An enforce rule with operator {\tt :<} (alternative~\ref{enforce del}) keeps the population of $r$ smaller than or equal to \id{expression}.
-   When the expression is losing pairs that are in $r$, the engine will spontaneously remove such pairs from $r$ to keep $r$ smaller than or equal to \id{expression}.
-   An enforce rule with operator {\tt >:} (alternative~\ref{enforce ins}) keeps the population of $r$ larger than or equal to \id{expression}.
-   When the expression is growing with pairs that are in $r$, the engine will spontaneously insert such pairs in $r$ to keep $r$ greater than or equal to \id{expression}.
-   An enforce rule with operator {\tt :=} (alternative~\ref{enforce eq}) keeps the population of $r$ equal to \id{expression}.
-   It does so by inserting and deleting pairs in $r$ to keep it equal to \id{expression}.
-   Summarizing, an enforce rule can be read as an invariant that keeps one of $r\leq\id{expression}$, $r=\id{expression}$, or $r\geq\id{expression}$ satisfied by changing the population of $r$.
-
-   It would be tempting to state that a relation from the example system, say {\tt takes}, could be copied by the following code fragment%
-\footnote{The prefix {\tt old.} refers to concepts, relations, and rules from the existing system (section~\ref{sct:Example existing IS}).
-The prefix {\tt new.} refers to items from the desired information system (this section).}:
-\begin{verbatim}
-   ENFORCE new.takes >: old.takes
-\end{verbatim}
-   This would mean, however, that all pairs from {\tt old.takes} will be in {\tt new.takes} forever.
-   An extra relation fixes this problem by remembering which pairs have been copied:
-
-\begin{verbatim}
-   RELATION copied_takes[Student*Course]
-   ENFORCE new.takes >: old.takes - copied_takes
-   ENFORCE copied_takes >: new.takes
-\end{verbatim}
-   Initially, when {\tt copied\_takes} is still empty, the difference between {\tt old.takes} and {\tt copied\_takes} equals {\tt old.takes}.
-   So, {\tt new.takes} will start filling up with the pairs of {\tt old.takes}.
-   Then, the second enforce rule will start filling {\tt copied\_takes} with pairs from {\tt new.takes},
-   and the difference between {\tt old.takes} and {\tt copied\_takes} gets smaller and smaller, until {\tt old.takes} and {\tt copied\_takes} are equal.
-   Then, the enforce rule that fills {\tt new.takes} will be satisfied forever and will change nothing more to {\tt new.takes}.
-
-   So, the generator that ``writes'' the code for the migration system must introduce an extra relation for each relation to be copied.
-   After the migration has finished, this extra relation and both enforce rules can be taken out of production.
-   For this purpose, the generator writes a rule that signals the migration engineer when copying is done.
-
 \subsection{Migration system}
-   The generated migration system is:
+   The generated migration system, $\migrsys$, is:
 \begin{verbatim}
@@
 
-\subsection{Moment of Transition}
-   Let us assume the population just before the moment of transition is.
-\begin{verbatim}
-   RELATION takes[Student*Course]
+\subsection{Preparing for the Moment of Transition}
+   Just before loading the desired system with data,
+   the data in the existing system is no longer the data we saw in section~\ref{sct:Example existing IS}.
+   Let us assume the actual population at this time is:
+\begin{verbatim}
+   RELATION takes[Student*Course] =
+      [ ("Peter", "Management")
+      ; ("Susan", "Business IT")
+      ; ("John", "Business IT")
+      ]
    RELATION isPartOf[Module*Course] [UNI]=
-      [ ("IT-Governance", "Business IT") ]
+      [ ("Finance", "Management")
+      ; ("Business Rules", "Business IT")
+      ; ("Business Analytics", "Business IT")
+      ; ("IT-Governance", "Management")
+      ; ("IT-Governance", "Business IT")
+      ]
    RELATION isEnrolledFor [Student*Module] =

### 2023-01-08 2022 Migration/articleMigration.tex
WIP

@@
 
-   These new pairs are copied by the same rule.
-   However, {\tt old.takes} may also lose pairs, which are 
-   Then, the enforce rule that fills {\tt new.takes} will be satisfied forever and will change nothing more to {\tt new.takes}.
-
-   But what if (in the meantime) someone removes a pair from {\tt old.takes} that has already been inserted in {\tt new.takes}?
-\begin{verbatim}
-   RELATION added_takes[Student*Course]
-   RELATION deleted_takes[Student*Course]
-   ENFORCE new.takes >: old.takes - added_takes
-   ENFORCE added_takes >: new.takes
-\end{verbatim}
-   So, the generator that ``writes'' the code for the migration system must introduce an extra relation for each relation to be copied.
-   After the migration has finished, this extra relation and both enforce rules can be taken out of production.
-
 \subsection{Example}

### 2023-01-11 Generative Software/proposal.tex
qualification of researchers added

\documentclass{elsarticle}
% \usepackage{natbib}
% \usepackage{graphicx}
% \usepackage{footmisc}
% \usepackage{amstext}
% \usepackage{amsmath}
% \usepackage{amssymb}
% \usepackage{amsthm}
\usepackage[english]{babel}
\usepackage{libertine}
\usepackage{inconsolata}
% \usepackage[official,right]{eurosym}
\selectlanguage{english}
\usepackage{hyperref}
\frenchspacing

\begin{document}
% \include{preamble}

\title{Generative Software Development\\{\em\normalsize Research Offering}}
\author{prof.dr.ir.\ Stef Joosten\footnote{ORCID 0000-0001-8308-0189}}
\author{dr.\ Tim Steenvoorden\footnote{ORCID 0000-0002-8436-2054}}
\address{Open Universiteit Nederland, Heerlen, the Netherlands}

\begin{abstract}
    This paper summarizes the capability of the Open Universiteit in generative software development.
    We offer this enabling technology as a component of a larger effort that involves designing distributed information systems.
    Generative software development offers attractive benefits for correctness, security, and resilience.
    Also, it enables a product to be demonstrated live during its creation,
    hereby supporting agile software development methods.
    %every 2 or 3 weeks as it develops into an end product.
\end{abstract}

\begin{keyword}
    generative software, software engineering, security, prototyping, correctness, Ampersand, TopHat
\end{keyword}
\maketitle

\section{Requirements (remove this section before submitting)}
Reviewers will judge this proposal by the following requirements.
\subsection{Scientific quality}
\begin{enumerate}
   \item 	To what extent is the proposed research original and how would you rate the innovative elements?
   \item 	What is your assessment of the design of the project, including the goals, hypotheses, research methods, and scientific feasibility?
   \item 	What is your assessment of the coherence and time schedule of the proposed lines of research?
   \item 	Is the research group competent enough to carry out the research? Does the group have a relevant position in the international scientific community? Is the available infrastructure adequate?
   \item 	Are the number and category of requested personnel, budget for materials, investments, and foreign travel adequate?
   \item 	Does the proposed research fit in the Open Technology Programme; does it focus on engineering sciences research?
   \item 	What are the strong and weak points of the scientific part of the proposal?   
\end{enumerate}

\subsection{Knowledge Utilisation} (the application of the results of the research by third parties)
\begin{enumerate}
   \item  	What is your assessment of the description of the commercial and/or societal potential impacts of the research given in the proposal?
   \item 	What is your assessment of the contribution and commitment of the users and the proposed composition of the user committee?
   \item 	What is your assessment of the freedom to operate concerning commercial propositions, existing patents, eligibility or societal acceptance? 
   \item 	What are the prospects for collaboration with the industry and knowledge transfer, assuming the project is successful? Please address both aspects.
   \item 	What is your assessment of the research group's competence regarding the transfer and application of research results?
   \item 	What are the strong and weak points of the utilisation plan?   
\end{enumerate}

\section{Research question}
\label{sct:Research question}
   Data driven organizations that change frequently are facing the dilemma of buy vs. build.
   On one hand, flexible and fast adaptation requires immediate deployment of new software components, without interrupting the ongoing business.
   On the other hand, one-of-a-kind organizations require one-of-a-kind components that need to be developed first.
   Commercial off-the-shelf (COTS) components provide speedy deployments, but can be too generic.
   Homebrew components can satisfy specific requirements, but take time to build and maintain.
   The so called ``low code platforms'' are sometimes advocated as the solution
   because they generate user functionality from a graphical coding environment.
   However, practical experience shows that these platforms are not so ``low code'' after all,
   when the database structure is not properly aligned with the user functionality.

   In practice, we see many organizations that opt for homebrew solutions.
   Generative development can shorten the development time of homebrew software, regaining some of the desired flexibility.

\section{Plan}
   We want to participate in a larger research project by means of one researcher (PhD candidate, full-time, 4 years) and one scientific programmer.
   We intend to further the knowledge about generative software development by participating in a larger research effort.
   We bring the expertise~\cite{JoostenRAMiCS2017,Steenvoorden2022} and tools~\cite{Joosten-JLAMP2018,10.1145/3354166.3354182} on generative software development
   as an asset to that project at no extra cost.
   The authors also intend to participate in person in the role of supervisors of the researcher and the scientific programmer, each for 0.1 FTE.

   This capability offers the following benefits:
\begin{itemize}
    \item By generating software, we can produce live demonstrations periodically to keep the future user population informed of the progress.
    This produces quick results, which keeps funders happy.
    \item By automating software production, we can prevent a large class of mistakes and thereby reduce the project risk.
    \item By formally specifying the functionality, we can give valuable guarantees on the correctness of the results.
    \item Software that is correctly generated needs no testing. This yields significant savings on testing and test maintenance.
    \item The tooling that has been developed at the Open University is especially suitable for automating business processes and workflows.
\end{itemize}

\section{Qualification of researchers}
\label{sct:Qualification}
   Research into generative software has a long tradition at the OU.
   Results from this research comprise:
\begin{itemize}
   \item A generator, Ampersand, that produces a complete information system from a specification,
         and a generator, TopHat, that produces a workflow system from a specification,
         demonstrating that this research has tangible results~\cite{Michels2011}.
   \item Numerous publications in peer-reviewed scientific conferences and journals,
         demonstrating that this research is scientifically relevant.
         Some of the highlights:
         \begin{itemize}
            \item A type system for information systems to ensure that all data is meaningful (i.e. has an interpretation)~\cite{vdWoude2011}
            \item A book, "Rule Based Design", published as open educational content on the web~\cite{RBD}
            \item An application that checks modeling conventions in architectural specifications~\cite{iceis22}
            \item An application for identity and access management~\cite{SIAM2008}
            \item Domain analysis~\cite{Joosten2015}.
         \end{itemize}
   \item A course, Rule Based Design, which is in the Master's program of the Computer Science curriculum and the Business Process Management \& IT program of the OU,
         demonstrating that this research is valued by the OU.
   \item over 10 generated information systems in production, made by TNO and one made by the OU,
         demonstrating that this research has practical value for industry.
   \item in-kind support from Ordina,
         demonstrating that this research can draw positive attention from the IT industry.
\end{itemize}

Current work in our group is about supporting incremental changes to an information system generatively,
to support DevOps teams who need incremental deployments.
Supporting incremental deployments with a generator is likely to decrease deployment times,
to increase the number of releases per time unit, and to decrease the risk of errors propagating to production.
\section{Bibliography}
\bibliographystyle{elsarticle-harv}
\bibliography{doc}

\end{document}

### 2023-01-11 Generative Software/proposalBrief.tex
qualification of researchers added

@@
 
-\title{Generative Software Development\\{\em\normalsize Research Proposal}}
+\title{Generative Software Development\\{\em\normalsize Research Offering}}
 \author{prof.dr.ir.\ Stef Joosten\footnote{ORCID 0000-0001-8308-0189}}
@@
    Yet, research so far has yielded promising results with respect to validation and resilience.
-   A software generator enables the DevOps way of working, which is to demonstrate the end product at the end of every sprint.
+   Software generators that can deploy increments in small steps will enable the DevOps way of working.
    Thus, a way of software development emerges that allows for very fast incremental development of large information systems.
@@
 \section{Benefits}
-    The Open University offers its expertise~\cite{JoostenRAMiCS2017,Steenvoorden2022} and tools~\cite{Joosten-JLAMP2018,10.1145/3354166.3354182} as a building block for a larger research effort,
+    The Open University offers its expertise~\cite{JoostenRAMiCS2017,Steenvoorden2022} and tools~\cite{Joosten-JLAMP2018,10.1145/3354166.3354182}
+    as a building block for a larger research effort,
     in which generative software development is an enabler for the research proposed.
@@
     \item Software that is correctly generated needs no testing. This yields significant savings on testing and test maintenance.
+    \item The tooling that has been developed at the Open University is especially suitable for automating business processes and workflows.
 \end{itemize}

### 2023-01-11 Generative Software/proposal.tex
WIP

@@
    On one hand, flexible and fast adaptation requires immediate deployment of new software components, without interrupting the ongoing business.
-   On the other hand, one-of-a-kind organizations require one-of-a-kind components that need to be developed first.
-   Commercial off-the-shelf (COTS) components provide speedy deployments, but can be too generic.
-   Homebrew components can satisfy specific requirements, but take time to build and maintain.
-   The so called ``low code platforms'' are sometimes advocated as the solution
-   because they generate user functionality from a graphical coding environment.
-   However, practical experience shows that these platforms are not so ``low code'' after all,
-   when the database structure is not properly aligned with the user functionality.
+   Commercial off-the-shelf (COTS) components can solve that problem, but they can be too generic.
+   On the other hand, one-of-a-kind organizations require one-of-a-kind components that are not readily available.
+   The so called ``low code an no code platforms'' are sometimes advocated as the solution
+   because their (re-)sellers claim that the business itself can configure such systems.
+   However, we have seen some disappointing results of that in practice, because of integration problems and maintainability issues.
+   So homebrew solutions are still widespread in organizations with one-of-a-kind requirements.
+   They are still willing to invest the time to build and maintain information systems and take the risk of a failing IT-innovation.
 
-   In practice, we see many organizations that opt for homebrew solutions.
-   Generative development can shorten the development time of homebrew software, regaining some of the desired flexibility.
+   Generative development can shorten the development time and reduce the development risks of homebrew software,
+   regaining the desired flexibility and satisfying specific one-of-a-kind requirements at the same time.
+   Current research has shown that complete working information systems can be generated.
+   To get this technology ready for real use, research needs to be done on gaining more control over non-functional properties of generated systems,
+   such as scalability, robustness, and security.
 
 \section{Plan}
-   We want to participate in a larger research project by means of one researcher (PhD candidate, full-time, 4 years) and one scientific programmer.
+   We want to participate in a larger research project by means of one researcher (PhD candidate, full-time, 4 years) and one scientific programmer (professional, full-time, 4 years).
    We intend to further the knowledge about generative software development by participating in a larger research effort.

### 2023-02-06 2022 Migration/articleMigration.tex
fixed "rule 6" and fixed specialization

@@
    making Ampersand more useful for agile software development.
-   We propose a migration method that features zero down-time,
+   This paper proposes a migration method that features zero down-time,
    and features an undo after deployment.
@@
 
-   We believe in generating systems to prevent human programming errors.
-   For, every mistake not made will never propagate into production.
-   We also believe that information systems should be released into production in small increments,
-   to obtain shorter release cycles, more frequent releases, and thus a more agile software development process%
+   This paper is founded in the belief that generating systems can prevent human programming errors, which contributes to robust software.
+   Besides, information systems should be released into production in small increments.
+   That yields shorter release cycles, more frequent releases, and thus a more agile software development process%
    \footnote{These effects can be measured in terms of DevOps metrics such as
@@
    the users experience a system that evolves organically, without even noticing the updates.
-   Thus, we believe our theory will help to automate data migrations correctly and reliably.
+   Thus, this paper will help to automate data migrations correctly and reliably.
    The benefits work in two directions.
@@
 
-   In each increment, we distinguish an ``existing system'' and a ``desired system''.
+   Each increment has an ``existing system'' and a ``desired system''.
    A difference of schemas complicates the migration of data between the existing system and the desired system.
@@
    They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
-   In our work, however, we want to get rid of the existing system and not retain variations.
-   So, we need to look at a different approach.
-   Examples like these have convinced us to define a new method for data migration that is specifically meant for systems in production that are
+   This work, however, gets rid of the existing system without retaining the variations.
+   Examples like these have convinced the authors to define a new method for data migration that is specifically meant for systems in production that are
    deployed incrementally.
@@
    For our theory, zero down-time means that all human activity for the data migration can be done without taking the system out of production.
-   We consider any delay caused by automatically switching from the existing system to the desired one negligible,
-   assuming a modern deployment platform that does this with zero down-time.
+   Any delay caused by automatically switching from the existing system to the desired is considered negligible.
+   This assumes a modern deployment platform (such as Kubernetes) that updates applicatioins with zero down-time.
 
@@
 \end{itemize}
-   The theory we develop focuses on the correctness of the migration.
-   Efficiency of the migration is beyond the scope of this paper.
+   This paper focuses on the correctness of the migration.
+   Efficiency of the migration is beyond the scope.
 
@@
    and may work from different locations and on different moments.
-   This collective use by multiple users serves a purpose which we loosely call ``the business''.
+   This collective use by multiple users serves a purpose which we will loosely call ``the business''.
    To preserve meaning, the business maintains semantic constraints on the data,
@@
    In practice, actors ``talk to'' the system through an ingress mechanism, which connects each user to the right service(s).
-   We assume that the ingress function is provided by the deployment platform, so it is beyond the scope of this paper.
-   We also assume that every service runs independent of other services,
+   The ingress function is provided by the deployment platform, so it is beyond the scope of this paper.
+   Also, every service runs independent of other services,
    meaning that each service can be stopped, (re)started and substituted without disrupting the other services.
@@
 \end{figure}
-   To keep our theory technology independent, we will assume that a dataset contains triples.
+   To keep our theory technology independent, datasets are assumed to contains triples.
    This makes our theory valid for any kind of database that triples can represent,
    such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
-   We will also assume that the system's semantics are represented as constraints,
+   The system's semantics is represented as constraints,
    which most database management systems refer to as integrity rules.
@@
 
-   In this work, we use the Ampersand compiler~\cite{Joosten-JLAMP2018},
-   to validate our theory.
+   This work uses the Ampersand compiler~\cite{Joosten-JLAMP2018} to validate the theory.
    Ampersand is a tool to generate information systems.
@@
    which means that a classical database schema (i.e.\ data structure plus semantics) can be extracted from the Ampersand script.
-   For the purpose of this paper we can equate an Ampersand script with the schema of the generated system.
+   For the purpose of this paper an Ampersand script is considered equivelant to the schema of the generated system.
 
@@
    which may require human interventions to resolve.
-   We distinguish data pollution in the following categories:
+   Data pollution is distinguished in the following categories:
 \begin{itemize}
@@
    During the time between the MoT and MoN,
-   we propose to keep the existing system and the desired system alive, side by side, as shown in figure~\ref{fig:migration phase}.
+   the existing system and the desired system are kept alive, side by side, as shown in figure~\ref{fig:migration phase}.
 \begin{figure}[bht]
@@
 
-   In the following section we introduce the definitions required to migrate data from one system to another.
+   In the following section introduces the definitions required to migrate data from one system to another.
 
@@
    An {\em information system} is a combination of dataset, schema, and functionality.
-   For the purpose of this paper, we ignore the functionality because it does not impact the migration.
-   We start with datasets in section~\ref{sct:Datasets}, then we define schemas in section~\ref{sct:Schemas},
-   and then we can define information systems.
+   For the purpose of this paper, the functionality is ignored because it does not impact the migration.
+   Section~\ref{sct:Datasets} desribes datasets. Schemas are treated in section~\ref{sct:Schemas}.
+   Then section~\ref{sct:Information Systems} defines information systems.
 
@@
    A dataset $\dataset$ describes a set of structured data, which is typically stored persistently in a database of some kind.
-   We write $\dataset_{\infsys}$ to refer to the dataset of a particular information system $\infsys$.
+   The notation $\dataset_{\infsys}$ refers to the dataset of a particular information system $\infsys$.
    The purpose of a dataset is to describe the data of a system at one point in time. 
-   Before defining datasets, we must first define the constituent notions of atom, concept, relation, and triple.
+   Before defining datasets, let us first define the constituent notions of atom, concept, specialization, relation, and triple.
 
@@
    In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
-   We will use variables $A$, $B$, $C$, $D$ to represent concepts.
+   In the sequel, variables $A$, $B$, $C$, $D$ will represent concepts.
 
@@
    The expression $a\ \inst\ C$ means that atom $a$ is an \emph{instance} of concept $C$.
-   This relation is used in the type system, in which $\inst$ assigns a type to every atom in the dataset.
+   This relation is used in the type system, in which $\inst$ assigns a concept to every atom in the dataset.
+
+   Specialization, also known as {\em generalization} or {\em subtyping}, is a relation between concepts.
+   The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
+\begin{equation}
+   \label{eqn:specialization}
+   A\isa B\ \Leftrightarrow\ \forall a: a\ \inst\ A\rightarrow a\ \inst\ B
+\end{equation}
+   Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal'',
+   in which a user defines a pair of concepts in the relation $\isa$.
+   A compiler can construct $\isa$ as the transitive, reflexive closure of such user-defined pairs of concepts.
+   It must test for antisymmetry, ensuring that $\isa$ is a partial order of concepts.
+   Specialization causes atoms to have multiple concepts of which they can be an instance.
+   To ensure that this does not result in problems at run-time, a compiler can require users to declare all specializations explicitly.
+   As a result, if an atom is an instance of concept $A$ and $A\isa B$,
+   this atom has all properties that atoms of type $B$ have.
 
@@
    Every relation $r$ has a name, a source concept, and a target concept.
-   We write $r=\declare{n}{A}{B}$ to denote that relation $r$ has name $n$, source concept $A$, and target concept $B$.
+   The notation $r=\declare{n}{A}{B}$ denotes that relation $r$ has name $n$, source concept $A$, and target concept $B$.
 
    Triples serve to represent data.
-   A triple is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.
-
-   A dataset $\dataset$ is a tuple $\pair{\triples}{\inst}$ that satisfies:
+   A triple\footnote{Please note that this paper uses the word {\em triple} in a more restricted way than in natural language.}
+   is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.
+
+   A dataset $\dataset$ is a structure $\triple{\triples}{\isa}{\inst}$ that satisfies:
 \begin{eqnarray}
-   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B
+   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&
+      \begin{array}[t]{@{}l}
+         \exists C: C\isa A\ \wedge\ a\ \inst\ C\ \wedge\\
+         \exists D: D\isa B\ \wedge\ b\ \inst\ D
+      \end{array}
    \label{eqn:wellTypedEdge}
@@
    This ``meaning from practice'' has no consequences in the formal world.
-   We leave it entirely up to a user to attach a practical meaning to a triple.
-
-   In case we need to be explicit about the dataset,
-   we will write $\triples_{\dataset}$ and $\inst_{\dataset}$ to refer to $\triples$ and $\inst$ respectively.
+   Users are entirely free to attach a practical meaning to a triple.
+
+   The notations $\triples_{\dataset}$ and $\inst_{\dataset}$ are used to disambiguate $\triples$ and $\inst$ when necessary.
    Every dataset is an element of an infinite set called $\Dataset$.
-   To save writing in the sequel, we may write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\triples$.
+   To save writing in the sequel, the notation $a\ r\ b$ means that $\triple{a}{r}{b}\in\triples$.
 
@@
    as defined by the function $\id{pop}_r:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$.
-   It defines a set of pairs, which we call the population of $r$:
+   It defines a set of pairs, also known as the population of $r$:
 \begin{equation}
@@
 \label{sct:Schemas}
-   To design an information system, a developer defines a schema in which she defines concepts, relations, and rules.
-   Together, the concepts, relations, and rules capture the semantics of an information system~\cite{Spivak2012}.
+   To design an information system, a developer defines a schema in which she defines concepts, specializations, relations, and rules.
+   Together, these items capture the semantics of an information system~\cite{Spivak2012}.
    Each rule in a schema serves to constrain the dataset, to ensure its semantic integrity.
@@
    In this paper, variables $u$ and $v$ represent rules.
-   For every rule $u$ in the schema, we assume there is a predicate $\sat{u}{\Dataset}$
-   that says whether dataset $\dataset$ satisfies rule $u$.
+
+   A type system in Ampersand~\cite{vdWoude2011} identifies certain rules that are provably false.
+   It also ensures that all rules can be interpreted within the set of relations in the schema.
+   Due to this type system, the Ampersand compiler only generates code for scripts without type errors.
+   As a result, for every rule $u$ in the schema, there is a predicate $\sat{u}{\dataset}$ that says whether dataset $\dataset$ satisfies rule $u$.
+
 %    For every rule $u$ in the schema, we assume there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
@@
    i.e.\ constraints that the data in the database must satisfy at all times.
-   At times in which the dataset does not satisfy a rule, we say that the rule is {\em broken}.
+   At times in which the dataset does not satisfy a rule, that rule is said to be {\em broken}.
    In those cases, something or someone must take action to fix the situation and restore invariance of that rule.
 
-   A schema $\schema$ is a triple $\triple{\concepts}{\rels}{\rules}$,
+   A schema $\schema$ is a structure $\quadruple{\concepts}{\isa}{\rels}{\rules}$,
    in which $\concepts$ is a finite set of concepts,
+   $\isa$ is a partial order of concepts,
    $\rels$ is a finite set of relations,
    and $\rules$ is a finite set of rules.
-   We write $\schema_{\infsys}$ to refer to the schema of information system $\infsys$.
-   If needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$ to refer to $\concepts$, $\rels$, and $\rules$ respectively.
+   The notation $\schema_{\infsys}$ refers to the schema of information system $\infsys$.
+   Disambiguation sometimes requires to write $\concepts_{\schema}$, $\isa_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$
+   rather than $\concepts$, $\isa$, $\rels$, and $\rules$ respectively.
 %  Every schema is an element of an infinite set called $\Schema$.
@@
 \begin{eqnarray}
+   A\ \isa\ B&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
+   \label{eqn:isasIntroduceConcepts}\\
    \declare{n}{A}{B}\in\rels&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
@@
 \end{eqnarray}
-   If a triple $\triple{a}{r}{b}$ satisfies requirement~\ref{eqn:define R} and its relation $r$ satisfies requirement~{eqn:relationsIntroduceConcepts},
-   we say that it {\em befits} schema $\triple{\concepts}{\rels}{\rules}$.
-   We assume a special kind of rule that is used for specialization.
-   The rule $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
-\begin{equation}
-   \label{eqn:specialization}
-   \sat{A\isa B}{\dataset}\ \Leftrightarrow\ \forall a: a\ \inst\ A\rightarrow a\ \inst\ B
-\end{equation}
-   We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
-   Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.
-   The specialization rules in a schema form a partial order $\isa$,
-   meaning that $\isa$ is reflexive, transitive, and antisymmetric.
-   Specialization rules are special for two reasons:
-   Firstly, the system keeps $A\isa B$ satisfied automatically by changing $\inst$ whenever needed.
-   This means that a system that contains rule $A\isa B$ can never be brought into a state in which $\id{sat}_{A\isa B}$ is False.
-   This allows for certain run-time optimizations, which are beyond the scope of this paper.
-   A second property that sets specialization rules apart, is that they play a role in the Ampersand compiler.
-   For example, code that determines equality between atoms only works for atoms that are an instance of the same concept.
-   To ensure that this does not result in problems at run-time, users are required to write all specialization rules explicitly.
-
-   As a result, if an atom is an instance of concept $A$ and $A\isa B$,
-   this atom has all properties that atoms of type $B$ have:
-\begin{equation}
-   \begin{array}[b]{ll}
-      &a\ \inst\ A\wedge b\ \inst\ B\wedge A\isa B\\
-      \wedge&a=b\\
-      \wedge&b\ r\ x
-   \end{array}
-      \ \Rightarrow\ a\ r\ x
-\end{equation}
+   If a triple $\triple{a}{r}{b}$ satisfies requirement~\ref{eqn:define R} and its relation $r$ satisfies requirement~\ref{eqn:relationsIntroduceConcepts},
+   this triple is said to {\em befit} schema $\quadruple{\concepts}{\isa}{\rels}{\rules}$.
 
@@
    This section defines the notion of information system.
-   As before, we will suffix the elements of this definition if the situation requires it.
+   As before, a suffix disambiguates the elements of this definition when needed.
 \begin{definition}[information system]
 \label{def:information system}
-\item An information system $\infsys$ is a tuple $\la\dataset,\schema,\roles,\maintain\ra$, in which
+\item An information system $\infsys$ is a structure $\la\dataset,\schema,\roles,\maintain\ra$, in which
 \begin{itemize}
-   \item dataset $\dataset=\pair{\triples}{\inst}$ is defined as in section~\ref{sct:Datasets};
-   \item schema $\schema=\triple{\concepts}{\rels}{\rules}$ is defined as in section~\ref{sct:Schemas};
+   \item dataset $\dataset=\triple{\triples}{\isa}{\inst}$ is defined as in section~\ref{sct:Datasets};
+   \item schema $\schema=\quadruple{\concepts}{\isa}{\rels}{\rules}$ is defined as in section~\ref{sct:Schemas};
    \item $\roles$ is a set of roles;
@@
 \end{definition}
+   Note that dataset $\dataset$ and schema $\schema$ have the same specialization relation $\isa$.
    A \define{role} is a name that identifies a group of users.
@@
    The purpose of a role is to mention an individual user (human) or an automated actor (bot) without knowing who that user is.
-   In the sequel, when we talk about a role we actually mean an actor who fulfills that role in the system.
 
@@
 
-   In this paper we assume two ways of enforcing satisfaction of rules: automatic or manual.
+   This paper assumes two ways of enforcing satisfaction of rules: automatic or manual.
    Manual enforcement is specified by the developer by assigning a rule $u$ to a role $o$.
@@
    can restore invariance of rule $u$ by inserting or deleting the right triples.
-   For this reason, we assume that every rule is assigned to at least one role by means of the relation $\maintain$.
+   For this reason, every rule is assigned to at least one role by means of the relation $\maintain$.
 \begin{equation}
@@
    An enforce rule specifies not only the rule, but also the way to restore invariance.
-   We assume that the system features an engine that restores invariance of all enforce rules without unneccessary delay.
+   The system features an engine that restores invariance of all enforce rules without unneccessary delay.
    Whenever the dataset violates an enforce rule,
@@
    Having defined an information system in mathematical terms, let us discuss a small example.
-   For this purpose we use the language Ampersand
-   because it makes the examples more appealing to read.
+   It is written in the language Ampersand to make it more appealing to read.
    Let us first define a dataset of just a handful of triples and three relations.
@@
    The third and last relation states which students are enrolled for which module.
-   We leave it empty for now.
+   It is left empty for now.
 \begin{verbatim}
@@
 
-   To complete the semantics,
-   we also define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that student takes.
+   A rule, {\tt EnrollRule} completes the semantics
+   It states that a student can enroll for any module that is part of a course he or she takes.
    In Ampersand, which is a syntactically sugared form of relation algebra~\cite{JoostenRAMiCS2017},
-   we give each rule a name and declare a role to maintain its invariance:
+   each rule has a name and each rule has a role to maintain its invariance:
 \begin{verbatim}
@@
    \end{array}\]
-   The relation $\id{inst}$ contains the pairs:
+   The relation $\inst$ contains the pairs:
    \[\begin{array}{l}
@@
    \end{array}\]
-   The tuple $\pair{\triples}{\inst}$ satisfies requirement~\ref{eqn:wellTypedEdge} so this is a dataset $\dataset$ as introduced in section~\ref{sct:Datasets}.
+   The pair $\pair{\triples}{\inst}$ satisfies requirement~\ref{eqn:wellTypedEdge} so this is a dataset $\dataset$ as introduced in section~\ref{sct:Datasets}.
 
@@
    Requirement~\ref{eqn:specialization} is satisfied because this example contains no specialization.
-   So, the schema $\schema=\triple{\concepts}{\rels}{\rules}$ satisfies the requirements from section~\ref{sct:Schemas}.
+   So, the schema $\schema=\quadruple{\concepts}{\isa}{\rels}{\rules}$ satisfies the requirements from section~\ref{sct:Schemas}.
 
@@
    This completes the requirements for $\infsys=\la\dataset,\schema,\roles,\maintain\ra$ from definition~\ref{def:information system}.
-   So we can conclude that $\infsys$ is an  information system.
+   This concludes the argument that $\infsys$ is an information system.
 
@@
          for instance to initialize new features on the moment of transition (MoT).
-         If schemas $\schema$ equals $\schema'$, we call the migration trivial because a data migration is not necessary.
+         If schemas $\schema$ equals $\schema'$, the data migration is trivial because it is not necessary.
    \item The generator generates a migration script, $\migrsys$,
          to migrate the data and define the semantics during the migration phase as far as a generator can reasonably predict.
-         It generates a schema, $\schema_{\migrsys}$, which contains the enforce rules that migrate existing data to the desired dataset.
+         It contains the enforce rules that migrate existing data to the desired dataset.
          It also contains data structures and rules that are needed for the duration of the migration phase.
+         The generated script must be free of type errors, to ensure that $\sat{u}{\Dataset}$ exists for every rule $u$ and dataset $\dataset$.
+         The proof is available \href{https://www.isa-afp.org/}{here}
    \item\label{item3}
@@
    \item To give the business an opportunity to evaluate the change,
-         we can require a go/no-go decision before the the migration engineer makes the migration irreversible.
+         there is room for a go/no-go decision before the the migration engineer makes the migration irreversible.
          This event marks the moment of no return (MoN).
@@
          In case of a no-go, the migration engineer restores the existing system by letting the ingress switch traffic back to $\infsys$.
-   \item In more complex cases, the migration system contains rules that must be satisfied during the migration.
+   \item In more complex cases, the migration script contains rules that must be satisfied during the migration.
          These rules can be necessary, for example to resolve data pollution that violates a constraint that was not in schema $\schema_{\infsys}$.
@@
          Or she can use the business to satisfy this rule, in case human intelligence is required.
-         Whatever the choice, we require that every rule from $\rules_{\migrsys'}$ is satisfied before the migration engineer takes down the migration system.
+         Whatever the choice, every rule from $\rules_{\migrsys'}$ must be satisfied before the migration engineer takes down the migration system.
    \item After taking down the existing system and the migration system, the desired system remains in production and the migration is finished.
@@
 \begin{eqnarray}
-   \triple{\concepts}{\rels}{\rules}\sqcup\triple{\concepts'}{\rels'}{\rules'}&=&\triple{\concepts\uplus\concepts}{\rels\uplus\rels'}{\rules\uplus\rules'}
+   \quadruple{\concepts}{\isa}{\rels}{\rules}\sqcup\quadruple{\concepts'}{\isa'}{\rels'}{\rules'}&=&\quadruple{\concepts\uplus\concepts}{\isa\uplus\isa'}{\rels\uplus\rels'}{\rules\uplus\rules'}
 \end{eqnarray}
@@
 
-To implement the disjoint union, we need to relabel 
+To implement the disjoint union, the following relabelling functions are needed:
 \begin{definition}[relabel concepts]
@@
    \begin{eqnarray}
-      \rlap{$\subst{C}{D}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
+      \rlap{$\subst{C}{D}{\quadruple{\concepts}{\isa}{\rels}{\rules}}$}\notag\\
       &=&\begin{array}[t]{l@{}l}
@@
    \begin{eqnarray}
-      \rlap{$\subst{r'}{m'}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
-      &=&\la\concepts,\{\subst{r'}{m'}{r} \mid r\in\rels\}, \{\subst{r'}{m'}{u} \mid u\in\rules\}\ \ra\notag
+      \rlap{$\subst{r'}{m'}{\quadruple{\concepts}{\isa}{\rels}{\rules}}$}\notag\\
+      &=&\quadruple{\concepts}{\isa}{\{\subst{r'}{m'}{r} \mid r\in\rels\}}{\{\subst{r'}{m'}{u} \mid u\in\rules\}}\notag
    \end{eqnarray}
@@
 \subsection{Loading Data}
-   For loading the data in the desired database, we use enforce rules.
+   Enforce rules can load the data in the desired database.
    In Ampersand, an enforce rule comes in three flavors:
@@
    Let us proceed to specify a desired system, $\infsys'$, to illustrate a (toy) migration.
-   We will use the example in section~\ref{sct:Example existing IS} as the existing system
-   and call that $\infsys$.
+   The example in section~\ref{sct:Example existing IS}, called $\infsys$, serves as the existing system.
    Its population reflects the state of the system just before the migration.
@@
 \end{verbatim}
-   We can see that the relations and the rule of the existing system remain part of the desired system%
+   In this example, the relations and the rule of the existing system remain part of the desired system%
 \footnote{Removal or changing of relations is allowed as well.}
@@
 
-\subsection{Migration system}
-   The generated migration system, $\migrsys$, is:
+\subsection{Migration script}
+   The generated migration script, $\migrsys$, is:
 \begin{verbatim}
@@
    Just before loading the desired system with data,
-   the data in the existing system is no longer the data we saw in section~\ref{sct:Example existing IS}.
+   the data in the existing system has changed.
+   So, the data we saw in section~\ref{sct:Example existing IS} has most likely changed.
    Let us assume the actual population at this time is:
@@
 \end{verbatim}
-   If we check this script, we get the following results:
+   When Ampersand checks this script, it produces the following results:
 \begin{verbatim}
@@
    Any data migration must anticipate any population in the existing system.
-   The only thing we know for sure about the existing population is that it satisfies the rules in $\infsys$.
+   The only certainty is that the existing population satisfies the rules in $\infsys$.
 
@@
 However, a problem that may occur is that regular users of the system, who are used to the existing system, might be adding data to the system that causes violations faster than they can be solved.
-To mitigate this, we propose two solutions.
+Two solutions can mitigate this.
 
@@
 The \verb=ENFORCE= statement at the end of this code snippet states that the relation recording existing violations should be shrunk whenever violations are solved.
-This way, we ensure that violations cannot re-occur.
+This way, violations cannot re-occur.
 As a whole, this means that \verb=migrationHelper= has a finite task.
 
-As a second solution, we observe that the data in the designed system does not have any violations.
+As a second solution, the data in the designed system does not have any violations.
 Consequentially, the cause of the violations comes from the triples in the existing system.
-Rather than taking the union of all triples, we can take the disjoint union instead;
+Rather than taking the union of all triples, the disjoint union is taken instead;
 \begin{verbatim}
@@
 In this second solution, we immediately move to a system in which the \verb=UNI= rule is maintained by the system.
-The work for the \verb=migrationHelper= is again finite: the tuples in \verb=isPartOf_old= need to be copied over or explicitly ignored.
-Once a tuple is in \verb=isPartOf=, the tuple is copied over to \verb=isPartOf_ignored= by the \verb=ENFORCE= statement, such that regular users can delete it without adding work to anyone with the \verb=migrationHelper= role.
+The work for the \verb=migrationHelper= is again finite: the pairs in \verb=isPartOf_old= need to be copied over or explicitly ignored.
+Once a pair is in \verb=isPartOf=, the pair is copied over to \verb=isPartOf_ignored= by the \verb=ENFORCE= statement, such that regular users can delete it without adding work to anyone with the \verb=migrationHelper= role.
 Of course, since this solution starts with fewer pairs in the \verb=isPartOf= relation, other rules could be violated as a result of choosing this solution.
@@
    % -> het migratie-systeem is typefout-vrij (opmerking: als we de type-checker buiten beschouwing willen laten, kunnen we dit niet beschrijven)
-   \item the migration system $\migrsys$ is free of type errors.
+   \item the migration script $\migrsys$ is free of type errors.
    % -> het migratie-systeem is vrij van overtredingen op regels die het migratie-systeem moet bewaken

### 2023-02-12 2023 Disambiguation/Disambiguation.tex
Article on disambiguation

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{bbold}
\usepackage[cal=boondoxo]{mathalfa}
\begin{document}

\title{What's in a name? Disambiguation strategies for relational descriptions}
\author{Sebastiaan Joosten}
\date{Work in progress}
\maketitle

\section{Motivation}
Notations matter: They help communicate ideas by establishing expectations.
When programming, it is recommended to choose function symbols, variable names, etcetera, in a way that it helps the reader of the program.
Notations refer to a certain semantical notion, and choosing a good notation can help establish that.
Regardless of expectations, the thing a notation refers to must always be clear.

Programming languages greatly influence the notations a programmer uses through the rules on what a notation might refer to.
This paper looks at how identifiers that refer to binary relations might be used in a semantic modeling language.
In particular, we look at the use of identifiers in the programming language Ampersand, and ways in which changes to the language might effect the use of identifiers.

Modeling in Ampersand happens via the use of binary relations, which are declared by stating a name, a source concept, and a target concept.
When referring to binary relations, we typically only refer to them by their name.
It is desirable to allow different binary relations to have the same name:
Allowing this gives the programmer greater flexibility in which names to use.

\subsection{Running example}
An application can be described by binary relations over concepts, a description of what the concepts and relations mean, stating invariants and process rules that determine the behavior of the system, and describing interfaces to interact with the system.
For the purposes of this article, we just declare binary relations and some invariants.

We give the following planning application as an example in which binary relations share a name:
\begin{verbatim}
before : Task * Task
before : Time * Time
scheduled : Task * Time
\end{verbatim}

An invariant in this system is:

\[
\mathrm{before}_{\mathrm{Task * Task}};\mathrm{scheduled}_{\mathrm{Task * Time}} \subseteq \mathrm{scheduled};\mathrm{before}_{\mathrm{Time * Time}}
\]

Such invariants are well-typed: a Task is only compared to a Task, and a Time is only compared to a Time, and the only way in which a Task and a Time are related is through explicitly declared relations.
We can consider the same expression in which types are dropped: $\mathrm{before};\mathrm{scheduled} \subseteq \mathrm{scheduled};\mathrm{before}$.
If we require that invariants are well-typed, there is only one way of annotating such expressions with types, regardless of our interpretation.

In Ampersand's implementation, Ampersand does not require writing the types: the use of the possibly ambiguous relation name $\mathrm{before}$ is allowed as long as it is clear which relation is intended, as is the case here.
We refer to the problem of determining which relation is intended as the disambiguation problem, which we investigate more closely in a separate section.

\subsection{Generalization}
A common task in semantic modeling is to identify that some concepts are generalizations or specializations of other concepts.
This section motivates the need for a special language feature to deal with generalizations.

As an example, our planning application allows regular working hours to be set, which means that some times fall during regular hours.
This means that such times would be a specialization of a time.
We can model this with a relation called $\mathrm{worktime}$ and an invariant: $\mathrm{worktime} \subseteq \mathbb{1}_{Time}$, where $\mathbb{1}_{A} = \{(a,a) \mid a\in A \}$ is the relational identity.
Another way to model this, is to introduce a concept $\mathrm{Worktime}$ and use $\mathbb{1}_\mathrm{Worktime}$ in the place of the relation $\mathrm{worktime}$, along with the invariant $\mathbb{1}_\mathrm{Worktime} \subseteq \mathbb{1}_{Time}$.
However, this latter invariant is ill-typed: the expression on the left of the subset ($\subseteq$) is of type $(\mathrm{Worktime},\mathrm{Worktime})$, while the expression on the right is of type $(\mathrm{Time},\mathrm{Time})$.
To allow expressions like $\mathbb{1}_\mathrm{Worktime} \subseteq \mathbb{1}_{Time}$ in a programming language, we need a more refined notion of which expressions are `type correct'.
A paper by van der Woude and Joosten shows how one can do this~\cite{Woude11}.
However, that paper does not consider the interplay with disambiguation which we consider here.

\subsection{Two interpretations.}
There are two different ways in which we can think of the two binary relations called $\mathrm{before}$.
In one interpretation, we can say that $\mathrm{before}$ is a single binary relation, defined over the (disjoint) union between Task and Time.
Whenever we use $\mathrm{before}$, we restrict its type to either \verb=Task * Task= or \verb=Time * Time= to get two completely separate behaviors.
This means that we read $\mathrm{before}_{\mathrm{Task * Task}}$ as $\mathrm{before} \cap (\mathrm{Task}\times\mathrm{Task})$.

Another interpretation is to say that there are two entirely separate relations, $\mathrm{before}_{\mathrm{Task * Task}}$ and $\mathrm{before}_{\mathrm{Time * Time}}$.
In this interpretation, we can consider ${\mathrm{Task * Task}}$ as a part of the name of the relation.

Note our interpretation does not matter for the behavior we get from well-typed invariants like the one we saw before.
However, a problem arises if we add generalization.

If we were to state - for some obscure reason - that every Task is a Time, in our first interpretation we would expect that $\mathrm{before}_{\mathrm{Task * Task}} = \mathrm{before}_{\mathrm{Time * Time}} \cap (\mathrm{Task}\times\mathrm{Task})$, since $(\mathrm{Task}\times\mathrm{Task}) \subseteq (\mathrm{Time}\times\mathrm{Time})$.
In our second interpretation, we would not expect the behavior to change.

Observe that we are free to add $\mathrm{before}_{\mathrm{Task * Task}} \subseteq \mathrm{before}_{\mathrm{Time * Time}}$, or the stronger statement $\mathrm{before}_{\mathrm{Task * Task}} = \mathrm{before}_{\mathrm{Time * Time}} \cap (\mathrm{Task}\times\mathrm{Task})$, as an invariant under the second interpretation.
Therefore, we conclude that working with the second interpretation is most flexible and propose to follow this interpretation.

At the time of writing, Ampersand infers the invariant $\mathrm{before}_{\mathrm{Task * Task}} \subseteq \mathrm{before}_{\mathrm{Time * Time}}$ without requiring that $\mathrm{before}_{\mathrm{Task * Task}} = \mathrm{before}_{\mathrm{Time * Time}} \cap (\mathrm{Task}\times\mathrm{Task})$. This can be seen as a sort of middle ground between the two interpretations. However, we foresee that Ampersand will change its implementation and do not follow its behavior here.

\section{Solving the disambiguation problem without generalization}
As an aside, we ask whether a compiler can be expected to figure out which relations are intended in reasonable time. In this section, we answer this question for a very simple setting in the affirmative by giving a simple efficient algorithm.

We can define the disambiguation problem as follows:

A disambiguation problem is given by an expression $E$ using the operations $\cap$ and $;$, and relation names from a set of names $\mathcal{N}$, and a set of possible types of each name: $\mathcal{t}(n) \subseteq \mathcal{T}\times \mathcal{T}$ for $n\in \mathcal{N}$.
A solution to the disambiguation problem is a well-typed expression $S$ in which the annotations to relation names corresponds to an element in $\mathcal{t}(n)$, such that dropping the types from the expression yields $E$.

As an example, $E = (\mathrm{before};\mathrm{scheduled}) \cap (\mathrm{scheduled};\mathrm{before})$ with types $\mathcal{t}(\mathrm{before}) = \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ and $\mathcal{t}(\mathrm{scheduled}) = \{(\mathrm{Task}, \mathrm{Time})\}$, is a disambiguation problem.
A solution would be:
\[\left(\mathrm{before}_{\mathrm{Task * Task}};\mathrm{scheduled}_{\mathrm{Task * Time}}\right) \cap \left(\mathrm{scheduled};\mathrm{before}_{\mathrm{Time * Time}}\right)\]

We can solve the disambiguation problem using two passes through $E$.
In the first pass, we decorate the expression tree for $E$ with all the possible types for each sub-expression bottom up.
Let $p(e)$ denote the possible types, then $p(n) = \mathcal{t}(n)$, and $p(e_1;e_2) = \{(x,y) \mid \exists z. (x,z) \in p(e_1) \wedge (z,y) \in p(e_2))\}$, and $p(e_1 \cap e_2) = p(e_1) \cap p(e_2)$.
If $p(E)$ is the empty set, then there is no solution to the disambiguation problem, and we are done.
If $p(E)$ contains multiple elements, then there are multiple solutions to the disambiguation problem.
We can arbitrarily select a single element from $p(E)$ to get an arbitrary solution\footnote{In practice, we might want to indicate that $E$ is ambiguous, which could require running the next step twice: once for two different elements in $p(E)$.}.

For the top-down step, let $d(e,t)$ denote the disambiguated expression $e$ with the type $t$.
For the top-level call of $d(e,t)$, let $e=E$ and $t\in p(E)$.
Then $d(n,t) = n_{[t]}$, $d(e_1 \cap e_2, t) = d(e_1,t) \cap d(e_2,t)$ and $d(e_1 ; e_2, (x,y)) = d(e_1, (x,z)) ; d(e_2, (z,y))$ where $z$ is picked such that $(x,z)\in p(e_1)$ and $(z,y)\in p(e_2)$.
Such a $z$ is guaranteed to exist by how $p$ is defined.

As an example, let's look at $E = (\mathrm{before};\mathrm{scheduled}) \cap (\mathrm{scheduled};\mathrm{before})$.
Then $p(\mathrm{before}) =  \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ following $\mathcal{t}$ and similarly $p(\mathrm{scheduled}) = \{(\mathrm{Task}, \mathrm{Time})\}$.
This gives subterms $p(\mathrm{before};\mathrm{scheduled}) = p(\mathrm{scheduled};\mathrm{before}) = \{(\mathrm{Task}, \mathrm{Time})\}$ so $p(E) = \{(\mathrm{Task}, \mathrm{Time})\}$.

Top-down, $d(E,t) = d(\mathrm{before};\mathrm{scheduled},t) \cap d(\mathrm{scheduled};\mathrm{before},t)$ with $t=(\mathrm{Task}, \mathrm{Time})$.
Since $E$ is not ambiguous, we have only one possible way to continue as we solve the disambiguation problem:
\begin{align*}d(\mathrm{before};\mathrm{scheduled},t) &= d(\mathrm{before},(\mathrm{Task}, \mathrm{Task})) ;  d(\mathrm{scheduled},(\mathrm{Task}, \mathrm{Time})) \\
&=\mathrm{before}_{\mathrm{Task} * \mathrm{Task}}; \mathrm{scheduled}_{\mathrm{Task}* \mathrm{Time}}\\
d(\mathrm{scheduled};\mathrm{before},t) &=  d(\mathrm{scheduled},(\mathrm{Task}, \mathrm{Time}));d(\mathrm{before},(\mathrm{Time}, \mathrm{Time}))\\
&=\mathrm{scheduled}_{\mathrm{Time} * \mathrm{Task}}; \mathrm{before}_{\mathrm{Time}* \mathrm{Time}}\\
\end{align*}

The disambiguation problem and the corresponding algorithm can be extended over all operations in relation algebra.

\section{Interplay between Generalization and Disambiguation}

% van der Woude paper: https://link.springer.com/chapter/10.1007/978-3-642-21070-9_25
\end{document}

### 2023-02-12 Generative Software/proposal.tex
wip

@@
     hereby supporting agile software development methods.
+
+    The ambition of this research is to contribute to solving the problem of large IT-projects.
+    By generating fully functional information systems, we aim to reduce the effort and innovation risk of such projects.
+    The direction of research goes toward more frequent deployments,
+    which enables large projects to deliver incremental results in days rather than months.
     %every 2 or 3 weeks as it develops into an end product.
@@
 
-\section{Requirements (remove this section before submitting)}
-Reviewers will judge this proposal by the following requirements.
-\subsection{Scientific quality}
-\begin{enumerate}
-   \item 	To what extent is the proposed research original and how would you rate the innovative elements?
-   \item 	What is your assessment of the design of the project, including the goals, hypotheses, research methods, and scientific feasibility?
-   \item 	What is your assessment of the coherence and time schedule of the proposed lines of research?
-   \item 	Is the research group competent enough to carry out the research? Does the group have a relevant position in the international scientific community? Is the available infrastructure adequate?
-   \item 	Are the number and category of requested personnel, budget for materials, investments, and foreign travel adequate?
-   \item 	Does the proposed research fit in the Open Technology Programme; does it focus on engineering sciences research?
-   \item 	What are the strong and weak points of the scientific part of the proposal?   
-\end{enumerate}
+% \section{Requirements (remove this section before submitting)}
+% Reviewers will judge this proposal by the following requirements.
+% \subsection{Scientific quality}
+% \begin{enumerate}
+%    \item 	To what extent is the proposed research original and how would you rate the innovative elements?
+%    \item 	What is your assessment of the design of the project, including the goals, hypotheses, research methods, and scientific feasibility?
+%    \item 	What is your assessment of the coherence and time schedule of the proposed lines of research?
+%    \item 	Is the research group competent enough to carry out the research?
+Does the group have a relevant position in the international scientific community?
+Is the available infrastructure adequate?
+%    \item 	Are the number and category of requested personnel, budget for materials, investments, and foreign travel adequate?
+%    \item 	Does the proposed research fit in the Open Technology Programme; does it focus on engineering sciences research?
+%    \item 	What are the strong and weak points of the scientific part of the proposal?   
+% \end{enumerate}
 
-\subsection{Knowledge Utilisation} (the application of the results of the research by third parties)
-\begin{enumerate}
-   \item  	What is your assessment of the description of the commercial and/or societal potential impacts of the research given in the proposal?
-   \item 	What is your assessment of the contribution and commitment of the users and the proposed composition of the user committee?
-   \item 	What is your assessment of the freedom to operate concerning commercial propositions, existing patents, eligibility or societal acceptance? 
-   \item 	What are the prospects for collaboration with the industry and knowledge transfer, assuming the project is successful? Please address both aspects.
-   \item 	What is your assessment of the research group's competence regarding the transfer and application of research results?
-   \item 	What are the strong and weak points of the utilisation plan?   
-\end{enumerate}
+% \subsection{Knowledge Utilisation} (the application of the results of the research by third parties)
+% \begin{enumerate}
+%    \item  	What is your assessment of the description of the commercial and/or societal potential impacts of the research given in the proposal?
+%    \item 	What is your assessment of the contribution and commitment of the users and the proposed composition of the user committee?
+%    \item 	What is your assessment of the freedom to operate concerning commercial propositions, existing patents, eligibility or societal acceptance? 
+%    \item 	What are the prospects for collaboration with the industry and knowledge transfer, assuming the project is successful? Please address both aspects.
+%    \item 	What is your assessment of the research group's competence regarding the transfer and application of research results?
+%    \item 	What are the strong and weak points of the utilisation plan?   
+% \end{enumerate}
 
@@
    On the other hand, one-of-a-kind organizations require one-of-a-kind components that are not readily available.
-   The so called ``low code an no code platforms'' are sometimes advocated as the solution
+   The so called ``low code and no code platforms'' are sometimes advocated as the solution
    because their (re-)sellers claim that the business itself can configure such systems.
    However, we have seen some disappointing results of that in practice, because of integration problems and maintainability issues.
-   So homebrew solutions are still widespread in organizations with one-of-a-kind requirements.
+   So custom-built solutions are still widespread in organizations with one-of-a-kind requirements.
    They are still willing to invest the time to build and maintain information systems and take the risk of a failing IT-innovation.
 
-   Generative development can shorten the development time and reduce the development risks of homebrew software,
+   Generative development can shorten the development time and reduce the development risks of custom-built software,
    regaining the desired flexibility and satisfying specific one-of-a-kind requirements at the same time.

### 2023-02-12 2022 Migration/articleMigration.tex
wip

@@
 \begin{abstract}
-   To support incremental software development with a software generator bears the potential of decreasing the time to deploy changes.
-   However, an increment that changes the database schema causes a nontrivial data migration problem.
+   To support incremental software development with a software generator bears the potential of saving work and decreasing the time to deploy changes.
+   However, increments that change the database schema cause nontrivial data migration issues.
    The problem is how to preserve the semantics of the data as much as possible and satisfy elementary business requirements at the same time.

### 2023-02-20 2023 Disambiguation/Disambiguation.tex
Bas, kun jij verifieren of ik je goed snap?

@@
 \usepackage[cal=boondoxo]{mathalfa}
+\newcommand{\id}[1]{\text{\it #1\/}}
+\newcommand{\Relations}{\mathbb{R}}
+\newcommand{\inst}{\mathbin{\id{inst}}}
+\newcommand{\pair}[2]{\langle{#1},{#2}\rangle}
 \begin{document}
@@
 This paper looks at how identifiers that refer to binary relations might be used in a semantic modeling language.
-In particular, we look at the use of identifiers in the programming language Ampersand, and ways in which changes to the language might effect the use of identifiers.
+In particular, we look at the use of identifiers in the programming language Ampersand, and ways in which changes to the language might affect the use of identifiers.
 
@@
 
-\[
-\mathrm{before}_{\mathrm{Task * Task}};\mathrm{scheduled}_{\mathrm{Task * Time}} \subseteq \mathrm{scheduled};\mathrm{before}_{\mathrm{Time * Time}}
-\]
+\begin{eqnarray}
+    \mathrm{before}_{\mathrm{Task * Task}};\mathrm{scheduled}_{\mathrm{Task * Time}}&\subseteq&\mathrm{scheduled};\mathrm{before}_{\mathrm{Time * Time}}
+    \label{eqn:well-typed example}
+\end{eqnarray}
 
@@
 However, this latter invariant is ill-typed: the expression on the left of the subset ($\subseteq$) is of type $(\mathrm{Worktime},\mathrm{Worktime})$, while the expression on the right is of type $(\mathrm{Time},\mathrm{Time})$.
-To allow expressions like $\mathbb{1}_\mathrm{Worktime} \subseteq \mathbb{1}_{Time}$ in a programming language, we need a more refined notion of which expressions are `type correct'.
-A paper by van der Woude and Joosten shows how one can do this~\cite{Woude11}.
+To allow this in a programming language, we let the user declare that $\mathrm{Worktime}$ is a specialization of $\mathrm{Time}$
+\begin{verbatim}
+CLASSIFY Worktime ISA Time
+\end{verbatim}
+The meaning of this statement is:
+\begin{eqnarray}
+    \forall t&:&t\in{\tt Worktime}\Rightarrow t\in{\tt Time}
+\end{eqnarray}
+Specialization requires a more refined notion of which expressions are `type correct'.
+A paper by van der Woude and Joosten shows how one can do this~\cite{vdWoude2011}.
 However, that paper does not consider the interplay with disambiguation which we consider here.
@@
 \subsection{Two interpretations.}
+Let us go back to the running example, in which we have:
+\begin{verbatim}
+before : Task * Task
+before : Time * Time
+\end{verbatim}
 There are two different ways in which we can think of the two binary relations called $\mathrm{before}$.
-In one interpretation, we can say that $\mathrm{before}$ is a single binary relation, defined over the (disjoint) union between Task and Time.
+In one interpretation, we can say that $\mathrm{before}$ is a single binary relation, say $r$, defined over the (disjoint) union between Task and Time.
 Whenever we use $\mathrm{before}$, we restrict its type to either \verb=Task * Task= or \verb=Time * Time= to get two completely separate behaviors.
@@
 
-Another interpretation is to say that there are two entirely separate relations, $\mathrm{before}_{\mathrm{Task * Task}}$ and $\mathrm{before}_{\mathrm{Time * Time}}$.
+Bas, ik lees dat als:
+\begin{eqnarray}
+    \pair{a}{b}\in r&\Rightarrow&a\in\mathrm{Time}\cup\mathrm{Task}\wedge b\in\mathrm{Time}\cup\mathrm{Task}\\
+    \mathrm{before}_{\mathrm{Time * Time}}&=&r\cap (\mathrm{Time}\times\mathrm{Time})\\
+    \mathrm{before}_{\mathrm{Task * Task}}&=&r\cap (\mathrm{Task}\times\mathrm{Task})
+\end{eqnarray}
+Another interpretation is to say that there are two entirely separate relations,
+$r:\mathrm{before}_{\mathrm{Time * Time}}$ and $r':\mathrm{before}_{\mathrm{Task * Task}}$.
 In this interpretation, we can consider ${\mathrm{Task * Task}}$ as a part of the name of the relation.
 
-Note our interpretation does not matter for the behavior we get from well-typed invariants like the one we saw before.
+Bas, ik lees dat als:
+\begin{eqnarray}
+    \pair{a}{b}\in r&\Rightarrow&a\in\mathrm{Time}\wedge b\in\mathrm{Time}\\
+    \pair{a}{b}\in r'&\Rightarrow&a\in\mathrm{Task}\wedge b\in\mathrm{Task}\\
+    \mathrm{before}_{\mathrm{Time * Time}}&=&r\\
+    \mathrm{before}_{\mathrm{Task * Task}}&=&r'
+\end{eqnarray}
+
+Note our interpretation does not matter for the behavior we get from well-typed invariants like equation~\ref{eqn:well-typed example}.
 However, a problem arises if we add generalization.
 
-If we were to state - for some obscure reason - that every Task is a Time, in our first interpretation we would expect that $\mathrm{before}_{\mathrm{Task * Task}} = \mathrm{before}_{\mathrm{Time * Time}} \cap (\mathrm{Task}\times\mathrm{Task})$, since $(\mathrm{Task}\times\mathrm{Task}) \subseteq (\mathrm{Time}\times\mathrm{Time})$.
+If we were to state - for some obscure reason - that every Task is a Time, this adds:
+\begin{equation}
+    t\in\mathrm{Task}\ \Rightarrow\ t\in\mathrm{Time}
+\end{equation}
+In our first interpretation we would expect that $\mathrm{before}_{\mathrm{Task * Task}} = \mathrm{before}_{\mathrm{Time * Time}} \cap (\mathrm{Task}\times\mathrm{Task})$, since $(\mathrm{Task}\times\mathrm{Task}) \subseteq (\mathrm{Time}\times\mathrm{Time})$.
 In our second interpretation, we would not expect the behavior to change.
@@
 
+\section{Bibliography}
+\bibliographystyle{elsarticle-harv}
+\bibliography{doc}
+
 % van der Woude paper: https://link.springer.com/chapter/10.1007/978-3-642-21070-9_25

### 2023-02-25 2023 Disambiguation/Disambiguation.tex
Update disambiguation paper: interplay section

--- 2023-02-12 2023 Disambiguation/Disambiguation.tex
@@
 \documentclass[12pt]{article}
-\usepackage{amsmath}
-\usepackage{amssymb}
+\usepackage[utf8]{inputenc}
+\usepackage{graphicx,stmaryrd}
+\usepackage{amsmath,amsfonts,amssymb}
+\usepackage{bbold,bbm}
+\makeatletter
+\def\amsbb{\use@mathgroup \M@U \symAMSb}
+\makeatother
+\usepackage[cal=boondoxo]{mathalfa}
+\newcommand{\typesemi}{\mathbin{\text{\textbb{;}}}}
+\newcommand{\typesubset}{\mathbin{\mathbbm{\sqsubseteq}}}
+\newcommand{\typeunion}{\mathbin{\mathbb{U}}}
+\newcommand{\typeinter}{\mathbin{\text{\rotatebox[origin=c]{180}{$\typeunion$}}}}
+\newcommand{\typevee}{\mathbbm{T}}
+\newcommand{\typenil}{\mathbbm{F}}
+\newcommand{\typecomp}[1]{\neg{#1}}
+\newcommand{\typeconv}[1]{{#1}^\smallsmile}
+\newcommand{\typeident}{\mathbb{1}}
+\newcommand{\typetyped}[2]{{#1}_{\llbracket #2 \rrbracket}}
+\newcommand{\conv}[1]{{#1}^{-1}}
 \usepackage{algorithmic}
-\usepackage{bbold}
-\usepackage[cal=boondoxo]{mathalfa}
 \begin{document}
@@
 
-\section{Motivation}
+\section{Introduction}
 Notations matter: They help communicate ideas by establishing expectations.
@@
 
-\subsection{Running example}
+\subsection*{Running example}
 An application can be described by binary relations over concepts, a description of what the concepts and relations mean, stating invariants and process rules that determine the behavior of the system, and describing interfaces to interact with the system.
@@
 
-We give the following planning application as an example in which binary relations share a name:
+We give the following planning application as an example in which two binary relations share a name:
 \begin{verbatim}
@@
 before : Time * Time
-scheduled : Task * Time
+scheduled : Event * Time
 \end{verbatim}
@@
 \[
-\mathrm{before}_{\mathrm{Task * Task}};\mathrm{scheduled}_{\mathrm{Task * Time}} \subseteq \mathrm{scheduled};\mathrm{before}_{\mathrm{Time * Time}}
+\mathrm{before}_{\mathrm{Task * Task}}\typesemi\mathrm{scheduled}_{\mathrm{Event * Time}} \typesubset \mathrm{scheduled}_\mathrm{Event * Time}\typesemi\mathrm{before}_{\mathrm{Time * Time}}
 \]
 
-Such invariants are well-typed: a Task is only compared to a Task, and a Time is only compared to a Time, and the only way in which a Task and a Time are related is through explicitly declared relations.
-We can consider the same expression in which types are dropped: $\mathrm{before};\mathrm{scheduled} \subseteq \mathrm{scheduled};\mathrm{before}$.
-If we require that invariants are well-typed, there is only one way of annotating such expressions with types, regardless of our interpretation.
+In heterogeneous relation algebra, where composition is partial, such an invariant would not be well-defined: A Task cannot be compared to an Event, so there is a type error in $\mathrm{before}_{\mathrm{Task * Task}}\typesemi\mathrm{scheduled}_{\mathrm{Event * Time}}$.
+However, Ampersand allows sub-typing.
+The term $\mathrm{before}_{\mathrm{Task * Task}}\typesemi\mathrm{scheduled}_{\mathrm{Event * Time}}$ will be accepted if there is a least upper bound for the concepts Event and Task.
+In this example, we will say that Event is that least upper bound by stating that every Task is an Event.
+In Ampersand Syntax:
 
-In Ampersand's implementation, Ampersand does not require writing the types: the use of the possibly ambiguous relation name $\mathrm{before}$ is allowed as long as it is clear which relation is intended, as is the case here.
-We refer to the problem of determining which relation is intended as the disambiguation problem, which we investigate more closely in a separate section.
+\begin{verbatim}
+CLASSIFY Task ISA Event
+\end{verbatim}
 
-\subsection{Generalization}
-A common task in semantic modeling is to identify that some concepts are generalizations or specializations of other concepts.
-This section motivates the need for a special language feature to deal with generalizations.
-
-As an example, our planning application allows regular working hours to be set, which means that some times fall during regular hours.
-This means that such times would be a specialization of a time.
-We can model this with a relation called $\mathrm{worktime}$ and an invariant: $\mathrm{worktime} \subseteq \mathbb{1}_{Time}$, where $\mathbb{1}_{A} = \{(a,a) \mid a\in A \}$ is the relational identity.
-Another way to model this, is to introduce a concept $\mathrm{Worktime}$ and use $\mathbb{1}_\mathrm{Worktime}$ in the place of the relation $\mathrm{worktime}$, along with the invariant $\mathbb{1}_\mathrm{Worktime} \subseteq \mathbb{1}_{Time}$.
-However, this latter invariant is ill-typed: the expression on the left of the subset ($\subseteq$) is of type $(\mathrm{Worktime},\mathrm{Worktime})$, while the expression on the right is of type $(\mathrm{Time},\mathrm{Time})$.
-To allow expressions like $\mathbb{1}_\mathrm{Worktime} \subseteq \mathbb{1}_{Time}$ in a programming language, we need a more refined notion of which expressions are `type correct'.
-A paper by van der Woude and Joosten shows how one can do this~\cite{Woude11}.
-However, that paper does not consider the interplay with disambiguation which we consider here.
-
-\subsection{Two interpretations.}
-There are two different ways in which we can think of the two binary relations called $\mathrm{before}$.
-In one interpretation, we can say that $\mathrm{before}$ is a single binary relation, defined over the (disjoint) union between Task and Time.
-Whenever we use $\mathrm{before}$, we restrict its type to either \verb=Task * Task= or \verb=Time * Time= to get two completely separate behaviors.
-This means that we read $\mathrm{before}_{\mathrm{Task * Task}}$ as $\mathrm{before} \cap (\mathrm{Task}\times\mathrm{Task})$.
-
-Another interpretation is to say that there are two entirely separate relations, $\mathrm{before}_{\mathrm{Task * Task}}$ and $\mathrm{before}_{\mathrm{Time * Time}}$.
-In this interpretation, we can consider ${\mathrm{Task * Task}}$ as a part of the name of the relation.
-
-Note our interpretation does not matter for the behavior we get from well-typed invariants like the one we saw before.
-However, a problem arises if we add generalization.
-
-If we were to state - for some obscure reason - that every Task is a Time, in our first interpretation we would expect that $\mathrm{before}_{\mathrm{Task * Task}} = \mathrm{before}_{\mathrm{Time * Time}} \cap (\mathrm{Task}\times\mathrm{Task})$, since $(\mathrm{Task}\times\mathrm{Task}) \subseteq (\mathrm{Time}\times\mathrm{Time})$.
-In our second interpretation, we would not expect the behavior to change.
-
-Observe that we are free to add $\mathrm{before}_{\mathrm{Task * Task}} \subseteq \mathrm{before}_{\mathrm{Time * Time}}$, or the stronger statement $\mathrm{before}_{\mathrm{Task * Task}} = \mathrm{before}_{\mathrm{Time * Time}} \cap (\mathrm{Task}\times\mathrm{Task})$, as an invariant under the second interpretation.
-Therefore, we conclude that working with the second interpretation is most flexible and propose to follow this interpretation.
-
-At the time of writing, Ampersand infers the invariant $\mathrm{before}_{\mathrm{Task * Task}} \subseteq \mathrm{before}_{\mathrm{Time * Time}}$ without requiring that $\mathrm{before}_{\mathrm{Task * Task}} = \mathrm{before}_{\mathrm{Time * Time}} \cap (\mathrm{Task}\times\mathrm{Task})$.
-This can be seen as a sort of middle ground between the two interpretations.
-However, we foresee that Ampersand will change its implementation and do not follow its behavior here.
+In Ampersand's implementation, Ampersand does not require writing the types when writing Invariants.
+We can consider expressing the same invariant as earlier without writing types: $\mathrm{before}\typesemi \mathrm{scheduled} \typesubset \mathrm{scheduled}\typesemi \mathrm{before}$.
+If we require that invariants are well-typed, there is only one way of annotating such expressions with types.
+The use of the possibly ambiguous relation name $\mathrm{before}$ is allowed as long as it is clear which relation is intended, as is the case here.
+We refer to the problem of determining which relation is intended as the disambiguation problem.%, which we investigate more closely in a separate section.
+%
+%\subsection{Generalization}
+%A common task in semantic modeling is to identify that some concepts are generalizations or specializations of other concepts.
+%This section motivates the need for a special language feature to deal with generalizations.
+%
+%As an example, our planning application allows regular working hours to be set, which means that some times fall during regular hours.
+%This means that such times would be a specialization of a time.
+%We can model this with a relation called $\mathrm{worktime}$ and an invariant: $\mathrm{worktime} \subseteq \mathbb{1}_{Time}$, where $\mathbb{1}_{A} = \{(a,a) \mid a\in A \}$ is the relational identity.
+%Another way to model this, is to introduce a concept $\mathrm{Worktime}$ and use $\mathbb{1}_\mathrm{Worktime}$ in the place of the relation $\mathrm{worktime}$, along with the invariant $\mathbb{1}_\mathrm{Worktime} \subseteq \mathbb{1}_{Time}$.
+%However, this latter invariant is ill-typed: the expression on the left of the subset ($\subseteq$) is of type $(\mathrm{Worktime},\mathrm{Worktime})$, while the expression on the right is of type $(\mathrm{Time},\mathrm{Time})$.
+%To allow expressions like $\mathbb{1}_\mathrm{Worktime} \subseteq \mathbb{1}_{Time}$ in a programming language, we need a more refined notion of which expressions are `type correct'.
+%A paper by van der Woude and Joosten shows how one can do this~\cite{Woude11}.
+%This paper has largely the same notion of type correctness.
+%However, that paper does not consider the interplay with disambiguation which we consider here.
+%
+%\subsection{Two interpretations.}
+%There are two different ways in which we can think of the two binary relations called $\mathrm{before}$.
+%In one interpretation, we can say that $\mathrm{before}$ is a single binary relation, defined over the (disjoint) union between Task and Time.
+%Whenever we use $\mathrm{before}$, we restrict its type to either \verb=Task * Task= or \verb=Time * Time= to get two completely separate behaviors.
+%This means that we read $\mathrm{before}_{\mathrm{Task * Task}}$ as $\mathrm{before} \cap (\mathrm{Task}\times\mathrm{Task})$.
+%
+%Another interpretation is to say that there are two entirely separate relations, $\mathrm{before}_{\mathrm{Task * Task}}$ and $\mathrm{before}_{\mathrm{Time * Time}}$.
+%In this interpretation, we can consider ${\mathrm{Task * Task}}$ as a part of the name of the relation.
+%
+%Note our interpretation does not matter for the behavior we get from well-typed invariants like the one we saw before.
+%However, a problem arises if we add generalization.
+%
+%If we were to state - for some obscure reason - that every Task is a Time, in our first interpretation we would expect that $\mathrm{before}_{\mathrm{Task * Task}} = \mathrm{before}_{\mathrm{Time * Time}} \cap (\mathrm{Task}\times\mathrm{Task})$, since $(\mathrm{Task}\times\mathrm{Task}) \subseteq (\mathrm{Time}\times\mathrm{Time})$.
+%In our second interpretation, we would not expect the behavior to change.
+%
+%Observe that we are free to add $\mathrm{before}_{\mathrm{Task * Task}} = \mathrm{before}_{\mathrm{Time * Time}} \cap (\mathrm{Task}\times\mathrm{Task})$ as an invariant under the second interpretation, mimicking the behavior of the first interpretation under the second interpretation.
+%Therefore, we conclude that working with the second interpretation is most flexible and propose to follow this interpretation.
 
@@
 
-A disambiguation problem is given by an expression $E$ using the operations $\cap$ and $;$, and relation names from a set of names $\mathcal{N}$, and a set of possible types of each name: $\mathcal{t}(n) \subseteq \mathcal{T}\times \mathcal{T}$ for $n\in \mathcal{N}$.
+A disambiguation problem is given by an expression $E$ using the operations $\cup$ and $;$, and relation names from a set of names $\mathcal{N}$, and a set of possible types of each name: $\mathcal{t}(n) \subseteq \mathcal{T}\times \mathcal{T}$ for $n\in \mathcal{N}$.
 A solution to the disambiguation problem is a well-typed expression $S$ in which the annotations to relation names corresponds to an element in $\mathcal{t}(n)$, such that dropping the types from the expression yields $E$.
 
-As an example, $E = (\mathrm{before};\mathrm{scheduled}) \cap (\mathrm{scheduled};\mathrm{before})$ with types $\mathcal{t}(\mathrm{before}) = \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ and $\mathcal{t}(\mathrm{scheduled}) = \{(\mathrm{Task}, \mathrm{Time})\}$, is a disambiguation problem.
-A solution would be:
-\[\left(\mathrm{before}_{\mathrm{Task * Task}};\mathrm{scheduled}_{\mathrm{Task * Time}}\right) \cap \left(\mathrm{scheduled};\mathrm{before}_{\mathrm{Time * Time}}\right)\]
+As an example, $E = (\mathrm{before};\mathrm{scheduled}) \cup (\mathrm{scheduled};\mathrm{before})$ with types $\mathcal{t}(\mathrm{before}) = \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ and $\mathcal{t}(\mathrm{scheduled}) = \{(\mathrm{Task}, \mathrm{Time})\}$, is a disambiguation problem.
+It's (here unique) solution would be:
+\[\left(\mathrm{before}_{\mathrm{Task * Task}};\mathrm{scheduled}_{\mathrm{Task * Time}}\right) \cup \left(\mathrm{scheduled}_{\mathrm{Task * Time}};\mathrm{before}_{\mathrm{Time * Time}}\right)\]
 
@@
 In the first pass, we decorate the expression tree for $E$ with all the possible types for each sub-expression bottom up.
-Let $p(e)$ denote the possible types, then $p(n) = \mathcal{t}(n)$, and $p(e_1;e_2) = \{(x,y) \mid \exists z. (x,z) \in p(e_1) \wedge (z,y) \in p(e_2))\}$, and $p(e_1 \cap e_2) = p(e_1) \cap p(e_2)$.
+Let $p(e)$ denote the possible types, then $p(n) = \mathcal{t}(n)$, and $p(e_1;e_2) = \{(x,y) \mid \exists z. (x,z) \in p(e_1) \wedge (z,y) \in p(e_2))\}$, and $p(e_1 \cup e_2) = p(e_1) \cap p(e_2)$.
 If $p(E)$ is the empty set, then there is no solution to the disambiguation problem, and we are done.
@@
 For the top-level call of $d(e,t)$, let $e=E$ and $t\in p(E)$.
-Then $d(n,t) = n_{[t]}$, $d(e_1 \cap e_2, t) = d(e_1,t) \cap d(e_2,t)$ and $d(e_1 ; e_2, (x,y)) = d(e_1, (x,z)) ; d(e_2, (z,y))$ where $z$ is picked such that $(x,z)\in p(e_1)$ and $(z,y)\in p(e_2)$.
+Then $d(n,t) = n_{[t]}$, $d(e_1 \cup e_2, t) = d(e_1,t) \cup d(e_2,t)$ and $d(e_1 ; e_2, (x,y)) = d(e_1, (x,z)) ; d(e_2, (z,y))$ where $z$ is picked such that $(x,z)\in p(e_1)$ and $(z,y)\in p(e_2)$.
 Such a $z$ is guaranteed to exist by how $p$ is defined.
 
-As an example, let's look at $E = (\mathrm{before};\mathrm{scheduled}) \cap (\mathrm{scheduled};\mathrm{before})$.
+As an example, let's look at $E = (\mathrm{before};\mathrm{scheduled}) \cup (\mathrm{scheduled};\mathrm{before})$.
 Then $p(\mathrm{before}) =  \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ following $\mathcal{t}$ and similarly $p(\mathrm{scheduled}) = \{(\mathrm{Task}, \mathrm{Time})\}$.
@@
 
-Top-down, $d(E,t) = d(\mathrm{before};\mathrm{scheduled},t) \cap d(\mathrm{scheduled};\mathrm{before},t)$ with $t=(\mathrm{Task}, \mathrm{Time})$.
+Top-down, $d(E,t) = d(\mathrm{before};\mathrm{scheduled},t) \cup d(\mathrm{scheduled};\mathrm{before},t)$ with $t=(\mathrm{Task}, \mathrm{Time})$.
 Since $E$ is not ambiguous, we have only one possible way to continue as we solve the disambiguation problem:
@@
 d(\mathrm{scheduled};\mathrm{before},t) &=  d(\mathrm{scheduled},(\mathrm{Task}, \mathrm{Time}));d(\mathrm{before},(\mathrm{Time}, \mathrm{Time}))\\
-&=\mathrm{scheduled}_{\mathrm{Time} * \mathrm{Task}}; \mathrm{before}_{\mathrm{Time}* \mathrm{Time}}\\
+&=\mathrm{scheduled}_{\mathrm{Task} * \mathrm{Time}}; \mathrm{before}_{\mathrm{Time}* \mathrm{Time}}\\
 \end{align*}
@@
 The disambiguation problem and the corresponding algorithm can be extended over all operations in relation algebra.
+It can even be used to disambiguate special relations like the identity relation.
+Let's take $E_2 = \mathrm{scheduled};(\mathrm{before} \cup \mathbb{1})$ as an expression.
+The types are as in the previous example, with $\mathcal{t}(\mathbb{1}) = \left\{(t,t)\mid t\in \mathcal{T}\right\}$.
+This gives $p(E_2) = \{\mathrm{Task} * \mathrm{Time}\}$ and $d(\{\mathrm{Task} * \mathrm{Time}\},E_2) = \mathrm{scheduled}_{\mathrm{Task} * \mathrm{Time}};(\mathrm{before}_{\mathrm{Time} * \mathrm{Time}}\ \cup\ \mathbb{1}_{\mathrm{Time}})$, as desired.
 
 \section{Interplay between Generalization and Disambiguation}
+Before giving a solution to this problem, we first observe that the disambiguation algorithm discussed earlier fails for our running example.
+Following the declarations of the script, we let $\mathcal{t}(\mathrm{before}) = \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ and $\mathcal{t}(\mathrm{scheduled}) = \{(\mathrm{Event}, \mathrm{Time})\}$.
+Now the above algorithm would conclude with an error on $E = (\mathrm{before};\mathrm{scheduled}) \cup (\mathrm{scheduled};\mathrm{before})$.
+
+The paper by van der Woude and Joosten~\cite{Woude11} introduced an embedding relation $\epsilon_{A*B} : A \times B$ to write sentences in heterogeneous relation algebra with sub-typing in terms of heterogeneous relation algebra without it.
+It can be thought of as defined by $\epsilon_{A*B} = \{(a,a) \mid a\in A \wedge a\in B\}$: it is the largest subset of the identity relation that fits the type $A\times B$.
+We deviate slightly from the ideas by van der Woude and Joosten: Their $\epsilon$ relation is a subtyping relation, so $\epsilon_{A*B}$ would exist only if $A$ is a subtype of $B$.
+In this paper, we relax that requirement.
+
+Let $\mathcal{S}$ be the language of heterogeneous relation algebra with subtyping.
+It is defined by the following grammar, where $n\in\mathcal{N}$ and $A,B\in\mathcal{T}$:
+\[
+\varphi,\psi\in\mathcal{S} \mathbin{:=} n \mid \typetyped{\varphi}{A*B} \mid \typevee \mid \typeident \mid \typenil \mid \varphi \typeunion \psi \mid \varphi \typeinter \psi \mid \varphi \typesemi \psi \mid \typecomp{\psi}\mid \typeconv{\psi}
+\]
+The grammar $\mathcal{S}$ allows the programmer to write relations with their type, $\typetyped{n}{A*B}$ and $\typetyped{\typevee}{A*B}$, as usual.
+However, the grammar is slightly more general, allowing type annotations for arbitrary terms $\typetyped{\varphi}{A*B}$.
+
+Our aim is to disambiguate expressions written in this language into `normal' heterogeneous relation algebra, that is: without subtyping.
+We use the following grammar to describe the language $\mathcal{R}$ for that, where $n_{A*B}$ satisfies $(A,B)\in \mathcal{t}(n)$, and $A,B\in\mathcal{T}$ for all other occurrences of $A$ and $B$:
+\[
+\varphi,\psi\in\mathcal{R} \mathbin{:=} n_{A*B} \mid \epsilon_{A*B} \mid T_{A*B} \mid 1_{A} \mid F_{A*B} \mid \varphi \cup \psi \mid \varphi \cap \psi \mid \varphi ; \psi \mid \overline{\psi}\mid \conv{\psi}
+\]
+
+The type of a term in $\mathcal{R}$ is given by a partial typing function, which we define inductively via the relation symbol $:$ as follows:
+\begin{align*}
+x_{A*B} &: A * B & \text{For $x \in \mathcal{N} \cup \{\epsilon, T, F\}$}\\
+1_{A} &: A * A&\\
+\varphi \odot \psi &: A * B &\quad\text{For $\odot \in \{\cup,\cap\}$, $\varphi : A * B$ and $\psi : A * B$}\\
+\varphi ; \psi &: A * B &\text{For $\varphi : A * C$ and $\psi : C * B$}\\
+\overline{\psi} &: A * B &\text{For $\psi : A * B$}\\
+\conv{\psi} &: A * B &\text{For $\psi : B * A$}
+\end{align*}
+We say that a term $r\in \mathcal{R}$ is \emph{well-typed} if $r : A*B$ for some $A$ and $B$.
+Univalence of $:$ follows by induction on the structure of $r$.
+
+Rather than translating terms in $\mathcal{S}$ into $\mathcal{R}$ directly, we create an intermediate language $\mathcal{D}$, which contains $\mathcal{R}$ (that is: $\mathcal{R} \subseteq \mathcal{D}$).
+This will allow us to use a purely syntactic translation from $\mathcal{S}$ to $\mathcal{D}$, after which we run a disambiguation algorithm to turn it into a term in $\mathcal{R}$.
+The language $\mathcal{D}$ is defined as follows, where $\xi \in \mathcal{R}$, $\Xi \subseteq \mathcal{R}$, and $A,B\in\mathcal{T}$:
+\[
+\varphi,\psi\in\mathcal{D} \mathbin{:=} \xi \mid \varphi \cup \psi \mid \varphi \cap \psi \mid \varphi ; \psi \mid \overline{\psi}\mid \conv{\psi} \mid \typetyped{\varphi}{A*B} \mid \Xi
+\]
+
+The term $\Xi$ is used to indicate that elements in $\Xi$ are terms that might be filled in on this position.
+For example, the ambiguous term $\mathrm{before}$ in $\mathcal{S}$ would be represented in $\mathcal{D}$ as $\{\mathrm{before}_{Task*Task},\mathrm{before}_{Time*Time}\}$.
+We syntactically distinguish terms of the shape $\Xi$ by writing set brackets $\{\}$ around them.
+Type annotations from $\mathcal{S}$ are kept in $\mathcal{D}$, but the structure is otherwise the same as that of $\mathcal{R}$.
+
+The translation $s : \mathcal{S} \to \mathcal{D}$ is as follows:
+\begin{align*}
+s(n) &= \{n_{A*B} \mid (A,B)\in\mathcal{t}(n)\}\\
+s(\typetyped{\varphi}{A*B}) &= \typetyped{s(\varphi)}{A*B}\\
+s(\typevee) &= \{T_{A*B} \mid A,B\in\mathcal{T}\}\\
+s(\typenil) &= \{F_{A*B} \mid A,B\in\mathcal{T}\}\\
+s(\typeident) &= \{1_{A} \mid A\in\mathcal{T}\}\\
+s(\varphi \typeunion \psi) &= ..?\\
+s(\varphi \typeinter \psi) &= ..?\\
+s(\varphi \typesemi \psi) &= ..?\\
+s(\typecomp{\varphi}) &= ..?\\
+s(\typeconv{\varphi}) &= ..?
+\end{align*}
 

### 2023-02-25 2022 Migration/articleMigration.tex

@@
    e.g.~\cite{Gholami2016,Bisbal1999},
-   migration engineers may exclude new functionality to avoid introducing new errors in an otherwise error-prone migration process.
+   migration engineers may deliberately exclude new functionality to avoid introducing new errors in an otherwise error-prone migration process.
    In our work, however, introducing new functionality and removing obsolete functionality is the very purpose of an increment.
@@
    This work, however, gets rid of the existing system without retaining the variations.
+   So this is clearly a different notion of increment.
    Examples like these have convinced the authors to define a new method for data migration that is specifically meant for systems in production that are
@@
    The relation $\inst:\Pair{\Atoms}{\Concepts}$ is the instance relation between atoms and concepts.
-   The expression $a\ \inst\ C$ means that atom $a$ is an \emph{instance} of concept $C$.
-   This relation is used in the type system, in which $\inst$ assigns a concept to every atom in the dataset.
+   The expression $a\inst C$ means that atom $a$ is an \emph{instance} of concept $C$.
+   This relation is used in the type system, in which $\inst$ assigns one or more concepts to every atom in the dataset.
 
@@
    \label{eqn:specialization}
-   A\isa B\ \Leftrightarrow\ \forall a: a\ \inst\ A\rightarrow a\ \inst\ B
+   A\isa B\ \Leftrightarrow\ \forall a: a\inst A\rightarrow a\inst B
 \end{equation}
-   Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal'',
-   in which a user defines a pair of concepts in the relation $\isa$.
-   A compiler can construct $\isa$ as the transitive, reflexive closure of such user-defined pairs of concepts.
-   It must test for antisymmetry, ensuring that $\isa$ is a partial order of concepts.
+   Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.
+   In her script, a user can declare a specialization as a pair of concepts in the relation $\isa$.
+   A compiler can construct $\isa$ as the transitive, reflexive closure of all user-defined specializations.
+   It must make sure that $\isa$ is antisymmetric, so that $\isa$ is a partial order of concepts.
    Specialization causes atoms to have multiple concepts of which they can be an instance.
-   To ensure that this does not result in problems at run-time, a compiler can require users to declare all specializations explicitly.
    As a result, if an atom is an instance of concept $A$ and $A\isa B$,
@@
 
-   A dataset $\dataset$ is a structure $\triple{\triples}{\isa}{\inst}$ that satisfies:
+   A dataset $\dataset$ is a tuple $\pair{\triples}{\inst}$ that satisfies:
 \begin{eqnarray}
-   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&
-      \begin{array}[t]{@{}l}
-         \exists C: C\isa A\ \wedge\ a\ \inst\ C\ \wedge\\
-         \exists D: D\isa B\ \wedge\ b\ \inst\ D
-      \end{array}
+   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&a\inst A\ \wedge\ b\inst B
    \label{eqn:wellTypedEdge}
@@
    The notations $\triples_{\dataset}$ and $\inst_{\dataset}$ are used to disambiguate $\triples$ and $\inst$ when necessary.
-   Every dataset is an element of an infinite set called $\Dataset$.
+   Every dataset is an element of an infinite set called $\Dataset$,
+   which is disjoint from $\Rels$, $\Concepts$ and $\Atoms$.
    To save writing in the sequel, the notation $a\ r\ b$ means that $\triple{a}{r}{b}\in\triples$.
@@
    Equation~\ref{eqn:wellTypedEdge} implies that for every dataset $\dataset$:
-\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ a\ \inst\ A\ \wedge\ b\ \inst\ B\]
+\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ a\inst A\ \wedge\ b\inst B\]
    For a developer, this means that the type of an atom depends only on the relation in which it resides; not on the actual population of the database.
    This allows for static typing, which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.
+
+\subsection{Changing data}
+\label{sct:Changing data}
+   Events can cause data to change.
+   For triples, we distinguish just two types of event: insert and delete events.
+   An event is a label (ins or del) with a dataset that is to be inserted in or deleted from another dataset.
+   So we shall write
+\begin{eqnarray}
+   \ins{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}\pair{\triples}{\inst}&=&\pair{\triples\cup\triples_{\id{ev}}}{\inst\cup\inst_{\id{ev}}}\\
+   \del{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}\pair{\triples}{\inst}&=&\pair{\triples-\triples_{\id{ev}}}{\inst-\inst_{\id{ev}}}
+\end{eqnarray}
+   Insert events are represented as $\ins{\dataset}$ and delete events by $\del{\dataset}$.
+   Both are functions $\Dataset\rightarrow\Dataset$, which are defined as:
+   Since datasets are sets, we will use set union ($\cup$) and set difference ($-$) to make changes in datasets.
 
@@
 % \begin{eqnarray}
-%    \sign{u}=\pair{A}{B}\ \wedge\ \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\ \inst\ A\wedge b\ \inst\ B
+%    \sign{u}=\pair{A}{B}\ \wedge\ \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\inst A\wedge b\inst B
 %    \label{eqn:wellTypedEdged violations}
@@
 
-   A schema $\schema$ is a structure $\quadruple{\concepts}{\isa}{\rels}{\rules}$,
+   A schema $\schema$ is a tuple $\quadruple{\concepts}{\isa}{\rels}{\rules}$,
    in which $\concepts$ is a finite set of concepts,
@@
 \begin{eqnarray}
+   \declare{n}{A}{B}\in\rels&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
+   \label{eqn:relationsIntroduceConcepts}\\
    A\ \isa\ B&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
    \label{eqn:isasIntroduceConcepts}\\
-   \declare{n}{A}{B}\in\rels&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
-   \label{eqn:relationsIntroduceConcepts}\\
    \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
@@
 \label{def:information system}
-\item An information system $\infsys$ is a structure $\la\dataset,\schema,\roles,\maintain\ra$, in which
+\item An information system $\infsys$ is a tuple $\quadruple{\dataset}{\schema}{\roles}{\maintain}$, in which
 \begin{itemize}
-   \item dataset $\dataset=\triple{\triples}{\isa}{\inst}$ is defined as in section~\ref{sct:Datasets};
+   \item dataset $\dataset=\pair{\triples}{\inst}$ is defined as in section~\ref{sct:Datasets};
    \item schema $\schema=\quadruple{\concepts}{\isa}{\rels}{\rules}$ is defined as in section~\ref{sct:Schemas};
    \item $\roles$ is a set of roles;
-   \item $\maintain : \roles\times\rules$ is the maintainance relation between roles and rules.
+   \item ${\maintain} : \roles\times\rules$ is the maintainance relation between roles and rules.
 \end{itemize}
 \end{definition}
-   Note that dataset $\dataset$ and schema $\schema$ have the same specialization relation $\isa$.
    A \define{role} is a name that identifies a group of users.
@@
 \begin{equation}
-   \forall u\in\rules\ \exists o\in\roles:\ o\ \maintain\ u\label{eqn:maintain}
+   \forall u\in\rules\ \exists o\in\roles:\ o\maintain u\label{eqn:maintain}
 \end{equation}
-   $o\ \maintain\ u$ means that role $o$ must keep rule $u$ satisfied, i.e.\ to keep $\sat{u}{\dataset}$ true.
+   $o\maintain u$ means that role $o$ must keep rule $u$ satisfied, i.e.\ to keep $\sat{u}{\dataset}$ true.
    As long as rule $u$ is broken, the system should notify actors in the role(s) that are administered in $\maintain$ to do something about it.
@@
    An enforce rule specifies not only the rule, but also the way to restore invariance.
-   The system features an engine that restores invariance of all enforce rules without unneccessary delay.
+   The system features an engine that restores invariance of all enforce rules without delay.
    Whenever the dataset violates an enforce rule,
@@
    Now let us check the requirements to verify that this example defines an information system.
+   Requirement~\ref{eqn:specialization} is satisfied because this example contains no specialization.
    The Ampersand compiler generates a dataset $\dataset$, which contains a set of triples and a relation $\inst$.
@@
 
-   The Ampersand compiler generates a schema $\schema$, which contains concepts, relations, and rules.
+   The Ampersand compiler generates a schema $\schema$, which contains concepts, a specialization relation, relations, and rules.
    It defines the set of concepts to satisfy requirement~\ref{eqn:relationsIntroduceConcepts}:
    \[\concepts=\{ {\tt Module}, {\tt Course}, {\tt Student}\}\]
-   It defines the set of relations to satisfy requirement~\ref{eqn:define R}:
+   The user has defined no specializations.
+   This means that $\isa$ equals the identity relation on $\concepts$ and requirement~\ref{eqn:isasIntroduceConcepts} is satisfied.
+   The set of relations, $\rels$, satisfies requirement~\ref{eqn:define R}:
    \[\begin{array}{rcl}
@@
    And, it defines the set of rules $\rules$ to contain just one rule: \verb-EnrollRule-.
-   Requirement~\ref{eqn:specialization} is satisfied because this example contains no specialization.
    So, the schema $\schema=\quadruple{\concepts}{\isa}{\rels}{\rules}$ satisfies the requirements from section~\ref{sct:Schemas}.
@@
    Requirement~\ref{eqn:satisfaction} is satisfied because the only rule, {\tt EnrollRule} is satisfied.
-   This completes the requirements for $\infsys=\la\dataset,\schema,\roles,\maintain\ra$ from definition~\ref{def:information system}.
+   This completes the requirements for $\infsys=\quadruple{\dataset}{\schema}{\roles}{\maintain}$ from definition~\ref{def:information system}.
    This concludes the argument that $\infsys$ is an information system.
@@
          These rules can be necessary, for example to resolve data pollution that violates a constraint that was not in schema $\schema_{\infsys}$.
-         By putting it in the schema $\schema_{\migrsys'}$,
-         the migration engineer can fix such problems.
+         By putting such rules in the schema $\schema_{\migrsys'}$,
+         the migration engineer can fix these problems.
          She can do this in the form of an enforce rule, which means that the problem gets fixed automatically.
@@
 %    \omit\rlap{$\la\atoms,\concepts,\inst,\isa,\rels,\dataset\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\dataset'\ra$}\notag\\
-%    &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\ \inst\uplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \dataset\uplus^5\dataset'\ra\notag\\
+%    &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\instuplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \dataset\uplus^5\dataset'\ra\notag\\
 %       X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
@@
 \begin{definition}[relabel concepts in $\inst$]
-   \[a\ \subst{C}{D}{\inst}\ \subst{C}{D}{A}\ \Leftrightarrow\ a\ \inst\ A\]
+   \[a\ \subst{C}{D}{\inst}\ \subst{C}{D}{A}\ \Leftrightarrow\ a\inst A\]
 \end{definition}
@@
    \begin{eqnarray}
-      \infsys\sqcup\infsys'&=&\la\roles',\rules',\maintain',\dataset\sqcup\dataset'\ra
+      \infsys\sqcup\infsys'&=&\quadruple{\roles'}{\rules'}{\maintain'}{\dataset\sqcup\dataset'}
    \end{eqnarray}

### 2023-02-25 2023 Disambiguation/Disambiguation.tex

@@
 %Therefore, we conclude that working with the second interpretation is most flexible and propose to follow this interpretation.
+
+%
+%
+%Bas, ik lees dat als:
+%\begin{eqnarray}
+%    \pair{a}{b}\in r&\Rightarrow&a\in\mathrm{Time}\cup\mathrm{Task}\wedge b\in\mathrm{Time}\cup\mathrm{Task}\\
+%    \mathrm{before}_{\mathrm{Time * Time}}&=&r\cap (\mathrm{Time}\times\mathrm{Time})\\
+%    \mathrm{before}_{\mathrm{Task * Task}}&=&r\cap (\mathrm{Task}\times\mathrm{Task})
+%\end{eqnarray}
+%Another interpretation is to say that there are two entirely separate relations,
+%$r:\mathrm{before}_{\mathrm{Time * Time}}$ and $r':\mathrm{before}_{\mathrm{Task * Task}}$.
+%In this interpretation, we can consider ${\mathrm{Task * Task}}$ as a part of the name of the relation.
+%
+%Bas, ik lees dat als:
+%\begin{eqnarray}
+%    \pair{a}{b}\in r&\Rightarrow&a\in\mathrm{Time}\wedge b\in\mathrm{Time}\\
+%    \pair{a}{b}\in r'&\Rightarrow&a\in\mathrm{Task}\wedge b\in\mathrm{Task}\\
+%    \mathrm{before}_{\mathrm{Time * Time}}&=&r\\
+%    \mathrm{before}_{\mathrm{Task * Task}}&=&r'
+%\end{eqnarray}
 

### 2023-02-25 2023 Disambiguation/Disambiguation.tex
Initial stab at the syntactic translation bit

@@
 Our aim is to disambiguate expressions written in this language into `normal' heterogeneous relation algebra, that is: without subtyping.
-We use the following grammar to describe the language $\mathcal{R}$ for that, where $n_{A*B}$ satisfies $(A,B)\in \mathcal{t}(n)$, and $A,B\in\mathcal{T}$ for all other occurrences of $A$ and $B$:
+To signal certain type errors, we add an additional element $\bot$ to the set of concepts, that we will assume is different from every element in $\mathcal{T}$.
+Since $\mathcal{T}$ is referred to as the set of concepts, we will refer to $\mathcal{T}\cup\{\bot\}$ as the set of pseudo-concepts.
+We use the following grammar to describe the language $\mathcal{R}$ for that, where $n_{A*B}$ satisfies $(A,B)\in \mathcal{t}(n)$, and $A,B\in\mathcal{T}\cup\{\bot\}$ for all other occurrences of $A$ and $B$:
 \[
@@
 \begin{align*}
-x_{A*B} &: A * B & \text{For $x \in \mathcal{N} \cup \{\epsilon, T, F\}$}\\
+x_{A*B} &: A * B & \text{For $x \in \mathcal{N} \cup \{\epsilon, T, F\}$ and $A,B\in \mathcal{T}$}\\
 1_{A} &: A * A&\\
@@
 We say that a term $r\in \mathcal{R}$ is \emph{well-typed} if $r : A*B$ for some $A$ and $B$.
-Univalence of $:$ follows by induction on the structure of $r$.
+By induction on the inductive definition of $\mathcal{R}$, the relation $:$ is between $\mathcal{R}$ and types $\mathcal{T}\times\mathcal{T}$ and univalent.
+In other words: well-typed terms in $\mathcal{R}$ have precisely one type, all of its sub-terms are well-typed, and $\bot$ does not occur as a pseudo-concept.
 
@@
 
+The translation from $\mathcal{S}$ to $\mathcal{D}$ will depend on the users's \verb=CLASSIFY= statements.
+We write $A\preceq B$ to indicate that the pair $A,B$ is in the transitive closure of the pairs written \verb=CLASSIFY _ ISA _=, meaning that all elements in $A$ must occur in $B$ whenever $A\preceq B$.
+Moreover, we let $\bot$ be a least element in this preorder: $\bot \preceq A$ for all $A$.
+Since there are no classify statements containing $\bot$ (the user cannot write them), the least element in $\preceq$ is unique.
+
 The translation $s : \mathcal{S} \to \mathcal{D}$ is as follows:
@@
 s(\typeident) &= \{1_{A} \mid A\in\mathcal{T}\}\\
-s(\varphi \typeunion \psi) &= ..?\\
-s(\varphi \typeinter \psi) &= ..?\\
-s(\varphi \typesemi \psi) &= ..?\\
-s(\typecomp{\varphi}) &= ..?\\
-s(\typeconv{\varphi}) &= ..?
+s(\varphi \typeunion \psi) &= \conv{\varepsilon};s(\varphi);\varepsilon\ \cup\ \conv{\varepsilon};s(\psi);\varepsilon&\quad\text{with $\varepsilon = \{\epsilon_{A*B} \mid A\preceq B \}$} \\
+s(\varphi \typeinter \psi) &= \conv{\varepsilon};s(\varphi);\varepsilon\ \cap\ \conv{\varepsilon};s(\psi);\varepsilon&\quad\text{with $\varepsilon = \{\epsilon_{A*B} \mid B\preceq A \}$}\\
+s(\varphi \typesemi \psi) &= s(\varphi);\varepsilon;s(\psi)&\quad\text{with $\varepsilon = \{\epsilon_{A*B} \mid B\preceq A \vee A\preceq B \}$}\\
+s(\typecomp{\varphi}) &= \overline{s(\varphi)}\\
+s(\typeconv{\varphi}) &= \conv{s(\varphi)}
 \end{align*}

### 2023-02-25 2023 Disambiguation/Disambiguation.tex
Bibliography and subsections

@@
 
-\section{Interplay between Generalization and Disambiguation}
+\section{Solving Disambiguation with Generalization}
 Before giving a solution to this problem, we first observe that the disambiguation algorithm discussed earlier fails for our running example.
@@
 It can be thought of as defined by $\epsilon_{A*B} = \{(a,a) \mid a\in A \wedge a\in B\}$: it is the largest subset of the identity relation that fits the type $A\times B$.
-We deviate slightly from the ideas by van der Woude and Joosten: Their $\epsilon$ relation is a subtyping relation, so $\epsilon_{A*B}$ would exist only if $A$ is a subtype of $B$.
-In this paper, we relax that requirement.
+
+This paper follows a similar approach: We define a language to be used by the Ampersand programmer that allows writing relations without their types as well as subtyping.
+Next, we define a language that is interpreted as a heterogeneous relation algebra with a typical set-interpretation.
+The semantics of the language with subtyping will be determined by what the translation process gives us, but we validate the translation by inspecting certain examples and examining the validity of Tarski's axioms for this language under the resulting interpretation.
+%We deviate slightly from the ideas by van der Woude and Joosten: Their $\epsilon$ relation is a subtyping relation, so $\epsilon_{A*B}$ would exist only if $A$ is a subtype of $B$.
+%In this paper, we relax that requirement.
+
+\subsection{Language definitions}
+In this section, we introduce the language in which an Ampersand programmer can write terms, as well as the language for which the disambiguation problem is considered `solved'.
+The intended interpretation of these symbols is that of the regular relation algebraic operations, but we delay talking about semantics until later in this paper.
 
@@
 
+\subsection{Translation process}
 Rather than translating terms in $\mathcal{S}$ into $\mathcal{R}$ directly, we create an intermediate language $\mathcal{D}$, which contains $\mathcal{R}$ (that is: $\mathcal{R} \subseteq \mathcal{D}$).
@@
 s(\varphi \typeinter \psi) &= \conv{\varepsilon};s(\varphi);\varepsilon\ \cap\ \conv{\varepsilon};s(\psi);\varepsilon&\quad\text{with $\varepsilon = \{\epsilon_{A*B} \mid B\preceq A \}$}\\
-s(\varphi \typesemi \psi) &= s(\varphi);\varepsilon;s(\psi)&\quad\text{with $\varepsilon = \{\epsilon_{A*B} \mid B\preceq A \vee A\preceq B \}$}\\
+s(\varphi \typesemi \psi) &= s(\varphi);\varepsilon;s(\psi)&\text{with $\varepsilon = \{\epsilon_{A*B} \mid B\preceq A \vee A\preceq B \}$}\\
 s(\typecomp{\varphi}) &= \overline{s(\varphi)}\\
@@
 
+\paragraph{[Aside],} The Ampersand language also allows for \verb=CLASSIFY A IS B /\ C= statements, meaning that elements are in $A$ if and only if they are in $B$ and in $C$.
+The information about this captured in $\preceq$ is just that $A \preceq B$ and $A\preceq C$, which means that information is lost.
+We can consider the lost information when dealing with the $\typesemi$ operation by changing the $\epsilon$ relations introduced there.
+However, it is an open problem to deal with this with the $\typeinter$ operation, so we give a simpler presentation that ignores this additional information in \verb=CLASSIFY _ IS _= statements for now.
+
 \section{Bibliography}
 \bibliographystyle{elsarticle-harv}
-\bibliography{doc}
+\bibliography{Disambiguation}
 

### 2023-03-08 2022 Migration/articleMigration.tex
WIP

@@
    the users experience a system that evolves organically, without even noticing the updates.
-   Thus, this paper will help to automate data migrations correctly and reliably.
-   The benefits work in two directions.
-   Automating data migrations saves effort and errors,
-   contributing to more frequent and smaller releases.
+   By automating data migrations correctly and reliably, migration engineers save effort and errors.
+   As a result, they can release smaller increments more frequently, which contributes to a more agile software development process.
    Smaller and faster releases also mean a smaller difference between the existing system and the desired system.
-   The smaller that difference, the simpler the migration can be and the more of it can be automated.
-
-   Each increment has an ``existing system'' and a ``desired system''.
-   A difference of schemas complicates the migration of data between the existing system and the desired system.
+   This yields even simpler migrations, leveraging even more frequent releases.
+
+   Differences of schemas of an existing system and a desired system complicate the migration of data.
    The purpose of incremental deployment poses specific requirements to data migration,
@@
    e.g.~\cite{Gholami2016,Bisbal1999},
-   migration engineers may deliberately exclude new functionality to avoid introducing new errors in an otherwise error-prone migration process.
-   In our work, however, introducing new functionality and removing obsolete functionality is the very purpose of an increment.
+   migration engineers may deliberately avoid schema differences and functionality changes to avoid introducing new errors in an otherwise error-prone migration process.
+   For the purpose of software updates, however, introducing new functionality, removing obsolete functionality, and schema differences are the very reason for deploying an increment.
    Data migration literature sometimes uses a different notion of increment.
@@
    They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
-   This work, however, gets rid of the existing system without retaining the variations.
-   So this is clearly a different notion of increment.
+   In a software update, however, the variations are not preserved.
+   So this notion of increment in\cite{Ataei2021} is obviously different.
    Examples like these have convinced the authors to define a new method for data migration that is specifically meant for systems in production that are
@@
    zero down-time updates and ways to cope with data pollution.
+   Zero down-time means that all human activity for the data migration can be done without taking the system out of production.
    In practice, more frequent updates require that updates come with zero down-time.
-   For our theory, zero down-time means that all human activity for the data migration can be done without taking the system out of production.
    Any delay caused by automatically switching from the existing system to the desired is considered negligible.
-   This assumes a modern deployment platform (such as Kubernetes) that updates applicatioins with zero down-time.
+   This assumes a modern deployment platform (such as Kubernetes) that updates applications with zero down-time.
 
@@
 \section{Analysis}
+\label{sct:Analysis}
    The purpose of an information system is to make data meaningful to its users.
+   The analysis of the problem starts with a formal definition of information systems,
+   in which the schema contains all information that makes data meaningful.
+   We then define a migration procedure from one information system to another,
+   which preserves the meaning of the data as much as possible.
+
    Users have their own tasks and responsibilities
@@
    This collective use by multiple users serves a purpose which we will loosely call ``the business''.
-   To preserve meaning, the business maintains semantic constraints on the data,
-   amidst of all changes that are going on around them.
+   It is this business that depends on the semantics of the data to draw the right conclusions and carry out their tasks.
+   This paper uses semantic constraints to represent this meaning.
 

### 2023-03-08 2023 Disambiguation/Disambiguation.tex
WIP

@@
 \section{Introduction}
+% @Bas, jouw introduction poogt via een omweg bij het op te lossen probleem te komen.
+% Een meer directe inleiding zou zijn:
+
+% In a heterogeneous relation algebra~\cite{Hattensperger}, all relations have a type.
+% To refer to a relation, one must write $r_{[A\times B]}$ instead of just $r$.
+% When the heterogeneous relation algebra is used as a programming language, e.g.~Ampersand~\cite{Ampersand}, identifiers are involved.
+% This yields verbose expressions such as:
+% \begin{verbatim}
+%     account[Person*Bankaccount];creditline[Bankaccount*Amount]
+% \end{verbatim}
+% A programmer, however, might just want to write \verb-account;creditline-.
+
+% This paper proposes a way to derive the types in expressions of heterogeneous relation algebra,
+% to enable more compact references to relations.
+% The idea is that type information is only necessary if an expression would otherwise be ambiguous.
+% If the programmer already uses the notation $r_{[A\times B]}$, there is no ambiguity.
+% But if the programmer refers to a relation just by $r$, and there are multiple relations with that name, ambiguity lures.
+% This paper tries to resolve as many ambiguities as possible.
+
 Notations matter: They help communicate ideas by establishing expectations.
@@
 %
-%Bas, ik lees dat als:
+% @Bas, ik lees dat als:
 %\begin{eqnarray}
@@
 
+% @Bas, deze paragraaf is wellicht leesbaarder als je gewoon de functie à la Haskell uitschrijft.
 For the top-down step, let $d(e,t)$ denote the disambiguated expression $e$ with the type $t$.
@@
 \section{Solving Disambiguation with Generalization}
-Before giving a solution to this problem, we first observe that the disambiguation algorithm discussed earlier fails for our running example.
+Before giving a solution to this % @Bas, een referentie over de grens van een paragraaf is taalkundig al niet fraai, maar over de grens van een section is niet slim.
+Als lezer wil ik iets meer duidelijkheid dan "this" als het gaat over welk probleem je naar verwijst.
+problem, we first observe that the disambiguation algorithm discussed earlier fails for our running example.
 Following the declarations of the script, we let $\mathcal{t}(\mathrm{before}) = \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ and $\mathcal{t}(\mathrm{scheduled}) = \{(\mathrm{Event}, \mathrm{Time})\}$.
@@
 Next, we define a language that is interpreted as a heterogeneous relation algebra with a typical set-interpretation.
+% @Bas, de volgende zin snap ik niet...
 The semantics of the language with subtyping will be determined by what the translation process gives us, but we validate the translation by inspecting certain examples and examining the validity of Tarski's axioms for this language under the resulting interpretation.

### 2023-03-12 2022 Migration/articleMigration.tex
define constraints for the type checker

@@
 
-\subsection{Changing data}
-\label{sct:Changing data}
-   Events can cause data to change.
-   For triples, we distinguish just two types of event: insert and delete events.
-   An event is a label (ins or del) with a dataset that is to be inserted in or deleted from another dataset.
-   So we shall write
-\begin{eqnarray}
-   \ins{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}\pair{\triples}{\inst}&=&\pair{\triples\cup\triples_{\id{ev}}}{\inst\cup\inst_{\id{ev}}}\\
-   \del{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}\pair{\triples}{\inst}&=&\pair{\triples-\triples_{\id{ev}}}{\inst-\inst_{\id{ev}}}
-\end{eqnarray}
-   Insert events are represented as $\ins{\dataset}$ and delete events by $\del{\dataset}$.
-   Both are functions $\Dataset\rightarrow\Dataset$, which are defined as:
-   Since datasets are sets, we will use set union ($\cup$) and set difference ($-$) to make changes in datasets.
-
 \subsection{Schemas}
@@
    Together, these items capture the semantics of an information system~\cite{Spivak2012}.
+   A schema $\schema$ is a tuple $\quadruple{\concepts}{\isa}{\rels}{\rules}$,
+   in which $\concepts$ is a finite set of concepts,
+   $\isa$ is a partial order of concepts,
+   $\rels$ is a finite set of relations,
+   and $\rules$ is a finite set of rules.
    Each rule in a schema serves to constrain the dataset, to ensure its semantic integrity.
@@
    In this paper, variables $u$ and $v$ represent rules.
-
-   A type system in Ampersand~\cite{vdWoude2011} identifies certain rules that are provably false.
-   It also ensures that all rules can be interpreted within the set of relations in the schema.
-   Due to this type system, the Ampersand compiler only generates code for scripts without type errors.
-   As a result, for every rule $u$ in the schema, there is a predicate $\sat{u}{\dataset}$ that says whether dataset $\dataset$ satisfies rule $u$.
+   The notation $\schema_{\infsys}$ refers to the schema of information system $\infsys$.
+   Disambiguation sometimes requires to write $\concepts_{\schema}$, $\isa_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$
+   rather than $\concepts$, $\isa$, $\rels$, and $\rules$ respectively.
+   For every rule $u$ in the schema, there is a predicate $\sat{u}{\dataset}$ that says whether or not dataset $\dataset$ satisfies rule $u$.
+%  Every schema is an element of an infinite set called $\Schema$.
+
+   A schema must ensure that triples ``make sense'' in the semantics defined by the schema.
+   This means that it must satisfy the following requirements.
+\begin{eqnarray}
+   \declare{n}{A}{B}\in\rels&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
+   \label{eqn:relationsIntroduceConcepts}\\
+   A\ \isa\ B&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
+   \label{eqn:isasIntroduceConcepts}\\
+   \isa&&\text{is a partial order on $\concepts$}
+   \label{eqn:isasPartialOrder}\\
+   \forall u\in\rules,\exists\dataset\in\Dataset-\emptyset&:&\sat{u}{\dataset}
+   \label{eqn:trueifyableRules}
+\end{eqnarray}
+   Note that these requirements do not depend on any particular dataset,
+   so they can be checked statically.
+   The type system of Ampersand~\cite{vdWoude2011} ensures that every information system it generates satisfies these requirements.
 
@@
 
-   Rules serve as invariants,
-   i.e.\ constraints that the data in the database must satisfy at all times.
-   At times in which the dataset does not satisfy a rule, that rule is said to be {\em broken}.
-   In those cases, something or someone must take action to fix the situation and restore invariance of that rule.
-
-   A schema $\schema$ is a tuple $\quadruple{\concepts}{\isa}{\rels}{\rules}$,
-   in which $\concepts$ is a finite set of concepts,
-   $\isa$ is a partial order of concepts,
-   $\rels$ is a finite set of relations,
-   and $\rules$ is a finite set of rules.
-   The notation $\schema_{\infsys}$ refers to the schema of information system $\infsys$.
-   Disambiguation sometimes requires to write $\concepts_{\schema}$, $\isa_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$
-   rather than $\concepts$, $\isa$, $\rels$, and $\rules$ respectively.
-%  Every schema is an element of an infinite set called $\Schema$.
-
-   In order to do static type checking,
-   all concepts and relations must be ``known'' in the schema:
-\begin{eqnarray}
-   \declare{n}{A}{B}\in\rels&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
-   \label{eqn:relationsIntroduceConcepts}\\
-   A\ \isa\ B&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
-   \label{eqn:isasIntroduceConcepts}\\
-   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
-   \label{eqn:define R}
-\end{eqnarray}
-   If a triple $\triple{a}{r}{b}$ satisfies requirement~\ref{eqn:define R} and its relation $r$ satisfies requirement~\ref{eqn:relationsIntroduceConcepts},
-   this triple is said to {\em befit} schema $\quadruple{\concepts}{\isa}{\rels}{\rules}$.
-
 \subsection{Information Systems}
@@
 
+   The information system enforces the semantics by ensuring that the following requirements remain true at all times:
+\begin{eqnarray}
+   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
+   \label{eqn:define R}\\
+   \forall u\in\rules&:&\sat{u}{\dataset}
+   \label{eqn:satisfaction}
+\end{eqnarray}
+   Requirement~\ref{eqn:define R} ensures that the information system only works with triples whose relation is defined in the schema.
+   Requirement~\ref{eqn:satisfaction} ensures that the dataset satisfies all rules in the schema.
+
+   The maintainance relation between roles and rules is used to determine who is allowed to change the dataset.
+   For every rule $u$ in the schema, there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
+   that represents the set of violations%
+
+   The rules in the schema serve as invariants,
+   i.e.\ constraints that the data in the database must satisfy at all times.
+   At times in which the dataset does not satisfy a rule, that rule is said to be {\em broken}.
+   In those cases, something or someone must take action to fix the situation and restore invariance of that rule.
+
+   If a triple $\triple{a}{r}{b}$ satisfies requirement~\ref{eqn:define R} and its relation $r$ satisfies requirement~\ref{eqn:relationsIntroduceConcepts},
+   this triple is said to {\em befit} schema $\quadruple{\concepts}{\isa}{\rels}{\rules}$.
+
    The system is {\em at rest} when every rule in the schema is satisfied:
 \begin{equation}
-   \forall u\in\rules:\ \sat{u}{\dataset}\label{eqn:satisfaction}
 \end{equation}
@@
    so every rule has at least one role in the relation $\maintain$, means that all roles together are keeping all rules satisfied.
+
+   \subsection{Changing data}
+   \label{sct:Changing data}
+   @Bas, In het onderstaande heb ik mij er nog wat te makkelijk vanaf gemaakt.
+   Aandacht nodig.
+
+      Events can cause data to change.
+      So, we model events as functions $\Dataset\rightarrow\Dataset$.
+      For triples, we distinguish just two types of event: insert and delete events.
+      An event is a label (ins or del) with a dataset that is to be inserted in or deleted from another dataset.
+      We denote events as $\ins{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}$ and $\del{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}$.
+      They are defined by:
+   \begin{eqnarray}
+      \ins{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}(\dataset)&=&\pair{\triples_{\dataset}\cup\triples_{\id{ev}}}{\inst_{\dataset}\cup\inst_{\id{ev}}}\\
+      \del{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}(\dataset)&=&\pair{\triples_{\dataset}-\triples_{\id{ev}}}{\inst_{\dataset}-\inst_{\id{ev}}}
+   \end{eqnarray}
+      Insert events are represented as $\ins{\dataset}$ and delete events by $\del{\dataset}$.
+      Both are functions $\Dataset\rightarrow\Dataset$, which are defined as:
+      Since datasets are sets, we will use set union ($\cup$) and set difference ($-$) to make changes in datasets.
 

### 2023-03-13 2022 Migration/articleMigration.tex
WIP

@@
    Requirement~\ref{eqn:satisfaction} ensures that the dataset satisfies all rules in the schema.
-
-   The maintainance relation between roles and rules is used to determine who is allowed to change the dataset.
-   For every rule $u$ in the schema, there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
-   that represents the set of violations%
-
    The rules in the schema serve as invariants,
@@
    In those cases, something or someone must take action to fix the situation and restore invariance of that rule.
-
-   If a triple $\triple{a}{r}{b}$ satisfies requirement~\ref{eqn:define R} and its relation $r$ satisfies requirement~\ref{eqn:relationsIntroduceConcepts},
-   this triple is said to {\em befit} schema $\quadruple{\concepts}{\isa}{\rels}{\rules}$.
-
-   The system is {\em at rest} when every rule in the schema is satisfied:
+   The system is {\em at rest} when every rule in the schema is satisfied.
+
+   % The maintainance relation between roles and rules is used to determine who is allowed to change the dataset.
+   % For every rule $u$ in the schema, there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
+   % that represents the set of violations.%
+
+\subsection{Changing data}
+\label{sct:Changing data}
+   Events can cause a dataset to change.
+   So, we model events as functions $\Dataset\rightarrow\Dataset$.
+   We will use variable $e$, $e'$, $e''$, and so on to represent events.
+   We distinguish just two types of event: insert and delete events.
+   An event is a label (\id{ins} or \id{del}) with a dataset that is to be inserted or deleted.
+   If $\triples$ is the set of triples inserted or deleted by an event $e$,
+   we denote that event either by $\ins{\triples}$ or $\del{\triples}$.
+   They are defined by:
+\begin{eqnarray}
+   e(\dataset)&=&\id{case}\ e\ \id{of}\\
+   &&\begin{array}[t]{@{}rcl}
+      \ins{\triples}&=&\pair{\triples_{\dataset}\cup\triples}{\id{insInst}_{\triples}(\inst_{\dataset})}\\
+      \del{\triples}&=&\pair{\triples_{\dataset}-\triples}{\id{delInst}_{\triples}(\inst_{\dataset})}
+   \end{array}
+\end{eqnarray}
+   in which $\id{insInst}_{\triples}$ transforms $\inst_{\dataset}$ such that $\ins{\triples}(\dataset)$ complies with
+   equations~\ref{eqn:specialization} and~\ref{eqn:wellTypedEdge},
+   and likewise for $\id{delInst}_{\triples}$.
+
+   As the system registers events, it inserts or deletes triples in its dataset.
+   This may violate requirement~\ref{eqn:satisfaction} when one or more rules are broken,
+   i.e.\ if there is a rule $u$ for which $\sat{u}{e(\dataset)}$ does not hold.
+   In that case, requirement~\ref{eqn:satisfaction} prohibits the system to be in state $e(\dataset)$,
+   so either the system stays in the state $\dataset$ or there is a reaction $e'$ that restores invariance of the broken rules:
 \begin{equation}
+   \sat{u}{\dataset}\Rightarrow\sat{u}{e'(e(\dataset))}
+   \label{eqn:restoring}
 \end{equation}
-   However, the system continually registers events that insert or delete triples in its dataset,
-   which may cause rules to be broken.
-   This requires a reaction to restore invariance.
-   Example: if a rule says that every request must be acknowledged, the receipt of a request (which is an event) will violate this rule.
-   Sending a message that acknowledges this receipt (another event) will cause the rule to be satisfied again.
-   During the time between the receipt of the request and sending the acknowledgment, the rule is (temporarily) not satisfied.
-
-   This paper assumes two ways of enforcing satisfaction of rules: automatic or manual.
-   Manual enforcement is specified by the developer by assigning a rule $u$ to a role $o$.
-   Any actor with role $o$, whether person or machine,
+   The combination of events $e$ and $e'$ is called a {\em transaction}.
+
+   In practice, there are different ways of enforcing satisfaction of rules.
+   One way is to let a user or a computer ``invent'' $e'$.
+   A developer can specify this by assigning a rule $u$ to a role $o$.
+   Any actor with role $o$
    can restore invariance of rule $u$ by inserting or deleting the right triples.
@@
 
+   If $\maintain$ specifies no role for a given rule $u$, the system can prevent $e$ from having any effect.
+   This implements blocking behaviour, and must be accompanied by an error message.
+
    Automatic enforcement is specified by the developer with a special syntax,
@@
    so every rule has at least one role in the relation $\maintain$, means that all roles together are keeping all rules satisfied.
-
-   \subsection{Changing data}
-   \label{sct:Changing data}
-   @Bas, In het onderstaande heb ik mij er nog wat te makkelijk vanaf gemaakt.
-   Aandacht nodig.
-
-      Events can cause data to change.
-      So, we model events as functions $\Dataset\rightarrow\Dataset$.
-      For triples, we distinguish just two types of event: insert and delete events.
-      An event is a label (ins or del) with a dataset that is to be inserted in or deleted from another dataset.
-      We denote events as $\ins{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}$ and $\del{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}$.
-      They are defined by:
-   \begin{eqnarray}
-      \ins{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}(\dataset)&=&\pair{\triples_{\dataset}\cup\triples_{\id{ev}}}{\inst_{\dataset}\cup\inst_{\id{ev}}}\\
-      \del{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}(\dataset)&=&\pair{\triples_{\dataset}-\triples_{\id{ev}}}{\inst_{\dataset}-\inst_{\id{ev}}}
-   \end{eqnarray}
-      Insert events are represented as $\ins{\dataset}$ and delete events by $\del{\dataset}$.
-      Both are functions $\Dataset\rightarrow\Dataset$, which are defined as:
-      Since datasets are sets, we will use set union ($\cup$) and set difference ($-$) to make changes in datasets.
 
@@
 The ${\tt ENFORCE\ }r{\tt\ :=\ }e$ statements add triples $(x,r,y)$ for all $(x,y)\in e$.
-The expressions $e$ in these statements are such that $x\ (\inst \compose \kleenestar{\flip{\isa}})\ (A',1)$ and $y \ (\inst \compose \kleenestar{\flip{\isa}})\ (B',1)$, thus preserving the property that the result is a dataset.
+The terms $t$ in these statements are such that $x\ (\inst \compose \kleenestar{\flip{\isa}})\ (A',1)$ and $y \ (\inst \compose \kleenestar{\flip{\isa}})\ (B',1)$, thus preserving the property that the result is a dataset.
 

### 2023-03-19 2022 Migration/articleMigration.tex

@@
    The notation $r=\declare{n}{A}{B}$ denotes that relation $r$ has name $n$, source concept $A$, and target concept $B$.
+   The part $\pair{A}{B}$ is called the {\em signature} of the relation.
+
+   Specialization distributes in a straightforward way over signatures:
+\begin{equation}
+   \pair{A}{B}\isa\pair{C}{D}\ \Leftrightarrow\ A\isa C\ \wedge\ B\isa D
+\label{eqn:isa-signature}
+\end{equation}
 
@@
    \pop{r}{\dataset}\ =\ \{ \pair{a}{b}\mid\ \triple{a}{r}{b}\in\triples_{\dataset}\}
-\label{eqn:pop}
+\label{eqn:pop-rel}
 \end{equation}
    Equation~\ref{eqn:wellTypedEdge} implies that for every dataset $\dataset$:
-\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ a\inst A\ \wedge\ b\inst B\]
+\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ a\inst_{\dataset}A\ \wedge\ b\inst_{\dataset}B\]
    For a developer, this means that the type of an atom depends only on the relation in which it resides; not on the actual population of the database.
-   This allows for static typing, which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.
+
+   Concepts too can be seen as containers of atoms,
+   defined by the function $\id{pop}_C:\Dataset\rightarrow\powerset{\Atoms}$.
+\begin{equation}
+   \pop{C}{\dataset}\ =\ \{ x\mid\ x\ \inst_{\dataset}\ C\}
+\label{eqn:pop-concept}
+\end{equation}
 
@@
    \label{eqn:isasIntroduceConcepts}\\
-   \isa&&\text{is a partial order on $\concepts$}
+   \pair{A}{B}\isa\pair{C}{D}\ \wedge\ \declare{n}{C}{D}\in\rels&\Rightarrow&
+   \begin{array}[t]{@{}l}
+      \declare{n}{A}{B}\in\rels\ \wedge\\
+      \pop{\declare{n}{A}{B}}{\dataset}\subseteq\pop{\declare{n}{C}{D}}{\dataset}
+   \end{array}
+   \label{eqn:subrelations} \\
+   \isa\ \text{is a partial order}
    \label{eqn:isasPartialOrder}\\
-   \forall u\in\rules,\exists\dataset\in\Dataset-\emptyset&:&\sat{u}{\dataset}
+   \exists\dataset\in\Dataset-\emptyset:\ \sat{u}{\dataset}
    \label{eqn:trueifyableRules}
 \end{eqnarray}
+   The type system of Ampersand~\cite{vdWoude2011} ensures that every information system it generates satisfies these requirements.
    Note that these requirements do not depend on any particular dataset,
-   so they can be checked statically.
-   The type system of Ampersand~\cite{vdWoude2011} ensures that every information system it generates satisfies these requirements.
+   so they can be checked at compile time.
+   This is also known as static typing,
+   which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.
 
@@
    Requirement~\ref{eqn:satisfaction} ensures that the dataset satisfies all rules in the schema.
-   The rules in the schema serve as invariants,
-   i.e.\ constraints that the data in the database must satisfy at all times.
-   At times in which the dataset does not satisfy a rule, that rule is said to be {\em broken}.
-   In those cases, something or someone must take action to fix the situation and restore invariance of that rule.
-   The system is {\em at rest} when every rule in the schema is satisfied.
+   Note that these requirements are enforced at runtime.
+
+\subsection{Changing data}
+\label{sct:Changing data}
+   Every rule in the schema serves as invariant,
+   i.e.\ a constraint that the data in the database must satisfy at all times.
+   If an actor tries to change the dataset in a way that violates a rule,
+   the system will either reject that change or help the actor to find additional changes to keep that rule satisfied.
 
@@
 
-\subsection{Changing data}
-\label{sct:Changing data}
    Events can cause a dataset to change.
    So, we model events as functions $\Dataset\rightarrow\Dataset$.
-   We will use variable $e$, $e'$, $e''$, and so on to represent events.
+   We will use variables $e$, $e'$, $e''$, and so on to represent events.
    We distinguish just two types of event: insert and delete events.
@@
    we denote that event either by $\ins{\triples}$ or $\del{\triples}$.
-   They are defined by:
-\begin{eqnarray}
-   e(\dataset)&=&\id{case}\ e\ \id{of}\\
+   Equation~\ref{eqn:event} defines the effect of an event on a dataset.
+\begin{equation}
+   \begin{array}{rcl}
+      e(\dataset)&=&\text{case}\ e\ \text{of}\\
       &&\begin{array}[t]{@{}rcl}
@@
       \end{array}
-\end{eqnarray}
+   \end{array}
+   \label{eqn:event}
+\end{equation}
    in which $\id{insInst}_{\triples}$ transforms $\inst_{\dataset}$ such that $\ins{\triples}(\dataset)$ complies with
@@
    As the system registers events, it inserts or deletes triples in its dataset.
-   This may violate requirement~\ref{eqn:satisfaction} when one or more rules are broken,
-   i.e.\ if there is a rule $u$ for which $\sat{u}{e(\dataset)}$ does not hold.
-   In that case, requirement~\ref{eqn:satisfaction} prohibits the system to be in state $e(\dataset)$,
-   so either the system stays in the state $\dataset$ or there is a reaction $e'$ that restores invariance of the broken rules:
+   If such an event would violate requirement~\ref{eqn:satisfaction},
+   either the system stays in the state $\dataset$ or there is a reaction $e'$ that restores invariance of the affected rules:
 \begin{equation}
@@
 \end{equation}
+   In this way, the system can restore invariance of rules and keep requirement~\ref{eqn:satisfaction} satisfied.
    The combination of events $e$ and $e'$ is called a {\em transaction}.
 
-   In practice, there are different ways of enforcing satisfaction of rules.
-   One way is to let a user or a computer ``invent'' $e'$.
-   A developer can specify this by assigning a rule $u$ to a role $o$.
-   Any actor with role $o$
-   can restore invariance of rule $u$ by inserting or deleting the right triples.
-   For this reason, every rule is assigned to at least one role by means of the relation $\maintain$.
-\begin{equation}
-   \forall u\in\rules\ \exists o\in\roles:\ o\maintain u\label{eqn:maintain}
-\end{equation}
-   $o\maintain u$ means that role $o$ must keep rule $u$ satisfied, i.e.\ to keep $\sat{u}{\dataset}$ true.
-   As long as rule $u$ is broken, the system should notify actors in the role(s) that are administered in $\maintain$ to do something about it.
-
-   If $\maintain$ specifies no role for a given rule $u$, the system can prevent $e$ from having any effect.
+   In practice, there are different ways of enforcing satisfaction of rules: automatic enforcement and manual enforcement.
+
+   Automatic enforcement is specified by the developer with a special syntax:
+   the {\em enforce rule}.
+   An enforce rule specifies not only the rule,
+   but it implicitly defines event $e'$ to restore the invariance.
+   The system must have an engine that restores invariance of all enforce rules,
+   so users will not notice any violations.
+
+   Manual enforcement means that the system (temporarily) allows violations,
+   to let the user ``invent'' a reaction $e'$ that restores invariance.
+   For this purpose, Ampersand relaxes requirement~\ref{eqn:satisfaction} as follows:
+\begin{eqnarray}
+   \forall u\in\rules&:&(\exists o\in\roles:\ o\maintain u)\ \vee\ \sat{u}{\dataset}
+   \label{eqn:satisfaction relaxed}
+\end{eqnarray}
+   The purpose of this requirement is to allow the compiler to generate code for users with role $o$,
+   to restore invariance.
+   In the Ampersand language, a developer explicitly assigns a rule $u$ to a role $o$,
+   to keep control over the type of persons that are allowed to ``break rules temporarily''.
+   An actor, who attempts an event that violates a rule, should receive notification of the system with an appropriate error message.
+
+   If a rule $u$ is not assigned to any role and it is not an enforce rule,
+   the system must prevent any event $e$ from having any effect on the database.
+   So, if $e$ would violate rule $u$, the system must reject $e$.
    This implements blocking behaviour, and must be accompanied by an error message.
-
-   Automatic enforcement is specified by the developer with a special syntax,
-   the {\em enforce rule}.
-   An enforce rule specifies not only the rule, but also the way to restore invariance.
-   The system features an engine that restores invariance of all enforce rules without delay.
-   Whenever the dataset violates an enforce rule,
-   this engine will make other changes to the dataset to satisfy that rule.
-   Hence, the rule is not satisfied during a small, finite amount of time.
-
-   The system contains a dedicated role for the enforcement engine,
-   so every rule has at least one role in the relation $\maintain$, means that all roles together are keeping all rules satisfied.
 

### 2023-03-19 2023 Disambiguation/Disambiguation.tex
work in progress on Disambiguation article

@@
 \usepackage[cal=boondoxo]{mathalfa}
+
+\DeclareMathAlphabet{\pazocal}{OMS}{zplm}{m}{n}
+\newcommand{\powerset}{\pazocal{P}}
+
 \newcommand{\typesemi}{\mathbin{\text{\textbb{;}}}}
@@
 \begin{align*}
-x_{A*B} &: A * B & \text{For $x \in \mathcal{N} \cup \{\epsilon, T, F\}$ and $A,B\in \mathcal{T}$}\\
+x_{A*B} &: A * B & \text{For $x \in \mathcal{N} \cup \{\epsilon, T, F\}$}\\
 1_{A} &: A * A&\\
@@
 \end{align*}
-We say that a term $r\in \mathcal{R}$ is \emph{well-typed} if $r : A*B$ for some $A$ and $B$.
-By induction on the inductive definition of $\mathcal{R}$, the relation $:$ is between $\mathcal{R}$ and types $\mathcal{T}\times\mathcal{T}$ and univalent.
+We say that a term $r\in \mathcal{R}$ is \emph{well-typed} if $r : A*B$ for $A,B\in \mathcal{T}$ and every sub-term is well-typed.
+By induction on the inductive definition of $\mathcal{R}$, the relation $:$ is univalent.
 In other words: well-typed terms in $\mathcal{R}$ have precisely one type, all of its sub-terms are well-typed, and $\bot$ does not occur as a pseudo-concept.
@@
 
-\paragraph{[Aside],} The Ampersand language also allows for \verb=CLASSIFY A IS B /\ C= statements, meaning that elements are in $A$ if and only if they are in $B$ and in $C$.
+\paragraph{[Aside]} The Ampersand language also allows for \verb=CLASSIFY A IS B /\ C= statements, meaning that elements are in $A$ if and only if they are in $B$ and in $C$.
 The information about this captured in $\preceq$ is just that $A \preceq B$ and $A\preceq C$, which means that information is lost.
@@
 However, it is an open problem to deal with this with the $\typeinter$ operation, so we give a simpler presentation that ignores this additional information in \verb=CLASSIFY _ IS _= statements for now.
+
+Translating a term from $\mathcal{D}$ to $\mathcal{R}$ proceeds by considering which choices are type correct.
+However, if we were to simply implement the simple algorithm proposed earlier, too many terms would be deemed ambiguous:
+the term $\typetyped{\typeident}{\mathrm{Event}} \typesemi \typevee \typesemi \typetyped{\mathrm{scheduled}}{\mathrm{Event}*\mathrm{Time}}$ with $\mathrm{Task}\preceq\mathrm{Event}$ would not disambiguate to the obvious term, since $\typetyped{\typeident}{\mathrm{Event}} \typesemi \typetyped{\typevee}{\mathrm{Task}*\mathrm{Task}} \typesemi \typetyped{\mathrm{scheduled}}{\mathrm{Event}*\mathrm{Time}}$ is also well-typed.
+To deal with this, we prefer terms in which the $\epsilon$ relations have the same source and target.
+
+Given a term $D\in\mathcal{D}$, the function $p(D)$ will return a set of types and a preference relation between those types.
+The preference relation is a pre-order.
+We will write a tree node as $t \mapsfrom \{c_1,\ldots,c_n\}$ where the child-nodes $c_i$ are themselves written as trees, and $t$ is a set of types that the corresponding term could have.
+For example, if $p(D) = t_1 \mapsfrom \{t_2 \mapsfrom \{t_3 \mapsfrom \{\}\},t_4 \mapsfrom \{\}\}$, then types of $t_1$ are preferred over those in $t_2$, $t_3$ and $t_4$, those of $t_2$ are preferred over $t_3$, and those types in $t_4$, are incomparable to those in $t_2$ and $t_3$.
+
+We define a binary operation $\times$ on such trees as follows:
+\[
+(a \mapsfrom A) \times (b \mapsfrom B)
+= a \cap b \mapsfrom 
+\]
+
+We define $p$ as follows:
+\begin{align*}
+p(\xi) &= \{{A*B} \mid \xi : (A,B)\}\mapsfrom\{\}\\
+p(\Xi) &= \{{A*A} \mid \xi = \epsilon_{A*A}\}\mapsfrom \{\{{A*B} \mid \xi : {A*B} \wedge (A\neq B \vee \xi \neq \epsilon_{A*B})\} \mapsfrom\{\}\}\\
+p(\typetyped{\varphi}{A*B}) &= \{A*B \mid A*B\in p(\varphi)[i] \text{ for some $i$}\}\mapsfrom\{\}\\
+p(\varphi \cup \psi) &= ..?\\
+p(\varphi \cap \psi) &= p(\varphi \cup \psi)\\
+p(\varphi ; \psi) &= ..?\\
+p(\overline{\varphi}) &= p(\varphi)\\
+p(\conv{\varphi}) &= [\{B*A\mid A*B\in p(\varphi)[i]\}]_i
+\end{align*}
+
+\paragraph{[Aside]}
+Without optimizations, the complexity of the algorithm for constructing these trees is exponential: the combination of two trees for $\cap$, $\cup$ and $;$ will be of a size that is the product of the two.
+However, we hope to be able to reduce the size of trees by looking for equivalent trees that express the same preferences.
+If trees can be reduced so that the number of nodes never exceeds the number of types in them, this should give the tree construction polynomial complexity.
 

### 2023-03-22 2022 Migration/articleMigration.tex
WIP

@@
 
-   The relation $\inst:\Pair{\Atoms}{\Concepts}$ is the instance relation between atoms and concepts.
+   The relation $\inst:\Pair{\Atoms}{\Concepts}$ relates atoms to concepts.
    The expression $a\inst C$ means that atom $a$ is an \emph{instance} of concept $C$.
@@
 
-   Specialization distributes in a straightforward way over signatures:
-\begin{equation}
-   \pair{A}{B}\isa\pair{C}{D}\ \Leftrightarrow\ A\isa C\ \wedge\ B\isa D
-\label{eqn:isa-signature}
-\end{equation}
+%    The following is for wide tables only
+%    Specialization distributes in a straightforward way over signatures:
+% \begin{equation}
+%    \pair{A}{B}\isa\pair{C}{D}\ \Leftrightarrow\ A\isa C\ \wedge\ B\isa D
+% \label{eqn:isa-signature}
+% \end{equation}
 
@@
    This ``meaning from practice'' has no consequences in the formal world.
-   Users are entirely free to attach a practical meaning to a triple.
+   Users are free to attach any practical meaning to a triple.
 
@@
 \label{sct:Schemas}
-   To design an information system, a developer defines a schema in which she defines concepts, specializations, relations, and rules.
-   Together, these items capture the semantics of an information system~\cite{Spivak2012}.
+   Schemas serve to capture the semantics of an information system~\cite{Spivak2012}.
+   A schema defines concepts, specializations, relations, and rules.
+   A software engineer defines a schema on design time, so that semantic checks can be implemented at compile time.
+
    A schema $\schema$ is a tuple $\quadruple{\concepts}{\isa}{\rels}{\rules}$,
@@
    and $\rules$ is a finite set of rules.
-   Each rule in a schema serves to constrain the dataset, to ensure its semantic integrity.
+   Each rule in a schema serves to constrain the dataset at runtime, to ensure its semantic integrity.
    Every rule is an element of an infinite set called $\Rules$,
@@
    \label{eqn:isasIntroduceConcepts}\\
-   \pair{A}{B}\isa\pair{C}{D}\ \wedge\ \declare{n}{C}{D}\in\rels&\Rightarrow&
-   \begin{array}[t]{@{}l}
-      \declare{n}{A}{B}\in\rels\ \wedge\\
-      \pop{\declare{n}{A}{B}}{\dataset}\subseteq\pop{\declare{n}{C}{D}}{\dataset}
-   \end{array}
-   \label{eqn:subrelations} \\
+%    The following is for wide tables only
+%    \pair{A}{B}\isa\pair{C}{D}\ \wedge\ \declare{n}{C}{D}\in\rels&\Rightarrow&\declare{n}{A}{B}\in\rels
+%    \label{eqn:subrelations}\\
    \isa\ \text{is a partial order}
@@
    \exists\dataset\in\Dataset-\emptyset:\ \sat{u}{\dataset}
-   \label{eqn:trueifyableRules}
+   \label{eqn:consistentRules}
 \end{eqnarray}
+   Requirement~\ref{eqn:relationsIntroduceConcepts} ensures that concepts mentioned in a relation are defined in the schema.
+   Requirement~\ref{eqn:isasIntroduceConcepts} ensures that the specialization relation works on concepts of $\concepts$.
+%    The following is for wide tables only
+%    Requirement~\ref{eqn:subrelations} ensures that all specializations of a relation exist.
+   Requirement~\ref{eqn:isasPartialOrder} ensures that the specialization relation contains no cycles.
+   Requirement~\ref{eqn:consistentRules} ensures that every rule is consistent.
+
    The type system of Ampersand~\cite{vdWoude2011} ensures that every information system it generates satisfies these requirements.
@@
    \label{eqn:satisfaction}
+%    The following is for wide tables only
+%    \pair{A}{B}\isa\pair{C}{D}\ \wedge\ \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\triple{a}{\declare{n}{C}{D}}{b}\in\triples
+%    \label{eqn:subpopulations}
 \end{eqnarray}
@@
    Requirement~\ref{eqn:satisfaction} ensures that the dataset satisfies all rules in the schema.
+%    The following is for wide tables only
+%    Requirement~\ref{eqn:subpopulations} ensures that specialization works across relations.
    Note that these requirements are enforced at runtime.
-
-\subsection{Changing data}
-\label{sct:Changing data}
-   Every rule in the schema serves as invariant,
-   i.e.\ a constraint that the data in the database must satisfy at all times.
-   If an actor tries to change the dataset in a way that violates a rule,
-   the system will either reject that change or help the actor to find additional changes to keep that rule satisfied.
-
-   % The maintainance relation between roles and rules is used to determine who is allowed to change the dataset.
-   % For every rule $u$ in the schema, there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
-   % that represents the set of violations.%
-
-   Events can cause a dataset to change.
-   So, we model events as functions $\Dataset\rightarrow\Dataset$.
-   We will use variables $e$, $e'$, $e''$, and so on to represent events.
-   We distinguish just two types of event: insert and delete events.
-   An event is a label (\id{ins} or \id{del}) with a dataset that is to be inserted or deleted.
-   If $\triples$ is the set of triples inserted or deleted by an event $e$,
-   we denote that event either by $\ins{\triples}$ or $\del{\triples}$.
-   Equation~\ref{eqn:event} defines the effect of an event on a dataset.
-\begin{equation}
-   \begin{array}{rcl}
-      e(\dataset)&=&\text{case}\ e\ \text{of}\\
-      &&\begin{array}[t]{@{}rcl}
-         \ins{\triples}&=&\pair{\triples_{\dataset}\cup\triples}{\id{insInst}_{\triples}(\inst_{\dataset})}\\
-         \del{\triples}&=&\pair{\triples_{\dataset}-\triples}{\id{delInst}_{\triples}(\inst_{\dataset})}
-      \end{array}
-   \end{array}
-   \label{eqn:event}
-\end{equation}
-   in which $\id{insInst}_{\triples}$ transforms $\inst_{\dataset}$ such that $\ins{\triples}(\dataset)$ complies with
-   equations~\ref{eqn:specialization} and~\ref{eqn:wellTypedEdge},
-   and likewise for $\id{delInst}_{\triples}$.
-
-   As the system registers events, it inserts or deletes triples in its dataset.
-   If such an event would violate requirement~\ref{eqn:satisfaction},
-   either the system stays in the state $\dataset$ or there is a reaction $e'$ that restores invariance of the affected rules:
-\begin{equation}
-   \sat{u}{\dataset}\Rightarrow\sat{u}{e'(e(\dataset))}
-   \label{eqn:restoring}
-\end{equation}
-   In this way, the system can restore invariance of rules and keep requirement~\ref{eqn:satisfaction} satisfied.
-   The combination of events $e$ and $e'$ is called a {\em transaction}.
-
-   In practice, there are different ways of enforcing satisfaction of rules: automatic enforcement and manual enforcement.
-
-   Automatic enforcement is specified by the developer with a special syntax:
-   the {\em enforce rule}.
-   An enforce rule specifies not only the rule,
-   but it implicitly defines event $e'$ to restore the invariance.
-   The system must have an engine that restores invariance of all enforce rules,
-   so users will not notice any violations.
-
-   Manual enforcement means that the system (temporarily) allows violations,
-   to let the user ``invent'' a reaction $e'$ that restores invariance.
-   For this purpose, Ampersand relaxes requirement~\ref{eqn:satisfaction} as follows:
-\begin{eqnarray}
-   \forall u\in\rules&:&(\exists o\in\roles:\ o\maintain u)\ \vee\ \sat{u}{\dataset}
-   \label{eqn:satisfaction relaxed}
-\end{eqnarray}
-   The purpose of this requirement is to allow the compiler to generate code for users with role $o$,
-   to restore invariance.
-   In the Ampersand language, a developer explicitly assigns a rule $u$ to a role $o$,
-   to keep control over the type of persons that are allowed to ``break rules temporarily''.
-   An actor, who attempts an event that violates a rule, should receive notification of the system with an appropriate error message.
-
-   If a rule $u$ is not assigned to any role and it is not an enforce rule,
-   the system must prevent any event $e$ from having any effect on the database.
-   So, if $e$ would violate rule $u$, the system must reject $e$.
-   This implements blocking behaviour, and must be accompanied by an error message.
 
@@
    The third and last relation states which students are enrolled for which module.
-   It is left empty for now.
+   It is left without population for now.
 \begin{verbatim}
@@
 
-   A rule, {\tt EnrollRule} completes the semantics
+   A rule, {\tt EnrollRule} completes the script.
    It states that a student can enroll for any module that is part of a course he or she takes.
@@
    The Ampersand compiler generates a schema $\schema$, which contains concepts, a specialization relation, relations, and rules.
-   It defines the set of concepts to satisfy requirement~\ref{eqn:relationsIntroduceConcepts}:
-   \[\concepts=\{ {\tt Module}, {\tt Course}, {\tt Student}\}\]
-   The user has defined no specializations.
-   This means that $\isa$ equals the identity relation on $\concepts$ and requirement~\ref{eqn:isasIntroduceConcepts} is satisfied.
-   The set of relations, $\rels$, satisfies requirement~\ref{eqn:define R}:
+   It defines the sets of concepts and relations to satisfy requirement~\ref{eqn:relationsIntroduceConcepts}:
 \[\begin{array}{rcl}
+   \concepts&=&\{ {\tt Module}, {\tt Course}, {\tt Student}\}\\
    \rels&=&\{\begin{array}[t]{l}
@@
 \]
-   And, it defines the set of rules $\rules$ to contain just one rule: \verb-EnrollRule-.
-   So, the schema $\schema=\quadruple{\concepts}{\isa}{\rels}{\rules}$ satisfies the requirements from section~\ref{sct:Schemas}.
+   The user has defined no specializations.
+   This means that $\isa$ equals the identity relation on $\concepts$ and requirements~\ref{eqn:isasIntroduceConcepts} and~\ref{eqn:isasPartialOrder} are satisfied.
+   The script defines the set of rules $\rules$ to contain just one rule: \verb-EnrollRule-.
+   The (nonempty) dataset satisfies this rule, which meets requirement~\ref{eqn:consistentRules}.
+   So, the schema $\schema=\quadruple{\concepts}{\isa}{\rels}{\rules}$ satisfies all requirements from section~\ref{sct:Schemas}.
 
@@
    This concludes the argument that $\infsys$ is an information system.
+
+\subsection{Changing data}
+\label{sct:Changing data}
+   Every rule in the schema serves as invariant,
+   i.e.\ a constraint that the data in the database must satisfy at all times.
+   If an actor tries to change the dataset in a way that violates a rule,
+   the system will either reject that change or help the actor to find additional changes to keep that rule satisfied.
+
+   % The maintainance relation between roles and rules is used to determine who is allowed to change the dataset.
+   % For every rule $u$ in the schema, there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
+   % that represents the set of violations.%
+
+   Events can cause a dataset to change.
+   So, we model events as functions $\Dataset\rightarrow\Dataset$.
+   We will use variables $e$, $e'$, $e''$, and so on to represent events.
+   We distinguish just two types of event: insert and delete events.
+   An event is a label (\id{ins} or \id{del}) with a dataset that is to be inserted or deleted.
+   If $\triples$ is the set of triples inserted or deleted by an event $e$,
+   we denote that event either by $\ins{\triples}$ or $\del{\triples}$.
+   Equation~\ref{eqn:event} defines the effect of an event on a dataset.
+\begin{equation}
+   \begin{array}{rcl}
+      e(\dataset)&=&\text{case}\ e\ \text{of}\\
+      &&\begin{array}[t]{@{}rcl}
+         \ins{\triples}&=&\pair{\triples_{\dataset}\cup\triples}{\id{insInst}_{\triples}(\inst_{\dataset})}\\
+         \del{\triples}&=&\pair{\triples_{\dataset}-\triples}{\id{delInst}_{\triples}(\inst_{\dataset})}
+      \end{array}
+   \end{array}
+   \label{eqn:event}
+\end{equation}
+   in which $\id{insInst}_{\triples}$ transforms $\inst_{\dataset}$ such that $\ins{\triples}(\dataset)$ complies with
+   equations~\ref{eqn:specialization} and~\ref{eqn:wellTypedEdge},
+   and likewise for $\id{delInst}_{\triples}$.
+
+   As the system registers events, it inserts or deletes triples in its dataset.
+   If such an event would violate requirement~\ref{eqn:satisfaction},
+   either the system stays in the state $\dataset$ or there is a reaction $e'$ that restores invariance of the affected rules:
+\begin{equation}
+   \sat{u}{\dataset}\Rightarrow\sat{u}{e'(e(\dataset))}
+   \label{eqn:restoring}
+\end{equation}
+   In this way, the system can restore invariance of rules and keep requirement~\ref{eqn:satisfaction} satisfied.
+   The combination of events $e$ and $e'$ is called a {\em transaction}.
+
+   In practice, there are different ways of enforcing satisfaction of rules: automatic enforcement and manual enforcement.
+
+   Automatic enforcement is specified by the developer with a special syntax:
+   the {\em enforce rule}.
+   An enforce rule specifies not only the rule,
+   but it implicitly defines event $e'$ to restore the invariance.
+   The system must have an engine that restores invariance of all enforce rules,
+   so users will not notice any violations.
+
+   Manual enforcement means that the system (temporarily) allows violations,
+   to let the user ``invent'' a reaction $e'$ that restores invariance.
+   For this purpose, Ampersand relaxes requirement~\ref{eqn:satisfaction} as follows:
+\begin{eqnarray}
+   \forall u\in\rules&:&(\exists o\in\roles:\ o\maintain u)\ \vee\ \sat{u}{\dataset}
+   \label{eqn:satisfaction relaxed}
+\end{eqnarray}
+   The purpose of this requirement is to allow the compiler to generate code for users with role $o$,
+   to restore invariance ``by hand''.
+   In the Ampersand language, a developer explicitly assigns a rule $u$ to a role $o$,
+   to keep control over the type of persons that are allowed to ``restore broken rules''.
+   As a consequence, a mechanism is needed to notify those persons of the work to be done.
+
+   If a rule $u$ is not assigned to any role and it is not an enforce rule,
+   the system must prevent any event $e$ from having any effect on the database.
+   So, if $e$ would violate rule $u$, the system must reject $e$.
+   This implements blocking behaviour, and must be accompanied by an error message.
 
@@
 \end{eqnarray}
+   Understanding what the engine does to satisfy these rules might help the reader, so let us look what happens at runtime.
    Initially, when {\tt copied\_takes} is still empty, the difference between {\tt old.takes} and {\tt copied\_takes} equals {\tt old.takes}.
    So, {\tt new.takes} will start filling up with the pairs of {\tt old.takes} by rule~\ref*{difference}.
-   Then, the engine will start filling {\tt copied\_takes} with pairs from {\tt new.takes} because of rule~\ref*{fill copies with new.takes}.
+   Then, the engine starts filling {\tt copied\_takes} with pairs from {\tt new.takes} because of rule~\ref*{fill copies with new.takes}.
    As a result, the difference between {\tt old.takes} and {\tt copied\_takes} gets smaller and smaller, until {\tt old.takes} and {\tt copied\_takes} are equal.
@@
    RELATION copied_isEnrolledFor[Student*Module]
-   ENFORCE new.isEnrolledFor >: old.isEnrolledFor - copied_isEnrolledFor
+   ENFORCE new.isEnrolledFor >:
+           old.isEnrolledFor - copied_isEnrolledFor
    ENFORCE copied_isEnrolledFor >: new.isEnrolledFor
 
-   ROLE MigrationEngineer MAINTAINS migrating_takes, migrating_isPartOf, migrating_isEnrolledFor
+   ROLE MigrationEngineer MAINTAINS migrating_takes,
+                                    migrating_isPartOf,
+                                    migrating_isEnrolledFor
    RULE migrating_takes : old.takes-copied_takes = -V
    RULE migrating_isPartOf : old.isPartOf-copied_isPartOf = -V
-   RULE migrating_isEnrolledFor : old.isEnrolledFor-copied_isEnrolledFor = -V
+   RULE migrating_isEnrolledFor :
+        old.isEnrolledFor-copied_isEnrolledFor = -V
 \end{verbatim}

### 2023-03-25 2022 Migration/articleMigration.tex
fixed some textual inconsistencies.

@@
 \footnote{Allegories are a specific type of categories, which the reader does not need to understand for the purpose of this paper}~\cite{Zielinski2013}.
-   Each rule in Ampersand is a constraint (aka invariant), which the Ampersand run-time engine keeps satisfied as long as that rule lives%
-\footnote{The dotted lines in the diagrams show which set of rules the engine keeps satisfied.}.
+   Each rule in Ampersand is a constraint (aka invariant), which are kept satisfied as long as that rule lives%
+\footnote{The dotted lines in the diagrams show which set of rules is enforced.}.
    So, the user spends no effort to extract constraints from specifications or (even more laborious) from code
@@
    Such violations do not occur in our theory because the existing system was generated by Ampersand,
-   so the existing system has kept all constraints from the schema satisfied.
+   so the dataset satisfies all constraints in the schema.
    This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
@@
    so the migration engineer can change everything she wants to suit the particulars of the data migration.
-
-%    A naive migration might proceed as follows:
-%    Launch the desired system in parallel to the existing one, copy data from the existing to the desired system, and have everyone use the desired system.
-%    However, numerous issues might impede that plan.
-%    The following examples illustrate the practical issues that may occur:
-% \begin{enumerate}
-% \item Data required in the desired system is missing in the existing system.
-%    There may be no way in the existing system to enter that data.
-%    An example could be that every reimbursement form needs to have an address associated to it to mail the check to, but address information is not stored in the existing system:
-%    The existing system required the reimbursement office to look up employee's addresses from a hand-written list they had on their desk.
-
-%    This issue will require somebody to insert the missing addresses in the desired system.
-% \item Data in the existing system is wrong but cannot be corrected there due to how the existing system was designed.
-%    An example is if the existing system only allows approvals to be entered as the current user, and the CEO has always insisted that her administrative staff enters the approvals into the system for her.
-%    This may result in approvals being entered as admin staff, where it was really the CEO making the approval.
-
-%    This issue will require that the incorrect registration of staff members is corrected in the desired system.
-% \item Data in the existing system does not satisfy invariants of the desired system.
-%    There may be no way in the existing system of making the data satisfy those invariants.
-%    We can use the same example as in the previous bullet point,
-%    but add the requirement (in the desired system) that every purchase above a certain amount needs to be approved by the CEO.
-
-%    This issue will require somebody to change the data to satisfy the broken invariants in the desired system.
-% \item A way of entering data into the existing system is missing in the desired system.
-%    People or automated processes might rely on these ways of entering data.
-%    An example could be that when employees turned their computers on or off,
-%    an ad-hoc script would automatically check them in- and out to determine the number of hours they worked.
-%    A handful of employees still relies on this.
-
-%    This issue will require that users of obsolete functionality are informed and given a way to cope with the missing functionality where appropriate.
-% \item Data present in the existing system cannot be stored in the desired system.
-%    An example could be that references to physical locations where original receipts are kept are stored in the existing system,
-%    but the desired system relies on scans of receipts and allows the originals to be destroyed or not submitted.
-
-%    This issue will require that the existing data is kept until the original receipts have been scanned.
-% \end{enumerate}
-
-% (@Bas, het volgende zit al in de oplossingen sfeer. Willen we dat hier al doen?)
-%    To mitigate these issues, we:
-
-%    \begin{enumerate}
-%    \item Allow `missing' triples requirements to be ignored during migration.
-%    \item Allow triples to be migrated while being marked as needing correction.
-%    \item Allow invariants in the desired system to be ignored for certain triples during migration.
-%    \item Allow continued use of interfaces of the existing system, data entered into the existing system via those interfaces needs to be continuously copied to the desired system.
-%    \item Retain data in the existing system until it can be marked as ready to be phased out.
-%    \end{enumerate}
 
@@
    rather than $\concepts$, $\isa$, $\rels$, and $\rules$ respectively.
-   For every rule $u$ in the schema, there is a predicate $\sat{u}{\dataset}$ that says whether or not dataset $\dataset$ satisfies rule $u$.
+   For every rule $u$ in the schema, there is a predicate $\sat{u}{\dataset}$ that constrains dataset $\dataset$.
 %  Every schema is an element of an infinite set called $\Schema$.
@@
 %    that represents the set of violations%
-% \footnote{The definition of ``violation'' is not needed in this paper. We only use the absence of violations to show satisfaction of a rule.}
+% \footnote{The definition of ``violation'' is not needed in this paper. We only use the absence of violations to show satisfaction of a constraint.}
 %    of rule $u$ in any dataset.
@@
 % \end{eqnarray}
-%    To determine whether rule $u$ is satisfied in dataset $\dataset$,
+%    To determine whether rule $u$ satisfies dataset $\dataset$,
 %    we define $\sat{u}{\Dataset}$:
@@
 % \end{eqnarray}
-%    Violations are used to produce meaningful error messages and to trigger changes in the dataset to restore satisfaction of that rule.
+%    Violations are used to produce meaningful error messages and to trigger changes in the dataset to restore invariance of that rule.
 
@@
    The purpose of a role is to mention an individual user (human) or an automated actor (bot) without knowing who that user is.
-
-   The information system enforces the semantics by ensuring that the following requirements remain true at all times:
+   The expression $o\ \maintain\ u$ means that role $o$ is responsible for maintaining rule $u$, i.e. that rule $u$ is kept satisfied at all times.
+
+   The information system itself enforces the semantics by ensuring that the following requirements remain true at all times:
 \begin{eqnarray}
@@
    \label{eqn:define R}\\
-   \forall u\in\rules&:&\sat{u}{\dataset}
+   \forall u\in\rules&:&(\exists o\in\roles: o\ \maintain\ u)\ \vee\ \sat{u}{\dataset}
    \label{eqn:satisfaction}
@@
    Requirement~\ref{eqn:define R} ensures that the information system only works with triples whose relation is defined in the schema.
-   Requirement~\ref{eqn:satisfaction} ensures that the dataset satisfies all rules in the schema.
+   Requirement~\ref{eqn:satisfaction} ensures that the dataset satisfies all rules that are not maintained by a role.
 %    The following is for wide tables only
 %    Requirement~\ref{eqn:subpopulations} ensures that specialization works across relations.
-   Note that these requirements are enforced at runtime.
+   Note that these requirements apply at runtime.
+   So, as long as a rule $u$ ``lives'' in an information system, the dataset will satisfy that rule.
 
@@
 RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
-ROLE Administrator MAINTAINS EnrollRule
 \end{verbatim}
@@
 \end{equation}
-   Rule {\tt EnrollRule} is satisfied because relation $\declare{\tt isEnrolledFor}{\tt Student}{\tt Module}$ is empty.
+   As relation $\declare{\tt isEnrolledFor}{\tt Student}{\tt Module}$ is empty, rule {\tt EnrollRule} is satisfied in this example.
 
@@
    \concepts&=&\{ {\tt Module}, {\tt Course}, {\tt Student}\}\\
-   \rels&=&\{\begin{array}[t]{l}
+   \rels&=&\{\begin{array}[t]{@{}l@{}}
                \declare{\tt takes}{\tt Student}{\tt Course},\\
                \declare{\tt isPartOf}{\tt Module}{\tt Course},\\
-               \declare{\tt isEnrolledFor}{\tt Student}{\tt Module}\ \}
+               \declare{\tt isEnrolledFor}{\tt Student}{\tt Module}\}
              \end{array}
@@
    Now let us check the definition of information system.
+   Requirement~\ref{eqn:satisfaction} is satisfied because the only rule, {\tt EnrollRule} is satisfied.
+   The set of roles is empty and so is the relation $\maintain$.
+   This meets all requirements for $\infsys=\quadruple{\dataset}{\schema}{\roles}{\maintain}$ from definition~\ref{def:information system}.
+   This concludes the argument that $\infsys$ is an information system.
+
+   Now suppose that the script also contains the following line:
+\begin{verbatim}
+ROLE Administrator MAINTAINS EnrollRule
+\end{verbatim}
    Ampersand generates a set of roles $\roles=\{{\tt Administrator}\}$.
-   The relation $\maintain$ contains one pair only, $\pair{{\tt Administrator}}{{\tt EnrollRule}}$,
-   instructing the run-time engine to forward all violations of {\tt EnrollRule} to an administrator so she can keep {\tt EnrollRule} satisfied.
-   Requirement~\ref{eqn:satisfaction} is satisfied because the only rule, {\tt EnrollRule} is satisfied.
-   This completes the requirements for $\infsys=\quadruple{\dataset}{\schema}{\roles}{\maintain}$ from definition~\ref{def:information system}.
-   This concludes the argument that $\infsys$ is an information system.
+   The relation $\maintain$ contains one pair only, $\pair{{\tt Administrator}}{{\tt EnrollRule}}$.
+   Requirement~\ref{eqn:satisfaction} is still satisfied because now there is a role to maintain rule {\tt EnrollRule}.
+   The run-time engine will no longer enforce satisfaction of rule {\tt EnrollRule}, but leave it up to an administrator.
 
@@
 \label{sct:Changing data}
-   Every rule in the schema serves as invariant,
-   i.e.\ a constraint that the data in the database must satisfy at all times.
-   If an actor tries to change the dataset in a way that violates a rule,
-   the system will either reject that change or help the actor to find additional changes to keep that rule satisfied.
+   Every rule in the schema represents a constraint on the dataset.
+   The rule serves as invariant,
+   which means that the data in the database must satisfy the rule at all times.
+   If an actor tries to change the dataset in a way that violates this rule,
+   the system will either reject that change or help the actor to find additional changes to keep it satisfied.
+   The phrase ``restore an invariant'' or ``restore invariance of a rule''
+   means that an actor changes the dataset to satisfy the constraint that rule represents.
 
@@
    If such an event would violate requirement~\ref{eqn:satisfaction},
-   either the system stays in the state $\dataset$ or there is a reaction $e'$ that restores invariance of the affected rules:
+   either the system stays in the state $\dataset$ or there is a reaction $e'$ that restores the affected invariants:
 \begin{equation}
@@
 \end{equation}
-   In this way, the system can restore invariance of rules and keep requirement~\ref{eqn:satisfaction} satisfied.
+   In this way, the system can restore invariants and keep requirement~\ref{eqn:satisfaction} satisfied.
    The combination of events $e$ and $e'$ is called a {\em transaction}.
 
-   In practice, there are different ways of enforcing satisfaction of rules: automatic enforcement and manual enforcement.
+   In practice, there are different ways of enforcing rules: automatic enforcement and manual enforcement.
 
@@
    The system must have an engine that restores invariance of all enforce rules,
-   so users will not notice any violations.
+   so users will experience that these rules are always satisfied.
 
    Manual enforcement means that the system (temporarily) allows violations,
-   to let the user ``invent'' a reaction $e'$ that restores invariance.
-   For this purpose, Ampersand relaxes requirement~\ref{eqn:satisfaction} as follows:
-\begin{eqnarray}
-   \forall u\in\rules&:&(\exists o\in\roles:\ o\maintain u)\ \vee\ \sat{u}{\dataset}
-   \label{eqn:satisfaction relaxed}
-\end{eqnarray}
-   The purpose of this requirement is to allow the compiler to generate code for users with role $o$,
-   to restore invariance ``by hand''.
+   to let a user ``invent'' a reaction $e'$ that restores invariance.
+   That is why the term $(\exists o\in\roles:\ o\maintain u)$ is added to requirement~\ref{eqn:satisfaction}.
+   It allows the compiler to generate code for users with role $o$,
+   so they can restore the invariant manually.
    In the Ampersand language, a developer explicitly assigns a rule $u$ to a role $o$,
@@
          A developer defines a desired system $\infsys'$, which contains a dataset of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.} as illustrated in figure~\ref{fig:post-migration}.
-         After the migration, the system will contain the triples from $\dataset$ that befit $\schema'$,
+         After the migration, the system will contain the triples from $\dataset$ that satisfy $\schema'$,
          except if the migration engineer decides otherwise (see step~\ref{item3}).
@@
          for instance to initialize new features on the moment of transition (MoT).
-         If schemas $\schema$ equals $\schema'$, the data migration is trivial because it is not necessary.
+         If schemas $\schema$ equals $\schema'$, the data migration is unnecessary, so in the sequel we assume they are different.
    \item The generator generates a migration script, $\migrsys$,
@@
          For that purpose, the generator generates the migration script as an Ampersand script,
-         to let the migration engineer accommodate $\migrsys$ to the non-standard requirements.
-         The resulting migration script, $\migrsys'$, can be tested separately and can be used to take the desired system to production.
+         so the migration engineer can accommodate such requirements.
+         This can be necessary, for example to resolve data pollution that violates a constraint that was not in schema $\schema_{\infsys}$.
+         She can either implement enforce rules or use the business to resolve the issues.
+         She can test the resulting migration script, $\migrsys'$ separately before taking the desired system to production.
          The migration engineer does not alter the schemas of the existing system or the desired system.
          All migration functionality is concentrated in $\migrsys'$.
-   \item The migration engineer deploys system $\migrsys'$.
+   \item Then, the migration engineer deploys system $\migrsys'$.
          This event starts the copying of data from $\dataset$ to $\dataset'$ (figure~\ref{fig:migration phase}).
          During this phase, the ingress is still referring users to the existing system,
-         so they will not notice that the desired system is being fired up and loaded with data.
+         so they will hardly notice that the desired system is being fired up and loaded with data.
          Users are still changing $\dataset$ (both inserts and deletes),
@@
          In case of a go-decision, the migration engineer can take down the existing system, which by this time has become unreachable for users.
+         The resulting dataset must satisfy every constraint from $\rules_{\migrsys'}$ before the migration engineer takes down the migration system.
          In case of a no-go, the migration engineer restores the existing system by letting the ingress switch traffic back to $\infsys$.
-   \item In more complex cases, the migration script contains rules that must be satisfied during the migration.
-         These rules can be necessary, for example to resolve data pollution that violates a constraint that was not in schema $\schema_{\infsys}$.
-         By putting such rules in the schema $\schema_{\migrsys'}$,
-         the migration engineer can fix these problems.
-         She can do this in the form of an enforce rule, which means that the problem gets fixed automatically.
-         Or she can use the business to satisfy this rule, in case human intelligence is required.
-         Whatever the choice, every rule from $\rules_{\migrsys'}$ must be satisfied before the migration engineer takes down the migration system.
    \item After taking down the existing system and the migration system, the desired system remains in production and the migration is finished.
@@
 \end{eqnarray}
-   Every enforce rule is enforced by an engine,
-   which changes the population of relation $r$ to satisfy the rule.
+   An engine enforces all enforce rules by changing the population of relation $r$.
    An enforce rule with operator {\tt :<} (alternative~\ref{enforce del}) keeps the population of $r$ smaller than or equal to \id{expression}.
@@
 \end{eqnarray}
-   Understanding what the engine does to satisfy these rules might help the reader, so let us look what happens at runtime.
+   Understanding what the engine does to enforce these rules might help the reader, so let us look what happens at runtime.
    Initially, when {\tt copied\_takes} is still empty, the difference between {\tt old.takes} and {\tt copied\_takes} equals {\tt old.takes}.

### 2023-04-01 2022 Migration/articleMigration.tex
redone the def. of information system

@@
 
-   Specialization, also known as {\em generalization} or {\em subtyping}, is a relation between concepts.
-   The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
-\begin{equation}
-   \label{eqn:specialization}
-   A\isa B\ \Leftrightarrow\ \forall a: a\inst A\rightarrow a\inst B
-\end{equation}
-   Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.
-   In her script, a user can declare a specialization as a pair of concepts in the relation $\isa$.
-   A compiler can construct $\isa$ as the transitive, reflexive closure of all user-defined specializations.
-   It must make sure that $\isa$ is antisymmetric, so that $\isa$ is a partial order of concepts.
-   Specialization causes atoms to have multiple concepts of which they can be an instance.
-   As a result, if an atom is an instance of concept $A$ and $A\isa B$,
-   this atom has all properties that atoms of type $B$ have.
+%    Specialization, also known as {\em generalization} or {\em subtyping}, is a relation between concepts.
+%    The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
+% \begin{equation}
+%    \label{eqn:specialization}
+%    A\isa B\ \Leftrightarrow\ \forall a: a\inst A\rightarrow a\inst B
+% \end{equation}
+%    Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.
+%    In her script, a user can declare a specialization as a pair of concepts in the relation $\isa$.
+%    A compiler can construct $\isa$ as the transitive, reflexive closure of all user-defined specializations.
+%    It must make sure that $\isa$ is antisymmetric, so that $\isa$ is a partial order of concepts.
+%    Specialization causes atoms to have multiple concepts of which they can be an instance.
+%    As a result, if an atom is an instance of concept $A$ and $A\isa B$,
+%    this atom has all properties that atoms of type $B$ have.
 
@@
    is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.
-
+   For example, $\triple{\text{\atom{Peter}}}{\declare{\id{phone}}{\tt Person}{\tt TelephoneNumber}}{\text{\atom{074238991}}}$ is a triple.
+
+   Let $\Dataset$ be the set of all pairs $\pair{\triples}{\inst}$, so
+   $\Dataset$ is $\Pair{(\Triple{\Atoms}{\Rels}{\Atoms})}{(\Pair{\Atoms}{\Concepts})}$.
    A dataset $\dataset$ is a tuple $\pair{\triples}{\inst}$ that satisfies:
@@
 \end{eqnarray}
-   For example, $\triple{\text{\atom{Peter}}}{\declare{\id{phone}}{\tt Person}{\tt TelephoneNumber}}{\text{\atom{074238991}}}$ is a triple.
-   Equation~\ref{eqn:wellTypedEdge} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
+   Looking at the example,
+   equation~\ref{eqn:wellTypedEdge} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
    In practice, users can say that Peter has telephone number 074238991.
@@
    The notations $\triples_{\dataset}$ and $\inst_{\dataset}$ are used to disambiguate $\triples$ and $\inst$ when necessary.
-   Every dataset is an element of an infinite set called $\Dataset$,
-   which is disjoint from $\Rels$, $\Concepts$ and $\Atoms$.
    To save writing in the sequel, the notation $a\ r\ b$ means that $\triple{a}{r}{b}\in\triples$.
@@
 
-   A schema $\schema$ is a tuple $\quadruple{\concepts}{\isa}{\rels}{\rules}$,
+   A schema $\schema$ is a tuple $\triple{\concepts}{\rels}{\rules}$,
    in which $\concepts$ is a finite set of concepts,
-   $\isa$ is a partial order of concepts,
    $\rels$ is a finite set of relations,
@@
    The notation $\schema_{\infsys}$ refers to the schema of information system $\infsys$.
-   Disambiguation sometimes requires to write $\concepts_{\schema}$, $\isa_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$
-   rather than $\concepts$, $\isa$, $\rels$, and $\rules$ respectively.
+   Disambiguation sometimes requires to write $\concepts_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$
+   rather than $\concepts$, $\rels$, and $\rules$ respectively.
    For every rule $u$ in the schema, there is a predicate $\sat{u}{\dataset}$ that constrains dataset $\dataset$.
@@
    A schema must ensure that triples ``make sense'' in the semantics defined by the schema.
-   This means that it must satisfy the following requirements.
+   This means that it must satisfy:
 \begin{eqnarray}
    \declare{n}{A}{B}\in\rels&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
-   \label{eqn:relationsIntroduceConcepts}\\
-   A\ \isa\ B&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
-   \label{eqn:isasIntroduceConcepts}\\
+   \label{eqn:relationsIntroduceConcepts}
+%    A\ \isa\ B&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
+%    \label{eqn:isasIntroduceConcepts}\\
 %    The following is for wide tables only
@@
 %    \label{eqn:subrelations}\\
-   \isa\ \text{is a partial order}
-   \label{eqn:isasPartialOrder}\\
-   \exists\dataset\in\Dataset-\emptyset:\ \sat{u}{\dataset}
-   \label{eqn:consistentRules}
+%    \isa\ \text{is a partial order}
+%    \label{eqn:isasPartialOrder}\\
+%    \exists\dataset\in\Dataset-\emptyset:\ \sat{u}{\dataset}
+%    \label{eqn:consistentRules}
 \end{eqnarray}
    Requirement~\ref{eqn:relationsIntroduceConcepts} ensures that concepts mentioned in a relation are defined in the schema.
-   Requirement~\ref{eqn:isasIntroduceConcepts} ensures that the specialization relation works on concepts of $\concepts$.
+%    Requirement~\ref{eqn:isasIntroduceConcepts} ensures that the specialization relation works on concepts of $\concepts$.
 %    The following is for wide tables only
 %    Requirement~\ref{eqn:subrelations} ensures that all specializations of a relation exist.
-   Requirement~\ref{eqn:isasPartialOrder} ensures that the specialization relation contains no cycles.
-   Requirement~\ref{eqn:consistentRules} ensures that every rule is consistent.
-
-   The type system of Ampersand~\cite{vdWoude2011} ensures that every information system it generates satisfies these requirements.
+%    Requirement~\ref{eqn:isasPartialOrder} ensures that the specialization relation contains no cycles.
+%    Requirement~\ref{eqn:consistentRules} ensures that every rule is consistent.
+
+   The type system of Ampersand~\cite{vdWoude2011} ensures that every information system it generates satisfies this requirement.
    Note that these requirements do not depend on any particular dataset,
@@
 
+   % The maintainance relation between roles and rules is used to determine who is allowed to change the dataset.
+   % For every rule $u$ in the schema, there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
+   % that represents the set of violations.%
+
+   % In practice, there are different ways of enforcing rules: automatic enforcement and manual enforcement.
+
+   % Automatic enforcement is specified by the developer with a special syntax:
+   % the {\em enforce rule}.
+   % An enforce rule specifies not only the rule,
+   % but it implicitly defines event $e'$ to restore the invariance.
+   % The system must have an engine that restores invariance of all enforce rules,
+   % so users will experience that these rules are always satisfied.
+
+   % Manual enforcement means that the system (temporarily) allows violations,
+   % to let a user ``invent'' a reaction $e'$ that restores invariance.
+   % That is why the term $(\exists o\in\roles:\ o\maintain u)$ is added to requirement~\ref{eqn:satisfaction}.
+   % It allows the compiler to generate code for users with role $o$,
+   % so they can restore the invariant manually.
+   % In the Ampersand language, a developer explicitly assigns a rule $u$ to a role $o$,
+   % to keep control over the type of persons that are allowed to ``restore broken rules''.
+   % As a consequence, a mechanism is needed to notify those persons of the work to be done.
+
+   % If a rule $u$ is not assigned to any role and it is not an enforce rule,
+   % the system must prevent any event $e$ from having any effect on the database.
+   % So, if $e$ would violate rule $u$, the system must reject $e$.
+   % This implements blocking behaviour, and must be accompanied by an error message.
+
 \subsection{Information Systems}
 \label{sct:Information Systems}
-   This section defines the notion of information system.
-   As before, a suffix disambiguates the elements of this definition when needed.
+   An information system has a changing dataset because of events that occur.
+   When the system ``observes'' an event, it attempts to change its dataset to reflect that event.
+   However, every rule in the schema represents a constraint,
+   which the dataset must invariably satisfy.
+   That is why rules are also known as ``invariants''.
+
+   Events are taken from an infinite set, $\Events$.
+   An event is denoted as $\evt{d^-}{d^+}$, in which $d^-$ and $d^+$ are datasets.
+   Every event is a function $\Dataset\rightarrow\Dataset$, so $\evt{d^-}{d^+}(\dataset)$ is a dataset.
+   Equation~\ref{eqn:event} defines the effect of an event on a dataset.
+\begin{eqnarray}
+      \evt{d^-}{d^+}(\dataset)&=&\la(\triples_{\dataset}-\triples_{d^-})\cup\triples_{d^+},(\inst_{\dataset}-\inst_{d^-})\cup\inst_{d^+}\ra
+\label{eqn:event}
+\end{eqnarray}
+   In the sequel, variables $e$, $e'$, $e''$, etc. represent events.
+   As soon as a system $\infsys$ observes an event $e$,
+   it applies $e$ to its dataset $\dataset$.
+   Yet, if $e(\dataset)$ does not satisfy a rule $u$ from schema $\schema_\infsys$,
+   another event, $e'$, is needed so that $\sat{u}{e'(e(\dataset))}$.
+   The phrase ``restore an invariant'' or ``restore invariance of a rule''
+   means that the system will change $e(\dataset)$ by applying another event $e'$ to satisfy requirement~\ref{eqn:enforce}.
+   The notation $\enf_{u,e}$ represents this combination of events $e$ and $e'$.
+   So, $\enf_{u,e}$ must satisfy:
+\begin{eqnarray}
+   \sat{u}{\dataset}&\Rightarrow&\sat{u}{\enforce{u,e}{\dataset}}
+\label{eqn:enforce}
+\end{eqnarray}
+
+   Let us now define the notion of an information system.
 \begin{definition}[information system]
 \label{def:information system}
-\item An information system $\infsys$ is a tuple $\quadruple{\dataset}{\schema}{\roles}{\maintain}$, in which
+\item An information system $\infsys$ is a tuple $\triple{\dataset}{\schema}{\enf}$, in which
 \begin{itemize}
    \item dataset $\dataset=\pair{\triples}{\inst}$ is defined as in section~\ref{sct:Datasets};
-   \item schema $\schema=\quadruple{\concepts}{\isa}{\rels}{\rules}$ is defined as in section~\ref{sct:Schemas};
-   \item $\roles$ is a set of roles;
-   \item ${\maintain} : \roles\times\rules$ is the maintainance relation between roles and rules.
+   \item schema $\schema=\triple{\concepts}{\rels}{\rules}$ is defined as in section~\ref{sct:Schemas};
+   \item enforce function ${\enf}:\Rules\rightarrow\Events\rightarrow\Dataset$ satisfies requirement~\ref{eqn:enforce}.
 \end{itemize}
 \end{definition}
-   A \define{role} is a name that identifies a group of users.
-   It serves as a placeholder for a person or a machine (i.e. an actor) that works with the dataset (i.e. create, read, update, or delete triples).
-   The purpose of a role is to mention an individual user (human) or an automated actor (bot) without knowing who that user is.
-   The expression $o\ \maintain\ u$ means that role $o$ is responsible for maintaining rule $u$, i.e. that rule $u$ is kept satisfied at all times.
+   % A \define{role} is a name that identifies a group of users.
+   % It serves as a placeholder for a person or a machine (i.e. an actor) that works with the dataset (i.e. create, read, update, or delete triples).
+   % The purpose of a role is to mention an individual user (human) or an automated actor (bot) without knowing who that user is.
 
@@
    \label{eqn:define R}\\
-   \forall u\in\rules&:&(\exists o\in\roles: o\ \maintain\ u)\ \vee\ \sat{u}{\dataset}
+   \forall u\in\rules&:&\sat{u}{\dataset}
    \label{eqn:satisfaction}
+% TODO:
+% De engine heeft een eigen rol. Alle regels die aan de engine zijn gekoppeld, worden automatisch onderhouden.
 %    The following is for wide tables only
@@
    Requirement~\ref{eqn:define R} ensures that the information system only works with triples whose relation is defined in the schema.
-   Requirement~\ref{eqn:satisfaction} ensures that the dataset satisfies all rules that are not maintained by a role.
+   Requirement~\ref{eqn:satisfaction} ensures that the dataset satisfies all rules.
 %    The following is for wide tables only
@@
    So, as long as a rule $u$ ``lives'' in an information system, the dataset will satisfy that rule.
+
+   As the system registers events, it inserts or deletes triples in its dataset.
+   If such an event would violate requirement~\ref{eqn:satisfaction},
+   either the system stays in the state $\dataset$ or there is a reaction $e'$ that restores the affected invariants:
+\begin{equation}
+   \sat{u}{\dataset}\Rightarrow\sat{u}{e'_{d'}(e_{d}(\dataset))}
+   \label{eqn:restoring}
+\end{equation}
+   In this way, the system can restore invariants and keep requirement~\ref{eqn:satisfaction} satisfied.
+   The combination of events $e$ and $e'$ is called a {\em transaction}.
 
@@
    Now let us check the requirements to verify that this example defines an information system.
-   Requirement~\ref{eqn:specialization} is satisfied because this example contains no specialization.
+   % Requirement~\ref{eqn:specialization} is satisfied because this example contains no specialization.
    The Ampersand compiler generates a dataset $\dataset$, which contains a set of triples and a relation $\inst$.
@@
 \]
-   The user has defined no specializations.
-   This means that $\isa$ equals the identity relation on $\concepts$ and requirements~\ref{eqn:isasIntroduceConcepts} and~\ref{eqn:isasPartialOrder} are satisfied.
-   The script defines the set of rules $\rules$ to contain just one rule: \verb-EnrollRule-.
-   The (nonempty) dataset satisfies this rule, which meets requirement~\ref{eqn:consistentRules}.
-   So, the schema $\schema=\quadruple{\concepts}{\isa}{\rels}{\rules}$ satisfies all requirements from section~\ref{sct:Schemas}.
+   % The user has defined no specializations.
+   % This means that $\isa$ equals the identity relation on $\concepts$ and requirements~\ref{eqn:isasIntroduceConcepts} and~\ref{eqn:isasPartialOrder} are satisfied.
+   % The script defines the set of rules $\rules$ to contain just one rule: \verb-EnrollRule-.
+   % The (nonempty) dataset satisfies this rule, which meets requirement~\ref{eqn:consistentRules}.
+   So, the schema $\schema=\triple{\concepts}{\rels}{\rules}$ satisfies all requirements from section~\ref{sct:Schemas}.
 
@@
    The set of roles is empty and so is the relation $\maintain$.
-   This meets all requirements for $\infsys=\quadruple{\dataset}{\schema}{\roles}{\maintain}$ from definition~\ref{def:information system}.
+   This meets all requirements for $\infsys=\triple{\dataset}{\schema}{\roles}{\maintain}$ from definition~\ref{def:information system}.
    This concludes the argument that $\infsys$ is an information system.
@@
    The run-time engine will no longer enforce satisfaction of rule {\tt EnrollRule}, but leave it up to an administrator.
-
-\subsection{Changing data}
-\label{sct:Changing data}
-   Every rule in the schema represents a constraint on the dataset.
-   The rule serves as invariant,
-   which means that the data in the database must satisfy the rule at all times.
-   If an actor tries to change the dataset in a way that violates this rule,
-   the system will either reject that change or help the actor to find additional changes to keep it satisfied.
-   The phrase ``restore an invariant'' or ``restore invariance of a rule''
-   means that an actor changes the dataset to satisfy the constraint that rule represents.
-
-   % The maintainance relation between roles and rules is used to determine who is allowed to change the dataset.
-   % For every rule $u$ in the schema, there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
-   % that represents the set of violations.%
-
-   Events can cause a dataset to change.
-   So, we model events as functions $\Dataset\rightarrow\Dataset$.
-   We will use variables $e$, $e'$, $e''$, and so on to represent events.
-   We distinguish just two types of event: insert and delete events.
-   An event is a label (\id{ins} or \id{del}) with a dataset that is to be inserted or deleted.
-   If $\triples$ is the set of triples inserted or deleted by an event $e$,
-   we denote that event either by $\ins{\triples}$ or $\del{\triples}$.
-   Equation~\ref{eqn:event} defines the effect of an event on a dataset.
-\begin{equation}
-   \begin{array}{rcl}
-      e(\dataset)&=&\text{case}\ e\ \text{of}\\
-      &&\begin{array}[t]{@{}rcl}
-         \ins{\triples}&=&\pair{\triples_{\dataset}\cup\triples}{\id{insInst}_{\triples}(\inst_{\dataset})}\\
-         \del{\triples}&=&\pair{\triples_{\dataset}-\triples}{\id{delInst}_{\triples}(\inst_{\dataset})}
-      \end{array}
-   \end{array}
-   \label{eqn:event}
-\end{equation}
-   in which $\id{insInst}_{\triples}$ transforms $\inst_{\dataset}$ such that $\ins{\triples}(\dataset)$ complies with
-   equations~\ref{eqn:specialization} and~\ref{eqn:wellTypedEdge},
-   and likewise for $\id{delInst}_{\triples}$.
-
-   As the system registers events, it inserts or deletes triples in its dataset.
-   If such an event would violate requirement~\ref{eqn:satisfaction},
-   either the system stays in the state $\dataset$ or there is a reaction $e'$ that restores the affected invariants:
-\begin{equation}
-   \sat{u}{\dataset}\Rightarrow\sat{u}{e'(e(\dataset))}
-   \label{eqn:restoring}
-\end{equation}
-   In this way, the system can restore invariants and keep requirement~\ref{eqn:satisfaction} satisfied.
-   The combination of events $e$ and $e'$ is called a {\em transaction}.
-
-   In practice, there are different ways of enforcing rules: automatic enforcement and manual enforcement.
-
-   Automatic enforcement is specified by the developer with a special syntax:
-   the {\em enforce rule}.
-   An enforce rule specifies not only the rule,
-   but it implicitly defines event $e'$ to restore the invariance.
-   The system must have an engine that restores invariance of all enforce rules,
-   so users will experience that these rules are always satisfied.
-
-   Manual enforcement means that the system (temporarily) allows violations,
-   to let a user ``invent'' a reaction $e'$ that restores invariance.
-   That is why the term $(\exists o\in\roles:\ o\maintain u)$ is added to requirement~\ref{eqn:satisfaction}.
-   It allows the compiler to generate code for users with role $o$,
-   so they can restore the invariant manually.
-   In the Ampersand language, a developer explicitly assigns a rule $u$ to a role $o$,
-   to keep control over the type of persons that are allowed to ``restore broken rules''.
-   As a consequence, a mechanism is needed to notify those persons of the work to be done.
-
-   If a rule $u$ is not assigned to any role and it is not an enforce rule,
-   the system must prevent any event $e$ from having any effect on the database.
-   So, if $e$ would violate rule $u$, the system must reject $e$.
-   This implements blocking behaviour, and must be accompanied by an error message.
 
@@
    Let us define the instruments needed to describe the derivation of $\schema_{\migrsys}$.
+\begin{definition}[union of datasets]
+\label{def:graph_union_wellTyped_if_parts_wellTyped}
+    \[\pair{\triples_1}{\inst_1}\cup\pair{\triples_2}{\inst_2}\ =\ \pair{\triples_1\cup\triples_2}{\inst_1\cup\inst_2}\]
+\end{definition}
+
+\begin{lemma}
+\label{lemma:graph_union_wellTyped_if_parts_wellTyped}
+   If $\dataset_1$ and $\dataset_2$ are datasets, then $\dataset_1\cup\dataset_2$ is a dataset.
+\end{lemma}
+
+\begin{definition}[map relation names in datasets]
+\label{def:map_labels_preserves_wellTypedness}
+   Let $h:\Rels\rightarrow\Rels$, then
+   \[\begin{array}{rcl}
+      \maprel{h}{\pair{\triples}{\inst}}&=&\pair{\maprel{h}{\triples}}{\inst}\\
+      \maprel{h}{\triple{a}{r}{b}}&=&\triple{a}{h(r)}{b}
+   \end{array}\]
+\end{definition}
+
+\begin{lemma}
+\label{lemma:map_labels_preserves_wellTypedness}
+   If $\dataset$ is a dataset and $h:\Rels\rightarrow\Rels$ is an injective function, then $\maprel{h}{\dataset}$ is a dataset.
+\end{lemma}
+
 \begin{definition}[disjoint union of datasets]
@@
       X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
-      X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}\\
+      X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}
 \end{eqnarray}
@@
 \begin{eqnarray}
-   \quadruple{\concepts}{\isa}{\rels}{\rules}\sqcup\quadruple{\concepts'}{\isa'}{\rels'}{\rules'}&=&\quadruple{\concepts\uplus\concepts}{\isa\uplus\isa'}{\rels\uplus\rels'}{\rules\uplus\rules'}
+   \triple{\concepts}{\rels}{\rules}\sqcup\triple{\concepts'}{\rels'}{\rules'}&=&\triple{\concepts\uplus\concepts}{\rels\uplus\rels'}{\rules\uplus\rules'}
 \end{eqnarray}
@@
    \begin{eqnarray}
-      \rlap{$\subst{C}{D}{\quadruple{\concepts}{\isa}{\rels}{\rules}}$}\notag\\
+      \rlap{$\subst{C}{D}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
       &=&\begin{array}[t]{l@{}l}
@@
    \begin{eqnarray}
-      \rlap{$\subst{r'}{m'}{\quadruple{\concepts}{\isa}{\rels}{\rules}}$}\notag\\
-      &=&\quadruple{\concepts}{\isa}{\{\subst{r'}{m'}{r} \mid r\in\rels\}}{\{\subst{r'}{m'}{u} \mid u\in\rules\}}\notag
+      \rlap{$\subst{r'}{m'}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
+      &=&\triple{\concepts}{\{\subst{r'}{m'}{r} \mid r\in\rels\}}{\{\subst{r'}{m'}{u} \mid u\in\rules\}}\notag
    \end{eqnarray}
@@
    \begin{eqnarray}
-      \infsys\sqcup\infsys'&=&\quadruple{\roles'}{\rules'}{\maintain'}{\dataset\sqcup\dataset'}
+      \infsys\sqcup\infsys'&=&\triple{\roles'}{\rules'}{\maintain'}{\dataset\sqcup\dataset'}
    \end{eqnarray}

### 2023-04-02 2022 Migration/articleMigration.tex
wip

@@
 \begin{abstract}
-   To support incremental software development with a software generator bears the potential of saving work and decreasing the time to deploy changes.
-   However, increments that change the database schema cause nontrivial data migration issues.
-   The problem is how to preserve the semantics of the data as much as possible and satisfy elementary business requirements at the same time.
-
-   To generate an information system in a non-incremental manner is already
-   provided for by the Ampersand project.
-   This paper is about changing an already generated system in production.
-   It develops a theory for deploying an incremental change,
-   making Ampersand more useful for agile software development.
-   This paper proposes a migration method that features zero down-time,
-   and features an undo after deployment.
-   It also leaves room for the business to finish ongoing work
-   after the moment the change has been deployed.
+   Incremental software development with a software generator can save work and reduce the time to deploy changes,
+   but data migrations that change the database schema can be challenging.
+   The problem is how to preserve the semantics of the data under a changing schema.
+
+   This paper proposes a theory for deploying an incremental change in an already generated system in production using the Ampersand project.
+   Our approach aims to preserve data semantics and satisfy elementary business requirements
+   while featuring zero down-time and an undo option after deployment.
+   We derive a migration schema from both the existing system and a changed schema that specifies the desired system.
+   This work is based on the assumptions that
+   the migration is meant to deploy a software increment in production,
+   the existing data set may be polluted,
+   human interaction may be required,
+   data meaning must be preserved,
+   the business continues during the migration without interruption,
+   and there is a compiler to generate an information system from a given schema.
+   The correctness of the migration is the focus of this paper, while efficiency is beyond its scope.
 \end{abstract}
@@
 
-   This work uses the Ampersand compiler~\cite{Joosten-JLAMP2018} to validate the theory.
-   Ampersand is a tool to generate information systems.
-   It works with relations in a way similar to Alloy~\cite{Alloy2006} and allegories%
+   This work uses the Ampersand compiler~\cite{Joosten-JLAMP2018, JoostenRAMiCS2017} to validate the theory.
+   Ampersand is a language to generate information systems.
+   It uses a syntactically sugared form of heterogeneous relation algebra,
+   which works with relations in a way similar to Alloy~\cite{Alloy2006} and allegories%
 \footnote{Allegories are a specific type of categories, which the reader does not need to understand for the purpose of this paper}~\cite{Zielinski2013}.
@@
    An information system has a changing dataset because of events that occur.
-   When the system ``observes'' an event, it attempts to change its dataset to reflect that event.
+   Whenever the system ``observes'' an event, it attempts to change its dataset accordingly.
    However, every rule in the schema represents a constraint,
@@
    Yet, if $e(\dataset)$ does not satisfy a rule $u$ from schema $\schema_\infsys$,
-   another event, $e'$, is needed so that $\sat{u}{e'(e(\dataset))}$.
-   The phrase ``restore an invariant'' or ``restore invariance of a rule''
+   another event is needed, say $e'$, to ensure that $\sat{u}{e'(e(\dataset))}$.
+   The phrase ``to restore an invariant'' or ``to restore invariance of a rule''
    means that the system will change $e(\dataset)$ by applying another event $e'$ to satisfy requirement~\ref{eqn:enforce}.
@@
 \end{verbatim}
-   The third and last relation states which students are enrolled for which module.
+   The third relation states which students are enrolled for which module.
    It is left without population for now.
@@
    It states that a student can enroll for any module that is part of a course he or she takes.
-   In Ampersand, which is a syntactically sugared form of relation algebra~\cite{JoostenRAMiCS2017},
-   each rule has a name and each rule has a role to maintain its invariance:
+   % In Ampersand, which is a syntactically sugared form of relation algebra~\cite{JoostenRAMiCS2017},
+   % each rule has a name and each rule has a role to maintain its invariance:
 \begin{verbatim}
@@
 \end{verbatim}
-   The semantics of this rule defines $\sat{\tt EnrollRule}{\dataset}$ as:
+   The Ampersand compiler defines $\sat{\tt EnrollRule}{\dataset}$ as:
 \begin{equation}
@@
    Requirement~\ref{eqn:satisfaction} is satisfied because the only rule, {\tt EnrollRule} is satisfied.
-   The set of roles is empty and so is the relation $\maintain$.
-   This meets all requirements for $\infsys=\triple{\dataset}{\schema}{\roles}{\maintain}$ from definition~\ref{def:information system}.
-   This concludes the argument that $\infsys$ is an information system.
-
-   Now suppose that the script also contains the following line:
-\begin{verbatim}
-ROLE Administrator MAINTAINS EnrollRule
-\end{verbatim}
-   Ampersand generates a set of roles $\roles=\{{\tt Administrator}\}$.
-   The relation $\maintain$ contains one pair only, $\pair{{\tt Administrator}}{{\tt EnrollRule}}$.
-   Requirement~\ref{eqn:satisfaction} is still satisfied because now there is a role to maintain rule {\tt EnrollRule}.
-   The run-time engine will no longer enforce satisfaction of rule {\tt EnrollRule}, but leave it up to an administrator.
+   This satisfies definition~\ref{def:information system}, which establishes that $\infsys$ is an information system.
+   % The set of roles is empty and so is the relation $\maintain$.
+   % This meets all requirements for $\infsys=\triple{\dataset}{\schema}{\roles}{\maintain}$ from definition~\ref{def:information system}.
+   % This concludes the argument that $\infsys$ is an information system.
+
+%    Now suppose that the script also contains the following line:
+% \begin{verbatim}
+% ROLE Administrator MAINTAINS EnrollRule
+% \end{verbatim}
+%    Ampersand generates a set of roles $\roles=\{{\tt Administrator}\}$.
+%    The relation $\maintain$ contains one pair only, $\pair{{\tt Administrator}}{{\tt EnrollRule}}$.
+%    Requirement~\ref{eqn:satisfaction} is still satisfied because now there is a role to maintain rule {\tt EnrollRule}.
+%    The run-time engine will no longer enforce satisfaction of rule {\tt EnrollRule}, but leave it up to an administrator.
 
@@
          After the migration, the system will contain the triples from $\dataset$ that satisfy $\schema'$,
-         except if the migration engineer decides otherwise (see step~\ref{item3}).
+         with additional changes made by the migration engineer (see step~\ref{item3}).
          $\dataset'$ may contain some new data,
@@
    RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
-   ROLE Administrator MAINTAINS EnrollRule
 
@@
       ]
-   ROLE Student MAINTAINS ExamRule1, ExamRule2
    RULE ExamRule1: student~;course |- takes

### 2023-04-30 2022 Migration/articleMigration.tex
Results of April 30th, 2023.

@@
 \begin{abstract}
-   Incremental software development with a software generator can save work and reduce the time to deploy changes,
-   but data migrations that change the database schema can be challenging.
-   The problem is how to preserve the semantics of the data under a changing schema.
-
-   This paper proposes a theory for deploying an incremental change in an already generated system in production using the Ampersand project.
-   Our approach aims to preserve data semantics and satisfy elementary business requirements
-   while featuring zero down-time and an undo option after deployment.
+   Incremental software deployment is meant to reduce the time to deploy changes.
+   However, an increment might involve a change in the database schema.
+   This poses the challenge to preserve already deployed data.
+   The problem is how to preserve the semantics of that data under a changing schema.
+   This paper proposes a theory for deploying an incremental change,
+   meant to be implemented in a software generator called Ampersand.
+
+   This paper aims to preserve the semantics of data, satisfy elementary business requirements,
+   and featuring a zero down-time option after deployment.
    We derive a migration schema from both the existing system and a changed schema that specifies the desired system.
@@
 \begin{keyword}
-generative software\sep incremental software development\sep data migration\sep relation algebra\sep Ampersand\sep schema change
+generative software\sep incremental software deployment\sep data migration\sep relation algebra\sep Ampersand\sep schema change
 \end{keyword}
@@
    change failure rate~\cite{DevOps2021}.}.
-   By installing upgrades automatically, frequently, in small increments, and without down-time,
-   the users experience a system that evolves organically, without even noticing the updates.
-   By automating data migrations correctly and reliably, migration engineers save effort and errors.
-   As a result, they can release smaller increments more frequently, which contributes to a more agile software development process.
-   Smaller and faster releases also mean a smaller difference between the existing system and the desired system.
-   This yields even simpler migrations, leveraging even more frequent releases.
-
-   Differences of schemas of an existing system and a desired system complicate the migration of data.
-   The purpose of incremental deployment poses specific requirements to data migration,
-   which sets it apart from data migrations for other purposes.
+   By deploying frequently, in small increments, and without down-time,
+   users get a system that evolves organically.
+   This requires a robust data migration process in which the deployment succeeds without failure.
+   Automation is required too, to prevent human errors in the process.
+   As a result, we expect that migration engineers can deploy smaller increments more frequently, which contributes to a more agile software development process.
+
+   This paper assumes the the existing system and the desired system have different schemas,
+   which occurs in incremental software deployment.
+   The existing system is the system that is already deployed in production,
+   which makes it necessary to preserve the existing data.
+
+   Data migration for other purposes has been described in the literature.
    For instance, if a data migration is done for switching to another platform or to different technology,
@@
    migration engineers may deliberately avoid schema differences and functionality changes to avoid introducing new errors in an otherwise error-prone migration process.
-   For the purpose of software updates, however, introducing new functionality, removing obsolete functionality, and schema differences are the very reason for deploying an increment.
-   Data migration literature sometimes uses a different notion of increment.
    For example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014} define an increment as a variation between two data structures.
    They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
-   In a software update, however, the variations are not preserved.
-   So this notion of increment in\cite{Ataei2021} is obviously different.
+   Obviously, such data migrations are not suitable for incremental software deployment,
+   if the schema changes and the old functionality has to be replaced.
    Examples like these have convinced the authors to define a new method for data migration that is specifically meant for systems in production that are
@@
    The contribution of this paper is to derive a migration schema from two artifacts: an existing system
-   (which has its own schema and dataset) and a new schema (which specifies the desired system).
+   (which has its own schema and data set) and a new schema (which specifies the desired system).
    The purpose of our theory is to develop tools.
@@
 
-   Every system contains a dataset, which represents the state of the system.
+   Every system contains a data set, which represents the state of the system.
    Every service produces and consumes events that may change the state of the system.
@@
    \begin{center}
-     \includegraphics[scale=.45]{datamigration-Pre-migration.drawio.png}
+     \includegraphics[scale=.45]{figures/datamigration-Pre-migration.drawio.png}
    \end{center}
@@
 \end{figure}
-   To keep our theory technology independent, datasets are assumed to contains triples.
+   To keep our theory technology independent, data sets are assumed to contains triples.
    This makes our theory valid for any kind of database that triples can represent,
@@
    which most database management systems refer to as integrity rules.
-   The purpose of integrity rules is to keep the constraints they represent satisfied.
-
-   This work uses the Ampersand compiler~\cite{Joosten-JLAMP2018, JoostenRAMiCS2017} to validate the theory.
-   Ampersand is a language to generate information systems.
+   The purpose of these rules is to keep all constraints satisfied.
+
+   This work uses the Ampersand compiler~\cite{Joosten-JLAMP2018, JoostenRAMiCS2017} to generate information systems.
    It uses a syntactically sugared form of heterogeneous relation algebra,
@@
 \footnote{Allegories are a specific type of categories, which the reader does not need to understand for the purpose of this paper}~\cite{Zielinski2013}.
-   Each rule in Ampersand is a constraint (aka invariant), which are kept satisfied as long as that rule lives%
+   Each rule in Ampersand is a constraint (aka invariant), which the system keeps satisfied as long as that rule lives%
 \footnote{The dotted lines in the diagrams show which set of rules is enforced.}.
@@
    because all constraints are explicitly available as rules in the code.
-   This, and the absence of imperative code in an Ampersand script, makes Ampersand a suitable platform to validate the theory.
+   This, and the absence of imperative code in an Ampersand script, makes Ampersand a suitable platform on which to implement the theory.
    It also allows us to be explicit about ``preserving the meaning as much as possible''.
@@
    \item data pollution that violates a constraint that is not in the schema.
-   \item data pollution that cannot be captured in constraints on the dataset,
+   \item data pollution that cannot be captured in constraints on the data set,
    such as a street address that has become obsolete because someone has moved without notification.
@@
    Such violations do not occur in our theory because the existing system was generated by Ampersand,
-   so the dataset satisfies all constraints in the schema.
+   so the data set satisfies all constraints in the schema.
    This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
@@
    \item the moment of transition (the MoT), i.e. the moment the desired system is made available to users;
-   \item the moment of no return (the MoN), i.e. the moment that the migration cannot be undone other than by a backup-procedure%
+   \item the moment of cleanup (the MoC), i.e. the moment that the old software and data are removed.%
 \footnote{Backup mechanisms are outside the scope of this paper.}.
@@
 
-   During the time between the MoT and MoN,
+   During the time between the MoT and MoC,
    the existing system and the desired system are kept alive, side by side, as shown in figure~\ref{fig:migration phase}.
@@
    \begin{center}
-     \includegraphics[scale=.35]{datamigration-Migration phase.drawio.png}
+     \includegraphics[scale=.35]{figures/datamigration-Migration phase.drawio.png}
    \end{center}
@@
    The rules of the migration schema will cause the right data to be copied correctly into the desired system.
-   This migration schema preserves data from the existing system, it may introduce new data automatically (generated from the existing dataset),
+   This migration schema preserves data from the existing system, it may introduce new data automatically (generated from the existing data set),
    and it may require users to introduce new data (which will take time). 
@@
    only after all migration rules are at rest.
-   At the MoN, the migration engineer takes down everything but the desired system,
+   At the MoC, the migration engineer takes down everything but the desired system,
    leaving the business with a successful migration.
@@
    \begin{center}
-     \includegraphics[scale=.35]{datamigration-Post-migration.drawio.png}
+     \includegraphics[scale=.35]{figures/datamigration-Post-migration.drawio.png}
    \end{center}
@@
 \label{sct:Terminology}
-   An {\em information system} is a combination of dataset, schema, and functionality.
+   An {\em information system} is a combination of data set, schema, and functionality.
    For the purpose of this paper, the functionality is ignored because it does not impact the migration.
-   Section~\ref{sct:Datasets} desribes datasets. Schemas are treated in section~\ref{sct:Schemas}.
+   Section~\ref{sct:Data sets} desribes data sets. Schemas are treated in section~\ref{sct:Schemas}.
    Then section~\ref{sct:Information Systems} defines information systems.
 
-\subsection{Datasets}
-\label{sct:Datasets}
-   A dataset $\dataset$ describes a set of structured data, which is typically stored persistently in a database of some kind.
-   The notation $\dataset_{\infsys}$ refers to the dataset of a particular information system $\infsys$.
-   The purpose of a dataset is to describe the data of a system at one point in time. 
-   Before defining datasets, let us first define the constituent notions of atom, concept, specialization, relation, and triple.
+\subsection{Data sets}
+\label{sct:Data sets}
+   A data set $\dataset$ describes a set of structured data, which is typically stored persistently in a database of some kind.
+   The notation $\dataset_{\infsys}$ refers to the data set of a particular information system $\infsys$.
+   The purpose of a data set is to describe the data of a system at one point in time. 
+   Before defining data sets, let us first define the constituent notions of atom, concept, specialization, relation, and triple.
 
@@
    The expression $a\inst C$ means that atom $a$ is an \emph{instance} of concept $C$.
-   This relation is used in the type system, in which $\inst$ assigns one or more concepts to every atom in the dataset.
+   This relation is used in the type system, in which $\inst$ assigns one or more concepts to every atom in the data set.
+   Since $\inst$ is a relation and every relation is a set of pairs,
+   set operators $\cup$, $\cap$, and $-$ can be used on $\inst$.
 
@@
    $\Dataset$ is $\Pair{(\Triple{\Atoms}{\Rels}{\Atoms})}{(\Pair{\Atoms}{\Concepts})}$.
-   A dataset $\dataset$ is a tuple $\pair{\triples}{\inst}$ that satisfies:
+   A data set $\dataset$ is a tuple $\pair{\triples}{\inst}$ that satisfies:
 \begin{eqnarray}
@@
 \end{equation}
-   Equation~\ref{eqn:wellTypedEdge} implies that for every dataset $\dataset$:
+   Equation~\ref{eqn:wellTypedEdge} implies that for every data set $\dataset$:
 \[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ a\inst_{\dataset}A\ \wedge\ b\inst_{\dataset}B\]
@@
    and $\rules$ is a finite set of rules.
-   Each rule in a schema serves to constrain the dataset at runtime, to ensure its semantic integrity.
+   Each rule in a schema serves to constrain the data set at runtime, to ensure its semantic integrity.
    Every rule is an element of an infinite set called $\Rules$,
@@
    rather than $\concepts$, $\rels$, and $\rules$ respectively.
-   For every rule $u$ in the schema, there is a predicate $\sat{u}{\dataset}$ that constrains dataset $\dataset$.
+   For every rule $u$ in the schema, there is a predicate $\sat{u}{\dataset}$ that constrains data set $\dataset$.
 %  Every schema is an element of an infinite set called $\Schema$.
@@
    The type system of Ampersand~\cite{vdWoude2011} ensures that every information system it generates satisfies this requirement.
-   Note that these requirements do not depend on any particular dataset,
+   Note that these requirements do not depend on any particular data set,
    so they can be checked at compile time.
@@
 % \footnote{The definition of ``violation'' is not needed in this paper. We only use the absence of violations to show satisfaction of a constraint.}
-%    of rule $u$ in any dataset.
+%    of rule $u$ in any data set.
 %    We also assume a function $\id{sign} : \Rules\rightarrow\Pair{\Concepts}{\Concepts}$
@@
 % \end{eqnarray}
-%    To determine whether rule $u$ satisfies dataset $\dataset$,
+%    To determine whether rule $u$ satisfies data set $\dataset$,
 %    we define $\sat{u}{\Dataset}$:
@@
 % \end{eqnarray}
-%    Violations are used to produce meaningful error messages and to trigger changes in the dataset to restore invariance of that rule.
-
-   % The maintainance relation between roles and rules is used to determine who is allowed to change the dataset.
+%    Violations are used to produce meaningful error messages and to trigger changes in the data set to restore invariance of that rule.
+
+   % The maintainance relation between roles and rules is used to determine who is allowed to change the data set.
    % For every rule $u$ in the schema, there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
@@
 \label{sct:Information Systems}
-   An information system has a changing dataset because of events that occur.
-   Whenever the system ``observes'' an event, it attempts to change its dataset accordingly.
+   An information system has a changing data set because of events that occur.
+   Whenever the system ``observes'' an event, it attempts to change its data set accordingly.
    However, every rule in the schema represents a constraint,
-   which the dataset must invariably satisfy.
+   which the data set must invariably satisfy.
    That is why rules are also known as ``invariants''.
@@
    Events are taken from an infinite set, $\Events$.
-   An event is denoted as $\evt{d^-}{d^+}$, in which $d^-$ and $d^+$ are datasets.
-   Every event is a function $\Dataset\rightarrow\Dataset$, so $\evt{d^-}{d^+}(\dataset)$ is a dataset.
-   Equation~\ref{eqn:event} defines the effect of an event on a dataset.
+   An event is denoted as $\evt{d^-}{d^+}$, in which $d^-$ and $d^+$ are data sets.
+   Every event is a function $\Dataset\rightarrow\Dataset$, so $\evt{d^-}{d^+}(\dataset)$ is a data set.
+   Equation~\ref{eqn:event} defines the effect of an event on a data set.
 \begin{eqnarray}
@@
    As soon as a system $\infsys$ observes an event $e$,
-   it applies $e$ to its dataset $\dataset$.
+   it applies $e$ to its data set $\dataset$.
    Yet, if $e(\dataset)$ does not satisfy a rule $u$ from schema $\schema_\infsys$,
@@
 \begin{itemize}
-   \item dataset $\dataset=\pair{\triples}{\inst}$ is defined as in section~\ref{sct:Datasets};
+   \item data set $\dataset=\pair{\triples}{\inst}$ is defined as in section~\ref{sct:Data sets};
    \item schema $\schema=\triple{\concepts}{\rels}{\rules}$ is defined as in section~\ref{sct:Schemas};
@@
    % A \define{role} is a name that identifies a group of users.
-   % It serves as a placeholder for a person or a machine (i.e. an actor) that works with the dataset (i.e. create, read, update, or delete triples).
+   % It serves as a placeholder for a person or a machine (i.e. an actor) that works with the data set (i.e. create, read, update, or delete triples).
    % The purpose of a role is to mention an individual user (human) or an automated actor (bot) without knowing who that user is.
@@
    Requirement~\ref{eqn:define R} ensures that the information system only works with triples whose relation is defined in the schema.
-   Requirement~\ref{eqn:satisfaction} ensures that the dataset satisfies all rules.
+   Requirement~\ref{eqn:satisfaction} ensures that the data set satisfies all rules.
 %    The following is for wide tables only
@@
    Note that these requirements apply at runtime.
-   So, as long as a rule $u$ ``lives'' in an information system, the dataset will satisfy that rule.
-
-   As the system registers events, it inserts or deletes triples in its dataset.
+   So, as long as a rule $u$ ``lives'' in an information system, the data set will satisfy that rule.
+
+   As the system registers events, it inserts or deletes triples in its data set.
    If such an event would violate requirement~\ref{eqn:satisfaction},
@@
    It is written in the language Ampersand to make it more appealing to read.
-   Let us first define a dataset of just a handful of triples and three relations.
+   Let us first define a data set of just a handful of triples and three relations.
 \begin{verbatim}
@@
    % Requirement~\ref{eqn:specialization} is satisfied because this example contains no specialization.
-   The Ampersand compiler generates a dataset $\dataset$, which contains a set of triples and a relation $\inst$.
+   The Ampersand compiler generates a data set $\dataset$, which contains a set of triples and a relation $\inst$.
    It defines the set of triples $\triples$ as:
@@
 \end{array}\]
-   The pair $\pair{\triples}{\inst}$ satisfies requirement~\ref{eqn:wellTypedEdge} so this is a dataset $\dataset$ as introduced in section~\ref{sct:Datasets}.
+   The pair $\pair{\triples}{\inst}$ satisfies requirement~\ref{eqn:wellTypedEdge} so this is a data set $\dataset$ as introduced in section~\ref{sct:Data sets}.
 
    The Ampersand compiler generates a schema $\schema$, which contains concepts, a specialization relation, relations, and rules.
-   It defines the sets of concepts and relations to satisfy requirement~\ref{eqn:relationsIntroduceConcepts}:
+   It defines the sets of concepts, relations, and rules to obtain a schema $\schema$:
 \[\begin{array}{rcl}
@@
                \declare{\tt isEnrolledFor}{\tt Student}{\tt Module}\}
-             \end{array}
+             \end{array}\\
+   \rules&=&\{ {\tt EnrollRule}\}
   \end{array}
@@
    % The script defines the set of rules $\rules$ to contain just one rule: \verb-EnrollRule-.
-   % The (nonempty) dataset satisfies this rule, which meets requirement~\ref{eqn:consistentRules}.
+   % The (nonempty) data set satisfies this rule, which meets requirement~\ref{eqn:consistentRules}.
    So, the schema $\schema=\triple{\concepts}{\rels}{\rules}$ satisfies all requirements from section~\ref{sct:Schemas}.
@@
    Sections~\ref{sct:Information Systems} and~\ref{sct:Example existing IS} introduce a single information system.
-   The transition from an existing system to the desired one is the topic of this section.
+   The transition from an existing system $\infsys$ to the desired one $\infsys'$ is the topic of this section.
    Let us first sketch the steps to migrate system $\infsys$ to $\infsys'$.
 \begin{enumerate}
-   \item Let the existing system $\infsys$ have state $\dataset$ and schema $\schema$.
-         A developer defines a desired system $\infsys'$, which contains a dataset of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.} as illustrated in figure~\ref{fig:post-migration}.
-         After the migration, the system will contain the triples from $\dataset$ that satisfy $\schema'$,
-         with additional changes made by the migration engineer (see step~\ref{item3}).
+   \item Let $\infsys$ have state $\dataset$ and schema $\schema$.
+         A developer defines $\infsys'$, which contains a data set of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.} as illustrated in figure~\ref{fig:post-migration}.
+         After the migration, the resulting system will be $\infsys'$, enhanced with the triples from $\dataset$ that satisfy $\schema'$,
+         and with additional changes made by the migration engineer (see step~\ref{item3}) and by users who have been using the system during the migration.
          $\dataset'$ may contain some new data,
@@
          If schemas $\schema$ equals $\schema'$, the data migration is unnecessary, so in the sequel we assume they are different.
-   \item The generator generates a migration script, $\migrsys$,
-         to migrate the data and define the semantics during the migration phase as far as a generator can reasonably predict.
-         It contains the enforce rules that migrate existing data to the desired dataset.
+   \item The migration engineer uses a software generator to generate a migration script, $\migrsys$,
+         This is a system that includes both $\infsys$ and $\infsys'$.
+         $\migrsys$ is meant to migrate the data and define the semantics during the migration phase.
+         It contains enforce rules that migrate existing data to the desired data set.
          It also contains data structures and rules that are needed for the duration of the migration phase.
-         The generated script must be free of type errors, to ensure that $\sat{u}{\Dataset}$ exists for every rule $u$ and dataset $\dataset$.
+         The generated script must be free of type errors, to ensure that $\sat{u}{\Dataset}$ exists for every rule $u$ and data set $\dataset$.
          The proof is available \href{https://www.isa-afp.org/}{here}
@@
          She can test the resulting migration script, $\migrsys'$ separately before taking the desired system to production.
-         The migration engineer does not alter the schemas of the existing system or the desired system.
-         All migration functionality is concentrated in $\migrsys'$.
+         The migration engineer does not alter $\schema$ equals $\schema'$ while adapting $\migrsys'$ to the specific needs of the current migration.
    \item Then, the migration engineer deploys system $\migrsys'$.
@@
          This step is complete when all data from $\dataset$ has been copied to $\dataset'$ and
-         both datasets are in sync with respect to the ongoing transactions.
+         both data sets are in sync with respect to the ongoing transactions.
    \item At this point, the ingress switches the traffic from $\infsys$ to $\migrsys'$,
@@
          This event marks the MoT.
-         From this point onward, users work with the desired functionality in dataset $\dataset'$.
-   \item To give the business an opportunity to evaluate the change,
-         there is room for a go/no-go decision before the the migration engineer makes the migration irreversible.
-         This event marks the moment of no return (MoN).
-         In case of a go-decision, the migration engineer can take down the existing system, which by this time has become unreachable for users.
-         The resulting dataset must satisfy every constraint from $\rules_{\migrsys'}$ before the migration engineer takes down the migration system.
-         In case of a no-go, the migration engineer restores the existing system by letting the ingress switch traffic back to $\infsys$.
+         From this point onward, users work with the desired functionality in data set $\dataset'$.
    \item After taking down the existing system and the migration system, the desired system remains in production and the migration is finished.
@@
    Let us define the instruments needed to describe the derivation of $\schema_{\migrsys}$.
-\begin{definition}[union of datasets]
+\begin{definition}[union of data sets]
 \label{def:graph_union_wellTyped_if_parts_wellTyped}
@@
 \label{lemma:graph_union_wellTyped_if_parts_wellTyped}
-   If $\dataset_1$ and $\dataset_2$ are datasets, then $\dataset_1\cup\dataset_2$ is a dataset.
+   If $\dataset_1$ and $\dataset_2$ are data sets, then $\dataset_1\cup\dataset_2$ is a data set.
 \end{lemma}
 
-\begin{definition}[map relation names in datasets]
+\begin{definition}[map relation names in data sets]
 \label{def:map_labels_preserves_wellTypedness}
@@
 \label{lemma:map_labels_preserves_wellTypedness}
-   If $\dataset$ is a dataset and $h:\Rels\rightarrow\Rels$ is an injective function, then $\maprel{h}{\dataset}$ is a dataset.
+   If $\dataset$ is a data set and $h:\Rels\rightarrow\Rels$ is an injective function, then $\maprel{h}{\dataset}$ is a data set.
 \end{lemma}
 
-\begin{definition}[disjoint union of datasets]
+\begin{definition}[disjoint union of data sets]
 \begin{eqnarray}
-   \rlap{$\pair{\triples}{\inst}\sqcup\pair{\triples'}{\inst'}$}\notag\\
-   &=&\pair{\triples\uplus\triples'}{\inst\uplus^2\inst'}\notag\\
+   \pair{\triples}{\inst}\sqcup\pair{\triples'}{\inst'}&=&\pair{\triples\uplus\triples'}{\inst\uplus^2\inst'}\\
       X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
@@
 \end{definition}
-
+   In a disjoint union, the instances of the two data sets are relabeled to avoid name clashes.
+   In practice, we will not use 0 and 1 as labels for $x$ and $y$, but rather extend their names by prefixing.
+   For example, \texttt{old.Person} and \texttt{new.Address} might be the names of concepts \texttt{Person} and \texttt{Address} from contexts \texttt{old} and \texttt{new}, respectively.
 \begin{lemma}
-   If $\dataset$ and $\dataset'$ are datasets, then so is $\dataset\sqcup\dataset'$.
+   If $\dataset$ and $\dataset'$ are data sets, then so is $\dataset\sqcup\dataset'$.
 \end{lemma}
@@
 \end{definition}
-% \begin{definition}[disjoint union of datasets]
+% \begin{definition}[disjoint union of data sets]
 % \begin{eqnarray}
@@
 \end{definition}
-\begin{definition}[relabel concepts in datasets]
+\begin{definition}[relabel concepts in data sets]
    \[\begin{array}{rcl}
@@
 \end{definition}
-\begin{definition}[relabel relations in datasets]
+\begin{definition}[relabel relations in data sets]
    \[\begin{array}{rcl}
@@
 
+\section*{Generating a Migration Script}
 \subsection{Loading Data}
-   Enforce rules can load the data in the desired database.
+   Before explaining how to migrate data from $\dataset$ to $\dataset'$, we need to explain {\tt ENFORCE} rules.
    In Ampersand, an enforce rule comes in three flavors:
@@
 
+   Now, let us study how to migrate data from $\dataset$ to $\dataset'$.
    It would be tempting to state that a relation from the example system, say {\tt takes}, could be copied by the following code fragment%
@@
 
-\begin{definition}[migration dataset]
+\begin{definition}[migration data set]
    \begin{eqnarray}
@@
 $\pair{(C',1)}{(C,0)}$ and $\pair{(C,0)}{(C',1)}$.
-Note that if $\dataset$ is a dataset, and $\dataset'$ is that dataset with pairs added%
-\footnote{TODO: het woord `add' is informeel, dit kan beter} to the $\isa$ relation, then the result is again a dataset.
+Note that if $\dataset$ is a data set, and $\dataset'$ is that data set with pairs added%
+\footnote{TODO: het woord `add' is informeel, dit kan beter} to the $\isa$ relation, then the result is again a data set.
 The ${\tt ENFORCE\ }r{\tt\ :=\ }e$ statements add triples $(x,r,y)$ for all $(x,y)\in e$.
-The terms $t$ in these statements are such that $x\ (\inst \compose \kleenestar{\flip{\isa}})\ (A',1)$ and $y \ (\inst \compose \kleenestar{\flip{\isa}})\ (B',1)$, thus preserving the property that the result is a dataset.
+The terms $t$ in these statements are such that $x\ (\inst \compose \kleenestar{\flip{\isa}})\ (A',1)$ and $y \ (\inst \compose \kleenestar{\flip{\isa}})\ (B',1)$, thus preserving the property that the result is a data set.
 
@@
 \subsection{Changes}
-% The purpose of this section is to explain why a dataset is structured the way it is.
-   In the migration of a dataset we deal with changes to the elements that are not data:
+% The purpose of this section is to explain why a data set is structured the way it is.
+   In the migration of a data set we deal with changes to the elements that are not data:
    $\inst$, $\concepts$, and $\rels$.
    Such changes have further reaching consequences, however.
-   Changes to $\inst$, $\concepts$, $\rels$ change the data structure in a dataset.
-   We define a \define{migration of a dataset} $\la\atoms,\concepts,\inst,\rels,\dataset\ra$ as a change in which one or more of $\inst$, $\concepts$, or $\rels$ change.
-   To satisfy the definition of datasets, $\dataset$ must change too,
+   Changes to $\inst$, $\concepts$, $\rels$ change the data structure in a data set.
+   We define a \define{migration of a data set} $\la\atoms,\concepts,\inst,\rels,\dataset\ra$ as a change in which one or more of $\inst$, $\concepts$, or $\rels$ change.
+   To satisfy the definition of data sets, $\dataset$ must change too,
    but the migration should try to preserve the maximal amount of data.
@@
    % -> het migratie-systeem bevat alle oude data, ihb nog steeds na het toepassen van de enforce regels
-   \item $\migrsys$ contains the existing data, insofar its enforce rules have not deleted it.
+   \item $\migrsys$ contains all data of the existing system. So, $\dataset_\infsys\subseteq\dataset_\migrsys$.
    % -> er is een pad naar ingebruikname van het nieuwe systeem (vanaf de initiële toestand van het migratie-systeem)
-   \item there is a path to releasing the desired system, starting with the initial state of the $\migrsys$.
+   \item there is a well-founded progress measure towards releasing the desired system.
    % -> corollary: op het moment van ingebruikname van het nieuwe systeem, is het migratie-systeem vrij van overtredingen op regels die het nieuwe systeem moet bewaken
-   \item As a result, the desired system $\infsys'$ is void of violations at the start of its production life.
+   \item When the progress measure indicates that further steps are not possible, the desired system $\infsys'$ is void of violations.
+   The invariants of $\infsys'$ are cork rules in $\migrsys$.
+   So, once all these rules are satisfied (and secured with a cork),
+   all invariants in $\infsys'$ are satisfied and ready for deployment.
    % -> optioneel: na een begrensd aantal 'voortgangs-stappen' kan het nieuwe systeem in gebruik genomen worden
-   \item Optionally, the desired system can be used in production after a limited number of steps.
-   % -> optioneel: er bestaat een migratie-systeem dat de oude functionaliteit behoudt (mogelijk uitbreidt) totdat het nieuwe systeem in gebruik genomen is?
-   \item Optionally, there is a migration systen that preserves the functionality of the existing system until the desired system has been put in production.
 \end{itemize}
 % Bewijsverplichtingen:
+
+\section{Software architecture}
+   Since functionality is excluded from this paper, we discuss the software architecture of the back-end of the migration system.
+   For this research, we have used the Ampersand compiler to generate the system(s).
+   Figure~\ref{fig:initial} shows the starting point of the migration.
+\begin{figure}
+   \centering
+   \includegraphics[width=0.8\textwidth]{figures/Before migration.png}
+   \caption{Initial state of the migration}
+   \label{fig:initial}
+\end{figure}
+   For the purpose of data migration, the existing system must make its schema available through its API.
+   For this purpose, the back-end offers a service ``retrieve schema'', which provides the schema as Ampersand source-code.
+   As a result, the migration engineer can rely on the schema of any generated application to be accurate.
+   For this reason, we can safely assume that the schema this service provides has not been changed since it was deployed.
+   Its presence allows the generator to build various front-ends and migration scripts during the lifetime of the existing system%
+
+   After the migration engineer has produced and tested a migration script, the Ampersand compiler produces a migration system.
+   It includes the schemas of both the existing system and the desired system, besides having a schema of its own.
+   This is implemented by Ampersand's INCLUDE mechanism, which boils down to taking the disjoint union of the schemas.
+\begin{figure}
+   \centering
+   \includegraphics[width=0.8\textwidth]{figures/During migration.png}
+   \caption{During the migration}
+   \label{fig:during}
+\end{figure}
+   Deployment of the migration system (figure~\ref{fig:during}) leaves the existing system intact.
+   It enhances the system with the data set needed by the migration system, which includes the data set needed by the new system.
+   So, users will notice no difference in the existing system when the migration system is deployed.
+
+   When all rules in the migration script have been satisfied, the migration system is ready to be deployed.
+   At this point, the ingress switches traffic from the existing system to the migration system.
+   Until this moment, users have been updating data in the existing system, which is continuously being migrated to the new system.
+   After the ingress switch is made, users have the new functionality at their disposal and make their changes in the new system.
+   If for some reason, there is remaining migration work, the migration system will execute this.
+   Since there is no more input in the existing system, the migration system will eventually be consistent.
+   At this point, the existing system can be removed, which leaves the migration system in place.
+
+   The main purpose of the migration script is to define the migration rules that are used to migrate data from the existing system to the desired system.
+   During this process, users continu to use the existing system.
+
+\begin{figure}
+   \centering
+   \includegraphics[width=0.8\textwidth]{figures/After migration.png}
+   \caption{After the migration}
+   \label{fig:after}
+\end{figure}
 

### 2023-05-14 2023 Disambiguation/Disambiguation.tex
Update changes

--- 2023-03-08 2023 Disambiguation/Disambiguation.tex
@@
 
-An invariant in this system is:
+Note that `type' can refer to both a single concept, like \verb=Task=, \verb=Time= and \verb=Event=, and pairs thereof.
+To avoid ambiguity, we refer to pairs of concepts (such as \verb=Task * Task=, \verb=Time * Time= and \verb=Even * Time=) as signatures.
+We use `type' to refer to both.
 
-\[
-\mathrm{before}_{\mathrm{Task * Task}}\typesemi\mathrm{scheduled}_{\mathrm{Event * Time}} \typesubset \mathrm{scheduled}_\mathrm{Event * Time}\typesemi\mathrm{before}_{\mathrm{Time * Time}}
-\]
+Some invariants in this system are:
+
+\begin{align}
+\mathrm{before}_{\mathrm{Task * Task}}\typesemi\mathrm{scheduled}_{\mathrm{Event * Time}} &\typesubset \mathrm{scheduled}_\mathrm{Event * Time}\typesemi\mathrm{before}_{\mathrm{Time * Time}} \\
+\mathrm{before}_{\mathrm{Task * Task}} \typeinter \typeident_\mathrm{Task} &\typesubset \typenil_\mathrm{Task * Task}
+\end{align}
+
+The intended meaning behind these invariants is: 1) The set of tasks that needs to be performed before some task $x$, all need to be scheduled at times before the time at which $x$ itself is scheduled.
+2) It makes no sense to require a task to be performed before itself.
 
@@
 
-In Ampersand's implementation, Ampersand does not require writing the types when writing Invariants.
+In Ampersand's implementation, Ampersand does not require writing types when writing Invariants.
 We can consider expressing the same invariant as earlier without writing types: $\mathrm{before}\typesemi \mathrm{scheduled} \typesubset \mathrm{scheduled}\typesemi \mathrm{before}$.
-If we require that invariants are well-typed, there is only one way of annotating such expressions with types.
+If we require that invariants are well-typed, there is only one possible intended meaning.
 The use of the possibly ambiguous relation name $\mathrm{before}$ is allowed as long as it is clear which relation is intended, as is the case here.
+
+For the other expression, we cannot omit writing all types: $\mathrm{before} \typeinter \typeident \typesubset \typenil$ could be interpreted as:
+$\mathrm{before}_{\mathrm{Time * Time}} \typeinter \typeident_{\mathrm{Time}} \typesubset \typenil_{\mathrm{Time * Time}}$.
+To avoid this issue, one might write \emph{some} type information: $\mathrm{before} \typeinter \typeident \typesubset \typenil_{\mathrm{Task * Task}}$.
+From this, it should become clear which relations were intended (namely $\mathrm{before}_\mathrm{Task * Task}$ and $\typeident_\mathrm{Task}$), even though they appear elsewhere.
+
 We refer to the problem of determining which relation is intended as the disambiguation problem.%, which we investigate more closely in a separate section.
@@
 
-A disambiguation problem is given by an expression $E$ using the operations $\cup$ and $;$, and relation names from a set of names $\mathcal{N}$, and a set of possible types of each name: $\mathcal{t}(n) \subseteq \mathcal{T}\times \mathcal{T}$ for $n\in \mathcal{N}$.
+A disambiguation problem is given by an expression $E$ using the operations $\cup$ and $;$, and relation names from a set of names $\mathcal{N}$, and a set of possible signatures of each name: $\mathcal{t}(n) \subseteq \mathcal{T}\times \mathcal{T}$ for $n\in \mathcal{N}$.
 A solution to the disambiguation problem is a well-typed expression $S$ in which the annotations to relation names corresponds to an element in $\mathcal{t}(n)$, such that dropping the types from the expression yields $E$.
@@
 % @Bas, deze paragraaf is wellicht leesbaarder als je gewoon de functie à la Haskell uitschrijft.
+% @Stef, zo staat het er al.
 For the top-down step, let $d(e,t)$ denote the disambiguated expression $e$ with the type $t$.
@@
 Let's take $E_2 = \mathrm{scheduled};(\mathrm{before} \cup \mathbb{1})$ as an expression.
-The types are as in the previous example, with $\mathcal{t}(\mathbb{1}) = \left\{(t,t)\mid t\in \mathcal{T}\right\}$.
+The signatures are as in the previous example, with $\mathcal{t}(\mathbb{1}) = \left\{(t,t)\mid t\in \mathcal{T}\right\}$.
 This gives $p(E_2) = \{\mathrm{Task} * \mathrm{Time}\}$ and $d(\{\mathrm{Task} * \mathrm{Time}\},E_2) = \mathrm{scheduled}_{\mathrm{Task} * \mathrm{Time}};(\mathrm{before}_{\mathrm{Time} * \mathrm{Time}}\ \cup\ \mathbb{1}_{\mathrm{Time}})$, as desired.
@@
 \section{Solving Disambiguation with Generalization}
-Before giving a solution to this % @Bas, een referentie over de grens van een paragraaf is taalkundig al niet fraai, maar over de grens van een section is niet slim.
-Als lezer wil ik iets meer duidelijkheid dan "this" als het gaat over welk probleem je naar verwijst.
-problem, we first observe that the disambiguation algorithm discussed earlier fails for our running example.
+We first observe that the disambiguation algorithm discussed above fails for our running example.
 Following the declarations of the script, we let $\mathcal{t}(\mathrm{before}) = \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ and $\mathcal{t}(\mathrm{scheduled}) = \{(\mathrm{Event}, \mathrm{Time})\}$.
-Now the above algorithm would conclude with an error on $E = (\mathrm{before};\mathrm{scheduled}) \cup (\mathrm{scheduled};\mathrm{before})$.
+Now the apply the algorithm on $E = (\mathrm{before};\mathrm{scheduled}) \cup (\mathrm{scheduled};\mathrm{before})$.
+Since the disambiguation algorithm does not consider sub-typing, it would conclude there are is no way to well-type this expression.
+In this section we address this shortcoming.
 
@@
 
-This paper follows a similar approach: We define a language to be used by the Ampersand programmer that allows writing relations without their types as well as subtyping.
-Next, we define a language that is interpreted as a heterogeneous relation algebra with a typical set-interpretation.
-% @Bas, de volgende zin snap ik niet...
-The semantics of the language with subtyping will be determined by what the translation process gives us, but we validate the translation by inspecting certain examples and examining the validity of Tarski's axioms for this language under the resulting interpretation.
-%We deviate slightly from the ideas by van der Woude and Joosten: Their $\epsilon$ relation is a subtyping relation, so $\epsilon_{A*B}$ would exist only if $A$ is a subtype of $B$.
-%In this paper, we relax that requirement.
+%This paper follows a similar approach: We define a language to be used by the Ampersand programmer that allows writing relations without their types as well as subtyping.
+%Next, we define a language that is interpreted as a heterogeneous relation algebra with a typical set-interpretation.
+%% @Bas, de volgende zin snap ik niet...
+%The semantics of the language with subtyping will be determined by what the translation process gives us, but we validate the translation by inspecting certain examples and examining the validity of Tarski's axioms for this language under the resulting interpretation.
+%%We deviate slightly from the ideas by van der Woude and Joosten: Their $\epsilon$ relation is a subtyping relation, so $\epsilon_{A*B}$ would exist only if $A$ is a subtype of $B$.
+%%In this paper, we relax that requirement.
 
@@
 
-\paragraph{[Aside],} The Ampersand language also allows for \verb=CLASSIFY A IS B /\ C= statements, meaning that elements are in $A$ if and only if they are in $B$ and in $C$.
-The information about this captured in $\preceq$ is just that $A \preceq B$ and $A\preceq C$, which means that information is lost.
-We can consider the lost information when dealing with the $\typesemi$ operation by changing the $\epsilon$ relations introduced there.
-However, it is an open problem to deal with this with the $\typeinter$ operation, so we give a simpler presentation that ignores this additional information in \verb=CLASSIFY _ IS _= statements for now.
+%\paragraph{[Aside]} The Ampersand language also allows for \verb=CLASSIFY A IS B /\ C= statements, meaning that elements are in $A$ if and only if they are in $B$ and in $C$.
+%The information about this captured in $\preceq$ is just that $A \preceq B$ and $A\preceq C$, which means that information is lost.
+%We can consider the lost information when dealing with the $\typesemi$ operation by changing the $\epsilon$ relations introduced there.
+%However, it is an open problem to deal with this with the $\typeinter$ operation, so we give a simpler presentation that ignores this additional information in \verb=CLASSIFY _ IS _= statements for now.
+
+Translating a term from $\mathcal{D}$ to $\mathcal{R}$ can be thought of as a search problem:
+For each $\Xi$-term in $\mathcal{D}$, we can substitute it for an element in $\Xi$ to get a term in $\mathcal{R}$.
+Our choice should satisfy the following conditions:
+\begin{itemize}
+\item The chosen term should be well-typed.
+\item Depending on the choices that remain after applying these criteria, we need to decide to pick an arbitrary one or raise an error to the user.
+\item We prefer some terms over others. In particular, we prefer to use $\epsilon_{A*B}$ relations where $A = B$ over those where $A\neq B$.
+\end{itemize}
+
+We illustrate these conditions one by one.
+The first condition is per design: we aim to disambiguate terms by using that they are well-typed.
+The second and third condition are best illustrated by giving an example of what could go wrong without them.
+
+To illustrate why we prefer to use $\varepsilon_{A*B}$ relations where $A=B$, consider the expression: $(\mathrm{before} \typeinter \typeident) \typeunion \typetyped{\typenil}{\mathrm{Task * Task}}$.
+This expression is similar to one of the terms from our running example, but written as a proper $\mathcal{S}$-term.
+Possible disambiguated terms include:
+\begin{itemize}
+\item $\varepsilon_T;(\varepsilon_T;\mathrm{before}_T;\varepsilon_T \cap \varepsilon_T;\typeident_\mathrm{Task};\varepsilon_T);\varepsilon_T \cup \varepsilon_T;F_T;\varepsilon_T$ with $T = \mathrm{Task*Task}$
+\item
+\item
+\end{itemize}
+
+the term $\typetyped{\typeident}{\mathrm{Event}} \typesemi \typevee \typesemi \typetyped{\mathrm{scheduled}}{\mathrm{Event}*\mathrm{Time}}$ with $\mathrm{Task}\preceq\mathrm{Event}$ would not disambiguate to the obvious term, since $\typetyped{\typeident}{\mathrm{Event}} \typesemi \typetyped{\typevee}{\mathrm{Task}*\mathrm{Task}} \typesemi \typetyped{\mathrm{scheduled}}{\mathrm{Event}*\mathrm{Time}}$ is also well-typed.
+To deal with this, we prefer terms in which the $\epsilon$ relations have the same source and target.
 

### 2023-05-19 2023 Disambiguation/Disambiguation.tex
Wrote a selection of disambiguation algorithms we can choose from.

@@
 \usepackage{graphicx,stmaryrd}
-\usepackage{amsmath,amsfonts,amssymb}
+\usepackage{amsmath,amsfonts,amssymb,mathtools}
 \usepackage{bbold,bbm}
@@
 \begin{align}
-\mathrm{before}_{\mathrm{Task * Task}}\typesemi\mathrm{scheduled}_{\mathrm{Event * Time}} &\typesubset \mathrm{scheduled}_\mathrm{Event * Time}\typesemi\mathrm{before}_{\mathrm{Time * Time}} \\
-\mathrm{before}_{\mathrm{Task * Task}} \typeinter \typeident_\mathrm{Task} &\typesubset \typenil_\mathrm{Task * Task}
+\mathrm{before}_{\mathrm{Task * Task}}\typesemi\mathrm{scheduled}_{\mathrm{Event * Time}} &\typesubset \mathrm{scheduled}_\mathrm{Event * Time}\typesemi\mathrm{before}_{\mathrm{Time * Time}} \label{eqn:schedule_sync}
+ \\
+\typecomp{\typeident_\mathrm{Task}} &\typesubset \typecomp{\left(\mathrm{before}_{\mathrm{Task * Task}} \typeinter \typeconv{{\mathrm{before}_{\mathrm{Task * Task}}}}\right)}
+\label{eqn:before_antisym}
 \end{align}
 
-The intended meaning behind these invariants is: 1) The set of tasks that needs to be performed before some task $x$, all need to be scheduled at times before the time at which $x$ itself is scheduled.
-2) It makes no sense to require a task to be performed before itself.
+The operations used to describe these invariants have their typical set-theoretic interpretation.
+Invariant~\ref{eqn:schedule_sync} intends to express: The set of tasks that needs to be performed before some task $x$, all need to be scheduled at times before the time at which $x$ itself is scheduled.
+Invariant~\ref{eqn:before_antisym} intends to express: A task cannot be scheduled both before and after some other task.
+It is the contrapositive of the antisymmetry property: ${\mathrm{before}_{\mathrm{Task * Task}}} \typeinter {\typeconv{{\mathrm{before}_{\mathrm{Task * Task}}}}} \typesubset \typeident_\mathrm{Task}$.
+We focus on this somewhat clumsy formulation of this property to help explain the choices we can make in disambiguation algorithms.
 
@@
 
-For the other expression, we cannot omit writing all types: $\mathrm{before} \typeinter \typeident \typesubset \typenil$ could be interpreted as:
-$\mathrm{before}_{\mathrm{Time * Time}} \typeinter \typeident_{\mathrm{Time}} \typesubset \typenil_{\mathrm{Time * Time}}$.
-To avoid this issue, one might write \emph{some} type information: $\mathrm{before} \typeinter \typeident \typesubset \typenil_{\mathrm{Task * Task}}$.
-From this, it should become clear which relations were intended (namely $\mathrm{before}_\mathrm{Task * Task}$ and $\typeident_\mathrm{Task}$), even though they appear elsewhere.
+For the other expression, we cannot omit writing types.
+The expression $\typecomp\typeident \typesubset \typecomp{\left(\mathrm{before} \typeinter \typeconv{{\mathrm{before}}}\right)}$ is truly ambiguous, as it could be interpreted as:
+$\typecomp{\typeident_{\mathrm{Time}}} \typesubset \typecomp{\left(\mathrm{before}_{\mathrm{Time * Time}} \typeinter \typeconv{\mathrm{before}_{\mathrm{Time * Time}}}\right)}$.
+To avoid this issue, one might write \emph{some} type information: $\typecomp\typeident_\mathrm{Task} \typesubset \typecomp{\left(\mathrm{before} \typeinter \typeconv{{\mathrm{before}}}\right)}$.
+From this, it should become clear which relations were intended (namely $\mathrm{before}_\mathrm{Task * Task}$), even though these appear elsewhere.
 
@@
 
-\section{Solving the disambiguation problem without generalization}
-As an aside, we ask whether a compiler can be expected to figure out which relations are intended in reasonable time.
-In this section, we answer this question for a very simple setting in the affirmative by giving a simple efficient algorithm.
-
-We can define the disambiguation problem as follows:
-
-A disambiguation problem is given by an expression $E$ using the operations $\cup$ and $;$, and relation names from a set of names $\mathcal{N}$, and a set of possible signatures of each name: $\mathcal{t}(n) \subseteq \mathcal{T}\times \mathcal{T}$ for $n\in \mathcal{N}$.
-A solution to the disambiguation problem is a well-typed expression $S$ in which the annotations to relation names corresponds to an element in $\mathcal{t}(n)$, such that dropping the types from the expression yields $E$.
-
-As an example, $E = (\mathrm{before};\mathrm{scheduled}) \cup (\mathrm{scheduled};\mathrm{before})$ with types $\mathcal{t}(\mathrm{before}) = \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ and $\mathcal{t}(\mathrm{scheduled}) = \{(\mathrm{Task}, \mathrm{Time})\}$, is a disambiguation problem.
-It's (here unique) solution would be:
-\[\left(\mathrm{before}_{\mathrm{Task * Task}};\mathrm{scheduled}_{\mathrm{Task * Time}}\right) \cup \left(\mathrm{scheduled}_{\mathrm{Task * Time}};\mathrm{before}_{\mathrm{Time * Time}}\right)\]
-
-We can solve the disambiguation problem using two passes through $E$.
-In the first pass, we decorate the expression tree for $E$ with all the possible types for each sub-expression bottom up.
-Let $p(e)$ denote the possible types, then $p(n) = \mathcal{t}(n)$, and $p(e_1;e_2) = \{(x,y) \mid \exists z. (x,z) \in p(e_1) \wedge (z,y) \in p(e_2))\}$, and $p(e_1 \cup e_2) = p(e_1) \cap p(e_2)$.
-If $p(E)$ is the empty set, then there is no solution to the disambiguation problem, and we are done.
-If $p(E)$ contains multiple elements, then there are multiple solutions to the disambiguation problem.
-We can arbitrarily select a single element from $p(E)$ to get an arbitrary solution\footnote{In practice, we might want to indicate that $E$ is ambiguous, which could require running the next step twice: once for two different elements in $p(E)$.}.
-
-% @Bas, deze paragraaf is wellicht leesbaarder als je gewoon de functie à la Haskell uitschrijft.
-% @Stef, zo staat het er al.
-For the top-down step, let $d(e,t)$ denote the disambiguated expression $e$ with the type $t$.
-For the top-level call of $d(e,t)$, let $e=E$ and $t\in p(E)$.
-Then $d(n,t) = n_{[t]}$, $d(e_1 \cup e_2, t) = d(e_1,t) \cup d(e_2,t)$ and $d(e_1 ; e_2, (x,y)) = d(e_1, (x,z)) ; d(e_2, (z,y))$ where $z$ is picked such that $(x,z)\in p(e_1)$ and $(z,y)\in p(e_2)$.
-Such a $z$ is guaranteed to exist by how $p$ is defined.
-
-As an example, let's look at $E = (\mathrm{before};\mathrm{scheduled}) \cup (\mathrm{scheduled};\mathrm{before})$.
-Then $p(\mathrm{before}) =  \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ following $\mathcal{t}$ and similarly $p(\mathrm{scheduled}) = \{(\mathrm{Task}, \mathrm{Time})\}$.
-This gives subterms $p(\mathrm{before};\mathrm{scheduled}) = p(\mathrm{scheduled};\mathrm{before}) = \{(\mathrm{Task}, \mathrm{Time})\}$ so $p(E) = \{(\mathrm{Task}, \mathrm{Time})\}$.
-
-Top-down, $d(E,t) = d(\mathrm{before};\mathrm{scheduled},t) \cup d(\mathrm{scheduled};\mathrm{before},t)$ with $t=(\mathrm{Task}, \mathrm{Time})$.
-Since $E$ is not ambiguous, we have only one possible way to continue as we solve the disambiguation problem:
-\begin{align*}d(\mathrm{before};\mathrm{scheduled},t) &= d(\mathrm{before},(\mathrm{Task}, \mathrm{Task})) ;  d(\mathrm{scheduled},(\mathrm{Task}, \mathrm{Time})) \\
-&=\mathrm{before}_{\mathrm{Task} * \mathrm{Task}}; \mathrm{scheduled}_{\mathrm{Task}* \mathrm{Time}}\\
-d(\mathrm{scheduled};\mathrm{before},t) &=  d(\mathrm{scheduled},(\mathrm{Task}, \mathrm{Time}));d(\mathrm{before},(\mathrm{Time}, \mathrm{Time}))\\
-&=\mathrm{scheduled}_{\mathrm{Task} * \mathrm{Time}}; \mathrm{before}_{\mathrm{Time}* \mathrm{Time}}\\
-\end{align*}
-
-The disambiguation problem and the corresponding algorithm can be extended over all operations in relation algebra.
-It can even be used to disambiguate special relations like the identity relation.
-Let's take $E_2 = \mathrm{scheduled};(\mathrm{before} \cup \mathbb{1})$ as an expression.
-The signatures are as in the previous example, with $\mathcal{t}(\mathbb{1}) = \left\{(t,t)\mid t\in \mathcal{T}\right\}$.
-This gives $p(E_2) = \{\mathrm{Task} * \mathrm{Time}\}$ and $d(\{\mathrm{Task} * \mathrm{Time}\},E_2) = \mathrm{scheduled}_{\mathrm{Task} * \mathrm{Time}};(\mathrm{before}_{\mathrm{Time} * \mathrm{Time}}\ \cup\ \mathbb{1}_{\mathrm{Time}})$, as desired.
-
-\section{Solving Disambiguation with Generalization}
+\section{Solving Disambiguation}
 We first observe that the disambiguation algorithm discussed above fails for our running example.
@@
 
-%This paper follows a similar approach: We define a language to be used by the Ampersand programmer that allows writing relations without their types as well as subtyping.
+%This paper follows a similar approach: We define a language to be used by the Ampersand programmer that allows writing relations without their types as well as sub-typing.
 %Next, we define a language that is interpreted as a heterogeneous relation algebra with a typical set-interpretation.
 %% @Bas, de volgende zin snap ik niet...
-%The semantics of the language with subtyping will be determined by what the translation process gives us, but we validate the translation by inspecting certain examples and examining the validity of Tarski's axioms for this language under the resulting interpretation.
-%%We deviate slightly from the ideas by van der Woude and Joosten: Their $\epsilon$ relation is a subtyping relation, so $\epsilon_{A*B}$ would exist only if $A$ is a subtype of $B$.
+%The semantics of the language with sub-typing will be determined by what the translation process gives us, but we validate the translation by inspecting certain examples and examining the validity of Tarski's axioms for this language under the resulting interpretation.
+%%We deviate slightly from the ideas by van der Woude and Joosten: Their $\epsilon$ relation is a sub-typing relation, so $\epsilon_{A*B}$ would exist only if $A$ is a sub-type of $B$.
 %%In this paper, we relax that requirement.
@@
 
-Let $\mathcal{S}$ be the language of heterogeneous relation algebra with subtyping.
+Let $\mathcal{S}$ be the language of heterogeneous relation algebra with sub-typing.
 It is defined by the following grammar, where $n\in\mathcal{N}$ and $A,B\in\mathcal{T}$:
@@
 
-Our aim is to disambiguate expressions written in this language into `normal' heterogeneous relation algebra, that is: without subtyping.
-To signal certain type errors, we add an additional element $\bot$ to the set of concepts, that we will assume is different from every element in $\mathcal{T}$.
-Since $\mathcal{T}$ is referred to as the set of concepts, we will refer to $\mathcal{T}\cup\{\bot\}$ as the set of pseudo-concepts.
-We use the following grammar to describe the language $\mathcal{R}$ for that, where $n_{A*B}$ satisfies $(A,B)\in \mathcal{t}(n)$, and $A,B\in\mathcal{T}\cup\{\bot\}$ for all other occurrences of $A$ and $B$:
+Our aim is to disambiguate expressions written in this language into `normal' heterogeneous relation algebra, that is: without sub-typing.
+%To signal certain type errors, we add an additional element $\bot$ to the set of concepts, that we will assume is different from every element in $\mathcal{T}$.
+%Since $\mathcal{T}$ is referred to as the set of concepts, we will refer to $\mathcal{T}\cup\{\bot\}$ as the set of pseudo-concepts.
+We use the following grammar to describe the language $\mathcal{R}$ for that, where $n_{A*B}$ satisfies $(A,B)\in \mathcal{t}(n)$, and $A,B\in\mathcal{T}%\cup\{\bot\}
+$ for all other occurrences of $A$ and $B$:
 \[
@@
 \end{align*}
-We say that a term $r\in \mathcal{R}$ is \emph{well-typed} if $r : A*B$ for $A,B\in \mathcal{T}$ and every sub-term is well-typed.
-By induction on the inductive definition of $\mathcal{R}$, the relation $:$ is univalent.
-In other words: well-typed terms in $\mathcal{R}$ have precisely one type, all of its sub-terms are well-typed, and $\bot$ does not occur as a pseudo-concept.
+We say that a term $r\in \mathcal{R}$ is \emph{well-typed} if $r : A*B$ for $A,B\in \mathcal{T}$.
+As a consequence of this definition all sub-terms of a well-typed term are well-typed.
+Moreover, the relation $:$ is univalent (that is: well-typed terms in $\mathcal{R}$ have precisely one type).%, and $\bot$ does not occur as a pseudo-concept.
 
 \subsection{Translation process}
-Rather than translating terms in $\mathcal{S}$ into $\mathcal{R}$ directly, we create an intermediate language $\mathcal{D}$, which contains $\mathcal{R}$ (that is: $\mathcal{R} \subseteq \mathcal{D}$).
+Rather than translating terms in $\mathcal{S}$ into $\mathcal{R}$ directly, we create an intermediate language $\mathcal{D}$.
 This will allow us to use a purely syntactic translation from $\mathcal{S}$ to $\mathcal{D}$, after which we run a disambiguation algorithm to turn it into a term in $\mathcal{R}$.
-The language $\mathcal{D}$ is defined as follows, where $\xi \in \mathcal{R}$, $\Xi \subseteq \mathcal{R}$, and $A,B\in\mathcal{T}$:
+The language $\mathcal{D}$ is defined as follows, where $\Xi \subseteq \mathcal{R}$, and $A,B\in\mathcal{T}$:
 \[
-\varphi,\psi\in\mathcal{D} \mathbin{:=} \xi \mid \varphi \cup \psi \mid \varphi \cap \psi \mid \varphi ; \psi \mid \overline{\psi}\mid \conv{\psi} \mid \typetyped{\varphi}{A*B} \mid \Xi
+\varphi,\psi\in\mathcal{D} \mathbin{:=} \varphi \cup \psi \mid \varphi \cap \psi \mid \varphi ; \psi \mid \overline{\psi}\mid \conv{\psi} \mid \typetyped{\varphi}{A*B} \mid \Xi
 \]
@@
 We write $A\preceq B$ to indicate that the pair $A,B$ is in the transitive closure of the pairs written \verb=CLASSIFY _ ISA _=, meaning that all elements in $A$ must occur in $B$ whenever $A\preceq B$.
-Moreover, we let $\bot$ be a least element in this preorder: $\bot \preceq A$ for all $A$.
-Since there are no classify statements containing $\bot$ (the user cannot write them), the least element in $\preceq$ is unique.
+%Moreover, we let $\bot$ be a least element in this preorder: $\bot \preceq A$ for all $A$.
+%Since there are no classify statements containing $\bot$ (the user cannot write them), the least element in $\preceq$ is unique.
 
@@
 s(\varphi \typeunion \psi) &= \conv{\varepsilon};s(\varphi);\varepsilon\ \cup\ \conv{\varepsilon};s(\psi);\varepsilon&\quad\text{with $\varepsilon = \{\epsilon_{A*B} \mid A\preceq B \}$} \\
-s(\varphi \typeinter \psi) &= \conv{\varepsilon};s(\varphi);\varepsilon\ \cap\ \conv{\varepsilon};s(\psi);\varepsilon&\quad\text{with $\varepsilon = \{\epsilon_{A*B} \mid B\preceq A \}$}\\
+s(\varphi \typeinter \psi) &= \conv{\varepsilon};s(\varphi);\varepsilon\ \cap\ \conv{\varepsilon};s(\psi);\varepsilon&\quad\text{with $\varepsilon = \{\epsilon_{A*B} \mid A\preceq B \}$}\\
 s(\varphi \typesemi \psi) &= s(\varphi);\varepsilon;s(\psi)&\text{with $\varepsilon = \{\epsilon_{A*B} \mid B\preceq A \vee A\preceq B \}$}\\
@@
 
+\section{Solving the disambiguation problem}
 Translating a term from $\mathcal{D}$ to $\mathcal{R}$ can be thought of as a search problem:
@@
 \item The chosen term should be well-typed.
+\item We prefer some terms over others. In particular, we prefer to use $\epsilon_{A*B}$ relations where $A = B$ over those where $A\neq B$.
 \item Depending on the choices that remain after applying these criteria, we need to decide to pick an arbitrary one or raise an error to the user.
-\item We prefer some terms over others. In particular, we prefer to use $\epsilon_{A*B}$ relations where $A = B$ over those where $A\neq B$.
 \end{itemize}
@@
 
-To illustrate why we prefer to use $\varepsilon_{A*B}$ relations where $A=B$, consider the expression: $(\mathrm{before} \typeinter \typeident) \typeunion \typetyped{\typenil}{\mathrm{Task * Task}}$.
-This expression is similar to one of the terms from our running example, but written as a proper $\mathcal{S}$-term.
+To illustrate why we prefer to use $\epsilon_{A*B}$ relations where $A=B$, consider the expression: $\typecomp{\typeident} \typeunion \typecomp{\left(\typetyped{\mathrm{before}}{Task*Task} \typeinter \typeconv{\mathrm{before}}\right)}$.
+This expression is similar to Invariant~\ref{eqn:before_antisym} from our running example, but written as a proper $\mathcal{S}$-term.
 Possible disambiguated terms include:
+\begin{align}
+\epsilon_{t_0};\overline{1_{\mathrm{Task}}};\epsilon_{t_0} &\cup \epsilon_{t_0};\overline{\left(\epsilon_{t_0};\typetyped{\mathrm{before}_{t_0}}{t_0};\epsilon_{t_0} \cap \epsilon_{t_0};\conv{\mathrm{before}_{t_0}};\epsilon_{t_0} \right)};\epsilon_{t_0}\label{eqn:disambGood}\\
+\epsilon_{t_3};\overline{1_{\mathrm{Event}}};\epsilon_{t_3} &\cup \epsilon_{t_1};\overline{\left(\epsilon_{t_0};\typetyped{\mathrm{before}_{t_0}}{t_0};\epsilon_{t_0} \cap \epsilon_{t_0};\conv{\mathrm{before}_{t_0}};\epsilon_{t_0} \right)};\epsilon_{t_2}\label{eqn:disambBad}\\
+\shortintertext{\flushright{ with ${t_0} = \mathrm{Task*Task}$, ${t_1} = \mathrm{Event*Task}$, ${t_2} = \mathrm{Task*Event}$, and ${t_3} = \mathrm{Event*Event}$}}\nonumber{}
+\end{align}
+
+The intended term is Term~\ref{eqn:disambGood} here.
+Its meaning is different from that of Term~\ref{eqn:disambBad}, since the term $\epsilon_{t_3};\typecomp{1_{\mathrm{Event}}};\epsilon_{t_3}$ can contain pairs in $\mathrm{Event}$ that $\epsilon_{t_0};\typecomp{1_{\mathrm{Task}}};\epsilon_{t_0}$ does not.
+We could argue that the user must specify more type information here, since the meanings for these terms differ.
+However, the term can be interpreted as one that does not use sub-typing in any meaningful way.
+Since the purpose of a disambiguation system is to reduce verbosity, we should prefer the intended term over Term~\ref{eqn:disambBad}.
+A criterium that distinguishes the two terms is that the intended disambiguation uses $\epsilon_{t_0}$ throughout, which will be the preferred term if we prefer $\epsilon_{A*B}$'s that do not employ sub-typing (that is: those where $A=B$).
+
+What if the user writes $\typecomp{\typetyped{\typeident}{\mathrm{Event*Event}}} \typeunion \typecomp{\left(\mathrm{before} \typeinter \typeconv{\mathrm{before}}\right)}$, a term forces the use of sub-typing?
+Now possible disambiguated terms include:
+\begin{align}
+\epsilon_{t_3};\overline{\typetyped{1_{\mathrm{Event}}}{t_3}};\epsilon_{t_3} &\cup \epsilon_{t_3};\overline{\left(\epsilon_{t_1};\mathrm{before}_{t_0};\epsilon_{t_2} \cap \epsilon_{t_1};\conv{\mathrm{before}_{t_0}};\epsilon_{t_2} \right)};\epsilon_{t_3}\label{eqn:disambOption}\\
+\epsilon_{t_3};\overline{\typetyped{1_{\mathrm{Event}}}{t_3}};\epsilon_{t_3} &\cup \epsilon_{t_1};\overline{\left(\epsilon_{t_0};\mathrm{before}_{t_0};\epsilon_{t_0} \cap \epsilon_{t_0};\conv{\mathrm{before}_{t_0}};\epsilon_{t_0} \right)};\epsilon_{t_2}\label{eqn:disambOld}\\
+\shortintertext{\flushright{ with ${t_0} = \mathrm{Task*Task}$, ${t_1} = \mathrm{Event*Task}$, ${t_2} = \mathrm{Task*Event}$, and ${t_3} = \mathrm{Event*Event}$}}\nonumber{}
+\end{align}
+
+Observe that again, the two terms have different semantics, due to the complement being computed over a different type.
+In previous publications on the Ampersand type system, where sub-typing did not play a role, Term~\ref{eqn:disambOld} would be chosen over Term~\ref{eqn:disambOption}.
+The rational here is that types are computed bottom up.
+However, we have observed that sub-typing around complements can be error-prone, and the Term~\ref{eqn:disambOption} might well have been intended.
+We therefore argue that an error should be raised here.
+The algorithm presented can be configured either way: accepting the term as Term~\ref{eqn:disambOld}, or throwing an error.
+%As a consequence, we argue that it is desirable to require more information from the user here.
+%terms like $\typecomp{\typetyped{\typeident}{\mathrm{Event*Event}}} \typeunion \typecomp{\left(\mathrm{before} \typeinter \typeconv{\mathrm{before}}\right)}$ are
+
+\section{Solving the disambiguation problem}
+%We define the disambiguation problem as follows:
+%
+%A disambiguation problem is given by an expression $E$ using the operations $\cup$ and $;$, and relation names from a set of names $\mathcal{N}$, and a set of possible signatures of each name: $\mathcal{t}(n) \subseteq \mathcal{T}\times \mathcal{T}$ for $n\in \mathcal{N}$.
+%A solution to the disambiguation problem is a well-typed expression $S$ in which the annotations to relation names corresponds to an element in $\mathcal{t}(n)$, such that dropping the types from the expression yields $E$.
+%
+%As an example, $E = (\mathrm{before};\mathrm{scheduled}) \cup (\mathrm{scheduled};\mathrm{before})$ with types $\mathcal{t}(\mathrm{before}) = \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ and $\mathcal{t}(\mathrm{scheduled}) = \{(\mathrm{Task}, \mathrm{Time})\}$, is a disambiguation problem.
+%It's (here unique) solution would be:
+%\[\left(\mathrm{before}_{\mathrm{Task * Task}};\mathrm{scheduled}_{\mathrm{Task * Time}}\right) \cup \left(\mathrm{scheduled}_{\mathrm{Task * Time}};\mathrm{before}_{\mathrm{Time * Time}}\right)\]
+
+The question of whether a type-correct version of a term exists can be answered by a simple bottom-up calculation of the possible types for that term.
+We define $p : \mathcal{D} \rightarrow \powerset{\left(\mathcal{T}\times \mathcal{T}\right)}$, for `possible types' by:
+\begin{align*}
+p(\Xi) &= \{t \mid r : t \text{ for $r\in\Xi$}\}\\
+p(e_1; e_2) &= \{(x,y) \mid \exists z. (x,z) \in p(e_1) \wedge (z,y) \in p(e_2))\}\\
+p(e_1\cup e_2) = p(e_1\cap e_2) &= \{(x,y) \mid (x,y) \in p(e_1) \wedge (x,y)\in p(e_2)\}\\
+p(\overline{e}) &= p(e)\\
+p(\typetyped{e}{t}) &= \begin{cases}\{t\}\quad\text{if $t\in p(e_1)$}\\\emptyset\quad\text{otherwise}\end{cases}\\
+p(\conv{e}) &= \{(y,x) \mid (x,y)\in p(e)\}
+\end{align*}
+
+If $p(e) = \emptyset$ then $e$ has a type error, and a smallest sub-term $s$ of $e$ for which $p(s)=\emptyset$ can be used to give a meaningful type error.
+We can compute $p(e)$ in time that is at most quadratic in terms of $\left|\mathcal{T}\times \mathcal{T}\right|$, so computing $p$ can be done in polynomial time in $\left|\mathcal{T}\right|$.
+For Ampersand, $\left|\mathcal{T}\right|$ is at most linear in the size of the user's script.
+
+Aside from type errors, we can use $p$ to find ambiguous terms:
+If there is a sub-term $\Xi$ of $e$ that can be replaced by both $x\in \Xi$ and $y\in Xi$ for some $x \neq y$, then $p(e')$ and $p(e'')$ are nonempty for $e'$ and $e''$ equal to $e$ under the substitution of $\Xi$ for $\{x\}$ and $\{y\}$ respectively.
+Furthermore, we can use $p$ to find out whether a negation can have multiple types.
+
+A second operation at our disposal is one that selects preferred $\mathcal{R}$-terms for each $\Xi$ as a way to avoid ambiguous terms.
+This operation fulfils our requirement that we should prefer some terms over others.
+Let $f : \powerset{\mathcal{R}}\to\powerset{\mathcal{R}}$ be an idempotent function (a function such that $f(f(x))=f(x)$ for all $x$) that returns a subset of its argument, indicating which elements would be preferred.
+In particular, choosing $f(\Xi) = \left\{\epsilon_{A*B} \mid \epsilon_{A*B}\in \Xi \wedge A=B\right\}$ would have us pick $\epsilon_{A*B}$ relations with $A=B$.
+Given such a function $f$, we say that $e'$ is more preferable to $e$ if $p(e')\neq \emptyset$ and $e'$ can be obtained from $e$ by substituting some $\Xi$ terms for $f(\Xi)$.
+We say that $e$ is {\emph maximally preferable} if $p(e)\neq \emptyset$ and for every occurrence of a $\Xi$-term in $e$, substituting it for $f(\Xi)$ would either leave $e$ unchanged (as $\Xi=f(\Xi)$ or result in a term $e'$ such that $p(e')=\emptyset$.
+Terms that are more preferable to $e$ form a lattice\footnote{Proof: `more preferable' is the subset relation for the set of $\Xi$-term occurrences that have been changed wrt $e$.}, and we call the least upper bound to all maximally preferable terms the term that is generally preferred to $e$ under $f$.
+We believe\footnote{Sketch: Create a maximally preferable term using an eager algorithm, then mark all the $\Xi$-terms that are not in such maximally preferable terms as not in the generally preferred term.
+Keep going starting from $\Xi$-terms that are known not to be in the generally preferred term.
+Each iteration a new term that is not in the generally preferred term should be discovered: if not, we are done and the remaining terms are those that can be replaced in the generally preferred term.} we can find the generally preferred term in polynomial time.
+
+With these operations, we can create several strategies for disambiguating terms:
 \begin{itemize}
-\item $\varepsilon_T;(\varepsilon_T;\mathrm{before}_T;\varepsilon_T \cap \varepsilon_T;\typeident_\mathrm{Task};\varepsilon_T);\varepsilon_T \cup \varepsilon_T;F_T;\varepsilon_T$ with $T = \mathrm{Task*Task}$
-\item
-\item
+\item Find all potential ambiguities: declared relations, full relations, identity relations and complements should only have one legal type. (Don't use the generally preferred term.)
+\item Find all declared relations that are ambiguous first, then pick the generally preferred term, and then check that full relations, identity relations and complements can only have one type.
+\item Find the generally preferred term, and check for ambiguities: declared relations, full relations, identity relations and complements should only have one legal type.
+\item Find the generally preferred term, and while ambiguities remain, repeat with different functions for generally preferred terms.
 \end{itemize}
 
-the term $\typetyped{\typeident}{\mathrm{Event}} \typesemi \typevee \typesemi \typetyped{\mathrm{scheduled}}{\mathrm{Event}*\mathrm{Time}}$ with $\mathrm{Task}\preceq\mathrm{Event}$ would not disambiguate to the obvious term, since $\typetyped{\typeident}{\mathrm{Event}} \typesemi \typetyped{\typevee}{\mathrm{Task}*\mathrm{Task}} \typesemi \typetyped{\mathrm{scheduled}}{\mathrm{Event}*\mathrm{Time}}$ is also well-typed.
-To deal with this, we prefer terms in which the $\epsilon$ relations have the same source and target.
+%
+%
+%
+%%TODO: hier verder
+%
+%For the top-down step, let $d(e,t)$ denote the disambiguated expression $e$ with the type $t$.
+%For the top-level call of $d(e,t)$, let $e=E$ and $t\in p(E)$.
+%Then $d(n,t) = n_{[t]}$, $d(e_1 \cup e_2, t) = d(e_1,t) \cup d(e_2,t)$ and $d(e_1 ; e_2, (x,y)) = d(e_1, (x,z)) ; d(e_2, (z,y))$ where $z$ is picked such that $(x,z)\in p(e_1)$ and $(z,y)\in p(e_2)$.
+%Such a $z$ is guaranteed to exist by how $p$ is defined.
+%
+%As an example, let's look at $E = (\mathrm{before};\mathrm{scheduled}) \cup (\mathrm{scheduled};\mathrm{before})$.
+%Then $p(\mathrm{before}) =  \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ following $\mathcal{t}$ and similarly $p(\mathrm{scheduled}) = \{(\mathrm{Task}, \mathrm{Time})\}$.
+%This gives subterms $p(\mathrm{before};\mathrm{scheduled}) = p(\mathrm{scheduled};\mathrm{before}) = \{(\mathrm{Task}, \mathrm{Time})\}$ so $p(E) = \{(\mathrm{Task}, \mathrm{Time})\}$.
+%
+%Top-down, $d(E,t) = d(\mathrm{before};\mathrm{scheduled},t) \cup d(\mathrm{scheduled};\mathrm{before},t)$ with $t=(\mathrm{Task}, \mathrm{Time})$.
+%Since $E$ is not ambiguous, we have only one possible way to continue as we solve the disambiguation problem:
+%\begin{align*}d(\mathrm{before};\mathrm{scheduled},t) &= d(\mathrm{before},(\mathrm{Task}, \mathrm{Task})) ;  d(\mathrm{scheduled},(\mathrm{Task}, \mathrm{Time})) \\
+%&=\mathrm{before}_{\mathrm{Task} * \mathrm{Task}}; \mathrm{scheduled}_{\mathrm{Task}* \mathrm{Time}}\\
+%d(\mathrm{scheduled};\mathrm{before},t) &=  d(\mathrm{scheduled},(\mathrm{Task}, \mathrm{Time}));d(\mathrm{before},(\mathrm{Time}, \mathrm{Time}))\\
+%&=\mathrm{scheduled}_{\mathrm{Task} * \mathrm{Time}}; \mathrm{before}_{\mathrm{Time}* \mathrm{Time}}\\
+%\end{align*}
+%
+%The disambiguation problem and the corresponding algorithm can be extended over all operations in relation algebra.
+%It can even be used to disambiguate special relations like the identity relation.
+%Let's take $E_2 = \mathrm{scheduled};(\mathrm{before} \cup \mathbb{1})$ as an expression.
+%The signatures are as in the previous example, with $\mathcal{t}(\mathbb{1}) = \left\{(t,t)\mid t\in \mathcal{T}\right\}$.
+%This gives $p(E_2) = \{\mathrm{Task} * \mathrm{Time}\}$ and $d(\{\mathrm{Task} * \mathrm{Time}\},E_2) = \mathrm{scheduled}_{\mathrm{Task} * \mathrm{Time}};(\mathrm{before}_{\mathrm{Time} * \mathrm{Time}}\ \cup\ \mathbb{1}_{\mathrm{Time}})$, as desired.
+%
 

### 2023-06-04 2022 Migration/articleMigration.tex
little things

@@
 \begin{enumerate}
-   \item Let $\infsys$ have state $\dataset$ and schema $\schema$.
+   \item Let $\infsys$ have data set $\dataset$ and schema $\schema$.
          A developer defines $\infsys'$, which contains a data set of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.} as illustrated in figure~\ref{fig:post-migration}.
@@
          It also contains data structures and rules that are needed for the duration of the migration phase.
-         The generated script must be free of type errors, to ensure that $\sat{u}{\Dataset}$ exists for every rule $u$ and data set $\dataset$.
+         The generated script must be free of type errors, to ensure that $\sat{u}{\dataset}$ exists for every rule $u$ and data set $\dataset$.
          The proof is available \href{https://www.isa-afp.org/}{here}
@@
 \end{eqnarray}
-   An engine enforces all enforce rules by changing the population of relation $r$.
+   The run time environment contains an engine that enforces such rules.
+   The engine will change the population of relation $r$ by the least set of pairs needed to satisfy the rule.
    An enforce rule with operator {\tt :<} (alternative~\ref{enforce del}) keeps the population of $r$ smaller than or equal to \id{expression}.
@@
    When the population of this expression is growing with pairs that are not in $r$, the engine will insert such pairs in $r$ to keep $r$ greater than or equal to \id{expression}.
-   An enforce rule with operator {\tt :=} (alternative~\ref{enforce eq}) keeps the population of $r$ equal to \id{expression}.
-   It does so by inserting and deleting pairs in $r$ to keep it equal to \id{expression}.
+   An enforce rule with operator {\tt :=} (alternative~\ref{enforce eq}) keeps the population of $r$ equal to \id{expression} by inserting and deleting the right pairs in $r$.
    Summarizing, an enforce rule can be read as an invariant that keeps one of $r\leq\id{expression}$, $r=\id{expression}$, or $r\geq\id{expression}$ satisfied by changing the population of $r$.
@@
 \end{verbatim}
+   The relation `old.takes' resides in the existing system and will not change anymore.
+
    This would mean, however, that all pairs from {\tt old.takes} will be in {\tt new.takes} forever.

### 2023-06-27 2022 Migration/articleMigration.tex
Changed file extensions so that latex doesn't complain. Note: Please add the missing files to git

@@
    \begin{center}
-     \includegraphics[scale=.45]{figures/datamigration-Pre-migration.drawio.png}
+     \includegraphics[scale=.45]{figures/datamigration-Pre-migration.png}
    \end{center}
@@
    \begin{center}
-     \includegraphics[scale=.35]{figures/datamigration-Migration phase.drawio.png}
+     \includegraphics[scale=.35]{figures/datamigration-Migration phase.png}
    \end{center}
@@
    \begin{center}
-     \includegraphics[scale=.35]{figures/datamigration-Post-migration.drawio.png}
+     \includegraphics[scale=.35]{figures/datamigration-Post-migration.png}
    \end{center}

### 2023-06-27 2022 Migration/articleMigration.tex
WIP on the migration process

@@
    % Automatic enforcement is specified by the developer with a special syntax:
-   % the {\em enforce rule}.
-   % An enforce rule specifies not only the rule,
+   % the {\em enforcement rule}.
+   % An enforcement rule specifies not only the rule,
    % but it implicitly defines event $e'$ to restore the invariance.
-   % The system must have an engine that restores invariance of all enforce rules,
+   % The system must have an engine that restores invariance of all enforcement rules,
    % so users will experience that these rules are always satisfied.
@@
 
-   % If a rule $u$ is not assigned to any role and it is not an enforce rule,
+   % If a rule $u$ is not assigned to any role and it is not an enforcement rule,
    % the system must prevent any event $e$ from having any effect on the database.
@@
 \begin{enumerate}
-   \item Let $\infsys$ have data set $\dataset$ and schema $\schema$.
-         A developer defines $\infsys'$, which contains a data set of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.} as illustrated in figure~\ref{fig:post-migration}.
+   \item Let the existing information system $\infsys$ have data set $\dataset$ and schema $\schema$.
+         A developer defines a new information system, $\infsys'$, which contains a data set of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.} as illustrated in figure~\ref{fig:post-migration}.
          After the migration, the resulting system will be $\infsys'$, enhanced with the triples from $\dataset$ that satisfy $\schema'$,
@@
          $\dataset'$ may contain some new data,
-         for instance to initialize new features on the moment of transition (MoT).
+         for instance, to initialize new features at the moment of transition (MoT).
          If schemas $\schema$ equals $\schema'$, the data migration is unnecessary, so in the sequel we assume they are different.
@@
          $\migrsys$ is meant to migrate the data and define the semantics during the migration phase.
-         It contains enforce rules that migrate existing data to the desired data set.
+         It contains enforcement rules that migrate existing data to the desired data set.
          It also contains data structures and rules that are needed for the duration of the migration phase.
@@
          so the migration engineer can accommodate such requirements.
-         This can be necessary, for example to resolve data pollution that violates a constraint that was not in schema $\schema_{\infsys}$.
-         She can either implement enforce rules or use the business to resolve the issues.
+         This can be necessary, for example, to resolve data pollution that violates a constraint that was not in schema $\schema_{\infsys}$.
+         She can either implement enforcement rules or use the business to resolve the issues.
          She can test the resulting migration script, $\migrsys'$ separately before taking the desired system to production.
-         The migration engineer does not alter $\schema$ equals $\schema'$ while adapting $\migrsys'$ to the specific needs of the current migration.
+         The migration engineer alters $\schema$ nor $\schema'$ while adapting $\migrsys'$ to the specific needs of the current migration.
    \item Then, the migration engineer deploys system $\migrsys'$.
-         This event starts the copying of data from $\dataset$ to $\dataset'$ (figure~\ref{fig:migration phase}).
+         This event starts the copying of data from $\dataset$ to $\dataset'$, as specified in $\migrsys'$ (figure~\ref{fig:migration phase}).
          During this phase, the ingress is still referring users to the existing system,
@@
          but these changes are being transferred to $\dataset'$ as well.
-         This step is complete when all data from $\dataset$ has been copied to $\dataset'$ and
-         both data sets are in sync with respect to the ongoing transactions.
-   \item At this point, the ingress switches the traffic from $\infsys$ to $\migrsys'$,
-         so users will now experience the new functionality.
+         This step is complete when all data from $\dataset$ has been copied to $\dataset'$ and all invariants of the migration system are true.
+   \item At this moment, the ingress switches the incoming traffic from $\infsys$ to $\migrsys'$,
+         so $\migrsys'$ effectively takes over.
          This event marks the MoT.
-         From this point onward, users work with the desired functionality in data set $\dataset'$.
-   \item After taking down the existing system and the migration system, the desired system remains in production and the migration is finished.
+         Since $\infsys$ may still have some violations due to last-minute edits,
+         it may take some (short but finite) time until the invariants of $\infsys$ are true.
+         After that, dataset $\dataset$ will no longer change.
+   \item From this moment, users experience the functionality of the migration system $\migrsys'$.
+         This system must ensure that all invariants of $\schema'$ are true before the migration is complete.
+         It no longer ensures the rules of $\schema$, which govern the existing system.
+         Since $\dataset$ is no longer changing, the migration engineer can now focus on the remaining violations of $\schema'$.
+         Users of the migration system will deal with process rules in the migration system that are still violated.
+         Once all invariants of $\migrsys'$ are true, the migration is complete.
+         This state marks the moment of cleanup (MoC)
+   \item After the MoC, the ingress switches the incoming traffic from $\migrsys'$ to $\infsys'$.
+         The migration system is no longer needed and can be taken out of production, together with the existing system $\infsys$.
+         The desired system remains in production and the migration is finished.
          This establishes the post-migration situation (figure~\ref{fig:post-migration}).
@@
    Then the substitution $[C\rightarrow D]$ means that every atom $a$ that used to be of concept $C$ in the existing system, is now not only a $D$ but also an $A$.
-   For example by relabeling the concept {\tt Hotel} to {\tt Motel}, in a situation where ${\tt Motel}\isa{\tt Parking}$,
+   For example, by relabeling the concept {\tt Hotel} to {\tt Motel}, in a situation where ${\tt Motel}\isa{\tt Parking}$,
    all atoms that represent hotels might suddenly be expected to have parking lots.
@@
    Before explaining how to migrate data from $\dataset$ to $\dataset'$, we need to explain {\tt ENFORCE} rules.
-   In Ampersand, an enforce rule comes in three flavors:
+   In Ampersand, an enforcement rule comes in three flavors:
 \begin{eqnarray}
@@
    The engine will change the population of relation $r$ by the least set of pairs needed to satisfy the rule.
-   An enforce rule with operator {\tt :<} (alternative~\ref{enforce del}) keeps the population of $r$ smaller than or equal to \id{expression}.
+   An enforcement rule with operator {\tt :<} (alternative~\ref{enforce del}) keeps the population of $r$ smaller than or equal to \id{expression}.
    When the population of this expression is losing pairs that are in $r$, the engine will remove such pairs from $r$ to keep $r$ smaller than or equal to \id{expression}.
-   An enforce rule with operator {\tt >:} (alternative~\ref{enforce ins}) keeps the population of $r$ larger than or equal to \id{expression}.
+   An enforcement rule with operator {\tt >:} (alternative~\ref{enforce ins}) keeps the population of $r$ larger than or equal to \id{expression}.
    When the population of this expression is growing with pairs that are not in $r$, the engine will insert such pairs in $r$ to keep $r$ greater than or equal to \id{expression}.
-   An enforce rule with operator {\tt :=} (alternative~\ref{enforce eq}) keeps the population of $r$ equal to \id{expression} by inserting and deleting the right pairs in $r$.
-   Summarizing, an enforce rule can be read as an invariant that keeps one of $r\leq\id{expression}$, $r=\id{expression}$, or $r\geq\id{expression}$ satisfied by changing the population of $r$.
+   An enforcement rule with operator {\tt :=} (alternative~\ref{enforce eq}) keeps the population of $r$ equal to \id{expression} by inserting and deleting the right pairs in $r$.
+   Summarizing, an enforcement rule can be read as an invariant that keeps one of $r\leq\id{expression}$, $r=\id{expression}$, or $r\geq\id{expression}$ satisfied by changing the population of $r$.
 

### 2023-06-28 2022 Migration/articleMigration.tex
WIP introduction

@@
 \begin{abstract}
-   Incremental software deployment is meant to reduce the time to deploy changes.
-   However, an increment might involve a change in the database schema.
-   This poses the challenge to preserve already deployed data.
-   The problem is how to preserve the semantics of that data under a changing schema.
+   Software generators can help to increase the frequency of releases and their reliability.
+   They save on time spent on development and time spent on fixing human-induced mistakes by compiling a specification into a working information system.
+   However, many generators do not support data migrations, which are necessary when the schema of a system changes during its operational life.
+   As a consequence, developers often migrate data ``by hand'', which slows down the pace of incremental software development.
+
+   To address this problem, this paper proposes a theory for data migrations.
+   The problem at large is how to preserve the semantics of that data under a changing schema.
    This paper proposes a theory for deploying an incremental change,
@@
 
-   This paper aims to preserve the semantics of data, satisfy elementary business requirements,
-   and featuring a zero down-time option after deployment.
+   This paper aims to preserve the semantics of data by satisfying concrete business requirements.
    We derive a migration schema from both the existing system and a changed schema that specifies the desired system.
@@
    data meaning must be preserved,
-   the business continues during the migration without interruption,
+   the business continues during the migration without interruption (zero down-time),
    and there is a compiler to generate an information system from a given schema.
-   The correctness of the migration is the focus of this paper, while efficiency is beyond its scope.
+   The correctness of the migration is the focus of this paper, while efficiency is outside its scope.
 \end{abstract}
@@
    This paper is founded in the belief that generating systems can prevent human programming errors, which contributes to robust software.
-   Besides, information systems should be released into production in small increments.
-   That yields shorter release cycles, more frequent releases, and thus a more agile software development process%
+   Developers can use their efforts saved by generating systems to decrease the time to release successive increments.
+   More frequent data migrations are then to be expected.
+   A problem is that software generators typically do not support the data migration process, especially when the schema changes.
+   With such generators at hand, we can expect developers to spend a larger portion of their time on data migrations,
+   which limits the pace of deploying successive releases%
 \footnote{These effects can be measured in terms of DevOps metrics such as
@@
    change failure rate~\cite{DevOps2021}.}.
-   By deploying frequently, in small increments, and without down-time,
-   users get a system that evolves organically.
+
+   This paper proposes a theory for data migrations that is meant to be implemented in a software generator called Ampersand~\cite{Joosten-JLAMP2018, JoostenRAMiCS2017}.
+   It aims at automating data migrations, so incremental changes can be done more reliably and faster.
+   This contributes to a more agile software development process.
+   We believe that users will experience a more organic evolution of the system because increments are smaller, occur more frequently and more reliably.
    This requires a robust data migration process in which the deployment succeeds without failure.
-   Automation is required too, to prevent human errors in the process.
-   As a result, we expect that migration engineers can deploy smaller increments more frequently, which contributes to a more agile software development process.
 
@@
 \label{sct:Analysis}
+\subsection{Information Systems}
    The purpose of an information system is to make data meaningful to its users.
@@
 
-   This work uses the Ampersand compiler~\cite{Joosten-JLAMP2018, JoostenRAMiCS2017} to generate information systems.
-   It uses a syntactically sugared form of heterogeneous relation algebra,
+   Like most other software generators, Ampersand~\cite{Joosten-JLAMP2018, JoostenRAMiCS2017}
+   is a compiler that generates information systems from a script.
+   Once a generated system is in production, changes that affect the schema are expensive because developers must migrate data ``by hand''.
+   This is an error-prone process which slows down the deployment of successive increments.
+
+   Ampersand uses a syntactically sugared form of heterogeneous relation algebra,
    which works with relations in a way similar to Alloy~\cite{Alloy2006} and allegories%
@@
 \footnote{The dotted lines in the diagrams show which set of rules is enforced.}.
-   So, the user spends no effort to extract constraints from specifications or (even more laborious) from code
+   So, developers spend no effort to extract constraints from specifications or (even more laborious) from code
    because all constraints are explicitly available as rules in the code.
+
    This, and the absence of imperative code in an Ampersand script, makes Ampersand a suitable platform on which to implement the theory.
@@
    which means that a classical database schema (i.e.\ data structure plus semantics) can be extracted from the Ampersand script.
-   For the purpose of this paper an Ampersand script is considered equivelant to the schema of the generated system.
-
+
+\subsection{Data Migrations}
    Data migration occurs when a desired system replaces an existing one,
@@
 
-   Complications may arise as a result of data pollution,
-   which may require human interventions to resolve.
-   Data pollution is distinguished in the following categories:
+   In practice, data migrations are done by deploying the desired system and the existing system side-by-side,
+   while transferring data in a controlled fashion.
+   A migration strategy (e.g. case by case, customer by customer, or in batches) must ensure the quality of the migration.
+   To assume that this can be done fully automatically is unrealistic in many cases%
+\footnote{Note that smaller increments improve the chance that a fully automic migration is possible,
+although there will always be cases where human intervention is required.}.
+   Manual interventions can be necessary to resolve data pollution, new business rules, or to fix known issues in the old system.
+   Consequently, a migration engineer must have the freedom to change the migration script to suit the particulars of the migration.
+   Such complications may arise mostly as a result of data pollution.
+   We distinguish data pollution in the following categories:
 \begin{itemize}
@@
 \begin{enumerate}
-   \item the moment of transition (the MoT), i.e. the moment the desired system is made available to users;
-   \item the moment of cleanup (the MoC), i.e. the moment that the old software and data are removed.%
+   \item the moment of transition (the MoT), i.e. the moment that changed functionality is made available to users;
+   \item the moment of cleanup (the MoC), i.e. the moment after which the old software and data can be removed.%
 \footnote{Backup mechanisms are outside the scope of this paper.}.
@@
 
-   In the following section introduces the definitions required to migrate data from one system to another.
+   The following section introduces the definitions required to migrate data from one system to another.
 
@@
    Deployment of the migration system (figure~\ref{fig:during}) leaves the existing system intact.
-   It enhances the system with the data set needed by the migration system, which includes the data set needed by the new system.
+   It enhances the system with the data set needed by the migration system, which includes the data set needed by the desired system.
    So, users will notice no difference in the existing system when the migration system is deployed.
@@
    At this point, the ingress switches traffic from the existing system to the migration system.
-   Until this moment, users have been updating data in the existing system, which is continuously being migrated to the new system.
-   After the ingress switch is made, users have the new functionality at their disposal and make their changes in the new system.
+   Until this moment, users have been updating data in the existing system, which is continuously being migrated to the desired system.
+   After the ingress switch is made, users have the new functionality at their disposal and make their changes in the desired system.
    If for some reason, there is remaining migration work, the migration system will execute this.

### 2023-06-28 2022 Migration/articleMigration.tex
Minor

@@
    \item Let the existing information system $\infsys$ have data set $\dataset$ and schema $\schema$.
-         A developer defines a new information system, $\infsys'$, which contains a data set of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.} as illustrated in figure~\ref{fig:post-migration}.
+         A developer defines a new information system, $\infsys'$, which contains a data set of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{In this paper, we only consider the functionality captured by rules.} as illustrated in figure~\ref{fig:post-migration}.
          After the migration, the resulting system will be $\infsys'$, enhanced with the triples from $\dataset$ that satisfy $\schema'$,
@@
          It also contains data structures and rules that are needed for the duration of the migration phase.
-         The generated script must be free of type errors, to ensure that $\sat{u}{\dataset}$ exists for every rule $u$ and data set $\dataset$.
-         The proof is available \href{https://www.isa-afp.org/}{here}
+%         The generated script must be free of type errors, to ensure that $\sat{u}{\dataset}$ exists for every rule $u$ and data set $\dataset$.
+%         The proof is available \href{https://www.isa-afp.org/}{here}
    \item\label{item3}
@@
 
-\subsection{Formalization}
-   Let us define the instruments needed to describe the derivation of $\schema_{\migrsys}$.
-\begin{definition}[union of data sets]
-\label{def:graph_union_wellTyped_if_parts_wellTyped}
-    \[\pair{\triples_1}{\inst_1}\cup\pair{\triples_2}{\inst_2}\ =\ \pair{\triples_1\cup\triples_2}{\inst_1\cup\inst_2}\]
-\end{definition}
-
-\begin{lemma}
-\label{lemma:graph_union_wellTyped_if_parts_wellTyped}
-   If $\dataset_1$ and $\dataset_2$ are data sets, then $\dataset_1\cup\dataset_2$ is a data set.
-\end{lemma}
-
-\begin{definition}[map relation names in data sets]
-\label{def:map_labels_preserves_wellTypedness}
-   Let $h:\Rels\rightarrow\Rels$, then
-   \[\begin{array}{rcl}
-      \maprel{h}{\pair{\triples}{\inst}}&=&\pair{\maprel{h}{\triples}}{\inst}\\
-      \maprel{h}{\triple{a}{r}{b}}&=&\triple{a}{h(r)}{b}
-   \end{array}\]
-\end{definition}
-
-\begin{lemma}
-\label{lemma:map_labels_preserves_wellTypedness}
-   If $\dataset$ is a data set and $h:\Rels\rightarrow\Rels$ is an injective function, then $\maprel{h}{\dataset}$ is a data set.
-\end{lemma}
-
-\begin{definition}[disjoint union of data sets]
-\begin{eqnarray}
-   \pair{\triples}{\inst}\sqcup\pair{\triples'}{\inst'}&=&\pair{\triples\uplus\triples'}{\inst\uplus^2\inst'}\\
-      X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
-      X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}
-\end{eqnarray}
-\end{definition}
-   In a disjoint union, the instances of the two data sets are relabeled to avoid name clashes.
-   In practice, we will not use 0 and 1 as labels for $x$ and $y$, but rather extend their names by prefixing.
-   For example, \texttt{old.Person} and \texttt{new.Address} might be the names of concepts \texttt{Person} and \texttt{Address} from contexts \texttt{old} and \texttt{new}, respectively.
-\begin{lemma}
-   If $\dataset$ and $\dataset'$ are data sets, then so is $\dataset\sqcup\dataset'$.
-\end{lemma}
-   This lemma is being proven in Isabelle/HOL~\cite{Isabelle}. The proof is published \href{location.domain}{here}
-
-\begin{definition}[disjoint union of schemas]
-\begin{eqnarray}
-   \triple{\concepts}{\rels}{\rules}\sqcup\triple{\concepts'}{\rels'}{\rules'}&=&\triple{\concepts\uplus\concepts}{\rels\uplus\rels'}{\rules\uplus\rules'}
-\end{eqnarray}
-\end{definition}
-% \begin{definition}[disjoint union of data sets]
+%SJC on Formalization: it's often better to put these inline where we first use them. Leaving this as a comment for easy copy-paste.
+%\subsection{Formalization}
+%   Let us define the instruments needed to describe the derivation of $\schema_{\migrsys}$.
+%\begin{definition}[union of data sets]
+%\label{def:graph_union_wellTyped_if_parts_wellTyped}
+%    \[\pair{\triples_1}{\inst_1}\cup\pair{\triples_2}{\inst_2}\ =\ \pair{\triples_1\cup\triples_2}{\inst_1\cup\inst_2}\]
+%\end{definition}
+%
+%\begin{lemma}
+%\label{lemma:graph_union_wellTyped_if_parts_wellTyped}
+%   If $\dataset_1$ and $\dataset_2$ are data sets, then $\dataset_1\cup\dataset_2$ is a data set.
+%\end{lemma}
+%
+%\begin{definition}[map relation names in data sets]
+%\label{def:map_labels_preserves_wellTypedness}
+%   Let $h:\Rels\rightarrow\Rels$, then
+%   \[\begin{array}{rcl}
+%      \maprel{h}{\pair{\triples}{\inst}}&=&\pair{\maprel{h}{\triples}}{\inst}\\
+%      \maprel{h}{\triple{a}{r}{b}}&=&\triple{a}{h(r)}{b}
+%   \end{array}\]
+%\end{definition}
+%
+%\begin{lemma}
+%\label{lemma:map_labels_preserves_wellTypedness}
+%   If $\dataset$ is a data set and $h:\Rels\rightarrow\Rels$ is an injective function, then $\maprel{h}{\dataset}$ is a data set.
+%\end{lemma}
+%
+%\begin{definition}[disjoint union of data sets]
+%\begin{eqnarray}
+%   \pair{\triples}{\inst}\sqcup\pair{\triples'}{\inst'}&=&\pair{\triples\uplus\triples'}{\inst\uplus^2\inst'}\\
+%      X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
+%      X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}
+%\end{eqnarray}
+%\end{definition}
+%   In a disjoint union, the instances of the two data sets are relabeled to avoid name clashes.
+%   In practice, we will not use 0 and 1 as labels for $x$ and $y$, but rather extend their names by prefixing.
+%   For example, \texttt{old.Person} and \texttt{new.Address} might be the names of concepts \texttt{Person} and \texttt{Address} from contexts \texttt{old} and \texttt{new}, respectively.
+%\begin{lemma}
+%   If $\dataset$ and $\dataset'$ are data sets, then so is $\dataset\sqcup\dataset'$.
+%\end{lemma}
+%   This lemma is proven in Isabelle/HOL~\cite{Isabelle}. % TODO: point to formalization %The proof is published \href{location.domain}{here}
+%
+%\begin{definition}[disjoint union of schemas]
+%\begin{eqnarray}
+%   \triple{\concepts}{\rels}{\rules}\sqcup\triple{\concepts'}{\rels'}{\rules'}&=&\triple{\concepts\uplus\concepts}{\rels\uplus\rels'}{\rules\uplus\rules'}
+%\end{eqnarray}
+%\end{definition}
+%% \begin{definition}[disjoint union of data sets]
+%% \begin{eqnarray}
+%%    \omit\rlap{$\la\atoms,\concepts,\inst,\isa,\rels,\dataset\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\dataset'\ra$}\notag\\
+%%    &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\instuplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \dataset\uplus^5\dataset'\ra\notag\\
+%%       X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
+%%       X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}\\
+%%       X\uplus^3 Y&=&\begin{array}[t]{@{}l}\{\declare{(n,0)}{(A,0)}{(B,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\declare{(n,1)}{(A,1)}{(B,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}\\
+%%       X\uplus^5 Y&=&\begin{array}[t]{@{}l}\{\triple{(a,0)}{\declare{(n,0)}{(A,0)}{(B,0)}}{(b,0)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in X\}\ \cup\\ \{\triple{(a,1)}{\declare{(n,1)}{(A,1)}{(B,1)}}{(b,1)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in Y\}\end{array}
+%% \end{eqnarray}
+%% \end{definition}
+%\begin{lemma}
+%   If $\schema$ and $\schema'$ are schemas, then so is $\schema\sqcup\schema'$.
+%\end{lemma}
+%The proof of this lemma is published \href{location.domain}{here}
+%
+%To implement the disjoint union, the following relabelling functions are needed:
+%\begin{definition}[relabel concepts]
+%   \[\subst{C}{D}{A}\ =\ \text{\bf if}\ C=A\ \text{\bf then}\ D\ \text{\bf else}\ A\]
+%\end{definition}
+%\begin{definition}[relabel concepts in triples]
+%   \[\subst{C}{D}{\triple{a}{r}{b}} = \triple{a}{\subst{C}{D}{r}}{b}\]
+%\end{definition}
+%\begin{definition}[relabel concepts in relations]
+%   \[\subst{C}{D}{(\declare{n}{A}{B})} = \declare{n}{\subst{C}{D}{A}}{\subst{C}{D}{B}}\]
+%\end{definition}
+%\begin{definition}[relabel concepts in $\inst$]
+%   \[a\ \subst{C}{D}{\inst}\ \subst{C}{D}{A}\ \Leftrightarrow\ a\inst A\]
+%\end{definition}
+%\begin{definition}[relabel concepts in data sets]
+%   \[\begin{array}{rcl}
+%      \rlap{$\subst{C}{D}{\pair{\triples}{\inst}}$}\\
+%      &=&\pair{\{\subst{C}{D}{t}\mid t\in\triples\}}{\subst{C}{D}{\inst}}
+%   \end{array}\]
+%\end{definition}
+%   The relabeling of concepts in rules must satisfy:
+%\begin{equation}
+%   \sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{C}{D}{u}}{\subst{C}{D}{\dataset}}
+%\end{equation}
+%\begin{definition}[relabel concepts in schemas]
 %   \begin{eqnarray}
-%    \omit\rlap{$\la\atoms,\concepts,\inst,\isa,\rels,\dataset\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\dataset'\ra$}\notag\\
-%    &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\instuplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \dataset\uplus^5\dataset'\ra\notag\\
-%       X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
-%       X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}\\
-%       X\uplus^3 Y&=&\begin{array}[t]{@{}l}\{\declare{(n,0)}{(A,0)}{(B,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\declare{(n,1)}{(A,1)}{(B,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}\\
-%       X\uplus^5 Y&=&\begin{array}[t]{@{}l}\{\triple{(a,0)}{\declare{(n,0)}{(A,0)}{(B,0)}}{(b,0)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in X\}\ \cup\\ \{\triple{(a,1)}{\declare{(n,1)}{(A,1)}{(B,1)}}{(b,1)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in Y\}\end{array}
+%      \rlap{$\subst{C}{D}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
+%      &=&\begin{array}[t]{l@{}l}
+%         \la&\{\subst{C}{D}{c}\mid c\in\concepts\}\\
+%         ,&\{\subst{C}{D}{r}\mid r\in\rels\}\\
+%         ,&\{\subst{C}{D}{u}\mid u\in\rules\}\ \ra\notag
+%         \end{array}
 %   \end{eqnarray}
-% \end{definition}
-\begin{lemma}
-   If $\schema$ and $\schema'$ are schemas, then so is $\schema\sqcup\schema'$.
-\end{lemma}
-The proof of this lemma is published \href{location.domain}{here}
-
-To implement the disjoint union, the following relabelling functions are needed:
-\begin{definition}[relabel concepts]
-   \[\subst{C}{D}{A}\ =\ \text{\bf if}\ C=A\ \text{\bf then}\ D\ \text{\bf else}\ A\]
-\end{definition}
-\begin{definition}[relabel concepts in triples]
-   \[\subst{C}{D}{\triple{a}{r}{b}} = \triple{a}{\subst{C}{D}{r}}{b}\]
-\end{definition}
-\begin{definition}[relabel concepts in relations]
-   \[\subst{C}{D}{(\declare{n}{A}{B})} = \declare{n}{\subst{C}{D}{A}}{\subst{C}{D}{B}}\]
-\end{definition}
-\begin{definition}[relabel concepts in $\inst$]
-   \[a\ \subst{C}{D}{\inst}\ \subst{C}{D}{A}\ \Leftrightarrow\ a\inst A\]
-\end{definition}
-\begin{definition}[relabel concepts in data sets]
-   \[\begin{array}{rcl}
-      \rlap{$\subst{C}{D}{\pair{\triples}{\inst}}$}\\
-      &=&\pair{\{\subst{C}{D}{t}\mid t\in\triples\}}{\subst{C}{D}{\inst}}
-   \end{array}\]
-\end{definition}
-   The relabeling of concepts in rules must satisfy:
-\begin{equation}
-   \sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{C}{D}{u}}{\subst{C}{D}{\dataset}}
-\end{equation}
-\begin{definition}[relabel concepts in schemas]
-   \begin{eqnarray}
-      \rlap{$\subst{C}{D}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
-      &=&\begin{array}[t]{l@{}l}
-         \la&\{\subst{C}{D}{c}\mid c\in\concepts\}\\
-         ,&\{\subst{C}{D}{r}\mid r\in\rels\}\\
-         ,&\{\subst{C}{D}{u}\mid u\in\rules\}\ \ra\notag
-         \end{array}
-   \end{eqnarray}
-\end{definition}
-\begin{definition}[relabel relations]
-   \[\subst{\declare{m}{C}{D}}{m'}{{\declare{n}{A}{B}}}\ =\ \text{\bf if}\ n=m\wedge A=C\wedge B=D\ \text{\bf then}\ \declare{m'}{A}{B}\ \text{\bf else}\ \declare{m}{A}{B}\]
-\end{definition}
-\begin{definition}[relabel relations in triples]
-   \[\subst{r'}{m'}{\triple{a}{r}{b}} = \triple{a}{\subst{r'}{m'}{r}}{b}\]
-\end{definition}
-\begin{definition}[relabel relations in data sets]
-   \[\begin{array}{rcl}
-      \rlap{$\subst{r'}{m'}{\pair{\triples}{\inst}}$}\\
-      &=&\pair{\{\subst{r'}{m'}{t}\mid t\in\triples\}}{\inst}
-   \end{array}\]
-\end{definition}
-   The relabeling of relation names in rules must satisfy:
-\begin{equation}
-   \sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{r'}{m'}{u}}{\subst{r'}{m'}{\dataset}}
-\end{equation}
-\begin{definition}[relabel relations in schemas]
-   \begin{eqnarray}
-      \rlap{$\subst{r'}{m'}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
-      &=&\triple{\concepts}{\{\subst{r'}{m'}{r} \mid r\in\rels\}}{\{\subst{r'}{m'}{u} \mid u\in\rules\}}\notag
-   \end{eqnarray}
-\end{definition}
-
-   The relabeling of concepts is not trivial because of specialization.
-   To illustrate this point, suppose $D\isa A$.
-   Then the substitution $[C\rightarrow D]$ means that every atom $a$ that used to be of concept $C$ in the existing system, is now not only a $D$ but also an $A$.
-   For example, by relabeling the concept {\tt Hotel} to {\tt Motel}, in a situation where ${\tt Motel}\isa{\tt Parking}$,
-   all atoms that represent hotels might suddenly be expected to have parking lots.
-   So, whereas such examples may fly in a technical sense, the developer must use renaming with care to preserve the intended semantics.
+%\end{definition}
+%\begin{definition}[relabel relations]
+%   \[\subst{\declare{m}{C}{D}}{m'}{{\declare{n}{A}{B}}}\ =\ \text{\bf if}\ n=m\wedge A=C\wedge B=D\ \text{\bf then}\ \declare{m'}{A}{B}\ \text{\bf else}\ \declare{m}{A}{B}\]
+%\end{definition}
+%\begin{definition}[relabel relations in triples]
+%   \[\subst{r'}{m'}{\triple{a}{r}{b}} = \triple{a}{\subst{r'}{m'}{r}}{b}\]
+%\end{definition}
+%\begin{definition}[relabel relations in data sets]
+%   \[\begin{array}{rcl}
+%      \rlap{$\subst{r'}{m'}{\pair{\triples}{\inst}}$}\\
+%      &=&\pair{\{\subst{r'}{m'}{t}\mid t\in\triples\}}{\inst}
+%   \end{array}\]
+%\end{definition}
+%   The relabeling of relation names in rules must satisfy:
+%\begin{equation}
+%   \sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{r'}{m'}{u}}{\subst{r'}{m'}{\dataset}}
+%\end{equation}
+%\begin{definition}[relabel relations in schemas]
+%   \begin{eqnarray}
+%      \rlap{$\subst{r'}{m'}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
+%      &=&\triple{\concepts}{\{\subst{r'}{m'}{r} \mid r\in\rels\}}{\{\subst{r'}{m'}{u} \mid u\in\rules\}}\notag
+%   \end{eqnarray}
+%\end{definition}
+%   
+%   The relabeling of concepts is not trivial because of specialization.
+%   To illustrate this point, suppose $D\isa A$.
+%   Then the substitution $[C\rightarrow D]$ means that every atom $a$ that used to be of concept $C$ in the existing system, is now not only a $D$ but also an $A$.
+%   For example, by relabeling the concept {\tt Hotel} to {\tt Motel}, in a situation where ${\tt Motel}\isa{\tt Parking}$,
+%   all atoms that represent hotels might suddenly be expected to have parking lots.
+%   So, whereas such examples may fly in a technical sense, the developer must use renaming with care to preserve the intended semantics.
 
@@
 
-   ROLE MigrationEngineer MAINTAINS migrating_takes,
-                                    migrating_isPartOf,
-                                    migrating_isEnrolledFor
-   RULE migrating_takes : old.takes-copied_takes = -V
-   RULE migrating_isPartOf : old.isPartOf-copied_isPartOf = -V
-   RULE migrating_isEnrolledFor :
-        old.isEnrolledFor-copied_isEnrolledFor = -V
+   RULE ExamRule1: new.student~;new.course |- new.takes
+   RULE ExamRule2: new.student~;new.course;new.isPartOf~ 
+                 |- new.isEnrolledFor
 \end{verbatim}

### 2023-06-28 2022Migration/articleMigration.tex
elaborate migration example

--- 2023-06-28 2022 Migration/articleMigration.tex
@@
 
-\title{Data migration under a Changing Schema}
+\title{Data Migration under a Changing Schema}
 \author[ou,ordina]{Stef Joosten\fnref{fn1}}
@@
    They save on time spent on development and time spent on fixing human-induced mistakes by compiling a specification into a working information system.
-   However, many generators do not support data migrations, which are necessary when the schema of a system changes during its operational life.
-   As a consequence, developers often migrate data ``by hand'', which slows down the pace of incremental software development.
-
-   To address this problem, this paper proposes a theory for data migrations.
+   However, many generators do not support data migrations.
+   A data migration is necessary when an incremental deployment changes the schema of the system.
+   As a consequence, developers tend to avoid migrations or migrate data ``by hand''.
+
+   To address this problem, this paper proposes a theory for data migrations aimed at automating the migration process.
    The problem at large is how to preserve the semantics of that data under a changing schema.
-   This paper proposes a theory for deploying an incremental change,
-   meant to be implemented in a software generator called Ampersand.
+   This paper proposes a theory for deploying an incremental change.
+   The theory is applicable in general, but will be implemented in a software generator called Ampersand.
 
    This paper aims to preserve the semantics of data by satisfying concrete business requirements.
-   We derive a migration schema from both the existing system and a changed schema that specifies the desired system.
-   This work is based on the assumptions that
-   the migration is meant to deploy a software increment in production,
+   The migration process is based on the assumptions that
+   software is deployed incrementally,
    the existing data set may be polluted,
    human interaction may be required,
-   data meaning must be preserved,
+   the meaning of data must be preserved,
    the business continues during the migration without interruption (zero down-time),
@@
 
-   This paper is founded in the belief that generating systems can prevent human programming errors, which contributes to robust software.
-   Developers can use their efforts saved by generating systems to decrease the time to release successive increments.
-   More frequent data migrations are then to be expected.
+   This paper is founded in the belief that software generators help developers to increase the frequency of deploying successive increments.
+   They also prevent some human-induced errors, both in the process and in the resulting software.
+   We believe that software generation can thus lead to a higher pace of deployments, a more predictable deployment process, and more reliable information systems.
+
    A problem is that software generators typically do not support the data migration process, especially when the schema changes.
-   With such generators at hand, we can expect developers to spend a larger portion of their time on data migrations,
-   which limits the pace of deploying successive releases%
+   The ``manual'' effort required to migrate data slows down the development process and introduces mistakes%
 \footnote{These effects can be measured in terms of DevOps metrics such as
@@
 
-   This paper proposes a theory for data migrations that is meant to be implemented in a software generator called Ampersand~\cite{Joosten-JLAMP2018, JoostenRAMiCS2017}.
+   This paper proposes a theory for data migrations that is meant to be implemented in a software generator called Ampersand~\cite{JoostenRAMiCS2017, Joosten-JLAMP2018}.
    It aims at automating data migrations, so incremental changes can be done more reliably and faster.
@@
    We believe that users will experience a more organic evolution of the system because increments are smaller, occur more frequently and more reliably.
-   This requires a robust data migration process in which the deployment succeeds without failure.
-
-   This paper assumes the the existing system and the desired system have different schemas,
-   which occurs in incremental software deployment.
+
+   This paper assumes that the existing system and the desired system have different schemas.
    The existing system is the system that is already deployed in production,
    which makes it necessary to preserve the existing data.
+   The desired system contains a number of changes and is meant to replace the existing system.
 
@@
    Examples like these have convinced the authors to define a new method for data migration that is specifically meant for systems in production that are
-   deployed incrementally.
+   deployed in increments.
 
    The contribution of this paper is to derive a migration schema from two artifacts: an existing system
-   (which has its own schema and data set) and a new schema (which specifies the desired system).
-   The purpose of our theory is to develop tools.
-   These tools must satisfy such practical requirements as
-   zero down-time updates and ways to cope with data pollution.
-   Zero down-time means that all human activity for the data migration can be done without taking the system out of production.
-   In practice, more frequent updates require that updates come with zero down-time.
-   Any delay caused by automatically switching from the existing system to the desired is considered negligible.
-   This assumes a modern deployment platform (such as Kubernetes) that updates applications with zero down-time.
-
+   (which has its own schema and data set) and the specification of a desired system (which has a different schema and will ``import'' relevant parts of the existing data).   
    Our theory is based on the following assumptions and requirements:
 \begin{itemize}
-   \item the data migration is meant to deploy a software increment in production;
-   \item the existing data set may be polluted, but it satisfies its schema;
-   \item the data migration may require human interaction, which may take time;
-   \item the meaning of data must be preserved;
-   \item the business continues during the migration without interruption;
-   \item there is a compiler to generate an information system from a given schema.
+   \item {\em incremental deployment}\\The data migration is meant to deploy a software increment in production.
+   \item {\em data pollution}\\The existing data set may be polluted, but it satisfies its schema.
+   \item {\em human intervention}\\The data migration may require human interaction, which may take time.
+   \item {\em semantic continuity}\\the meaning of data must be preserved.
+   \item {\em zero down-time}\\The business continues during the migration without interruption.
+   \item {\em generative software development}\\A compiler exists to generate an information system from a given schema.
 \end{itemize}
@@
 
-   Users have their own tasks and responsibilities
+   Every user has her\footnote{Throughout this paper, we ask the reader to read ``she'' or ``her'' to include the masculin form as well.} own tasks and responsibilities
    and may work from different locations and on different moments.
@@
    meaning that each service can be stopped, (re)started and substituted without disrupting the other services.
+   For practical purposes, this presumes a platform of managed services, such as Kubernetes,
+   even though the theory in this paper is not specific to any platform.
 
@@
 
-   Like most other software generators, Ampersand~\cite{Joosten-JLAMP2018, JoostenRAMiCS2017}
-   is a compiler that generates information systems from a script.
-   Once a generated system is in production, changes that affect the schema are expensive because developers must migrate data ``by hand''.
-   This is an error-prone process which slows down the deployment of successive increments.
-
+   Like most other software generators, Ampersand is a compiler that generates information systems from a script.
    Ampersand uses a syntactically sugared form of heterogeneous relation algebra,
@@
    while transferring data in a controlled fashion.
-   A migration strategy (e.g. case by case, customer by customer, or in batches) must ensure the quality of the migration.
-   To assume that this can be done fully automatically is unrealistic in many cases%
+   Typically, a migration strategy (e.g. case by case, customer by customer, or in batches) is in place
+   to ensure the preservation of valuable data and the quality of the migration.
+   To automate this process completely is unrealistic in many cases%
 \footnote{Note that smaller increments improve the chance that a fully automic migration is possible,
@@
    The relation $\inst:\Pair{\Atoms}{\Concepts}$ relates atoms to concepts.
-   The expression $a\inst C$ means that atom $a$ is an \emph{instance} of concept $C$.
+   The term $a\inst C$ means that atom $a$ is an \emph{instance} of concept $C$.
    This relation is used in the type system, in which $\inst$ assigns one or more concepts to every atom in the data set.
@@
 \end{eqnarray}
+   In English, we would say that an event $\evt{d^-}{d^+}$ removes the triples in $d^-$ from $\dataset$ and then inserts the triples in $d^+$ into the result.
    In the sequel, variables $e$, $e'$, $e''$, etc. represent events.
@@
    A rule, {\tt EnrollRule} completes the script.
-   It states that a student can enroll for any module that is part of a course he or she takes.
+   It states that a student can enroll for any module that is part of a course she takes.
    % In Ampersand, which is a syntactically sugared form of relation algebra~\cite{JoostenRAMiCS2017},
@@
    In practice, we will not use 0 and 1 as labels for $x$ and $y$, but rather extend their names by prefixing.
-   For example, \texttt{old.Person} and \texttt{new.Address} might be the names of concepts \texttt{Person} and \texttt{Address} from contexts \texttt{old} and \texttt{new}, respectively.
 \begin{lemma}
@@
 \begin{eqnarray}
-   \text{\tt ENFORCE}\ r\ \text{\tt :<}\ \id{expression}\label{enforce del}\\
-   \text{\tt ENFORCE}\ r\ \text{\tt :=}\ \id{expression}\label{enforce eq}\\
-   \text{\tt ENFORCE}\ r\ \text{\tt >:}\ \id{expression}\label{enforce ins}
+   \text{\tt ENFORCE}\ r\ \text{\tt :<}\ \id{term}\label{enforce del}\\
+   \text{\tt ENFORCE}\ r\ \text{\tt :=}\ \id{term}\label{enforce eq}\\
+   \text{\tt ENFORCE}\ r\ \text{\tt >:}\ \id{term}\label{enforce ins}
 \end{eqnarray}
@@
    The engine will change the population of relation $r$ by the least set of pairs needed to satisfy the rule.
-   An enforcement rule with operator {\tt :<} (alternative~\ref{enforce del}) keeps the population of $r$ smaller than or equal to \id{expression}.
-   When the population of this expression is losing pairs that are in $r$, the engine will remove such pairs from $r$ to keep $r$ smaller than or equal to \id{expression}.
-   An enforcement rule with operator {\tt >:} (alternative~\ref{enforce ins}) keeps the population of $r$ larger than or equal to \id{expression}.
-   When the population of this expression is growing with pairs that are not in $r$, the engine will insert such pairs in $r$ to keep $r$ greater than or equal to \id{expression}.
-   An enforcement rule with operator {\tt :=} (alternative~\ref{enforce eq}) keeps the population of $r$ equal to \id{expression} by inserting and deleting the right pairs in $r$.
-   Summarizing, an enforcement rule can be read as an invariant that keeps one of $r\leq\id{expression}$, $r=\id{expression}$, or $r\geq\id{expression}$ satisfied by changing the population of $r$.
+   An enforcement rule with operator {\tt :<} (alternative~\ref{enforce del}) keeps the population of $r$ smaller than or equal to \id{term}.
+   When the population of this term is losing pairs that are in $r$, the engine will remove such pairs from $r$ to keep $r$ smaller than or equal to \id{term}.
+   An enforcement rule with operator {\tt >:} (alternative~\ref{enforce ins}) keeps the population of $r$ larger than or equal to \id{term}.
+   When the population of this term is growing with pairs that are not in $r$, the engine will insert such pairs in $r$ to keep $r$ greater than or equal to \id{term}.
+   An enforcement rule with operator {\tt :=} (alternative~\ref{enforce eq}) keeps the population of $r$ equal to \id{term} by inserting and deleting the right pairs in $r$.
+   Summarizing, an enforcement rule can be read as an invariant that keeps one of $r\leq\id{term}$, $r=\id{term}$, or $r\geq\id{term}$ satisfied by changing the population of $r$.
 
@@
 
-   This would mean, however, that all pairs from {\tt old.takes} will be in {\tt new.takes} forever.
+   This means, however, that all pairs from {\tt old.takes} will be in {\tt new.takes} forever.
+   This is undesirable because data in the existing system may still change.
    An extra relation (\ref{extraRelation}) fixes this problem by remembering which pairs have been copied:
@@
    Then, the engine starts filling {\tt copied\_takes} with pairs from {\tt new.takes} because of rule~\ref*{fill copies with new.takes}.
-   As a result, the difference between {\tt old.takes} and {\tt copied\_takes} gets smaller and smaller, until {\tt old.takes} and {\tt copied\_takes} are equal.
+   As a result, the difference between {\tt old.takes} and {\tt copied\_takes} gets smaller and smaller.
+   Once {\tt old.takes} and {\tt copied\_takes} are equal, rule~\ref*{difference} will no longer insert pairs in {\tt new.takes}.
    While this copying process is going on, {\tt old.takes} may accumulate more pairs or lose pairs.
-\begin{itemize}
-\item If pair $p$ is inserted in {\tt old.takes} and $p$ is not in {\tt copied\_takes},
-      then rule~\ref{difference} causes the engine to insert $p$ in {\tt new.takes}.
-      Subsequently, rule~\ref{fill copies with new.takes} causes the engine to insert $p$ in {\tt copied\_takes} as well.
-\item If pair $p$ is inserted in {\tt old.takes} and $p$ is already in {\tt copied\_takes},
-      then nothing happens.
-\item If pair $p$ is deleted from {\tt old.takes} and $p$ is in {\tt copied\_takes},
-      then rule~\ref{del copies from old.takes} causes the engine to delete $p$ from {\tt copied\_takes}.
-\item If pair $p$ is deleted from {\tt old.takes} and $p$ is not in {\tt copied\_takes},
-      then nothing happens.
-\item If pair $p$ is inserted in {\tt new.takes},
-      rule~\ref{fill copies with new.takes} causes the engine to insert $p$ into {\tt copied\_takes}.
-\item If pair $p$ is deleted from {\tt new.takes},
-      rule~\ref{fill copies with new.takes} causes the engine to delete $p$ from {\tt copied\_takes}.
-      If this pair is in {\tt old.takes}, rule~\ref{difference} causes the engine to insert $p$ back into {\tt new.takes}.
-\end{itemize}
+
+% Het volgende is wellicht iets teveel van het goede. Het lijkt me dat de lezer het voorgaande al wel snapt.
+%    The following table summarizes what happens when pairs are inserted or deleted from {\tt old.takes} and {\tt new.takes}.
+% \begin{itemize}
+% \item If pair $p$ is inserted in {\tt old.takes} and $p$ is not in {\tt copied\_takes},
+%       then rule~\ref{difference} causes the engine to insert $p$ in {\tt new.takes}.
+%       Subsequently, rule~\ref{fill copies with new.takes} causes the engine to insert $p$ in {\tt copied\_takes} as well.
+% \item If pair $p$ is inserted in {\tt old.takes} and $p$ is already in {\tt copied\_takes},
+%       then nothing happens.
+% \item If pair $p$ is deleted from {\tt old.takes} and $p$ is in {\tt copied\_takes},
+%       then rule~\ref{del copies from old.takes} causes the engine to delete $p$ from {\tt copied\_takes}.
+% \item If pair $p$ is deleted from {\tt old.takes} and $p$ is not in {\tt copied\_takes},
+%       then nothing happens.
+% \item If pair $p$ is inserted in {\tt new.takes},
+%       rule~\ref{fill copies with new.takes} causes the engine to insert $p$ into {\tt copied\_takes}.
+% \item If pair $p$ is deleted from {\tt new.takes},
+%       rule~\ref{fill copies with new.takes} causes the engine to delete $p$ from {\tt copied\_takes}.
+%       If this pair is in {\tt old.takes}, rule~\ref{difference} causes the engine to insert $p$ back into {\tt new.takes}.
+% \end{itemize}
 
@@
 \end{array}\]
-   Both rules must be maintained by the student.
-   So, a student can register for an exam only if (s)he is taking the course for which (s)he registers and if (s)he is enrolled for every module of that course.
+   Both rules are invariants.
+   So, a student can register for an exam only if she is taking the course for which she registers and if she is enrolled for every module of that course.
    % Ampersand also derives the following violation sets:
@@
 \subsection{Migration script}
-   The generated migration script, $\migrsys$, is:
+   The generated migration script, $\migrsys$, consists of the schema of the existing system $\infsys$,
+   the migration system, and the desired system $\infsys'$.
+   The migration system copies the data of the existing system to the desired system by the following rules:
 \begin{verbatim}
+-- The existing system (everything is prefixed with "old.")
+   RELATION old.takes[Student*Course]
+   RELATION old.isPartOf[Module*Course]
+   RELATION old.isEnrolledFor[Student*Module]
+
+   RULE old.EnrollRule:
+      old.isEnrolledFor |- old.takes;old.isPartOf~
+
+-- Migration system
    RELATION copied_takes[Student*Course]
@@
 
-   ROLE MigrationEngineer MAINTAINS migrating_takes,
-                                    migrating_isPartOf,
-                                    migrating_isEnrolledFor
-   RULE migrating_takes : old.takes-copied_takes = -V
-   RULE migrating_isPartOf : old.isPartOf-copied_isPartOf = -V
-   RULE migrating_isEnrolledFor :
-        old.isEnrolledFor-copied_isEnrolledFor = -V
+-- The desired system (everything is prefixed with "new.")
+   RELATION new.takes[Student*Course]
+   RELATION new.isPartOf[Module*Course] [UNI]
+   RELATION new.isEnrolledFor[Student*Module]
+   RULE new.EnrollRule:
+      new.isEnrolledFor |- new.takes;new.isPartOf~
+
+   RELATION new.course[ExamReg*Course] [UNI] =
+      [ ("ER1", "Management")
+      ; ("ER2", "Business IT")
+      ; ("ER3", "Business IT")
+      ]
+   RELATION new.student[ExamReg*Student] [UNI] =
+      [ ("ER1", "Peter")
+      ; ("ER2", "Susan")
+      ]
+
+   RULE new.ExamRule1 :
+      new.student~;new.course |- new.takes
+   RULE new.ExamRule2 :
+      new.student~;new.course;new.isPartOf~ |- new.isEnrolledFor
 \end{verbatim}
@@
 \subsection{Preparing for the Moment of Transition}
-   Just before loading the desired system with data,
-   the data in the existing system has changed.
-   So, the data we saw in section~\ref{sct:Example existing IS} has most likely changed.
-   Let us assume the actual population at this time is:
+   Just before starting the migration,
+   the data in the existing system typically differs from the data at its inception (section~\ref{sct:Example existing IS}).
+   Let us assume the actual population at the start of the migration is:
 \begin{verbatim}
-   RELATION takes[Student*Course] =
+   POPULATION takes[Student*Course] CONTAINS
       [ ("Peter", "Management")
@@
       ]
-   RELATION isPartOf[Module*Course] [UNI]=
+   POPULATION isPartOf[Module*Course] CONTAINS
       [ ("Finance", "Management")
@@
       ]
-   RELATION isEnrolledFor [Student*Module] =
+   POPULATION isEnrolledFor [Student*Module] CONTAINS
       [ ("Susan", "Business Analytics")
@@
 
-   So let us now turn to the migration.
-   The intention of any migration is to preserve as much of the data from $\infsys$ as possible into $\infsys'$,
-   while adding the new triples to $\infsys'$.
-   So the first thing to do is to take the disjoint union of both systems.
-   This yields:
-\begin{verbatim}
-   RELATION takes[Student*Course] =
-      [ ("Peter", "Management")
-      ; ("Susan", "Business IT")
-      ; ("John", "Business IT")
-      ]
-
-   RELATION isPartOf[Module*Course] [UNI]=
-      [ ("Finance", "Management")
-      ; ("Business Rules", "Business IT")
-      ; ("Business Analytics", "Business IT")
-      ; ("IT-Governance", "Management")
-      ; ("IT-Governance", "Business IT")
-      ]
-
-   RELATION isEnrolledFor [Student*Module] =
-      [ ("Susan", "Business Analytics")
-      ; ("Susan", "IT-Governance")
-      ; ("Susan", "Business Rules")
-      ]
-
-   RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
-
-   RELATION course[ExamReg*Course] [UNI] =
-      [ ("ER1", "Management")
-      ; ("ER2", "Business IT")
-      ; ("ER3", "Business IT")
-      ]
-   RELATION student[ExamReg*Student] [UNI] =
-      [ ("ER1", "Peter")
-      ; ("ER2", "Susan")
-      ]
-   RULE ExamRule1: student~;course |- takes
-   RULE ExamRule2: student~;course;isPartOf~ |- isEnrolledFor
-\end{verbatim}
-   When Ampersand checks this script, it produces the following results:
-\begin{verbatim}
-   There are 2 violations of RULE "ExamRule2":
-      ("Peter", "IT-Governance")
-      ("Peter", "Finance")
-   ==============================
-   There is one violation of RULE "UNI isPartOf[Module*Course]":
-      ("IT-Governance", "IT-Governance")
-\end{verbatim}
-   Any data migration must anticipate any population in the existing system.
-   The only certainty is that the existing population satisfies the rules in $\infsys$.
+   Upon initialization, the migration system evaluates all enforce rules in an attempt to satisfy all invariants.
+   In this example, it leaves two violations behind that it cannot resolve by enforce rules.
+   The pair {\tt ("Peter", "Management")} is the first one; it violates rule {\tt new.ExamRule2}.
+   This is because Peter has registered for an examination of the course Management,
+   which contains the module Finance for which Peter is not enrolled.
+   The other is {\tt "IT-Governance"}, which violates the univalence of relation {\tt new.isPartOf}.
+   Since rule {\tt new.ExamRule2} and the univalence of {\tt new.isPartOf} are both defined as invariants,
+   the desired system cannot be taken into production until these violations are resolved.
+   Until that moment, the migration system remains operational to allow a user to resolve this violation. 
+   The moment of cleanup is marked by the event that all invariants of the Migration system and the desired system are satisfied.
+   At that moment, the migration system is no longer needed and can be removed.
 

### 2023-06-28 2022Migration/articleMigration.tex

@@
 
-\subsection{Formalization}
-   Let us define the instruments needed to describe the derivation of $\schema_{\migrsys}$.
-\begin{definition}[union of data sets]
-\label{def:graph_union_wellTyped_if_parts_wellTyped}
-    \[\pair{\triples_1}{\inst_1}\cup\pair{\triples_2}{\inst_2}\ =\ \pair{\triples_1\cup\triples_2}{\inst_1\cup\inst_2}\]
-\end{definition}
-
-\begin{lemma}
-\label{lemma:graph_union_wellTyped_if_parts_wellTyped}
-   If $\dataset_1$ and $\dataset_2$ are data sets, then $\dataset_1\cup\dataset_2$ is a data set.
-\end{lemma}
-
-\begin{definition}[map relation names in data sets]
-\label{def:map_labels_preserves_wellTypedness}
-   Let $h:\Rels\rightarrow\Rels$, then
-   \[\begin{array}{rcl}
-      \maprel{h}{\pair{\triples}{\inst}}&=&\pair{\maprel{h}{\triples}}{\inst}\\
-      \maprel{h}{\triple{a}{r}{b}}&=&\triple{a}{h(r)}{b}
-   \end{array}\]
-\end{definition}
-
-\begin{lemma}
-\label{lemma:map_labels_preserves_wellTypedness}
-   If $\dataset$ is a data set and $h:\Rels\rightarrow\Rels$ is an injective function, then $\maprel{h}{\dataset}$ is a data set.
-\end{lemma}
-
-\begin{definition}[disjoint union of data sets]
-\begin{eqnarray}
-   \pair{\triples}{\inst}\sqcup\pair{\triples'}{\inst'}&=&\pair{\triples\uplus\triples'}{\inst\uplus^2\inst'}\\
-      X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
-      X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}
-\end{eqnarray}
-\end{definition}
-   In a disjoint union, the instances of the two data sets are relabeled to avoid name clashes.
-   In practice, we will not use 0 and 1 as labels for $x$ and $y$, but rather extend their names by prefixing.
-\begin{lemma}
-   If $\dataset$ and $\dataset'$ are data sets, then so is $\dataset\sqcup\dataset'$.
-\end{lemma}
-   This lemma is being proven in Isabelle/HOL~\cite{Isabelle}. The proof is published \href{location.domain}{here}
-
-\begin{definition}[disjoint union of schemas]
-\begin{eqnarray}
-   \triple{\concepts}{\rels}{\rules}\sqcup\triple{\concepts'}{\rels'}{\rules'}&=&\triple{\concepts\uplus\concepts}{\rels\uplus\rels'}{\rules\uplus\rules'}
-\end{eqnarray}
-\end{definition}
+%SJC on Formalization: it's often better to put these inline where we first use them. Leaving this as a comment for easy copy-paste.
+%    
+%    \subsection{Formalization}
+%    Let us define the instruments needed to describe the derivation of $\schema_{\migrsys}$.
+%    \begin{definition}[union of data sets]
+%    \label{def:graph_union_wellTyped_if_parts_wellTyped}
+%    \[\pair{\triples_1}{\inst_1}\cup\pair{\triples_2}{\inst_2}\ =\ \pair{\triples_1\cup\triples_2}{\inst_1\cup\inst_2}\]
+%    \end{definition}
+%    
+%    \begin{lemma}
+%    \label{lemma:graph_union_wellTyped_if_parts_wellTyped}
+%    If $\dataset_1$ and $\dataset_2$ are data sets, then $\dataset_1\cup\dataset_2$ is a data set.
+%    \end{lemma}
+%    
+%    \begin{definition}[map relation names in data sets]
+%    \label{def:map_labels_preserves_wellTypedness}
+%    Let $h:\Rels\rightarrow\Rels$, then
+%    \[\begin{array}{rcl}
+%      \maprel{h}{\pair{\triples}{\inst}}&=&\pair{\maprel{h}{\triples}}{\inst}\\
+%      \maprel{h}{\triple{a}{r}{b}}&=&\triple{a}{h(r)}{b}
+%    \end{array}\]
+%    \end{definition}
+%    
+%    \begin{lemma}
+%    \label{lemma:map_labels_preserves_wellTypedness}
+%    If $\dataset$ is a data set and $h:\Rels\rightarrow\Rels$ is an injective function, then $\maprel{h}{\dataset}$ is a data set.
+%    \end{lemma}
+%    
 %    \begin{definition}[disjoint union of data sets]
 %    \begin{eqnarray}
-%    \omit\rlap{$\la\atoms,\concepts,\inst,\isa,\rels,\dataset\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\dataset'\ra$}\notag\\
-%    &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\instuplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \dataset\uplus^5\dataset'\ra\notag\\
+%    \pair{\triples}{\inst}\sqcup\pair{\triples'}{\inst'}&=&\pair{\triples\uplus\triples'}{\inst\uplus^2\inst'}\\
 %      X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
-%       X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}\\
-%       X\uplus^3 Y&=&\begin{array}[t]{@{}l}\{\declare{(n,0)}{(A,0)}{(B,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\declare{(n,1)}{(A,1)}{(B,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}\\
-%       X\uplus^5 Y&=&\begin{array}[t]{@{}l}\{\triple{(a,0)}{\declare{(n,0)}{(A,0)}{(B,0)}}{(b,0)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in X\}\ \cup\\ \{\triple{(a,1)}{\declare{(n,1)}{(A,1)}{(B,1)}}{(b,1)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in Y\}\end{array}
+%      X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}
 %    \end{eqnarray}
 %    \end{definition}
-\begin{lemma}
-   If $\schema$ and $\schema'$ are schemas, then so is $\schema\sqcup\schema'$.
-\end{lemma}
-The proof of this lemma is published \href{location.domain}{here}
-
-To implement the disjoint union, the following relabelling functions are needed:
-\begin{definition}[relabel concepts]
-   \[\subst{C}{D}{A}\ =\ \text{\bf if}\ C=A\ \text{\bf then}\ D\ \text{\bf else}\ A\]
-\end{definition}
-\begin{definition}[relabel concepts in triples]
-   \[\subst{C}{D}{\triple{a}{r}{b}} = \triple{a}{\subst{C}{D}{r}}{b}\]
-\end{definition}
-\begin{definition}[relabel concepts in relations]
-   \[\subst{C}{D}{(\declare{n}{A}{B})} = \declare{n}{\subst{C}{D}{A}}{\subst{C}{D}{B}}\]
-\end{definition}
-\begin{definition}[relabel concepts in $\inst$]
-   \[a\ \subst{C}{D}{\inst}\ \subst{C}{D}{A}\ \Leftrightarrow\ a\inst A\]
-\end{definition}
-\begin{definition}[relabel concepts in data sets]
-   \[\begin{array}{rcl}
-      \rlap{$\subst{C}{D}{\pair{\triples}{\inst}}$}\\
-      &=&\pair{\{\subst{C}{D}{t}\mid t\in\triples\}}{\subst{C}{D}{\inst}}
-   \end{array}\]
-\end{definition}
-   The relabeling of concepts in rules must satisfy:
-\begin{equation}
-   \sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{C}{D}{u}}{\subst{C}{D}{\dataset}}
-\end{equation}
-\begin{definition}[relabel concepts in schemas]
-   \begin{eqnarray}
-      \rlap{$\subst{C}{D}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
-      &=&\begin{array}[t]{l@{}l}
-         \la&\{\subst{C}{D}{c}\mid c\in\concepts\}\\
-         ,&\{\subst{C}{D}{r}\mid r\in\rels\}\\
-         ,&\{\subst{C}{D}{u}\mid u\in\rules\}\ \ra\notag
-         \end{array}
-   \end{eqnarray}
-\end{definition}
-\begin{definition}[relabel relations]
-   \[\subst{\declare{m}{C}{D}}{m'}{{\declare{n}{A}{B}}}\ =\ \text{\bf if}\ n=m\wedge A=C\wedge B=D\ \text{\bf then}\ \declare{m'}{A}{B}\ \text{\bf else}\ \declare{m}{A}{B}\]
-\end{definition}
-\begin{definition}[relabel relations in triples]
-   \[\subst{r'}{m'}{\triple{a}{r}{b}} = \triple{a}{\subst{r'}{m'}{r}}{b}\]
-\end{definition}
-\begin{definition}[relabel relations in data sets]
-   \[\begin{array}{rcl}
-      \rlap{$\subst{r'}{m'}{\pair{\triples}{\inst}}$}\\
-      &=&\pair{\{\subst{r'}{m'}{t}\mid t\in\triples\}}{\inst}
-   \end{array}\]
-\end{definition}
-   The relabeling of relation names in rules must satisfy:
-\begin{equation}
-   \sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{r'}{m'}{u}}{\subst{r'}{m'}{\dataset}}
-\end{equation}
-\begin{definition}[relabel relations in schemas]
-   \begin{eqnarray}
-      \rlap{$\subst{r'}{m'}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
-      &=&\triple{\concepts}{\{\subst{r'}{m'}{r} \mid r\in\rels\}}{\{\subst{r'}{m'}{u} \mid u\in\rules\}}\notag
-   \end{eqnarray}
-\end{definition}
-
-   The relabeling of concepts is not trivial because of specialization.
-   To illustrate this point, suppose $D\isa A$.
-   Then the substitution $[C\rightarrow D]$ means that every atom $a$ that used to be of concept $C$ in the existing system, is now not only a $D$ but also an $A$.
-   For example, by relabeling the concept {\tt Hotel} to {\tt Motel}, in a situation where ${\tt Motel}\isa{\tt Parking}$,
-   all atoms that represent hotels might suddenly be expected to have parking lots.
-   So, whereas such examples may fly in a technical sense, the developer must use renaming with care to preserve the intended semantics.
+%    In a disjoint union, the instances of the two data sets are relabeled to avoid name clashes.
+%    In practice, we will not use 0 and 1 as labels for $x$ and $y$, but rather extend their names by prefixing.
+%    \begin{lemma}
+%    If $\dataset$ and $\dataset'$ are data sets, then so is $\dataset\sqcup\dataset'$.
+%    \end{lemma}
+%    This lemma is being proven in Isabelle/HOL~\cite{Isabelle}. The proof is published \href{location.domain}{here}
+%    
+%    \begin{definition}[disjoint union of schemas]
+%    \begin{eqnarray}
+%    \triple{\concepts}{\rels}{\rules}\sqcup\triple{\concepts'}{\rels'}{\rules'}&=&\triple{\concepts\uplus\concepts}{\rels\uplus\rels'}{\rules\uplus\rules'}
+%    \end{eqnarray}
+%    \end{definition}
+%    % \begin{definition}[disjoint union of data sets]
+%    % \begin{eqnarray}
+%    %    \omit\rlap{$\la\atoms,\concepts,\inst,\isa,\rels,\dataset\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\dataset'\ra$}\notag\\
+%    %    &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\instuplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \dataset\uplus^5\dataset'\ra\notag\\
+%    %       X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
+%    %       X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}\\
+%    %       X\uplus^3 Y&=&\begin{array}[t]{@{}l}\{\declare{(n,0)}{(A,0)}{(B,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\declare{(n,1)}{(A,1)}{(B,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}\\
+%    %       X\uplus^5 Y&=&\begin{array}[t]{@{}l}\{\triple{(a,0)}{\declare{(n,0)}{(A,0)}{(B,0)}}{(b,0)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in X\}\ \cup\\ \{\triple{(a,1)}{\declare{(n,1)}{(A,1)}{(B,1)}}{(b,1)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in Y\}\end{array}
+%    % \end{eqnarray}
+%    % \end{definition}
+%    \begin{lemma}
+%    If $\schema$ and $\schema'$ are schemas, then so is $\schema\sqcup\schema'$.
+%    \end{lemma}
+%    The proof of this lemma is published \href{location.domain}{here}
+%    
+%    To implement the disjoint union, the following relabelling functions are needed:
+%    \begin{definition}[relabel concepts]
+%    \[\subst{C}{D}{A}\ =\ \text{\bf if}\ C=A\ \text{\bf then}\ D\ \text{\bf else}\ A\]
+%    \end{definition}
+%    \begin{definition}[relabel concepts in triples]
+%    \[\subst{C}{D}{\triple{a}{r}{b}} = \triple{a}{\subst{C}{D}{r}}{b}\]
+%    \end{definition}
+%    \begin{definition}[relabel concepts in relations]
+%    \[\subst{C}{D}{(\declare{n}{A}{B})} = \declare{n}{\subst{C}{D}{A}}{\subst{C}{D}{B}}\]
+%    \end{definition}
+%    \begin{definition}[relabel concepts in $\inst$]
+%    \[a\ \subst{C}{D}{\inst}\ \subst{C}{D}{A}\ \Leftrightarrow\ a\inst A\]
+%    \end{definition}
+%    \begin{definition}[relabel concepts in data sets]
+%    \[\begin{array}{rcl}
+%      \rlap{$\subst{C}{D}{\pair{\triples}{\inst}}$}\\
+%      &=&\pair{\{\subst{C}{D}{t}\mid t\in\triples\}}{\subst{C}{D}{\inst}}
+%    \end{array}\]
+%    \end{definition}
+%    The relabeling of concepts in rules must satisfy:
+%    \begin{equation}
+%    \sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{C}{D}{u}}{\subst{C}{D}{\dataset}}
+%    \end{equation}
+%    \begin{definition}[relabel concepts in schemas]
+%    \begin{eqnarray}
+%      \rlap{$\subst{C}{D}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
+%      &=&\begin{array}[t]{l@{}l}
+%         \la&\{\subst{C}{D}{c}\mid c\in\concepts\}\\
+%         ,&\{\subst{C}{D}{r}\mid r\in\rels\}\\
+%         ,&\{\subst{C}{D}{u}\mid u\in\rules\}\ \ra\notag
+%         \end{array}
+%    \end{eqnarray}
+%    \end{definition}
+%    \begin{definition}[relabel relations]
+%    \[\subst{\declare{m}{C}{D}}{m'}{{\declare{n}{A}{B}}}\ =\ \text{\bf if}\ n=m\wedge A=C\wedge B=D\ \text{\bf then}\ \declare{m'}{A}{B}\ \text{\bf else}\ \declare{m}{A}{B}\]
+%    \end{definition}
+%    \begin{definition}[relabel relations in triples]
+%    \[\subst{r'}{m'}{\triple{a}{r}{b}} = \triple{a}{\subst{r'}{m'}{r}}{b}\]
+%    \end{definition}
+%    \begin{definition}[relabel relations in data sets]
+%    \[\begin{array}{rcl}
+%      \rlap{$\subst{r'}{m'}{\pair{\triples}{\inst}}$}\\
+%      &=&\pair{\{\subst{r'}{m'}{t}\mid t\in\triples\}}{\inst}
+%    \end{array}\]
+%    \end{definition}
+%    The relabeling of relation names in rules must satisfy:
+%    \begin{equation}
+%    \sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{r'}{m'}{u}}{\subst{r'}{m'}{\dataset}}
+%    \end{equation}
+%    \begin{definition}[relabel relations in schemas]
+%    \begin{eqnarray}
+%      \rlap{$\subst{r'}{m'}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
+%      &=&\triple{\concepts}{\{\subst{r'}{m'}{r} \mid r\in\rels\}}{\{\subst{r'}{m'}{u} \mid u\in\rules\}}\notag
+%    \end{eqnarray}
+%    \end{definition}
+%    
+%    The relabeling of concepts is not trivial because of specialization.
+%    To illustrate this point, suppose $D\isa A$.
+%    Then the substitution $[C\rightarrow D]$ means that every atom $a$ that used to be of concept $C$ in the existing system, is now not only a $D$ but also an $A$.
+%    For example, by relabeling the concept {\tt Hotel} to {\tt Motel}, in a situation where ${\tt Motel}\isa{\tt Parking}$,
+%    all atoms that represent hotels might suddenly be expected to have parking lots.
+%    So, whereas such examples may fly in a technical sense, the developer must use renaming with care to preserve the intended semantics.
 

### 2023-06-30 2022Migration/articleMigration.tex
WIP on chp 5 software architecture

@@
 \section{Software architecture}
-   Since functionality is excluded from this paper, we discuss the software architecture of the back-end of the migration system.
+   The theory presented above roughly corresponds to the back-end of a migration system.
+   This section describes its software architecture.
    For this research, we have used the Ampersand compiler to generate the system(s).
@@
    \centering
-   \includegraphics[width=0.8\textwidth]{figures/Before migration.png}
+   \includegraphics[scale=0.1]{figures/Before migration.png}
    \caption{Initial state of the migration}
@@
    \centering
-   \includegraphics[width=0.8\textwidth]{figures/During migration.png}
+   \includegraphics[scale=0.1]{figures/During migration.png}
    \caption{During the migration}
@@
    \centering
-   \includegraphics[width=0.8\textwidth]{figures/After migration.png}
+   \includegraphics[scale=0.1]{figures/After migration.png}
    \caption{After the migration}

### 2023-06-30 2022Migration/articleMigration.tex
Validation

--- 2023-06-28 2022Migration/articleMigration.tex
@@
 \begin{verbatim}
-   ENFORCE new.takes >: old.takes
+   ENFORCE new.takes := old.takes
 \end{verbatim}
@@
 
-   This means, however, that all pairs from {\tt old.takes} will be in {\tt new.takes} forever.
-   This is undesirable because data in the existing system may still change.
-   An extra relation (\ref{extraRelation}) fixes this problem by remembering which pairs have been copied:
+   This means, however, that {\tt old.takes} determines {\tt new.takes} forever.
+   This is undesirable because it prevents us from changing data in the migration system.
+   The desired behavior is that all pairs in {\tt old.takes} are copied to {\tt new.takes} when the system is first initialized, but that {\tt new.takes} can be changed freely after that.
+   An extra relation (\ref{extraRelation}) achieves this behavior by remembering which pairs have been copied:
 \begin{eqnarray}
@@
 However, a problem that may occur is that regular users of the system, who are used to the existing system, might be adding data to the system that causes violations faster than they can be solved.
-Two solutions can mitigate this.
-
-One solution is to state that the only violations that may occur in the relation \verb=isPartOfUNI= are those that were present when the migration started:
+To avoid this problem, we restrict the behavior of the system such that incremental progress will lead to eventually reaching the goal.
+
+Our solution is to state that the only violations that may occur in the relation \verb=isPartOfUNI= are those that were present when the migration started:
 \begin{verbatim}
@@
 
-In this first solution, the first rule is the same as with the simple solution.
+In this solution, the first rule is the same as with the simple solution.
 The second rule prevents any new violation from occurring.
@@
 
-As a second solution, the data in the designed system does not have any violations.
-Consequentially, the cause of the violations comes from the triples in the existing system.
-Rather than taking the union of all triples, the disjoint union is taken instead;
-\begin{verbatim}
-   RELATION isPartOf_old[Module*Course] [UNI]=
-      [ ("Finance", "Management")
-      ; ("Business Rules", "Business IT")
-      ; ("Business Analytics", "Business IT")
-      ; ("IT-Governance", "Management")
-      ]
-   RELATION isPartOf[Module*Course] [UNI]=
-      [ ("IT-Governance", "Business IT") ]
-   RELATION isPartOf_ignored[Module*Course]
-
-   RULE isPartOf_isUnion:
-         isPartOf_old |- isPartOf \/ isPartOf_ignored
-   ROLE migrationHelper MAINTAINS isPartOf_isUnion
-
-   ENFORCE isPartOf_ignored :> isPartOf
-\end{verbatim}
-
-\begin{verbatim}
-   RELATION isPartOf_old[Module*Course] [UNI]=
-      [ ("Finance", "Management")
-      ; ("Business Rules", "Business IT")
-      ; ("Business Analytics", "Business IT")
-      ; ("IT-Governance", "Management")
-      ]
-   RELATION isPartOf[Module*Course] [UNI]=
-      [ ("IT-Governance", "Business IT") ]
-   RELATION isPartOf_ignored[Module*Course]
-
-   ENFORCE isPartOf :> isPartOf_old - isPartOf_ignored
-   ENFORCE isPartOf_ignored :> isPartOf
-\end{verbatim}
-In this second solution, we immediately move to a system in which the \verb=UNI= rule is maintained by the system.
-The work for the \verb=migrationHelper= is again finite: the pairs in \verb=isPartOf_old= need to be copied over or explicitly ignored.
-Once a pair is in \verb=isPartOf=, the pair is copied over to \verb=isPartOf_ignored= by the \verb=ENFORCE= statement, such that regular users can delete it without adding work to anyone with the \verb=migrationHelper= role.
-Of course, since this solution starts with fewer pairs in the \verb=isPartOf= relation, other rules could be violated as a result of choosing this solution.
-Indeed, ExamRule2 initially has more violations in this scenario.
-
-\begin{definition}[migration data set]
-   \begin{eqnarray}
-      &&\begin{array}[t]{@{}l}
-         \{ {\tt ENFORCE\ }\declare{(nm',1)}{(A',1)}{(B',1)}\\\quad{\tt\ :=\ }\ident{(A',1)};\declare{(nm,0)}{(A,0)}{(B,0)};\ident{(B',1)}\\
-            \mid\ \begin{array}[t]{@{}l}
-               \declare{nm}{A}{B}\in\rels\wedge\declare{nm'}{A'}{B'}\in\rels'\wedge\id{nm}=\id{nm'}\ \wedge\\
-               \{\pair{A}{A'},\pair{B}{B'}\}\subseteq\isa\cup\isa'\cup\flip{\isa}\cup\flip{\isa'}\}
-               \end{array}
-           \end{array}\\
-           &\cup&\begin{array}[t]{@{}l}
-            \{ {\tt CLASSIFY\ }(C',1){\tt\ IS\ }(C,0)\mid\ C\in\concepts,C'\in\concepts',C=C'\}\\
-           \end{array}
-   \end{eqnarray}
-\end{definition}
-
-The ${\tt CLASSIFY\ }(C',1){\tt\ IS\ }(C,0)$ statement adds two pairs to the $\isa$ relation:
-$\pair{(C',1)}{(C,0)}$ and $\pair{(C,0)}{(C',1)}$.
-Note that if $\dataset$ is a data set, and $\dataset'$ is that data set with pairs added%
-\footnote{TODO: het woord `add' is informeel, dit kan beter} to the $\isa$ relation, then the result is again a data set.
-The ${\tt ENFORCE\ }r{\tt\ :=\ }e$ statements add triples $(x,r,y)$ for all $(x,y)\in e$.
-The terms $t$ in these statements are such that $x\ (\inst \compose \kleenestar{\flip{\isa}})\ (A',1)$ and $y \ (\inst \compose \kleenestar{\flip{\isa}})\ (B',1)$, thus preserving the property that the result is a data set.
-
-\begin{definition}[]
-   \begin{eqnarray}
-      \infsys\sqcup\infsys'&=&\triple{\roles'}{\rules'}{\maintain'}{\dataset\sqcup\dataset'}
-   \end{eqnarray}
-\end{definition}
+% SJC: TODO voor Stef. Misschien is het de moeite waard om te zeggen hoe isPartOfUNI_violations initieel gevuld kan worden.
+% (Dit werkt precies als het initieel vullen van andere relaties, zoals in de vorige sectie).
+
+%As a second solution, the data in the designed system does not have any violations.
+%Consequentially, the cause of the violations comes from the triples in the existing system.
+%Rather than taking the union of all triples, the disjoint union is taken instead;
+%\begin{verbatim}
+%   RELATION isPartOf_old[Module*Course] [UNI]=
+%      [ ("Finance", "Management")
+%      ; ("Business Rules", "Business IT")
+%      ; ("Business Analytics", "Business IT")
+%      ; ("IT-Governance", "Management")
+%      ]
+%   RELATION isPartOf[Module*Course] [UNI]=
+%      [ ("IT-Governance", "Business IT") ]
+%   RELATION isPartOf_ignored[Module*Course]
+%   
+%   RULE isPartOf_isUnion:
+%         isPartOf_old |- isPartOf \/ isPartOf_ignored
+%   ROLE migrationHelper MAINTAINS isPartOf_isUnion
+%   
+%   ENFORCE isPartOf_ignored :> isPartOf
+%\end{verbatim}
+%
+%\begin{verbatim}
+%   RELATION isPartOf_old[Module*Course] [UNI]=
+%      [ ("Finance", "Management")
+%      ; ("Business Rules", "Business IT")
+%      ; ("Business Analytics", "Business IT")
+%      ; ("IT-Governance", "Management")
+%      ]
+%   RELATION isPartOf[Module*Course] [UNI]=
+%      [ ("IT-Governance", "Business IT") ]
+%   RELATION isPartOf_ignored[Module*Course]
+%
+%   ENFORCE isPartOf :> isPartOf_old - isPartOf_ignored
+%   ENFORCE isPartOf_ignored :> isPartOf
+%\end{verbatim}
+%In this second solution, we immediately move to a system in which the \verb=UNI= rule is maintained by the system.
+%The work for the \verb=migrationHelper= is again finite: the pairs in \verb=isPartOf_old= need to be copied over or explicitly ignored.
+%Once a pair is in \verb=isPartOf=, the pair is copied over to \verb=isPartOf_ignored= by the \verb=ENFORCE= statement, such that regular users can delete it without adding work to anyone with the \verb=migrationHelper= role.
+%Of course, since this solution starts with fewer pairs in the \verb=isPartOf= relation, other rules could be violated as a result of choosing this solution.
+%Indeed, ExamRule2 initially has more violations in this scenario.
+%
+%\begin{definition}[migration data set]
+%   \begin{eqnarray}
+%      &&\begin{array}[t]{@{}l}
+%         \{ {\tt ENFORCE\ }\declare{(nm',1)}{(A',1)}{(B',1)}\\\quad{\tt\ :=\ }\ident{(A',1)};\declare{(nm,0)}{(A,0)}{(B,0)};\ident{(B',1)}\\
+%            \mid\ \begin{array}[t]{@{}l}
+%               \declare{nm}{A}{B}\in\rels\wedge\declare{nm'}{A'}{B'}\in\rels'\wedge\id{nm}=\id{nm'}\ \wedge\\
+%               \{\pair{A}{A'},\pair{B}{B'}\}\subseteq\isa\cup\isa'\cup\flip{\isa}\cup\flip{\isa'}\}
+%               \end{array}
+%           \end{array}\\
+%           &\cup&\begin{array}[t]{@{}l}
+%            \{ {\tt CLASSIFY\ }(C',1){\tt\ IS\ }(C,0)\mid\ C\in\concepts,C'\in\concepts',C=C'\}\\
+%           \end{array}
+%   \end{eqnarray}
+%\end{definition}
+%
+%The ${\tt CLASSIFY\ }(C',1){\tt\ IS\ }(C,0)$ statement adds two pairs to the $\isa$ relation:
+%$\pair{(C',1)}{(C,0)}$ and $\pair{(C,0)}{(C',1)}$.
+%Note that if $\dataset$ is a data set, and $\dataset'$ is that data set with pairs added%
+%\footnote{TODO: het woord `add' is informeel, dit kan beter} to the $\isa$ relation, then the result is again a data set.
+%The ${\tt ENFORCE\ }r{\tt\ :=\ }e$ statements add triples $(x,r,y)$ for all $(x,y)\in e$.
+%The terms $t$ in these statements are such that $x\ (\inst \compose \kleenestar{\flip{\isa}})\ (A',1)$ and $y \ (\inst \compose \kleenestar{\flip{\isa}})\ (B',1)$, thus preserving the property that the result is a data set.
+%
+%\begin{definition}[]
+%   \begin{eqnarray}
+%      \infsys\sqcup\infsys'&=&\triple{\roles'}{\rules'}{\maintain'}{\dataset\sqcup\dataset'}
+%   \end{eqnarray}
+%\end{definition}
 
@@
 \section{Validation}
-   To validate the theory, we have to prove that:
-\begin{itemize}
-   % -> het migratie-systeem is typefout-vrij (opmerking: als we de type-checker buiten beschouwing willen laten, kunnen we dit niet beschrijven)
-   \item the migration script $\migrsys$ is free of type errors.
-   % -> het migratie-systeem is vrij van overtredingen op regels die het migratie-systeem moet bewaken
-   \item the migration system $\migrsys$ is an information system (definition~\ref{def:information system}),
-         which implies that its population yields no violations.
-   % -> het migratie-systeem bevat alle oude data, ihb nog steeds na het toepassen van de enforce regels
-   \item $\migrsys$ contains all data of the existing system. So, $\dataset_\infsys\subseteq\dataset_\migrsys$.
-   % -> er is een pad naar ingebruikname van het nieuwe systeem (vanaf de initiële toestand van het migratie-systeem)
-   \item there is a well-founded progress measure towards releasing the desired system.
-   % -> corollary: op het moment van ingebruikname van het nieuwe systeem, is het migratie-systeem vrij van overtredingen op regels die het nieuwe systeem moet bewaken
-   \item When the progress measure indicates that further steps are not possible, the desired system $\infsys'$ is void of violations.
-   The invariants of $\infsys'$ are cork rules in $\migrsys$.
-   So, once all these rules are satisfied (and secured with a cork),
-   all invariants in $\infsys'$ are satisfied and ready for deployment.
-   % -> optioneel: na een begrensd aantal 'voortgangs-stappen' kan het nieuwe systeem in gebruik genomen worden
-\end{itemize}
-% Bewijsverplichtingen:
+   We claim that the information system does not need to be paused in order for the migration to happen successfully.
+   By this, we mean that computations that take a considerable amount of time, as well as human actions, can be performed incrementally while the system is in use.
+   In order for us to prove this claim, we model the behavior of our information system as a series of systems $\migrsys_0,\migrsys_1,\ldots,\migrsys_n$ with $\migrsys = \migrsys_0$, where only the dataset of the system changes.
+   We consider two types of updates to the dataset.
+
+   The first type of update helps the migration to go forward: we'll write $\rightsquigarrow$ for those.
+   The second type of update is that which other users of the system do (possibly none), indicated as $\dashrightarrow$.
+   The two updates alternate: $\migrsys_0 \dashrightarrow \migrsys_1 \rightsquigarrow \migrsys_2 \dashrightarrow \ldots \rightsquigarrow \migrsys_{2i} \dashrightarrow \migrsys_{2i+1}\rightsquigarrow\migrsys_{2i+2}\dashrightarrow\ldots\migrsys_n$.
+
+   We validate our claim that the information system will reach the MoC by arguing about two situations:
+   The system between deployment and the MoT can reach the MoT by firing the ENFORCE rules, and the system between the MoT and the MoC can reach the MoC after a finite number of actions by the migration engineer.
+   More concretely, this means the following:
+
+When the default migration system $\migrsys$ is deployed, the system will always be in a state such that continually running any violated ENFORCE rules (in any order) leads to the MoT.
+   This means that $A\rightsquigarrow B$ is such that the dataset of $B$ is that of $A$ with an update applied according to one or more violated enforce rules. % TODO: define 'violated'
+   If there is no such $B$ because no enforce rules are violated, we are at the moment of transition.
+   We say that the system $A$ is in a state such that the MoT can be reached if there are no infinite $\rightsquigarrow$-paths from $A$.
+   The constraint on $A\dashrightarrow B$ is that only the old system can change at this point.
+   Note that this does not prove that the moment of transition to never be reached, as more enforce rules may get violated at every $\dashrightarrow$ step.
+   We expect this not to be a problem in practice since $\dashrightarrow$ transitions are triggered by users while $\rightsquigarrow$ is triggered automatically through ENFORCE rules: the ENFORCE rules should outperform the $\dashrightarrow$ actions by a big margin.
+
+After the moment of transition, there is a finite upper-bound on the amount of work the migration engineer has to do:
+   $A\rightsquigarrow B$ is such that the dataset of $B$ is that of $A$ with one of the violations shown to a migration engineer resolved (including the firing of corresponding ENFORCE statements), and $\dashrightarrow$ arbitrary.
+   In both $A\rightsquigarrow B$ and $A\dashrightarrow B$ we assume that the relations that were specifically created for the bookkeeping (suffixed by \verb=_progress= and \verb=_violations=) can only be updated through the firing of the ENFORCE statements.
+   These assumptions make the total number of pairs in \verb=_violations= relations decline in every $\rightsquigarrow$-step, proving that the MoC will be reached after a finite number of $\rightsquigarrow$ steps (assuming that the set of violations is finite).
+   Moreover, we show that the desired system with the dataset taken from the migration system (after dropping the old system and the bookkeeping) is free of violations when all \verb=_violations= relations are empty.
 

### 2023-07-01 2022Migration/articleMigration.tex
Work on the process description (new chp 4)

@@
 
-\subsection{Process of Data Migration}
+\section{Process of Data Migration}
    Sections~\ref{sct:Information Systems} and~\ref{sct:Example existing IS} introduce a single information system.
    The transition from an existing system $\infsys$ to the desired one $\infsys'$ is the topic of this section.
-   Let us first sketch the steps to migrate system $\infsys$ to $\infsys'$.
+\begin{figure}
+   \centering
+   \includegraphics[scale=0.1]{figures/Before migration.png}
+   \caption{Initial state of the migration}
+   \label{fig:initial}
+\end{figure}
+   Figure~\ref{fig:initial} shows the initial state of the migration,
+   in which the existing information system $\infsys$ has data set $\dataset$ and schema $\schema$.
+   At the start of this migration, $\dataset$ satisfies all rules of $\schema$ (by requirement~\ref{eqn:satisfaction}).
 \begin{enumerate}
-   \item Let the existing information system $\infsys$ have data set $\dataset$ and schema $\schema$.
-         A developer defines a new information system, $\infsys'$, which contains a data set of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{In this paper, we only consider the functionality captured by rules.} as illustrated in figure~\ref{fig:post-migration}.
-         After the migration, the resulting system will be $\infsys'$, enhanced with the triples from $\dataset$ that satisfy $\schema'$,
-         and with additional changes made by the migration engineer (see step~\ref{item3}) and by users who have been using the system during the migration.
-         $\dataset'$ may contain some new data,
-         for instance, to initialize new features at the moment of transition (MoT).
-         If schemas $\schema$ equals $\schema'$, the data migration is unnecessary, so in the sequel we assume they are different.
-   \item The migration engineer uses a software generator to generate a migration script, $\migrsys$,
-         This is a system that includes both $\infsys$ and $\infsys'$.
-         $\migrsys$ is meant to migrate the data and define the semantics during the migration phase.
-         It contains enforcement rules that migrate existing data to the desired data set.
-         It also contains data structures and rules that are needed for the duration of the migration phase.
-%         The generated script must be free of type errors, to ensure that $\sat{u}{\dataset}$ exists for every rule $u$ and data set $\dataset$.
-%         The proof is available \href{https://www.isa-afp.org/}{here}
+\item The migration starts with a developer%
+\footnote{The developer role may stand for a team of developers, a single developer, a migration engineer, or whoever defines the desired system.}
+      who specifies a new information system, $\infsys'$.
+      Let us call this moment the start of the migration (SoM).
+      $\infsys'$ contains a data set of its own, $\dataset'$, a schema $\schema'$, and some functionality%
+\footnote{In this paper, we only consider functionality that is captured by rules.}.
+      Before replacing $\infsys$ by $\infsys'$, the data from $\dataset$ must be migrated to $\dataset'$
+      and all invariants of $\schema'$ must be satisfied.
+      We assume that schemas $\schema$ and $\schema'$ are different, or else the migration would be unnecessary.
+\item From schemas $\schema$ and $\schema'$, a migration engineer%
+\footnote{The migration engineer may also stand for a team, which may overlap with the developer role.}
+      generates a migration script, using the Ampersand compiler.
+      This script specifies the migration system, $\migrsys$, that includes $\infsys$, $\infsys'$ and a number of migration rules.
+      $\migrsys$ is meant to migrate the data and ensure that $\dataset'$ satisfies all invariants of $\schema'$.
 \item\label{item3}
-         In many cases, the schema $\schema_{\migrsys}$ will be enough to perform the migration.
+      In specific cases, the schema $\schema_{\migrsys}$ will be enough to perform the migration automatically.
+      As soon as the data is transferred and $\migrsys$ has satisfied the invariants of $\schema'$,
+      $\infsys'$ can take care of production on its own without the help of $\migrsys$.
       However, a migration engineer may have specific requirements that call for changes in $\migrsys$.
-         For that purpose, the generator generates the migration script as an Ampersand script,
-         so the migration engineer can accommodate such requirements.
+      For that purpose, she may want to alter the migration script to accommodate such requirements.
       This can be necessary, for example, to resolve data pollution that violates a constraint that was not in schema $\schema_{\infsys}$.
-         She can either implement enforcement rules or use the business to resolve the issues.
+      She can implement such adaptations as enforcement rules, which are executed automatically.
+      But she can also ask business users to resolve issues by specifying process rules.
+      Human involvement is often needed to implement a desired migration strategy.
+      In that case the migration system must ensure that there are finitely many steps to complete the migration.
+      The migration engineer must ensure that $\migrsys'$ can satisfy all invariants in $\schema'$, either by enforcement rules or by process rules.
       She can test the resulting migration script, $\migrsys'$ separately before taking the desired system to production.
-         The migration engineer alters $\schema$ nor $\schema'$ while adapting $\migrsys'$ to the specific needs of the current migration.
-   \item Then, the migration engineer deploys system $\migrsys'$.
-         This event starts the copying of data from $\dataset$ to $\dataset'$, as specified in $\migrsys'$ (figure~\ref{fig:migration phase}).
+      When this work is done, the migration engineer has a migration system $\migrsys'$ with schema $\schema'$ and an initial dataset $\dataset'$
+      that is ready to deploy.
+\item\label{item4}
+      Then, the migration engineer deploys system $\migrsys'$.
+\begin{figure}
+   \centering
+   \includegraphics[scale=0.1]{figures/4MoD.png}
+   \caption{During the migration}
+   \label{fig:4MoD}
+\end{figure}
+      Let us call this moment the moment of deployment (MoD).
+      This event starts the copying of data from $\dataset$ to $\dataset'$, as specified in $\migrsys'$.
       During this phase, the ingress is still referring users to the existing system,
       so they will hardly notice that the desired system is being fired up and loaded with data.
-         Users are still changing $\dataset$ (both inserts and deletes),
-         but these changes are being transferred to $\dataset'$ as well.
+      During this phase, users can still change $\dataset$ (both inserts and deletes) while $\infsys$ is keeping its invariants true.
+      Yet, the migration system also transfers these ``new'' changes to $\dataset'$ by means of its enforcement rules.
       This step is complete when all data from $\dataset$ has been copied to $\dataset'$ and all invariants of the migration system are true.
-   \item At this moment, the ingress switches the incoming traffic from $\infsys$ to $\migrsys'$,
+\item After completing step~\ref{item4}, the ingress switches the incoming traffic from $\infsys$ to $\migrsys'$,
       so $\migrsys'$ effectively takes over.
+\begin{figure}
+   \centering
+   \includegraphics[scale=0.1]{figures/5MoT.png}
+   \caption{Situation after MoT}
+   \label{fig:during}
+\end{figure}
+      Simultaneously, all clients must switch to the functionality of $\migrsys'$,
+      so users will experience the new functionality.
       This event marks the MoT.
@@
       it may take some (short but finite) time until the invariants of $\infsys$ are true.
-         After that, dataset $\dataset$ will no longer change.
-   \item From this moment, users experience the functionality of the migration system $\migrsys'$.
-         This system must ensure that all invariants of $\schema'$ are true before the migration is complete.
-         It no longer ensures the rules of $\schema$, which govern the existing system.
-         Since $\dataset$ is no longer changing, the migration engineer can now focus on the remaining violations of $\schema'$.
-         Users of the migration system will deal with process rules in the migration system that are still violated.
+      After that, the existing dataset $\dataset$ will no longer change.
+\item Starting at MoT, $\migrsys'$ is working towards satisfying the migration rules.
+      It executes its enforce rules to migrate data.
+      It also executes enforce rules and process rules to ensure that all invariants of $\infsys'$ are true.
+      Since $\migrsys'$ also contains $\schema$,
+      it will also keep the invariants of $\infsys$ as the final transactions trickle into the new system.
+      However, the traffic into $\infsys$ has stopped, so the changes in $\dataset$ will quickly dry up.
       Once all invariants of $\migrsys'$ are true, the migration is complete.
-         This state marks the moment of cleanup (MoC)
-   \item After the MoC, the ingress switches the incoming traffic from $\migrsys'$ to $\infsys'$.
-         The migration system is no longer needed and can be taken out of production, together with the existing system $\infsys$.
-         The desired system remains in production and the migration is finished.
+\item Let us designate the moment of completion as MoC.
+      After the MoC, the ingress can switch the incoming traffic from $\migrsys'$ to $\infsys'$ because all invariants of $\infsys'$ are true.
+      $\infsys$ and $\migrsys'$ are no longer needed and can be taken out of production.
+      $\infsys'$ remains in production and the migration is finished.
       This establishes the post-migration situation (figure~\ref{fig:post-migration}).
@@
 
-   Upon initialization, the migration system evaluates all enforce rules in an attempt to satisfy all invariants.
-   In this example, it leaves two violations behind that it cannot resolve by enforce rules.
+   Upon initialization, the migration system evaluates all enforcement rules to satisfy all invariants.
+   In this example, it leaves two violations behind that it cannot resolve by enforcement rules.
    The pair {\tt ("Peter", "Management")} is the first one; it violates rule {\tt new.ExamRule2}.
@@
 When the default migration system $\migrsys$ is deployed, the system will always be in a state such that continually running any violated ENFORCE rules (in any order) leads to the MoT.
-   This means that $A\rightsquigarrow B$ is such that the dataset of $B$ is that of $A$ with an update applied according to one or more violated enforce rules. % TODO: define 'violated'
-   If there is no such $B$ because no enforce rules are violated, we are at the moment of transition.
+   This means that $A\rightsquigarrow B$ is such that the dataset of $B$ is that of $A$ with an update applied according to one or more violated enforcement rules. % TODO: define 'violated'
+   If there is no such $B$ because no enforcement rules are violated, we are at the moment of transition.
    We say that the system $A$ is in a state such that the MoT can be reached if there are no infinite $\rightsquigarrow$-paths from $A$.
    The constraint on $A\dashrightarrow B$ is that only the old system can change at this point.
-   Note that this does not prove that the moment of transition to never be reached, as more enforce rules may get violated at every $\dashrightarrow$ step.
+   Note that this does not prove that the moment of transition to never be reached, as more enforcement rules may get violated at every $\dashrightarrow$ step.
    We expect this not to be a problem in practice since $\dashrightarrow$ transitions are triggered by users while $\rightsquigarrow$ is triggered automatically through ENFORCE rules: the ENFORCE rules should outperform the $\dashrightarrow$ actions by a big margin.
@@
    For this research, we have used the Ampersand compiler to generate the system(s).
-   Figure~\ref{fig:initial} shows the starting point of the migration.
-\begin{figure}
-   \centering
-   \includegraphics[scale=0.1]{figures/Before migration.png}
-   \caption{Initial state of the migration}
-   \label{fig:initial}
-\end{figure}
    For the purpose of data migration, the existing system must make its schema available through its API.
@@
    This is implemented by Ampersand's INCLUDE mechanism, which boils down to taking the disjoint union of the schemas.
-\begin{figure}
-   \centering
-   \includegraphics[scale=0.1]{figures/During migration.png}
-   \caption{During the migration}
-   \label{fig:during}
-\end{figure}
-   Deployment of the migration system (figure~\ref{fig:during}) leaves the existing system intact.
+   Deployment of the migration system leaves the existing system intact.
    It enhances the system with the data set needed by the migration system, which includes the data set needed by the desired system.

### 2023-07-06 2022Migration/articleMigrationFACS.tex
FACS article without conclusion.

--- 2023-06-30 2022Migration/articleMigration.tex
@@
-\documentclass{elsarticle}
-% \usepackage{graphicx}
+% This is samplepaper.tex, a sample chapter demonstrating the
+% LLNCS macro package for Springer Computer Science proceedings;
+% Version 2.21 of 2022/01/12
+%
+\documentclass[runningheads]{llncs}
+%
+\usepackage[T1]{fontenc}
+% T1 fonts will be used to generate the final print and online PDFs,
+% so please use T1 fonts in your manuscript whenever possible.
+% Other font encondings may result in incorrect characters.
+%
+\usepackage{graphicx}
+% Used for displaying a sample figure. If possible, figure files should
+% be included in EPS format.
+%
+
 \usepackage{hyperref}
@@
 %\usepackage{footmisc}
-\usepackage{amstext}
+%\usepackage{amstext}
 \usepackage{amsmath}
@@
 \usepackage{mathrsfs}
-\usepackage{amsthm}
+\usepackage{mathtools}
+%\usepackage{amsthm}
 \usepackage[english]{babel}
@@
 \hyphenation{ExecEngine}
-\newtheorem{lemma}{Lemma}
+%\newtheorem{lemma}{Lemma}
+
+% Ampersand -----------------------------------------------------------
+
+%\def\id#1{\text{\it #1\/}}
+\newcommand{\xrightarrowdbl}[2][]{%
+  \xrightarrow[#1]{#2}\mathrel{\mkern-14mu}\rightarrow
+}
+\newcommand{\id}[1]{\text{\it #1\/}}
+\newcommand{\code}[1]{\text{\tt\small #1}}
+\newcommand{\stmtText}[1]{``{\small\tt #1}''}
+\newcommand{\dom}[1]{\id{dom}(#1)}
+\newcommand{\cod}[1]{\id{cod}(#1)}
+%\renewcommand{\int}[2]{\id{inter}(#1,#2)}
+\newcommand{\relsIn}[1]{\id{relsIn}(#1)}    % maps a Term to a set of Relations
+\newcommand{\pop}[2]{\id{pop}_{#1}(#2)}
+\newcommand{\maintain}{\mathbin{\id{maint}}}
+\newcommand{\enf}{\mathbin{{\tt enforce}}}
+\newcommand{\enforce}[2]{{\tt enforce}_{#1}(#2)}
+\newcommand{\instance}{\mathbin{\id{inst}}}
+\newcommand{\relname}[1]{\id{relname}(#1)}
+\newcommand{\evt}[2]{\id{event}_{#1,#2}}
+\newcommand{\src}[1]{\id{src}(#1)}
+\newcommand{\tgt}[1]{\id{tgt}(#1)}
+\newcommand{\sat}[2]{\id{sat}_{#1}(#2)}
+\newcommand{\viol}[2]{\violC{#1}(#2)}
+\newcommand{\violC}[1]{\id{viol}_{#1}}
+\newcommand{\sign}[1]{\id{sign}(#1)}
+\newcommand{\powerset}[1]{\cal{P}\{#1\}}
+\newcommand{\la}{\langle}
+\newcommand{\ra}{\rangle}
+\newcommand{\full}{V}
+\newcommand{\declare}[3]{\id{#1}_{\pair{#2}{#3}}}
+\newcommand{\subst}[3]{#3_{[#1\rightarrow #2]}}
+\newcommand{\fullt}[2]{V_{\pair{#1}{#2}}}
+\newcommand{\iden}{I}
+\newcommand{\ident}[1]{I_{\id{#1}}}
+\newcommand{\expr}[3]{(#1)_{#2\times #3}}
+\newcommand{\pair}[2]{\langle{#1},{#2}\rangle}
+\newcommand{\maprel}[2]{{\tt maprel}_{#1}({#2})}
+\newcommand{\Pair}[2]{#1\times#2}
+\newcommand{\pairs}[1]{\id{pairs}(#1)}
+\newcommand{\triple}[3]{\langle{#1},{#2},{#3}\rangle}
+\newcommand{\quadruple}[4]{\langle{#1},{#2},{#3},{#4}\rangle}
+\newcommand{\atom}[1]{{\tt\small #1}}
+\newcommand{\atoms}{\mathcal{A}}
+\newcommand{\Atoms}{\mathbb{A}}
+\newcommand{\events}{\mathcal{E}}
+\newcommand{\Events}{\mathbb{E}}
+\newcommand{\concept}[1]{{\tt\small #1}}
+\newcommand{\concepts}{\mathcal{C}}
+\newcommand{\Concepts}{\mathbb{C}}
+\newcommand{\decls}{\mathcal{D}}  %% names of relations
+\newcommand{\rels}{\mathcal{R}}   %% all relations
+\newcommand{\Rels}{\mathbb{R}}   %% all relations
+\newcommand{\relations}{\mathcal{M}} % representing terms. M is a subset of R.
+\newcommand{\triples}{\mathcal{T}}
+\newcommand{\Triples}{\mathbb{T}}
+\newcommand{\Triple}[3]{#1\times#2\times#3}
+\newcommand{\vertices}{N}
+\newcommand{\rules}{\mathcal{U}}
+\newcommand{\Rules}{\mathbb{U}}
+\newcommand{\specrules}{\mathcal{S}}
+\newcommand{\roles}{\mathcal{O}}
+\newcommand{\dataset}{\mathscr{D}}
+\newcommand{\Dataset}{\mathbb{D}}
+\newcommand{\schema}{\mathscr{Z}}
+\newcommand{\functionality}{\mathscr{F}}
+\newcommand{\select}[2]{\id{select}_{#1}\{{#2}\}}
+\newcommand{\migrsys}{\mathscr{M}}
+\newcommand{\infsys}{\mathscr{S}}
+\newcommand{\tf}[1]{\mathscr{T}(#1)}
+\newcommand{\ptf}[1]{\mathscr{T}'(#1)}
+\newcommand{\ti}[1]{\mathscr{I}(#1)}
+\newcommand{\tic}[1]{I_{\cal C}(#1)}
+\newcommand{\relAdd}{\dagger}
+\newcommand{\flip}[1]{{#1}^\smallsmile} %formerly:  {#1}^\backsim
+\newcommand{\kleeneplus}[1]{{#1}^+}
+\newcommand{\kleenestar}[1]{{#1}^*}
+\newcommand{\cmpl}[1]{\overline{#1}}
+\newcommand{\rel}{\times}
+\newcommand{\compose}{;}
+\newcommand{\subs}{\subseteq}%{\models}
+\newcommand{\fun}{\rightarrow}
+\newcommand{\isa}{\preceq}
+%\newcommand{\isaClos}{\sqsubseteq}
+\newcommand{\typetest}{?}
+\newcommand{\meet}{\sqcap}
+\newcommand{\join}{\sqcup}
+\newcommand{\Meet}{\bigsqcap}
+\newcommand{\Moin}{\bigsqcup} % because LaTeX has already defined command \Join.
+\newcommand{\order}{\ominus}
+\newcommand{\anything}{\top}
+\newcommand{\nothing}{\bot}
+\newcommand{\rewriteto}{\rightarrow}
+\newcommand{\calc}{\implies}
+\newcommand{\alland}{\bigwedge}
+\newcommand{\mph}[3]{#1_{#2\times #3}}
+\newcommand{\mphu}[1]{#1_{\univ\times\univ}}
+
+%-----------------------------------------
+\newcommand{\kse}{\hspace*{1.7em}}
+\newcommand{\ksf}{\hspace*{1em}}
+\newcommand{\ksg}{\hspace*{1em}}
+\newenvironment{derivation}{\begin{tabbing}\kse \= \ksf \= \ksg \= \kill}{\end{tabbing}}
+%\newtheorem{definition}{Definition}
+\newcommand{\term}[1]{\>\>\(#1\)\\[1ex]}
+\newcommand{\rela}[2]{\>\(#1\)\>\>\{ \ #2 \ \}\\[1ex]}
+\newcommand{\weg}[1]{}
+
+\def\define#1{\label{dfn:#1}{\em #1}\index{#1}}
+\def\definem#1{\label{dfnm:#1}{\em #1}\index{#1}\marge{#1}}
+\newcommand{\marg}[1]{\index{#1}\marge{#1}}
+
+\usepackage{color}
+\renewcommand\UrlFont{\color{blue}\rmfamily}
+%
+
 \begin{document}
-\include{preambleMigrations}
-
-% TODO Algemeen:
-% - type van triple-set en atom-set definieren zodat viol_u ?->P(?x?)
-%   (Bas: ik heb foute types weggehaald, maar dit nog niet gedefinieerd)
-% - nieuwe populatie na de disjoint union zou de populatie met ENFORCE regels toegepast moeten zijn
-%  (en in het bijzonder de ISA regels)
-
-% TODO Stef:
-% - stukje literatuuronderzoek
-% https://ieeexplore.ieee.org/abstract/document/7445334?casa_token=ECzi6XeV2ncAAAAA:KhWzB8XBFOUJ0C6AD-XjX_ryuA9ARvTd3gm6RR-ZNiR8sZ1858FJpQ7zKQhkAZDlv8IjPdgD
-% https://ieeexplore.ieee.org/abstract/document/8549944?casa_token=9qiGqNzh2Q0AAAAA:C-cYogExB35nGxQdxLcdBh4JoLNvM0OHedAMhCbB5V4kb4_6nzHUvc23xSJbeoBu67LSiz-Y
-% https://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.651.9298&rep=rep1&type=pdf
-% https://www.scirp.org/html/4-7800724_106592.htm
-% https://www.researchgate.net/profile/Ranjana-Badre/publication/318665687_GUI_for_Data_Migration_and_Query_Conversion/links/5b45bbea0f7e9b1c722386e5/GUI-for-Data-Migration-and-Query-Conversion.pdf
-% https://journal3.uin-alauddin.ac.id/index.php/literatify/article/view/12567
-
-% Bewijsverplichtingen:
-% -> het migratie-systeem is typefout-vrij (opmerking: als we de type-checker buiten beschouwing willen laten, kunnen we dit niet beschrijven)
-% -> het migratie-systeem is vrij van overtredingen op regels die het migratie-systeem moet bewaken
-% -> het migratie-systeem bevat alle oude data, ihb nog steeds na het toepassen van de enforce regels
-% -> er is een pad naar ingebruikname van het nieuwe systeem (vanaf de initiële toestand van het migratie-systeem)
-% -> corollary: op het moment van ingebruikname van het nieuwe systeem, is het migratie-systeem vrij van overtredingen op regels die het nieuwe systeem moet bewaken
-% -> optioneel: na een begrensd aantal 'voortgangs-stappen' kan het nieuwe systeem in gebruik genomen worden
-% -> optioneel: er bestaat een migratie-systeem dat de oude functionaliteit behoudt (mogelijk uitbreid) totdat het nieuwe systeem in gebruik genomen is?
+%
 
 \title{Data Migration under a Changing Schema}
-\author[ou,ordina]{Stef Joosten\fnref{fn1}}
-\ead{stef.joosten@ou.nl}
-\author[umn]{Sebastiaan Joosten\fnref{fn2}}
-\address[ou]{Open Universiteit Nederland, Heerlen, the Netherlands}
-\address[ordina]{Ordina NV, Nieuwegein, the Netherlands}
-\address[umn]{University of Minnesota, Minneapolis, USA}
-\fntext[fn1]{ORCID 0000-0001-8308-0189}
-\fntext[fn2]{ORCID 0000-0002-6590-6220}
-
+%
+%\titlerunning{Abbreviated paper title}
+% If the paper title is too long for the running head, you can set
+% an abbreviated paper title here
+%
+\author{Sebastiaan Joosten\inst{1}\orcidID{0000-0002-6590-6220}\\ \and
+Stef Joosten\inst{2,3}\orcidID{0000-0001-8308-0189}}
+%
+\authorrunning{S. Joosten}
+% First names are abbreviated in the running head.
+% If there are more than two authors, 'et al.' is used.
+%
+\institute{University of Minnesota, Minneapolis, USA \and Open Universiteit Nederland, Heerlen, the Netherlands\\ 
+\email{stef.joosten@ou.nl} \and
+Ordina NV, Nieuwegein, the Netherlands}
+%
+\maketitle              % typeset the header of the contribution
+%
 \begin{abstract}
 Software generators can help to increase the frequency of releases and their reliability.
-   They save on time spent on development and time spent on fixing human-induced mistakes by compiling a specification into a working information system.
+They save on time spent on development and fixing human-induced mistakes by compiling a specification into a working information system.
 However, many generators do not support data migrations.
-   A data migration is necessary when an incremental deployment changes the schema of the system.
-   As a consequence, developers tend to avoid migrations or migrate data ``by hand''.
+Data migration is necessary when an incremental deployment changes the system's schema.
+Consequently, developers tend to avoid migrations or migrate data ``by hand''.
 
@@
 The problem at large is how to preserve the semantics of that data under a changing schema.
-   This paper proposes a theory for deploying an incremental change.
-   The theory is applicable in general, but will be implemented in a software generator called Ampersand.
+This paper proposes a theory for deploying incremental change.
+The theory is applicable in general but will be implemented in a software generator called Ampersand.
 
 This paper aims to preserve the semantics of data by satisfying concrete business requirements.
-   The migration process is based on the assumptions that
-   software is deployed incrementally,
-   the existing data set may be polluted,
-   human interaction may be required,
-   the meaning of data must be preserved,
-   the business continues during the migration without interruption (zero down-time),
-   and there is a compiler to generate an information system from a given schema.
+The migration process is based on the assumption that software is deployed incrementally, the existing data set may be polluted, human interaction may be required, the meaning of data must be preserved, the business continues during the migration without interruption (zero downtime), and there is a compiler to generate an information system from a given schema.
 The correctness of the migration is the focus of this paper, while efficiency is outside its scope.
+
+\keywords{Generative software \and Incremental software deployment \and Data migration \and Relation algebra \and Ampersand \and Schema change.}
 \end{abstract}
-
-\begin{keyword}
-generative software\sep incremental software deployment\sep data migration\sep relation algebra\sep Ampersand\sep schema change
-\end{keyword}
-\maketitle
-
+%
+%
+%
 \section{Introduction}
@@
 
-   This paper is founded in the belief that software generators help developers to increase the frequency of deploying successive increments.
-   They also prevent some human-induced errors, both in the process and in the resulting software.
-   We believe that software generation can thus lead to a higher pace of deployments, a more predictable deployment process, and more reliable information systems.
-
    A problem is that software generators typically do not support the data migration process, especially when the schema changes.
    The ``manual'' effort required to migrate data slows down the development process and introduces mistakes%
-\footnote{These effects can be measured in terms of DevOps metrics such as
-   deployment frequency,
-   reliability of deployments, and
-   change failure rate~\cite{DevOps2021}.}.
-
-   This paper proposes a theory for data migrations that is meant to be implemented in a software generator called Ampersand~\cite{JoostenRAMiCS2017, Joosten-JLAMP2018}.
+   Such effects can be measured in terms of DevOps metrics such as deployment frequency, reliability of deployments, and change failure rate~\cite{DevOps2021}.
+
+   This paper proposes a theory for data migrations that is meant to be implemented in a software generator called Ampersand~\cite{JoostenRAMiCS2017,Joosten-JLAMP2018}.
    It aims at automating data migrations, so incremental changes can be done more reliably and faster.
@@
    They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
-   Obviously, such data migrations are not suitable for incremental software deployment,
-   if the schema changes and the old functionality has to be replaced.
+   Such data migrations do not allow for schema changes and do not fit our need for incremental software deployment.
    Examples like these have convinced the authors to define a new method for data migration that is specifically meant for systems in production that are
@@
 \subsection{Information Systems}
-   The purpose of an information system is to make data meaningful to its users.
-   The analysis of the problem starts with a formal definition of information systems,
-   in which the schema contains all information that makes data meaningful.
-   We then define a migration procedure from one information system to another,
-   which preserves the meaning of the data as much as possible.
-
-   Every user has her\footnote{Throughout this paper, we ask the reader to read ``she'' or ``her'' to include the masculin form as well.} own tasks and responsibilities
-   and may work from different locations and on different moments.
+   The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
+   This section defines information systems, after which we define a migration procedure from one information system to another.
+
+   Every user has her own tasks and responsibilities and may work from different locations and on different moments.
    This collective use by multiple users serves a purpose which we will loosely call ``the business''.
@@
 
-   Information systems are typically used by actors (both users and computers) who are distributed and work with data all the time.
+   Information systems are typically used by actors (both users and computers) who are distributed and work with data.
    As a consequence, the data in a system changes continually.
-   In practice, actors ``talk to'' the system through an ingress mechanism, which connects each user to the right service(s).
-   The ingress function is provided by the deployment platform, so it is beyond the scope of this paper.
-   Also, every service runs independent of other services,
-   meaning that each service can be stopped, (re)started and substituted without disrupting the other services.
-   For practical purposes, this presumes a platform of managed services, such as Kubernetes,
-   even though the theory in this paper is not specific to any platform.
-
-   Every system contains a data set, which represents the state of the system.
-   Every service produces and consumes events that may change the state of the system.
-   This state is represented in a persistent store, aka the database%
-\footnote{Whether the database is local, remote, or globally distributed is immaterial for the theory in this paper.}.
+   The state of the system is represented by a data set, typically represented in a persistent store or database.
    Events that the system detects may cause the state to change.
@@
 \end{figure}
-   To keep our theory technology independent, data sets are assumed to contains triples.
+   To keep our theory technology independent, data sets are assumed to contain triples.
    This makes our theory valid for any kind of database that triples can represent,
-   such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
-   The system's semantics is represented as constraints,
-   which most database management systems refer to as integrity rules.
-   The purpose of these rules is to keep all constraints satisfied.
-
-   Like most other software generators, Ampersand is a compiler that generates information systems from a script.
-   Ampersand uses a syntactically sugared form of heterogeneous relation algebra,
-   which works with relations in a way similar to Alloy~\cite{Alloy2006} and allegories%
-\footnote{Allegories are a specific type of categories, which the reader does not need to understand for the purpose of this paper}~\cite{Zielinski2013}.
-   Each rule in Ampersand is a constraint (aka invariant), which the system keeps satisfied as long as that rule lives%
-\footnote{The dotted lines in the diagrams show which set of rules is enforced.}.
-   So, developers spend no effort to extract constraints from specifications or (even more laborious) from code
-   because all constraints are explicitly available as rules in the code.
+   including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
+   An important aspect of the system's semantics are the constraints, sometimes known as invariants or integrity rules.
+   %The purpose of these rules is to keep all constraints satisfied.
+
+   Like most software generators, Ampersand is a compiler that generates information systems from a script.
+   Ampersand uses a form of heterogeneous relation algebra,
+   which works with relations in a way similar to Alloy~\cite{Alloy2006}.
+   A system is then described by a set of rules (aka constraint / invariant), which the system keeps satisfied.
+   So all constraints are explicitly available as rules in the Ampersand code.
 
@@
    while preserving the meaning of the present data as much as possible~\cite{Spivak2012}.
-   Just copying the set of data from the existing system to the desired system is obviously wrong if the schemas of both systems differ.
-
+   %Just copying the set of data from the existing system to the desired system is obviously wrong if the schemas of both systems differ.
+   %
    In practice, data migrations are done by deploying the desired system and the existing system side-by-side,
@@
    to ensure the preservation of valuable data and the quality of the migration.
-   To automate this process completely is unrealistic in many cases%
-\footnote{Note that smaller increments improve the chance that a fully automic migration is possible,
-although there will always be cases where human intervention is required.}.
+   Automating this process completely is unrealistic in many cases:
+   While smaller increments improve the chance that a fully automatic migration is possible, there will always be cases where human intervention is required.
    Manual interventions can be necessary to resolve data pollution, new business rules, or to fix known issues in the old system.
-   Consequently, a migration engineer must have the freedom to change the migration script to suit the particulars of the migration.
-   Such complications may arise mostly as a result of data pollution.
+   Consequently, a migration engineer must have the freedom to tailor any automatic approach to migration to suit the particulars.
+
+   Complications that require manual intervention arise mostly as a result of data pollution.
    We distinguish data pollution in the following categories:
@@
    The first category is about violations of constraints from the schema.
-   Such violations do not occur in our theory because the existing system was generated by Ampersand,
+   Such violations do not occur in our theory because we assume the existing system was generated by Ampersand,
    so the data set satisfies all constraints in the schema.
@@
    So, such constraints may help to move this type of pollution to one of the other categories.
-   The last category of data pollution, erroneous constraints, must be fixed by replacing them by the correct constraints in the desired system.
+   The last category of data pollution, erroneous constraints, must be fixed by removing them or replacing them by the correct constraints in the desired system.
 
@@
    \item the moment of transition (the MoT), i.e. the moment that changed functionality is made available to users;
-   \item the moment of cleanup (the MoC), i.e. the moment after which the old software and data can be removed.%
-\footnote{Backup mechanisms are outside the scope of this paper.}.
+   \item the moment of cleanup (the MoC), i.e. the moment after which the old software and data can be removed.
+   This leaves the business with the intended system: a successful migration.
 \end{enumerate}
@@
    During the time between the MoT and MoC,
-   the existing system and the desired system are kept alive, side by side, as shown in figure~\ref{fig:migration phase}.
+   the existing system and the desired system are kept alive, side by side, as shown in Figure~\ref{fig:migration phase}.
 \begin{figure}[bht]
    \begin{center}
-     \includegraphics[scale=.35]{figures/datamigration-Migration phase.png}
+     \includegraphics[scale=.35]{figures/datamigration-Migration-phase.png}
    \end{center}
@@
    This migration schema preserves data from the existing system, it may introduce new data automatically (generated from the existing data set),
-   and it may require users to introduce new data (which will take time). 
-   So the idea of a migration schema is to compute the disjoint union of the existing and the desired schemas
-   and to present it to the migration engineer in the form of source code,
+   and it may require users to introduce new data (after the MoT and before the MoC).
+   So the migration schema is a (disjoint) union of the existing and the desired schemas,
+   which is presented to the migration engineer in the form of source code,
    so the migration engineer can change everything she wants to suit the particulars of the data migration.
 
-   During the migration phase, transactions in the existing system are allowed because the migration engine will transport them to the desired system.
-   This allows the business to finish transactions in the existing system while the desired system is already up and running.
-   Similarly, the existing system must execute transactions from the desired system, just in case the business calls off the migration.
-   Rules in the migration system that require user interaction are given time by requiring that the migration phase ends
-   only after all migration rules are at rest.
-   At the MoC, the migration engineer takes down everything but the desired system,
-   leaving the business with a successful migration.
+%During the migration phase, transactions in the existing system are allowed because the migration engine will transport them to the desired system.
+%This allows the business to finish transactions in the existing system while the desired system is up and running.
+%Similarly, the existing system must execute transactions from the desired system, just in case the business calls off the migration.
+%Rules in the migration system that require user interaction are given time by requiring that the migration phase ends
+%only after all migration rules are at rest.
+
 \begin{figure}[bht]
@@
    An {\em information system} is a combination of data set, schema, and functionality.
-   For the purpose of this paper, the functionality is ignored because it does not impact the migration.
+   For the purpose of this paper, functionality captured in user interfaces are ignored because it does not impact the migration.
    Section~\ref{sct:Data sets} desribes data sets. Schemas are treated in section~\ref{sct:Schemas}.
@@
    Atoms serve as data elements.
-   Atoms are values without internal structure of interest, meant to represent atomic data elements (e.g. dates, strings, numbers, etc.) in a database.
-   From a business perspective, atoms represent concrete items of the world,
-   such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
-   By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ represent \emph{atoms}.
+   %Atoms are values without internal structure of interest, meant to represent atomic data elements (e.g. dates, strings, numbers, etc.) in a database.
+   %From a business perspective, atoms represent concrete items of the world,
+   %such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
+   %By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ represent \emph{atoms}.
    All atoms are taken from an infinite set called $\Atoms$.
-
+   %
    Concepts are names that group atoms of the same type.
    All concepts are taken from an infinite set $\Concepts$.
-   $\Concepts$ and $\Atoms$ are disjoint.
+   %$\Concepts$ and $\Atoms$ are disjoint.
    For example, a developer might choose to classify \atom{Peter} and \atom{Melissa} as \concept{Person},
@@
    In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
-   In the sequel, variables $A$, $B$, $C$, $D$ will represent concepts.
-
-   The relation $\inst:\Pair{\Atoms}{\Concepts}$ relates atoms to concepts.
-   The term $a\inst C$ means that atom $a$ is an \emph{instance} of concept $C$.
-   This relation is used in the type system, in which $\inst$ assigns one or more concepts to every atom in the data set.
-   Since $\inst$ is a relation and every relation is a set of pairs,
-   set operators $\cup$, $\cap$, and $-$ can be used on $\inst$.
+   Moreover, \atom{Peter}, \atom{Melissa} and \atom{074238991} are atoms.
+   In the sequel, variables $A$, $B$, $C$, $D$ will represent concepts, and variables $a$, $b$, and $c$ represent \emph{atoms}.
+   %
+   The relation $\instance:\Pair{\Atoms}{\Concepts}$ relates atoms to concepts.
+   The term $a\instance C$ means that atom $a$ is an \emph{instance} of concept $C$.
+   %This relation is used in the type system, in which $\instance$ assigns one or more concepts to every atom in the data set.
+   %Since $\instance$ is a relation and every relation is a set of pairs,
+   %set operators $\cup$, $\cap$, and $-$ can be used on $\instance$.
 
@@
 %    \label{eqn:specialization}
-%    A\isa B\ \Leftrightarrow\ \forall a: a\inst A\rightarrow a\inst B
+%    A\isa B\ \Leftrightarrow\ \forall a: a\instance A\rightarrow a\instance B
 % \end{equation}
@@
    Relations serve to organize and store data, to allow a developer to represent facts.
-   In this paper, variables $r$, $s$, and $d$ represent relations.
+   In this paper, variables $r$, $s$, and $d$ represent relations\footnote{Some readers might like to read `relation symbol' where we write `relation'.}.
    All relations are taken from an infinite set $\Rels$.
@@
 
-%    The following is for wide tables only
-%    Specialization distributes in a straightforward way over signatures:
-% \begin{equation}
-%    \pair{A}{B}\isa\pair{C}{D}\ \Leftrightarrow\ A\isa C\ \wedge\ B\isa D
-% \label{eqn:isa-signature}
-% \end{equation}
-
    Triples serve to represent data.
-   A triple\footnote{Please note that this paper uses the word {\em triple} in a more restricted way than in natural language.}
+   A triple %\footnote{Please note that this paper uses the word {\em triple} in a more restricted way than in natural language.}
    is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.
@@
 
-   Let $\Dataset$ be the set of all pairs $\pair{\triples}{\inst}$, so
-   $\Dataset$ is $\Pair{(\Triple{\Atoms}{\Rels}{\Atoms})}{(\Pair{\Atoms}{\Concepts})}$.
-   A data set $\dataset$ is a tuple $\pair{\triples}{\inst}$ that satisfies:
+   \begin{definition}[Data set]
+   A data set $\dataset$ is a tuple $\pair{\triples}{\instance}$ with $\triples \subseteq {\Triple{\Atoms}{\Rels}{\Atoms}}$ and $\instance \subseteq  {\Pair{\Atoms}{\Concepts}}$ that satisfies:
 \begin{eqnarray}
-   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&a\inst A\ \wedge\ b\inst B
+   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&a\instance A\ \wedge\ b\instance B
    \label{eqn:wellTypedEdge}
 \end{eqnarray}
+\end{definition}
    Looking at the example,
    equation~\ref{eqn:wellTypedEdge} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
-   In practice, users can say that Peter has telephone number 074238991.
+   In practice, users can say that the Person Peter has telephone number 074238991.
    So, the ``thing'' that \atom{Peter} refers to (which is Peter) has \atom{074238991} as a telephone number.
-   This ``meaning from practice'' has no consequences in the formal world.
-   Users are free to attach any practical meaning to a triple.
-
-   The notations $\triples_{\dataset}$ and $\inst_{\dataset}$ are used to disambiguate $\triples$ and $\inst$ when necessary.
+   The notations $\triples_{\dataset}$ and $\instance_{\dataset}$ are used to disambiguate $\triples$ and $\instance$ when necessary.
    To save writing in the sequel, the notation $a\ r\ b$ means that $\triple{a}{r}{b}\in\triples$.
@@
 \end{equation}
-   Equation~\ref{eqn:wellTypedEdge} implies that for every data set $\dataset$:
-\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ a\inst_{\dataset}A\ \wedge\ b\inst_{\dataset}B\]
-   For a developer, this means that the type of an atom depends only on the relation in which it resides; not on the actual population of the database.
-
+%   Equation~\ref{eqn:wellTypedEdge} implies that for every data set $\dataset$:
+%\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ a\instance_{\dataset}A\ \wedge\ b\instance_{\dataset}B\]
+%   For a developer, this means that the type of an atom depends only on the relation in which it resides; not on the actual population of the database.
+%
    Concepts too can be seen as containers of atoms,
@@
 \begin{equation}
-   \pop{C}{\dataset}\ =\ \{ x\mid\ x\ \inst_{\dataset}\ C\}
+   \pop{C}{\dataset}\ =\ \{ x\mid\ x\ \instance_{\dataset}\ C\}
 \label{eqn:pop-concept}
@@
    Schemas serve to capture the semantics of an information system~\cite{Spivak2012}.
-   A schema defines concepts, specializations, relations, and rules.
+   We consider a schema to represent the static part of an information system, defining concepts, relations, and rules.
    A software engineer defines a schema on design time, so that semantic checks can be implemented at compile time.
 
-   A schema $\schema$ is a tuple $\triple{\concepts}{\rels}{\rules}$,
-   in which $\concepts$ is a finite set of concepts,
-   $\rels$ is a finite set of relations,
-   and $\rules$ is a finite set of rules.
+   We describe a schema $\schema$ via a tuple $\triple{\concepts}{\rels}{\rules}$,
+   in which $\concepts\subseteq \Concepts$ is a finite set of concepts,
+   $\rels\subseteq \Rels$ is a finite set of relations,
+   and $\rules\subseteq \Rules$ is a finite set of rules.
+
    Each rule in a schema serves to constrain the data set at runtime, to ensure its semantic integrity.
-   Every rule is an element of an infinite set called $\Rules$,
-   which is disjoint from $\Atoms$, $\Concepts$, $\Rels$, and $\Dataset$.
+   Every rule is an element of an infinite set called $\Rules$.
    In this paper, variables $u$ and $v$ represent rules.
+   For every rule $u$ in the schema, we assume there is a function $\violC{u}$ such that $\viol{u}{\dataset}$ represents any violations to $u$ in data set $\dataset$.
+   If there are no violations, that is: $\viol{u}{\dataset} = \emptyset{}$, then we say that $u$ is satisfied.
+
    The notation $\schema_{\infsys}$ refers to the schema of information system $\infsys$.
-   Disambiguation sometimes requires to write $\concepts_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$
-   rather than $\concepts$, $\rels$, and $\rules$ respectively.
-   For every rule $u$ in the schema, there is a predicate $\sat{u}{\dataset}$ that constrains data set $\dataset$.
-%  Every schema is an element of an infinite set called $\Schema$.
-
-   A schema must ensure that triples ``make sense'' in the semantics defined by the schema.
-   This means that it must satisfy:
+   When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$
+   for $\concepts$, $\rels$, and $\rules$ corresponding to $\schema$.
+
+   \begin{definition}[Schema]
+   A schema is a tuple $\triple{\concepts}{\rels}{\rules}$ that satisfies:
 \begin{eqnarray}
@@
    \label{eqn:relationsIntroduceConcepts}
-%    A\ \isa\ B&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
-%    \label{eqn:isasIntroduceConcepts}\\
-%    The following is for wide tables only
-%    \pair{A}{B}\isa\pair{C}{D}\ \wedge\ \declare{n}{C}{D}\in\rels&\Rightarrow&\declare{n}{A}{B}\in\rels
-%    \label{eqn:subrelations}\\
-%    \isa\ \text{is a partial order}
-%    \label{eqn:isasPartialOrder}\\
-%    \exists\dataset\in\Dataset-\emptyset:\ \sat{u}{\dataset}
-%    \label{eqn:consistentRules}
 \end{eqnarray}
+   \end{definition}
    Requirement~\ref{eqn:relationsIntroduceConcepts} ensures that concepts mentioned in a relation are defined in the schema.
-%    Requirement~\ref{eqn:isasIntroduceConcepts} ensures that the specialization relation works on concepts of $\concepts$.
-%    The following is for wide tables only
-%    Requirement~\ref{eqn:subrelations} ensures that all specializations of a relation exist.
-%    Requirement~\ref{eqn:isasPartialOrder} ensures that the specialization relation contains no cycles.
-%    Requirement~\ref{eqn:consistentRules} ensures that every rule is consistent.
-
    The type system of Ampersand~\cite{vdWoude2011} ensures that every information system it generates satisfies this requirement.
-   Note that these requirements do not depend on any particular data set,
-   so they can be checked at compile time.
-   This is also known as static typing,
-   which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.
+   This is also known as static typing, which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.
+
+%   Note that these requirements do not depend on any particular data set,
+%   so they can be checked at compile time.
+%   This is also known as static typing,
+%   which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.
 
@@
 % \begin{eqnarray}
-%    \sign{u}=\pair{A}{B}\ \wedge\ \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\inst A\wedge b\inst B
+%    \sign{u}=\pair{A}{B}\ \wedge\ \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\instance A\wedge b\instance B
 %    \label{eqn:wellTypedEdged violations}
@@
 \label{sct:Information Systems}
-   An information system has a changing data set because of events that occur.
-   Whenever the system ``observes'' an event, it attempts to change its data set accordingly.
-   However, every rule in the schema represents a constraint,
-   which the data set must invariably satisfy.
-   That is why rules are also known as ``invariants''.
-
-   Events are taken from an infinite set, $\Events$.
-   An event is denoted as $\evt{d^-}{d^+}$, in which $d^-$ and $d^+$ are data sets.
-   Every event is a function $\Dataset\rightarrow\Dataset$, so $\evt{d^-}{d^+}(\dataset)$ is a data set.
-   Equation~\ref{eqn:event} defines the effect of an event on a data set.
-\begin{eqnarray}
-      \evt{d^-}{d^+}(\dataset)&=&\la(\triples_{\dataset}-\triples_{d^-})\cup\triples_{d^+},(\inst_{\dataset}-\inst_{d^-})\cup\inst_{d^+}\ra
-\label{eqn:event}
-\end{eqnarray}
-   In English, we would say that an event $\evt{d^-}{d^+}$ removes the triples in $d^-$ from $\dataset$ and then inserts the triples in $d^+$ into the result.
-   In the sequel, variables $e$, $e'$, $e''$, etc. represent events.
-   As soon as a system $\infsys$ observes an event $e$,
-   it applies $e$ to its data set $\dataset$.
-   Yet, if $e(\dataset)$ does not satisfy a rule $u$ from schema $\schema_\infsys$,
-   another event is needed, say $e'$, to ensure that $\sat{u}{e'(e(\dataset))}$.
-   The phrase ``to restore an invariant'' or ``to restore invariance of a rule''
-   means that the system will change $e(\dataset)$ by applying another event $e'$ to satisfy requirement~\ref{eqn:enforce}.
-   The notation $\enf_{u,e}$ represents this combination of events $e$ and $e'$.
-   So, $\enf_{u,e}$ must satisfy:
-\begin{eqnarray}
-   \sat{u}{\dataset}&\Rightarrow&\sat{u}{\enforce{u,e}{\dataset}}
-\label{eqn:enforce}
-\end{eqnarray}
+%   An information system has a changing data set because of events that occur.
+%   Whenever the system ``observes'' an event, it attempts to change its data set accordingly.
+%   However, every rule in the schema represents a constraint,
+%   which the data set must invariably satisfy.
+%   That is why rules are also known as ``invariants''.
+%
+%   Events are taken from an infinite set, $\Events$.
+%   An event is denoted as $\evt{d^-}{d^+}$, in which $d^-$ and $d^+$ are data sets.
+%   Every event is a function $\Dataset\rightarrow\Dataset$, so $\evt{d^-}{d^+}(\dataset)$ is a data set.
+%   Equation~\ref{eqn:event} defines the effect of an event on a data set.
+%\begin{eqnarray}
+%      \evt{d^-}{d^+}(\dataset)&=&\la(\triples_{\dataset}-\triples_{d^-})\cup\triples_{d^+},(\instance_{\dataset}-\instance_{d^-})\cup\instance_{d^+}\ra
+%\label{eqn:event}
+%\end{eqnarray}
+%   In English, we would say that an event $\evt{d^-}{d^+}$ removes the triples in $d^-$ from $\dataset$ and then inserts the triples in $d^+$ into the result.
+%   In the sequel, variables $e$, $e'$, $e''$, etc. represent events.
+%   As soon as a system $\infsys$ observes an event $e$,
+%   it applies $e$ to its data set $\dataset$.
+%   Yet, if $e(\dataset)$ does not satisfy a rule $u$ from schema $\schema_\infsys$,
+%   another event is needed, say $e'$, to ensure that $\sat{u}{e'(e(\dataset))}$.
+%   The phrase ``to restore an invariant'' or ``to restore invariance of a rule''
+%   means that the system will change $e(\dataset)$ by applying another event $e'$ to satisfy requirement~\ref{eqn:enforce}.
+%   The notation $\enf_{u,e}$ represents this combination of events $e$ and $e'$.
+%   So, $\enf_{u,e}$ must satisfy:
+%\begin{eqnarray}
+%   \sat{u}{\dataset}&\Rightarrow&\sat{u}{\enforce{u,e}{\dataset}}
+%\label{eqn:enforce}
+%\end{eqnarray}
 
@@
 \begin{itemize}
-   \item data set $\dataset=\pair{\triples}{\inst}$ is defined as in section~\ref{sct:Data sets};
-   \item schema $\schema=\triple{\concepts}{\rels}{\rules}$ is defined as in section~\ref{sct:Schemas};
-   \item enforce function ${\enf}:\Rules\rightarrow\Events\rightarrow\Dataset$ satisfies requirement~\ref{eqn:enforce}.
+   \item $\dataset=\pair{\triples}{\instance}$ is a data set (satisfies Equation~\ref{eqn:wellTypedEdge});
+   \item $\schema=\triple{\concepts}{\rels}{\rules}$ is a schema (satisfies Equation~\ref{eqn:relationsIntroduceConcepts});
+   \item all rules are satisfied:
+   \begin{eqnarray}
+   \forall u\in\rules&:\viol{u}{\dataset}=\emptyset
+   \label{eqn:satisfaction}
+   \end{eqnarray}
+   \item triples in the data set have their relation mentioned in the schema:
+   \begin{eqnarray}
+   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
+   \label{eqn:define R}
+   \end{eqnarray}
 \end{itemize}
@@
 
-   The information system itself enforces the semantics by ensuring that the following requirements remain true at all times:
-\begin{eqnarray}
-   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
-   \label{eqn:define R}\\
-   \forall u\in\rules&:&\sat{u}{\dataset}
-   \label{eqn:satisfaction}
-% TODO:
-% De engine heeft een eigen rol. Alle regels die aan de engine zijn gekoppeld, worden automatisch onderhouden.
-%    The following is for wide tables only
-%    \pair{A}{B}\isa\pair{C}{D}\ \wedge\ \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\triple{a}{\declare{n}{C}{D}}{b}\in\triples
-%    \label{eqn:subpopulations}
-\end{eqnarray}
-   Requirement~\ref{eqn:define R} ensures that the information system only works with triples whose relation is defined in the schema.
-   Requirement~\ref{eqn:satisfaction} ensures that the data set satisfies all rules.
-%    The following is for wide tables only
-%    Requirement~\ref{eqn:subpopulations} ensures that specialization works across relations.
-   Note that these requirements apply at runtime.
-   So, as long as a rule $u$ ``lives'' in an information system, the data set will satisfy that rule.
-
-   As the system registers events, it inserts or deletes triples in its data set.
-   If such an event would violate requirement~\ref{eqn:satisfaction},
-   either the system stays in the state $\dataset$ or there is a reaction $e'$ that restores the affected invariants:
-\begin{equation}
-   \sat{u}{\dataset}\Rightarrow\sat{u}{e'_{d'}(e_{d}(\dataset))}
-   \label{eqn:restoring}
-\end{equation}
-   In this way, the system can restore invariants and keep requirement~\ref{eqn:satisfaction} satisfied.
-   The combination of events $e$ and $e'$ is called a {\em transaction}.
+   As the system responds to events, it changes its data set, while the schema remains the same.
+   In this paper, it suffices to define an event as a pair of information systems for which the schema stays constant.
+   Events are categorized by what part of the data set changes.
+
+   \begin{definition}[Event]
+   Let $R \subseteq \Rels$ and $I \subseteq \Rels$ be sets of relations, and let $\infsys = \pair{\schema}{\dataset}$ and $\infsys' = \pair{\schema}{\dataset'}$ be information systems (so $\infsys$ and $\infsys'$ have the same schema), and:
+      \begin{align}
+      \triples_{\dataset} - (\Triple{\Atoms}{(R \cup I)}{\Atoms}) &= \triples_{\dataset'} - (\Triple{\Atoms}{(R \cup I)}{\Atoms})
+   \label{eqn:eventUnchanged}\\
+      \triples_{\dataset} - (\Triple{\Atoms}{R}{\Atoms}) &\subseteq \triples_{\dataset'} - (\Triple{\Atoms}{R}{\Atoms})
+   \label{eqn:eventInsert}
+   \end{align}
+   Then we say that $\pair{\infsys}{\infsys'}$ is an event with scope $R$ inserting in $I$.
+   We use the notation $\infsys \xrightarrow[I]{R} \infsys'$ to indicate this.
+   \end{definition}
+
+   Equation~\ref{eqn:eventUnchanged} states that the triples of those for relations in $I$ or $R$ are the only ones to have changed, and Equation~\ref{eqn:eventInsert} states that of the ones that are only in $I$, no triples were removed.
+   If $I$ or $R$ is the empty set, we omit it from the arrow, so $\infsys \xrightarrow{R} \infsys'$ is a notation for $\infsys \xrightarrow[\emptyset]{R} \infsys'$.
+
+   \begin{definition}[Inserting event]
+   We write $\infsys \xrightarrowdbl[I]{R} \infsys'$ to state that some triple was inserted whose relation was in $I$, that is:
+   $t \in \triples_{\dataset'} \cap (\Triple{\Atoms}{I}{\Atoms})$ and $t \not\in \triples_{\dataset}$ for some triple $t$.
+   \end{definition}
+
+   We use inserting events to model the effect of a type of rule that is particular to Ampersand, known as an enforce rule.
+   The syntax is as follows:
+
+\begin{align}
+      \text{\tt ENFORCE}\ r\ \text{\tt >:}\ \id{term}\label{enforce ins}
+\end{align}
+
+   Adding an enforce rule to the script of an information system specifies a rule $u$ with $\viol{u}{\dataset} = \pop{\id{term}}{\dataset} - \pop{r}{\dataset}$, and the information system ensures that the rule is satisfied by adding a triple $\triple{x}{r}{y}$ for each $\pair{x}{y} \in \viol{u}{\dataset}$, thus maintaining $\pop{r}{\dataset} \supseteq \pop{\id{term}}{\dataset}$ as an invariant.
+   If $R$ is a set of relations that covers those that occur in $\id{term}$, then changes to an information system $\infsys$ can be described by the event $\infsys \xrightarrow[\{r\}]{R} \infsys'$.
 
@@
    % Requirement~\ref{eqn:specialization} is satisfied because this example contains no specialization.
-   The Ampersand compiler generates a data set $\dataset$, which contains a set of triples and a relation $\inst$.
+   The Ampersand compiler generates a data set $\dataset$, which contains a set of triples and a relation $\instance$.
    It defines the set of triples $\triples$ as:
@@
 \end{array}\]
-The relation $\inst$ contains the pairs:
+The relation $\instance$ contains the pairs:
 \[\begin{array}{l}
@@
 \end{array}\]
-   The pair $\pair{\triples}{\inst}$ satisfies requirement~\ref{eqn:wellTypedEdge} so this is a data set $\dataset$ as introduced in section~\ref{sct:Data sets}.
+   The pair $\pair{\triples}{\instance}$ satisfies requirement~\ref{eqn:wellTypedEdge} so this is a data set $\dataset$ as introduced in section~\ref{sct:Data sets}.
 
@@
 
-\subsection{Process of Data Migration}
-   Sections~\ref{sct:Information Systems} and~\ref{sct:Example existing IS} introduce a single information system.
-   The transition from an existing system $\infsys$ to the desired one $\infsys'$ is the topic of this section.
-   Let us first sketch the steps to migrate system $\infsys$ to $\infsys'$.
-\begin{enumerate}
-   \item Let the existing information system $\infsys$ have data set $\dataset$ and schema $\schema$.
-         A developer defines a new information system, $\infsys'$, which contains a data set of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{In this paper, we only consider the functionality captured by rules.} as illustrated in figure~\ref{fig:post-migration}.
-         After the migration, the resulting system will be $\infsys'$, enhanced with the triples from $\dataset$ that satisfy $\schema'$,
-         and with additional changes made by the migration engineer (see step~\ref{item3}) and by users who have been using the system during the migration.
-         $\dataset'$ may contain some new data,
-         for instance, to initialize new features at the moment of transition (MoT).
-         If schemas $\schema$ equals $\schema'$, the data migration is unnecessary, so in the sequel we assume they are different.
-   \item The migration engineer uses a software generator to generate a migration script, $\migrsys$,
-         This is a system that includes both $\infsys$ and $\infsys'$.
-         $\migrsys$ is meant to migrate the data and define the semantics during the migration phase.
-         It contains enforcement rules that migrate existing data to the desired data set.
-         It also contains data structures and rules that are needed for the duration of the migration phase.
-%         The generated script must be free of type errors, to ensure that $\sat{u}{\dataset}$ exists for every rule $u$ and data set $\dataset$.
-%         The proof is available \href{https://www.isa-afp.org/}{here}
-   \item\label{item3}
-         In many cases, the schema $\schema_{\migrsys}$ will be enough to perform the migration.
-         However, a migration engineer may have specific requirements that call for changes in $\migrsys$.
-         For that purpose, the generator generates the migration script as an Ampersand script,
-         so the migration engineer can accommodate such requirements.
-         This can be necessary, for example, to resolve data pollution that violates a constraint that was not in schema $\schema_{\infsys}$.
-         She can either implement enforcement rules or use the business to resolve the issues.
-         She can test the resulting migration script, $\migrsys'$ separately before taking the desired system to production.
-         The migration engineer alters $\schema$ nor $\schema'$ while adapting $\migrsys'$ to the specific needs of the current migration.
-   \item Then, the migration engineer deploys system $\migrsys'$.
-         This event starts the copying of data from $\dataset$ to $\dataset'$, as specified in $\migrsys'$ (figure~\ref{fig:migration phase}).
-         During this phase, the ingress is still referring users to the existing system,
-         so they will hardly notice that the desired system is being fired up and loaded with data.
-         Users are still changing $\dataset$ (both inserts and deletes),
-         but these changes are being transferred to $\dataset'$ as well.
-         This step is complete when all data from $\dataset$ has been copied to $\dataset'$ and all invariants of the migration system are true.
-   \item At this moment, the ingress switches the incoming traffic from $\infsys$ to $\migrsys'$,
-         so $\migrsys'$ effectively takes over.
-         This event marks the MoT.
-         Since $\infsys$ may still have some violations due to last-minute edits,
-         it may take some (short but finite) time until the invariants of $\infsys$ are true.
-         After that, dataset $\dataset$ will no longer change.
-   \item From this moment, users experience the functionality of the migration system $\migrsys'$.
-         This system must ensure that all invariants of $\schema'$ are true before the migration is complete.
-         It no longer ensures the rules of $\schema$, which govern the existing system.
-         Since $\dataset$ is no longer changing, the migration engineer can now focus on the remaining violations of $\schema'$.
-         Users of the migration system will deal with process rules in the migration system that are still violated.
-         Once all invariants of $\migrsys'$ are true, the migration is complete.
-         This state marks the moment of cleanup (MoC)
-   \item After the MoC, the ingress switches the incoming traffic from $\migrsys'$ to $\infsys'$.
-         The migration system is no longer needed and can be taken out of production, together with the existing system $\infsys$.
-         The desired system remains in production and the migration is finished.
-         This establishes the post-migration situation (figure~\ref{fig:post-migration}).
-\end{enumerate}
+%\subsection{Process of Data Migration}
+%   Section~\ref{sct:Information Systems} introduces a single information system.
+%   The transition from an existing system $\infsys$ to the desired one $\infsys'$ is the topic of this section.
+%   Let us first sketch the steps to migrate system $\infsys$ to $\infsys'$.
+%\begin{enumerate}
+%   \item Let the existing information system $\infsys$ have data set $\dataset$ and schema $\schema$.
+%         A developer defines a new information system, $\infsys'$, which contains a data set of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{In this paper, we only consider the functionality captured by rules.} as illustrated in Figure~\ref{fig:post-migration}.
+%         After the migration, the resulting system will be $\infsys'$, enhanced with the triples from $\dataset$ that satisfy $\schema'$,
+%         and with additional changes made by the migration engineer (see step~\ref{item3}) and by users who have been using the system during the migration.
+%         $\dataset'$ may contain some new data,
+%         for instance, to initialize new features at the moment of transition (MoT).
+%         If schemas $\schema$ equals $\schema'$, the data migration is unnecessary, so in the sequel we assume they are different.
+%   \item The migration engineer uses a software generator to generate a migration script, $\migrsys$,
+%         This is a system that includes both $\infsys$ and $\infsys'$.
+%         $\migrsys$ is meant to migrate the data and define the semantics during the migration phase.
+%         It contains enforcement rules that migrate existing data to the desired data set.
+%         It also contains data structures and rules that are needed for the duration of the migration phase.
+%%         The generated script must be free of type errors, to ensure that $\sat{u}{\dataset}$ exists for every rule $u$ and data set $\dataset$.
+%%         The proof is available \href{https://www.isa-afp.org/}{here}
+%   \item\label{item3}
+%         In many cases, the schema $\schema_{\migrsys}$ will be enough to perform the migration.
+%         However, a migration engineer may have specific requirements that call for changes in $\migrsys$.
+%         For that purpose, the generator generates the migration script as an Ampersand script,
+%         so the migration engineer can accommodate such requirements.
+%         This can be necessary, for example, to resolve data pollution that violates a constraint that was not in schema $\schema_{\infsys}$.
+%         She can either implement enforcement rules or use the business to resolve the issues.
+%         She can test the resulting migration script, $\migrsys'$ separately before taking the desired system to production.
+%         The migration engineer alters $\schema$ nor $\schema'$ while adapting $\migrsys'$ to the specific needs of the current migration.
+%   \item Then, the migration engineer deploys system $\migrsys'$.
+%         This event starts the copying of data from $\dataset$ to $\dataset'$, as specified in $\migrsys'$ (Figure~\ref{fig:migration phase}).
+%         During this phase, the ingress is still referring users to the existing system,
+%         so they will hardly notice that the desired system is being fired up and loaded with data.
+%         Users are still changing $\dataset$ (both inserts and deletes),
+%         but these changes are being transferred to $\dataset'$ as well.
+%         This step is complete when all data from $\dataset$ has been copied to $\dataset'$ and all invariants of the migration system are true.
+%   \item At this moment, the ingress switches the incoming traffic from $\infsys$ to $\migrsys'$,
+%         so $\migrsys'$ effectively takes over.
+%         This event marks the MoT.
+%         Since $\infsys$ may still have some violations due to last-minute edits,
+%         it may take some (short but finite) time until the invariants of $\infsys$ are true.
+%         After that, data set $\dataset$ will no longer change.
+%   \item From this moment, users experience the functionality of the migration system $\migrsys'$.
+%         This system must ensure that all invariants of $\schema'$ are true before the migration is complete.
+%         It no longer ensures the rules of $\schema$, which govern the existing system.
+%         Since $\dataset$ is no longer changing, the migration engineer can now focus on the remaining violations of $\schema'$.
+%         Users of the migration system will deal with process rules in the migration system that are still violated.
+%         Once all invariants of $\migrsys'$ are true, the migration is complete.
+%         This state marks the moment of cleanup (MoC)
+%   \item After the MoC, the ingress switches the incoming traffic from $\migrsys'$ to $\infsys'$.
+%         The migration system is no longer needed and can be taken out of production, together with the existing system $\infsys$.
+%         The desired system remains in production and the migration is finished.
+%         This establishes the post-migration situation (Figure~\ref{fig:post-migration}).
+%\end{enumerate}
 
@@
 %    \label{def:graph_union_wellTyped_if_parts_wellTyped}
-%    \[\pair{\triples_1}{\inst_1}\cup\pair{\triples_2}{\inst_2}\ =\ \pair{\triples_1\cup\triples_2}{\inst_1\cup\inst_2}\]
+%    \[\pair{\triples_1}{\instance_1}\cup\pair{\triples_2}{\instance_2}\ =\ \pair{\triples_1\cup\triples_2}{\instance_1\cup\instance_2}\]
 %    \end{definition}
@@
 %    \[\begin{array}{rcl}
-%      \maprel{h}{\pair{\triples}{\inst}}&=&\pair{\maprel{h}{\triples}}{\inst}\\
+%      \maprel{h}{\pair{\triples}{\instance}}&=&\pair{\maprel{h}{\triples}}{\instance}\\
 %      \maprel{h}{\triple{a}{r}{b}}&=&\triple{a}{h(r)}{b}
@@
 %    \begin{eqnarray}
-%    \pair{\triples}{\inst}\sqcup\pair{\triples'}{\inst'}&=&\pair{\triples\uplus\triples'}{\inst\uplus^2\inst'}\\
+%    \pair{\triples}{\instance}\sqcup\pair{\triples'}{\instance'}&=&\pair{\triples\uplus\triples'}{\instance\uplus^2\instance'}\\
 %      X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
@@
 %    % \begin{eqnarray}
-%    %    \omit\rlap{$\la\atoms,\concepts,\inst,\isa,\rels,\dataset\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\dataset'\ra$}\notag\\
-%    %    &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\instuplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \dataset\uplus^5\dataset'\ra\notag\\
+%    %    \omit\rlap{$\la\atoms,\concepts,\instance,\isa,\rels,\dataset\ra\sqcup\la\atoms',\concepts',\instance',\isa',\rels',\dataset'\ra$}\notag\\
+%    %    &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\instanceuplus^2\instance',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \dataset\uplus^5\dataset'\ra\notag\\
 %    %       X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
@@
 %    \end{definition}
-%    \begin{definition}[relabel concepts in $\inst$]
-%    \[a\ \subst{C}{D}{\inst}\ \subst{C}{D}{A}\ \Leftrightarrow\ a\inst A\]
+%    \begin{definition}[relabel concepts in $\instance$]
+%    \[a\ \subst{C}{D}{\instance}\ \subst{C}{D}{A}\ \Leftrightarrow\ a\instance A\]
 %    \end{definition}
@@
 %    \[\begin{array}{rcl}
-%      \rlap{$\subst{C}{D}{\pair{\triples}{\inst}}$}\\
-%      &=&\pair{\{\subst{C}{D}{t}\mid t\in\triples\}}{\subst{C}{D}{\inst}}
+%      \rlap{$\subst{C}{D}{\pair{\triples}{\instance}}$}\\
+%      &=&\pair{\{\subst{C}{D}{t}\mid t\in\triples\}}{\subst{C}{D}{\instance}}
 %    \end{array}\]
@@
 %    \[\begin{array}{rcl}
-%      \rlap{$\subst{r'}{m'}{\pair{\triples}{\inst}}$}\\
-%      &=&\pair{\{\subst{r'}{m'}{t}\mid t\in\triples\}}{\inst}
+%      \rlap{$\subst{r'}{m'}{\pair{\triples}{\instance}}$}\\
+%      &=&\pair{\{\subst{r'}{m'}{t}\mid t\in\triples\}}{\instance}
 %    \end{array}\]
@@
 
-\section*{Generating a Migration Script}
+\section{Generating a Migration Script}
+
 \subsection{Loading Data}
-   Before explaining how to migrate data from $\dataset$ to $\dataset'$, we need to explain {\tt ENFORCE} rules.
-   In Ampersand, an enforcement rule comes in three flavors:
-\begin{eqnarray}
-   \text{\tt ENFORCE}\ r\ \text{\tt :<}\ \id{term}\label{enforce del}\\
-   \text{\tt ENFORCE}\ r\ \text{\tt :=}\ \id{term}\label{enforce eq}\\
-   \text{\tt ENFORCE}\ r\ \text{\tt >:}\ \id{term}\label{enforce ins}
-\end{eqnarray}
-   The run time environment contains an engine that enforces such rules.
-   The engine will change the population of relation $r$ by the least set of pairs needed to satisfy the rule.
-   An enforcement rule with operator {\tt :<} (alternative~\ref{enforce del}) keeps the population of $r$ smaller than or equal to \id{term}.
-   When the population of this term is losing pairs that are in $r$, the engine will remove such pairs from $r$ to keep $r$ smaller than or equal to \id{term}.
-   An enforcement rule with operator {\tt >:} (alternative~\ref{enforce ins}) keeps the population of $r$ larger than or equal to \id{term}.
-   When the population of this term is growing with pairs that are not in $r$, the engine will insert such pairs in $r$ to keep $r$ greater than or equal to \id{term}.
-   An enforcement rule with operator {\tt :=} (alternative~\ref{enforce eq}) keeps the population of $r$ equal to \id{term} by inserting and deleting the right pairs in $r$.
-   Summarizing, an enforcement rule can be read as an invariant that keeps one of $r\leq\id{term}$, $r=\id{term}$, or $r\geq\id{term}$ satisfied by changing the population of $r$.
-
-   Now, let us study how to migrate data from $\dataset$ to $\dataset'$.
-   It would be tempting to state that a relation from the example system, say {\tt takes}, could be copied by the following code fragment%
+   Let us study how to migrate data from $\dataset$ to $\dataset'$.
+   It would be tempting to state that a relation from the example system, say {\tt isPartOf}, could be copied by the following code fragment%
 \footnote{The prefix {\tt old.} refers to concepts, relations, and rules from the existing system (section~\ref{sct:Example existing IS}).
@@
 \begin{verbatim}
-   ENFORCE new.takes := old.takes
+   ENFORCE new.isPartOf >: old.isPartOf
 \end{verbatim}
-   The relation `old.takes' resides in the existing system and will not change anymore.
-
-   This means, however, that {\tt old.takes} determines {\tt new.takes} forever.
+   The relation `old.isPartOf' resides in the existing system and will not change after the MoT.
+
+   This means, however, that anything in {\tt old.isPartOf} cannot be deleted from {\tt new.isPartOf}.
    This is undesirable because it prevents us from changing data in the migration system.
-   The desired behavior is that all pairs in {\tt old.takes} are copied to {\tt new.takes} when the system is first initialized, but that {\tt new.takes} can be changed freely after that.
+   The desired behavior is that all pairs in {\tt old.isPartOf} are copied to {\tt new.isPartOf} when the system is first initialized, but that {\tt new.isPartOf} can be changed freely after that.
    An extra relation (\ref{extraRelation}) achieves this behavior by remembering which pairs have been copied:
 \begin{eqnarray}
-   &&\verb#RELATION copied_takes[Student*Course]#\label{extraRelation}\\
-   &&\verb#ENFORCE new.takes >: old.takes - copied_takes#\label{difference}\\
-   &&\verb#ENFORCE copied_takes := new.takes#\label{fill copies with new.takes}\\
-   &&\verb#ENFORCE copied_takes :< old.takes#\label{del copies from old.takes}
+   &&\verb#RELATION copied_isPartOf[Student*Course]#\label{extraRelation}\\
+   &&\verb#ENFORCE new.isPartOf >: old.isPartOf - copied_isPartOf#\label{difference}\\
+   &&\verb#ENFORCE copied_isPartOf >: new.isPartOf#\label{fill copies with new.isPartOf}
 \end{eqnarray}
-   Understanding what the engine does to enforce these rules might help the reader, so let us look what happens at runtime.
-   Initially, when {\tt copied\_takes} is still empty, the difference between {\tt old.takes} and {\tt copied\_takes} equals {\tt old.takes}.
-   So, {\tt new.takes} will start filling up with the pairs of {\tt old.takes} by rule~\ref*{difference}.
-   Then, the engine starts filling {\tt copied\_takes} with pairs from {\tt new.takes} because of rule~\ref*{fill copies with new.takes}.
-   As a result, the difference between {\tt old.takes} and {\tt copied\_takes} gets smaller and smaller.
-   Once {\tt old.takes} and {\tt copied\_takes} are equal, rule~\ref*{difference} will no longer insert pairs in {\tt new.takes}.
-   While this copying process is going on, {\tt old.takes} may accumulate more pairs or lose pairs.
+
+   Let us look what happens at runtime.
+   Initially, when {\tt copied\_isPartOf} is still empty, the difference between {\tt old.isPartOf} and {\tt copied\_isPartOf} equals {\tt old.isPartOf}.
+   So, {\tt new.isPartOf} will contain the pairs of {\tt old.isPartOf} by rule~\ref*{difference}.
+   Then, the engine inserts the same pairs into {\tt copied\_isPartOf} because of rule~\ref*{fill copies with new.isPartOf}.
+   After this, triples with the relation {\tt new.isPartOf} can be inserted and deleted freely.
+
+   \subsection{Dealing with violations on invariants}
+   Suppose the script for the desired system adds an invariant that is not present in the existing system:
+
+   \begin{verbatim}
+   RELATION isPartOf[Module*Course] [UNI]
+   \end{verbatim}
+
+   The \verb=UNI= annotation requires that each \verb=Module= is part of at most one \verb=Course=.
+   The violations are given by:
+   \begin{eqnarray}
+   \viol{\tt UNI}{\dataset} = \{\pair{x}{y} \mid x \neq y \wedge \exists z.~ \pair{z}{x}, \pair{z}{y}\in\pop{\tt isPartOf}{\dataset} \}
+   \end{eqnarray}
+   In Ampersand, the term that gives these violations is: \verb=isPartOf~;isPartOf - I=.
+
+   While the \verb=UNI= requirement holds for the population in the original script, that population may have changed.
+   Therefore, we must assume that we cannot expect that \verb=new.isPartOf= satisfies the \verb=UNI= requirement in the desired system.
+   The migration system allows for this by simply not requiring that they hold at the MoT.
+   A challenge lies in ensuring that they do hold at the MoC, and that users that help the migration forward can achieve a state in which all requirements hold.
+
+   Two extra relations are used: one to track the violations that occur in the original data (\ref{violRelation}), and one to track the violations that have been fixed (\ref{fixedRelation}) which prevents them from recurring.
+   \begin{align}
+   &\text{\tt RELATION viols\_UNI[Course*Course]}\label{violRelation}\\
+   &\text{\tt RELATION fixed\_UNI[Course*Course]}\label{fixedRelation}\\
+   &\text{\tt ENFORCE viols\_UNI >: old.isPartOf\textasciitilde;old.isPartOf - I}\label{fill viols}\\
+   &\text{\tt ENFORCE fixed\_UNI >: viols\_UNI - (old.isPartOf\textasciitilde;old.isPartOf - I)}\label{fill fixed}\\
+   &\begin{aligned}
+   \text{\tt RULE isPartOf\_UNI: old.isPartOf\textasciitilde;old.isPartOf}\\
+   \quad\text{\tt |- I \textbackslash/ (viols\_UNI - fixed\_UNI)}
+   \end{aligned}\label{rule uni}
+   \end{align}
+
+   We again look at what happens at runtime:
+   Rule~\ref{fill viols} will adds all violations present in the old system to {\tt viols\_UNI}.
+   Until the MoT, rule~\ref{fill fixed} does nothing, since {\tt viols\_UNI} equals $\viol{\tt UNI}{\dataset}$.
+   At that point, rule~\ref{rule uni} is satisfied, since {\tt fixed\_UNI} is empty.
+   As violations are fixed, rule~\ref{fill fixed} records this by adding to {\tt fixed\_UNI}.
+   This way, previously fixed errors cannot reoccur.
+   Once {\tt viols\_UNI} is equal to {\tt fixed\_UNI}, the \verb=UNI= requirement holds and the information system is ready for the MoC in regard to this rule.
 
@@
 
-\subsubsection{Strategies for restoring invariance}
-
-We allow ExamRule2 to be broken during the transition to the desired system, and expect people with the Administrator role to deal with them.
-
-However, the violations that arise when taking the (non-disjoint) union of the two scripts include violations of rules that the system should maintain.
-This means that if we cannot generate software based on the disjoint union as it is.
-To resolve this, the simplest solution is to change the roles assigned to the rules that are violated in the disjoint union:
-\begin{verbatim}
-   RELATION isPartOf[Module*Course]
-   RULE isPartOfUNI: isPartOf;isPartOf~ |- I
-   ROLE migrationHelper MAINTAINS isPartOfUNI
-\end{verbatim}
-
-Naturally, this solution requires effort from people with the role migrationHelper.
-This in itself is not an issue: problems need to be solved one way or another.
-However, a problem that may occur is that regular users of the system, who are used to the existing system, might be adding data to the system that causes violations faster than they can be solved.
-To avoid this problem, we restrict the behavior of the system such that incremental progress will lead to eventually reaching the goal.
-
-Our solution is to state that the only violations that may occur in the relation \verb=isPartOfUNI= are those that were present when the migration started:
-\begin{verbatim}
-   RELATION isPartOf[Module*Course]
-   RULE isPartOfUNI: isPartOf;isPartOf~ |- I
-   ROLE migrationHelper MAINTAINS isPartOfUNI
-
-   RELATION isPartOfUNI_violations[Module*Module]
-     = [("IT-Governance", "IT-Governance")]
-   RULE isPartOfUNI_progress: -(isPartOf;isPartOf~) /\ I
-         |- isPartOfUNI_violations
-   SYSTEM MAINTAINS isPartOfUNI_progress
-
-   ENFORCE isPartOfUNI_violations :< -(isPartOf;isPartOf~) /\ I  
-\end{verbatim}
-
-In this solution, the first rule is the same as with the simple solution.
-The second rule prevents any new violation from occurring.
-Existing violations are recorded in the \verb=isPartOfUNI_violations= relation, and the rule \verb=isPartOfUNI_progress= prevents the set of violations to \verb=isPartOfUNI= from growing beyond this.
-The \verb=ENFORCE= statement at the end of this code snippet states that the relation recording existing violations should be shrunk whenever violations are solved.
-This way, violations cannot re-occur.
-As a whole, this means that \verb=migrationHelper= has a finite task.
+%\subsubsection{Strategies for restoring invariants}
+%
+%We allow ExamRule2 to be broken during the transition to the desired system, and expect people with the Administrator role to deal with them.
+%
+%However, the violations that arise when taking the (non-disjoint) union of the two scripts include violations of rules that the system should maintain.
+%This means that if we cannot generate software based on the disjoint union as it is.
+%To resolve this, the simplest solution is to change the roles assigned to the rules that are violated in the disjoint union:
+%\begin{verbatim}
+%   RELATION isPartOf[Module*Course]
+%   RULE isPartOfUNI: isPartOf;isPartOf~ |- I
+%   ROLE migrationHelper MAINTAINS isPartOfUNI
+%\end{verbatim}
+%
+%Naturally, this solution requires effort from people with the role migrationHelper.
+%This in itself is not an issue: problems need to be solved one way or another.
+%However, a problem that may occur is that regular users of the system, who are used to the existing system, might be adding data to the system that causes violations faster than they can be solved.
+%To avoid this problem, we restrict the behavior of the system such that incremental progress will lead to eventually reaching the goal.
+%
+%Our solution is to state that the only violations that may occur in the relation \verb=isPartOfUNI= are those that were present when the migration started:
+%\begin{verbatim}
+%   RELATION isPartOf[Module*Course]
+%   RULE isPartOfUNI: isPartOf;isPartOf~ |- I
+%   ROLE migrationHelper MAINTAINS isPartOfUNI
+%   
+%   RELATION isPartOfUNI_violations[Module*Module]
+%     = [("IT-Governance", "IT-Governance")]
+%   RULE isPartOfUNI_progress: -(isPartOf;isPartOf~) /\ I
+%         |- isPartOfUNI_violations
+%   SYSTEM MAINTAINS isPartOfUNI_progress
+%   
+%   ENFORCE isPartOfUNI_violations :< -(isPartOf;isPartOf~) /\ I  
+%\end{verbatim}
+%
+%In this solution, the first rule is the same as with the simple solution.
+%The second rule prevents any new violation from occurring.
+%Existing violations are recorded in the \verb=isPartOfUNI_violations= relation, and the rule \verb=isPartOfUNI_progress= prevents the set of violations to \verb=isPartOfUNI= from growing beyond this.
+%The \verb=ENFORCE= statement at the end of this code snippet states that the relation recording existing violations should be shrunk whenever violations are solved.
+%This way, violations cannot re-occur.
+%As a whole, this means that \verb=migrationHelper= has a finite task.
 
@@
 %The ${\tt ENFORCE\ }r{\tt\ :=\ }e$ statements add triples $(x,r,y)$ for all $(x,y)\in e$.
-%The terms $t$ in these statements are such that $x\ (\inst \compose \kleenestar{\flip{\isa}})\ (A',1)$ and $y \ (\inst \compose \kleenestar{\flip{\isa}})\ (B',1)$, thus preserving the property that the result is a data set.
+%The terms $t$ in these statements are such that $x\ (\instance \compose \kleenestar{\flip{\isa}})\ (A',1)$ and $y \ (\instance \compose \kleenestar{\flip{\isa}})\ (B',1)$, thus preserving the property that the result is a data set.
 %
@@
 
-\subsection{Changes}
-% The purpose of this section is to explain why a data set is structured the way it is.
-   In the migration of a data set we deal with changes to the elements that are not data:
-   $\inst$, $\concepts$, and $\rels$.
-   Such changes have further reaching consequences, however.
-   Changes to $\inst$, $\concepts$, $\rels$ change the data structure in a data set.
-   We define a \define{migration of a data set} $\la\atoms,\concepts,\inst,\rels,\dataset\ra$ as a change in which one or more of $\inst$, $\concepts$, or $\rels$ change.
-   To satisfy the definition of data sets, $\dataset$ must change too,
-   but the migration should try to preserve the maximal amount of data.
-
-\section{Validation}
-   We claim that the information system does not need to be paused in order for the migration to happen successfully.
-   By this, we mean that computations that take a considerable amount of time, as well as human actions, can be performed incrementally while the system is in use.
-   In order for us to prove this claim, we model the behavior of our information system as a series of systems $\migrsys_0,\migrsys_1,\ldots,\migrsys_n$ with $\migrsys = \migrsys_0$, where only the dataset of the system changes.
-   We consider two types of updates to the dataset.
-
-   The first type of update helps the migration to go forward: we'll write $\rightsquigarrow$ for those.
-   The second type of update is that which other users of the system do (possibly none), indicated as $\dashrightarrow$.
-   The two updates alternate: $\migrsys_0 \dashrightarrow \migrsys_1 \rightsquigarrow \migrsys_2 \dashrightarrow \ldots \rightsquigarrow \migrsys_{2i} \dashrightarrow \migrsys_{2i+1}\rightsquigarrow\migrsys_{2i+2}\dashrightarrow\ldots\migrsys_n$.
-
-   We validate our claim that the information system will reach the MoC by arguing about two situations:
-   The system between deployment and the MoT can reach the MoT by firing the ENFORCE rules, and the system between the MoT and the MoC can reach the MoC after a finite number of actions by the migration engineer.
-   More concretely, this means the following:
-
-When the default migration system $\migrsys$ is deployed, the system will always be in a state such that continually running any violated ENFORCE rules (in any order) leads to the MoT.
-   This means that $A\rightsquigarrow B$ is such that the dataset of $B$ is that of $A$ with an update applied according to one or more violated enforce rules. % TODO: define 'violated'
-   If there is no such $B$ because no enforce rules are violated, we are at the moment of transition.
-   We say that the system $A$ is in a state such that the MoT can be reached if there are no infinite $\rightsquigarrow$-paths from $A$.
-   The constraint on $A\dashrightarrow B$ is that only the old system can change at this point.
-   Note that this does not prove that the moment of transition to never be reached, as more enforce rules may get violated at every $\dashrightarrow$ step.
-   We expect this not to be a problem in practice since $\dashrightarrow$ transitions are triggered by users while $\rightsquigarrow$ is triggered automatically through ENFORCE rules: the ENFORCE rules should outperform the $\dashrightarrow$ actions by a big margin.
-
-After the moment of transition, there is a finite upper-bound on the amount of work the migration engineer has to do:
-   $A\rightsquigarrow B$ is such that the dataset of $B$ is that of $A$ with one of the violations shown to a migration engineer resolved (including the firing of corresponding ENFORCE statements), and $\dashrightarrow$ arbitrary.
-   In both $A\rightsquigarrow B$ and $A\dashrightarrow B$ we assume that the relations that were specifically created for the bookkeeping (suffixed by \verb=_progress= and \verb=_violations=) can only be updated through the firing of the ENFORCE statements.
-   These assumptions make the total number of pairs in \verb=_violations= relations decline in every $\rightsquigarrow$-step, proving that the MoC will be reached after a finite number of $\rightsquigarrow$ steps (assuming that the set of violations is finite).
-   Moreover, we show that the desired system with the dataset taken from the migration system (after dropping the old system and the bookkeeping) is free of violations when all \verb=_violations= relations are empty.
-
-\section{Software architecture}
-   The theory presented above roughly corresponds to the back-end of a migration system.
-   This section describes its software architecture.
-   For this research, we have used the Ampersand compiler to generate the system(s).
-   Figure~\ref{fig:initial} shows the starting point of the migration.
-\begin{figure}
-   \centering
-   \includegraphics[scale=0.1]{figures/Before migration.png}
-   \caption{Initial state of the migration}
-   \label{fig:initial}
-\end{figure}
-   For the purpose of data migration, the existing system must make its schema available through its API.
-   For this purpose, the back-end offers a service ``retrieve schema'', which provides the schema as Ampersand source-code.
-   As a result, the migration engineer can rely on the schema of any generated application to be accurate.
-   For this reason, we can safely assume that the schema this service provides has not been changed since it was deployed.
-   Its presence allows the generator to build various front-ends and migration scripts during the lifetime of the existing system%
-
-   After the migration engineer has produced and tested a migration script, the Ampersand compiler produces a migration system.
-   It includes the schemas of both the existing system and the desired system, besides having a schema of its own.
-   This is implemented by Ampersand's INCLUDE mechanism, which boils down to taking the disjoint union of the schemas.
-\begin{figure}
-   \centering
-   \includegraphics[scale=0.1]{figures/During migration.png}
-   \caption{During the migration}
-   \label{fig:during}
-\end{figure}
-   Deployment of the migration system (figure~\ref{fig:during}) leaves the existing system intact.
-   It enhances the system with the data set needed by the migration system, which includes the data set needed by the desired system.
-   So, users will notice no difference in the existing system when the migration system is deployed.
-
-   When all rules in the migration script have been satisfied, the migration system is ready to be deployed.
-   At this point, the ingress switches traffic from the existing system to the migration system.
-   Until this moment, users have been updating data in the existing system, which is continuously being migrated to the desired system.
-   After the ingress switch is made, users have the new functionality at their disposal and make their changes in the desired system.
-   If for some reason, there is remaining migration work, the migration system will execute this.
-   Since there is no more input in the existing system, the migration system will eventually be consistent.
-   At this point, the existing system can be removed, which leaves the migration system in place.
-
-   The main purpose of the migration script is to define the migration rules that are used to migrate data from the existing system to the desired system.
-   During this process, users continu to use the existing system.
-
-\begin{figure}
-   \centering
-   \includegraphics[scale=0.1]{figures/After migration.png}
-   \caption{After the migration}
-   \label{fig:after}
-\end{figure}
+%\subsection{Changes}
+%% The purpose of this section is to explain why a data set is structured the way it is.
+%   In the migration of a data set we deal with changes to the elements that are not data:
+%   $\instance$, $\concepts$, and $\rels$.
+%   Such changes have further reaching consequences, however.
+%   Changes to $\instance$, $\concepts$, $\rels$ change the data structure in a data set.
+%   We define a \define{migration of a data set} $\la\atoms,\concepts,\instance,\rels,\dataset\ra$ as a change in which one or more of $\instance$, $\concepts$, or $\rels$ change.
+%   To satisfy the definition of data sets, $\dataset$ must change too,
+%   but the migration should try to preserve the maximal amount of data.
+
+%\section{Validation}
+%   We claim that the information system does not need to be paused in order for the migration to happen successfully.
+%   By this, we mean that computations that take a considerable amount of time, as well as human actions, can be performed incrementally while the system is in use.
+%   In order for us to prove this claim, we model the behavior of our information system as a series of systems $\migrsys_0,\migrsys_1,\ldots,\migrsys_n$ with $\migrsys = \migrsys_0$, where only the data set of the system changes.
+%   We consider two types of updates to the data set.
+%   
+%   The first type of update helps the migration to go forward: we'll write $\rightsquigarrow$ for those.
+%   The second type of update is that which other users of the system do (possibly none), indicated as $\dashrightarrow$.
+%   The two updates alternate: $\migrsys_0 \dashrightarrow \migrsys_1 \rightsquigarrow \migrsys_2 \dashrightarrow \ldots \rightsquigarrow \migrsys_{2i} \dashrightarrow \migrsys_{2i+1}\rightsquigarrow\migrsys_{2i+2}\dashrightarrow\ldots\migrsys_n$.
+%   
+%   We validate our claim that the information system will reach the MoC by arguing about two situations:
+%   The system between deployment and the MoT can reach the MoT by firing the ENFORCE rules, and the system between the MoT and the MoC can reach the MoC after a finite number of actions by the migration engineer.
+%   More concretely, this means the following:
+%   
+%When the default migration system $\migrsys$ is deployed, the system will always be in a state such that continually running any violated ENFORCE rules (in any order) leads to the MoT.
+%   This means that $A\rightsquigarrow B$ is such that the data set of $B$ is that of $A$ with an update applied according to one or more violated enforce rules. % TODO: define 'violated'
+%   If there is no such $B$ because no enforce rules are violated, we are at the moment of transition.
+%   We say that the system $A$ is in a state such that the MoT can be reached if there are no infinite $\rightsquigarrow$-paths from $A$.
+%   The constraint on $A\dashrightarrow B$ is that only the old system can change at this point.
+%   Note that this does not prove that the moment of transition to never be reached, as more enforce rules may get violated at every $\dashrightarrow$ step.
+%   We expect this not to be a problem in practice since $\dashrightarrow$ transitions are triggered by users while $\rightsquigarrow$ is triggered automatically through ENFORCE rules: the ENFORCE rules should outperform the $\dashrightarrow$ actions by a big margin.
+%
+%After the moment of transition, there is a finite upper-bound on the amount of work the migration engineer has to do:
+%   $A\rightsquigarrow B$ is such that the data set of $B$ is that of $A$ with one of the violations shown to a migration engineer resolved (including the firing of corresponding ENFORCE statements), and $\dashrightarrow$ arbitrary.
+%   In both $A\rightsquigarrow B$ and $A\dashrightarrow B$ we assume that the relations that were specifically created for the bookkeeping (suffixed by \verb=_progress= and \verb=_violations=) can only be updated through the firing of the ENFORCE statements.
+%   These assumptions make the total number of pairs in \verb=_violations= relations decline in every $\rightsquigarrow$-step, proving that the MoC will be reached after a finite number of $\rightsquigarrow$ steps (assuming that the set of violations is finite).
+%   Moreover, we show that the desired system with the data set taken from the migration system (after dropping the old system and the bookkeeping) is free of violations when all \verb=_violations= relations are empty.
+%
+%
+%\section{Software architecture}
+%   The theory presented above roughly corresponds to the back-end of a migration system.
+%   This section describes its software architecture.
+%   For this research, we have used the Ampersand compiler to generate the system(s).
+%   Figure~\ref{fig:initial} shows the starting point of the migration.
+%\begin{figure}
+%   \centering
+%   \includegraphics[scale=0.1]{figures/Before migration.png}
+%   \caption{Initial state of the migration}
+%   \label{fig:initial}
+%\end{figure}
+%   For the purpose of data migration, the existing system must make its schema available through its API.
+%   For this purpose, the back-end offers a service ``retrieve schema'', which provides the schema as Ampersand source-code.
+%   As a result, the migration engineer can rely on the schema of any generated application to be accurate.
+%   For this reason, we can safely assume that the schema this service provides has not been changed since it was deployed.
+%   Its presence allows the generator to build various front-ends and migration scripts during the lifetime of the existing system%
+%
+%   After the migration engineer has produced and tested a migration script, the Ampersand compiler produces a migration system.
+%   It includes the schemas of both the existing system and the desired system, besides having a schema of its own.
+%   This is implemented by Ampersand's INCLUDE mechanism, which boils down to taking the disjoint union of the schemas.
+%\begin{figure}
+%   \centering
+%   \includegraphics[scale=0.1]{figures/During-migration.png}
+%   \caption{During the migration}
+%   \label{fig:during}
+%\end{figure}
+%   Deployment of the migration system (Figure~\ref{fig:during}) leaves the existing system intact.
+%   It enhances the system with the data set needed by the migration system, which includes the data set needed by the desired system.
+%   So, users will notice no difference in the existing system when the migration system is deployed.
+%
+%   When all rules in the migration script have been satisfied, the migration system is ready to be deployed.
+%   At this point, the ingress switches traffic from the existing system to the migration system.
+%   Until this moment, users have been updating data in the existing system, which is continuously being migrated to the desired system.
+%   After the ingress switch is made, users have the new functionality at their disposal and make their changes in the desired system.
+%   If for some reason, there is remaining migration work, the migration system will execute this.
+%   Since there is no more input in the existing system, the migration system will eventually be consistent.
+%   At this point, the existing system can be removed, which leaves the migration system in place.
+%
+%
+%   The main purpose of the migration script is to define the migration rules that are used to migrate data from the existing system to the desired system.
+%   During this process, users continu to use the existing system.
+%
+%\begin{figure}
+%   \centering
+%   \includegraphics[scale=0.1]{figures/After migration.png}
+%   \caption{After the migration}
+%   \label{fig:after}
+%\end{figure}
 
@@
 \end{itemize}
-\section{Bibliography}
-\bibliographystyle{elsarticle-harv}
+%\section{Bibliography}
+\bibliographystyle{splncs04}
 \bibliography{doc}

### 2023-07-06 2022Migration/articleMigration.tex

@@
    As a consequence, the data in a system changes continually.
-   In practice, actors ``talk to'' the system through an ingress mechanism, which connects each user to the right service(s).
-   The ingress function is provided by the deployment platform, so it is beyond the scope of this paper.
-   Also, every service runs independent of other services,
+   In practice, actors ``talk to'' the system through an ingress mechanism,
+   which routes the traffic between each user and the required application service(s).
+   The ingress function is used during migration to switch users from the old to new functionality.
+   In this research we assume that the deployment platform provides the ingress mechanism,
+   so developers merely have to configure it.
+   Also, we assume that every service runs independent of other services,
    meaning that each service can be stopped, (re)started and substituted without disrupting the other services.
@@
    \centering
-   \includegraphics[scale=0.1]{figures/Before migration.png}
+   \includegraphics[scale=0.09]{figures/Before migration.png}
    \caption{Initial state of the migration}
@@
       We assume that schemas $\schema$ and $\schema'$ are different, or else the migration would be unnecessary.
-\item From schemas $\schema$ and $\schema'$, a migration engineer%
+\item A migration engineer%
 \footnote{The migration engineer may also stand for a team, which may overlap with the developer role.}
-      generates a migration script, using the Ampersand compiler.
-      This script specifies the migration system, $\migrsys$, that includes $\infsys$, $\infsys'$ and a number of migration rules.
-      $\migrsys$ is meant to migrate the data and ensure that $\dataset'$ satisfies all invariants of $\schema'$.
+      uses the Ampersand compiler to generate a migration script from schemas $\schema$ and $\schema'$.
+      This script specifies a migration system, $\migrsys$, that includes $\infsys$, $\infsys'$ and a number of migration rules.
+      $\migrsys$ is meant to migrate the data and ensure that $\dataset'$ satisfies all invariants of $\schema'$
+      before taking $\infsys'$ to production.
 \item\label{item3}
-      In specific cases, the schema $\schema_{\migrsys}$ will be enough to perform the migration automatically.
-      As soon as the data is transferred and $\migrsys$ has satisfied the invariants of $\schema'$,
-      $\infsys'$ can take care of production on its own without the help of $\migrsys$.
+      In specific cases, $\migrsys$ suffices to perform the migration automatically.
       However, a migration engineer may have specific requirements that call for changes in $\migrsys$.
-      For that purpose, she may want to alter the migration script to accommodate such requirements.
+      For that purpose, she may want to alter the migration script to accommodate such requirements, yielding another migration script $\migrsys'$.
       This can be necessary, for example, to resolve data pollution that violates a constraint that was not in schema $\schema_{\infsys}$.
       She can implement such adaptations as enforcement rules, which are executed automatically.
-      But she can also ask business users to resolve issues by specifying process rules.
+      But she can also choose to specify process rules that require business users to resolve issues interactively.
       Human involvement is often needed to implement a desired migration strategy.
@@
       The migration engineer must ensure that $\migrsys'$ can satisfy all invariants in $\schema'$, either by enforcement rules or by process rules.
-      She can test the resulting migration script, $\migrsys'$ separately before taking the desired system to production.
+      She can test $\migrsys'$ separately before taking the desired system to production.
       When this work is done, the migration engineer has a migration system $\migrsys'$ with schema $\schema'$ and an initial dataset $\dataset'$
@@
 \item\label{item4}
-      Then, the migration engineer deploys system $\migrsys'$.
-\begin{figure}
+      At this moment, let us call it the moment of deployment (MoD),
+\begin{figure}[h]
    \centering
-   \includegraphics[scale=0.1]{figures/4MoD.png}
+   \includegraphics[scale=0.09]{figures/4MoD.png}
    \caption{During the migration}
@@
 \end{figure}
-      Let us call this moment the moment of deployment (MoD).
+      the migration engineer deploys system $\migrsys'$.
       This event starts the copying of data from $\dataset$ to $\dataset'$, as specified in $\migrsys'$.
-      During this phase, the ingress is still referring users to the existing system,
+      During this phase, the ingress still refers users to the existing system,
       so they will hardly notice that the desired system is being fired up and loaded with data.
-      During this phase, users can still change $\dataset$ (both inserts and deletes) while $\infsys$ is keeping its invariants true.
+      Users can still change data in $\dataset$, while $\infsys$ is keeping its invariants true.
       Yet, the migration system also transfers these ``new'' changes to $\dataset'$ by means of its enforcement rules.
@@
       so $\migrsys'$ effectively takes over.
-\begin{figure}
+\begin{figure}[h]
    \centering
-   \includegraphics[scale=0.1]{figures/5MoT.png}
+   \includegraphics[scale=0.09]{figures/5MoT.png}
    \caption{Situation after MoT}
@@
       After the MoC, the ingress can switch the incoming traffic from $\migrsys'$ to $\infsys'$ because all invariants of $\infsys'$ are true.
+      Simultaneously, all clients must switch to the functionality of $\infsys'$.
       $\infsys$ and $\migrsys'$ are no longer needed and can be taken out of production.

### 2023-07-07 2022Migration/articleMigration.tex
WIP

@@
    Software generators can help to increase the frequency of releases and their reliability.
-   They save on time spent on development and time spent on fixing human-induced mistakes by compiling a specification into a working information system.
+   They save on time spent on development and fixing human-induced mistakes by compiling a specification into a working information system.
    However, many generators do not support data migrations.
-   A data migration is necessary when an incremental deployment changes the schema of the system.
-   As a consequence, developers tend to avoid migrations or migrate data ``by hand''.
+   Data migration is necessary when an incremental deployment changes the system's schema.
+   Consequently, developers tend to avoid migrations or migrate data ``by hand''.
 
@@
    The problem at large is how to preserve the semantics of that data under a changing schema.
-   This paper proposes a theory for deploying an incremental change.
-   The theory is applicable in general, but will be implemented in a software generator called Ampersand.
+   This paper proposes a theory for deploying incremental change.
+   The theory is applicable in general but will be implemented in a software generator called Ampersand.
 
    This paper aims to preserve the semantics of data by satisfying concrete business requirements.
-   The migration process is based on the assumptions that
+   The migration process is based on the assumption that
    software is deployed incrementally,
@@
    the meaning of data must be preserved,
-   the business continues during the migration without interruption (zero down-time),
+   the business continues during the migration without interruption (zero downtime),
    and there is a compiler to generate an information system from a given schema.
@@
 
-   This paper proposes a theory for data migrations that is meant to be implemented in a software generator called Ampersand~\cite{JoostenRAMiCS2017, Joosten-JLAMP2018}.
-   It aims at automating data migrations, so incremental changes can be done more reliably and faster.
+   This paper proposes a theory for data migrations in general. 
+   It will be implemented in a software generator called Ampersand~\cite{JoostenRAMiCS2017, Joosten-JLAMP2018},
+   but the theory is kept general to make it applicable in other contexts as well.
+   It aims at automating data migrations, so developers can change systems in production more reliably and faster.
    This contributes to a more agile software development process.
-   We believe that users will experience a more organic evolution of the system because increments are smaller, occur more frequently and more reliably.
+   We believe that end users will experience a more organic evolution of the system because increments are smaller, occur more frequently and more reliably.
 
@@
 \subsection{Information Systems}
-   The purpose of an information system is to make data meaningful to its users.
-   The analysis of the problem starts with a formal definition of information systems,
-   in which the schema contains all information that makes data meaningful.
+   Our analysis of data migration starts with a definition of information systems.
+   This comprises a schema, which contains all information that makes data meaningful.
    We then define a migration procedure from one information system to another,
@@
 
-   Every user has her\footnote{Throughout this paper, we ask the reader to read ``she'' or ``her'' to include the masculin form as well.} own tasks and responsibilities
-   and may work from different locations and on different moments.
-   This collective use by multiple users serves a purpose which we will loosely call ``the business''.
-   It is this business that depends on the semantics of the data to draw the right conclusions and carry out their tasks.
-   This paper uses semantic constraints to represent this meaning.
-
-   Information systems are typically used by actors (both users and computers) who are distributed and work with data all the time.
+   The purpose of an information system is to make data meaningful to its users.
+   These users work from different locations and on different moments. They work with data that the system keeps (syn: stores, maintains).
    As a consequence, the data in a system changes continually.
+   Other computers can use the system too. So, we will use the word ``actor'' to refer to both users and computers.
+   The collective use by multiple actors serves a purpose which we will loosely call ``the business''.
+   It is this business that relies on the semantics of the data for achieving its goals and fulfilling its purpose.
+
+   Besides data, the system also contains services that actors can use to change the data.
+   For the data migration we need services that can be switched on and off independently of each other,
+   so we can switch old and new functionalities on and off as needed in the process of migrating the data.
    In practice, actors ``talk to'' the system through an ingress mechanism,
    which routes the traffic between each user and the required application service(s).
-   The ingress function is used during migration to switch users from the old to new functionality.
+   We need the ingress function during a migration to switch users from the old to new functionality.
    In this research we assume that the deployment platform provides the ingress mechanism,
    so developers merely have to configure it.
-   Also, we assume that every service runs independent of other services,
-   meaning that each service can be stopped, (re)started and substituted without disrupting the other services.
-   For practical purposes, this presumes a platform of managed services, such as Kubernetes,
-   even though the theory in this paper is not specific to any platform.
-
-   Every system contains a data set, which represents the state of the system.
+
+   In this work, we represent the semantics of an information system as a set of constraints on the data in the system.
+   Every constraint a service keeps satisfied is called an invariant.
+   So, deploying a service means that the system will keep the invariants of that service satisfied as long as it is alive.
+   It also means that the invariants of a service are prerequisites for its deployment.
+   These constraints are part of the schema of the information system.
+   The data contained in a system represents the state of that system.
    Every service produces and consumes events that may change the state of the system.
@@
 
+   To try out and implement our theory, we use a software generator called Ampersand~\cite{JoostenRAMiCS2017, Joosten-JLAMP2018}.
    Like most other software generators, Ampersand is a compiler that generates information systems from a script.
+   The choice for Ampersand is motivated by its proximity to the theory in this paper, which makes it easy to implement the theory.
    Ampersand uses a syntactically sugared form of heterogeneous relation algebra,
@@
 \footnote{The dotted lines in the diagrams show which set of rules is enforced.}.
-   So, developers spend no effort to extract constraints from specifications or (even more laborious) from code
-   because all constraints are explicitly available as rules in the code.
-
-   This, and the absence of imperative code in an Ampersand script, makes Ampersand a suitable platform on which to implement the theory.
-   It also allows us to be explicit about ``preserving the meaning as much as possible''.
-   An Ampersand script contains just enough information to generate a complete system,
-   which means that a classical database schema (i.e.\ data structure plus semantics) can be extracted from the Ampersand script.
+   Ampersand generates all program code needed in the system from a script, the Ampersand script,
+   in which a developer only specifies rules (semantic constraints). The generator does the rest of the programming.
+   So, developers need notag extract constraints from specifications or (even more laborious) from program code
+   because they have already specified every constraint explicitly as a rule in the Ampersand script.
+
+   Our theory can be implemented in an ordinary programming language as well, although it might require more effort.
 
@@
    and to present it to the migration engineer in the form of source code,
-   so the migration engineer can change everything she wants to suit the particulars of the data migration.
+   so the migration engineer can change everything she\footnote{Throughout this paper, we ask the reader to read ``she'' or ``her'' to include the masculin form as well.} wants to suit the particulars of the data migration.
 

### 2023-07-07 2022Migration/articleMigrationFACS.tex
Dealing with violations update: added warnings

@@
    &\begin{aligned}
-   \text{\tt RULE isPartOf\_UNI: old.isPartOf\textasciitilde;old.isPartOf}\\
+   \text{\tt RULE isPartOf\_UNI: new.isPartOf\textasciitilde;new.isPartOf}\\
    \quad\text{\tt |- I \textbackslash/ (viols\_UNI - fixed\_UNI)}
@@
    Once {\tt viols\_UNI} is equal to {\tt fixed\_UNI}, the \verb=UNI= requirement holds and the information system is ready for the MoC in regard to this rule.
+
+   We also wish to alert users with the role `migration helper' to violations that need to be fixed.
+   For the original \verb=UNI= requirement, this means specifying:
+
+   \begin{align}
+   \begin{aligned}
+   \text{\tt RULE isPartOf\_UNI\_warn: new.isPartOf\textasciitilde;new.isPartOf |- I}\\
+   \quad\text{\tt RULE MigrationHelper}
+   \end{aligned}\label{warnings}
+   \end{align}
+
+   Despite the syntax, Line~{\ref{warnings}} is not a rule of the generated information system:
+   By assigning a role, the responsibility of maintaining the rule no longer lies with the information system.
+   Instead, users with the migration helper role can see the violations and act on it as they see fit.
 

### 2023-07-10 2022Migration/articleMigrationFACS.tex
wip

@@
    A problem is that software generators typically do not support the data migration process, especially when the schema changes.
-   The ``manual'' effort required to migrate data slows down the development process and introduces mistakes%
+   The ``manual'' effort required to migrate data slows down the development process and introduces mistakes.
    Such effects can be measured in terms of DevOps metrics such as deployment frequency, reliability of deployments, and change failure rate~\cite{DevOps2021}.

### 2023-07-10 2022Migration/articleMigrationFACS.tex
sync Migration FACS article

--- 2023-07-07 2022Migration/articleMigrationFACS.tex
@@
    \item {\em human intervention}\\The data migration may require human interaction, which may take time.
-   \item {\em semantic continuity}\\the meaning of data must be preserved.
+   %\item {\em semantic continuity}\\the meaning of data must be preserved.
    \item {\em zero down-time}\\The business continues during the migration without interruption.
@@
 \label{def:information system}
-\item An information system $\infsys$ is a tuple $\triple{\dataset}{\schema}{\enf}$, in which
+\item An information system $\infsys$ is a tuple $\pair{\dataset}{\schema}$, in which
 \begin{itemize}
-   \item $\dataset=\pair{\triples}{\instance}$ is a data set (satisfies Equation~\ref{eqn:wellTypedEdge});
-   \item $\schema=\triple{\concepts}{\rels}{\rules}$ is a schema (satisfies Equation~\ref{eqn:relationsIntroduceConcepts});
+   \item $\dataset=\pair{\triples}{\instance}$ is a data set (satisfies Equation~\ref{eqn:wellTypedEdge}), we write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$;
+   \item $\schema=\triple{\concepts}{\rels}{\rules}$ is a schema (satisfies Equation~\ref{eqn:relationsIntroduceConcepts}), we write $\concepts_\infsys = \concepts$, $\rels_\infsys = \rels$ and $\rules_\infsys=\rules$;
    \item all rules are satisfied:
@@
 \section{Generating a Migration Script}
+   A migration script takes care of two things: copying data and ensuring that the migration system is a proper information system.
+   We first illustrate how this works by example, using Ampersand syntax.
+   We can then describe what this means in terms of the definitions of information system.
 
@@
    \text{\tt RULE isPartOf\_UNI\_warn: new.isPartOf\textasciitilde;new.isPartOf |- I}\\
-   \quad\text{\tt RULE MigrationHelper}
+   \quad\text{\tt ROLE MigrationHelper}
    \end{aligned}\label{warnings}
@@
 
-\subsection{Example}
-   Let us proceed to specify a desired system, $\infsys'$, to illustrate a (toy) migration.
-   The example in section~\ref{sct:Example existing IS}, called $\infsys$, serves as the existing system.
-   Its population reflects the state of the system just before the migration.
-
-   Information system $\infsys'$, the desired system, is specified by:
-\begin{verbatim}
-   RELATION takes[Student*Course]
-   RELATION isPartOf[Module*Course] [UNI]
-   RELATION isEnrolledFor[Student*Module]
-   RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
-
-   RELATION course[ExamReg*Course] [UNI] =
-      [ ("ER1", "Management")
-      ; ("ER2", "Business IT")
-      ; ("ER3", "Business IT")
-      ]
-   RELATION student[ExamReg*Student] [UNI] =
-      [ ("ER1", "Peter")
-      ; ("ER2", "Susan")
-      ]
-   RULE ExamRule1: student~;course |- takes
-   RULE ExamRule2: student~;course;isPartOf~ |- isEnrolledFor
-\end{verbatim}
-   In this example, the relations and the rule of the existing system remain part of the desired system%
-\footnote{Removal or changing of relations is allowed as well.}
-   because the developer has copied these items into the script of the desired system.
-   However, she has imposed the restriction \verb-UNI- on the relation {\tt isPartOf}, stating that this relation is univalent.
-   This means that every \verb-Module- is part of at most one \verb-Course-.
-   Besides, two new relations and two rules have appeared.
-
-   The desired system, $\infsys'$, contains exam registrations (concept \verb-ExamReg-), which is new.
-   In an exam registration, a student registers for the examination of a course.
-   $\infsys'$ contains two extra rules: \verb-ExamRule1- and \verb-ExamRule2-.
-   The first one says that an exam registration requires that a student actually takes the course.
-   In logic, the constraint $p_{\tt ExamRule1}$ is written as:
-\[\begin{array}{l}\forall s\in\text{\tt Student}, e\in\text{\tt ExamReg}, c\in\text{\tt Course}:\\
-   e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \rightarrow\ s\ \text{\tt takes}\ c\\
-\end{array}\]
-   Another requirement, \verb-ExamRule2-, is that the student is enrolled for every module that is part of the course.
-   Its constraint, $p_{\tt ExamRule2}$, is written in logic as:
-\[\begin{array}{l}\forall s\in\text{\tt Student}, e\in\text{\tt ExamReg}, m\in\text{\tt Module}, c\in\text{\tt Course}:\\
-   e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\ \rightarrow\ s\ \text{\tt isEnrolledFor}\ m
-\end{array}\]
-   Both rules are invariants.
-   So, a student can register for an exam only if she is taking the course for which she registers and if she is enrolled for every module of that course.
-   % Ampersand also derives the following violation sets:
-% \[\begin{array}{l}
-   % \viol{\text{\tt ExamRule1}}{\dataset}\\
-   % \hspace{1cm}=\{\pair{s}{c}\mid\exists e:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\neg(s\ \text{\tt takes}\ c)\}\\
-   % \viol{\text{\tt ExamRule2}}{\dataset}\\
-   % \hspace{1cm}=\{\pair{s}{m}\mid\exists e,c:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\ \wedge\neg(s\ \text{\tt isEnrolledFor}\ m)\}
-% \end{array}\]
-
-\subsection{Migration script}
-   The generated migration script, $\migrsys$, consists of the schema of the existing system $\infsys$,
-   the migration system, and the desired system $\infsys'$.
-   The migration system copies the data of the existing system to the desired system by the following rules:
-\begin{verbatim}
--- The existing system (everything is prefixed with "old.")
-   RELATION old.takes[Student*Course]
-   RELATION old.isPartOf[Module*Course]
-   RELATION old.isEnrolledFor[Student*Module]
-
-   RULE old.EnrollRule:
-      old.isEnrolledFor |- old.takes;old.isPartOf~
-
--- Migration system
-   RELATION copied_takes[Student*Course]
-   ENFORCE new.takes >: old.takes - copied_takes
-   ENFORCE copied_takes >: new.takes
-
-   RELATION copied_isPartOf[Module*Course]
-   ENFORCE new.isPartOf >: old.isPartOf - copied_isPartOf
-   ENFORCE copied_isPartOf >: new.isPartOf
-
-   RELATION copied_isEnrolledFor[Student*Module]
-   ENFORCE new.isEnrolledFor >:
-           old.isEnrolledFor - copied_isEnrolledFor
-   ENFORCE copied_isEnrolledFor >: new.isEnrolledFor
-
--- The desired system (everything is prefixed with "new.")
-   RELATION new.takes[Student*Course]
-   RELATION new.isPartOf[Module*Course] [UNI]
-   RELATION new.isEnrolledFor[Student*Module]
-   RULE new.EnrollRule:
-      new.isEnrolledFor |- new.takes;new.isPartOf~
-
-   RELATION new.course[ExamReg*Course] [UNI] =
-      [ ("ER1", "Management")
-      ; ("ER2", "Business IT")
-      ; ("ER3", "Business IT")
-      ]
-   RELATION new.student[ExamReg*Student] [UNI] =
-      [ ("ER1", "Peter")
-      ; ("ER2", "Susan")
-      ]
-
-   RULE new.ExamRule1 :
-      new.student~;new.course |- new.takes
-   RULE new.ExamRule2 :
-      new.student~;new.course;new.isPartOf~ |- new.isEnrolledFor
-\end{verbatim}
-
-\subsection{Preparing for the Moment of Transition}
-   Just before starting the migration,
-   the data in the existing system typically differs from the data at its inception (section~\ref{sct:Example existing IS}).
-   Let us assume the actual population at the start of the migration is:
-\begin{verbatim}
-   POPULATION takes[Student*Course] CONTAINS
-      [ ("Peter", "Management")
-      ; ("Susan", "Business IT")
-      ; ("John", "Business IT")
-      ]
-   POPULATION isPartOf[Module*Course] CONTAINS
-      [ ("Finance", "Management")
-      ; ("Business Rules", "Business IT")
-      ; ("Business Analytics", "Business IT")
-      ; ("IT-Governance", "Management")
-      ; ("IT-Governance", "Business IT")
-      ]
-   POPULATION isEnrolledFor [Student*Module] CONTAINS
-      [ ("Susan", "Business Analytics")
-      ; ("Susan", "IT-Governance")
-      ; ("Susan", "Business Rules")
-      ]
-\end{verbatim}
-
-   Upon initialization, the migration system evaluates all enforce rules in an attempt to satisfy all invariants.
-   In this example, it leaves two violations behind that it cannot resolve by enforce rules.
-   The pair {\tt ("Peter", "Management")} is the first one; it violates rule {\tt new.ExamRule2}.
-   This is because Peter has registered for an examination of the course Management,
-   which contains the module Finance for which Peter is not enrolled.
-   The other is {\tt "IT-Governance"}, which violates the univalence of relation {\tt new.isPartOf}.
-   Since rule {\tt new.ExamRule2} and the univalence of {\tt new.isPartOf} are both defined as invariants,
-   the desired system cannot be taken into production until these violations are resolved.
-   Until that moment, the migration system remains operational to allow a user to resolve this violation. 
-   The moment of cleanup is marked by the event that all invariants of the Migration system and the desired system are satisfied.
-   At that moment, the migration system is no longer needed and can be removed.
+\subsection{General migration script}
+   So far, we illustrated how to do migration by giving two concrete examples.
+   This section generalizes those ideas by giving a general approach for generating a migration script.
+
+   Let $\pair{\dataset}{\schema}$ be the existing system, and let ${\schema'}$ be the schema of the desired system.
+   Then we define the migration systems $\migrsys$ and the events as follows:
+   \begin{enumerate}
+   \item We take a disjoint union of the relations: $\rels_\migrsys = \overleftarrow\rels_{\schema} \cup \overrightarrow\rels_{\schema'}$
+     where we define $\overleftarrow~$ and $\overrightarrow~$ to generate new names: $\overleftarrow{\declare{n}{x}{y}}=\declare{(old.n)}{x}{y}$ and $\overrightarrow{\declare{n}{x}{y}}=\declare{(new.n)}{x}{y}$,
+     overloading the notations on sets pointwise: $\overleftarrow{X} = \{\overleftarrow{x}\mid x\in X\}$ and $\overrightarrow{X} = \{\overrightarrow{x}\mid x\in X\}$.
+   \item We copy the relations 
+   \end{enumerate}
+
+%   Let us proceed to specify a desired system, $\infsys'$, to illustrate a (toy) migration.
+%   The example in section~\ref{sct:Example existing IS}, called $\infsys$, serves as the existing system.
+%   Its population reflects the state of the system just before the migration.
+%
+%   Information system $\infsys'$, the desired system, is specified by:
+%\begin{verbatim}
+%   RELATION takes[Student*Course]
+%   RELATION isPartOf[Module*Course] [UNI]
+%   RELATION isEnrolledFor[Student*Module]
+%   RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
+%   
+%   RELATION course[ExamReg*Course] [UNI] =
+%      [ ("ER1", "Management")
+%      ; ("ER2", "Business IT")
+%      ; ("ER3", "Business IT")
+%      ]
+%   RELATION student[ExamReg*Student] [UNI] =
+%      [ ("ER1", "Peter")
+%      ; ("ER2", "Susan")
+%      ]
+%   RULE ExamRule1: student~;course |- takes
+%   RULE ExamRule2: student~;course;isPartOf~ |- isEnrolledFor
+%\end{verbatim}
+%   In this example, the relations and the rule of the existing system remain part of the desired system%
+%\footnote{Removal or changing of relations is allowed as well.}
+%   because the developer has copied these items into the script of the desired system.
+%   However, she has imposed the restriction \verb-UNI- on the relation {\tt isPartOf}, stating that this relation is univalent.
+%   This means that every \verb-Module- is part of at most one \verb-Course-.
+%   Besides, two new relations and two rules have appeared.
+%
+%   The desired system, $\infsys'$, contains exam registrations (concept \verb-ExamReg-), which is new.
+%   In an exam registration, a student registers for the examination of a course.
+%   $\infsys'$ contains two extra rules: \verb-ExamRule1- and \verb-ExamRule2-.
+%   The first one says that an exam registration requires that a student actually takes the course.
+%   In logic, the constraint $p_{\tt ExamRule1}$ is written as:
+%\[\begin{array}{l}\forall s\in\text{\tt Student}, e\in\text{\tt ExamReg}, c\in\text{\tt Course}:\\
+%   e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \rightarrow\ s\ \text{\tt takes}\ c\\
+%\end{array}\]
+%   Another requirement, \verb-ExamRule2-, is that the student is enrolled for every module that is part of the course.
+%   Its constraint, $p_{\tt ExamRule2}$, is written in logic as:
+%\[\begin{array}{l}\forall s\in\text{\tt Student}, e\in\text{\tt ExamReg}, m\in\text{\tt Module}, c\in\text{\tt Course}:\\
+%   e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\ \rightarrow\ s\ \text{\tt isEnrolledFor}\ m
+%\end{array}\]
+%   Both rules are invariants.
+%   So, a student can register for an exam only if she is taking the course for which she registers and if she is enrolled for every module of that course.
+%   % Ampersand also derives the following violation sets:
+%% \[\begin{array}{l}
+%   % \viol{\text{\tt ExamRule1}}{\dataset}\\
+%   % \hspace{1cm}=\{\pair{s}{c}\mid\exists e:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\neg(s\ \text{\tt takes}\ c)\}\\
+%   % \viol{\text{\tt ExamRule2}}{\dataset}\\
+%   % \hspace{1cm}=\{\pair{s}{m}\mid\exists e,c:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\ \wedge\neg(s\ \text{\tt isEnrolledFor}\ m)\}
+%% \end{array}\]
+%
+%\subsection{Migration script}
+%   The generated migration script, $\migrsys$, consists of the schema of the existing system $\infsys$,
+%   the migration system, and the desired system $\infsys'$.
+%   The migration system copies the data of the existing system to the desired system by the following rules:
+%\begin{verbatim}
+%-- The existing system (everything is prefixed with "old.")
+%   RELATION old.takes[Student*Course]
+%   RELATION old.isPartOf[Module*Course]
+%   RELATION old.isEnrolledFor[Student*Module]
+%
+%   RULE old.EnrollRule:
+%      old.isEnrolledFor |- old.takes;old.isPartOf~
+%      
+%-- Migration system
+%   RELATION copied_takes[Student*Course]
+%   ENFORCE new.takes >: old.takes - copied_takes
+%   ENFORCE copied_takes >: new.takes
+%
+%   RELATION copied_isPartOf[Module*Course]
+%   ENFORCE new.isPartOf >: old.isPartOf - copied_isPartOf
+%   ENFORCE copied_isPartOf >: new.isPartOf
+%
+%   RELATION copied_isEnrolledFor[Student*Module]
+%   ENFORCE new.isEnrolledFor >:
+%           old.isEnrolledFor - copied_isEnrolledFor
+%   ENFORCE copied_isEnrolledFor >: new.isEnrolledFor
+%
+%-- The desired system (everything is prefixed with "new.")
+%   RELATION new.takes[Student*Course]
+%   RELATION new.isPartOf[Module*Course] [UNI]
+%   RELATION new.isEnrolledFor[Student*Module]
+%   RULE new.EnrollRule:
+%      new.isEnrolledFor |- new.takes;new.isPartOf~
+%      
+%   RELATION new.course[ExamReg*Course] [UNI] =
+%      [ ("ER1", "Management")
+%      ; ("ER2", "Business IT")
+%      ; ("ER3", "Business IT")
+%      ]
+%   RELATION new.student[ExamReg*Student] [UNI] =
+%      [ ("ER1", "Peter")
+%      ; ("ER2", "Susan")
+%      ]
+%
+%   RULE new.ExamRule1 :
+%      new.student~;new.course |- new.takes
+%   RULE new.ExamRule2 :
+%      new.student~;new.course;new.isPartOf~ |- new.isEnrolledFor
+%\end{verbatim}
+%
+%\subsection{Preparing for the Moment of Transition}
+%   Just before starting the migration,
+%   the data in the existing system typically differs from the data at its inception (section~\ref{sct:Example existing IS}).
+%   Let us assume the actual population at the start of the migration is:
+%\begin{verbatim}
+%   POPULATION takes[Student*Course] CONTAINS
+%      [ ("Peter", "Management")
+%      ; ("Susan", "Business IT")
+%      ; ("John", "Business IT")
+%      ]
+%   POPULATION isPartOf[Module*Course] CONTAINS
+%      [ ("Finance", "Management")
+%      ; ("Business Rules", "Business IT")
+%      ; ("Business Analytics", "Business IT")
+%      ; ("IT-Governance", "Management")
+%      ; ("IT-Governance", "Business IT")
+%      ]
+%   POPULATION isEnrolledFor [Student*Module] CONTAINS
+%      [ ("Susan", "Business Analytics")
+%      ; ("Susan", "IT-Governance")
+%      ; ("Susan", "Business Rules")
+%      ]
+%\end{verbatim}
+%
+%   Upon initialization, the migration system evaluates all enforce rules in an attempt to satisfy all invariants.
+%   In this example, it leaves two violations behind that it cannot resolve by enforce rules.
+%   The pair {\tt ("Peter", "Management")} is the first one; it violates rule {\tt new.ExamRule2}.
+%   This is because Peter has registered for an examination of the course Management,
+%   which contains the module Finance for which Peter is not enrolled.
+%   The other is {\tt "IT-Governance"}, which violates the univalence of relation {\tt new.isPartOf}.
+%   Since rule {\tt new.ExamRule2} and the univalence of {\tt new.isPartOf} are both defined as invariants,
+%   the desired system cannot be taken into production until these violations are resolved.
+%   Until that moment, the migration system remains operational to allow a user to resolve this violation. 
+%   The moment of cleanup is marked by the event that all invariants of the Migration system and the desired system are satisfied.
+%   At that moment, the migration system is no longer needed and can be removed.
 

### 2023-07-10 2022Migration/articleMigrationFACS.tex
First mathematical enforce rule

@@
 \usepackage{amsmath}
-\usepackage{amssymb}
+\usepackage{amssymb,stmaryrd}
 \usepackage{mathrsfs}
@@
 
-   The notation $\schema_{\infsys}$ refers to the schema of information system $\infsys$.
+   %The notation $\schema_{\infsys}$ refers to the schema of information system $\infsys$.
    When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$
@@
    \begin{definition}[Event]
-   Let $R \subseteq \Rels$ and $I \subseteq \Rels$ be sets of relations, and let $\infsys = \pair{\schema}{\dataset}$ and $\infsys' = \pair{\schema}{\dataset'}$ be information systems (so $\infsys$ and $\infsys'$ have the same schema), and:
+   Let $R \subseteq \Rels$ and $I \subseteq \Rels$ be sets of relations, and let $\infsys' = \pair{\schema}{\dataset'}$ be an information system, and:
       \begin{align}
@@
    \end{align}
-   Then we say that $\pair{\infsys}{\infsys'}$ is an event with scope $R$ inserting in $I$.
-   We use the notation $\infsys \xrightarrow[I]{R} \infsys'$ to indicate this.
+   Then we say that $\pair{\pair{\schema}{\dataset}}{\infsys'}$ is an event with scope $R$ inserting in $I$.
+   We use the notation $\pair{\schema}{\dataset} \xrightarrow[I]{R} \infsys'$ to indicate this.
    \end{definition}
@@
    Equation~\ref{eqn:eventUnchanged} states that the triples of those for relations in $I$ or $R$ are the only ones to have changed, and Equation~\ref{eqn:eventInsert} states that of the ones that are only in $I$, no triples were removed.
-   If $I$ or $R$ is the empty set, we omit it from the arrow, so $\infsys \xrightarrow{R} \infsys'$ is a notation for $\infsys \xrightarrow[\emptyset]{R} \infsys'$.
+   %If $I$ or $R$ is the empty set, we omit it from the arrow, so $\infsys \xrightarrow{R} \infsys'$ is a notation for $\infsys \xrightarrow[\emptyset]{R} \infsys'$.
 
@@
 
-\subsection{Loading Data}
+\subsection{Loading Data}\label{sec:loading}
    Let us study how to migrate data from $\dataset$ to $\dataset'$.
@@
    \begin{enumerate}
-   \item We take a disjoint union of the relations: $\rels_\migrsys = \overleftarrow\rels_{\schema} \cup \overrightarrow\rels_{\schema'}$
-     where we define $\overleftarrow~$ and $\overrightarrow~$ to generate new names: $\overleftarrow{\declare{n}{x}{y}}=\declare{(old.n)}{x}{y}$ and $\overrightarrow{\declare{n}{x}{y}}=\declare{(new.n)}{x}{y}$,
-     overloading the notations on sets pointwise: $\overleftarrow{X} = \{\overleftarrow{x}\mid x\in X\}$ and $\overrightarrow{X} = \{\overrightarrow{x}\mid x\in X\}$.
-   \item We copy the relations 
+   \item We take a disjoint union of the relations, defining $\overleftarrow~$ and $\overrightarrow~$ as (overloaded) notation to generate new names:
+   \begin{align}
+   \overleftarrow{\declare{n}{x}{y}}&=\declare{(\id{old}.n)}{x}{y}&\overleftarrow{X} &= \{\overleftarrow{x}\mid x\in X\}\\
+   \overrightarrow{\declare{n}{x}{y}}&=\declare{(\id{new}.n)}{x}{y}& \overrightarrow{X} &= \{\overrightarrow{x}\mid x\in X\}\\
+   \rels_1 &= \overleftarrow\rels_{\schema} \cup \overrightarrow\rels_{\schema'}
+   \end{align}
+   \item We create ENFORCE rules to copy equal relations, corresponding to (\ref{extraRelation}) and (\ref{difference}) in Section~\ref{sec:loading}:
+   \begin{align}
+   \widetilde{\declare{n}{x}{y}} &= \declare{(\id{copied}.n)}{x}{y}\\
+   \rels_2 &= \{\widetilde{r}\mid r \in \rels_\schema \cap \rels_{\schema'}\}\\
+   \triples_\migrsys &= \{\triple{a}{\overleftarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_\infsys\}\\
+   \mathcal{E}_2 &= \{\overrightarrow{r}\mapsfrom (\overleftarrow{r} - \widetilde{r}) \mid \widetilde{r} \in \rels_2\} \cup \{\widetilde{r}\mapsfrom \overrightarrow{r} \mid \widetilde{r} \in \rels_2\}
+   \end{align}
+   \item 
    \end{enumerate}

### 2023-07-10 2022Migration/articleMigrationFACS.tex
WIP on abstract

@@
 %
-\begin{abstract}
+\begin{abstract}  % There are 176 words in the abstract
 Software generators can help to increase the frequency of releases and their reliability.
@@
 
-This paper aims to preserve the semantics of data by satisfying concrete business requirements.
-The migration process is based on the assumption that software is deployed incrementally, the existing data set may be polluted, human interaction may be required, the meaning of data must be preserved, the business continues during the migration without interruption (zero downtime), and there is a compiler to generate an information system from a given schema.
-The correctness of the migration is the focus of this paper, while efficiency is outside its scope.
+This paper defines a migration process that preserves the semantics of data by means of rules (constraints).
+The migration process is based assumptions and requirements that aim for a large class of data migrations in practice.
+This paper focuses on correctness of the migration. Efficiency of the migration is outside its scope.
 
@@
    to ensure the preservation of valuable data and the quality of the migration.
-   Automating this process completely is unrealistic in many cases:
-   While smaller increments improve the chance that a fully automatic migration is possible, there will always be cases where human intervention is required.
+   To fully automate every incremental data migration is unrealistic.
+   Smaller increments stand a higher chance of being automated, but there will always be cases where some human intervention is required.
    Manual interventions can be necessary to resolve data pollution, new business rules, or to fix known issues in the old system.

### 2023-07-11 2022Migration/articleMigrationFACS.tex
Rules are functions

@@
 \newcommand{\sat}[2]{\id{sat}_{#1}(#2)}
-\newcommand{\viol}[2]{\violC{#1}(#2)}
-\newcommand{\violC}[1]{\id{viol}_{#1}}
+\newcommand{\viol}[2]{{#1}(#2)}
+%\newcommand{\violC}[1]{\id{viol}_{#1}}
 \newcommand{\sign}[1]{\id{sign}(#1)}
@@
    $\rels\subseteq \Rels$ is a finite set of relations,
-   and $\rules\subseteq \Rules$ is a finite set of rules.
+   and $\rules\subseteq \left(\Dataset \to \Pair{\Atoms}{\Atoms}\right)$ is a finite set of rules.
 
@@
    In this paper, variables $u$ and $v$ represent rules.
-   For every rule $u$ in the schema, we assume there is a function $\violC{u}$ such that $\viol{u}{\dataset}$ represents any violations to $u$ in data set $\dataset$.
+   A rule $u$ in the schema is a function such that $\viol{u}{\dataset}$ represents any violations to $u$ in data set $\dataset$.
    If there are no violations, that is: $\viol{u}{\dataset} = \emptyset{}$, then we say that $u$ is satisfied.
@@
    \begin{enumerate}
-   \item We take a disjoint union of the relations, defining $\overleftarrow~$ and $\overrightarrow~$ as (overloaded) notation to generate new names:
+   \item We take a disjoint union of the relations, defining $\overleftarrow~$ and $\overrightarrow~$ as (overloaded) notation to generate new names, keeping the dataset of the old system and the rules of the new system:
    \begin{align}
@@
    \overrightarrow{\declare{n}{x}{y}}&=\declare{(\id{new}.n)}{x}{y}& \overrightarrow{X} &= \{\overrightarrow{x}\mid x\in X\}\\
-   \rels_1 &= \overleftarrow\rels_{\schema} \cup \overrightarrow\rels_{\schema'}
+   \rels_\theenumi &= \overleftarrow\rels_{\schema} \cup \overrightarrow\rels_{\schema'}\\
+   \triples_\migrsys &= \{\triple{a}{\overleftarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_\infsys\}&
+   \instance_\migrsys &= \instance_\infsys\\
+   \rules_\theenumi &= \rules_\infsys
    \end{align}
@@
    \widetilde{\declare{n}{x}{y}} &= \declare{(\id{copied}.n)}{x}{y}\\
-   \rels_2 &= \{\widetilde{r}\mid r \in \rels_\schema \cap \rels_{\schema'}\}\\
-   \triples_\migrsys &= \{\triple{a}{\overleftarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_\infsys\}\\
-   \mathcal{E}_2 &= \{\overrightarrow{r}\mapsfrom (\overleftarrow{r} - \widetilde{r}) \mid \widetilde{r} \in \rels_2\} \cup \{\widetilde{r}\mapsfrom \overrightarrow{r} \mid \widetilde{r} \in \rels_2\}
+   \rels_\theenumi &= \{\widetilde{r}\mid r \in \rels_\schema \cap \rels_{\schema'}\}\\
+   \mathcal{E}_\theenumi &= \{\overrightarrow{r}\mapsfrom (\overleftarrow{r} - \widetilde{r}) \mid \widetilde{r} \in \rels_2\} \cup \{\widetilde{r}\mapsfrom \overrightarrow{r} \mid \widetilde{r} \in \rels_2\}
    \end{align}
-   \item 
+   \item For each rule, we generate two helper relations, viols and fixed. Note that Rule~\ref{warnings} only influences the user interface and therefore is not modeled here.
+   \begin{align}
+   \widetilde{\dataset'} &= \pair{\{\triple{x}{\widetilde{r}}{y}\mid \triple{x}{r}{y} \in \triples_{\dataset'}\}}{\instance_{\dataset'}}\\
+   \overrightarrow{\dataset'} &= \pair{\{\triple{a}{\overrightarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_\infsys\}}{\instance_{\dataset'}}\\
+   \mathcal{E}_{\theenumi} &=\{TODO\}\\
+   \rules_\theenumi &= \{\lambda {\dataset'}. \viol{u}{\widetilde{\dataset'}}\}
+   \end{align}
    \end{enumerate}

### 2023-07-11 2022Migration/articleMigrationFACS.tex
General migration script

@@
 \newcommand{\relsIn}[1]{\id{relsIn}(#1)}    % maps a Term to a set of Relations
-\newcommand{\pop}[2]{\id{pop}_{#1}(#2)}
+\newcommand{\popF}[1]{\id{pop}_{#1}}
+\newcommand{\pop}[2]{\popF{#1}(#2)}
 \newcommand{\maintain}{\mathbin{\id{maint}}}
@@
    A relation $r$ can serve as a container of pairs,
-   as defined by the function $\id{pop}_r:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$.
+   as defined by the function $\popF{r}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$.
    It defines a set of pairs, also known as the population of $r$:
@@
 %
-   Concepts too can be seen as containers of atoms,
-   defined by the function $\id{pop}_C:\Dataset\rightarrow\powerset{\Atoms}$.
-\begin{equation}
-   \pop{C}{\dataset}\ =\ \{ x\mid\ x\ \instance_{\dataset}\ C\}
-\label{eqn:pop-concept}
-\end{equation}
+   We overload the notation $\popF{\cdot}$ so we can use it on concepts $\popF{C}:\Dataset\rightarrow\powerset{\Atoms}$
+   and expressions.
+\begin{align}
+   \pop{C}{\dataset}&= \{ x\mid\ x\ \instance_{\dataset}\ C\}
+\label{eqn:pop-concept}\\
+   \pop{A - B}{\dataset}&= \pop{A}{\dataset} - \pop{B}{\dataset}
+\label{eqn:pop-expr}
+\end{align}
 
@@
    $\rels\subseteq \Rels$ is a finite set of relations,
-   and $\rules\subseteq \left(\Dataset \to \Pair{\Atoms}{\Atoms}\right)$ is a finite set of rules.
+   and $\rules\subseteq \left(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}\right)$ is a finite set of rules.
 
@@
    \rels_\theenumi &= \overleftarrow\rels_{\schema} \cup \overrightarrow\rels_{\schema'}\\
-   \triples_\migrsys &= \{\triple{a}{\overleftarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_\infsys\}&
-   \instance_\migrsys &= \instance_\infsys\\
-   \rules_\theenumi &= \rules_\infsys
+   \triples_\migrsys &= \{\triple{a}{\overleftarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_\dataset\}&
+   \instance_\migrsys &= \instance_\dataset\\
+   \rules_\theenumi &= \rules_\schema
    \end{align}
@@
    \rels_\theenumi &= \{\widetilde{r}\mid r \in \rels_\schema \cap \rels_{\schema'}\}\\
-   \mathcal{E}_\theenumi &= \{\overrightarrow{r}\mapsfrom (\overleftarrow{r} - \widetilde{r}) \mid \widetilde{r} \in \rels_2\} \cup \{\widetilde{r}\mapsfrom \overrightarrow{r} \mid \widetilde{r} \in \rels_2\}
+   \mathcal{E}_\theenumi &= \{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r} - \widetilde{r}} \mid \widetilde{r} \in \rels_2\} \cup \{\widetilde{r}\mapsfrom \popF{\overrightarrow{r}} \mid \widetilde{r} \in \rels_2\}
    \end{align}
@@
    \widetilde{\dataset'} &= \pair{\{\triple{x}{\widetilde{r}}{y}\mid \triple{x}{r}{y} \in \triples_{\dataset'}\}}{\instance_{\dataset'}}\\
-   \overrightarrow{\dataset'} &= \pair{\{\triple{a}{\overrightarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_\infsys\}}{\instance_{\dataset'}}\\
-   \mathcal{E}_{\theenumi} &=\{TODO\}\\
-   \rules_\theenumi &= \{\lambda {\dataset'}. \viol{u}{\widetilde{\dataset'}}\}
+   \overrightarrow{\dataset'} &= \pair{\{\triple{a}{\overrightarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_\dataset\}}{\instance_{\dataset'}}\\
+   \mathcal{E}_{\theenumi} &=\{v_u \mapsfrom \lambda \dataset'.~ \viol{u}{\widetilde{\dataset'}}\} \cup
+    \{f_u \mapsfrom \lambda \dataset'.~ v_u - \viol{u}{\overrightarrow{\dataset'}}\} \\
+   \rels_\theenumi &= \{v_u\mid u \in \rules_{\schema'}\} \cup \{f_u\mid u \in \rules_{\schema'}\}\\
+   \rules_\theenumi &= \{\lambda {\dataset'}.~ \viol{u}{\widetilde{\dataset'}} - (v_u - f_u) \mid u\in\rules_{\schema'}\}
+   \end{align}
+   \item Combining the above into a single migration system:
+   \begin{align}
+   \rels_\migrsys &= \rels_1\cup\rels_2\cup \rels_3\\
+   \rules_\migrsys &= \rules_1\cup \rules_3\\
+   \mathcal{E}_\migrsys &= \mathcal{E}_1\cup\mathcal{E}_2\cup\mathcal{E}_3\\
+   \concepts_\migrsys &= \concepts_\dataset \cup  \concepts_{\dataset'}
    \end{align}

### 2023-07-11 2022Migration/articleMigrationFACS.tex
Finished Validation chapter

@@
 \newcommand{\Atoms}{\mathbb{A}}
-\newcommand{\events}{\mathcal{E}}
-\newcommand{\Events}{\mathbb{E}}
+%\newcommand{\events}{\mathcal{E}}
+%\newcommand{\Events}{\mathbb{E}}
 \newcommand{\concept}[1]{{\tt\small #1}}
@@
 \newcommand{\vertices}{N}
+\newcommand{\enforces}{\mathcal{E}}
 \newcommand{\rules}{\mathcal{U}}
@@
 
-   %The notation $\schema_{\infsys}$ refers to the schema of information system $\infsys$.
-   When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$
-   for $\concepts$, $\rels$, and $\rules$ corresponding to $\schema$.
+   For some rules, the system is told what to do in order to fix it.
+   Such rules are called enforce rules.
+   We just need to consider one type of enforce rule in this work, its syntax is as follows:
+
+\begin{align}
+      \text{\tt ENFORCE}\ r\ \text{\tt >:}\ \id{term}\label{enforce ins}
+\end{align}
+
+   Adding an enforce rule to the script of an information system specifies a rule $u$ with $\viol{u}{\dataset} = \pop{\id{term} - r}{\dataset}$, and the information system ensures that the rule is satisfied by adding a triple $\triple{x}{r}{y}$ for each $\pair{x}{y} \in \viol{u}{\dataset}$, thus maintaining $\pop{r}{\dataset} \supseteq \pop{\id{term}}{\dataset}$ as an invariant.
+
+   Formally, enforce rules are pairs with $u$ a rule and $r$ a relation, written $r \mapsfrom u$ or equivalently $r \mapsfrom \lambda \dataset.~ u(\dataset)$.
+   The set of enforce rules in a schema is $\enforces$.
 
    \begin{definition}[Schema]
-   A schema is a tuple $\triple{\concepts}{\rels}{\rules}$ that satisfies:
-\begin{eqnarray}
-   \declare{n}{A}{B}\in\rels&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
-   \label{eqn:relationsIntroduceConcepts}
-\end{eqnarray}
+   A schema is a tuple $\quadruple{\concepts}{\rels}{\rules}{\enforces}$ that satisfies:
+\begin{align}
+   \declare{n}{A}{B}\in\rels&~\Rightarrow~ A\in\concepts\ \wedge\ B\in\concepts
+   \label{eqn:relationsIntroduceConcepts}\\
+   \pair{u}{r}\in\mathcal\enforces&~\Rightarrow~ r \in \rels\label{eqn:enforceRulesRels}
+\end{align}
    \end{definition}
@@
    This is also known as static typing, which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.
+
+   %The notation $\schema_{\infsys}$ refers to the schema of information system $\infsys$.
+   When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, $\rules_{\schema}$, and $\enforces_{\schema}$
+   for $\concepts$, $\rels$, $\rules$, and $\enforces$ corresponding to $\schema = \quadruple{\concepts}{\rels}{\rules}{\enforces}$.
 
@@
    \item $\dataset=\pair{\triples}{\instance}$ is a data set (satisfies Equation~\ref{eqn:wellTypedEdge}), we write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$;
-   \item $\schema=\triple{\concepts}{\rels}{\rules}$ is a schema (satisfies Equation~\ref{eqn:relationsIntroduceConcepts}), we write $\concepts_\infsys = \concepts$, $\rels_\infsys = \rels$ and $\rules_\infsys=\rules$;
+   \item $\schema=\triple{\concepts}{\rels}{\rules}$ is a schema (satisfies Equation~\ref{eqn:relationsIntroduceConcepts} and~\ref{eqn:enforceRulesRels}), we write $\concepts_\infsys = \concepts$, $\rels_\infsys = \rels$ and $\rules_\infsys=\rules$;
    \item all rules are satisfied:
@@
 
-   We use inserting events to model the effect of a type of rule that is particular to Ampersand, known as an enforce rule.
-   The syntax is as follows:
-
-\begin{align}
-      \text{\tt ENFORCE}\ r\ \text{\tt >:}\ \id{term}\label{enforce ins}
-\end{align}
-
-   Adding an enforce rule to the script of an information system specifies a rule $u$ with $\viol{u}{\dataset} = \pop{\id{term}}{\dataset} - \pop{r}{\dataset}$, and the information system ensures that the rule is satisfied by adding a triple $\triple{x}{r}{y}$ for each $\pair{x}{y} \in \viol{u}{\dataset}$, thus maintaining $\pop{r}{\dataset} \supseteq \pop{\id{term}}{\dataset}$ as an invariant.
-   If $R$ is a set of relations that covers those that occur in $\id{term}$, then changes to an information system $\infsys$ can be described by the event $\infsys \xrightarrow[\{r\}]{R} \infsys'$.
+   We use inserting events to model the effect of enforce rules.
+%   If $R$ is a set of relations that covers those that occur in the rule, then changes to an information system $\infsys$ can be described by the event $\infsys \xrightarrow[\{r\}]{R} \infsys'$.
 
@@
 
-   Despite the syntax, Line~{\ref{warnings}} is not a rule of the generated information system:
+   Despite the syntax, Line~{\ref{warnings}} is not a rule of the generated information system in the formal sense:
    By assigning a role, the responsibility of maintaining the rule no longer lies with the information system.
-   Instead, users with the migration helper role can see the violations and act on it as they see fit.
+   Instead, users with the migration helper role can see the violations and act on it as they see fit, so the only effect of Line~{\ref{warnings}} is that such users will be shown its violations.
 
@@
    \overrightarrow{\declare{n}{x}{y}}&=\declare{(\id{new}.n)}{x}{y}& \overrightarrow{X} &= \{\overrightarrow{x}\mid x\in X\}\\
-   \rels_\theenumi &= \overleftarrow\rels_{\schema} \cup \overrightarrow\rels_{\schema'}\\
    \triples_\migrsys &= \{\triple{a}{\overleftarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_\dataset\}&
    \instance_\migrsys &= \instance_\dataset\\
-   \rules_\theenumi &= \rules_\schema
+   \rels_\theenumi &= \overleftarrow\rels_{\schema} \cup \overrightarrow\rels_{\schema'}
+   %&\rules_\theenumi &= \rules_\schema
    \end{align}
@@
    \rels_\theenumi &= \{\widetilde{r}\mid r \in \rels_\schema \cap \rels_{\schema'}\}\\
-   \mathcal{E}_\theenumi &= \{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r} - \widetilde{r}} \mid \widetilde{r} \in \rels_2\} \cup \{\widetilde{r}\mapsfrom \popF{\overrightarrow{r}} \mid \widetilde{r} \in \rels_2\}
+   \enforces_\theenumi &= \{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r} - \widetilde{r}} \mid \widetilde{r} \in \rels_2\} \cup \{\widetilde{r}\mapsfrom \popF{\overrightarrow{r}} \mid \widetilde{r} \in \rels_2\}
    \end{align}
@@
    \widetilde{\dataset'} &= \pair{\{\triple{x}{\widetilde{r}}{y}\mid \triple{x}{r}{y} \in \triples_{\dataset'}\}}{\instance_{\dataset'}}\\
-   \overrightarrow{\dataset'} &= \pair{\{\triple{a}{\overrightarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_\dataset\}}{\instance_{\dataset'}}\\
-   \mathcal{E}_{\theenumi} &=\{v_u \mapsfrom \lambda \dataset'.~ \viol{u}{\widetilde{\dataset'}}\} \cup
-    \{f_u \mapsfrom \lambda \dataset'.~ v_u - \viol{u}{\overrightarrow{\dataset'}}\} \\
+   \overrightarrow{\dataset'} &= \pair{\{\triple{a}{\overrightarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_\dataset\}}{\instance_{\dataset'}} \\
    \rels_\theenumi &= \{v_u\mid u \in \rules_{\schema'}\} \cup \{f_u\mid u \in \rules_{\schema'}\}\\
-   \rules_\theenumi &= \{\lambda {\dataset'}.~ \viol{u}{\widetilde{\dataset'}} - (v_u - f_u) \mid u\in\rules_{\schema'}\}
+   \rules_\migrsys &= \{\lambda {\dataset'}.~ \viol{u}{\widetilde{\dataset'}} - \pop{v_u - f_u}{\dataset'} \mid u\in\rules_{\schema'}\}\label{eqn:migrsysrules}\\
+   \enforces_{\theenumi} &=\{v_u \mapsfrom \lambda \dataset'.~ \viol{u}{\widetilde{\dataset'}}\} \cup
+    \{f_u \mapsfrom \lambda \dataset'.~ v_u - \viol{u}{\overrightarrow{\dataset'}}\}\label{eqn:enforceForRules}
    \end{align}
@@
    \begin{align}
+   \concepts_\migrsys &= \concepts_\dataset \cup  \concepts_{\dataset'}\\
    \rels_\migrsys &= \rels_1\cup\rels_2\cup \rels_3\\
-   \rules_\migrsys &= \rules_1\cup \rules_3\\
-   \mathcal{E}_\migrsys &= \mathcal{E}_1\cup\mathcal{E}_2\cup\mathcal{E}_3\\
-   \concepts_\migrsys &= \concepts_\dataset \cup  \concepts_{\dataset'}
+   %\rules_\migrsys &= \rules_1\cup \rules_3\\
+   \enforces_\migrsys &= \enforces_1\cup\enforces_2\cup\enforces_3
    \end{align}
    \end{enumerate}
+
+   \subsection{Validation}
+   We first observe that $\migrsys = \pair{\pair{\triples_\migrsys}{\instance_\migrsys}}{\quadruple{\concepts_\migrsys}{\rels_\migrsys}{\rules_\migrsys}{\enforces_\migrsys}}$ is an information system after application of the enforce rules:
+   the only rules are those introduced in Eqution~\ref{eqn:migrsysrules}, and enforce rules from Equation~\ref{eqn:enforceForRules} ensure that $\popF{v_u} = u$ initially, while $\popF{f_u}$ is empty.
+   Hence, the MoT can happen after the enforce rules have been applied, giving us $\migrsys'$.
+   The event $\migrsys \xrightarrow[\rels_2 \cup \rels_3]{\overrightarrow\rels_{\schema'}}\migrsys'$ represents the MoT.
+
+   The MoC is reached for dataset $\dataset'$ whenever $\pop{v_u - f_u}{\dataset'}=\emptyset$ for all rules.
+   This should happen after a finite number of inserting events: let $I = \{f_u \mid u\in\rules_{\schema'}\}$ be the relations that keep track of which violations have been fixed, then there is no infinite chain of systems $\migrsys' = \migrsys_0, \migrsys_1,\ldots$ such that $\migrsys_i\xrightarrowdbl[I]{\overrightarrow\rels_{\schema'}}\migrsys_{i+1}$.
+   This shows that when users get rid of violations, there is real progress: they can only do this a finite number of times.
+   Next we ask ourselves: does this mean we always reach the MoC?
+
+   Suppose there is a system in which all violations have been fixed, that is, suppose there is a $\migrsys^C$ such that $\pop{f_u}{\dataset_{\migrsys^C}}=\pop{v_u}{\dataset_{\migrsys^C}}$ and $\migrsys'\xrightarrow[I]{\overrightarrow\rels_{\schema'}} \migrsys^C$.
+   In that case, after any steps done by users (both those with and without the migration helper role), we can reach it, that is: for any $\migrsys_i$, $\migrsys'\xrightarrow[I]{\overrightarrow\rels_{\schema'}} \migrsys_i$ implies $\migrsys_i\xrightarrow[I]{\overrightarrow\rels_{\schema'}} \migrsys^C$.
+   This implies that if the MoC is initially reachable through $\xrightarrow[I]{\overrightarrow\rels_{\schema'}}$ events, then it will remain reachable.
+   So is the MoC initially reachable?
+
+   For this last question, we need to add a small condition:
+   There needs to be an information system that has the desired schema $\schema'$.
+   If rules in $\schema'$ are such that they cannot all be satisfied, regardless of the dataset, no such system exists and we have no hope of reaching it.
+   However, let $\pair{\dataset'}{\schema'}$ be an information system, then we can get to it through our migration system.
+   To show this, we construct a migration system $\migrsys^C$ satisfying $\pop{f_u}{\dataset_{\migrsys^C}}=\pop{v_u}{\dataset_{\migrsys^C}}$ and $\migrsys'\xrightarrow[I]{\overrightarrow\rels_{\schema'}} \migrsys^C$.
+   We can do so by simply stating its triples, since the schema is determined by $\migrsys'$:
+
+   \begin{align}
+   \begin{aligned}
+   \triples_{\migrsys^C} &= \triples_\migrsys 
+   \cup \{\triple{a}{\overrightarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_{\dataset'}\}\\
+   &\cup \{\triple{a}{v_u}{b} \mid \pair{a}{b}\in \viol{u}{\dataset}, u\in \rules_{\schema'} \}
+   \cup \{\triple{a}{f_u}{b} \mid \pair{a}{b}\in \viol{u}{\dataset}, u\in \rules_{\schema'} \}
+   \end{aligned}
+   \end{align}
+
+   Collectively, this shows that the MoC is and remains reachable after a finite number of inserting events by users in the migration helper role.
 

### 2023-07-11 Generative Software/proposal2.tex
2nd version (proposal under construction)

--- 2023-02-12 Generative Software/proposal.tex
@@
 
-\title{Generative Software Development\\{\em\normalsize Research Offering}}
+\title{Enhancing Software Development with Artificial Intelligence}
 \author{prof.dr.ir.\ Stef Joosten\footnote{ORCID 0000-0001-8308-0189}}
@@
 \begin{abstract}
-    This paper summarizes the capability of the Open Universiteit in generative software development.
-    We offer this enabling technology as a component of a larger effort that involves designing distributed information systems.
-    Generative software development offers attractive benefits for correctness, security, and resilience.
-    Also, it enables a product to be demonstrated live during its creation,
-    hereby supporting agile software development methods.
+    This proposal aims at supporting software engineers to develop software correctly and reliably by construction.
+    The threshold of learning formal specification languages has kept scores of software engineers from using them in practice.
+    Recent progress in Artificial Intelligence fuels the hope that AI might leverage the use of formal specification languages.
+    This proposal aims at investigating how.
 
-    The ambition of this research is to contribute to solving the problem of large IT-projects.
-    By generating fully functional information systems, we aim to reduce the effort and innovation risk of such projects.
-    The direction of research goes toward more frequent deployments,
-    which enables large projects to deliver incremental results in days rather than months.
-    %every 2 or 3 weeks as it develops into an end product.
+    The ambition of this research is to contribute to solving the problem of large IT-projects
+    along the line of correctness and reliability.
 \end{abstract}
@@
 \begin{keyword}
-    generative software, software engineering, security, prototyping, correctness, Ampersand, TopHat
+    generative software, AI enhanced software engineering, reliability by construction, prototyping, correctness by construction, Ampersand, TopHat
 \end{keyword}

### 2023-07-11 Generative Software/proposalBrief2.tex
2nd version (proposal under construction)

\documentclass{elsarticle}
% \usepackage{graphicx}
% \usepackage{footmisc}
% \usepackage{amstext}
% \usepackage{amsmath}
% \usepackage{amssymb}
% \usepackage{amsthm}
\usepackage[english]{babel}
\usepackage{libertine}
\usepackage{inconsolata}
% \usepackage[official,right]{eurosym}
\selectlanguage{english}
\frenchspacing

\begin{document}
% \include{preamble}

\title{Enhancing Software Development with Artificial Intelligence}
\author{prof.dr.ir.\ Stef Joosten\footnote{ORCID 0000-0001-8308-0189}}
\author{dr.\ Tim Steenvoorden\footnote{ORCID 0000-0002-8436-2054}}
\address{Open Universiteit Nederland, Heerlen, the Netherlands}

\begin{abstract}
\end{abstract}

\begin{keyword}
    generative software, AI enhanced software engineering, reliability by construction, prototyping, correctness by construction, Ampersand, TopHat
\end{keyword}
\maketitle

\section{Problem area}
    Recent developments in Artificial Intelligence (e.g.~\cite{Chen2021EvaluatingLL,Ernst2022AIDrivenDI,Imai2022}) have attracted
    the attention of the software engineering community to unleash its potential for software engineering.
    This proposal addresses the problem of developing software correctly and reliably by construction,
    leveraging the advances in AI for augmented software engineering for solving an important issue of large IT-projects in practice.

    We hold it to be self-evident that correctness and reliability by construction requires software to be generated from a formal specification.
    Contemporary low-code platforms, SaaS-solutions, and model-driven development environments are examples of software generators,
    all of which require some human programming to suit the deployed artifacts to the concrete requirements from practice.
    Therefore, human errors cannot be ruled out.
    Such software generators are typically based on informal methods, so their reliability and correctness depends on testing practices.
    In contrast, formal specification languages, such as~\cite{Alloy2006,CSP,Z}, give a concrete perspective on correctness.
    However, they typically don't generate full-fledged information systems.
    As far as we know today, Ampersand~\cite{JoostenRAMiCS2017} is the only formal specification language
    that generates complete information systems from formal specification.
    Yet, formal specification languages suffer from the problem that most software engineers lack the education and skills
    to make formal specifications in real-life situations.
    These languages are typically difficult to learn and use, so they are not widely used in practice.

\section{Vision}
    Our vision consists of four beliefs:
\begin{itemize}
\item 
    We believe that formal specifications are prerequisite to the correctness and reliability of software systems.
\item 
    We believe that generating software can eliminate many opportunities for human error.
    For this reason, software generation is a neccessary ingredient of correctness and reliability by construction.
\item 
    We believe that AI can be used to make formal specification languages more accessible to software engineers,
    so they can develop correct and reliable software by construction.
\item 
    We believe that AI might cause a breakthrough in the use of formal specification languages by software engineers.
    If formal methods can be made more accessible to the software engineering community,
    we may have a game changer with a large potential for the quality of governmental, financial, and industrial data in practice.
\end{itemize}
    That summarizes our vision underlying this proposal.

    Our proposal builds on the extensive experience of the Open Universiteit with
    generating information systems~\cite{JoostenRAMiCS2017,Joosten-JLAMP2018,Steenvoorden2022,10.1145/3354166.3354182},
    combined with the AI expertise of the new research group of artificial intelligence.
    For tooling, the proposal builds exclusively on free open source software
    and the results will be available as free open source software to the public.

\section{Benefits}
    This research aims to contribute to 
\begin{itemize}
    \item Correctness by construction by working from a formal specification, enabling software engineers to use proofs besides testing.
    \item Reliability by construction by preventing human programming mistakes.
    \item Productivity of software engineers by eliminating the need to program the software.
    \item Diminished project risk, by automating software production and thus preventing a large class of mistakes.
    \item Short release cycles, so live demonstrations can keep the future user population informed and keep funders happy.
    \item Guarantees on the correctness of software, by eliciting business processes and business rules into formal specifications.
    \item Savings are expected on software development (by generating), testing (by error prevention), and maintenance of both software and tests.
\end{itemize}

\section{Researchers}
    The authors are acknowledged scientists, all with a proven track record in generative software development and workflow modelling.

\newpage

\section{Bibliography}
    \bibliographystyle{elsarticle-harv}
    \bibliography{doc}

\end{document}

### 2023-07-11 2022Migration/articleMigrationFACS.tex
Conclusion

@@
 
-   For some rules, the system is told what to do in order to fix it.
-   Such rules are called enforce rules.
+   For a different kind of constraint, the system is told what to do in order to fix it.
+   Such constraints are called enforce rules.
    We just need to consider one type of enforce rule in this work, its syntax is as follows:
@@
 
-   Adding an enforce rule to the script of an information system specifies a rule $u$ with $\viol{u}{\dataset} = \pop{\id{term} - r}{\dataset}$, and the information system ensures that the rule is satisfied by adding a triple $\triple{x}{r}{y}$ for each $\pair{x}{y} \in \viol{u}{\dataset}$, thus maintaining $\pop{r}{\dataset} \supseteq \pop{\id{term}}{\dataset}$ as an invariant.
-
-   Formally, enforce rules are pairs with $u$ a rule and $r$ a relation, written $r \mapsfrom u$ or equivalently $r \mapsfrom \lambda \dataset.~ u(\dataset)$.
+   Adding an enforce rule to the script of an information system specifies a function $u : \Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}$ with $\viol{u}{\dataset} = \pop{\id{term} - r}{\dataset}$, and the information system ensures that the rule is satisfied by adding a triple $\triple{x}{r}{y}$ for each $\pair{x}{y} \in \viol{u}{\dataset}$, thus maintaining $\pop{r}{\dataset} \supseteq \pop{\id{term}}{\dataset}$ as an invariant.
+
+   Formally, enforce rules are pairs with $u$ a function and $r$ a relation, written $r \mapsfrom u$ or equivalently $r \mapsfrom \lambda \dataset.~ u(\dataset)$.
    The set of enforce rules in a schema is $\enforces$.
@@
    \item $\schema=\triple{\concepts}{\rels}{\rules}$ is a schema (satisfies Equation~\ref{eqn:relationsIntroduceConcepts} and~\ref{eqn:enforceRulesRels}), we write $\concepts_\infsys = \concepts$, $\rels_\infsys = \rels$ and $\rules_\infsys=\rules$;
-   \item all rules are satisfied:
-   \begin{eqnarray}
-   \forall u\in\rules&:\viol{u}{\dataset}=\emptyset
+   \item all rules and enforce rules are satisfied:
+   \begin{align}
+   \forall u\in\rules&:\viol{u}{\dataset}=\emptyset\\
+   \forall (r\mapsfrom u)\in\enforces&:\viol{u}{\dataset} - \pop{r}{\dataset}=\emptyset
    \label{eqn:satisfaction}
-   \end{eqnarray}
+   \end{align}
    \item triples in the data set have their relation mentioned in the schema:
@@
 
-   For this last question, we need to add a small condition:
+   For this last question, we need to add a condition:
    There needs to be an information system that has the desired schema $\schema'$.
    If rules in $\schema'$ are such that they cannot all be satisfied, regardless of the dataset, no such system exists and we have no hope of reaching it.
-   However, let $\pair{\dataset'}{\schema'}$ be an information system, then we can get to it through our migration system.
+   We assume that the rules of $\schema'$ are intended to prevent data pollution, such that a dataset $\dataset'$ describing the `real world' would be one that satisfies them.
+   So let $\pair{\dataset'}{\schema'}$ be an information system, then we can get to it through our migration system.
    To show this, we construct a migration system $\migrsys^C$ satisfying $\pop{f_u}{\dataset_{\migrsys^C}}=\pop{v_u}{\dataset_{\migrsys^C}}$ and $\migrsys'\xrightarrow[I]{\overrightarrow\rels_{\schema'}} \migrsys^C$.
@@
 \section{Conclusions}
-\begin{itemize}
-   \item the part of the data that does not change can be migrated automatically;
-   \item the part of the migration that can be automated is usually not sufficient;
-         it takes additional human creativity to complete the migration specification;
-\end{itemize}
+
+In this paper, we describe the data migration as going from an existing information system to a desired one, where the schema changes.
+As Ampersand generates information systems, creating a new information system can be a small task, allowing for incremental deployment of new features.
+We describe the parts of an information system that have an effect on data pollution.
+We assume that the existing system does not violate any constraints of its schema, but address other forms of data pollution:
+constraints that are not in the schema but are in the desired schema are initially relaxed such that the business can start using the migration system, after which this form of data pollution needs to be addressed by human intervention.
+We propose a method for doing migration such that only a finite amount of human intervention is needed.
+Our method allows a system similar to the desired system to be used while the intervention takes place.
+
+Our proposed migration is certainly not the only approach one could think of.
+However, we have not come across other approaches that allow changing the schema in the presence of constraints.
+As such, we cannot compare our approach against other approaches.
+We envision that one day there will be multiple approaches for migration under a changing schema to choose from.
+For now, our next step is to implement the approach shown here into Ampersand.
+
+This work does not consider what to do about (user) interfaces.
+Instead, it models events by assuming that any change to the dataset can be achieved.
+In practice, such changes need to be achieved through interfaces.
+Most Ampersand systems indeed allow the users of the information system to edit the dataset quite freely through the interfaces.
+However, some interfaces may require certain constraints to be satisfied, which means that interfaces of the desired system might break.
+In the spirit of the approach outlined here, we hope to generate migration interfaces that can replace any broken interfaces until the Moment of Transition.
+How to do this is future work.
+
 %\section{Bibliography}

### 2023-07-12 2022Migration/articleMigrationFACS.tex

@@
 \newcommand{\tgt}[1]{\id{tgt}(#1)}
-\newcommand{\sat}[2]{\id{sat}_{#1}(#2)}
 \newcommand{\viol}[2]{{#1}(#2)}
@@
    The purpose of this work is to automate the data migrations that come with incrementally developing and deploying information systems.
-   This paper proposes a theory for such data migrations in the context of generated information systems%
-\footnote{In the sequel, the word ``system'' refers to the phrase ``information system''. This simplifies the language a little. }.
+   This paper proposes a theory for such data migrations in the context of generated information systems.
+   In the sequel, the word ``system'' refers to the phrase ``information system'', to simplify the language a little.
 
@@
    \item {\em zero down-time}\\The business continues during the migration without interruption.
-   \item {\em generative software development}\\A compiler exists to generate an information system from a given schema.
+   \item {\em generative software development}\\A compiler exists to generate a system from a given schema.
 \end{itemize}
@@
    The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
-   This section defines information systems, after which we define a migration procedure from one information system to another.
+   This section defines information systems, after which we define a migration procedure from one system to another.
 
@@
    A data set $\dataset$ describes a set of structured data, which is typically stored persistently in a database of some kind.
-   The notation $\dataset_{\infsys}$ refers to the data set of a particular information system $\infsys$.
+   The notation $\dataset_{\infsys}$ refers to the data set of a particular system $\infsys$.
    The purpose of a data set is to describe the data of a system at one point in time. 
@@
    %set operators $\cup$, $\cap$, and $-$ can be used on $\instance$.
-
-%    Specialization, also known as {\em generalization} or {\em subtyping}, is a relation between concepts.
-%    The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
-% \begin{equation}
-%    \label{eqn:specialization}
-%    A\isa B\ \Leftrightarrow\ \forall a: a\instance A\rightarrow a\instance B
-% \end{equation}
-%    Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.
-%    In her script, a user can declare a specialization as a pair of concepts in the relation $\isa$.
-%    A compiler can construct $\isa$ as the transitive, reflexive closure of all user-defined specializations.
-%    It must make sure that $\isa$ is antisymmetric, so that $\isa$ is a partial order of concepts.
-%    Specialization causes atoms to have multiple concepts of which they can be an instance.
-%    As a result, if an atom is an instance of concept $A$ and $A\isa B$,
-%    this atom has all properties that atoms of type $B$ have.
 
@@
    Schemas serve to capture the semantics of an information system~\cite{Spivak2012}.
-   We consider a schema to represent the static part of an information system, defining concepts, relations, and rules.
+   We consider a schema to represent the static part of a system, defining concepts, relations, and rules.
    A software engineer defines a schema on design time, so that semantic checks can be implemented at compile time.
@@
    $\rels\subseteq \Rels$ is a finite set of relations,
-   and $\rules\subseteq \left(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}\right)$ is a finite set of rules.
+   and $\rules\subseteq \powerset{\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}}$ is a finite set of rules.
 
@@
    For a different kind of constraint, the system is told what to do in order to fix it.
-   Such constraints are called enforce rules.
-   We just need to consider one type of enforce rule in this work, its syntax is as follows:
-
+   Such constraints are called enforcement rules.
+   We just use one type of enforcement rule\footnote{Ampersand has several types of enforcement rules.} in this work.
+   Its syntax is as follows:
 \begin{align}
@@
 
-   Adding an enforce rule to the script of an information system specifies a function $u : \Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}$ with $\viol{u}{\dataset} = \pop{\id{term} - r}{\dataset}$, and the information system ensures that the rule is satisfied by adding a triple $\triple{x}{r}{y}$ for each $\pair{x}{y} \in \viol{u}{\dataset}$, thus maintaining $\pop{r}{\dataset} \supseteq \pop{\id{term}}{\dataset}$ as an invariant.
-
-   Formally, enforce rules are pairs with $u$ a function and $r$ a relation, written $r \mapsfrom u$ or equivalently $r \mapsfrom \lambda \dataset.~ u(\dataset)$.
-   The set of enforce rules in a schema is $\enforces$.
+   Adding an enforcement rule to the script of a system specifies a function $u : \Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}$
+   with $\viol{u}{\dataset} = \pop{\id{term} - r}{\dataset}$.
+   The system ensures that the rule is satisfied by adding a triple $\triple{x}{r}{y}$ for each $\pair{x}{y} \in \viol{u}{\dataset}$.
+   Thus the system maintains $\pop{r}{\dataset} \supseteq \pop{\id{term}}{\dataset}$ as an invariant.
+
+   Formally, enforcement rules are pairs with $u$ a function and $r$ a relation, written $r \mapsfrom u$ or equivalently $r \mapsfrom \lambda \dataset.~ u(\dataset)$.
+   The set of enforcement rules in a schema is $\enforces$.
 
@@
    Requirement~\ref{eqn:relationsIntroduceConcepts} ensures that concepts mentioned in a relation are defined in the schema.
-   The type system of Ampersand~\cite{vdWoude2011} ensures that every information system it generates satisfies this requirement.
+   The type system of Ampersand~\cite{vdWoude2011} ensures that every system it generates satisfies this requirement.
    This is also known as static typing, which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.
 
-   %The notation $\schema_{\infsys}$ refers to the schema of information system $\infsys$.
+   %The notation $\schema_{\infsys}$ refers to the schema of system $\infsys$.
    When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, $\rules_{\schema}$, and $\enforces_{\schema}$
@@
 
-%   Note that these requirements do not depend on any particular data set,
-%   so they can be checked at compile time.
-%   This is also known as static typing,
-%   which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.
-
-%    For every rule $u$ in the schema, we assume there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
-%    that represents the set of violations%
-% \footnote{The definition of ``violation'' is not needed in this paper. We only use the absence of violations to show satisfaction of a constraint.}
-%    of rule $u$ in any data set.
-%    We also assume a function $\id{sign} : \Rules\rightarrow\Pair{\Concepts}{\Concepts}$
-%    that represents the signature of a rule, i.e. the type of atoms in violations:
-% \begin{eqnarray}
-%    \sign{u}=\pair{A}{B}\ \wedge\ \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\instance A\wedge b\instance B
-%    \label{eqn:wellTypedEdged violations}
-% \end{eqnarray}
-%    To determine whether rule $u$ satisfies data set $\dataset$,
-%    we define $\sat{u}{\Dataset}$:
-% \begin{eqnarray}
-%    \sat{u}{\dataset}&\Leftrightarrow&\viol{u}{\dataset}=\emptyset
-%    \label{eqn:sat}
-% \end{eqnarray}
-%    Violations are used to produce meaningful error messages and to trigger changes in the data set to restore invariance of that rule.
-
-   % The maintainance relation between roles and rules is used to determine who is allowed to change the data set.
-   % For every rule $u$ in the schema, there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
-   % that represents the set of violations.%
-
-   % In practice, there are different ways of enforcing rules: automatic enforcement and manual enforcement.
-
-   % Automatic enforcement is specified by the developer with a special syntax:
-   % the {\em enforcement rule}.
-   % An enforcement rule specifies not only the rule,
-   % but it implicitly defines event $e'$ to restore the invariance.
-   % The system must have an engine that restores invariance of all enforcement rules,
-   % so users will experience that these rules are always satisfied.
-
-   % Manual enforcement means that the system (temporarily) allows violations,
-   % to let a user ``invent'' a reaction $e'$ that restores invariance.
-   % That is why the term $(\exists o\in\roles:\ o\maintain u)$ is added to requirement~\ref{eqn:satisfaction}.
-   % It allows the compiler to generate code for users with role $o$,
-   % so they can restore the invariant manually.
-   % In the Ampersand language, a developer explicitly assigns a rule $u$ to a role $o$,
-   % to keep control over the type of persons that are allowed to ``restore broken rules''.
-   % As a consequence, a mechanism is needed to notify those persons of the work to be done.
-
-   % If a rule $u$ is not assigned to any role and it is not an enforcement rule,
-   % the system must prevent any event $e$ from having any effect on the database.
-   % So, if $e$ would violate rule $u$, the system must reject $e$.
-   % This implements blocking behaviour, and must be accompanied by an error message.
-
 \subsection{Information Systems}
 \label{sct:Information Systems}
-%   An information system has a changing data set because of events that occur.
-%   Whenever the system ``observes'' an event, it attempts to change its data set accordingly.
-%   However, every rule in the schema represents a constraint,
-%   which the data set must invariably satisfy.
-%   That is why rules are also known as ``invariants''.
-%
-%   Events are taken from an infinite set, $\Events$.
-%   An event is denoted as $\evt{d^-}{d^+}$, in which $d^-$ and $d^+$ are data sets.
-%   Every event is a function $\Dataset\rightarrow\Dataset$, so $\evt{d^-}{d^+}(\dataset)$ is a data set.
-%   Equation~\ref{eqn:event} defines the effect of an event on a data set.
-%\begin{eqnarray}
-%      \evt{d^-}{d^+}(\dataset)&=&\la(\triples_{\dataset}-\triples_{d^-})\cup\triples_{d^+},(\instance_{\dataset}-\instance_{d^-})\cup\instance_{d^+}\ra
-%\label{eqn:event}
-%\end{eqnarray}
-%   In English, we would say that an event $\evt{d^-}{d^+}$ removes the triples in $d^-$ from $\dataset$ and then inserts the triples in $d^+$ into the result.
-%   In the sequel, variables $e$, $e'$, $e''$, etc. represent events.
-%   As soon as a system $\infsys$ observes an event $e$,
-%   it applies $e$ to its data set $\dataset$.
-%   Yet, if $e(\dataset)$ does not satisfy a rule $u$ from schema $\schema_\infsys$,
-%   another event is needed, say $e'$, to ensure that $\sat{u}{e'(e(\dataset))}$.
-%   The phrase ``to restore an invariant'' or ``to restore invariance of a rule''
-%   means that the system will change $e(\dataset)$ by applying another event $e'$ to satisfy requirement~\ref{eqn:enforce}.
-%   The notation $\enf_{u,e}$ represents this combination of events $e$ and $e'$.
-%   So, $\enf_{u,e}$ must satisfy:
-%\begin{eqnarray}
-%   \sat{u}{\dataset}&\Rightarrow&\sat{u}{\enforce{u,e}{\dataset}}
-%\label{eqn:enforce}
-%\end{eqnarray}
-
    Let us now define the notion of an information system.
@@
 \begin{itemize}
-   \item $\dataset=\pair{\triples}{\instance}$ is a data set (satisfies Equation~\ref{eqn:wellTypedEdge}), we write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$;
-   \item $\schema=\triple{\concepts}{\rels}{\rules}$ is a schema (satisfies Equation~\ref{eqn:relationsIntroduceConcepts} and~\ref{eqn:enforceRulesRels}), we write $\concepts_\infsys = \concepts$, $\rels_\infsys = \rels$ and $\rules_\infsys=\rules$;
-   \item all rules and enforce rules are satisfied:
-   \begin{align}
-   \forall u\in\rules&:\viol{u}{\dataset}=\emptyset\\
-   \forall (r\mapsfrom u)\in\enforces&:\viol{u}{\dataset} - \pop{r}{\dataset}=\emptyset
+   \item $\dataset=\pair{\triples}{\instance}$ is a data set (satisfies equation~\ref{eqn:wellTypedEdge}), we write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$;
+   \item $\schema=\quadruple{\concepts}{\rels}{\rules}{\enforces}$ is a schema (satisfies equation~\ref{eqn:relationsIntroduceConcepts} and~\ref{eqn:enforceRulesRels}), we write $\concepts_\infsys = \concepts$, $\rels_\infsys = \rels$, $\rules_\infsys=\rules$, and $\enforces_\infsys=\enforces$;
+   \item all rules and enforcement rules are satisfied:
+   \begin{align}
+   \forall u\in\rules&.~\viol{u}{\dataset}=\emptyset\\
+   \forall (r\mapsfrom u)\in\enforces&.~\viol{u}{\dataset} - \pop{r}{\dataset}=\emptyset
    \label{eqn:satisfaction}
@@
    As the system responds to events, it changes its data set, while the schema remains the same.
-   In this paper, it suffices to define an event as a pair of information systems for which the schema stays constant.
+   In this paper, it suffices to define an event as a pair of systems for which the schema stays constant.
    Events are categorized by what part of the data set changes.
@@
 \begin{definition}[Event]
-   Let $R \subseteq \Rels$ and $I \subseteq \Rels$ be sets of relations, and let $\infsys' = \pair{\schema}{\dataset'}$ be an information system, and:
+   Let $R \subseteq \Rels$ and $I \subseteq \Rels$ be sets of relations, and let $\infsys=\pair{\schema}{\dataset}$ and  $\infsys'=\pair{\schema}{\dataset'}$ be information systems, and:
       \begin{align}
@@
    \end{align}
-   Then we say that $\pair{\pair{\schema}{\dataset}}{\infsys'}$ is an event with scope $R$ inserting in $I$.
-   We use the notation $\pair{\schema}{\dataset} \xrightarrow[I]{R} \infsys'$ to indicate this.
+%   Then we say that $\pair{\pair{\schema}{\dataset}}{\infsys'}$ is an event with scope $R$ inserting in $I$.
+   Then we say that $\infsys\xrightarrow[I]{R} \infsys'$ is an event
+   that brings the system $\infsys$ to $\infsys'$ inserting pairs in relations from $I$ and updating (i.e. deleting and inserting) pairs in $R$.
+%   We use the notation $\pair{\schema}{\dataset} \xrightarrow[I]{R} \infsys'$ to indicate this.
 \end{definition}
 
-   Equation~\ref{eqn:eventUnchanged} states that the triples of those for relations in $I$ or $R$ are the only ones to have changed, and Equation~\ref{eqn:eventInsert} states that of the ones that are only in $I$, no triples were removed.
+   equation~\ref{eqn:eventUnchanged} states that the triples of those for relations in $I$ or $R$ are the only ones to have changed, and equation~\ref{eqn:eventInsert} states that of the ones that are only in $I$, no triples were removed.
    %If $I$ or $R$ is the empty set, we omit it from the arrow, so $\infsys \xrightarrow{R} \infsys'$ is a notation for $\infsys \xrightarrow[\emptyset]{R} \infsys'$.
@@
 
-   We use inserting events to model the effect of enforce rules.
-%   If $R$ is a set of relations that covers those that occur in the rule, then changes to an information system $\infsys$ can be described by the event $\infsys \xrightarrow[\{r\}]{R} \infsys'$.
+   We use inserting events to model the effect of enforcement rules.
+%   If $R$ is a set of relations that covers those that occur in the rule, then changes to an system $\infsys$ can be described by the event $\infsys \xrightarrow[\{r\}]{R} \infsys'$.
 
@@
 \label{sct:Example existing IS}
-   Having defined an information system in mathematical terms, let us discuss a small example.
+   Having defined information systems in mathematical terms, let us discuss a small example.
    It is written in the language Ampersand to make it more appealing to read.
@@
 \end{verbatim}
-   The Ampersand compiler defines $\sat{\tt EnrollRule}{\dataset}$ as:
+   The Ampersand compiler defines $\viol{\tt EnrollRule}{\dataset}$ as
+\begin{equation}
+   \pop{\tt isEnrolledFor}{\dataset}-\{\pair{s}{m} \mid\exists c.~s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\}
+\end{equation}
+   This rule is satisfied if $\viol{\tt EnrollRule}{\dataset}=\emptyset$, which means that
 \begin{equation}
    \begin{array}{l}
-   \forall \pair{s}{m}\in\pop{\tt isEnrolledFor}{\dataset}\ \exists c\in\text{\tt Course}:\\
+   \forall \pair{s}{m}\in\pop{\tt isEnrolledFor}{\dataset}.~ \exists c\in\text{\tt Course}~.\\
    s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
@@
 \end{equation}
-   As relation $\declare{\tt isEnrolledFor}{\tt Student}{\tt Module}$ is empty, rule {\tt EnrollRule} is satisfied in this example.
+   In this example, relation $\declare{\tt isEnrolledFor}{\tt Student}{\tt Module}$ is empty, so rule {\tt EnrollRule} is satisfied.
 
@@
              \end{array}\\
-   \rules&=&\{ {\tt EnrollRule}\}
+   \rules&=&\{ {\tt EnrollRule}\}\\
+   \enforces&=&\emptyset
   \end{array}
@@
    % The (nonempty) data set satisfies this rule, which meets requirement~\ref{eqn:consistentRules}.
-   So, the schema $\schema=\triple{\concepts}{\rels}{\rules}$ satisfies all requirements from section~\ref{sct:Schemas}.
+   So, the schema $\schema=\quadruple{\concepts}{\rels}{\rules}{\enforces}$ satisfies all requirements from section~\ref{sct:Schemas}.
 
@@
 
-%    Now suppose that the script also contains the following line:
-% \begin{verbatim}
-% ROLE Administrator MAINTAINS EnrollRule
-% \end{verbatim}
-%    Ampersand generates a set of roles $\roles=\{{\tt Administrator}\}$.
-%    The relation $\maintain$ contains one pair only, $\pair{{\tt Administrator}}{{\tt EnrollRule}}$.
-%    Requirement~\ref{eqn:satisfaction} is still satisfied because now there is a role to maintain rule {\tt EnrollRule}.
-%    The run-time engine will no longer enforce satisfaction of rule {\tt EnrollRule}, but leave it up to an administrator.
-
-%\subsection{Process of Data Migration}
-%   Section~\ref{sct:Information Systems} introduces a single information system.
-%   The transition from an existing system $\infsys$ to the desired one $\infsys'$ is the topic of this section.
-%   Let us first sketch the steps to migrate system $\infsys$ to $\infsys'$.
-%\begin{enumerate}
-%   \item Let the existing information system $\infsys$ have data set $\dataset$ and schema $\schema$.
-%         A developer defines a new information system, $\infsys'$, which contains a data set of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{In this paper, we only consider the functionality captured by rules.} as illustrated in Figure~\ref{fig:post-migration}.
-%         After the migration, the resulting system will be $\infsys'$, enhanced with the triples from $\dataset$ that satisfy $\schema'$,
-%         and with additional changes made by the migration engineer (see step~\ref{item3}) and by users who have been using the system during the migration.
-%         $\dataset'$ may contain some new data,
-%         for instance, to initialize new features at the moment of transition (MoT).
-%         If schemas $\schema$ equals $\schema'$, the data migration is unnecessary, so in the sequel we assume they are different.
-%   \item The migration engineer uses a software generator to generate a migration script, $\migrsys$,
-%         This is a system that includes both $\infsys$ and $\infsys'$.
-%         $\migrsys$ is meant to migrate the data and define the semantics during the migration phase.
-%         It contains enforcement rules that migrate existing data to the desired data set.
-%         It also contains data structures and rules that are needed for the duration of the migration phase.
-%%         The generated script must be free of type errors, to ensure that $\sat{u}{\dataset}$ exists for every rule $u$ and data set $\dataset$.
-%%         The proof is available \href{https://www.isa-afp.org/}{here}
-%   \item\label{item3}
-%         In many cases, the schema $\schema_{\migrsys}$ will be enough to perform the migration.
-%         However, a migration engineer may have specific requirements that call for changes in $\migrsys$.
-%         For that purpose, the generator generates the migration script as an Ampersand script,
-%         so the migration engineer can accommodate such requirements.
-%         This can be necessary, for example, to resolve data pollution that violates a constraint that was not in schema $\schema_{\infsys}$.
-%         She can either implement enforcement rules or use the business to resolve the issues.
-%         She can test the resulting migration script, $\migrsys'$ separately before taking the desired system to production.
-%         The migration engineer alters $\schema$ nor $\schema'$ while adapting $\migrsys'$ to the specific needs of the current migration.
-%   \item Then, the migration engineer deploys system $\migrsys'$.
-%         This event starts the copying of data from $\dataset$ to $\dataset'$, as specified in $\migrsys'$ (Figure~\ref{fig:migration phase}).
-%         During this phase, the ingress is still referring users to the existing system,
-%         so they will hardly notice that the desired system is being fired up and loaded with data.
-%         Users are still changing $\dataset$ (both inserts and deletes),
-%         but these changes are being transferred to $\dataset'$ as well.
-%         This step is complete when all data from $\dataset$ has been copied to $\dataset'$ and all invariants of the migration system are true.
-%   \item At this moment, the ingress switches the incoming traffic from $\infsys$ to $\migrsys'$,
-%         so $\migrsys'$ effectively takes over.
-%         This event marks the MoT.
-%         Since $\infsys$ may still have some violations due to last-minute edits,
-%         it may take some (short but finite) time until the invariants of $\infsys$ are true.
-%         After that, data set $\dataset$ will no longer change.
-%   \item From this moment, users experience the functionality of the migration system $\migrsys'$.
-%         This system must ensure that all invariants of $\schema'$ are true before the migration is complete.
-%         It no longer ensures the rules of $\schema$, which govern the existing system.
-%         Since $\dataset$ is no longer changing, the migration engineer can now focus on the remaining violations of $\schema'$.
-%         Users of the migration system will deal with process rules in the migration system that are still violated.
-%         Once all invariants of $\migrsys'$ are true, the migration is complete.
-%         This state marks the moment of cleanup (MoC)
-%   \item After the MoC, the ingress switches the incoming traffic from $\migrsys'$ to $\infsys'$.
-%         The migration system is no longer needed and can be taken out of production, together with the existing system $\infsys$.
-%         The desired system remains in production and the migration is finished.
-%         This establishes the post-migration situation (Figure~\ref{fig:post-migration}).
-%\end{enumerate}
-
-%SJC on Formalization: it's often better to put these inline where we first use them. Leaving this as a comment for easy copy-paste.
-%    
-%    \subsection{Formalization}
-%    Let us define the instruments needed to describe the derivation of $\schema_{\migrsys}$.
-%    \begin{definition}[union of data sets]
-%    \label{def:graph_union_wellTyped_if_parts_wellTyped}
-%    \[\pair{\triples_1}{\instance_1}\cup\pair{\triples_2}{\instance_2}\ =\ \pair{\triples_1\cup\triples_2}{\instance_1\cup\instance_2}\]
-%    \end{definition}
-%    
-%    \begin{lemma}
-%    \label{lemma:graph_union_wellTyped_if_parts_wellTyped}
-%    If $\dataset_1$ and $\dataset_2$ are data sets, then $\dataset_1\cup\dataset_2$ is a data set.
-%    \end{lemma}
-%    
-%    \begin{definition}[map relation names in data sets]
-%    \label{def:map_labels_preserves_wellTypedness}
-%    Let $h:\Rels\rightarrow\Rels$, then
-%    \[\begin{array}{rcl}
-%      \maprel{h}{\pair{\triples}{\instance}}&=&\pair{\maprel{h}{\triples}}{\instance}\\
-%      \maprel{h}{\triple{a}{r}{b}}&=&\triple{a}{h(r)}{b}
-%    \end{array}\]
-%    \end{definition}
-%    
-%    \begin{lemma}
-%    \label{lemma:map_labels_preserves_wellTypedness}
-%    If $\dataset$ is a data set and $h:\Rels\rightarrow\Rels$ is an injective function, then $\maprel{h}{\dataset}$ is a data set.
-%    \end{lemma}
-%    
-%    \begin{definition}[disjoint union of data sets]
-%    \begin{eqnarray}
-%    \pair{\triples}{\instance}\sqcup\pair{\triples'}{\instance'}&=&\pair{\triples\uplus\triples'}{\instance\uplus^2\instance'}\\
-%      X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
-%      X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}
-%    \end{eqnarray}
-%    \end{definition}
-%    In a disjoint union, the instances of the two data sets are relabeled to avoid name clashes.
-%    In practice, we will not use 0 and 1 as labels for $x$ and $y$, but rather extend their names by prefixing.
-%    \begin{lemma}
-%    If $\dataset$ and $\dataset'$ are data sets, then so is $\dataset\sqcup\dataset'$.
-%    \end{lemma}
-%    This lemma is being proven in Isabelle/HOL~\cite{Isabelle}. The proof is published \href{location.domain}{here}
-%    
-%    \begin{definition}[disjoint union of schemas]
-%    \begin{eqnarray}
-%    \triple{\concepts}{\rels}{\rules}\sqcup\triple{\concepts'}{\rels'}{\rules'}&=&\triple{\concepts\uplus\concepts}{\rels\uplus\rels'}{\rules\uplus\rules'}
-%    \end{eqnarray}
-%    \end{definition}
-%    % \begin{definition}[disjoint union of data sets]
-%    % \begin{eqnarray}
-%    %    \omit\rlap{$\la\atoms,\concepts,\instance,\isa,\rels,\dataset\ra\sqcup\la\atoms',\concepts',\instance',\isa',\rels',\dataset'\ra$}\notag\\
-%    %    &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\instanceuplus^2\instance',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \dataset\uplus^5\dataset'\ra\notag\\
-%    %       X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
-%    %       X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}\\
-%    %       X\uplus^3 Y&=&\begin{array}[t]{@{}l}\{\declare{(n,0)}{(A,0)}{(B,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\declare{(n,1)}{(A,1)}{(B,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}\\
-%    %       X\uplus^5 Y&=&\begin{array}[t]{@{}l}\{\triple{(a,0)}{\declare{(n,0)}{(A,0)}{(B,0)}}{(b,0)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in X\}\ \cup\\ \{\triple{(a,1)}{\declare{(n,1)}{(A,1)}{(B,1)}}{(b,1)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in Y\}\end{array}
-%    % \end{eqnarray}
-%    % \end{definition}
-%    \begin{lemma}
-%    If $\schema$ and $\schema'$ are schemas, then so is $\schema\sqcup\schema'$.
-%    \end{lemma}
-%    The proof of this lemma is published \href{location.domain}{here}
-%    
-%    To implement the disjoint union, the following relabelling functions are needed:
-%    \begin{definition}[relabel concepts]
-%    \[\subst{C}{D}{A}\ =\ \text{\bf if}\ C=A\ \text{\bf then}\ D\ \text{\bf else}\ A\]
-%    \end{definition}
-%    \begin{definition}[relabel concepts in triples]
-%    \[\subst{C}{D}{\triple{a}{r}{b}} = \triple{a}{\subst{C}{D}{r}}{b}\]
-%    \end{definition}
-%    \begin{definition}[relabel concepts in relations]
-%    \[\subst{C}{D}{(\declare{n}{A}{B})} = \declare{n}{\subst{C}{D}{A}}{\subst{C}{D}{B}}\]
-%    \end{definition}
-%    \begin{definition}[relabel concepts in $\instance$]
-%    \[a\ \subst{C}{D}{\instance}\ \subst{C}{D}{A}\ \Leftrightarrow\ a\instance A\]
-%    \end{definition}
-%    \begin{definition}[relabel concepts in data sets]
-%    \[\begin{array}{rcl}
-%      \rlap{$\subst{C}{D}{\pair{\triples}{\instance}}$}\\
-%      &=&\pair{\{\subst{C}{D}{t}\mid t\in\triples\}}{\subst{C}{D}{\instance}}
-%    \end{array}\]
-%    \end{definition}
-%    The relabeling of concepts in rules must satisfy:
-%    \begin{equation}
-%    \sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{C}{D}{u}}{\subst{C}{D}{\dataset}}
-%    \end{equation}
-%    \begin{definition}[relabel concepts in schemas]
-%    \begin{eqnarray}
-%      \rlap{$\subst{C}{D}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
-%      &=&\begin{array}[t]{l@{}l}
-%         \la&\{\subst{C}{D}{c}\mid c\in\concepts\}\\
-%         ,&\{\subst{C}{D}{r}\mid r\in\rels\}\\
-%         ,&\{\subst{C}{D}{u}\mid u\in\rules\}\ \ra\notag
-%         \end{array}
-%    \end{eqnarray}
-%    \end{definition}
-%    \begin{definition}[relabel relations]
-%    \[\subst{\declare{m}{C}{D}}{m'}{{\declare{n}{A}{B}}}\ =\ \text{\bf if}\ n=m\wedge A=C\wedge B=D\ \text{\bf then}\ \declare{m'}{A}{B}\ \text{\bf else}\ \declare{m}{A}{B}\]
-%    \end{definition}
-%    \begin{definition}[relabel relations in triples]
-%    \[\subst{r'}{m'}{\triple{a}{r}{b}} = \triple{a}{\subst{r'}{m'}{r}}{b}\]
-%    \end{definition}
-%    \begin{definition}[relabel relations in data sets]
-%    \[\begin{array}{rcl}
-%      \rlap{$\subst{r'}{m'}{\pair{\triples}{\instance}}$}\\
-%      &=&\pair{\{\subst{r'}{m'}{t}\mid t\in\triples\}}{\instance}
-%    \end{array}\]
-%    \end{definition}
-%    The relabeling of relation names in rules must satisfy:
-%    \begin{equation}
-%    \sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{r'}{m'}{u}}{\subst{r'}{m'}{\dataset}}
-%    \end{equation}
-%    \begin{definition}[relabel relations in schemas]
-%    \begin{eqnarray}
-%      \rlap{$\subst{r'}{m'}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
-%      &=&\triple{\concepts}{\{\subst{r'}{m'}{r} \mid r\in\rels\}}{\{\subst{r'}{m'}{u} \mid u\in\rules\}}\notag
-%    \end{eqnarray}
-%    \end{definition}
-%    
-%    The relabeling of concepts is not trivial because of specialization.
-%    To illustrate this point, suppose $D\isa A$.
-%    Then the substitution $[C\rightarrow D]$ means that every atom $a$ that used to be of concept $C$ in the existing system, is now not only a $D$ but also an $A$.
-%    For example, by relabeling the concept {\tt Hotel} to {\tt Motel}, in a situation where ${\tt Motel}\isa{\tt Parking}$,
-%    all atoms that represent hotels might suddenly be expected to have parking lots.
-%    So, whereas such examples may fly in a technical sense, the developer must use renaming with care to preserve the intended semantics.
-
 \section{Generating a Migration Script}
    A migration script takes care of two things: copying data and ensuring that the migration system is a proper information system.
-   We first illustrate how this works by example, using Ampersand syntax.
-   We can then describe what this means in terms of the definitions of information system.
+   Sections~\ref{sec:loading} and~\ref{sec:violations on invariants} illustrate how this works by example, using the Ampersand syntax.
+   Then, section~\ref{General migration script} defines what this means in general terms.
 
@@
 \footnote{The prefix {\tt old.} refers to concepts, relations, and rules from the existing system (section~\ref{sct:Example existing IS}).
-The prefix {\tt new.} refers to items from the desired information system.}:
+The prefix {\tt new.} refers to items from the desired system.}:
 \begin{verbatim}
@@
 \end{verbatim}
-   The relation `old.isPartOf' resides in the existing system and will not change after the MoT.
+   The relation {\tt old.isPartOf} resides in the existing system and will not change after the MoT.
 
@@
 
-   \subsection{Dealing with violations on invariants}
+\subsection{Dealing with violations on invariants}\label{sec:violations on invariants}
    Suppose the script for the desired system adds an invariant that is not present in the existing system:
@@
    We again look at what happens at runtime:
-   Rule~\ref{fill viols} will adds all violations present in the old system to {\tt viols\_UNI}.
+   Enforcement rule~\ref{fill viols} adds all violations present in the old system to {\tt viols\_UNI}.
    Until the MoT, rule~\ref{fill fixed} does nothing, since {\tt viols\_UNI} equals $\viol{\tt UNI}{\dataset}$.
-   At that point, rule~\ref{rule uni} is satisfied, since {\tt fixed\_UNI} is empty.
+   At that point, rule~\ref{rule uni} is satisfied, since the right-hand side of equation~\ref{rule uni} is empty.
    As violations are fixed, rule~\ref{fill fixed} records this by adding to {\tt fixed\_UNI}.
    This way, previously fixed errors cannot reoccur.
-   Once {\tt viols\_UNI} is equal to {\tt fixed\_UNI}, the \verb=UNI= requirement holds and the information system is ready for the MoC in regard to this rule.
+   Once {\tt viols\_UNI} is equal to {\tt fixed\_UNI}, the \verb=UNI= requirement holds and the system is ready for the MoC in regard to this rule.
 
@@
 
-   Despite the syntax, Line~{\ref{warnings}} is not a rule of the generated information system in the formal sense:
-   By assigning a role, the responsibility of maintaining the rule no longer lies with the information system.
+   Despite the syntax, Line~{\ref{warnings}} is not a rule of the generated system in the formal sense:
+   By assigning a role, the responsibility of maintaining the rule no longer lies with the system.
    Instead, users with the migration helper role can see the violations and act on it as they see fit, so the only effect of Line~{\ref{warnings}} is that such users will be shown its violations.
 
-% Het volgende is wellicht iets teveel van het goede. Het lijkt me dat de lezer het voorgaande al wel snapt.
-%    The following table summarizes what happens when pairs are inserted or deleted from {\tt old.takes} and {\tt new.takes}.
-% \begin{itemize}
-% \item If pair $p$ is inserted in {\tt old.takes} and $p$ is not in {\tt copied\_takes},
-%       then rule~\ref{difference} causes the engine to insert $p$ in {\tt new.takes}.
-%       Subsequently, rule~\ref{fill copies with new.takes} causes the engine to insert $p$ in {\tt copied\_takes} as well.
-% \item If pair $p$ is inserted in {\tt old.takes} and $p$ is already in {\tt copied\_takes},
-%       then nothing happens.
-% \item If pair $p$ is deleted from {\tt old.takes} and $p$ is in {\tt copied\_takes},
-%       then rule~\ref{del copies from old.takes} causes the engine to delete $p$ from {\tt copied\_takes}.
-% \item If pair $p$ is deleted from {\tt old.takes} and $p$ is not in {\tt copied\_takes},
-%       then nothing happens.
-% \item If pair $p$ is inserted in {\tt new.takes},
-%       rule~\ref{fill copies with new.takes} causes the engine to insert $p$ into {\tt copied\_takes}.
-% \item If pair $p$ is deleted from {\tt new.takes},
-%       rule~\ref{fill copies with new.takes} causes the engine to delete $p$ from {\tt copied\_takes}.
-%       If this pair is in {\tt old.takes}, rule~\ref{difference} causes the engine to insert $p$ back into {\tt new.takes}.
-% \end{itemize}
-
-\subsection{General migration script}
+\subsection{General migration script}\label{General migration script}
    So far, we illustrated how to do migration by giving two concrete examples.
@@
    \end{align}
-   \item For each rule, we generate two helper relations, viols and fixed. Note that Rule~\ref{warnings} only influences the user interface and therefore is not modeled here.
+\item For each rule, we generate two helper relations, {\tt viols} and {\tt fixed}. Note that Rule~\ref{warnings} only influences the user interface and therefore is not modeled here.
    \begin{align}
@@
    \overrightarrow{\dataset'} &= \pair{\{\triple{a}{\overrightarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_\dataset\}}{\instance_{\dataset'}} \\
-   \rels_\theenumi &= \{v_u\mid u \in \rules_{\schema'}\} \cup \{f_u\mid u \in \rules_{\schema'}\}\\
-   \rules_\migrsys &= \{\lambda {\dataset'}.~ \viol{u}{\widetilde{\dataset'}} - \pop{v_u - f_u}{\dataset'} \mid u\in\rules_{\schema'}\}\label{eqn:migrsysrules}\\
-   \enforces_{\theenumi} &=\{v_u \mapsfrom \lambda \dataset'.~ \viol{u}{\widetilde{\dataset'}}\} \cup
-    \{f_u \mapsfrom \lambda \dataset'.~ v_u - \viol{u}{\overrightarrow{\dataset'}}\}\label{eqn:enforceForRules}
+   \rels_\theenumi &= \{{\tt viols}_u\mid u \in \rules_{\schema'}\} \cup \{{\tt fixed}_u\mid u \in \rules_{\schema'}\}\\
+   \rules_\migrsys &= \{\lambda {\dataset'}.~ \viol{u}{\widetilde{\dataset'}} - \pop{{\tt viols}_u - {\tt fixed}_u}{\dataset'} \mid u\in\rules_{\schema'}\}\label{eqn:migrsysrules}\\
+   \enforces_{\theenumi} &=\{{\tt viols}_u \mapsfrom \lambda \dataset'.~ \viol{u}{\widetilde{\dataset'}}\} \cup
+    \{{\tt fixed}_u \mapsfrom \lambda \dataset'.~ {\tt viols}_u - \viol{u}{\overrightarrow{\dataset'}}\}\label{eqn:enforceForRules}
    \end{align}
@@
 \subsection{Validation}
-   We first observe that $\migrsys = \pair{\pair{\triples_\migrsys}{\instance_\migrsys}}{\quadruple{\concepts_\migrsys}{\rels_\migrsys}{\rules_\migrsys}{\enforces_\migrsys}}$ is an information system after application of the enforce rules:
-   the only rules are those introduced in Eqution~\ref{eqn:migrsysrules}, and enforce rules from Equation~\ref{eqn:enforceForRules} ensure that $\popF{v_u} = u$ initially, while $\popF{f_u}$ is empty.
-   Hence, the MoT can happen after the enforce rules have been applied, giving us $\migrsys'$.
+   We first observe that $\migrsys = \pair{\pair{\triples_\migrsys}{\instance_\migrsys}}{\quadruple{\concepts_\migrsys}{\rels_\migrsys}{\rules_\migrsys}{\enforces_\migrsys}}$ is a system after application of the enforcement rules:
+   the only rules are those introduced in equation~\ref{eqn:migrsysrules}, and enforcement rules from equation~\ref{eqn:enforceForRules} ensure that $\popF{{\tt viols}_u} = u$ initially, while $\popF{{\tt fixed}_u}$ is empty.
+   Hence, the MoT can happen after the enforcement rules have been applied, giving us $\migrsys'$.
    The event $\migrsys \xrightarrow[\rels_2 \cup \rels_3]{\overrightarrow\rels_{\schema'}}\migrsys'$ represents the MoT.
 
-   The MoC is reached for dataset $\dataset'$ whenever $\pop{v_u - f_u}{\dataset'}=\emptyset$ for all rules.
-   This should happen after a finite number of inserting events: let $I = \{f_u \mid u\in\rules_{\schema'}\}$ be the relations that keep track of which violations have been fixed, then there is no infinite chain of systems $\migrsys' = \migrsys_0, \migrsys_1,\ldots$ such that $\migrsys_i\xrightarrowdbl[I]{\overrightarrow\rels_{\schema'}}\migrsys_{i+1}$.
+   The MoC is reached for dataset $\dataset'$ whenever $\pop{{\tt viols}_u - {\tt fixed}_u}{\dataset'}=\emptyset$ for all rules.
+   This should happen after a finite number of inserting events: let $I = \{{\tt fixed}_u \mid u\in\rules_{\schema'}\}$ be the relations that keep track of which violations have been fixed, then there is no infinite chain of systems $\migrsys' = \migrsys_0, \migrsys_1,\ldots$ such that $\migrsys_i\xrightarrowdbl[I]{\overrightarrow\rels_{\schema'}}\migrsys_{i+1}$.
    This shows that when users get rid of violations, there is real progress: they can only do this a finite number of times.
@@
 
-   Suppose there is a system in which all violations have been fixed, that is, suppose there is a $\migrsys^C$ such that $\pop{f_u}{\dataset_{\migrsys^C}}=\pop{v_u}{\dataset_{\migrsys^C}}$ and $\migrsys'\xrightarrow[I]{\overrightarrow\rels_{\schema'}} \migrsys^C$.
+   Suppose there is a system in which all violations have been fixed, that is, suppose there is a $\migrsys^C$ such that $\pop{{\tt fixed}_u}{\dataset_{\migrsys^C}}=\pop{{\tt viols}_u}{\dataset_{\migrsys^C}}$ and $\migrsys'\xrightarrow[I]{\overrightarrow\rels_{\schema'}} \migrsys^C$.
    In that case, after any steps done by users (both those with and without the migration helper role), we can reach it, that is: for any $\migrsys_i$, $\migrsys'\xrightarrow[I]{\overrightarrow\rels_{\schema'}} \migrsys_i$ implies $\migrsys_i\xrightarrow[I]{\overrightarrow\rels_{\schema'}} \migrsys^C$.
@@
    For this last question, we need to add a condition:
-   There needs to be an information system that has the desired schema $\schema'$.
+   There needs to be a system that has the desired schema $\schema'$.
    If rules in $\schema'$ are such that they cannot all be satisfied, regardless of the dataset, no such system exists and we have no hope of reaching it.
    We assume that the rules of $\schema'$ are intended to prevent data pollution, such that a dataset $\dataset'$ describing the `real world' would be one that satisfies them.
-   So let $\pair{\dataset'}{\schema'}$ be an information system, then we can get to it through our migration system.
-   To show this, we construct a migration system $\migrsys^C$ satisfying $\pop{f_u}{\dataset_{\migrsys^C}}=\pop{v_u}{\dataset_{\migrsys^C}}$ and $\migrsys'\xrightarrow[I]{\overrightarrow\rels_{\schema'}} \migrsys^C$.
+   So let $\pair{\dataset'}{\schema'}$ be a system, then we can get to it through our migration system.
+   To show this, we construct a migration system $\migrsys^C$ satisfying $\pop{{\tt fixed}_u}{\dataset_{\migrsys^C}}=\pop{{\tt viols}_u}{\dataset_{\migrsys^C}}$ and $\migrsys'\xrightarrow[I]{\overrightarrow\rels_{\schema'}} \migrsys^C$.
    We can do so by simply stating its triples, since the schema is determined by $\migrsys'$:
@@
    \cup \{\triple{a}{\overrightarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_{\dataset'}\}\\
-   &\cup \{\triple{a}{v_u}{b} \mid \pair{a}{b}\in \viol{u}{\dataset}, u\in \rules_{\schema'} \}
-   \cup \{\triple{a}{f_u}{b} \mid \pair{a}{b}\in \viol{u}{\dataset}, u\in \rules_{\schema'} \}
+   &\cup \{\triple{a}{{\tt viols}_u}{b} \mid \pair{a}{b}\in \viol{u}{\dataset}, u\in \rules_{\schema'} \}
+   \cup \{\triple{a}{{\tt fixed}_u}{b} \mid \pair{a}{b}\in \viol{u}{\dataset}, u\in \rules_{\schema'} \}
    \end{aligned}
@@
 
-%   Let us proceed to specify a desired system, $\infsys'$, to illustrate a (toy) migration.
-%   The example in section~\ref{sct:Example existing IS}, called $\infsys$, serves as the existing system.
-%   Its population reflects the state of the system just before the migration.
-%
-%   Information system $\infsys'$, the desired system, is specified by:
-%\begin{verbatim}
-%   RELATION takes[Student*Course]
-%   RELATION isPartOf[Module*Course] [UNI]
-%   RELATION isEnrolledFor[Student*Module]
-%   RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
-%   
-%   RELATION course[ExamReg*Course] [UNI] =
-%      [ ("ER1", "Management")
-%      ; ("ER2", "Business IT")
-%      ; ("ER3", "Business IT")
-%      ]
-%   RELATION student[ExamReg*Student] [UNI] =
-%      [ ("ER1", "Peter")
-%      ; ("ER2", "Susan")
-%      ]
-%   RULE ExamRule1: student~;course |- takes
-%   RULE ExamRule2: student~;course;isPartOf~ |- isEnrolledFor
-%\end{verbatim}
-%   In this example, the relations and the rule of the existing system remain part of the desired system%
-%\footnote{Removal or changing of relations is allowed as well.}
-%   because the developer has copied these items into the script of the desired system.
-%   However, she has imposed the restriction \verb-UNI- on the relation {\tt isPartOf}, stating that this relation is univalent.
-%   This means that every \verb-Module- is part of at most one \verb-Course-.
-%   Besides, two new relations and two rules have appeared.
-%
-%   The desired system, $\infsys'$, contains exam registrations (concept \verb-ExamReg-), which is new.
-%   In an exam registration, a student registers for the examination of a course.
-%   $\infsys'$ contains two extra rules: \verb-ExamRule1- and \verb-ExamRule2-.
-%   The first one says that an exam registration requires that a student actually takes the course.
-%   In logic, the constraint $p_{\tt ExamRule1}$ is written as:
-%\[\begin{array}{l}\forall s\in\text{\tt Student}, e\in\text{\tt ExamReg}, c\in\text{\tt Course}:\\
-%   e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \rightarrow\ s\ \text{\tt takes}\ c\\
-%\end{array}\]
-%   Another requirement, \verb-ExamRule2-, is that the student is enrolled for every module that is part of the course.
-%   Its constraint, $p_{\tt ExamRule2}$, is written in logic as:
-%\[\begin{array}{l}\forall s\in\text{\tt Student}, e\in\text{\tt ExamReg}, m\in\text{\tt Module}, c\in\text{\tt Course}:\\
-%   e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\ \rightarrow\ s\ \text{\tt isEnrolledFor}\ m
-%\end{array}\]
-%   Both rules are invariants.
-%   So, a student can register for an exam only if she is taking the course for which she registers and if she is enrolled for every module of that course.
-%   % Ampersand also derives the following violation sets:
-%% \[\begin{array}{l}
-%   % \viol{\text{\tt ExamRule1}}{\dataset}\\
-%   % \hspace{1cm}=\{\pair{s}{c}\mid\exists e:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\neg(s\ \text{\tt takes}\ c)\}\\
-%   % \viol{\text{\tt ExamRule2}}{\dataset}\\
-%   % \hspace{1cm}=\{\pair{s}{m}\mid\exists e,c:\ e\ \text{\tt student}\ s\ \wedge\ \ e\ \text{\tt course}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\ \wedge\neg(s\ \text{\tt isEnrolledFor}\ m)\}
-%% \end{array}\]
-%
-%\subsection{Migration script}
-%   The generated migration script, $\migrsys$, consists of the schema of the existing system $\infsys$,
-%   the migration system, and the desired system $\infsys'$.
-%   The migration system copies the data of the existing system to the desired system by the following rules:
-%\begin{verbatim}
-%-- The existing system (everything is prefixed with "old.")
-%   RELATION old.takes[Student*Course]
-%   RELATION old.isPartOf[Module*Course]
-%   RELATION old.isEnrolledFor[Student*Module]
-%
-%   RULE old.EnrollRule:
-%      old.isEnrolledFor |- old.takes;old.isPartOf~
-%      
-%-- Migration system
-%   RELATION copied_takes[Student*Course]
-%   ENFORCE new.takes >: old.takes - copied_takes
-%   ENFORCE copied_takes >: new.takes
-%
-%   RELATION copied_isPartOf[Module*Course]
-%   ENFORCE new.isPartOf >: old.isPartOf - copied_isPartOf
-%   ENFORCE copied_isPartOf >: new.isPartOf
-%
-%   RELATION copied_isEnrolledFor[Student*Module]
-%   ENFORCE new.isEnrolledFor >:
-%           old.isEnrolledFor - copied_isEnrolledFor
-%   ENFORCE copied_isEnrolledFor >: new.isEnrolledFor
-%
-%-- The desired system (everything is prefixed with "new.")
-%   RELATION new.takes[Student*Course]
-%   RELATION new.isPartOf[Module*Course] [UNI]
-%   RELATION new.isEnrolledFor[Student*Module]
-%   RULE new.EnrollRule:
-%      new.isEnrolledFor |- new.takes;new.isPartOf~
-%      
-%   RELATION new.course[ExamReg*Course] [UNI] =
-%      [ ("ER1", "Management")
-%      ; ("ER2", "Business IT")
-%      ; ("ER3", "Business IT")
-%      ]
-%   RELATION new.student[ExamReg*Student] [UNI] =
-%      [ ("ER1", "Peter")
-%      ; ("ER2", "Susan")
-%      ]
-%
-%   RULE new.ExamRule1 :
-%      new.student~;new.course |- new.takes
-%   RULE new.ExamRule2 :
-%      new.student~;new.course;new.isPartOf~ |- new.isEnrolledFor
-%\end{verbatim}
-%
-%\subsection{Preparing for the Moment of Transition}
-%   Just before starting the migration,
-%   the data in the existing system typically differs from the data at its inception (section~\ref{sct:Example existing IS}).
-%   Let us assume the actual population at the start of the migration is:
-%\begin{verbatim}
-%   POPULATION takes[Student*Course] CONTAINS
-%      [ ("Peter", "Management")
-%      ; ("Susan", "Business IT")
-%      ; ("John", "Business IT")
-%      ]
-%   POPULATION isPartOf[Module*Course] CONTAINS
-%      [ ("Finance", "Management")
-%      ; ("Business Rules", "Business IT")
-%      ; ("Business Analytics", "Business IT")
-%      ; ("IT-Governance", "Management")
-%      ; ("IT-Governance", "Business IT")
-%      ]
-%   POPULATION isEnrolledFor [Student*Module] CONTAINS
-%      [ ("Susan", "Business Analytics")
-%      ; ("Susan", "IT-Governance")
-%      ; ("Susan", "Business Rules")
-%      ]
-%\end{verbatim}
-%
-%   Upon initialization, the migration system evaluates all enforce rules in an attempt to satisfy all invariants.
-%   In this example, it leaves two violations behind that it cannot resolve by enforce rules.
-%   The pair {\tt ("Peter", "Management")} is the first one; it violates rule {\tt new.ExamRule2}.
-%   This is because Peter has registered for an examination of the course Management,
-%   which contains the module Finance for which Peter is not enrolled.
-%   The other is {\tt "IT-Governance"}, which violates the univalence of relation {\tt new.isPartOf}.
-%   Since rule {\tt new.ExamRule2} and the univalence of {\tt new.isPartOf} are both defined as invariants,
-%   the desired system cannot be taken into production until these violations are resolved.
-%   Until that moment, the migration system remains operational to allow a user to resolve this violation. 
-%   The moment of cleanup is marked by the event that all invariants of the Migration system and the desired system are satisfied.
-%   At that moment, the migration system is no longer needed and can be removed.
-
-%\subsubsection{Strategies for restoring invariants}
-%
-%We allow ExamRule2 to be broken during the transition to the desired system, and expect people with the Administrator role to deal with them.
-%
-%However, the violations that arise when taking the (non-disjoint) union of the two scripts include violations of rules that the system should maintain.
-%This means that if we cannot generate software based on the disjoint union as it is.
-%To resolve this, the simplest solution is to change the roles assigned to the rules that are violated in the disjoint union:
-%\begin{verbatim}
-%   RELATION isPartOf[Module*Course]
-%   RULE isPartOfUNI: isPartOf;isPartOf~ |- I
-%   ROLE migrationHelper MAINTAINS isPartOfUNI
-%\end{verbatim}
-%
-%Naturally, this solution requires effort from people with the role migrationHelper.
-%This in itself is not an issue: problems need to be solved one way or another.
-%However, a problem that may occur is that regular users of the system, who are used to the existing system, might be adding data to the system that causes violations faster than they can be solved.
-%To avoid this problem, we restrict the behavior of the system such that incremental progress will lead to eventually reaching the goal.
-%
-%Our solution is to state that the only violations that may occur in the relation \verb=isPartOfUNI= are those that were present when the migration started:
-%\begin{verbatim}
-%   RELATION isPartOf[Module*Course]
-%   RULE isPartOfUNI: isPartOf;isPartOf~ |- I
-%   ROLE migrationHelper MAINTAINS isPartOfUNI
-%   
-%   RELATION isPartOfUNI_violations[Module*Module]
-%     = [("IT-Governance", "IT-Governance")]
-%   RULE isPartOfUNI_progress: -(isPartOf;isPartOf~) /\ I
-%         |- isPartOfUNI_violations
-%   SYSTEM MAINTAINS isPartOfUNI_progress
-%   
-%   ENFORCE isPartOfUNI_violations :< -(isPartOf;isPartOf~) /\ I  
-%\end{verbatim}
-%
-%In this solution, the first rule is the same as with the simple solution.
-%The second rule prevents any new violation from occurring.
-%Existing violations are recorded in the \verb=isPartOfUNI_violations= relation, and the rule \verb=isPartOfUNI_progress= prevents the set of violations to \verb=isPartOfUNI= from growing beyond this.
-%The \verb=ENFORCE= statement at the end of this code snippet states that the relation recording existing violations should be shrunk whenever violations are solved.
-%This way, violations cannot re-occur.
-%As a whole, this means that \verb=migrationHelper= has a finite task.
-
-% SJC: TODO voor Stef. Misschien is het de moeite waard om te zeggen hoe isPartOfUNI_violations initieel gevuld kan worden.
-% (Dit werkt precies als het initieel vullen van andere relaties, zoals in de vorige sectie).
-
-%As a second solution, the data in the designed system does not have any violations.
-%Consequentially, the cause of the violations comes from the triples in the existing system.
-%Rather than taking the union of all triples, the disjoint union is taken instead;
-%\begin{verbatim}
-%   RELATION isPartOf_old[Module*Course] [UNI]=
-%      [ ("Finance", "Management")
-%      ; ("Business Rules", "Business IT")
-%      ; ("Business Analytics", "Business IT")
-%      ; ("IT-Governance", "Management")
-%      ]
-%   RELATION isPartOf[Module*Course] [UNI]=
-%      [ ("IT-Governance", "Business IT") ]
-%   RELATION isPartOf_ignored[Module*Course]
-%   
-%   RULE isPartOf_isUnion:
-%         isPartOf_old |- isPartOf \/ isPartOf_ignored
-%   ROLE migrationHelper MAINTAINS isPartOf_isUnion
-%   
-%   ENFORCE isPartOf_ignored :> isPartOf
-%\end{verbatim}
-%
-%\begin{verbatim}
-%   RELATION isPartOf_old[Module*Course] [UNI]=
-%      [ ("Finance", "Management")
-%      ; ("Business Rules", "Business IT")
-%      ; ("Business Analytics", "Business IT")
-%      ; ("IT-Governance", "Management")
-%      ]
-%   RELATION isPartOf[Module*Course] [UNI]=
-%      [ ("IT-Governance", "Business IT") ]
-%   RELATION isPartOf_ignored[Module*Course]
-%
-%   ENFORCE isPartOf :> isPartOf_old - isPartOf_ignored
-%   ENFORCE isPartOf_ignored :> isPartOf
-%\end{verbatim}
-%In this second solution, we immediately move to a system in which the \verb=UNI= rule is maintained by the system.
-%The work for the \verb=migrationHelper= is again finite: the pairs in \verb=isPartOf_old= need to be copied over or explicitly ignored.
-%Once a pair is in \verb=isPartOf=, the pair is copied over to \verb=isPartOf_ignored= by the \verb=ENFORCE= statement, such that regular users can delete it without adding work to anyone with the \verb=migrationHelper= role.
-%Of course, since this solution starts with fewer pairs in the \verb=isPartOf= relation, other rules could be violated as a result of choosing this solution.
-%Indeed, ExamRule2 initially has more violations in this scenario.
-%
-%\begin{definition}[migration data set]
-%   \begin{eqnarray}
-%      &&\begin{array}[t]{@{}l}
-%         \{ {\tt ENFORCE\ }\declare{(nm',1)}{(A',1)}{(B',1)}\\\quad{\tt\ :=\ }\ident{(A',1)};\declare{(nm,0)}{(A,0)}{(B,0)};\ident{(B',1)}\\
-%            \mid\ \begin{array}[t]{@{}l}
-%               \declare{nm}{A}{B}\in\rels\wedge\declare{nm'}{A'}{B'}\in\rels'\wedge\id{nm}=\id{nm'}\ \wedge\\
-%               \{\pair{A}{A'},\pair{B}{B'}\}\subseteq\isa\cup\isa'\cup\flip{\isa}\cup\flip{\isa'}\}
-%               \end{array}
-%           \end{array}\\
-%           &\cup&\begin{array}[t]{@{}l}
-%            \{ {\tt CLASSIFY\ }(C',1){\tt\ IS\ }(C,0)\mid\ C\in\concepts,C'\in\concepts',C=C'\}\\
-%           \end{array}
-%   \end{eqnarray}
-%\end{definition}
-%
-%The ${\tt CLASSIFY\ }(C',1){\tt\ IS\ }(C,0)$ statement adds two pairs to the $\isa$ relation:
-%$\pair{(C',1)}{(C,0)}$ and $\pair{(C,0)}{(C',1)}$.
-%Note that if $\dataset$ is a data set, and $\dataset'$ is that data set with pairs added%
-%\footnote{TODO: het woord `add' is informeel, dit kan beter} to the $\isa$ relation, then the result is again a data set.
-%The ${\tt ENFORCE\ }r{\tt\ :=\ }e$ statements add triples $(x,r,y)$ for all $(x,y)\in e$.
-%The terms $t$ in these statements are such that $x\ (\instance \compose \kleenestar{\flip{\isa}})\ (A',1)$ and $y \ (\instance \compose \kleenestar{\flip{\isa}})\ (B',1)$, thus preserving the property that the result is a data set.
-%
-%\begin{definition}[]
-%   \begin{eqnarray}
-%      \infsys\sqcup\infsys'&=&\triple{\roles'}{\rules'}{\maintain'}{\dataset\sqcup\dataset'}
-%   \end{eqnarray}
-%\end{definition}
-
-%\subsection{Changes}
-%% The purpose of this section is to explain why a data set is structured the way it is.
-%   In the migration of a data set we deal with changes to the elements that are not data:
-%   $\instance$, $\concepts$, and $\rels$.
-%   Such changes have further reaching consequences, however.
-%   Changes to $\instance$, $\concepts$, $\rels$ change the data structure in a data set.
-%   We define a \define{migration of a data set} $\la\atoms,\concepts,\instance,\rels,\dataset\ra$ as a change in which one or more of $\instance$, $\concepts$, or $\rels$ change.
-%   To satisfy the definition of data sets, $\dataset$ must change too,
-%   but the migration should try to preserve the maximal amount of data.
-
-%\section{Validation}
-%   We claim that the information system does not need to be paused in order for the migration to happen successfully.
-%   By this, we mean that computations that take a considerable amount of time, as well as human actions, can be performed incrementally while the system is in use.
-%   In order for us to prove this claim, we model the behavior of our information system as a series of systems $\migrsys_0,\migrsys_1,\ldots,\migrsys_n$ with $\migrsys = \migrsys_0$, where only the data set of the system changes.
-%   We consider two types of updates to the data set.
-%   
-%   The first type of update helps the migration to go forward: we'll write $\rightsquigarrow$ for those.
-%   The second type of update is that which other users of the system do (possibly none), indicated as $\dashrightarrow$.
-%   The two updates alternate: $\migrsys_0 \dashrightarrow \migrsys_1 \rightsquigarrow \migrsys_2 \dashrightarrow \ldots \rightsquigarrow \migrsys_{2i} \dashrightarrow \migrsys_{2i+1}\rightsquigarrow\migrsys_{2i+2}\dashrightarrow\ldots\migrsys_n$.
-%   
-%   We validate our claim that the information system will reach the MoC by arguing about two situations:
-%   The system between deployment and the MoT can reach the MoT by firing the ENFORCE rules, and the system between the MoT and the MoC can reach the MoC after a finite number of actions by the migration engineer.
-%   More concretely, this means the following:
-%   
-%When the default migration system $\migrsys$ is deployed, the system will always be in a state such that continually running any violated ENFORCE rules (in any order) leads to the MoT.
-%   This means that $A\rightsquigarrow B$ is such that the data set of $B$ is that of $A$ with an update applied according to one or more violated enforce rules. % TODO: define 'violated'
-%   If there is no such $B$ because no enforce rules are violated, we are at the moment of transition.
-%   We say that the system $A$ is in a state such that the MoT can be reached if there are no infinite $\rightsquigarrow$-paths from $A$.
-%   The constraint on $A\dashrightarrow B$ is that only the old system can change at this point.
-%   Note that this does not prove that the moment of transition to never be reached, as more enforce rules may get violated at every $\dashrightarrow$ step.
-%   We expect this not to be a problem in practice since $\dashrightarrow$ transitions are triggered by users while $\rightsquigarrow$ is triggered automatically through ENFORCE rules: the ENFORCE rules should outperform the $\dashrightarrow$ actions by a big margin.
-%
-%After the moment of transition, there is a finite upper-bound on the amount of work the migration engineer has to do:
-%   $A\rightsquigarrow B$ is such that the data set of $B$ is that of $A$ with one of the violations shown to a migration engineer resolved (including the firing of corresponding ENFORCE statements), and $\dashrightarrow$ arbitrary.
-%   In both $A\rightsquigarrow B$ and $A\dashrightarrow B$ we assume that the relations that were specifically created for the bookkeeping (suffixed by \verb=_progress= and \verb=_violations=) can only be updated through the firing of the ENFORCE statements.
-%   These assumptions make the total number of pairs in \verb=_violations= relations decline in every $\rightsquigarrow$-step, proving that the MoC will be reached after a finite number of $\rightsquigarrow$ steps (assuming that the set of violations is finite).
-%   Moreover, we show that the desired system with the data set taken from the migration system (after dropping the old system and the bookkeeping) is free of violations when all \verb=_violations= relations are empty.
-%
-%
-%\section{Software architecture}
-%   The theory presented above roughly corresponds to the back-end of a migration system.
-%   This section describes its software architecture.
-%   For this research, we have used the Ampersand compiler to generate the system(s).
-%   Figure~\ref{fig:initial} shows the starting point of the migration.
-%\begin{figure}
-%   \centering
-%   \includegraphics[scale=0.1]{figures/Before migration.png}
-%   \caption{Initial state of the migration}
-%   \label{fig:initial}
-%\end{figure}
-%   For the purpose of data migration, the existing system must make its schema available through its API.
-%   For this purpose, the back-end offers a service ``retrieve schema'', which provides the schema as Ampersand source-code.
-%   As a result, the migration engineer can rely on the schema of any generated application to be accurate.
-%   For this reason, we can safely assume that the schema this service provides has not been changed since it was deployed.
-%   Its presence allows the generator to build various front-ends and migration scripts during the lifetime of the existing system%
-%
-%   After the migration engineer has produced and tested a migration script, the Ampersand compiler produces a migration system.
-%   It includes the schemas of both the existing system and the desired system, besides having a schema of its own.
-%   This is implemented by Ampersand's INCLUDE mechanism, which boils down to taking the disjoint union of the schemas.
-%\begin{figure}
-%   \centering
-%   \includegraphics[scale=0.1]{figures/During-migration.png}
-%   \caption{During the migration}
-%   \label{fig:during}
-%\end{figure}
-%   Deployment of the migration system (Figure~\ref{fig:during}) leaves the existing system intact.
-%   It enhances the system with the data set needed by the migration system, which includes the data set needed by the desired system.
-%   So, users will notice no difference in the existing system when the migration system is deployed.
-%
-%   When all rules in the migration script have been satisfied, the migration system is ready to be deployed.
-%   At this point, the ingress switches traffic from the existing system to the migration system.
-%   Until this moment, users have been updating data in the existing system, which is continuously being migrated to the desired system.
-%   After the ingress switch is made, users have the new functionality at their disposal and make their changes in the desired system.
-%   If for some reason, there is remaining migration work, the migration system will execute this.
-%   Since there is no more input in the existing system, the migration system will eventually be consistent.
-%   At this point, the existing system can be removed, which leaves the migration system in place.
-%
-%
-%   The main purpose of the migration script is to define the migration rules that are used to migrate data from the existing system to the desired system.
-%   During this process, users continu to use the existing system.
-%
-%\begin{figure}
-%   \centering
-%   \includegraphics[scale=0.1]{figures/After migration.png}
-%   \caption{After the migration}
-%   \label{fig:after}
-%\end{figure}
-
 \section{Conclusions}
 
-In this paper, we describe the data migration as going from an existing information system to a desired one, where the schema changes.
-As Ampersand generates information systems, creating a new information system can be a small task, allowing for incremental deployment of new features.
-We describe the parts of an information system that have an effect on data pollution.
+   In this paper, we describe the data migration as going from an existing system to a desired one, where the schema changes.
+   As Ampersand generates information systems, creating a new system can be a small task, allowing for incremental deployment of new features.
+   We describe the parts of a system that have an effect on data pollution.
    We assume that the existing system does not violate any constraints of its schema, but address other forms of data pollution:
@@
    In practice, such changes need to be achieved through interfaces.
-Most Ampersand systems indeed allow the users of the information system to edit the dataset quite freely through the interfaces.
+   Most Ampersand systems indeed allow the users of the system to edit the dataset quite freely through the interfaces.
    However, some interfaces may require certain constraints to be satisfied, which means that interfaces of the desired system might break.

### 2023-07-12 2023 Disambiguation/Disambiguation.tex

@@
 The intended term is Term~\ref{eqn:disambGood} here.
-Its meaning is different from that of Term~\ref{eqn:disambBad}, since the term $\epsilon_{t_3};\typecomp{1_{\mathrm{Event}}};\epsilon_{t_3}$ can contain pairs in $\mathrm{Event}$ that $\epsilon_{t_0};\typecomp{1_{\mathrm{Task}}};\epsilon_{t_0}$ does not.
+Its meaning is different from that of Term~\ref{eqn:disambBad}, since the term $\epsilon_{t_3};\overline{1_{\mathrm{Event}}};\epsilon_{t_3}$ can contain pairs in $\mathrm{Event}$ that $\epsilon_{t_0};\overline{1_{\mathrm{Task}}};\epsilon_{t_0}$ does not.
 We could argue that the user must specify more type information here, since the meanings for these terms differ.
-However, the term can be interpreted as one that does not use sub-typing in any meaningful way.
+However, the term can be interpreted as one that does not use sub-typing.
 Since the purpose of a disambiguation system is to reduce verbosity, we should prefer the intended term over Term~\ref{eqn:disambBad}.
@@
 
-What if the user writes $\typecomp{\typetyped{\typeident}{\mathrm{Event*Event}}} \typeunion \typecomp{\left(\mathrm{before} \typeinter \typeconv{\mathrm{before}}\right)}$, a term forces the use of sub-typing?
+What if the user writes $\typecomp{\typetyped{\typeident}{\mathrm{Event*Event}}} \typeunion \typecomp{\left(\mathrm{before} \typeinter \typeconv{\mathrm{before}}\right)}$, a term that forces the use of sub-typing?
 Now possible disambiguated terms include:
@@
 In previous publications on the Ampersand type system, where sub-typing did not play a role, Term~\ref{eqn:disambOld} would be chosen over Term~\ref{eqn:disambOption}.
-The rational here is that types are computed bottom up.
+The rationale here is that types are computed bottom up.
 However, we have observed that sub-typing around complements can be error-prone, and the Term~\ref{eqn:disambOption} might well have been intended.
@@
 Aside from type errors, we can use $p$ to find ambiguous terms:
-If there is a sub-term $\Xi$ of $e$ that can be replaced by both $x\in \Xi$ and $y\in Xi$ for some $x \neq y$, then $p(e')$ and $p(e'')$ are nonempty for $e'$ and $e''$ equal to $e$ under the substitution of $\Xi$ for $\{x\}$ and $\{y\}$ respectively.
+If there is a sub-term $\Xi$ of $e$ that can be replaced by both $x\in \Xi$ and $y\in \Xi$ for some $x \neq y$, then $p(e')$ and $p(e'')$ are nonempty for $e'$ and $e''$ equal to $e$ under the substitution of $\Xi$ for $\{x\}$ and $\{y\}$ respectively.
 Furthermore, we can use $p$ to find out whether a negation can have multiple types.
@@
 This operation fulfils our requirement that we should prefer some terms over others.
-Let $f : \powerset{\mathcal{R}}\to\powerset{\mathcal{R}}$ be an idempotent function (a function such that $f(f(x))=f(x)$ for all $x$) that returns a subset of its argument, indicating which elements would be preferred.
-In particular, choosing $f(\Xi) = \left\{\epsilon_{A*B} \mid \epsilon_{A*B}\in \Xi \wedge A=B\right\}$ would have us pick $\epsilon_{A*B}$ relations with $A=B$.
+Let $f : \powerset{(\mathcal{R})}\to\powerset{(\mathcal{R})}$ be an idempotent function (a function such that $f(f(x))=f(x)$ for all $x$) that returns a subset of its argument, indicating which elements would be preferred.
+In particular, choosing $f(\Xi) = \left\{\epsilon_{A*B} \mid \epsilon_{A*B}\in \Xi \wedge A=B\right\}$ would have us prefer $\epsilon_{A*B}$ relations with $A=B$.
 Given such a function $f$, we say that $e'$ is more preferable to $e$ if $p(e')\neq \emptyset$ and $e'$ can be obtained from $e$ by substituting some $\Xi$ terms for $f(\Xi)$.
-We say that $e$ is {\emph maximally preferable} if $p(e)\neq \emptyset$ and for every occurrence of a $\Xi$-term in $e$, substituting it for $f(\Xi)$ would either leave $e$ unchanged (as $\Xi=f(\Xi)$ or result in a term $e'$ such that $p(e')=\emptyset$.
+We say that $e$ is {\emph maximally preferable} if $p(e)\neq \emptyset$ and for every occurrence of a $\Xi$-term in $e$, substituting it for $f(\Xi)$ would either leave $e$ unchanged (as $\Xi=f(\Xi)$) or result in a term $e'$ such that $p(e')=\emptyset$.
 Terms that are more preferable to $e$ form a lattice\footnote{Proof: `more preferable' is the subset relation for the set of $\Xi$-term occurrences that have been changed wrt $e$.}, and we call the least upper bound to all maximally preferable terms the term that is generally preferred to $e$ under $f$.

### 2023-07-13 2022Migration/articleMigrationFACS.tex
Made the definition of Event more readable.

@@
 \begin{definition}[Event]
-   Let $R \subseteq \Rels$ and $I \subseteq \Rels$ be sets of relations, and let $\infsys=\pair{\schema}{\dataset}$ and  $\infsys'=\pair{\schema}{\dataset'}$ be information systems, and:
+   Let $R \subseteq \Rels$ and $I \subseteq \Rels$ be sets of relations,
+   and let $\infsys=\pair{\schema}{\dataset}$ and  $\infsys'=\pair{\schema}{\dataset'}$ be information systems,
+   then $\infsys\xrightarrow[I]{R} \infsys'$ is an event if and only if:
    \begin{align}
-      \triples_{\dataset} - (\Triple{\Atoms}{(R \cup I)}{\Atoms}) &= \triples_{\dataset'} - (\Triple{\Atoms}{(R \cup I)}{\Atoms})
+      \triple{a}{r}{b}\in\triples_{\dataset}-\triples_{\dataset'}&\Rightarrow\ r\in R
    \label{eqn:eventUnchanged}\\
-      \triples_{\dataset} - (\Triple{\Atoms}{R}{\Atoms}) &\subseteq \triples_{\dataset'} - (\Triple{\Atoms}{R}{\Atoms})
+      \triple{a}{r}{b}\in\triples_{\dataset'}-\triples_{\dataset}&\Rightarrow\ r\in(R \cup I)
    \label{eqn:eventInsert}
    \end{align}
-%   Then we say that $\pair{\pair{\schema}{\dataset}}{\infsys'}$ is an event with scope $R$ inserting in $I$.
-   Then we say that $\infsys\xrightarrow[I]{R} \infsys'$ is an event
-   that brings the system $\infsys$ to $\infsys'$ inserting pairs in relations from $I$ and updating (i.e. deleting and inserting) pairs in $R$.
-%   We use the notation $\pair{\schema}{\dataset} \xrightarrow[I]{R} \infsys'$ to indicate this.
+   % was:
+   % \begin{align}
+   %    \triples_{\dataset} - (\Triple{\Atoms}{(R \cup I)}{\Atoms}) &= \triples_{\dataset'} - (\Triple{\Atoms}{(R \cup I)}{\Atoms})
+   % \label{eqn:eventUnchanged}\\
+   %    \triples_{\dataset} - (\Triple{\Atoms}{R}{\Atoms}) &\subseteq \triples_{\dataset'} - (\Triple{\Atoms}{R}{\Atoms})
+   % \label{eqn:eventInsert}
+   % \end{align}
+   We say that event $\infsys\xrightarrow[I]{R} \infsys'$ brings the system $\infsys$ to $\infsys'$ inserting pairs in relations from $I$ and deleting pairs in relations from $R$.
 \end{definition}
 
-   equation~\ref{eqn:eventUnchanged} states that the triples of those for relations in $I$ or $R$ are the only ones to have changed, and equation~\ref{eqn:eventInsert} states that of the ones that are only in $I$, no triples were removed.
+   Equation~\ref{eqn:eventUnchanged} states that the triples of those for relations in $I$ or $R$ are the only ones to have changed, and equation~\ref{eqn:eventInsert} states that of the ones that are only in $I$, no triples were removed.
    %If $I$ or $R$ is the empty set, we omit it from the arrow, so $\infsys \xrightarrow{R} \infsys'$ is a notation for $\infsys \xrightarrow[\emptyset]{R} \infsys'$.

### 2023-07-13 2022Migration/articleMigrationFACS.tex

@@
 %
-\begin{abstract}  % There are 176 words in the abstract
-Software generators can help to increase the frequency of releases and their reliability.
+\begin{abstract}
+Software generators can help increase the frequency of releases and their reliability.
 They save on time spent on development and fixing human-induced mistakes by compiling a specification into a working information system.
@@
 This paper defines a migration process that preserves the semantics of data by means of rules (constraints).
-The migration process is based assumptions and requirements that aim for a large class of data migrations in practice.
+The migration process is based on assumptions and requirements that aim to capture a large class of data migrations in practice.
 This paper focuses on correctness of the migration. Efficiency of the migration is outside its scope.
@@
    It is this business that depends on the semantics of the data to draw the right conclusions and carry out their tasks.
-   This paper uses semantic constraints to represent this meaning.
+   %This paper uses semantic constraints to represent this meaning.
 
@@
    An Ampersand script contains just enough information to generate a complete system,
-   which means that a classical database schema (i.e.\ data structure plus semantics) can be extracted from the Ampersand script.
+   which means that a classical database schema (i.e.\ data structure plus constraints) can be extracted from the Ampersand script.
 
@@
    The purpose of a data set is to describe the data of a system at one point in time. 
-   Before defining data sets, let us first define the constituent notions of atom, concept, specialization, relation, and triple.
+   Before defining data sets, let us first define the constituent notions of atom, concept, relation, and triple.
 
@@
    \begin{definition}[Data set]
-   A data set $\dataset$ is a tuple $\pair{\triples}{\instance}$ with $\triples \subseteq {\Triple{\Atoms}{\Rels}{\Atoms}}$ and $\instance \subseteq  {\Pair{\Atoms}{\Concepts}}$ that satisfies:
+   A data set $\dataset$ is a tuple $\pair{\triples}{\instance}$ with finite $\triples \subseteq {\Triple{\Atoms}{\Rels}{\Atoms}}$ and $\instance \subseteq  {\Pair{\Atoms}{\Concepts}}$ that satisfies:
 \begin{eqnarray}
@@
    The notations $\triples_{\dataset}$ and $\instance_{\dataset}$ are used to disambiguate $\triples$ and $\instance$ when necessary.
-   To save writing in the sequel, the notation $a\ r\ b$ means that $\triple{a}{r}{b}\in\triples$.
+   %To save writing in the sequel, the notation $a\ r\ b$ means that $\triple{a}{r}{b}\in\triples$.
 
@@
 %
-   We overload the notation $\popF{\cdot}$ so we can use it on concepts $\popF{C}:\Dataset\rightarrow\powerset{\Atoms}$
+   We overload the notation $\popF{}$ so we can use it on concepts $\popF{C}:\Dataset\rightarrow\powerset{\Atoms}$
    and expressions.
@@
 \label{sct:Schemas}
-   Schemas serve to capture the semantics of an information system~\cite{Spivak2012}.
+   Schemas serve to capture the semantics of a system~\cite{Spivak2012}.
    We consider a schema to represent the static part of a system, defining concepts, relations, and rules.
@@
 
-   We describe a schema $\schema$ via a tuple $\triple{\concepts}{\rels}{\rules}$,
+   We describe a schema $\schema$ via a tuple $\quadruple{\concepts}{\rels}{\rules}{\enforces}$,
    in which $\concepts\subseteq \Concepts$ is a finite set of concepts,
    $\rels\subseteq \Rels$ is a finite set of relations,
-   and $\rules\subseteq \powerset{\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}}$ is a finite set of rules.
+   $\rules\subseteq {\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}}$ is a finite set of rules,
+   and $\enforces\subseteq \Pair{\Rels}{\left(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}\right)}$ is a finite set of enforce rules.
 
@@
    Such constraints are called enforcement rules.
-   We just use one type of enforcement rule\footnote{Ampersand has several types of enforcement rules.} in this work.
+   We use just one type of enforcement rule\footnote{Ampersand has several types of enforcement rules.} in this work.
    Its syntax is as follows:
@@
    \item $\schema=\quadruple{\concepts}{\rels}{\rules}{\enforces}$ is a schema (satisfies equation~\ref{eqn:relationsIntroduceConcepts} and~\ref{eqn:enforceRulesRels}), we write $\concepts_\infsys = \concepts$, $\rels_\infsys = \rels$, $\rules_\infsys=\rules$, and $\enforces_\infsys=\enforces$;
-   \item all rules and enforcement rules are satisfied:
+   \item All rules and enforcement rules are satisfied:
    \begin{align}
@@
    \end{align}
-   \item triples in the data set have their relation mentioned in the schema:
+   \item Triples in the data set have their relation mentioned in the schema:
    \begin{eqnarray}
@@
    In this paper, it suffices to define an event as a pair of systems for which the schema stays constant.
-   Events are categorized by what part of the data set changes.
+   Events are categorized by the part of the data set that changes.
 
@@
 
-   Equation~\ref{eqn:eventUnchanged} states that the triples of those for relations in $I$ or $R$ are the only ones to have changed, and equation~\ref{eqn:eventInsert} states that of the ones that are only in $I$, no triples were removed.
+   Equation~\ref{eqn:eventUnchanged} and~\ref{eqn:eventInsert} state that the triples of those for relations in $I$ or $R$ are the only ones that can be changed, and that only triples in $R$ can be removed.
    %If $I$ or $R$ is the empty set, we omit it from the arrow, so $\infsys \xrightarrow{R} \infsys'$ is a notation for $\infsys \xrightarrow[\emptyset]{R} \infsys'$.
@@
 \[\begin{array}{l}
-   \pair{\tt "Finance"}{\tt Module}\\
-   \pair{\tt "Business Rules"}{\tt Module}\\
-   \pair{\tt "Business Analytics"}{\tt Module}\\
-   \pair{\tt "IT-Governance"}{\tt Module}\\
-   \pair{\tt "Management"}{\tt Course}\\
-   \pair{\tt "Business IT"}{\tt Course}\\
-   \pair{\tt "Peter"}{\tt Student}\\
-   \pair{\tt "Susan"}{\tt Student}\\
-   \pair{\tt "John"}{\tt Student}
+   \pair{\text{\tt "Finance"}}{\tt Module}\\
+   \pair{\text{\tt "Business Rules"}}{\tt Module}\\
+   \pair{\text{\tt "Business Analytics"}}{\tt Module}\\
+   \pair{\text{\tt "IT-Governance"}}{\tt Module}\\
+   \pair{\text{\tt "Management"}}{\tt Course}\\
+   \pair{\text{\tt "Business IT"}}{\tt Course}\\
+   \pair{\text{\tt "Peter"}}{\tt Student}\\
+   \pair{\text{\tt "Susan"}}{\tt Student}\\
+   \pair{\text{\tt "John"}}{\tt Student}
 \end{array}\]
@@
    As violations are fixed, rule~\ref{fill fixed} records this by adding to {\tt fixed\_UNI}.
-   This way, previously fixed errors cannot reoccur.
+   This way, previously fixed errors cannot recur.
    Once {\tt viols\_UNI} is equal to {\tt fixed\_UNI}, the \verb=UNI= requirement holds and the system is ready for the MoC in regard to this rule.
@@
    \triples_{\migrsys^C} &= \triples_\migrsys 
-   \cup \{\triple{a}{\overrightarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_{\dataset'}\}\\
-   &\cup \{\triple{a}{{\tt viols}_u}{b} \mid \pair{a}{b}\in \viol{u}{\dataset}, u\in \rules_{\schema'} \}
-   \cup \{\triple{a}{{\tt fixed}_u}{b} \mid \pair{a}{b}\in \viol{u}{\dataset}, u\in \rules_{\schema'} \}
+   \cup  \{\triple{a}{{\tt viols}_u}{b} \mid \pair{a}{b}\in \viol{u}{\dataset}, u\in \rules_{\schema'} \} \\
+   &\cup \{\triple{a}{{\tt fixed}_u}{b} \mid \pair{a}{b}\in \viol{u}{\dataset}, u\in \rules_{\schema'} \}
+   \cup \{\triple{a}{\overrightarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_{\dataset'}\}   
    \end{aligned}

### 2023-07-13 2022Migration/articleMigrationFACS.tex
enforce -> enforcement

@@
    $\rules\subseteq {\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}}$ is a finite set of rules,
-   and $\enforces\subseteq \Pair{\Rels}{\left(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}\right)}$ is a finite set of enforce rules.
+   and $\enforces\subseteq \Pair{\Rels}{\left(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}\right)}$ is a finite set of enforcement rules.
 

### 2023-07-13 2022Migration/articleMigrationFACS.tex
Fixed 4.2 and 4.3

@@
    &\text{\tt ENFORCE viols\_UNI >: old.isPartOf\textasciitilde;old.isPartOf - I}\label{fill viols}\\
-   &\text{\tt ENFORCE fixed\_UNI >: viols\_UNI - (old.isPartOf\textasciitilde;old.isPartOf - I)}\label{fill fixed}\\
+   &\text{\tt ENFORCE fixed\_UNI >: viols\_UNI - (new.isPartOf\textasciitilde;new.isPartOf - I)}\label{fill fixed}\\
    &\begin{aligned}
@@
    \begin{align}
-   \widetilde{\dataset'} &= \pair{\{\triple{x}{\widetilde{r}}{y}\mid \triple{x}{r}{y} \in \triples_{\dataset'}\}}{\instance_{\dataset'}}\\
-   \overrightarrow{\dataset'} &= \pair{\{\triple{a}{\overrightarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_\dataset\}}{\instance_{\dataset'}} \\
+   \overleftarrow{\dataset'} &= \pair{\{\triple{x}{{r}}{y}\mid \triple{x}{\overleftarrow{r}}{y} \in \triples_{\dataset'}\}}{\instance_{\dataset'}}\\
+   \overrightarrow{\dataset'} &= \pair{\{\triple{a}{{r}}{b} \mid \triple{a}{\overrightarrow{r}}{b}\in\triples_\dataset\}}{\instance_{\dataset'}} \\
    \rels_\theenumi &= \{{\tt viols}_u\mid u \in \rules_{\schema'}\} \cup \{{\tt fixed}_u\mid u \in \rules_{\schema'}\}\\
-   \rules_\migrsys &= \{\lambda {\dataset'}.~ \viol{u}{\widetilde{\dataset'}} - \pop{{\tt viols}_u - {\tt fixed}_u}{\dataset'} \mid u\in\rules_{\schema'}\}\label{eqn:migrsysrules}\\
-   \enforces_{\theenumi} &=\{{\tt viols}_u \mapsfrom \lambda \dataset'.~ \viol{u}{\widetilde{\dataset'}}\} \cup
+   \rules_\migrsys &= \{\lambda {\dataset'}.~ \viol{u}{\overrightarrow{\dataset'}} - \pop{{\tt viols}_u - {\tt fixed}_u}{\dataset'} \mid u\in\rules_{\schema'}\}\label{eqn:migrsysrules}\\
+   \enforces_{\theenumi} &=\{{\tt viols}_u \mapsfrom \lambda \dataset'.~ \viol{u}{\overleftarrow{\dataset'}}\} \cup
     \{{\tt fixed}_u \mapsfrom \lambda \dataset'.~ {\tt viols}_u - \viol{u}{\overrightarrow{\dataset'}}\}\label{eqn:enforceForRules}

### 2023-07-14 2022Migration/articleMigrationFACS.tex
Added references to tie 4.3 to 4.2.

@@
    \end{align}
-\item We create ENFORCE rules to copy equal relations, corresponding to (\ref{extraRelation}) and (\ref{difference}) in Section~\ref{sec:loading}:
+\item We create ENFORCE rules to copy equal relations, corresponding to equation~\ref{extraRelation} through~\ref{fill copies with new.isPartOf} in Section~\ref{sec:loading}:
    \begin{align}
@@
    \end{align}
-\item For each rule, we generate two helper relations, {\tt viols} and {\tt fixed}. Note that Rule~\ref{warnings} only influences the user interface and therefore is not modeled here.
+\item For each rule, we generate two helper relations, {\tt viols} and {\tt fixed}.
+Note that Rule~\ref{warnings} only influences the user interface and therefore is not modeled here.
+This corresponds to equation~\ref{violRelation} through~\ref{rule uni}.
    \begin{align}

### 2023-07-30 2022Migration/articleMigrationFACS.tex
progress July 30th

@@
 \newcommand{\tgt}[1]{\id{tgt}(#1)}
-\newcommand{\viol}[2]{{#1}(#2)}
-%\newcommand{\violC}[1]{\id{viol}_{#1}}
+\newcommand{\viol}[2]{\violC{#1}(#2)}
+\newcommand{\violC}[1]{\id{viol}_{#1}}
 \newcommand{\sign}[1]{\id{sign}(#1)}
@@
 \label{eqn:pop-concept}\\
-   \pop{A - B}{\dataset}&= \pop{A}{\dataset} - \pop{B}{\dataset}
+   \pop{x-y}{\dataset}&= \pop{x}{\dataset} - \pop{y}{\dataset}
 \label{eqn:pop-expr}
@@
    $\rels\subseteq \Rels$ is a finite set of relations,
-   $\rules\subseteq {\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}}$ is a finite set of rules,
+   $\rules\subseteq \Triple{(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}})}{\Concepts}{\Concepts}$ is a finite set of rules,
    and $\enforces\subseteq \Pair{\Rels}{\left(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}\right)}$ is a finite set of enforcement rules.
@@
    In this paper, variables $u$ and $v$ represent rules.
-   A rule $u$ in the schema is a function such that $\viol{u}{\dataset}$ represents any violations to $u$ in data set $\dataset$.
+   A rule $u$ in the schema is a triple $\triple{\violC{u}}{A}{B}$ such that $\viol{u}{\dataset}$ represents any violations to $u$ in data set $\dataset$.
+   In Ampersand, rules are well-typed, just like relations:
+   Every violation $\pair{a}{b}$ of a rule $\triple{\violC{u}}{A}{B}$ must satisfy $a\instance A$ and $b\instance B$.
+   % For every data set $\dataset$, the set $\viol{u}{\dataset}$ is a subset of $\pop{A}{\dataset}\times\pop{B}{\dataset}$.
    If there are no violations, that is: $\viol{u}{\dataset} = \emptyset{}$, then we say that $u$ is satisfied.
@@
 
-   Adding an enforcement rule to the script of a system specifies a function $u : \Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}$
+   Adding an enforcement rule to the script of a system specifies a function $\violC{u} : \Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}$
    with $\viol{u}{\dataset} = \pop{\id{term} - r}{\dataset}$.
@@
 
-   Formally, enforcement rules are pairs with $u$ a function and $r$ a relation, written $r \mapsfrom u$ or equivalently $r \mapsfrom \lambda \dataset.~ u(\dataset)$.
+   Formally, enforcement rules are pairs with $u$ a function and $r$ a relation, written $r \mapsfrom t$ or equivalently $r \mapsfrom \lambda \dataset.~ t(\dataset)$.
    The set of enforcement rules in a schema is $\enforces$.
@@
 \begin{align}
-   \declare{n}{A}{B}\in\rels&~\Rightarrow~ A\in\concepts\ \wedge\ B\in\concepts
+   \declare{n}{A}{B}\in\rels&~\Rightarrow~ A\in\concepts\,\wedge\, B\in\concepts
    \label{eqn:relationsIntroduceConcepts}\\
-   \pair{u}{r}\in\mathcal\enforces&~\Rightarrow~ r \in \rels\label{eqn:enforceRulesRels}
+   \triple{\violC{u}}{A}{B}\in\rules&~\Rightarrow~A\in\concepts\,\wedge\, B\in\concepts\label{eqn:rule-has-type}\\
+   \triple{\violC{u}}{A}{B}\in\rules&~\Rightarrow~\forall\dataset.\ \viol{u}{\dataset}\subseteq\Pair{\pop{A}{\dataset}}{\pop{B}{\dataset}}\label{eqn:rule-well-typed}\\
+   (\declare{n}{A}{B}\mapsfrom t)\in\mathcal\enforces&~\Rightarrow~ \declare{n}{A}{B}\in \rels\label{eqn:enforcement-has-type}\\
+   (\declare{n}{A}{B}\mapsfrom t)\in\mathcal\enforces&~\Rightarrow~\forall\dataset.\ \pop{t}{\dataset}\subseteq\Pair{\pop{A}{\dataset}}{\pop{B}{\dataset}}\label{eqn:enforcement-well-typed}
 \end{align}
@@
    Requirement~\ref{eqn:relationsIntroduceConcepts} ensures that concepts mentioned in a relation are defined in the schema.
-   The type system of Ampersand~\cite{vdWoude2011} ensures that every system it generates satisfies this requirement.
    This is also known as static typing, which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.
-
-   %The notation $\schema_{\infsys}$ refers to the schema of system $\infsys$.
+   Requirements~\ref{eqn:rule-has-type} and~\ref{eqn:rule-well-typed} ensure that violations in a rule correspond to the type of that rule. 
+   Requirements~\ref{eqn:enforcement-has-type} and~\ref{eqn:enforcement-well-typed} ensure the same for enforcement rules. 
+   The type system of Ampersand~\cite{vdWoude2011} ensures that every system it generates satisfies requirements~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-well-typed}.
    When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, $\rules_{\schema}$, and $\enforces_{\schema}$
@@
    \item $\dataset=\pair{\triples}{\instance}$ is a data set (satisfies equation~\ref{eqn:wellTypedEdge}), we write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$;
-   \item $\schema=\quadruple{\concepts}{\rels}{\rules}{\enforces}$ is a schema (satisfies equation~\ref{eqn:relationsIntroduceConcepts} and~\ref{eqn:enforceRulesRels}), we write $\concepts_\infsys = \concepts$, $\rels_\infsys = \rels$, $\rules_\infsys=\rules$, and $\enforces_\infsys=\enforces$;
+   \item $\schema=\quadruple{\concepts}{\rels}{\rules}{\enforces}$ is a schema (satisfies equation~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-well-typed}), we write $\concepts_\infsys = \concepts$, $\rels_\infsys = \rels$, $\rules_\infsys=\rules$, and $\enforces_\infsys=\enforces$;
    \item All rules and enforcement rules are satisfied:
@@
    \forall u\in\rules&.~\viol{u}{\dataset}=\emptyset\\
-   \forall (r\mapsfrom u)\in\enforces&.~\viol{u}{\dataset} - \pop{r}{\dataset}=\emptyset
+   \forall (r\mapsfrom t)\in\enforces&.~\viol{u}{\dataset} - \pop{r}{\dataset}=\emptyset
    \label{eqn:satisfaction}

### 2023-08-06 2022Migration/articleMigrationFACS.tex
Simplifying the explanation

@@
    In practice, data migrations are done by deploying the desired system and the existing system side-by-side,
-   while transferring data in a controlled fashion.
-   Typically, a migration strategy (e.g. case by case, customer by customer, or in batches) is in place
-   to ensure the preservation of valuable data and the quality of the migration.
-   To fully automate every incremental data migration is unrealistic.
-   Smaller increments stand a higher chance of being automated, but there will always be cases where some human intervention is required.
-   Manual interventions can be necessary to resolve data pollution, new business rules, or to fix known issues in the old system.
-   Consequently, a migration engineer must have the freedom to tailor any automatic approach to migration to suit the particulars.
-
-   Complications that require manual intervention arise mostly as a result of data pollution.
-   We distinguish data pollution in the following categories:
-\begin{itemize}
-   \item data pollution that violates a constraint in the schema.
-   \item data pollution that violates a constraint that is not in the schema.
-   \item data pollution that cannot be captured in constraints on the data set,
-   such as a street address that has become obsolete because someone has moved without notification.
-   \item data pollution that is a consequence of an erroneous constraint in the schema.
-\end{itemize}
-   The first category is about violations of constraints from the schema.
-   Such violations do not occur in our theory because we assume the existing system was generated by Ampersand,
-   so the data set satisfies all constraints in the schema.
-   This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
-   The second category, data pollution that is not captured by any of the constraints mentioned in the schema,
-   can be dealt with by adding constraints in the desired system.
-   In that case, a migration engineer must ensure that the desired system can start with data that satisfies all constraints in its schema.
-   So some human intervention may be necessary in specific cases.
-   The third category of pollution must be dealt with by working procedures, to prevent such pollution as much as possible.
-   In some cases, constraints on data can be formulated,
-   so the system can help the user and simplify (or eliminate) the corresponding working procedure.
-   So, such constraints may help to move this type of pollution to one of the other categories.
-   The last category of data pollution, erroneous constraints, must be fixed by removing them or replacing them by the correct constraints in the desired system.
-
-   This illustrates that automating data migrations may require user intervention,
-   either by a migration engineer or by end-users.
-   In our approach, such user interventions are given some time by
-   defining two distinct moments:
-\begin{enumerate}
-   \item the moment of transition (the MoT), i.e. the moment that changed functionality is made available to users;
-   \item the moment of cleanup (the MoC), i.e. the moment after which the old software and data can be removed.
-   This leaves the business with the intended system: a successful migration.
-\end{enumerate}
-
-   During the time between the MoT and MoC,
-   the existing system and the desired system are kept alive, side by side, as shown in Figure~\ref{fig:migration phase}.
+   while transferring data in a controlled fashion, as shown in Figure~\ref{fig:migration phase}.
 \begin{figure}[bht]
@@
 \end{figure}
+   To automate the migration as much as possible,
+   we also deploy a migration system,
+   which has its own schema to specify the migration itself.
+   The migration system comprises the schemas of both the existing system and the desired system,
+   to implement the copying of data.
+   It also implements the migration strategy,
+   automated policies to resolve data pollution,
+   and other rules at the discretion of the migration engineer.
+   To fully automate every data migration is unrealistic.
+   A migration engineer can minimize manual work by minimizing the size of the increments,
+   but she cannot eliminate it completely.
+   Data pollution, new business rules, or known issues in the existing system
+   may occasionally require the migration engineer to tailor the migration script to specific needs.
 
-   The migration requires a third schema, which specifies the migration itself.
-   The migration schema comprises the schemas of both the existing system and the desired system.
-   The rules of the migration schema will cause the right data to be copied correctly into the desired system.
-   This migration schema preserves data from the existing system, it may introduce new data automatically (generated from the existing data set),
-   and it may require users to introduce new data (after the MoT and before the MoC).
-   So the migration schema is a (disjoint) union of the existing and the desired schemas,
-   which is presented to the migration engineer in the form of source code,
-   so the migration engineer can change everything she wants to suit the particulars of the data migration.
+   A data migration has the following distinct moments:
+\begin{enumerate}
+   \item the moment of transition (the MoT), i.e. the moment that users experience changed functionality;
+   \item the moment of completion (the MoC), i.e. the migration engineer removes the migration system and the old system, including its data.
+\end{enumerate}
 
-%During the migration phase, transactions in the existing system are allowed because the migration engine will transport them to the desired system.
-%This allows the business to finish transactions in the existing system while the desired system is up and running.
-%Similarly, the existing system must execute transactions from the desired system, just in case the business calls off the migration.
-%Rules in the migration system that require user interaction are given time by requiring that the migration phase ends
-%only after all migration rules are at rest.
+   Before the migration, the existing system satisfies its invariants (by definition).
+   When the migration system compiles correctly, it satisfies its own invariants,
+   so it can be deployed in production.
+   That is the MoT.
+   The task of the migration system is to satisfy the invariants of the desired system.
+   Some of these invariants can be satisfied automatically by means of enforce rules.
+   There may be invariants, however, that require human interaction to resolve.
+   Once all invariants of the desired system are satisfied, the migration is ready for the MoC.
+   So, the schema of the migration system must contain the rules to satisfy all invariants of the desired system.
+
+   During the migration phase, transactions in the existing system are allowed because the migration engine will transport them to the desired system.
+   However, after the MoT there must be no new changes in the existing system,
+   to allow the system to finish transactions while the desired system is up and running.
 
@@
    The notations $\triples_{\dataset}$ and $\instance_{\dataset}$ are used to disambiguate $\triples$ and $\instance$ when necessary.
-   %To save writing in the sequel, the notation $a\ r\ b$ means that $\triple{a}{r}{b}\in\triples$.
+   To save writing in the sequel, the notation $a\ r\ b$ means that $\triple{a}{r}{b}\in\triples$.
 
@@
    We use just one type of enforcement rule\footnote{Ampersand has several types of enforcement rules.} in this work.
-   Its syntax is as follows:
+   Its syntax in Ampersand is as follows:
 \begin{align}
@@
    After this, triples with the relation {\tt new.isPartOf} can be inserted and deleted freely.
+   Once all triples have been copied, {\tt copied\_isPartOf} will contain all pairs of {\tt old.isPartOf} and rule~\ref*{difference} will no longer insert any triples in {\tt new.isPartOf}.
+   So, this rule can be removed after the copying has taken place.
+   Consequently, declarations~\ref{extraRelation} thru~\ref{fill copies with new.isPartOf} can be removed from the migration script at the MoC.
 

### 2023-08-12 2022Migration/articleMigrationFACS.tex
minor

@@
    to implement the copying of data.
-   It also implements the migration strategy,
-   automated policies to resolve data pollution,
-   and other rules at the discretion of the migration engineer.
    To fully automate every data migration is unrealistic.
@@
 
-   A data migration has the following distinct moments:
+   A data migration has the following distinct moments\footnote{Completion can also be done gradually, but that is outside the scope of this paper.}:
 \begin{enumerate}
-   \item the moment of transition (the MoT), i.e. the moment that users experience changed functionality;
-   \item the moment of completion (the MoC), i.e. the migration engineer removes the migration system and the old system, including its data.
+   \item the moment of transition (the MoT), after which users experience changed functionality;
+   \item the moment of completion (the MoC), after which the migration engineer removes the migration system and the old system, including its data.
 \end{enumerate}
@@
    That is the MoT.
-   The task of the migration system is to satisfy the invariants of the desired system.
-   Some of these invariants can be satisfied automatically by means of enforce rules.
+   The task of the migration system is to satisfy the invariants of the desired system
+   so that it can be deployed in production.
+   For some of these invariants, enforcement rules can satisfy them automatically.
    There may be invariants, however, that require human interaction to resolve.
@@
    \end{align}
-\item We create ENFORCE rules to copy equal relations, corresponding to equation~\ref{extraRelation} through~\ref{fill copies with new.isPartOf} in Section~\ref{sec:loading}:
+\item We create enforcement rules to copy equal relations, corresponding to equation~\ref{extraRelation} through~\ref{fill copies with new.isPartOf} in Section~\ref{sec:loading}:
    \begin{align}

### 2023-09-10 Generative Software/proposal2.tex
WIP

@@
     This proposal aims at supporting software engineers to develop software correctly and reliably by construction.
-    The threshold of learning formal specification languages has kept scores of software engineers from using them in practice.
-    Recent progress in Artificial Intelligence fuels the hope that AI might leverage the use of formal specification languages.
+    The threshold of learning specification languages has kept scores of software engineers from using them in practice.
+    Recent progress in Artificial Intelligence fuels the hope that AI might leverage the use of specification languages.
     This proposal aims at investigating how.
@@
 
-   This capability offers the following benefits:
+   We expect the following benefits:
 \begin{itemize}

### 2023-09-30 2022Migration/articleMigrationFACS.tex
minor
WIP on the introduction

@@
 \selectlanguage{english}
-\hyphenation{ExecEngine}
+\hyphenation{ExecEngine DevOps}
 %\newtheorem{lemma}{Lemma}
@@
 
-\title{Data Migration under a Changing Schema}
+\title{Data Migration under a Changing Schema in Ampersand}
 %
@@
 %
-\institute{University of Minnesota, Minneapolis, USA \and Open Universiteit Nederland, Heerlen, the Netherlands\\ 
+\institute{University of Minnesota, Minneapolis, USA \and Open Universiteit, Heerlen, the Netherlands\\ 
 \email{stef.joosten@ou.nl} \and
@@
 \label{sct:Introduction}
-   The purpose of this work is to automate the data migrations that come with incrementally developing and deploying information systems.
-   This paper proposes a theory for such data migrations in the context of generated information systems.
-   In the sequel, the word ``system'' refers to the phrase ``information system'', to simplify the language a little.
+   In practice, information systems%
+   \footnote{In the sequel, the word ``system'' refers to the phrase ``information system'', to simplify the language a little.}
+   may live for many years.
+   After they are built, they need to be upgraded at an ever-increasing pace to keep up with changing requirements in a dynamically evolving environment.
+   Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
+   Obviously, these deployments are mostly upgrades of existing systems.
 
-   A problem is that software generators typically do not support the data migration process, especially when the schema changes.
-   The ``manual'' effort required to migrate data slows down the development process and introduces mistakes.
-   Such effects can be measured in terms of DevOps metrics such as deployment frequency, reliability of deployments, and change failure rate~\cite{DevOps2021}.
+   A problem occurs when an upgrade involves a change in the schema of the deployed system:
+   The production data of the existing system must be migrated to fit the schema of the migrated system.
+   Let us call that a "schema changing data migration" (SCDM).
+   For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
+   Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
+   For good reasons, most development teams will therefore try to avoid schema changes at all costs, fearing the risk and effort of the data migration.
 
-   This paper proposes a theory for data migrations that is meant to be implemented in a software generator called Ampersand~\cite{JoostenRAMiCS2017,Joosten-JLAMP2018}.
+   During the lifetime of a system, schema changing data migrations cannot always be avoided, however.
+   It is this problem that we address in this paper.
+   Data migration for other purposes has been described in the literature.
+   For instance, if a data migration is done for switching to another platform or to different technology,
+   e.g.~\cite{Gholami2016,Bisbal1999},
+   migration engineers can and will avoid schema changes and functionality changes to avoid introducing new errors in an otherwise error-prone migration process.
+   Another example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014}, defines a migration as a variation between two data structures.
+   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
+   This does not cater for schema changes, however.
+   Then there are SCDMs in situations without a schema or an implicit schema, e.g.~\cite{Hillenbrand2022}.
+   If the schema is not available at compile time, the work will have to be done at runtime.
+   This leads to versioned storage and an overhead in performance.
+   This paper focuses on SCDMs in situations with an explicit schema that changes with the data migration.
+   The prototypical use case for that is to frequently release upgrades of information systems in production,
+   where the semantic integrity of data must be preserved across schema changes.
+   Another use case is application integration for multiple dispersed data sources with explicit schemas.
+
+   To analyze this situation, we need to define the notion of schema and how it constrains the data.
+   To facilitate a formal analysis of the situation,
+   we define an information system (section~\ref{sct:Information Systems}) as a pair of a schema and a dataset,
+   so that we have the schema explicitly at our disposal.
+   The dataset is separate, so we may consider an analysis of the schema only as a compile-time activity.
+   Section~\ref{sct:Schemas} formalizes schemas as a combination of concepts, relations and rules.
+   Relations are carriers of the data and concepts are needed to define them.
+   Rules serve as invariants on the dataset; they constrain the data to impose semantics on the data.
+   For experimental validation we use the language Ampersand~\cite{JoostenRAMiCS2017,Joosten-JLAMP2018}
+   because its syntax and semantics correspond directly to these definitions.
+   It has a notion of rule, which implements the invariants on the dataset.
+   In this way, prototypes built in Ampersand have helped to refine the theory.
+
+   In practice, data migrations typically follow Extract-Transform-Load (ETL) patterns~\cite{Theodorou2017},
+   for which many tools are available.
+   However, invariants that change yield extra work for software engineers to write the transform code,
+   for which ETL tools typically provide little support.
+   If automation can improve that, the frequency of SCDM's might well increase,
+   and downtime for the sake of migration will become less acceptable.
+   So, our research aims for zero-downtime as perceived by end-users.
+   Another practical problem is that of data quality.
+   Migrations typically suffer from a backlog of deteriorated data, incurring work to clean up data.
+   Some of that must be done before the migration; some can wait till after the migration.
+   We can define part of the data quality as satisfying semantic constraints.
+   E.g. the constraint that the combination of street name, house number, postal code, and city occurs in
+   a registration of valid addresses can be checked automatically.
+   We can actually use invariants to help find and correct data pollution.
+   We cannot use always use invariants, however.
+   An example is when a person has deliberately specified a false name, without violating any of the existing invariants.
+
+   To make the case for zero-downtime, we must distinguish between three types of constraint:
+\begin{enumerate}
+\item Transactional invariant\\
+   A \define{transactional invariant} is a constraint is always true in a system.
+   This is the classical database transaction; it may be violated temporarily, but the outside world will never notice.
+   Thus, the zero-downtime requirement holds in the eyes of an end user.
+\item Business invariant\\
+   A \define{business invariant} is a constraint that that can be violated temporarily until the computer restores it.
+   Example: All purchase orders are reflected on an order sheet, but different representations of the order sheet may be out of sync.
+   On my smartphone, I see 7 purchase orders, but in my ERP system I see 6.
+   The inconsistency arises because the back end application is still working on the consistency of the two.
+   As a user, I must trust that the two will eventually be consistent.
+   In this paper, we assume that this delay can be kept short enough to be negligible for users,
+   so that the zero-downtime requirement is satisfied in their perception.
+\item Business constraint\\
+   A \define{business constraint} is a constraint that can be violated temporarily until a user restores it.
+   Example: ``An authorized manager has to sign every purchase order.''
+   We consider business constraints not to be invariants because users can violate them temporarily.
+   This delay is definitely noticable for users.
+   To satisfy the zero-downtime requirement, the system must be up and running to let users restore business constraints.
+\end{enumerate}
+
+Summarizing, the following requirements apply to SCDM'S
+\begin{enumerate}
+\item users must experience zero-downtime.
+   Delays caused by transactional invariants or business invariants are not considered as downtime.
+\item users must be able to fix business constraints, so they can be used
+\item 
+\end{enumerate}
+
+   We can also use invariants to signal data pollution and make the migration system fix that.
+   For that purpose we need to compute violations of invariants, so that we can restore them.
+   The most important use of invariants, however, is to ensure that the migration system preserves the semantics of the data as much as possible.
+   That prevents operational disruptions in the business caused by mistakes during the migration.
+
+------------
+oud
+
+   This paper proposes a theory for data migrations that is meant to be implemented in a software generator called Ampersand.
    It aims at automating data migrations, so incremental changes can be done more reliably and faster.
@@
 
-   This paper assumes that the existing system and the desired system have different schemas.
-   The existing system is the system that is already deployed in production,
-   which makes it necessary to preserve the existing data.
-   The desired system contains a number of changes and is meant to replace the existing system.
-
-   Data migration for other purposes has been described in the literature.
-   For instance, if a data migration is done for switching to another platform or to different technology,
-   e.g.~\cite{Gholami2016,Bisbal1999},
-   migration engineers may deliberately avoid schema differences and functionality changes to avoid introducing new errors in an otherwise error-prone migration process.
-   For example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014} define an increment as a variation between two data structures.
-   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
-   Such data migrations do not allow for schema changes and do not fit our need for incremental software deployment.
+   We have not found any literature on schema changing data migrations.
    Examples like these have convinced the authors to define a new method for data migration that is specifically meant for systems in production that are
@@
    %\item {\em semantic continuity}\\the meaning of data must be preserved.
-   \item {\em zero down-time}\\The business continues during the migration without interruption.
+   \item {\em zero-downtime}\\The business continues during the migration without interruption.
    \item {\em generative software development}\\A compiler exists to generate a system from a given schema.
@@
 \subsection{Information Systems}
+   This section defines information systems, so we can formally define the migration procedure from one system to another.
+
    The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
-   This section defines information systems, after which we define a migration procedure from one system to another.
-
-   Every user has her own tasks and responsibilities and may work from different locations and on different moments.
-   This collective use by multiple users serves a purpose which we will loosely call ``the business''.
-   It is this business that depends on the semantics of the data to draw the right conclusions and carry out their tasks.
+   Multiple users, working from different locations and on different moments, constitute what we will loosely call ``the business''.
+   The data in the system constitutes the collective memory of the business,
+   which relies on the semantics of the data to draw the right conclusions and carry out their tasks.
    %This paper uses semantic constraints to represent this meaning.
 
-   Information systems are typically used by actors (both users and computers) who are distributed and work with data.
-   As a consequence, the data in a system changes continually.
+   Actors (both users and computers) are changing the data in a system continually.
    The state of the system is represented by a data set, typically represented in a persistent store or database.

### 2023-10-01 2022Migration/articleMigrationFACS.tex
new introduction is ready for Bas' review

@@
    Obviously, these deployments are mostly upgrades of existing systems.
-
-   A problem occurs when an upgrade involves a change in the schema of the deployed system:
-   The production data of the existing system must be migrated to fit the schema of the migrated system.
+   The need to do a data migration arises only if the schema of a new system differs from the existing one.
    Let us call that a "schema changing data migration" (SCDM).
@@
    Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
-   For good reasons, most development teams will therefore try to avoid schema changes at all costs, fearing the risk and effort of the data migration.
+   We can expect more SCDMs as the deployment frequency increases.
 
-   During the lifetime of a system, schema changing data migrations cannot always be avoided, however.
+   For good reasons, most development teams will try to avoid schema changes at all costs, fearing the risk and effort of the data migration.
+   During the lifetime of a system, SCDMs cannot always be avoided, however.
    It is this problem that we address in this paper.
@@
 
-   To analyze this situation, we need to define the notion of schema and how it constrains the data.
+   To analyze data migrations with changing schemas, we need to define the notion of schema and how it constrains the data.
    To facilitate a formal analysis of the situation,
@@
    so that we have the schema explicitly at our disposal.
-   The dataset is separate, so we may consider an analysis of the schema only as a compile-time activity.
+   The dataset is separate, so we may consider the analysis of just a schema as a compile-time activity.
    Section~\ref{sct:Schemas} formalizes schemas as a combination of concepts, relations and rules.
    Relations are carriers of the data and concepts are needed to define them.
-   Rules serve as invariants on the dataset; they constrain the data to impose semantics on the data.
-   For experimental validation we use the language Ampersand~\cite{JoostenRAMiCS2017,Joosten-JLAMP2018}
+   Rules constrain the data to impose semantics on the data.
+   These constraints must be satisfied all the time, so we call them {\em invariant}.
+   For experimental validation we have used the language Ampersand~\cite{JoostenRAMiCS2017,Joosten-JLAMP2018}
    because its syntax and semantics correspond directly to these definitions.
-   It has a notion of rule, which implements the invariants on the dataset.
+   Ampersand has a notion of rule, which implements the invariants on the dataset.
    In this way, prototypes built in Ampersand have helped to refine the theory.
@@
    for which ETL tools typically provide little support.
-   If automation can improve that, the frequency of SCDM's might well increase,
-   and downtime for the sake of migration will become less acceptable.
+   If automation can improve that, the frequency of SCDMs might well increase.
+   As a consequence, users will accept downtime for the sake of migration much less.
    So, our research aims for zero-downtime as perceived by end-users.
@@
 
-   To make the case for zero-downtime, we must distinguish between three types of constraint:
+   To make the case for zero-downtime, we must distinguish between two types of constraint:
 \begin{enumerate}
-\item Transactional invariant\\
-   A \define{transactional invariant} is a constraint is always true in a system.
-   This is the classical database transaction; it may be violated temporarily, but the outside world will never notice.
-   Thus, the zero-downtime requirement holds in the eyes of an end user.
-\item Business invariant\\
-   A \define{business invariant} is a constraint that that can be violated temporarily until the computer restores it.
-   Example: All purchase orders are reflected on an order sheet, but different representations of the order sheet may be out of sync.
-   On my smartphone, I see 7 purchase orders, but in my ERP system I see 6.
-   The inconsistency arises because the back end application is still working on the consistency of the two.
-   As a user, I must trust that the two will eventually be consistent.
-   In this paper, we assume that this delay can be kept short enough to be negligible for users,
-   so that the zero-downtime requirement is satisfied in their perception.
+\item Invariant\\
+   An \define{invariant} is a constraint that is always true in a system.
+   This is the classical database transaction.
+   Violation of an invariant means it has to be restored without the outside world noticing.
+   This requires either human intervention or an automated procedure.
+   If it is restored, the system commits to the new state, which satisfies the constraint.
+   If it cannot be restored, the system refuses the transaction and keeps the state as it was before starting the transaction.
+   In both cases, the invariant remains true.
 \item Business constraint\\
@@
    Example: ``An authorized manager has to sign every purchase order.''
-   We consider business constraints not to be invariants because users can violate them temporarily.
-   This delay is definitely noticable for users.
-   To satisfy the zero-downtime requirement, the system must be up and running to let users restore business constraints.
+   We consider business constraints not to be invariants because users can violate them.
 \end{enumerate}
+   Suppose some invariant of the desired system does not apply to the existing system,
+   and suppose it would be violated if applied to the data of the existing system.
+   However, spinning up the desired system on the data set of the existing system requires
+   that every constraint of the desired system is satisfied.
+   That is a problem.
+   In many cases, clearing out all violations before spinning up the desired system might take too much time and new violations will arise in the process.
+   So, keeping the existing system until the invariants of the desired system are satisfied is not an option.
+   To solve this problem, we implement every invariant in the migration system initially as a business constraint,
+   letting users restore invariance while the new system is running.
+   For this purpose, we must define an intermediate system, the \define{migration system},
+   which implements the invariants of the desired system as business constraints.
+   This migration system ensures that every violation that is fixed will not return as a violation.
+   In this way, the business constraint gradually turns into an invariant.
+   Since the number of violations is finite, it is up to the business to resolve these violations.
+   New violations that occur are under control of the desired system,
+   so the business experiences the migration system as the desired system.
+   In this way, our research realizes zero-downtime.
 
-Summarizing, the following requirements apply to SCDM'S
+   Summarizing, the following requirements apply to SCDMs
 \begin{enumerate}
-\item users must experience zero-downtime.
-   Delays caused by transactional invariants or business invariants are not considered as downtime.
-\item users must be able to fix business constraints, so they can be used
-\item 
+\item users must experience zero-downtime, to enable more frequent SCDMs.
+\item users must be able to fix business constraints, so they can be used to spin up a migration system and deliver zero-downtime.
+\item the number of violations that users must fix is finite and decreases (monotonically) over time.
 \end{enumerate}
-
-   We can also use invariants to signal data pollution and make the migration system fix that.
-   For that purpose we need to compute violations of invariants, so that we can restore them.
-   The most important use of invariants, however, is to ensure that the migration system preserves the semantics of the data as much as possible.
-   That prevents operational disruptions in the business caused by mistakes during the migration.
-
-------------
-oud
-
-   This paper proposes a theory for data migrations that is meant to be implemented in a software generator called Ampersand.
-   It aims at automating data migrations, so incremental changes can be done more reliably and faster.
-   This contributes to a more agile software development process.
-   We believe that users will experience a more organic evolution of the system because increments are smaller, occur more frequently and more reliably.
-
-   We have not found any literature on schema changing data migrations.
-   Examples like these have convinced the authors to define a new method for data migration that is specifically meant for systems in production that are
-   deployed in increments.
-
-   The contribution of this paper is to derive a migration schema from two artifacts: an existing system
-   (which has its own schema and data set) and the specification of a desired system (which has a different schema and will ``import'' relevant parts of the existing data).   
-   Our theory is based on the following assumptions and requirements:
-\begin{itemize}
-   \item {\em incremental deployment}\\The data migration is meant to deploy a software increment in production.
-   \item {\em data pollution}\\The existing data set may be polluted, but it satisfies its schema.
-   \item {\em human intervention}\\The data migration may require human interaction, which may take time.
-   %\item {\em semantic continuity}\\the meaning of data must be preserved.
-   \item {\em zero-downtime}\\The business continues during the migration without interruption.
-   \item {\em generative software development}\\A compiler exists to generate a system from a given schema.
-\end{itemize}
-   This paper focuses on the correctness of the migration.
-   Efficiency of the migration is beyond the scope.
+   An information system that is defined by constraints on a data set
+   can signal certain forms of data pollution by adding the proper constraints.
+   The mechanism of business constraints can help to clean that pollution.
 

### 2023-10-02 2022Migration/articleMigrationFACS.tex
refine some formulations (minor)

@@
    Obviously, these deployments are mostly upgrades of existing systems.
-   The need to do a data migration arises only if the schema of a new system differs from the existing one.
-   Let us call that a "schema changing data migration" (SCDM).
+   If an upgrade does not change the schema, we don't need a data migration.
+   Yet, schema changes cannot always be avoided, so a {\em schema changing data migration} (SCDM) will be neccessary from time to time.
    For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
    Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
-   We can expect more SCDMs as the deployment frequency increases.
+   So, development teams will try to avoid schema changes at all costs, fearing the risk and effort of the data migration,
+   and for a good reason.
 
-   For good reasons, most development teams will try to avoid schema changes at all costs, fearing the risk and effort of the data migration.
-   During the lifetime of a system, SCDMs cannot always be avoided, however.
-   It is this problem that we address in this paper.
-   Data migration for other purposes has been described in the literature.
+   Data migration for other purposes than schema change has been described in the literature.
    For instance, if a data migration is done for switching to another platform or to different technology,
@@
    To facilitate a formal analysis of the situation,
-   we define an information system (section~\ref{sct:Information Systems}) as a pair of a schema and a dataset,
+   we define an information system (in section~\ref{sct:Information Systems}) as a pair of a schema and a dataset,
    so that we have the schema explicitly at our disposal.
@@
    for which ETL tools typically provide little support.
-   If automation can improve that, the frequency of SCDMs might well increase.
-   As a consequence, users will accept downtime for the sake of migration much less.
-   So, our research aims for zero-downtime as perceived by end-users.
+   The pressure on increasing deploy times calls for further automation of SCDMs.
+   Hence, we can expect more frequent SCDMs, and downtime for the sake of migration becomes less acceptable.
+   So, our research aims for zero-downtime.
    Another practical problem is that of data quality.
@@
    a registration of valid addresses can be checked automatically.
-   We can actually use invariants to help find and correct data pollution.
-   We cannot use always use invariants, however.
-   An example is when a person has deliberately specified a false name, without violating any of the existing invariants.
+   An information system that is defined by constraints on a data set
+   can signal this type of data pollution by adding such constraints.
+   Some forms of data pollution cannot be detected in this way, however.
+   An example is when a person has deliberately specified a false name without violating any conceivable constraint.
 
@@
    An \define{invariant} is a constraint that is always true in a system.
-   This is the classical database transaction.
+   This corresponds to the classical database transaction.
    Violation of an invariant means it has to be restored without the outside world noticing.
@@
    We consider business constraints not to be invariants because users can violate them.
+   They are not true all the time.
 \end{enumerate}
-   Suppose some invariant of the desired system does not apply to the existing system,
+   Suppose some invariant, $u$, of the {\em desired system} does not apply to the {\em existing system},
    and suppose it would be violated if applied to the data of the existing system.
-   However, spinning up the desired system on the data set of the existing system requires
-   that every constraint of the desired system is satisfied.
-   That is a problem.
+   Now, when we spin up the desired system on the data set of the existing system, we have a problem.
+   Every invariant of the desired system must be satisfied but $u$ isn't.
+   So we cannot spin it up.
    In many cases, clearing out all violations before spinning up the desired system might take too much time and new violations will arise in the process.
-   So, keeping the existing system until the invariants of the desired system are satisfied is not an option.
    To solve this problem, we implement every invariant in the migration system initially as a business constraint,
@@
    so the business experiences the migration system as the desired system.
-   In this way, our research realizes zero-downtime.
+   In this way, the migration system bridges the gap and users get a zero-downtime SCDM.
 
@@
 \item users must experience zero-downtime, to enable more frequent SCDMs.
-\item users must be able to fix business constraints, so they can be used to spin up a migration system and deliver zero-downtime.
-\item the number of violations that users must fix is finite and decreases (monotonically) over time.
+\item users must be able to fix business constraints, so we need an intermediate ``migration system'' to deliver zero-downtime.
+\item the number of violations that users must fix is finite and decreases (monotonically) over time,
+   to ensure that the migration system can be removed after the migration is done.
 \end{enumerate}
-   An information system that is defined by constraints on a data set
-   can signal certain forms of data pollution by adding the proper constraints.
-   The mechanism of business constraints can help to clean that pollution.
 

### 2023-10-07 2022Migration/articleMigrationFACS.tex
minor, to avoid an overfull box
violations

@@
 
-This paper defines a migration process that preserves the semantics of data by means of rules (constraints).
+   This paper defines a migration process that preserves the semantics of data by means of invariants (constraints)
+   with zero downtime for users.
    The migration process is based on assumptions and requirements that aim to capture a large class of data migrations in practice.
    This paper focuses on correctness of the migration. Efficiency of the migration is outside its scope.
+\keywords{Generative software \and Incremental software deployment \and Data migration \and Relation algebra \and Ampersand \and Schema change. \and Invariants \and Zero downtime.}
+\end{abstract}
 
-\keywords{Generative software \and Incremental software deployment \and Data migration \and Relation algebra \and Ampersand \and Schema change.}
-\end{abstract}
-%
-%
-%
 \section{Introduction}
@@
    may live for many years.
-   After they are built, they need to be upgraded at an ever-increasing pace to keep up with changing requirements in a dynamically evolving environment.
+   After they are built, they need to be upgraded regularly to keep up with changing requirements in a dynamically evolving environment.
    Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
    Obviously, these deployments are mostly upgrades of existing systems.
-   If an upgrade does not change the schema, we don't need a data migration.
-   Yet, schema changes cannot always be avoided, so a {\em schema changing data migration} (SCDM) will be neccessary from time to time.
+   In upgrades, schema changes cannot always be avoided, so a {\em schema changing data migration} (SCDM) will be neccessary from time to time.
    For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
    Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
-   So, development teams will try to avoid schema changes at all costs, fearing the risk and effort of the data migration,
-   and for a good reason.
+   The risk and effort of such data migrations explains why development teams try to avoid schema changes at all costs.
 
@@
    If the schema is not available at compile time, the work will have to be done at runtime.
-   This leads to versioned storage and an overhead in performance.
+   This requires versioned storage of production data and an overhead in performance.
    This paper focuses on SCDMs in situations with an explicit schema that changes with the data migration.
@@
    To facilitate a formal analysis of the situation,
-   we define an information system (in section~\ref{sct:Information Systems}) as a pair of a schema and a dataset,
+   we define an information system (in section~\ref{sct:Information Systems}) as a pair of a schema and a data set,
    so that we have the schema explicitly at our disposal.
-   The dataset is separate, so we may consider the analysis of just a schema as a compile-time activity.
-   Section~\ref{sct:Schemas} formalizes schemas as a combination of concepts, relations and rules.
+   The data set is separate, so we may consider the analysis of just a schema as a compile-time activity.
+   Section~\ref{sct:Schemas} formalizes schemas as a combination of concepts, relations and invariants.
    Relations are carriers of the data and concepts are needed to define them.
-   Rules constrain the data to impose semantics on the data.
-   These constraints must be satisfied all the time, so we call them {\em invariant}.
+   Invariants are constraints on the data to impose semantics.
+   These constraints must be satisfied all the time.
    For experimental validation we have used the language Ampersand~\cite{JoostenRAMiCS2017,Joosten-JLAMP2018}
    because its syntax and semantics correspond directly to these definitions.
-   Ampersand has a notion of rule, which implements the invariants on the dataset.
+   Ampersand has a notion of rule, which implements the invariants on the data set.
    In this way, prototypes built in Ampersand have helped to refine the theory.
@@
    for which ETL tools typically provide little support.
-   The pressure on increasing deploy times calls for further automation of SCDMs.
-   Hence, we can expect more frequent SCDMs, and downtime for the sake of migration becomes less acceptable.
-   So, our research aims for zero-downtime.
+   The pressure to decrease deploy times calls for further automation of SCDMs.
+   Hence, we can expect more frequent SCDMs, so downtime for the sake of migration becomes less acceptable.
+   That is why our research aims for zero downtime.
+
    Another practical problem is that of data quality.
@@
    Some of that must be done before the migration; some can wait till after the migration.
-   We can define part of the data quality as satisfying semantic constraints.
+   We can capture part of the data quality problem as a requirement to satisfy semantic constraints.
    E.g. the constraint that the combination of street name, house number, postal code, and city occurs in
    a registration of valid addresses can be checked automatically.
-   An information system that is defined by constraints on a data set
-   can signal this type of data pollution by adding such constraints.
+   In a formalism like Ampersand, which allows us to express such constraints, we can add constraints for data quality to the schema.
+   This allows us to signal the data pollution at runtime.
    Some forms of data pollution cannot be detected in this way, however.
@@
 
-   To make the case for zero-downtime, we must distinguish between two types of constraint:
+   To make the case for zero downtime, we must distinguish between two types of constraint:
 \begin{enumerate}
@@
    This requires either human intervention or an automated procedure.
-   If it is restored, the system commits to the new state, which satisfies the constraint.
+   Once it is restored, the system commits to the new state, which satisfies the constraint.
    If it cannot be restored, the system refuses the transaction and keeps the state as it was before starting the transaction.
    In both cases, the invariant remains true.
+
+   A special case of an invariant is the \define{enforcement rule}.
+   As soon as an enforce rule is violated, the system restores it automatically, without human intervention.
 \item Business constraint\\
@@
    and suppose it would be violated if applied to the data of the existing system.
-   Now, when we spin up the desired system on the data set of the existing system, we have a problem.
+   Now, if we try to spin up the desired system on the data set of the existing system, we have a problem.
    Every invariant of the desired system must be satisfied but $u$ isn't.
    So we cannot spin it up.
-   In many cases, clearing out all violations before spinning up the desired system might take too much time and new violations will arise in the process.
+   In many cases, clearing out all violations before spinning up the desired system might take too much time and new violations will arise in the meantime.
    To solve this problem, we implement every invariant in the migration system initially as a business constraint,
@@
    which implements the invariants of the desired system as business constraints.
-   This migration system ensures that every violation that is fixed will not return as a violation.
+   This migration system must ensure that every violation that is fixed will not return as a violation.
    In this way, the business constraint gradually turns into an invariant.
-   Since the number of violations is finite, it is up to the business to resolve these violations.
+   Since the number of violations is finite, the business can resolve these violations in finite time.
    New violations that occur are under control of the desired system,
    so the business experiences the migration system as the desired system.
-   In this way, the migration system bridges the gap and users get a zero-downtime SCDM.
+   In this way, the migration system bridges the gap and users get a zero downtime SCDM.
 
@@
 \begin{enumerate}
-\item users must experience zero-downtime, to enable more frequent SCDMs.
-\item users must be able to fix business constraints, so we need an intermediate ``migration system'' to deliver zero-downtime.
+\item users must experience zero downtime, to enable more frequent SCDMs.
+\item users must be able to fix business constraints, so we need an intermediate ``migration system'' to deliver zero downtime.
 \item the number of violations that users must fix is finite and decreases (monotonically) over time,
@@
    Actors (both users and computers) are changing the data in a system continually.
-   The state of the system is represented by a data set, typically represented in a persistent store or database.
+   The state of the system is represented by a data set, typically represented in some form of persistent store such as a database.
    Events that the system detects may cause the state to change.
@@
 \end{figure}
-   To keep our theory technology independent, data sets are assumed to contain triples.
+   To keep our theory technology independent, we assume data sets to contain triples.
    This makes our theory valid for any kind of database that triples can represent,
    including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
-   An important aspect of the system's semantics are the constraints, sometimes known as invariants or integrity rules.
+   An important aspect of the system's semantics are the invariants, sometimes known as integrity rules.
    %The purpose of these rules is to keep all constraints satisfied.
 
-   Like most software generators, Ampersand is a compiler that generates information systems from a script.
-   Ampersand uses a form of heterogeneous relation algebra,
+   Ampersand is a compiler that generates information systems from a script.
+   It uses a form of heterogeneous relation algebra,
    which works with relations in a way similar to Alloy~\cite{Alloy2006}.
    A system is then described by a set of rules (aka constraint / invariant), which the system keeps satisfied.
-   So all constraints are explicitly available as rules in the Ampersand code.
+   As a result, all invariants are explicit in the source code of a system specified in Ampersand.
 
@@
 \end{figure}
-   To automate the migration as much as possible,
-   we also deploy a migration system,
-   which has its own schema to specify the migration itself.
-   The migration system comprises the schemas of both the existing system and the desired system,
-   to implement the copying of data.
-   To fully automate every data migration is unrealistic.
-   A migration engineer can minimize manual work by minimizing the size of the increments,
-   but she cannot eliminate it completely.
-   Data pollution, new business rules, or known issues in the existing system
-   may occasionally require the migration engineer to tailor the migration script to specific needs.
+   To automate the migration as much as possible and to achieve zero downtime,
+   we also deploy a migration system.
+   The migration system has its own schema.
+   It comprises the schemas of both the existing system and the desired system,
+   so the migration system can copy the data from the existing to the desired system.
+   Some of the migration work, however, cannot be automated.
+   A migration engineer can minimize manual work by minimizing the size of the increments.
+   But data pollution, new business rules, or known issues in the existing system
+   may occasionally require tailoring the migration script to specific needs
+   that require action of users in production.
+   In Ampersand, she can use business constraints to implement these needs.
 
@@
    An {\em information system} is a combination of data set, schema, and functionality.
-   For the purpose of this paper, functionality captured in user interfaces are ignored because it does not impact the migration.
+   For the purpose of this paper, we ignore functionality captured in user interfaces because it does not impact the migration.
    Section~\ref{sct:Data sets} desribes data sets. Schemas are treated in section~\ref{sct:Schemas}.
@@
 
-   Relations serve to organize and store data, to allow a developer to represent facts.
+   Relations serve to organize and store data, allowing developers to represent facts.
    In this paper, variables $r$, $s$, and $d$ represent relations\footnote{Some readers might like to read `relation symbol' where we write `relation'.}.
@@
    We overload the notation $\popF{}$ so we can use it on concepts $\popF{C}:\Dataset\rightarrow\powerset{\Atoms}$
-   and expressions.
+   and expressions. We also define the difference of populations, both for relations and concepts:
 \begin{align}
@@
 
+\subsection{Violations}
+\label{sct:Violations}
+   To satisfy a constraint $u$ means that it is true for all triples in the data set.
+   Let us assume that we can represent each violation of a constraint $u$ as a pair
+   and let the function $\violC{u}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$ produce these violations.
+   Note that this assumption implies that rules have a signature, just like a relations.
+   So, we can represent an invariant $u$ as a triple $\triple{\violC{u}}{A}{B}$
+   that satisfies:
+\begin{equation}
+   \pair{a}{b}\in\viol{\triple{\violC{u}}{A}{B}}{\dataset}\ \Rightarrow\ a\instance A\ \wedge\ b\instance B
+\end{equation}
+   We represent an enforce rule $u$ as a pair $\pair{\declare{n}{A}{B}}{\violC{u}}$,
+   where relation $\declare{n}{A}{B}$ is used to restore invariance of the enforce rule.
+   The enforce rule satisfies:
+\begin{equation}
+   \pair{a}{b}\in\viol{u}{\dataset}\ \Rightarrow\ a\instance A\ \wedge\ b\instance B\ \wedge\ \viol{u}{\dataset\cup\triple{a}{\declare{n}{A}{B}}{b}}=\emptyset
+\end{equation}
+
 \subsection{Schemas}
@@
    Schemas serve to capture the semantics of a system~\cite{Spivak2012}.
-   We consider a schema to represent the static part of a system, defining concepts, relations, and rules.
-   A software engineer defines a schema on design time, so that semantic checks can be implemented at compile time.
+   It defines concepts, relations, and invariants.
+   We assume that a software engineer defines a schema on design time, and a compiler checks whether the semantics are consistent.
+   Error messages from the compiler are prohibitive for generating code,
+   which prevents a substantial class of mistakes to ever reach end users.
 
@@
    $\rels\subseteq \Rels$ is a finite set of relations,
-   $\rules\subseteq \Triple{(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}})}{\Concepts}{\Concepts}$ is a finite set of rules,
+   $\rules\subseteq \Triple{(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}})}{\Concepts}{\Concepts}$ is a finite set of invariants,
    and $\enforces\subseteq \Pair{\Rels}{\left(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}\right)}$ is a finite set of enforcement rules.
 
-   Each rule in a schema serves to constrain the data set at runtime, to ensure its semantic integrity.
-   Every rule is an element of an infinite set called $\Rules$.
+   Each invariant in a schema serves to constrain the data set at runtime, to ensure its semantic integrity.
+   Every invariant is an element of an infinite set called $\Rules$.
    In this paper, variables $u$ and $v$ represent rules.
-   A rule $u$ in the schema is a triple $\triple{\violC{u}}{A}{B}$ such that $\viol{u}{\dataset}$ represents any violations to $u$ in data set $\dataset$.
+   A invariant $u$ in the schema is a triple $\triple{\violC{u}}{A}{B}$ such that $\viol{u}{\dataset}$ represents any violations to $u$ in data set $\dataset$.
    In Ampersand, rules are well-typed, just like relations:
-   Every violation $\pair{a}{b}$ of a rule $\triple{\violC{u}}{A}{B}$ must satisfy $a\instance A$ and $b\instance B$.
+   Every violation $\pair{a}{b}$ of a invariant $\triple{\violC{u}}{A}{B}$ must satisfy $a\instance A$ and $b\instance B$.
    % For every data set $\dataset$, the set $\viol{u}{\dataset}$ is a subset of $\pop{A}{\dataset}\times\pop{B}{\dataset}$.
@@
 \end{equation}
-   In this example, relation $\declare{\tt isEnrolledFor}{\tt Student}{\tt Module}$ is empty, so rule {\tt EnrollRule} is satisfied.
+   In this example, relation $\declare{\tt isEnrolledFor}{\tt Student}{\tt Module}$ is empty, so the rule is satisfied.
 
@@
 \begin{enumerate}
-\item We take a disjoint union of the relations, defining $\overleftarrow~$ and $\overrightarrow~$ as (overloaded) notation to generate new names, keeping the dataset of the old system and the rules of the new system:
+\item We take a disjoint union of the relations, defining $\overleftarrow~$ and $\overrightarrow~$ as (overloaded) notation to generate new names, keeping the data set of the old system and the rules of the new system:
    \begin{align}
@@
 
-   The MoC is reached for dataset $\dataset'$ whenever $\pop{{\tt viols}_u - {\tt fixed}_u}{\dataset'}=\emptyset$ for all rules.
+   The MoC is reached for data set $\dataset'$ whenever $\pop{{\tt viols}_u - {\tt fixed}_u}{\dataset'}=\emptyset$ for all rules.
    This should happen after a finite number of inserting events: let $I = \{{\tt fixed}_u \mid u\in\rules_{\schema'}\}$ be the relations that keep track of which violations have been fixed, then there is no infinite chain of systems $\migrsys' = \migrsys_0, \migrsys_1,\ldots$ such that $\migrsys_i\xrightarrowdbl[I]{\overrightarrow\rels_{\schema'}}\migrsys_{i+1}$.
@@
    There needs to be a system that has the desired schema $\schema'$.
-   If rules in $\schema'$ are such that they cannot all be satisfied, regardless of the dataset, no such system exists and we have no hope of reaching it.
-   We assume that the rules of $\schema'$ are intended to prevent data pollution, such that a dataset $\dataset'$ describing the `real world' would be one that satisfies them.
+   If rules in $\schema'$ are such that they cannot all be satisfied, regardless of the data set, no such system exists and we have no hope of reaching it.
+   We assume that the rules of $\schema'$ are intended to prevent data pollution, such that a data set $\dataset'$ describing the `real world' would be one that satisfies them.
    So let $\pair{\dataset'}{\schema'}$ be a system, then we can get to it through our migration system.
@@
    This work does not consider what to do about (user) interfaces.
-   Instead, it models events by assuming that any change to the dataset can be achieved.
+   Instead, it models events by assuming that any change to the data set can be achieved.
    In practice, such changes need to be achieved through interfaces.
-   Most Ampersand systems indeed allow the users of the system to edit the dataset quite freely through the interfaces.
+   Most Ampersand systems indeed allow the users of the system to edit the data set quite freely through the interfaces.
    However, some interfaces may require certain constraints to be satisfied, which means that interfaces of the desired system might break.

### 2023-10-08 2022Migration/articleMigrationFACS.tex
WIP

@@
    may live for many years.
-   After they are built, they need to be upgraded regularly to keep up with changing requirements in a dynamically evolving environment.
+   After they are built, they need to be updated regularly to keep up with changing requirements in a dynamically evolving environment.
    Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
-   Obviously, these deployments are mostly upgrades of existing systems.
-   In upgrades, schema changes cannot always be avoided, so a {\em schema changing data migration} (SCDM) will be neccessary from time to time.
+   Obviously, these deployments are mostly updates of existing systems.
+   Schema changes cannot always be avoided when updating software, so a {\em schema changing data migration} (SCDM) will be neccessary from time to time.
    For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
@@
    This paper focuses on SCDMs in situations with an explicit schema that changes with the data migration.
-   The prototypical use case for that is to frequently release upgrades of information systems in production,
+   The prototypical use case for that is to release updates of information systems in production,
    where the semantic integrity of data must be preserved across schema changes.
    Another use case is application integration for multiple dispersed data sources with explicit schemas.
+
+   In practice, data migrations typically follow Extract-Transform-Load (ETL) patterns~\cite{Theodorou2017},
+   for which many tools are available.
+   However, invariants that change yield extra work for software engineers to write the transform code,
+   for which ETL tools typically provide little support.
+   The ubiquitous pressure to decrease deploy times and increase release frequencies calls for further automation of SCDMs.
+   Hence, any downtime for the sake of migration becomes less and less acceptable.
+   That is why our research aims for zero downtime.
+
+   Another practical problem is that of data quality.
+   Migrations typically suffer from a backlog of deteriorated data, incurring work to clean it up.
+   Some of that work must be done before the migration; some can wait till after the migration.
+   We can capture part of the data quality problem as a requirement to satisfy semantic constraints.
+   E.g. the constraint that the combination of street name, house number, postal code, and city occurs in
+   a registration of valid addresses can be checked automatically.
+   In a formalism like Ampersand, which allows us to express such constraints, we can add constraints for data quality to the schema.
+   This allows us to signal the data pollution at runtime.
+   Some forms of data pollution cannot be detected in this way, however.
+   An example is when a person has deliberately specified a false name without violating any constraint in the system.
 
@@
 
-   In practice, data migrations typically follow Extract-Transform-Load (ETL) patterns~\cite{Theodorou2017},
-   for which many tools are available.
-   However, invariants that change yield extra work for software engineers to write the transform code,
-   for which ETL tools typically provide little support.
-   The pressure to decrease deploy times calls for further automation of SCDMs.
-   Hence, we can expect more frequent SCDMs, so downtime for the sake of migration becomes less acceptable.
-   That is why our research aims for zero downtime.
+\section{Analysis}
+\label{sct:Analysis}
+   This section analyzes information systems, so we can formally define the migration procedure from one system to another.
+\subsection{Information Systems}
+   The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
+   Multiple users, working from different locations and on different moments, constitute what we will loosely call ``the business''.
+   The data in the system constitutes the collective memory of the business,
+   which relies on the semantics of the data to draw the right conclusions and carry out their tasks.
+   %This paper uses semantic constraints to represent this meaning.
+\begin{figure}[bht]
+   \begin{center}
+     \includegraphics[scale=.45]{figures/datamigration-Pre-migration.png}
+   \end{center}
+\caption{Anatomy of an information system}
+\label{fig:pre-migration}
+\end{figure}
 
-   Another practical problem is that of data quality.
-   Migrations typically suffer from a backlog of deteriorated data, incurring work to clean up data.
-   Some of that must be done before the migration; some can wait till after the migration.
-   We can capture part of the data quality problem as a requirement to satisfy semantic constraints.
-   E.g. the constraint that the combination of street name, house number, postal code, and city occurs in
-   a registration of valid addresses can be checked automatically.
-   In a formalism like Ampersand, which allows us to express such constraints, we can add constraints for data quality to the schema.
-   This allows us to signal the data pollution at runtime.
-   Some forms of data pollution cannot be detected in this way, however.
-   An example is when a person has deliberately specified a false name without violating any conceivable constraint.
+   Actors (both users and computers) are changing the data in a system continually.
+   The state of the system is represented by a data set, typically represented in some form of persistent store such as a database.
+   Events that the system detects may cause the state to change.
+   To keep our theory technology independent, we assume data sets to contain triples.
+   This makes our theory valid for any kind of database that triples can represent,
+   including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
 
-   To make the case for zero downtime, we must distinguish between two types of constraint:
+   We assume that the semantics of the data is ensured by means of constraints.
+   In figure~\ref{fig:pre-migration} they are depicted in the schema s rules.
+   We distinguish two types of constraint:
 \begin{enumerate}
@@
    An \define{invariant} is a constraint that is always true in a system.
-   This corresponds to the classical database transaction.
-   Violation of an invariant means it has to be restored without the outside world noticing.
-   This requires either human intervention or an automated procedure.
-   Once it is restored, the system commits to the new state, which satisfies the constraint.
-   If it cannot be restored, the system refuses the transaction and keeps the state as it was before starting the transaction.
-   In both cases, the invariant remains true.
+   It serves to constrain the data set at runtime, to ensure its semantic integrity.
+   % Violation of an invariant means it has to be restored without the outside world noticing.
+   % This requires either human intervention or an automated procedure.
+   % Once it is restored, the system commits to the new state, which satisfies the constraint.
+   % If it cannot be restored, the system refuses the transaction and keeps the state as it was before starting the transaction.
+   % In both cases, the invariant remains true.
 
    A special case of an invariant is the \define{enforcement rule}.
-   As soon as an enforce rule is violated, the system restores it automatically, without human intervention.
+   This corresponds to the classical database transaction.
+   As soon as an enforcement rule is violated, the system restores it automatically, without human intervention,
+   so the outside world will never notice that it was violated.
 \item Business constraint\\
@@
 \end{enumerate}
-   Suppose some invariant, $u$, of the {\em desired system} does not apply to the {\em existing system},
-   and suppose it would be violated if applied to the data of the existing system.
-   Now, if we try to spin up the desired system on the data set of the existing system, we have a problem.
-   Every invariant of the desired system must be satisfied but $u$ isn't.
-   So we cannot spin it up.
-   In many cases, clearing out all violations before spinning up the desired system might take too much time and new violations will arise in the meantime.
-   To solve this problem, we implement every invariant in the migration system initially as a business constraint,
+
+\subsection{Zero downtime}
+   To make the case for zero downtime, consider this problem:
+   Suppose we have an invariant, $u$, in the {\em desired system}, which is absent in the {\em existing system}.
+   And suppose the data in the existing system does not satisfy $u$.
+   When spinning up the desired system on the data set of the existing system, $u$ must be satisfied.
+   If $u$ is an enforcement rule, the desired system will restore it automatically.
+   But otherwise, the system will refuse to spin up.
+   In many cases, clearing out all violations before spinning up the desired system might take too long, while new violations may arise in the meantime.
+   So, the problem lies with invariants that are not enforced automatically.
+   To solve this problem, we must implement such invariants initially as a business constraint,
    letting users restore invariance while the new system is running.
    For this purpose, we must define an intermediate system, the \define{migration system},
-   which implements the invariants of the desired system as business constraints.
-   This migration system must ensure that every violation that is fixed will not return as a violation.
-   In this way, the business constraint gradually turns into an invariant.
+   which implements the invariants of the desired system as business constraints, unless they are enforcement rules.
+   This migration system must also ensure that every violation that is fixed will not return as a violation.
+   In this way, the business constraint eventually becomes an invariant, satisfying the specification of the desired system.
    Since the number of violations is finite, the business can resolve these violations in finite time.
@@
 \end{enumerate}
-
-\section{Analysis}
-\label{sct:Analysis}
-\subsection{Information Systems}
-   This section defines information systems, so we can formally define the migration procedure from one system to another.
-
-   The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
-   Multiple users, working from different locations and on different moments, constitute what we will loosely call ``the business''.
-   The data in the system constitutes the collective memory of the business,
-   which relies on the semantics of the data to draw the right conclusions and carry out their tasks.
-   %This paper uses semantic constraints to represent this meaning.
-
-   Actors (both users and computers) are changing the data in a system continually.
-   The state of the system is represented by a data set, typically represented in some form of persistent store such as a database.
-   Events that the system detects may cause the state to change.
-\begin{figure}[bht]
-   \begin{center}
-     \includegraphics[scale=.45]{figures/datamigration-Pre-migration.png}
-   \end{center}
-\caption{Anatomy of an information system}
-\label{fig:pre-migration}
-\end{figure}
-   To keep our theory technology independent, we assume data sets to contain triples.
-   This makes our theory valid for any kind of database that triples can represent,
-   including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
-   An important aspect of the system's semantics are the invariants, sometimes known as integrity rules.
-   %The purpose of these rules is to keep all constraints satisfied.
 
@@
 
-   Before the migration, the existing system satisfies its invariants (by definition).
+   Before the migration, the data in the existing system satisfies its invariants (by definition).
    When the migration system compiles correctly, it satisfies its own invariants,
@@
    and let the function $\violC{u}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$ produce these violations.
-   Note that this assumption implies that rules have a signature, just like a relations.
-   So, we can represent an invariant $u$ as a triple $\triple{\violC{u}}{A}{B}$
-   that satisfies:
+   Note that $\viol{u}{\dataset}=\emptyset$ means that $\dataset$ satisfies $u$.
+   Now we can represent an invariant $u$ as a triple $\triple{\violC{u}}{A}{B}$.
+   It must be well-typed, just like a relation:
 \begin{equation}
@@
 \end{equation}
-   We represent an enforce rule $u$ as a pair $\pair{\declare{n}{A}{B}}{\violC{u}}$,
-   where relation $\declare{n}{A}{B}$ is used to restore invariance of the enforce rule.
-   The enforce rule satisfies:
+   We represent an enforcement rule $u$ as a pair $\pair{\declare{n}{A}{B}}{\violC{u}}$,
+   where relation $\declare{n}{A}{B}$ is used to restore invariance of the enforcement rule.
+   The enforcement rule must be well-typed:
 \begin{equation}
-   \pair{a}{b}\in\viol{u}{\dataset}\ \Rightarrow\ a\instance A\ \wedge\ b\instance B\ \wedge\ \viol{u}{\dataset\cup\triple{a}{\declare{n}{A}{B}}{b}}=\emptyset
+   \pair{a}{b}\in\viol{\pair{\declare{n}{A}{B}}{\violC{u}}}{\dataset}\ \Rightarrow\ a\instance A\ \wedge\ b\instance B
 \end{equation}
+   Besides, an enforcement rule must be automatically restorable by adding the violations to relation $\declare{n}{A}{B}$:
+\begin{equation}
+   \viol{u}{\dataset\cup\{\triple{a}{\declare{n}{A}{B}}{b}\mid\pair{a}{b}\in\viol{u}{\dataset}\}}=\emptyset
+\end{equation}
+   The language Ampersand has more types of enforcement rules than just this one,
+   but to define this one only is sufficient for this paper.
 
@@
    Schemas serve to capture the semantics of a system~\cite{Spivak2012}.
-   It defines concepts, relations, and invariants.
+   They define concepts, relations, and invariants.
    We assume that a software engineer defines a schema on design time, and a compiler checks whether the semantics are consistent.
@@
 
-   Each invariant in a schema serves to constrain the data set at runtime, to ensure its semantic integrity.
    Every invariant is an element of an infinite set called $\Rules$.
    In this paper, variables $u$ and $v$ represent rules.
-   A invariant $u$ in the schema is a triple $\triple{\violC{u}}{A}{B}$ such that $\viol{u}{\dataset}$ represents any violations to $u$ in data set $\dataset$.
    In Ampersand, rules are well-typed, just like relations:

### 2023-10-15 2022Migration/articleMigrationFACS.tex
Adapt the informal chapters 1 and 2.

@@
    for which many tools are available.
-   However, invariants that change yield extra work for software engineers to write the transform code,
+   However, invariants that change require writing code in the transform phase,
    for which ETL tools typically provide little support.
-   The ubiquitous pressure to decrease deploy times and increase release frequencies calls for further automation of SCDMs.
-   Hence, any downtime for the sake of migration becomes less and less acceptable.
+   This yields extra work for software engineers, and it increases the risk of software errors.
+   The ubiquitous pressure to decrease deploy times calls for further automation of SCDMs.
+   With the release frequencies increasing and deploy times decreasing, any downtime for the sake of migration becomes less and less acceptable.
    That is why our research aims for zero downtime.
@@
 
-   To analyze data migrations with changing schemas, we need to define the notion of schema and how it constrains the data.
-   To facilitate a formal analysis of the situation,
-   we define an information system (in section~\ref{sct:Information Systems}) as a pair of a schema and a data set,
+   The next chapter analyzes SCDMs with an eye on zero downtime and data quality.
+   To prevent mistakes, this analysis takes a formal approach.
+   We define an information system (in section~\ref{sct:Information Systems}) as a pair of a schema and a data set,
    so that we have the schema explicitly at our disposal.
@@
 \label{sct:Analysis}
-   This section analyzes information systems, so we can formally define the migration procedure from one system to another.
+   This section analyzes information systems qualitatively, to prepare for a formal treatment in section~\ref{sct:Definitions}.
+   The current section yields a procedure for migrating data from one system to another.
 \subsection{Information Systems}
@@
    To keep our theory technology independent, we assume data sets to contain triples.
-   This makes our theory valid for any kind of database that triples can represent,
+   This makes our theory valid for every kind of database that triples can represent,
    including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
@@
    We assume that the semantics of the data is ensured by means of constraints.
-   In figure~\ref{fig:pre-migration} they are depicted in the schema s rules.
-   We distinguish two types of constraint:
+   Figure~\ref{fig:pre-migration} depicts them as rules.
+   Some constraints must be satisfied through human intervention, while others are being satisfied by the system.
+   So, we distinguish between two types of constraint:
 \begin{enumerate}
-\item Invariant\\
-   An \define{invariant} is a constraint that is always true in a system.
+\item Blocking invariant\\
+   A \define{blocking invariant} is a constraint that is always true in a system.
    It serves to constrain the data set at runtime, to ensure its semantic integrity.
-   % Violation of an invariant means it has to be restored without the outside world noticing.
-   % This requires either human intervention or an automated procedure.
-   % Once it is restored, the system commits to the new state, which satisfies the constraint.
-   % If it cannot be restored, the system refuses the transaction and keeps the state as it was before starting the transaction.
-   % In both cases, the invariant remains true.
-
-   A special case of an invariant is the \define{enforcement rule}.
-   This corresponds to the classical database transaction.
-   As soon as an enforcement rule is violated, the system restores it automatically, without human intervention,
+   When it is violated, the system produces an error message and refuses the transaction.
+\item Enforced invariant\\
+   An enforced invariant corresponds to the classical database transaction.
+   As soon as the data violates an enforced invariant, the system restores it automatically without human intervention,
    so the outside world will never notice that it was violated.
 \item Business constraint\\
-   A \define{business constraint} is a constraint that can be violated temporarily until a user restores it.
+   A \define{business constraint} is a constraint that users can violate temporarily until someone restores it.
    Example: ``An authorized manager has to sign every purchase order.''
-   We consider business constraints not to be invariants because users can violate them.
-   They are not true all the time.
+   A business constraint is not true all the time, so we do not consider it to be an invariant%
+\footnote{The business may impose a business constraint, but that doesn't make it satisfied.
+   Every violation requires some form of human action to satisfy the business constraint (e.g. "sign the purchase order").}.
 \end{enumerate}
+   Summarizing, in our notion of information systems, concepts, relations, and constraints carry the business semantics.
+
+\subsection{Ampersand}
+   Ampersand is a language that specifies information systems as a system of concepts, relations, and constraints.
+   It features the types of constraints that we need for this paper, so we have used it to try our theory in practice.
+   Ampersand has a compiler that generates the information system from a script.
+   A developer specifies constraints in heterogeneous relation algebra~\cite{Hattensperger1999,Alloy2006}.
+   The generated system keeps the invariants satisfied and signals violations of business constraints to a user.
+   Having all invariants explicit in the Ampersand source code makes Ampersand a suitable platform on which to implement our theory.
+   The absence of imperative code in an Ampersand script makes it easier to reason about the system.
+   Constraints carry the business semantics, so they allow us to be explicit about ``preserving the meaning as much as possible''.
+   An Ampersand script contains just enough information to generate a complete system,
+   which means that a classical database schema (i.e.\ data structure plus constraints) can be extracted from the Ampersand script.
 
@@
    To make the case for zero downtime, consider this problem:
-   Suppose we have an invariant, $u$, in the {\em desired system}, which is absent in the {\em existing system}.
-   And suppose the data in the existing system does not satisfy $u$.
-   When spinning up the desired system on the data set of the existing system, $u$ must be satisfied.
-   If $u$ is an enforcement rule, the desired system will restore it automatically.
-   But otherwise, the system will refuse to spin up.
-   In many cases, clearing out all violations before spinning up the desired system might take too long, while new violations may arise in the meantime.
-   So, the problem lies with invariants that are not enforced automatically.
-   To solve this problem, we must implement such invariants initially as a business constraint,
-   letting users restore invariance while the new system is running.
+   Suppose we have an invariant, $u$, in the {\em desired system}, which is not required in the {\em existing system}.
+   In the sequel, let us call this a {\em new invariant}.
+   Now, suppose the data in the existing system does not satisfy $u$.
+   If $u$ is an enforced invariant, the desired system will restore it automatically.
+   But if it is a blocking invariant, the desired system cannot spin up because all of its invariants must be satisfied.
+   To avoid downtime, we must implement new invariants initially as a business constraint,
+   to let users do the work.
+
    For this purpose, we must define an intermediate system, the \define{migration system},
-   which implements the invariants of the desired system as business constraints, unless they are enforcement rules.
+   which implements the blocking invariants of the desired system as business constraints.
    This migration system must also ensure that every violation that is fixed will not return as a violation.
@@
 \item users must experience zero downtime, to enable more frequent SCDMs.
-\item users must be able to fix business constraints, so we need an intermediate ``migration system'' to deliver zero downtime.
+\item users must be able to fix invariants in the new system.
+   So, we need an intermediate ``migration system'' that implements blocking constraints of the desired system as business constraints,
+   in order to deliver zero downtime.
 \item the number of violations that users must fix is finite and decreases (monotonically) over time,
-   to ensure that the migration system can be removed after the migration is done.
+   to ensure that the migration system does not need to live ever after.
 \end{enumerate}
-
-   Ampersand is a compiler that generates information systems from a script.
-   It uses a form of heterogeneous relation algebra,
-   which works with relations in a way similar to Alloy~\cite{Alloy2006}.
-   A system is then described by a set of rules (aka constraint / invariant), which the system keeps satisfied.
-   As a result, all invariants are explicit in the source code of a system specified in Ampersand.
-
-   This, and the absence of imperative code in an Ampersand script, makes Ampersand a suitable platform on which to implement the theory.
-   It also allows us to be explicit about ``preserving the meaning as much as possible''.
-   An Ampersand script contains just enough information to generate a complete system,
-   which means that a classical database schema (i.e.\ data structure plus constraints) can be extracted from the Ampersand script.
 
@@
    %
-   In practice, data migrations are done by deploying the desired system and the existing system side-by-side,
+   In practice, data migrations typically deploy the existing system and the desired system side-by-side,
    while transferring data in a controlled fashion, as shown in Figure~\ref{fig:migration phase}.
@@
    To automate the migration as much as possible and to achieve zero downtime,
-   we also deploy a migration system.
-   The migration system has its own schema.
+   we must deploy a third system: the migration system.
+   This system has its own schema.
    It comprises the schemas of both the existing system and the desired system,
    so the migration system can copy the data from the existing to the desired system.
-   Some of the migration work, however, cannot be automated.
-   A migration engineer can minimize manual work by minimizing the size of the increments.
-   But data pollution, new business rules, or known issues in the existing system
+   It uses enforced invariants to copy the data and to resolve some forms of data pollution.
+   Not all migration work, however, can be automated.
+   Data pollution, new business rules, or known issues in the existing system
    may occasionally require tailoring the migration script to specific needs
    that require action of users in production.
-   In Ampersand, she can use business constraints to implement these needs.
+   For that purpose, the migration engineer specifies business constraints in the migration system.
+   In Ampersand, a developer can use business constraints to implement such needs.
 
-   A data migration has the following distinct moments\footnote{Completion can also be done gradually, but that is outside the scope of this paper.}:
-\begin{enumerate}
-   \item the moment of transition (the MoT), after which users experience changed functionality;
-   \item the moment of completion (the MoC), after which the migration engineer removes the migration system and the old system, including its data.
-\end{enumerate}
+   Before a migration starts we assume that the existing system is up and running and all of its invariants are satisfied.
+   The ingress is directing traffic to the existing system.
+   The data migration has three distinct steps:
+\begin{description}
+\item[Pre-deploy]
+      The migration engineer deploys both the migration system and the desired system with their initial data, to allow both systems to satisfy their initial invariants before going live.
+      The existing data is still in the existing system and the ingress still directs all traffic to the existing system.
+      So, users notice no difference yet.
+      The migration system starts copying data from the existing system.
+\item[moment of transition (MoT)]
+      After the migration system is done copying data, the migration engineer switches all traffic to the migration system.
+      From this moment on, users experience the changed functionality on the existing data.
+      However, the invariants of the desired system may not yet be satisfied.
+      So, users are working on the migration system, which looks like the desired system but relaxes some of its invariants.
+      Since the existing system receives no more traffic, its activity will come to a halt and its data will become static.
+      The migration system stays active until all invariants of the desired system are satisfied and it can take over the work from the migration system.
+\item[moment of completion (MoC)]
+      Once the invariants of the desired system are satisfied, the migration engineer switches all traffic to the desired system.
+      After that, she can safely remove the migration system and the existing system, including their data.
+\end{description}
 
-   Before the migration, the data in the existing system satisfies its invariants (by definition).
-   When the migration system compiles correctly, it satisfies its own invariants,
-   so it can be deployed in production.
-   That is the MoT.
-   The task of the migration system is to satisfy the invariants of the desired system
-   so that it can be deployed in production.
-   For some of these invariants, enforcement rules can satisfy them automatically.
-   There may be invariants, however, that require human interaction to resolve.
-   Once all invariants of the desired system are satisfied, the migration is ready for the MoC.
-   So, the schema of the migration system must contain the rules to satisfy all invariants of the desired system.
-
-   During the migration phase, transactions in the existing system are allowed because the migration engine will transport them to the desired system.
-   However, after the MoT there must be no new changes in the existing system,
-   to allow the system to finish transactions while the desired system is up and running.
-
-\begin{figure}[bht]
-   \begin{center}
-     \includegraphics[scale=.35]{figures/datamigration-Post-migration.png}
-   \end{center}
-\caption{The system after the data migration}
-\label{fig:post-migration}
-\end{figure}
+   Transactions in the existing system that happen during the time that the migration system is copying data cause no problem,
+   because these changes are copied by the migration system too.
+   However, after the MoT there must be no new changes in the existing system
+   to avoid violations of new invariants that the migration system has already fixed.
 
@@
 
-\section{Terminology}
-\label{sct:Terminology}
+\section{Definitions}
+\label{sct:Definitions}
    An {\em information system} is a combination of data set, schema, and functionality.
    For the purpose of this paper, we ignore functionality captured in user interfaces because it does not impact the migration.
-   Section~\ref{sct:Data sets} desribes data sets. Schemas are treated in section~\ref{sct:Schemas}.
+   Section~\ref{sct:Data sets} describes how we define data sets. Section~\ref{sct:Violations} defines violations. Schemas are treated in section~\ref{sct:Schemas}.
    Then section~\ref{sct:Information Systems} defines information systems.
@@
    We overload the notation $\popF{}$ so we can use it on concepts $\popF{C}:\Dataset\rightarrow\powerset{\Atoms}$
-   and expressions. We also define the difference of populations, both for relations and concepts:
+   and expressions. We also define the difference of populations, in equation~\ref{eqn:pop-expr}, both for relations and concepts:
 \begin{align}
@@
 \end{equation}
-   We represent an enforcement rule $u$ as a pair $\pair{\declare{n}{A}{B}}{\violC{u}}$,
-   where relation $\declare{n}{A}{B}$ is used to restore invariance of the enforcement rule.
-   The enforcement rule must be well-typed:
+   We represent an enforced invariant $u$ as a pair $\pair{\declare{n}{A}{B}}{\violC{u}}$,
+   where relation $\declare{n}{A}{B}$ is used to restore invariance of the enforced invariant.
+   The enforced invariant must be well-typed:
 \begin{equation}
@@
 \end{equation}
-   Besides, an enforcement rule must be automatically restorable by adding the violations to relation $\declare{n}{A}{B}$:
+   Besides, an enforced invariant must be automatically restorable by adding the violations to relation $\declare{n}{A}{B}$:
 \begin{equation}
@@
 \end{equation}
-   The language Ampersand has more types of enforcement rules than just this one,
+   (TODO: define $\cup$ for data sets.)
+   The language Ampersand has more types of enforced invariants than just this one,
    but to define this one only is sufficient for this paper.
@@
    We assume that a software engineer defines a schema on design time, and a compiler checks whether the semantics are consistent.
-   Error messages from the compiler are prohibitive for generating code,
-   which prevents a substantial class of mistakes to ever reach end users.
+   Errors the compiler detects are prohibitive for generating code,
+   to prevent a substantial class of mistakes to ever reach end users.
 
-   We describe a schema $\schema$ via a tuple $\quadruple{\concepts}{\rels}{\rules}{\enforces}$,
+   We describe a schema $\schema$ as a tuple $\quadruple{\concepts}{\rels}{\rules}{\enforces}$,
    in which $\concepts\subseteq \Concepts$ is a finite set of concepts,
@@
    $\rules\subseteq \Triple{(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}})}{\Concepts}{\Concepts}$ is a finite set of invariants,
-   and $\enforces\subseteq \Pair{\Rels}{\left(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}\right)}$ is a finite set of enforcement rules.
+   and $\enforces\subseteq \Pair{\Rels}{\left(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}\right)}$ is a finite set of enforced invariants.
 
    Every invariant is an element of an infinite set called $\Rules$.
-   In this paper, variables $u$ and $v$ represent rules.
-   In Ampersand, rules are well-typed, just like relations:
-   Every violation $\pair{a}{b}$ of a invariant $\triple{\violC{u}}{A}{B}$ must satisfy $a\instance A$ and $b\instance B$.
-   % For every data set $\dataset$, the set $\viol{u}{\dataset}$ is a subset of $\pop{A}{\dataset}\times\pop{B}{\dataset}$.
-   If there are no violations, that is: $\viol{u}{\dataset} = \emptyset{}$, then we say that $u$ is satisfied.
+   In this paper, variables $u$ and $v$ represent constraints, both invariants and business constraints.
 
    For a different kind of constraint, the system is told what to do in order to fix it.
-   Such constraints are called enforcement rules.
-   We use just one type of enforcement rule\footnote{Ampersand has several types of enforcement rules.} in this work.
+   Such constraints are called enforced invariants.
+   We use just one type of enforced invariant\footnote{Ampersand has several types of enforced invariants.} in this work.
    Its syntax in Ampersand is as follows:
@@
 
-   Adding an enforcement rule to the script of a system specifies a function $\violC{u} : \Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}$
+   Adding an enforced invariant to the script of a system specifies a function $\violC{u} : \Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}$
    with $\viol{u}{\dataset} = \pop{\id{term} - r}{\dataset}$.
@@
 
-   Formally, enforcement rules are pairs with $u$ a function and $r$ a relation, written $r \mapsfrom t$ or equivalently $r \mapsfrom \lambda \dataset.~ t(\dataset)$.
-   The set of enforcement rules in a schema is $\enforces$.
+   Formally, enforced invariants are pairs with $u$ a function and $r$ a relation, written $r \mapsfrom t$ or equivalently $r \mapsfrom \lambda \dataset.~ t(\dataset)$.
+   The set of enforced invariants in a schema is $\enforces$.
 
@@
    Requirements~\ref{eqn:rule-has-type} and~\ref{eqn:rule-well-typed} ensure that violations in a rule correspond to the type of that rule. 
-   Requirements~\ref{eqn:enforcement-has-type} and~\ref{eqn:enforcement-well-typed} ensure the same for enforcement rules. 
+   Requirements~\ref{eqn:enforcement-has-type} and~\ref{eqn:enforcement-well-typed} ensure the same for enforced invariants. 
    The type system of Ampersand~\cite{vdWoude2011} ensures that every system it generates satisfies requirements~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-well-typed}.
@@
    \item $\schema=\quadruple{\concepts}{\rels}{\rules}{\enforces}$ is a schema (satisfies equation~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-well-typed}), we write $\concepts_\infsys = \concepts$, $\rels_\infsys = \rels$, $\rules_\infsys=\rules$, and $\enforces_\infsys=\enforces$;
-   \item All rules and enforcement rules are satisfied:
+   \item All rules and enforced invariants are satisfied:
    \begin{align}
@@
 
-   We use inserting events to model the effect of enforcement rules.
+   We use inserting events to model the effect of enforced invariants.
 %   If $R$ is a set of relations that covers those that occur in the rule, then changes to an system $\infsys$ can be described by the event $\infsys \xrightarrow[\{r\}]{R} \infsys'$.
@@
    \end{align}
-\item We create enforcement rules to copy equal relations, corresponding to equation~\ref{extraRelation} through~\ref{fill copies with new.isPartOf} in Section~\ref{sec:loading}:
+\item We create enforced invariants to copy equal relations, corresponding to equation~\ref{extraRelation} through~\ref{fill copies with new.isPartOf} in Section~\ref{sec:loading}:
    \begin{align}
@@
 \subsection{Validation}
-   We first observe that $\migrsys = \pair{\pair{\triples_\migrsys}{\instance_\migrsys}}{\quadruple{\concepts_\migrsys}{\rels_\migrsys}{\rules_\migrsys}{\enforces_\migrsys}}$ is a system after application of the enforcement rules:
-   the only rules are those introduced in equation~\ref{eqn:migrsysrules}, and enforcement rules from equation~\ref{eqn:enforceForRules} ensure that $\popF{{\tt viols}_u} = u$ initially, while $\popF{{\tt fixed}_u}$ is empty.
-   Hence, the MoT can happen after the enforcement rules have been applied, giving us $\migrsys'$.
+   We first observe that $\migrsys = \pair{\pair{\triples_\migrsys}{\instance_\migrsys}}{\quadruple{\concepts_\migrsys}{\rels_\migrsys}{\rules_\migrsys}{\enforces_\migrsys}}$ is a system after application of the enforced invariants:
+   the only rules are those introduced in equation~\ref{eqn:migrsysrules}, and enforced invariants from equation~\ref{eqn:enforceForRules} ensure that $\popF{{\tt viols}_u} = u$ initially, while $\popF{{\tt fixed}_u}$ is empty.
+   Hence, the MoT can happen after the enforced invariants have been applied, giving us $\migrsys'$.
    The event $\migrsys \xrightarrow[\rels_2 \cup \rels_3]{\overrightarrow\rels_{\schema'}}\migrsys'$ represents the MoT.

### 2023-10-15 2022Migration/articleMigrationFACS.tex
WIP

@@
 \newcommand{\violC}[1]{\id{viol}_{#1}}
-\newcommand{\sign}[1]{\id{sign}(#1)}
+\newcommand{\sign}[1]{\id{sign}_{#1}}
+\newcommand{\enfRel}[1]{\id{enf}_{#1}}
+\newcommand{\signature}[2]{\langle{#1},{#2}\rangle}
 \newcommand{\powerset}[1]{\cal{P}\{#1\}}
@@
    \begin{definition}[Data set]
-   A data set $\dataset$ is a tuple $\pair{\triples}{\instance}$ with finite $\triples \subseteq {\Triple{\Atoms}{\Rels}{\Atoms}}$ and $\instance \subseteq  {\Pair{\Atoms}{\Concepts}}$ that satisfies:
+   A data set $\dataset$ is a tuple $\pair{\triples}{\instance}$ with finite $\triples \subseteq {\Triple{\Atoms}{\Rels}{\Atoms}}$ and $\instance\subseteq\Pair{\Atoms}{\Concepts}$ that satisfies:
 \begin{eqnarray}
@@
    To satisfy a constraint $u$ means that it is true for all triples in the data set.
-   Let us assume that we can represent each violation of a constraint $u$ as a pair
-   and let the function $\violC{u}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$ produce these violations.
+   Let us assume that we can represent each violation of a constraint $u$ as a pair.
+   Let the function $\violC{u}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$ produce these violations
+   and let the function $\sign{u}:\Pair{\Concepts}{\Concepts}$ tell their signature:
+\begin{equation}
+   \pair{a}{b}\in\viol{u}{\dataset}\ \wedge\ \sign{u}=\pair{A}{B}\Rightarrow\ a\instance A\ \wedge\ b\instance B
+\end{equation}
    Note that $\viol{u}{\dataset}=\emptyset$ means that $\dataset$ satisfies $u$.
-   Now we can represent an invariant $u$ as a triple $\triple{\violC{u}}{A}{B}$.
-   It must be well-typed, just like a relation:
+
+   In case $u$ is an enforced invariant,
+   the system will keep it satisfied by adding the violations to relation $\declare{n}{A}{B}$:
 \begin{equation}
-   \pair{a}{b}\in\viol{\triple{\violC{u}}{A}{B}}{\dataset}\ \Rightarrow\ a\instance A\ \wedge\ b\instance B
+   \viol{u}{\pair{\triples\cup\{\triple{a}{\declare{n}{A}{B}}{b}\mid\pair{a}{b}\in\viol{u}{\dataset}\}}{\instance\cup\instance'}}=\emptyset
 \end{equation}
-   We represent an enforced invariant $u$ as a pair $\pair{\declare{n}{A}{B}}{\violC{u}}$,
-   where relation $\declare{n}{A}{B}$ is used to restore invariance of the enforced invariant.
-   The enforced invariant must be well-typed:
+   Let us call $\declare{n}{A}{B}$ the {enforced relation} of $u$, defined by $\enfRel$:
 \begin{equation}
-   \pair{a}{b}\in\viol{\pair{\declare{n}{A}{B}}{\violC{u}}}{\dataset}\ \Rightarrow\ a\instance A\ \wedge\ b\instance B
+   \enfRel{u}=\declare{n}{A}{B}
 \end{equation}
-   Besides, an enforced invariant must be automatically restorable by adding the violations to relation $\declare{n}{A}{B}$:
-\begin{equation}
-   \viol{u}{\dataset\cup\{\triple{a}{\declare{n}{A}{B}}{b}\mid\pair{a}{b}\in\viol{u}{\dataset}\}}=\emptyset
-\end{equation}
-   (TODO: define $\cup$ for data sets.)
    The language Ampersand has more types of enforced invariants than just this one,

### 2023-10-16 2022Migration/articleMigrationFACS.tex
Adapt schemas (sct. 3.3) to new terminology

@@
 \newcommand{\vertices}{N}
-\newcommand{\enforces}{\mathcal{E}}
+\newcommand{\transactions}{\mathcal{E}}
 \newcommand{\rules}{\mathcal{U}}
-\newcommand{\Rules}{\mathbb{U}}
+\newcommand{\Constraints}{\mathbb{U}}
 \newcommand{\specrules}{\mathcal{S}}
@@
 
-   We assume that the semantics of the data is ensured by means of constraints.
+   We assume that constraints implement the semantics of the data.
    Figure~\ref{fig:pre-migration} depicts them as rules.
    Some constraints must be satisfied through human intervention, while others are being satisfied by the system.
-   So, we distinguish between two types of constraint:
+   So, we distinguish different kinds of constraint:
 \begin{enumerate}
@@
    When it is violated, the system produces an error message and refuses the transaction.
-\item Enforced invariant\\
-   An enforced invariant corresponds to the classical database transaction.
-   As soon as the data violates an enforced invariant, the system restores it automatically without human intervention,
-   so the outside world will never notice that it was violated.
+\item Transaction\\
+   The classical database transaction can be seen as an invariant,
+   which the system keeps satisfied by adding or deleting triples to the dataset.
+   As soon as the data violates a transaction, the system restores it without human intervention.
+   So, the outside world experiences a constraint that is always true, i.e.~an invariant.
 \item Business constraint\\
@@
    Example: ``An authorized manager has to sign every purchase order.''
-   A business constraint is not true all the time, so we do not consider it to be an invariant%
-\footnote{The business may impose a business constraint, but that doesn't make it satisfied.
-   Every violation requires some form of human action to satisfy the business constraint (e.g. "sign the purchase order").}.
+   Every violation requires some form of human action to satisfy the business constraint (e.g. "sign the purchase order").
+   That takes some time, during which the constraint is violated.
+   So, we do not consider business constraints to be invariant.
 \end{enumerate}
    Summarizing, in our notion of information systems, concepts, relations, and constraints carry the business semantics.
+   Of the three types of constraint, only blocking invariants and transactions are invariants.
 
@@
    The absence of imperative code in an Ampersand script makes it easier to reason about the system.
+   The type system of Ampersand~\cite{vdWoude2011} features static typing,
+   which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.
    Constraints carry the business semantics, so they allow us to be explicit about ``preserving the meaning as much as possible''.
@@
    To make the case for zero downtime, consider this problem:
-   Suppose we have an invariant, $u$, in the {\em desired system}, which is not required in the {\em existing system}.
+   Suppose we have an invariant, $u$, in the {\em desired system}, which is not part of the {\em existing system}.
    In the sequel, let us call this a {\em new invariant}.
    Now, suppose the data in the existing system does not satisfy $u$.
-   If $u$ is an enforced invariant, the desired system will restore it automatically.
+   If $u$ is a transaction, the desired system will restore it automatically.
    But if it is a blocking invariant, the desired system cannot spin up because all of its invariants must be satisfied.
-   To avoid downtime, we must implement new invariants initially as a business constraint,
+   To avoid downtime, we must implement new blocking invariants initially as a business constraint,
    to let users do the work.
 
-   For this purpose, we must define an intermediate system, the \define{migration system},
-   which implements the blocking invariants of the desired system as business constraints.
-   This migration system must also ensure that every violation that is fixed will not return as a violation.
-   In this way, the business constraint eventually becomes an invariant, satisfying the specification of the desired system.
+   For this purpose, we define an intermediate system, the \define{migration system},
+   which contains all concepts, relations, and constraints of the desired system.
+   However, it implements the blocking invariants of the desired system as business constraints.
+   This migration system must also ensure that every violation that is fixed is blocked from recurring.
+   In this way, the business constraint gradually turns into a blocking invariant, satisfying the specification of the desired system.
    Since the number of violations is finite, the business can resolve these violations in finite time.
-   New violations that occur are under control of the desired system,
-   so the business experiences the migration system as the desired system.
    In this way, the migration system bridges the gap and users get a zero downtime SCDM.
@@
    so the migration system can copy the data from the existing to the desired system.
-   It uses enforced invariants to copy the data and to resolve some forms of data pollution.
+   It uses transactions to copy the data and to resolve some forms of data pollution.
    Not all migration work, however, can be automated.
@@
 \item[Pre-deploy]
-      The migration engineer deploys both the migration system and the desired system with their initial data, to allow both systems to satisfy their initial invariants before going live.
+      The migration engineer deploys both the migration system and the desired system with their initial data,
+      to allow both systems to satisfy their initial invariants before going live.
       The existing data is still in the existing system and the ingress still directs all traffic to the existing system.
@@
 \item[moment of transition (MoT)]
-      After the migration system is done copying data, the migration engineer switches all traffic to the migration system.
-      From this moment on, users experience the changed functionality on the existing data.
-      However, the invariants of the desired system may not yet be satisfied.
-      So, users are working on the migration system, which looks like the desired system but relaxes some of its invariants.
+      After the migration system is done copying data, the migration engineer switches all traffic to the migration system
+      and deploys the new functionality at the same time.
+      This is the moment users will notice the difference.
+      Even though the migration system may look like the desired system in the eyes of an average user,
+      it relaxes the blocking invariants and keeps the system operational.
       Since the existing system receives no more traffic, its activity will come to a halt and its data will become static.
@@
       Once the invariants of the desired system are satisfied, the migration engineer switches all traffic to the desired system.
-      After that, she can safely remove the migration system and the existing system, including their data.
+      After that, she can safely remove both the migration system and the existing system, including their data.
 \end{description}
@@
    Transactions in the existing system that happen during the time that the migration system is copying data cause no problem,
-   because these changes are copied by the migration system too.
+   because their changes are copied by the migration system too.
    However, after the MoT there must be no new changes in the existing system
@@
    For the purpose of this paper, we ignore functionality captured in user interfaces because it does not impact the migration.
-   Section~\ref{sct:Data sets} describes how we define data sets. Section~\ref{sct:Violations} defines violations. Schemas are treated in section~\ref{sct:Schemas}.
+   Section~\ref{sct:Data sets} describes how we define data sets.
+   Section~\ref{sct:Constraints} defines constraints and their violations.
+   Schemas are treated in section~\ref{sct:Schemas}.
    Then section~\ref{sct:Information Systems} defines information systems.
@@
 
-\subsection{Violations}
-\label{sct:Violations}
-   To satisfy a constraint $u$ means that it is true for all triples in the data set.
-   Let us assume that we can represent each violation of a constraint $u$ as a pair.
-   Let the function $\violC{u}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$ produce these violations
-   and let the function $\sign{u}:\Pair{\Concepts}{\Concepts}$ tell their signature:
+\subsection{Constraints}
+\label{sct:Constraints}
+   Every constraint is an element of an infinite set called $\Constraints$.
+   In this paper, variables $u$ and $v$ represent all three types of constraints.
+   We say that a constraint, $u$, is satisfied when $u$ is true for all triples in the data set.
+   For every constraint $u$, function $\violC{u}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$ produces the violations of $u$,
+   and $\sign{u}:\Pair{\Concepts}{\Concepts}$ yields the signature of $u$.
+   The definition of $\violC{u}$ implies the assumption that we represent each violation as a pair.
+   Every constraint must satisfy:
 \begin{equation}
-   \pair{a}{b}\in\viol{u}{\dataset}\ \wedge\ \sign{u}=\pair{A}{B}\Rightarrow\ a\instance A\ \wedge\ b\instance B
+   \pair{a}{b}\in\viol{u}{\dataset}\ \wedge\ \sign{u}=\pair{A}{B}\ \Rightarrow\ a\instance A\ \wedge\ b\instance B
+\label{eqn:wellTypedViolation}
 \end{equation}
@@
 
-   In case $u$ is an enforced invariant,
-   the system will keep it satisfied by adding the violations to relation $\declare{n}{A}{B}$:
+   In case $u$ is a transaction,
+   the system will keep it satisfied by adding the violations to a specific relation $\declare{n}{A}{B}$.
 \begin{equation}
-   \viol{u}{\pair{\triples\cup\{\triple{a}{\declare{n}{A}{B}}{b}\mid\pair{a}{b}\in\viol{u}{\dataset}\}}{\instance\cup\instance'}}=\emptyset
+   \begin{array}[t]{l}
+      \viol{u}{\pair{\triples\cup\{\triple{a}{\declare{n}{A}{B}}{b}\mid\pair{a}{b}\in\viol{u}{\dataset}\}}{\instance\cup\instance'}}=\emptyset\\
+      \text{\bf where}\ \instance'=\{ \pair{a}{A}\mid\pair{a}{b}\in\viol{u}{\dataset}\}\cup\{ \pair{b}{B}\mid\pair{a}{b}\in\viol{u}{\dataset}\}
+   \end{array}
+\label{eqn:transaction}
 \end{equation}
-   Let us call $\declare{n}{A}{B}$ the {enforced relation} of $u$, defined by $\enfRel$:
+   It is obvious that not every conceivable constraint can satisfy this equation.
+   So, we assume that the compiler restricts the set of transactions to those that satisfy equation~\ref{eqn:transaction}.
+   As $\declare{n}{A}{B}$ is specific for $u$, we can write $\enfRel{u}$ for it.
+   We call this the {\em enforced relation} of transaction $u$:
 \begin{equation}
@@
 \end{equation}
-   The language Ampersand has more types of enforced invariants than just this one,
+   The language Ampersand has more types of transactions than just this one,
    but to define this one only is sufficient for this paper.
@@
    Schemas serve to capture the semantics of a system~\cite{Spivak2012}.
-   They define concepts, relations, and invariants.
+   They define concepts, relations, and constraints.
    We assume that a software engineer defines a schema on design time, and a compiler checks whether the semantics are consistent.
@@
 
-   We describe a schema $\schema$ as a tuple $\quadruple{\concepts}{\rels}{\rules}{\enforces}$,
+   We describe a schema $\schema$ as a tuple $\quadruple{\concepts}{\rels}{\rules}{\transactions}$,
    in which $\concepts\subseteq \Concepts$ is a finite set of concepts,
-   $\rels\subseteq \Rels$ is a finite set of relations,
-   $\rules\subseteq \Triple{(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}})}{\Concepts}{\Concepts}$ is a finite set of invariants,
-   and $\enforces\subseteq \Pair{\Rels}{\left(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}\right)}$ is a finite set of enforced invariants.
+   $\rels\subseteq\Rels$ is a finite set of relations,
+   $\rules\subseteq\Constraints$ is a finite set of blocking invariants,
+   and $\transactions\subseteq\Constraints$ is a finite set of transactions.
 
-   Every invariant is an element of an infinite set called $\Rules$.
-   In this paper, variables $u$ and $v$ represent constraints, both invariants and business constraints.
-
-   For a different kind of constraint, the system is told what to do in order to fix it.
-   Such constraints are called enforced invariants.
-   We use just one type of enforced invariant\footnote{Ampersand has several types of enforced invariants.} in this work.
-   Its syntax in Ampersand is as follows:
-\begin{align}
-      \text{\tt ENFORCE}\ r\ \text{\tt >:}\ \id{term}\label{enforce ins}
-\end{align}
-
-   Adding an enforced invariant to the script of a system specifies a function $\violC{u} : \Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}$
-   with $\viol{u}{\dataset} = \pop{\id{term} - r}{\dataset}$.
-   The system ensures that the rule is satisfied by adding a triple $\triple{x}{r}{y}$ for each $\pair{x}{y} \in \viol{u}{\dataset}$.
-   Thus the system maintains $\pop{r}{\dataset} \supseteq \pop{\id{term}}{\dataset}$ as an invariant.
-
-   Formally, enforced invariants are pairs with $u$ a function and $r$ a relation, written $r \mapsfrom t$ or equivalently $r \mapsfrom \lambda \dataset.~ t(\dataset)$.
-   The set of enforced invariants in a schema is $\enforces$.
+   Let us denote a transaction as $r \mapsfrom u$ or equivalently $r \mapsfrom \lambda \dataset.~ u(\dataset)$,
+   in which $r=\enfRel{u}$.
+   The set of transactions in a schema is $\transactions$.
 
    \begin{definition}[Schema]
-   A schema is a tuple $\quadruple{\concepts}{\rels}{\rules}{\enforces}$ that satisfies:
+   A schema is a tuple $\quadruple{\concepts}{\rels}{\rules}{\transactions}$ that satisfies:
 \begin{align}
@@
    \label{eqn:relationsIntroduceConcepts}\\
-   \triple{\violC{u}}{A}{B}\in\rules&~\Rightarrow~A\in\concepts\,\wedge\, B\in\concepts\label{eqn:rule-has-type}\\
-   \triple{\violC{u}}{A}{B}\in\rules&~\Rightarrow~\forall\dataset.\ \viol{u}{\dataset}\subseteq\Pair{\pop{A}{\dataset}}{\pop{B}{\dataset}}\label{eqn:rule-well-typed}\\
-   (\declare{n}{A}{B}\mapsfrom t)\in\mathcal\enforces&~\Rightarrow~ \declare{n}{A}{B}\in \rels\label{eqn:enforcement-has-type}\\
-   (\declare{n}{A}{B}\mapsfrom t)\in\mathcal\enforces&~\Rightarrow~\forall\dataset.\ \pop{t}{\dataset}\subseteq\Pair{\pop{A}{\dataset}}{\pop{B}{\dataset}}\label{eqn:enforcement-well-typed}
+   u\in\rules\cup\transactions\ \wedge\ \sign{u}=\pair{A}{B}&~\Rightarrow~A\in\concepts\,\wedge\, B\in\concepts\label{eqn:invariant-has-type}\\
+   u\in\rules\cup\transactions\wedge\sign{u}=\pair{A}{B}\wedge\pair{a}{b}\in\viol{u}{\dataset}&~\Rightarrow~a\instance A\ \wedge\ b\instance B\label{eqn:rule-has-type}\\
+   (\declare{n}{A}{B}\mapsfrom t)\in\mathcal\transactions&~\Rightarrow~ \declare{n}{A}{B}\in \rels\label{eqn:enforcement-has-type}
 \end{align}
    \end{definition}
-   Requirement~\ref{eqn:relationsIntroduceConcepts} ensures that concepts mentioned in a relation are defined in the schema.
-   This is also known as static typing, which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.
-   Requirements~\ref{eqn:rule-has-type} and~\ref{eqn:rule-well-typed} ensure that violations in a rule correspond to the type of that rule. 
-   Requirements~\ref{eqn:enforcement-has-type} and~\ref{eqn:enforcement-well-typed} ensure the same for enforced invariants. 
-   The type system of Ampersand~\cite{vdWoude2011} ensures that every system it generates satisfies requirements~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-well-typed}.
-   When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, $\rules_{\schema}$, and $\enforces_{\schema}$
-   for $\concepts$, $\rels$, $\rules$, and $\enforces$ corresponding to $\schema = \quadruple{\concepts}{\rels}{\rules}{\enforces}$.
+   Requirements~\ref{eqn:relationsIntroduceConcepts} and~\ref{eqn:invariant-has-type} ensure that concepts mentioned in relations and in the signature of rules are part of the schema.
+   Requirement~\ref{eqn:rule-has-type} ensures that violations in a rule correspond to the type of that rule. 
+   Requirement~\ref{eqn:enforcement-has-type} ensures the enforced relation of a transaction is declared in the schema. 
+   When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, $\rules_{\schema}$, and $\transactions_{\schema}$
+   for $\concepts$, $\rels$, $\rules$, and $\transactions$ corresponding to $\schema = \quadruple{\concepts}{\rels}{\rules}{\transactions}$.
 
@@
    \item $\dataset=\pair{\triples}{\instance}$ is a data set (satisfies equation~\ref{eqn:wellTypedEdge}), we write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$;
-   \item $\schema=\quadruple{\concepts}{\rels}{\rules}{\enforces}$ is a schema (satisfies equation~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-well-typed}), we write $\concepts_\infsys = \concepts$, $\rels_\infsys = \rels$, $\rules_\infsys=\rules$, and $\enforces_\infsys=\enforces$;
-   \item All rules and enforced invariants are satisfied:
+   \item $\schema=\quadruple{\concepts}{\rels}{\rules}{\transactions}$ is a schema (satisfies equation~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}), we write $\concepts_\infsys = \concepts$, $\rels_\infsys = \rels$, $\rules_\infsys=\rules$, and $\transactions_\infsys=\transactions$;
+   \item All rules and transactions are satisfied:
    \begin{align}
    \forall u\in\rules&.~\viol{u}{\dataset}=\emptyset\\
-   \forall (r\mapsfrom t)\in\enforces&.~\viol{u}{\dataset} - \pop{r}{\dataset}=\emptyset
+   \forall (r\mapsfrom t)\in\transactions&.~\viol{u}{\dataset} - \pop{r}{\dataset}=\emptyset
    \label{eqn:satisfaction}
@@
 
-   We use inserting events to model the effect of enforced invariants.
+   We use inserting events to model the effect of transactions.
 %   If $R$ is a set of relations that covers those that occur in the rule, then changes to an system $\infsys$ can be described by the event $\infsys \xrightarrow[\{r\}]{R} \infsys'$.
@@
    \rules&=&\{ {\tt EnrollRule}\}\\
-   \enforces&=&\emptyset
+   \transactions&=&\emptyset
   \end{array}
@@
    % The (nonempty) data set satisfies this rule, which meets requirement~\ref{eqn:consistentRules}.
-   So, the schema $\schema=\quadruple{\concepts}{\rels}{\rules}{\enforces}$ satisfies all requirements from section~\ref{sct:Schemas}.
+   So, the schema $\schema=\quadruple{\concepts}{\rels}{\rules}{\transactions}$ satisfies all requirements from section~\ref{sct:Schemas}.
 
@@
    \end{align}
-\item We create enforced invariants to copy equal relations, corresponding to equation~\ref{extraRelation} through~\ref{fill copies with new.isPartOf} in Section~\ref{sec:loading}:
+\item We create transactions to copy equal relations, corresponding to equation~\ref{extraRelation} through~\ref{fill copies with new.isPartOf} in Section~\ref{sec:loading}:
    \begin{align}
@@
    \rels_\theenumi &= \{\widetilde{r}\mid r \in \rels_\schema \cap \rels_{\schema'}\}\\
-   \enforces_\theenumi &= \{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r} - \widetilde{r}} \mid \widetilde{r} \in \rels_2\} \cup \{\widetilde{r}\mapsfrom \popF{\overrightarrow{r}} \mid \widetilde{r} \in \rels_2\}
+   \transactions_\theenumi &= \{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r} - \widetilde{r}} \mid \widetilde{r} \in \rels_2\} \cup \{\widetilde{r}\mapsfrom \popF{\overrightarrow{r}} \mid \widetilde{r} \in \rels_2\}
    \end{align}
@@
    \rules_\migrsys &= \{\lambda {\dataset'}.~ \viol{u}{\overrightarrow{\dataset'}} - \pop{{\tt viols}_u - {\tt fixed}_u}{\dataset'} \mid u\in\rules_{\schema'}\}\label{eqn:migrsysrules}\\
-   \enforces_{\theenumi} &=\{{\tt viols}_u \mapsfrom \lambda \dataset'.~ \viol{u}{\overleftarrow{\dataset'}}\} \cup
+   \transactions_{\theenumi} &=\{{\tt viols}_u \mapsfrom \lambda \dataset'.~ \viol{u}{\overleftarrow{\dataset'}}\} \cup
     \{{\tt fixed}_u \mapsfrom \lambda \dataset'.~ {\tt viols}_u - \viol{u}{\overrightarrow{\dataset'}}\}\label{eqn:enforceForRules}
@@
    %\rules_\migrsys &= \rules_1\cup \rules_3\\
-   \enforces_\migrsys &= \enforces_1\cup\enforces_2\cup\enforces_3
+   \transactions_\migrsys &= \transactions_1\cup\transactions_2\cup\transactions_3
    \end{align}
@@
 \subsection{Validation}
-   We first observe that $\migrsys = \pair{\pair{\triples_\migrsys}{\instance_\migrsys}}{\quadruple{\concepts_\migrsys}{\rels_\migrsys}{\rules_\migrsys}{\enforces_\migrsys}}$ is a system after application of the enforced invariants:
-   the only rules are those introduced in equation~\ref{eqn:migrsysrules}, and enforced invariants from equation~\ref{eqn:enforceForRules} ensure that $\popF{{\tt viols}_u} = u$ initially, while $\popF{{\tt fixed}_u}$ is empty.
-   Hence, the MoT can happen after the enforced invariants have been applied, giving us $\migrsys'$.
+   We first observe that $\migrsys = \pair{\pair{\triples_\migrsys}{\instance_\migrsys}}{\quadruple{\concepts_\migrsys}{\rels_\migrsys}{\rules_\migrsys}{\transactions_\migrsys}}$ is a system after application of the transactions:
+   the only rules are those introduced in equation~\ref{eqn:migrsysrules}, and transactions from equation~\ref{eqn:enforceForRules} ensure that $\popF{{\tt viols}_u} = u$ initially, while $\popF{{\tt fixed}_u}$ is empty.
+   Hence, the MoT can happen after the transactions have been applied, giving us $\migrsys'$.
    The event $\migrsys \xrightarrow[\rels_2 \cup \rels_3]{\overrightarrow\rels_{\schema'}}\migrsys'$ represents the MoT.

### 2023-10-17 2022Migration/articleMigrationFACS.tex
working towards redefining the migration script

@@
    u\in\rules\cup\transactions\ \wedge\ \sign{u}=\pair{A}{B}&~\Rightarrow~A\in\concepts\,\wedge\, B\in\concepts\label{eqn:invariant-has-type}\\
-   u\in\rules\cup\transactions\wedge\sign{u}=\pair{A}{B}\wedge\pair{a}{b}\in\viol{u}{\dataset}&~\Rightarrow~a\instance A\ \wedge\ b\instance B\label{eqn:rule-has-type}\\
    (\declare{n}{A}{B}\mapsfrom t)\in\mathcal\transactions&~\Rightarrow~ \declare{n}{A}{B}\in \rels\label{eqn:enforcement-has-type}
@@
    Requirements~\ref{eqn:relationsIntroduceConcepts} and~\ref{eqn:invariant-has-type} ensure that concepts mentioned in relations and in the signature of rules are part of the schema.
-   Requirement~\ref{eqn:rule-has-type} ensures that violations in a rule correspond to the type of that rule. 
    Requirement~\ref{eqn:enforcement-has-type} ensures the enforced relation of a transaction is declared in the schema. 
@@
 \label{sct:Information Systems}
-   Let us now define the notion of an information system.
+   Let us now define information systems by enumerating the requirements.
 \begin{definition}[information system]
@@
 \begin{itemize}
-   \item $\dataset=\pair{\triples}{\instance}$ is a data set (satisfies equation~\ref{eqn:wellTypedEdge}), we write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$;
-   \item $\schema=\quadruple{\concepts}{\rels}{\rules}{\transactions}$ is a schema (satisfies equation~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}), we write $\concepts_\infsys = \concepts$, $\rels_\infsys = \rels$, $\rules_\infsys=\rules$, and $\transactions_\infsys=\transactions$;
-   \item All rules and transactions are satisfied:
-   \begin{align}
-   \forall u\in\rules&.~\viol{u}{\dataset}=\emptyset\\
-   \forall (r\mapsfrom t)\in\transactions&.~\viol{u}{\dataset} - \pop{r}{\dataset}=\emptyset
-   \label{eqn:satisfaction}
-   \end{align}
-   \item Triples in the data set have their relation mentioned in the schema:
+\item $\dataset=\pair{\triples}{\instance}$ is a data set (so it must satisfy equation~\ref{eqn:wellTypedEdge}). We write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$;
+\item Triples in the data set must have their relation mentioned in the schema:
    \begin{eqnarray}
@@
    \end{eqnarray}
+\item $\schema=\quadruple{\concepts}{\rels}{\rules}{\transactions}$ is a schema (so it must satisfy equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}).
+We write $\concepts_\infsys = \concepts$, $\rels_\infsys = \rels$, $\rules_\infsys=\rules$, and $\transactions_\infsys=\transactions$;
+\item All violations must have a type, which follows from  (\ref{eqn:wellTypedViolation}).
+\item All blocking invariants and transactions must remain satisfied:
+   \begin{align}
+      \forall u\in\rules\cup\transactions&.~\viol{u}{\dataset}=\emptyset
+   \label{eqn:satisfaction}
+   \end{align}
+\item Transactions remain satisfied by adding violations to a specific relation (\ref{eqn:transaction}).
 \end{itemize}
@@
    % \end{align}
-   We say that event $\infsys\xrightarrow[I]{R} \infsys'$ brings the system $\infsys$ to $\infsys'$ inserting pairs in relations from $I$ and deleting pairs in relations from $R$.
+   We say that event $\infsys\xrightarrow[I]{R} \infsys'$ brings the system $\infsys$ to $\infsys'$ by inserting pairs in relations from $I$ and deleting pairs in relations from $R$.
 \end{definition}
@@
 %   If $R$ is a set of relations that covers those that occur in the rule, then changes to an system $\infsys$ can be described by the event $\infsys \xrightarrow[\{r\}]{R} \infsys'$.
-
-\subsection{Example}
-\label{sct:Example existing IS}
-   Having defined information systems in mathematical terms, let us discuss a small example.
-   It is written in the language Ampersand to make it more appealing to read.
-   Let us first define a data set of just a handful of triples and three relations.
-\begin{verbatim}
-RELATION takes[Student*Course] =
-[ ("Peter", "Management")
-; ("Susan", "Business IT")
-; ("John", "Business IT")
-]
-\end{verbatim}
-   This declaration introduces a relation with the name \verb#takes#,
-   source concept \verb#Student#, and
-   target concept \verb#Course#.
-   The informal meaning of this relation is that it states which students are taking which courses.
-
-   The example system also has a second relation that states which modules are part of which course.
-\begin{verbatim}
-RELATION isPartOf[Module*Course] =
-[ ("Finance", "Management")
-; ("Business Rules", "Business IT")
-; ("Business Analytics", "Business IT")
-; ("IT-Governance", "Management")
-]
-\end{verbatim}
-   The third relation states which students are enrolled for which module.
-   It is left without population for now.
-\begin{verbatim}
-RELATION isEnrolledFor[Student*Module]
-\end{verbatim}
-
-   A rule, {\tt EnrollRule} completes the script.
-   It states that a student can enroll for any module that is part of a course she takes.
-   % In Ampersand, which is a syntactically sugared form of relation algebra~\cite{JoostenRAMiCS2017},
-   % each rule has a name and each rule has a role to maintain its invariance:
-\begin{verbatim}
-RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
-\end{verbatim}
-   The Ampersand compiler defines $\viol{\tt EnrollRule}{\dataset}$ as
-\begin{equation}
-   \pop{\tt isEnrolledFor}{\dataset}-\{\pair{s}{m} \mid\exists c.~s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\}
-\end{equation}
-   This rule is satisfied if $\viol{\tt EnrollRule}{\dataset}=\emptyset$, which means that
-\begin{equation}
-   \begin{array}{l}
-   \forall \pair{s}{m}\in\pop{\tt isEnrolledFor}{\dataset}.~ \exists c\in\text{\tt Course}~.\\
-   s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
-   \end{array}
-\label{eqn:example isEnrolledFor}
-\end{equation}
-   In this example, relation $\declare{\tt isEnrolledFor}{\tt Student}{\tt Module}$ is empty, so the rule is satisfied.
-
-   Now let us check the requirements to verify that this example defines an information system.
-   % Requirement~\ref{eqn:specialization} is satisfied because this example contains no specialization.
-   The Ampersand compiler generates a data set $\dataset$, which contains a set of triples and a relation $\instance$.
-   It defines the set of triples $\triples$ as:
-\[\begin{array}[t]{l}
-   \triple{\text{\tt "Peter"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Management"}}\\
-   \triple{\text{\tt "Susan"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
-   \triple{\text{\tt "John"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
-   \triple{\text{\tt "Finance"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Management"}}\\
-   \triple{\text{\tt "Business Rules"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
-   \triple{\text{\tt "Business Analytics"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
-   \triple{\text{\tt "IT-Governance"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Management"}}
-\end{array}\]
-The relation $\instance$ contains the pairs:
-\[\begin{array}{l}
-   \pair{\text{\tt "Finance"}}{\tt Module}\\
-   \pair{\text{\tt "Business Rules"}}{\tt Module}\\
-   \pair{\text{\tt "Business Analytics"}}{\tt Module}\\
-   \pair{\text{\tt "IT-Governance"}}{\tt Module}\\
-   \pair{\text{\tt "Management"}}{\tt Course}\\
-   \pair{\text{\tt "Business IT"}}{\tt Course}\\
-   \pair{\text{\tt "Peter"}}{\tt Student}\\
-   \pair{\text{\tt "Susan"}}{\tt Student}\\
-   \pair{\text{\tt "John"}}{\tt Student}
-\end{array}\]
-   The pair $\pair{\triples}{\instance}$ satisfies requirement~\ref{eqn:wellTypedEdge} so this is a data set $\dataset$ as introduced in section~\ref{sct:Data sets}.
-
-   The Ampersand compiler generates a schema $\schema$, which contains concepts, a specialization relation, relations, and rules.
-   It defines the sets of concepts, relations, and rules to obtain a schema $\schema$:
-\[\begin{array}{rcl}
-   \concepts&=&\{ {\tt Module}, {\tt Course}, {\tt Student}\}\\
-   \rels&=&\{\begin{array}[t]{@{}l@{}}
-               \declare{\tt takes}{\tt Student}{\tt Course},\\
-               \declare{\tt isPartOf}{\tt Module}{\tt Course},\\
-               \declare{\tt isEnrolledFor}{\tt Student}{\tt Module}\}
-             \end{array}\\
-   \rules&=&\{ {\tt EnrollRule}\}\\
-   \transactions&=&\emptyset
-  \end{array}
-\]
-   % The user has defined no specializations.
-   % This means that $\isa$ equals the identity relation on $\concepts$ and requirements~\ref{eqn:isasIntroduceConcepts} and~\ref{eqn:isasPartialOrder} are satisfied.
-   % The script defines the set of rules $\rules$ to contain just one rule: \verb-EnrollRule-.
-   % The (nonempty) data set satisfies this rule, which meets requirement~\ref{eqn:consistentRules}.
-   So, the schema $\schema=\quadruple{\concepts}{\rels}{\rules}{\transactions}$ satisfies all requirements from section~\ref{sct:Schemas}.
-
-   Now let us check the definition of information system.
-   Requirement~\ref{eqn:satisfaction} is satisfied because the only rule, {\tt EnrollRule} is satisfied.
-   This satisfies definition~\ref{def:information system}, which establishes that $\infsys$ is an information system.
-   % The set of roles is empty and so is the relation $\maintain$.
-   % This meets all requirements for $\infsys=\triple{\dataset}{\schema}{\roles}{\maintain}$ from definition~\ref{def:information system}.
-   % This concludes the argument that $\infsys$ is an information system.
 
@@
 \subsection{General migration script}\label{General migration script}
-   So far, we illustrated how to do migration by giving two concrete examples.
-   This section generalizes those ideas by giving a general approach for generating a migration script.
+   This section presents a general approach for generating a migration script.
+   Let $\overleftarrow{\infsys}$ be the existing system, and let $\overrightarrow{\infsys}$ be the desired system.
+   Consequently, we define $\pair{\overleftarrow{\dataset}}{\overleftarrow{\schema}}=\overleftarrow{\infsys}$ and $\pair{\overrightarrow{\dataset}}{\overrightarrow{\schema}}=\overrightarrow{\infsys}$ as the data set and schema of the existing and desired system, respectively.
+   We also define $\quadruple{\overleftarrow{\concepts}}{\overleftarrow{\rels}}{\overleftarrow{\rules}}{\overleftarrow{\transactions}}=\overleftarrow{\schema}$
+   and $\quadruple{\overrightarrow{\concepts}}{\overrightarrow{\rels}}{\overrightarrow{\rules}}{\overrightarrow{\transactions}}=\overrightarrow{\schema}$.
+
+   Then we define the migration schema $\schema_\migrsys$ as follows:
 

### 2023-10-17 2022Migration/articleMigrationFACS.tex
Remove the example migration script.

@@
    A migration script takes care of two things: copying data and ensuring that the migration system is a proper information system.
-   Sections~\ref{sec:loading} and~\ref{sec:violations on invariants} illustrate how this works by example, using the Ampersand syntax.
-   Then, section~\ref{General migration script} defines what this means in general terms.
-
-\subsection{Loading Data}\label{sec:loading}
-   Let us study how to migrate data from $\dataset$ to $\dataset'$.
-   It would be tempting to state that a relation from the example system, say {\tt isPartOf}, could be copied by the following code fragment%
-\footnote{The prefix {\tt old.} refers to concepts, relations, and rules from the existing system (section~\ref{sct:Example existing IS}).
-The prefix {\tt new.} refers to items from the desired system.}:
-\begin{verbatim}
-   ENFORCE new.isPartOf >: old.isPartOf
-\end{verbatim}
-   The relation {\tt old.isPartOf} resides in the existing system and will not change after the MoT.
-
-   This means, however, that anything in {\tt old.isPartOf} cannot be deleted from {\tt new.isPartOf}.
-   This is undesirable because it prevents us from changing data in the migration system.
-   The desired behavior is that all pairs in {\tt old.isPartOf} are copied to {\tt new.isPartOf} when the system is first initialized, but that {\tt new.isPartOf} can be changed freely after that.
-   An extra relation (\ref{extraRelation}) achieves this behavior by remembering which pairs have been copied:
-\begin{eqnarray}
-   &&\verb#RELATION copied_isPartOf[Student*Course]#\label{extraRelation}\\
-   &&\verb#ENFORCE new.isPartOf >: old.isPartOf - copied_isPartOf#\label{difference}\\
-   &&\verb#ENFORCE copied_isPartOf >: new.isPartOf#\label{fill copies with new.isPartOf}
-\end{eqnarray}
-
-   Let us look what happens at runtime.
-   Initially, when {\tt copied\_isPartOf} is still empty, the difference between {\tt old.isPartOf} and {\tt copied\_isPartOf} equals {\tt old.isPartOf}.
-   So, {\tt new.isPartOf} will contain the pairs of {\tt old.isPartOf} by rule~\ref*{difference}.
-   Then, the engine inserts the same pairs into {\tt copied\_isPartOf} because of rule~\ref*{fill copies with new.isPartOf}.
-   After this, triples with the relation {\tt new.isPartOf} can be inserted and deleted freely.
-   Once all triples have been copied, {\tt copied\_isPartOf} will contain all pairs of {\tt old.isPartOf} and rule~\ref*{difference} will no longer insert any triples in {\tt new.isPartOf}.
-   So, this rule can be removed after the copying has taken place.
-   Consequently, declarations~\ref{extraRelation} thru~\ref{fill copies with new.isPartOf} can be removed from the migration script at the MoC.
-
-\subsection{Dealing with violations on invariants}\label{sec:violations on invariants}
-   Suppose the script for the desired system adds an invariant that is not present in the existing system:
-
-\begin{verbatim}
-   RELATION isPartOf[Module*Course] [UNI]
-\end{verbatim}
-
-   The \verb=UNI= annotation requires that each \verb=Module= is part of at most one \verb=Course=.
-   The violations are given by:
-\begin{eqnarray}
-   \viol{\tt UNI}{\dataset} = \{\pair{x}{y} \mid x \neq y \wedge \exists z.~ \pair{z}{x}, \pair{z}{y}\in\pop{\tt isPartOf}{\dataset} \}
-\end{eqnarray}
-   In Ampersand, the term that gives these violations is: \verb=isPartOf~;isPartOf - I=.
-
-   While the \verb=UNI= requirement holds for the population in the original script, that population may have changed.
-   Therefore, we must assume that we cannot expect that \verb=new.isPartOf= satisfies the \verb=UNI= requirement in the desired system.
-   The migration system allows for this by simply not requiring that they hold at the MoT.
-   A challenge lies in ensuring that they do hold at the MoC, and that users that help the migration forward can achieve a state in which all requirements hold.
-
-   Two extra relations are used: one to track the violations that occur in the original data (\ref{violRelation}), and one to track the violations that have been fixed (\ref{fixedRelation}) which prevents them from recurring.
-\begin{align}
-   &\text{\tt RELATION viols\_UNI[Course*Course]}\label{violRelation}\\
-   &\text{\tt RELATION fixed\_UNI[Course*Course]}\label{fixedRelation}\\
-   &\text{\tt ENFORCE viols\_UNI >: old.isPartOf\textasciitilde;old.isPartOf - I}\label{fill viols}\\
-   &\text{\tt ENFORCE fixed\_UNI >: viols\_UNI - (new.isPartOf\textasciitilde;new.isPartOf - I)}\label{fill fixed}\\
-   &\begin{aligned}
-   \text{\tt RULE isPartOf\_UNI: new.isPartOf\textasciitilde;new.isPartOf}\\
-   \quad\text{\tt |- I \textbackslash/ (viols\_UNI - fixed\_UNI)}
-   \end{aligned}\label{rule uni}
-\end{align}
-
-   We again look at what happens at runtime:
-   Enforcement rule~\ref{fill viols} adds all violations present in the old system to {\tt viols\_UNI}.
-   Until the MoT, rule~\ref{fill fixed} does nothing, since {\tt viols\_UNI} equals $\viol{\tt UNI}{\dataset}$.
-   At that point, rule~\ref{rule uni} is satisfied, since the right-hand side of equation~\ref{rule uni} is empty.
-   As violations are fixed, rule~\ref{fill fixed} records this by adding to {\tt fixed\_UNI}.
-   This way, previously fixed errors cannot recur.
-   Once {\tt viols\_UNI} is equal to {\tt fixed\_UNI}, the \verb=UNI= requirement holds and the system is ready for the MoC in regard to this rule.
-
-   We also wish to alert users with the role `migration helper' to violations that need to be fixed.
-   For the original \verb=UNI= requirement, this means specifying:
-
-\begin{align}
-   \begin{aligned}
-   \text{\tt RULE isPartOf\_UNI\_warn: new.isPartOf\textasciitilde;new.isPartOf |- I}\\
-   \quad\text{\tt ROLE MigrationHelper}
-   \end{aligned}\label{warnings}
-\end{align}
-
-   Despite the syntax, Line~{\ref{warnings}} is not a rule of the generated system in the formal sense:
-   By assigning a role, the responsibility of maintaining the rule no longer lies with the system.
-   Instead, users with the migration helper role can see the violations and act on it as they see fit, so the only effect of Line~{\ref{warnings}} is that such users will be shown its violations.
-
-\subsection{General migration script}\label{General migration script}
    This section presents a general approach for generating a migration script.
@@
    \end{align}
-\item We create transactions to copy equal relations, corresponding to equation~\ref{extraRelation} through~\ref{fill copies with new.isPartOf} in Section~\ref{sec:loading}:
+\item We create transactions to copy equal relations:
    \begin{align}
@@
 \item For each rule, we generate two helper relations, {\tt viols} and {\tt fixed}.
-Note that Rule~\ref{warnings} only influences the user interface and therefore is not modeled here.
-This corresponds to equation~\ref{violRelation} through~\ref{rule uni}.
    \begin{align}

### 2023-10-17 2022Migration/articleMigrationFACS.tex
making corrections in the migration script...

@@
    A migration script takes care of two things: copying data and ensuring that the migration system is a proper information system.
-   This section presents a general approach for generating a migration script.
-   Let $\overleftarrow{\infsys}$ be the existing system, and let $\overrightarrow{\infsys}$ be the desired system.
-   Consequently, we define $\pair{\overleftarrow{\dataset}}{\overleftarrow{\schema}}=\overleftarrow{\infsys}$ and $\pair{\overrightarrow{\dataset}}{\overrightarrow{\schema}}=\overrightarrow{\infsys}$ as the data set and schema of the existing and desired system, respectively.
-   We also define $\quadruple{\overleftarrow{\concepts}}{\overleftarrow{\rels}}{\overleftarrow{\rules}}{\overleftarrow{\transactions}}=\overleftarrow{\schema}$
-   and $\quadruple{\overrightarrow{\concepts}}{\overrightarrow{\rels}}{\overrightarrow{\rules}}{\overrightarrow{\transactions}}=\overrightarrow{\schema}$.
+\subsection{Generating a migration script}
+   In the migration system, we need to refer to the items (concepts, relations, and constraints) of both the existing system and the desired system.
+   For this purpose, we relabel those items.
+   We use a left arrow to denote relabeling by prefixing the name of the item with ``old.'', and a right arrow to denote prefixing with ``new.''.
+\begin{equation}
+   \begin{array}[m]{rcl}
+      \overleftarrow{\pair{\dataset}{\schema}}&=&\pair{\overleftarrow{\dataset}}{\overleftarrow{\schema}}\\
+      \overleftarrow{\pair{\triples}{\instance}}&=&\pair{\overleftarrow{\triples}}{\instance}\\
+      \overleftarrow{\quadruple{\concepts}{\rels}{\rules}{\transactions}}&=&\quadruple{\concepts}{\overleftarrow{\rels}}{\overleftarrow{\rules}}{\overleftarrow{\transactions}}\\
+      \overleftarrow{\triples}&=&\{\triple{a}{\overleftarrow{r}}{b}\mid\triple{a}{r}{b}\in\triples\}\\
+      \overleftarrow{\declare{n}{A}{B}}&=&\declare{old.n}{A}{B}\\
+      \overleftarrow{X}&=&\{\overleftarrow{x}\mid x\in X\}\\
+      \viol{\overleftarrow{u}}{\overleftarrow{\dataset}}&=&\viol{u}{\dataset}\\
+      \sign{\overleftarrow{u}}&=&\sign{u}\\
+      \enfRel{\overleftarrow{u}}&=&\overleftarrow{\enfRel{u}}
+   \end{array}
+\end{equation}
+   The notation $\overrightarrow{\infsys}$ is defined similarly, using prefix ``new.'' instead of ``old.''.
 
-   Then we define the migration schema $\schema_\migrsys$ as follows:
-
-   Let $\pair{\dataset}{\schema}$ be the existing system, and let ${\schema'}$ be the schema of the desired system.
-   Then we define the migration systems $\migrsys$ and the events as follows:
+   Then we define the migration system $\migrsys$ as follows:
+   Let $\pair{\dataset}{\schema}$ be the existing system, and let $\pair{\dataset'}{\schema'}$ be the desired system in its initial state.
 \begin{enumerate}
-\item We take a disjoint union of the relations, defining $\overleftarrow~$ and $\overrightarrow~$ as (overloaded) notation to generate new names, keeping the data set of the old system and the rules of the new system:
+\item We take a disjoint union of the data sets by relabeling relation names:
    \begin{align}
-   \overleftarrow{\declare{n}{x}{y}}&=\declare{(\id{old}.n)}{x}{y}&\overleftarrow{X} &= \{\overleftarrow{x}\mid x\in X\}\\
-   \overrightarrow{\declare{n}{x}{y}}&=\declare{(\id{new}.n)}{x}{y}& \overrightarrow{X} &= \{\overrightarrow{x}\mid x\in X\}\\
-   \triples_\migrsys &= \{\triple{a}{\overleftarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_\dataset\}&
-   \instance_\migrsys &= \instance_\dataset\\
-   \rels_\theenumi &= \overleftarrow\rels_{\schema} \cup \overrightarrow\rels_{\schema'}
-   %&\rules_\theenumi &= \rules_\schema
+      \dataset_\migrsys={}&\overleftarrow{\dataset}\cup\overrightarrow{\dataset'}
    \end{align}
@@
    \begin{align}
-   \widetilde{\declare{n}{x}{y}} &= \declare{(\id{copied}.n)}{x}{y}\\
-   \rels_\theenumi &= \{\widetilde{r}\mid r \in \rels_\schema \cap \rels_{\schema'}\}\\
-   \transactions_\theenumi &= \{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r} - \widetilde{r}} \mid \widetilde{r} \in \rels_2\} \cup \{\widetilde{r}\mapsfrom \popF{\overrightarrow{r}} \mid \widetilde{r} \in \rels_2\}
+      \widetilde{\declare{n}{A}{B}}={}&\declare{(\id{copied}.n)}{A}{B}\\
+      \begin{split}
+         \transactions_1={}&\{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r} - \widetilde{r}} \mid r \in \rels_\schema \cap \rels_{\schema'}\}\\
+         &\cup\{\widetilde{r}\mapsfrom \popF{\overrightarrow{r}} \mid r \in \rels_\schema \cap \rels_{\schema'}\}
+      \end{split}
    \end{align}
-\item For each rule, we generate two helper relations, {\tt viols} and {\tt fixed}.
+\item For each new blocking invariant $u$, we generate two helper relations: ${\tt viols}_u$ and ${\tt fixed}_u$, and two corresponding transactions.
    \begin{align}
-   \overleftarrow{\dataset'} &= \pair{\{\triple{x}{{r}}{y}\mid \triple{x}{\overleftarrow{r}}{y} \in \triples_{\dataset'}\}}{\instance_{\dataset'}}\\
-   \overrightarrow{\dataset'} &= \pair{\{\triple{a}{{r}}{b} \mid \triple{a}{\overrightarrow{r}}{b}\in\triples_\dataset\}}{\instance_{\dataset'}} \\
-   \rels_\theenumi &= \{{\tt viols}_u\mid u \in \rules_{\schema'}\} \cup \{{\tt fixed}_u\mid u \in \rules_{\schema'}\}\\
-   \rules_\migrsys &= \{\lambda {\dataset'}.~ \viol{u}{\overrightarrow{\dataset'}} - \pop{{\tt viols}_u - {\tt fixed}_u}{\dataset'} \mid u\in\rules_{\schema'}\}\label{eqn:migrsysrules}\\
-   \transactions_{\theenumi} &=\{{\tt viols}_u \mapsfrom \lambda \dataset'.~ \viol{u}{\overleftarrow{\dataset'}}\} \cup
-    \{{\tt fixed}_u \mapsfrom \lambda \dataset'.~ {\tt viols}_u - \viol{u}{\overrightarrow{\dataset'}}\}\label{eqn:enforceForRules}
+      \rels_\theenumi={}& \{{\tt viols}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\} \cup \{{\tt fixed}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
+      \begin{split}
+         \transactions_2={}& \{{\tt viols}_u \mapsfrom \violC{\overrightarrow{u}}\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
+         &\cup \{{\tt fixed}_u \mapsfrom \lambda \dataset'.~ {\tt viols}_u - \viol{\overrightarrow{u}}{\dataset'}\mid u \in \rules_{\schema'}-\rules_{\schema}\}
+      \end{split}\label{eqn:enforceForRules}
    \end{align}
-\item Combining the above into a single migration system:
+\item Let us combine the above into a single migration schema, $\schema_\migrsys$:
    \begin{align}
-   \concepts_\migrsys &= \concepts_\dataset \cup  \concepts_{\dataset'}\\
-   \rels_\migrsys &= \rels_1\cup\rels_2\cup \rels_3\\
-   %\rules_\migrsys &= \rules_1\cup \rules_3\\
-   \transactions_\migrsys &= \transactions_1\cup\transactions_2\cup\transactions_3
+      \concepts_\migrsys ={}& \concepts_\dataset \cup  \concepts_{\dataset'}\\
+      \rels_\migrsys ={}& \overleftarrow\rels_{\schema} \cup \overrightarrow\rels_{\schema'} \cup \rels_3\\
+      \rules_\migrsys={}& \emptyset\label{eqn:migrsysrules}\\
+      \transactions_\migrsys ={}& \transactions_1\cup\transactions_2
    \end{align}
@@
 \subsection{Validation}
-   We first observe that $\migrsys = \pair{\pair{\triples_\migrsys}{\instance_\migrsys}}{\quadruple{\concepts_\migrsys}{\rels_\migrsys}{\rules_\migrsys}{\transactions_\migrsys}}$ is a system after application of the transactions:
+   We first observe that $\migrsys = \pair{\dataset_\migrsys}{\quadruple{\concepts_\migrsys}{\rels_\migrsys}{\rules_\migrsys}{\transactions_\migrsys}}$ is a system after application of the transactions:
    the only rules are those introduced in equation~\ref{eqn:migrsysrules}, and transactions from equation~\ref{eqn:enforceForRules} ensure that $\popF{{\tt viols}_u} = u$ initially, while $\popF{{\tt fixed}_u}$ is empty.

### 2023-10-18 2022Migration/articleMigrationFACS.tex
sct 4.1: generating migration script needs review by Bas

@@
 \item We create transactions to copy equal relations:
+% \begin{verbatim}
+%    RELATION copied_r[A*B]
+%    RELATION new_r[A*B] [UNI]
+%    ENFORCE copied_r >: new_r /\ old_r
+%    ENFORCE new_r >: old_r - copied_r
+% \end{verbatim}
    \begin{align}
@@
       \begin{split}
-         \transactions_1={}&\{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r} - \widetilde{r}} \mid r \in \rels_\schema \cap \rels_{\schema'}\}\\
-         &\cup\{\widetilde{r}\mapsfrom \popF{\overrightarrow{r}} \mid r \in \rels_\schema \cap \rels_{\schema'}\}
+         \transactions_1={}&\{\widetilde{r}\mapsfrom \popF{\overrightarrow{r}\cap\overleftarrow{r}} \mid r \in \rels_\schema \cap \rels_{\schema'}\}\\
+         &\cup\{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r} - \widetilde{r}} \mid r \in \rels_\schema \cap \rels_{\schema'}\}
       \end{split}
    \end{align}
-\item For each new blocking invariant $u$, we generate two helper relations: ${\tt viols}_u$ and ${\tt fixed}_u$, and two corresponding transactions.
+\item For each new blocking invariant $u$, we generate a helper relation: ${\tt fixed}_u$, and a transaction to produce its population:
+% \begin{verbatim}
+%    RELATION fixed_TOTr[A*A]
+%    ENFORCE fixed_TOTr >: I /\ new_r;new_r~
+%    RULE Block_TOTr : fixed_TOTr |- new_r;new_r~
+%    MESSAGE "Relation r[A*B] must remain total."
+%    VIOLATION (TXT "Atom ", SRC I, TXT " must remain paired with an atom from B.")
+
+%    ROLE User MAINTAINS TOTr
+%    RULE TOTr : I - fixed_TOTr |- new_r;new_r~
+%    MESSAGE "Please, make relation r[A*B] total."
+%    VIOLATION (TXT "Fix ", SRC I, TXT " by pairing it with an (any) atom from B.")
+% \end{verbatim}
    \begin{align}
-      \rels_\theenumi={}& \{{\tt viols}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\} \cup \{{\tt fixed}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
-      \begin{split}
-         \transactions_2={}& \{{\tt viols}_u \mapsfrom \violC{\overrightarrow{u}}\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
-         &\cup \{{\tt fixed}_u \mapsfrom \lambda \dataset'.~ {\tt viols}_u - \viol{\overrightarrow{u}}{\dataset'}\mid u \in \rules_{\schema'}-\rules_{\schema}\}
-      \end{split}\label{eqn:enforceForRules}
+      \rels_\theenumi={}&\{{\tt fixed}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
+      \transactions_2={}&\{{\tt fixed}_u\mapsfrom\lambda\dataset'.~\viol{\overleftarrow{u}}{\dataset'}-\viol{\overrightarrow{u}}{\dataset'}\mid u\in\rules_{\schema'}-\rules_{\schema}\}\label{eqn:enforceForRules}
    \end{align}
@@
 \end{enumerate}
+   After the migration system has been deployed, we must wait until the business has resolved all violations.
+   This is something the system must signal, so the migration engineer can switch all traffic to the desired system:
+% \begin{verbatim}
+   % ROLE User MAINTAINS CleanUp_TOTr
+   % RULE CleanUp_TOTr : V[ONE*A] ; (I - fixed_TOTr) ; V[A*ONE]
+   % MESSAGE "Now you can remove the migration system because relation r[A*B] is total."
+% \end{verbatim}
+\begin{equation}
+   \forall u\in\rules_{\schema'}-\rules_{\schema}.\ \viol{\overleftarrow{u}}{\dataset'}={\tt fixed}_u
+\end{equation}
+   After this, the migration system is no longer needed and can be removed.
 

### 2023-10-18 2022Migration/articleMigrationFACS.tex
some corrections on the previous

@@
    \end{align}
-\item For each new blocking invariant $u$, we generate a helper relation: ${\tt fixed}_u$, and a transaction to produce its population:
+\item The new blocking invariants are $\rules_{\schema'}-\rules_{\schema}$.
+   For each new blocking invariant $u$, we generate a helper relation: ${\tt fixed}_u$, and a transaction to produce its population:
 % \begin{verbatim}
@@
 %    ENFORCE fixed_TOTr >: I /\ new_r;new_r~
-%    RULE Block_TOTr : fixed_TOTr |- new_r;new_r~
-%    MESSAGE "Relation r[A*B] must remain total."
-%    VIOLATION (TXT "Atom ", SRC I, TXT " must remain paired with an atom from B.")
 
@@
       \rels_\theenumi={}&\{{\tt fixed}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
-      \transactions_2={}&\{{\tt fixed}_u\mapsfrom\lambda\dataset'.~\viol{\overleftarrow{u}}{\dataset'}-\viol{\overrightarrow{u}}{\dataset'}\mid u\in\rules_{\schema'}-\rules_{\schema}\}\label{eqn:enforceForRules}
+      \transactions_2={}&\{{\tt fixed}_u\mapsfrom\lambda\dataset.~\viol{\overleftarrow{u}}{\dataset}-\viol{\overrightarrow{u}}{\dataset}\mid u\in\rules_{\schema'}-\rules_{\schema}\}\label{eqn:enforceForRules}
    \end{align}
-\item Let us combine the above into a single migration schema, $\schema_\migrsys$:
+\item For each new blocking invariant $u$, we generate another blocking invariant $v$ in the migration system that blocks fixed violations from recurring:
+% \begin{verbatim}
+%    RULE Block_TOTr : fixed_TOTr |- new_r;new_r~
+%    MESSAGE "Relation r[A*B] must remain total."
+%    VIOLATION (TXT "Atom ", SRC I, TXT " must remain paired with an atom from B.")
+% \end{verbatim}
    \begin{align}
-      \concepts_\migrsys ={}& \concepts_\dataset \cup  \concepts_{\dataset'}\\
-      \rels_\migrsys ={}& \overleftarrow\rels_{\schema} \cup \overrightarrow\rels_{\schema'} \cup \rels_3\\
-      \rules_\migrsys={}& \emptyset\label{eqn:migrsysrules}\\
-      \transactions_\migrsys ={}& \transactions_1\cup\transactions_2
+      \rules_\text{block}={}&\{v\ 
+      \begin{array}[t]{l}
+         \text{\bf where}\\
+         \sign{v}=\sign{u}\\
+         \viol{v}{\dataset}={\tt fixed}_u\cap\viol{\overrightarrow{u}}{\dataset}
+      \end{array}\\
+      &\mid u\in\rules_{\schema'}-\rules_{\schema}\}\notag
+   \end{align}
+\item Let us combine the above into a single migration schema:
+   \begin{align}
+      \schema_\migrsys=\quadruple{\concepts_\dataset\cup\concepts_{\dataset'}}{\overleftarrow\rels_{\schema}\cup\overrightarrow\rels_{\schema'}\cup\rels_3}{\rules_\text{block}}{\transactions_1\cup\transactions_2}
    \end{align}
 \end{enumerate}
-   After the migration system has been deployed, we must wait until the business has resolved all violations.
-   This is something the system must signal, so the migration engineer can switch all traffic to the desired system:
+   The migration engineer can switch all traffic to the desired system
+   after resolving the violations that prohibit deploying the desired system.
+   That is the case when violations of new invariants on the old population have all been fixed:
 % \begin{verbatim}
@@
 \begin{equation}
-   \forall u\in\rules_{\schema'}-\rules_{\schema}.\ \viol{\overleftarrow{u}}{\dataset'}={\tt fixed}_u
+   \forall u\in\rules_{\schema'}-\rules_{\schema}.\ \viol{\overleftarrow{u}}{\dataset_\migrsys}={\tt fixed}_u
 \end{equation}
-   After this, the migration system is no longer needed and can be removed.
+   After this, the migration engineer can remove the migration system and the old system.
 
 \subsection{Validation}
-   We first observe that $\migrsys = \pair{\dataset_\migrsys}{\quadruple{\concepts_\migrsys}{\rels_\migrsys}{\rules_\migrsys}{\transactions_\migrsys}}$ is a system after application of the transactions:
+   We first observe that $\migrsys = \pair{\dataset_\migrsys}{\schema_\migrsys}$ is a system after application of the transactions:
    the only rules are those introduced in equation~\ref{eqn:migrsysrules}, and transactions from equation~\ref{eqn:enforceForRules} ensure that $\popF{{\tt viols}_u} = u$ initially, while $\popF{{\tt fixed}_u}$ is empty.

### 2023-10-18 2022Migration/articleMigrationFACS.tex
add business constraints to the schema

@@
 \newcommand{\triple}[3]{\langle{#1},{#2},{#3}\rangle}
-\newcommand{\quadruple}[4]{\langle{#1},{#2},{#3},{#4}\rangle}
+\newcommand{\quintuple}[5]{\langle{#1},{#2},{#3},{#4},{#5}\rangle}
 \newcommand{\atom}[1]{{\tt\small #1}}
@@
 \newcommand{\vertices}{N}
+\newcommand{\rules}{\mathcal{U}}
 \newcommand{\transactions}{\mathcal{E}}
-\newcommand{\rules}{\mathcal{U}}
+\newcommand{\busConstraints}{\mathcal{B}}
 \newcommand{\Constraints}{\mathbb{U}}
@@
 
-   We describe a schema $\schema$ as a tuple $\quadruple{\concepts}{\rels}{\rules}{\transactions}$,
+   We describe a schema $\schema$ as a tuple $\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$,
    in which $\concepts\subseteq \Concepts$ is a finite set of concepts,
@@
    $\rules\subseteq\Constraints$ is a finite set of blocking invariants,
-   and $\transactions\subseteq\Constraints$ is a finite set of transactions.
+   $\transactions\subseteq\Constraints$ is a finite set of transactions,
+   and $\busConstraints\subseteq\Constraints$ is a finite set of business constraints.
 
@@
    in which $r=\enfRel{u}$.
-   The set of transactions in a schema is $\transactions$.
 
    \begin{definition}[Schema]
-   A schema is a tuple $\quadruple{\concepts}{\rels}{\rules}{\transactions}$ that satisfies:
+   A schema is a tuple $\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$ that satisfies:
 \begin{align}
@@
    \label{eqn:relationsIntroduceConcepts}\\
-   u\in\rules\cup\transactions\ \wedge\ \sign{u}=\pair{A}{B}&~\Rightarrow~A\in\concepts\,\wedge\, B\in\concepts\label{eqn:invariant-has-type}\\
+   u\in\rules\cup\transactions\cup\busConstraints\ \wedge\ \sign{u}=\pair{A}{B}&~\Rightarrow~A\in\concepts\,\wedge\, B\in\concepts\label{eqn:invariant-has-type}\\
    (\declare{n}{A}{B}\mapsfrom t)\in\mathcal\transactions&~\Rightarrow~ \declare{n}{A}{B}\in \rels\label{eqn:enforcement-has-type}
@@
    Requirement~\ref{eqn:enforcement-has-type} ensures the enforced relation of a transaction is declared in the schema. 
-   When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, $\rules_{\schema}$, and $\transactions_{\schema}$
-   for $\concepts$, $\rels$, $\rules$, and $\transactions$ corresponding to $\schema = \quadruple{\concepts}{\rels}{\rules}{\transactions}$.
+   When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, $\rules_{\schema}$, $\transactions_{\schema}$, and $\busConstraints_{\schema}$
+   for $\concepts$, $\rels$, $\rules$, $\transactions$, and $\busConstraints$ corresponding to $\schema = \quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$.
 
@@
    \end{eqnarray}
-\item $\schema=\quadruple{\concepts}{\rels}{\rules}{\transactions}$ is a schema (so it must satisfy equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}).
-We write $\concepts_\infsys = \concepts$, $\rels_\infsys = \rels$, $\rules_\infsys=\rules$, and $\transactions_\infsys=\transactions$;
+\item $\schema=\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$ is a schema (so it must satisfy equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}).
 \item All violations must have a type, which follows from  (\ref{eqn:wellTypedViolation}).
@@
    In the migration system, we need to refer to the items (concepts, relations, and constraints) of both the existing system and the desired system.
-   For this purpose, we relabel those items.
-   We use a left arrow to denote relabeling by prefixing the name of the item with ``old.'', and a right arrow to denote prefixing with ``new.''.
+   We use a prime symbol to denote the elements of the desired system, so $\infsys$ is the existing system and $\infsys'$ is the desired system.
+   We also have to relabel items with prefixes to avoid name clashes in the migration system.
+   We use a left arrow to denote relabeling by prefixing the name of the item with ``old.''.
 \begin{equation}
@@
       \overleftarrow{\pair{\triples}{\instance}}&=&\pair{\overleftarrow{\triples}}{\instance}\\
-      \overleftarrow{\quadruple{\concepts}{\rels}{\rules}{\transactions}}&=&\quadruple{\concepts}{\overleftarrow{\rels}}{\overleftarrow{\rules}}{\overleftarrow{\transactions}}\\
+      \overleftarrow{\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}}&=&\quintuple{\concepts}{\overleftarrow{\rels}}{\overleftarrow{\rules}}{\overleftarrow{\transactions}}{\overleftarrow{\busConstraints}}\\
       \overleftarrow{\triples}&=&\{\triple{a}{\overleftarrow{r}}{b}\mid\triple{a}{r}{b}\in\triples\}\\
@@
    Then we define the migration system $\migrsys$ as follows:
-   Let $\pair{\dataset}{\schema}$ be the existing system, and let $\pair{\dataset'}{\schema'}$ be the desired system in its initial state.
+   Let $\pair{\dataset}{\schema}$ be the existing system.
+   Let $\pair{\dataset'}{\schema'}$ be the desired system in its initial state.
 \begin{enumerate}
@@
 %    ENFORCE fixed_TOTr >: I /\ new_r;new_r~
-
-%    ROLE User MAINTAINS TOTr
-%    RULE TOTr : I - fixed_TOTr |- new_r;new_r~
-%    MESSAGE "Please, make relation r[A*B] total."
-%    VIOLATION (TXT "Fix ", SRC I, TXT " by pairing it with an (any) atom from B.")
 % \end{verbatim}
@@
          \sign{v}=\sign{u}\\
-         \viol{v}{\dataset}={\tt fixed}_u\cap\viol{\overrightarrow{u}}{\dataset}
+         \viol{v}{\dataset}=\viol{\overrightarrow{u}}{\dataset}\cap{\tt fixed}_u
+      \end{array}\\
+      &\mid u\in\rules_{\schema'}-\rules_{\schema}\}\notag
+   \end{align}
+\item To signal users that there are violations that need to be fixed, we generate a business constraint for each new blocking invariant $u$:
+% \begin{verbatim}
+%    ROLE User MAINTAINS TOTr
+%    RULE TOTr : I - fixed_TOTr |- new_r;new_r~
+%    MESSAGE "Please, make relation r[A*B] total."
+%    VIOLATION (TXT "Fix ", SRC I, TXT " by pairing it with an (any) atom from B.")
+% \end{verbatim}
+   \begin{align}
+      \busConstraints_\text{fix}={}&\{v\ 
+      \begin{array}[t]{l}
+         \text{\bf where}\\
+         \sign{v}=\sign{u}\\
+         \viol{v}{\dataset}=\viol{\overrightarrow{u}}{\dataset}-{\tt fixed}_u
       \end{array}\\
@@
    \begin{align}
-      \schema_\migrsys=\quadruple{\concepts_\dataset\cup\concepts_{\dataset'}}{\overleftarrow\rels_{\schema}\cup\overrightarrow\rels_{\schema'}\cup\rels_3}{\rules_\text{block}}{\transactions_1\cup\transactions_2}
+      \schema_\migrsys=\quintuple{\concepts_\dataset\cup\concepts_{\dataset'}}{\overleftarrow\rels_{\schema}\cup\overrightarrow\rels_{\schema'}\cup\rels_3}{\rules_\text{block}}{\transactions_1\cup\transactions_2}{\busConstraints_\text{fix}\cup\busConstraints_{\schema'}}
    \end{align}

### 2023-10-18 2022Migration/articleMigrationFACS.tex
minor

@@
    In the migration system, we need to refer to the items (concepts, relations, and constraints) of both the existing system and the desired system.
-   We use a prime symbol to denote the elements of the desired system, so $\infsys$ is the existing system and $\infsys'$ is the desired system.
-   We also have to relabel items with prefixes to avoid name clashes in the migration system.
+   We have to relabel items with prefixes to avoid name clashes in the migration system.
    We use a left arrow to denote relabeling by prefixing the name of the item with ``old.''.

### 2023-10-18 2022Migration/articleMigrationFACS.tex
correct a mistake

@@
    A migration script takes care of two things: copying data and ensuring that the migration system is a proper information system.
-\subsection{Generating a migration script}
+\subsection{Algorithm for generating a migration script}
    In the migration system, we need to refer to the items (concepts, relations, and constraints) of both the existing system and the desired system.
@@
 \item We create transactions to copy equal relations:
+For every relation $r$ shared by the existing and desired systems, we generate a helper relation: ${\tt copy}_r$, and a transaction to produce its population.
+We also generate a transaction to populate the relation $\overrightarrow{r}$ (in the desired system).
 % \begin{verbatim}
@@
    \begin{align}
-      \widetilde{\declare{n}{A}{B}}={}&\declare{(\id{copied}.n)}{A}{B}\\
-      \begin{split}
-         \transactions_1={}&\{\widetilde{r}\mapsfrom \popF{\overrightarrow{r}\cap\overleftarrow{r}} \mid r \in \rels_\schema \cap \rels_{\schema'}\}\\
-         &\cup\{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r} - \widetilde{r}} \mid r \in \rels_\schema \cap \rels_{\schema'}\}
-      \end{split}
+      \rels_1={}&\{{\tt copy}_r\mid r \in \rels_{\schema'}\cap\rels_{\schema}\}\\
+      \transactions_1={}&\{{\tt copy}_r\mapsfrom \popF{\overrightarrow{r}\cap\overleftarrow{r}} \mid r \in\rels_1\}\cup\{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r}-{\tt copy}_r} \mid r \in\rels_1\}
    \end{align}
+   The helper relation is needed to ensure that the copying process terminates.
 \item The new blocking invariants are $\rules_{\schema'}-\rules_{\schema}$.
-   For each new blocking invariant $u$, we generate a helper relation: ${\tt fixed}_u$, and a transaction to produce its population:
+   For each new blocking invariant $u$, we generate a helper relation: ${\tt fixed}_u$, to register all violations that are fixed.
+   We also need a transaction to produce its population:
 % \begin{verbatim}
@@
    \begin{align}
-      \rels_\theenumi={}&\{{\tt fixed}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
+      \rels_2={}&\{{\tt fixed}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
       \transactions_2={}&\{{\tt fixed}_u\mapsfrom\lambda\dataset.~\viol{\overleftarrow{u}}{\dataset}-\viol{\overrightarrow{u}}{\dataset}\mid u\in\rules_{\schema'}-\rules_{\schema}\}\label{eqn:enforceForRules}
@@
    \end{align}
+   Note that the blocking invariants from $\schema'\cap\schema$ are also used in $\migrsys$ to ensure continuity.
 \item To signal users that there are violations that need to be fixed, we generate a business constraint for each new blocking invariant $u$:
@@
    \begin{align}
-      \schema_\migrsys=\quintuple{\concepts_\dataset\cup\concepts_{\dataset'}}{\overleftarrow\rels_{\schema}\cup\overrightarrow\rels_{\schema'}\cup\rels_3}{\rules_\text{block}}{\transactions_1\cup\transactions_2}{\busConstraints_\text{fix}\cup\busConstraints_{\schema'}}
+      \schema_\migrsys=\langle{}&\concepts_\dataset\cup\concepts_{\dataset'},\\
+      &\overleftarrow{\rels_{\schema}}\cup\overrightarrow{\rels_{\schema'}}\cup\rels_1\cup\rels_2,\notag\\
+      &\rules_\text{block}\cup(\rules_{\schema}\cap\rules_{\schema'}),\notag\\
+      &\transactions_1\cup\transactions_2,\notag\\
+      &\busConstraints_\text{fix}\cup\busConstraints_{\schema'}\rangle\notag
    \end{align}

### 2023-10-18 2022Migration/articleMigrationFACS.tex
Introduce the Proof Obligations

@@
    Requirement~\ref{eqn:enforcement-has-type} ensures the enforced relation of a transaction is declared in the schema. 
-   When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, $\rules_{\schema}$, $\transactions_{\schema}$, and $\busConstraints_{\schema}$
-   for $\concepts$, $\rels$, $\rules$, $\transactions$, and $\busConstraints$ corresponding to $\schema = \quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$.
+   When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, $\rules_{\schema}$, $\transactions_{\schema}$, $\busConstraints_{\schema}$
+   for $\concepts$, $\rels$, $\rules$, $\transactions$, $\busConstraints$ corresponding to $\schema = \quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$.
 
@@
 \begin{itemize}
-\item $\dataset=\pair{\triples}{\instance}$ is a data set (so it must satisfy equation~\ref{eqn:wellTypedEdge}). We write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$;
+\item $\dataset=\pair{\triples}{\instance}$ is a data set (so it must satisfy equation~\ref{eqn:wellTypedEdge}).
+   We write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$ if needed;
 \item Triples in the data set must have their relation mentioned in the schema:
@@
    \end{align}
-   The helper relation is needed to ensure that the copying process terminates.
+   The helper relation is needed to ensure that the copying process terminates,
+   which is the case when:
+   \begin{align}
+      \forall r\in\rels_1.~\overleftarrow{r}={\tt copy}_r\label{eqn:copyingTerminates}
+   \end{align}
+
 \item The new blocking invariants are $\rules_{\schema'}-\rules_{\schema}$.
@@
       \transactions_2={}&\{{\tt fixed}_u\mapsfrom\lambda\dataset.~\viol{\overleftarrow{u}}{\dataset}-\viol{\overrightarrow{u}}{\dataset}\mid u\in\rules_{\schema'}-\rules_{\schema}\}\label{eqn:enforceForRules}
+   \end{align}
+   The helper relation is needed to ensure that the fixing of violations terminates,
+   which is the case when:
+   \begin{align}
+      \forall u\in\rules_{\schema'}-\rules_{\schema}.~\viol{\overrightarrow{u}}{\dataset}={\tt fixed}_u\label{eqn:fixingTerminates}
    \end{align}
@@
       &\rules_\text{block}\cup(\rules_{\schema}\cap\rules_{\schema'}),\notag\\
-      &\transactions_1\cup\transactions_2,\notag\\
+      &\transactions_1\cup\transactions_2\cup\transactions_{\schema'},\notag\\
       &\busConstraints_\text{fix}\cup\busConstraints_{\schema'}\rangle\notag
@@
 
+\subsection{Proof Obligations}
+\begin{enumerate}
+\item Assume that $\infsys$ and $\infsys'$ are information systems, so they satisfy definition~\ref{def:information system}.
+      This means that equations~\ref{eqn:wellTypedEdge}, \ref{eqn:wellTypedViolation}, \ref{eqn:transaction}, \ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}, \ref{eqn:define R}, \ref{eqn:satisfaction} hold for both systems.
+\item Prove that $\migrsys=\pair{\dataset_\migrsys}{\schema_\migrsys}$ is an information system,
+      so the migration engineer can deploy the migration system.
+\item Prove that $\overrightarrow{r}$ contains all triples from the existing data set in a finite time after deploying $\migrsys$,
+      to ensure that the existing data is migrated to the desired system.
+\item Prove that equation~\ref{eqn:satisfaction} holds for $\migrsys$ when the copying of relations is done (just before MoT),
+      to ensure that a valid information system is presented to users at the moment of transition.
+\item Prove that equation~\ref{eqn:copyingTerminates} represents the condition for terminating the copying process.
+\item Prove that equation~\ref{eqn:fixingTerminates} represents the condition for terminating the fixing of violations.
+\item Prove that the migration engineer can switch the ingress from the existing system to the migration system
+      when equations~\ref{eqn:copyingTerminates} and~\ref{eqn:fixingTerminates} are true,
+      so the computer can let the migration engineer know that it is safe to switch.
+\item Assuming that the business can fix each violation in finite time,
+      prove that the number of violations to be fixed by the business is finite,
+      and that it decreases monotonically, so the business can reach the MoC in finite time.
+\item Prove that $\rules_\text{block}$ is free of violations, so it is redundant after the MoC.
+\item Prove that $\transactions_1$ will never more add any violations to any ${\tt copy}_u$ at the MoC, so $\rels_1$ and $\transactions_1$ are redundant after the MoC.
+\item Prove that $\transactions_2$ will never more add any violations to any ${\tt fixed}_u$ at the MoC, so that $\rels_2$ and $\transactions_2$ are redundant after the MoC.
+\item Prove that $\busConstraints_\text{fix}$ is free of violations, so it is redundant after the MoC.
+\item Prove that $\concepts_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$ are redundant after the MoC.
+\item Prove that, given the redundancies proven above, the migration system is equivalent to $\infsys'$ after the MoC,
+      so the migration engineer can switch the ingress to the desired system and remove the existing system and the migration system.
+\end{enumerate}
 \subsection{Validation}

### 2023-10-18 2022Migration/articleMigrationFACS.tex
gluing sections with intermediate texts

@@
 
-   The next chapter analyzes SCDMs with an eye on zero downtime and data quality.
-   To prevent mistakes, this analysis takes a formal approach.
-   We define an information system (in section~\ref{sct:Information Systems}) as a pair of a schema and a data set,
-   so that we have the schema explicitly at our disposal.
-   The data set is separate, so we may consider the analysis of just a schema as a compile-time activity.
-   Section~\ref{sct:Schemas} formalizes schemas as a combination of concepts, relations and invariants.
-   Relations are carriers of the data and concepts are needed to define them.
-   Invariants are constraints on the data to impose semantics.
-   These constraints must be satisfied all the time.
+   The next section analyzes SCDMs with an eye on zero downtime and data quality.
+   It sketches the outline of a procedure for SCDMs.
+   Section~\ref{sct:Definitions} formalizes the concepts that we need to define the procedure.
+   Section~\ref{sct:Generating} defines the algoritm for generating a migration system, to automate SCDMs.
+   This algorithm has been proven correct in Isabelle/HOL.
+   The outline of this proof is discussed in section~\ref{sct:Proof} and the proof itself is published on the internet.
+
    For experimental validation we have used the language Ampersand~\cite{JoostenRAMiCS2017,Joosten-JLAMP2018}
-   because its syntax and semantics correspond directly to these definitions.
-   Ampersand has a notion of rule, which implements the invariants on the data set.
+   because its syntax and semantics correspond directly to the definitions in section~\ref{sct:Definitions}.
    In this way, prototypes built in Ampersand have helped to refine the theory.
@@
    which relies on the semantics of the data to draw the right conclusions and carry out their tasks.
-   %This paper uses semantic constraints to represent this meaning.
 \begin{figure}[bht]
@@
 
-   We assume that constraints implement the semantics of the data.
+   We assume that constraints implement the business semantics of the data.
+   Constraints represent business concerns formally, so they can be checked automatically and be used to generate software.
    Figure~\ref{fig:pre-migration} depicts them as rules.
@@
    A \define{blocking invariant} is a constraint that is always true in a system.
-   It serves to constrain the data set at runtime, to ensure its semantic integrity.
-   When it is violated, the system produces an error message and refuses the transaction.
+   It serves to constrain the data set at runtime.
+   When it is about to be violated, the system produces an error message and refuses the transaction.
 \item Transaction\\
@@
    which the system keeps satisfied by adding or deleting triples to the dataset.
-   As soon as the data violates a transaction, the system restores it without human intervention.
+   As soon as the data violates this constraint, the system restores it without human intervention.
    So, the outside world experiences a constraint that is always true, i.e.~an invariant.
@@
    For this purpose, we define an intermediate system, the \define{migration system},
-   which contains all concepts, relations, and constraints of the desired system.
+   which contains all concepts, relations, and constraints of both the existing and the desired system.
    However, it implements the blocking invariants of the desired system as business constraints.
@@
    We write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$ if needed;
+\item $\schema=\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$ is a schema (so it must satisfy equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}).
 \item Triples in the data set must have their relation mentioned in the schema:
@@
    \end{eqnarray}
-\item $\schema=\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$ is a schema (so it must satisfy equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}).
 \item All violations must have a type, which follows from  (\ref{eqn:wellTypedViolation}).
+\item Transactions remain satisfied by adding violations to a specific relation (\ref{eqn:transaction}).
 \item All blocking invariants and transactions must remain satisfied:
@@
    \end{align}
-\item Transactions remain satisfied by adding violations to a specific relation (\ref{eqn:transaction}).
 \end{itemize}
 \end{definition}
-   % A \define{role} is a name that identifies a group of users.
-   % It serves as a placeholder for a person or a machine (i.e. an actor) that works with the data set (i.e. create, read, update, or delete triples).
-   % The purpose of a role is to mention an individual user (human) or an automated actor (bot) without knowing who that user is.
-
-   As the system responds to events, it changes its data set, while the schema remains the same.
-   In this paper, it suffices to define an event as a pair of systems for which the schema stays constant.
-   Events are categorized by the part of the data set that changes.
-
-\begin{definition}[Event]
-   Let $R \subseteq \Rels$ and $I \subseteq \Rels$ be sets of relations,
-   and let $\infsys=\pair{\schema}{\dataset}$ and  $\infsys'=\pair{\schema}{\dataset'}$ be information systems,
-   then $\infsys\xrightarrow[I]{R} \infsys'$ is an event if and only if:
-   \begin{align}
-      \triple{a}{r}{b}\in\triples_{\dataset}-\triples_{\dataset'}&\Rightarrow\ r\in R
-   \label{eqn:eventUnchanged}\\
-      \triple{a}{r}{b}\in\triples_{\dataset'}-\triples_{\dataset}&\Rightarrow\ r\in(R \cup I)
-   \label{eqn:eventInsert}
-   \end{align}
-   % was:
-   % \begin{align}
-   %    \triples_{\dataset} - (\Triple{\Atoms}{(R \cup I)}{\Atoms}) &= \triples_{\dataset'} - (\Triple{\Atoms}{(R \cup I)}{\Atoms})
-   % \label{eqn:eventUnchanged}\\
-   %    \triples_{\dataset} - (\Triple{\Atoms}{R}{\Atoms}) &\subseteq \triples_{\dataset'} - (\Triple{\Atoms}{R}{\Atoms})
-   % \label{eqn:eventInsert}
-   % \end{align}
-   We say that event $\infsys\xrightarrow[I]{R} \infsys'$ brings the system $\infsys$ to $\infsys'$ by inserting pairs in relations from $I$ and deleting pairs in relations from $R$.
-\end{definition}
-
-   Equation~\ref{eqn:eventUnchanged} and~\ref{eqn:eventInsert} state that the triples of those for relations in $I$ or $R$ are the only ones that can be changed, and that only triples in $R$ can be removed.
-   %If $I$ or $R$ is the empty set, we omit it from the arrow, so $\infsys \xrightarrow{R} \infsys'$ is a notation for $\infsys \xrightarrow[\emptyset]{R} \infsys'$.
-
-\begin{definition}[Inserting event]
-   We write $\infsys \xrightarrowdbl[I]{R} \infsys'$ to state that some triple was inserted whose relation was in $I$, that is:
-   $t \in \triples_{\dataset'} \cap (\Triple{\Atoms}{I}{\Atoms})$ and $t \not\in \triples_{\dataset}$ for some triple $t$.
-\end{definition}
-
-   We use inserting events to model the effect of transactions.
-%   If $R$ is a set of relations that covers those that occur in the rule, then changes to an system $\infsys$ can be described by the event $\infsys \xrightarrow[\{r\}]{R} \infsys'$.
 
 \section{Generating a Migration Script}
-   A migration script takes care of two things: copying data and ensuring that the migration system is a proper information system.
+\label{sct:Generating}
+   The complexity of data migration results in a reasonable chance of making mistakes in writing code for data migration.
+   For that reason, we want to generate this code.
+   To ensure the correctness of the generator is even more challenging.
+   That is why we have proven the correctness of the generator in Isabelle/HOL.
+
+   This section starts with a presentation of the algorithm.
+   Then we discuss its proof of correctness.
+
 \subsection{Algorithm for generating a migration script}
@@
    \end{align}
-\item We create transactions to copy equal relations:
+\item We create transactions to copy the population of relations from $\dataset$ to $\dataset'$:
 For every relation $r$ shared by the existing and desired systems, we generate a helper relation: ${\tt copy}_r$, and a transaction to produce its population.
@@
    \end{align}
+   In some cases, a migration engineer can invent ways to satisfy these invariants automatically.
+   For this purpose, the generator should produce source code (as opposed to compiled code) to allow the migration engineer to replace a business constraint with transactions of her own making.
 \item Let us combine the above into a single migration schema:
@@
 
-\subsection{Proof Obligations}
+\subsection{Proof}
+\label{sct:Proof}
+   A theorem prover, Isabelle/HOL, is used to establish the correctness of the presented algorithm.
+   This section discusses the proof obligations. The proof is available online.
+   Assume that $\infsys$ and $\infsys'$ are information systems, so they satisfy definition~\ref{def:information system}.
+   This means that equations~\ref{eqn:wellTypedEdge}, \ref{eqn:wellTypedViolation}, \ref{eqn:transaction}, \ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}, \ref{eqn:define R}, \ref{eqn:satisfaction} hold for both systems.
 \begin{enumerate}
-\item Assume that $\infsys$ and $\infsys'$ are information systems, so they satisfy definition~\ref{def:information system}.
-      This means that equations~\ref{eqn:wellTypedEdge}, \ref{eqn:wellTypedViolation}, \ref{eqn:transaction}, \ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}, \ref{eqn:define R}, \ref{eqn:satisfaction} hold for both systems.
-\item Prove that $\migrsys=\pair{\dataset_\migrsys}{\schema_\migrsys}$ is an information system,
+\item Prove that $\migrsys$ is an information system,
       so the migration engineer can deploy the migration system.
-\item Prove that $\overrightarrow{r}$ contains all triples from the existing data set in a finite time after deploying $\migrsys$,
+\item For every relation $r\in\rels_{\schema}\cap\rels_{\schema'}$, prove that $\overrightarrow{r}$ contains all triples from the existing data set in a finite time after deploying $\migrsys$,
       to ensure that the existing data is migrated to the desired system.
@@
 \end{enumerate}
-\subsection{Validation}
+\section{Validation}
+   As the system responds to events, it changes its data set, while the schema remains the same.
+   In this paper, it suffices to define an event as a pair of systems for which the schema stays constant.
+   Events are categorized by the part of the data set that changes.
+
+\begin{definition}[Event]
+   Let $R \subseteq \Rels$ and $I \subseteq \Rels$ be sets of relations,
+   and let $\infsys=\pair{\schema}{\dataset}$ and  $\infsys'=\pair{\schema}{\dataset'}$ be information systems,
+t  hen $\infsys\xrightarrow[I]{R} \infsys'$ is an event if and only if:
+\begin{align}
+   \triple{a}{r}{b}\in\triples_{\dataset}-\triples_{\dataset'}&\Rightarrow\ r\in R\label{eqn:eventUnchanged}\\
+   \triple{a}{r}{b}\in\triples_{\dataset'}-\triples_{\dataset}&\Rightarrow\ r\in(R \cup I)\label{eqn:eventInsert}
+\end{align}
+% was:
+% \begin{align}
+%    \triples_{\dataset} - (\Triple{\Atoms}{(R \cup I)}{\Atoms}) &= \triples_{\dataset'} - (\Triple{\Atoms}{(R \cup I)}{\Atoms})
+% \label{eqn:eventUnchanged}\\
+%    \triples_{\dataset} - (\Triple{\Atoms}{R}{\Atoms}) &\subseteq \triples_{\dataset'} - (\Triple{\Atoms}{R}{\Atoms})
+% \label{eqn:eventInsert}
+% \end{align}
+   We say that event $\infsys\xrightarrow[I]{R} \infsys'$ brings the system $\infsys$ to $\infsys'$ by inserting pairs in relations from $I$ and deleting pairs in relations from $R$.
+\end{definition}
+
+   Equation~\ref{eqn:eventUnchanged} and~\ref{eqn:eventInsert} state that the triples of those for relations in $I$ or $R$ are the only ones that can be changed, and that only triples in $R$ can be removed.
+%If $I$ or $R$ is the empty set, we omit it from the arrow, so $\infsys \xrightarrow{R} \infsys'$ is a notation for $\infsys \xrightarrow[\emptyset]{R} \infsys'$.
+
+\begin{definition}[Inserting event]
+   We write $\infsys \xrightarrowdbl[I]{R} \infsys'$ to state that some triple was inserted whose relation was in $I$, that is:
+   $t \in \triples_{\dataset'} \cap (\Triple{\Atoms}{I}{\Atoms})$ and $t \not\in \triples_{\dataset}$ for some triple $t$.
+\end{definition}
+
+W  e use inserting events to model the effect of transactions.
+%   If $R$ is a set of relations that covers those that occur in the rule, then changes to an system $\infsys$ can be described by the event $\infsys \xrightarrow[\{r\}]{R} \infsys'$.
+
    We first observe that $\migrsys = \pair{\dataset_\migrsys}{\schema_\migrsys}$ is a system after application of the transactions:
@@
 \section{Conclusions}
-
+\label{sct:Conclusions}
    In this paper, we describe the data migration as going from an existing system to a desired one, where the schema changes.

### 2023-10-29 2022Migration/articleMigrationFACS.tex
cleaning up
minor
remark to Bas included

@@
 \end{definition}
+   We assume that a deployment will fail if these requirements are not satisfied.
+   We will write $\text{isIS}(\infsys)$ to denote that $\infsys$ is an information system.
 
@@
 \label{sct:Generating}
-   The complexity of data migration results in a reasonable chance of making mistakes in writing code for data migration.
+   The complexity of migrating data to date yields expensive and error-prone migration projects.
    For that reason, we want to generate this code.
@@
 \subsection{Algorithm for generating a migration script}
+   (\@Bas, I have added the relevant portions of {\tt Kurk.adl} as \LaTeX\ comments near every equation in the \LaTeX\ source,
+   so we can guard their correspondence.)
+
    In the migration system, we need to refer to the items (concepts, relations, and constraints) of both the existing system and the desired system.
@@
    \end{align}
-\item We create transactions to copy the population of relations from $\dataset$ to $\dataset'$:
+\item\label{step2} We create transactions to copy the population of relations from $\dataset$ to $\dataset'$:
 For every relation $r$ shared by the existing and desired systems, we generate a helper relation: ${\tt copy}_r$, and a transaction to produce its population.
@@
 % \begin{verbatim}
-%    RELATION copied_r[A*B]
-%    RELATION new_r[A*B] [UNI]
-%    ENFORCE copied_r >: new_r /\ old_r
-%    ENFORCE new_r >: old_r - copied_r
+%    RELATION copy.r[A*B]
+%    RELATION new.r[A*B] [UNI]
+%    ENFORCE copy.r >: new.r /\ old.r
+%    ENFORCE new.r >: old.r - copy.r
 % \end{verbatim}
@@
    \end{align}
+   (Step~\ref{step2} contains a problem: How can we be sure that last minute deletes in the existing system are propagated to the desired system?)
 
@@
 %    RELATION fixed_TOTr[A*A]
-%    ENFORCE fixed_TOTr >: I /\ new_r;new_r~
+%    ENFORCE fixed_TOTr >: I /\ new.r;new.r~
 % \end{verbatim}
@@
 % \begin{verbatim}
-%    RULE Block_TOTr : fixed_TOTr |- new_r;new_r~
+%    RULE Block_TOTr : fixed_TOTr |- new.r;new.r~
 %    MESSAGE "Relation r[A*B] must remain total."
@@
 %    ROLE User MAINTAINS TOTr
-%    RULE TOTr : I - fixed_TOTr |- new_r;new_r~
+%    RULE TOTr : I - fixed_TOTr |- new.r;new.r~
 %    MESSAGE "Please, make relation r[A*B] total."

### 2023-10-30 2022Migration/articleMigrationFACS.tex
working on more concrete proof obligations

@@
 
-\subsection{Proof}
+\subsection{Proof Obligations}
 \label{sct:Proof}
@@
 \item Prove that $\migrsys$ is an information system,
-      so the migration engineer can deploy the migration system.
-\item For every relation $r\in\rels_{\schema}\cap\rels_{\schema'}$, prove that $\overrightarrow{r}$ contains all triples from the existing data set in a finite time after deploying $\migrsys$,
-      to ensure that the existing data is migrated to the desired system.
-\item Prove that equation~\ref{eqn:satisfaction} holds for $\migrsys$ when the copying of relations is done (just before MoT),
-      to ensure that a valid information system is presented to users at the moment of transition.
-\item Prove that equation~\ref{eqn:copyingTerminates} represents the condition for terminating the copying process.
-\item Prove that equation~\ref{eqn:fixingTerminates} represents the condition for terminating the fixing of violations.
-\item Prove that the migration engineer can switch the ingress from the existing system to the migration system
-      when equations~\ref{eqn:copyingTerminates} and~\ref{eqn:fixingTerminates} are true,
-      so the computer can let the migration engineer know that it is safe to switch.
+      so the migration engineer can be sure it deploys without failure.
+   \begin{equation}
+      \text{isIS}\pair{\dataset_\infsys}{\schema_\infsys}\ \Rightarrow\ \text{isIS}\pair{\dataset_\infsys}{\schema_\migrsys}
+   \end{equation}
+\item For every relation $r\in\rels_1$ (i.e.
+in $\rels_{\schema}\cap\rels_{\schema'}$), prove that $\overrightarrow{r}$ contains all triples from the existing data set in a finite time after deploying $\migrsys$,
+      to ensure that the existing data is migrated to the desired system:
+   \begin{equation}
+      \triple{a}{\overleftarrow{r}}{b}\in\triples_\infsys\ \Rightarrow\ \triple{a}{\overrightarrow{r}}{b}\in\triples_\migrsys
+   \end{equation}
+\item Prove that the copying process leaves all rules from $\transactions_1$ satisfied forever:
+   \begin{equation}
+      \forall u\in\transactions_1\ \Rightarrow\ \viol{u}{\dataset_\migrsys}=\emptyset
+   \end{equation}
+      This implies that transactions in $\transactions_1$ will never more add any violations to any ${\tt copy}_u$ at the MoC,
+      and they need not exist in the desired system.
+      So, $\rels_1$ and $\transactions_1$ are redundant after the MoC.
+      (\@Bas, is dat voldoende bewijs dat $\rels_1$ redundant is na de MoT?)
+\item Prove that equation~\ref{eqn:copyingTerminates} represents the condition for terminating the copying process,
+      so we can use this condition to safeguard that the migration system is ready for the MoT.
+\item $\rels_2$ contains all violations of new blocking invariants.
+      Prove that the fixing process leaves all rules from $\transactions_2$ satisfied forever:
+   \begin{equation}
+      \forall u\in\transactions_2\ \Rightarrow\ \viol{u}{\dataset_\migrsys}=\emptyset
+   \end{equation}
+      This implies that transactions in $\transactions_2$ need not exist in the desired system.
+      So, $\rels_2$ and $\transactions_2$ are redundant after the MoC.
+\item Prove that equation~\ref{eqn:fixingTerminates} represents the condition for terminating the fixing of violations,
+      so we can use this condition to safeguard that the migration system and desired system are ready for the MoC.
+\item Prove that after the MoC, for every blocking invariant $u\in\rules_{\schema'}$ of the desired system,
+      there is a blocking invariant $v\in\rules_\text{block}\cup(\rules_{\schema}\cap\rules_{\schema'})$
+      with exactly the same signature and violations in the data set of the desired system:
+   \begin{equation}
+      \viol{u}{\dataset'}=\viol{v}{\dataset'}
+   \end{equation}
+      And also vice versa, for every blocking invariant $v\in\rules_\text{block}\cup(\rules_{\schema}\cap\rules_{\schema'})$
+      there must be a blocking invariant $u\in:\rules_{\schema'}$ such that $\viol{u}{\dataset'}=\viol{v}{\dataset'}$.
+      This implies that $\rules_{\schema'}$ can replace $\rules_\text{block}\cup(\rules_{\schema}\cap\rules_{\schema'})$ on the MoC
+      without users noticing the difference.
 \item Assuming that the business can fix each violation in finite time,
@@
       and that it decreases monotonically, so the business can reach the MoC in finite time.
-\item Prove that $\rules_\text{block}$ is free of violations, so it is redundant after the MoC.
-\item Prove that $\transactions_1$ will never more add any violations to any ${\tt copy}_u$ at the MoC, so $\rels_1$ and $\transactions_1$ are redundant after the MoC.
+\item Prove that the copying process leaves all rules from $\rules_\text{block}$ satisfied forever,
+      so these rules need not exist in the desired system:
+   \begin{equation}
+      \forall u\in\rules_\text{block}\ \Rightarrow\ \viol{u}{\dataset_\migrsys}=\emptyset
+   \end{equation}
+      Note that
+\item Prove that $\rels_2$ is redundant after the MoT,
+      so the desired system can do without it.
+   \begin{equation}
+      TBD
+   \end{equation}
+\item Prove that $\overleftarrow{\rels_{\schema}}$ is redundant after the MoT,
+      so the desired system can do without it.
+   \begin{equation}
+      TBD
+   \end{equation}
+\item 
 \item Prove that $\transactions_2$ will never more add any violations to any ${\tt fixed}_u$ at the MoC, so that $\rels_2$ and $\transactions_2$ are redundant after the MoC.
@@
       so the migration engineer can switch the ingress to the desired system and remove the existing system and the migration system.
+   \begin{equation}
+      \text{isIS}\pair{\dataset}{\schema_\migrsys}\ \Rightarrow\ \text{isIS}\pair{\dataset}{\schema'}
+   \end{equation}
 \end{enumerate}

### 2023-10-30 2022Migration/articleMigrationFACS.tex
minor

@@
    The complexity of migrating data to date yields expensive and error-prone migration projects.
-   For that reason, we want to generate this code.
-   To ensure the correctness of the generator is even more challenging.
-   That is why we have proven the correctness of the generator in Isabelle/HOL.
+   By generating this code we can eliminate human induced errors.
+   We enhance the quality of the generator further by proving its correctness in Isabelle/HOL.
 
@@
 \subsection{Algorithm for generating a migration script}
-   (\@Bas, I have added the relevant portions of {\tt Kurk.adl} as \LaTeX\ comments near every equation in the \LaTeX\ source,
-   so we can guard their correspondence.)
+   (\@Bas, ik heb de relevante delen van {\tt Kurk.adl} als \LaTeX\ commentaar bij elke equation in de \LaTeX\ broncode toegevoegd,
+   zodat we in de gaten kunnen houden dat ze met elkaar overeenkomen.)
 

### 2023-10-31 2022Migration/articleMigrationFACS.tex
fix some mistakes

@@
       &\overleftarrow{\rels_{\schema}}\cup\overrightarrow{\rels_{\schema'}}\cup\rels_1\cup\rels_2,\notag\\
-      &\rules_\text{block}\cup(\rules_{\schema}\cap\rules_{\schema'}),\notag\\
-      &\transactions_1\cup\transactions_2\cup\transactions_{\schema'},\notag\\
+      &\rules_\text{block}\cup\overrightarrow{\rules_{\schema}\cap\rules_{\schema'}},\notag\\
+      &\transactions_1\cup\transactions_2\cup\overrightarrow{\transactions_{\schema'}},\notag\\
       &\busConstraints_\text{fix}\cup\busConstraints_{\schema'}\rangle\notag
@@
    This section discusses the proof obligations. The proof is available online.
-   Assume that $\infsys$ and $\infsys'$ are information systems, so they satisfy definition~\ref{def:information system}.
+   Assume that $\infsys=\pair{\dataset}{\schema}$ and $\infsys'=\pair{\dataset'}{\schema'}$ are information systems, so they satisfy definition~\ref{def:information system}.
    This means that equations~\ref{eqn:wellTypedEdge}, \ref{eqn:wellTypedViolation}, \ref{eqn:transaction}, \ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}, \ref{eqn:define R}, \ref{eqn:satisfaction} hold for both systems.
+   We also assume that $\migrsys=\pair{\dataset_\migrsys}{\schema_\migrsys}$ is the migration system generated from $\infsys$ and $\infsys'$.
 \begin{enumerate}
@@
    \begin{equation}
-      \text{isIS}\pair{\dataset_\infsys}{\schema_\infsys}\ \Rightarrow\ \text{isIS}\pair{\dataset_\infsys}{\schema_\migrsys}
+      \text{isIS}\pair{\dataset}{\schema}\ \Rightarrow\ \text{isIS}\pair{\dataset}{\schema_\migrsys}
    \end{equation}
@@
    \begin{equation}
-      \triple{a}{\overleftarrow{r}}{b}\in\triples_\infsys\ \Rightarrow\ \triple{a}{\overrightarrow{r}}{b}\in\triples_\migrsys
+      \triple{a}{r}{b}\in\triples_\infsys\ \Rightarrow\ \triple{a}{\overrightarrow{r}}{b}\in\triples_\migrsys
    \end{equation}
@@
    \end{equation}
-      Note that
-\item Prove that $\rels_2$ is redundant after the MoT,
-      so the desired system can do without it.
-   \begin{equation}
-      TBD
-   \end{equation}
-\item Prove that $\overleftarrow{\rels_{\schema}}$ is redundant after the MoT,
-      so the desired system can do without it.
-   \begin{equation}
-      TBD
-   \end{equation}
-\item 
 \item Prove that $\transactions_2$ will never more add any violations to any ${\tt fixed}_u$ at the MoC, so that $\rels_2$ and $\transactions_2$ are redundant after the MoC.
@@
    \begin{equation}
-      \text{isIS}\pair{\dataset}{\schema_\migrsys}\ \Rightarrow\ \text{isIS}\pair{\dataset}{\schema'}
+      \forall u\in\rules_{\schema'}-\rules_{\schema}.~\viol{\overrightarrow{u}}{\dataset}={\tt fixed}_u\ \Rightarrow\ (\text{isIS}\pair{\dataset}{\schema_\migrsys}\Leftrightarrow\text{isIS}\pair{\dataset}{\schema'})
    \end{equation}

### 2023-11-01 2022Migration/articleMigrationFACS.tex
fixed mistake in the notation of a transactional invariant

@@
    It serves to constrain the data set at runtime.
-   When it is about to be violated, the system produces an error message and refuses the transaction.
-\item Transaction\\
-   The classical database transaction can be seen as an invariant,
+   When the data set changes in a way that violates a blocking invariant, the system produces an error message and refuses the change.
+\item Transactional invariant\\
+   The classical database transaction can be understood as an invariant,
    which the system keeps satisfied by adding or deleting triples to the dataset.
@@
    Summarizing, in our notion of information systems, concepts, relations, and constraints carry the business semantics.
-   Of the three types of constraint, only blocking invariants and transactions are invariants.
+   Of the three types of constraint, only two are invariants.
 
@@
    Now, suppose the data in the existing system does not satisfy $u$.
-   If $u$ is a transaction, the desired system will restore it automatically.
+   If $u$ is a transactional invariant, the desired system will restore it automatically.
    But if it is a blocking invariant, the desired system cannot spin up because all of its invariants must be satisfied.
@@
 
-   In case $u$ is a transaction,
+   In case $u$ is a transactional invariant,
    the system will keep it satisfied by adding the violations to a specific relation $\declare{n}{A}{B}$.
@@
    It is obvious that not every conceivable constraint can satisfy this equation.
-   So, we assume that the compiler restricts the set of transactions to those that satisfy equation~\ref{eqn:transaction}.
+   So, we assume that the compiler restricts the set of transactional invariants to those that satisfy equation~\ref{eqn:transaction}.
    As $\declare{n}{A}{B}$ is specific for $u$, we can write $\enfRel{u}$ for it.
-   We call this the {\em enforced relation} of transaction $u$:
+   We call this the {\em enforced relation} of transactional invariant $u$:
 \begin{equation}
@@
 \end{equation}
-   The language Ampersand has more types of transactions than just this one,
-   but to define this one only is sufficient for this paper.
+   Let us denote a transactional invariant as $r\mapsfrom\violC{u}$ or equivalently $r \mapsfrom \lambda \dataset.~\viol{u}{\dataset}$,
+   in which $r=\enfRel{u}$.
+
+   The language Ampersand has more types of transactional invariants than just this one,
+   but this is sufficient for this paper.
 
@@
    We describe a schema $\schema$ as a tuple $\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$,
-   in which $\concepts\subseteq \Concepts$ is a finite set of concepts,
+   in which $\concepts\subseteq\Concepts$ is a finite set of concepts,
    $\rels\subseteq\Rels$ is a finite set of relations,
    $\rules\subseteq\Constraints$ is a finite set of blocking invariants,
-   $\transactions\subseteq\Constraints$ is a finite set of transactions,
+   $\transactions\subseteq\Constraints$ is a finite set of transactional invariants,
    and $\busConstraints\subseteq\Constraints$ is a finite set of business constraints.
-
-   Let us denote a transaction as $r \mapsfrom u$ or equivalently $r \mapsfrom \lambda \dataset.~ u(\dataset)$,
-   in which $r=\enfRel{u}$.
 
@@
    Requirements~\ref{eqn:relationsIntroduceConcepts} and~\ref{eqn:invariant-has-type} ensure that concepts mentioned in relations and in the signature of rules are part of the schema.
-   Requirement~\ref{eqn:enforcement-has-type} ensures the enforced relation of a transaction is declared in the schema. 
+   Requirement~\ref{eqn:enforcement-has-type} ensures the enforced relation of a transactional invariant is declared in the schema. 
    When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, $\rules_{\schema}$, $\transactions_{\schema}$, $\busConstraints_{\schema}$
@@
 \item All violations must have a type, which follows from  (\ref{eqn:wellTypedViolation}).
-\item Transactions remain satisfied by adding violations to a specific relation (\ref{eqn:transaction}).
-\item All blocking invariants and transactions must remain satisfied:
+\item Transactional invariants remain satisfied by adding violations to a specific relation (\ref{eqn:transaction}).
+\item All invariants must remain satisfied:
    \begin{align}
@@
    \end{align}
-\item\label{step2} We create transactions to copy the population of relations from $\dataset$ to $\dataset'$:
-For every relation $r$ shared by the existing and desired systems, we generate a helper relation: ${\tt copy}_r$, and a transaction to produce its population.
-We also generate a transaction to populate the relation $\overrightarrow{r}$ (in the desired system).
+\item\label{step2} We create transactional invariants to copy the population of relations from $\dataset$ to $\dataset'$:
+For every relation $r$ shared by the existing and desired systems, we generate a helper relation: ${\tt copy}_r$, and a transactional invariant to produce its population.
+We also generate a transactional invariant to populate the relation $\overrightarrow{r}$ (in the desired system).
 % \begin{verbatim}
@@
    For each new blocking invariant $u$, we generate a helper relation: ${\tt fixed}_u$, to register all violations that are fixed.
-   We also need a transaction to produce its population:
+   We also need a transactional invariant to produce its population:
 % \begin{verbatim}
@@
    In some cases, a migration engineer can invent ways to satisfy these invariants automatically.
-   For this purpose, the generator should produce source code (as opposed to compiled code) to allow the migration engineer to replace a business constraint with transactions of her own making.
+   For this purpose, the generator should produce source code (as opposed to compiled code) to allow the migration engineer to replace a business constraint with transactional invariants of her own making.
 \item Let us combine the above into a single migration schema:
@@
    \end{equation}
-      This implies that transactions in $\transactions_1$ will never more add any violations to any ${\tt copy}_u$ at the MoC,
+      This implies that invariants in $\transactions_1$ will never more add any violations to any ${\tt copy}_u$ at the MoC,
       and they need not exist in the desired system.
@@
    \end{equation}
-      This implies that transactions in $\transactions_2$ need not exist in the desired system.
+      This implies that invariants in $\transactions_2$ need not exist in the desired system.
       So, $\rels_2$ and $\transactions_2$ are redundant after the MoC.
@@
 
-W  e use inserting events to model the effect of transactions.
+We use inserting events to model the effect of transactional invariants.
 %   If $R$ is a set of relations that covers those that occur in the rule, then changes to an system $\infsys$ can be described by the event $\infsys \xrightarrow[\{r\}]{R} \infsys'$.
 
-   We first observe that $\migrsys = \pair{\dataset_\migrsys}{\schema_\migrsys}$ is a system after application of the transactions:
-   the only rules are those introduced in equation~\ref{eqn:migrsysrules}, and transactions from equation~\ref{eqn:enforceForRules} ensure that $\popF{{\tt viols}_u} = u$ initially, while $\popF{{\tt fixed}_u}$ is empty.
-   Hence, the MoT can happen after the transactions have been applied, giving us $\migrsys'$.
+   We first observe that $\migrsys = \pair{\dataset_\migrsys}{\schema_\migrsys}$ is a system after application of the transactional invariants:
+   the only rules are those introduced in equation~\ref{eqn:migrsysrules}, and transactional invariants from equation~\ref{eqn:enforceForRules} ensure that $\popF{{\tt viols}_u} = u$ initially, while $\popF{{\tt fixed}_u}$ is empty.
+   Hence, the MoT can happen after the transactional invariants have been applied, giving us $\migrsys'$.
    The event $\migrsys \xrightarrow[\rels_2 \cup \rels_3]{\overrightarrow\rels_{\schema'}}\migrsys'$ represents the MoT.

### 2023-11-05 2022Migration/articleMigrationFACS.tex
minor

@@
 \begin{abstract}
-   Software generators can help increase the frequency of releases and their reliability.
-   They save on time spent on development and fixing human-induced mistakes by compiling a specification into a working information system.
+   Software generators compile a specification into a working information system.
+   This can help increase the frequency of releases and their reliability
+   because the generator saves on development time and prevents human-induced mistakes.
    However, many generators do not support data migrations.
    Data migration is necessary when an incremental deployment changes the system's schema.
-   Consequently, developers tend to avoid migrations or migrate data ``by hand''.
+   As a result, developers tend to avoid migrations or migrate data ``by hand''.
 
@@
    This paper focuses on correctness of the migration. Efficiency of the migration is outside its scope.
-\keywords{Generative software \and Incremental software deployment \and Data migration \and Relation algebra \and Ampersand \and Schema change. \and Invariants \and Zero downtime.}
+\keywords{Generative software \and Incremental software deployment \and Data migration \and Relation algebra \and Ampersand \and Schema change \and Invariants \and Zero downtime}
 \end{abstract}
@@
    Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
-   The risk and effort of such data migrations explains why development teams try to avoid schema changes at all costs.
+   The risk and effort of such data migrations explains why development teams try to avoid schema changes.
 
@@
    and let $\infsys=\pair{\schema}{\dataset}$ and  $\infsys'=\pair{\schema}{\dataset'}$ be information systems,
-t  hen $\infsys\xrightarrow[I]{R} \infsys'$ is an event if and only if:
+   then $\infsys\xrightarrow[I]{R} \infsys'$ is an event if and only if:
 \begin{align}

### 2023-11-12 2022Migration/articleMigrationFACS.tex
Retitled chapter for SJC
WIP
add Proof of Concept (to be elaborated further)

@@
       \viol{u}{\pair{\triples\cup\{\triple{a}{\declare{n}{A}{B}}{b}\mid\pair{a}{b}\in\viol{u}{\dataset}\}}{\instance\cup\instance'}}=\emptyset\\
-      \text{\bf where}\ \instance'=\{ \pair{a}{A}\mid\pair{a}{b}\in\viol{u}{\dataset}\}\cup\{ \pair{b}{B}\mid\pair{a}{b}\in\viol{u}{\dataset}\}
+      \text{\bf with}\ \instance'=\{ \pair{a}{A}\mid\pair{a}{b}\in\viol{u}{\dataset}\}\cup\{ \pair{b}{B}\mid\pair{a}{b}\in\viol{u}{\dataset}\}
    \end{array}
@@
       \begin{array}[t]{l}
-         \text{\bf where}\\
+         \text{\bf with}\\
          \sign{v}=\sign{u}\\
@@
 %    ROLE User MAINTAINS TOTr
-%    RULE TOTr : I - fixed_TOTr |- new.r;new.r~
+%    RULE TOTr : I |- new.r;new.r~
 %    MESSAGE "Please, make relation r[A*B] total."
@@
    \begin{align}
-      \schema_\migrsys=\langle{}&\concepts_\dataset\cup\concepts_{\dataset'},\\
+      \schema_\migrsys=\langle{}&\concepts_\dataset\cup\concepts_{\dataset'},\label{eqn:schema migrsys}\\
       &\overleftarrow{\rels_{\schema}}\cup\overrightarrow{\rels_{\schema'}}\cup\rels_1\cup\rels_2,\notag\\
@@
 \begin{equation}
-   \forall u\in\rules_{\schema'}-\rules_{\schema}.\ \viol{\overleftarrow{u}}{\dataset_\migrsys}={\tt fixed}_u
+   \forall u\in\rules_{\schema'}-\rules_{\schema}.\ \viol{\overleftarrow{u}}{\dataset_\migrsys}={\tt fixed}_u\label{eqn:readyForMoC}
 \end{equation}
@@
 
-\subsection{Proof Obligations}
+\subsection{System properties}
+% TODO: SJC
 \label{sct:Proof}
@@
 \end{enumerate}
-\section{Validation}
+\paragraph{Validation}
    As the system responds to events, it changes its data set, while the schema remains the same.
@@
 
+\section{Proof of Concept}
+\label{sct:PoC}
+   By way of proof of concept (PoC), we have built a migration system in Ampersand.
+   To demonstrate it in the context of this paper, the existing system, $\pair{\dataset}{\schema}$, is rather trivial.
+   It has no constraints and just one relation, $\declare{r}{A}{B}$.
+   Its population is $A=\{a_1,a_2,a_3\}$, $B=\{b_1\}$, and $\pop{r}{\dataset}=\{\pair{a_1}{b_1}\}$.
+   The schema of the migration system, $\schema_\migrsys$, follows from definition~\ref{eqn:schema migrsys}.
+   Figure~\ref{fig:PoC} exhibits four successive screenshots,
+   featuring $\overleftarrow{r}$ as {\small\verb#old_r#},
+   $\overrightarrow{r}$ as {\small\verb#new_r#}.
+\begin{figure}[bht]
+   \begin{center}
+      \makebox[\textwidth][c]{\includegraphics[scale=.07]{figures/screenshots.drawio.png}}%
+   \end{center}
+\caption{Four successive screenshots in the PoC}
+\label{fig:PoC}
+\end{figure}
+
+   Exhibit A shows the migration system just after deployment, at the MoT.
+   It shows that the copying of {\small\verb#old_r#} to {\small\verb#new_r#} has worked.
+   The desired system contains one blocking invariant, which is the totality of $\declare{r}{A}{B}$.
+   The yellow message in exhibit A indicates that a user needs to fix totality for $a_2$ and $a_3$.
+   The column {\small\verb#fixed_u#} contains the elements for which totality is satisfied, so these fields are blocked from becoming empty.
+
+   Exhibit B shows an attempt to remove $\pair{a_1}{b_1}$ from {\small\verb#new_r#}
+   The red message blocks this from happening and the user gets a ``Cancel'' button to roll back the action.
+   Note that the fields for $a_2$ and $a_3$ are empty, which is fine because they will not be blocked until they are given some value.
+
+   Exhibit C shows that the user fills in ``Jill'', which means that $\pair{a_1}{Jill}$ is added to {\small\verb#new_r#}.
+   When the last atom of {tt A} is paired with an atom from {\tt B} (exhibit D), requirement~\ref{eqn:readyForMoC} is satisfied and the prototype informs the user to remove the migration system.
+
 \section{Conclusions}

### 2023-11-13 2022Migration/articleMigrationFACS.tex
two new figures

@@
    Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
-   The risk and effort of such data migrations explains why development teams try to avoid schema changes.
+   In practice, data migrations typically follow Extract-Transform-Load (ETL) patterns~\cite{Theodorou2017},
+   for which many tools are available.
+   However, ETL tools typically provide little support for invariants that change, forcing development teams to write code.
+   The risk and effort of such data migrations explains why these teams try to avoid schema changes.
+   Our research aims at (partly) automating SCDMs to make them less risky and less costly,
+   so development teams can increase the frequency of releases and do schema changes with zero downtime.
 
@@
 
-   In practice, data migrations typically follow Extract-Transform-Load (ETL) patterns~\cite{Theodorou2017},
-   for which many tools are available.
-   However, invariants that change require writing code in the transform phase,
-   for which ETL tools typically provide little support.
-   This yields extra work for software engineers, and it increases the risk of software errors.
-   The ubiquitous pressure to decrease deploy times calls for further automation of SCDMs.
-   With the release frequencies increasing and deploy times decreasing, any downtime for the sake of migration becomes less and less acceptable.
-   That is why our research aims for zero downtime.
-
    Another practical problem is that of data quality.
@@
    Some of that work must be done before the migration; some can wait till after the migration.
-   We can capture part of the data quality problem as a requirement to satisfy semantic constraints.
+   We can capture the automatable part of the data quality problem by regarding it as a requirement to satisfy semantic constraints.
    E.g. the constraint that the combination of street name, house number, postal code, and city occurs in
@@
    This allows us to signal the data pollution at runtime.
-   Some forms of data pollution cannot be detected in this way, however.
+   Some forms of data pollution are not automatable, however.
    An example is when a person has deliberately specified a false name without violating any constraint in the system.
@@
       \begin{center}
-     \includegraphics[scale=.45]{figures/datamigration-Pre-migration.png}
+        \includegraphics[scale=0.8]{figures/existing system.pdf}
       \end{center}
@@
 
+   Figure~\ref{fig:pre-migration} depicts the situation before migration.
+   An existing application service ingests traffic through an ingress and persists data in a data set, which is typically a database.
+   This research assumes that the structure and business semantics are represented in a schema, from which the system is generated.
    Actors (both users and computers) are changing the data in a system continually.
@@
    \begin{center}
-     \includegraphics[scale=.35]{figures/datamigration-Migration-phase.png}
+     \includegraphics[scale=.8]{figures/migration system deployed.pdf}
    \end{center}

### 2023-11-18 2022Migration/articleMigrationFACS.tex
improved the abstract

@@
 \begin{abstract}
-   Software generators compile a specification into a working information system.
-   This can help increase the frequency of releases and their reliability
-   because the generator saves on development time and prevents human-induced mistakes.
-   However, many generators do not support data migrations.
-   Data migration is necessary when an incremental deployment changes the system's schema.
-   As a result, developers tend to avoid migrations or migrate data ``by hand''.
+   Software generators that compile and deploy a specification into a functional information system
+   can help to increase the frequency of releases and their reliability.
+   They achieve this by reducing development time and minimizing human-induced errors.
+   However, a common drawback is the lack of support for data migration,
+   which becomes essential during incremental deployments that alter the system's schema.
+   Consequently, schema-changing data migrations often face challenges, leading developers to resort to manual migration or employ workarounds.
 
-   To address this problem, this paper proposes a theory for data migrations aimed at automating the migration process.
-   The problem at large is how to preserve the semantics of that data under a changing schema.
-   This paper proposes a theory for deploying incremental change.
-   The theory is applicable in general but will be implemented in a software generator called Ampersand.
-
-   This paper defines a migration process that preserves the semantics of data by means of invariants (constraints)
-   with zero downtime for users.
-   The migration process is based on assumptions and requirements that aim to capture a large class of data migrations in practice.
-   This paper focuses on correctness of the migration. Efficiency of the migration is outside its scope.
+   To address this issue, this paper proposes a foundational theory for data migration,
+   aiming to generate migration scripts for automating the migration process.
+   The overarching challenge is preserving the business semantics of data amidst schema changes.
+   Specifically, the paper tackles the task of generating a migration script based on the schemas of both the existing and the desired system with zero downtime.
+   The proposed solution was validated by a prototype demonstrating its efficacy.
+   Notably, the theory is technology-independent, articulating systems in terms of invariants, thereby ensuring applicability across various scenarios.
+   The migration script generator will be implemented in a software generator named Ampersand.
 \keywords{Generative software \and Incremental software deployment \and Data migration \and Relation algebra \and Ampersand \and Schema change \and Invariants \and Zero downtime}
@@
 
-   Another practical problem is that of data quality.
-   Migrations typically suffer from a backlog of deteriorated data, incurring work to clean it up.
+   A practical complication in many data migration projects is the presence of deteriorated data.
+   To clean it up may incur much work.
    Some of that work must be done before the migration; some can wait till after the migration.
+   In all cases, data pollution in an existing system requires careful analysis and planning.
    We can capture the automatable part of the data quality problem by regarding it as a requirement to satisfy semantic constraints.
@@
    a registration of valid addresses can be checked automatically.
-   In a formalism like Ampersand, which allows us to express such constraints, we can add constraints for data quality to the schema.
+   In a formalism like Ampersand~\cite{JoostenRAMiCS2017,Joosten-JLAMP2018}, which allows us to express such constraints, we can add constraints for data quality to the schema.
    This allows us to signal the data pollution at runtime.
@@
 
+   The complexity of data migration has prompted us to develop a theory first,
+   which we present in this contribution.
+   We have validated the theory by prototyping because a formal proof of correctness is beyond our reach.
+   This theory perceives an information system as a data set with constraints, so
+   The existence of a software generator
+
+   for changing schemas in production systems before automating SCDMs.
    The next section analyzes SCDMs with an eye on zero downtime and data quality.
@@
    Section~\ref{sct:Generating} defines the algoritm for generating a migration system, to automate SCDMs.
-   This algorithm has been proven correct in Isabelle/HOL.
-   The outline of this proof is discussed in section~\ref{sct:Proof} and the proof itself is published on the internet.
-
-   For experimental validation we have used the language Ampersand~\cite{JoostenRAMiCS2017,Joosten-JLAMP2018}
+   Section~\ref{sct:PoC} demonstrates the prototype of a migration system, which we used to validate our theory experimentally.
+   For this purpose we have used the language Ampersand
    because its syntax and semantics correspond directly to the definitions in section~\ref{sct:Definitions}.
-   In this way, prototypes built in Ampersand have helped to refine the theory.
 
@@
    An existing application service ingests traffic through an ingress and persists data in a data set, which is typically a database.
-   This research assumes that the structure and business semantics are represented in a schema, from which the system is generated.
+   Our research assumes that the structure and business semantics are represented in a schema, from which the system is generated.
    Actors (both users and computers) are changing the data in a system continually.
@@
    We assume that constraints implement the business semantics of the data.
-   Constraints represent business concerns formally, so they can be checked automatically and be used to generate software.
-   Figure~\ref{fig:pre-migration} depicts them as rules.
-   Some constraints must be satisfied through human intervention, while others are being satisfied by the system.
-   So, we distinguish different kinds of constraint:
+   Constraints represent business concerns formally, so they can be checked automatically and can be used to generate software.
+   Some of the constraints require human intervention, while others require a system to intervene.
+   In this paper, we distinguish three different kinds of constraint:
 \begin{enumerate}
@@
 \item Transactional invariant\\
-   The classical database transaction can be understood as an invariant,
+   The classical database transaction can be understood as a \define{transactional invariant},
    which the system keeps satisfied by adding or deleting triples to the dataset.
    As soon as the data violates this constraint, the system restores it without human intervention.
-   So, the outside world experiences a constraint that is always true, i.e.~an invariant.
+   So, the outside world experiences this as a constraint that is always true, i.e.~an invariant.
 \item Business constraint\\
@@
 \subsection{Ampersand}
+   We use Ampersand as a prototyping language to demonstrate our theory.
+   More importantly, we intend to enhance the Ampersand compiler with the theory presented in this paper,
+   to generate migration systems automatically.
+
    Ampersand is a language that specifies information systems as a system of concepts, relations, and constraints.
-   It features the types of constraints that we need for this paper, so we have used it to try our theory in practice.
-   Ampersand has a compiler that generates the information system from a script.
-   A developer specifies constraints in heterogeneous relation algebra~\cite{Hattensperger1999,Alloy2006}.
+   It features the kinds of constraint we need for this paper, so we have used it to try our theory in practice.
+   Ampersand has a compiler that generates the information system from a script, so it satisfies our research requirement.
+   A developer specifies constraints in heterogeneous relation algebra~\cite{Hattensperger1999, Alloy2006}.
    The generated system keeps the invariants satisfied and signals violations of business constraints to a user.
@@
    The type system of Ampersand~\cite{vdWoude2011} features static typing,
-   which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.
+   which has well-established advantages for the software engineering process~\cite{HanenbergKRTS14, Petersen2014}.
    Constraints carry the business semantics, so they allow us to be explicit about ``preserving the meaning as much as possible''.
@@
    which means that a classical database schema (i.e.\ data structure plus constraints) can be extracted from the Ampersand script.
+
+   Ampersand has been used in practice both in education (Open University of the Netherlands)
+	and in industry (Ordina and TNO-ICT).
+	For example, Ordina designed a proof-of-concept in 2007 of the INDIGO-system.
+	This design was based on Ampersand, to obtain correct, detailed results in the least amount of time.
+	Today INDIGO is in use as the core information system of the Dutch immigration authority, IND.
+	More recently, Ampersand was used to design an information system called DTV for the Dutch Food Authority, NVWA.
+	A prototype of DTV was built in Ampersand and was used as a model to build the actual system.
+	TNO-ICT, a major Dutch industrial research laboratory, is using Ampersand for research purposes.
+	For example, TNO-ICT did a study of international standardization efforts such as
+	RBAC (Role Based Access Control) in 2003 and architecture (IEEE 1471-2000)~\cite{IEEE1471} in 2004.
+	Several inconsistencies were found in the last (draft) RBAC standard~\cite{RBAC}.
+	TNO-ICT has also used the technique in conceiving several patents%
+\footnote{e.g. patents DE60218042D, WO2006126875, EP1727327, WO2004046848, EP1563361, NL1023394C, EP1420323, WO03007571, and NL1013450C.}.
+	At the Open University of the Netherlands, Ampersand is being taught in a course called Rule-based Design~\cite{RBD}.
+	In this course, students use a platform called RAP, which has been built in Ampersand~\cite{Michels2015}.
+	RAP has been the first Ampersand application that has run in production.
 
@@
    By generating this code we can eliminate human induced errors.
-   We enhance the quality of the generator further by proving its correctness in Isabelle/HOL.
 
    This section starts with a presentation of the algorithm.
-   Then we discuss its proof of correctness.
 

### 2023-11-19 2022Migration/articleMigrationFACS.tex
reformulate Ampersand paragraph

@@
 %
-
 \begin{document}
@@
    Software generators that compile and deploy a specification into a functional information system
-   can help to increase the frequency of releases and their reliability.
+   can help to increase the frequency of releases in the software process.
    They achieve this by reducing development time and minimizing human-induced errors.
-   However, a common drawback is the lack of support for data migration,
-   which becomes essential during incremental deployments that alter the system's schema.
+   However, many software generators lack support for data migration.
+   This can inhibit a steady pace of releases, especially for increments that alter the system's schema in production.
    Consequently, schema-changing data migrations often face challenges, leading developers to resort to manual migration or employ workarounds.
@@
    The overarching challenge is preserving the business semantics of data amidst schema changes.
-   Specifically, the paper tackles the task of generating a migration script based on the schemas of both the existing and the desired system with zero downtime.
+   Specifically, this paper tackles the task of generating a migration script based on the schemas of both the existing and the desired system,
+   under the condition of zero downtime.
    The proposed solution was validated by a prototype demonstrating its efficacy.
@@
    After they are built, they need to be updated regularly to keep up with changing requirements in a dynamically evolving environment.
-   Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
-   Obviously, these deployments are mostly updates of existing systems.
-   Schema changes cannot always be avoided when updating software, so a {\em schema changing data migration} (SCDM) will be neccessary from time to time.
+   Schema changes cannot always be avoided when updating software, so a \define{schema-changing data migration} (SCDM) will be neccessary from time to time.
    For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
@@
    However, ETL tools typically provide little support for invariants that change, forcing development teams to write code.
-   The risk and effort of such data migrations explains why these teams try to avoid schema changes.
-   Our research aims at (partly) automating SCDMs to make them less risky and less costly,
-   so development teams can increase the frequency of releases and do schema changes with zero downtime.
+   In a world where automation of the software process is resulting in higher productivity, more frequent releases, and better quality,
+   SCDM's should not stay behind.
+   Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
+   Obviously, these deployments are mostly updates of existing systems.
+   The risk and effort of SCDMs explains why these teams try to avoid schema changes in the first place.
+   Our research aims at automating SCDMs to make them less risky and less costly,
+   so development teams can deploy schema changes with zero downtime.
 
@@
    e.g.~\cite{Gholami2016,Bisbal1999},
-   migration engineers can and will avoid schema changes and functionality changes to avoid introducing new errors in an otherwise error-prone migration process.
-   Another example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014}, defines a migration as a variation between two data structures.
+   migration engineers can and will avoid schema changes and functionality changes
+   to avoid introducing new errors in an otherwise error-prone migration process.
+   In another example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014} define a migration as a variation between two data structures.
    They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
@@
    Then there are SCDMs in situations without a schema or an implicit schema, e.g.~\cite{Hillenbrand2022}.
-   If the schema is not available at compile time, the work will have to be done at runtime.
+   Such situations lack the error preventing power that explicit schemas bring during the development of software.
+   All errors a schema can prevent at compile time must then be compensated by runtime checks,
+   which increases the likelyhood of end-users getting error messages.
    This requires versioned storage of production data and an overhead in performance.
-   This paper focuses on SCDMs in situations with an explicit schema that changes with the data migration.
+   That is why this paper focuses on SCDMs for systems with an explicit schema.
    The prototypical use case for that is to release updates of information systems in production,
@@
    which we present in this contribution.
-   We have validated the theory by prototyping because a formal proof of correctness is beyond our reach.
-   This theory perceives an information system as a data set with constraints, so
-   The existence of a software generator
+   We have validated the theory by prototyping because a formal proof of correctness is currently beyond our reach.
+   This theory perceives an information system as a data set with constraints,
+   so we can represent invariants (and thus the business semantics) directly as constraints.
 
-   for changing schemas in production systems before automating SCDMs.
    The next section analyzes SCDMs with an eye on zero downtime and data quality.
@@
    Events that the system detects may cause the state to change.
-   To keep our theory technology independent, we assume data sets to contain triples.
+   To keep our theory technology independent, we assume that data sets contain triples.
    This makes our theory valid for every kind of database that triples can represent,
@@
    Some of the constraints require human intervention, while others require a system to intervene.
-   In this paper, we distinguish three different kinds of constraint:
+   In this paper, we distinguish three different kinds of constraints:
 \begin{enumerate}
@@
    That takes some time, during which the constraint is violated.
-   So, we do not consider business constraints to be invariant.
+   So, we do not consider business constraints to be invariants.
 \end{enumerate}
@@
 \subsection{Ampersand}
-   We use Ampersand as a prototyping language to demonstrate our theory.
-   More importantly, we intend to enhance the Ampersand compiler with the theory presented in this paper,
+   We employ Ampersand as a prototyping language to demonstrate our theory.
+   More significantly, our intention is to augment the Ampersand compiler with the theory outlined in this paper
    to generate migration systems automatically.
 
-   Ampersand is a language that specifies information systems as a system of concepts, relations, and constraints.
-   It features the kinds of constraint we need for this paper, so we have used it to try our theory in practice.
-   Ampersand has a compiler that generates the information system from a script, so it satisfies our research requirement.
-   A developer specifies constraints in heterogeneous relation algebra~\cite{Hattensperger1999, Alloy2006}.
-   The generated system keeps the invariants satisfied and signals violations of business constraints to a user.
-   Having all invariants explicit in the Ampersand source code makes Ampersand a suitable platform on which to implement our theory.
-   The absence of imperative code in an Ampersand script makes it easier to reason about the system.
-   The type system of Ampersand~\cite{vdWoude2011} features static typing,
-   which has well-established advantages for the software engineering process~\cite{HanenbergKRTS14, Petersen2014}.
-   Constraints carry the business semantics, so they allow us to be explicit about ``preserving the meaning as much as possible''.
-   An Ampersand script contains just enough information to generate a complete system,
-   which means that a classical database schema (i.e.\ data structure plus constraints) can be extracted from the Ampersand script.
+   Ampersand serves as a language for specifying information systems through a framework of concepts, relations, and constraints.
+   It encompasses the necessary constraints discussed in this paper, making it an ideal platform for practical testing of our theory.
+   In Ampersand, developers articulate constraints using heterogeneous relation algebra~\cite{Hattensperger1999,Alloy2006}
+   The systems they generate maintain invariants and alert users to violations of business constraints.
+   Ampersand's explicit representation of all invariants in its source code makes it well-suited for implementing our theory.
+   The absence of imperative code in Ampersand scripts enhances system reasoning,
+   while its static typing~\cite{vdWoude2011} yields the established benefits in software engineering processes~\cite{HanenbergKRTS14,Petersen2014}.
+   Constraints, which carry business semantics, make ``preserving the meaning as much as possible'' explicit.
+   An Ampersand script provides just enough information to generate a complete system,
+   allowing extraction of a classical database schema (i.e., data structure plus constraints) from the Ampersand script.
 
-   Ampersand has been used in practice both in education (Open University of the Netherlands)
-	and in industry (Ordina and TNO-ICT).
-	For example, Ordina designed a proof-of-concept in 2007 of the INDIGO-system.
-	This design was based on Ampersand, to obtain correct, detailed results in the least amount of time.
-	Today INDIGO is in use as the core information system of the Dutch immigration authority, IND.
-	More recently, Ampersand was used to design an information system called DTV for the Dutch Food Authority, NVWA.
-	A prototype of DTV was built in Ampersand and was used as a model to build the actual system.
-	TNO-ICT, a major Dutch industrial research laboratory, is using Ampersand for research purposes.
-	For example, TNO-ICT did a study of international standardization efforts such as
-	RBAC (Role Based Access Control) in 2003 and architecture (IEEE 1471-2000)~\cite{IEEE1471} in 2004.
-	Several inconsistencies were found in the last (draft) RBAC standard~\cite{RBAC}.
-	TNO-ICT has also used the technique in conceiving several patents%
-\footnote{e.g. patents DE60218042D, WO2006126875, EP1727327, WO2004046848, EP1563361, NL1023394C, EP1420323, WO03007571, and NL1013450C.}.
-	At the Open University of the Netherlands, Ampersand is being taught in a course called Rule-based Design~\cite{RBD}.
-	In this course, students use a platform called RAP, which has been built in Ampersand~\cite{Michels2015}.
-	RAP has been the first Ampersand application that has run in production.
+   Ampersand has seen practical use both in education (Open University of the Netherlands) and industry (Ordina and TNO-ICT).
+   For instance, Ordina developed a proof-of-concept of the INDIGO-system in 2007,
+   leveraging Ampersand for accurate results under tight deadlines.
+   Today, INDIGO serves as the core information system for the Dutch immigration authority (IND).
+   More recently, Ampersand played a role in designing the DTV information system for the Dutch Food Authority (NVWA),
+   with a prototype built in Ampersand serving as a model for the actual system.
+   TNO-ICT, a prominent Dutch industrial research laboratory, has utilized Ampersand for various research purposes,
+   including a study of international standardization efforts of RBAC (Role-Based Access Control) in 2003,
+   and a study of IT architecture (IEEE 1471-2000)\cite{IEEE1471} in 2004.
+   Ampersand has also been employed at the Open University of the Netherlands,
+   where it is taught in a course called Rule-Based Design\cite{RBD}.
+   Students in this course utilize a platform named RAP, constructed in Ampersand~\cite{Michels2015},
+   which represents the first Ampersand application to run in production.
 
@@
    To make the case for zero downtime, consider this problem:
-   Suppose we have an invariant, $u$, in the {\em desired system}, which is not part of the {\em existing system}.
-   In the sequel, let us call this a {\em new invariant}.
+   Suppose we have an invariant, $u$, in the \define{desired system}, which is not part of the \define{existing system}.
+   In the sequel, let us call this a \define{new invariant}.
    Now, suppose the data in the existing system does not satisfy $u$.
@@
    To avoid downtime, we must implement new blocking invariants initially as a business constraint,
-   to let users do the work.
+   to let users satisfy them.
+   The moment the last violation of $u$ is fixed, the business constraint can be removed and $u$ can be implemented as a blocking invariant.
+   This is the core idea of our theory.
 
-   For this purpose, we define an intermediate system, the \define{migration system},
+   The \define{migration system} to be generated is an intermediate system,
    which contains all concepts, relations, and constraints of both the existing and the desired system.
@@
 \label{sct:Definitions}
-   An {\em information system} is a combination of data set, schema, and functionality.
+   An \define{information system} is a combination of data set, schema, and functionality.
    For the purpose of this paper, we ignore functionality captured in user interfaces because it does not impact the migration.
@@
    The notation $r=\declare{n}{A}{B}$ denotes that relation $r$ has name $n$, source concept $A$, and target concept $B$.
-   The part $\pair{A}{B}$ is called the {\em signature} of the relation.
+   The part $\pair{A}{B}$ is called the \define{signature} of the relation.
 
@@
    As $\declare{n}{A}{B}$ is specific for $u$, we can write $\enfRel{u}$ for it.
-   We call this the {\em enforced relation} of transactional invariant $u$:
+   We call this the \define{enforced relation} of transactional invariant $u$:
 \begin{equation}

### 2023-11-19 2022Migration/articleMigrationFACS.tex
minor

@@
 \item[moment of transition (MoT)]
-      After the migration system is done copying data, the migration engineer switches all traffic to the migration system
-      and deploys the new functionality at the same time.
-      This is the moment users will notice the difference.
-      Even though the migration system may look like the desired system in the eyes of an average user,
-      it relaxes the blocking invariants and keeps the system operational.
+      After the migration system is done copying data, the migration engineer switches all traffic to the migration system.
+      This is the moment users will notice the difference because the traffic switch also deploys the functionality of the desired system.
+      So, in the eyes of an average user, the migration system may look like the desired system.
+      However, the migration system relaxes the blocking invariants of the desired system until users resolve their violations.
       Since the existing system receives no more traffic, its activity will come to a halt and its data will become static.
@@
       Once the invariants of the desired system are satisfied, the migration engineer switches all traffic to the desired system.
-      After that, she can safely remove both the migration system and the existing system, including their data.
+      The blocking invariants of the desired system are now in effect, so users cannot violate them anymore.
+      After this switch, the migration engineer can safely remove both the migration system and the existing system.
 \end{description}
@@
    By way of proof of concept (PoC), we have built a migration system in Ampersand.
-   To demonstrate it in the context of this paper, the existing system, $\pair{\dataset}{\schema}$, is rather trivial.
+   To demonstrate it in the context of this paper, the existing system, $\pair{\dataset}{\schema}$, is almost trivial.
    It has no constraints and just one relation, $\declare{r}{A}{B}$.

### 2023-11-19 2022Migration/articleMigrationFACS.tex
WIP

@@
 \item\label{step2} We create transactional invariants to copy the population of relations from $\dataset$ to $\dataset'$:
-For every relation $r$ shared by the existing and desired systems, we generate a helper relation: ${\tt copy}_r$, and a transactional invariant to produce its population.
-We also generate a transactional invariant to populate the relation $\overrightarrow{r}$ (in the desired system).
+   For every relation $r$ shared by the existing and desired systems, we generate a helper relation: ${\tt copy}_r$, and two transactional invariants.
+   The first transactional invariant populates relation $r$ and the second populates ${\tt copy}_r$.
+   The helper relation ${\tt copy}_r$ contains the triples that have been copied.
+   We use the helper relation to keep the transactional invariants from immediately repopulating $\overrightarrow{r}$ when a user deletes triples in it.
 % \begin{verbatim}
@@
       \rels_1={}&\{{\tt copy}_r\mid r \in \rels_{\schema'}\cap\rels_{\schema}\}\\
-      \transactions_1={}&\{{\tt copy}_r\mapsfrom \popF{\overrightarrow{r}\cap\overleftarrow{r}} \mid r \in\rels_1\}\cup\{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r}-{\tt copy}_r} \mid r \in\rels_1\}
+      \transactions_1={}&\{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r}-{\tt copy}_r} \mid r \in\rels_1\}\cup\{{\tt copy}_r\mapsfrom \popF{\overrightarrow{r}\cap\overleftarrow{r}} \mid r \in\rels_1\}
    \end{align}
-   The helper relation is needed to ensure that the copying process terminates,
-   which is the case when:
+   The copying process terminates when:
    \begin{align}
@@
    \end{align}
-   (Step~\ref{step2} contains a problem: How can we be sure that last minute deletes in the existing system are propagated to the desired system?)
 
-\item The new blocking invariants are $\rules_{\schema'}-\rules_{\schema}$.
-   For each new blocking invariant $u$, we generate a helper relation: ${\tt fixed}_u$, to register all violations that are fixed.
-   We also need a transactional invariant to produce its population:
-% \begin{verbatim}
-%    RELATION fixed_TOTr[A*A]
-%    ENFORCE fixed_TOTr >: I /\ new.r;new.r~
-% \end{verbatim}
-   \begin{align}
-      \rels_2={}&\{{\tt fixed}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
-      \transactions_2={}&\{{\tt fixed}_u\mapsfrom\lambda\dataset.~\viol{\overleftarrow{u}}{\dataset}-\viol{\overrightarrow{u}}{\dataset}\mid u\in\rules_{\schema'}-\rules_{\schema}\}\label{eqn:enforceForRules}
-   \end{align}
-   The helper relation is needed to ensure that the fixing of violations terminates,
-   which is the case when:
-   \begin{align}
-      \forall u\in\rules_{\schema'}-\rules_{\schema}.~\viol{\overrightarrow{u}}{\dataset}={\tt fixed}_u\label{eqn:fixingTerminates}
-   \end{align}
-\item For each new blocking invariant $u$, we generate another blocking invariant $v$ in the migration system that blocks fixed violations from recurring:
+\item\label{step3} For each new blocking invariant $u$, we generate another blocking invariant $v$ in the migration system that blocks fixed violations from recurring:
 % \begin{verbatim}
@@
    \begin{align}
+      \rels_2={}&\{{\tt fixed}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
       \rules_\text{block}={}&\{v\ 
@@
    Note that the blocking invariants from $\schema'\cap\schema$ are also used in $\migrsys$ to ensure continuity.
-\item To signal users that there are violations that need to be fixed, we generate a business constraint for each new blocking invariant $u$:
+\item\label{step4} The new blocking invariants are $\rules_{\schema'}-\rules_{\schema}$.
+   For each new blocking invariant $u$, we generate a helper relation: ${\tt fixed}_u$, to register all violations that are fixed.
+   We also need a transactional invariant to produce its population:
+% \begin{verbatim}
+%    RELATION fixed_TOTr[A*A]
+%    ENFORCE fixed_TOTr >: I /\ new.r;new.r~
+% \end{verbatim}
+   \begin{align}
+      \transactions_2={}&\{{\tt fixed}_u\mapsfrom\lambda\dataset.~\viol{\overleftarrow{u}}{\dataset}-\viol{\overrightarrow{u}}{\dataset}\mid u\in\rules_{\schema'}-\rules_{\schema}\}\label{eqn:enforceForRules}
+   \end{align}
+   The helper relation is needed to keep violations from reoccurring after they have been fixed.
+   The process is complete when:
+   \begin{align}
+      \forall u\in\rules_{\schema'}-\rules_{\schema}.~\viol{\overrightarrow{u}}{\dataset}={\tt fixed}_u\label{eqn:fixingTerminates}
+   \end{align}
+\item\label{step5} To signal users that there are violations that need to be fixed, we generate a business constraint for each new blocking invariant $u$:
 % \begin{verbatim}
@@
    By way of proof of concept (PoC), we have built a migration system in Ampersand.
-   To demonstrate it in the context of this paper, the existing system, $\pair{\dataset}{\schema}$, is almost trivial.
+   To demonstrate it in the context of this paper, the existing system, $\pair{\dataset}{\schema}$, is rather trivial.
    It has no constraints and just one relation, $\declare{r}{A}{B}$.

### 2023-11-19 2022Migration/articleMigrationFACS.tex
Rename file so that it compiles with SJC
WIP

@@
    Section~\ref{sct:Data sets} describes how we define data sets.
+   Since data sets are sets, we will use set operators $\cup$ (union), $\cap$ (intersect), $-$ (set difference), and an overline $\cmpl{x}$ as complement operator on sets.
    Section~\ref{sct:Constraints} defines constraints and their violations.
@@
 \end{equation}
+   Note that the phrase ``pair $\pair{a}{b}$ is in relation $r$'' means that there is a dataset $\dataset$ in which $\pair{a}{b}\in\pop{r}{\dataset}$,
+   so the phrase ``triple $\triple{a}{r}{b}$ is in $\dataset$'' means the same thing.
 %   Equation~\ref{eqn:wellTypedEdge} implies that for every data set $\dataset$:
@@
 \subsection{Algorithm for generating a migration script}
-   (\@Bas, ik heb de relevante delen van {\tt Kurk.adl} als \LaTeX\ commentaar bij elke equation in de \LaTeX\ broncode toegevoegd,
-   zodat we in de gaten kunnen houden dat ze met elkaar overeenkomen.)
-
    In the migration system, we need to refer to the items (concepts, relations, and constraints) of both the existing system and the desired system.
@@
 \begin{enumerate}
-\item We take a disjoint union of the data sets by relabeling relation names:
+\item We take a disjoint union of the data sets by relabeling relation names, so the migration script can refer to relations from both systems:
    \begin{align}
@@
    For every relation $r$ shared by the existing and desired systems, we generate a helper relation: ${\tt copy}_r$, and two transactional invariants.
-   The first transactional invariant populates relation $r$ and the second populates ${\tt copy}_r$.
-   The helper relation ${\tt copy}_r$ contains the triples that have been copied.
-   We use the helper relation to keep the transactional invariants from immediately repopulating $\overrightarrow{r}$ when a user deletes triples in it.
+   The first transactional invariant populates relation $\overrightarrow{r}$ and the second populates ${\tt copy}_r$.
+   The helper relation ${\tt copy}_r$ contains the pairs that have been copied.
+   We use the helper relation to keep the transactional invariants from immediately repopulating $\overrightarrow{r}$ when a user deletes triples in $\dataset'$.
 % \begin{verbatim}
@@
 
-\item\label{step3} For each new blocking invariant $u$, we generate another blocking invariant $v$ in the migration system that blocks fixed violations from recurring:
+\item\label{step3}
+   The new blocking invariants are $\rules_{\schema'}-\rules_{\schema}$.
+   For each new blocking invariant $u$, we generate a helper relation: ${\tt fixed}_u$, to register all violations that are fixed,
+   and a blocking invariant $v$ in the migration system that blocks fixed violations from recurring:
 % \begin{verbatim}
+%    RELATION fixed_TOTr[A*A]
 %    RULE Block_TOTr : fixed_TOTr |- new.r;new.r~
@@
          \viol{v}{\dataset}=\viol{\overrightarrow{u}}{\dataset}\cap{\tt fixed}_u
-      \end{array}\\
+      \end{array}\label{eqn:blockRule}\\
       &\mid u\in\rules_{\schema'}-\rules_{\schema}\}\notag
    \end{align}
-   Note that the blocking invariants from $\schema'\cap\schema$ are also used in $\migrsys$ to ensure continuity.
-\item\label{step4} The new blocking invariants are $\rules_{\schema'}-\rules_{\schema}$.
-   For each new blocking invariant $u$, we generate a helper relation: ${\tt fixed}_u$, to register all violations that are fixed.
-   We also need a transactional invariant to produce its population:
+   % Note that the blocking invariants from $\schema'\cap\schema$ are also used in $\migrsys$ to ensure continuity.
+\item\label{step4}
+   We use a transactional invariant to produce the population of the helper relation ${\tt fixed}_u$.
 % \begin{verbatim}
-%    RELATION fixed_TOTr[A*A]
 %    ENFORCE fixed_TOTr >: I /\ new.r;new.r~
 % \end{verbatim}
-   \begin{align}
-      \transactions_2={}&\{{\tt fixed}_u\mapsfrom\lambda\dataset.~\viol{\overleftarrow{u}}{\dataset}-\viol{\overrightarrow{u}}{\dataset}\mid u\in\rules_{\schema'}-\rules_{\schema}\}\label{eqn:enforceForRules}
-   \end{align}
-   The helper relation is needed to keep violations from reoccurring after they have been fixed.
-   The process is complete when:
-   \begin{align}
-      \forall u\in\rules_{\schema'}-\rules_{\schema}.~\viol{\overrightarrow{u}}{\dataset}={\tt fixed}_u\label{eqn:fixingTerminates}
-   \end{align}
+   \begin{equation}
+     \transactions_2\ =\ \{{\tt fixed}_u\mapsfrom\lambda\dataset.~\cmpl{\viol{\overrightarrow{u}}{\dataset}}\mid u\in\rules_{\schema'}-\rules_{\schema}\}\label{eqn:enforceForRules}
+   \end{equation}
 \item\label{step5} To signal users that there are violations that need to be fixed, we generate a business constraint for each new blocking invariant $u$:
@@
    In some cases, a migration engineer can invent ways to satisfy these invariants automatically.
-   For this purpose, the generator should produce source code (as opposed to compiled code) to allow the migration engineer to replace a business constraint with transactional invariants of her own making.
+   For this purpose, the generator must produce source code (as opposed to compiled code) to allow the migration engineer
+   to replace a business constraint with transactional invariants of her own making.
+   After all violations are fixed, i.e. when equation~\ref{eqn:fixingTerminates} is satisfied,
+   the migration engineer can switch the ingress to the desired system.
+   This occurs at MoC and
+   replaces $\rules_\text{block}$ in the migration system by the blocking invariants of the desired system.
+   This moment arrives when:
+   \begin{align}
+      \forall u\in\rules_{\schema'}-\rules_{\schema}.~\viol{\overrightarrow{u}}{\dataset}\subseteq{\tt fixed}_u\label{eqn:fixingTerminates}
+   \end{align}
+   After this, the migration engineer can remove the migration system and the old system.
+
 \item Let us combine the above into a single migration schema:
@@
    \end{align}
+   This schema represents the migration system.
+   In our reasoning, we have only used information from the schemas of the existing system and the desired system.
+   This shows that it can be generated from these schemas without using any knowledge of the data sets.
 \end{enumerate}
-   The migration engineer can switch all traffic to the desired system
-   after resolving the violations that prohibit deploying the desired system.
-   That is the case when violations of new invariants on the old population have all been fixed:
-% \begin{verbatim}
-%    ROLE User MAINTAINS CleanUp_TOTr
-%    RULE CleanUp_TOTr : V[ONE*A] ; (I - fixed_TOTr) ; V[A*ONE]
-%    MESSAGE "Now you can remove the migration system because relation r[A*B] is total."
-% \end{verbatim}
-\begin{equation}
-   \forall u\in\rules_{\schema'}-\rules_{\schema}.\ \viol{\overleftarrow{u}}{\dataset_\migrsys}={\tt fixed}_u\label{eqn:readyForMoC}
-\end{equation}
-   After this, the migration engineer can remove the migration system and the old system.
 
@@
    \begin{center}
-      \makebox[\textwidth][c]{\includegraphics[scale=.07]{figures/screenshots.drawio.png}}%
+      \makebox[\textwidth][c]{\includegraphics[scale=.07]{figures/screenshots.png}}%
    \end{center}

### 2023-11-25 2022Migration/articleMigrationFoIKS.tex
renamed for FoIKS

--- 2023-11-19 2022Migration/articleMigrationFACS.tex
@@
    \begin{align}
-      \rels_1={}&\{{\tt copy}_r\mid r \in \rels_{\schema'}\cap\rels_{\schema}\}\\
+      \rels_1={}&\{{\tt copy}_r\mid r \in \rels_{\schema'}\cap\rels_{\schema}\}\label{eqn:copy relations}\\
       \transactions_1={}&\{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r}-{\tt copy}_r} \mid r \in\rels_1\}\cup\{{\tt copy}_r\mapsfrom \popF{\overrightarrow{r}\cap\overleftarrow{r}} \mid r \in\rels_1\}
@@
       \begin{array}[t]{l}
-         \text{\bf with}\\
+         \text{\bf with}\label{eqn:Bfix}\\
          \sign{v}=\sign{u}\\
@@
    to replace a business constraint with transactional invariants of her own making.
-   After all violations are fixed, i.e. when equation~\ref{eqn:fixingTerminates} is satisfied,
+   After all violations are fixed, i.e. when equation~\ref{eqn:readyForMoC} is satisfied,
    the migration engineer can switch the ingress to the desired system.
@@
    \begin{align}
-      \forall u\in\rules_{\schema'}-\rules_{\schema}.~\viol{\overrightarrow{u}}{\dataset}\subseteq{\tt fixed}_u\label{eqn:fixingTerminates}
+      \forall u\in\rules_{\schema'}-\rules_{\schema}.~\viol{\overrightarrow{u}}{\dataset}\subseteq{\tt fixed}_u
+   \label{eqn:readyForMoC}
    \end{align}
@@
 \end{enumerate}
-
-\subsection{System properties}
-% TODO: SJC
-\label{sct:Proof}
-   A theorem prover, Isabelle/HOL, is used to establish the correctness of the presented algorithm.
-   This section discusses the proof obligations. The proof is available online.
-   Assume that $\infsys=\pair{\dataset}{\schema}$ and $\infsys'=\pair{\dataset'}{\schema'}$ are information systems, so they satisfy definition~\ref{def:information system}.
-   This means that equations~\ref{eqn:wellTypedEdge}, \ref{eqn:wellTypedViolation}, \ref{eqn:transaction}, \ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}, \ref{eqn:define R}, \ref{eqn:satisfaction} hold for both systems.
-   We also assume that $\migrsys=\pair{\dataset_\migrsys}{\schema_\migrsys}$ is the migration system generated from $\infsys$ and $\infsys'$.
-\begin{enumerate}
-\item Prove that $\migrsys$ is an information system,
-      so the migration engineer can be sure it deploys without failure.
-   \begin{equation}
-      \text{isIS}\pair{\dataset}{\schema}\ \Rightarrow\ \text{isIS}\pair{\dataset}{\schema_\migrsys}
-   \end{equation}
-\item For every relation $r\in\rels_1$ (i.e.
-in $\rels_{\schema}\cap\rels_{\schema'}$), prove that $\overrightarrow{r}$ contains all triples from the existing data set in a finite time after deploying $\migrsys$,
-      to ensure that the existing data is migrated to the desired system:
-   \begin{equation}
-      \triple{a}{r}{b}\in\triples_\infsys\ \Rightarrow\ \triple{a}{\overrightarrow{r}}{b}\in\triples_\migrsys
-   \end{equation}
-\item Prove that the copying process leaves all rules from $\transactions_1$ satisfied forever:
-   \begin{equation}
-      \forall u\in\transactions_1\ \Rightarrow\ \viol{u}{\dataset_\migrsys}=\emptyset
-   \end{equation}
-      This implies that invariants in $\transactions_1$ will never more add any violations to any ${\tt copy}_u$ at the MoC,
-      and they need not exist in the desired system.
-      So, $\rels_1$ and $\transactions_1$ are redundant after the MoC.
-      (\@Bas, is dat voldoende bewijs dat $\rels_1$ redundant is na de MoT?)
-\item Prove that equation~\ref{eqn:copyingTerminates} represents the condition for terminating the copying process,
-      so we can use this condition to safeguard that the migration system is ready for the MoT.
-\item $\rels_2$ contains all violations of new blocking invariants.
-      Prove that the fixing process leaves all rules from $\transactions_2$ satisfied forever:
-   \begin{equation}
-      \forall u\in\transactions_2\ \Rightarrow\ \viol{u}{\dataset_\migrsys}=\emptyset
-   \end{equation}
-      This implies that invariants in $\transactions_2$ need not exist in the desired system.
-      So, $\rels_2$ and $\transactions_2$ are redundant after the MoC.
-\item Prove that equation~\ref{eqn:fixingTerminates} represents the condition for terminating the fixing of violations,
-      so we can use this condition to safeguard that the migration system and desired system are ready for the MoC.
-\item Prove that after the MoC, for every blocking invariant $u\in\rules_{\schema'}$ of the desired system,
-      there is a blocking invariant $v\in\rules_\text{block}\cup(\rules_{\schema}\cap\rules_{\schema'})$
-      with exactly the same signature and violations in the data set of the desired system:
-   \begin{equation}
-      \viol{u}{\dataset'}=\viol{v}{\dataset'}
-   \end{equation}
-      And also vice versa, for every blocking invariant $v\in\rules_\text{block}\cup(\rules_{\schema}\cap\rules_{\schema'})$
-      there must be a blocking invariant $u\in:\rules_{\schema'}$ such that $\viol{u}{\dataset'}=\viol{v}{\dataset'}$.
-      This implies that $\rules_{\schema'}$ can replace $\rules_\text{block}\cup(\rules_{\schema}\cap\rules_{\schema'})$ on the MoC
-      without users noticing the difference.
-\item Assuming that the business can fix each violation in finite time,
-      prove that the number of violations to be fixed by the business is finite,
-      and that it decreases monotonically, so the business can reach the MoC in finite time.
-\item Prove that the copying process leaves all rules from $\rules_\text{block}$ satisfied forever,
-      so these rules need not exist in the desired system:
-   \begin{equation}
-      \forall u\in\rules_\text{block}\ \Rightarrow\ \viol{u}{\dataset_\migrsys}=\emptyset
-   \end{equation}
-\item Prove that $\transactions_2$ will never more add any violations to any ${\tt fixed}_u$ at the MoC, so that $\rels_2$ and $\transactions_2$ are redundant after the MoC.
-\item Prove that $\busConstraints_\text{fix}$ is free of violations, so it is redundant after the MoC.
-\item Prove that $\concepts_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$ are redundant after the MoC.
-\item Prove that, given the redundancies proven above, the migration system is equivalent to $\infsys'$ after the MoC,
-      so the migration engineer can switch the ingress to the desired system and remove the existing system and the migration system.
-   \begin{equation}
-      \forall u\in\rules_{\schema'}-\rules_{\schema}.~\viol{\overrightarrow{u}}{\dataset}={\tt fixed}_u\ \Rightarrow\ (\text{isIS}\pair{\dataset}{\schema_\migrsys}\Leftrightarrow\text{isIS}\pair{\dataset}{\schema'})
-   \end{equation}
-\end{enumerate}
-\paragraph{Validation}
-   As the system responds to events, it changes its data set, while the schema remains the same.
-   In this paper, it suffices to define an event as a pair of systems for which the schema stays constant.
-   Events are categorized by the part of the data set that changes.
-
-\begin{definition}[Event]
-   Let $R \subseteq \Rels$ and $I \subseteq \Rels$ be sets of relations,
-   and let $\infsys=\pair{\schema}{\dataset}$ and  $\infsys'=\pair{\schema}{\dataset'}$ be information systems,
-   then $\infsys\xrightarrow[I]{R} \infsys'$ is an event if and only if:
-\begin{align}
-   \triple{a}{r}{b}\in\triples_{\dataset}-\triples_{\dataset'}&\Rightarrow\ r\in R\label{eqn:eventUnchanged}\\
-   \triple{a}{r}{b}\in\triples_{\dataset'}-\triples_{\dataset}&\Rightarrow\ r\in(R \cup I)\label{eqn:eventInsert}
-\end{align}
-% was:
-% \begin{align}
-%    \triples_{\dataset} - (\Triple{\Atoms}{(R \cup I)}{\Atoms}) &= \triples_{\dataset'} - (\Triple{\Atoms}{(R \cup I)}{\Atoms})
-% \label{eqn:eventUnchanged}\\
-%    \triples_{\dataset} - (\Triple{\Atoms}{R}{\Atoms}) &\subseteq \triples_{\dataset'} - (\Triple{\Atoms}{R}{\Atoms})
-% \label{eqn:eventInsert}
-% \end{align}
-   We say that event $\infsys\xrightarrow[I]{R} \infsys'$ brings the system $\infsys$ to $\infsys'$ by inserting pairs in relations from $I$ and deleting pairs in relations from $R$.
-\end{definition}
-
-   Equation~\ref{eqn:eventUnchanged} and~\ref{eqn:eventInsert} state that the triples of those for relations in $I$ or $R$ are the only ones that can be changed, and that only triples in $R$ can be removed.
-%If $I$ or $R$ is the empty set, we omit it from the arrow, so $\infsys \xrightarrow{R} \infsys'$ is a notation for $\infsys \xrightarrow[\emptyset]{R} \infsys'$.
-
-\begin{definition}[Inserting event]
-   We write $\infsys \xrightarrowdbl[I]{R} \infsys'$ to state that some triple was inserted whose relation was in $I$, that is:
-   $t \in \triples_{\dataset'} \cap (\Triple{\Atoms}{I}{\Atoms})$ and $t \not\in \triples_{\dataset}$ for some triple $t$.
-\end{definition}
-
-We use inserting events to model the effect of transactional invariants.
-%   If $R$ is a set of relations that covers those that occur in the rule, then changes to an system $\infsys$ can be described by the event $\infsys \xrightarrow[\{r\}]{R} \infsys'$.
-
-   We first observe that $\migrsys = \pair{\dataset_\migrsys}{\schema_\migrsys}$ is a system after application of the transactional invariants:
-   the only rules are those introduced in equation~\ref{eqn:migrsysrules}, and transactional invariants from equation~\ref{eqn:enforceForRules} ensure that $\popF{{\tt viols}_u} = u$ initially, while $\popF{{\tt fixed}_u}$ is empty.
-   Hence, the MoT can happen after the transactional invariants have been applied, giving us $\migrsys'$.
-   The event $\migrsys \xrightarrow[\rels_2 \cup \rels_3]{\overrightarrow\rels_{\schema'}}\migrsys'$ represents the MoT.
-
-   The MoC is reached for data set $\dataset'$ whenever $\pop{{\tt viols}_u - {\tt fixed}_u}{\dataset'}=\emptyset$ for all rules.
-   This should happen after a finite number of inserting events: let $I = \{{\tt fixed}_u \mid u\in\rules_{\schema'}\}$ be the relations that keep track of which violations have been fixed, then there is no infinite chain of systems $\migrsys' = \migrsys_0, \migrsys_1,\ldots$ such that $\migrsys_i\xrightarrowdbl[I]{\overrightarrow\rels_{\schema'}}\migrsys_{i+1}$.
-   This shows that when users get rid of violations, there is real progress: they can only do this a finite number of times.
-   Next we ask ourselves: does this mean we always reach the MoC?
-
-   Suppose there is a system in which all violations have been fixed, that is, suppose there is a $\migrsys^C$ such that $\pop{{\tt fixed}_u}{\dataset_{\migrsys^C}}=\pop{{\tt viols}_u}{\dataset_{\migrsys^C}}$ and $\migrsys'\xrightarrow[I]{\overrightarrow\rels_{\schema'}} \migrsys^C$.
-   In that case, after any steps done by users (both those with and without the migration helper role), we can reach it, that is: for any $\migrsys_i$, $\migrsys'\xrightarrow[I]{\overrightarrow\rels_{\schema'}} \migrsys_i$ implies $\migrsys_i\xrightarrow[I]{\overrightarrow\rels_{\schema'}} \migrsys^C$.
-   This implies that if the MoC is initially reachable through $\xrightarrow[I]{\overrightarrow\rels_{\schema'}}$ events, then it will remain reachable.
-   So is the MoC initially reachable?
-
-   For this last question, we need to add a condition:
-   There needs to be a system that has the desired schema $\schema'$.
-   If rules in $\schema'$ are such that they cannot all be satisfied, regardless of the data set, no such system exists and we have no hope of reaching it.
-   We assume that the rules of $\schema'$ are intended to prevent data pollution, such that a data set $\dataset'$ describing the `real world' would be one that satisfies them.
-   So let $\pair{\dataset'}{\schema'}$ be a system, then we can get to it through our migration system.
-   To show this, we construct a migration system $\migrsys^C$ satisfying $\pop{{\tt fixed}_u}{\dataset_{\migrsys^C}}=\pop{{\tt viols}_u}{\dataset_{\migrsys^C}}$ and $\migrsys'\xrightarrow[I]{\overrightarrow\rels_{\schema'}} \migrsys^C$.
-   We can do so by simply stating its triples, since the schema is determined by $\migrsys'$:
-
-\begin{align}
-   \begin{aligned}
-   \triples_{\migrsys^C} &= \triples_\migrsys 
-   \cup  \{\triple{a}{{\tt viols}_u}{b} \mid \pair{a}{b}\in \viol{u}{\dataset}, u\in \rules_{\schema'} \} \\
-   &\cup \{\triple{a}{{\tt fixed}_u}{b} \mid \pair{a}{b}\in \viol{u}{\dataset}, u\in \rules_{\schema'} \}
-   \cup \{\triple{a}{\overrightarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_{\dataset'}\}   
-   \end{aligned}
-\end{align}
-
-   Collectively, this shows that the MoC is and remains reachable after a finite number of inserting events by users in the migration helper role.
 
@@
 
+\subsection{Validation}
+   The proof of concept gives but one example of something that works.
+   To validate that this works in all cases, we must verify that:
+\begin{enumerate}
+\item The predeployment yields a migration system that copies the designated triples to the desired system in finite time.
+\item The migration system satisfies the definition of information system (\ref{def:information system}) at MoT, especially that it has no blocking violations (requirement~\ref{eqn:satisfaction}) to ensure a successful deployment.
+\item The violations that business actors must fix are finite in number and decrease monotonically, so the business can fulfil the condition for the MoC (requirement~\ref{eqn:readyForMoC}) in finite time.
+\item The behavior of the migration system and the desired system is identical, so moving the ingress from the migration system to the desired system is not noticable for end-users.
+\end{enumerate}
+   Let us discuss these points one by one.
+
+   The relations to be copied from the old system are those relations that the desired system retains ($\rels_{\schema'}\cap\rels_{\schema}$).
+   For each $r$, $\rels_1$ contains a relation ${\tt copy}_r$ (eqn.~\ref{eqn:copy relations}).
+   After the MoT, no more changes will occur in the existing system.
+   In other words, the population of every relation in $\rels_{\schema}$ becomes stable and so does every ${\tt copy}_r$.
+   When the migration system has copied those relations,
+   every ${\tt copy}_r$ is populated and eqn.~\ref{eqn:copyingTerminates}) is true.
+   Hence, the migration system will make no more changes to ${\tt copy}_r$ and the transactional invariants in $\transactions_1$ will cause no more changes.
+Effectively, $\transactions_1$ becomes redundant once the copying is done.
+
+   Then, the definition of the migration system, def.~\ref{dfn:migration system}, contains only blocking invariants that exist in the existing system as well.
+   For this reason, all blocking invariants of the migration system are satisfied on MoT.
+   Since it contains no other blocking invariants, the migration system satisfies requirement~\ref{eqn:satisfaction}.
+
+   Thirdly, the problematic constraints are the blocking invariants of the desired system that were not in the existing system ($\rules_{\schema'}-\rules_{\schema}$ in def.~\ref{eqn:Bfix}).
+   The migration engineer might introduce transactional invariants to satisfy these invariants automatically,
+   but otherwise, definition~\ref{eqn:Bfix} defines them as business constraints to be resolved by the business.
+   Each violation in those rules that a business actor resolves will never reappear,
+   because it is registered in ${\tt fixed}_r$.
+   So when all violations are resolved, the constraints in ... have effectively become blocking constraints.
 \section{Conclusions}

### 2023-11-25 2022Migration/articleMigrationFoIKS.tex
wip

@@
    \begin{align}
-      \rels_2={}&\{{\tt fixed}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
+      \rels_2={}&\{{\tt fixed}_u\mid u \in\overrightarrow{\rules_{\schema'}-\rules_{\schema}}\}\\
       \rules_\text{block}={}&\{v\ 
@@
          \sign{v}=\sign{u}\\
-         \viol{v}{\dataset}=\viol{\overrightarrow{u}}{\dataset}\cap{\tt fixed}_u
+         \viol{v}{\dataset}=\viol{u}{\dataset}\cap{\tt fixed}_u
       \end{array}\label{eqn:blockRule}\\
-      &\mid u\in\rules_{\schema'}-\rules_{\schema}\}\notag
+      &\mid u\in\overrightarrow{\rules_{\schema'}-\rules_{\schema}}\}\notag
    \end{align}
@@
    \begin{equation}
-     \transactions_2\ =\ \{{\tt fixed}_u\mapsfrom\lambda\dataset.~\cmpl{\viol{\overrightarrow{u}}{\dataset}}\mid u\in\rules_{\schema'}-\rules_{\schema}\}\label{eqn:enforceForRules}
+     \transactions_2\ =\ \{{\tt fixed}_u\mapsfrom\lambda\dataset.~\cmpl{\viol{u}{\dataset}}\mid u\in\overrightarrow{\rules_{\schema'}-\rules_{\schema}}\}\label{eqn:enforceForRules}
    \end{equation}
@@
          \sign{v}=\sign{u}\\
-         \viol{v}{\dataset}=\viol{\overrightarrow{u}}{\dataset}-{\tt fixed}_u
+         \viol{v}{\dataset}=\viol{u}{\dataset}-{\tt fixed}_u
       \end{array}\\
-      &\mid u\in\rules_{\schema'}-\rules_{\schema}\}\notag
+      &\mid u\in\overrightarrow{\rules_{\schema'}-\rules_{\schema}}\}\notag
    \end{align}
@@
       &\transactions_1\cup\transactions_2\cup\overrightarrow{\transactions_{\schema'}},\notag\\
-      &\busConstraints_\text{fix}\cup\busConstraints_{\schema'}\rangle\notag
+      &\busConstraints_\text{fix}\cup\overrightarrow{\busConstraints_{\schema'}}\rangle\notag
    \end{align}
@@
 \begin{enumerate}
-\item The predeployment yields a migration system that copies the designated triples to the desired system in finite time.
-\item The migration system satisfies the definition of information system (\ref{def:information system}) at MoT, especially that it has no blocking violations (requirement~\ref{eqn:satisfaction}) to ensure a successful deployment.
+\item The predeployment yields a migration system, $\migrsys$, that copies the designated triples to the desired system in finite time.
+\item $\migrsys$ satisfies the definition of information system (\ref{def:information system}) at MoT, especially that it has no violations of blocking invariants (requirement~\ref{eqn:satisfaction}) to ensure a successful deployment.
 \item The violations that business actors must fix are finite in number and decrease monotonically, so the business can fulfil the condition for the MoC (requirement~\ref{eqn:readyForMoC}) in finite time.
-\item The behavior of the migration system and the desired system is identical, so moving the ingress from the migration system to the desired system is not noticable for end-users.
+\item The behavior of $\migrsys$ and the desired system is identical, so moving the ingress from $\migrsys$ to the desired system is not noticable for end-users.
 \end{enumerate}
@@
    The relations to be copied from the old system are those relations that the desired system retains ($\rels_{\schema'}\cap\rels_{\schema}$).
-   For each $r$, $\rels_1$ contains a relation ${\tt copy}_r$ (eqn.~\ref{eqn:copy relations}).
-   After the MoT, no more changes will occur in the existing system.
+   For each $r$ to be copied, $\migrsys$ contains a relation ${\tt copy}_r$ in $\rels_1$ (eqn.~\ref{eqn:copy relations}).
+   After the MoT, the ingress sends all change events to $\migrsys$, so the existing system can finish the work it is doing for transactional invariants and will not change after that.
    In other words, the population of every relation in $\rels_{\schema}$ becomes stable and so does every ${\tt copy}_r$.
-   When the migration system has copied those relations,
-   every ${\tt copy}_r$ is populated and eqn.~\ref{eqn:copyingTerminates}) is true.
-   Hence, the migration system will make no more changes to ${\tt copy}_r$ and the transactional invariants in $\transactions_1$ will cause no more changes.
+   At that point in time, eqn.~\ref{eqn:copyingTerminates} is true and stays true.
    Effectively, $\transactions_1$ becomes redundant once the copying is done.
 
-   Then, the definition of the migration system, def.~\ref{dfn:migration system}, contains only blocking invariants that exist in the existing system as well.
-   For this reason, all blocking invariants of the migration system are satisfied on MoT.
-   Since it contains no other blocking invariants, the migration system satisfies requirement~\ref{eqn:satisfaction}.
+   Then, $\migrsys$ contains only blocking invariants that exist in the existing system as well (def.~\ref{dfn:migration system}).
+   For this reason, all of these blocking invariants are satisfied on MoT.
+   Since $\migrsys$ contains no other blocking invariants, it satisfies requirement~\ref{eqn:satisfaction}.
 
@@
    The migration engineer might introduce transactional invariants to satisfy these invariants automatically,
-   but otherwise, definition~\ref{eqn:Bfix} defines them as business constraints to be resolved by the business.
-   Each violation in those rules that a business actor resolves will never reappear,
-   because it is registered in ${\tt fixed}_r$.
-   So when all violations are resolved, the constraints in ... have effectively become blocking constraints.
+   but otherwise, definition~\ref{eqn:Bfix} defines them as business constraints, $\busConstraints_\text{fix}$, to be resolved by the business.
+   Each violation in $\busConstraints_\text{fix}$ that a business actor resolves,
+   will never reappear because it is registered in ${\tt fixed}_r$ by the transactional invariants of $\transactions_2$ (eqn.~\ref{eqn:enforceForRules}).
+
+   So finally, when all violations are resolved, the constraints in $\busConstraints_\text{fix}$ have effectively become blocking invariants.
+   The blocking invariants in the desired system consist of $\rules_{\schema}$ and $\busConstraints_\text{fix}$, which is equivalent to $\rules_{\schema'}$.
+   Hence we can replace $\busConstraints_\text{fix}\cup\rules_{\schema}$ with $\rules_{\schema'}$ after condition~\ref{eqn:readyForMoC} is satisfied.
+
+   Now we can assemble the results.
+   Every rule in $\rules_\text{block}$ has become blocking, so after the MoC $\rules_\text{block}\cup\overrightarrow{\rules_{\schema}\cap\rules_{\schema'}}=\overrightarrow{\rules_{\schema'}}$ in the migration system, which is $\rules_{\schema'}$ in the desired system.
+   Since $\transactions_1$ and $\transactions_2$ are redundant after the MoC,
+   we can retain $\overrightarrow{\transactions_{\schema'}}$ in $\migrsys$,
+   which is equivalent to $\transactions_{\schema'}$ in the desired system.
+   Likewise, $\busConstraints_\text{fix}$ has become redundant, so $\migrsys$ can do with just $\overrightarrow{\busConstraints_{\schema'}}$.
+   In the desired system, that is equivalent to $\busConstraints_{\schema'}$.
+   So, the constraints in the desired system after the MoC are equivalent to the constraints in the MoC
+
 \section{Conclusions}

### 2023-11-28 2022Migration/articleMigrationFoIKS.tex
in essence submittable

@@
 \newcommand{\pop}[2]{\popF{#1}(#2)}
-\newcommand{\maintain}{\mathbin{\id{maint}}}
 \newcommand{\enf}{\mathbin{{\tt enforce}}}
 \newcommand{\enforce}[2]{{\tt enforce}_{#1}(#2)}
+\newcommand{\enforceC}[1]{{\tt enforce}_{#1}}
 \newcommand{\instance}{\mathbin{\id{inst}}}
@@
       \begin{center}
-        \includegraphics[scale=0.8]{figures/existing system.pdf}
+        \includegraphics[scale=0.8]{figures/existing_system.pdf}
       \end{center}
@@
 \item Blocking invariant\\
-   A \define{blocking invariant} is a constraint that is always true in a system.
+   A \define{blocking invariant} is a constraint that is always satisfied in a system.
    It serves to constrain the data set at runtime.
@@
    As soon as the data violates this constraint, the system restores it without human intervention.
-   So, the outside world experiences this as a constraint that is always true, i.e.~an invariant.
+   So, the outside world experiences this as a constraint that is always satisfied, i.e.~an invariant.
 \item Business constraint\\
@@
    Ampersand serves as a language for specifying information systems through a framework of concepts, relations, and constraints.
-   It encompasses the necessary constraints discussed in this paper, making it an ideal platform for practical testing of our theory.
+   It comprises the three types of constraints discussed in this paper, making it an ideal platform for practical testing of our theory.
    In Ampersand, developers articulate constraints using heterogeneous relation algebra~\cite{Hattensperger1999,Alloy2006}
-   The systems they generate maintain invariants and alert users to violations of business constraints.
-   Ampersand's explicit representation of all invariants in its source code makes it well-suited for implementing our theory.
-   The absence of imperative code in Ampersand scripts enhances system reasoning,
+   The systems they generate keep invariants satisfied and alert users to violations of business constraints.
+   The absence of imperative code in Ampersand scripts enhances reasoning about the system,
    while its static typing~\cite{vdWoude2011} yields the established benefits in software engineering processes~\cite{HanenbergKRTS14,Petersen2014}.
-   Constraints, which carry business semantics, make ``preserving the meaning as much as possible'' explicit.
+   Constraints carry business semantics, which makes ``preserving the meaning as much as possible'' explicit.
    An Ampersand script provides just enough information to generate a complete system,
@@
 \item users must experience zero downtime, to enable more frequent SCDMs.
-\item users must be able to fix invariants in the new system.
+\item users must be able to restore invariants in the new system.
    So, we need an intermediate ``migration system'' that implements blocking constraints of the desired system as business constraints,
@@
 \item the number of violations that users must fix is finite and decreases (monotonically) over time,
-   to ensure that the migration system does not need to live ever after.
+   to ensure that the migration system does not need to be kept alive infinitely.
 \end{enumerate}
@@
    \begin{center}
-     \includegraphics[scale=.8]{figures/migration system deployed.pdf}
+     \includegraphics[scale=.8]{figures/migration_system_deployed.pdf}
    \end{center}
@@
    It uses transactions to copy the data and to resolve some forms of data pollution.
-   Not all migration work, however, can be automated.
+   Not all of the work, however, can be automated.
    Data pollution, new business rules, or known issues in the existing system
@@
    An \define{information system} is a combination of data set, schema, and functionality.
-   For the purpose of this paper, we ignore functionality captured in user interfaces because it does not impact the migration.
+   For the purpose of this paper, we ignore functionality captured in user interfaces to focus on the transition of business semantics in the data.
    Section~\ref{sct:Data sets} describes how we define data sets.
@@
    In this paper, variables $u$ and $v$ represent all three types of constraints.
-   We say that a constraint, $u$, is satisfied when $u$ is true for all triples in the data set.
+   We say that a constraint, $u$, is satisfied when it is true for all triples in the data set.
    For every constraint $u$, function $\violC{u}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$ produces the violations of $u$,
@@
       \viol{u}{\pair{\triples\cup\{\triple{a}{\declare{n}{A}{B}}{b}\mid\pair{a}{b}\in\viol{u}{\dataset}\}}{\instance\cup\instance'}}=\emptyset\\
-      \text{\bf with}\ \instance'=\{ \pair{a}{A}\mid\pair{a}{b}\in\viol{u}{\dataset}\}\cup\{ \pair{b}{B}\mid\pair{a}{b}\in\viol{u}{\dataset}\}
+      \text{\bf with\ }\begin{array}[t]{lcl}
+         \instance&=&\{\pair{a}{A}\mid\pair{a}{b}\in\viol{u}{\dataset}\}\cup\{ \pair{b}{B}\mid\pair{a}{b}\in\viol{u}{\dataset}\}\\
+         \pair{A}{B}&=&\sign{u}
+      \end{array}
    \end{array}
@@
 \item All violations must have a type, which follows from  (\ref{eqn:wellTypedViolation}).
-\item Transactional invariants remain satisfied by adding violations to a specific relation (\ref{eqn:transaction}).
+\item The system keeps any transactional invariant $u$ satisfied by adding violations to the relation $\enforceC{u}$ (\ref{eqn:transaction}).
 \item All invariants must remain satisfied:
@@
    We assume that a deployment will fail if these requirements are not satisfied.
-   We will write $\text{isIS}(\infsys)$ to denote that $\infsys$ is an information system.
 
@@
    The complexity of migrating data to date yields expensive and error-prone migration projects.
-   By generating this code we can eliminate human induced errors.
+   By generating this code we can prevent many human induced errors.
 
@@
    The notation $\overrightarrow{\infsys}$ is defined similarly, using prefix ``new.'' instead of ``old.''.
+   This relabeling does not have any effect on the behavior of the system, i.e. $\infsys$, $\overleftarrow{\infsys}$, and $\overrightarrow{\infsys}$ are indistinguishable for end-users.
 
@@
    \begin{align}
-      \rels_1={}&\{{\tt copy}_r\mid r \in \rels_{\schema'}\cap\rels_{\schema}\}\label{eqn:copy relations}\\
-      \transactions_1={}&\{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r}-{\tt copy}_r} \mid r \in\rels_1\}\cup\{{\tt copy}_r\mapsfrom \popF{\overrightarrow{r}\cap\overleftarrow{r}} \mid r \in\rels_1\}
+      \rels_1={}&\{{\tt copy}_r\mid r\in\rels_{\schema'}\cap\rels_{\schema}\}\label{eqn:copy relations}\\
+      \transactions_1={}&\{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r}-{\tt copy}_r} \mid r\in\rels_{\schema'}\cap\rels_{\schema}\}\ \cup\\
+      {}&\{{\tt copy}_r\mapsfrom \popF{\overrightarrow{r}\cap\overleftarrow{r}} \mid r\in\rels_{\schema'}\cap\rels_{\schema}\}\notag
    \end{align}
@@
    \begin{align}
-      \forall r\in\rels_1.~\overleftarrow{r}={\tt copy}_r\label{eqn:copyingTerminates}
+      \forall r\in\rels_{\schema'}\cap\rels_{\schema}.~\overleftarrow{r}={\tt copy}_r\label{eqn:copyingTerminates}
    \end{align}
@@
    \begin{align}
-      \forall u\in\rules_{\schema'}-\rules_{\schema}.~\viol{\overrightarrow{u}}{\dataset}\subseteq{\tt fixed}_u
+      \forall u\in\overrightarrow{\rules_{\schema'}-\rules_{\schema}}.~\cmpl{\viol{u}{\dataset}}\subseteq{\tt fixed}_u
    \label{eqn:readyForMoC}
@@
    Its population is $A=\{a_1,a_2,a_3\}$, $B=\{b_1\}$, and $\pop{r}{\dataset}=\{\pair{a_1}{b_1}\}$.
+   The desired system contains one blocking invariant, which is the totality of $\declare{r}{A}{B}$.
    The schema of the migration system, $\schema_\migrsys$, follows from definition~\ref{eqn:schema migrsys}.
+
    Figure~\ref{fig:PoC} exhibits four successive screenshots,
@@
    \begin{center}
-      \makebox[\textwidth][c]{\includegraphics[scale=.07]{figures/screenshots.png}}%
+      \includegraphics[scale=0.9]{figures/screenshots2x2.pdf}
    \end{center}
@@
    It shows that the copying of {\small\verb#old_r#} to {\small\verb#new_r#} has worked.
-   The desired system contains one blocking invariant, which is the totality of $\declare{r}{A}{B}$.
    The yellow message in exhibit A indicates that a user needs to fix totality for $a_2$ and $a_3$.
@@
    Exhibit C shows that the user fills in ``Jill'', which means that $\pair{a_1}{Jill}$ is added to {\small\verb#new_r#}.
-   When the last atom of {tt A} is paired with an atom from {\tt B} (exhibit D), requirement~\ref{eqn:readyForMoC} is satisfied and the prototype informs the user to remove the migration system.
+
+   Exhibit D: When the last atom of {tt A} is paired with an atom from {\tt B}, requirement~\ref{eqn:readyForMoC} is satisfied and the prototype informs the user to remove the migration system.
 
@@
    The proof of concept gives but one example of something that works.
-   To validate that this works in all cases, we must verify that:
+   In fact, having built prototypes increases our confidence, but cannot serve as a proof.
+   This section argues the validity of our method.
+   We take advantage of the formal definition of the generator (section~\ref{sct:Generating}) to precisely state the assumptions and requirements of its validity.
+
+   The initial situation consists of one existing system $\infsys$ and one desired system $\infsys'$ of which we have a schema, $\schema_{\infsys'}$ and an initial dataset, $\dataset_{\infsys'}$.
+   We may assume that $\infsys$ satisfies definition~\ref{def:information system} because it is a system in production.
+   Also, $\schema_{\infsys'}$ satisfies equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type} because the script of the desired system is type-correct.
+   Together, the schema and the intial dataset forms the desired system $\pair{\schema_{\infsys'}}{\dataset_{\infsys'}}$, which satisfies definition~\ref{def:information system}.
+   With these assumptions in place,
+   we must verify that:
 \begin{enumerate}
-\item The predeployment yields a migration system, $\migrsys$, that copies the designated triples to the desired system in finite time.
-\item $\migrsys$ satisfies the definition of information system (\ref{def:information system}) at MoT, especially that it has no violations of blocking invariants (requirement~\ref{eqn:satisfaction}) to ensure a successful deployment.
-\item The violations that business actors must fix are finite in number and decrease monotonically, so the business can fulfil the condition for the MoC (requirement~\ref{eqn:readyForMoC}) in finite time.
-\item The behavior of $\migrsys$ and the desired system is identical, so moving the ingress from $\migrsys$ to the desired system is not noticable for end-users.
+\item After predeployment, the migration system $\migrsys$ copies the designated triples to the desired system in finite time.
+\item At MoT, $\migrsys$ satisfies the definition of information system (\ref{def:information system}), especially that it has no violations of blocking invariants (requirement~\ref{eqn:satisfaction}) to ensure a successful deployment.
+\item Once business actors have fixed the new business invariants, i.e. when the condition for the MoC (requirement~\ref{eqn:readyForMoC}) is satisfied,
+   the behavior of $\migrsys$ and the desired system is identical, so moving the ingress from $\migrsys$ to the desired system is not noticable for end-users.
+\item The old system has become redundant, so we can remove $\infsys$ and $\migrsys$.
 \end{enumerate}
@@
 
-   The relations to be copied from the old system are those relations that the desired system retains ($\rels_{\schema'}\cap\rels_{\schema}$).
-   For each $r$ to be copied, $\migrsys$ contains a relation ${\tt copy}_r$ in $\rels_1$ (eqn.~\ref{eqn:copy relations}).
+   The relations to be copied from $\infsys$ are those relations that the desired system retains: $\rels_{\schema'}\cap\rels_{\schema}$.
+   For each $r$ to be copied from $\infsys$, $\migrsys$ contains a relation ${\tt copy}_r$ in $\rels_1$ (eqn.~\ref{eqn:copy relations}).
    After the MoT, the ingress sends all change events to $\migrsys$, so the existing system can finish the work it is doing for transactional invariants and will not change after that.
    In other words, the population of every relation in $\rels_{\schema}$ becomes stable and so does every ${\tt copy}_r$.
-   At that point in time, eqn.~\ref{eqn:copyingTerminates} is true and stays true.
+   At that point in time, eqn.~\ref{eqn:copyingTerminates} is satisfied and stays satisfied.
    Effectively, $\transactions_1$ becomes redundant once the copying is done.
@@
    Since $\migrsys$ contains no other blocking invariants, it satisfies requirement~\ref{eqn:satisfaction}.
+   This implies that $\migrsys$ works and the migration engineer may safely switch the ingress from $\infsys$ to $\migrsys$.
 
    Thirdly, the problematic constraints are the blocking invariants of the desired system that were not in the existing system ($\rules_{\schema'}-\rules_{\schema}$ in def.~\ref{eqn:Bfix}).
-   The migration engineer might introduce transactional invariants to satisfy these invariants automatically,
-   but otherwise, definition~\ref{eqn:Bfix} defines them as business constraints, $\busConstraints_\text{fix}$, to be resolved by the business.
+   $\migrsys$ features $\busConstraints_\text{fix}$ to represent these rules in the guise of business constraints.
+   This lets business actors resolve the violations.
    Each violation in $\busConstraints_\text{fix}$ that a business actor resolves,
    will never reappear because it is registered in ${\tt fixed}_r$ by the transactional invariants of $\transactions_2$ (eqn.~\ref{eqn:enforceForRules}).
+   When all violations are fixed, every rule in $\rules_\text{block}$ has become blocking.
+   So, after the MoC, $\rules_\text{block}\cup\overrightarrow{\rules_{\schema}\cap\rules_{\schema'}}$ equals $\overrightarrow{\rules_{\schema'}}$ in the migration system.
+   % Since $\overrightarrow{\rules_{\schema'}}$ and $\rules_{\schema'}$ have the same behaviour for the end-user, the migration engineer can safely switch the ingress from $\migrsys$ to $\infsys'$ (the desired system).
+
+   The rules in the desired system are partly written to resolve data pollution.
+   In some cases, the migration engineer wants users to get rid of that pollution and turn the rule in a blocking invariant as described above.
+   However, some of these constraints might be satisfiable automatically.
+   In that case, the migration engineer might substitute such constraints from $\busConstraints_\text{fix}$ by transactional invariants in $\transactions_{\migrsys}$.
+   These invariants don't need the mechanism described above, because the migration system itself will take care that all constraints in $\transactions_{\migrsys}$ are satisfied.
+   Both cases need to be resolved at MoC.
 
@@
 
-   Now we can assemble the results.
-   Every rule in $\rules_\text{block}$ has become blocking, so after the MoC $\rules_\text{block}\cup\overrightarrow{\rules_{\schema}\cap\rules_{\schema'}}=\overrightarrow{\rules_{\schema'}}$ in the migration system, which is $\rules_{\schema'}$ in the desired system.
+   Now we can assemble the results at MoC.
+   The above reasoning shows that at MoC
+   $\langle\concepts_\dataset\cup\concepts_{\dataset'},
+   \overleftarrow{\rels_{\schema}}\cup\overrightarrow{\rels_{\schema'}}\cup\rels_1\cup\rels_2,
+   \rules_\text{block}\cup\overrightarrow{\rules_{\schema}\cap\rules_{\schema'}},
+   \transactions_1\cup\transactions_2\cup\overrightarrow{\transactions_{\schema'}},
+   \busConstraints_\text{fix}\cup\overrightarrow{\busConstraints_{\schema'}}\rangle$ is equivalent to 
+   $\langle\concepts_{\dataset'}, \overrightarrow{\rels_{\schema'}}, \overrightarrow{\rules_{\schema'}}, \overrightarrow{\transactions_{\schema'}}, \overrightarrow{\busConstraints_{\schema'}}\rangle$,
+   which is equal to $\overrightarrow{\schema'}$.
+   So from MoC onwards, $\pair{\dataset_{\migrsys}}{\schema_{\migrsys}}$ is equivalent to $\pair{\dataset_{\migrsys}}{\overrightarrow{\schema'}}$.
+   Hence, we can tell that $\dataset_{\migrsys}$ is a valid dataset for the desired system, so we can switch the ingress from the migration system to the desired system without users noticing the difference.
+   Then, the migration system gets no more inputs, so it can be removed. 
+   Since this is equivalent to 
    Since $\transactions_1$ and $\transactions_2$ are redundant after the MoC,

### 2023-12-07 2022Migration/articleMigrationFoIKS.tex
typo

@@
 
-   Exhibit D: When the last atom of {tt A} is paired with an atom from {\tt B}, requirement~\ref{eqn:readyForMoC} is satisfied and the prototype informs the user to remove the migration system.
+   Exhibit D: When the last atom of {\tt A} is paired with an atom from {\tt B}, requirement~\ref{eqn:readyForMoC} is satisfied and the prototype informs the user to remove the migration system.
 

### 2023-12-10 2022Migration/articleMigrationFoIKS.tex
Final touch

@@
 %\def\id#1{\text{\it #1\/}}
+\newcommand{\ourtheory}{approach}
+\newcommand{\foundtheory}{foundational approach}
+
 \newcommand{\xrightarrowdbl}[2][]{%
@@
 \newcommand{\sign}[1]{\id{sign}_{#1}}
-\newcommand{\enfRel}[1]{\id{enf}_{#1}}
+\newcommand{\enfRel}[1]{\enforceC{#1}}
 \newcommand{\signature}[2]{\langle{#1},{#2}\rangle}
@@
 \begin{abstract}
-   Software generators that compile and deploy a specification into a functional information system
+   Software generators that compile and deploy a specification into a functioning information system
    can help to increase the frequency of releases in the software process.
@@
 
-   To address this issue, this paper proposes a foundational theory for data migration,
+   To address this issue, this paper proposes a \foundtheory{} for data migration,
    aiming to generate migration scripts for automating the migration process.
@@
    The proposed solution was validated by a prototype demonstrating its efficacy.
-   Notably, the theory is technology-independent, articulating systems in terms of invariants, thereby ensuring applicability across various scenarios.
+   Notably, the \ourtheory{} is technology-independent, articulating systems in terms of invariants, thereby ensuring applicability across various scenarios.
    The migration script generator will be implemented in a software generator named Ampersand.
@@
    After they are built, they need to be updated regularly to keep up with changing requirements in a dynamically evolving environment.
-   Schema changes cannot always be avoided when updating software, so a \define{schema-changing data migration} (SCDM) will be neccessary from time to time.
+   Schema changes cannot always be avoided when updating software, so a \define{schema-changing data migration} (SCDM) will be necessary from time to time.
    For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
@@
    The risk and effort of SCDMs explains why these teams try to avoid schema changes in the first place.
-   Our research aims at automating SCDMs to make them less risky and less costly,
-   so development teams can deploy schema changes with zero downtime.
+   We believe that software teams that realize high update frequencies need better tools to perform SCDM's reliably with zero-downtime.
+   We want then to design, implement, and deploy SCDMs in the flow of their DevOps practice.
 
@@
 
-   The complexity of data migration has prompted us to develop a theory first,
+   The complexity of data migration has prompted us to develop an \ourtheory{} first,
    which we present in this contribution.
-   We have validated the theory by prototyping because a formal proof of correctness is currently beyond our reach.
-   This theory perceives an information system as a data set with constraints,
+   We have validated the \ourtheory{} by prototyping because a formal proof of correctness is currently beyond our reach.
+   This \ourtheory{} perceives an information system as a data set with constraints,
    so we can represent invariants (and thus the business semantics) directly as constraints.
@@
    Section~\ref{sct:Generating} defines the algoritm for generating a migration system, to automate SCDMs.
-   Section~\ref{sct:PoC} demonstrates the prototype of a migration system, which we used to validate our theory experimentally.
+   Section~\ref{sct:PoC} demonstrates the prototype of a migration system, which we used to validate our \ourtheory{} experimentally.
    For this purpose we have used the language Ampersand
@@
    Events that the system detects may cause the state to change.
-   To keep our theory technology independent, we assume that data sets contain triples.
-   This makes our theory valid for every kind of database that triples can represent,
+   To keep our \ourtheory{} technology independent, we assume that data sets contain triples.
+   This makes our \ourtheory{} valid for every kind of database that triples can represent,
    including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
@@
    The classical database transaction can be understood as a \define{transactional invariant},
-   which the system keeps satisfied by adding or deleting triples to the dataset.
+   which the system keeps satisfied by adding or deleting triples to the data set.
    As soon as the data violates this constraint, the system restores it without human intervention.
@@
 \subsection{Ampersand}
-   We employ Ampersand as a prototyping language to demonstrate our theory.
-   More significantly, our intention is to augment the Ampersand compiler with the theory outlined in this paper
-   to generate migration systems automatically.
-
+   We employ Ampersand as a prototyping language to demonstrate our \ourtheory{}.
+   %More significantly, our intention is to augment the Ampersand compiler with the \ourtheory{} outlined in this paper
+   %to generate migration systems automatically.
+   % Opmerking van SJC: ik denk dat deze intentie hier het best achterwege gelaten kan worden,
+   %   het zet de lezer direct op het spoor van: en waarom is dit nog niet in Ampersand geimplementeerd dan? Is dit niet af?
+   %
    Ampersand serves as a language for specifying information systems through a framework of concepts, relations, and constraints.
-   It comprises the three types of constraints discussed in this paper, making it an ideal platform for practical testing of our theory.
+   It comprises the three types of constraints discussed in this paper, making it an ideal platform for practical testing of our \ourtheory{}.
    In Ampersand, developers articulate constraints using heterogeneous relation algebra~\cite{Hattensperger1999,Alloy2006}
@@
    The moment the last violation of $u$ is fixed, the business constraint can be removed and $u$ can be implemented as a blocking invariant.
-   This is the core idea of our theory.
+   This is the core idea of our \ourtheory{}.
 
@@
    However, it implements the blocking invariants of the desired system as business constraints.
-   This migration system must also ensure that every violation that is fixed is blocked from recurring.
+   This migration system must also prevent every violation that is fixed from recurring.
    In this way, the business constraint gradually turns into a blocking invariant, satisfying the specification of the desired system.
@@
       The migration system starts copying data from the existing system.
-\item[moment of transition (MoT)]
-      After the migration system is done copying data, the migration engineer switches all traffic to the migration system.
-      This is the moment users will notice the difference because the traffic switch also deploys the functionality of the desired system.
-      So, in the eyes of an average user, the migration system may look like the desired system.
-      However, the migration system relaxes the blocking invariants of the desired system until users resolve their violations.
+\item[Moment of transition (MoT)]
+      After the migration system is done copying data, the migration engineer switches the ingress to relay all traffic to the migration system instead of the existing system.
+      This is the moment users will notice the difference because users experience the functionality of the migration system.
+      The migration system relaxes the blocking invariants of the desired system until users resolve their violations.
+      So it behaves like the desired system in the eyes of an average user, with some violations yet to be resolved.
       Since the existing system receives no more traffic, its activity will come to a halt and its data will become static.
-      The migration system stays active until all invariants of the desired system are satisfied and it can take over the work from the migration system.
-\item[moment of completion (MoC)]
+      The migration system stays active until the desired system can take over.
+\item[Moment of completion (MoC)]
       Once the invariants of the desired system are satisfied, the migration engineer switches all traffic to the desired system.
@@
    because their changes are copied by the migration system too.
-   However, after the MoT there must be no new changes in the existing system
+   However, after the MoT there must be no new changes in the existing system (by then the ``old'' system)
    to avoid violations of new invariants that the migration system has already fixed.
@@
    Atoms serve as data elements.
-   %Atoms are values without internal structure of interest, meant to represent atomic data elements (e.g. dates, strings, numbers, etc.) in a database.
-   %From a business perspective, atoms represent concrete items of the world,
-   %such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
-   %By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ represent \emph{atoms}.
    All atoms are taken from an infinite set called $\Atoms$.
@@
    Moreover, \atom{Peter}, \atom{Melissa} and \atom{074238991} are atoms.
-   In the sequel, variables $A$, $B$, $C$, $D$ will represent concepts, and variables $a$, $b$, and $c$ represent \emph{atoms}.
+   In the sequel, variables $A$, $B$, and $C$ will represent concepts, and variables $a$, and $b$ represent \emph{atoms}.
    %
@@
    Relations serve to organize and store data, allowing developers to represent facts.
-   In this paper, variables $r$, $s$, and $d$ represent relations\footnote{Some readers might like to read `relation symbol' where we write `relation'.}.
+   In this paper, variable $r$ represents relations\footnote{Some readers might like to read `relation symbol' where we write `relation'.}.
    All relations are taken from an infinite set $\Rels$.
@@
 \end{equation}
-   Note that the phrase ``pair $\pair{a}{b}$ is in relation $r$'' means that there is a dataset $\dataset$ in which $\pair{a}{b}\in\pop{r}{\dataset}$,
+   Note that the phrase ``pair $\pair{a}{b}$ is in relation $r$'' means that there is a data set $\dataset$ in which $\pair{a}{b}\in\pop{r}{\dataset}$,
    so the phrase ``triple $\triple{a}{r}{b}$ is in $\dataset$'' means the same thing.
@@
    In case $u$ is a transactional invariant,
-   the system will keep it satisfied by adding the violations to a specific relation $\declare{n}{A}{B}$.
+   the system will keep it satisfied by adding the violations to a specific relation $\declare{n}{A}{B}$ such that $\pair{A}{B}=\sign{u}$.
+   This requires that adding the pair to that relation solves the violation:
 \begin{equation}
    \begin{array}[t]{l}
-      \viol{u}{\pair{\triples\cup\{\triple{a}{\declare{n}{A}{B}}{b}\mid\pair{a}{b}\in\viol{u}{\dataset}\}}{\instance\cup\instance'}}=\emptyset\\
-      \text{\bf with\ }\begin{array}[t]{lcl}
-         \instance&=&\{\pair{a}{A}\mid\pair{a}{b}\in\viol{u}{\dataset}\}\cup\{ \pair{b}{B}\mid\pair{a}{b}\in\viol{u}{\dataset}\}\\
-         \pair{A}{B}&=&\sign{u}
+      (a,b)\in \viol{u}{\pair{\triples}{\instance}}\, \Longrightarrow \\
+      \quad \viol{u}{\pair{\triples \cup \{\triple{a}{\declare{n}{A}{B}}{b}\}}{\instance}} = \viol{u}{\pair{\triples}{\instance}} - \{(a,b)\}
    \end{array}
-   \end{array}
+   % SJC's opmerking:  \instance \cup \{\pair{a}{A},\pair{b}{B}\} = \instance ivm eqn:wellTypedViolation
 \label{eqn:transaction}
@@
    The language Ampersand has more types of transactional invariants than just this one,
-   but this is sufficient for this paper.
+   but this one is sufficient for this paper.
 
@@
    Errors the compiler detects are prohibitive for generating code,
-   to prevent a substantial class of mistakes to ever reach end users.
+   to prevent a substantial class of mistakes to ever reach end-users.
 
@@
 \label{sct:Generating}
-   The complexity of migrating data to date yields expensive and error-prone migration projects.
-   By generating this code we can prevent many human induced errors.
+   To prevent expensive and error-prone migration projects,
+   we define a migration script that can be generated by a compiler.
 
-   This section starts with a presentation of the algorithm.
-
-\subsection{Algorithm for generating a migration script}
    In the migration system, we need to refer to the items (concepts, relations, and constraints) of both the existing system and the desired system.
@@
          \sign{v}=\sign{u}\\
-         \viol{v}{\dataset}=\viol{u}{\dataset}\cap{\tt fixed}_u
+         \viol{v}{\dataset}=\viol{u}{\dataset}\cap\pop{{\tt fixed}_u}{\dataset}
       \end{array}\label{eqn:blockRule}\\
@@
    \begin{equation}
-     \transactions_2\ =\ \{{\tt fixed}_u\mapsfrom\lambda\dataset.~\cmpl{\viol{u}{\dataset}}\mid u\in\overrightarrow{\rules_{\schema'}-\rules_{\schema}}\}\label{eqn:enforceForRules}
+     \transactions_2\ =\ \{{\tt fixed}_u\mapsfrom\lambda\dataset.~\cmpl{\viol{u}{\dataset} \cup \pop{{\tt fixed}_u}{\dataset}}\mid u\in\overrightarrow{\rules_{\schema'}-\rules_{\schema}}\}\label{eqn:enforceForRules}
+     % Opmerking van SJC: de {\tt fixed}_u aan de rechterkant is nodig om aan (\ref{eqn:transaction}) te voldoen.
    \end{equation}
@@
          \sign{v}=\sign{u}\\
-         \viol{v}{\dataset}=\viol{u}{\dataset}-{\tt fixed}_u
+         \viol{v}{\dataset}=\viol{u}{\dataset} % -{\tt fixed}_u %opmerking van SJC: de - fixed is overbodig omdat eqn:blockRule geldt
       \end{array}\\
@@
    \begin{align}
-      \forall u\in\overrightarrow{\rules_{\schema'}-\rules_{\schema}}.~\cmpl{\viol{u}{\dataset}}\subseteq{\tt fixed}_u
+      \forall u\in\overrightarrow{\rules_{\schema'}-\rules_{\schema}}.~{\viol{u}{\dataset}}\subseteq \pop{{\tt fixed}_u}{\dataset}
    \label{eqn:readyForMoC}
    \end{align}
+   Equivalently, $\forall u\in\overrightarrow{\rules_{\schema'}-\rules_{\schema}}.~\viol{u}{\dataset} = \emptyset$.
 After this, the migration engineer can remove the migration system and the old system.
@@
    This section argues the validity of our method.
-   We take advantage of the formal definition of the generator (section~\ref{sct:Generating}) to precisely state the assumptions and requirements of its validity.
+   We take advantage of the formal definition of the generated migration system (section~\ref{sct:Generating}) to precisely state the assumptions and requirements of its validity.
 
-   The initial situation consists of one existing system $\infsys$ and one desired system $\infsys'$ of which we have a schema, $\schema_{\infsys'}$ and an initial dataset, $\dataset_{\infsys'}$.
+   The initial situation consists of one existing system $\infsys$ and one desired system $\infsys'$ of which we have a schema, $\schema_{\infsys'}$ and an initial data set, $\dataset_{\infsys'}$.
    We may assume that $\infsys$ satisfies definition~\ref{def:information system} because it is a system in production.
    Also, $\schema_{\infsys'}$ satisfies equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type} because the script of the desired system is type-correct.
-   Together, the schema and the intial dataset forms the desired system $\pair{\schema_{\infsys'}}{\dataset_{\infsys'}}$, which satisfies definition~\ref{def:information system}.
+   Together, the schema and the intial data set forms the desired system $\pair{\schema_{\infsys'}}{\dataset_{\infsys'}}$, which satisfies definition~\ref{def:information system}.
    With these assumptions in place,
@@
 
-   Thirdly, the problematic constraints are the blocking invariants of the desired system that were not in the existing system ($\rules_{\schema'}-\rules_{\schema}$ in def.~\ref{eqn:Bfix}).
+   Thirdly, the constraints that may need human intervention are the blocking invariants of the desired system that were not in the existing system ($\rules_{\schema'}-\rules_{\schema}$ in def.~\ref{eqn:Bfix}).
    $\migrsys$ features $\busConstraints_\text{fix}$ to represent these rules in the guise of business constraints.
@@
    The rules in the desired system are partly written to resolve data pollution.
-   In some cases, the migration engineer wants users to get rid of that pollution and turn the rule in a blocking invariant as described above.
-   However, some of these constraints might be satisfiable automatically.
-   In that case, the migration engineer might substitute such constraints from $\busConstraints_\text{fix}$ by transactional invariants in $\transactions_{\migrsys}$.
-   These invariants don't need the mechanism described above, because the migration system itself will take care that all constraints in $\transactions_{\migrsys}$ are satisfied.
-   Both cases need to be resolved at MoC.
+   In some cases, the migration engineer wants users to get rid of that pollution, as described above.
+   However, some data pollution might be satisfiable automatically.
+   In that case, the migration engineer might substitute the generated business constraint from $\busConstraints_\text{fix}$ by a transactional invariant in $\transactions_{\migrsys}$ to resolve it without bothering the business.
+   Of course, both cases need to be resolved at MoC.
 
@@
    So from MoC onwards, $\pair{\dataset_{\migrsys}}{\schema_{\migrsys}}$ is equivalent to $\pair{\dataset_{\migrsys}}{\overrightarrow{\schema'}}$.
-   Hence, we can tell that $\dataset_{\migrsys}$ is a valid dataset for the desired system, so we can switch the ingress from the migration system to the desired system without users noticing the difference.
+   Hence, we can tell that $\dataset_{\migrsys}$ is a valid data set for the desired system, so we can switch the ingress from the migration system to the desired system without users noticing the difference.
    Then, the migration system gets no more inputs, so it can be removed.
-   Since this is equivalent to 
-   Since $\transactions_1$ and $\transactions_2$ are redundant after the MoC,
-   we can retain $\overrightarrow{\transactions_{\schema'}}$ in $\migrsys$,
-   which is equivalent to $\transactions_{\schema'}$ in the desired system.
-   Likewise, $\busConstraints_\text{fix}$ has become redundant, so $\migrsys$ can do with just $\overrightarrow{\busConstraints_{\schema'}}$.
-   In the desired system, that is equivalent to $\busConstraints_{\schema'}$.
-   So, the constraints in the desired system after the MoC are equivalent to the constraints in the MoC
 

### 2023-12-10 2022Migration/articleMigrationFoIKS.tex
WIP

@@
    Section~\ref{sct:Generating} defines the algoritm for generating a migration system, to automate SCDMs.
-   Section~\ref{sct:PoC} demonstrates the prototype of a migration system, which we used to validate our \ourtheory{} experimentally.
-   For this purpose we have used the language Ampersand
+   Section~\ref{sct:PoC} demonstrates the prototype of a migration system, which we used to verify our \ourtheory{} experimentally.
+   We have used the language Ampersand for this purpose,
    because its syntax and semantics correspond directly to the definitions in section~\ref{sct:Definitions}.
+   Finally, section~\ref{sct:Validation} discusses the validation of our \ourtheory{} by showing that all requirements are met.
 
@@
 
-   Exhibit D: When the last atom of {\tt A} is paired with an atom from {\tt B}, requirement~\ref{eqn:readyForMoC} is satisfied and the prototype informs the user to remove the migration system.
+   Exhibit D: After the last atom of {\tt A} is paired with an atom from {\tt B}, requirement~\ref{eqn:readyForMoC} is satisfied and the prototype informs the user to remove the migration system.
 
-\subsection{Validation}
+\section{Validation}
+\label{sct:Validation}
    The proof of concept gives but one example of something that works.
@@
 
-\section{Conclusions}
+\section{Conclusion}
 \label{sct:Conclusions}
@@
 
-   Our proposed migration is certainly not the only approach one could think of.
+   The data migration we propose is certainly not the only approach one could think of.
    However, we have not come across other approaches that allow changing the schema in the presence of constraints.

### 2023-12-11 2022Migration/articleMigrationFoIKS.tex

@@
 % Ampersand -----------------------------------------------------------
-
-%\def\id#1{\text{\it #1\/}}
-\newcommand{\ourtheory}{approach}
-\newcommand{\foundtheory}{foundational approach}
 
@@
 
-   To address this issue, this paper proposes a \foundtheory{} for data migration,
+   To address this issue, this paper proposes a foundational approach for data migration,
    aiming to generate migration scripts for automating the migration process.
@@
    The proposed solution was validated by a prototype demonstrating its efficacy.
-   Notably, the \ourtheory{} is technology-independent, articulating systems in terms of invariants, thereby ensuring applicability across various scenarios.
+   Notably, the approach is technology-independent, articulating systems in terms of invariants, thereby ensuring applicability across various scenarios.
    The migration script generator will be implemented in a software generator named Ampersand.
@@
 
-   The complexity of data migration has prompted us to develop an \ourtheory{} first,
+   The complexity of data migration has prompted us to develop an approach first,
    which we present in this contribution.
-   We have validated the \ourtheory{} by prototyping because a formal proof of correctness is currently beyond our reach.
-   This \ourtheory{} perceives an information system as a data set with constraints,
+   We have validated the approach by prototyping because a formal proof of correctness is currently beyond our reach.
+   This approach perceives an information system as a data set with constraints,
    so we can represent invariants (and thus the business semantics) directly as constraints.
@@
    Section~\ref{sct:Definitions} formalizes the concepts that we need to define the procedure.
-   Section~\ref{sct:Generating} defines the algoritm for generating a migration system, to automate SCDMs.
-   Section~\ref{sct:PoC} demonstrates the prototype of a migration system, which we used to verify our \ourtheory{} experimentally.
+   Section~\ref{sct:Generating} defines the migration system that automates SCDMs.
+   Section~\ref{sct:PoC} demonstrates the prototype of a migration system, which we used to verify our approach experimentally.
    We have used the language Ampersand for this purpose,
    because its syntax and semantics correspond directly to the definitions in section~\ref{sct:Definitions}.
-   Finally, section~\ref{sct:Validation} discusses the validation of our \ourtheory{} by showing that all requirements are met.
+   Finally, section~\ref{sct:Validation} discusses the validation of our approach by showing that all requirements are met.
 
@@
    Events that the system detects may cause the state to change.
-   To keep our \ourtheory{} technology independent, we assume that data sets contain triples.
-   This makes our \ourtheory{} valid for every kind of database that triples can represent,
+   To keep our approach technology independent, we assume that data sets contain triples.
+   This makes our approach valid for every kind of database that triples can represent,
    including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
@@
 \subsection{Ampersand}
-   We employ Ampersand as a prototyping language to demonstrate our \ourtheory{}.
-   %More significantly, our intention is to augment the Ampersand compiler with the \ourtheory{} outlined in this paper
-   %to generate migration systems automatically.
-   % Opmerking van SJC: ik denk dat deze intentie hier het best achterwege gelaten kan worden,
-   %   het zet de lezer direct op het spoor van: en waarom is dit nog niet in Ampersand geimplementeerd dan? Is dit niet af?
-   %
+   We employ Ampersand as a prototyping language to demonstrate our approach.
    Ampersand serves as a language for specifying information systems through a framework of concepts, relations, and constraints.
-   It comprises the three types of constraints discussed in this paper, making it an ideal platform for practical testing of our \ourtheory{}.
+   It comprises the three types of constraints discussed in this paper, making it an ideal platform for practical testing of our approach.
    In Ampersand, developers articulate constraints using heterogeneous relation algebra~\cite{Hattensperger1999,Alloy2006}
@@
    The moment the last violation of $u$ is fixed, the business constraint can be removed and $u$ can be implemented as a blocking invariant.
-   This is the core idea of our \ourtheory{}.
+   This is the core idea of our approach.
 
@@
    \end{array}
-   % SJC's opmerking:  \instance \cup \{\pair{a}{A},\pair{b}{B}\} = \instance ivm eqn:wellTypedViolation
+   % SJC's remark:  \instance \cup \{\pair{a}{A},\pair{b}{B}\} = \instance ivm eqn:wellTypedViolation
 \label{eqn:transaction}
@@
          \sign{v}=\sign{u}\\
-         \viol{v}{\dataset}=\viol{u}{\dataset} % -{\tt fixed}_u %opmerking van SJC: de - fixed is overbodig omdat eqn:blockRule geldt
+         \viol{v}{\dataset}=\viol{u}{\dataset}
       \end{array}\\
@@
    \begin{align}
-      \forall u\in\overrightarrow{\rules_{\schema'}-\rules_{\schema}}.~{\viol{u}{\dataset}}\subseteq \pop{{\tt fixed}_u}{\dataset}
+      \forall u\in\overrightarrow{\rules_{\schema'}-\rules_{\schema}}.~{\viol{u}{\dataset}} = \emptyset
    \label{eqn:readyForMoC}
    \end{align}
-   Equivalently, $\forall u\in\overrightarrow{\rules_{\schema'}-\rules_{\schema}}.~\viol{u}{\dataset} = \emptyset$.
    After this, the migration engineer can remove the migration system and the old system.
 
-\item Let us combine the above into a single migration schema:
+\item\label{step6} Let us combine the above into a single migration schema:
    \begin{align}
@@
 \label{sct:Conclusions}
-   In this paper, we describe the data migration as going from an existing system to a desired one, where the schema changes.
-   As Ampersand generates information systems, creating a new system can be a small task, allowing for incremental deployment of new features.
-   We describe the parts of a system that have an effect on data pollution.
-   We assume that the existing system does not violate any constraints of its schema, but address other forms of data pollution:
-   constraints that are not in the schema but are in the desired schema are initially relaxed such that the business can start using the migration system, after which this form of data pollution needs to be addressed by human intervention.
-   We propose a method for doing migration such that only a finite amount of human intervention is needed.
-   Our method allows a system similar to the desired system to be used while the intervention takes place.
+   We have shown that it is possible to generate a migration system from the schemas of an existing system and a desired system.
+   We have defined a migration system such that:
+   \begin{itemize}
+   \item after deploying the migration system, data from the existing system will be copied automatically,
+   \item the migration system behaves as a relaxed version of the desired system,
+   \item the migration engineer can introduce transactional invariants in the migration system to address data pollution,
+   \item there is no down time in switching from the existing system to the migration system at the MoT,
+   \item the number of violations the business needs to fix is finite and known at the MoT,
+   \item there is no down time in switching from the migration system to the desired system at the MoC,
+   \item the migration system is equivalent to the desired system at the MoC.
+   \end{itemize}
 
@@
    However, we have not come across other approaches that allow changing the schema in the presence of constraints.
-   As such, we cannot compare our approach against other approaches.
+   As such, we cannot compare our approach against other approaches yet.
    We envision that one day there will be multiple approaches for migration under a changing schema to choose from.
@@
 
-   This work does not consider what to do about (user) interfaces.
+   This work does not consider what to do about (user) interfaces yet.
    Instead, it models events by assuming that any change to the data set can be achieved.
@@
    Most Ampersand systems indeed allow the users of the system to edit the data set quite freely through the interfaces.
-   However, some interfaces may require certain constraints to be satisfied, which means that interfaces of the desired system might break.
-   In the spirit of the approach outlined here, we hope to generate migration interfaces that can replace any broken interfaces until the Moment of Transition.
-   How to do this is future work.
+   This has allowed us to set aside interfaces in this paper for now.
+   Nevertheless, we expect that an analysis of interfaces may yield insights about event streams,
+   which are further building blocks to a foundational theory of information systems.
+   But that is future work.
 

### 2024-03-23 2022Migration/articleMigrationRaMiCS.tex
Ramics version of article

--- 2023-12-10 2022Migration/articleMigrationFoIKS.tex
@@
 \begin{abstract}
-   Software generators that compile and deploy a specification into a functioning information system
+   Software generators that compile and deploy a specification into a functional information system
    can help to increase the frequency of releases in the software process.
@@
    The risk and effort of SCDMs explains why these teams try to avoid schema changes in the first place.
-   We believe that software teams that realize high update frequencies need better tools to perform SCDM's reliably with zero-downtime.
-   We want then to design, implement, and deploy SCDMs in the flow of their DevOps practice.
+   Our research aims at automating SCDMs to make them less risky and less costly,
+   so development teams can deploy schema changes with zero downtime.
 
@@
 
-   The complexity of data migration has prompted us to develop an \ourtheory{} first,
+   The complexity of data migration has prompted us to develop a \ourtheory{} first,
    which we present in this contribution.
@@
    The classical database transaction can be understood as a \define{transactional invariant},
-   which the system keeps satisfied by adding or deleting triples to the data set.
+   which the system keeps satisfied by adding or deleting triples to the dataset.
    As soon as the data violates this constraint, the system restores it without human intervention.
@@
    However, it implements the blocking invariants of the desired system as business constraints.
-   This migration system must also prevent every violation that is fixed from recurring.
+   This migration system must also ensure that every violation that is fixed is blocked from recurring.
    In this way, the business constraint gradually turns into a blocking invariant, satisfying the specification of the desired system.
@@
 \item[Moment of transition (MoT)]
-      After the migration system is done copying data, the migration engineer switches the ingress to relay all traffic to the migration system instead of the existing system.
-      This is the moment users will notice the difference because users experience the functionality of the migration system.
-      The migration system relaxes the blocking invariants of the desired system until users resolve their violations.
-      So it behaves like the desired system in the eyes of an average user, with some violations yet to be resolved.
+      After the migration system is done copying data, the migration engineer switches all traffic to the migration system.
+      This is the moment users will notice the difference because the traffic switch also deploys the functionality of the desired system.
+      So, in the eyes of an average user, the migration system may look like the desired system.
+      However, the migration system relaxes the blocking invariants of the desired system until users resolve their violations.
       Since the existing system receives no more traffic, its activity will come to a halt and its data will become static.
-      The migration system stays active until the desired system can take over.
+      The migration system stays active until all invariants of the desired system are satisfied and it can take over the work from the migration system.
 \item[Moment of completion (MoC)]
@@
    because their changes are copied by the migration system too.
-   However, after the MoT there must be no new changes in the existing system (by then the ``old'' system)
+   However, after the MoT there must be no new changes in the existing system
    to avoid violations of new invariants that the migration system has already fixed.
@@
    Atoms serve as data elements.
+   %Atoms are values without internal structure of interest, meant to represent atomic data elements (e.g. dates, strings, numbers, etc.) in a database.
+   %From a business perspective, atoms represent concrete items of the world,
+   %such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
+   %By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ represent \emph{atoms}.
    All atoms are taken from an infinite set called $\Atoms$.
@@
    Moreover, \atom{Peter}, \atom{Melissa} and \atom{074238991} are atoms.
-   In the sequel, variables $A$, $B$, and $C$ will represent concepts, and variables $a$, and $b$ represent \emph{atoms}.
+   In the sequel, variables $A$, $B$, $C$, $D$ will represent concepts, and variables $a$, $b$, and $c$ represent \emph{atoms}.
    %
@@
    Relations serve to organize and store data, allowing developers to represent facts.
-   In this paper, variable $r$ represents relations\footnote{Some readers might like to read `relation symbol' where we write `relation'.}.
+   In this paper, variables $r$, $s$, and $d$ represent relations\footnote{Some readers might like to read `relation symbol' where we write `relation'.}.
    All relations are taken from an infinite set $\Rels$.
@@
 \end{equation}
-   Note that the phrase ``pair $\pair{a}{b}$ is in relation $r$'' means that there is a data set $\dataset$ in which $\pair{a}{b}\in\pop{r}{\dataset}$,
+   Note that the phrase ``pair $\pair{a}{b}$ is in relation $r$'' means that there is a dataset $\dataset$ in which $\pair{a}{b}\in\pop{r}{\dataset}$,
    so the phrase ``triple $\triple{a}{r}{b}$ is in $\dataset$'' means the same thing.
@@
    The language Ampersand has more types of transactional invariants than just this one,
-   but this one is sufficient for this paper.
+   but this is sufficient for this paper.
 
@@
 \label{sct:Generating}
-   To prevent expensive and error-prone migration projects,
-   we define a migration script that can be generated by a compiler.
+   The complexity of migrating data to date yields expensive and error-prone migration projects.
+   By generating this code we can prevent many human induced errors.
 
+   This section starts with a presentation of the migration script that is to be used.
+
+\subsection{Generating a migration script}
    In the migration system, we need to refer to the items (concepts, relations, and constraints) of both the existing system and the desired system.
@@
 
-   The initial situation consists of one existing system $\infsys$ and one desired system $\infsys'$ of which we have a schema, $\schema_{\infsys'}$ and an initial data set, $\dataset_{\infsys'}$.
+   The initial situation consists of one existing system $\infsys$ and one desired system $\infsys'$ of which we have a schema, $\schema_{\infsys'}$ and an initial dataset, $\dataset_{\infsys'}$.
    We may assume that $\infsys$ satisfies definition~\ref{def:information system} because it is a system in production.
    Also, $\schema_{\infsys'}$ satisfies equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type} because the script of the desired system is type-correct.
-   Together, the schema and the intial data set forms the desired system $\pair{\schema_{\infsys'}}{\dataset_{\infsys'}}$, which satisfies definition~\ref{def:information system}.
+   Together, the schema and the intial dataset forms the desired system $\pair{\schema_{\infsys'}}{\dataset_{\infsys'}}$, which satisfies definition~\ref{def:information system}.
    With these assumptions in place,
@@
    The rules in the desired system are partly written to resolve data pollution.
-   In some cases, the migration engineer wants users to get rid of that pollution, as described above.
-   However, some data pollution might be satisfiable automatically.
-   In that case, the migration engineer might substitute the generated business constraint from $\busConstraints_\text{fix}$ by a transactional invariant in $\transactions_{\migrsys}$ to resolve it without bothering the business.
-   Of course, both cases need to be resolved at MoC.
+   In some cases, the migration engineer wants users to get rid of that pollution and turn the rule in a blocking invariant as described above.
+   However, some of these constraints might be satisfiable automatically.
+   In that case, the migration engineer might substitute such constraints from $\busConstraints_\text{fix}$ by transactional invariants in $\transactions_{\migrsys}$.
+   These invariants don't need the mechanism described above, because the migration system itself will take care that all constraints in $\transactions_{\migrsys}$ are satisfied.
+   Both cases need to be resolved at MoC.
 
@@
    So from MoC onwards, $\pair{\dataset_{\migrsys}}{\schema_{\migrsys}}$ is equivalent to $\pair{\dataset_{\migrsys}}{\overrightarrow{\schema'}}$.
-   Hence, we can tell that $\dataset_{\migrsys}$ is a valid data set for the desired system, so we can switch the ingress from the migration system to the desired system without users noticing the difference.
+   Hence, we can tell that $\dataset_{\migrsys}$ is a valid dataset for the desired system, so we can switch the ingress from the migration system to the desired system without users noticing the difference.
    Then, the migration system gets no more inputs, so it can be removed. 
+   Since this is equivalent to 
+   Since $\transactions_1$ and $\transactions_2$ are redundant after the MoC,
+   we can retain $\overrightarrow{\transactions_{\schema'}}$ in $\migrsys$,
+   which is equivalent to $\transactions_{\schema'}$ in the desired system.
+   Likewise, $\busConstraints_\text{fix}$ has become redundant, so $\migrsys$ can do with just $\overrightarrow{\busConstraints_{\schema'}}$.
+   In the desired system, that is equivalent to $\busConstraints_{\schema'}$.
+   So, the constraints in the desired system after the MoC are equivalent to the constraints in the MoC
 

### 2024-05-25 2024_RAMiCS_Migration/articleMigrationRAMiCS.tex
transfer Migration paper to RAMiCS

--- 2023-12-11 2022Migration/articleMigrationFoIKS.tex
@@
    In a world where automation of the software process is resulting in higher productivity, more frequent releases, and better quality,
-   SCDM's should not stay behind.
+   SCDMs should not stay behind.
    Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
@@
    The risk and effort of SCDMs explains why these teams try to avoid schema changes in the first place.
-   We believe that software teams that realize high update frequencies need better tools to perform SCDM's reliably with zero-downtime.
-   We want then to design, implement, and deploy SCDMs in the flow of their DevOps practice.
+   We believe that software teams that realize high update frequencies need better tools to perform SCDMs reliably with zero-downtime.
+   We want them to design, implement, and deploy SCDMs in the flow of their DevOps practice.
 
@@
    This does not cater for schema changes, however.
-   Then there are SCDMs in situations without a schema or an implicit schema, e.g.~\cite{Hillenbrand2022}.
-   Such situations lack the error preventing power that explicit schemas bring during the development of software.
+   Then there are SCDMs in situations without a schema or with an implicit schema, e.g.~\cite{Hillenbrand2022}.
+   Such situations lack the error-preventing power that explicit schemas bring during the development of software.
    All errors a schema can prevent at compile time must then be compensated by runtime checks,
-   which increases the likelyhood of end-users getting error messages.
+   which increases the likelihood of end-users getting error messages.
    This requires versioned storage of production data and an overhead in performance.
@@
    Section~\ref{sct:Generating} defines the migration system that automates SCDMs.
-   Section~\ref{sct:PoC} demonstrates the prototype of a migration system, which we used to verify our approach experimentally.
+   Section~\ref{sct:PoC} exhibits the prototype of a migration system, which we used to verify our approach experimentally.
    We have used the language Ampersand for this purpose,
@@
    The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
-   Multiple users, working from different locations and on different moments, constitute what we will loosely call ``the business''.
-   The data in the system constitutes the collective memory of the business,
-   which relies on the semantics of the data to draw the right conclusions and carry out their tasks.
-   \begin{figure}[bht]
-      \begin{center}
-        \includegraphics[scale=0.8]{figures/existing_system.pdf}
-      \end{center}
-   \caption{Anatomy of an information system}
-   \label{fig:pre-migration}
-   \end{figure}
+   Multiple users, working from different locations and at different moments, collectively form what we can call ``the business''.
+   The data in the system constitutes the collective memory of the business.
 
-   Figure~\ref{fig:pre-migration} depicts the situation before migration.
-   An existing application service ingests traffic through an ingress and persists data in a data set, which is typically a database.
-   Our research assumes that the structure and business semantics are represented in a schema, from which the system is generated.
-   Actors (both users and computers) are changing the data in a system continually.
-   The state of the system is represented by a data set, typically represented in some form of persistent store such as a database.
-   Events that the system detects may cause the state to change.
-   To keep our approach technology independent, we assume that data sets contain triples.
-   This makes our approach valid for every kind of database that triples can represent,
-   including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
+   Constraints represent the elementary truths of a business,
+   for example: ``{\em To submit a comment to the website of ACME Inc., the user must be registered.}''
+   By formally defining constraints, designers can restrict a data set.
+   We call a constraint {\em invariant} when it is implemented in an information system
+   to ensure that the data set satisfies this constraint at all times as long as it is deployed.
+   Thus, constraints implement the business semantics of the data.
 
-   We assume that constraints implement the business semantics of the data.
-   Constraints represent business concerns formally, so they can be checked automatically and can be used to generate software.
-   Some of the constraints require human intervention, while others require a system to intervene.
-   In this paper, we distinguish three different kinds of constraints:
+   There are many ways to make sure that the data set satisfies a constraint.
+   In this paper, we distinguish just three different kinds of enforcement:
 \begin{enumerate}
@@
 
+   \begin{figure}[bht]
+      \begin{center}
+        \includegraphics[scale=0.8]{figures/existing_system.pdf}
+      \end{center}
+   \caption{Anatomy of an information system}
+   \label{fig:pre-migration}
+   \end{figure}
+
+   Figure~\ref{fig:pre-migration} depicts the situation before migration.
+   An existing application service ingests traffic through an ingress and persists data in a data set, which is typically a database.
+   Our research assumes that the structure and business semantics are represented in a schema, from which the system is generated.
+   Actors (both users and computers) are changing the data in a system continually.
+   The state of the system is represented by a data set, typically represented in some form of persistent store such as a database.
+   Events that the system detects may cause the state to change.
+   To keep our approach technology-independent, we assume that data sets contain triples.
+   This makes our approach valid for every kind of database that can be represented by triples,
+   including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
 \subsection{Ampersand}
@@
    and \atom{074238991} as a \concept{TelephoneNumber}.
-   In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
-   Moreover, \atom{Peter}, \atom{Melissa} and \atom{074238991} are atoms.
+   In this example, \atom{Peter}, \atom{Melissa} and \atom{074238991} are atoms and
+   \concept{Person} and \concept{TelephoneNumber} are concepts.
    In the sequel, variables $A$, $B$, and $C$ will represent concepts, and variables $a$, and $b$ represent \emph{atoms}.
@@
    Relations serve to organize and store data, allowing developers to represent facts.
-   In this paper, variable $r$ represents relations\footnote{Some readers might like to read `relation symbol' where we write `relation'.}.
+   In this paper, variables $r$ and $s$ represent relations\footnote{Some readers might like to read `relation symbol' where we write `relation'.}.
    All relations are taken from an infinite set $\Rels$.
@@
 \end{definition}
+   All datasets are taken from an infinite set $\Dataset$.
    Looking at the example,
@@
 \label{eqn:pop-concept}\\
-   \pop{x-y}{\dataset}&= \pop{x}{\dataset} - \pop{y}{\dataset}
+   \pop{r-s}{\dataset}&= \pop{r}{\dataset} - \pop{s}{\dataset}
 \label{eqn:pop-expr}
@@
    \begin{array}[t]{l}
-      (a,b)\in \viol{u}{\pair{\triples}{\instance}}\, \Longrightarrow \\
-      \quad \viol{u}{\pair{\triples \cup \{\triple{a}{\declare{n}{A}{B}}{b}\}}{\instance}} = \viol{u}{\pair{\triples}{\instance}} - \{(a,b)\}
+      \pair{a}{b}\in \viol{u}{\pair{\triples}{\instance}}\, \Longrightarrow \\
+      \quad \viol{u}{\pair{\triples \cup \{\triple{a}{\declare{n}{A}{B}}{b}\}}{\instance}} = \viol{u}{\pair{\triples}{\instance}} - \{\pair{a}{b}\}
    \end{array}
@@
    We write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$ if needed;
-\item $\schema=\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$ is a schema (so it must satisfy equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}).
+\item $\schema=\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$ is a schema (so it must satisfy equations~\ref{eqn:relationsIntroduceConcepts} through~\ref{eqn:enforcement-has-type}).
 \item Triples in the data set must have their relation mentioned in the schema:
@@
 \label{sct:PoC}
-   By way of proof of concept (PoC), we have built a migration system in Ampersand.
-   To demonstrate it in the context of this paper, the existing system, $\pair{\dataset}{\schema}$, is rather trivial.
-   It has no constraints and just one relation, $\declare{r}{A}{B}$.
+   By way of proof of concept (PoC), we have built a tiny migration system in Ampersand.
+   It is tiny only to fit it in the page limit of this paper.
+   The existing system, $\pair{\dataset}{\schema}$, has no constraints and just one relation, $\declare{r}{A}{B}$.
    Its population is $A=\{a_1,a_2,a_3\}$, $B=\{b_1\}$, and $\pop{r}{\dataset}=\{\pair{a_1}{b_1}\}$.
@@
    We may assume that $\infsys$ satisfies definition~\ref{def:information system} because it is a system in production.
-   Also, $\schema_{\infsys'}$ satisfies equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type} because the script of the desired system is type-correct.
+   Also, $\schema_{\infsys'}$ satisfies equations~\ref{eqn:relationsIntroduceConcepts} through~\ref{eqn:enforcement-has-type} because the script of the desired system is type-correct.
    Together, the schema and the intial data set forms the desired system $\pair{\schema_{\infsys'}}{\dataset_{\infsys'}}$, which satisfies definition~\ref{def:information system}.
@@
 \label{sct:Conclusions}
-   We have shown that it is possible to generate a migration system from the schemas of an existing system and a desired system.
+   We have shown that it is possible in theory to generate a migration system from the schemas of an existing system and a desired system.
    We have defined a migration system such that:
@@
 
-   The data migration we propose is certainly not the only approach one could think of.
-   However, we have not come across other approaches that allow changing the schema in the presence of constraints.
-   As such, we cannot compare our approach against other approaches yet.
-   We envision that one day there will be multiple approaches for migration under a changing schema to choose from.
-   For now, our next step is to implement the approach shown here into Ampersand.
-
-   This work does not consider what to do about (user) interfaces yet.
-   Instead, it models events by assuming that any change to the data set can be achieved.
-   In practice, such changes need to be achieved through interfaces.
-   Most Ampersand systems indeed allow the users of the system to edit the data set quite freely through the interfaces.
-   This has allowed us to set aside interfaces in this paper for now.
-   Nevertheless, we expect that an analysis of interfaces may yield insights about event streams,
-   which are further building blocks to a foundational theory of information systems.
-   But that is future work.
+   We intend to address the following issues in future work:
+   \begin{itemize}
+   \item Building the migration system generator in Ampersand, so it can be used in practice.
+   \item Investigating the possibility of in-situ migration with changing schemas, to enable incremental development in production.
+   \item Extending the theory to incorporate interfaces, so we can reason about the behavior of the system as a whole.
+   \item Keep looking for other approaches to migration under a changing schema, to compare them with the approach shown here.
+   So far, we have not found any, so we cannot compare our approach against other approaches yet.
+   \end{itemize}
 

### 2024-06-06 2022Migration/articleMigrationRaMiCS.tex
Going through reviews. Done with the exception of two TODOs (in latex comments)

@@
 \newcommand{\weg}[1]{}
+\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar$\hfil\cr$\to$\cr}}}
 
@@
    This allows us to signal the data pollution at runtime.
-   Some forms of data pollution are not automatable, however.
-   An example is when a person has deliberately specified a false name without violating any constraint in the system.
+   Some forms of data pollution will need to remain out of scope.
+   An example is when a person has specified a false name without violating any constraint in the system.
 
-   The complexity of data migration has prompted us to develop a \ourtheory{} first,
+   The complexity of data migration has prompted us to develop an \ourtheory{} first,
    which we present in this contribution.
@@
    The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
-   Multiple users, working from different locations and on different moments, constitute what we will loosely call ``the business''.
+   Multiple users, working from different locations and at different moments, constitute what we will loosely call ``the business''.
    The data in the system constitutes the collective memory of the business,
@@
    Figure~\ref{fig:pre-migration} depicts the situation before migration.
-   An existing application service ingests traffic through an ingress and persists data in a data set, which is typically a database.
+   The state of the system is represented by a data set, typically represented in some form of persistent store such as a database.
+   An existing application service ingests traffic through an ingress and persists data in a data set.
    Our research assumes that the structure and business semantics are represented in a schema, from which the system is generated.
    Actors (both users and computers) are changing the data in a system continually.
-   The state of the system is represented by a data set, typically represented in some form of persistent store such as a database.
    Events that the system detects may cause the state to change.
@@
 \item Transactional invariant\\
-   The classical database transaction can be understood as a \define{transactional invariant},
-   which the system keeps satisfied by adding or deleting triples to the dataset.
-   As soon as the data violates this constraint, the system restores it without human intervention.
+   A constraint that can is kept satisfied automatically by taking corrective actions is called a \define{transactional invariant}.
+   The system keeps these satisfied by adding or deleting triples to the dataset, typically using wrapped inside a classical database transaction to avoid issues with concurrency.
+   As soon as the data violates a transactional constraint, the system restores it without human intervention.
    So, the outside world experiences this as a constraint that is always satisfied, i.e.~an invariant.
@@
    It comprises the three types of constraints discussed in this paper, making it an ideal platform for practical testing of our \ourtheory{}.
-   In Ampersand, developers articulate constraints using heterogeneous relation algebra~\cite{Hattensperger1999,Alloy2006}
+   In Ampersand, developers articulate constraints using heterogeneous relation algebra~\cite{Hattensperger1999,Alloy2006}.
    The systems they generate keep invariants satisfied and alert users to violations of business constraints.
@@
    Data pollution, new business rules, or known issues in the existing system
-   may occasionally require tailoring the migration script to specific needs
+   may occasionally require tailoring a script describing the migration system to specific needs
    that require action of users in production.
@@
       So, in the eyes of an average user, the migration system may look like the desired system.
-      However, the migration system relaxes the blocking invariants of the desired system until users resolve their violations.
+      However, the migration system relaxes the blocking invariants of the desired system until users resolve the violations of the new blocking invariants.
       Since the existing system receives no more traffic, its activity will come to a halt and its data will become static.
-      The migration system stays active until all invariants of the desired system are satisfied and it can take over the work from the migration system.
+      The migration system stays active until all invariants of the desired system are satisfied and the desired system can take over the work from the migration system.
 \item[Moment of completion (MoC)]
@@
    Transactions in the existing system that happen during the time that the migration system is copying data cause no problem,
-   because their changes are copied by the migration system too.
+   because their changes are copied by the migration system, too.
    However, after the MoT there must be no new changes in the existing system
@@
    %By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ represent \emph{atoms}.
-   All atoms are taken from an infinite set called $\Atoms$.
+   All atoms are taken from a set called $\Atoms$.
    %
-   Concepts are names that group atoms of the same type.
-   All concepts are taken from an infinite set $\Concepts$.
+   Concepts (concept symbols) can be understood as (names for) types, and our definitions do not exclude sub-typing.
+   All concept symbols are taken from a set $\Concepts$.
    %$\Concepts$ and $\Atoms$ are disjoint.
@@
    and \atom{074238991} as a \concept{TelephoneNumber}.
-   In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
+   In this example, \concept{Person} and \concept{TelephoneNumber} are concept symbols.
    Moreover, \atom{Peter}, \atom{Melissa} and \atom{074238991} are atoms.
-   In the sequel, variables $A$, $B$, $C$, $D$ will represent concepts, and variables $a$, $b$, and $c$ represent \emph{atoms}.
+   In the sequel, variables $A$, $B$, $C$, $D$ will represent concept symbols, and variables $a$, $b$, and $c$ represent \emph{atoms}.
    %
-   The relation $\instance:\Pair{\Atoms}{\Concepts}$ relates atoms to concepts.
-   The term $a\instance C$ means that atom $a$ is an \emph{instance} of concept $C$.
+   The relation $\instance:\Pair{\Atoms}{\Concepts}$ relates atoms to concept symbols.
+   The term $a\instance C$ means that atom $a$ is an \emph{instance} of a concept denoted by $C$.
    %This relation is used in the type system, in which $\instance$ assigns one or more concepts to every atom in the data set.
@@
    Relations serve to organize and store data, allowing developers to represent facts.
-   In this paper, variables $r$, $s$, and $d$ represent relations\footnote{Some readers might like to read `relation symbol' where we write `relation'.}.
-   All relations are taken from an infinite set $\Rels$.
+   In this paper, variables $r$, $s$, and $d$ represent relation symbols.
+   All relation symbols are taken from a set $\Rels$.
    $\Rels$ is disjoint from $\Concepts$ and $\Atoms$.
-   Every relation $r$ has a name, a source concept, and a target concept.
-   The notation $r=\declare{n}{A}{B}$ denotes that relation $r$ has name $n$, source concept $A$, and target concept $B$.
-   The part $\pair{A}{B}$ is called the \define{signature} of the relation.
+   Every relation symbol $r$ has a name, a source concept, and a target concept.
+   The notation $r=\declare{n}{A}{B}$ denotes that relation symbol $r$ has name $n$, source concept $A$, and target concept $B$.
+   The part $\pair{A}{B}$ is called the \define{signature} of the relation symbol.
+   When the signature is clear from the context, or not important, we use $r$ and $n$ interchangeably.
 
@@
    \begin{definition}[Data set]
+   %TODO: SJC weet niet wat hij met dit commentaar aan moet (Review 2):
+   % 
+%Before Def. 1: You could group the implicit parameters of Def. 1
+%into a ``data vocabulary'' $(\AA, \CC, \RR, \src, \trg)$.
    A data set $\dataset$ is a tuple $\pair{\triples}{\instance}$ with finite $\triples \subseteq {\Triple{\Atoms}{\Rels}{\Atoms}}$ and $\instance\subseteq\Pair{\Atoms}{\Concepts}$ that satisfies:
@@
    equation~\ref{eqn:wellTypedEdge} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
-   In practice, users can say that the Person Peter has telephone number 074238991.
+   In practice, users can say that the person Peter has telephone number 074238991.
    So, the ``thing'' that \atom{Peter} refers to (which is Peter) has \atom{074238991} as a telephone number.
@@
    To save writing in the sequel, the notation $a\ r\ b$ means that $\triple{a}{r}{b}\in\triples$.
+   We'll use $\Dataset$ to denote the set of all possible data sets.
 
-   A relation $r$ can serve as a container of pairs,
+   A relation symbol $r$ can serve as a container of pairs,
    as defined by the function $\popF{r}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$.
@@
 \end{equation}
-   Note that the phrase ``pair $\pair{a}{b}$ is in relation $r$'' means that there is a dataset $\dataset$ in which $\pair{a}{b}\in\pop{r}{\dataset}$,
-   so the phrase ``triple $\triple{a}{r}{b}$ is in $\dataset$'' means the same thing.
+   Note that the phrase ``pair $\pair{a}{b}$ is in relation $r$'' means that $\pair{a}{b}\in\pop{r}{\dataset}$ where $\dataset$ is clear from the context.
 %   Equation~\ref{eqn:wellTypedEdge} implies that for every data set $\dataset$:
@@
 %
-   We overload the notation $\popF{}$ so we can use it on concepts $\popF{C}:\Dataset\rightarrow\powerset{\Atoms}$
-   and expressions. We also define the difference of populations, in equation~\ref{eqn:pop-expr}, both for relations and concepts:
+   We overload the notation $\popF{}$ so we can use it on concept symbols $\popF{C}:\Dataset\rightarrow\powerset{\Atoms}$
+   and expressions. We also define the difference of populations, in equation~\ref{eqn:pop-expr}, both for relation and concept symbols:
 \begin{align}
@@
 \label{sct:Constraints}
-   Every constraint is an element of an infinite set called $\Constraints$.
-   In this paper, variables $u$ and $v$ represent all three types of constraints.
-   We say that a constraint, $u$, is satisfied when it is true for all triples in the data set.
+   Every constraint symbol is an element of a set called $\Constraints$.
+   In this paper, variables $u$ and $v$ represent symbols for all three types of constraints.
    For every constraint $u$, function $\violC{u}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$ produces the violations of $u$,
@@
 \end{equation}
-   Note that $\viol{u}{\dataset}=\emptyset$ means that $\dataset$ satisfies $u$.
+   Note that $\viol{u}{\dataset}=\emptyset$ means that $\dataset$ satisfies the constraint whose symbol is $u$.
+   We'll say that $u$ is satisfied in such cases.
 
-   In case $u$ is a transactional invariant,
-   the system will keep it satisfied by adding the violations to a specific relation $\declare{n}{A}{B}$ such that $\pair{A}{B}=\sign{u}$.
+   In order to guarantee that the work required for migration is finite, it suffices to require that $\viol{u}{\dataset}$ is a finite set.
+   The language Ampersand implements many of the sets described so far as finite sets, causing any constraint that can be specified in it to satisfy that $\viol{u}{\dataset}$ is finite.
+
+   In the current paper, we will, for the sake of simplicity, only
+consider transactional invariants for which violations can be repaired
+by inserting them into a single designated relation.
+   The language Ampersand has more types of transactional invariants than just this one,
+   but this is sufficient for this paper.
+
+   In case $u$ denotes a transactional invariant,
+   the system will keep it satisfied by adding the violations to a specific relation denoted by $\declare{n}{A}{B}$ such that $\pair{A}{B}=\sign{u}$.
    This requires that adding the pair to that relation solves the violation:
@@
    As $\declare{n}{A}{B}$ is specific for $u$, we can write $\enfRel{u}$ for it.
-   We call this the \define{enforced relation} of transactional invariant $u$:
+   We call this the symbol for the \define{enforced relation} of the transactional invariant denoted by $u$:
 \begin{equation}
@@
    in which $r=\enfRel{u}$.
-
-   The language Ampersand has more types of transactional invariants than just this one,
-   but this is sufficient for this paper.
 
@@
    We describe a schema $\schema$ as a tuple $\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$,
-   in which $\concepts\subseteq\Concepts$ is a finite set of concepts,
-   $\rels\subseteq\Rels$ is a finite set of relations,
-   $\rules\subseteq\Constraints$ is a finite set of blocking invariants,
-   $\transactions\subseteq\Constraints$ is a finite set of transactional invariants,
-   and $\busConstraints\subseteq\Constraints$ is a finite set of business constraints.
+   in which $\concepts\subseteq\Concepts$ is a finite set of concept symbols,
+   $\rels\subseteq\Rels$ is a finite set of relation symbols,
+   $\rules\subseteq\Constraints$ is a finite set of symbols for blocking invariants,
+   $\transactions\subseteq\Constraints\pfun\Rels$ is a finite set of symbols for transactional invariants,
+   and $\busConstraints\subseteq\Constraints$ is a finite set of symbols for business constraints.
 
@@
    \end{definition}
-   Requirements~\ref{eqn:relationsIntroduceConcepts} and~\ref{eqn:invariant-has-type} ensure that concepts mentioned in relations and in the signature of rules are part of the schema.
-   Requirement~\ref{eqn:enforcement-has-type} ensures the enforced relation of a transactional invariant is declared in the schema. 
+   Requirements~\ref{eqn:relationsIntroduceConcepts} and~\ref{eqn:invariant-has-type} ensure that concept symbols mentioned in relations and in the signature of constraints are part of the schema.
+   Requirement~\ref{eqn:enforcement-has-type} ensures the enforced relation symbol of a transactional invariant is declared in the schema. 
    When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, $\rules_{\schema}$, $\transactions_{\schema}$, $\busConstraints_{\schema}$
@@
 \item $\schema=\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$ is a schema (so it must satisfy equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}).
-\item Triples in the data set must have their relation mentioned in the schema:
+\item Triples in the data set must have their relation symbol mentioned in the schema:
    \begin{eqnarray}
@@
 \item All violations must have a type, which follows from  (\ref{eqn:wellTypedViolation}).
-\item The system keeps any transactional invariant $u$ satisfied by adding violations to the relation $\enforceC{u}$ (\ref{eqn:transaction}).
+\item The system keeps any transactional invariant denoted by $u$ satisfied by adding violations to the relation denoted by $\enforceC{u}$ (\ref{eqn:transaction}).
 \item All invariants must remain satisfied:
@@
    The complexity of migrating data to date yields expensive and error-prone migration projects.
-   By generating this code we can prevent many human induced errors.
+   By generating the migration system we can prevent many human induced errors.
+   However, to allow for human tailoring, we generate a script that describes this migration system (from which the system can be generated automatically).
 
@@
    We have to relabel items with prefixes to avoid name clashes in the migration system.
-   We use a left arrow to denote relabeling by prefixing the name of the item with ``old.''.
+   We use a left arrow to denote relabeling by prefixing the name of the item with ``old.'' (or some other prefix that avoids name clashes).
 \begin{equation}
@@
    \end{align}
+
+%TODO:
+% SJC begrijpt niet helemaal wat reviewer 2 hier bedoelt:
+    %(16) Transactional invariants should be elements of $\UU$.
+    %    No such elements can be seen in (16).
+    %
+    %    Are transactional invariants repaired via transactions
+    %    with any specified granularity?
+    %    (16) looks like offering lots of opportunities for race
+    %    conditions during pre-deploy (``starts copying data'')
+    %    if the old system can also delete...
+    %    In particular, it appears that (17) would then be reachable
+    %    in states where even if $\mathcal{D}'$ was empty,
+    %    $new.r \neq old.r = copy_r$,
+    %    because for some pairs $p$ it could happen that:
+    %    1. $p$ is propagated to new.r according to the first part of (16)
+    %    2. $p$ is deleted in old.r via the ingress,
+    %    3. $p$ is not in old.r anymore, and will therefore not be
+    %       copied to copy_r
+    %    4. (17) is reached with $p$ in new.r, but not in old.r and copy.r
+    %    5. The deletion of $p$ being lost invalidates the migration.
 
@@
    The desired system contains one blocking invariant, which is the totality of $\declare{r}{A}{B}$.
+   Its violations are $\pair{a_2}{a_2}$ and $\pair{a_3}{a_3}$.
    The schema of the migration system, $\schema_\migrsys$, follows from definition~\ref{eqn:schema migrsys}.
@@
    We may assume that $\infsys$ satisfies definition~\ref{def:information system} because it is a system in production.
-   Also, $\schema_{\infsys'}$ satisfies equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type} because the script of the desired system is type-correct.
+   Also, $\schema_{\infsys'}$ satisfies equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type} because the schema of the desired system is type-correct.
    Together, the schema and the intial dataset forms the desired system $\pair{\schema_{\infsys'}}{\dataset_{\infsys'}}$, which satisfies definition~\ref{def:information system}.
@@
    The relations to be copied from $\infsys$ are those relations that the desired system retains: $\rels_{\schema'}\cap\rels_{\schema}$.
-   For each $r$ to be copied from $\infsys$, $\migrsys$ contains a relation ${\tt copy}_r$ in $\rels_1$ (eqn.~\ref{eqn:copy relations}).
+   For each $r$ to be copied from $\infsys$, $\migrsys$ contains a relation symbol ${\tt copy}_r$ in $\rels_1$ (eqn.~\ref{eqn:copy relations}).
    After the MoT, the ingress sends all change events to $\migrsys$, so the existing system can finish the work it is doing for transactional invariants and will not change after that.
-   In other words, the population of every relation in $\rels_{\schema}$ becomes stable and so does every ${\tt copy}_r$.
+   In other words, the population of every relation symbol in $\rels_{\schema}$ becomes stable and so does every ${\tt copy}_r$.
    At that point in time, eqn.~\ref{eqn:copyingTerminates} is satisfied and stays satisfied.
@@
    Thirdly, the constraints that may need human intervention are the blocking invariants of the desired system that were not in the existing system ($\rules_{\schema'}-\rules_{\schema}$ in def.~\ref{eqn:Bfix}).
-   $\migrsys$ features $\busConstraints_\text{fix}$ to represent these rules in the guise of business constraints.
+   $\migrsys$ features $\busConstraints_\text{fix}$ to represent these invariants in the guise of business constraints.
    This lets business actors resolve the violations.
@@
 
-   The rules in the desired system are partly written to resolve data pollution.
+   The constraints in the desired system are partly written to resolve data pollution.
    In some cases, the migration engineer wants users to get rid of that pollution and turn the rule in a blocking invariant as described above.
@@
    Then, the migration system gets no more inputs, so it can be removed. 
-   Since this is equivalent to 
    Since $\transactions_1$ and $\transactions_2$ are redundant after the MoC,
@@
    In the desired system, that is equivalent to $\busConstraints_{\schema'}$.
-   So, the constraints in the desired system after the MoC are equivalent to the constraints in the MoC
+   So, the constraints in the desired system after the MoC are equivalent to the constraints in the MoC.
 
@@
    We envision that one day there will be multiple approaches for migration under a changing schema to choose from.
-   For now, our next step is to implement the approach shown here into Ampersand.
+   For now, our next step is to automate the generation of migration scripts as an extension to Ampersand.
 
@@
    Most Ampersand systems indeed allow the users of the system to edit the data set quite freely through the interfaces.
-   However, some interfaces may require certain constraints to be satisfied, which means that interfaces of the desired system might break.
+   However, some interfaces may require certain constraints to be satisfied, which means that interfaces of the desired system might break when used through the migration system.
    In the spirit of the approach outlined here, we hope to generate migration interfaces that can replace any broken interfaces until the Moment of Transition.

### 2024-06-07 2022Migration/articleMigrationRaMiCS.tex
minor textual edits

@@
 \begin{enumerate}
-\item We take a disjoint union of the data sets by relabeling relation names, so the migration script can refer to relations from both systems:
+\item We take a disjoint union of the data sets by relabeling relation symbols, so the migration script can refer to relations from both systems:
    \begin{align}
@@
    \end{equation}
-\item\label{step5} To signal users that there are violations that need to be fixed, we generate a business constraint for each new blocking invariant $u$:
+\item\label{step5} To signal users that there are violations that need to be fixed, we generate a business constraint for each new blocking invariant denoted by $u$:
 % \begin{verbatim}
@@
    In some cases, a migration engineer can invent ways to satisfy these invariants automatically.
-   For this purpose, the generator must produce source code (as opposed to compiled code) to allow the migration engineer
+   This is one of the places where it is useful for the generator to produce source code (as opposed to compiled code) to allow the migration engineer
    to replace a business constraint with transactional invariants of her own making.

### 2024-06-22 2024_RAMiCS_Migration/articleMigrationRAMiCS.tex
preprint submitted 2024/06/21

--- 2023-12-10 2022Migration/articleMigrationFoIKS.tex
@@
    This allows us to signal the data pollution at runtime.
-   Some forms of data pollution are not automatable, however.
-   An example is when a person has deliberately specified a false name without violating any constraint in the system.
+   Some forms of data pollution will need to remain out of scope.
+   An example is when a person has specified a false name without violating any constraint in the system.
 
@@
    Section~\ref{sct:Generating} defines the algoritm for generating a migration system, to automate SCDMs.
-   Section~\ref{sct:PoC} demonstrates the prototype of a migration system, which we used to verify our \ourtheory{} experimentally.
-   We have used the language Ampersand for this purpose,
+   Section~\ref{sct:PoC} demonstrates the prototype of a migration system, which we used to validate our \ourtheory{} experimentally.
+   For this purpose we have used the language Ampersand
    because its syntax and semantics correspond directly to the definitions in section~\ref{sct:Definitions}.
-   Finally, section~\ref{sct:Validation} discusses the validation of our \ourtheory{} by showing that all requirements are met.
 
@@
 \item Transactional invariant\\
-   The classical database transaction can be understood as a \define{transactional invariant},
-   which the system keeps satisfied by adding or deleting triples to the data set.
-   As soon as the data violates this constraint, the system restores it without human intervention.
+   A constraint that can is kept satisfied automatically by taking corrective actions is called a \define{transactional invariant}.
+   The system keeps these satisfied by adding or deleting triples to the dataset, typically using wrapped inside a classical database transaction to avoid issues with concurrency.
+   As soon as the data violates a transactional constraint, the system restores it without human intervention.
    So, the outside world experiences this as a constraint that is always satisfied, i.e.~an invariant.
@@
 \item[Moment of transition (MoT)]
-      After the migration system is done copying data, the migration engineer switches the ingress to relay all traffic to the migration system instead of the existing system.
-      This is the moment users will notice the difference because users experience the functionality of the migration system.
-      The migration system relaxes the blocking invariants of the desired system until users resolve their violations.
-      So it behaves like the desired system in the eyes of an average user, with some violations yet to be resolved.
+      After the migration system is done copying data, the migration engineer switches all traffic to the migration system.
+      This is the moment users will notice the difference because the traffic switch also deploys the functionality of the desired system.
+      So, in the eyes of an average user, the migration system may look like the desired system.
+      However, the migration system relaxes the blocking invariants of the desired system until users resolve the violations of the new blocking invariants.
       Since the existing system receives no more traffic, its activity will come to a halt and its data will become static.
-      The migration system stays active until the desired system can take over.
+      The migration system stays active until all invariants of the desired system are satisfied and the desired system can take over the work from the migration system.
 \item[Moment of completion (MoC)]
@@
    Transactions in the existing system that happen during the time that the migration system is copying data cause no problem,
-   because their changes are copied by the migration system too.
-   However, after the MoT there must be no new changes in the existing system (by then the ``old'' system)
+   because their changes are copied by the migration system, too.
+   However, after the MoT there must be no new changes in the existing system
    to avoid violations of new invariants that the migration system has already fixed.
@@
    Atoms serve as data elements.
-   All atoms are taken from an infinite set called $\Atoms$.
+   %Atoms are values without internal structure of interest, meant to represent atomic data elements (e.g. dates, strings, numbers, etc.) in a database.
+   %From a business perspective, atoms represent concrete items of the world,
+   %such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
+   %By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ represent \emph{atoms}.
+   All atoms are taken from a set called $\Atoms$.
    %
-   Concepts are names that group atoms of the same type.
-   All concepts are taken from an infinite set $\Concepts$.
+   Concepts (concept symbols) can be understood as (names for) types, and our definitions do not exclude sub-typing.
+   All concept symbols are taken from a set $\Concepts$.
    %$\Concepts$ and $\Atoms$ are disjoint.
@@
    and \atom{074238991} as a \concept{TelephoneNumber}.
-   In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
+   In this example, \concept{Person} and \concept{TelephoneNumber} are concept symbols.
    Moreover, \atom{Peter}, \atom{Melissa} and \atom{074238991} are atoms.
-   In the sequel, variables $A$, $B$, and $C$ will represent concepts, and variables $a$, and $b$ represent \emph{atoms}.
+   In the sequel, variables $A$, $B$, $C$, $D$ will represent concept symbols, and variables $a$, $b$, and $c$ represent \emph{atoms}.
    %
-   The relation $\instance:\Pair{\Atoms}{\Concepts}$ relates atoms to concepts.
-   The term $a\instance C$ means that atom $a$ is an \emph{instance} of concept $C$.
+   The relation $\instance:\Pair{\Atoms}{\Concepts}$ relates atoms to concept symbols.
+   The term $a\instance C$ means that atom $a$ is an \emph{instance} of a concept denoted by $C$.
    %This relation is used in the type system, in which $\instance$ assigns one or more concepts to every atom in the data set.
@@
    Relations serve to organize and store data, allowing developers to represent facts.
-   In this paper, variable $r$ represents relations\footnote{Some readers might like to read `relation symbol' where we write `relation'.}.
-   All relations are taken from an infinite set $\Rels$.
+   In this paper, variables $r$, $s$, and $d$ represent relation symbols.
+   All relation symbols are taken from a set $\Rels$.
    $\Rels$ is disjoint from $\Concepts$ and $\Atoms$.
-   Every relation $r$ has a name, a source concept, and a target concept.
-   The notation $r=\declare{n}{A}{B}$ denotes that relation $r$ has name $n$, source concept $A$, and target concept $B$.
-   The part $\pair{A}{B}$ is called the \define{signature} of the relation.
+   Every relation symbol $r$ has a name, a source concept, and a target concept.
+   The notation $r=\declare{n}{A}{B}$ denotes that relation symbol $r$ has name $n$, source concept $A$, and target concept $B$.
+   The part $\pair{A}{B}$ is called the \define{signature} of the relation symbol.
+   When the signature is clear from the context, or not important, we use $r$ and $n$ interchangeably.
 
@@
 \end{equation}
-   Note that the phrase ``pair $\pair{a}{b}$ is in relation $r$'' means that there is a data set $\dataset$ in which $\pair{a}{b}\in\pop{r}{\dataset}$,
-   so the phrase ``triple $\triple{a}{r}{b}$ is in $\dataset$'' means the same thing.
+   Note that the phrase ``pair $\pair{a}{b}$ is in relation $r$'' means that $\pair{a}{b}\in\pop{r}{\dataset}$ where $\dataset$ is clear from the context.
 %   Equation~\ref{eqn:wellTypedEdge} implies that for every data set $\dataset$:
@@
    in which $r=\enfRel{u}$.
-
-   The language Ampersand has more types of transactional invariants than just this one,
-   but this one is sufficient for this paper.
+   The symbol $u\in\mathbb{U}$ that refers to $\enfRel{u}\mapsfrom\violC{u}$ is written as $[\enfRel{u}\mapsfrom\violC{u}]$.
 
@@
    We describe a schema $\schema$ as a tuple $\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$,
-   in which $\concepts\subseteq\Concepts$ is a finite set of concepts,
-   $\rels\subseteq\Rels$ is a finite set of relations,
-   $\rules\subseteq\Constraints$ is a finite set of blocking invariants,
-   $\transactions\subseteq\Constraints$ is a finite set of transactional invariants,
-   and $\busConstraints\subseteq\Constraints$ is a finite set of business constraints.
+   in which $\concepts\subseteq\Concepts$ is a finite set of concept symbols,
+   $\rels\subseteq\Rels$ is a finite set of relation symbols,
+   $\rules\subseteq\Constraints$ is a finite set of symbols for blocking invariants,
+   $\transactions\subseteq \Constraints$ is a finite set of symbols denoting transactional invariants,
+   and $\busConstraints\subseteq\Constraints$ is a finite set of symbols for business constraints.
 
@@
    u\in\rules\cup\transactions\cup\busConstraints\ \wedge\ \sign{u}=\pair{A}{B}&~\Rightarrow~A\in\concepts\,\wedge\, B\in\concepts\label{eqn:invariant-has-type}\\
-   (\declare{n}{A}{B}\mapsfrom t)\in\mathcal\transactions&~\Rightarrow~ \declare{n}{A}{B}\in \rels\label{eqn:enforcement-has-type}
+   u\in\mathcal\transactions&~\Rightarrow~ \enfRel{u}\in \rels\label{eqn:enforcement-has-type}
 \end{align}
    \end{definition}
-   Requirements~\ref{eqn:relationsIntroduceConcepts} and~\ref{eqn:invariant-has-type} ensure that concepts mentioned in relations and in the signature of rules are part of the schema.
-   Requirement~\ref{eqn:enforcement-has-type} ensures the enforced relation of a transactional invariant is declared in the schema. 
+   Requirements~\ref{eqn:relationsIntroduceConcepts} and~\ref{eqn:invariant-has-type} ensure that concept symbols mentioned in relations and in the signature of constraints are part of the schema.
+   Requirement~\ref{eqn:enforcement-has-type} ensures the enforced relation symbol of a transactional invariant is declared in the schema. 
    When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, $\rules_{\schema}$, $\transactions_{\schema}$, $\busConstraints_{\schema}$
@@
 \label{sct:Generating}
-   To prevent expensive and error-prone migration projects,
-   we define a migration script that can be generated by a compiler.
+   The complexity of migrating data to date yields expensive and error-prone migration projects.
+   By generating the migration system we can prevent many human induced errors.
+   However, to allow for human tailoring, we generate a script that describes this migration system (from which the system can be generated automatically).
 
+   This section starts with a presentation of the migration script that is to be used.
+
+\subsection{Generating a migration script}
    In the migration system, we need to refer to the items (concepts, relations, and constraints) of both the existing system and the desired system.
    We have to relabel items with prefixes to avoid name clashes in the migration system.
-   We use a left arrow to denote relabeling by prefixing the name of the item with ``old.''.
+   We use a left arrow to denote relabeling by prefixing the name of the item with ``old.'' (or some other prefix that avoids name clashes).
 \begin{equation}
@@
       \rels_1={}&\{{\tt copy}_r\mid r\in\rels_{\schema'}\cap\rels_{\schema}\}\label{eqn:copy relations}\\
-      \transactions_1={}&\{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r}-{\tt copy}_r} \mid r\in\rels_{\schema'}\cap\rels_{\schema}\}\ \cup\\
-      {}&\{{\tt copy}_r\mapsfrom \popF{\overrightarrow{r}\cap\overleftarrow{r}} \mid r\in\rels_{\schema'}\cap\rels_{\schema}\}\notag
+      \transactions_1={}&\left\{\left[\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r}-{\tt copy}_r}\right] \Bigm\vert r\in\rels_{\schema'}\cap\rels_{\schema}\right\}\ \cup\\
+      {}&\{\left[{\tt copy}_r\mapsfrom \popF{\overrightarrow{r}\cap\overleftarrow{r}}\right] \mid r\in\rels_{\schema'}\cap\rels_{\schema}\}\notag
    \end{align}
@@
    \end{align}
+   Since the enforce rules only insert triples, the copying process is guaranteed to terminate.
+   However, deletions in the old system that happen during this copying process might not propagate into the migration system.
+   This may pose a risk to the business with respect to data quality.
 
@@
    The desired system contains one blocking invariant, which is the totality of $\declare{r}{A}{B}$.
+   Its violations are $\pair{a_2}{a_2}$ and $\pair{a_3}{a_3}$.
    The schema of the migration system, $\schema_\migrsys$, follows from definition~\ref{eqn:schema migrsys}.
 
-   Figure~\ref{fig:PoC} exhibits four successive screenshots,
+   Figure~\ref{fig:PoC} shows four successive screenshots,
    featuring $\overleftarrow{r}$ as {\small\verb#old_r#},
    $\overrightarrow{r}$ as {\small\verb#new_r#}.
-\begin{figure}[bht]
+
+\begin{figure}[!htb]
    \begin{center}
@@
 
-   Exhibit D: After the last atom of {\tt A} is paired with an atom from {\tt B}, requirement~\ref{eqn:readyForMoC} is satisfied and the prototype informs the user to remove the migration system.
+   Exhibit D: When the last atom of {\tt A} is paired with an atom from {\tt B}, requirement~\ref{eqn:readyForMoC} is satisfied and the prototype informs the user to remove the migration system.
 
-\section{Validation}
-\label{sct:Validation}
+\subsection{Validation}
    The proof of concept gives but one example of something that works.
@@
 
-   The initial situation consists of one existing system $\infsys$ and one desired system $\infsys'$ of which we have a schema, $\schema_{\infsys'}$ and an initial data set, $\dataset_{\infsys'}$.
+   The initial situation consists of one existing system $\infsys$ and one desired system $\infsys'$ of which we have a schema, $\schema_{\infsys'}$ and an initial dataset, $\dataset_{\infsys'}$.
    We may assume that $\infsys$ satisfies definition~\ref{def:information system} because it is a system in production.
-   Also, $\schema_{\infsys'}$ satisfies equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type} because the script of the desired system is type-correct.
-   Together, the schema and the intial data set forms the desired system $\pair{\schema_{\infsys'}}{\dataset_{\infsys'}}$, which satisfies definition~\ref{def:information system}.
+   Also, $\schema_{\infsys'}$ satisfies equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type} because the schema of the desired system is type-correct.
+   Together, the schema and the intial dataset forms the desired system $\pair{\schema_{\infsys'}}{\dataset_{\infsys'}}$, which satisfies definition~\ref{def:information system}.
    With these assumptions in place,
@@
 
-   The rules in the desired system are partly written to resolve data pollution.
-   In some cases, the migration engineer wants users to get rid of that pollution, as described above.
-   However, some data pollution might be satisfiable automatically.
-   In that case, the migration engineer might substitute the generated business constraint from $\busConstraints_\text{fix}$ by a transactional invariant in $\transactions_{\migrsys}$ to resolve it without bothering the business.
-   Of course, both cases need to be resolved at MoC.
+   The constraints in the desired system are partly written to resolve data pollution.
+   In some cases, the migration engineer wants users to get rid of that pollution and turn the rule in a blocking invariant as described above.
+   However, some of these constraints might be satisfiable automatically.
+   In that case, the migration engineer might substitute such constraints from $\busConstraints_\text{fix}$ by transactional invariants in $\transactions_{\migrsys}$.
+   These invariants don't need the mechanism described above, because the migration system itself will take care that all constraints in $\transactions_{\migrsys}$ are satisfied.
+   Both cases need to be resolved at MoC.
 
@@
    So from MoC onwards, $\pair{\dataset_{\migrsys}}{\schema_{\migrsys}}$ is equivalent to $\pair{\dataset_{\migrsys}}{\overrightarrow{\schema'}}$.
-   Hence, we can tell that $\dataset_{\migrsys}$ is a valid data set for the desired system, so we can switch the ingress from the migration system to the desired system without users noticing the difference.
+   Hence, we can tell that $\dataset_{\migrsys}$ is a valid dataset for the desired system, so we can switch the ingress from the migration system to the desired system without users noticing the difference.
    Then, the migration system gets no more inputs, so it can be removed. 
+   Since $\transactions_1$ and $\transactions_2$ are redundant after the MoC,
+   we can retain $\overrightarrow{\transactions_{\schema'}}$ in $\migrsys$,
+   which is equivalent to $\transactions_{\schema'}$ in the desired system.
+   Likewise, $\busConstraints_\text{fix}$ has become redundant, so $\migrsys$ can do with just $\overrightarrow{\busConstraints_{\schema'}}$.
+   In the desired system, that is equivalent to $\busConstraints_{\schema'}$.
+   So, the constraints in the desired system after the MoC are equivalent to the constraints in the MoC.
 
-\section{Conclusion}
+\section{Conclusions}
 \label{sct:Conclusions}
@@
 
-   The data migration we propose is certainly not the only approach one could think of.
+   Our proposed migration is certainly not the only approach one could think of.
    However, we have not come across other approaches that allow changing the schema in the presence of constraints.

### 2024-06-22 Generative Software/proposal.tex

--- 2023-01-11 Generative Software/proposal.tex
@@
 \documentclass{elsarticle}
-% \usepackage{natbib}
 % \usepackage{graphicx}
@@
 \usepackage[english]{babel}
-\usepackage{libertine}
-\usepackage{inconsolata}
 % \usepackage[official,right]{eurosym}
 \selectlanguage{english}
-\usepackage{hyperref}
-\frenchspacing
-
 \begin{document}
@@
 
-\title{Generative Software Development\\{\em\normalsize Research Offering}}
-\author{prof.dr.ir.\ Stef Joosten\footnote{ORCID 0000-0001-8308-0189}}
-\author{dr.\ Tim Steenvoorden\footnote{ORCID 0000-0002-8436-2054}}
+\title{Generation of user interfaces\\{\em\normalsize Research Proposal}}
+\author{Stef Joosten\footnote{ORCID 0000-0001-8308-0189}}
+\author{Tim Steenvoorden\footnote{ORCID 0000-0002-8436-2054}}
 \address{Open Universiteit Nederland, Heerlen, the Netherlands}
@@
 \begin{abstract}
-    This paper summarizes the capability of the Open Universiteit in generative software development.
-    We offer this enabling technology as a component of a larger effort that involves designing distributed information systems.
-    Generative software development offers attractive benefits for correctness, security, and resilience.
-    Also, it enables a product to be demonstrated live during its creation,
-    hereby supporting agile software development methods.
-    %every 2 or 3 weeks as it develops into an end product.
+Disciplines: software engineering, 
 \end{abstract}
@@
 \begin{keyword}
-    generative software, software engineering, security, prototyping, correctness, Ampersand, TopHat
 \end{keyword}
@@
 
-\section{Requirements (remove this section before submitting)}
-Reviewers will judge this proposal by the following requirements.
+\section{Requirements}
 \subsection{Scientific quality}
@@
 \section{Research question}
-\label{sct:Research question}
-   Data driven organizations that change frequently are facing the dilemma of buy vs. build.
+   Data driven organizations that need to adapt frequently to changing circumstances face a software dilemma.
    On one hand, flexible and fast adaptation requires immediate deployment of new software components, without interrupting the ongoing business.
    On the other hand, one-of-a-kind organizations require one-of-a-kind components that need to be developed first.
-   Commercial off-the-shelf (COTS) components provide speedy deployments, but can be too generic.
-   Homebrew components can satisfy specific requirements, but take time to build and maintain.
-   The so called ``low code platforms'' are sometimes advocated as the solution
-   because they generate user functionality from a graphical coding environment.
-   However, practical experience shows that these platforms are not so ``low code'' after all,
-   when the database structure is not properly aligned with the user functionality.
+   If the required component is generic enough, commercial off-the-shelf (COTS) components solve that problem.
+   However, if very specific components are not on the market, organizations must revert to homebrew components.
+   They either sacrifice the specific support for their business and choose a generic but readily available solution,
+   or they sacrifice their flexibility and enter into a software project to achieve the specific support they desire.
 
    In practice, we see many organizations that opt for homebrew solutions.
-   Generative development can shorten the development time of homebrew software, regaining some of the desired flexibility.
+   In these situations, generative software can shorten the development time and reclaim some of the desired flexibility.
+   Software generators are popular for many different purposes.
+   There are generators that build the database structure from a (textual or graphical) specification.
+   The so called ``low code platforms'' generate user functionality from a graphical coding environment.
+   When the database structure is not in-sync with the user functionality, custom code is needed to link the two together.
+   Generating user interfaces
+   The state of the art in generative software is that databases can be generated from specifications.
 
-\section{Plan}
-   We want to participate in a larger research project by means of one researcher (PhD candidate, full-time, 4 years) and one scientific programmer.
-   We intend to further the knowledge about generative software development by participating in a larger research effort.
-   We bring the expertise~\cite{JoostenRAMiCS2017,Steenvoorden2022} and tools~\cite{Joosten-JLAMP2018,10.1145/3354166.3354182} on generative software development
-   as an asset to that project at no extra cost.
-   The authors also intend to participate in person in the role of supervisors of the researcher and the scientific programmer, each for 0.1 FTE.
-
-   This capability offers the following benefits:
-\begin{itemize}
-    \item By generating software, we can produce live demonstrations periodically to keep the future user population informed of the progress.
-    This produces quick results, which keeps funders happy.
-    \item By automating software production, we can prevent a large class of mistakes and thereby reduce the project risk.
-    \item By formally specifying the functionality, we can give valuable guarantees on the correctness of the results.
-    \item Software that is correctly generated needs no testing. This yields significant savings on testing and test maintenance.
-    \item The tooling that has been developed at the Open University is especially suitable for automating business processes and workflows.
-\end{itemize}
-
-\section{Qualification of researchers}
-\label{sct:Qualification}
-   Research into generative software has a long tradition at the OU.
-   Results from this research comprise:
-\begin{itemize}
-   \item A generator, Ampersand, that produces a complete information system from a specification,
-         and a generator, TopHat, that produces a workflow system from a specification,
-         demonstrating that this research has tangible results~\cite{Michels2011}.
-   \item Numerous publications in peer-reviewed scientific conferences and journals,
-         demonstrating that this research is scientifically relevant.
-         Some of the highlights:
-         \begin{itemize}
-            \item A type system for information systems to ensure that all data is meaningful (i.e. has an interpretation)~\cite{vdWoude2011}
-            \item A book, "Rule Based Design", published as open educational content on the web~\cite{RBD}
-            \item An application that checks modeling conventions in architectural specifications~\cite{iceis22}
-            \item An application for identity and access management~\cite{SIAM2008}
-            \item Domain analysis~\cite{Joosten2015}.
-         \end{itemize}
-   \item A course, Rule Based Design, which is in the Master's program of the Computer Science curriculum and the Business Process Management \& IT program of the OU,
-         demonstrating that this research is valued by the OU.
-   \item over 10 generated information systems in production, made by TNO and one made by the OU,
-         demonstrating that this research has practical value for industry.
-   \item in-kind support from Ordina,
-         demonstrating that this research can draw positive attention from the IT industry.
-\end{itemize}
-
-Current work in our group is about supporting incremental changes to an information system generatively,
-to support DevOps teams who need incremental deployments.
-Supporting incremental deployments with a generator is likely to decrease deployment times,
-to increase the number of releases per time unit, and to decrease the risk of errors propagating to production.
+\label{sct:Introduction}
 \section{Bibliography}
